### 2016-10-19 manuscript.tex
Initial commit

\documentclass[12pt]{article}

\usepackage{graphicx} 
\usepackage{amsmath}
%\usepackage{a4wide}
%\usepackage{multicol,caption}
%\setlength{\textwidth}{16cm}
%\setlength{\textheight}{22cm}

\begin{document}\sloppy

\title{A preliminary rcK derivation}
\author{J Bertram and J Masel}
\maketitle

``...the concept of fitness is probably too complex to allow of a useful mathematical development. Since it enters fundamentally into many population genetics considerations, it is remarkable how little attention has been paid to it.'' --- Warren J. Ewens, Mathematical Population Genetics I, 2004

\section{Introduction}

Evolutionary models differ greatly in their treatment of fitness. In models of genetic evolution, genotypes are typically assigned constant (or frequency-dependent) selection coefficients describing the change in their relative frequencies over time due to differences in viability. This considerably simplifies the mathematics of selection, facilitating greater genetic realism, and can be justified over sufficiently short time intervals \cite[p. 276]{ewens_2012}. However, selection can have very different effects when operating on different types of traits, and evolutionary changes in one population can lead to complicated ecological responses.

By contrast models of phenotypic trait evolution represent the change in phenotypic abundances over time using absolute fitness functions which describe how those traits affect survival and reproduction in particular ecological scenarios. This approach is powerful enough to model eco-evolutionary feedbacks between co-evolving traits, but is generally problem-specific and restricted to only a few traits at a time.

Far less work has been done to model fitness in more general terms than particular traits or ecological scenarios, while still capturing key distinctions between different forms of selection. Perhaps this is not surprising given that fitness is such a complex quantity, dependent on all of a phenotype's functional traits \cite{violle_2007} as well as its biotic and abiotic environment. In most cases, a detailed, trait-based, predictive model of fitness would be enormously complicated and have narrow applicability. It is therefore easy to doubt the feasibility of a simplified, general mathematical treatment of fitness \cite[p. 276]{ewens_2012}. Even MacArthur's famous r/K selection scheme is now almost exclusively known as a framework for understanding life-history traits, and judged on its failure in that role \cite{pianka_1970,stearns_1977,boyce_1984,reznick_2002}. In spite of the r/K scheme's original purpose as an extension of the existing population-genetic treatment of selection to account for population density \cite{macarthur_1962}, comparatively few attempts have been made to develop it further as a mathematical analysis of the major different forms of selection.

Nevertheless, there are strong indications there are broader principles governing the operation of selection. In many groups of organisms (including corals \cite{darling_2012}, insects \cite{southwood_1977}, fishes \cite{winemiller_1992} and plants \cite{grime_1988}), species can be partitioned into a small number of distinct trait clusters corresponding to fundamentally distinct ``primary strategies'' \cite{winemiller_2015}. The most famous example is Grime's plant trait classification scheme \cite{grime_1974,grime_1977,grime_1988}. Grime considered two broad determinants of population density: stress (persistent hardship e.g. due to resource scarcity, unfavorable temperatures or toxins) and disturbance (intermittent destruction of vegetation e.g. due to trampling, herbivory, pathogens, extreme weather or fire).  The extremes of these two factors define three primary strategies denoted by C/S/R respectively: competitors ``C'' excel in low stress, low disturbance environments; stress tolerators ``S'' excel in high stress, low disturbance environments; and ruderals ``R''  excel in low stress, high disturbance environments. Survival is not possible in high-stress, high-disturbance environments. Grime showed that measures of C, S and R across a wide range of plant species are anti-correlated, so that strong C-strategists are weak S and R strategists, and so on. Thus, plant species can be classified on a triangular C/S/R ternary plot \cite{grime_1974}. Trait classification schemes for other organisms closely parallel Grime's scheme \cite{winemiller_2015}.

Trait classification schemes show empirically that, beneath the complicated details of trait variation, even among closely-related species, fitness is predominantly determined by a few key factors such as intrinsic reproducive rate or stress-tolerance. However, while trait classification schemes are firmly grounded in trait data, they are verbal and descriptive rather than mathematical, a recognized hinderance to their broader applicability (e.g. \cite{tilman_2007}).

The aim of this paper is explore the interplay between some major dimensions of fitness in a simplified, spatially-homogeneous model of genotype growth, dispersal and competition. Building on the earlier r/K and C/S/R schemes, a central question is how fitness depends on the interaction between population density, intrinsic birth/death rates and competitive ability.

We broadly follow the spirit of MacArthur's r/K selection scheme in that our model is intended to account for fundamentally different forms of selection without getting  entangled in the intricacies of particular ecological scenarios. However, rather than building directly on MacArthur's formalism and its later extensions [cite alpha stuff] (e.g. using Lotka-Volterra equations), our model is devised primarily with Grime's C/S/R scheme in mind, and represents a quantitative formalization of how C/S/R manifests at the level of genotype evolution (as opposed to divergence between species). This choice is motivated in part by the substantial empirical support for the C/S/R scheme, and in part by the failings of the r/K low/high density dichotomy --- many growth ability traits will confer advantages at both low and high densities, in which case r- and K- selection will effectively coincide (empirically, positive correlations between measures of r and K are common, both between species and strains \cite{luckinbill_1979,kuno_1991,hendriks_2005,fitzsimmons_2010}, and as a result of experimental evolution \cite{luckinbill_1979,luckinbill_1978}).

In section

\section{Model}

We assume that each individual in a population requires its own territory to survive and reproduce (a site-occupancy model). All territories are identical, and the total number of territories is $K$. Time $t$ advances in discrete steps, each representing one generation. In generation $t$, the number of mature individuals (henceforth called ``adults'') of the $i$'th genotype is $n_i(t)$, the total number of adults is $N(t)=\sum_i n_i(t)$, and the number of unoccupied territories $U(t)=K-N(t)$.

Each generation, adults produce $m_i$ new offspring (henceforth called ``propagules'') which disperse at random over the $U$ unnocupied territories. We ignore propagules that land on occupied territories --- these are assumed to have no chance of survival and are not included in $m_i$. For simplicity, we assume $m_i=b_i n_i$, where $b_i$ is a constant, genotype-specific birth rate.

The number of individuals of the $i$'th genotype landing in any particular territory is denoted $x_i$. Random dispersal implies that in the limit $K\rightarrow \infty$, with $n_i/K$ held fixed, $x_i$ is Poisson distributed with mean territorial propagule density $l_i=m_i/U$. Although $K$ is finite in our model, we assume that $K$ and the $n_i$ are large enough that $x_i$ is Poisson-distributed to a good approximation (details in Appendix A). This Poisson distribution is denoted $p_i(x_i)$. In particular, we do not consider the stochastic behavior that occurs at very low $n_i$ due to the lottery nature of territorial acquisition (this will be discussed further when we consider the behavior of mutant lineages).

When multiple propagules land on the same territory, they compete to secure the territory as they develop. This territorial contest is modeled as a weighted lottery: the probability that genotype $i$ wins a given territory by the next generation is $c_i x_i/\sum_j c_j x_j$ where $c_i$ is a constant representing relative competitive ability.

The increase in $n_i$ over one generation due to territorial acquisition, $\Delta_+ n_i$, is the sum of genotype $i$'s victories over all $U$ unoccupied territories. Since $p_1(x_1)\ldots p_G(x_G)$ is equal to the proportion of unoccupied territories with $x_1,\ldots,x_G$ of the respective propagules (again, we assume that $K$ is large enough that fluctuations in this proportion are negligible), this sum can be replaced by an expectation over the $p_i$. This gives
\begin{equation}
\Delta_+ n_i(t)=U(t)\sum_{x_1,\ldots,x_G} \frac{c_i x_i}{\sum_j c_j x_j} p_1(x_1)\ldots p_G(x_G). \label{eq:growthsumuncoupled}
\end{equation}

In addition to propagule birth and competition, occupied territories become unoccupied when adults die. We assume that these adult deaths occur at a constant, genotype-specific rate $d_i$. Thus, the overall change in genotype abundances is
\begin{equation}
\Delta n_i(t)=\Delta_+ n_i(t)-d_i n_i(t). \label{eq:delttot}
\end{equation}

The change in total population size $N$, obtained by summing Eq. \eqref{eq:delttot} over genotypes, is
\begin{equation}
\Delta N=U(1-e^{-L})-\sum_i d_i n_i=(K-N)(1-e^{-L})-\overline{d}N,
\end{equation}
where $\overline{d}=\sum_i d_i n_i/N$, which is independent of the competitive ability coefficients $c_i$. Thus, the $c_i$ represent a strictly relative aspect of fitness in the sense that they only influence $N$ indirectly by changing genotype frequencies, which may in turn change population mean birth and death rates.

\section{Results}

\subsection{Mean Field Approximation}

Eq. \eqref{eq:delttot} gives little intuition about the dynamics of density-dependent lottery competition, since \eqref{eq:growthsumuncoupled} involves an expectation over the random dispersal distributions $p_i$, which depend on how the $n_i$ change over time. We now evaluate this expectation approximately, which allows us to develop deeper intuitions about the dynamics of density dependent lottery competition.

When propagule densities are high ($l_i\gg 1$ for all genotypes), the fluctuations in the $x_i$ are negligible compared to the corresponding expectations $l_i$ (since the $x_i$ are Poisson distributed), and we can make the replacement $x_i\rightarrow l_i$ in Eq. \eqref{eq:growthsumuncoupled}. Eq. \eqref{eq:growthsumuncoupled} then reduces to the classic lottery model \cite{chesson_1981},
\begin{equation}
\Delta_+ n_i(t)=U(t)\frac{c_i m_i}{\sum_j c_j m_j}= b_i n_i\frac{1}{L}\frac{c_i}{\overline{c}}, \label{eq:lottery}
\end{equation}
where $L=\sum_j l_j=M/U$ is the overall propagule density,  $\overline{c}=\sum_j c_j m_j/M$ is the expected competitive ability for randomly selected propagules, and $M=\sum_j m_j$ is the total number of propagules.

When $l_i\ll 1$, $x_i$ cannot be replaced by $l_i$ in Eq. \eqref{eq:growthsumuncoupled}, because $x_i=1$ in the sparsely scattered territories where $i$ propagules do land, meaning that all of the contributions to $\Delta_+ n_i$ are large deviations from the mean. Eq. \eqref{eq:lottery} is clearly nonsensical in this case: genotype $i$ can win at most $m_i$ territories, yet Eq. \eqref{eq:lottery} demands a fraction $c_i m_i/\sum_j c_j m_j$ of the unoccupied territories $U$, no matter how large $U$ is.

A similar mean field approximation can be applied to Eq. \eqref{eq:growthsumuncoupled} for arbitrary propagule densities provided that the $x_i=1$ terms for the focal genotype are handled separately. This approximation relies on the the relative fluctuations in the lottery competition denominator $\sum c_j x_j$ being small, allowing the sum in Eq. \eqref{eq:growthsumuncoupled} the be evaluated under a much broader range of conditions than $l_i\gg 1$ for all genotypes. In Appendix B we show that this is generally the case, unless there large discrepancy in competitive ability ($c_i/c_j\gg 1$). This exception will be discussed below; for now we assume that no such large discrepancy is present. We obtain (Appendix B)
\begin{equation}
\Delta_+ n_i(t)\approx b_i n_i\left[e^{-L}+(R_i+A_i)\frac{c_i}{\overline{c}}\right]. \label{eq:master}
\end{equation}
where
\begin{equation}
R_i=\frac{\overline{c}e^{-l_i}(1-e^{-(L-l_i)})}{c_i +\frac{\overline{c}L- c_il_i}{1-(1+L)e^{-L}}\frac{L-1+e^{-L}}{L-l_i}},\label{eq:Dr}
\end{equation}
and
\begin{equation}
A_i=\frac{\overline{c}(1-e^{-l_i})}{c_il_i\frac{1-e^{-l_i}}{1-(1+l_i)e^{-l_i}}+\sum_{j\neq i}\frac{c_jl_j}{L-l_j}\left(L\frac{1-e^{-L}}{1-(1+L)e^{-L}}-l_j\frac{1-e^{-l_j}}{1-(1+l_j)e^{-l_j}}\right)}.\label{eq:Da}
\end{equation}

Comparing Eq. \eqref{eq:master} to Eq. \eqref{eq:lottery}, the classic lottery per-propagule success rate $c_i/\overline{c}L$ has been replaced by three separate terms. The first, $e^{-L}$, accounts for propagules which land alone on unoccupied territories; these territories are won without contest. The second term, $R_i c_i/\overline{c}$ represents competitive victories when the $i$ genotype is a rare invader in a high density population: from Eq. \eqref{eq:Dr}, $R_i\rightarrow 0$ when the $i$ genotype is abundant ($l_i\gg 1$), or other genotypes are collectively rare ($L-l_i\ll 1$). The third term, $A_ic_i/\overline{c}$, represents competitive victories when the $i$ genotype is abundant: $A_i\rightarrow 0$ if $l_i\ll 1$. The relative importance of these three terms changes with both the overall propagule density $L$ and the relative propagule frequencies. In the case that all $l_i\gg 1$ for all genotypes, only the last term $A_ic_i/\overline{c}$ remains, and $A_i\rightarrow 1/L$, recovering the classic lottery model.

Fig. \ref{fig:simcomp} shows that Eq. \eqref{eq:master} (and its components) closely approximate direct simulations of random dispersal with lottery competition over a wide range of propagule densities (obtained by varying $U$). A low-frequency genotype with higher $c_i$ is attempting to invade an otherwise identical high-frequency genotype. The low-density invader term $R_i c_i/\overline{c}$ is crucially important for the low-frequency genotype (Fig. \ref{fig:simcomp}b), but is negligible for the high-frequency one (Fig. \ref{fig:simcomp}d). The high density competition term $A_i c_i/\overline{c}$ is crucial for the high-frequency genotype, and can also be important for the invader if $U$ is small enough.

\begin{figure}
\centering
\includegraphics[scale=0.7]{simulationcomparison.pdf}
\caption{\label{fig:simcomp} Eq. \eqref{eq:master} is a good approximation of the density dependent lottery model. $\Delta_+ n_i/m_i$ from Eq. \eqref{eq:master} (and its separate components) are shown, along with direct simulations of random dispersal and lottery competition over one generation over a range of propagule densities (varied by changing $U$ with the $m_i$ fixed). Two genotypes are present: a low-frequency, better competitor ($m_1/M=0.1$, $c_1=1.5$), in (a) and (b); and a high-frequency, worse competitor ($m_2/M=0.9$, $c_2=1$) in (c) and (d).} 
\end{figure}

Thus, Eq. \ref{eq:master} is a density-dependent generalization of the classic lottery model. In particular, Eq. \ref{eq:master} accounts for the growth (or decline) of initially rare mutant lineages, which is essential for modeling evolution.

%with the exception of large discrepancies in competitive ability since it is extremely unlikely that a large discrepancy competitive would arise in the absence of migration.

\subsection{Invasion and coexistence}

Now suppose that a mutation occurs, producing a new genotype $j$ which differs from $i$ in at least one of the traits $b_i$, $c_i$ and $d_i$. The new mutant lineage starts with one individual, and remains at low abundance for many generations. During this period, the mutant lineage abundance $n_j$ will behave stochastically, and Eq. \eqref{eq:master} does not apply. However, if $n_j$ becomes large enough, its behavior will become effectively deterministic, and governed by Eq. \eqref{eq:master}. For mutants with fitness greater than the population mean fitness, this process is known as ``establishment'', and occurs when $n_j$ is of order $1/s$, where $s$ is the mutant's fitness advantage relative to the mean \cite{desai_2007}.

Here we do not analyze the early stochastic behavior of new mutations, and restrict our attention to the earliest deterministic behavior of rare genotypes. In particular, for beneficial mutants we only consider the case where $s$ is large enough that initially we have $n_j \ll N$.

Suppose that a population with a single genotype $i$ is in equilibrium. Then $R^{(i)}_i=0$, $\overline{c}=c_i$ and $\Delta n_i = 0$, which implies
\begin{equation}
b_i\left(e^{-L}+A_i\right)-d_i=0.\label{eq:equil}
\end{equation}
Now suppose that new genotype $j$, which is initially rare, appears in the population. Then $A_j\ll 1$, $l_j\ll L$ and $\overline{c}\approx c_i$, and so initial growth of $j$ occurs if 
\begin{equation}
b_j \left(e^{-L^{(i)}}+R^{(i)}_j\frac{c_j}{c_i}\right)-d_j>0.\label{eq:invad}
\end{equation}

If $j$ is superior in one trait, but otherwise identical to $i$, it will eventually displace $i$. However, stable coexistence is possible between genotypes that are superior in different traits. To illustrate, suppose that $j$ is better at securing territories ($c_j>c_i$), that $i$ is better at producing propagules ($b_i>b_j$), and that $d_i=d_j$. Coexistence occurs if $j$ will invade an $i$-dominated population, but $i$ will also invade a $j$-dominated population (``mutual invasion''). It is not hard to show that this is possible, since if $b_i$ so large that $L\gg 1$ if $i$ is dominant, while $b_j$ is so small that $L\ll 1$ when $j$ is dominant, then, combining Eqs. \eqref{eq:equil} and \eqref{eq:invad}, we find that $j$ invades $i$ if
\begin{equation}
b_jc_jR^{(i)}_j-b_i c_i A^{(i)}_i>0, \label{eq:jinvadcoex}
\end{equation}
while $i$ invades $j$ if
\begin{equation}
b_i-b_j>0.\label{eq:iinvadcoex}
\end{equation}
Eq. \eqref{eq:iinvadcoex} is already satisfied by assumption, while \eqref{eq:jinvadcoex} is satisfied for large enough $c_j$. The mechanism for coexistence is that territorial contests are important in $i$-dominated populations (high $L$), ensuring that $j$ is not excluded by $i$, yet territorial contests are irrelevant in $j$-dominated populations (low $L$), ensuring that $i$ is not excluded by $j$.

Fig. \ref{fig:coex} shows an example of coexistence between $b$ and $c$ specialists.

 A similar argument applies for coexistence between high-$c$ and low-$d$ specialists.

\begin{figure}
\centering
\includegraphics[scale=0.7]{coex.pdf}
\caption{\label{fig:coex} Coexistence between $b$ ($c_i=1$, $b_i=1$) and $c$ ($c_j=2$, $b_j=0.7$) specialists, where $d_i=d_j=0.3$. Frequency of the $c$-specialist predicted by Eq. \eqref{eq:master} is shown.} 
\end{figure}

\subsection{Environmental archetypes and Grime's triangle}

We can then determine which mutational changes in the traits $b_i, c_i$ and $d_i$ will be favored under different environmental conditions. Of particular interest are Grime's ``disturbance'', ``stress'' and ``ideal'' environmental archetypes.

To proceed, we map these archetypes to different parameter regimes in our model. The ideal archetype is defined by low extrinsic mortality $d_i\ll 1$ and high birth rates, such that $b_i$ significantly exceeds $d_i$,   population density $N/K$ is high, and thus so is the overall propagule density $L\gg 1$. In this case, competition between propagules is important.

The disturbance archetype is defined by unavoidably high extrinsic mortality, usually due to physical destruction of individual organisms, to the extent that the population is kept at low densities $N/K\ll 1$. This means that, even though $b_i$ may be quite high, it cannot greatly exceed $d_i$, and that  propagule densities are low $l_i\ll 1$. The terms proportional to $c_i/\overline{c}$ in Eq. \eqref{eq:master} are then negligible, and $\Delta_+ n_i$ depends primarily on $b_i$.

The stress archetype is more ambiguous, and has been the subject of an extensive debate in the plant ecology literature. In Grime's view, the archetypal stressful environmental imposes such severe challenges that success depends primarily on the ability to tolerate the stressors. Population density $N/K$ is suppressed to such low levels that competition between individuals is not important. In our model, this again corresponds to $b_i$ only slightly exceeding $d_i$ and $l_i\ll 1$, similar to the disturbance archetype, except that stress now constrains $b_i$ to be small.

The alternative view is that the stress archetype should rather be interpreted as a large reduction in the maximum number of individuals that can be supported \cite{taylor_1990}. For example, in the case that the stress is induced by a scarcity of consumable resources, competition for resources would likely be intense, and the stressed population should actually be regarded as having a high population density. In our model, this would imply a large reduction in $K$ (greater per-individual territorial requirement), such that even though fewer individuals are supported, density $N/K$ is still high.

The mapping of environmental achetypes to our model parameters is summarized in the first three rows of Fig. \ref{fig:table}. Also shown is the approximate dependence of $\Delta_+ n_i$ on $b_i$ and $c_i$ for each archetype (fourth row), which can be used infer the expected direction of evolution for the traits $b_i$, $c_i$ and $d_i$ (fifth row).

The latter is inferred as follows. As noted in the previous section, if beneficial mutants can survive the low-abundance stochastic regime, their behavior is governed deterministically by Eq. \eqref{eq:master}. If their advantage in $b$, $c$ or $d$ is large enough to overcome the lottery rarity disadvantage, they will proceed to grow deterministically (establishment). The probability of establishment increases with the mutant fitness advantage, and is of order $1/N$ for neutral mutations. Consequently, the direction of evolutionary change is determined by which trait changes confer an appreciable benefit, subject to the constraints imposed by the environmental.

For example, in Grime's version of the stress archetype, population density is low, so competition is not important, and so only mutants with greater $b$ or lower $d$ will have an appreciably greater $\Delta n_i$. Mutations in $c$ are effectively neutral, and will rarely establish. However, by definition of the stress archetype, $b$ is constrained to be small. Thus, while some rare mutations may produce small improvements in $b$, it is much more likely that mutations will arise that lower $d$, making this the expected direction of evolutionary change for Grime's stress archetype.

\begin{figure}
\centering
\begin{tabular}{*{5}{c}}
  & Ideal & Disturbance & Stress (G) & Stress (K) \\ \hline
  Parameter- & $d_i \ll 1$ & $d_i \gg 1$ & $b_i \ll 1$ & $b_i \ll 1$ \\
  regime & $b_i>d_i$ & $b_i\approx d_i$ & $b_i\approx d_i$ & $b_i>  d_i$ \\
  Density  & High & Low & Low & High \\
  $\Delta_+ n_i\propto$ & $b_i c_i$ & $b_i$ & $b_i$ & $b_i c_i$ \\
  Evolution & $\uparrow b$, $ \uparrow c$ & $\uparrow b$ & $\downarrow d$ & $\uparrow c$
\end{tabular}
\caption{\label{fig:table} The realization of Grime's environmental archetypes in our model, as well as the low-$K$ variant of the stress archetype. Shown are the mapping to our parameters of each archetype, the approximate dependence of $\Delta_+ n_i$ on $b_i$ and $c_i$, as well as the corresponding expected evolutionary changes in $b_i$, $c_i$ and $d_i$.}
\end{figure}

Following Grime's original argument for a triangular scheme \cite{grime_1977}, Fig. \ref{fig:axes} represents each environmental archetype schematically as a vertex on a triangular space defined by perpendicular stress and disturbance axes. The ideal archetype lies at the origin (no stress or disturbance), while the stress and disturbance archetypes lie at the limits of survival on their respective axes. The hypotenuse connecting the stress and disturbance endpoints represents the limits of survival in the presence of a combination of stress and disturbance. The direction of evolutionary change is different at each vertex, leading to the emergence of different trait clusters or ``primary strategies''.

\begin{figure}
\centering
\includegraphics[scale=1]{axes.pdf}
\caption{\label{fig:axes} The realization of Grime's triangle in our model. Schematic representation of the triangular space bounded by the low/high extremes of stress/disturbance. The low-$K$ interpretation of stress is also shown. The vertices of the triangles correspond to environmental archetypes. Selection favors different traits at each vertex, leading to different trait clusters.} 
\end{figure}

\section{Discussion}

Grime's triangle without constraints

r-K correlation, meaning of K selection

Actual K selection

caveats: large c discrepancy

\bibliographystyle{unsrt}
\bibliography{reference}

\section*{Appendix A: Poisson approximation}

The propagule numbers $x_i$ in different territories are not independent random variables. To determine the dispersal outcomes in all unoccupied territories exactly, we would need to proceed territory-by-territory as follows. In the first territory we evaluate, $x_i$ drawn from a binomial distribution with $m_i$ trials and success probability $1/U$. In the second, $x_i$ is drawn from a binomial distribution with $m_i-x$ trials and success probability $1/(U-1)$, where $x$ is the number of propagules that landed in the first territory. And so on.

For sufficiently large $K$, holding $n_i/K$ fixed, the Poisson limit theorem implies that the binomial distributions for $x_i$ at each successive stage of this procedure are all closely approximated by a Poisson distribution with mean $l_i$, where we have used the fact that large $K$ implies large $U$ except in the biologically uninteresting case that there is vanishing population turnover $d_i \sim 1/K$.

Under the Poisson approximation, the total number of genotype $i$ propagules $\sum x_i$ (summed over unoccupied territories) will deviate about its mean value $m_i$. Since the coefficient of variation of $\sum x_i$ is proportional to $1/\sqrt{m_i}$, these deviations are negligible unless $m_i$ is very small (say of order $100$ or less).

\section*{Appendix B: Derivation of growth equation}

We separate the right hand side of Eq. \eqref{eq:growthsumuncoupled} into three components $\Delta_+ n_i = \Delta_u n_i+\Delta_r n_i+\Delta_a n_i$ which vary in relative magnitude depending on the propagule densities $l_i$. Following the notation in the main text, the Poisson distributions for the $x_i$ (or some subset of the $x_i$) will be denoted $p$; for instance $p(x_1,\ldots,x_G)=p_1(x_1)\ldots p_G(x_G)$ and $p(x_1,\ldots,x_{i-1},x_{i+1}\ldots,x_G)=p_1(x_1)\ldots p_{i-1}(x_{i-1})p_{i+1}(x_{i+1})\ldots p_G(x_G)$. We use $P$ as a general shorthand for the probability of particular outcomes.

\subsection*{Growth without competition}

The first component, $\Delta_u n_i$, accounts for territories where only one focal propagule is present $x_i=1$ and $x_j=0$ for $j\neq i$ ($u$ stands for ``uncontested''). The proportion of territories where this occurs is $l_i e^{-L}$, and so 
\begin{equation}
\Delta_u n_i=Ul_i e^{-L}=m_i e^{-L}.
\end{equation}

\subsection*{Competition when rare}

The second component, $\Delta_r n_i$, accounts for territories where a single focal genotype propagule is present along with at least one non-focal propagule ($r$ stands for ``rare'') i.e. $x_i=1$ and $\sum_{j\neq i} x_j\geq 1$. The number of territories where this occurs is $Up_i(1)P(\sum_{j\neq i} x_j\geq 1)=b_i n_i e^{-l_i}(1-e^{-(L-l_i)})$. Thus 
\begin{equation}
\Delta_r n_i = m_i e^{-l_i}P\left\langle  \frac{c_i}{c_i +\sum_{j\neq i} c_j x_j } \right\rangle_{\tilde{p}},  \label{eq:deltr}
\end{equation}
where $\langle \rangle_{\tilde{p}}$ denotes the expectation with respect to $\tilde{p}$, and $\tilde{p}$ is the probability distribution of nonfocal propagaule abundances $x_j$ \textit{after} dispersal, in those territories where exactly one focal propagule, and at least one non-focal propagule, landed.

We now show that, with respect to $\tilde{p}$, the standard deviation in $\sum_{j\neq i} c_j x_j$, $\sigma(\sum_{j\neq i} c_j x_j)$, is much smaller than its mean $\langle\sum_{j\neq i} c_j x_j\rangle_{\tilde{p}}$. Then $x_j$ can be replaced by its mean in the last term in Eq. \eqref{eq:deltr},
\begin{equation}
\left\langle\frac{c_i}{c_i +\sum_{j\neq i} c_j x_j}\right\rangle_{\tilde{p}}\approx \frac{c_i}{c_i +\sum_{j\neq i} c_j \langle x_j\rangle_{\tilde{p}}},\label{eq:meanfieldr}
\end{equation}
which will give us Eq. \eqref{eq:Dr}.

The exact expression for $\langle x_j \rangle_{\tilde{p}}$ is somewhat complicated. Letting $k$ denote the total number of propagules in a territory, and ${\mathbf x_i}=(x_1,\ldots,x_{i-1},x_{i+1}\ldots,x_G)$ denote the vector of non-focal abundances, $\tilde{p}$ can be written as
\begin{align}
\tilde{p}({\mathbf x_i})&=p({\mathbf x_i}|k\geq 2,x_i=1),\nonumber\\
&=\frac{P(k\geq 2|{\mathbf x_i},x_i=1) p({\mathbf x_i}|x_i=1)}{P(k\geq 2)},\nonumber\\
&=\frac{p({\mathbf x_i}|x_i=1)}{1-(1+L)e^{-L}},\nonumber\\
&=\frac{1}{1-(1+L)e^{-L}}\sum_{k=2}^{\infty} P(k) p({\mathbf x_i}|\sum_{j\neq i} x_j=k-1),\nonumber\\
&=\frac{1}{1-(1+L)e^{-L}}\sum_{k=2}^{\infty} \frac{P(k)\delta^{\sum_{j\neq i} x_j}_{k-1}}{P(\sum_{j\neq i} x_j=k-1)} p({\mathbf x_i}),\nonumber\\
&=\frac{Le^{-l_i}}{1-(1+L)e^{-L}}\sum_{k=1}^{\infty} \left(\frac{L}{L-l_i}\right)^k\frac{\delta^{\sum_{j\neq i} x_j}_k}{k+1} p({\mathbf x_i}),\label{eq:raremonster}
\end{align}
where $\delta^{\sum_{j\neq i} x_j}_k=1$ if $\sum_{j\neq i} x_j=k$, and equals zero otherwise. Then, since 
\begin{align}
\sum_{\mathbf x_i}\delta^{\sum_{j\neq i} x_j}_k p({\mathbf x_i})x_j&= \frac{l_j}{L-L_j}kP(\sum_{j\neq i} x_j=k)\nonumber\\
&=l_j P(\sum_{j\neq i} x_j=k-1),
\end{align}
after some algebra we obtain,
\begin{equation}
\langle x_j \rangle_{\tilde{p}}=\frac{l_j}{1-(1+L)e^{-L}}\frac{L-1+e^{-L}}{L-l_i}.\label{eq:meanxjrare}
\end{equation}

To calculate the relative fluctuations in $\sum_{j\neq i} c_j x_j$, we use the following approximation, which gives considerably simpler expressions for the means, variances and covariances of the $x_j$ compared with the exact expressions using $\tilde{p}$. Rather than evaluating the situation in each territory after dispersal as above, we let $\tilde{p}$ instead be the ${\mathbf x_i}$ dispersal probabilities in a territory where one focal propagule is assumed to be present, conditional on $\sum_{j\neq i} x_j>1$. This gives $\langle x_j \rangle_{\tilde{p}}=l_j/C$, 
\begin{equation}
\sigma^2(x_j)=\frac{l_j^2}{C}\left(1-\frac{1}{C}\right)+\frac{l_j}{C},\label{eq:varr}
\end{equation}
and 
\begin{equation}
\sigma(x_j,x_k)=\frac{l_j l_k}{C}\left(1-\frac{1}{C}\right),\label{eq:covr}
\end{equation}
where $C=1-e^{-(L-l_i)}$ (note the difference from Eq. \eqref{eq:meanxjrare} for $\langle x_j \rangle_{\tilde{p}}$). Then, since
\begin{equation}
\sigma^2(\sum_{j\neq i} c_j x_j)=\sum_{j\neq i}\left[c_j^2\sigma^2(x_j)+2\sum_{k>j}c_j c_k\sigma(x_j,x_k)\right],\label{eq:vartotr}
\end{equation}
and $1/C>1$, we have
\begin{equation}
\frac{\sigma(\sum_{j\neq i} c_j x_j)}{\langle\sum_{j\neq i} c_j x_j\rangle}<C^{1/2}\frac{\left(\sum_{j\neq i}c_j^2 l_j\right)^{1/2}}{\sum_{j\neq i}c_j l_j}. \label{eq:cvr}
\end{equation}

Without loss of generality, we restrict attention to the case that the total nonfocal density $L-l_i$ is of order $1$ or larger (otherwise $\Delta_r n_i$ does not contribute significantly to $\Delta_+ n_i$ because $\Delta_r n_i$ is proportional to $C=1-e^{-(L-l_i)}$).

When at least some of the nonfocal propagule densities are large $l_j\gg 1$, then the RHS of Eq. \eqref{eq:cvr} is $\ll 1$, as desired. This is also the case if none of the nonfocal genotype densities are large and the $c_j$ are all of similar magnitude (their ratios are of order one); the worst case scenario occurs when $(L-l_i)\sim O(1)$, in which case the negative covariances (Eq. \eqref{eq:covr}) which were neglected in the RHS of Eq. \eqref{eq:cvr} significantly reduce the overall variance $\sigma^2(\sum_{j\neq i} c_j x_j)$.

However, the relative fluctuations in $\sum_{j\neq i} c_j x_j$ can be large if some of the $c_j$ are much larger than the others. Specifically, if $c_j l_j\gg c_k l_k$ ($j,k\neq i$, $j\neq k$) and $l_j\ll 1$ (i.e. in the presence of a rare, extremely strong competitor), then we cannot make the replacement Eq. \eqref{eq:meanfieldr}.

Substituting Eqs. \eqref{eq:meanfieldr} and \eqref{eq:meanxjrare} into Eq. \eqref{eq:deltr}, we obtain
\begin{equation}
\Delta_r n_i\approx m_i R_i\frac{c_i}{\overline{c}}, \label{eq:deltrfinal}
\end{equation}
where $R_i$ is defined in Eq. \eqref{eq:Dr}.

\subsection*{Competition when abundant}

The final contribution, $\Delta_a n_i$, accounts for territories where two or more focal propagules are present ($a$ stands for ``abundant"). Similarly to Eq. \eqref{eq:deltr}, we have 
\begin{equation}
\Delta_a n_i=U(1-(1+l_i)e^{l_i})\left\langle \frac{c_i x_i}{\sum_j c_j x_j} \right\rangle_{\hat{p}}\label{eq:delta}
\end{equation}
where $\hat{p}$ is the probability distribution of both focal and nonfocal propagaule abundances \textit{after} dispersal in those territories where at least two focal propagules landed.

Again, we wish to show that the relative fluctuations in $\sum c_j x_j$ are much smaller than $1$ (with respect to $\hat{p}$), so that we have 
\begin{equation}
\left\langle \frac{c_i x_i}{\sum_j c_j x_j} \right\rangle_{\hat{p}}\approx  \frac{c_i \langle x_i \rangle_{\hat{p}}}{\sum_j c_j \langle x_j\rangle_{\hat{p}}}.\label{eq:meanfielda}
\end{equation}
Following a similar procedure as for $\Delta_r n_i$, where the vector of propagule abundances is denoted ${\mathbf x}$, we have
\begin{align}
\langle x_j \rangle_{\hat{p}}&=\sum_{\mathbf x} x_j p(\mathbf x|x_i\geq 2)\nonumber \\
&=\sum_{k}P(k|x_i\geq 2)\sum_{x_i} \sum_{\mathbf x_i} x_j p(\mathbf x_i|\sum_{j\neq i} x_j=k-x_i)p(x_i|x_i\geq 2,k)\nonumber\\
&=\sum_{k}P(k|x_i\geq 2)\sum_{x_i} \frac{l_j(k-x_i)}{L-l_j} p(x_i|x_i\geq 2,k)\nonumber\\
&=\frac{l_j}{L-l_j}\left( L\frac{1-e^{-L}}{1-(1+L)e^{-L}}- l_j\frac{1-e^{-l_j}}{1-(1+l_j)e^{-l_j}}\right) 
\end{align}
for $j\neq i$, and 
\begin{equation}
\langle x_i\rangle_{\hat{p}}=l_i\frac{1-e^{-l_i}}{1-(1+l_i)e^{-l_i}}.
\end{equation}

To calculate the relative fluctuations in $\sum_{j\neq i} c_j x_j$, we use a similar approximation as for $\Delta_r n_i$: $\tilde{p}$ is approximated by the ${\mathbf x}$ dispersal probabilities in a territory where at least two focal propagule is assumed to be present. All covariances are now zero, so that $\sigma^2(\sum c_j x_j)=\sum c_j^2 \sigma^2(x_j)$, where $\sigma^2(x_j)=l_j$ for $j\neq i$. The expression for $\sigma^2(x_i)$ is more complicated. We assume $p(x_i=0)\approx 0$ without loss of generality (since otherwise $D\gg 1$ and $\Delta n_a$ is negligible). Then
\begin{equation}
\sigma^2(x_i)=\frac{l_i^2}{D}\left(1-\frac{1}{D}\right)+\frac{l_i}{D},
\end{equation}
where $D=1-(1+l_i)e^{-l_i}$, analogous to Eq. \eqref{eq:varr}, and 
\begin{equation}
\frac{\sigma(\sum c_j x_j)}{\langle\sum c_j x_j\rangle} \approx\frac{\left(\sum_{j\neq i} c_j^2 l_j + c_i^2 \sigma^2(x_i)\right)^{1/2}}{\sum_{j\neq i} c_j l_j + c_i l_i/D} \label{eq:cva}.
\end{equation}

Similarly to Eq. \eqref{eq:cvr}, the RHS of \eqref{eq:cva} will not be $\ll1$ if there is a nonfocal genotype $j$ with $l_j\ll 1$ and $c_j l_j\gg c_k l_k$ for $j,k\neq i$, $j\neq k$. When this is not the case, then since $l_i$ must be of order $1$ or larger for $\Delta_a n$ to make an appreciable contribution to $\Delta_+ n_i$, the RHS of Eq. \eqref{eq:cva} is $\ll 1$ as desired.

Combining Eqs. \eqref{eq:delta} and \eqref{eq:meanfielda}, we obtain
\begin{equation}
\Delta_a n_i=m_i A_i \frac{c_i}{\overline{c}},
\end{equation}
where $A_i$ is defined in Eq. \eqref{eq:Da}.

%The logistic equation for a single genotype or species is given by
%\begin{equation}
%\frac{dN}{dt}=r\left(1-\frac{N}{K}\right)N \label{eq:logistic}
%\end{equation}
%where $N$ is the total population abundance or biomass, $r$ is the growth rate at  low densities, and the ``carrying capacity" $K$ is the value of $N$ where growth ceases.
%
%a simple, ubiquitous model of density-dependent population growth
%declines linearly with density,
%
%Eq. \eqref{eq:logistic} is often interpreted as a linear approximation of a more complicated growth equation
%\begin{equation}
%\frac{dN}{dt}=r(N)N,
%\end{equation}
%where the growth rate $r(N)$ is some unknown, possibly complicated, function of $N$.

%\subsection{Logistic growth and classic nesting sites model}

%In the nesting sites model, it is assumed that propagules landing on occupied territory do not survive, while those landing on unoccupied territory immediately occupy the territory. Thus, the probability that a propagule survives is simply $(1-N/K)$, where $N=\sum_i n_i$ is the total population size. Consequently, $n_i$ increases by $1$ over the interval $[t,t+\Delta t]$ with probability $b_i n_i (1-N/K)\Delta t$. Provided that the abundances $n_i$ are large enough that demographic stochasticity is negligible (i.e. they have established), we can treat $n_i(t)$ as a continuous variable with growth given by the expectation of $b_i n_i (1-N/K)\Delta t$, and so, taking $\Delta t \rightarrow 0$ we obtain
%\begin{equation}
%\frac{dn_i}{dt}=b_i\left(1-\frac{N}{K}\right)n_i. \label{eq:logistic}
%\end{equation}

%\section{Territorial availability}
%
%In Sec. \ref{sec:c} we assumed that adults have the same territorial requirement, regardless of genotype. This can be generalized to allow genotypes to differ in their territorial requirements. Each adult from genotype $i$ now requires $t_i$ units of territory, $T$ is the total territory available to the population, and $T-\sum_i t_i$ units of territory are unoccupied. 
%
%Competition between propagules with different territorial requirements is potentially much more complicated than the model in Sec. \ref{sec:c}, because competition is no longer neatly partitioned into a set of identical territories.  The larger a propagule's territorial requirement, the more neighboring propagules it will interact with during its development to adulthood. 
%
%
%A simple way to avoid these complications is to partition the unoccupied territory into uniform units of territory with size given by the largest territorial requirement $t^*={\rm max}_i t_i$. The uniform territory approach used in Sec.  \ref{sec:c}
%
%The biological intuition behind this 
%
%Thus $U=(T-\sum_i t_i n_i)/t^*$.

\end{document}

### 2016-10-19 simulations.py

@@ -1,0 +1,219 @@
+# -*- coding: utf-8 -*-
+"""
+Created on Thu Mar 24 12:01:54 2016
+
+@author: jbertram
+"""
+
+from numpy import *
+from pylab import *
+import bisect
+
+def expCisim(Ms,U,cs):
+    scatter=[[0 for x in range(len(Ms))] for y in range(U)];
+    for i in range(len(Ms)):
+        for y in xrange(Ms[i]):
+            cell=int(U*rand());
+            scatter[cell][i]=scatter[cell][i]+1;
+    
+    winsnocomp=zeros(len(Ms)); wins1=zeros(len(Ms)); wins2=zeros(len(Ms));
+    comp=zeros(U);
+    for i in range(U):
+        comp[i]=sum(scatter[i])
+        if comp[i]>0:            
+            lotterycmf=cumsum(array(scatter[i])*cs)
+            victor=bisect.bisect(lotterycmf,rand()*lotterycmf[-1])
+            
+            if scatter[i][victor]==1 and comp[i]==1:
+                winsnocomp[victor]=winsnocomp[victor]+1
+            elif scatter[i][victor]==1:
+                wins1[victor]=wins1[victor]+1
+            else: 
+                wins2[victor]=wins2[victor]+1
+        
+    return winsnocomp,wins1,wins2
+
+def deltnu(Ms,U,cs,i):
+    return Ms[i]*exp(-sum(Ms)/float(U))
+
+def deltnr(Ms,U,cs,i):
+    l=Ms[i]/float(U);notl=sum(Ms)/float(U)-l; L=float(l+notl);
+    #return Ms[i]*exp(-l)*(1-exp(-notl))*cs[i]/(sum([Ms[j]*cs[j]/(U*(1-exp(-notl))) for j in range(len(Ms)) if (j!=i)])+cs[i])
+    return Ms[i]*exp(-l)*(1-exp(-notl))*cs[i]/(sum([Ms[j]*cs[j]/(U*(1-(1+L)*exp(-L))) for j in range(len(Ms)) if (j!=i)])*(L-1+exp(-L))/notl+cs[i])
+    
+#def deltna(Ms,U,cs,i):
[…]

### 2016-10-20 manuscript.tex
latest

@@
 \begin{equation}
-\Delta N=U(1-e^{-L})-\sum_i d_i n_i=(K-N)(1-e^{-L})-\overline{d}N,
+\Delta N=U(1-e^{-L})-\sum_i d_i n_i,
 \end{equation}
-where $\overline{d}=\sum_i d_i n_i/N$, which is independent of the competitive ability coefficients $c_i$.
-Thus, the $c_i$ represent a strictly relative aspect of fitness in the sense that they only influence $N$ indirectly by changing genotype frequencies, which may in turn change population mean birth and death rates.
+where $L=\sum_j l_j$ is the overall propagule density.
+$\Delta N$ is independent of the competitive ability coefficients $c_i$; thus, the $c_i$ represent a strictly relative aspect of fitness in the sense that they can only influence $N$ indirectly by changing genotype frequencies (which may change the mean birth and death rates in the population).
 
@@
 \eqref{eq:delttot} gives little intuition about the dynamics of density-dependent lottery competition, since \eqref{eq:growthsumuncoupled} involves an expectation over the random dispersal distributions $p_i$, which depend on how the $n_i$ change over time.
-We now evaluate this expectation approximately, which allows us to develop deeper intuitions about the dynamics of density dependent lottery competition.
+We now evaluate this expectation using a ``mean field'' approximation; the intuition behind this approximation is as follows.
 
-When propagule densities are high ($l_i\gg 1$ for all genotypes), the fluctuations in the $x_i$ are negligible compared to the corresponding expectations $l_i$ (since the $x_i$ are Poisson distributed), and we can make the replacement $x_i\rightarrow l_i$ in Eq.
+When the unoccupied territories are saturated with propagules from every genotype ($l_i\gg 1$ for all genotypes), the fluctuations in the $x_i$ are small compared to their means $l_i$ (since the $x_i$ are Poisson distributed), and so the composition of propagules in a territory will only rarely be appreciably different from the mean composition $l_1,l_2,\ldots,l_G$.
+Consequently, we can replace $x_i$ with $l_i$ in Eq.
 \eqref{eq:growthsumuncoupled}.
-Eq.
-\eqref{eq:growthsumuncoupled} then reduces to the classic lottery model \cite{chesson_1981},
+This gives the classic lottery model \cite{chesson_1981},
 \begin{equation}
@@
 \end{equation}
-where $L=\sum_j l_j=M/U$ is the overall propagule density,  $\overline{c}=\sum_j c_j m_j/M$ is the expected competitive ability for randomly selected propagules, and $M=\sum_j m_j$ is the total number of propagules.
+where $\overline{c}=\sum_j c_j m_j/M$ is the mean propagule competitive ability for a randomly selected propagule (here $M=\sum_j m_j$ is the total number of propagules).
 
-When $l_i\ll 1$, $x_i$ cannot be replaced by $l_i$ in Eq.
-\eqref{eq:growthsumuncoupled}, because $x_i=1$ in the sparsely scattered territories where $i$ propagules do land, meaning that all of the contributions to $\Delta_+ n_i$ are large deviations from the mean.
-Eq.
-\eqref{eq:lottery} is clearly nonsensical in this case: genotype $i$ can win at most $m_i$ territories, yet Eq.
+If the $l_i$ are not all large, the $x_i$ cannot simply be replaced by their means in Eq.
+\eqref{eq:growthsumuncoupled}.
+Indeed, Eq.
+\eqref{eq:lottery} is nonsensical if $l_i$ is sufficiently small: genotype $i$ can win at most $m_i$ territories, yet Eq.
 \eqref{eq:lottery} demands a fraction $c_i m_i/\sum_j c_j m_j$ of the unoccupied territories $U$, no matter how large $U$ is.
+The source of this pathological behavior when $l_i\ll 1$ is that $x_i=1$ in the few territories where $i$ propagules do land, and so $i$'s growth comes entirely from territories which deviate appreciably from the mean.
 
-A similar mean field approximation can be applied to Eq.
-\eqref{eq:growthsumuncoupled} for arbitrary propagule densities provided that the $x_i=1$ terms for the focal genotype are handled separately.
-This approximation relies on the the relative fluctuations in the lottery competition denominator $\sum c_j x_j$ being small, allowing the sum in Eq.
-\eqref{eq:growthsumuncoupled} the be evaluated under a much broader range of conditions than $l_i\gg 1$ for all genotypes.
-In Appendix B we show that this is generally the case, unless there large discrepancy in competitive ability ($c_i/c_j\gg 1$).
-This exception will be discussed below; for now we assume that no such large discrepancy is present.
-We obtain (Appendix B)
+Our mean field approximation is similar to the approximation leading to Eq.
+\eqref{eq:lottery}, except that territories with a single propagule from the focal genotype, which are critical at low densities, are handled separately.
+In place of the requirement of negligible relative fluctuations in $x_i$ for each genotype, we only require that there are no large discrepancies in competitive ability (discussed further below).
+The details of our approximation are contained in Appendix B.
+We obtain
 \begin{equation}
@@
 The third term, $A_ic_i/\overline{c}$, represents competitive victories when the $i$ genotype is abundant: $A_i\rightarrow 0$ if $l_i\ll 1$.
-The relative importance of these three terms changes with both the overall propagule density $L$ and the relative propagule frequencies.
-In the case that all $l_i\gg 1$ for all genotypes, only the last term $A_ic_i/\overline{c}$ remains, and $A_i\rightarrow 1/L$, recovering the classic lottery model.
+The relative importance of these three terms varies with both the overall propagule density $L$ and the relative propagule frequencies $l_i/L$.
+If $l_i\gg 1$ for all genotypes, we recover the classic lottery model (only the $A_ic_i/\overline{c}$ term remains, and $A_i\rightarrow 1/L$).
 
@@
 \eqref{eq:master} (and its components) closely approximate direct simulations of random dispersal with lottery competition over a wide range of propagule densities (obtained by varying $U$).
-A low-frequency genotype with higher $c_i$ is attempting to invade an otherwise identical high-frequency genotype.
+
+A low-frequency genotype with higher $c$ is attempting to invade an otherwise identical high-frequency genotype.
 The low-density invader term $R_i c_i/\overline{c}$ is crucially important for the low-frequency genotype (Fig.

### 2016-10-21 manuscript.tex
latest

@@
 
-\section{Model}
+\section{Model}\label{sec:model}
 
@@
 
-When the unoccupied territories are saturated with propagules from every genotype ($l_i\gg 1$ for all genotypes), the fluctuations in the $x_i$ are small compared to their means $l_i$ (since the $x_i$ are Poisson distributed), and so the composition of propagules in a territory will only rarely be appreciably different from the mean composition $l_1,l_2,\ldots,l_G$.
+If the unoccupied territories are saturated with propagules from every genotype ($l_i\gg 1$ for all genotypes), the fluctuations in the $x_i$ are small compared to their means $l_i$ (since the $x_i$ are Poisson distributed), and so the composition of propagules in a territory will only rarely differ appreciably from the mean composition $l_1,l_2,\ldots,l_G$.
 Consequently, we can replace $x_i$ with $l_i$ in Eq.
@@
 \end{equation}
-where $\overline{c}=\sum_j c_j m_j/M$ is the mean propagule competitive ability for a randomly selected propagule (here $M=\sum_j m_j$ is the total number of propagules).
+where $\overline{c}=\sum_j c_j m_j/M$ is the mean propagule competitive ability for a randomly selected propagule ($M=\sum_j m_j$ is the total number of propagules).
 
-If the $l_i$ are not all large, the $x_i$ cannot simply be replaced by their means in Eq.
+However, in general the $l_i$ are not all large, and the $x_i$ cannot simply be replaced by their means in Eq.
 \eqref{eq:growthsumuncoupled}.
@@
 
-Our mean field approximation is similar to the approximation leading to Eq.
-\eqref{eq:lottery}, except that territories with a single propagule from the focal genotype, which are critical at low densities, are handled separately.
-In place of the requirement of negligible relative fluctuations in $x_i$ for each genotype, we only require that there are no large discrepancies in competitive ability (discussed further below).
-The details of our approximation are contained in Appendix B.
-We obtain
+Our mean field approximation is similar to the high-$l_i$ approximation leading to Eq.
+\eqref{eq:lottery} in that we replace the $x_i$ with appropriate mean values.
+The key distinction is that territories with a single propagule from the focal genotype, which are critical at low densities, are handled separately.
+In place of the requirement of $l_i\gg 1$ for all $i$, our approximation only requires that there are no large discrepancies in competitive ability (discussed further below).
+We obtain (details in Appendix B)
 \begin{equation}
-\Delta_+ n_i(t)\approx b_i n_i\left[e^{-L}+(R_i+A_i)\frac{c_i}{\overline{c}}\right]. \label{eq:master}
+\Delta_+ n_i(t)\approx b_i n_i\left[e^{-L}+(R_i+A_i)\frac{c_i}{\overline{c}}\right], \label{eq:master}
 \end{equation}
@@
 \begin{equation}
-R_i=\frac{\overline{c}e^{-l_i}(1-e^{-(L-l_i)})}{c_i +\frac{\overline{c}L- c_il_i}{1-(1+L)e^{-L}}\frac{L-1+e^{-L}}{L-l_i}},\label{eq:Dr}
+R_i=\frac{\overline{c}e^{-l_i}(1-e^{-(L-l_i)})}{c_i +\frac{L-1+e^{-L}}{1-(1+L)e^{-L}}\frac{\overline{c}L- c_il_i}{L-l_i}},\label{eq:Dr}
 \end{equation}
@@
 If $l_i\gg 1$ for all genotypes, we recover the classic lottery model (only the $A_ic_i/\overline{c}$ term remains, and $A_i\rightarrow 1/L$).
+Thus, Eq.
+\eqref{eq:master} generalizes the classic lottery model to account for arbitrary propagule densities for each genotype.
 
@@
 \ref{fig:simcomp} shows that Eq.
-\eqref{eq:master} (and its components) closely approximate direct simulations of random dispersal with lottery competition over a wide range of propagule densities (obtained by varying $U$).
-
-A low-frequency genotype with higher $c$ is attempting to invade an otherwise identical high-frequency genotype.
-The low-density invader term $R_i c_i/\overline{c}$ is crucially important for the low-frequency genotype (Fig.
-\ref{fig:simcomp}b), but is negligible for the high-frequency one (Fig.
+\eqref{eq:master} (and its components) closely approximate direct simulations of random dispersal and lottery competition over a wide range of propagule densities (obtained by varying $U$).
+Two genotypes are present, one of which has a $c$-advantage and is at low frequency.
+The growth of the low-frequency genotype relies crucially on the low-density competition term $R_i c_i/\overline{c}$, and also to a lesser extent on the high density competition term $A_i c_i/\overline{c}$ if $l_1$ is large enough (Fig.
+\ref{fig:simcomp}b).
+On the other hand, $R_i c_i/\overline{c}$ is negligible for the high-frequency genotype, which depends instead on high density territorial victories (Fig.
 \ref{fig:simcomp}d).
-The high density competition term $A_i c_i/\overline{c}$ is crucial for the high-frequency genotype, and can also be important for the invader if $U$ is small enough.
 
@@
 \includegraphics[scale=0.7]{simulationcomparison.pdf}
-\caption{\label{fig:simcomp} Eq.
-\eqref{eq:master} is a good approximation of the density dependent lottery model.
+\caption{\label{fig:simcomp} The change in genotype abundances in a density dependent lottery model is closely approximated by Eq.
+\eqref{eq:master}.
 $\Delta_+ n_i/m_i$ from Eq.
 \eqref{eq:master} (and its separate components) are shown, along with direct simulations of random dispersal and lottery competition over one generation over a range of propagule densities (varied by changing $U$ with the $m_i$ fixed).
-Two genotypes are present: a low-frequency, better competitor ($m_1/M=0.1$, $c_1=1.5$), in (a) and (b); and a high-frequency, worse competitor ($m_2/M=0.9$, $c_2=1$) in (c) and (d).}
+Two genotypes are present.
+(a) and (b) show low-frequency genotype with $c$-advantage ($m_1/M=0.1$, $c_1=1.5$), (c) and (d) show the high-frequency predominant genotype ($m_2/M=0.9$, $c_2=1$).}
 \end{figure}
-
-Thus, Eq.
-\ref{eq:master} is a density-dependent generalization of the classic lottery model.
-In particular, Eq.
-\ref{eq:master} accounts for the growth (or decline) of initially rare mutant lineages, which is essential for modeling evolution.
 
@@
 
-\subsection{Invasion and coexistence}
+\subsection{Invasion of rare genotypes and coexistence}
 
-Now suppose that a mutation occurs, producing a new genotype $j$ which differs from $i$ in at least one of the traits $b_i$, $c_i$ and $d_i$.
-The new mutant lineage starts with one individual, and remains at low abundance for many generations.
-During this period, the mutant lineage abundance $n_j$ will behave stochastically, and Eq.
-\eqref{eq:master} does not apply.
-However, if $n_j$ becomes large enough, its behavior will become effectively deterministic, and governed by Eq.
+To determine how $b$, $c$ and $d$ will evolve in a population where those traits are being modified by mutations, we need to know whether mutant lineages will grow (or decline) starting from low densities.
+In this section we discuss the behavior of rare genotypes predicted by Eq.
 \eqref{eq:master}.
-For mutants with fitness greater than the population mean fitness, this process is known as ``establishment'', and occurs when $n_j$ is of order $1/s$, where $s$ is the mutant's fitness advantage relative to the mean \cite{desai_2007}.
 
-Here we do not analyze the early stochastic behavior of new mutations, and restrict our attention to the earliest deterministic behavior of rare genotypes.
-In particular, for beneficial mutants we only consider the case where $s$ is large enough that initially we have $n_j \ll N$.
-
-Suppose that a population with a single genotype $i$ is in equilibrium. Then $R^{(i)}_i=0$, $\overline{c}=c_i$ and $\Delta n_i = 0$, which implies
+Suppose that a population with a single genotype $i$ is in equilibrium. Then $R_i=0$, $\overline{c}=c_i$ and $\Delta n_i = 0$, and so Eq. \eqref{eq:master} gives
 \begin{equation}
@@
 \end{equation}
-Now suppose that new genotype $j$, which is initially rare, appears in the population. Then $A_j\ll 1$, $l_j\ll L$ and $\overline{c}\approx c_i$, and so initial growth of $j$ occurs if 
+Now suppose that a new genotype $j$, which is initially rare, appears in the population.
+Then $A_j\ll 1$, $l_j\ll L$ and $\overline{c}\approx c_i$, and so, from Eq.
+\eqref{eq:master}, $n_j$ will increase if
 \begin{equation}
-b_j \left(e^{-L^{(i)}}+R^{(i)}_j\frac{c_j}{c_i}\right)-d_j>0.\label{eq:invad}
+b_j \left(e^{-L}+R_j\frac{c_j}{c_i}\right)-d_j>0.\label{eq:invad}
 \end{equation}
 
-If $j$ is superior in one trait, but otherwise identical to $i$, it will eventually displace $i$.
+Combining Eqs.
+\eqref{eq:equil} and \eqref{eq:invad}, it is easily verified that if $j$ is superior in one trait, but otherwise identical to $i$, it will invade.
+Moreover, $j$ will eventually exclude $i$, since it is strictly superior.
 However, stable coexistence is possible between genotypes that are superior in different traits.
@@
 Coexistence occurs if $j$ will invade an $i$-dominated population, but $i$ will also invade a $j$-dominated population (``mutual invasion'').
-It is not hard to show that this is possible, since if $b_i$ so large that $L\gg 1$ if $i$ is dominant, while $b_j$ is so small that $L\ll 1$ when $j$ is dominant, then, combining Eqs.
+It is not hard to show that this is possible, since if $b_i$ is so large that $L\gg 1$ when $i$ is dominant, and $b_j$ is so small that $L\ll 1$ when $j$ is dominant, then, combining Eqs.
 \eqref{eq:equil} and \eqref{eq:invad}, we find that $j$ invades $i$ if
 \begin{equation}
-b_jc_jR^{(i)}_j-b_i c_i A^{(i)}_i>0, \label{eq:jinvadcoex}
+b_jc_jR_j-b_i c_i A_i>0, \label{eq:jinvadcoex}
 \end{equation}
@@
 Eq.
-\eqref{eq:iinvadcoex} is already satisfied by assumption, while \eqref{eq:jinvadcoex} is satisfied for large enough $c_j$.
-The mechanism for coexistence is that territorial contests are important in $i$-dominated populations (high $L$), ensuring that $j$ is not excluded by $i$, yet territorial contests are irrelevant in $j$-dominated populations (low $L$), ensuring that $i$ is not excluded by $j$.
-
-Fig. \ref{fig:coex} shows an example of coexistence between $b$ and $c$ specialists.
-
- A similar argument applies for coexistence between high-$c$ and low-$d$ specialists.
+\eqref{eq:iinvadcoex} is satisfied by assumption, while Eq.
+\eqref{eq:jinvadcoex} can always be satisfied with large enough $c_j$.
+The mechanism for coexistence is that territorial contests are important in $i$-dominated populations (high $L$), ensuring that the $c$-specialist $j$ is not excluded, yet territorial contests are irrelevant in $j$-dominated populations (low $L$), ensuring that the $b$-specialist $i$ is not excluded.
+Fig.
+\ref{fig:coex} shows an example of this coexistence between $b$ and $c$ specialists.
 
@@
 \caption{\label{fig:coex} Coexistence between $b$ ($c_i=1$, $b_i=1$) and $c$ ($c_j=2$, $b_j=0.7$) specialists, where $d_i=d_j=0.3$.
-Frequency of the $c$-specialist predicted by Eq.
-\eqref{eq:master} is shown.}
+Vertical axis shows frequency of the $c$-specialist predicted by Eq.
+\eqref{eq:master}.}
 \end{figure}
+
+A similar argument applies for coexistence between high-$c$ and low-$d$ specialists; again coexistence occurs because the importance of territorial contests declines along with propagule density $L$ as the $c$-specialist increases in frequency.
+Coexistence is technically possible between $b$- and $d$-specialists which exactly satisfy $b_i/d_i=b_j/d_j$ (this follows from the fact that all propagules have the same success rate when $c_i=c_j$ i.e.
+$A_i+R_i=A_j+R_j$).
+However, this coexistence scenario is not biologically relevant, since the tiniest deviation from $b_i/d_i=b_j/d_j$ will lead to the eventual exclusion of the genotype with greater $b_i/d_i$.
+
+If the rare genotype $j$ arises due to mutation, then it's initial low-density behavior is more complicated than the above invasion analysis suggests.
+The mutant lineage starts with one individual $n_j=1$, and remains at low abundance for many generations after its initial appearance.
+During this period, the mutant abundance $n_j$ will behave stochastically, and the deterministic equations \eqref{eq:growthsumuncoupled} and \eqref{eq:master} do not apply (Section \ref{sec:model}).
+However, if $n_j$ becomes large enough, its behavior will become effectively deterministic, and governed by Eq.
+\eqref{eq:master}.
+For mutants with fitness greater than the population mean fitness, this process is known as ``establishment'', and occurs when $n_j$ is of order $1/s$, where $s$ is the mutant's fitness advantage relative to the mean \cite{desai_2007}.
+Here we do not consider the initial stochastic behavior of novel mutants, and have restricted our attention to the earliest deterministic behavior of rare genotypes.
+In particular, for beneficial mutations we have only considered the case where $s$ is large enough that deterministic behavior starts when $n_j \ll N$.
 
@@
 Actual K selection
+
+Significance of stage structure
 

### 2016-10-21 simulations.py

@@ -154,8 +154,8 @@ plot(nt[:,1],'k--')
 
 #coexistence 
 #=================================================================
-K=100000; bs=array([0.7,1.]); cs=array([2.,1.]); ds=array([.3,.3]);
-ns0=(1-ds)*array([1000,99000]);gens=1000;
+K=100000; bs=array([0.4,.2]); cs=array([1.,1.]); ds=array([.2,.1]);
+ns0=(1-ds)*array([1000,99000]);gens=100;
 
 nhistapprox=[]; Us=[]; Rs=[]; As=[]; ns=ns0
 for i in range(gens):

### 2016-10-21 manuscript.tex
latest

@@
 It is not hard to show that this is possible, since if $b_i$ is so large that $L\gg 1$ when $i$ is dominant, and $b_j$ is so small that $L\ll 1$ when $j$ is dominant, then, combining Eqs.
-\eqref{eq:equil} and \eqref{eq:invad}, we find that $j$ invades $i$ if
+\eqref{eq:equil} and \eqref{eq:invad}, we find that $i$ invades $j$ because $b_i>b_j$, while $j$ invades $i$ provided that
 \begin{equation}
-b_jc_jR_j-b_i c_i A_i>0, \label{eq:jinvadcoex}
+b_jc_jR_j-b_i c_i A_i>0. \label{eq:jinvadcoex}
 \end{equation}
-while $i$ invades $j$ if
-\begin{equation}
-b_i-b_j>0.\label{eq:iinvadcoex}
-\end{equation}
-Eq.
-\eqref{eq:iinvadcoex} is satisfied by assumption, while Eq.
-\eqref{eq:jinvadcoex} can always be satisfied with large enough $c_j$.
-The mechanism for coexistence is that territorial contests are important in $i$-dominated populations (high $L$), ensuring that the $c$-specialist $j$ is not excluded, yet territorial contests are irrelevant in $j$-dominated populations (low $L$), ensuring that the $b$-specialist $i$ is not excluded.
+Thus, coexistence occurs if $c_j$ is large enough.
+Intuitively, the mechanism for coexistence is that territorial contests are important in an $i$-dominated population (high $L$), ensuring that the $c$-specialist $j$ is not excluded, yet territorial contests are irrelevant in a $j$-dominated population (low $L$), ensuring that the $b$-specialist $i$ is not excluded.
 Fig.
@@
 A similar argument applies for coexistence between high-$c$ and low-$d$ specialists; again coexistence occurs because the importance of territorial contests declines along with propagule density $L$ as the $c$-specialist increases in frequency.
-Coexistence is technically possible between $b$- and $d$-specialists which exactly satisfy $b_i/d_i=b_j/d_j$ (this follows from the fact that all propagules have the same success rate when $c_i=c_j$ i.e.
+Coexistence is technically possible between $b$- and $d$-specialists which exactly satisfy $b_i/d_i=b_j/d_j$ (this follows from the fact that all propagules have the same probability of success when $c_i=c_j$ i.e.
 $A_i+R_i=A_j+R_j$).
@@
 
-We can then determine which mutational changes in the traits $b_i, c_i$ and $d_i$ will be favored under different environmental conditions.
+We can now determine which mutational changes in the traits $b, c$ and $d$ will be favored under different environmental conditions.
 Of particular interest are Grime's ``disturbance'', ``stress'' and ``ideal'' environmental archetypes.
@@
 To proceed, we map these archetypes to different parameter regimes in our model.
-The ideal archetype is defined by low extrinsic mortality $d_i\ll 1$ and high birth rates, such that $b_i$ significantly exceeds $d_i$,   population density $N/K$ is high, and thus so is the overall propagule density $L\gg 1$.
-In this case, competition between propagules is important.
+The ideal archetype is defined by low extrinsic mortality $d_i\ll 1$ and high birth rates, such that $b_i$ significantly exceeds $d_i$, and so  population density $N/K$ is high, and $L\gg 1$.
+In this case, territorial contests are important.
 
-The disturbance archetype is defined by unavoidably high extrinsic mortality, usually due to physical destruction of individual organisms, to the extent that the population is kept at low densities $N/K\ll 1$.
+The disturbance archetype is defined by unavoidably high extrinsic mortality due to direct physical destruction of individuals, to the extent that population density is kept low $N/K\ll 1$.
 This means that, even though $b_i$ may be quite high, it cannot greatly exceed $d_i$, and that  propagule densities are low $l_i\ll 1$.

### 2016-10-21 manuscript.tex
latest

@@
 
-The change in total population size $N$, obtained by summing Eq. \eqref{eq:delttot} over genotypes, is
+Note that the competitive ability coefficients $c_i$ represent a strictly relative aspect of fitness in the sense that they only influence population size $N$ indirectly by changing genotype frequencies, which may in turn change the population mean birth and death rates.
+This can be seen by summing Eq.
+\eqref{eq:delttot} over genotypes to get the  change in population size $N$,
 \begin{equation}
@@
 \end{equation}
-where $L=\sum_j l_j$ is the overall propagule density.
-$\Delta N$ is independent of the competitive ability coefficients $c_i$; thus, the $c_i$ represent a strictly relative aspect of fitness in the sense that they can only influence $N$ indirectly by changing genotype frequencies (which may change the mean birth and death rates in the population).
+where $L=\sum_j l_j$ is the overall propagule density; $\Delta N$ is independent of the $c_i$.
 
@@
 
-We can now determine which mutational changes in the traits $b, c$ and $d$ will be favored under different environmental conditions.
+We now discuss which mutational changes in the traits $b, c$ and $d$ will be favored under different environmental conditions.
 Of particular interest are Grime's ``disturbance'', ``stress'' and ``ideal'' environmental archetypes.
+To proceed, we map these archetypes to different parameter regimes in our model.
 
-To proceed, we map these archetypes to different parameter regimes in our model.
-The ideal archetype is defined by low extrinsic mortality $d_i\ll 1$ and high birth rates, such that $b_i$ significantly exceeds $d_i$, and so  population density $N/K$ is high, and $L\gg 1$.
-In this case, territorial contests are important.
+The disturbance archetype is characterized by unavoidably high extrinsic mortality caused by the direct physical destruction of individuals, to the extent that most territories are not used most of the time ($N/K\ll 1$).
 
-The disturbance archetype is defined by unavoidably high extrinsic mortality due to direct physical destruction of individuals, to the extent that population density is kept low $N/K\ll 1$.
+, and reproduction is a race against time
+
 This means that, even though $b_i$ may be quite high, it cannot greatly exceed $d_i$, and that  propagule densities are low $l_i\ll 1$.
@@
 \eqref{eq:master} are then negligible, and $\Delta_+ n_i$ depends primarily on $b_i$.
+
+The ideal environmental archetype is characterized by the absence of stress and disturbance, low extrinsic mortality and high [BLAH] such that $b_i$ significantly exceeds $d_i$, and so $N/K$ is high, and $L\gg 1$.
+In this case, territorial contests are important.
 

### 2016-10-24 manuscript.tex
latest

@@
 All territories are identical, and the total number of territories is $K$.
-Time $t$ advances in discrete steps, each representing one generation.
-In generation $t$, the number of mature individuals (henceforth called ``adults'') of the $i$'th genotype is $n_i(t)$, the total number of adults is $N(t)=\sum_i n_i(t)$, and the number of unoccupied territories $U(t)=K-N(t)$.
+Time $t$ advances in discrete steps, each step representing one generation.
+In generation $t$, the number of mature individuals (henceforth called ``adults'') of the $i$'th genotype is $n_i(t)$, the total number of adults is $N(t)=\sum_i n_i(t)$, and the number of unoccupied territories is $U(t)=K-N(t)$.
 
-Each generation, adults produce $m_i$ new offspring (henceforth called ``propagules'') which disperse at random over the $U$ unnocupied territories.
+Each generation, adults produce $m_i$ new offspring (henceforth called ``propagules'') which disperse at random over the $U$ unnocupied territories (there is no dispersal limitation).
 We ignore propagules that land on occupied territories --- these are assumed to have no chance of survival and are not included in $m_i$.
@@
 Although $K$ is finite in our model, we assume that $K$ and the $n_i$ are large enough that $x_i$ is Poisson-distributed to a good approximation (details in Appendix A).
-This Poisson distribution is denoted $p_i(x_i)$.
-In particular, we do not consider the stochastic behavior that occurs at very low $n_i$ due to the lottery nature of territorial acquisition (this will be discussed further when we consider the behavior of mutant lineages).
+This dispersal Poisson distribution is denoted $p_i(x_i)$.
+Note that the large $n_i$, large $K$ approximation places no restrictions on our densities $n_i/K$, but it does preclude consideration of demographic stochasticity when $n_i$ itself is very small (this will be discussed further in Section \ref{sec:invas}).
 
@@
 
-In addition to propagule birth and competition, occupied territories become unoccupied when adults die.
-We assume that these adult deaths occur at a constant, genotype-specific rate $d_i$.
-Thus, the overall change in genotype abundances is
+In addition to propagule birth and competition, occupied territories become unoccupied due to mortality.
+For the vast majority of this manuscript we assume that mortality only occurs in adults, and at a constant, genotype-specific per-capita rate $d_i$, so that the overall change in genotype abundances is
 \begin{equation}
@@
 \end{equation}
+We will introduce a different mortality model when we consider the effects of disturbances (Section \ref{sec:grime}), which will also affect competing juveniles.
 
-Note that the competitive ability coefficients $c_i$ represent a strictly relative aspect of fitness in the sense that they only influence population size $N$ indirectly by changing genotype frequencies, which may in turn change the population mean birth and death rates.
+Note that the competitive ability coefficients $c_i$ represent a strictly relative aspect of fitness in the sense that they only influence population size $N$ indirectly by changing genotype frequencies; that may in turn change the population mean birth and death rates.
 This can be seen by summing Eq.
@@
 \begin{equation}
-\Delta N=U(1-e^{-L})-\sum_i d_i n_i,
+\Delta N=U(1-e^{-L})-\sum_i d_i n_i,\label{eq:deltN}
 \end{equation}
-where $L=\sum_j l_j$ is the overall propagule density; $\Delta N$ is independent of the $c_i$.
+which is independent of $c_i$ (here $L=\sum_j l_j$ is the overall propagule density).
 
@@
 
-\subsection{Invasion of rare genotypes and coexistence}
+\subsection{Invasion of rare genotypes and coexistence}\label{sec:invas}
 
@@
 
-\subsection{Environmental archetypes and Grime's triangle}
+\subsection{Environmental archetypes and Grime's triangle}\label{sec:grime}
 
-We now discuss which mutational changes in the traits $b, c$ and $d$ will be favored under different environmental conditions.
+We now discuss which changes in the traits $b, c$ and $d$ will be most favored under different environmental conditions.
 Of particular interest are Grime's ``disturbance'', ``stress'' and ``ideal'' environmental archetypes.
-To proceed, we map these archetypes to different parameter regimes in our model.
+To proceed, we need to map these verbal archetypes to quantitative parameter regimes in our model.
 
-The disturbance archetype is characterized by unavoidably high extrinsic mortality caused by the direct physical destruction of individuals, to the extent that most territories are not used most of the time ($N/K\ll 1$).
+The ideal environmental archetype is characterized by the near-absence of stress and disturbance.
+Consequently, $d_i\ll 1$, whereas $b_i$ can easily be of order $1$ or larger.
+From Eq.
+\eqref{eq:deltN}, the equilibrium value of $L$ only depends on the ratio of birth and death rates.
+For one genotype, $L/(1-e^{-L})=b_i/d_i$, and so the propagule density is high $L\approx b_i/d_i\gg 1$.
+Moreover, since $L=b_i N/(N-K)$ by definition, population density is also high $N/K\approx 1$.
+Thus, territorial contests are decisively important.
 
-, and reproduction is a race against time
-
-This means that, even though $b_i$ may be quite high, it cannot greatly exceed $d_i$, and that  propagule densities are low $l_i\ll 1$.
+The disturbance archetype is characterized by unavoidably high extrinsic mortality caused by physical destruction.
+Disturbances do not only affect adults as in Eq.
+\eqref{eq:delttot}, but also juveniles in the process of territorial contest.
+These juvenile deaths can be represented as a fractional reduction in the number of territories secured.
+To illustrate, we assume that the disturbance is equally damaging to adults and juveniles, so that only $(1-d_i)\Delta_+ n_i$ rather than $\Delta_+ n_i$ territories are secured by genotype $i$ each generation.
+Then, the disturbance archetype is characterized by $d_i$ being close to $1$ (almost all adults and juveniles are killed each generation), which can only be tolerated by those organisms with large enough $b_i$ to ensure population persistence.
+The single genotype equilibrium then gives $L\approx 2(1-d_i/[(1-d_i)b_i])$, and we have $L\ll 1$ and $N/K\ll 1$.
 The terms proportional to $c_i/\overline{c}$ in Eq.
@@
 
-The ideal environmental archetype is characterized by the absence of stress and disturbance, low extrinsic mortality and high [BLAH] such that $b_i$ significantly exceeds $d_i$, and so $N/K$ is high, and $L\gg 1$.
-In this case, territorial contests are important.
-
-The stress archetype is more ambiguous, and has been the subject of an extensive debate in the plant ecology literature.
+The stress archetype is more ambiguous, and has been the subject of an extensive debate in the plant ecology literature (the ``Grime-Tilman'' debate \cite{aerts_1999}).
 In Grime's view, the archetypal stressful environmental imposes such severe challenges that success depends primarily on the ability to tolerate the stressors.
 Population density $N/K$ is suppressed to such low levels that competition between individuals is not important.
-In our model, this again corresponds to $b_i$ only slightly exceeding $d_i$ and $l_i\ll 1$, similar to the disturbance archetype, except that stress now constrains $b_i$ to be small.
+In our model, this corresponds to $b_i\ll 1$ and $b_i/d_i\approx 1$.
+Thus $L\ll 1$ and $N/K\ll 1$ similar to the disturbance archetype, except that stress now constrains $b_i$ to be small.
 
-The alternative view is that the stress archetype should rather be interpreted as a large reduction in the maximum number of individuals that can be supported \cite{taylor_1990}.
+The alternative view is that the stress archetype should rather be interpreted as a large reduction in the maximum number of individuals that can be supported by the environment \cite{taylor_1990}.
 For example, in the case that the stress is induced by a scarcity of consumable resources, competition for resources would likely be intense, and the stressed population should actually be regarded as having a high population density.
-In our model, this would imply a large reduction in $K$ (greater per-individual territorial requirement), such that even though fewer individuals are supported, density $N/K$ is still high.
+In our model, this would imply a large reduction in $K$ (greater per-individual territorial requirement).
+That is, $N$ under stress is much lower than  under ideal conditions, but it is not much lower than $K$ for the stressful environment.
+Since our model accommodates both of these alternatives, we include them both here.
 
-The mapping of environmental achetypes to our model parameters is summarized in the first three rows of Fig.
+The mapping of environmental achetypes to our model parameters is summarized in the first two rows of Fig.
 \ref{fig:table}.
-Also shown is the approximate dependence of $\Delta_+ n_i$ on $b_i$ and $c_i$ for each archetype (fourth row), which can be used infer the expected direction of evolution for the traits $b_i$, $c_i$ and $d_i$ (fifth row).
+Also shown is the approximate dependence of $\Delta_+ n_i$ on $b_i$ and $c_i$ for each archetype (third row), which can be used infer the expected direction of evolution for the traits $b$, $c$ and $d$ (fourth row).
 
-The latter is inferred as follows.
-As noted in the previous section, if beneficial mutants can survive the low-abundance stochastic regime, their behavior is governed deterministically by Eq.
+The latter is obtained as follows.
+As noted in the previous section, if beneficial mutations can survive the low-abundance stochastic regime, their behavior is governed deterministically by Eq.
 \eqref{eq:master}.
-If their advantage in $b$, $c$ or $d$ is large enough to overcome the lottery rarity disadvantage, they will proceed to grow deterministically (establishment).
-The probability of establishment increases with the mutant fitness advantage, and is of order $1/N$ for neutral mutations.
-Consequently, the direction of evolutionary change is determined by which trait changes confer an appreciable benefit, subject to the constraints imposed by the environmental.
+They will then proceed to grow deterministically (establishment).
+The probability of establishment increases with the mutant fitness advantage, and is therefore typically on the order of a few percent, whereas the fixation of neutral mutations is exceedingly unlikely (probability of order $1/N$).
+Consequently, the direction of evolutionary change is determined by which trait changes confer an appreciable benefit, subject to the constraints imposed by the environment.
 
 For example, in Grime's version of the stress archetype, population density is low, so competition is not important, and so only mutants with greater $b$ or lower $d$ will have an appreciably greater $\Delta n_i$.
-Mutations in $c$ are effectively neutral, and will rarely establish.
-However, by definition of the stress archetype, $b$ is constrained to be small.
+Mutations in $c$ are effectively neutral, and will rarely fix.
+However, by definition of the stress archetype, $b$ is constrained to be very small.
 Thus, while some rare mutations may produce small improvements in $b$, it is much more likely that mutations will arise that lower $d$, making this the expected direction of evolutionary change for Grime's stress archetype.
@@
 \begin{tabular}{*{5}{c}}
-  & Ideal & Disturbance & Stress (G) & Stress (K) \\ \hline
-  Parameter- & $d_i \ll 1$ & $d_i \gg 1$ & $b_i \ll 1$ & $b_i \ll 1$ \\
-  regime & $b_i>d_i$ & $b_i\approx d_i$ & $b_i\approx d_i$ & $b_i>  d_i$ \\
-  Density  & High & Low & Low & High \\
+  & Ideal & Disturbance* & Stress (G) & Stress (K) \\ \hline
+  Parameter- & $d_i \ll 1$ & $d_i \approx 1$ & $b_i \ll 1$ & $b_i \ll 1$ \\
+  regime & $b_i/d_i\gg 1$ & $b_i/d_i\gg 1$ & $b_i\approx d_i$ & $b_i>  d_i$ \\
+  Density $N/K$  & High & Low & Low & High \\
   $\Delta_+ n_i\propto$ & $b_i c_i$ & $b_i$ & $b_i$ & $b_i c_i$ \\
-  Evolution & $\uparrow b$, $ \uparrow c$ & $\uparrow b$ & $\downarrow d$ & $\uparrow c$
+  Evolution for & $\uparrow b$, $ \uparrow c$ & $\uparrow b$ & $\downarrow d$ & $\uparrow c$
 \end{tabular}
 \caption{\label{fig:table} The realization of Grime's environmental archetypes in our model, as well as the low-$K$ variant of the stress archetype.
-Shown are the mapping to our parameters of each archetype, the approximate dependence of $\Delta_+ n_i$ on $b_i$ and $c_i$, as well as the corresponding expected evolutionary changes in $b_i$, $c_i$ and $d_i$.}
+Shown are the mapping to our parameters of each archetype, the approximate dependence of $\Delta_+ n_i$ on $b_i$ and $c_i$, as well as the corresponding expected evolutionary changes in $b_i$, $c_i$ and $d_i$.
+*Mortality affects both adults and juveniles in the disturbance archetype, with $\Delta_+ n_i$ replaced by $(1-d_i)\Delta_+ n_i$ in Eq.
+\eqref{eq:delttot}.}
 \end{figure}

### 2016-10-25 manuscript.tex
latest

@@
 
-By contrast models of phenotypic trait evolution represent the change in phenotypic abundances over time using absolute fitness functions which describe how those traits affect survival and reproduction in particular ecological scenarios.
+By contrast, models of phenotypic trait evolution represent the change in phenotypic abundances over time using absolute fitness functions which describe how those traits affect survival and reproduction in particular ecological scenarios.
 This approach is powerful enough to model eco-evolutionary feedbacks between co-evolving traits, but is generally problem-specific and restricted to only a few traits at a time.
@@
 We broadly follow the spirit of MacArthur's r/K selection scheme in that our model is intended to account for fundamentally different forms of selection without getting  entangled in the intricacies of particular ecological scenarios.
-However, rather than building directly on MacArthur's formalism and its later extensions [cite alpha stuff] (e.g.
+However, rather than building directly on MacArthur's formalism and its later extensions \cite{joshi_2001} (e.g.
 using Lotka-Volterra equations), our model is devised primarily with Grime's C/S/R scheme in mind, and represents a quantitative formalization of how C/S/R manifests at the level of genotype evolution (as opposed to divergence between species).
-This choice is motivated in part by the substantial empirical support for the C/S/R scheme, and in part by the failings of the r/K low/high density dichotomy --- many growth ability traits will confer advantages at both low and high densities, in which case r- and K- selection will effectively coincide (empirically, positive correlations between measures of r and K are common, both between species and strains \cite{luckinbill_1979,kuno_1991,hendriks_2005,fitzsimmons_2010}, and as a result of experimental evolution \cite{luckinbill_1979,luckinbill_1978}).
+This choice is motivated in part by the substantial empirical support for the C/S/R scheme, and in part by the failings of the r/K low/high density dichotomy --- many growth ability traits will confer advantages at both low and high densities, in which case r- and K- selection will effectively coincide (more details in the Discussion).
 
@@
 
+Our model differs from both Grime's C/S/R and MacArthur's $r$/$K$ schemes in the role of the propagule production rate $b$, a measure of intrinsic fecundity closely related to the growth rate at low densities $r=b-d$.
+In both of those schemes, the essential feature of life at high densities is competition.
+This is less  in the $r$/$K$ scheme, which does not explicitly  it is better to have a
+
+contributes just as much to
+
+In the Introduction, we noted that the $r$-$K$ dichotomy is not consistent with ``$K$-selection'' (i.e. selection at high densities) for growth ability traits.
+
+Specifically, positive correlations between measures of $r$ and $K$ are common, both between species and strains \cite{luckinbill_1979,kuno_1991,hendriks_2005,fitzsimmons_2010}, and as a result of experimental evolution \cite{luckinbill_1979,luckinbill_1978}).
+From the perspective of our model, this correlation is not at all surprising;
+
 r-K correlation, meaning of K selection

### 2016-10-25 manuscript.tex
latest

@@
-\documentclass[12pt]{article}
+\documentclass[11pt]{article}
+\usepackage[sc]{mathpazo}
+\usepackage{amsmath}
+\usepackage{fullpage}
+\usepackage[authoryear,sectionbib,sort]{natbib}
+\linespread{1.7}
+\usepackage[utf8]{inputenc}
+\usepackage{lineno}
 
 \usepackage{graphicx} 
-\usepackage{amsmath}
+%\usepackage{amsmath}
 %\usepackage{a4wide}
@@
 
-\begin{document}\sloppy
+\title{Density-dependent selection in evolutionary genetics: a lottery model of Grime's triangle}
+\author{Jason Bertram $^{1,\ast}$ \\ 
+Joanna Masel $^{1}$}
 
-\title{A preliminary rcK derivation}
-\author{J Bertram and J Masel}
+\date{}
+
+\begin{document}
+
 \maketitle
+
+\noindent{}1. Department of Ecology and Evolutionary Biology, University of Arizona, Tucson, AZ 85721.
+
+\noindent{}$\ast$ Corresponding author; e-mail: amnat@uchicago.edu.
+
+\bigskip
+
+\textit{Manuscript elements}: Figure~1, figure~2, table~1, online
+appendices~A and B (including figure~A1 and figure~A2). Figure~2 is to
+print in color.
+
+\bigskip
+
+\textit{Keywords}: Examples, model, template, guidelines.
+
+\bigskip
+
+\textit{Manuscript type}: Article. 
+% Or e-article, note, e-note, natural history miscellany,
+% e-natural history miscellany, comment, reply, symposium, or
+% countdown to 150.
+
+\bigskip
+
+\noindent{\footnotesize Prepared using the suggested \LaTeX{} 
+template for \textit{Am.\ Nat.}}
+
+\linenumbers{}
+\modulolinenumbers[3]
+
+\newpage{}
+
+\section*{Abstract}
+
+Some abstract stuff that runs over multiple lines so that I can see the quote formatting.
+
+\newpage{}
 
@@
 
-\section{Introduction}
+\section*{Introduction}
 
@@
 We broadly follow the spirit of MacArthur's r/K selection scheme in that our model is intended to account for fundamentally different forms of selection without getting  entangled in the intricacies of particular ecological scenarios.
-However, rather than building directly on MacArthur's formalism and its later extensions \cite{joshi_2001} (e.g.
-using Lotka-Volterra equations), our model is devised primarily with Grime's C/S/R scheme in mind, and represents a quantitative formalization of how C/S/R manifests at the level of genotype evolution (as opposed to divergence between species).
+However, rather than building directly on MacArthur's formalism and its later extensions using Lotka-Volterra equations to incorporate competition (``$\alpha$-selection'') \cite{gill_1974,case_1974,joshi_2001}, our model is devised primarily with Grime's C/S/R scheme in mind, and represents a quantitative formalization of how C/S/R manifests at the level of genotype evolution (as opposed to divergence between species).
 This choice is motivated in part by the substantial empirical support for the C/S/R scheme, and in part by the failings of the r/K low/high density dichotomy --- many growth ability traits will confer advantages at both low and high densities, in which case r- and K- selection will effectively coincide (more details in the Discussion).
@@
 All territories are identical, and the total number of territories is $K$.
-Time $t$ advances in discrete steps, each step representing one generation.
-In generation $t$, the number of mature individuals (henceforth called ``adults'') of the $i$'th genotype is $n_i(t)$, the total number of adults is $N(t)=\sum_i n_i(t)$, and the number of unoccupied territories is $U(t)=K-N(t)$.
+Time $t$ advances in discrete iterations, each representing the average time from birth to reproductive maturity.
+In iteration $t$, the number of reproductively mature individuals (henceforth called ``adults'') of the $i$'th genotype is $n_i(t)$, the total number of adults is $N(t)=\sum_i n_i(t)$, and the number of unoccupied territories is $U(t)=K-N(t)$.
 
-Each generation, adults produce $m_i$ new offspring (henceforth called ``propagules'') which disperse at random over the $U$ unnocupied territories (there is no dispersal limitation).
-We ignore propagules that land on occupied territories --- these are assumed to have no chance of survival and are not included in $m_i$.
+Each iteration, adults produce $m_i$ new offspring (henceforth called ``propagules'') which disperse at random over the $U$ unnocupied territories (no dispersal limitation).
+We assume adults cannot be ousted from occupied territories, so only propagules landing on occupied territories are included in $m_i$.
+More generally, $m_i$ does not include propagules which never even begin the development cycle.
 For simplicity, we assume $m_i=b_i n_i$, where $b_i$ is a constant, genotype-specific birth rate.
@@
 When multiple propagules land on the same territory, they compete to secure the territory as they develop.
-This territorial contest is modeled as a weighted lottery: the probability that genotype $i$ wins a given territory by the next generation is $c_i x_i/\sum_j c_j x_j$ where $c_i$ is a constant representing relative competitive ability.
+This territorial contest is modeled as a weighted lottery: the probability that genotype $i$ wins a given territory by the next iteration is $c_i x_i/\sum_j c_j x_j$ where $c_i$ is a constant representing relative competitive ability.
 
-The increase in $n_i$ over one generation due to territorial acquisition, $\Delta_+ n_i$, is the sum of genotype $i$'s victories over all $U$ unoccupied territories.
+The increase in $n_i$ over one iteration due to territorial acquisition, $\Delta_+ n_i$, is the sum of genotype $i$'s victories over all $U$ unoccupied territories.
 Since $p_1(x_1)\ldots p_G(x_G)$ is equal to the proportion of unoccupied territories with $x_1,\ldots,x_G$ of the respective propagules (again, we assume that $K$ is large enough that fluctuations in this proportion are negligible), this sum can be replaced by an expectation over the $p_i$.
@@
 $\Delta_+ n_i/m_i$ from Eq.
-\eqref{eq:master} (and its separate components) are shown, along with direct simulations of random dispersal and lottery competition over one generation over a range of propagule densities (varied by changing $U$ with the $m_i$ fixed).
+\eqref{eq:master} (and its separate components) are shown, along with direct simulations of random dispersal and lottery competition over one iteration over a range of propagule densities (varied by changing $U$ with the $m_i$ fixed).
 Two genotypes are present.
@@
 The ideal environmental archetype is characterized by the near-absence of stress and disturbance.
-Consequently, $d_i\ll 1$, whereas $b_i$ can easily be of order $1$ or larger.
+Consequently, $d_i\ll 1$, whereas $b_i$ is potentially much larger than $1$.
 From Eq.
@@
 These juvenile deaths can be represented as a fractional reduction in the number of territories secured.
-To illustrate, we assume that the disturbance is equally damaging to adults and juveniles, so that only $(1-d_i)\Delta_+ n_i$ rather than $\Delta_+ n_i$ territories are secured by genotype $i$ each generation.
-Then, the disturbance archetype is characterized by $d_i$ being close to $1$ (almost all adults and juveniles are killed each generation), which can only be tolerated by those organisms with large enough $b_i$ to ensure population persistence.
-The single genotype equilibrium then gives $L\approx 2(1-d_i/[(1-d_i)b_i])$, and we have $L\ll 1$ and $N/K\ll 1$.
+To illustrate, we assume that the disturbance is equally damaging to adults and juveniles, so that only $(1-d_i)\Delta_+ n_i$ rather than $\Delta_+ n_i$ territories are secured by genotype $i$ each iteration.
+Then, the disturbance archetype is characterized by $d_i$ being close to $1$ for all genotypes (almost all adults and juveniles are killed each iteration).
+The single genotype equilibrium then gives $L\approx 2(1-d_i/[(1-d_i)b_i])$, where $b_i$ must be exceptionally large to ensure population persistence, and we have $L\ll 1$ and $N/K\ll 1$.
 The terms proportional to $c_i/\overline{c}$ in Eq.
@@
 The stress archetype is more ambiguous, and has been the subject of an extensive debate in the plant ecology literature (the ``Grime-Tilman'' debate \cite{aerts_1999}).
-In Grime's view, the archetypal stressful environmental imposes such severe challenges that success depends primarily on the ability to tolerate the stressors.
-Population density $N/K$ is suppressed to such low levels that competition between individuals is not important.
-In our model, this corresponds to $b_i\ll 1$ and $b_i/d_i\approx 1$.
-Thus $L\ll 1$ and $N/K\ll 1$ similar to the disturbance archetype, except that stress now constrains $b_i$ to be small.
+In Grime's view, stressful environments impose such severe challenges that surviving the stressors at all is the primary challenge.
+In our model, this can be expressed by $b_i\ll 1$ and $b_i/d_i\approx 1$.
+The propagule density $L$ (as well as $N/K$) are suppressed to such low levels that there is essentially no competition between individuals.
+The severity of the stress is such that mutations which appreciably improve $b_i$ are extremely rare, so $b_i$ is constrained to remain low.
 
 The alternative view is that the stress archetype should rather be interpreted as a large reduction in the maximum number of individuals that can be supported by the environment \cite{taylor_1990}.
-For example, in the case that the stress is induced by a scarcity of consumable resources, competition for resources would likely be intense, and the stressed population should actually be regarded as having a high population density.
+For example, in the commonly cited case that the stress is induced by a scarcity of consumable resources, competition for resources would likely be intense, and the stressed population should actually be regarded as having a high population density.
 In our model, this would imply a large reduction in $K$ (greater per-individual territorial requirement).
@@
   $\Delta_+ n_i\propto$ & $b_i c_i$ & $b_i$ & $b_i$ & $b_i c_i$ \\
-  Evolution for & $\uparrow b$, $ \uparrow c$ & $\uparrow b$ & $\downarrow d$ & $\uparrow c$
+  Evolution for & $\uparrow b$, $ \uparrow c$ & $\uparrow b$, $\downarrow d$ & $\downarrow d$ & $\uparrow c$
 \end{tabular}
@@
 
+How does Fig. 4 compare to Grime's C/S/R strategies? In disturbed environments, the ruderal (R) strategy is characterized by high fecundity, effective dispersal and short generation times.
+
+We predict evolution for higher $b$, but not higher $c$ or lower $d$ (constrained to be large), amounts to focusing on fecundity.
+While there is no dispersal limitation in our model, propagules must be minimally effective at dispersal since our model only includes propagules which will enter territorial contests on unoccupied sites;
+
+ Shorter generation times are thought to be a mechanism for increasing the chance that the reproductive cycle completes before disturbance-induced death
+
+The evolution of the generation time itself is beyond the scope of our model, but it could
+
 \section{Discussion}
-
-Grime's triangle without constraints
 

### 2016-10-26 manuscript.tex
latest
merge hell

@@
 In models of genetic evolution, genotypes are typically assigned constant (or frequency-dependent) selection coefficients describing the change in their relative frequencies over time due to differences in viability.
-This considerably simplifies the mathematics of selection, facilitating greater genetic realism, and can be justified over sufficiently short time intervals \cite[p.
+This considerably simplifies the mathematics of selection, facilitating greater genetic realism, and can be justified over sufficiently short time intervals \citep[p.
 276]{ewens_2012}.
@@
 Far less work has been done to model fitness in more general terms than particular traits or ecological scenarios, while still capturing key distinctions between different forms of selection.
-Perhaps this is not surprising given that fitness is such a complex quantity, dependent on all of a phenotype's functional traits \cite{violle_2007} as well as its biotic and abiotic environment.
+Perhaps this is not surprising given that fitness is such a complex quantity, dependent on all of a phenotype's functional traits \citep{violle_2007} as well as its biotic and abiotic environment.
 In most cases, a detailed, trait-based, predictive model of fitness would be enormously complicated and have narrow applicability.
-It is therefore easy to doubt the feasibility of a simplified, general mathematical treatment of fitness \cite[p.
+It is therefore easy to doubt the feasibility of a simplified, general mathematical treatment of fitness \citep[p.
 276]{ewens_2012}.
-Even MacArthur's famous r/K selection scheme is now almost exclusively known as a framework for understanding life-history traits, and judged on its failure in that role \cite{pianka_1970,stearns_1977,boyce_1984,reznick_2002}.
-In spite of the r/K scheme's original purpose as an extension of the existing population-genetic treatment of selection to account for population density \cite{macarthur_1962}, comparatively few attempts have been made to develop it further as a mathematical analysis of the major different forms of selection.
+Even MacArthur's famous r/K selection scheme is now almost exclusively known as a framework for understanding life-history traits, and judged on its failure in that role \citep{pianka_1970,stearns_1977,boyce_1984,reznick_2002}.
+In spite of the r/K scheme's original purpose as an extension of the existing population-genetic treatment of selection to account for population density \citep{macarthur_1962}, comparatively few attempts have been made to develop it further as a mathematical analysis of the major different forms of selection.
 
 Nevertheless, there are strong indications there are broader principles governing the operation of selection.
-In many groups of organisms (including corals \cite{darling_2012}, insects \cite{southwood_1977}, fishes \cite{winemiller_1992} and plants \cite{grime_1988}), species can be partitioned into a small number of distinct trait clusters corresponding to fundamentally distinct ``primary strategies'' \cite{winemiller_2015}.
-The most famous example is Grime's plant trait classification scheme \cite{grime_1974,grime_1977,grime_1988}.
+In many groups of organisms (including corals \citep{darling_2012}, insects \citep{southwood_1977}, fishes \citep{winemiller_1992} and plants \citep{grime_1988}), species can be partitioned into a small number of distinct trait clusters corresponding to fundamentally distinct ``primary strategies'' \citep{winemiller_2015}.
+The most famous example is Grime's plant trait classification scheme \citep{grime_1974,grime_1977,grime_1988}.
 Grime considered two broad determinants of population density: stress (persistent hardship e.g.
@@
 Grime showed that measures of C, S and R across a wide range of plant species are anti-correlated, so that strong C-strategists are weak S and R strategists, and so on.
-Thus, plant species can be classified on a triangular C/S/R ternary plot \cite{grime_1974}.
-Trait classification schemes for other organisms closely parallel Grime's scheme \cite{winemiller_2015}.
+Thus, plant species can be classified on a triangular C/S/R ternary plot \citep{grime_1974}.
+Trait classification schemes for other organisms closely parallel Grime's scheme \citep{winemiller_2015}.
 
@@
 However, while trait classification schemes are firmly grounded in trait data, they are verbal and descriptive rather than mathematical, a recognized hinderance to their broader applicability (e.g.
-\cite{tilman_2007}).
+\citep{tilman_2007}).
 
@@
 We broadly follow the spirit of MacArthur's r/K selection scheme in that our model is intended to account for fundamentally different forms of selection without getting  entangled in the intricacies of particular ecological scenarios.
-However, rather than building directly on MacArthur's formalism and its later extensions using Lotka-Volterra equations to incorporate competition (``$\alpha$-selection'') \cite{gill_1974,case_1974,joshi_2001}, our model is devised primarily with Grime's C/S/R scheme in mind, and represents a quantitative formalization of how C/S/R manifests at the level of genotype evolution (as opposed to divergence between species).
+However, rather than building directly on MacArthur's formalism and its later extensions using Lotka-Volterra equations to incorporate competition (``$\alpha$-selection'') \citep{gill_1974,case_1974,joshi_2001}, our model is devised primarily with Grime's C/S/R scheme in mind, and represents a quantitative formalization of how C/S/R manifests at the level of genotype evolution (as opposed to divergence between species).
 This choice is motivated in part by the substantial empirical support for the C/S/R scheme, and in part by the failings of the r/K low/high density dichotomy --- many growth ability traits will confer advantages at both low and high densities, in which case r- and K- selection will effectively coincide (more details in the Discussion).
@@
 
-\section{Model}\label{sec:model}
+\section*{The Model}
 
@@
 
-\section{Results}
+\section*{Results}
 
-\subsection{Mean Field Approximation}
+\subsection*{Mean Field Approximation}
 
@@
 \eqref{eq:growthsumuncoupled}.
-This gives the classic lottery model \cite{chesson_1981},
+This gives the classic lottery model \citep{chesson_1981},
 \begin{equation}
@@
 
-\subsection{Invasion of rare genotypes and coexistence}\label{sec:invas}
+\subsection*{Invasion of rare genotypes and coexistence}\label{sec:invas}
 
@@
 \eqref{eq:master}.
-For mutants with fitness greater than the population mean fitness, this process is known as ``establishment'', and occurs when $n_j$ is of order $1/s$, where $s$ is the mutant's fitness advantage relative to the mean \cite{desai_2007}.
+For mutants with fitness greater than the population mean fitness, this process is known as ``establishment'', and occurs when $n_j$ is of order $1/s$, where $s$ is the mutant's fitness advantage relative to the mean \citep{desai_2007}.
 Here we do not consider the initial stochastic behavior of novel mutants, and have restricted our attention to the earliest deterministic behavior of rare genotypes.
@@
 
-\subsection{Environmental archetypes and Grime's triangle}\label{sec:grime}
+\subsection*{Environmental archetypes and Grime's triangle}
 
@@
 
-The stress archetype is more ambiguous, and has been the subject of an extensive debate in the plant ecology literature (the ``Grime-Tilman'' debate \cite{aerts_1999}).
+The stress archetype is more ambiguous, and has been the subject of an extensive debate in the plant ecology literature (the ``Grime-Tilman'' debate \citep{aerts_1999}).
 In Grime's view, stressful environments impose such severe challenges that surviving the stressors at all is the primary challenge.
@@
 
-The alternative view is that the stress archetype should rather be interpreted as a large reduction in the maximum number of individuals that can be supported by the environment \cite{taylor_1990}.
+The alternative view is that the stress archetype should rather be interpreted as a large reduction in the maximum number of individuals that can be supported by the environment \citep{taylor_1990}.
 For example, in the commonly cited case that the stress is induced by a scarcity of consumable resources, competition for resources would likely be intense, and the stressed population should actually be regarded as having a high population density.
@@
 
-Following Grime's original argument for a triangular scheme \cite{grime_1977}, Fig.
+Following Grime's original argument for a triangular scheme \citep{grime_1977}, Fig.
 \ref{fig:axes} represents each environmental archetype schematically as a vertex on a triangular space defined by perpendicular stress and disturbance axes.
@@
 
-How does Fig. 4 compare to Grime's C/S/R strategies? In disturbed environments, the ruderal (R) strategy is characterized by high fecundity, effective dispersal and short generation times.
+How does Fig. 4 compare to Grime's C/S/R strategies? In disturbed environments, the ruderal (R) strategy is characterized by high fecundity, wide dispersal and short generation times.
 
@@
 
-\section{Discussion}
+\section*{Discussion}
 
@@
 
-Specifically, positive correlations between measures of $r$ and $K$ are common, both between species and strains \cite{luckinbill_1979,kuno_1991,hendriks_2005,fitzsimmons_2010}, and as a result of experimental evolution \cite{luckinbill_1979,luckinbill_1978}).
+Specifically, positive correlations between measures of $r$ and $K$ are common, both between species and strains \citep{luckinbill_1979,kuno_1991,hendriks_2005,fitzsimmons_2010}, and as a result of experimental evolution \citep{luckinbill_1979,luckinbill_1978}).
 From the perspective of our model, this correlation is not at all surprising;
@@
 
-\bibliographystyle{unsrt}
+\bibliographystyle{amnatnat}
 \bibliography{reference}

### 2016-10-27 manuscript.tex
latest

@@
 \usepackage{graphicx} 
-%\usepackage{amsmath}
 %\usepackage{a4wide}
@@
 
-\section*{The Model}
+\section{Model}\label{sec:model}
 
@@
 
-\section*{Results}
+\section{Results}
 
-\subsection*{Mean Field Approximation}
+\subsection{Mean Field Approximation}
 
@@
 
-\subsection*{Invasion of rare genotypes and coexistence}\label{sec:invas}
+\subsection{Invasion of rare genotypes and coexistence}\label{sec:invas}
 
@@
 
-\subsection*{Environmental archetypes and Grime's triangle}
+\subsection{Grime's triangle}\label{sec:grime}
 
@@
 
-How does Fig. 4 compare to Grime's C/S/R strategies? In disturbed environments, the ruderal (R) strategy is characterized by high fecundity, wide dispersal and short generation times.
+How does Fig.
+4 compare to Grime's C/S/R strategies?
+In disturbed environments, the ruderal (R) strategy is characterized by high fecundity, effective dispersal and rapid development to reproductive maturity.
+We predict evolution for higher $b$ and lower $d$, but not higher $c$.
+Higher $b$ means higher fecundity, and while our model does not account for differences in dispersal effectiveness, $b$ cannot be increased at the cost of poorer dispersal since $b$ represents only those propagules which end up contesting unoccupied territories.
+To some extent, lower $d$ could mean improved individual resistance to physical destruction, but it is hard to avoid the severe forms physical destruction that characterize the disturbance archetype.
+A more effective method to reduce $d$ is to shorten the time to reproductive maturity; this shortens the iteration time in our model and thereby reduces the chance of death per iteration $d$.
 
-We predict evolution for higher $b$, but not higher $c$ or lower $d$ (constrained to be large), amounts to focusing on fecundity.
-While there is no dispersal limitation in our model, propagules must be minimally effective at dispersal since our model only includes propagules which will enter territorial contests on unoccupied sites;
-
- Shorter generation times are thought to be a mechanism for increasing the chance that the reproductive cycle completes before disturbance-induced death
-
-The evolution of the generation time itself is beyond the scope of our model, but it could
-
-\section*{Discussion}
+\section{Discussion}
 

### 2016-10-27 manuscript.tex
latest

--- 2016-10-26 manuscript.tex
@@
 Nevertheless, there are strong indications there are broader principles governing the operation of selection.
-In many groups of organisms (including corals \citep{darling_2012}, insects \citep{southwood_1977}, fishes \citep{winemiller_1992} and plants \citep{grime_1988}), species can be partitioned into a small number of distinct trait clusters corresponding to fundamentally distinct ``primary strategies'' \citep{winemiller_2015}.
+In many groups of organisms, including corals \citep{darling_2012}, insects \citep{southwood_1977}, fishes \citep{winemiller_1992}, zooplankton \citep{allan_76} and plants \citep{grime_1988}, different species can be divided into a small number of distinct trait clusters corresponding to fundamentally distinct ``primary strategies'' \citep{winemiller_2015}.
 The most famous example is Grime's plant trait classification scheme \citep{grime_1974,grime_1977,grime_1988}.
@@
 
-\section*{The Model}
+\section*{Model}\label{sec:model}
 
@@
 
-\subsection*{Environmental archetypes and Grime's triangle}
+\subsection*{Grime's triangle}\label{sec:grime}
 
@@
 The stress archetype is more ambiguous, and has been the subject of an extensive debate in the plant ecology literature (the ``Grime-Tilman'' debate \citep{aerts_1999}).
-In Grime's view, stressful environments impose such severe challenges that surviving the stressors at all is the primary challenge.
+In Grime's view, stressful environments impose such severe challenges that they are only barely survivable.
+
 In our model, this can be expressed by $b_i\ll 1$ and $b_i/d_i\approx 1$.
@@
 
-The alternative view is that the stress archetype should rather be interpreted as a large reduction in the maximum number of individuals that can be supported by the environment \citep{taylor_1990}.
-For example, in the commonly cited case that the stress is induced by a scarcity of consumable resources, competition for resources would likely be intense, and the stressed population should actually be regarded as having a high population density.
-In our model, this would imply a large reduction in $K$ (greater per-individual territorial requirement).
+The alternative view is that the stress archetype should rather be interpreted as a reduction in the maximum number of individuals that can be supported by the environment \citep{taylor_1990}; in our model, this means a greater per-individual territorial requirement represented by a lower $K$ for a given amount of space.
+The stress implies that $b$ is low
+
+For example, in the commonly cited case that the stress is induced by a scarcity of consumable resources, it is argued that competition for resources would likely be intense, and the stressed population should actually be regarded as having a high population density \citep{taylor_1990}.
+
 That is, $N$ under stress is much lower than  under ideal conditions, but it is not much lower than $K$ for the stressful environment.
@@
 \ref{fig:table}.
-Also shown is the approximate dependence of $\Delta_+ n_i$ on $b_i$ and $c_i$ for each archetype (third row), which can be used infer the expected direction of evolution for the traits $b$, $c$ and $d$ (fourth row).
+Also shown is the approximate dependence of $\Delta_+ n_i$ on $b_i$ and $c_i$ for each archetype (third row).
+These can be used infer the expected direction of evolution for the traits $b$, $c$ and $d$ (fourth row) as follows.
 
-The latter is obtained as follows.
-As noted in the previous section, if beneficial mutations can survive the low-abundance stochastic regime, their behavior is governed deterministically by Eq.
+As noted in the previous section, if beneficial mutations establish (i.e survive the low-abundance stochastic regime), they will proceed to grow deterministically according to Eq.
 \eqref{eq:master}.
-They will then proceed to grow deterministically (establishment).
-The probability of establishment increases with the mutant fitness advantage, and is therefore typically on the order of a few percent, whereas the fixation of neutral mutations is exceedingly unlikely (probability of order $1/N$).
+The probability of establishment increases with the mutant fitness advantage, and is therefore typically on the order of one percent, whereas the fixation of neutral mutations is exceedingly unlikely (probability of order $1/N$).
 Consequently, the direction of evolutionary change is determined by which trait changes confer an appreciable benefit, subject to the constraints imposed by the environment.
 
-For example, in Grime's version of the stress archetype, population density is low, so competition is not important, and so only mutants with greater $b$ or lower $d$ will have an appreciably greater $\Delta n_i$.
+For example, in Grime's version of the stress archetype, $L$ is low, so competition is not important, and only mutants with greater $b$ or lower $d$ will have an appreciably greater $\Delta n_i$.
 Mutations in $c$ are effectively neutral, and will rarely fix.
-However, by definition of the stress archetype, $b$ is constrained to be very small.
+However, by definition of the stress archetype, $b$ is constrained to be small.
 Thus, while some rare mutations may produce small improvements in $b$, it is much more likely that mutations will arise that lower $d$, making this the expected direction of evolutionary change for Grime's stress archetype.
@@
   & Ideal & Disturbance* & Stress (G) & Stress (K) \\ \hline
-  Parameter- & $d_i \ll 1$ & $d_i \approx 1$ & $b_i \ll 1$ & $b_i \ll 1$ \\
-  regime & $b_i/d_i\gg 1$ & $b_i/d_i\gg 1$ & $b_i\approx d_i$ & $b_i>  d_i$ \\
+  Parameter- & $d \ll 1$ & $d \approx 1$ & $b \ll 1$ & $b \ll 1$ \\
+  regime & $b/d\gg 1$ & $b/d\gg 1$ & $b\approx d$ & $b>d$ \\
   Density $N/K$  & High & Low & Low & High \\
@@
 
-How does Fig. 4 compare to Grime's C/S/R strategies? In disturbed environments, the ruderal (R) strategy is characterized by high fecundity, wide dispersal and short generation times.
+How does Fig.
+4 compare to Grime's C/S/R strategies?
+In disturbed environments, we predict evolution for higher $b$ and lower $d$, but not higher $c$.
+Higher $b$ means higher fecundity, but not necessarily mass propagule production: $b$ represents only those propagules which sucessfully develop into juveniles in unoccupied territories.
+This is broadly consistent with the ruderal primary strategy.
+Plant ruderals devote a large proportion of their productivity to seed production \cite{grime_1977}, whereas the analogous ``opportunistic'' strategists in fishes have large intrinsic growth rates \citep{winemiller_1992}.
+In corals, a distinguishing feature of the ruderal cluster is brood spawning (rather than broadcast spawning), which corresponds to higher parental investment and lower overall propagule production, but may be important for successful propagation in low density, disturbed environments due to Allee effects \citep{darling_2012}.
+Lower $d$ could be achieved by improved individual resistance to physical destruction, but it is hard to reduce mortality in the face of severe disturbances.
+Given this constraint, shortening the time to reproductive maturity (the iteration time in our model) is an effective way of reducing the chance of death per iteration $d$.
+A short life cycle is probably the most consistent characteristic of ruderals \cite{grime_1977,winemiller_1992,darling_2012}.
 
-We predict evolution for higher $b$, but not higher $c$ or lower $d$ (constrained to be large), amounts to focusing on fecundity.
-While there is no dispersal limitation in our model, propagules must be minimally effective at dispersal since our model only includes propagules which will enter territorial contests on unoccupied sites;
+In stressful environments, we predict evolution either for lower $d$ or higher $c$ depending on how the stress interpreted.
 
- Shorter generation times are thought to be a mechanism for increasing the chance that the reproductive cycle completes before disturbance-induced death
-
-The evolution of the generation time itself is beyond the scope of our model, but it could
+In ideal environments, we predict evolution for higher $b$ and $c$, but not lower $d$ (which is already small).
+In plants and corals, the ability to secure territories $c$ is strongly associated with rapid individual growth, and rapid individual growth is a defining trait of the competitor trait cluster \cite{grime_1977,winemiller_1992,darling_2012}.
 

### 2016-10-28 manuscript.tex
latest

@@
 
-The stress archetype is more ambiguous, and has been the subject of an extensive debate in the plant ecology literature (the ``Grime-Tilman'' debate \citep{aerts_1999}).
-In Grime's view, stressful environments impose such severe challenges that they are only barely survivable.
+The stress archetype is more ambiguous, and has been the subject of an extensive debate in the plant ecology literature (the ``Grime-Tilman'' debate; \citealt{aerts_1999}).
+Stressful environments severely restrict growth and reproduction, so that $b_i\ll 1$ \cite{grime_1974,grime_1977}.
+Mutations which appreciably improve $b_i$ will be either non-existent or extremely unlikely, so $b_i$ is constrained to remain low.
+In Grime's view, under these conditions the rate at which propagules successfully develop to adulthood cannot appreciably exceed the mortality rate.
+This implies $b_i/d_i\approx 1$ in our model, and so the propagule density $L$ is suppressed to such low levels that there are essentially no territorial contests occurring.
 
-In our model, this can be expressed by $b_i\ll 1$ and $b_i/d_i\approx 1$.
-The propagule density $L$ (as well as $N/K$) are suppressed to such low levels that there is essentially no competition between individuals.
-The severity of the stress is such that mutations which appreciably improve $b_i$ are extremely rare, so $b_i$ is constrained to remain low.
-
-The alternative view is that the stress archetype should rather be interpreted as a reduction in the maximum number of individuals that can be supported by the environment \citep{taylor_1990}; in our model, this means a greater per-individual territorial requirement represented by a lower $K$ for a given amount of space.
-The stress implies that $b$ is low
-
-For example, in the commonly cited case that the stress is induced by a scarcity of consumable resources, it is argued that competition for resources would likely be intense, and the stressed population should actually be regarded as having a high population density \citep{taylor_1990}.
-
-That is, $N$ under stress is much lower than  under ideal conditions, but it is not much lower than $K$ for the stressful environment.
-Since our model accommodates both of these alternatives, we include them both here.
+The alternative view is that stressful environments simply have a lower carrying capacity \citep{taylor_1990}; in our model, this means a greater per-individual territorial requirement represented by a lower $K$ for a given amount of space.
+In particular, it is argued that when stress is induced by a scarcity of consumable resources, competition for those resources would likely be intense.
+Thus, $b_i$ need not be particularly close to $d_i$, but contests for consumable resources at the juvenile phase would kill off most propagules before adulthood.
+In other words, the stressed population is a high density population where competition is important \citep{taylor_1990} (that is, high density relative to $K$, not relative to ideal conditions).
+We include both of these alternatives in our discussion here.
 

### 2016-10-28 manuscript.tex
latest

@@
 Thus, plant species can be classified on a triangular C/S/R ternary plot \citep{grime_1974}.
-Trait classification schemes for other organisms closely parallel Grime's scheme \citep{winemiller_2015}.
+Trait classification schemes for other organisms are broadly analogous to Grime's scheme \citep{winemiller_2015}.
 
@@
 In other words, the stressed population is a high density population where competition is important \citep{taylor_1990} (that is, high density relative to $K$, not relative to ideal conditions).
-We include both of these alternatives in our discussion here.
 
@@
   $\Delta_+ n_i\propto$ & $b_i c_i$ & $b_i$ & $b_i$ & $b_i c_i$ \\
-  Evolution for & $\uparrow b$, $ \uparrow c$ & $\uparrow b$, $\downarrow d$ & $\downarrow d$ & $\uparrow c$
+  Evolution for & $\uparrow b$, $ \uparrow c$ & $\uparrow b$, $\downarrow d$ & $\downarrow d$ & $\uparrow c, \downarrow d$
 \end{tabular}
@@
 4 compare to Grime's C/S/R strategies?
+In our comparison we will stick to fishes, corals and plants, for which three-way primary strategy schemes are well developed \citep{grime_1977,winemiller_1992,darling_2012}.
+The connection of our model to fish strategies is necessarily more tentative, given that fishes are motile and not all territorial and the starting assumption of our model is site-occupancy .
+
 In disturbed environments, we predict evolution for higher $b$ and lower $d$, but not higher $c$.
@@
 Plant ruderals devote a large proportion of their productivity to seed production \cite{grime_1977}, whereas the analogous ``opportunistic'' strategists in fishes have large intrinsic growth rates \citep{winemiller_1992}.
-In corals, a distinguishing feature of the ruderal cluster is brood spawning (rather than broadcast spawning), which corresponds to higher parental investment and lower overall propagule production, but may be important for successful propagation in low density, disturbed environments due to Allee effects \citep{darling_2012}.
+In corals, a distinguishing feature of the ruderal cluster is brood spawning (rather than broadcast spawning).
+This corresponds to higher parental investment and lower overall propagule production, but potentially also higher $b$ at low densities, since broadcast spawners are vulnerable to a powerful Allee effect at the egg fertilization stage \citep{knowlton_2001}.
 Lower $d$ could be achieved by improved individual resistance to physical destruction, but it is hard to reduce mortality in the face of severe disturbances.
 Given this constraint, shortening the time to reproductive maturity (the iteration time in our model) is an effective way of reducing the chance of death per iteration $d$.
-A short life cycle is probably the most consistent characteristic of ruderals \cite{grime_1977,winemiller_1992,darling_2012}.
+An exceptionally short life cycle is probably the most defining characteristic of ruderals \citep{grime_1977,winemiller_1992,darling_2012}.
 
-In stressful environments, we predict evolution either for lower $d$ or higher $c$ depending on how the stress interpreted.
+In stressful environments, we predict evolution for lower $d$, and also  for higher $c$ in the low-$K$ interpretation of the stress archetype.
+Lowering $d$ is obviously essential when $b\ll 1$, and stress tolerant plants and corals have long life spans, allowing for long intervals between successful recruitments (and episodic broadcast spawning in corals).
+For fishes, the ``equilibrium'' strategy is the analogue of Grime's stress tolerator.
+This strategy is associated with resource limitation, and is also characterized by long life span, as well as high parental investment in tiny broods.
+This may reflect a high-$c$ strategy in the face of intense competition for severely limited resources (the low-$K$ interpretation).
 
 In ideal environments, we predict evolution for higher $b$ and $c$, but not lower $d$ (which is already small).
-In plants and corals, the ability to secure territories $c$ is strongly associated with rapid individual growth, and rapid individual growth is a defining trait of the competitor trait cluster \cite{grime_1977,winemiller_1992,darling_2012}.
+In plants and corals, the ability to secure territories $c$ is strongly associated with rapid individual growth, and rapid individual growth is a defining feature of the competitor trait cluster \cite{grime_1977,darling_2012}.
+Evolution for higher $b$ (which is closely related to the intrinsic low-density growth rate $r=b-d$ ) is not usually thought to be the expected outcome in high-density competitive environments.
+We will discuss this further below in the context of MacArthur's $r$/$K$ scheme.
+For now we simply note that the analogue of Grime's ideal environment for fishes occurs when resources limitations are not severe and the environment varies predictably over large temporal and spatial scales \cite{winemiller_1992,winemiller_2015}.
+The associated ``periodic'' strategy is characterized by enormous brood sizes, suggesting a strategy of ensuring that more propagules actually end up contesting areas favorable for development (higher $b$).
+
+Indeed, the classic lottery model was originally applied to territorial reef fishes, and consequently the $l_i\gg 1$ assumption was
 

### 2016-10-28 manuscript.tex
latest

@@
 
-\subsection*{Grime's triangle}\label{sec:grime}
+\subsection*{Primary strategies and Grime's triangle}
 
@@
 
-In ideal environments, we predict evolution for higher $b$ and $c$, but not lower $d$ (which is already small).
-In plants and corals, the ability to secure territories $c$ is strongly associated with rapid individual growth, and rapid individual growth is a defining feature of the competitor trait cluster \cite{grime_1977,darling_2012}.
-Evolution for higher $b$ (which is closely related to the intrinsic low-density growth rate $r=b-d$ ) is not usually thought to be the expected outcome in high-density competitive environments.
-We will discuss this further below in the context of MacArthur's $r$/$K$ scheme.
-For now we simply note that the analogue of Grime's ideal environment for fishes occurs when resources limitations are not severe and the environment varies predictably over large temporal and spatial scales \cite{winemiller_1992,winemiller_2015}.
+In ideal environments, we predict evolution for higher $b$ and $c$, but not lower $d$.
+In plants and corals, the ability to secure territories $c$ is strongly associated with rapid individual growth; not surprisingly, rapid individual growth is a defining feature of the competitor trait cluster \cite{grime_1977,darling_2012}.
+The analogue of Grime's ideal environment for fishes occurs when resources limitations are not severe and the environment varies predictably over large temporal and spatial scales \cite{winemiller_1992,winemiller_2015}.
 The associated ``periodic'' strategy is characterized by enormous brood sizes, suggesting a strategy of ensuring that more propagules actually end up contesting areas favorable for development (higher $b$).
+Evolution for higher $b$ (which is closely related to the intrinsic low-density growth rate $r=b-d$) will be discussed further in the Discussion in the context of MacArthur's $r$/$K$ scheme.
+
+\section*{Discussion}
 
 Indeed, the classic lottery model was originally applied to territorial reef fishes, and consequently the $l_i\gg 1$ assumption was
-
-\section*{Discussion}
 
@@
 
-To calculate the relative fluctuations in $\sum_{j\neq i} c_j x_j$, we use a similar approximation as for $\Delta_r n_i$: $\tilde{p}$ is approximated by the ${\mathbf x}$ dispersal probabilities in a territory where at least two focal propagule is assumed to be present.
+To calculate the relative fluctuations in $\sum_{j\neq i} c_j x_j$, we use a similar approximation as for $\Delta_r n_i$: $\hat{p}$ is approximated by the ${\mathbf x}$ dispersal probabilities in a territory where at least two focal propagule is assumed to be present.
 All covariances are now zero, so that $\sigma^2(\sum c_j x_j)=\sum c_j^2 \sigma^2(x_j)$, where $\sigma^2(x_j)=l_j$ for $j\neq i$.

### 2016-10-29 manuscript.tex
latest

@@
 In ideal environments, we predict evolution for higher $b$ and $c$, but not lower $d$.
-In plants and corals, the ability to secure territories $c$ is strongly associated with rapid individual growth; not surprisingly, rapid individual growth is a defining feature of the competitor trait cluster \cite{grime_1977,darling_2012}.
-The analogue of Grime's ideal environment for fishes occurs when resources limitations are not severe and the environment varies predictably over large temporal and spatial scales \cite{winemiller_1992,winemiller_2015}.
-The associated ``periodic'' strategy is characterized by enormous brood sizes, suggesting a strategy of ensuring that more propagules actually end up contesting areas favorable for development (higher $b$).
-Evolution for higher $b$ (which is closely related to the intrinsic low-density growth rate $r=b-d$) will be discussed further in the Discussion in the context of MacArthur's $r$/$K$ scheme.
+In plants and corals, a key mechanism for winning territorial contests is rapidly outgrowing and ``shading out'' competitors; not surprisingly, rapid individual growth is a defining feature of the competitor trait cluster \citep{grime_1977,darling_2012}.
+Evolution for higher $b$ under high-density, competitive conditions may seem counter-intuitive given that $b$ is closely related to the low-density reproductive rate $r=b-d$, and the opposite outcome is expected in MacArthur's $r$/$K$ dichotomy \citep{macarthur_1962}.
+Neither particularly high nor low $b$ have been associated with the competitor strategy in plants and corals.
+However, for fishes, the analogous ``periodic'' strategy is characterized by enormous brood sizes as well as rapid development \citep{winemiller_1992,winemiller_2015}, suggesting a strategy of ensuring that many propagules actually end up contesting areas favorable for development (higher $b$).
+The evolution of $b$ in ideal environments will be discussed further in the Discussion.
 
 \section*{Discussion}
+
+The model developed here is a density-dependent generalization of the classic lottery model (Results section).
 

### 2016-10-31 manuscript.tex
latest

@@
 In plants and corals, a key mechanism for winning territorial contests is rapidly outgrowing and ``shading out'' competitors; not surprisingly, rapid individual growth is a defining feature of the competitor trait cluster \citep{grime_1977,darling_2012}.
-Evolution for higher $b$ under high-density, competitive conditions may seem counter-intuitive given that $b$ is closely related to the low-density reproductive rate $r=b-d$, and the opposite outcome is expected in MacArthur's $r$/$K$ dichotomy \citep{macarthur_1962}.
+Evolution for higher $b$ under high-density, competitive conditions may seem counter-intuitive.
 Neither particularly high nor low $b$ have been associated with the competitor strategy in plants and corals.
@@
 
+Unlike Grime's classic ternary plot for observed traits \citep{grime_1974}, our realization of Grime's triangle (Fig.
+4) refers only to the direction of trait evolution.
+Whereas a ternary plot represents anti-correlation between traits relevant for success in different environmental archetypes, our triangle reflects the underlying environmental limits of survival.
+
+We have made no appeal to, or attempted to explain, trade-offs between traits .
+
+given that $b$ is closely related to the low-density reproductive rate $r=b-d$, and the opposite outcome is expected in MacArthur's $r$/$K$ dichotomy \citep{macarthur_1962}.
+
+Apart from the prediction of evolution for higher $b$ in ideal environments,
+
 The model developed here is a density-dependent generalization of the classic lottery model (Results section).
-
-Indeed, the classic lottery model was originally applied to territorial reef fishes, and consequently the $l_i\gg 1$ assumption was
 

### 2016-10-31 manuscript.tex
latest

--- 2016-10-29 manuscript.tex
@@
 Nevertheless, there are strong indications there are broader principles governing the operation of selection.
-In many groups of organisms, including corals \citep{darling_2012}, insects \citep{southwood_1977}, fishes \citep{winemiller_1992}, zooplankton \citep{allan_76} and plants \citep{grime_1988}, different species can be divided into a small number of distinct trait clusters corresponding to fundamentally distinct ``primary strategies'' \citep{winemiller_2015}.
+In many groups of organisms, including corals \citep{darling_2012}, insects \citep{southwood_1977}, fishes \citep{winemiller_1992}, zooplankton \citep{allan_76} and plants \citep{grime_1988,westoby_1998}, different species can be divided into a small number of distinct trait clusters corresponding to fundamentally distinct ``primary strategies'' \citep{winemiller_2015}.
 The most famous example is Grime's plant trait classification scheme \citep{grime_1974,grime_1977,grime_1988}.
@@
 
-We broadly follow the spirit of MacArthur's r/K selection scheme in that our model is intended to account for fundamentally different forms of selection without getting  entangled in the intricacies of particular ecological scenarios.
+We broadly follow the spirit of MacArthur's $r$/$K$ selection scheme in that our model is intended to account for fundamentally different forms of selection without getting  entangled in the intricacies of particular ecological scenarios.
 However, rather than building directly on MacArthur's formalism and its later extensions using Lotka-Volterra equations to incorporate competition (``$\alpha$-selection'') \citep{gill_1974,case_1974,joshi_2001}, our model is devised primarily with Grime's C/S/R scheme in mind, and represents a quantitative formalization of how C/S/R manifests at the level of genotype evolution (as opposed to divergence between species).
-This choice is motivated in part by the substantial empirical support for the C/S/R scheme, and in part by the failings of the r/K low/high density dichotomy --- many growth ability traits will confer advantages at both low and high densities, in which case r- and K- selection will effectively coincide (more details in the Discussion).
+This choice is motivated in part by the substantial empirical support for the C/S/R scheme, and in part by the failings of the $r$/$K$ low/high density dichotomy --- many growth ability traits will confer advantages at both low and high densities (more details in the Discussion).
 
@@
 
-The model developed here is a density-dependent generalization of the classic lottery model (Results section).
+Unlike Grime's classic ternary plot \citep{grime_1974}, which represents anti-correlations between traits relevant for success in different environmental archetypes, our realization of Grime's triangle (Fig.
+4) refers instead to the direction of trait evolution under different regimes of stress and disturbance.
+As discussed in section ``Primary strategies and Grime's triangle'', over time our predicted trait evolution should lead to trait values consistent with Grime's scheme.
+In making these predictions, we have made no reference to any kind of trade-offs or pleitropy, even though trade-offs are often invoked to explain primary strategy schemes \citep{macarthur_1962,winemiller_1992,aerts_1999}.
+Thus, while trade-offs may amplify specialization, they are not necessary for it.
+For example, corals which rapidly out-shade neighbors have a tall, branched morphology which is vulnerable to disturbances and so, all else being equal, ideal environment $c$-strategists will suffer higher mortality from disturbances.
+Yet Fig.
+\ref{fig:axes} gives the same conclusion; mutations which change disturbance vulnerability are essentially neutral under ideal conditions, whereas higher $c$ is strongly favored.
 
-Indeed, the classic lottery model was originally applied to territorial reef fishes, and consequently the $l_i\gg 1$ assumption was
+Our prediction of evolution for higher $b$ in ideal environments is contrary to the expectations of MacArthur's $r$/$K$ dichotomy \citep{macarthur_1962} since $b$ is closely related to the maximal, low-density growth rate $b-d$, and ideal environments support high population densities which should be subject to ``$K$-selection''.
+However, in the Introduction, we noted that the $r$-$K$ dichotomy is not consistent with empirical studies showing that maximal growth rate and saturation density are positively correlated, both between species/strains \citep{luckinbill_1979,kuno_1991,hendriks_2005,fitzsimmons_2010}, and as a result of experimental evolution \citep{luckinbill_1978,luckinbill_1979}.
+From the perspective of our model, these correlations are not surprising, since the saturation density, which is determined by a balance between births and deaths, increases with $b$.
+In reality there may be trade-offs between $b$ and $c$ e.g.
+due to body size, which we have not accounted for, so that higher $b$ correlates with lower overall competitive success \cite{luckinbill_1979}.
+Our higher-$b$ predictions simply reflects the fact that, all else being equal, producing more propagules is always advantageous, regardless of population density, a fact lost in the simple logistic interpretation of the $r$/$K$ scheme.
 
-Our model differs from both Grime's C/S/R and MacArthur's $r$/$K$ schemes in the role of the propagule production rate $b$, a measure of intrinsic fecundity closely related to the growth rate at low densities $r=b-d$.
-In both of those schemes, the essential feature of life at high densities is competition.
-This is less  in the $r$/$K$ scheme, which does not explicitly  it is better to have a
-
-contributes just as much to
-
-In the Introduction, we noted that the $r$-$K$ dichotomy is not consistent with ``$K$-selection'' (i.e. selection at high densities) for growth ability traits.
-
-Specifically, positive correlations between measures of $r$ and $K$ are common, both between species and strains \citep{luckinbill_1979,kuno_1991,hendriks_2005,fitzsimmons_2010}, and as a result of experimental evolution \citep{luckinbill_1979,luckinbill_1978}).
-From the perspective of our model, this correlation is not at all surprising;
-
-r-K correlation, meaning of K selection
+The importance of $b$ at all densities is a general feature of lottery competition.
+Indeed, as can be seen from Eq.
+\eqref{eq:lottery}, in the classic lottery model $b_i$ and $c_i$ are essentially equivalent in that only the products $b_i c_i$ matter \citep{chesson_1981}.
+This is no longer the case in our density-dependent generalization of the classic lottery model, where stable co-existence is possible between $b$ and $c$ strategists.
+This may seem surprising, given that the classic lottery model was specifically developed for studying species co-existence questions, but the focus in that case was on the role of environmental fluctuations rather than stable coexistence \citep{chesson_1981}.
+It is not clear whether correctly accounting for the behavior of species at low density will significantly alter the conclusions of \cite{chesson_1981}, but, in any case, species co-existence questions are well beyond the scope of this manuscript.
+The foremost advantage of our generalized lottery model is that it can account for the growth of rare mutants, making it possible to study evolution in the context of lottery competition.
+In particular, we are primarily interested in conspecific genotypes rather than the interactions between different species, but there is nothing preventing our model from
 

### 2016-10-31 manuscript.tex
latest

@@
 Yet Fig.
-\ref{fig:axes} gives the same conclusion; mutations which change disturbance vulnerability are essentially neutral under ideal conditions, whereas higher $c$ is strongly favored.
+\ref{fig:axes} gives the same conclusion; mutations which change disturbance vulnerability are essentially neutral under ideal conditions, whereas improvements in $c$ yield an appreciable fitness benefit.
 
@@
 This may seem surprising, given that the classic lottery model was specifically developed for studying species co-existence questions, but the focus in that case was on the role of environmental fluctuations rather than stable coexistence \citep{chesson_1981}.
-It is not clear whether correctly accounting for the behavior of species at low density will significantly alter the conclusions of \cite{chesson_1981}, but, in any case, species co-existence questions are well beyond the scope of this manuscript.
-The foremost advantage of our generalized lottery model is that it can account for the growth of rare mutants, making it possible to study evolution in the context of lottery competition.
-In particular, we are primarily interested in conspecific genotypes rather than the interactions between different species, but there is nothing preventing our model from
+It is not clear whether correctly accounting for the behavior of species at low density will significantly alter the conclusions of \cite{chesson_1981}, but, in any case, species co-existence questions are beyond the scope of this manuscript.
+The point of generalizing the classic lottery model here is to account for the growth of rare mutants, making it possible to study evolution in the context of lottery competition.
+There is nothing preventing our model from also being applied to study species invasion, but our focus here is on genotypes within a population rather than the interactions between different species.
+
+Given
 

### 2016-10-31 manuscript.tex
latest

--- 2016-10-31 manuscript.tex
@@
 The key distinction is that territories with a single propagule from the focal genotype, which are critical at low densities, are handled separately.
-In place of the requirement of $l_i\gg 1$ for all $i$, our approximation only requires that there are no large discrepancies in competitive ability (discussed further below).
+In place of the requirement of $l_i\gg 1$ for all $i$, our approximation only requires that there are no large discrepancies in competitive ability (specifically, that we do not have $c_i/c_j\gg 1$ for any two genotypes; further discussion in section ``Discussion'').
 We obtain (details in Appendix B)
@@
 
-Fig.
-\ref{fig:simcomp} shows that Eq.
+Fig.~\ref{fig:simcomp} shows that Eq.
 \eqref{eq:master} (and its components) closely approximate direct simulations of random dispersal and lottery competition over a wide range of propagule densities (obtained by varying $U$).
 Two genotypes are present, one of which has a $c$-advantage and is at low frequency.
-The growth of the low-frequency genotype relies crucially on the low-density competition term $R_i c_i/\overline{c}$, and also to a lesser extent on the high density competition term $A_i c_i/\overline{c}$ if $l_1$ is large enough (Fig.
-\ref{fig:simcomp}b).
-On the other hand, $R_i c_i/\overline{c}$ is negligible for the high-frequency genotype, which depends instead on high density territorial victories (Fig.
-\ref{fig:simcomp}d).
+The growth of the low-frequency genotype relies crucially on the low-density competition term $R_i c_i/\overline{c}$, and also to a lesser extent on the high density competition term $A_i c_i/\overline{c}$ if $l_1$ is large enough (Fig.~\ref{fig:simcomp}b).
+On the other hand, $R_i c_i/\overline{c}$ is negligible for the high-frequency genotype, which depends instead on high density territorial victories (Fig.~\ref{fig:simcomp}d).
 
@@
 Intuitively, the mechanism for coexistence is that territorial contests are important in an $i$-dominated population (high $L$), ensuring that the $c$-specialist $j$ is not excluded, yet territorial contests are irrelevant in a $j$-dominated population (low $L$), ensuring that the $b$-specialist $i$ is not excluded.
-Fig.
-\ref{fig:coex} shows an example of this coexistence between $b$ and $c$ specialists.
+Fig.~\ref{fig:coex} shows an example of this coexistence between $b$ and $c$ specialists.
 
@@
 
-The mapping of environmental achetypes to our model parameters is summarized in the first two rows of Fig.
-\ref{fig:table}.
+The mapping of environmental achetypes to our model parameters is summarized in the first two rows of Fig.~\ref{fig:table}.
 Also shown is the approximate dependence of $\Delta_+ n_i$ on $b_i$ and $c_i$ for each archetype (third row).
@@
 
-Following Grime's original argument for a triangular scheme \citep{grime_1977}, Fig.
-\ref{fig:axes} represents each environmental archetype schematically as a vertex on a triangular space defined by perpendicular stress and disturbance axes.
+Following Grime's original argument for a triangular scheme \citep{grime_1977}, Fig.~\ref{fig:axes} represents each environmental archetype schematically as a vertex on a triangular space defined by perpendicular stress and disturbance axes.
 The ideal archetype lies at the origin (no stress or disturbance), while the stress and disturbance archetypes lie at the limits of survival on their respective axes.
@@
 
-How does Fig.
-4 compare to Grime's C/S/R strategies?
+How does Fig.~4 compare to Grime's C/S/R strategies?
 In our comparison we will stick to fishes, corals and plants, for which three-way primary strategy schemes are well developed \citep{grime_1977,winemiller_1992,darling_2012}.
@@
 
-Unlike Grime's classic ternary plot \citep{grime_1974}, which represents anti-correlations between traits relevant for success in different environmental archetypes, our realization of Grime's triangle (Fig.
-4) refers instead to the direction of trait evolution under different regimes of stress and disturbance.
+Unlike Grime's classic ternary plot \citep{grime_1974}, which represents anti-correlations between traits relevant for success in different environmental archetypes, our realization of Grime's triangle (Fig.~4) refers instead to the direction of trait evolution under different regimes of stress and disturbance.
 As discussed in section ``Primary strategies and Grime's triangle'', over time our predicted trait evolution should lead to trait values consistent with Grime's scheme.
@@
 Thus, while trade-offs may amplify specialization, they are not necessary for it.
-For example, corals which rapidly out-shade neighbors have a tall, branched morphology which is vulnerable to disturbances and so, all else being equal, ideal environment $c$-strategists will suffer higher mortality from disturbances.
-Yet Fig.
-\ref{fig:axes} gives the same conclusion; mutations which change disturbance vulnerability are essentially neutral under ideal conditions, whereas higher $c$ is strongly favored.
+For example, corals which rapidly out-shade neighbors have a tall, branched morphology which is vulnerable to disturbances, and so, all else being equal, ideal environment $c$-strategists will suffer higher mortality from disturbances.
+Yet Fig.~\ref{fig:axes} gives the same conclusion without invoking trade-offs; mutations which reduce disturbance vulnerability are essentially neutral under ideal conditions, leading to no improvements in mortality from disturbances, whereas $c$ will tend to increase over time.
 
@@
 From the perspective of our model, these correlations are not surprising, since the saturation density, which is determined by a balance between births and deaths, increases with $b$.
+Our higher-$b$ predictions simply reflects the fact that, all else being equal, producing more propagules is always advantageous, regardless of population density, a fact lost in the simple logistic interpretation of the $r$/$K$ scheme.
+
+Confusingly, the term ``$K$-selection'' has sometimes been used to refer generally to selection at high density; this encompasses both selection for higher saturation density --- ``$K$'' in the logistic equation --- as well as selection for competitive ability.
+The latter form of selection has been termed $\alpha$-selection after the competition coefficients in the Lotka-Volterra equation \citep{gill_1974,case_1974,joshi_2001}.
+Unlike the correlation
+
+In contrast to the problems with a strict $r$/$K$ dichotomy, there are
+
 In reality there may be trade-offs between $b$ and $c$ e.g.
 due to body size, which we have not accounted for, so that higher $b$ correlates with lower overall competitive success \cite{luckinbill_1979}.
-Our higher-$b$ predictions simply reflects the fact that, all else being equal, producing more propagules is always advantageous, regardless of population density, a fact lost in the simple logistic interpretation of the $r$/$K$ scheme.
 
-The importance of $b$ at all densities is a general feature of lottery competition.
+Actual K selection,
+
+Lotka Volterra with alpha,
+
+resource comp.
+
+significance of stage structure
+
+The importance of $b$ in securing territories is a general feature of lottery competition.
 Indeed, as can be seen from Eq.
@@
 This is no longer the case in our density-dependent generalization of the classic lottery model, where stable co-existence is possible between $b$ and $c$ strategists.
-This may seem surprising, given that the classic lottery model was specifically developed for studying species co-existence questions, but the focus in that case was on the role of environmental fluctuations rather than stable coexistence \citep{chesson_1981}.
-It is not clear whether correctly accounting for the behavior of species at low density will significantly alter the conclusions of \cite{chesson_1981}, but, in any case, species co-existence questions are well beyond the scope of this manuscript.
-The foremost advantage of our generalized lottery model is that it can account for the growth of rare mutants, making it possible to study evolution in the context of lottery competition.
-In particular, we are primarily interested in conspecific genotypes rather than the interactions between different species, but there is nothing preventing our model from
+Given that the classic lottery model was specifically developed for studying species co-existence questions, this may seem surprising, but the focus in that case was on the role of environmental fluctuations in promoting co-existence rather than stable coexistence \citep{chesson_1981}.
+It is not clear whether correctly accounting for the behavior of species with low density would significantly alter the conclusions of \cite{chesson_1981}, but, in any case, species co-existence questions are beyond the scope of this manuscript.
 
-Actual K selection
+More broadly, while our model can be applied to questions at a species level (e.g.
+ecological invasions), our focus here is on the evolution of genotype frequencies within a population.
+This is only possible because the model accounts for the growth of mutants from low densities.
+Given this focus, our assumption that there are no large $c$ discrepancies (section ``Mean field approximation'') amounts to a restriction on the amount of genetic variation in $c$ that will be sustained in the population.
+Since beneficial mutation effect sizes will typically not be much larger than a few percent, large $c$ discrepancies can only arise if the mutation rate is extremely large, and so the assumption will not be violated in most cases.
+However, more care would be needed if we were looking species interactions instead of genotype evolution.
 
-Significance of stage structure
-
-caveats: large c discrepancy
+eco-evo stuff
 
@@
 
-We separate the right hand side of Eq.
-\eqref{eq:growthsumuncoupled} into three components $\Delta_+ n_i = \Delta_u n_i+\Delta_r n_i+\Delta_a n_i$ which vary in relative magnitude depending on the propagule densities $l_i$.
+We separate the right hand side of Eq.~\eqref{eq:growthsumuncoupled} into three components $\Delta_+ n_i = \Delta_u n_i+\Delta_r n_i+\Delta_a n_i$ which vary in relative magnitude depending on the propagule densities $l_i$.
 Following the notation in the main text, the Poisson distributions for the $x_i$ (or some subset of the $x_i$) will be denoted $p$; for instance $p(x_1,\ldots,x_G)=p_1(x_1)\ldots p_G(x_G)$ and $p(x_1,\ldots,x_{i-1},x_{i+1}\ldots,x_G)=p_1(x_1)\ldots p_{i-1}(x_{i-1})p_{i+1}(x_{i+1})\ldots p_G(x_G)$.
@@
 We now show that, with respect to $\tilde{p}$, the standard deviation in $\sum_{j\neq i} c_j x_j$, $\sigma(\sum_{j\neq i} c_j x_j)$, is much smaller than its mean $\langle\sum_{j\neq i} c_j x_j\rangle_{\tilde{p}}$.
-Then $x_j$ can be replaced by its mean in the last term in Eq.
-\eqref{eq:deltr},
+Then $x_j$ can be replaced by its mean in the last term in Eq.~\eqref{eq:deltr},
 \begin{equation}
@@
 \end{equation}
-which will give us Eq. \eqref{eq:Dr}.
+which will give us Eq.~\eqref{eq:Dr}.
 
@@
 \end{equation}
-where $C=1-e^{-(L-l_i)}$ (note the difference from Eq. \eqref{eq:meanxjrare} for $\langle x_j \rangle_{\tilde{p}}$). Then, since
+where $C=1-e^{-(L-l_i)}$ (note the difference from Eq.~\eqref{eq:meanxjrare} for $\langle x_j \rangle_{\tilde{p}}$). Then, since
 \begin{equation}
@@
 
-When at least some of the nonfocal propagule densities are large $l_j\gg 1$, then the RHS of Eq.
-\eqref{eq:cvr} is $\ll 1$, as desired.
-This is also the case if none of the nonfocal genotype densities are large and the $c_j$ are all of similar magnitude (their ratios are of order one); the worst case scenario occurs when $(L-l_i)\sim O(1)$, in which case the negative covariances (Eq.
-\eqref{eq:covr}) which were neglected in the RHS of Eq.
-\eqref{eq:cvr} significantly reduce the overall variance $\sigma^2(\sum_{j\neq i} c_j x_j)$.
+When at least some of the nonfocal propagule densities are large $l_j\gg 1$, then the RHS of Eq.~\eqref{eq:cvr} is $\ll 1$, as desired.
+This is also the case if none of the nonfocal genotype densities are large and the $c_j$ are all of similar magnitude (their ratios are of order one); the worst case scenario occurs when $(L-l_i)\sim O(1)$, in which case the negative covariances (Eq.~\eqref{eq:covr}) which were neglected in the RHS of Eq.~\eqref{eq:cvr} significantly reduce the overall variance $\sigma^2(\sum_{j\neq i} c_j x_j)$.
 
@@
 Specifically, if $c_j l_j\gg c_k l_k$ ($j,k\neq i$, $j\neq k$) and $l_j\ll 1$ (i.e.
-in the presence of a rare, extremely strong competitor), then we cannot make the replacement Eq.
-\eqref{eq:meanfieldr}.
+in the presence of a rare, extremely strong competitor), then we cannot make the replacement Eq.~\eqref{eq:meanfieldr}.
 
-Substituting Eqs. \eqref{eq:meanfieldr} and \eqref{eq:meanxjrare} into Eq. \eqref{eq:deltr}, we obtain
+Substituting Eqs. \eqref{eq:meanfieldr} and \eqref{eq:meanxjrare} into Eq.~\eqref{eq:deltr}, we obtain
 \begin{equation}
@@
 \end{equation}
-where $R_i$ is defined in Eq. \eqref{eq:Dr}.
+where $R_i$ is defined in Eq.~\eqref{eq:Dr}.
 
@@
 
-The final contribution, $\Delta_a n_i$, accounts for territories where two or more focal propagules are present ($a$ stands for ``abundant"). Similarly to Eq. \eqref{eq:deltr}, we have 
+The final contribution, $\Delta_a n_i$, accounts for territories where two or more focal propagules are present ($a$ stands for ``abundant"). Similarly to Eq.~\eqref{eq:deltr}, we have 
 \begin{equation}
@@
 \end{equation}
-where $D=1-(1+l_i)e^{-l_i}$, analogous to Eq. \eqref{eq:varr}, and 
+where $D=1-(1+l_i)e^{-l_i}$, analogous to Eq.~\eqref{eq:varr}, and 
 \begin{equation}
@@
 
-Similarly to Eq.
-\eqref{eq:cvr}, the RHS of \eqref{eq:cva} will not be $\ll1$ if there is a nonfocal genotype $j$ with $l_j\ll 1$ and $c_j l_j\gg c_k l_k$ for $j,k\neq i$, $j\neq k$.
-When this is not the case, then since $l_i$ must be of order $1$ or larger for $\Delta_a n$ to make an appreciable contribution to $\Delta_+ n_i$, the RHS of Eq.
-\eqref{eq:cva} is $\ll 1$ as desired.
+Similarly to Eq.~\eqref{eq:cvr}, the RHS of \eqref{eq:cva} will not be $\ll1$ if there is a nonfocal genotype $j$ with $l_j\ll 1$ and $c_j l_j\gg c_k l_k$ for $j,k\neq i$, $j\neq k$.
+When this is not the case, then since $l_i$ must be of order $1$ or larger for $\Delta_a n$ to make an appreciable contribution to $\Delta_+ n_i$, the RHS of Eq.~\eqref{eq:cva} is $\ll 1$ as desired.
 
@@
 \end{equation}
-where $A_i$ is defined in Eq. \eqref{eq:Da}.
+where $A_i$ is defined in Eq.~\eqref{eq:Da}.
 
@@
 %
-%Eq. \eqref{eq:logistic} is often interpreted as a linear approximation of a more complicated growth equation
+%Eq.~\eqref{eq:logistic} is often interpreted as a linear approximation of a more complicated growth equation
 %\begin{equation}

### 2016-11-01 manuscript.tex
latest

@@
 
-Some abstract stuff that runs over multiple lines so that I can see the quote formatting.
+The classic lottery model is generalized to accomodate arbitrary densities.
+We use the model to predict the direction of trait evolution under different environmental conditions and thereby provide a mathematical underpinning for Grime's verbal theory of plant primary strategies.
+We evaluate the density-dependent lottery model as a general-purpose model of density-dependent selection in the spirit of MacArthur's $r$/$K$ theory.
 
@@
 
-In section
+In section ``Model'' we introduce the basic assumptions of our generalized lottery model.
+Analytical expressions for the change in genotype abundances over time are introduced in section ``Mean field approximation'', with mathematical details relegated to the Appendices.
+The following two sections discuss the behaviour of rare mutants and our treatment of Grime's triangle.
 
@@
 We assume that each individual in a population requires its own territory to survive and reproduce (a site-occupancy model).
-All territories are identical, and the total number of territories is $K$.
+All territories are identical, and the total number of territories is $T$.
 Time $t$ advances in discrete iterations, each representing the average time from birth to reproductive maturity.
-In iteration $t$, the number of reproductively mature individuals (henceforth called ``adults'') of the $i$'th genotype is $n_i(t)$, the total number of adults is $N(t)=\sum_i n_i(t)$, and the number of unoccupied territories is $U(t)=K-N(t)$.
+In iteration $t$, the number of reproductively mature individuals (henceforth called ``adults'') of the $i$'th genotype is $n_i(t)$, the total number of adults is $N(t)=\sum_i n_i(t)$, and the number of unoccupied territories is $U(t)=T-N(t)$.
 
@@
 Random dispersal implies that in the limit $K\rightarrow \infty$, with $n_i/K$ held fixed, $x_i$ is Poisson distributed with mean territorial propagule density $l_i=m_i/U$.
-Although $K$ is finite in our model, we assume that $K$ and the $n_i$ are large enough that $x_i$ is Poisson-distributed to a good approximation (details in Appendix A).
+Although $T$ is finite in our model, we assume that $K$ and the $n_i$ are large enough that $x_i$ is Poisson-distributed to a good approximation (details in Appendix A).
 This dispersal Poisson distribution is denoted $p_i(x_i)$.
-Note that the large $n_i$, large $K$ approximation places no restrictions on our densities $n_i/K$, but it does preclude consideration of demographic stochasticity when $n_i$ itself is very small (this will be discussed further in Section \ref{sec:invas}).
+Note that the large $n_i$, large $T$ approximation places no restrictions on our densities $n_i/T$, but it does preclude consideration of demographic stochasticity when $n_i$ itself is very small (this will be discussed further in Section \ref{sec:invas}).
 
@@
 The increase in $n_i$ over one iteration due to territorial acquisition, $\Delta_+ n_i$, is the sum of genotype $i$'s victories over all $U$ unoccupied territories.
-Since $p_1(x_1)\ldots p_G(x_G)$ is equal to the proportion of unoccupied territories with $x_1,\ldots,x_G$ of the respective propagules (again, we assume that $K$ is large enough that fluctuations in this proportion are negligible), this sum can be replaced by an expectation over the $p_i$.
+Since $p_1(x_1)\ldots p_G(x_G)$ is equal to the proportion of unoccupied territories with $x_1,\ldots,x_G$ of the respective propagules (again, we assume that $T$ is large enough that fluctuations in this proportion are negligible), this sum can be replaced by an expectation over the $p_i$.
 This gives
@@
 For one genotype, $L/(1-e^{-L})=b_i/d_i$, and so the propagule density is high $L\approx b_i/d_i\gg 1$.
-Moreover, since $L=b_i N/(N-K)$ by definition, population density is also high $N/K\approx 1$.
+Moreover, since $L=b_i N/(N-T)$ by definition, population density is also high $N/T\approx 1$.
 Thus, territorial contests are decisively important.
@@
 Then, the disturbance archetype is characterized by $d_i$ being close to $1$ for all genotypes (almost all adults and juveniles are killed each iteration).
-The single genotype equilibrium then gives $L\approx 2(1-d_i/[(1-d_i)b_i])$, where $b_i$ must be exceptionally large to ensure population persistence, and we have $L\ll 1$ and $N/K\ll 1$.
+The single genotype equilibrium then gives $L\approx 2(1-d_i/[(1-d_i)b_i])$, where $b_i$ must be exceptionally large to ensure population persistence, and we have $L\ll 1$ and $N/T\ll 1$.
 The terms proportional to $c_i/\overline{c}$ in Eq.
@@
 
-The alternative view is that stressful environments simply have a lower carrying capacity \citep{taylor_1990}; in our model, this means a greater per-individual territorial requirement represented by a lower $K$ for a given amount of space.
+The alternative view is that stressful environments simply have a lower carrying capacity \citep{taylor_1990}; in our model, this means a greater per-individual territorial requirement represented by a lower $T$ for a given amount of space.
 In particular, it is argued that when stress is induced by a scarcity of consumable resources, competition for those resources would likely be intense.
 Thus, $b_i$ need not be particularly close to $d_i$, but contests for consumable resources at the juvenile phase would kill off most propagules before adulthood.
-In other words, the stressed population is a high density population where competition is important \citep{taylor_1990} (that is, high density relative to $K$, not relative to ideal conditions).
+In other words, the stressed population is a high density population where competition is important \citep{taylor_1990} (that is, high density relative to $T$, not relative to ideal conditions).
 
@@
 \begin{tabular}{*{5}{c}}
-  & Ideal & Disturbance* & Stress (G) & Stress (K) \\ \hline
+  & Ideal & Disturbance* & Stress (G) & Stress (T) \\ \hline
   Parameter- & $d \ll 1$ & $d \approx 1$ & $b \ll 1$ & $b \ll 1$ \\
   regime & $b/d\gg 1$ & $b/d\gg 1$ & $b\approx d$ & $b>d$ \\
-  Density $N/K$  & High & Low & Low & High \\
+  Density $N/T$  & High & Low & Low & High \\
   $\Delta_+ n_i\propto$ & $b_i c_i$ & $b_i$ & $b_i$ & $b_i c_i$ \\
@@
 \end{tabular}
-\caption{\label{fig:table} The realization of Grime's environmental archetypes in our model, as well as the low-$K$ variant of the stress archetype.
+\caption{\label{fig:table} The realization of Grime's environmental archetypes in our model, as well as the low-$T$ variant of the stress archetype.
 Shown are the mapping to our parameters of each archetype, the approximate dependence of $\Delta_+ n_i$ on $b_i$ and $c_i$, as well as the corresponding expected evolutionary changes in $b_i$, $c_i$ and $d_i$.
@@
 Schematic representation of the triangular space bounded by the low/high extremes of stress/disturbance.
-The low-$K$ interpretation of stress is also shown.
+The low-$T$ interpretation of stress is also shown.
 The vertices of the triangles correspond to environmental archetypes.
@@
 
-In stressful environments, we predict evolution for lower $d$, and also  for higher $c$ in the low-$K$ interpretation of the stress archetype.
+In stressful environments, we predict evolution for lower $d$, and also  for higher $c$ in the low-$T$ interpretation of the stress archetype.
 Lowering $d$ is obviously essential when $b\ll 1$, and stress tolerant plants and corals have long life spans, allowing for long intervals between successful recruitments (and episodic broadcast spawning in corals).
@@
 This strategy is associated with resource limitation, and is also characterized by long life span, as well as high parental investment in tiny broods.
-This may reflect a high-$c$ strategy in the face of intense competition for severely limited resources (the low-$K$ interpretation).
+This may reflect a high-$c$ strategy in the face of intense competition for severely limited resources (the low-$T$ interpretation).
 
@@
 For example, corals which rapidly out-shade neighbors have a tall, branched morphology which is vulnerable to disturbances, and so, all else being equal, ideal environment $c$-strategists will suffer higher mortality from disturbances.
-Yet Fig.~\ref{fig:axes} gives the same conclusion without invoking trade-offs; mutations which reduce disturbance vulnerability are essentially neutral under ideal conditions, leading to no improvements in mortality from disturbances, whereas $c$ will tend to increase over time.
+Fig.~\ref{fig:axes} gives the same conclusion without invoking trade-offs; mutations which reduce disturbance vulnerability are essentially neutral under ideal conditions, leading to no improvements in mortality from disturbances, whereas $c$ will tend to increase over time.
 
-Our prediction of evolution for higher $b$ in ideal environments is contrary to the expectations of MacArthur's $r$/$K$ dichotomy \citep{macarthur_1962} since $b$ is closely related to the maximal, low-density growth rate $b-d$, and ideal environments support high population densities which should be subject to ``$K$-selection''.
-However, in the Introduction, we noted that the $r$-$K$ dichotomy is not consistent with empirical studies showing that maximal growth rate and saturation density are positively correlated, both between species/strains \citep{luckinbill_1979,kuno_1991,hendriks_2005,fitzsimmons_2010}, and as a result of experimental evolution \citep{luckinbill_1978,luckinbill_1979}.
-From the perspective of our model, these correlations are not surprising, since the saturation density, which is determined by a balance between births and deaths, increases with $b$.
-Our higher-$b$ predictions simply reflects the fact that, all else being equal, producing more propagules is always advantageous, regardless of population density, a fact lost in the simple logistic interpretation of the $r$/$K$ scheme.
+Our prediction of evolution for higher $b$ in ideal environments is counter to the expectations of MacArthur's $r$/$K$ dichotomy \citep{macarthur_1962} since $b$ is closely related to the maximal, low-density growth rate $b-d$, and ideal environments support high population densities which should be subject to ``$K$-selection''.
+However, in the Introduction, we noted that the $r$-$K$ dichotomy is not consistent with empirical studies showing that maximal growth rate and saturation density (measured by abundance) are positively correlated, both between species/strains \citep{luckinbill_1979,kuno_1991,hendriks_2005,fitzsimmons_2010}, and as a result of experimental evolution \citep{luckinbill_1978,luckinbill_1979}.
+From the perspective of our model, these correlations are not surprising since the saturation density, which is determined by a balance between births and deaths, increases with $b$.
+Our higher-$b$ prediction simply reflects the fact that, all else being equal, producing more propagules is always advantageous, regardless of population density, a fact lost in the simple logistic interpretation of the $r$/$K$ scheme.
 
 Confusingly, the term ``$K$-selection'' has sometimes been used to refer generally to selection at high density; this encompasses both selection for higher saturation density --- ``$K$'' in the logistic equation --- as well as selection for competitive ability.
-The latter form of selection has been termed $\alpha$-selection after the competition coefficients in the Lotka-Volterra equation \citep{gill_1974,case_1974,joshi_2001}.
-Unlike the correlation
+To avoid this ambiguity, the latter form of selection has been called ``$\alpha$-selection'' after the competition coefficients in the Lotka-Volterra equation \citep{gill_1974,case_1974,joshi_2001}.
+Unlike saturation density, there is support for a negative relationship between competitive success at high densities and maximal growth rate \cite{luckinbill_1979}; this could be driven by a tradeoff between individual size and reproductive rate.
+However, competitive success as measured by $\alpha$ (i.e.
+the per-capita effect of one genotype on another genotype's growth rate) is only partly determined by individual competitive ability --- in the presence of age-structure and territoriality, it also includes the ability of each genotype to produce contestants i.e.
+$b$ in our model.
+In contrast, our $c$ is strictly competitive ability only --- as such, it does not directly affect population density (section ``Model'').
 
-In contrast to the problems with a strict $r$/$K$ dichotomy, there are
-
-In reality there may be trade-offs between $b$ and $c$ e.g.
-due to body size, which we have not accounted for, so that higher $b$ correlates with lower overall competitive success \cite{luckinbill_1979}.
-
-Actual K selection,
-
-Lotka Volterra with alpha,
-
-resource comp.
-
-significance of stage structure
+$K$-selection in the sense of selection for a greater environmental carrying capacity for given birth and death rates, sometimes referred to as ``efficiency'' \citep{macarthur_1962}, would be represented in our model by smaller individual territorial requirements.
+To a first approximation, two co-occurring genotypes which differ by a small amount in their territorial requirements only should have the same fitness since the costs or benefits of a change in the amount of unocupied territory is shared equally among genotypes via the propagule density $L$.
+The situation is more complicated if those genotypes differ in multiple traits, and when the differences in territorial requirements become large enough that territorial contests can occur on different scales.
+We leave these complications for future work.
 
@@
 Since beneficial mutation effect sizes will typically not be much larger than a few percent, large $c$ discrepancies can only arise if the mutation rate is extremely large, and so the assumption will not be violated in most cases.
-However, more care would be needed if we were looking species interactions instead of genotype evolution.
+However, this restriction could become important when looking at species interactions rather than genotype evolution.
 
-eco-evo stuff
+In our view, the density dependent lottery model has two major limitations as a general-purpose model of density-dependent selection: a reliance on interference competition for durable resources (territoriality), and the restriction of competition to juveniles (lottery recruitment to adulthood).
+In some respects this is the complement of resource competition models, which restrict their attention to exploitation competition, typically without age structure \citep{tilman_1982}.
+In the particular case that resources are spatially localized (e.g.
+due to restricted movement through soils), then resource competition and territorial acquisition effectively coincide, and in principle a competitive ability $c$ should be derivable from resource competition.
+The situation is more complicated if the resources are well-mixed, since, in general, resource levels then need to be explicitly tracked.
+It is possible that explicit resource tracking is in fact not necessary when the focus is on the evolution of similar genotypes rather than the stable co-existence of widely differing species.
+We are not aware of any attempts to delineate the conditions under which explicit resource tracking is not required, let alone to connect those results to the density-dependent selection literature [need to check Tilman's book].
+
+On the other hand, the model does remarkably well in capturing apparently general features of selection under different environmental conditions using only three trait parameters: $b$, $c$ and $d$.
+The clean separation of a strictly-relative $c$ parameter is particularly useful from an evolutionary genetics perspective, essentially embedding a relative fitness trait within an absolute fitness model.
+This could have interesting applications for modeling the impacts of intra-specific competition on species extinction, without arbitrarily assigning an absolute fitness cost to that competition.
+[citations to sexual selection stuff?]
 
@@
 
-For sufficiently large $K$, holding $n_i/K$ fixed, the Poisson limit theorem implies that the binomial distributions for $x_i$ at each successive stage of this procedure are all closely approximated by a Poisson distribution with mean $l_i$, where we have used the fact that large $K$ implies large $U$ except in the biologically uninteresting case that there is vanishing population turnover $d_i \sim 1/K$.
+For sufficiently large $K$, holding $n_i/T$ fixed, the Poisson limit theorem implies that the binomial distributions for $x_i$ at each successive stage of this procedure are all closely approximated by a Poisson distribution with mean $l_i$, where we have used the fact that large $T$ implies large $U$ except in the biologically uninteresting case that there is vanishing population turnover $d_i \sim 1/T$.
 
 Under the Poisson approximation, the total number of genotype $i$ propagules $\sum x_i$ (summed over unoccupied territories) will deviate about its mean value $m_i$.
-Since the coefficient of variation of $\sum x_i$ is proportional to $1/\sqrt{m_i}$, these deviations are negligible unless $m_i$ is very small (say of order $100$ or less).
+Since the coefficient of variation of $\sum x_i$ is proportional to $1/\sqrt{m_i}$, these deviations are negligible unless $m_i$ is very small (say of order $10^2$ or less).
 

### 2016-11-04 manuscript.tex
latest

@@
 We separate the right hand side of Eq.~\eqref{eq:growthsumuncoupled} into three components $\Delta_+ n_i = \Delta_u n_i+\Delta_r n_i+\Delta_a n_i$ which vary in relative magnitude depending on the propagule densities $l_i$.
-Following the notation in the main text, the Poisson distributions for the $x_i$ (or some subset of the $x_i$) will be denoted $p$; for instance $p(x_1,\ldots,x_G)=p_1(x_1)\ldots p_G(x_G)$ and $p(x_1,\ldots,x_{i-1},x_{i+1}\ldots,x_G)=p_1(x_1)\ldots p_{i-1}(x_{i-1})p_{i+1}(x_{i+1})\ldots p_G(x_G)$.
-We use $P$ as a general shorthand for the probability of particular outcomes.
+Following the notation in the main text, the Poisson distributions for the $x_i$ (or some subset of the $x_i$) will be denoted $p$, and we use $P$ as a general shorthand for the probability of particular outcomes.
 
@@
 \begin{equation}
-\Delta_r n_i = m_i e^{-l_i}P\left\langle  \frac{c_i}{c_i +\sum_{j\neq i} c_j x_j } \right\rangle_{\tilde{p}},  \label{eq:deltr}
+\Delta_r n_i = m_i e^{-l_i}(1-e^{-(L-l_i)})\left\langle  \frac{c_i}{c_i +\sum_{j\neq i} c_j x_j } \right\rangle_{\tilde{p}},  \label{eq:deltr}
 \end{equation}
-where $\langle \rangle_{\tilde{p}}$ denotes the expectation with respect to $\tilde{p}$, and $\tilde{p}$ is the probability distribution of nonfocal propagaule abundances $x_j$ \textit{after} dispersal, in those territories where exactly one focal propagule, and at least one non-focal propagule, landed.
+where $\langle \rangle_{\tilde{p}}$ denotes the expectation with respect to $\tilde{p}$, and $\tilde{p}$ is the probability distribution of nonfocal propagule abundances $x_j$ \textit{after} dispersal, in those territories where exactly one focal propagule, and at least one non-focal propagule, landed.
 
-We now show that, with respect to $\tilde{p}$, the standard deviation in $\sum_{j\neq i} c_j x_j$, $\sigma(\sum_{j\neq i} c_j x_j)$, is much smaller than its mean $\langle\sum_{j\neq i} c_j x_j\rangle_{\tilde{p}}$.
+We now show that, with respect to $\tilde{p}$, the standard deviation $\sigma_{\tilde{p}}(\sum_{j\neq i} c_j x_j)$, is much smaller than its mean $\langle\sum_{j\neq i} c_j x_j\rangle_{\tilde{p}}$.
 Then $x_j$ can be replaced by its mean in the last term in Eq.~\eqref{eq:deltr},
 \begin{equation}
-\left\langle\frac{c_i}{c_i +\sum_{j\neq i} c_j x_j}\right\rangle_{\tilde{p}}\approx \frac{c_i}{c_i +\sum_{j\neq i} c_j \langle x_j\rangle_{\tilde{p}}},\label{eq:meanfieldr}
+\left\langle\frac{c_i}{c_i +\sum_{j\neq i} c_j x_j}\right\rangle_{\tilde{p}}\approx \frac{c_i}{c_i +\sum_{j\neq i} c_j \langle x_j\rangle_{\tilde{p}}}.\label{eq:meanfieldr}
 \end{equation}
-which will give us Eq.~\eqref{eq:Dr}.
 
 The exact expression for $\langle x_j \rangle_{\tilde{p}}$ is somewhat complicated.
-Letting $k$ denote the total number of propagules in a territory, and ${\mathbf x_i}=(x_1,\ldots,x_{i-1},x_{i+1}\ldots,x_G)$ denote the vector of non-focal abundances, $\tilde{p}$ can be written as
+Let $k$ denote the total number of propagules in a territory, and ${\mathbf x_i}=(x_1,\ldots,x_{i-1},x_{i+1}\ldots,x_G)$ denote the vector of non-focal abundances, so that $p({\mathbf x_i})=p_1(x_1)\ldots p_{i-1}(x_{i-1})p_{i+1}(x_{i+1})\ldots p_G(x_G)$.
+Then, applying Bayes' theorem, $\tilde{p}$ can be written as
 \begin{align}
 \tilde{p}({\mathbf x_i})&=p({\mathbf x_i}|k\geq 2,x_i=1),\nonumber\\
-&=\frac{P(k\geq 2|{\mathbf x_i},x_i=1) p({\mathbf x_i}|x_i=1)}{P(k\geq 2)},\nonumber\\
-&=\frac{p({\mathbf x_i}|x_i=1)}{1-(1+L)e^{-L}},\nonumber\\
+&=\frac{P({\mathbf x_i},k\geq 2|x_i=1)}{P(k\geq 2)},\nonumber\\
 &=\frac{1}{1-(1+L)e^{-L}}\sum_{k=2}^{\infty} P(k) p({\mathbf x_i}|\sum_{j\neq i} x_j=k-1),\nonumber\\
-&=\frac{1}{1-(1+L)e^{-L}}\sum_{k=2}^{\infty} \frac{P(k)\delta^{\sum_{j\neq i} x_j}_{k-1}}{P(\sum_{j\neq i} x_j=k-1)} p({\mathbf x_i}),\nonumber\\
-&=\frac{Le^{-l_i}}{1-(1+L)e^{-L}}\sum_{k=1}^{\infty} \left(\frac{L}{L-l_i}\right)^k\frac{\delta^{\sum_{j\neq i} x_j}_k}{k+1} p({\mathbf x_i}),\label{eq:raremonster}
+&=\frac{1}{1-(1+L)e^{-L}}\sum_{k=2}^{\infty} \frac{P(k)\delta^{\sum_{j\neq i} x_j}_{k-1}}{P(\sum_{j\neq i} x_j=k-1)} p({\mathbf x_i}),
+\label{eq:raremonster1}
 \end{align}
-where $\delta^{\sum_{j\neq i} x_j}_k=1$ if $\sum_{j\neq i} x_j=k$, and equals zero otherwise. Then, since 
-\begin{align}
-\sum_{\mathbf x_i}\delta^{\sum_{j\neq i} x_j}_k p({\mathbf x_i})x_j&= \frac{l_j}{L-L_j}kP(\sum_{j\neq i} x_j=k)\nonumber\\
-&=l_j P(\sum_{j\neq i} x_j=k-1),
-\end{align}
-after some algebra we obtain,
+where $P(k)=L^k e^{-L}/k!$, $P(\sum_{j\neq i} x_j=k-1)=(L-l_i)^{k-1} e^{-(L-l_i)}/(k-1)!$ and $\delta^{\sum_{j\neq i} x_j}_k=1$ if $\sum_{j\neq i} x_j=k$, and equals zero otherwise.
+This implies $P(k)/P(\sum_{j\neq i} x_j=k-1)=\frac{Le^{-l_i}}{k}\left(\frac{L}{L-l_i}\right)^{k-1}$, and so, changing the summation index to start from one, we obtain
 \begin{equation}
-\langle x_j \rangle_{\tilde{p}}=\frac{l_j}{1-(1+L)e^{-L}}\frac{L-1+e^{-L}}{L-l_i}.\label{eq:meanxjrare}
+\tilde{p}({\mathbf x_i})=\frac{Le^{-l_i}}{1-(1+L)e^{-L}}\sum_{k=1}^{\infty} \left(\frac{L}{L-l_i}\right)^k\frac{\delta^{\sum_{j\neq i} x_j}_k}{k+1} p({\mathbf x_i}).
+\label{eq:raremonster2}
 \end{equation}
 
-To calculate the relative fluctuations in $\sum_{j\neq i} c_j x_j$, we use the following approximation, which gives considerably simpler expressions for the means, variances and covariances of the $x_j$ compared with the exact expressions using $\tilde{p}$.
-Rather than evaluating the situation in each territory after dispersal as above, we let $\tilde{p}$ instead be the ${\mathbf x_i}$ dispersal probabilities in a territory where one focal propagule is assumed to be present, conditional on $\sum_{j\neq i} x_j>1$.
-This gives $\langle x_j \rangle_{\tilde{p}}=l_j/C$,
+To evaluate $\langle x_j \rangle_{\tilde{p}}=\sum_{\mathbf x_i} \tilde{p}({\mathbf x_i})x_j$, we first evaluate the sum $\sum_{\mathbf x_i}\delta^{\sum_{j\neq i} x_j}_k p({\mathbf x_i})x_j$, since the only term in Eq.
+\eqref{eq:raremonster2} which depends on ${\mathbf x_i}$ is $\delta^{\sum_{j\neq i} x_j}_k p({\mathbf x_i})$.
+By inspection, this sum is equal to the mean number of $j$ propagules in a territory which received $k$ nonfocal propagules, times the probability that a territory receives $k$ nonfocal propagules.
+Given that $k$ nonfocal propagules are present, on average genotype $j$ will make up $\frac{l_j}{L-l_j}k$ of them, and so
+\begin{align}
+\sum_{\mathbf x_i}\delta^{\sum_{j\neq i} x_j}_k p({\mathbf x_i})x_j&= \frac{l_j}{L-l_j}k \frac{(L-l_i)^{k} e^{-(L-l_i)}}{k!}\nonumber\\
+&=l_j \frac{(L-l_i)^{k-1} e^{-(L-l_i)}}{(k-1)!}.
+\end{align}
+The final step is to evaluate the sum over $k$, 
+\begin{align}
+\langle x_j \rangle_{\tilde{p}}&=\frac{l_j Le^{-l_i}}{1-(1+L)e^{-L}}\sum_{k=1}^{\infty} \left(\frac{L}{L-l_i}\right)^k\frac{(L-l_i)^{k-1} e^{-(L-l_i)}}{(k+1)(k-1)!}.\nonumber\\
+&=\frac{l_j Le^{-l_i}}{1-(1+L)e^{-L}}\frac{1}{L-l_i}\sum_{k=1}^{\infty} \frac{L^k e^{-(L-l_i)}}{(k+1)!}k.\nonumber\\
+&=\frac{l_j}{1-(1+L)e^{-L}}\frac{1}{L-l_i}\sum_{k=1}^{\infty} \frac{L^{k+1} e^{-L}}{(k+1)!}k.\nonumber\\
+&=\frac{l_j}{1-(1+L)e^{-L}}\frac{L-1+e^{-L}}{L-l_i},
+\label{eq:meanxjrare}
+\end{align}
+where the last line follows from the fact that $\sum_{k=1}^{\infty} \frac{L^{k+1} e^{-L}}{(k+1)!}k=\sum_{k=1}^{\infty} \frac{L^{k} e^{-L}}{k!}(k-1)=L-(1-e^{-L})$.
+
+For analyzing the relative fluctuations in $\sum_{j\neq i} c_j x_j$, Eq.
+\eqref{eq:meanxjrare} is unnecessarily complicated.
+We instead use the following approximation.
+Rather than evaluating the situation in each territory after dispersal as above, we replace $\tilde{p}$ by $\tilde{q}$, defined as the ${\mathbf x_i}$ dispersal probabilities, conditional on $\sum_{j\neq i} x_j>1$, in a territory where one focal propagule is assumed to already be present.
+This gives $\langle x_j \rangle_{\tilde{q}}=l_j/C$,
 \begin{equation}
-\sigma^2(x_j)=\frac{l_j^2}{C}\left(1-\frac{1}{C}\right)+\frac{l_j}{C},\label{eq:varr}
+\sigma_{\tilde{q}}^2(x_j)=\frac{l_j^2}{C}\left(1-\frac{1}{C}\right)+\frac{l_j}{C},\label{eq:varr}
 \end{equation}
@@
 \begin{equation}
-\sigma(x_j,x_k)=\frac{l_j l_k}{C}\left(1-\frac{1}{C}\right),\label{eq:covr}
+\sigma_{\tilde{q}}(x_j,x_k)=\frac{l_j l_k}{C}\left(1-\frac{1}{C}\right),\label{eq:covr}
 \end{equation}
-where $C=1-e^{-(L-l_i)}$ (note the difference from Eq.~\eqref{eq:meanxjrare} for $\langle x_j \rangle_{\tilde{p}}$). Then, since
+where $C=1-e^{-(L-l_i)}$.
+Comparing $\langle x_j \rangle_{\tilde{q}}=l_j/C$ and Eq.~\eqref{eq:meanxjrare}, the subtle difference in how we defined $\tilde{q}$ leads to considerably simpler expressions for the mean of $x_j$ (similarly for variances and covariances of the $x_j$, which we did not have the courage to calculate with $\tilde{p}$).
+To be sure, $\tilde{q}$ is incorrect: assuming that one propagule is already present biases total number of propagules by one.
+For example, if the focal genotype is rare and the propagule density is high ($l_j \approx L\gg 1$), Eq.
+\eqref{eq:meanxjrare} correctly gives $\langle x_j \rangle_{\tilde{p}}\approx L-1$, whereas $\langle x_j \rangle_{\tilde{q}}\approx L$.
+As a result, $\tilde{q}$ leads to pathological behavior for rare invaders (they have a rarity disadvantage), but its moments are quantitatively similar enough to $\tilde{p}$'s that it is sufficient for analyzing the relative fluctuations in $\sum_{j\neq i} c_j x_j$.
+
+Decomposing the variance in $\sum_{j\neq i} c_j x_j$,
 \begin{equation}
-\sigma^2(\sum_{j\neq i} c_j x_j)=\sum_{j\neq i}\left[c_j^2\sigma^2(x_j)+2\sum_{k>j}c_j c_k\sigma(x_j,x_k)\right],\label{eq:vartotr}
+\sigma_{\tilde{q}}^2(\sum_{j\neq i} c_j x_j)=\sum_{j\neq i}\left[c_j^2\sigma_{\tilde{q}}^2(x_j)+2\sum_{k>j}c_j c_k\sigma_{\tilde{q}}(x_j,x_k)\right],\label{eq:vartotr}
 \end{equation}
-and $1/C>1$, we have
+and using the fact that $1/C>1$ which implies that the first term in Eq. \eqref{eq:varr} and $\sigma_{\tilde{q}}(x_j,x_k)$ are negative, the relative fluctuations in $\sum_{j\neq i} c_j x_j$ satisfy
 \begin{equation}
@@
 When at least some of the nonfocal propagule densities are large $l_j\gg 1$, then the RHS of Eq.~\eqref{eq:cvr} is $\ll 1$, as desired.
-This is also the case if none of the nonfocal genotype densities are large and the $c_j$ are all of similar magnitude (their ratios are of order one); the worst case scenario occurs when $(L-l_i)\sim O(1)$, in which case the negative covariances (Eq.~\eqref{eq:covr}) which were neglected in the RHS of Eq.~\eqref{eq:cvr} significantly reduce the overall variance $\sigma^2(\sum_{j\neq i} c_j x_j)$.
+This is also the case if none of the nonfocal genotype densities are large and the $c_j$ are all of similar magnitude (their ratios are of order one); the worst case scenario occurs when $(L-l_i)\sim O(1)$, in which case the negative covariances (Eq.~\eqref{eq:covr}) which were neglected in the RHS of Eq.~\eqref{eq:cvr} significantly reduce the overall variance $\sigma_{\tilde{q}}^2(\sum_{j\neq i} c_j x_j)$.
 
 However, the relative fluctuations in $\sum_{j\neq i} c_j x_j$ can be large if some of the $c_j$ are much larger than the others.
-Specifically, if $c_j l_j\gg c_k l_k$ ($j,k\neq i$, $j\neq k$) and $l_j\ll 1$ (i.e.
+Specifically, if $c_j l_j\gg c_k l_k$ ($j,k\neq i$; $j\neq k$) and $l_j\ll 1$ (i.e.
 in the presence of a rare, extremely strong competitor), then we cannot make the replacement Eq.~\eqref{eq:meanfieldr}.
@@
 
-Again, we wish to show that the relative fluctuations in $\sum c_j x_j$ are much smaller than $1$ (with respect to $\hat{p}$), so that we have 
+Again, we show that the relative fluctuations in $\sum c_j x_j$ are much smaller than $1$ (with respect to $\hat{p}$), so that we have 
 \begin{equation}
@@
 \langle x_j \rangle_{\hat{p}}&=\sum_{\mathbf x} x_j p(\mathbf x|x_i\geq 2)\nonumber \\
+&=\sum_{k}P(k|x_i\geq 2)\sum_{\mathbf x} x_j p({\mathbf x}|x_i\geq 2,k)\nonumber\\
 &=\sum_{k}P(k|x_i\geq 2)\sum_{x_i} \sum_{\mathbf x_i} x_j p(\mathbf x_i|\sum_{j\neq i} x_j=k-x_i)p(x_i|x_i\geq 2,k)\nonumber\\

### 2016-11-04 manuscript.tex
latest

@@
 The exact expression for $\langle x_j \rangle_{\tilde{p}}$ is somewhat complicated.
-Let $k$ denote the total number of propagules in a territory, and ${\mathbf x_i}=(x_1,\ldots,x_{i-1},x_{i+1}\ldots,x_G)$ denote the vector of non-focal abundances, so that $p({\mathbf x_i})=p_1(x_1)\ldots p_{i-1}(x_{i-1})p_{i+1}(x_{i+1})\ldots p_G(x_G)$.
-Then, applying Bayes' theorem, $\tilde{p}$ can be written as
+Let $X$ denote the total number of propagules in a territory,  ${\mathbf x_i}=(x_1,\ldots,x_{i-1},x_{i+1}\ldots,x_G)$ denote the vector of non-focal abundances, so that $p({\mathbf x_i})=p_1(x_1)\ldots p_{i-1}(x_{i-1})p_{i+1}(x_{i+1})\ldots p_G(x_G)$, and let $X_i=X-x_i$ denote the total number of nonfocal propagules.
+Then, $\tilde{p}$ can be written as
 \begin{align}
-\tilde{p}({\mathbf x_i})&=p({\mathbf x_i}|k\geq 2,x_i=1),\nonumber\\
-&=\frac{P({\mathbf x_i},k\geq 2|x_i=1)}{P(k\geq 2)},\nonumber\\
-&=\frac{1}{1-(1+L)e^{-L}}\sum_{k=2}^{\infty} P(k) p({\mathbf x_i}|\sum_{j\neq i} x_j=k-1),\nonumber\\
-&=\frac{1}{1-(1+L)e^{-L}}\sum_{k=2}^{\infty} \frac{P(k)\delta^{\sum_{j\neq i} x_j}_{k-1}}{P(\sum_{j\neq i} x_j=k-1)} p({\mathbf x_i}),
+\tilde{p}({\mathbf x_i})&=p({\mathbf x_i}|X\geq 2,x_i=1)\nonumber\\
+&=\frac{P({\mathbf x_i},X\geq 2|x_i=1)}{P(X\geq 2)}\nonumber\\
+&=\frac{1}{1-(1+L)e^{-L}}\sum_{k=2}^{\infty} P(k) p({\mathbf x_i}|X_i=k-1),
+\end{align}
+and so
+\begin{align}
+\langle x_j \rangle_{\tilde{p}}&=\sum_{\mathbf x_i} \tilde{p}({\mathbf x_i})x_j,\nonumber\\
+&=\frac{1}{1-(1+L)e^{-L}}\sum_{k=2}^{\infty} P(k) \sum_{\mathbf x_i} p({\mathbf x_i}|X_i=k-1)x_j.
 \label{eq:raremonster1}
 \end{align}
-where $P(k)=L^k e^{-L}/k!$, $P(\sum_{j\neq i} x_j=k-1)=(L-l_i)^{k-1} e^{-(L-l_i)}/(k-1)!$ and $\delta^{\sum_{j\neq i} x_j}_k=1$ if $\sum_{j\neq i} x_j=k$, and equals zero otherwise.
-This implies $P(k)/P(\sum_{j\neq i} x_j=k-1)=\frac{Le^{-l_i}}{k}\left(\frac{L}{L-l_i}\right)^{k-1}$, and so, changing the summation index to start from one, we obtain
-\begin{equation}
-\tilde{p}({\mathbf x_i})=\frac{Le^{-l_i}}{1-(1+L)e^{-L}}\sum_{k=1}^{\infty} \left(\frac{L}{L-l_i}\right)^k\frac{\delta^{\sum_{j\neq i} x_j}_k}{k+1} p({\mathbf x_i}).
-\label{eq:raremonster2}
-\end{equation}
-
-To evaluate $\langle x_j \rangle_{\tilde{p}}=\sum_{\mathbf x_i} \tilde{p}({\mathbf x_i})x_j$, we first evaluate the sum $\sum_{\mathbf x_i}\delta^{\sum_{j\neq i} x_j}_k p({\mathbf x_i})x_j$, since the only term in Eq.
-\eqref{eq:raremonster2} which depends on ${\mathbf x_i}$ is $\delta^{\sum_{j\neq i} x_j}_k p({\mathbf x_i})$.
-By inspection, this sum is equal to the mean number of $j$ propagules in a territory which received $k$ nonfocal propagules, times the probability that a territory receives $k$ nonfocal propagules.
-Given that $k$ nonfocal propagules are present, on average genotype $j$ will make up $\frac{l_j}{L-l_j}k$ of them, and so
+The inner sum over ${\mathbf x_i}$ is the mean number of $j$ propagules that will be found in a territory which received $k-1$ nonfocal propagules, which is equal to $\frac{l_j}{L-l_j}(k-1)$. Thus, 
 \begin{align}
-\sum_{\mathbf x_i}\delta^{\sum_{j\neq i} x_j}_k p({\mathbf x_i})x_j&= \frac{l_j}{L-l_j}k \frac{(L-l_i)^{k} e^{-(L-l_i)}}{k!}\nonumber\\
-&=l_j \frac{(L-l_i)^{k-1} e^{-(L-l_i)}}{(k-1)!}.
-\end{align}
-The final step is to evaluate the sum over $k$, 
-\begin{align}
-\langle x_j \rangle_{\tilde{p}}&=\frac{l_j Le^{-l_i}}{1-(1+L)e^{-L}}\sum_{k=1}^{\infty} \left(\frac{L}{L-l_i}\right)^k\frac{(L-l_i)^{k-1} e^{-(L-l_i)}}{(k+1)(k-1)!}.\nonumber\\
-&=\frac{l_j Le^{-l_i}}{1-(1+L)e^{-L}}\frac{1}{L-l_i}\sum_{k=1}^{\infty} \frac{L^k e^{-(L-l_i)}}{(k+1)!}k.\nonumber\\
-&=\frac{l_j}{1-(1+L)e^{-L}}\frac{1}{L-l_i}\sum_{k=1}^{\infty} \frac{L^{k+1} e^{-L}}{(k+1)!}k.\nonumber\\
+\langle x_j \rangle_{\tilde{p}}&=\frac{l_j}{1-(1+L)e^{-L}}\frac{1}{L-l_i}\sum_{k=2}^{\infty} P(k) (k-1),\nonumber\\
 &=\frac{l_j}{1-(1+L)e^{-L}}\frac{L-1+e^{-L}}{L-l_i},
@@
 \end{align}
-where the last line follows from the fact that $\sum_{k=1}^{\infty} \frac{L^{k+1} e^{-L}}{(k+1)!}k=\sum_{k=1}^{\infty} \frac{L^{k} e^{-L}}{k!}(k-1)=L-(1-e^{-L})$.
+where the last line follows from the fact that $\sum_{k=2}^{\infty} P(k)(k-1)=\sum_{k=1}^{\infty} P(k)(k-1)=\sum_{k=1}^{\infty} P(k)k-\sum_{k=1}^{\infty}P(k)=L-(1-e^{-L})$.
 
 For analyzing the relative fluctuations in $\sum_{j\neq i} c_j x_j$, Eq.
-\eqref{eq:meanxjrare} is unnecessarily complicated.
-We instead use the following approximation.
-Rather than evaluating the situation in each territory after dispersal as above, we replace $\tilde{p}$ by $\tilde{q}$, defined as the ${\mathbf x_i}$ dispersal probabilities, conditional on $\sum_{j\neq i} x_j>1$, in a territory where one focal propagule is assumed to already be present.
+\eqref{eq:meanxjrare} is unnecessarily complicated, and so we instead use the following approximation.
+Rather than evaluating the situation in each territory after dispersal as above, we replace $\tilde{p}$ by $\tilde{q}$, defined as the ${\mathbf x_i}$ dispersal probabilities, conditional on $X_i>1$, in a territory where one focal propagule is assumed to already be present.
 This gives $\langle x_j \rangle_{\tilde{q}}=l_j/C$,

### 2016-11-04 manuscript.tex
latest

@@
 \begin{equation}
-A_i=\frac{\overline{c}(1-e^{-l_i})}{c_il_i\frac{1-e^{-l_i}}{1-(1+l_i)e^{-l_i}}+\sum_{j\neq i}\frac{c_jl_j}{L-l_j}\left(L\frac{1-e^{-L}}{1-(1+L)e^{-L}}-l_j\frac{1-e^{-l_j}}{1-(1+l_j)e^{-l_j}}\right)}.\label{eq:Da}
+A_i=\frac{\overline{c}(1-e^{-l_i})}{c_il_i\frac{1-e^{-l_i}}{1-(1+l_i)e^{-l_i}}+\left(L\frac{1-e^{-L}}{1-(1+L)e^{-L}}-l_i\frac{1-e^{-l_i}}{1-(1+l_i)e^{-l_i}}\right)\sum_{j\neq i}\frac{c_jl_j}{L-l_j}}.\label{eq:Da}
 \end{equation}
@@
 \end{align}
-where the last line follows from the fact that $\sum_{k=2}^{\infty} P(k)(k-1)=\sum_{k=1}^{\infty} P(k)(k-1)=\sum_{k=1}^{\infty} P(k)k-\sum_{k=1}^{\infty}P(k)=L-(1-e^{-L})$.
+where the last line follows from $\sum_{k=2}^{\infty} P(k)(k-1)=\sum_{k=1}^{\infty} P(k)(k-1)=\sum_{k=1}^{\infty} P(k)k-\sum_{k=1}^{\infty}P(k)$.
 
 For analyzing the relative fluctuations in $\sum_{j\neq i} c_j x_j$, Eq.
-\eqref{eq:meanxjrare} is unnecessarily complicated, and so we instead use the following approximation.
-Rather than evaluating the situation in each territory after dispersal as above, we replace $\tilde{p}$ by $\tilde{q}$, defined as the ${\mathbf x_i}$ dispersal probabilities, conditional on $X_i>1$, in a territory where one focal propagule is assumed to already be present.
+\eqref{eq:meanxjrare} is unnecessarily complicated.
+We instead use the following approximation.
+Rather than evaluating the situation in each territory after dispersal as above, we replace $\tilde{p}$ by $\tilde{q}$, defined as the ${\mathbf x_i}$ Poisson dispersal probabilities conditional on $X_i>1$.
 This gives $\langle x_j \rangle_{\tilde{q}}=l_j/C$,
@@
 where $C=1-e^{-(L-l_i)}$.
-Comparing $\langle x_j \rangle_{\tilde{q}}=l_j/C$ and Eq.~\eqref{eq:meanxjrare}, the subtle difference in how we defined $\tilde{q}$ leads to considerably simpler expressions for the mean of $x_j$ (similarly for variances and covariances of the $x_j$, which we did not have the courage to calculate with $\tilde{p}$).
-To be sure, $\tilde{q}$ is incorrect: assuming that one propagule is already present biases total number of propagules by one.
+Comparing $\langle x_j \rangle_{\tilde{q}}=l_j/C$ and Eq.~\eqref{eq:meanxjrare}, $\tilde{q}$ leads to considerably simpler expressions for the mean of $x_j$ (similarly for variances and covariances of the $x_j$, which we did not have the courage to calculate with $\tilde{p}$).
+However, $\tilde{q}$ does not correctly describe the situation after dispersal in territories which received one focal genotype: assuming that one propagule is already present biases total number of propagules by one.
 For example, if the focal genotype is rare and the propagule density is high ($l_j \approx L\gg 1$), Eq.
@@
 
-Again, we show that the relative fluctuations in $\sum c_j x_j$ are much smaller than $1$ (with respect to $\hat{p}$), so that we have 
+Again, we show that the relative fluctuations in $\sum c_j x_j$ are much smaller than $1$ (with respect to $\hat{p}$), so that,
 \begin{equation}
@@
 \end{equation}
-Following a similar procedure as for $\Delta_r n_i$, where the vector of propagule abundances is denoted ${\mathbf x}$, we have
+Following a similar procedure as for $\Delta_r n_i$, where the vector of propagule abundances is denoted ${\mathbf x}$, we have for nonfocal genotypes $j\neq i$
 \begin{align}
@@
 &=\sum_{k}P(k|x_i\geq 2)\sum_{\mathbf x} x_j p({\mathbf x}|x_i\geq 2,k)\nonumber\\
-&=\sum_{k}P(k|x_i\geq 2)\sum_{x_i} \sum_{\mathbf x_i} x_j p(\mathbf x_i|\sum_{j\neq i} x_j=k-x_i)p(x_i|x_i\geq 2,k)\nonumber\\
-&=\sum_{k}P(k|x_i\geq 2)\sum_{x_i} \frac{l_j(k-x_i)}{L-l_j} p(x_i|x_i\geq 2,k)\nonumber\\
-&=\frac{l_j}{L-l_j}\left( L\frac{1-e^{-L}}{1-(1+L)e^{-L}}- l_j\frac{1-e^{-l_j}}{1-(1+l_j)e^{-l_j}}\right) 
+&=\sum_{k}P(k|x_i\geq 2)\sum_{x_i} p(x_i|x_i\geq 2,k) \sum_{\mathbf x_i} x_j p(\mathbf x_i|X_i=k-x_i)\nonumber\\
+&=\sum_{k}P(k|x_i\geq 2)\sum_{x_i}p(x_i|x_i\geq 2,k) \frac{l_j(k-x_i)}{L-l_j} \nonumber\\
+&=\frac{l_j}{L-l_j}\left[\sum_{k}P(k|x_i\geq 2)k - \sum_{x_i}p(x_i|x_i\geq 2) x_i \right]\nonumber\\
+&=\frac{l_j}{L-l_j}\left( L\frac{1-e^{-L}}{1-(1+L)e^{-L}}- l_i\frac{1-e^{-l_i}}{1-(1+l_i)e^{-l_i}}\right) 
 \end{align}
-for $j\neq i$, and 
-\begin{equation}
-\langle x_i\rangle_{\hat{p}}=l_i\frac{1-e^{-l_i}}{1-(1+l_i)e^{-l_i}}.
-\end{equation}
+The mean focal genotype abundance is simpler, 
+\begin{align}
+\langle x_i \rangle_{\hat{p}}&=\sum_{\mathbf x} x_i p(\mathbf x|x_i\geq 2)\nonumber \\
+&=\sum_{x_i} x_i p(x_i|x_i\geq 2) \nonumber\\
+&=\frac{1}{1-(1+l_i)e^{-l_i}}\sum_{x_i\geq 2} p(x_i)x_i\nonumber\\
+&=l_i\frac{1-e^{-l_i}}{1-(1+l_i)e^{-l_i}}.
+\end{align}
 
-To calculate the relative fluctuations in $\sum_{j\neq i} c_j x_j$, we use a similar approximation as for $\Delta_r n_i$: $\hat{p}$ is approximated by the ${\mathbf x}$ dispersal probabilities in a territory where at least two focal propagule is assumed to be present.
+To calculate the relative fluctuations in $\sum_{j\neq i} c_j x_j$, we use a similar approximation as for $\Delta_r n_i$: $\hat{p}$ is approximated by the ${\mathbf x}$ dispersal probabilities in a territory where at least two focal propagules are assumed to be present.
 All covariances are now zero, so that $\sigma^2(\sum c_j x_j)=\sum c_j^2 \sigma^2(x_j)$, where $\sigma^2(x_j)=l_j$ for $j\neq i$.

### 2016-11-04 manuscript.tex
latest

@@
 
-We now show that, with respect to $\tilde{p}$, the standard deviation $\sigma_{\tilde{p}}(\sum_{j\neq i} c_j x_j)$, is much smaller than its mean $\langle\sum_{j\neq i} c_j x_j\rangle_{\tilde{p}}$.
+We will show that, with respect to $\tilde{p}$, the standard deviation $\sigma_{\tilde{p}}(\sum_{j\neq i} c_j x_j)$, is much smaller than its mean $\langle\sum_{j\neq i} c_j x_j\rangle_{\tilde{p}}$.
 Then $x_j$ can be replaced by its mean in the last term in Eq.~\eqref{eq:deltr},
@@
 &=\frac{P({\mathbf x_i},X\geq 2|x_i=1)}{P(X\geq 2)}\nonumber\\
-&=\frac{1}{1-(1+L)e^{-L}}\sum_{k=2}^{\infty} P(k) p({\mathbf x_i}|X_i=k-1),
+&=\frac{1}{1-(1+L)e^{-L}}\sum_{X=2}^{\infty} P(k) p({\mathbf x_i}|X_i=X-1),
 \end{align}
@@
 \langle x_j \rangle_{\tilde{p}}&=\sum_{\mathbf x_i} \tilde{p}({\mathbf x_i})x_j,\nonumber\\
-&=\frac{1}{1-(1+L)e^{-L}}\sum_{k=2}^{\infty} P(k) \sum_{\mathbf x_i} p({\mathbf x_i}|X_i=k-1)x_j.
+&=\frac{1}{1-(1+L)e^{-L}}\sum_{X=2}^{\infty} P(X) \sum_{\mathbf x_i} p({\mathbf x_i}|X_i=X-1)x_j.
 \label{eq:raremonster1}
 \end{align}
-The inner sum over ${\mathbf x_i}$ is the mean number of $j$ propagules that will be found in a territory which received $k-1$ nonfocal propagules, which is equal to $\frac{l_j}{L-l_j}(k-1)$. Thus, 
+The inner sum over ${\mathbf x_i}$ is the mean number of $j$ propagules that will be found in a territory which received $X-1$ nonfocal propagules, which is equal to $\frac{l_j}{L-l_j}(X-1)$. Thus, 
 \begin{align}
-\langle x_j \rangle_{\tilde{p}}&=\frac{l_j}{1-(1+L)e^{-L}}\frac{1}{L-l_i}\sum_{k=2}^{\infty} P(k) (k-1),\nonumber\\
+\langle x_j \rangle_{\tilde{p}}&=\frac{l_j}{1-(1+L)e^{-L}}\frac{1}{L-l_i}\sum_{k=2}^{\infty} P(X) (X-1),\nonumber\\
 &=\frac{l_j}{1-(1+L)e^{-L}}\frac{L-1+e^{-L}}{L-l_i},
@@
 \end{align}
-where the last line follows from $\sum_{k=2}^{\infty} P(k)(k-1)=\sum_{k=1}^{\infty} P(k)(k-1)=\sum_{k=1}^{\infty} P(k)k-\sum_{k=1}^{\infty}P(k)$.
+where the last line follows from $\sum_{X=2}^{\infty} P(X)(X-1)=\sum_{X=1}^{\infty} P(X)(X-1)=\sum_{X=1}^{\infty} P(X)X-\sum_{X=1}^{\infty}P(X)$.
 
@@
 We instead use the following approximation.
-Rather than evaluating the situation in each territory after dispersal as above, we replace $\tilde{p}$ by $\tilde{q}$, defined as the ${\mathbf x_i}$ Poisson dispersal probabilities conditional on $X_i>1$.
+Rather than evaluating the situation in each territory after dispersal as above, we replace $\tilde{p}$ by $\tilde{q}$, defined as the ${\mathbf x_i}$ Poisson dispersal probabilities conditional on $X_i>1$, independently of the outcome of $x_i$.
 This gives $\langle x_j \rangle_{\tilde{q}}=l_j/C$,
@@
 where $C=1-e^{-(L-l_i)}$.
-Comparing $\langle x_j \rangle_{\tilde{q}}=l_j/C$ and Eq.~\eqref{eq:meanxjrare}, $\tilde{q}$ leads to considerably simpler expressions for the mean of $x_j$ (similarly for variances and covariances of the $x_j$, which we did not have the courage to calculate with $\tilde{p}$).
-However, $\tilde{q}$ does not correctly describe the situation after dispersal in territories which received one focal genotype: assuming that one propagule is already present biases total number of propagules by one.
-For example, if the focal genotype is rare and the propagule density is high ($l_j \approx L\gg 1$), Eq.
-\eqref{eq:meanxjrare} correctly gives $\langle x_j \rangle_{\tilde{p}}\approx L-1$, whereas $\langle x_j \rangle_{\tilde{q}}\approx L$.
-As a result, $\tilde{q}$ leads to pathological behavior for rare invaders (they have a rarity disadvantage), but its moments are quantitatively similar enough to $\tilde{p}$'s that it is sufficient for analyzing the relative fluctuations in $\sum_{j\neq i} c_j x_j$.
+The distribution $\tilde{q}$ only approximates the situation after dispersal, since knowing that one focal genotype is among the propagules present restricts the possible outcomes for the $x_j$, so that the $x_j$ cannot strictly be treated as independent of $x_i$.
+This seemingly minor distinction has meaningful consequences.
+To illustrate, suppose that the focal genotype is rare and the propagule density is high ($l_j \approx L\gg 1$).
+Then Eq.
+\eqref{eq:meanxjrare} correctly predicts that there are on average $L-1$ nonfocal propagules $\langle x_j \rangle_{\tilde{p}}\approx L-1$, whereas $\tilde{q}$ predicts one extra $\langle x_j \rangle_{\tilde{q}}\approx L$.
+As a result, $\tilde{q}$ gives pathological behavior for rare invaders (they have a rarity disadvantage), but its moments are quantitatively similar enough to those of $\tilde{p}$ that it is sufficient for analyzing the relative fluctuations in $\sum_{j\neq i} c_j x_j$.
 
@@
 \end{equation}
-and using the fact that $1/C>1$ which implies that the first term in Eq. \eqref{eq:varr} and $\sigma_{\tilde{q}}(x_j,x_k)$ are negative, the relative fluctuations in $\sum_{j\neq i} c_j x_j$ satisfy
+and using the fact that $\sigma_{\tilde{q}}(x_j,x_k)$ and the first term in Eq.
+\eqref{eq:varr} are negative because $C<1$, we obtain an upper bound on the the relative fluctuations in $\sum_{j\neq i} c_j x_j$,
 \begin{equation}
@@
 
-When at least some of the nonfocal propagule densities are large $l_j\gg 1$, then the RHS of Eq.~\eqref{eq:cvr} is $\ll 1$, as desired.
-This is also the case if none of the nonfocal genotype densities are large and the $c_j$ are all of similar magnitude (their ratios are of order one); the worst case scenario occurs when $(L-l_i)\sim O(1)$, in which case the negative covariances (Eq.~\eqref{eq:covr}) which were neglected in the RHS of Eq.~\eqref{eq:cvr} significantly reduce the overall variance $\sigma_{\tilde{q}}^2(\sum_{j\neq i} c_j x_j)$.
+Then, when at least some of the nonfocal propagule densities are large $l_j\gg 1$, the RHS of Eq.~\eqref{eq:cvr} is $\ll 1$, as desired.
+This is also the case if none of the nonfocal genotype densities are large and the $c_j$ are all of similar magnitude (their ratios are of order one); the worst case scenario occurs when $L-l_i\sim O(1)$, in which case the negative covariances (Eq.~\eqref{eq:covr}) which were neglected in the RHS of Eq.~\eqref{eq:cvr} significantly reduce the overall variance $\sigma_{\tilde{q}}^2(\sum_{j\neq i} c_j x_j)$.
 
 However, the relative fluctuations in $\sum_{j\neq i} c_j x_j$ can be large if some of the $c_j$ are much larger than the others.
-Specifically, if $c_j l_j\gg c_k l_k$ ($j,k\neq i$; $j\neq k$) and $l_j\ll 1$ (i.e.
-in the presence of a rare, extremely strong competitor), then we cannot make the replacement Eq.~\eqref{eq:meanfieldr}.
+Specifically, in the presence of a rare, extremely strong competitor ($c_j l_j\gg c_{j'} l_{j'}$ for all other nonfocal genotypes $j'$, and $l_j\ll 1$), then the RHS of Eq.
+\eqref{eq:cvr} can be large and we cannot make the replacement Eq.~\eqref{eq:meanfieldr}.
 
@@
 \langle x_j \rangle_{\hat{p}}&=\sum_{\mathbf x} x_j p(\mathbf x|x_i\geq 2)\nonumber \\
-&=\sum_{k}P(k|x_i\geq 2)\sum_{\mathbf x} x_j p({\mathbf x}|x_i\geq 2,k)\nonumber\\
-&=\sum_{k}P(k|x_i\geq 2)\sum_{x_i} p(x_i|x_i\geq 2,k) \sum_{\mathbf x_i} x_j p(\mathbf x_i|X_i=k-x_i)\nonumber\\
-&=\sum_{k}P(k|x_i\geq 2)\sum_{x_i}p(x_i|x_i\geq 2,k) \frac{l_j(k-x_i)}{L-l_j} \nonumber\\
-&=\frac{l_j}{L-l_j}\left[\sum_{k}P(k|x_i\geq 2)k - \sum_{x_i}p(x_i|x_i\geq 2) x_i \right]\nonumber\\
+&=\sum_{X}P(X|x_i\geq 2)\sum_{\mathbf x} x_j p({\mathbf x}|x_i\geq 2,X)\nonumber\\
+&=\sum_{X}P(X|x_i\geq 2)\sum_{x_i} p(x_i|x_i\geq 2,X) \sum_{\mathbf x_i} x_j p(\mathbf x_i|X_i=X-x_i)\nonumber\\
+&=\sum_{X}P(X|x_i\geq 2)\sum_{x_i}p(x_i|x_i\geq 2,X) \frac{l_j(X-x_i)}{L-l_j} \nonumber\\
+&=\frac{l_j}{L-l_j}\left[\sum_{X}P(X|x_i\geq 2)X - \sum_{x_i}p(x_i|x_i\geq 2) x_i \right]\nonumber\\
 &=\frac{l_j}{L-l_j}\left( L\frac{1-e^{-L}}{1-(1+L)e^{-L}}- l_i\frac{1-e^{-l_i}}{1-(1+l_i)e^{-l_i}}\right) 
 \end{align}
-The mean focal genotype abundance is simpler, 
+The mean focal genotype abundance is, 
 \begin{align}
@@
 
-To calculate the relative fluctuations in $\sum_{j\neq i} c_j x_j$, we use a similar approximation as for $\Delta_r n_i$: $\hat{p}$ is approximated by the ${\mathbf x}$ dispersal probabilities in a territory where at least two focal propagules are assumed to be present.
-All covariances are now zero, so that $\sigma^2(\sum c_j x_j)=\sum c_j^2 \sigma^2(x_j)$, where $\sigma^2(x_j)=l_j$ for $j\neq i$.
-The expression for $\sigma^2(x_i)$ is more complicated.
-We assume $p(x_i=0)\approx 0$ without loss of generality (since otherwise $D\gg 1$ and $\Delta n_a$ is negligible).
-Then
+To calculate the relative fluctuations in $\sum_{j\neq i} c_j x_j$, we use a similar approximation as for $\Delta_r n_i$: $\hat{p}$ is approximated by $\hat{q}$, defined as the ${\mathbf x}$ dispersal probabilities in a territory conditional on $x_i>2$ (that is, treating the $x_j$ as independent of $x_i$).
+All covariances between nonfocal genotypes are now zero, so that $\sigma^2(\sum c_j x_j)=\sum c_j^2 \sigma^2(x_j)$, where $\sigma^2(x_j)=l_j$ for $j\neq i$.
+The expression for $\sigma^2(x_i)$ is more complicated, but in the relevant regime where $p(x_i=0)\approx 0$ (since otherwise $D\gg 1$ and $\Delta n_a$ is negligible), then
 \begin{equation}
-\sigma^2(x_i)=\frac{l_i^2}{D}\left(1-\frac{1}{D}\right)+\frac{l_i}{D},
+\sigma_{\hat{q}}^2(x_i)\approx\frac{l_i^2}{D}\left(1-\frac{1}{D}\right)+\frac{l_i}{D},
 \end{equation}
-where $D=1-(1+l_i)e^{-l_i}$, analogous to Eq.~\eqref{eq:varr}, and 
+where $D= 1-(1+l_i)e^{-l_i}$, analogous to Eq.~\eqref{eq:varr}, and 
 \begin{equation}
-\frac{\sigma(\sum c_j x_j)}{\langle\sum c_j x_j\rangle} \approx\frac{\left(\sum_{j\neq i} c_j^2 l_j + c_i^2 \sigma^2(x_i)\right)^{1/2}}{\sum_{j\neq i} c_j l_j + c_i l_i/D} \label{eq:cva}.
+\frac{\sigma_{\hat{q}}(\sum c_j x_j)}{\langle\sum c_j x_j\rangle} \approx\frac{\left(\sum_{j\neq i} c_j^2 l_j + c_i^2 \sigma_{\hat{q}}^2(x_i)\right)^{1/2}}{\sum_{j\neq i} c_j l_j + c_i l_i/D} \label{eq:cva}.
 \end{equation}
 
-Similarly to Eq.~\eqref{eq:cvr}, the RHS of \eqref{eq:cva} will not be $\ll1$ if there is a nonfocal genotype $j$ with $l_j\ll 1$ and $c_j l_j\gg c_k l_k$ for $j,k\neq i$, $j\neq k$.
+Similarly to Eq.~\eqref{eq:cvr}, the RHS of \eqref{eq:cva} will not be $\ll 1$ in the presence of a rare, extremely strong competitor.
 When this is not the case, then since $l_i$ must be of order $1$ or larger for $\Delta_a n$ to make an appreciable contribution to $\Delta_+ n_i$, the RHS of Eq.~\eqref{eq:cva} is $\ll 1$ as desired.

### 2016-11-07 manuscript.tex
latest

@@
 
-The second component, $\Delta_r n_i$, accounts for territories where a single focal genotype propagule is present along with at least one non-focal propagule ($r$ stands for ``rare'') i.e.
-$x_i=1$ and $\sum_{j\neq i} x_j\geq 1$.
-The number of territories where this occurs is $Up_i(1)P(\sum_{j\neq i} x_j\geq 1)=b_i n_i e^{-l_i}(1-e^{-(L-l_i)})$.
+The second component, $\Delta_r n_i$, accounts for territories where a single focal propagule is present along with at least one non-focal propagule ($r$ stands for ``rare'') i.e.
+$x_i=1$ and $X_i\geq 1$ where $X_i=\sum_{j\neq i} x_j$ is the number of nonfocal propagules.
+The number of territories where this occurs is $Up_i(1)P(X_i\geq 1)=b_i n_i e^{-l_i}(1-e^{-(L-l_i)})$.
 Thus
@@
 
-We will show that, with respect to $\tilde{p}$, the standard deviation $\sigma_{\tilde{p}}(\sum_{j\neq i} c_j x_j)$, is much smaller than its mean $\langle\sum_{j\neq i} c_j x_j\rangle_{\tilde{p}}$.
+We will show that, with respect to $\tilde{p}$, the standard deviation $\sigma_{\tilde{p}}(\sum_{j\neq i} c_j x_j)$, is much smaller than $\langle\sum_{j\neq i} c_j x_j\rangle_{\tilde{p}}$.
 Then $x_j$ can be replaced by its mean in the last term in Eq.~\eqref{eq:deltr},
@@
 
-The exact expression for $\langle x_j \rangle_{\tilde{p}}$ is somewhat complicated.
-Let $X$ denote the total number of propagules in a territory,  ${\mathbf x_i}=(x_1,\ldots,x_{i-1},x_{i+1}\ldots,x_G)$ denote the vector of non-focal abundances, so that $p({\mathbf x_i})=p_1(x_1)\ldots p_{i-1}(x_{i-1})p_{i+1}(x_{i+1})\ldots p_G(x_G)$, and let $X_i=X-x_i$ denote the total number of nonfocal propagules.
+We first calculate $\langle x_j \rangle_{\tilde{p}}$.
+Let $X=\sum_j x_j$ denote the total number of propagules in a territory and ${\mathbf x_i}=(x_1,\ldots,x_{i-1},x_{i+1}\ldots,x_G)$ denote the vector of non-focal abundances, so that $p({\mathbf x_i})=p_1(x_1)\ldots p_{i-1}(x_{i-1})p_{i+1}(x_{i+1})\ldots p_G(x_G)$.
 Then, $\tilde{p}$ can be written as
@@
 &=\frac{P({\mathbf x_i},X\geq 2|x_i=1)}{P(X\geq 2)}\nonumber\\
-&=\frac{1}{1-(1+L)e^{-L}}\sum_{X=2}^{\infty} P(k) p({\mathbf x_i}|X_i=X-1),
+&=\frac{1}{1-(1+L)e^{-L}}\sum_{X=2}^{\infty} P(X) p({\mathbf x_i}|X_i=X-1),
 \end{align}
@@
 \end{align}
-The inner sum over ${\mathbf x_i}$ is the mean number of $j$ propagules that will be found in a territory which received $X-1$ nonfocal propagules, which is equal to $\frac{l_j}{L-l_j}(X-1)$. Thus, 
+The inner sum over ${\mathbf x_i}$ is the mean number of propagules of a given nonfocal type $j$ that will be found in a territory which received $X-1$ nonfocal propagules in total, which is equal to $\frac{l_j}{L-l_i}(X-1)$.
+Thus,
 \begin{align}
@@
 Then Eq.
-\eqref{eq:meanxjrare} correctly predicts that there are on average $L-1$ nonfocal propagules $\langle x_j \rangle_{\tilde{p}}\approx L-1$, whereas $\tilde{q}$ predicts one extra $\langle x_j \rangle_{\tilde{q}}\approx L$.
+\eqref{eq:meanxjrare} correctly predicts that there are on average $L-1$ nonfocal propagules $\langle x_j \rangle_{\tilde{p}}\approx L-1$, with the focal propagule correctly excluded, whereas $\tilde{q}$ predicts one extra $\langle x_j \rangle_{\tilde{q}}\approx L$.
 As a result, $\tilde{q}$ gives pathological behavior for rare invaders (they have a rarity disadvantage), but its moments are quantitatively similar enough to those of $\tilde{p}$ that it is sufficient for analyzing the relative fluctuations in $\sum_{j\neq i} c_j x_j$.
@@
 and using the fact that $\sigma_{\tilde{q}}(x_j,x_k)$ and the first term in Eq.
-\eqref{eq:varr} are negative because $C<1$, we obtain an upper bound on the the relative fluctuations in $\sum_{j\neq i} c_j x_j$,
+\eqref{eq:varr} are negative because $C<1$, we obtain an upper bound on the relative fluctuations in $\sum_{j\neq i} c_j x_j$,
 \begin{equation}
@@
 \end{equation}
-Following a similar procedure as for $\Delta_r n_i$, where the vector of propagule abundances is denoted ${\mathbf x}$, we have for nonfocal genotypes $j\neq i$
-\begin{align}
-\langle x_j \rangle_{\hat{p}}&=\sum_{\mathbf x} x_j p(\mathbf x|x_i\geq 2)\nonumber \\
-&=\sum_{X}P(X|x_i\geq 2)\sum_{\mathbf x} x_j p({\mathbf x}|x_i\geq 2,X)\nonumber\\
-&=\sum_{X}P(X|x_i\geq 2)\sum_{x_i} p(x_i|x_i\geq 2,X) \sum_{\mathbf x_i} x_j p(\mathbf x_i|X_i=X-x_i)\nonumber\\
-&=\sum_{X}P(X|x_i\geq 2)\sum_{x_i}p(x_i|x_i\geq 2,X) \frac{l_j(X-x_i)}{L-l_j} \nonumber\\
-&=\frac{l_j}{L-l_j}\left[\sum_{X}P(X|x_i\geq 2)X - \sum_{x_i}p(x_i|x_i\geq 2) x_i \right]\nonumber\\
-&=\frac{l_j}{L-l_j}\left( L\frac{1-e^{-L}}{1-(1+L)e^{-L}}- l_i\frac{1-e^{-l_i}}{1-(1+l_i)e^{-l_i}}\right) 
-\end{align}
-The mean focal genotype abundance is, 
+Following a similar procedure as for $\Delta_r n_i$, where the vector of propagule abundances is denoted ${\mathbf x}$, the mean focal genotype abundance is, 
 \begin{align}
@@
 &=l_i\frac{1-e^{-l_i}}{1-(1+l_i)e^{-l_i}}.
+\end{align}
+For nonfocal genotypes $j\neq i$, we have
+\begin{align}
+\langle x_j \rangle_{\hat{p}}&=\sum_{\mathbf x} x_j p(\mathbf x|x_i\geq 2)\nonumber \\
+&=\sum_{X}P(X|x_i\geq 2)\sum_{\mathbf x} x_j p({\mathbf x}|x_i\geq 2,X)\nonumber\\
+&=\sum_{X}P(X|x_i\geq 2)\sum_{x_i} p(x_i|x_i\geq 2,X) \sum_{\mathbf x_i} x_j p(\mathbf x_i|X_i=X-x_i)\nonumber\\
+&=\sum_{X}P(X|x_i\geq 2)\sum_{x_i}p(x_i|x_i\geq 2,X) \frac{l_j(X-x_i)}{L-l_i} \nonumber\\
+&=\frac{l_j}{L-l_i}\left[\sum_{X}P(X|x_i\geq 2)X - \sum_{x_i}p(x_i|x_i\geq 2) x_i \right]\nonumber\\
+&=\frac{l_j}{L-l_i}\left( L\frac{1-e^{-L}}{1-(1+L)e^{-L}}- l_i\frac{1-e^{-l_i}}{1-(1+l_i)e^{-l_i}}\right). 
 \end{align}

### 2016-11-08 manuscript.tex
latest

@@
 
-\noindent{}$\ast$ Corresponding author; e-mail: amnat@uchicago.edu.
+\noindent{}$\ast$ Corresponding author; e-mail: jbertram@email.arizona.edu.
 
@@
 \begin{align}
-\langle x_j \rangle_{\tilde{p}}&=\sum_{\mathbf x_i} \tilde{p}({\mathbf x_i})x_j,\nonumber\\
+\langle x_j \rangle_{\tilde{p}}&=\sum_{\mathbf x_i} \tilde{p}({\mathbf x_i})x_j\nonumber\\
 &=\frac{1}{1-(1+L)e^{-L}}\sum_{X=2}^{\infty} P(X) \sum_{\mathbf x_i} p({\mathbf x_i}|X_i=X-1)x_j.
@@
 \begin{align}
-\langle x_j \rangle_{\tilde{p}}&=\frac{l_j}{1-(1+L)e^{-L}}\frac{1}{L-l_i}\sum_{k=2}^{\infty} P(X) (X-1),\nonumber\\
+\langle x_j \rangle_{\tilde{p}}&=\frac{l_j}{1-(1+L)e^{-L}}\frac{1}{L-l_i}\sum_{k=2}^{\infty} P(X) (X-1)\nonumber\\
 &=\frac{l_j}{1-(1+L)e^{-L}}\frac{L-1+e^{-L}}{L-l_i},
@@
 We instead use the following approximation.
-Rather than evaluating the situation in each territory after dispersal as above, we replace $\tilde{p}$ by $\tilde{q}$, defined as the ${\mathbf x_i}$ Poisson dispersal probabilities conditional on $X_i>1$, independently of the outcome of $x_i$.
-This gives $\langle x_j \rangle_{\tilde{q}}=l_j/C$,
-\begin{equation}
-\sigma_{\tilde{q}}^2(x_j)=\frac{l_j^2}{C}\left(1-\frac{1}{C}\right)+\frac{l_j}{C},\label{eq:varr}
-\end{equation}
+Rather than evaluating the situation in each territory after dispersal as above, we replace $\tilde{p}$ by $\tilde{q}$, defined as the ${\mathbf x_i}$ Poisson dispersal probabilities conditional on $X_i\geq1$, independently of the outcome of $x_i$.
+This gives $\langle x_j \rangle_{\tilde{q}}=\langle x_j \rangle_p/C=l_j/C$,
+\begin{align}
+\sigma_{\tilde{q}}^2(x_j)&=\langle x_j^2 \rangle_{\tilde{q}}-\langle x_j \rangle_{\tilde{q}}^2\nonumber\\
+&=\frac{1}{C}\langle x_j^2 \rangle_p-\frac{l_j^2}{C^2}\nonumber \\
+&=\frac{1}{C}(l_j^2 + l_j)-\frac{l_j^2}{C^2}\nonumber \\
+&=\frac{l_j^2}{C}\left(1-\frac{1}{C}\right)+\frac{l_j}{C},\label{eq:varr}
+\end{align}
 and 
-\begin{equation}
-\sigma_{\tilde{q}}(x_j,x_k)=\frac{l_j l_k}{C}\left(1-\frac{1}{C}\right),\label{eq:covr}
-\end{equation}
-where $C=1-e^{-(L-l_i)}$.
+\begin{align}
+\sigma_{\tilde{q}}(x_j,x_k)&=\langle x_j x_k \rangle_{\tilde{q}}-\langle x_j \rangle_{\tilde{q}}\langle x_k \rangle_{\tilde{q}}\nonumber\\
+&=\frac{1}{C}\langle x_j x_k \rangle_p-\frac{l_jl_k}{C^2}\nonumber\\
+&=\frac{l_j l_k}{C}\left(1-\frac{1}{C}\right),\label{eq:covr}
+\end{align}
+where $C=1-e^{-(L-l_i)}$ and $j\neq k$.
 The distribution $\tilde{q}$ only approximates the situation after dispersal, since knowing that one focal genotype is among the propagules present restricts the possible outcomes for the $x_j$, so that the $x_j$ cannot strictly be treated as independent of $x_i$.

### 2016-11-08 manuscript.tex
latest

@@
 276]{ewens_2012}.
-However, selection can have very different effects when operating on different types of traits, and evolutionary changes in one population can lead to complicated ecological responses.
+
+However, a genotype's selective advantage
+
+ can have very different effects when operating on different types of traits, and evolutionary changes in population can lead to complicated ecological responses.
 
@@
 
-Far less work has been done to model fitness in more general terms than particular traits or ecological scenarios, while still capturing key distinctions between different forms of selection.
+Far less work has been done to generalize beyond particular traits or ecological scenarios to models of fitness that still capture key distinctions between different forms of selection.
 Perhaps this is not surprising given that fitness is such a complex quantity, dependent on all of a phenotype's functional traits \citep{violle_2007} as well as its biotic and abiotic environment.
@@
 276]{ewens_2012}.
-Even MacArthur's famous r/K selection scheme is now almost exclusively known as a framework for understanding life-history traits, and judged on its failure in that role \citep{pianka_1970,stearns_1977,boyce_1984,reznick_2002}.
-In spite of the r/K scheme's original purpose as an extension of the existing population-genetic treatment of selection to account for population density \citep{macarthur_1962}, comparatively few attempts have been made to develop it further as a mathematical analysis of the major different forms of selection.
+For example, MacArthur's famous r/K selection scheme is now almost exclusively known as a framework for understanding life-history traits, and judged on its failure in that role \citep{pianka_1970,stearns_1977,boyce_1984,reznick_2002}.
+The r/K scheme's original purpose was as an extension of the existing population-genetic treatment of selection to account for population density \citep{macarthur_1962}, but few attempts have been made to develop it further as a mathematical analysis of the major different forms of selection.
 
@@
 However, while trait classification schemes are firmly grounded in trait data, they are verbal and descriptive rather than mathematical, a recognized hinderance to their broader applicability (e.g.
-\citep{tilman_2007}).
+\citealt{tilman_2007}).
 
@@
 We broadly follow the spirit of MacArthur's $r$/$K$ selection scheme in that our model is intended to account for fundamentally different forms of selection without getting  entangled in the intricacies of particular ecological scenarios.
-However, rather than building directly on MacArthur's formalism and its later extensions using Lotka-Volterra equations to incorporate competition (``$\alpha$-selection'') \citep{gill_1974,case_1974,joshi_2001}, our model is devised primarily with Grime's C/S/R scheme in mind, and represents a quantitative formalization of how C/S/R manifests at the level of genotype evolution (as opposed to divergence between species).
+However, rather than building directly on MacArthur's formalism and its later extensions using Lotka-Volterra equations to incorporate competition (``$\alpha$-selection'') \citep{gill_1974,case_1974,joshi_2001}, our model is devised primarily with Grime's C/S/R scheme in mind, and represents a quantitative formalization of how C/S/R manifests at the level of within-population genotype evolution (as opposed to phenotypic divergence between species).
 This choice is motivated in part by the substantial empirical support for the C/S/R scheme, and in part by the failings of the $r$/$K$ low/high density dichotomy --- many growth ability traits will confer advantages at both low and high densities (more details in the Discussion).
 
-In section ``Model'' we introduce the basic assumptions of our generalized lottery model.
+In the section ``Model'' we introduce the basic assumptions of our generalized lottery model.
 Analytical expressions for the change in genotype abundances over time are introduced in section ``Mean field approximation'', with mathematical details relegated to the Appendices.
@@
 We assume adults cannot be ousted from occupied territories, so only propagules landing on occupied territories are included in $m_i$.
-More generally, $m_i$ does not include propagules which never even begin the development cycle.
+More generally, $m_i$ only includes propagules which actually end up contesting unnocupied territories.
 For simplicity, we assume $m_i=b_i n_i$, where $b_i$ is a constant, genotype-specific birth rate.
@@
 The number of individuals of the $i$'th genotype landing in any particular territory is denoted $x_i$.
-Random dispersal implies that in the limit $K\rightarrow \infty$, with $n_i/K$ held fixed, $x_i$ is Poisson distributed with mean territorial propagule density $l_i=m_i/U$.
-Although $T$ is finite in our model, we assume that $K$ and the $n_i$ are large enough that $x_i$ is Poisson-distributed to a good approximation (details in Appendix A).
+Random dispersal implies that in the limit $T\rightarrow \infty$, with $n_i/T$ held fixed, $x_i$ is Poisson distributed with mean territorial propagule density $l_i=m_i/U$.
+Although $T$ is finite in our model, we assume that $T$ and the $n_i$ are large enough that $x_i$ is Poisson-distributed to a good approximation (details in Appendix A).
 This dispersal Poisson distribution is denoted $p_i(x_i)$.
-Note that the large $n_i$, large $T$ approximation places no restrictions on our densities $n_i/T$, but it does preclude consideration of demographic stochasticity when $n_i$ itself is very small (this will be discussed further in Section \ref{sec:invas}).
+Note that the large $n_i$, large $T$ approximation places no restrictions on our densities $n_i/T$, but it does preclude consideration of demographic stochasticity when $n_i$ itself is very small (this will be discussed further in Section ``Invasion of rare genotypes and coexistence'').
 
@@
 The increase in $n_i$ over one iteration due to territorial acquisition, $\Delta_+ n_i$, is the sum of genotype $i$'s victories over all $U$ unoccupied territories.
-Since $p_1(x_1)\ldots p_G(x_G)$ is equal to the proportion of unoccupied territories with $x_1,\ldots,x_G$ of the respective propagules (again, we assume that $T$ is large enough that fluctuations in this proportion are negligible), this sum can be replaced by an expectation over the $p_i$.
+Since $p_1(x_1)\ldots p_G(x_G)$ is equal to the proportion of unoccupied territories with $x_1,\ldots,x_G$ of the respective propagules (where $G$ is the number of genotypes present; and again, we assume that $T$ is large enough that fluctuations in this proportion are negligible), this sum can be replaced by an expectation over the $p_i$.
 This gives
@@
 In addition to propagule birth and competition, occupied territories become unoccupied due to mortality.
-For the vast majority of this manuscript we assume that mortality only occurs in adults, and at a constant, genotype-specific per-capita rate $d_i$, so that the overall change in genotype abundances is
+For the majority of this manuscript we assume that mortality only occurs in adults, and at a constant, genotype-specific per-capita rate $d_i$, so that the overall change in genotype abundances is
 \begin{equation}
@@
 \end{equation}
-We will introduce a different mortality model when we consider the effects of disturbances (Section \ref{sec:grime}), which will also affect competing juveniles.
+We will introduce a different mortality model when we consider the effects of disturbances (Section ``Primary strategies and Grime's triangle''), which will also affect competing juveniles.
 
@@
 \end{equation}
-which is independent of $c_i$ (here $L=\sum_j l_j$ is the overall propagule density).
+which is independent of $c_i$ (here $L=\sum_j l_j$ is the overalFaltinel propagule density).
 
@@
 \eqref{eq:lottery} in that we replace the $x_i$ with appropriate mean values.
-The key distinction is that territories with a single propagule from the focal genotype, which are critical at low densities, are handled separately.
+The key distinction is that territories with a single propagule from the focal genotype, whose behavior is critical at low densities, are handled separately.
 In place of the requirement of $l_i\gg 1$ for all $i$, our approximation only requires that there are no large discrepancies in competitive ability (specifically, that we do not have $c_i/c_j\gg 1$ for any two genotypes; further discussion in section ``Discussion'').
@@
 \begin{equation}
-A_i=\frac{\overline{c}(1-e^{-l_i})}{c_il_i\frac{1-e^{-l_i}}{1-(1+l_i)e^{-l_i}}+\left(L\frac{1-e^{-L}}{1-(1+L)e^{-L}}-l_i\frac{1-e^{-l_i}}{1-(1+l_i)e^{-l_i}}\right)\sum_{j\neq i}\frac{c_jl_j}{L-l_j}}.\label{eq:Da}
+A_i=\frac{\overline{c}(1-e^{-l_i})}{\frac{1-e^{-l_i}}{1-(1+l_i)e^{-l_i}}c_il_i+\frac{1}{L-l_i}\left(L\frac{1-e^{-L}}{1-(1+L)e^{-L}}-l_i\frac{1-e^{-l_i}}{1-(1+l_i)e^{-l_i}}\right)\sum_{j\neq i}c_jl_j}.\label{eq:Da}
 \end{equation}
@@
 The first, $e^{-L}$, accounts for propagules which land alone on unoccupied territories; these territories are won without contest.
-The second term, $R_i c_i/\overline{c}$ represents competitive victories when the $i$ genotype is a rare invader in a high density population: from Eq.
+The second, $R_i c_i/\overline{c}$ represents competitive victories when the $i$ genotype is a rare invader in a high density population: from Eq.
 \eqref{eq:Dr}, $R_i\rightarrow 0$ when the $i$ genotype is abundant ($l_i\gg 1$), or other genotypes are collectively rare ($L-l_i\ll 1$).
@@
 Two genotypes are present.
-(a) and (b) show low-frequency genotype with $c$-advantage ($m_1/M=0.1$, $c_1=1.5$), (c) and (d) show the high-frequency predominant genotype ($m_2/M=0.9$, $c_2=1$).}
+(a) and (b) show the low-frequency genotype with $c$-advantage ($m_1/M=0.1$, $c_1=1.5$), (c) and (d) show the high-frequency predominant genotype ($m_2/M=0.9$, $c_2=1$).
+Simulation points are almost invisible in (c) and (d) due to near exact agreement with Eq.
+\eqref{eq:master}.}
 \end{figure}
@@
 
-To determine how $b$, $c$ and $d$ will evolve in a population where those traits are being modified by mutations, we need to know whether mutant lineages will grow (or decline) starting from low densities.
+In our model (section ``Model''), each genotype is defined by three traits: $b$, $c$ and $d$.
+To determine how these will evolve in a population where they are being modified by mutations, we need to know whether mutant lineages will grow (or decline) starting from low densities.
 In this section we discuss the behavior of rare genotypes predicted by Eq.
@@
 \begin{equation}
-b_i\left(e^{-L}+A_i\right)-d_i=0.\label{eq:equil}
+b_i\left(e^{-L}+A_i\right)-d_i=0,\label{eq:equil}
 \end{equation}
+where $A_i=(1-(1+L)e^{-L})/L$.
 Now suppose that a new genotype $j$, which is initially rare, appears in the population.
-Then $A_j\ll 1$, $l_j\ll L$ and $\overline{c}\approx c_i$, and so, from Eq.
+Then $A_j\ll R_j$, $l_j\approx 0$ and $\overline{c}\approx c_i$, and so, from Eq.
 \eqref{eq:master}, $n_j$ will increase if
 \begin{equation}
-b_j \left(e^{-L}+R_j\frac{c_j}{c_i}\right)-d_j>0.\label{eq:invad}
+b_j \left(e^{-L}+R_j\frac{c_j}{c_i}\right)-d_j>0,\label{eq:invad}
 \end{equation}
+where $R_j\approx (1-e^{-L})/\left(\frac{c_j}{c_i}+\frac{L-1-e^{-L}}{1-(1+L)e^{-L}}\right)$.
 
 Combining Eqs.
-\eqref{eq:equil} and \eqref{eq:invad}, it is easily verified that if $j$ is superior in one trait, but otherwise identical to $i$, it will invade.
-Moreover, $j$ will eventually exclude $i$, since it is strictly superior.
+\eqref{eq:equil} and \eqref{eq:invad}, we see that $j$ will invade if it is superior in any one of the three traits, but is otherwise identical to $i$.
+If the new genotype has the same competitive ability $c_j=c_i$, then $R_j\approx A_i$ and Eqs.
+\eqref{eq:equil} and \eqref{eq:invad} imply that invasion occurs when $b_jd_i-b_id_j>0$, and in particular when $b_j>b_i$ with $d_j=d_i$, or when $d_j<d_i$ with $b_j=b_i$.
+In the case that the new genotype has a different competitive ability, Eqs.
+\eqref{eq:equil} and \eqref{eq:invad} imply that invasion occurs when $R_j c_j/c_i > A_i$; it is not hard to verify that this occurs if and only if $c_j>c_i$ using the simplified expressions for $A_i$ and $R_j$ given after Eqs.
+\eqref{eq:equil} and \eqref{eq:invad} respectively.
+Moreover, if $j$ invades in any of these cases, it will eventually exclude $i$, since it is strictly superior.
+
 However, stable coexistence is possible between genotypes that are superior in different traits.
@@
 \end{equation}
-Thus, coexistence occurs if $c_j$ is large enough.
+Thus, coexistence occurs if $c_j/c_i$ is large enough.
 Intuitively, the mechanism for coexistence is that territorial contests are important in an $i$-dominated population (high $L$), ensuring that the $c$-specialist $j$ is not excluded, yet territorial contests are irrelevant in a $j$-dominated population (low $L$), ensuring that the $b$-specialist $i$ is not excluded.
@@
 
-If the rare genotype $j$ arises due to mutation, then it's initial low-density behavior is more complicated than the above invasion analysis suggests.
+If the rare genotype $j$ arises due to mutation, then its initial low-density behavior is more complicated than the above invasion analysis suggests.
 The mutant lineage starts with one individual $n_j=1$, and remains at low abundance for many generations after its initial appearance.
-During this period, the mutant abundance $n_j$ will behave stochastically, and the deterministic equations \eqref{eq:growthsumuncoupled} and \eqref{eq:master} do not apply (Section \ref{sec:model}).
+During this period, the mutant abundance $n_j$ will behave stochastically, and the deterministic equations \eqref{eq:growthsumuncoupled} and \eqref{eq:master} do not apply (section ``Model'').
 However, if $n_j$ becomes large enough, its behavior will become effectively deterministic, and governed by Eq.
 \eqref{eq:master}.
-For mutants with fitness greater than the population mean fitness, this process is known as ``establishment'', and occurs when $n_j$ is of order $1/s$, where $s$ is the mutant's fitness advantage relative to the mean \citep{desai_2007}.
+For mutants with fitness greater than the population mean fitness, this process is known as ``establishment'', and occurs when $n_j$ is of order $1/s$ \citep{desai_2007}, where the selection coefficient $s$ is the mutant's fitness advantage (i.e.
+the mutant's fitness minus the population mean fitness).
 Here we do not consider the initial stochastic behavior of novel mutants, and have restricted our attention to the earliest deterministic behavior of rare genotypes.
@@
 
-We now discuss which changes in the traits $b, c$ and $d$ will be most favored under different environmental conditions.
-Of particular interest are Grime's ``disturbance'', ``stress'' and ``ideal'' environmental archetypes.
+We now discuss which changes in the traits $b, c$ and $d$ will be particularly favored under different environmental conditions.
+Of specific interest are Grime's ``disturbance'', ``stress'' and ``ideal'' environmental archetypes.
 To proceed, we need to map these verbal archetypes to quantitative parameter regimes in our model.
@@
 For one genotype, $L/(1-e^{-L})=b_i/d_i$, and so the propagule density is high $L\approx b_i/d_i\gg 1$.
-Moreover, since $L=b_i N/(N-T)$ by definition, population density is also high $N/T\approx 1$.
-Thus, territorial contests are decisively important.
+Moreover, since $L=b_i\frac{N}{N-T}=b_i\frac{1}{1-T/N}$ by definition, population density is also high $N/T\approx 1$.
+Thus, almost every unoccupied territory will be heavily contested.
 
@@
 Then, the disturbance archetype is characterized by $d_i$ being close to $1$ for all genotypes (almost all adults and juveniles are killed each iteration).
-The single genotype equilibrium then gives $L\approx 2(1-d_i/[(1-d_i)b_i])$, where $b_i$ must be exceptionally large to ensure population persistence, and we have $L\ll 1$ and $N/T\ll 1$.
+From Eq.
+\eqref{eq:deltN}, the single genotype equilibrium is given by $L/(1-e^{-L})=d_i/[(1-d_i)b_i]$, and since $L\ll 1$ and $N/T\ll 1$ due to high mortality, we have $L\approx 2(1-d_i/[(1-d_i)b_i])$.
+Clearly $b_i$ must be exceptionally large to ensure population persistence.
 The terms proportional to $c_i/\overline{c}$ in Eq.
@@
 The probability of establishment increases with the mutant fitness advantage, and is therefore typically on the order of one percent, whereas the fixation of neutral mutations is exceedingly unlikely (probability of order $1/N$).
-Consequently, the direction of evolutionary change is determined by which trait changes confer an appreciable benefit, subject to the constraints imposed by the environment.
+Consequently, the direction of evolutionary change is determined by which trait changes are both available, and confer an appreciable benefit, where availability is subject to constraints imposed by the environment.
 
@@
 
-How does Fig.~4 compare to Grime's C/S/R strategies?
+How does Fig.~4 compare to empirical analyses of Grime's C/S/R strategies?
 In our comparison we will stick to fishes, corals and plants, for which three-way primary strategy schemes are well developed \citep{grime_1977,winemiller_1992,darling_2012}.
-The connection of our model to fish strategies is necessarily more tentative, given that fishes are motile and not all territorial and the starting assumption of our model is site-occupancy .
+The connection of our model to fish strategies is necessarily more tentative, given that fishes are motile and not all territorial, and the starting assumption of our model is site-occupancy.
 
@@
 
-Unlike Grime's classic ternary plot \citep{grime_1974}, which represents anti-correlations between traits relevant for success in different environmental archetypes, our realization of Grime's triangle (Fig.~4) refers instead to the direction of trait evolution under different regimes of stress and disturbance.
+Unlike Grime's classic ternary plot \citep{grime_1974}, which represents anti-correlations between traits relevant for success in different environmental archetypes, our realization of Grime's triangle (Fig.~4) refers instead to the direction of adaptive trait evolution under different regimes of stress and disturbance.
 As discussed in section ``Primary strategies and Grime's triangle'', over time our predicted trait evolution should lead to trait values consistent with Grime's scheme.
@@
 Thus, while trade-offs may amplify specialization, they are not necessary for it.
-For example, corals which rapidly out-shade neighbors have a tall, branched morphology which is vulnerable to disturbances, and so, all else being equal, ideal environment $c$-strategists will suffer higher mortality from disturbances.
+As an example of a trade-off, corals which rapidly out-shade neighbors have a tall, branched morphology which is vulnerable to disturbances, and so, all else being equal, ideal environment $c$-strategists will suffer higher mortality from disturbances.
 Fig.~\ref{fig:axes} gives the same conclusion without invoking trade-offs; mutations which reduce disturbance vulnerability are essentially neutral under ideal conditions, leading to no improvements in mortality from disturbances, whereas $c$ will tend to increase over time.
 
-Our prediction of evolution for higher $b$ in ideal environments is counter to the expectations of MacArthur's $r$/$K$ dichotomy \citep{macarthur_1962} since $b$ is closely related to the maximal, low-density growth rate $b-d$, and ideal environments support high population densities which should be subject to ``$K$-selection''.
+Our prediction of evolution for higher $b$ in ideal environments is counter to the expectations of MacArthur's $r$/$K$ dichotomy \citep{macarthur_1962} since $b$ is closely related to the maximal, low-density growth rate $r=b-d$, and ideal environments support high population densities which should be subject to ``$K$-selection''.
 However, in the Introduction, we noted that the $r$-$K$ dichotomy is not consistent with empirical studies showing that maximal growth rate and saturation density (measured by abundance) are positively correlated, both between species/strains \citep{luckinbill_1979,kuno_1991,hendriks_2005,fitzsimmons_2010}, and as a result of experimental evolution \citep{luckinbill_1978,luckinbill_1979}.
@@
 To avoid this ambiguity, the latter form of selection has been called ``$\alpha$-selection'' after the competition coefficients in the Lotka-Volterra equation \citep{gill_1974,case_1974,joshi_2001}.
-Unlike saturation density, there is support for a negative relationship between competitive success at high densities and maximal growth rate \cite{luckinbill_1979}; this could be driven by a tradeoff between individual size and reproductive rate.
+Unlike saturation density, there is support for a negative relationship between competitive success at high densities and maximal growth rate \citep{luckinbill_1979}; this could be driven by a tradeoff between individual size and reproductive rate.
 However, competitive success as measured by $\alpha$ (i.e.
@@
 $b$ in our model.
-In contrast, our $c$ is strictly competitive ability only --- as such, it does not directly affect population density (section ``Model'').
+In contrast, our $c$ is strictly competitive ability only --- as such, changes in $c$ do not directly affect population density (section ``Model''), nor the ability to produce contestants.
 
@@
 \eqref{eq:lottery}, in the classic lottery model $b_i$ and $c_i$ are essentially equivalent in that only the products $b_i c_i$ matter \citep{chesson_1981}.
-This is no longer the case in our density-dependent generalization of the classic lottery model, where stable co-existence is possible between $b$ and $c$ strategists.
-Given that the classic lottery model was specifically developed for studying species co-existence questions, this may seem surprising, but the focus in that case was on the role of environmental fluctuations in promoting co-existence rather than stable coexistence \citep{chesson_1981}.
+This is no longer the case in our density- and frequency-dependent generalization of the classic lottery model, where stable co-existence is possible between $b$ and $c$ strategists.
+Given that the classic lottery model was specifically developed for studying species co-existence questions, this may seem surprising, but the focus in that case was on the role of environmental fluctuations in promoting co-existence rather than coexistence in a single, stable environment \citep{chesson_1981}.
 It is not clear whether correctly accounting for the behavior of species with low density would significantly alter the conclusions of \cite{chesson_1981}, but, in any case, species co-existence questions are beyond the scope of this manuscript.
 
-More broadly, while our model can be applied to questions at a species level (e.g.
+Rather, while our model can be applied to questions at an inter-species level (e.g.
 ecological invasions), our focus here is on the evolution of genotype frequencies within a population.
-This is only possible because the model accounts for the growth of mutants from low densities.
+Our ability to describe evolutionary processes is only possible because the model accounts for the growth of mutants from low densities.
 Given this focus, our assumption that there are no large $c$ discrepancies (section ``Mean field approximation'') amounts to a restriction on the amount of genetic variation in $c$ that will be sustained in the population.
@@
 
-In our view, the density dependent lottery model has two major limitations as a general-purpose model of density-dependent selection: a reliance on interference competition for durable resources (territoriality), and the restriction of competition to juveniles (lottery recruitment to adulthood).
+In our view, our density-dependent lottery model has two major limitations as a general-purpose model of density-dependent selection: a reliance on interference competition for durable resources (territoriality), and the restriction of competition to juveniles (lottery recruitment to adulthood).
 In some respects this is the complement of resource competition models, which restrict their attention to exploitation competition, typically without age structure \citep{tilman_1982}.

### 2016-11-10 manuscript.tex
latest

@@
 due to trampling, herbivory, pathogens, extreme weather or fire).
-The extremes of these two factors define three primary strategies denoted by C/S/R respectively: competitors ``C'' excel in low stress, low disturbance environments; stress tolerators ``S'' excel in high stress, low disturbance environments; and ruderals ``R''  excel in low stress, high disturbance environments.
+The extremes of these two factors define three primary strategies denoted by C/S/R respectively (Fig.
+\ref{fig:grimeschematic}): competitors ``C'' excel in low stress, low disturbance environments; stress tolerators ``S'' excel in high stress, low disturbance environments; and ruderals ``R''  excel in low stress, high disturbance environments.
 Survival is not possible in high-stress, high-disturbance environments.
@@
 Trait classification schemes for other organisms are broadly analogous to Grime's scheme \citep{winemiller_2015}.
+
+\begin{figure}
+\centering
+\includegraphics[scale=1]{grimeschematic.pdf}
+\caption{\label{fig:grimeschematic} Schematic of Grime's triangle.
+The two axes show increasing levels of environmental stress and disturbance, respectively.
+Survival is not possible if the combination of stress and disturbance is too large (dashed line).
+This creates a triangle, each corner of which corresponds to a ``primary strategy''.}
+\end{figure}
 
@@
 We broadly follow the spirit of MacArthur's $r$/$K$ selection scheme in that our model is intended to account for fundamentally different forms of selection without getting  entangled in the intricacies of particular ecological scenarios.
-However, rather than building directly on MacArthur's formalism and its later extensions using Lotka-Volterra equations to incorporate competition (``$\alpha$-selection'') \citep{gill_1974,case_1974,joshi_2001}, our model is devised primarily with Grime's C/S/R scheme in mind, and represents a quantitative formalization of how C/S/R manifests at the level of within-population genotype evolution (as opposed to phenotypic divergence between species).
-This choice is motivated in part by the substantial empirical support for the C/S/R scheme, and in part by the failings of the $r$/$K$ low/high density dichotomy --- many growth ability traits will confer advantages at both low and high densities (more details in the Discussion).
+However, rather than building directly on MacArthur's formalism and its later extensions using Lotka-Volterra equations to incorporate competition (``$\alpha$-selection'') \citep{gill_1974,case_1974,joshi_2001}, our model is devised more with Grime's C/S/R scheme in mind, and represents a quantitative formalization of how C/S/R manifests at the level of within-population genotype evolution (as opposed to phenotypic divergence between species).
+This choice is motivated in part by the substantial empirical support for C/S/R-like schemes, and in part by the failings of the $r$/$K$ low/high density dichotomy --- many growth ability traits will confer advantages at both low and high densities (more details in the Discussion).
 
-In the section ``Model'' we introduce the basic assumptions of our generalized lottery model.
+As we will see, a generalized version of the classic lottery model of \cite{chesson_1981} is well suited for this purpose.
+The lottery model gets its name from the way in which it represents competition: the growth of each type is determined by the number of contestants that it produces relative to the other types in the population (akin to a lottery),  weighted by a coefficient for each type representing its competitive ability.
+This is much simpler than having coefficients for the pairwise effects of types on each other (e.g.
+the $\alpha$ coefficients in the generalized Lotka-Volterra equations), or than modeling resource consumption explicitly \citep{tilman_1982}.
+The classic lottery model  breaks down at low larval densities (section ``Mean field approximation''), but this was not important for its original application to reef fishes, where a huge number of larvae from each type compete to secure territories each generation \citep{chesson_1981}.
+Here we extend the classic lottery model to correctly account for low density behavior.
+
+In the section ``Model'', we introduce the basic assumptions of our generalized lottery model.
 Analytical expressions for the change in genotype abundances over time are introduced in section ``Mean field approximation'', with mathematical details relegated to the Appendices.
-The following two sections discuss the behaviour of rare mutants and our treatment of Grime's triangle.
+The following two sections discuss the behavior of rare mutants and our treatment of Grime's triangle.
 
@@
 In iteration $t$, the number of reproductively mature individuals (henceforth called ``adults'') of the $i$'th genotype is $n_i(t)$, the total number of adults is $N(t)=\sum_i n_i(t)$, and the number of unoccupied territories is $U(t)=T-N(t)$.
+
+\begin{figure}
+\centering
+\includegraphics[scale=0.8]{lottery.pdf}
+\caption{\label{fig:lottery} Each iteration of our lottery model has three main elements.
+Each generation, propagules are produced by adults which are dispersed at random over the unoccupied territories.
+Lottery competition then occurs in each unoccupied territory: each genotype has a probability proportional to $b_i n_i c_i$ of securing the territory.
+Then death occurs.
+In Eq.
+\eqref{eq:delttot} and most of the paper, only adults can die (red crosses), but we will also consider the case where juveniles die (blue cross; section ``Primary strategies and Grime's triangle'').}
+\end{figure}
 
@@
 We will introduce a different mortality model when we consider the effects of disturbances (Section ``Primary strategies and Grime's triangle''), which will also affect competing juveniles.
+However, in the absence of severe disturbances, restricting mortality to adults (setting aside the deaths implicit in territorial contest) is a reasonable approximation  \label{eq:delttot}
 
@@
   & Ideal & Disturbance* & Stress (G) & Stress (T) \\ \hline
-  Parameter- & $d \ll 1$ & $d \approx 1$ & $b \ll 1$ & $b \ll 1$ \\
-  regime & $b/d\gg 1$ & $b/d\gg 1$ & $b\approx d$ & $b>d$ \\
+  Constraints & $d \ll 1$ & $d \approx 1$ & $b \ll 1$ & $b \ll 1$ \\
+  Other parameters & $b\gg d$ & $b\gg d$ & $b\approx d$ & $b>d$ \\
   Density $N/T$  & High & Low & Low & High \\

### 2016-11-11 manuscript.tex
latest

@@
 
-The classic lottery model is generalized to accomodate arbitrary densities.
-We use the model to predict the direction of trait evolution under different environmental conditions and thereby provide a mathematical underpinning for Grime's verbal theory of plant primary strategies.
-We evaluate the density-dependent lottery model as a general-purpose model of density-dependent selection in the spirit of MacArthur's $r$/$K$ theory.
+Fitness is typically represented in heavily simplified terms in evolutionary genetics, often using constant selection coefficients.
+This excludes fundamental ecological factors such as dynamic population size or density-dependence from our most genetically-realistic treatments of evolution, a problem that inspired MacArthur's influential but problematic $r$/$K$ theory.
+Following in the spirit of $r$/$K$-selection as a general-purpose theory of density-dependent selection, but grounding ourselves empirically in ``primary strategy'' trait classification schemes like Grime's triangle, we develop a new model of density-dependent selection which revolves around territorial contests.
+To do so, we generalize the classic lottery model of territorial acquisition, which has primarily been used for studying species co-existence questions, to accommodate arbitrary densities.
+We use this density-dependent lottery model to predict the direction of trait evolution under different environmental conditions and thereby provide a mathematical underpinning for Grime's verbal scheme.
+We revisit previous concepts of density-dependent selection, including $r$ and $K$ selection, and argue that our model distinguishes between different aspects of fitness in a more natural and intuitive manner.
 
@@
 276]{ewens_2012}.
-
-However, a genotype's selective advantage
-
- can have very different effects when operating on different types of traits, and evolutionary changes in population can lead to complicated ecological responses.
+However, the resulting picture of evolution does not include even basic elements of the ecological underpinnings of selection, including dynamic population size and density-dependence.
 
@@
 The classic lottery model  breaks down at low larval densities (section ``Mean field approximation''), but this was not important for its original application to reef fishes, where a huge number of larvae from each type compete to secure territories each generation \citep{chesson_1981}.
-Here we extend the classic lottery model to correctly account for low density behavior.
+Here we analytically extend the classic lottery model to correctly account for low density behavior.
 
@@
 Lottery competition then occurs in each unoccupied territory: each genotype has a probability proportional to $b_i n_i c_i$ of securing the territory.
-Then death occurs.
+Then occupied territories are freed up by adult mortality.
 In Eq.
@@
 In addition to propagule birth and competition, occupied territories become unoccupied due to mortality.
-For the majority of this manuscript we assume that mortality only occurs in adults, and at a constant, genotype-specific per-capita rate $d_i$, so that the overall change in genotype abundances is
+For the majority of this manuscript we assume that mortality only occurs in adults (setting aside the deaths implicit in territorial contest), and at a constant, genotype-specific per-capita rate $d_i$, so that the overall change in genotype abundances is
 \begin{equation}
@@
 \end{equation}
-We will introduce a different mortality model when we consider the effects of disturbances (Section ``Primary strategies and Grime's triangle''), which will also affect competing juveniles.
-However, in the absence of severe disturbances, restricting mortality to adults (setting aside the deaths implicit in territorial contest) is a reasonable approximation  \label{eq:delttot}
+This is reasonable approximation in the absence of disturbances; when we come to consider the effects of disturbances (Section ``Primary strategies and Grime's triangle''), we will incorporate disturbance-induced mortality in competing juveniles (Fig.~\ref{fig:lottery}).
 
@@
 Neither particularly high nor low $b$ have been associated with the competitor strategy in plants and corals.
-However, for fishes, the analogous ``periodic'' strategy is characterized by enormous brood sizes as well as rapid development \citep{winemiller_1992,winemiller_2015}, suggesting a strategy of ensuring that many propagules actually end up contesting areas favorable for development (higher $b$).
+However, for fishes, the analogous ``periodic'' strategy is characterized by enormous spawn sizes as well as rapid development \citep{winemiller_1992,winemiller_2015}, suggesting a strategy of ensuring that many propagules actually end up contesting areas favorable for development (higher $b$).
 The evolution of $b$ in ideal environments will be discussed further in the Discussion.

### 2016-11-12 manuscript.tex
latest

@@
 \usepackage[sc]{mathpazo}
-\usepackage{amsmath}
+\usepackage{amsmath,pifont}
 \usepackage{fullpage}
@@
 \usepackage{graphicx} 
-%\usepackage{a4wide}
-%\usepackage{multicol,caption}
-%\setlength{\textwidth}{16cm}
-%\setlength{\textheight}{22cm}
+\usepackage{tabularx,setspace}
 
@@
 \centering
-\begin{tabular}{*{5}{c}}
+\begin{tabular}{l*{4}{c}}
   & Ideal & Disturbance* & Stress (G) & Stress (T) \\ \hline
@@
 [citations to sexual selection stuff?]
+
+\begin{figure}
+\centering
+\begin{tabularx}{\linewidth}{lXXXXX}
+& Emeddable in flexible genetic scheme? & Ecologically meaningful? & Empirically-grounded trait scheme? & Generality beyond specific scenarios? & Formal model? \\ \hline
+Density-dependent lottery & \ding{51} & \ding{51} & \ding{51} & \ding{51} & \ding{51} \\  
+  MacArthur's $r$/$K$ + $\alpha$ & \ding{51} & \ding{51} & \ding{55}* & \ding{51} & \ding{51} \\
+  Grime's C/S/R & NA & \ding{51} & \ding{51} & \ding{51} & \ding{55} \\
+  Traditional pop. gen. & \ding{51} & \ding{55} & \ding{55} & \ding{51} & \ding{51} \\
+  Eco-evo:  \\
+  Adaptive dynamics & \ding{55} & \ding{51} & \ding{51} & \ding{55}** & \ding{51} \\
+  Brute force simulation & \ding{51} & \ding{51} & NA & \ding{55} & \ding{51}
+\end{tabularx}
+\caption{\label{fig:table2} Comparison of our density-dependent lottery model with related models/schemes in ecology and evolutionary biology.}
+\end{figure}
 

### 2016-11-13 manuscript.tex
latest

@@
 
-On the other hand, the model does remarkably well in capturing apparently general features of selection under different environmental conditions using only three trait parameters: $b$, $c$ and $d$.
+On the other hand, our model does remarkably well in capturing apparently general features of selection under different environmental conditions using only three trait parameters: $b$, $c$ and $d$.
 The clean separation of a strictly-relative $c$ parameter is particularly useful from an evolutionary genetics perspective, essentially embedding a relative fitness trait within an absolute fitness model.
@@
 
+In the broader context of eco-evolutionary synthesis [?
+Cite], we can imagine a spectrum of approaches ranging from population genetics, which is entirely evolutionary (no ecology) and has a strong emphasis on genetics, to pure ecological theory with no evolution.
+
 \begin{figure}
@@
 \begin{tabularx}{\linewidth}{lXXXXX}
-& Emeddable in flexible genetic scheme? & Ecologically meaningful? & Empirically-grounded trait scheme? & Generality beyond specific scenarios? & Formal model? \\ \hline
+&Formal model?  & Ecologically meaningful? & Empirically-grounded trait scheme? & Generality beyond specific scenarios? & Genetically flexible? \\ \hline
 Density-dependent lottery & \ding{51} & \ding{51} & \ding{51} & \ding{51} & \ding{51} \\  
   MacArthur's $r$/$K$ + $\alpha$ & \ding{51} & \ding{51} & \ding{55}* & \ding{51} & \ding{51} \\
-  Grime's C/S/R & NA & \ding{51} & \ding{51} & \ding{51} & \ding{55} \\
+  Grime's C/S/R & \ding{55} & \ding{51} & \ding{51} & \ding{51} & NA \\
   Traditional pop. gen. & \ding{51} & \ding{55} & \ding{55} & \ding{51} & \ding{51} \\
   Eco-evo:  \\
-  Adaptive dynamics & \ding{55} & \ding{51} & \ding{51} & \ding{55}** & \ding{51} \\
-  Brute force simulation & \ding{51} & \ding{51} & NA & \ding{55} & \ding{51}
+  \quad Adaptive dynamics & \ding{51} & \ding{51} & \ding{51} & \ding{55}** & \ding{55} \\
+  \quad Brute force simulation & \ding{51} & \ding{51} & NA & \ding{55} & \ding{51}
 \end{tabularx}
-\caption{\label{fig:table2} Comparison of our density-dependent lottery model with related models/schemes in ecology and evolutionary biology.}
+\caption{\label{fig:table2} Comparison of our density-dependent lottery model with related models and schemes in ecology and evolutionary biology.
+*MacArthur's $r$- and $K$-, as well as $\alpha$ selection, were all derived theoretically.
+Applications to traits come later, and with mixed success (see ``Introduction'' and ``Discussion'').
+**As a modeling framework, adaptive dynamics is not scenario-specific, and could be applied with any fitness model including our density-dependent lottery, but in practice most of the adaptive dynamics literature focuses on specific eco-evolutionary outcomes such as evolutionary ``branching'' or evolutionary suicide.}
 \end{figure}

### 2016-11-13 manuscript.tex
latest

@@
 
-In our view, our density-dependent lottery model has two major limitations as a general-purpose model of density-dependent selection: a reliance on interference competition for durable resources (territoriality), and the restriction of competition to juveniles (lottery recruitment to adulthood).
+In our view, our model has two major limitations as a general-purpose model of density-dependent selection: a reliance on interference competition for durable resources (territoriality), and the restriction of competition to juveniles (lottery recruitment to adulthood).
 In some respects this is the complement of resource competition models, which restrict their attention to exploitation competition, typically without age structure \citep{tilman_1982}.
@@
 The clean separation of a strictly-relative $c$ parameter is particularly useful from an evolutionary genetics perspective, essentially embedding a relative fitness trait within an absolute fitness model.
-This could have interesting applications for modeling the impacts of intra-specific competition on species extinction, without arbitrarily assigning an absolute fitness cost to that competition.
-[citations to sexual selection stuff?]
+This could have interesting applications for modeling the impacts of intra-specific competition on species extinction, for example due to linkage disequilibrium between $c$ on the one hand, and $b$ and $d$ on the other.
+Previous work on this question in the case of sexual conflict used problem-specific fitness models which directly imposed an absolute fitness cost on competition \citep{rankin_2011,vandoorn_2006}.
 
-In the broader context of eco-evolutionary synthesis [?
-Cite], we can imagine a spectrum of approaches ranging from population genetics, which is entirely evolutionary (no ecology) and has a strong emphasis on genetics, to pure ecological theory with no evolution.
+In Fig.
+\ref{fig:table2} we compare our model with some of the other models and schemes touched upon here.
+In a sense this comparison is unfair: for instance, Grime's scheme was developed for an entirely different purpose (species classification by traits).
+As such, Fig.
+\ref{fig:table2} is not exhaustive and should be read more as a summary of our model's purpose.
+Like MacArthur's $r$/$K$ scheme, our model is motivated by the need to expand the treatment of selection in population genetics \citep{macarthur_1962}, i.e.
+to capture some crucial ecological factors in our most genetically realistic models of evolution.
+Thus, viewing evolutionary ecology as a spectrum ranging from evolution-only to ecology-only, our model lies closer to the evolution-only end of the spectrum compared to more familiar approaches to evolutionary ecology such as adaptive dynamics --- essentially ecology plus mutant invasion --- which lies closer to the ecology-only end of the spectrum.
 
@@
 Applications to traits come later, and with mixed success (see ``Introduction'' and ``Discussion'').
-**As a modeling framework, adaptive dynamics is not scenario-specific, and could be applied with any fitness model including our density-dependent lottery, but in practice most of the adaptive dynamics literature focuses on specific eco-evolutionary outcomes such as evolutionary ``branching'' or evolutionary suicide.}
+**As a modeling framework, adaptive dynamics is not scenario-specific, and could be applied with any fitness model including our density-dependent lottery, but in practice most of the adaptive dynamics literature focuses on specific eco-evolutionary outcomes such as evolutionary ``branching''.}
 \end{figure}

### 2016-11-15 manuscript.tex
latest

@@
 
-\textit{Manuscript elements}: Figure~1, figure~2, table~1, online
-appendices~A and B (including figure~A1 and figure~A2). Figure~2 is to
-print in color.
+\textit{Manuscript elements}:
 
@@
 
-\textit{Keywords}: Examples, model, template, guidelines.
+\textit{Keywords}: r/K selection, absolute fitness, eco-evo, interference competition.
 
@@
 This territorial contest is modeled as a weighted lottery: the probability that genotype $i$ wins a given territory by the next iteration is $c_i x_i/\sum_j c_j x_j$ where $c_i$ is a constant representing relative competitive ability.
+This includes cases where territories are won via a sexual contest.
 
@@
 \end{equation}
-which is independent of $c_i$ (here $L=\sum_j l_j$ is the overalFaltinel propagule density).
+which is independent of $c_i$ (here $L=\sum_j l_j$ is the overall propagule density).
 
@@
 
-The stress archetype is more ambiguous, and has been the subject of an extensive debate in the plant ecology literature (the ``Grime-Tilman'' debate; \citealt{aerts_1999}).
+The stress archetype is more ambiguous, and has been the subject of an extensive debate in the plant ecology literature (the ``Grime-Tilman'' debate; \citealt{aerts_1999} and references therein).
 Stressful environments severely restrict growth and reproduction, so that $b_i\ll 1$ \cite{grime_1974,grime_1977}.
@@
 The alternative view is that stressful environments simply have a lower carrying capacity \citep{taylor_1990}; in our model, this means a greater per-individual territorial requirement represented by a lower $T$ for a given amount of space.
-In particular, it is argued that when stress is induced by a scarcity of consumable resources, competition for those resources would likely be intense.
+In particular, it is argued that when stress is induced by a scarcity of consumable resources, competition for those resources would likely be intense (for empirical support, see \citealt{davis_1998}).
 Thus, $b_i$ need not be particularly close to $d_i$, but contests for consumable resources at the juvenile phase would kill off most propagules before adulthood.
-In other words, the stressed population is a high density population where competition is important \citep{taylor_1990} (that is, high density relative to $T$, not relative to ideal conditions).
+In other words, the stressed population is a high density population where competition is important (that is, high density relative to $T$, not relative to ideal conditions; \citealt{taylor_1990}).
 
@@
 
-In our view, our model has two major limitations as a general-purpose model of density-dependent selection: a reliance on interference competition for durable resources (territoriality), and the restriction of competition to juveniles (lottery recruitment to adulthood).
-In some respects this is the complement of resource competition models, which restrict their attention to exploitation competition, typically without age structure \citep{tilman_1982}.
-In the particular case that resources are spatially localized (e.g.
-due to restricted movement through soils), then resource competition and territorial acquisition effectively coincide, and in principle a competitive ability $c$ should be derivable from resource competition.
-The situation is more complicated if the resources are well-mixed, since, in general, resource levels then need to be explicitly tracked.
-It is possible that explicit resource tracking is in fact not necessary when the focus is on the evolution of similar genotypes rather than the stable co-existence of widely differing species.
-We are not aware of any attempts to delineate the conditions under which explicit resource tracking is not required, let alone to connect those results to the density-dependent selection literature [need to check Tilman's book].
-
-On the other hand, our model does remarkably well in capturing apparently general features of selection under different environmental conditions using only three trait parameters: $b$, $c$ and $d$.
-The clean separation of a strictly-relative $c$ parameter is particularly useful from an evolutionary genetics perspective, essentially embedding a relative fitness trait within an absolute fitness model.
-This could have interesting applications for modeling the impacts of intra-specific competition on species extinction, for example due to linkage disequilibrium between $c$ on the one hand, and $b$ and $d$ on the other.
-Previous work on this question in the case of sexual conflict used problem-specific fitness models which directly imposed an absolute fitness cost on competition \citep{rankin_2011,vandoorn_2006}.
-
 In Fig.
@@
 Like MacArthur's $r$/$K$ scheme, our model is motivated by the need to expand the treatment of selection in population genetics \citep{macarthur_1962}, i.e.
-to capture some crucial ecological factors in our most genetically realistic models of evolution.
-Thus, viewing evolutionary ecology as a spectrum ranging from evolution-only to ecology-only, our model lies closer to the evolution-only end of the spectrum compared to more familiar approaches to evolutionary ecology such as adaptive dynamics --- essentially ecology plus mutant invasion --- which lies closer to the ecology-only end of the spectrum.
+to incorporate crucial ecological factors in our most genetically realistic models of evolution.
+Thus, viewing evolutionary ecology \citep{kokko_2007,pelletier_2009,schoener_2011} as a spectrum ranging from evolution-only to ecology-only, our model lies close to the understudied evolution-only end of the spectrum.
+By comparison, more familiar approaches to evolutionary ecology such as adaptive dynamics --- essentially ecology coupled with mutant invasion \citep{diekmann_2004} ---  lie close to the ecology-only end of the spectrum.
 
@@
   Eco-evo:  \\
-  \quad Adaptive dynamics & \ding{51} & \ding{51} & \ding{51} & \ding{55}** & \ding{55} \\
+  \quad Adaptive dynamics & \ding{51} & \ding{51} & \ding{51} & \ding{51}** & \ding{55} \\
   \quad Brute force simulation & \ding{51} & \ding{51} & NA & \ding{55} & \ding{51}
@@
 *MacArthur's $r$- and $K$-, as well as $\alpha$ selection, were all derived theoretically.
-Applications to traits come later, and with mixed success (see ``Introduction'' and ``Discussion'').
-**As a modeling framework, adaptive dynamics is not scenario-specific, and could be applied with any fitness model including our density-dependent lottery, but in practice most of the adaptive dynamics literature focuses on specific eco-evolutionary outcomes such as evolutionary ``branching''.}
+Applications to traits came later, and with mixed success (see ``Introduction'' and ``Discussion'').
+**In practice, most of the adaptive dynamics literature focuses on specific eco-evolutionary outcomes such as evolutionary ``branching'' \citep{geritz_1997} or mutualisms \citep{ferriere_2002}, but in principle it can be applied with any fitness model including our density-dependent lottery.}
 \end{figure}
+
+In our view, our model has two major limitations as a general-purpose model of density-dependent selection: a reliance on interference competition for durable resources (territoriality), and the restriction of competition to juveniles (lottery recruitment to adulthood).
+In some respects this is the complement of resource competition models, which restrict their attention to exploitation competition, typically without age structure \citep{tilman_1982}.
+In the particular case that resources are spatially localized (e.g.
+due to restricted movement through soils), then resource competition and territorial acquisition effectively coincide, and in principle resource competition could be represented by a competitive ability $c$ (or conversely, $c$ should be derivable from resource competition).
+The situation is more complicated if the resources are well-mixed, since, in general, resource levels then need to be explicitly tracked.
+It seems plausible that explicit resource tracking is not necessary when the focus is on the evolution of similar genotypes rather than the stable co-existence of widely differing species.
+We are not aware of any attempts to delineate conditions under which explicit resource tracking is unnecessary even if it is assumed that community structure is ultimately determined by competition for consumable resources.
+More work is needed connecting resource competition models to the density-dependent selection literature, since most of the existing work is focused on the narrower Grime-Tilman-debate issue of resolving the role of competition at different levels of resource availability \citep{aerts_1999,davis_1998,tilman_2007}.
+
+On the other hand, our model does remarkably well in capturing apparently general features of selection under different environmental conditions using only three trait parameters: $b$, $c$ and $d$.
+The clean separation of a strictly-relative $c$ parameter is particularly useful from an evolutionary genetics perspective, essentially embedding a zero-sum fitness trait within a non-zero-sum fitness model.
+This could have interesting applications for modeling the impacts of intra-specific competition on species extinction, for example due to clonal interference \citep{gerrish_1998,desai_2007} between $c$-strategists on the one hand, and $b$- and $d$- strategists on the other.
 
@@
 Under the Poisson approximation, the total number of genotype $i$ propagules $\sum x_i$ (summed over unoccupied territories) will deviate about its mean value $m_i$.
-Since the coefficient of variation of $\sum x_i$ is proportional to $1/\sqrt{m_i}$, these deviations are negligible unless $m_i$ is very small (say of order $10^2$ or less).
+Since the coefficient of variation of $\sum x_i$ is proportional to $1/\sqrt{m_i}$, these fluctuations are negligible unless $m_i$ is very small (say of order $10^2$ or less).
+These fluctuations in $m_i$ could be regarded as a feature rather than a flaw since having $m_i$ be exactly constant per generation (for given $b_i$ and $n_i$) is biologically unrealistic.
+In the canonical model of genetic drift, the Wright-Fisher model, the number of offspring per genotype fluctuates from generation to generation and is approximately Poisson distributed.
+Nevertheless, for simplicity and ease of comparison with the classic lottery model, we ignore fluctuations in $m_i$ and only account for Poisson fluctuations in the number of propagules landing in each territory.
 

### 2016-11-17 manuscript.tex
latest

@@
 276]{ewens_2012}.
-For example, MacArthur's famous r/K selection scheme is now almost exclusively known as a framework for understanding life-history traits, and judged on its failure in that role \citep{pianka_1970,stearns_1977,boyce_1984,reznick_2002}.
+For example, MacArthur's famous $r$/$K$ scheme \citep{macarthur_1962,macarthur_1967} is now almost exclusively known as a framework for understanding life-history traits, and judged on its failure in that role \citep{pianka_1970,stearns_1977,boyce_1984,reznick_2002}.
 The r/K scheme's original purpose was as an extension of the existing population-genetic treatment of selection to account for population density \citep{macarthur_1962}, but few attempts have been made to develop it further as a mathematical analysis of the major different forms of selection.
@@
 As discussed in section ``Primary strategies and Grime's triangle'', over time our predicted trait evolution should lead to trait values consistent with Grime's scheme.
-In making these predictions, we have made no reference to any kind of trade-offs or pleitropy, even though trade-offs are often invoked to explain primary strategy schemes \citep{macarthur_1962,winemiller_1992,aerts_1999}.
+In making these predictions, we have made no reference to any kind of trade-offs or pleitropy, even though trade-offs are often invoked to explain primary strategy schemes \citep{macarthur_1967,winemiller_1992,aerts_1999}.
 Thus, while trade-offs may amplify specialization, they are not necessary for it.
@@
 
-Our prediction of evolution for higher $b$ in ideal environments is counter to the expectations of MacArthur's $r$/$K$ dichotomy \citep{macarthur_1962} since $b$ is closely related to the maximal, low-density growth rate $r=b-d$, and ideal environments support high population densities which should be subject to ``$K$-selection''.
+Our prediction of evolution for higher $b$ in ideal environments is counter to the expectations of MacArthur's $r$/$K$ dichotomy \citep{macarthur_1967} since $b$ is closely related to the maximal, low-density growth rate $r=b-d$, and ideal environments support high population densities which should be subject to ``$K$-selection''.
 However, in the Introduction, we noted that the $r$-$K$ dichotomy is not consistent with empirical studies showing that maximal growth rate and saturation density (measured by abundance) are positively correlated, both between species/strains \citep{luckinbill_1979,kuno_1991,hendriks_2005,fitzsimmons_2010}, and as a result of experimental evolution \citep{luckinbill_1978,luckinbill_1979}.
@@
 
-$K$-selection in the sense of selection for a greater environmental carrying capacity for given birth and death rates, sometimes referred to as ``efficiency'' \citep{macarthur_1962}, would be represented in our model by smaller individual territorial requirements.
+$K$-selection in the sense of selection for a greater environmental carrying capacity for given birth and death rates, sometimes referred to as ``efficiency'' \citep{macarthur_1967}, would be represented in our model by smaller individual territorial requirements.
 To a first approximation, two co-occurring genotypes which differ by a small amount in their territorial requirements only should have the same fitness since the costs or benefits of a change in the amount of unocupied territory is shared equally among genotypes via the propagule density $L$.

### 2016-12-20 manuscript.tex
latest

@@
 As we will see, a generalized version of the classic lottery model of \cite{chesson_1981} is well suited for this purpose.
-The lottery model gets its name from the way in which it represents competition: the growth of each type is determined by the number of contestants that it produces relative to the other types in the population (akin to a lottery),  weighted by a coefficient for each type representing its competitive ability.
+The lottery model gets its name from the way in which it represents competition: the growth of each type depends on how many territories it secures from a fixed territorial stock, which is determined by the number of contestants that it produces relative to the other types in the population weighted by a coefficient for each type representing competitive ability.
 This is much simpler than having coefficients for the pairwise effects of types on each other (e.g.
 the $\alpha$ coefficients in the generalized Lotka-Volterra equations), or than modeling resource consumption explicitly \citep{tilman_1982}.
-The classic lottery model  breaks down at low larval densities (section ``Mean field approximation''), but this was not important for its original application to reef fishes, where a huge number of larvae from each type compete to secure territories each generation \citep{chesson_1981}.
+The classic lottery model  breaks down at low absolute abundances at the scale of territorial contests (section ``Mean field approximation''), but this was not important for its original application to reef fishes, where a huge number of larvae from each type compete to secure territories each generation \citep{chesson_1981}.
 Here we analytically extend the classic lottery model to correctly account for low density behavior.
@@
 
-Each iteration, adults produce $m_i$ new offspring (henceforth called ``propagules'') which disperse at random over the $U$ unnocupied territories (no dispersal limitation).
-We assume adults cannot be ousted from occupied territories, so only propagules landing on occupied territories are included in $m_i$.
+Each iteration, adults produce $m_i$ new offspring (henceforth called ``propagules'').
+We assume adults cannot be ousted from occupied territories, so only propagules landing on unoccupied territories are included in $m_i$.
+Dispersal is assumed to be unrestricted, so that all propagules have the same probability of landing on any of the $U$ unnocupied territories.
+Propagules disperse independently, and
+
  More generally, $m_i$ only includes propagules which actually end up contesting unnocupied territories.
@@
 When multiple propagules land on the same territory, they compete to secure the territory as they develop.
-This territorial contest is modeled as a weighted lottery: the probability that genotype $i$ wins a given territory by the next iteration is $c_i x_i/\sum_j c_j x_j$ where $c_i$ is a constant representing relative competitive ability.
-This includes cases where territories are won via a sexual contest.
+This territorial contest is modeled as a weighted lottery: the probability that genotype $i$ wins a territory by the next iteration, assuming that at least one of its propagules is present, is $c_i x_i/\sum_j c_j x_j$, where $c_i$ is a constant representing relative competitive ability.
 

### 2016-12-20 manuscript.tex
latest

@@
 We assume adults cannot be ousted from occupied territories, so only propagules landing on unoccupied territories are included in $m_i$.
-Dispersal is assumed to be unrestricted, so that all propagules have the same probability of landing on any of the $U$ unnocupied territories.
-Propagules disperse independently, and
-
- More generally, $m_i$ only includes propagules which actually end up contesting unnocupied territories.
-For simplicity, we assume $m_i=b_i n_i$, where $b_i$ is a constant, genotype-specific birth rate.
+Dispersal is random and unrestricted; each propagule has an equal probability of landing on any of the $U$ unnocupied territories.
+Loss of propagules during dispersal is subsumed into $m_i$ ($m_i$ only counts survivors which go on to contest the territory they land in), and we assume that propagules disperse independently of each other.
+In general, $m_i$ will increase with $n_i$, and will also depend on population density $N$.
+For example, if $b_i$ is the number of propagules produced per individual from genotype $i$, then $m_i=b_i(1-N/T)n_i$ would reflect the loss of propagules due to random dispersal to occupied territories.
+Nevertheless, for simplicity we will assume a linear function $m_i=b_i n_i$.
+Our justification is that the specific choice of $m_i$ makes little difference to our analysis here other than potentially complicating the mathematics.
+However, the choice of $m_i$ could matter in other applications, particularly when the mode of dispersal is a central concern [cite coexistence].
 

### 2016-12-21 manuscript.tex
latest

@@
 \caption{\label{fig:lottery} Each iteration of our lottery model has three main elements.
-Each generation, propagules are produced by adults which are dispersed at random over the unoccupied territories.
+First, propagules are produced by adults which are dispersed at random over the unoccupied territories (only propagules landing on unoccupied territories are shown).
 Lottery competition then occurs in each unoccupied territory: each genotype has a probability proportional to $b_i n_i c_i$ of securing the territory.
@@
 In general, $m_i$ will increase with $n_i$, and will also depend on population density $N$.
-For example, if $b_i$ is the number of propagules produced per individual from genotype $i$, then $m_i=b_i(1-N/T)n_i$ would reflect the loss of propagules due to random dispersal to occupied territories.
-Nevertheless, for simplicity we will assume a linear function $m_i=b_i n_i$.
-Our justification is that the specific choice of $m_i$ makes little difference to our analysis here other than potentially complicating the mathematics.
-However, the choice of $m_i$ could matter in other applications, particularly when the mode of dispersal is a central concern [cite coexistence].
+For example, if $b_i$ is the number of propagules produced per individual from genotype $i$, then $m_i=b_i(1-N/T)n_i$ would reflect the loss of propagules that disperse to occupied territories.
+Nevertheless, we will assume $m_i=b_i n_i$, meaning that all propagules land on unoccupied territories (a form of directed dispersal).
+This choice simplifies the mathematics without seriously restricting the generality of our analysis, since the results presented here are not sensitive to the specific functional form of $m_i$.
 
@@
 The increase in $n_i$ over one iteration due to territorial acquisition, $\Delta_+ n_i$, is the sum of genotype $i$'s victories over all $U$ unoccupied territories.
-Since $p_1(x_1)\ldots p_G(x_G)$ is equal to the proportion of unoccupied territories with $x_1,\ldots,x_G$ of the respective propagules (where $G$ is the number of genotypes present; and again, we assume that $T$ is large enough that fluctuations in this proportion are negligible), this sum can be replaced by an expectation over the $p_i$.
-This gives
+The expected proportion of unoccupied territories with $x_1,\ldots,x_G$ of the respective propagules is $p_1(x_1)\ldots p_G(x_G)$ ($G$ is the number of genotypes present), and the probability of $i$ winning in each of these territories is $c_i x_i/\sum_j c_j x_j$.
+As above, we assume that $T$ and the $n_i$ are large enough that we can ignore fluctuations in the proportion of unoccupied territories as well as the number of victories, which implies
 \begin{equation}
-\Delta_+ n_i(t)=U(t)\sum_{x_1,\ldots,x_G} \frac{c_i x_i}{\sum_j c_j x_j} p_1(x_1)\ldots p_G(x_G). \label{eq:growthsumuncoupled}
+\Delta_+ n_i(t)=U(t)\sum_{x_1,\ldots,x_G} \frac{c_i x_i}{\sum_j c_j x_j} p_1(x_1)\ldots p_G(x_G), \label{eq:growthsumuncoupled}
 \end{equation}
+where the sum only includes territories with at least one propagule present.
 
@@
 
-The propagule numbers $x_i$ in different territories are not independent random variables.
-To determine the dispersal outcomes in all unoccupied territories exactly, we would need to proceed territory-by-territory as follows.
-In the first territory we evaluate, $x_i$ drawn from a binomial distribution with $m_i$ trials and success probability $1/U$.
-In the second, $x_i$ is drawn from a binomial distribution with $m_i-x$ trials and success probability $1/(U-1)$, where $x$ is the number of propagules that landed in the first territory.
-And so on.
+The propagule numbers $x_i$ in different territories obey a multinomial distribution with
+
+are not independent random variables.
 

### 2016-12-21 manuscript.tex
latest

@@
 Each iteration, adults produce $m_i$ new offspring (henceforth called ``propagules'').
-We assume adults cannot be ousted from occupied territories, so only propagules landing on unoccupied territories are included in $m_i$.
-Dispersal is random and unrestricted; each propagule has an equal probability of landing on any of the $U$ unnocupied territories.
-Loss of propagules during dispersal is subsumed into $m_i$ ($m_i$ only counts survivors which go on to contest the territory they land in), and we assume that propagules disperse independently of each other.
+We assume adults cannot be ousted from occupied territories, and only propagules landing on unoccupied territories are included in $m_i$.
+Propagules disperse at random, independently of each other, and without spatial restrictions; each propagule has an equal probability of landing on any of the $U$ unnocupied territories.
+Thus, there is no interaction between propagules (e.g.
+avoidance of territories crowded with propagules).
+Loss of propagules during dispersal is subsumed into $m_i$ ($m_i$ only counts propagules which go on to contest the territory they land in).
 In general, $m_i$ will increase with $n_i$, and will also depend on population density $N$.
-For example, if $b_i$ is the number of propagules produced per individual from genotype $i$, then $m_i=b_i(1-N/T)n_i$ would reflect the loss of propagules that disperse to occupied territories.
-Nevertheless, we will assume $m_i=b_i n_i$, meaning that all propagules land on unoccupied territories (a form of directed dispersal).
-This choice simplifies the mathematics without seriously restricting the generality of our analysis, since the results presented here are not sensitive to the specific functional form of $m_i$.
+For example, if $b_i$ is the number of propagules produced per individual from genotype $i$, then $m_i=b_i(1-N/T)n_i$ would reflect the loss of propagules due to dispersal to occupied territories.
+We assume $m_i=b_i n_i$, meaning that all propagules land on unoccupied territories (a form of directed dispersal).
+This choice is not biologically motivated; it simplifies the mathematics without seriously restricting the generality of our analysis, since the results presented here are not sensitive to the specific functional form of $m_i$.
 
@@
 
-The propagule numbers $x_i$ in different territories obey a multinomial distribution with
-
-are not independent random variables.
-
-For sufficiently large $K$, holding $n_i/T$ fixed, the Poisson limit theorem implies that the binomial distributions for $x_i$ at each successive stage of this procedure are all closely approximated by a Poisson distribution with mean $l_i$, where we have used the fact that large $T$ implies large $U$ except in the biologically uninteresting case that there is vanishing population turnover $d_i \sim 1/T$.
+For each genotype's dispersal, the counts of propagules across the unnocupied territories follows a multinomial distribution with equal probabilities of ``success'' for all territories.
+Thus, the $x_i$ in different territories are not independent random variables.
+However, for sufficiently large $T$, holding $n_i/T$ fixed, the Poisson limit theorem implies that this multinomial distribution for the $x_i$ accross territories is closely approximated by a product of independent Poisson distributions for each territory, each with rate parameter $l_i$.
+Here we have used the fact that large $T$ implies large $U$ except in the biologically uninteresting case that there is vanishing population turnover $d_i \sim 1/T$.
 

### 2016-12-21 manuscript.tex
latest

@@
 The number of individuals of the $i$'th genotype landing in any particular territory is denoted $x_i$.
-Random dispersal implies that in the limit $T\rightarrow \infty$, with $n_i/T$ held fixed, $x_i$ is Poisson distributed with mean territorial propagule density $l_i=m_i/U$.
+Random dispersal implies that in the limit $T\rightarrow \infty$, with $n_i/T$ held fixed, $x_i$ is Poisson distributed with mean territorial propagule density $l_i=m_i/U$ (this dispersal Poisson distribution is denoted $p_i(x_i)=l_i^{x_i} e^{-l_i}/x_i!$).
 Although $T$ is finite in our model, we assume that $T$ and the $n_i$ are large enough that $x_i$ is Poisson-distributed to a good approximation (details in Appendix A).
-This dispersal Poisson distribution is denoted $p_i(x_i)$.
 Note that the large $n_i$, large $T$ approximation places no restrictions on our densities $n_i/T$, but it does preclude consideration of demographic stochasticity when $n_i$ itself is very small (this will be discussed further in Section ``Invasion of rare genotypes and coexistence'').
@@
 \eqref{eq:equil} and \eqref{eq:invad} imply that invasion occurs when $b_jd_i-b_id_j>0$, and in particular when $b_j>b_i$ with $d_j=d_i$, or when $d_j<d_i$ with $b_j=b_i$.
-In the case that the new genotype has a different competitive ability, Eqs.
+In the case that the new genotype has a different competitive ability but the same $b_i$ and $d_i$, Eqs.
 \eqref{eq:equil} and \eqref{eq:invad} imply that invasion occurs when $R_j c_j/c_i > A_i$; it is not hard to verify that this occurs if and only if $c_j>c_i$ using the simplified expressions for $A_i$ and $R_j$ given after Eqs.
@@
 
-\begin{figure}
-\centering
-\includegraphics[scale=0.7]{coex.pdf}
-\caption{\label{fig:coex} Coexistence between $b$ ($c_i=1$, $b_i=1$) and $c$ ($c_j=2$, $b_j=0.7$) specialists, where $d_i=d_j=0.3$.
-Vertical axis shows frequency of the $c$-specialist predicted by Eq.
-\eqref{eq:master}.}
-\end{figure}
-
 A similar argument applies for coexistence between high-$c$ and low-$d$ specialists; again coexistence occurs because the importance of territorial contests declines along with propagule density $L$ as the $c$-specialist increases in frequency.
-Coexistence is technically possible between $b$- and $d$-specialists which exactly satisfy $b_i/d_i=b_j/d_j$ (this follows from the fact that all propagules have the same probability of success when $c_i=c_j$ i.e.
+Mutual invasibility is not possible between $b$- and $d$-specialists (although genotypes with exactly $c_i=c_j$ and $b_i/d_i=b_j/d_j$ do not exclude each other --- this follows from the fact that all propagules have the same probability of success when $c_i=c_j$ i.e.
 $A_i+R_i=A_j+R_j$).
-However, this coexistence scenario is not biologically relevant, since the tiniest deviation from $b_i/d_i=b_j/d_j$ will lead to the eventual exclusion of the genotype with greater $b_i/d_i$.
 
@@
 \eqref{eq:master}.
-For mutants with fitness greater than the population mean fitness, this process is known as ``establishment'', and occurs when $n_j$ is of order $1/s$ \citep{desai_2007}, where the selection coefficient $s$ is the mutant's fitness advantage (i.e.
-the mutant's fitness minus the population mean fitness).
+For mutants with fitness greater than the population mean fitness, this occurs when $n_j$ is of order $1/s$ \citep{desai_2007}, where the selection coefficient $s$ is the mutant's fitness advantage (i.e.
+$s=\frac{\Delta n_i/n_i}{\sum_i\Delta n_i/n_i\times n_i/N}-1$).
 Here we do not consider the initial stochastic behavior of novel mutants, and have restricted our attention to the earliest deterministic behavior of rare genotypes.
@@
 
-For each genotype's dispersal, the counts of propagules across the unnocupied territories follows a multinomial distribution with equal probabilities of ``success'' for all territories.
+For each genotype's dispersal, the counts of propagules across unnocupied territories follows a multinomial distribution with equal probabilities of landing in each territory.
 Thus, the $x_i$ in different territories are not independent random variables.

### 2016-12-22 manuscript.tex
latest

@@
 First, propagules are produced by adults which are dispersed at random over the unoccupied territories (only propagules landing on unoccupied territories are shown).
-Lottery competition then occurs in each unoccupied territory: each genotype has a probability proportional to $b_i n_i c_i$ of securing the territory.
+Lottery competition then occurs in each unoccupied territory (only competition in one territory is illustrated): each genotype has a probability proportional to $b_i n_i c_i$ of securing the territory.
 Then occupied territories are freed up by adult mortality.
@@
 The stress archetype is more ambiguous, and has been the subject of an extensive debate in the plant ecology literature (the ``Grime-Tilman'' debate; \citealt{aerts_1999} and references therein).
-Stressful environments severely restrict growth and reproduction, so that $b_i\ll 1$ \cite{grime_1974,grime_1977}.
-Mutations which appreciably improve $b_i$ will be either non-existent or extremely unlikely, so $b_i$ is constrained to remain low.
+Stressful environments severely restrict growth and reproduction, so that $b\ll 1$ \cite{grime_1974,grime_1977}.
+Mutations which appreciably improve $b$ will be either non-existent or extremely unlikely, so $b$ is constrained to remain low.
 In Grime's view, under these conditions the rate at which propagules successfully develop to adulthood cannot appreciably exceed the mortality rate.
-This implies $b_i/d_i\approx 1$ in our model, and so the propagule density $L$ is suppressed to such low levels that there are essentially no territorial contests occurring.
+This implies $b/d\approx 1$ in our model, and so the propagule density $L$ is suppressed to such low levels that there are essentially no territorial contests occurring.
 
-The alternative view is that stressful environments simply have a lower carrying capacity \citep{taylor_1990}; in our model, this means a greater per-individual territorial requirement represented by a lower $T$ for a given amount of space.
-In particular, it is argued that when stress is induced by a scarcity of consumable resources, competition for those resources would likely be intense (for empirical support, see \citealt{davis_1998}).
-Thus, $b_i$ need not be particularly close to $d_i$, but contests for consumable resources at the juvenile phase would kill off most propagules before adulthood.
-In other words, the stressed population is a high density population where competition is important (that is, high density relative to $T$, not relative to ideal conditions; \citealt{taylor_1990}).
+The alternative view is that, while stressful environments imply lower $b$ and support a lower number of individuals per unit area compared what is attainable in ideal environments, stressed populations are actually at high densities relative to the environmental carrying capacity, and are highly competitive \citep{taylor_1990}.
+In the particular case that stress is caused by scarcity of consumable resources, it is argued that we should expect intense resource competition (for empirical support, see \citealt{davis_1998}).
+Thus, $b$ may actually appreciably exceed $d$ under stressful conditions, even though the absolute value of $b$ is small.
 

### 2017-01-05 manuscript.tex
latest

@@
 As we will see, a generalized version of the classic lottery model of \cite{chesson_1981} is well suited for this purpose.
-The lottery model gets its name from the way in which it represents competition: the growth of each type depends on how many territories it secures from a fixed territorial stock, which is determined by the number of contestants that it produces relative to the other types in the population weighted by a coefficient for each type representing competitive ability.
-This is much simpler than having coefficients for the pairwise effects of types on each other (e.g.
+In the classic lottery model, mature individuals (henceforth ``adults'') each require their own territory, whereas newborn individuals (henceforth ``propagules'') disperse to, and subsequently compete for, the territories made available by the death of adults each generation.
+Territorial contest among propagules leaves a single victorious adult per territory, with the victor chosen at random from the propagules present (akin to a lottery) with probabilities weighted by a coefficient for each type representing competitive ability.
+This lottery competition is much simpler than having coefficients for the pairwise effects of types on each other (e.g.
 the $\alpha$ coefficients in the generalized Lotka-Volterra equations), or than modeling resource consumption explicitly \citep{tilman_1982}.
-The classic lottery model  breaks down at low absolute abundances at the scale of territorial contests (section ``Mean field approximation''), but this was not important for its original application to reef fishes, where a huge number of larvae from each type compete to secure territories each generation \citep{chesson_1981}.
+
+In addition, the population-genetic Wright-Fisher model (each generation is a sample of constant size, with replacement, from the previous one) can be thought of as a special case of the classic lottery model.
+
+The classic lottery model assumed that the number of propagules and available territories is essentially infinitely large, where each available territory is swamped with competing propagules every generation (its original application was to reef fishes, where a huge number of larvae from each species compete to secure territories each generation; \citealt{chesson_1981}).
+
+As a result, type abundances change deterministically (in a constant environment). Nevertheless,
+
+which models gen can be interpreted as a type of lottery model
+
+Adults can persist over multiple generations, and so the classic lottery model allows for overlapping generations;
+
+this is essential for the coexistence mechanisms that
+
+propagule dispersal and competition is resolved over one generation.
+
+This is a feature
+
+ breaks down at low absolute abundances at the scale of territorial contests (section ``Mean field approximation''),
+
 Here we analytically extend the classic lottery model to correctly account for low density behavior.

### 2017-01-05 manuscript.tex
latest

--- 2016-12-22 manuscript.tex
@@
 
-As we will see, a generalized version of the classic lottery model of \cite{chesson_1981} is well suited for this purpose.
-The lottery model gets its name from the way in which it represents competition: the growth of each type depends on how many territories it secures from a fixed territorial stock, which is determined by the number of contestants that it produces relative to the other types in the population weighted by a coefficient for each type representing competitive ability.
-This is much simpler than having coefficients for the pairwise effects of types on each other (e.g.
+As we will see, a generalized version of the classic lottery model of \cite{chesson_1981} is well suited for our purposes.
+In the classic lottery model, mature individuals (henceforth ``adults'') each require their own territory, whereas newborn individuals (henceforth ``propagules'') disperse to, and subsequently compete for, the territories made available by the death of adults.
+Territorial contest among propagules leaves a single victorious adult per territory, with the victor chosen at random from the propagules present (akin to a lottery), with probabilities weighted by a coefficient for each type representing competitive ability.
+This representation of competition is much simpler than having coefficients for the pairwise effects of types on each other (e.g.
 the $\alpha$ coefficients in the generalized Lotka-Volterra equations), or than modeling resource consumption explicitly \citep{tilman_1982}.
-The classic lottery model  breaks down at low absolute abundances at the scale of territorial contests (section ``Mean field approximation''), but this was not important for its original application to reef fishes, where a huge number of larvae from each type compete to secure territories each generation \citep{chesson_1981}.
-Here we analytically extend the classic lottery model to correctly account for low density behavior.
+The classic lottery model also has similarities with the Wright-Fisher model in population-genetics (section BLAH).
+
+However, the classic lottery model breaks down at low densitites, specifically if there are only  a few propagules dispersing to each territory (section ``Mean field approximation'').
+This was not a limitation in its original application to reef fishes, where a huge number of larvae from each species compete to secure territories each generation \cite{chesson_1981}, but is a critical limitation here.
+We analytically extend the classic lottery model to correctly account for low density behavior.
+
+%(each generation is a sample of constant size, with replacement, from the previous one)
 

### 2017-01-09 manuscript.tex
latest

@@
 As we will see, a generalized version of the classic lottery model of \cite{chesson_1981} is well suited for our purposes.
-In the classic lottery model, mature individuals (henceforth ``adults'') each require their own territory, whereas newborn individuals (henceforth ``propagules'') disperse to, and subsequently compete for, the territories made available by the death of adults.
+In the classic lottery model, mature individuals (``adults'') each require their own territory, whereas newborn individuals (``propagules'') disperse to, and subsequently compete for, the territories made available by the death of adults.
 Territorial contest among propagules leaves a single victorious adult per territory, with the victor chosen at random from the propagules present (akin to a lottery), with probabilities weighted by a coefficient for each type representing competitive ability.
@@
 
-However, the classic lottery model breaks down at low densitites, specifically if there are only  a few propagules dispersing to each territory (section ``Mean field approximation'').
-This was not a limitation in its original application to reef fishes, where a huge number of larvae from each species compete to secure territories each generation \cite{chesson_1981}, but is a critical limitation here.
+However, the classic lottery model breaks down at low densitites (that is, if there are only a few propagules dispersing to each territory; section ``Mean field approximation'').
+This was not a limitation in the lottery model's original application to reef fishes, where a huge number of larvae from each species compete to secure territories each generation \citep{chesson_1981}, but is a critical limitation for studying density-dependent selection.
 We analytically extend the classic lottery model to correctly account for low density behavior.
@@
 Analytical expressions for the change in genotype abundances over time are introduced in section ``Mean field approximation'', with mathematical details relegated to the Appendices.
-The following two sections discuss the behavior of rare mutants and our treatment of Grime's triangle.
+In the following two sections, we then discuss the behavior of rare mutants and our treatment of Grime's triangle.
 
@@
 
-We assume that each individual in a population requires its own territory to survive and reproduce (a site-occupancy model).
+We assume that each individual in a population requires its own territory to survive and reproduce (a site-occupancy model; Fig.~\ref{fig:lottery}).
 All territories are identical, and the total number of territories is $T$.
@@
 First, propagules are produced by adults which are dispersed at random over the unoccupied territories (only propagules landing on unoccupied territories are shown).
-Lottery competition then occurs in each unoccupied territory (only competition in one territory is illustrated): each genotype has a probability proportional to $b_i n_i c_i$ of securing the territory.
+Lottery competition then occurs in each unoccupied territory (competition in only one territory is illustrated): each genotype has a probability proportional to $b_i n_i c_i$ of securing the territory.
 Then occupied territories are freed up by adult mortality.
@@
 We assume $m_i=b_i n_i$, meaning that all propagules land on unoccupied territories (a form of directed dispersal).
-This choice is not biologically motivated; it simplifies the mathematics without seriously restricting the generality of our analysis, since the results presented here are not sensitive to the specific functional form of $m_i$.
+This choice simplifies the mathematics without seriously restricting the generality of our analysis, since the results presented here are not sensitive to the specific functional form of $m_i$.
 
@@
 Although $T$ is finite in our model, we assume that $T$ and the $n_i$ are large enough that $x_i$ is Poisson-distributed to a good approximation (details in Appendix A).
-Note that the large $n_i$, large $T$ approximation places no restrictions on our densities $n_i/T$, but it does preclude consideration of demographic stochasticity when $n_i$ itself is very small (this will be discussed further in Section ``Invasion of rare genotypes and coexistence'').
+Note that the large $n_i$, large $T$ approximation places no restrictions on our densities $n_i/T$, but it does preclude consideration of demographic stochasticity when $n_i$ itself is very small (this will be discussed further in Section ``Invasion of rare genotypes and coexistence'' BLAH).
 
 When multiple propagules land on the same territory, they compete to secure the territory as they develop.
-This territorial contest is modeled as a weighted lottery: the probability that genotype $i$ wins a territory by the next iteration, assuming that at least one of its propagules is present, is $c_i x_i/\sum_j c_j x_j$, where $c_i$ is a constant representing relative competitive ability.
+This territorial contest is modeled as a weighted lottery: the probability that genotype $i$ wins a territory by the next iteration, assuming that at least one of its propagules is present, is $c_i x_i/\sum_j c_j x_j$, where $c_i$ is a constant representing relative competitive ability (Fig.
+\ref{fig:lottery}).
 
-The increase in $n_i$ over one iteration due to territorial acquisition, $\Delta_+ n_i$, is the sum of genotype $i$'s victories over all $U$ unoccupied territories.
-The expected proportion of unoccupied territories with $x_1,\ldots,x_G$ of the respective propagules is $p_1(x_1)\ldots p_G(x_G)$ ($G$ is the number of genotypes present), and the probability of $i$ winning in each of these territories is $c_i x_i/\sum_j c_j x_j$.
-As above, we assume that $T$ and the $n_i$ are large enough that we can ignore fluctuations in the proportion of unoccupied territories as well as the number of victories, which implies
+In the classic lottery model \citep{chesson_1981}, unoccupied territories are saturated with propagules from every genotype i.e.
+the mean territorial propagule density $l_i=m_i/U$ is much larger than one for all genotypes.
+The law of large numbers then implies that the composition of propagules in each territory will not deviate appreciably from the mean composition $l_1,l_2,\ldots,l_G$ ($G$ is the number of genotypes present), and so the probability that genotype $i$ wins any particular unoccupied territory is approximately $c_i l_i/\sum_j c_j l_j$.
+Since $T$ is large (by assumption), genotype $i$ wins approximately $c_i l_i/\sum_j c_j l_j$ of the $U$ available territories, and so the increase in $n_i$ due to territorial acquisition in the classic lottery model is given by
+\begin{equation}
+\Delta_+ n_i(t)=U(t)\frac{c_i l_i}{\sum_j c_j l_j}= b_i n_i\frac{1}{L}\frac{c_i}{\overline{c}}, \label{eq:lottery}
+\end{equation}
+where $\overline{c}=\sum_j c_j m_j/M$ is the mean propagule competitive ability for a randomly selected propagule, $L=M/U$ is the total propagule density and $M=\sum_j m_j$ is the total number of propagules.
+
+Unlike the classic lottery model, we do not restrict ourselves to high propagule densities.
+In our extension to arbitrary densities, we expect that $p_1(x_1)\ldots p_G(x_G)$ of the $U$ unoccupied territories will have the propagule composition $x_1,\ldots,x_G$.
+Within these $U(t) p_1(x_1)\ldots p_G(x_G)$ territories,
+
+Since $T$ and the $n_i$ are large enough that we can ignore fluctuations in this partitioning of the unoccupied territories as well as the number of victories, which implies
 \begin{equation}
@@
 
+%The expected Eq. \eqref{eq:lottery}
+%Thus, the numbers of territorial victories $\Delta_+ n_1,\Delta_+ n_2,\ldots$ follow a multinomial distribution with $U$ trials and event probabilities $\frac{c_1 l_1}{\sum_j c_j l_j},\frac{c_2 l_2}{\sum_j c_j l_j},\ldots$.
+
+%Consequently, we can replace $x_i$ with $l_i$ in Eq. \eqref{eq:growthsumuncoupled}.
+
 Eq.
@@
 We now evaluate this expectation using a ``mean field'' approximation; the intuition behind this approximation is as follows.
-
-If the unoccupied territories are saturated with propagules from every genotype ($l_i\gg 1$ for all genotypes), the fluctuations in the $x_i$ are small compared to their means $l_i$ (since the $x_i$ are Poisson distributed), and so the composition of propagules in a territory will only rarely differ appreciably from the mean composition $l_1,l_2,\ldots,l_G$.
-Consequently, we can replace $x_i$ with $l_i$ in Eq.
-\eqref{eq:growthsumuncoupled}.
-This gives the classic lottery model \citep{chesson_1981},
-\begin{equation}
-\Delta_+ n_i(t)=U(t)\frac{c_i m_i}{\sum_j c_j m_j}= b_i n_i\frac{1}{L}\frac{c_i}{\overline{c}}, \label{eq:lottery}
-\end{equation}
-where $\overline{c}=\sum_j c_j m_j/M$ is the mean propagule competitive ability for a randomly selected propagule ($M=\sum_j m_j$ is the total number of propagules).
 

### 2017-01-10 manuscript.tex
latest

@@
 
-In the classic lottery model \citep{chesson_1981}, unoccupied territories are saturated with propagules from every genotype i.e.
-the mean territorial propagule density $l_i=m_i/U$ is much larger than one for all genotypes.
-The law of large numbers then implies that the composition of propagules in each territory will not deviate appreciably from the mean composition $l_1,l_2,\ldots,l_G$ ($G$ is the number of genotypes present), and so the probability that genotype $i$ wins any particular unoccupied territory is approximately $c_i l_i/\sum_j c_j l_j$.
-Since $T$ is large (by assumption), genotype $i$ wins approximately $c_i l_i/\sum_j c_j l_j$ of the $U$ available territories, and so the increase in $n_i$ due to territorial acquisition in the classic lottery model is given by
+In the classic lottery model \citep{chesson_1981}, unoccupied territories are assumed to be saturated with propagules from every genotype i.e.
+the mean propagule density $l_i=m_i/U$ is much larger than one for all genotypes.
+From the law of large numbers, the composition of propagules in each territory will then not deviate appreciably from the mean composition $l_1,l_2,\ldots,l_G$ ($G$ is the number of genotypes present), and so the probability that genotype $i$ wins any particular unoccupied territory is approximately $c_i l_i/\sum_j c_j l_j$.
+Thus, genotype $i$ is expected to win $c_i l_i/\sum_j c_j l_j$ of the $U$ available territories, and deviations from this expected outcome are small (since $T$ is large by assumption).
+Consequently, the increase in $n_i$ due to territorial acquisition in the classic lottery model is given by
 \begin{equation}
-\Delta_+ n_i(t)=U(t)\frac{c_i l_i}{\sum_j c_j l_j}= b_i n_i\frac{1}{L}\frac{c_i}{\overline{c}}, \label{eq:lottery}
+\Delta_+ n_i(t)=\frac{c_i l_i}{\sum_j c_j l_j}U(t)=b_i n_i\frac{1}{L}\frac{c_i}{\overline{c}}, \label{eq:lottery}
 \end{equation}
@@
 
-Unlike the classic lottery model, we do not restrict ourselves to high propagule densities.
-In our extension to arbitrary densities, we expect that $p_1(x_1)\ldots p_G(x_G)$ of the $U$ unoccupied territories will have the propagule composition $x_1,\ldots,x_G$.
-Within these $U(t) p_1(x_1)\ldots p_G(x_G)$ territories,
+In our extension of the classic lottery model, we do not restrict ourselves to high propagule densities, and the propagule composition in different territories will deviate appreciably from $l_1,l_2,\ldots,l_G$.
+We expect that a fraction $p_1(x_1)\ldots p_G(x_G)$ of the $U$ unoccupied territories will have the propagule composition $x_1,\ldots,x_G$.
+Genotype $i$ is expected to win win $c_i x_i/\sum_j c_j x_j$ of these $p_1(x_1)\ldots p_G(x_G) U(t)$ territories.
 

### 2017-01-18 manuscript.tex
latest

@@
 
-However, the classic lottery model breaks down at low densitites (that is, if there are only a few propagules dispersing to each territory; section ``Mean field approximation'').
+However, the classic lottery model breaks down at low densities (that is, if there are only a few propagules dispersing to each territory; section ``Mean field approximation'').
 This was not a limitation in the lottery model's original application to reef fishes, where a huge number of larvae from each species compete to secure territories each generation \citep{chesson_1981}, but is a critical limitation for studying density-dependent selection.
@@
 In the classic lottery model \citep{chesson_1981}, unoccupied territories are assumed to be saturated with propagules from every genotype i.e.
-the mean propagule density $l_i=m_i/U$ is much larger than one for all genotypes.
+$l_i\gg 1$ where $l_i=m_i/U$ is the mean propagule density.
 From the law of large numbers, the composition of propagules in each territory will then not deviate appreciably from the mean composition $l_1,l_2,\ldots,l_G$ ($G$ is the number of genotypes present), and so the probability that genotype $i$ wins any particular unoccupied territory is approximately $c_i l_i/\sum_j c_j l_j$.
@@
 
-In our extension of the classic lottery model, we do not restrict ourselves to high propagule densities, and the propagule composition in different territories will deviate appreciably from $l_1,l_2,\ldots,l_G$.
-We expect that a fraction $p_1(x_1)\ldots p_G(x_G)$ of the $U$ unoccupied territories will have the propagule composition $x_1,\ldots,x_G$.
-Genotype $i$ is expected to win win $c_i x_i/\sum_j c_j x_j$ of these $p_1(x_1)\ldots p_G(x_G) U(t)$ territories.
-
-Since $T$ and the $n_i$ are large enough that we can ignore fluctuations in this partitioning of the unoccupied territories as well as the number of victories, which implies
+In our extension of the classic lottery model, we do not restrict ourselves to high propagule densities, and deviations from the mean propagule composition $l_1,l_2,\ldots,l_G$ will be important.
+Since dispersal is Poisson, we expect that a fraction $p_1(x_1)\ldots p_G(x_G)$ of the $U$ unoccupied territories will have the propagule composition $x_1,\ldots,x_G$.
+Genotype $i$ is expected to win $c_i x_i/\sum_j c_j x_j$ of these.
+Ignoring fluctuations about these two expectations (because $T$ is large), genotype $i$'s territorial acquisition is given by
 \begin{equation}
@@
 \end{equation}
-where the sum only includes territories with at least one propagule present.
+in our extended lottery model, where the sum only includes territories with at least one propagule present.
 
@@
 
-Note that the competitive ability coefficients $c_i$ represent a strictly relative aspect of fitness in the sense that they only influence population size $N$ indirectly by changing genotype frequencies; that may in turn change the population mean birth and death rates.
+Note that the competitive ability coefficients $c_i$ represent a strictly relative aspect of fitness in the sense that they only influence population size $N$ indirectly by changing genotype frequencies.
 This can be seen by summing Eq.
@@
 which is independent of $c_i$ (here $L=\sum_j l_j$ is the overall propagule density).
+Unlike the classic lottery model, not all $U$ unoccupied territories are claimed each iteration; a fraction $e^{-L}$ remain unoccupied.
 

### 2017-01-23 manuscript.tex
latest

@@
 
-We assume that each individual in a population requires its own territory to survive and reproduce (a site-occupancy model; Fig.~\ref{fig:lottery}).
+We assume that the individuals in a population each require their own territory to survive and reproduce (a site-occupancy model; Fig.~\ref{fig:lottery}).
 All territories are identical, and the total number of territories is $T$.
@@
 
-%Consequently, we can replace $x_i$ with $l_i$ in Eq. \eqref{eq:growthsumuncoupled}.
+Eq.
+\eqref{eq:growthsumuncoupled} gives little intuition about the dynamics of density-dependent lottery competition, since it involves an expectation over the dispersal distributions $p_i$, which depend on how the $n_i$ change over time.
+We now evaluate this expectation using a ``mean field'' approximation.
 
-Eq.
-\eqref{eq:delttot} gives little intuition about the dynamics of density-dependent lottery competition, since \eqref{eq:growthsumuncoupled} involves an expectation over the random dispersal distributions $p_i$, which depend on how the $n_i$ change over time.
-We now evaluate this expectation using a ``mean field'' approximation; the intuition behind this approximation is as follows.
+This approximation is similar to the high-$l_i$ approximation leading to Eq.
+\eqref{eq:lottery} in that we replace the $x_i$ with appropriate mean values.
+Replacing $x_i$ with $l_i$, as in the classic lottery model, is not valid for genotypes with low propagule densities $l_i\ll 1$ because $x_i=1$ in the few territories where $i$ propagules do land, and so $i$'s growth comes entirely from territories which deviate appreciably from the mean.
 
-However, in general the $l_i$ are not all large, and the $x_i$ cannot simply be replaced by their means in Eq.
-\eqref{eq:growthsumuncoupled}.
-Indeed, Eq.
+%Eq.
 \eqref{eq:lottery} is nonsensical if $l_i$ is sufficiently small: genotype $i$ can win at most $m_i$ territories, yet Eq.
 \eqref{eq:lottery} demands a fraction $c_i m_i/\sum_j c_j m_j$ of the unoccupied territories $U$, no matter how large $U$ is.
-The source of this pathological behavior when $l_i\ll 1$ is that $x_i=1$ in the few territories where $i$ propagules do land, and so $i$'s growth comes entirely from territories which deviate appreciably from the mean.
+The source of this pathological behavior when
 
-Our mean field approximation is similar to the high-$l_i$ approximation leading to Eq.
-\eqref{eq:lottery} in that we replace the $x_i$ with appropriate mean values.
 The key distinction is that territories with a single propagule from the focal genotype, whose behavior is critical at low densities, are handled separately.

### 2017-01-24 manuscript.tex
latest

@@
 In the classic lottery model, mature individuals (``adults'') each require their own territory, whereas newborn individuals (``propagules'') disperse to, and subsequently compete for, the territories made available by the death of adults.
-Territorial contest among propagules leaves a single victorious adult per territory, with the victor chosen at random from the propagules present (akin to a lottery), with probabilities weighted by a coefficient for each type representing competitive ability.
+Territorial contest among propagules leaves a single victorious adult per territory, with the victor chosen at random from the propagules present (akin to a lottery; \citealt{sale_77}), with probabilities weighted by a coefficient for each type representing competitive ability.
 This representation of competition is much simpler than having coefficients for the pairwise effects of types on each other (e.g.
@@
 In general, $m_i$ will increase with $n_i$, and will also depend on population density $N$.
-For example, if $b_i$ is the number of propagules produced per individual from genotype $i$, then $m_i=b_i(1-N/T)n_i$ would reflect the loss of propagules due to dispersal to occupied territories.
+For example, if $b_i$ is the number of propagules produced per individual from genotype $i$, then $m_i=b_i(1-N/T)n_i$ would reflect the loss of propagules due to dispersal to occupied territories \citep{levins_71}.
 We assume $m_i=b_i n_i$, meaning that all propagules land on unoccupied territories (a form of directed dispersal).
@@
 where $\overline{c}=\sum_j c_j m_j/M$ is the mean propagule competitive ability for a randomly selected propagule, $L=M/U$ is the total propagule density and $M=\sum_j m_j$ is the total number of propagules.
+
+Eq.
+\eqref{eq:lottery} is nonsensical at low densities, specifically if $l_i$ is sufficiently small: genotype $i$ can win at most $m_i$ territories, yet Eq.
+\eqref{eq:lottery} demands a fraction $c_i l_i/\sum_j c_j l_j$ of the unoccupied territories $U$, no matter how large $U$ is.
 
@@
 Eq.
-\eqref{eq:growthsumuncoupled} gives little intuition about the dynamics of density-dependent lottery competition, since it involves an expectation over the dispersal distributions $p_i$, which depend on how the $n_i$ change over time.
+\eqref{eq:growthsumuncoupled} involves an expectation over the time-dependent dispersal distributions $p_i$, and is thus too complicated to give intuition about the dynamics of density-dependent lottery competition.
 We now evaluate this expectation using a ``mean field'' approximation.
 
-This approximation is similar to the high-$l_i$ approximation leading to Eq.
-\eqref{eq:lottery} in that we replace the $x_i$ with appropriate mean values.
-Replacing $x_i$ with $l_i$, as in the classic lottery model, is not valid for genotypes with low propagule densities $l_i\ll 1$ because $x_i=1$ in the few territories where $i$ propagules do land, and so $i$'s growth comes entirely from territories which deviate appreciably from the mean.
-
-%Eq.
-\eqref{eq:lottery} is nonsensical if $l_i$ is sufficiently small: genotype $i$ can win at most $m_i$ territories, yet Eq.
-\eqref{eq:lottery} demands a fraction $c_i m_i/\sum_j c_j m_j$ of the unoccupied territories $U$, no matter how large $U$ is.
-The source of this pathological behavior when
-
-The key distinction is that territories with a single propagule from the focal genotype, whose behavior is critical at low densities, are handled separately.
+Our approximation is similar to the high-$l_i$ approximation behind the classic lottery model in that we replace the $x_i$ with appropriate mean values.
+However, we cannot simply replace $x_i$ with $l_i$ as in the classic lottery model.
+For a genotype with a low propagule density $l_i\ll 1$, we have $x_i=1$ in the few territories where its propagules land, and so its growth comes entirely from territories which deviate appreciably from $l_i$.
+In our more general approximation, territories with a single propagule from the focal genotype are handled separately.
 In place of the requirement of $l_i\gg 1$ for all $i$, our approximation only requires that there are no large discrepancies in competitive ability (specifically, that we do not have $c_i/c_j\gg 1$ for any two genotypes; further discussion in section ``Discussion'').
@@
 The third term, $A_ic_i/\overline{c}$, represents competitive victories when the $i$ genotype is abundant: $A_i\rightarrow 0$ if $l_i\ll 1$.
-The relative importance of these three terms varies with both the overall propagule density $L$ and the relative propagule frequencies $l_i/L$.
+The relative importance of these three terms varies with both the overall propagule density $L$ and the relative propagule frequencies $m_i/M$.
 If $l_i\gg 1$ for all genotypes, we recover the classic lottery model (only the $A_ic_i/\overline{c}$ term remains, and $A_i\rightarrow 1/L$).
@@
 On the other hand, $R_i c_i/\overline{c}$ is negligible for the high-frequency genotype, which depends instead on high density territorial victories (Fig.~\ref{fig:simcomp}d).
+Fig.
+3 also shows the breakdown of the classic lottery model at low densities (low $l_1$ and $l_2$).
 
@@
 Simulation points are almost invisible in (c) and (d) due to near exact agreement with Eq.
-\eqref{eq:master}.}
+\eqref{eq:master}.
+Dashed lines in (a) and (c) show the corresponding classic lottery model predictions.}
 \end{figure}
@@
 
-However, stable coexistence is possible between genotypes that are superior in different traits.
+Stable coexistence is possible between genotypes that are superior in different traits.
 To illustrate, suppose that $j$ is better at securing territories ($c_j>c_i$), that $i$ is better at producing propagules ($b_i>b_j$), and that $d_i=d_j$.
@@
 Intuitively, the mechanism for coexistence is that territorial contests are important in an $i$-dominated population (high $L$), ensuring that the $c$-specialist $j$ is not excluded, yet territorial contests are irrelevant in a $j$-dominated population (low $L$), ensuring that the $b$-specialist $i$ is not excluded.
-Fig.~\ref{fig:coex} shows an example of this coexistence between $b$ and $c$ specialists.
 

### 2017-01-24 simulations.py

@@ -55,6 +55,10 @@ def deltna(Ms,U,cs,i):
     
 def deltnp(Ms,U,cs):
     return array([deltnu(Ms,U,cs,j)+deltnr(Ms,U,cs,j)+deltna(Ms,U,cs,j) for j in range(len(Ms))])
+    
+def deltnpclassic(Ms,U,cs):
+    return array([float(U)*Ms[j]*cs[j]/sum([Ms[k]*cs[k] for k in range(len(Ms))]) for j in range(len(Ms))])
+    
 
 #Testing
 #=================================================================
@@ -68,7 +72,9 @@ Ms=map(int,array([10000,90000]))
 M=float(sum(Ms)); cs=array([1.5,1.]); G=len(Ms);
 maxl=2; Us=array([int(Ms[0]/x) for x in linspace(0.001,maxl,1000)]);
 exact=array([expCisim(Ms,U,cs) for U in Us]);
+totalclassic=array([deltnpclassic(Ms,U,cs) for U in Us]);
 
+#First genotype
 nocomp=array([deltnu(Ms,U,cs,0) for U in Us]);
 comp1=array([deltnr(Ms,U,cs,0) for U in Us]);
 comp2=array([deltna(Ms,U,cs,0) for U in Us]);
@@ -76,8 +82,9 @@ comp2=array([deltna(Ms,U,cs,0) for U in Us]);
 subplot(221)
 plot(float(Ms[0])/Us,(exact[:,0,0]+exact[:,1,0]+exact[:,2,0])/Ms[0],'k.',markersize=1.,label="Simulation")
 plot(float(Ms[0])/Us,(nocomp+comp1+comp2)/Ms[0],'k',label=r"$\Delta_+ n_1/m_1$")
-#ylim([0,1000])
+plot(float(Ms[0])/Us,totalclassic[:,0]/Ms[0],'k--',label=r"Classic lottery")
 xlim([0,maxl])
+ylim([0,1])
 gca().xaxis.set_label_coords(0.5, -0.09)
 gca().set_xticklabels(['0','','1','','2','','3'])
 xlabel(r"$l_1$",fontsize=14)
@@ -103,6 +110,7 @@ gca().set_yticklabels([])
 gca().annotate(r'$(b)$',xy=(0.9,0.9),xycoords='axes fraction',fontsize=12)
 legend(loc='upper center',prop={'size':10})
 
+#Second genotype
[…]

### 2017-01-25 manuscript.tex
latest

@@
 
-We assume that the individuals in a population each require their own territory to survive and reproduce (a site-occupancy model; Fig.~\ref{fig:lottery}).
+We assume that the reproductively mature individuals in a population (``adults'') each require their own territory to survive and reproduce (Fig.~\ref{fig:lottery}).
 All territories are identical, and the total number of territories is $T$.
 Time $t$ advances in discrete iterations, each representing the average time from birth to reproductive maturity.
-In iteration $t$, the number of reproductively mature individuals (henceforth called ``adults'') of the $i$'th genotype is $n_i(t)$, the total number of adults is $N(t)=\sum_i n_i(t)$, and the number of unoccupied territories is $U(t)=T-N(t)$.
+In iteration $t$, the number of adults of the $i$'th genotype is $n_i(t)$, the total number of adults is $N(t)=\sum_i n_i(t)$, and the number of unoccupied territories is $U(t)=T-N(t)$.
 
@@
 
-Each iteration, adults produce $m_i$ new offspring (henceforth called ``propagules'').
-We assume adults cannot be ousted from occupied territories, and only propagules landing on unoccupied territories are included in $m_i$.
-Propagules disperse at random, independently of each other, and without spatial restrictions; each propagule has an equal probability of landing on any of the $U$ unnocupied territories.
+Each iteration, adults produce new offspring (``propagules''), $m_i$ of which disperse to unoccupied territories.
+We assume that adults cannot be ousted from occupied territories, and so only propagules landing on unoccupied territories are included in $m_i$.
+Propagules disperse globally, at random, and independently of each other; all propagules have the same probability of landing on any of the $U$ unnocupied territories.
 Thus, there is no interaction between propagules (e.g.
 avoidance of territories crowded with propagules).
-Loss of propagules during dispersal is subsumed into $m_i$ ($m_i$ only counts propagules which go on to contest the territory they land in).
+Loss of propagules during dispersal is subsumed into $m_i$.
 In general, $m_i$ will increase with $n_i$, and will also depend on population density $N$.
-For example, if $b_i$ is the number of propagules produced per individual from genotype $i$, then $m_i=b_i(1-N/T)n_i$ would reflect the loss of propagules due to dispersal to occupied territories \citep{levins_71}.
-We assume $m_i=b_i n_i$, meaning that all propagules land on unoccupied territories (a form of directed dispersal).
+For example, if $b_i$ is the number of successfully dispersing propagules produced per genotype $i$ adult, then the loss of propagules due to dispersal to occupied territories implies $m_i=b_i(1-N/T)n_i$, akin to Levins' competition-colonization model \citep{levins_71,tilman_94}.
+Here we assume $m_i=b_i n_i$, where $b_i$ is a constant, meaning that all propagules land on unoccupied territories (a form of directed dispersal).
 This choice simplifies the mathematics without seriously restricting the generality of our analysis, since the results presented here are not sensitive to the specific functional form of $m_i$.
+Note that due to our assumption of global dispersal, the parameter $b_i$ can be thought of as a measure of colonization ability which combines fecundity and dispersal ability \citep{levins_71,tilman_94,bolker_99}.
 
@@
 where $\overline{c}=\sum_j c_j m_j/M$ is the mean propagule competitive ability for a randomly selected propagule, $L=M/U$ is the total propagule density and $M=\sum_j m_j$ is the total number of propagules.
-
 Eq.
@@
 \subsection*{Mean Field Approximation}
-
-%The expected Eq. \eqref{eq:lottery}
-%Thus, the numbers of territorial victories $\Delta_+ n_1,\Delta_+ n_2,\ldots$ follow a multinomial distribution with $U$ trials and event probabilities $\frac{c_1 l_1}{\sum_j c_j l_j},\frac{c_2 l_2}{\sum_j c_j l_j},\ldots$.
 
@@
 
-%with the exception of large discrepancies in competitive ability since it is extremely unlikely that a large discrepancy competitive would arise in the absence of migration.
-
 \subsection*{Invasion of rare genotypes and coexistence}\label{sec:invas}
@@
 Stable coexistence is possible between genotypes that are superior in different traits.
-To illustrate, suppose that $j$ is better at securing territories ($c_j>c_i$), that $i$ is better at producing propagules ($b_i>b_j$), and that $d_i=d_j$.
+Suppose that $j$ is better at securing territories ($c_j>c_i$), that $i$ is better at producing propagules ($b_i>b_j$), and that $d_i=d_j$.
 Coexistence occurs if $j$ will invade an $i$-dominated population, but $i$ will also invade a $j$-dominated population (``mutual invasion'').
-It is not hard to show that this is possible, since if $b_i$ is so large that $L\gg 1$ when $i$ is dominant, and $b_j$ is so small that $L\ll 1$ when $j$ is dominant, then, combining Eqs.
+If $b_i$ is so large that $L\gg 1$ when $i$ is dominant, and $b_j$ is so small that $L\ll 1$ when $j$ is dominant, then, combining Eqs.
 \eqref{eq:equil} and \eqref{eq:invad}, we find that $i$ invades $j$ because $b_i>b_j$, while $j$ invades $i$ provided that
@@
 Thus, coexistence occurs if $c_j/c_i$ is large enough.
-Intuitively, the mechanism for coexistence is that territorial contests are important in an $i$-dominated population (high $L$), ensuring that the $c$-specialist $j$ is not excluded, yet territorial contests are irrelevant in a $j$-dominated population (low $L$), ensuring that the $b$-specialist $i$ is not excluded.
-
-A similar argument applies for coexistence between high-$c$ and low-$d$ specialists; again coexistence occurs because the importance of territorial contests declines along with propagule density $L$ as the $c$-specialist increases in frequency.
-Mutual invasibility is not possible between $b$- and $d$-specialists (although genotypes with exactly $c_i=c_j$ and $b_i/d_i=b_j/d_j$ do not exclude each other --- this follows from the fact that all propagules have the same probability of success when $c_i=c_j$ i.e.
-$A_i+R_i=A_j+R_j$).
+This is a version of the classic competition-colonization trade-off \citep{tilman_94,levins_71}: the competitor ($c$-specialist) leaves many territories unoccupied (low $L$) due to its poor colonization ability (low $b$), which the colonizer ($b$-specialist) can then exploit.
+A similar argument applies for coexistence between high-$c$ and low-$d$ specialists; a ``competition-longevity'' trade-off \citep{tilman_94}.
+Mutual invasibility is not possible between $b$- and $d$-specialists.
 
@@
 
-Unlike Grime's classic ternary plot \citep{grime_1974}, which represents anti-correlations between traits relevant for success in different environmental archetypes, our realization of Grime's triangle (Fig.~4) refers instead to the direction of adaptive trait evolution under different regimes of stress and disturbance.
+W
+
+%Thus, the numbers of territorial victories $\Delta_+ n_1,\Delta_+ n_2,\ldots$ follow a multinomial distribution with $U$ trials and event probabilities $\frac{c_1 l_1}{\sum_j c_j l_j},\frac{c_2 l_2}{\sum_j c_j l_j},\ldots$.
+
+Unlike Grime's classic ternary plot \citep{grime_1974}, which represents anti-correlations between traits in different environmental archetypes, our realization of Grime's triangle (Fig.~4) refers instead to the direction of adaptive trait evolution under different regimes of stress and disturbance.
 As discussed in section ``Primary strategies and Grime's triangle'', over time our predicted trait evolution should lead to trait values consistent with Grime's scheme.

### 2017-01-26 manuscript.tex
latest

@@
 From the law of large numbers, the composition of propagules in each territory will then not deviate appreciably from the mean composition $l_1,l_2,\ldots,l_G$ ($G$ is the number of genotypes present), and so the probability that genotype $i$ wins any particular unoccupied territory is approximately $c_i l_i/\sum_j c_j l_j$.
-Thus, genotype $i$ is expected to win $c_i l_i/\sum_j c_j l_j$ of the $U$ available territories, and deviations from this expected outcome are small (since $T$ is large by assumption).
-Consequently, the increase in $n_i$ due to territorial acquisition in the classic lottery model is given by
+Let $\Delta_+ n_i$ denote the number of territories won by genotype $i$ in the current iteration.
+Then $\Delta_+ n_1,\Delta_+ n_2,\ldots,\Delta_+ n_G$ follow a multinomial distribution with $U$ trials and success probabilities $\frac{c_1 l_1}{\sum_j c_j l_j},\frac{c_2 l_2}{\sum_j c_j l_j},\ldots,\frac{c_G l_G}{\sum_j c_j l_j}$, respectively.
+Genotype $i$ is expected to win $c_i l_i/\sum_j c_j l_j$ of the $U$ available territories, and deviations from this expected outcome are small (since $T$ is large by assumption), giving
 \begin{equation}
@@
 \eqref{eq:lottery} demands a fraction $c_i l_i/\sum_j c_j l_j$ of the unoccupied territories $U$, no matter how large $U$ is.
+
+The classic lottery model is closely related to the Wright-Fisher model of genetic drift.
+In the Wright-Fisher model, genotype abundances in each generation follow a multinomial distribution with success probabilities $w_i n_i/\sum_j w_j n_j$, where $w$ is relative fitness and the $n_i$ are  genotype abundances in the preceding generation.
+Population size $N$ remains constant.
+This is mathematically equivalent to the classic lottery model with non-overlapping generations ($d_i=1$ for all $i$) and $w_i=b_i c_i$.
+The classic lottery model is more general in that it allows for overlapping generations, which crucially introduces ``storage'' effects \citep{svardal_2015}.
+Additionally, the relative fitnesses $w_i=b_i c_i$ in the classic lottery model have a mechanistic basis in terms of fecundity and territorial contests.
+However, the drift in genotype frequencies that are the focus of the Wright-Fisher model are omitted.
 
@@
 \section*{Discussion}
-
-W
-
-%Thus, the numbers of territorial victories $\Delta_+ n_1,\Delta_+ n_2,\ldots$ follow a multinomial distribution with $U$ trials and event probabilities $\frac{c_1 l_1}{\sum_j c_j l_j},\frac{c_2 l_2}{\sum_j c_j l_j},\ldots$.
 

### 2017-01-26 manuscript.tex
latest

@@
 
-Nevertheless, there are strong indications there are broader principles governing the operation of selection.
+Nevertheless, there are strong indications there exist broader principles governing the operation of selection.
 In many groups of organisms, including corals \citep{darling_2012}, insects \citep{southwood_1977}, fishes \citep{winemiller_1992}, zooplankton \citep{allan_76} and plants \citep{grime_1988,westoby_1998}, different species can be divided into a small number of distinct trait clusters corresponding to fundamentally distinct ``primary strategies'' \citep{winemiller_2015}.
@@
 
-As we will see, a generalized version of the classic lottery model of \cite{chesson_1981} is well suited for our purposes.
-In the classic lottery model, mature individuals (``adults'') each require their own territory, whereas newborn individuals (``propagules'') disperse to, and subsequently compete for, the territories made available by the death of adults.
-Territorial contest among propagules leaves a single victorious adult per territory, with the victor chosen at random from the propagules present (akin to a lottery; \citealt{sale_77}), with probabilities weighted by a coefficient for each type representing competitive ability.
+As we will see, a generalized version of the classic lottery model of \cite{chesson_1981} is a convenient starting point for ecologically-grounded models of selection in evolutionary genetics.
+In the classic lottery model, mature individuals (``adults'') each require their own territory, whereas newborn individuals (``propagules'') disperse to, and subsequently compete for, territories made available by the death of adults.
+Territorial contest among propagules leaves a single victorious adult per territory, the victor chosen at random from the propagules present (akin to a lottery; \citealt{sale_77}), with probabilities weighted by a coefficient for each type representing competitive ability.
 This representation of competition is much simpler than having coefficients for the pairwise effects of types on each other (e.g.
 the $\alpha$ coefficients in the generalized Lotka-Volterra equations), or than modeling resource consumption explicitly \citep{tilman_1982}.
-The classic lottery model also has similarities with the Wright-Fisher model in population-genetics (section BLAH).
+The classic lottery model is also closely connected to one of the central models of population genetics, the Wright-Fisher model of genetic drift.
 
-However, the classic lottery model breaks down at low densities (that is, if there are only a few propagules dispersing to each territory; section ``Mean field approximation'').
+However, the classic lottery model breaks down at low densities (that is, if there are only a few propagules dispersing to each territory; see ``Model'').
 This was not a limitation in the lottery model's original application to reef fishes, where a huge number of larvae from each species compete to secure territories each generation \citep{chesson_1981}, but is a critical limitation for studying density-dependent selection.
@@
 
-%(each generation is a sample of constant size, with replacement, from the previous one)
-
 In the section ``Model'', we introduce the basic assumptions of our generalized lottery model.
 Analytical expressions for the change in genotype abundances over time are introduced in section ``Mean field approximation'', with mathematical details relegated to the Appendices.
-In the following two sections, we then discuss the behavior of rare mutants and our treatment of Grime's triangle.
+In the following two sections, we then discuss the behavior of rare mutants (invasion and coexistence) and our formalization of Grime's triangle.
 
@@
 Although $T$ is finite in our model, we assume that $T$ and the $n_i$ are large enough that $x_i$ is Poisson-distributed to a good approximation (details in Appendix A).
-Note that the large $n_i$, large $T$ approximation places no restrictions on our densities $n_i/T$, but it does preclude consideration of demographic stochasticity when $n_i$ itself is very small (this will be discussed further in Section ``Invasion of rare genotypes and coexistence'' BLAH).
+Note that the large $n_i$, large $T$ approximation places no restrictions on our densities $n_i/T$, but it does preclude consideration of demographic stochasticity when $n_i$ itself is very small.
 
@@
 From the law of large numbers, the composition of propagules in each territory will then not deviate appreciably from the mean composition $l_1,l_2,\ldots,l_G$ ($G$ is the number of genotypes present), and so the probability that genotype $i$ wins any particular unoccupied territory is approximately $c_i l_i/\sum_j c_j l_j$.
-Let $\Delta_+ n_i$ denote the number of territories won by genotype $i$ in the current iteration.
+Let $\Delta_+ n_i$ denote the number of territories won by genotype $i$.
 Then $\Delta_+ n_1,\Delta_+ n_2,\ldots,\Delta_+ n_G$ follow a multinomial distribution with $U$ trials and success probabilities $\frac{c_1 l_1}{\sum_j c_j l_j},\frac{c_2 l_2}{\sum_j c_j l_j},\ldots,\frac{c_G l_G}{\sum_j c_j l_j}$, respectively.
@@
 where $\overline{c}=\sum_j c_j m_j/M$ is the mean propagule competitive ability for a randomly selected propagule, $L=M/U$ is the total propagule density and $M=\sum_j m_j$ is the total number of propagules.
-Eq.
-\eqref{eq:lottery} is nonsensical at low densities, specifically if $l_i$ is sufficiently small: genotype $i$ can win at most $m_i$ territories, yet Eq.
-\eqref{eq:lottery} demands a fraction $c_i l_i/\sum_j c_j l_j$ of the unoccupied territories $U$, no matter how large $U$ is.
 
-The classic lottery model is closely related to the Wright-Fisher model of genetic drift.
-In the Wright-Fisher model, genotype abundances in each generation follow a multinomial distribution with success probabilities $w_i n_i/\sum_j w_j n_j$, where $w$ is relative fitness and the $n_i$ are  genotype abundances in the preceding generation.
+There is a close connection between the classic lottery model and the Wright-Fisher model of genetic drift \citep{svardal_2015}.
+In the Wright-Fisher model, genotype abundances are sampled each generation from a multinomial distribution with success probabilities $w_i n_i/\sum_j w_j n_j$, where $w$ is relative fitness and the $n_i$ are  genotype abundances in the preceding generation.
 Population size $N$ remains constant.
 This is mathematically equivalent to the classic lottery model with non-overlapping generations ($d_i=1$ for all $i$) and $w_i=b_i c_i$.
-The classic lottery model is more general in that it allows for overlapping generations, which crucially introduces ``storage'' effects \citep{svardal_2015}.
-Additionally, the relative fitnesses $w_i=b_i c_i$ in the classic lottery model have a mechanistic basis in terms of fecundity and territorial contests.
-However, the drift in genotype frequencies that are the focus of the Wright-Fisher model are omitted.
 
-In our extension of the classic lottery model, we do not restrict ourselves to high propagule densities, and deviations from the mean propagule composition $l_1,l_2,\ldots,l_G$ will be important.
+ Unlike the abstract population-genetic relative fitnesses $w_i$, the lottery model $b_i$ and $c_i$ have a mechanistic interpretation in terms of fecundity and territorial contests, while overlapping generations allows for genotypic differences in longevity.
+Note, however, that the classic lottery model, and the present work, ignore the stochastic drift in type frequencies (large $T$ approximation), which is the focus of the Wright-Fisher model.
+
+In our extension of the classic lottery model, we do not restrict ourselves to high propagule densities.
+Eq.
+\eqref{eq:lottery} is nonsensical at low densities (if $l_i$ is small): genotype $i$ can win at most $m_i$ territories, yet Eq.
+\eqref{eq:lottery} demands $c_i l_i/\sum_j c_j l_j$ of the $U$ unoccupied territories, for any value of $U$.
+Thus, at low densities, deviations from the mean propagule composition $l_1,l_2,\ldots,l_G$ are important.
+
 Since dispersal is Poisson, we expect that a fraction $p_1(x_1)\ldots p_G(x_G)$ of the $U$ unoccupied territories will have the propagule composition $x_1,\ldots,x_G$.
 Genotype $i$ is expected to win $c_i x_i/\sum_j c_j x_j$ of these.
-Ignoring fluctuations about these two expectations (because $T$ is large), genotype $i$'s territorial acquisition is given by
+Ignoring fluctuations about these two expectations (large $T$ approximation), genotype $i$'s territorial acquisition is given by
 \begin{equation}
@@
 
-In addition to propagule birth and competition, occupied territories become unoccupied due to mortality.
-For the majority of this manuscript we assume that mortality only occurs in adults (setting aside the deaths implicit in territorial contest), and at a constant, genotype-specific per-capita rate $d_i$, so that the overall change in genotype abundances is
+For the majority of this manuscript we assume that mortality only occurs in adults (setting aside the juvenile deaths implicit in territorial contest), and at a constant, genotype-specific per-capita rate $d_i$, so that the overall change in genotype abundances is
 \begin{equation}
@@
 
-Note that the competitive ability coefficients $c_i$ represent a strictly relative aspect of fitness in the sense that they only influence population size $N$ indirectly by changing genotype frequencies.
+Note that the competitive ability coefficients $c_i$ represent a strictly relative aspect of fitness in the sense that they can only influence population size $N$ indirectly by changing genotype frequencies.
 This can be seen by summing Eq.
@@
 We now discuss which changes in the traits $b, c$ and $d$ will be particularly favored under different environmental conditions.
-Of specific interest are Grime's ``disturbance'', ``stress'' and ``ideal'' environmental archetypes.
-To proceed, we need to map these verbal archetypes to quantitative parameter regimes in our model.
+Of specific interest are Grime's ``disturbance'', ``stress'' and ``ideal'' environments.
+To proceed, we need to map these verbally-defined environments to quantitative parameter regimes in our model.
 
@@
 
+\section*{Acknowledgments}
+
+We thank Peter Chesson and Joachim Hermisson for many constructive comments on this manuscript.
+
 \bibliographystyle{amnatnat}
@@
 Thus, the $x_i$ in different territories are not independent random variables.
-However, for sufficiently large $T$, holding $n_i/T$ fixed, the Poisson limit theorem implies that this multinomial distribution for the $x_i$ accross territories is closely approximated by a product of independent Poisson distributions for each territory, each with rate parameter $l_i$.
-Here we have used the fact that large $T$ implies large $U$ except in the biologically uninteresting case that there is vanishing population turnover $d_i \sim 1/T$.
+However, for sufficiently large $T$, holding $n_i/T$ fixed, the Poisson limit theorem implies that this multinomial distribution for the $x_i$ accross territories is closely approximated by a product of independent Poisson distributions for each territory, each with rate parameter $l_i$ (large $T$ implies large $U$ except in the biologically uninteresting case that there is vanishing population turnover $d_i \sim 1/T$).
 
-Under the Poisson approximation, the total number of genotype $i$ propagules $\sum x_i$ (summed over unoccupied territories) will deviate about its mean value $m_i$.
-Since the coefficient of variation of $\sum x_i$ is proportional to $1/\sqrt{m_i}$, these fluctuations are negligible unless $m_i$ is very small (say of order $10^2$ or less).
-These fluctuations in $m_i$ could be regarded as a feature rather than a flaw since having $m_i$ be exactly constant per generation (for given $b_i$ and $n_i$) is biologically unrealistic.
-In the canonical model of genetic drift, the Wright-Fisher model, the number of offspring per genotype fluctuates from generation to generation and is approximately Poisson distributed.
-Nevertheless, for simplicity and ease of comparison with the classic lottery model, we ignore fluctuations in $m_i$ and only account for Poisson fluctuations in the number of propagules landing in each territory.
+Alternatively, we could have assumed a Poisson distribution for the $x_i$ as our model of dispersal from the outset.
+The total number of genotype $i$ propagules $\sum x_i$ (summed over unoccupied territories) would then no longer be a constant $m_i$, but would fluctuate between generations for a given mean $m_i$, which is more biologically realistic.
+Nevertheless, for simplicity and ease of comparison with the classic lottery model, we ignore the possibility of fluctuations in $m_i$ and focus instead on Poisson fluctuations propagule composition in each territory.
 
@@
 
-%The logistic equation for a single genotype or species is given by
-%\begin{equation}
-%\frac{dN}{dt}=r\left(1-\frac{N}{K}\right)N \label{eq:logistic}
-%\end{equation}
-%where $N$ is the total population abundance or biomass, $r$ is the growth rate at  low densities, and the ``carrying capacity" $K$ is the value of $N$ where growth ceases.
-%
-%a simple, ubiquitous model of density-dependent population growth
-%declines linearly with density,
-%
-%Eq.~\eqref{eq:logistic} is often interpreted as a linear approximation of a more complicated growth equation
-%\begin{equation}
-%\frac{dN}{dt}=r(N)N,
-%\end{equation}
-%where the growth rate $r(N)$ is some unknown, possibly complicated, function of $N$.
-
 %\subsection{Logistic growth and classic nesting sites model}

### 2017-01-27 manuscript.tex
latest

@@
 
- Unlike the abstract population-genetic relative fitnesses $w_i$, the lottery model $b_i$ and $c_i$ have a mechanistic interpretation in terms of fecundity and territorial contests, while overlapping generations allows for genotypic differences in longevity.
-Note, however, that the classic lottery model, and the present work, ignore the stochastic drift in type frequencies (large $T$ approximation), which is the focus of the Wright-Fisher model.
+The main advantage of the classic lottery model for our purposes here is that the abstract population-genetic relative fitnesses $w_i$ can be replaced with more ecologically-grounded fecundity, competitive ability and mortality parameters $b_i$, $c_i$ and $d_i$.
+In particular, differences in individual longevity are only possible because generations are overlapping.
+However, note that the classic lottery model, and the present work, ignore the stochastic drift in type frequencies (large $T$ approximation) which is the focus of the Wright-Fisher model.
 

### 2017-01-27 manuscript.tex
latest

@@
 This choice simplifies the mathematics without seriously restricting the generality of our analysis, since the results presented here are not sensitive to the specific functional form of $m_i$.
-Note that due to our assumption of global dispersal, the parameter $b_i$ can be thought of as a measure of colonization ability which combines fecundity and dispersal ability \citep{levins_71,tilman_94,bolker_99}.
+Note that due to our assumption of global dispersal, the parameter $b_i$ can be thought of as a measure of ``colonization ability'', which combines fecundity and dispersal ability \citep{levins_71,tilman_94,bolker_99}.
 
@@
 We now discuss which changes in the traits $b, c$ and $d$ will be particularly favored under different environmental conditions.
-Of specific interest are Grime's ``disturbance'', ``stress'' and ``ideal'' environments.
+Of specific interest are Grime's ``disturbed'', ``stressful'' and ``ideal'' environments.
 To proceed, we need to map these verbally-defined environments to quantitative parameter regimes in our model.
 
-The ideal environmental archetype is characterized by the near-absence of stress and disturbance.
+The ideal environment is characterized by the near-absence of stress and disturbance.
 Consequently, $d_i\ll 1$, whereas $b_i$ is potentially much larger than $1$.
@@
 \eqref{eq:deltN}, the equilibrium value of $L$ only depends on the ratio of birth and death rates.
-For one genotype, $L/(1-e^{-L})=b_i/d_i$, and so the propagule density is high $L\approx b_i/d_i\gg 1$.
-Moreover, since $L=b_i\frac{N}{N-T}=b_i\frac{1}{1-T/N}$ by definition, population density is also high $N/T\approx 1$.
-Thus, almost every unoccupied territory will be heavily contested.
+For one genotype, $L/(1-e^{-L})=b_i/d_i$, and so the propagule density is high $L\approx b_i/d_i\gg 1$, and every unoccupied territory will be heavily contested.
+The population density is also high $N/T\approx 1$ (since $L=b_iN/(N-T)=b_i/(1-T/N)$).
 
-The disturbance archetype is characterized by unavoidably high extrinsic mortality caused by physical destruction.
-Disturbances do not only affect adults as in Eq.
-\eqref{eq:delttot}, but also juveniles in the process of territorial contest.
-These juvenile deaths can be represented as a fractional reduction in the number of territories secured.
-To illustrate, we assume that the disturbance is equally damaging to adults and juveniles, so that only $(1-d_i)\Delta_+ n_i$ rather than $\Delta_+ n_i$ territories are secured by genotype $i$ each iteration.
-Then, the disturbance archetype is characterized by $d_i$ being close to $1$ for all genotypes (almost all adults and juveniles are killed each iteration).
+Disturbed environments are characterized by unavoidably high extrinsic mortality caused by physical destruction.
+Environmental variability, and disturbances in particular, can be modeled as stochastic fluctuations in $b$, $c$ and $d$ \citep{chesson_1981}.
+For simplicity, we do not pursue this more complicated stochastic approach.
+Instead, we represent disturbance by high constant mortality rates $d_i$, and extend this mortality to juveniles in the process of territorial contest (since disturbances do not only affect adults as in Eq.
+\eqref{eq:delttot}).
+We assume that disturbances are equally damaging to adults and juveniles, so that only $(1-d_i)\Delta_+ n_i$ rather than $\Delta_+ n_i$ territories are secured by genotype $i$ each iteration.
+[More justification?] Disturbed environments then correspond to $d_i$ being close to $1$ for all genotypes (almost all adults and juveniles are killed each iteration).
 From Eq.
@@
 
-The stress archetype is more ambiguous, and has been the subject of an extensive debate in the plant ecology literature (the ``Grime-Tilman'' debate; \citealt{aerts_1999} and references therein).
-Stressful environments severely restrict growth and reproduction, so that $b\ll 1$ \cite{grime_1974,grime_1977}.
+Stressful environments are more ambiguous, and have been the subject of an extensive debate in the plant ecology literature (the ``Grime-Tilman'' debate; \citealt{aerts_1999} and references therein).
+Severe stress inhibits growth and reproduction, so that $b\ll 1$ \cite{grime_1974,grime_1977}.
 Mutations which appreciably improve $b$ will be either non-existent or extremely unlikely, so $b$ is constrained to remain low.
@@
 The alternative view is that, while stressful environments imply lower $b$ and support a lower number of individuals per unit area compared what is attainable in ideal environments, stressed populations are actually at high densities relative to the environmental carrying capacity, and are highly competitive \citep{taylor_1990}.
-In the particular case that stress is caused by scarcity of consumable resources, it is argued that we should expect intense resource competition (for empirical support, see \citealt{davis_1998}).
+In the particular case that stress is caused by scarcity of consumable resources, we might expect intense resource competition (for empirical support, see \citealt{davis_1998}).
 Thus, $b$ may actually appreciably exceed $d$ under stressful conditions, even though the absolute value of $b$ is small.
 
-The mapping of environmental achetypes to our model parameters is summarized in the first two rows of Fig.~\ref{fig:table}.
-Also shown is the approximate dependence of $\Delta_+ n_i$ on $b_i$ and $c_i$ for each archetype (third row).
+The mapping of different environments to our model parameters is summarized in the first two rows of Fig.~\ref{fig:table}.
+Also shown is the approximate dependence of $\Delta_+ n_i$ on $b_i$ and $c_i$ for each environment (third row).
 These can be used infer the expected direction of evolution for the traits $b$, $c$ and $d$ (fourth row) as follows.
 
-As noted in the previous section, if beneficial mutations establish (i.e survive the low-abundance stochastic regime), they will proceed to grow deterministically according to Eq.
+As noted in the previous section, if beneficial mutations survive the low-abundance stochastic regime, they proceed to grow deterministically according to Eq.
 \eqref{eq:master}.
-The probability of establishment increases with the mutant fitness advantage, and is therefore typically on the order of one percent, whereas the fixation of neutral mutations is exceedingly unlikely (probability of order $1/N$).
+The probability of surviving to deterministic abundances increases with the mutant fitness advantage, and is therefore typically on the order of one percent, whereas the fixation of neutral mutations is exceedingly unlikely (probability of order $1/N$).
 Consequently, the direction of evolutionary change is determined by which trait changes are both available, and confer an appreciable benefit, where availability is subject to constraints imposed by the environment.
 
-For example, in Grime's version of the stress archetype, $L$ is low, so competition is not important, and only mutants with greater $b$ or lower $d$ will have an appreciably greater $\Delta n_i$.
+For example, in Grime's interpretation of stressful environments, $L$ is low, so competition is not important, and only mutants with greater $b$ or lower $d$ will have an appreciably greater $\Delta n_i$.
 Mutations in $c$ are effectively neutral, and will rarely fix.
-However, by definition of the stress archetype, $b$ is constrained to be small.
-Thus, while some rare mutations may produce small improvements in $b$, it is much more likely that mutations will arise that lower $d$, making this the expected direction of evolutionary change for Grime's stress archetype.
+However, $b$ is constrained to be small.
+Thus, while some rare mutations may produce small improvements in $b$, it is much more likely that mutations lower $d$, making this the expected direction of evolutionary change.
 
@@
 \begin{tabular}{l*{4}{c}}
-  & Ideal & Disturbance* & Stress (G) & Stress (T) \\ \hline
+  & Ideal & Disturbance* & Stress (G) & Stress (HD) \\ \hline
   Constraints & $d \ll 1$ & $d \approx 1$ & $b \ll 1$ & $b \ll 1$ \\
@@
 \end{tabular}
-\caption{\label{fig:table} The realization of Grime's environmental archetypes in our model, as well as the low-$T$ variant of the stress archetype.
-Shown are the mapping to our parameters of each archetype, the approximate dependence of $\Delta_+ n_i$ on $b_i$ and $c_i$, as well as the corresponding expected evolutionary changes in $b_i$, $c_i$ and $d_i$.
-*Mortality affects both adults and juveniles in the disturbance archetype, with $\Delta_+ n_i$ replaced by $(1-d_i)\Delta_+ n_i$ in Eq.
+\caption{\label{fig:table} The realization of Grime's three environmental extremes in our model, as well as the high-density variant of the stressful environment.
+Shown are the mapping of each environment to our parameters, the approximate dependence of $\Delta_+ n_i$ on $b_i$ and $c_i$, as well as the corresponding expected evolutionary changes in $b_i$, $c_i$ and $d_i$.
+*Mortality affects both adults and juveniles in the disturbed environment, with $\Delta_+ n_i$ replaced by $(1-d_i)\Delta_+ n_i$ in Eq.
 \eqref{eq:delttot}.}
@@
 
-Following Grime's original argument for a triangular scheme \citep{grime_1977}, Fig.~\ref{fig:axes} represents each environmental archetype schematically as a vertex on a triangular space defined by perpendicular stress and disturbance axes.
-The ideal archetype lies at the origin (no stress or disturbance), while the stress and disturbance archetypes lie at the limits of survival on their respective axes.
+Following Grime's original argument for a triangular scheme \citep{grime_1977}, Fig.~\ref{fig:axes} represents each environmental extreme schematically as a vertex on a triangular space defined by perpendicular stress and disturbance axes.
+The ideal environment lies at the origin (no stress or disturbance), while the stressful and disturbed environments lie at the limits of survival on their respective axes.
 The hypotenuse connecting the stress and disturbance endpoints represents the limits of survival in the presence of a combination of stress and disturbance.
@@
 The low-$T$ interpretation of stress is also shown.
-The vertices of the triangles correspond to environmental archetypes.
+The vertices of the triangles correspond to different environmental extremes.
 Selection favors different traits at each vertex, leading to different trait clusters.}
@@
 In our comparison we will stick to fishes, corals and plants, for which three-way primary strategy schemes are well developed \citep{grime_1977,winemiller_1992,darling_2012}.
-The connection of our model to fish strategies is necessarily more tentative, given that fishes are motile and not all territorial, and the starting assumption of our model is site-occupancy.
+The connection of our model to fish strategies is necessarily more tentative, given that fishes are motile and not all territorial.
 
 In disturbed environments, we predict evolution for higher $b$ and lower $d$, but not higher $c$.
-Higher $b$ means higher fecundity, but not necessarily mass propagule production: $b$ represents only those propagules which sucessfully develop into juveniles in unoccupied territories.
-This is broadly consistent with the ruderal primary strategy.
-Plant ruderals devote a large proportion of their productivity to seed production \cite{grime_1977}, whereas the analogous ``opportunistic'' strategists in fishes have large intrinsic growth rates \citep{winemiller_1992}.
-In corals, a distinguishing feature of the ruderal cluster is brood spawning (rather than broadcast spawning).
-This corresponds to higher parental investment and lower overall propagule production, but potentially also higher $b$ at low densities, since broadcast spawners are vulnerable to a powerful Allee effect at the egg fertilization stage \citep{knowlton_2001}.
+Higher $b$ means higher fecundity and/or dispersal ability (see ``Model'').
+This is consistent with a ruderal strategy.
+Plant ruderals devote a large proportion of their productivity to seed production \citep{grime_1977}, whereas the analogous ``opportunist'' fishes have large intrinsic growth rates \citep{winemiller_1992}.
+In corals, ruderals are distinguished by brood spawning (rather than broadcast spawning).
+This corresponds to higher parental investment and lower overall propagule production --- counter-intuitively, a stereotypical ``$K$-selected'', high-density trait \cite{pianka_1970}.
+However, since broadcast spawners are vulnerable to an Allee effect at the egg fertilization stage \citep{knowlton_2001}, brood spawning could actually be a way to ensure high $b$ at low densities \citep{darling_2012}.
 Lower $d$ could be achieved by improved individual resistance to physical destruction, but it is hard to reduce mortality in the face of severe disturbances.
-Given this constraint, shortening the time to reproductive maturity (the iteration time in our model) is an effective way of reducing the chance of death per iteration $d$.
+Alternatively, shortening the time to reproductive maturity (the iteration time in our model) is an effective way of reducing the chance of death per iteration, $d$, for a given frequency of disturbance.
 An exceptionally short life cycle is probably the most defining characteristic of ruderals \citep{grime_1977,winemiller_1992,darling_2012}.
+Note that if evolution manages to appreciably reduce $d$ for a given disturbance intensity, then the population no longer lies at the extreme disturbance vertex of Grime's triangle.
+Thus, ruderals are characterized by both high $b$ and high $d$, but there is a constant pressure for a shorter life cycle, extending the limits of disturbance that can be tolerated.
 
-In stressful environments, we predict evolution for lower $d$, and also  for higher $c$ in the low-$T$ interpretation of the stress archetype.
-Lowering $d$ is obviously essential when $b\ll 1$, and stress tolerant plants and corals have long life spans, allowing for long intervals between successful recruitments (and episodic broadcast spawning in corals).
+In stressful environments, we predict evolution for lower $d$, and also  for higher $c$ in the high-density interpretation of stressful environments.
+Low $d$ is essential when $b\ll 1$, and stress tolerant plants and corals have long life spans, allowing for long intervals between successful recruitments (and episodic broadcast spawning in corals).
 For fishes, the ``equilibrium'' strategy is the analogue of Grime's stress tolerator.
-This strategy is associated with resource limitation, and is also characterized by long life span, as well as high parental investment in tiny broods.
-This may reflect a high-$c$ strategy in the face of intense competition for severely limited resources (the low-$T$ interpretation).
+This strategy is associated with consumable resource limitation, and is also characterized by long life span, as well as high parental investment in tiny broods.
+This may reflect a high-$c$ strategy in the face of intense competition for severely limited resources (the high-density interpretation).
 
 In ideal environments, we predict evolution for higher $b$ and $c$, but not lower $d$.
-In plants and corals, a key mechanism for winning territorial contests is rapidly outgrowing and ``shading out'' competitors; not surprisingly, rapid individual growth is a defining feature of the competitor trait cluster \citep{grime_1977,darling_2012}.
+In plants and corals, a key mechanism for winning territorial contests is rapidly outgrowing and ``shading out'' competitors; empirically, rapid individual growth is a defining feature of the competitor trait cluster \citep{grime_1977,darling_2012}.
 Evolution for higher $b$ under high-density, competitive conditions may seem counter-intuitive.
@@
 However, for fishes, the analogous ``periodic'' strategy is characterized by enormous spawn sizes as well as rapid development \citep{winemiller_1992,winemiller_2015}, suggesting a strategy of ensuring that many propagules actually end up contesting areas favorable for development (higher $b$).
-The evolution of $b$ in ideal environments will be discussed further in the Discussion.
+The evolution of $b$ in ideal environments will be discussed further below.
 
@@
 
-Unlike Grime's classic ternary plot \citep{grime_1974}, which represents anti-correlations between traits in different environmental archetypes, our realization of Grime's triangle (Fig.~4) refers instead to the direction of adaptive trait evolution under different regimes of stress and disturbance.
-As discussed in section ``Primary strategies and Grime's triangle'', over time our predicted trait evolution should lead to trait values consistent with Grime's scheme.
-In making these predictions, we have made no reference to any kind of trade-offs or pleitropy, even though trade-offs are often invoked to explain primary strategy schemes \citep{macarthur_1967,winemiller_1992,aerts_1999}.
-Thus, while trade-offs may amplify specialization, they are not necessary for it.
-As an example of a trade-off, corals which rapidly out-shade neighbors have a tall, branched morphology which is vulnerable to disturbances, and so, all else being equal, ideal environment $c$-strategists will suffer higher mortality from disturbances.
+Our realization of Grime's triangle (Fig.~4) refers to the direction of adaptive trait evolution under different regimes of stress and disturbance.
+As discussed in ``Primary strategies and Grime's triangle'', over time our predicted trait evolution should produce traits consistent with Grime's scheme.
+In making these predictions, we have not assumed any kind of trade-offs or pleitropy between $b$, $c$ and $d$, only constraints imposed by the environment on the order of magnitude of $b$ and $d$.
+Thus, while trade-offs may amplify specialization, and are sometimes invoked to explain primary strategy schemes \citep{macarthur_1967,winemiller_1992,aerts_1999}, they are not necessary for it.
+As an example of a trade-off, corals which rapidly out-shade neighbors have a tall, branched morphology which is vulnerable to disturbances, and so, all else being equal, ideal environment $c$-strategists will suffer higher mortality from disturbances \citep{darling_2012}.
 Fig.~\ref{fig:axes} gives the same conclusion without invoking trade-offs; mutations which reduce disturbance vulnerability are essentially neutral under ideal conditions, leading to no improvements in mortality from disturbances, whereas $c$ will tend to increase over time.
 
-Our prediction of evolution for higher $b$ in ideal environments is counter to the expectations of MacArthur's $r$/$K$ dichotomy \citep{macarthur_1967} since $b$ is closely related to the maximal, low-density growth rate $r=b-d$, and ideal environments support high population densities which should be subject to ``$K$-selection''.
-However, in the Introduction, we noted that the $r$-$K$ dichotomy is not consistent with empirical studies showing that maximal growth rate and saturation density (measured by abundance) are positively correlated, both between species/strains \citep{luckinbill_1979,kuno_1991,hendriks_2005,fitzsimmons_2010}, and as a result of experimental evolution \citep{luckinbill_1978,luckinbill_1979}.
+Our prediction of evolution for higher $b$ in ideal environments is counter to the expectations of MacArthur's $r$/$K$ dichotomy \citep{macarthur_1967}, since $b$ is closely related to the maximal, low-density growth rate $r=b-d$ \citep{pianka_1972}, and ideal environments support high population densities which should be subject to ``$K$-selection''.
+However, in the Introduction, we noted that the $r$/$K$ dichotomy is not consistent with empirical studies showing that maximal growth rate and saturation density (measured by abundance) are positively correlated, both between species/strains \citep{luckinbill_1979,kuno_1991,hendriks_2005,fitzsimmons_2010}, and as a result of experimental evolution \citep{luckinbill_1978,luckinbill_1979}.
 From the perspective of our model, these correlations are not surprising since the saturation density, which is determined by a balance between births and deaths, increases with $b$.
+
 Our higher-$b$ prediction simply reflects the fact that, all else being equal, producing more propagules is always advantageous, regardless of population density, a fact lost in the simple logistic interpretation of the $r$/$K$ scheme.
 
-Confusingly, the term ``$K$-selection'' has sometimes been used to refer generally to selection at high density; this encompasses both selection for higher saturation density --- ``$K$'' in the logistic equation --- as well as selection for competitive ability.
+Confusingly, the term ``$K$-selection'' has sometimes been used to refer generally to selection at high density \citep{pianka_1972}; this encompasses both selection for higher saturation density --- ``$K$'' in the logistic equation --- as well as selection for competitive ability.
 To avoid this ambiguity, the latter form of selection has been called ``$\alpha$-selection'' after the competition coefficients in the Lotka-Volterra equation \citep{gill_1974,case_1974,joshi_2001}.

### 2017-01-27 manuscript.tex
latest

@@
 However, in the Introduction, we noted that the $r$/$K$ dichotomy is not consistent with empirical studies showing that maximal growth rate and saturation density (measured by abundance) are positively correlated, both between species/strains \citep{luckinbill_1979,kuno_1991,hendriks_2005,fitzsimmons_2010}, and as a result of experimental evolution \citep{luckinbill_1978,luckinbill_1979}.
-From the perspective of our model, these correlations are not surprising since the saturation density, which is determined by a balance between births and deaths, increases with $b$.
+From the perspective of our model, this correlation is not surprising since the saturation density, which is determined by a balance between births and deaths, increases with $b$.
 
-Our higher-$b$ prediction simply reflects the fact that, all else being equal, producing more propagules is always advantageous, regardless of population density, a fact lost in the simple logistic interpretation of the $r$/$K$ scheme.
+The importance of $b$ in securing territories is a general feature of lottery competition.
+Indeed, as can be seen from Eq.
+\eqref{eq:lottery}, in the classic lottery model $b_i$ and $c_i$ are essentially equivalent in that only the products $b_i c_i$ matter \citep{chesson_1981}.
+This is no longer the case in our density- and frequency-dependent generalization of the classic lottery model, where stable co-existence is possible between $b$ and $c$ strategists.
+Given that the classic lottery model was specifically developed for studying species co-existence questions, this may seem surprising, but the focus in that case was on the role of environmental fluctuations in promoting co-existence rather than coexistence in a single, stable environment \citep{chesson_1981}.
+It is not clear whether correctly accounting for the behavior of species with low density would significantly alter the conclusions of \cite{chesson_1981}, but, in any case, species co-existence questions are beyond the scope of this manuscript.
+
+Our higher-$b$ prediction simply reflects the fact that, all else being equal, producing more propagules , regardless of population density, a fact lost in the simple logistic interpretation of the $r$/$K$ scheme.
 
@@
 We leave these complications for future work.
-
-The importance of $b$ in securing territories is a general feature of lottery competition.
-Indeed, as can be seen from Eq.
-\eqref{eq:lottery}, in the classic lottery model $b_i$ and $c_i$ are essentially equivalent in that only the products $b_i c_i$ matter \citep{chesson_1981}.
-This is no longer the case in our density- and frequency-dependent generalization of the classic lottery model, where stable co-existence is possible between $b$ and $c$ strategists.
-Given that the classic lottery model was specifically developed for studying species co-existence questions, this may seem surprising, but the focus in that case was on the role of environmental fluctuations in promoting co-existence rather than coexistence in a single, stable environment \citep{chesson_1981}.
-It is not clear whether correctly accounting for the behavior of species with low density would significantly alter the conclusions of \cite{chesson_1981}, but, in any case, species co-existence questions are beyond the scope of this manuscript.
 

### 2017-01-27 manuscript.tex
latest

@@
 Our prediction of evolution for higher $b$ in ideal environments is counter to the expectations of MacArthur's $r$/$K$ dichotomy \citep{macarthur_1967}, since $b$ is closely related to the maximal, low-density growth rate $r=b-d$ \citep{pianka_1972}, and ideal environments support high population densities which should be subject to ``$K$-selection''.
-However, in the Introduction, we noted that the $r$/$K$ dichotomy is not consistent with empirical studies showing that maximal growth rate and saturation density (measured by abundance) are positively correlated, both between species/strains \citep{luckinbill_1979,kuno_1991,hendriks_2005,fitzsimmons_2010}, and as a result of experimental evolution \citep{luckinbill_1978,luckinbill_1979}.
+However, the $r$/$K$ dichotomy is also not consistent with empirical studies showing that maximal growth rate and saturation density (measured by abundance) are positively correlated, both between species/strains \citep{luckinbill_1979,kuno_1991,hendriks_2005,fitzsimmons_2010}, and as a result of experimental evolution \citep{luckinbill_1978,luckinbill_1979}.
 From the perspective of our model, this correlation is not surprising since the saturation density, which is determined by a balance between births and deaths, increases with $b$.
 
-The importance of $b$ in securing territories is a general feature of lottery competition.
-Indeed, as can be seen from Eq.
-\eqref{eq:lottery}, in the classic lottery model $b_i$ and $c_i$ are essentially equivalent in that only the products $b_i c_i$ matter \citep{chesson_1981}.
-This is no longer the case in our density- and frequency-dependent generalization of the classic lottery model, where stable co-existence is possible between $b$ and $c$ strategists.
-Given that the classic lottery model was specifically developed for studying species co-existence questions, this may seem surprising, but the focus in that case was on the role of environmental fluctuations in promoting co-existence rather than coexistence in a single, stable environment \citep{chesson_1981}.
+The importance of $b$ at high densities is a consequence of the unpredictable nature of competition at the territorial level.
+Even in the extreme case that one genotype is guaranteed to defeat all others in a ``fair'' contest (e.g.
+it is the most efficient exploiter of a limiting consumable resource; \citealt{tilman_1982}), inferior competitors can win any given territory by sheer luck.
+This could be because of which propagules happen to arrive first (a dispersal advantage), or other stochastic factors.
+The lottery model of competition captures this effect.
+Viewed through the lens of the Wright-Fisher model, the importance of $b$ at high densities is obvious --- the analogous effect is that fecundity confers a relative fitness advantage even though the population is not growing.
+
+In the classic lottery model, $b_i$ and $c_i$ are essentially equivalent in that only the product $b_i c_i$ matters (see ``Model'').
+This is no longer the case in our density- and frequency-dependent generalization of the classic lottery model, where there is a colonization-competition trade-off \citep{tilman_94}, and co-existence between $b$ and $c$ specialists.
+
 It is not clear whether correctly accounting for the behavior of species with low density would significantly alter the conclusions of \cite{chesson_1981}, but, in any case, species co-existence questions are beyond the scope of this manuscript.
-
-Our higher-$b$ prediction simply reflects the fact that, all else being equal, producing more propagules , regardless of population density, a fact lost in the simple logistic interpretation of the $r$/$K$ scheme.
 

### 2017-01-30 manuscript.tex
latest

--- 2017-01-27 manuscript.tex
@@
 In ideal environments, we predict evolution for higher $b$ and $c$, but not lower $d$.
-In plants and corals, a key mechanism for winning territorial contests is rapidly outgrowing and ``shading out'' competitors; empirically, rapid individual growth is a defining feature of the competitor trait cluster \citep{grime_1977,darling_2012}.
-Evolution for higher $b$ under high-density, competitive conditions may seem counter-intuitive.
-Neither particularly high nor low $b$ have been associated with the competitor strategy in plants and corals.
-However, for fishes, the analogous ``periodic'' strategy is characterized by enormous spawn sizes as well as rapid development \citep{winemiller_1992,winemiller_2015}, suggesting a strategy of ensuring that many propagules actually end up contesting areas favorable for development (higher $b$).
-The evolution of $b$ in ideal environments will be discussed further below.
+In plants and corals, a key mechanism for winning territorial contests (higher $c$) is rapidly outgrowing and ``shading out'' competitors; empirically, rapid individual growth is a defining feature of the competitor trait cluster \citep{grime_1977,darling_2012}.
+The situation for $b$ is more ambiguous.
+Competitor strategies in plants and corals span a range of $b$.
+For fishes, the analogous ``periodic'' strategy is characterized by enormous spawn sizes as well as rapid development \citep{winemiller_1992,winemiller_2015}.
+The role of $b$ in ideal environments will be discussed further below.
 
@@
 
-Our realization of Grime's triangle (Fig.~4) refers to the direction of adaptive trait evolution under different regimes of stress and disturbance.
-As discussed in ``Primary strategies and Grime's triangle'', over time our predicted trait evolution should produce traits consistent with Grime's scheme.
-In making these predictions, we have not assumed any kind of trade-offs or pleitropy between $b$, $c$ and $d$, only constraints imposed by the environment on the order of magnitude of $b$ and $d$.
+Adaptive evolution in the direction predicted by our model should produce traits consistent with Grime's scheme (``Primary strategies and Grime's triangle''), with the possible exception of evolution for higher $b$ at high-densities.
+The latter is also is counter to the expectations of MacArthur's $r$/$K$ dichotomy \citep{macarthur_1967}, since $b$ is closely related to the maximal, low-density growth rate $r=b-d$ \citep{pianka_1972}, and yet ideal environments support high population densities which should be subject to ``$K$-selection''.
+
+Intuitively, it makes sense that $b$ can matter at high densities.
+In our model (or any lottery model of competition), $b$ matters at high densities because territorial contests among juveniles are intrinsically unpredictable.
+This is a realistic feature of the model.
+Even if one genotype is guaranteed to win a territory in a ``fair'' contest (e.g.
+it is the most efficient exploiter of a limiting consumable resource; \citealt{tilman_1982}), inferior competitors can win by chance.
+For example, an inferior competitor's propagules may happen to arrive first, gaining a decisive developmental head start.
+First arrivals are more likely to occur for genotypes with a fecundity and/or dispersal advantage, as represented by higher $b$ in lottery models.
+The analogous intuition in terms of the Wright-Fisher model is that fecundity confers a relative fitness advantage, even though populaiton size is not changing.
+The logistic model for which $r$ and $K$ are named, does not capture this intuition.
+
+Confusingly, the term ``$K$-selection'' has sometimes been used to refer generally to selection at high density \citep{pianka_1972}; this encompasses both selection for higher saturation density --- ``$K$'' in the logistic equation --- as well as selection for competitive ability.
+Contrary to an $r$/$K$ dichotomy, empirical studies have shown that maximal growth rate and saturation density (measured by abundance) are positively correlated, both between species/strains \citep{luckinbill_1979,kuno_1991,hendriks_2005,fitzsimmons_2010}, and as a result of experimental evolution \citep{luckinbill_1978,luckinbill_1979}.
+From the perspective of our model, this correlation is not surprising since the saturation density, which is determined by a balance between births and deaths, increases with $b$.
+
+Unlike saturation density, and consistent with an $r$/$K$ dichotomy, there is support for a negative relationship between competitive success at high densities and maximal growth rate \citep{luckinbill_1979}; this could be driven by a tradeoff between individual size and reproductive rate.
+To avoid confusion with other forms of ''$K$-selection'', selection for competitive ability has been called ``$\alpha$-selection'' after the competition coefficients in the Lotka-Volterra equation \citep{gill_1974,case_1974,joshi_2001}.
+Competitive success as measured by $\alpha$ (i.e.
+the per-capita effect of one genotype on another genotype's growth rate) is only partly determined by individual competitive ability --- in the presence of age-structured competition and territoriality, it also includes the ability of each genotype to produce contestants i.e.
+$b$ in our model.
+Our $c$ is strictly competitive ability only --- as such, changes in $c$ do not directly affect population density (section ``Model''), nor the ability to produce contestants.
+
+$K$-selection in the narrow logistic sense of selection for a greater environmental carrying capacity for given birth and death rates, sometimes referred to as ``efficiency'' \citep{macarthur_1967}, could be represented in our model by smaller individual territorial requirements.
+To a first approximation, two co-occurring genotypes which differ by a small amount in their territorial requirements only should have the same fitness since the costs or benefits of a change in the amount of unocupied territory is shared equally among genotypes via the propagule density per territory $L$.
+The situation is more complicated when the differences in territorial requirements become large enough that territorial contests can occur on different scales for different genotypes.
+We leave these complications for future work.
+
+In the classic lottery model, $b_i$ and $c_i$ are essentially equivalent in the sense that the number of territorial victories only depends on the product $b_i c_i$ (see ``Model'').
+This is no longer the case in our density- and frequency-dependent generalization of the classic lottery model, where there is co-existence between $b$ and $c$ specialists.
+This trade-off is similar to the colonization-competition trade-off that is well known in the co-existence literature \citep{tilman_94}.
+It and similar forms of ``spatial co-existence'' in stable environments have previously been modeled either with Levin's  qualitative representation of competition \citep{levins_71,tilman_94}, as opposed to the quantitative $c$ of lottery competition, or with a more sophisticated treatment of space (non-uniform dispersal) \citep{shmida_84,bolker_99}.
+In fluctuating environments our model would likely give similar co-existence predictions as \cite{chesson_1981}, which revolve around the ``storage effect'', since we retain overlapping generations and competition at the juvenile phase only.
+
+While our model can be applied to questions at an inter-species level (e.g.
+ecological invasions), our focus here is on the evolution of genotype frequencies within a population.
+Our assumption that there are no large $c$ discrepancies (section ``Mean field approximation'') amounts to a restriction on the amount of genetic variation in $c$ in the population.
+Since beneficial mutation effect sizes will typically not be much larger than a few percent, large $c$ discrepancies can only arise if the mutation rate is extremely large, and so the assumption will not be violated in most cases.
+However, this restriction could become important when looking at species interactions rather than genotype evolution.
+
+Our realization of Grime's triangle (Fig.~4) differs from approaches which identify primary strategies as trait combinations which can co-exist \citep{bolker_99}. 
+Instead, refers to the direction of adaptive trait evolution under different regimes of stress and disturbance, which is closer in the spirit to Grime's original arguments.
+In making our predictions, we have not assumed any kind of trade-offs or pleitropy between $b$, $c$ and $d$, only constraints imposed by the environment on the order of magnitude of $b$ and $d$.
 Thus, while trade-offs may amplify specialization, and are sometimes invoked to explain primary strategy schemes \citep{macarthur_1967,winemiller_1992,aerts_1999}, they are not necessary for it.
@@
 Fig.~\ref{fig:axes} gives the same conclusion without invoking trade-offs; mutations which reduce disturbance vulnerability are essentially neutral under ideal conditions, leading to no improvements in mortality from disturbances, whereas $c$ will tend to increase over time.
-
-Our prediction of evolution for higher $b$ in ideal environments is counter to the expectations of MacArthur's $r$/$K$ dichotomy \citep{macarthur_1967}, since $b$ is closely related to the maximal, low-density growth rate $r=b-d$ \citep{pianka_1972}, and ideal environments support high population densities which should be subject to ``$K$-selection''.
-However, in the Introduction, we noted that the $r$/$K$ dichotomy is not consistent with empirical studies showing that maximal growth rate and saturation density (measured by abundance) are positively correlated, both between species/strains \citep{luckinbill_1979,kuno_1991,hendriks_2005,fitzsimmons_2010}, and as a result of experimental evolution \citep{luckinbill_1978,luckinbill_1979}.
-From the perspective of our model, this correlation is not surprising since the saturation density, which is determined by a balance between births and deaths, increases with $b$.
-
-The importance of $b$ in securing territories is a general feature of lottery competition.
-Indeed, as can be seen from Eq.
-\eqref{eq:lottery}, in the classic lottery model $b_i$ and $c_i$ are essentially equivalent in that only the products $b_i c_i$ matter \citep{chesson_1981}.
-This is no longer the case in our density- and frequency-dependent generalization of the classic lottery model, where stable co-existence is possible between $b$ and $c$ strategists.
-Given that the classic lottery model was specifically developed for studying species co-existence questions, this may seem surprising, but the focus in that case was on the role of environmental fluctuations in promoting co-existence rather than coexistence in a single, stable environment \citep{chesson_1981}.
-It is not clear whether correctly accounting for the behavior of species with low density would significantly alter the conclusions of \cite{chesson_1981}, but, in any case, species co-existence questions are beyond the scope of this manuscript.
-
-Our higher-$b$ prediction simply reflects the fact that, all else being equal, producing more propagules , regardless of population density, a fact lost in the simple logistic interpretation of the $r$/$K$ scheme.
-
-Confusingly, the term ``$K$-selection'' has sometimes been used to refer generally to selection at high density \citep{pianka_1972}; this encompasses both selection for higher saturation density --- ``$K$'' in the logistic equation --- as well as selection for competitive ability.
-To avoid this ambiguity, the latter form of selection has been called ``$\alpha$-selection'' after the competition coefficients in the Lotka-Volterra equation \citep{gill_1974,case_1974,joshi_2001}.
-Unlike saturation density, there is support for a negative relationship between competitive success at high densities and maximal growth rate \citep{luckinbill_1979}; this could be driven by a tradeoff between individual size and reproductive rate.
-However, competitive success as measured by $\alpha$ (i.e.
-the per-capita effect of one genotype on another genotype's growth rate) is only partly determined by individual competitive ability --- in the presence of age-structure and territoriality, it also includes the ability of each genotype to produce contestants i.e.
-$b$ in our model.
-In contrast, our $c$ is strictly competitive ability only --- as such, changes in $c$ do not directly affect population density (section ``Model''), nor the ability to produce contestants.
-
-$K$-selection in the sense of selection for a greater environmental carrying capacity for given birth and death rates, sometimes referred to as ``efficiency'' \citep{macarthur_1967}, would be represented in our model by smaller individual territorial requirements.
-To a first approximation, two co-occurring genotypes which differ by a small amount in their territorial requirements only should have the same fitness since the costs or benefits of a change in the amount of unocupied territory is shared equally among genotypes via the propagule density $L$.
-The situation is more complicated if those genotypes differ in multiple traits, and when the differences in territorial requirements become large enough that territorial contests can occur on different scales.
-We leave these complications for future work.
-
-Rather, while our model can be applied to questions at an inter-species level (e.g.
-ecological invasions), our focus here is on the evolution of genotype frequencies within a population.
-Our ability to describe evolutionary processes is only possible because the model accounts for the growth of mutants from low densities.
-Given this focus, our assumption that there are no large $c$ discrepancies (section ``Mean field approximation'') amounts to a restriction on the amount of genetic variation in $c$ that will be sustained in the population.
-Since beneficial mutation effect sizes will typically not be much larger than a few percent, large $c$ discrepancies can only arise if the mutation rate is extremely large, and so the assumption will not be violated in most cases.
-However, this restriction could become important when looking at species interactions rather than genotype evolution.
 

### 2017-01-30 manuscript.tex
latest

@@
 We assume that adults cannot be ousted from occupied territories, and so only propagules landing on unoccupied territories are included in $m_i$.
-Propagules disperse globally, at random, and independently of each other; all propagules have the same probability of landing on any of the $U$ unnocupied territories.
+Propagules disperse at random uniformly, and independently of each other; all propagules have the same probability of landing on any of the $U$ unnocupied territories.
 Thus, there is no interaction between propagules (e.g.
@@
 This choice simplifies the mathematics without seriously restricting the generality of our analysis, since the results presented here are not sensitive to the specific functional form of $m_i$.
-Note that due to our assumption of global dispersal, the parameter $b_i$ can be thought of as a measure of ``colonization ability'', which combines fecundity and dispersal ability \citep{levins_71,tilman_94,bolker_99}.
+Note that due to our assumption of uniform dispersal, the parameter $b_i$ can be thought of as a measure of ``colonization ability'', which combines fecundity and dispersal ability \citep{levins_71,tilman_94,bolker_99}.
 
@@
 
-The main advantage of the classic lottery model for our purposes here is that the abstract population-genetic relative fitnesses $w_i$ can be replaced with more ecologically-grounded fecundity, competitive ability and mortality parameters $b_i$, $c_i$ and $d_i$.
-In particular, differences in individual longevity are only possible because generations are overlapping.
-However, note that the classic lottery model, and the present work, ignore the stochastic drift in type frequencies (large $T$ approximation) which is the focus of the Wright-Fisher model.
+The classic lottery model allows us to replace the abstract Wright-Fisher relative fitnesses $w_i$ with more ecologically-grounded fecundity, competitive ability and mortality parameters $b_i$, $c_i$ and $d_i$, respectively.
+Since birth and death rates affect absolute abundances, this allows us to evaluate selection at different densities (after appropriate extensions are made), in an otherwise very similar model to the canonical Wright-Fisher model.
+Note that the classic lottery model, and the present work, both ignore the stochastic drift in type frequencies (large $T$ approximation) which is the focus of the Wright-Fisher model.
 
@@
 Eq.
-\eqref{eq:lottery} is nonsensical at low densities (if $l_i$ is small): genotype $i$ can win at most $m_i$ territories, yet Eq.
+\eqref{eq:lottery} is nonsensical at low densities ($l_i\ll 1$): genotype $i$ can win at most $m_i$ territories, yet Eq.
 \eqref{eq:lottery} demands $c_i l_i/\sum_j c_j l_j$ of the $U$ unoccupied territories, for any value of $U$.
-Thus, at low densities, deviations from the mean propagule composition $l_1,l_2,\ldots,l_G$ are important.
+Intuitively, the cause of this discrepancy is that individuals are discrete.
+Genotypes with few propagules depend on the outcome of contests in territories where they have at least one propagule present, not some small fraction of a propagule as would be implied by low propagule density $l$ in the classic lottery model.
+In other words, deviations from the mean propagule composition $l_1,l_2,\ldots,l_G$ are important at low density.
 
-Since dispersal is Poisson, we expect that a fraction $p_1(x_1)\ldots p_G(x_G)$ of the $U$ unoccupied territories will have the propagule composition $x_1,\ldots,x_G$.
+Dispersal is Poisson, and so we expect that a fraction $p_1(x_1)\ldots p_G(x_G)$ of the $U$ unoccupied territories will have the propagule composition $x_1,\ldots,x_G$.
 Genotype $i$ is expected to win $c_i x_i/\sum_j c_j x_j$ of these.
@@
 
-Adaptive evolution in the direction predicted by our model should produce traits consistent with Grime's scheme (``Primary strategies and Grime's triangle''), with the possible exception of evolution for higher $b$ at high-densities.
-The latter is also is counter to the expectations of MacArthur's $r$/$K$ dichotomy \citep{macarthur_1967}, since $b$ is closely related to the maximal, low-density growth rate $r=b-d$ \citep{pianka_1972}, and yet ideal environments support high population densities which should be subject to ``$K$-selection''.
+As discussed in ``Primary strategies and Grime's triangle'', adaptive evolution in the direction predicted by our generalized lottery model produces traits consistent with Grime's scheme, with the possible exception of selection for $b$ at high density, because the corresponding trait data  is ambiguous.
+Selection for $b$ at high density is also is counter to the expectations of MacArthur's $r$/$K$ dichotomy \citep{macarthur_1967}, since $b$ is closely related to the maximal, low-density growth rate $r=b-d$ \citep{pianka_1972}, yet in the $r$/$K$ scheme, high density populations should be subject to $K$, not $r$, selection.
 
-Intuitively, it makes sense that $b$ can matter at high densities.
+It is not surprising that $b$ can matter at high densities.
 In our model (or any lottery model of competition), $b$ matters at high densities because territorial contests among juveniles are intrinsically unpredictable.
@@
 it is the most efficient exploiter of a limiting consumable resource; \citealt{tilman_1982}), inferior competitors can win by chance.
-For example, an inferior competitor's propagules may happen to arrive first, gaining a decisive developmental head start.
+For example, an inferior competitor's propagules may happen to arrive first, gaining a decisive developmental advantage.
 First arrivals are more likely to occur for genotypes with a fecundity and/or dispersal advantage, as represented by higher $b$ in lottery models.
-The analogous intuition in terms of the Wright-Fisher model is that fecundity confers a relative fitness advantage, even though populaiton size is not changing.
+The analogous intuition in the Wright-Fisher model is that fecundity confers a relative fitness advantage, even though population size is not changing.
 The logistic model for which $r$ and $K$ are named, does not capture this intuition.
 
-Confusingly, the term ``$K$-selection'' has sometimes been used to refer generally to selection at high density \citep{pianka_1972}; this encompasses both selection for higher saturation density --- ``$K$'' in the logistic equation --- as well as selection for competitive ability.
+Confusingly, the term ``$K$-selection'' sometimes refers generally to selection at high density \citep{pianka_1972}, encompassing both selection for higher saturation density \citep{macarthur_1967} and competitive ability \citep{gill_1974}.
 Contrary to an $r$/$K$ dichotomy, empirical studies have shown that maximal growth rate and saturation density (measured by abundance) are positively correlated, both between species/strains \citep{luckinbill_1979,kuno_1991,hendriks_2005,fitzsimmons_2010}, and as a result of experimental evolution \citep{luckinbill_1978,luckinbill_1979}.
@@
 
-Unlike saturation density, and consistent with an $r$/$K$ dichotomy, there is support for a negative relationship between competitive success at high densities and maximal growth rate \citep{luckinbill_1979}; this could be driven by a tradeoff between individual size and reproductive rate.
+There is support for a negative relationship between competitive success at high density and maximal growth rate \citep{luckinbill_1979}, consistent with an $r$/$K$ dichotomy.
+This could be driven by a tradeoff between individual size and reproductive rate.
 To avoid confusion with other forms of ''$K$-selection'', selection for competitive ability has been called ``$\alpha$-selection'' after the competition coefficients in the Lotka-Volterra equation \citep{gill_1974,case_1974,joshi_2001}.
-Competitive success as measured by $\alpha$ (i.e.
+However, competitive success as measured by $\alpha$ (i.e.
 the per-capita effect of one genotype on another genotype's growth rate) is only partly determined by individual competitive ability --- in the presence of age-structured competition and territoriality, it also includes the ability of each genotype to produce contestants i.e.
 $b$ in our model.
-Our $c$ is strictly competitive ability only --- as such, changes in $c$ do not directly affect population density (section ``Model''), nor the ability to produce contestants.
+Our $c$ is strictly competitive ability only --- as such, changes in $c$ do not directly affect population density (section ``Model'').
+The clean separation of a strictly-relative $c$ parameter is particularly useful from an evolutionary genetics perspective, essentially embedding a zero-sum relative fitness trait within a non-zero-sum fitness model.
+This could have interesting applications for modeling the impacts of intra-specific competition on species extinction, for example due to clonal interference \citep{gerrish_1998,desai_2007} between $c$-strategists on the one hand, and $b$- and $d$- strategists on the other.
 
@@
 In the classic lottery model, $b_i$ and $c_i$ are essentially equivalent in the sense that the number of territorial victories only depends on the product $b_i c_i$ (see ``Model'').
-This is no longer the case in our density- and frequency-dependent generalization of the classic lottery model, where there is co-existence between $b$ and $c$ specialists.
-This trade-off is similar to the colonization-competition trade-off that is well known in the co-existence literature \citep{tilman_94}.
-It and similar forms of ``spatial co-existence'' in stable environments have previously been modeled either with Levin's  qualitative representation of competition \citep{levins_71,tilman_94}, as opposed to the quantitative $c$ of lottery competition, or with a more sophisticated treatment of space (non-uniform dispersal) \citep{shmida_84,bolker_99}.
+This is no longer the case in our density- and frequency-dependent generalization, where $b$ and $c$ specialists can co-exist.
+This ``colonization-competition trade-off'' is well known in the co-existence literature \citep{tilman_94}.
+It and similar forms of ``spatial co-existence'' in stable environments have previously been modeled either with Levin's  qualitative representation of competition \citep{levins_71,tilman_94}, as opposed to the quantitative $c$ of lottery competition, or with a more sophisticated treatment of space (non-uniform dispersal; \citealt{shmida_84,bolker_99}).
 In fluctuating environments our model would likely give similar co-existence predictions as \cite{chesson_1981}, which revolve around the ``storage effect'', since we retain overlapping generations and competition at the juvenile phase only.
 
-While our model can be applied to questions at an inter-species level (e.g.
-ecological invasions), our focus here is on the evolution of genotype frequencies within a population.
-Our assumption that there are no large $c$ discrepancies (section ``Mean field approximation'') amounts to a restriction on the amount of genetic variation in $c$ in the population.
-Since beneficial mutation effect sizes will typically not be much larger than a few percent, large $c$ discrepancies can only arise if the mutation rate is extremely large, and so the assumption will not be violated in most cases.
-However, this restriction could become important when looking at species interactions rather than genotype evolution.
-
-Our realization of Grime's triangle (Fig.~4) differs from approaches which identify primary strategies as trait combinations which can co-exist \citep{bolker_99}. 
-Instead, refers to the direction of adaptive trait evolution under different regimes of stress and disturbance, which is closer in the spirit to Grime's original arguments.
-In making our predictions, we have not assumed any kind of trade-offs or pleitropy between $b$, $c$ and $d$, only constraints imposed by the environment on the order of magnitude of $b$ and $d$.
-Thus, while trade-offs may amplify specialization, and are sometimes invoked to explain primary strategy schemes \citep{macarthur_1967,winemiller_1992,aerts_1999}, they are not necessary for it.
+Our realization of Grime's triangle (Fig.~4) differs from approaches which identify primary strategies as trait combinations which can co-exist \citep{bolker_99}, referring instead to the direction of adaptive trait evolution under different regimes of stress and disturbance, which is closer in the spirit to Grime's arguments \citep{grime_1974,grime_1977}.
+In addition, we have not assumed any kind of trade-offs or pleitropy between $b$, $c$ and $d$, only constraints imposed by the environment on the order of magnitude of $b$ and $d$.
 As an example of a trade-off, corals which rapidly out-shade neighbors have a tall, branched morphology which is vulnerable to disturbances, and so, all else being equal, ideal environment $c$-strategists will suffer higher mortality from disturbances \citep{darling_2012}.
 Fig.~\ref{fig:axes} gives the same conclusion without invoking trade-offs; mutations which reduce disturbance vulnerability are essentially neutral under ideal conditions, leading to no improvements in mortality from disturbances, whereas $c$ will tend to increase over time.
+Thus, while trade-offs may amplify specialization, and are sometimes invoked to explain primary strategy schemes \citep{macarthur_1967,winemiller_1992,aerts_1999}, they are not necessary for it.
 
@@
 \ref{fig:table2} we compare our model with some of the other models and schemes touched upon here.
-In a sense this comparison is unfair: for instance, Grime's scheme was developed for an entirely different purpose (species classification by traits).
+In a sense this an ``apples and oranges'' comparison: for instance, Grime's scheme was developed for an entirely different purpose (species classification by traits).
 As such, Fig.
@@
 By comparison, more familiar approaches to evolutionary ecology such as adaptive dynamics --- essentially ecology coupled with mutant invasion \citep{diekmann_2004} ---  lie close to the ecology-only end of the spectrum.
+
+[Inference paragraph] An important potential application of eco-evolutionary models close to the evolution-only end of the spectrum is improved population-genetic inference in expanding (e.g.
+humans) or seasonally-cycling (e.g.
+\textit{Drosophila}) populations \citep{bergland_14}.
 
@@
 
-In our view, our model has two major limitations as a general-purpose model of density-dependent selection: a reliance on interference competition for durable resources (territoriality), and the restriction of competition to juveniles (lottery recruitment to adulthood).
+Two major limitations of our model as a general-purpose model of density-dependent selection are the restriction to interference competition for durable resources (territoriality), and the restriction of competition to juveniles (lottery recruitment to adulthood).
 In some respects this is the complement of resource competition models, which restrict their attention to exploitation competition, typically without age structure \citep{tilman_1982}.
@@
 We are not aware of any attempts to delineate conditions under which explicit resource tracking is unnecessary even if it is assumed that community structure is ultimately determined by competition for consumable resources.
-More work is needed connecting resource competition models to the density-dependent selection literature, since most of the existing work is focused on the narrower Grime-Tilman-debate issue of resolving the role of competition at different levels of resource availability \citep{aerts_1999,davis_1998,tilman_2007}.
+More work is needed connecting resource competition models to the density-dependent selection literature, since most of the existing work is focused on narrower issue of the role of competition at low resource availability \citep{aerts_1999,davis_1998,tilman_2007}.
 
-On the other hand, our model does remarkably well in capturing apparently general features of selection under different environmental conditions using only three trait parameters: $b$, $c$ and $d$.
-The clean separation of a strictly-relative $c$ parameter is particularly useful from an evolutionary genetics perspective, essentially embedding a zero-sum fitness trait within a non-zero-sum fitness model.
-This could have interesting applications for modeling the impacts of intra-specific competition on species extinction, for example due to clonal interference \citep{gerrish_1998,desai_2007} between $c$-strategists on the one hand, and $b$- and $d$- strategists on the other.
+While our model can be applied to species rather than genotypes (e.g.
+ecological invasions), our focus is genotype evolution.
+Our assumption that there are no large $c$ discrepancies (section ``Mean field approximation'') amounts to a restriction on the amount of genetic variation in $c$ in the population.
+Since beneficial mutation effect sizes will typically not be much larger than a few percent, large $c$ discrepancies can only arise if the mutation rate is extremely large, and so the assumption will not be violated in most cases.
+However, this restriction could become important when looking at species interactions rather than genotype evolution.
 

### 2017-02-02 manuscript.tex
latest

@@
 \title{Density-dependent selection in evolutionary genetics: a lottery model of Grime's triangle}
-\author{Jason Bertram $^{1,\ast}$ \\ 
-Joanna Masel $^{1}$}
+%\author{Jason Bertram $^{1,\ast}$ \\ 
+%Joanna Masel $^{1}$}
 
@@
 
-\noindent{}1. Department of Ecology and Evolutionary Biology, University of Arizona, Tucson, AZ 85721.
+%\noindent{}1. Department of Ecology and Evolutionary Biology, University of Arizona, Tucson, AZ 85721.
 
-\noindent{}$\ast$ Corresponding author; e-mail: jbertram@email.arizona.edu.
+%\noindent{}$\ast$ Corresponding author; e-mail: jbertram@email.arizona.edu.
 
@@
 
-\textit{Keywords}: r/K selection, absolute fitness, eco-evo, interference competition.
+\textit{Keywords}: r/K selection, absolute fitness, eco-evo, interference competition, competition-colonization trade-off.
 
@@
 Fitness is typically represented in heavily simplified terms in evolutionary genetics, often using constant selection coefficients.
-This excludes fundamental ecological factors such as dynamic population size or density-dependence from our most genetically-realistic treatments of evolution, a problem that inspired MacArthur's influential but problematic $r$/$K$ theory.
+This excludes fundamental ecological factors such as dynamic population size or density-dependence from the most genetically-realistic treatments of evolution, a problem that inspired MacArthur's influential but problematic $r$/$K$ theory.
 Following in the spirit of $r$/$K$-selection as a general-purpose theory of density-dependent selection, but grounding ourselves empirically in ``primary strategy'' trait classification schemes like Grime's triangle, we develop a new model of density-dependent selection which revolves around territorial contests.
@@
 This considerably simplifies the mathematics of selection, facilitating greater genetic realism, and can be justified over sufficiently short time intervals \citep[p.
-276]{ewens_2012}.
+276]{ewens_2004}.
 However, the resulting picture of evolution does not include even basic elements of the ecological underpinnings of selection, including dynamic population size and density-dependence.
@@
 It is therefore easy to doubt the feasibility of a simplified, general mathematical treatment of fitness \citep[p.
-276]{ewens_2012}.
+276]{ewens_2004}.
 For example, MacArthur's famous $r$/$K$ scheme \citep{macarthur_1962,macarthur_1967} is now almost exclusively known as a framework for understanding life-history traits, and judged on its failure in that role \citep{pianka_1970,stearns_1977,boyce_1984,reznick_2002}.
@@
 
-Trait classification schemes show empirically that, beneath the complicated details of trait variation, even among closely-related species, fitness is predominantly determined by a few key factors such as intrinsic reproducive rate or stress-tolerance.
-However, while trait classification schemes are firmly grounded in trait data, they are verbal and descriptive rather than mathematical, a recognized hinderance to their broader applicability (e.g.
+Trait classification schemes show empirically that, beneath the complicated details of trait variation, even among closely-related species, fitness is predominantly determined by a few key factors such as intrinsic reproductive rate or stress-tolerance.
+However, while trait classification schemes are firmly grounded in trait data, they are verbal and descriptive rather than mathematical, a recognized hindrance to their broader applicability (e.g.
 \citealt{tilman_2007}).
 
-The aim of this paper is explore the interplay between some major dimensions of fitness in a simplified, spatially-homogeneous model of genotype growth, dispersal and competition.
+The aim of this paper is explore the interplay between some major dimensions of fitness in a simplified, territorial model of growth, dispersal and competition.
 Building on the earlier r/K and C/S/R schemes, a central question is how fitness depends on the interaction between population density, intrinsic birth/death rates and competitive ability.
@@
 We broadly follow the spirit of MacArthur's $r$/$K$ selection scheme in that our model is intended to account for fundamentally different forms of selection without getting  entangled in the intricacies of particular ecological scenarios.
-However, rather than building directly on MacArthur's formalism and its later extensions using Lotka-Volterra equations to incorporate competition (``$\alpha$-selection'') \citep{gill_1974,case_1974,joshi_2001}, our model is devised more with Grime's C/S/R scheme in mind, and represents a quantitative formalization of how C/S/R manifests at the level of within-population genotype evolution (as opposed to phenotypic divergence between species).
+However, rather than building directly on MacArthur's formalism and its later extensions using Lotka-Volterra equations to incorporate competition (``$\alpha$-selection'') \citep{gill_1974,case_1974,joshi_2001}, our model is devised more with Grime's C/S/R scheme in mind, and represents a quantitative formalization of how C/S/R manifests at the level of within-population genotypic evolution (as opposed to phenotypic divergence between species).
 This choice is motivated in part by the substantial empirical support for C/S/R-like schemes, and in part by the failings of the $r$/$K$ low/high density dichotomy --- many growth ability traits will confer advantages at both low and high densities (more details in the Discussion).
@@
 We assume that adults cannot be ousted from occupied territories, and so only propagules landing on unoccupied territories are included in $m_i$.
-Propagules disperse at random uniformly, and independently of each other; all propagules have the same probability of landing on any of the $U$ unnocupied territories.
+Propagules disperse at random uniformly, and independently of each other; all propagules have the same probability of landing on any of the $U$ unoccupied territories.
 Thus, there is no interaction between propagules (e.g.
@@
 Stressful environments are more ambiguous, and have been the subject of an extensive debate in the plant ecology literature (the ``Grime-Tilman'' debate; \citealt{aerts_1999} and references therein).
-Severe stress inhibits growth and reproduction, so that $b\ll 1$ \cite{grime_1974,grime_1977}.
+Severe stress inhibits growth and reproduction, so that $b\ll 1$ \citep{grime_1974,grime_1977}.
 Mutations which appreciably improve $b$ will be either non-existent or extremely unlikely, so $b$ is constrained to remain low.
@@
 The mapping of different environments to our model parameters is summarized in the first two rows of Fig.~\ref{fig:table}.
-Also shown is the approximate dependence of $\Delta_+ n_i$ on $b_i$ and $c_i$ for each environment (third row).
-These can be used infer the expected direction of evolution for the traits $b$, $c$ and $d$ (fourth row) as follows.
+Also shown is the approximate dependence of $\Delta_+ n_i$ on $b_i$ and $c_i$ for each environment (fourth row).
+These can be used infer the expected direction of evolution for the traits $b$, $c$ and $d$ (fifth row) as follows.
 
@@
 Plant ruderals devote a large proportion of their productivity to seed production \citep{grime_1977}, whereas the analogous ``opportunist'' fishes have large intrinsic growth rates \citep{winemiller_1992}.
-In corals, ruderals are distinguished by brood spawning (rather than broadcast spawning).
-This corresponds to higher parental investment and lower overall propagule production --- counter-intuitively, a stereotypical ``$K$-selected'', high-density trait \cite{pianka_1970}.
+In corals, ruderals are distinguished by brood spawning (rather than broadcast spawning; \citealt{darling_2012}).
+This corresponds to higher parental investment and lower overall propagule production --- counter-intuitively, a stereotypical ``$K$-selected'', high-density trait \citep{pianka_1970}.
 However, since broadcast spawners are vulnerable to an Allee effect at the egg fertilization stage \citep{knowlton_2001}, brood spawning could actually be a way to ensure high $b$ at low densities \citep{darling_2012}.
@@
 
-Our realization of Grime's triangle (Fig.~4) differs from approaches which identify primary strategies as trait combinations which can co-exist \citep{bolker_99}, referring instead to the direction of adaptive trait evolution under different regimes of stress and disturbance, which is closer in the spirit to Grime's arguments \citep{grime_1974,grime_1977}.
+Our realization of Grime's triangle (Fig.~4) differs from approaches which identify primary strategies as trait combinations which can co-exist \citep{bolker_99}, referring instead to the direction of adaptive trait evolution under different regimes of stress and disturbance, which is closer in spirit to Grime's arguments \citep{grime_1974,grime_1977}.
 In addition, we have not assumed any kind of trade-offs or pleitropy between $b$, $c$ and $d$, only constraints imposed by the environment on the order of magnitude of $b$ and $d$.
@@
 \ref{fig:table2} we compare our model with some of the other models and schemes touched upon here.
-In a sense this an ``apples and oranges'' comparison: for instance, Grime's scheme was developed for an entirely different purpose (species classification by traits).
+In a sense this is an ``apples and oranges'' comparison: for instance, Grime's scheme was developed for an entirely different purpose (species classification by traits).
 As such, Fig.
@@
 
-[Inference paragraph] An important potential application of eco-evolutionary models close to the evolution-only end of the spectrum is improved population-genetic inference in expanding (e.g.
+An important potential application of eco-evolutionary models close to the evolution-only end of the spectrum is improved population-genetic inference in expanding (e.g.
 humans) or seasonally-cycling (e.g.
 \textit{Drosophila}) populations \citep{bergland_14}.
+Current approaches treat demography and selection as separate questions, typically fitting a demographic model to data on ostensibly neutral sites, and then inferring selection against this demographic background, despite the recurrent problem that the two are confounded \citep{schrider_2016}.
+Eco-evolutionary models of selection may help to remedy this situation, since demography and selection are treated as inseparable, especially on shorter timescales, both simply being the outcome of patterns of births and deaths.
 
@@
 
-Two major limitations of our model as a general-purpose model of density-dependent selection are the restriction to interference competition for durable resources (territoriality), and the restriction of competition to juveniles (lottery recruitment to adulthood).
+One limitation of our model as a general-purpose model of density-dependent selection is the restriction of competition to interference competition between juveniles for durable resources (lottery recruitment to adulthood), analogous to the ubiquitous assumption of viability selection in population genetics \citep[p.
+45]{ewens_2004}.
 In some respects this is the complement of resource competition models, which restrict their attention to exploitation competition, typically without age structure \citep{tilman_1982}.
 In the particular case that resources are spatially localized (e.g.
-due to restricted movement through soils), then resource competition and territorial acquisition effectively coincide, and in principle resource competition could be represented by a competitive ability $c$ (or conversely, $c$ should be derivable from resource competition).
+due to restricted movement through soils), resource competition and territorial acquisition effectively coincide, and in principle resource competition could be represented by a competitive ability $c$ (or conversely, $c$ should be derivable from resource competition).
 The situation is more complicated if the resources are well-mixed, since, in general, resource levels then need to be explicitly tracked.
-It seems plausible that explicit resource tracking is not necessary when the focus is on the evolution of similar genotypes rather than the stable co-existence of widely differing species.
+It seems plausible that explicit resource tracking may not be necessary when the focus is on the evolution of similar genotypes rather than the stable co-existence of widely differing species \citep{ram_2016}.
 We are not aware of any attempts to delineate conditions under which explicit resource tracking is unnecessary even if it is assumed that community structure is ultimately determined by competition for consumable resources.
-More work is needed connecting resource competition models to the density-dependent selection literature, since most of the existing work is focused on narrower issue of the role of competition at low resource availability \citep{aerts_1999,davis_1998,tilman_2007}.
+More work is needed connecting resource competition models to the density-dependent selection literature, since most of the former has to date been focused on narrower issues of the role of competition at low resource availability \citep{aerts_1999,davis_1998,tilman_2007}.
 
@@
 
-\section*{Acknowledgments}
+%\section*{Acknowledgments}
 
-We thank Peter Chesson and Joachim Hermisson for many constructive comments on this manuscript.
+%We thank Peter Chesson and Joachim Hermisson for many constructive comments on this manuscript. This work was financially supported by the National Science Foundation (DEB-1348262).
 

### 2017-03-18 simulations.py
latest

@@ -9,19 +9,19 @@ from numpy import *
 from pylab import *
 import bisect
 
-def expCisim(Ms,U,cs):
-    scatter=[[0 for x in range(len(Ms))] for y in range(U)];
-    for i in range(len(Ms)):
-        for y in xrange(Ms[i]):
+def deltnplussim(m,c,U):
+    scatter=[[0 for x in range(len(m))] for y in range(U)];
+    for i in range(len(m)):
+        for y in xrange(m[i]):
             cell=int(U*rand());
             scatter[cell][i]=scatter[cell][i]+1;
     
-    winsnocomp=zeros(len(Ms)); wins1=zeros(len(Ms)); wins2=zeros(len(Ms));
+    winsnocomp=zeros(len(m)); wins1=zeros(len(m)); wins2=zeros(len(m));
     comp=zeros(U);
     for i in range(U):
         comp[i]=sum(scatter[i])
         if comp[i]>0:            
-            lotterycmf=cumsum(array(scatter[i])*cs)
+            lotterycmf=cumsum(array(scatter[i])*c)
             victor=bisect.bisect(lotterycmf,rand()*lotterycmf[-1])
             
             if scatter[i][victor]==1 and comp[i]==1:
@@ -33,56 +33,49 @@ def expCisim(Ms,U,cs):
         
     return winsnocomp,wins1,wins2
 
-def deltnu(Ms,U,cs,i):
-    return Ms[i]*exp(-sum(Ms)/float(U))
+def R(m,c,U):
+    l=m/float(U)
+    L=sum(l)
+    cbar=sum(m*c)/sum(m); 
+    return cbar*exp(-l)*(1-exp(-(L-l)))\
+            /(c + (L-1+exp(-L))/(1-(1+L)*exp(-L))*(cbar*L-c*l)/(L-l))
 
-def deltnr(Ms,U,cs,i):
-    l=Ms[i]/float(U);notl=sum(Ms)/float(U)-l; L=float(l+notl);
-    #return Ms[i]*exp(-l)*(1-exp(-notl))*cs[i]/(sum([Ms[j]*cs[j]/(U*(1-exp(-notl))) for j in range(len(Ms)) if (j!=i)])+cs[i])
[…]

### 2017-03-20 simulations.py
latest

@@ -49,13 +49,18 @@ def A(m,c,U):
             +(L*(1-exp(-L))/(1-(1+L)*exp(-L))-l*(1-exp(-l))/(1-(1+l)*exp(-l)))/(L-l)*(sum(c*l)-c*l))
     
 def deltnplus(m,c,U):
-    L=sum(m)/U
-    cbar=sum(m*c)/sum(m)
-    return m*(exp(-L)+(R(m,c,U)+A(m,c,U))*c/cbar)
+    if sum(m)>0:
+        L=sum(m)/U
+        cbar=sum(m*c)/sum(m)
+        return m*(exp(-L)+(R(m,c,U)+A(m,c,U))*c/cbar)
+    else:
+        return zeros(len(m))
     
 def deltnplusclassic(m,c,U):
-    return U*m*c/sum(m*c)
-    
+    if sum(m)>0:
+        return U*m*c/sum(m*c)
+    else:
+        return zeros(len(m))
 
 #Simulation comparison figure
 #================================================================
@@ -145,21 +150,37 @@ def b(t):
     return bparam[0]+bparam[1]*(1+sin(2*pi*t/g))
     
 def d(t):
-    return dparam[0]+dparam[1]*(1+sin(2*pi*t/g))
+    return dparam[0]+dparam[1]*(1+cos(2*pi*t/g))
 
-n=array([5000,5000])
+n=array([25000.,25000.])
 c=array([1.,1.])
 T=100000
-g=2.     #generations per seasonal cycle
-bparam=array([[0.1,0.1],[0.1,0.1]])
-dparam=array(array([[0.2,0.2],[0.,0.]]))
-    
-nhist=[]
-totaltime=20
+g=100.     #iterations per seasonal cycle
+bparam=array([[0.,0.],[.5,.3685]])
+dparam=array([[0.0,0.0],[0.2,0.15]])
+
+nhist=[n]
+totaltime=10000
 for t in range(totaltime):
-    print T-sum(n)
-    n=n+deltnplus(b(t)*n,c,T-sum(n))
+    U=T-sum(n)
+    n = n + deltnplus(b(t)*n*U/T,c,U)-d(t)*n
     nhist.append(list(n))
+
+figure()
+plot(nhist)
+figure()
+plot(transpose(transpose(array(nhist))/sum(nhist,1)))
+ylim([0,1])
+figure()
+plot(sum(nhist,1))
+ylim([0,T])
+
+'''
+figure()
+plot([b(t) for t in range(100)])
+figure()
[…]

### 2017-03-20 simulations.py
latest

@@ -153,14 +153,14 @@ def d(t):
     return dparam[0]+dparam[1]*(1+cos(2*pi*t/g))
 
 n=array([25000.,25000.])
-c=array([1.,1.])
+c=array([10.,1.])
 T=100000
-g=100.     #iterations per seasonal cycle
-bparam=array([[0.,0.],[.5,.3685]])
-dparam=array([[0.0,0.0],[0.2,0.15]])
+g=200.     #iterations per seasonal cycle
+bparam=array([[0.05,0.05],[1.,1.]])
+dparam=array([[0.0,0.0],[0.13,0.1]])
 
 nhist=[n]
-totaltime=10000
+totaltime=1000
 for t in range(totaltime):
     U=T-sum(n)
     n = n + deltnplus(b(t)*n*U/T,c,U)-d(t)*n

### 2017-03-21 manuscript.tex
latest

@@
 
+\section*{•}
+
+Current approaches to population genetic inference treat demography and selection separately, typically fitting a demographic model to data on ostensibly neutral sites, and then inferring selection against this demographic background, despite the recurrent problem that the two are confounded \citep{schrider_2016}.
+
+in expanding (e.g. humans) or seasonally-cycling (e.g. \textit{Drosophila}) populations \citep{bergland_14}.
+
+ Eco-evolutionary models of selection may help to remedy this situation, since demography and selection are treated as inseparable, especially on shorter timescales, both simply being the outcome of patterns of births and deaths.
+
 \section*{Discussion}
@@
 By comparison, more familiar approaches to evolutionary ecology such as adaptive dynamics --- essentially ecology coupled with mutant invasion \citep{diekmann_2004} ---  lie close to the ecology-only end of the spectrum.
-
-An important potential application of eco-evolutionary models close to the evolution-only end of the spectrum is improved population-genetic inference in expanding (e.g.
-humans) or seasonally-cycling (e.g.
-\textit{Drosophila}) populations \citep{bergland_14}.
-Current approaches treat demography and selection as separate questions, typically fitting a demographic model to data on ostensibly neutral sites, and then inferring selection against this demographic background, despite the recurrent problem that the two are confounded \citep{schrider_2016}.
-Eco-evolutionary models of selection may help to remedy this situation, since demography and selection are treated as inseparable, especially on shorter timescales, both simply being the outcome of patterns of births and deaths.
 

### 2017-03-24 simulations.py
latest

@@ -147,20 +147,29 @@ savefig('/home/jbertram/repos/densitydependentlottery/simulationcomparison.pdf')
 #=================================================================
 
 def b(t):
-    return bparam[0]+bparam[1]*(1+sin(2*pi*t/g))
+    #return bparam[0]+bparam[1]*(1+sin(2*pi*t/g))
+    if t%g<g/2:
+        return bparam[0]
+    else:
+        return bparam[1]
     
 def d(t):
-    return dparam[0]+dparam[1]*(1+cos(2*pi*t/g))
+    #return dparam[0]+dparam[1]*(1+cos(2*pi*t/g))
+    if t%g<g/2:
+        return dparam[0]
+    else:
+        return dparam[1]
+
 
 n=array([25000.,25000.])
 c=array([10.,1.])
 T=100000
-g=200.     #iterations per seasonal cycle
+g=10.     #iterations per seasonal cycle
 bparam=array([[0.05,0.05],[1.,1.]])
 dparam=array([[0.0,0.0],[0.13,0.1]])
 
 nhist=[n]
-totaltime=1000
+totaltime=100
 for t in range(totaltime):
     U=T-sum(n)
     n = n + deltnplus(b(t)*n*U/T,c,U)-d(t)*n

### 2017-03-25 simulations.py
latest

@@ -5,24 +5,24 @@ Created on Thu Mar 24 12:01:54 2016
 @author: jbertram
 """
 
-from numpy import *
-from pylab import *
+import numpy as np
+import matplotlib.pyplot as plt
 import bisect
 
 def deltnplussim(m,c,U):
     scatter=[[0 for x in range(len(m))] for y in range(U)];
     for i in range(len(m)):
         for y in xrange(m[i]):
-            cell=int(U*rand());
+            cell=int(U*np.random.rand());
             scatter[cell][i]=scatter[cell][i]+1;
     
-    winsnocomp=zeros(len(m)); wins1=zeros(len(m)); wins2=zeros(len(m));
-    comp=zeros(U);
+    winsnocomp=np.zeros(len(m)); wins1=np.zeros(len(m)); wins2=np.zeros(len(m));
+    comp=np.zeros(U);
     for i in range(U):
         comp[i]=sum(scatter[i])
         if comp[i]>0:            
-            lotterycmf=cumsum(array(scatter[i])*c)
-            victor=bisect.bisect(lotterycmf,rand()*lotterycmf[-1])
+            lotterycmf=np.cumsum(np.array(scatter[i])*c)
+            victor=bisect.bisect(lotterycmf,np.random.rand()*lotterycmf[-1])
             
             if scatter[i][victor]==1 and comp[i]==1:
                 winsnocomp[victor]=winsnocomp[victor]+1
@@ -31,56 +31,56 @@ def deltnplussim(m,c,U):
             else: 
                 wins2[victor]=wins2[victor]+1
         
-    return winsnocomp,wins1,wins2
+    return np.array([winsnocomp,wins1,wins2])
 
 def R(m,c,U):
     l=m/float(U)
     L=sum(l)
     cbar=sum(m*c)/sum(m); 
-    return cbar*exp(-l)*(1-exp(-(L-l)))\
-            /(c + (L-1+exp(-L))/(1-(1+L)*exp(-L))*(cbar*L-c*l)/(L-l))
+    return cbar*np.exp(-l)*(1-np.exp(-(L-l)))\
[…]

### 2017-04-01 simulations.py
latest

@@ -182,26 +182,30 @@ for t in range(totaltime):
 #    n = n + sum(deltnplussim((b(t)*n*U/T).astype(np.int),c,U.astype(np.int)))-d(t)*n
 #    nsimhist.append(list(n))
 
-plt.figure()
-plt.plot(nhist)
-#plt.plot(nclassichist)
-#plt.plot(nsimhist)
-plt.figure()
-plt.plot(np.array(nhist)[:,0]/np.sum(nhist,1))
-#plt.plot(np.array(nclassichist)[:,0]/np.sum(nclassichist,1))
-#plt.plot(np.array(nsimhist)[:,0]/np.sum(nsimhist,1))
-plt.ylim([0,1])
-plt.figure()
-plt.plot(np.sum(nsimhist,1))
-#plt.plot(np.sum(nclassichist,1))
-#plt.plot(np.sum(nsimhist,1))
-plt.ylim([0,T])
+fig=plt.subplot(311)
+fig.plot([b(t)[0] for t in range(totaltime)],'k',[b(t)[1] for t in range(totaltime)],'b',lw=2)
+fig.plot([d(t)[0] for t in range(totaltime)],'k--',[d(t)[1] for t in range(totaltime)],'b--',lw=2)
+fig.set_xticklabels([])
+fig.set_ylim([0,0.6])
+plt.ylabel(r"$b(t), d(t)$",fontsize=14)
+fig.annotate(r'$(a)$',xy=(0.01,0.9),xycoords='axes fraction',fontsize=12)
 
-'''
-plt.figure()
-plt.plot([b(t) for t in range(100)])
-plt.plot([d(t) for t in range(100)])
-'''
+fig=plt.subplot(312)
+plt.stackplot(range(totaltime+1),np.array(nhist)[:,1],np.array(nhist)[:,0],colors=['b','k'])
+fig.set_xticklabels([])
+fig.set_yticklabels(['0.0','0.2','0.4','0.6','0.8','1.0'])
+plt.ylabel(r"Abundance$/T$",fontsize=14)
+fig.annotate(r'$(b)$',xy=(0.01,0.9),xycoords='axes fraction',fontsize=12)
+
+fig=plt.subplot(313)
+plt.plot(np.array(nhist)[:,0]/np.sum(nhist,1),'k',lw=2)
+plt.ylim([0,1])
+plt.xlabel(r"$t$",fontsize=14)
+plt.ylabel("Frequency",fontsize=14)
[…]

### 2017-04-03 manuscript.tex
latest

@@
 
-\section*{•}
+\section*{Stable polymorphisms due to cyclical population density}
+
+We now consider the behaviour of Eq.
+\eqref{eq:master} when population density cycles periodically.
+For example, natural \textit{Drosophila}) populations undergo large seasonal density cycles, expanding rapidly in the warmer months when fruit is abundant, and being carried through the colder months by a much smaller number of dormant individuals.
+Allele frequencies at many loci [check] also cycle seasonally \citep{bergland_14}.
+
+This illustrates the density-dependent behaviour of Eq. \eqref{eq:master}.
 
@@
 
-in expanding (e.g. humans) or seasonally-cycling (e.g. \textit{Drosophila}) populations \citep{bergland_14}.
+in expanding (e.g. humans) or seasonally-cycling (e.g. .
 
  Eco-evolutionary models of selection may help to remedy this situation, since demography and selection are treated as inseparable, especially on shorter timescales, both simply being the outcome of patterns of births and deaths.
+
+\begin{figure}
+\centering
+\includegraphics[scale=1]{fluctuatingselection.pdf}
+\caption{\label{fig:axes} The realization of Grime's triangle in our model.
+Schematic representation of the triangular space bounded by the low/high extremes of stress/disturbance.
+The low-$T$ interpretation of stress is also shown.
+The vertices of the triangles correspond to different environmental extremes.
+Selection favors different traits at each vertex, leading to different trait clusters.}
+\end{figure}
 

### 2017-04-04 manuscript.tex
latest

--- 2017-03-21 manuscript.tex
@@
 
-\section*{•}
+\section*{Selection under cyclical population density}
+
+We now consider the situation when both selection and population density cycle periodically.
+For example, natural \textit{Drosophila} populations undergo seasonal density cycles, expanding rapidly in the warmer months when fruit is abundant, and being carried through the colder months by a much smaller number of dormant individuals.
+Within these density cycles, allele frequencies also cycle \citep{bergland_14}, and may potentially have been doing so for a long period of time [how many loci, how long].
+
+Figure \ref{fluctuatingselection} shows the behavior of Eq.
+\eqref{eq:master} for an example where $b$ and $d$ cycle between zero and positive values (``summers'' with rapid growth and no mortality, and ``winters'' with positive mortality and no growth).
+Selection alternatively favors the $b$ specialist
 
@@
 
-in expanding (e.g. humans) or seasonally-cycling (e.g. \textit{Drosophila}) populations \citep{bergland_14}.
+in expanding (e.g. humans) or seasonally-cycling (e.g. .
 
  Eco-evolutionary models of selection may help to remedy this situation, since demography and selection are treated as inseparable, especially on shorter timescales, both simply being the outcome of patterns of births and deaths.
+
+\begin{figure}
+\centering
+\includegraphics[scale=0.7]{fluctuatingselection.pdf}
+\caption{\label{fig:fluctuatingselection} Stable coexistence between $b$ and $d$ specialists when $b$ and $d$ cycle seasonally.
+(a) $b$ (solid lines) and $d$ (dashed lines) alternate between  zero and positive.
+The $b$ specialist (black) has higher $b$ and $d$ than the $d$ specialist.
+(b) Both types increase in abundance during the positive $b$ phase, and decline during the positive $d$ phase, but the $d$ specialist (blue area) does so at a slower rate than the $b$ specialist (black area).
+Total area (blue+black) shows the cycles in total population size $N$.
+(c) The positive $b$ and $d$ phases favor the $b$ and $d$ specialists, respectively.
+However, due to the storage effect neither has an advantage over a full environemntal cycle, and they stably coexist.}
+\end{figure}
 

### 2017-04-04 manuscript.tex
latest

--- 2017-03-21 manuscript.tex
@@
 
-\section*{•}
+\section*{Selection and population density under cyclical birth and death rates}
+
+We now consider periodically varying birth and death rates, as could occur due to seasonal cycles.
+For example, natural \textit{Drosophila} populations expand rapidly in the warmer months when fruit is abundant,  and are carried through the colder months by a much smaller number of dormant individuals.
+Population density undergoes seasonal cycles, and within these cycles, hundreds of polymorphisms also cycle in frequency, some of which may potentially have been doing so for millions of years \citep{bergland_14}.
+
+Figure \ref{fluctuatingselection} shows the behavior of Eq.
+\eqref{eq:master} for an example where $b$ and $d$ cycle between zero and positive values (``summers'' with rapid growth and no mortality, and ``winters'' with positive mortality and no growth).
+
+Selection alternatively favors the $b$ specialist
+
+. However, due to the storage effect neither has an advantage over a full environemntal cycle, and
 
@@
 
-in expanding (e.g. humans) or seasonally-cycling (e.g. \textit{Drosophila}) populations \citep{bergland_14}.
+in expanding (e.g. humans) or seasonally-cycling (e.g. .
 
  Eco-evolutionary models of selection may help to remedy this situation, since demography and selection are treated as inseparable, especially on shorter timescales, both simply being the outcome of patterns of births and deaths.
+
+\begin{figure}
+\centering
+\includegraphics[scale=0.7]{fluctuatingselection.pdf}
+\caption{\label{fig:fluctuatingselection} Stable coexistence between $b$ and $d$ specialists when $b$ and $d$ fluctuate.
+(a) $b$ (solid lines) and $d$ (dashed lines) alternate being nonzero.
+The $b$ specialist (black) has higher $b$ and $d$ than the $d$ specialist (blue).
+(b) Both types grow during the positive $b$ phase, and decline during the positive $d$ phase, but the $d$ specialist (blue) does so at a slower rate than the $b$ specialist (black).
+Total height (blue+black) is population size $N$.
+(c) The positive $b$ and $d$ phases favor the $b$ and $d$ specialists, respectively, and they stably coexist due to the storage effect.}
+\end{figure}
 

### 2017-04-04 simulations.py

@@ -164,16 +164,27 @@ def d(t):
 c=np.array([1.,1.])
 T=100000
 g=40.     #iterations per seasonal cycle
-bparam=np.array([[0.0,0.0],[.5,.22]])
+bparam=np.array([[0.0,0.0],[.5,.0423]])
 dparam=np.array([[0.0,0.0],[0.2,0.1]])
-totaltime=200
+totaltime=10000
 
+#n=np.array([60000.,25000.])
+#nhist=[n];
+#for t in range(totaltime):
+#    U=T-sum(n)
+#    n = n + deltnplus((b(t)*n*U/T).astype(np.int),c,U.astype(np.int))-d(t)*n
+#    nhist.append(list(n))
+    
 n=np.array([60000.,25000.])
-nhist=[n];
+nhistclassic=[n];
 for t in range(totaltime):
     U=T-sum(n)
-    n = n + deltnplus((b(t)*n*U/T).astype(np.int),c,U.astype(np.int))-d(t)*n
-    nhist.append(list(n))
+    n = n + deltnplusclassic((b(t)*n*U/T).astype(np.int),c,U.astype(np.int))-d(t)*n
+    nhistclassic.append(list(n))
+    
+#plt.stackplot(range(totaltime+1),np.array(nhistclassic)[:,1],np.array(nhistclassic)[:,0],colors=['b','k'])
+plt.figure()
+plt.plot(np.array(nhistclassic)[:,0]/np.sum(nhistclassic,1),'k',lw=2)
     
 #n=np.array([6000.,2500.])
 #nsimhist=[n]

### 2017-04-05 manuscript.tex
latest

@@
 
-We now consider periodically varying birth and death rates, as could occur due to seasonal cycles.
-For example, natural \textit{Drosophila} populations expand rapidly in the warmer months when fruit is abundant,  and are carried through the colder months by a much smaller number of dormant individuals.
-Population density undergoes seasonal cycles, and within these cycles, hundreds of polymorphisms also cycle in frequency, some of which may potentially have been doing so for millions of years \citep{bergland_14}.
+Sections ``'' and ``'' only depend on the behaviour of low-frequency invaders in Eq. \eqref{eq:master}. Here we consider an example of its full time-dendent behaviour.
+
+We consider the situation where birth and death rates vary periodically with amplitude sufficent to cause large changes in population density.
+For example, natural \textit{Drosophila} populations expand rapidly in the warmer months when fruit is abundant, but largely die off in the colder months.
+Within this seasonal population density cycle, hundreds of polymorphisms also cycle in frequency, some of which may be adaptive, and potentially millions of years old \citep{bergland_14}.
+In this situation, selection on allele frequencies occurs on the same time scale as population demography.
+
+In the classical population genetic treatment of fluctuating selection, environmental fluctuations do not promote coexistence.
+Allele frequencies are successively multiplied by relative fitness values for each environmental iteration, and so two alleles favored in different enviroments can only stably coexist if this product for  one allele exactly equals the product for the other; frequency dependent selection or heterozygote advantage are still required for coexistence [cite dempster].
+This argument implicitly assumes non-overlapping generations.
+If generations overlap and competition is restricted to juvenile recruitment (``viability selection'' in population genetic terminology), then environmental fluctuations do promote coexistence as a result of the storage effect [cite ellner and hairston].
 
@@
 \eqref{eq:master} for an example where $b$ and $d$ cycle between zero and positive values (``summers'' with rapid growth and no mortality, and ``winters'' with positive mortality and no growth).
+Two types are present, a $b$ specialist, which is better at rapidly growing in the summer (higher $b$), and a $d$ specialist which is better at surviving the winter (lower $d$).
+Due to the storage effect, neither type has an advantage over a full environmental cycle, and $b$ and $d$ types can stably coexist (which is not possible without environmental fluctuations - section ``'').
 
-Selection alternatively favors the $b$ specialist
-
-. However, due to the storage effect neither has an advantage over a full environemntal cycle, and
+While
 
@@
 in expanding (e.g. humans) or seasonally-cycling (e.g. .
-
- Eco-evolutionary models of selection may help to remedy this situation, since demography and selection are treated as inseparable, especially on shorter timescales, both simply being the outcome of patterns of births and deaths.
 

### 2017-04-05 manuscript.tex
latest

@@
 In the classical population genetic treatment of fluctuating selection, environmental fluctuations do not promote coexistence.
-Allele frequencies are successively multiplied by relative fitness values for each environmental iteration, and so two alleles favored in different enviroments can only stably coexist if this product for  one allele exactly equals the product for the other; frequency dependent selection or heterozygote advantage are still required for coexistence [cite dempster].
-This argument implicitly assumes non-overlapping generations.
-If generations overlap and competition is restricted to juvenile recruitment (``viability selection'' in population genetic terminology), then environmental fluctuations do promote coexistence as a result of the storage effect [cite ellner and hairston].
+Allele frequencies are successively multiplied by relative fitness values for each environmental iteration, and so two alleles favored in different enviroments can only stably coexist if this product for  one allele exactly equals the product for the other.
+This is exceedingly unlikely unless the  conditions for coexistence in a constant environment --- frequency dependent selection or heterozygote advantage --- are met [cite dempster].
+However, the classical argument implicitly assumes non-overlapping generations.
+If generations overlap and competition is restricted to juvenile recruitment (``viability selection'' in population genetic terminology), then environmental fluctuations via the storage effect [cite ellner and hairston].
 
-Figure \ref{fluctuatingselection} shows the behavior of Eq.
-\eqref{eq:master} for an example where $b$ and $d$ cycle between zero and positive values (``summers'' with rapid growth and no mortality, and ``winters'' with positive mortality and no growth).
+Figure \ref{fig:fluctuatingselection} shows the behavior of Eq.
+\eqref{eq:master} for an example where $b$ and $d$ cycle between zero and positive values (``summers'' with rapid growth and no mortality, and ``winters'' with mortality and no growth).
 Two types are present, a $b$ specialist, which is better at rapidly growing in the summer (higher $b$), and a $d$ specialist which is better at surviving the winter (lower $d$).
-Due to the storage effect, neither type has an advantage over a full environmental cycle, and $b$ and $d$ types can stably coexist (which is not possible without environmental fluctuations - section ``'').
-
-While
+Neither type has an advantage over a full environmental cycle, and they stably coexist.
+Stable coexistence was not possible in a constant environment (section ``''), and occurs in Fig.
+\ref{fig:fluctuatingselection} due to the storage effect.
 
 Current approaches to population genetic inference treat demography and selection separately, typically fitting a demographic model to data on ostensibly neutral sites, and then inferring selection against this demographic background, despite the recurrent problem that the two are confounded \citep{schrider_2016}.
-
-in expanding (e.g. humans) or seasonally-cycling (e.g. .
+in expanding (e.g.
+humans) or seasonally-cycling (e.g.
+.
 

### 2017-04-05 simulations.py

@@ -166,14 +166,15 @@ T=100000
 g=40.     #iterations per seasonal cycle
 bparam=np.array([[0.0,0.0],[.5,.0423]])
 dparam=np.array([[0.0,0.0],[0.2,0.1]])
-totaltime=10000
+totaltime=100000
 
-#n=np.array([60000.,25000.])
-#nhist=[n];
-#for t in range(totaltime):
-#    U=T-sum(n)
-#    n = n + deltnplus((b(t)*n*U/T).astype(np.int),c,U.astype(np.int))-d(t)*n
-#    nhist.append(list(n))
+n=np.array([60000.,25000.])
+nhist=[n];
+for t in range(totaltime):
+    U=T-sum(n)
+    #n = n + deltnplus((b(t)*n*U/T).astype(np.int),c,U.astype(np.int))-d(t)*n
+    n = n + deltnplus((b(t)*n).astype(np.int),c,U.astype(np.int))-d(t)*n
+    nhist.append(list(n))
     
 n=np.array([60000.,25000.])
 nhistclassic=[n];
@@ -181,10 +182,10 @@ for t in range(totaltime):
     U=T-sum(n)
     n = n + deltnplusclassic((b(t)*n*U/T).astype(np.int),c,U.astype(np.int))-d(t)*n
     nhistclassic.append(list(n))
-    
-#plt.stackplot(range(totaltime+1),np.array(nhistclassic)[:,1],np.array(nhistclassic)[:,0],colors=['b','k'])
-plt.figure()
+
+plt.stackplot(range(totaltime+1),np.array(nhistclassic)[:,1],np.array(nhistclassic)[:,0],colors=['b','k'])
 plt.plot(np.array(nhistclassic)[:,0]/np.sum(nhistclassic,1),'k',lw=2)
+    
     
 #n=np.array([6000.,2500.])
 #nsimhist=[n]

### 2017-04-08 manuscript.tex
latest

@@
 We instead use the following approximation.
-Rather than evaluating the situation in each territory after dispersal as above, we replace $\tilde{p}$ by $\tilde{q}$, defined as the ${\mathbf x_i}$ Poisson dispersal probabilities conditional on $X_i\geq1$, independently of the outcome of $x_i$.
-This gives $\langle x_j \rangle_{\tilde{q}}=\langle x_j \rangle_p/C=l_j/C$,
+Rather than evaluating the situation in each territory after dispersal as above, we replace $\tilde{p}$ with $\tilde{q}$, defined as the ${\mathbf x_i}$ Poisson dispersal probabilities conditional on $X_i\geq1$ (which are independent of the outcome of $x_i$).
+The distinction between $\tilde{p}$ with $\tilde{q}$ will be discussed further below.
+The moments of $\tilde{q}$ are simpler, but quantitatively similar enough to $\tilde{p}$'s moments that they can be used for analyzing the relative fluctuations in $\sum_{j\neq i} c_j x_j$.
+The $\tilde{q}$ approximation gives $\langle x_j \rangle_{\tilde{q}}=\langle x_j \rangle_p/C=l_j/C$,
 \begin{align}
@@
 where $C=1-e^{-(L-l_i)}$ and $j\neq k$.
-The distribution $\tilde{q}$ only approximates the situation after dispersal, since knowing that one focal genotype is among the propagules present restricts the possible outcomes for the $x_j$, so that the $x_j$ cannot strictly be treated as independent of $x_i$.
-This seemingly minor distinction has meaningful consequences.
-To illustrate, suppose that the focal genotype is rare and the propagule density is high ($l_j \approx L\gg 1$).
-Then Eq.
-\eqref{eq:meanxjrare} correctly predicts that there are on average $L-1$ nonfocal propagules $\langle x_j \rangle_{\tilde{p}}\approx L-1$, with the focal propagule correctly excluded, whereas $\tilde{q}$ predicts one extra $\langle x_j \rangle_{\tilde{q}}\approx L$.
-As a result, $\tilde{q}$ gives pathological behavior for rare invaders (they have a rarity disadvantage), but its moments are quantitatively similar enough to those of $\tilde{p}$ that it is sufficient for analyzing the relative fluctuations in $\sum_{j\neq i} c_j x_j$.
+
+The exact distribution $\tilde{p}$ assumes that exactly one of the  propagules present in a given site after dispersal belongs to the focal type, whereas $\tilde{q}$ assumes that there is a focal propagule present before non-focal dispersal commences.
+As a result, $\tilde{q}$ erroneously predicts that the mean propagule density is greater than $L$ in sites with only one focal propagule is present when the focal type is rare and the propagule density is high.
+Specifically, if $L-l_i \approx L\gg 1$, then the mean propagule density predicted by $\tilde{q}$ is approximately $L+1$.
+This is erroneous because by definition the mean number of propagules in every site is $L$.
+The discrepancy causes rare invaders to have an intrinsic rarity disadvantage (territorial contests under $\tilde{q}$ are more intense than they should be).
+In contrast, Eq.
+\eqref{eq:meanxjrare} correctly predicts that there are on average $\sum_{j\neq i}\langle x_j \rangle_{\tilde{p}}\approx L-1$ nonfocal propagules.
 

### 2017-04-11 manuscript.tex
latest

@@
 
-We will show that, with respect to $\tilde{p}$, the standard deviation $\sigma_{\tilde{p}}(\sum_{j\neq i} c_j x_j)$, is much smaller than $\langle\sum_{j\neq i} c_j x_j\rangle_{\tilde{p}}$.
-Then $x_j$ can be replaced by its mean in the last term in Eq.~\eqref{eq:deltr},
+Our ``mean field'' approximation is to replace $x_j$ with its mean in the last term in Eq.~\eqref{eq:deltr},
 \begin{equation}
@@
 \end{equation}
+Below we justify this replacement by arguing that the standard deviation $\sigma_{\tilde{p}}(\sum_{j\neq i} c_j x_j)$ (with respect to $\tilde{p}$), is much smaller than $\langle\sum_{j\neq i} c_j x_j\rangle_{\tilde{p}}$.
 
@@
 
-For analyzing the relative fluctuations in $\sum_{j\neq i} c_j x_j$, Eq.
-\eqref{eq:meanxjrare} is unnecessarily complicated.
-We instead use the following approximation.
-Rather than evaluating the situation in each territory after dispersal as above, we replace $\tilde{p}$ with $\tilde{q}$, defined as the ${\mathbf x_i}$ Poisson dispersal probabilities conditional on $X_i\geq1$ (which are independent of the outcome of $x_i$).
+The exact analysis of the fluctuations in $\sum_{j\neq i} c_j x_j$ is complicated because the $x_j$ are not independent with respect to $\tilde{p}$.
+These fluctuations are part of the ``drift'' in type abundances which we leave for future work.
+Here we use the following approximation to give some insight into the magnitude of these fluctuations and also the nature of the correlations between the $x_j$.
+We replace $\tilde{p}$ with $\tilde{q}$, defined as the ${\mathbf x_i}$ Poisson dispersal probabilities conditional on $X_i\geq1$ (which are independent).
 The distinction between $\tilde{p}$ with $\tilde{q}$ will be discussed further below.
-The moments of $\tilde{q}$ are simpler, but quantitatively similar enough to $\tilde{p}$'s moments that they can be used for analyzing the relative fluctuations in $\sum_{j\neq i} c_j x_j$.
 The $\tilde{q}$ approximation gives $\langle x_j \rangle_{\tilde{q}}=\langle x_j \rangle_p/C=l_j/C$,
@@
 The exact distribution $\tilde{p}$ assumes that exactly one of the  propagules present in a given site after dispersal belongs to the focal type, whereas $\tilde{q}$ assumes that there is a focal propagule present before non-focal dispersal commences.
-As a result, $\tilde{q}$ erroneously predicts that the mean propagule density is greater than $L$ in sites with only one focal propagule is present when the focal type is rare and the propagule density is high.
+As a result, $\tilde{q}$ predicts that the mean propagule density is greater than $L$ (in sites with only one focal propagule is present) when the focal type is rare and the propagule density is high.
+This is erroneous, because the mean number of propagules in every site is $L$ by definition.
 Specifically, if $L-l_i \approx L\gg 1$, then the mean propagule density predicted by $\tilde{q}$ is approximately $L+1$.
-This is erroneous because by definition the mean number of propagules in every site is $L$.
 The discrepancy causes rare invaders to have an intrinsic rarity disadvantage (territorial contests under $\tilde{q}$ are more intense than they should be).
 In contrast, Eq.
-\eqref{eq:meanxjrare} correctly predicts that there are on average $\sum_{j\neq i}\langle x_j \rangle_{\tilde{p}}\approx L-1$ nonfocal propagules.
+\eqref{eq:meanxjrare} correctly predicts that there are on average $\sum_{j\neq i}\langle x_j \rangle_{\tilde{p}}\approx L-1$ nonfocal propagules because $\tilde{p}$ accounts for potentially large negative covariances between the $x_j$ ``after dispersal''.
+By neglecting the latter covariences, $\tilde{q}$ overestimates the fluctuations in $\sum_{j\neq i} c_j x_j$; thus $\tilde{q}$ gives an upper bound on the fluctuations.
+The discrepancy between $\tilde{q}$ and $\tilde{p}$ will be largest when $L$ is of order $1$ or smaller, because then the propagule assumed to already be present under $\tilde{q}$ is comparable to, or greater than, the entire propgaule density.
 
@@
 \begin{equation}
-\sigma_{\tilde{q}}^2(\sum_{j\neq i} c_j x_j)=\sum_{j\neq i}\left[c_j^2\sigma_{\tilde{q}}^2(x_j)+2\sum_{k>j}c_j c_k\sigma_{\tilde{q}}(x_j,x_k)\right],\label{eq:vartotr}
+\sigma_{\tilde{q}}^2(\sum_{j\neq i} c_j x_j)=\sum_{j\neq i}\left[c_j^2\sigma_{\tilde{q}}^2(x_j)+2\sum_{k>j, k\neq i}c_j c_k\sigma_{\tilde{q}}(x_j,x_k)\right],\label{eq:vartotr}
 \end{equation}
@@
 \begin{equation}
-\frac{\sigma(\sum_{j\neq i} c_j x_j)}{\langle\sum_{j\neq i} c_j x_j\rangle}<C^{1/2}\frac{\left(\sum_{j\neq i}c_j^2 l_j\right)^{1/2}}{\sum_{j\neq i}c_j l_j}. \label{eq:cvr}
+\frac{\sigma(\sum_{j\neq i} c_j x_j)}{\langle\sum_{j\neq i} c_j x_j\rangle}=C^{1/2}\frac{\left(\sum_{j\neq i}c_j^2 l_j+(1-1/C)\left(\sum_{j\neq i}c_j l_j\right)^2 \right)^{1/2}}{\sum_{j\neq i}c_j l_j}<C^{1/2}\frac{\left(\sum_{j\neq i}c_j^2 l_j\right)^{1/2}}{\sum_{j\neq i}c_j l_j}.
+\label{eq:cvr}
 \end{equation}
 
-Without loss of generality, we restrict attention to the case that the total nonfocal density $L-l_i$ is of order $1$ or larger (otherwise $\Delta_r n_i$ does not contribute significantly to $\Delta_+ n_i$ because $\Delta_r n_i$ is proportional to $C=1-e^{-(L-l_i)}$).
-
-Then, when at least some of the nonfocal propagule densities are large $l_j\gg 1$, the RHS of Eq.~\eqref{eq:cvr} is $\ll 1$, as desired.
-This is also the case if none of the nonfocal genotype densities are large and the $c_j$ are all of similar magnitude (their ratios are of order one); the worst case scenario occurs when $L-l_i\sim O(1)$, in which case the negative covariances (Eq.~\eqref{eq:covr}) which were neglected in the RHS of Eq.~\eqref{eq:cvr} significantly reduce the overall variance $\sigma_{\tilde{q}}^2(\sum_{j\neq i} c_j x_j)$.
+Suppose that the $c_j$ are all of similar magnitude (their ratios are of order one).
+Then Eq.~\eqref{eq:cvr} is $\ll 1$ for the case when $L-l_i \ll 1$ (due to the factor of $C^{1/2}$), and also for the case when at least some of the nonfocal propagule densities are large $l_j\gg 1$ (since it is then of order $1/\sqrt{L-l_i}$).
+The worst case scenario occurs when $L-l_i$ is of order one.
+Then Eq.~\eqref{eq:cvr} gives a relative error of approximately $50\%$, which from our earlier discussion we know to be a substantial overestimate when $L$ is of order $1$.
+Our numerical results (Fig.
+\ref{fig:simcomp}) confirm that the relative errors are indeed small.
 
@@
 
-Again, we show that the relative fluctuations in $\sum c_j x_j$ are much smaller than $1$ (with respect to $\hat{p}$), so that,
+Again, we argue that the relative fluctuations in $\sum c_j x_j$ are much smaller than $1$ (with respect to $\hat{p}$), so that,
 \begin{equation}
@@
 
-To calculate the relative fluctuations in $\sum_{j\neq i} c_j x_j$, we use a similar approximation as for $\Delta_r n_i$: $\hat{p}$ is approximated by $\hat{q}$, defined as the ${\mathbf x}$ dispersal probabilities in a territory conditional on $x_i>2$ (that is, treating the $x_j$ as independent of $x_i$).
-All covariances between nonfocal genotypes are now zero, so that $\sigma^2(\sum c_j x_j)=\sum c_j^2 \sigma^2(x_j)$, where $\sigma^2(x_j)=l_j$ for $j\neq i$.
-The expression for $\sigma^2(x_i)$ is more complicated, but in the relevant regime where $p(x_i=0)\approx 0$ (since otherwise $D\gg 1$ and $\Delta n_a$ is negligible), then
+To calculate the relative fluctuations in $\sum_{j\neq i} c_j x_j$, we use a similar approximation as for $\Delta_r n_i$: $\hat{p}$ is approximated by $\hat{q}$, defined as the ${\mathbf x}$ dispersal probabilities in a territory conditional on $x_i>2$ (that is, treating the $x_j$ as indepenent).
+All covariances between nonfocal genotypes are now zero, so that $\sigma_{\hat{q}}^2(\sum c_j x_j)=\sum c_j^2 \sigma_{\hat{q}}^2(x_j)$, where $\sigma_{\hat{q}}^2(x_j)=l_j$ for $j\neq i$, and
 \begin{equation}
-\sigma_{\hat{q}}^2(x_i)\approx\frac{l_i^2}{D}\left(1-\frac{1}{D}\right)+\frac{l_i}{D},
+\sigma_{\hat{q}}^2(x_i)=\frac{l_i}{D}\left(l_i+1-e^{-l_i}-\frac{l_i}{D}\left(1-e^{-l_i}\right)^2\right),
 \end{equation}
-where $D= 1-(1+l_i)e^{-l_i}$, analogous to Eq.~\eqref{eq:varr}, and 
+where $D= 1-(1+l_i)e^{-l_i}$, and 
 \begin{equation}
-\frac{\sigma_{\hat{q}}(\sum c_j x_j)}{\langle\sum c_j x_j\rangle} \approx\frac{\left(\sum_{j\neq i} c_j^2 l_j + c_i^2 \sigma_{\hat{q}}^2(x_i)\right)^{1/2}}{\sum_{j\neq i} c_j l_j + c_i l_i/D} \label{eq:cva}.
+\frac{\sigma_{\hat{q}}(\sum c_j x_j)}{\langle\sum c_j x_j\rangle} = \frac{\left(\sum_{j\neq i} c_j^2 l_j + c_i^2 \sigma_{\hat{q}}^2(x_i)\right)^{1/2}}{\sum_{j\neq i} c_j l_j + c_i l_i (1-e^{-l_i})/D} \label{eq:cva}.
 \end{equation}
 
+Similarly to Eq.~\eqref{eq:cvr}, the RHS of Eq.
+\eqref{eq:cva} is $\ll 1$ for the case that $L \ll 1$ (due to a factor of $D^{1/2}$), and also for the case when at least some of the propagule densities (focal or nonfocal) are large --- provided that $c_i$ and the $c_j$ are all of similar magnitude.
+Again, the worst case scenario occurs when $l_i$ and $L-l_i$ are of order $1$, in which case Eq.
+\eqref{eq:cva} is around $35\%$, which is again where the $\hat{q}$ approximation produces the biggest overestimate of the fluctuations in ${\mathbf x}$.
 Similarly to Eq.~\eqref{eq:cvr}, the RHS of \eqref{eq:cva} will not be $\ll 1$ in the presence of a rare, extremely strong competitor.
-When this is not the case, then since $l_i$ must be of order $1$ or larger for $\Delta_a n$ to make an appreciable contribution to $\Delta_+ n_i$, the RHS of Eq.~\eqref{eq:cva} is $\ll 1$ as desired.
 

### 2017-04-11 manuscript.tex
latest

@@
 Ewens, Mathematical Population Genetics I, 2004
-
-\section*{Introduction}
 
@@
 
-\section*{Selection and population density under cyclical birth and death rates}
+\section*{Cyclical birth and death rates}
 
-Sections ``'' and ``'' only depend on the behaviour of low-frequency invaders in Eq. \eqref{eq:master}. Here we consider an example of its full time-dendent behaviour.
-
-We consider the situation where birth and death rates vary periodically with amplitude sufficent to cause large changes in population density.
-For example, natural \textit{Drosophila} populations expand rapidly in the warmer months when fruit is abundant, but largely die off in the colder months.
-Within this seasonal population density cycle, hundreds of polymorphisms also cycle in frequency, some of which may be adaptive, and potentially millions of years old \citep{bergland_14}.
-In this situation, selection on allele frequencies occurs on the same time scale as population demography.
+So we have only considered the low-frequency invasion behavior of Eq.
+\eqref{eq:master}.
+Here we start exploring its full time-dependent behaviour.
+We consider a situation where birth and death rates vary periodically with amplitude sufficent to cause large changes in population density.
+This is inspired by natural \textit{Drosophila} populations, which expand rapidly in the warmer months when fruit is abundant, but largely die off in the colder months.
+Within this seasonal population density cycle, hundreds of polymorphisms also cycle in frequency \citep{bergland_14}.
+Some of these polymorphisms may be adaptive and potentially millions of years old \citep{bergland_14}.
+In this situation, selection on allele frequencies occurs on the same time scale as population demography, a considerably more complicated scenario than classical sweeps in demographically stable populations \citep{messer_2016}.
 
 In the classical population genetic treatment of fluctuating selection, environmental fluctuations do not promote coexistence.
-Allele frequencies are successively multiplied by relative fitness values for each environmental iteration, and so two alleles favored in different enviroments can only stably coexist if this product for  one allele exactly equals the product for the other.
-This is exceedingly unlikely unless the  conditions for coexistence in a constant environment --- frequency dependent selection or heterozygote advantage --- are met [cite dempster].
-However, the classical argument implicitly assumes non-overlapping generations.
-If generations overlap and competition is restricted to juvenile recruitment (``viability selection'' in population genetic terminology), then environmental fluctuations via the storage effect [cite ellner and hairston].
+Allele frequencies are successively multiplied by relative fitness values for each environmental iteration, and so two alleles favored in different enviroments can only stably coexist if the product of fitnesses for one type exactly equals the product for the other \citep{dempster_1955}.
+Thus, stable coexistence still requires frequency dependent selection or heterozygote advantage (as is required in a constant environment).
+
+One general mechanism which by which fluctuating selection promotes coexistence is the ``storage effect'', which occurs when some individuals are protected from selection, and is would appear as a form of frequency dependent selection in the classical analysis of \cite{dempster_1955}.
+
+.
+In the lottery model, a certain fraction of the adult population ($(1-d_i)n_i$ of each type) does not experience selection in a given iteration.
+This favors rare types more than abundant ones, because abundant types
 
@@
 
-In Fig.
+%In Fig.
 \ref{fig:table2} we compare our model with some of the other models and schemes touched upon here.
@@
 
-\begin{figure}
-\centering
-\begin{tabularx}{\linewidth}{lXXXXX}
-&Formal model?  & Ecologically meaningful? & Empirically-grounded trait scheme? & Generality beyond specific scenarios? & Genetically flexible? \\ \hline
-Density-dependent lottery & \ding{51} & \ding{51} & \ding{51} & \ding{51} & \ding{51} \\  
-  MacArthur's $r$/$K$ + $\alpha$ & \ding{51} & \ding{51} & \ding{55}* & \ding{51} & \ding{51} \\
-  Grime's C/S/R & \ding{55} & \ding{51} & \ding{51} & \ding{51} & NA \\
-  Traditional pop. gen. & \ding{51} & \ding{55} & \ding{55} & \ding{51} & \ding{51} \\
-  Eco-evo:  \\
-  \quad Adaptive dynamics & \ding{51} & \ding{51} & \ding{51} & \ding{51}** & \ding{55} \\
-  \quad Brute force simulation & \ding{51} & \ding{51} & NA & \ding{55} & \ding{51}
-\end{tabularx}
-\caption{\label{fig:table2} Comparison of our density-dependent lottery model with related models and schemes in ecology and evolutionary biology.
+%\begin{figure}
+%\centering
+%\begin{tabularx}{\linewidth}{lXXXXX}
+%&Formal model?  & Ecologically meaningful? & Empirically-grounded trait scheme? & Generality beyond specific scenarios? & Genetically flexible? \\ \hline
+%Density-dependent lottery & \ding{51} & \ding{51} & \ding{51} & \ding{51} & \ding{51} \\  
+%  MacArthur's $r$/$K$ + $\alpha$ & \ding{51} & \ding{51} & \ding{55}* & \ding{51} & \ding{51} \\
+%  Grime's C/S/R & \ding{55} & \ding{51} & \ding{51} & \ding{51} & NA \\
+%  Traditional pop. gen. & \ding{51} & \ding{55} & \ding{55} & \ding{51} & \ding{51} \\
+%  Eco-evo:  \\
+%  \quad Adaptive dynamics & \ding{51} & \ding{51} & \ding{51} & \ding{51}** & \ding{55} \\
+%  \quad Brute force simulation & \ding{51} & \ding{51} & NA & \ding{55} & \ding{51}
+%\end{tabularx}
+%\caption{\label{fig:table2} Comparison of our density-dependent lottery model with related models and schemes in ecology and evolutionary biology.
 *MacArthur's $r$- and $K$-, as well as $\alpha$ selection, were all derived theoretically.
@@
 **In practice, most of the adaptive dynamics literature focuses on specific eco-evolutionary outcomes such as evolutionary ``branching'' \citep{geritz_1997} or mutualisms \citep{ferriere_2002}, but in principle it can be applied with any fitness model including our density-dependent lottery.}
-\end{figure}
+%\end{figure}
 

### 2017-04-12 manuscript.tex
latest

@@
 
-In the classical population genetic treatment of fluctuating selection, environmental fluctuations do not promote coexistence.
+The classical population genetic treatment of fluctuating selection suggests that environmental fluctuations do not promote coexistence.
 Allele frequencies are successively multiplied by relative fitness values for each environmental iteration, and so two alleles favored in different enviroments can only stably coexist if the product of fitnesses for one type exactly equals the product for the other \citep{dempster_1955}.
@@
 
-One general mechanism which by which fluctuating selection promotes coexistence is the ``storage effect'', which occurs when some individuals are protected from selection, and is would appear as a form of frequency dependent selection in the classical analysis of \cite{dempster_1955}.
+This classical argument overlooks two general mechanisms by which fluctuating selection promotes coexistence.
+The first is the ``storage effect'', which introduces a form of frequency dependent selection that promotes coexistence in the presence of environmental fluctuations but not in a constant environment.
+The storage effect occurs when some individuals are protected from selection.
+This is the case in both the classic lottery model and our extension of it: a fraction $(1-d_i)n_i$ of each type's adults do not experience selection in a given iteration.
+Protection from selection favors rare types in fluctuating environments because abundant types cannot fully exploit environmental periods that favor them (since only some of the rare type can be displaced), whereas rare types gain the full benefits of their favorable periods (far more abundant type individuals can be displaced than the rare types could possibly displace) \citep{chesson_1981}.
 
-.
-In the lottery model, a certain fraction of the adult population ($(1-d_i)n_i$ of each type) does not experience selection in a given iteration.
-This favors rare types more than abundant ones, because abundant types
+The second mechanism we will call the ``bounded density effect'', since it is a consequence of the inhibition of reproduction at high population densities (Dempster's (1955) argument ignores density-dependent effects).
+If there is growth from low to high density in each environmental cycle, then types that are abundant determine the time available for growth every cycle, with less time spent growing in cycles where they are favored, and more in cycles where they are not.
+This promotes coexistence even in the absence of frequency dependent selection \citep{yi_2013}.
 
@@
 \eqref{eq:master} for an example where $b$ and $d$ cycle between zero and positive values (``summers'' with rapid growth and no mortality, and ``winters'' with mortality and no growth).
+Both the storage effect (winter survivors are protected from selection when summer starts) and the bounded density effect (expansion to high density occurs every cycle) are present.
 Two types are present, a $b$ specialist, which is better at rapidly growing in the summer (higher $b$), and a $d$ specialist which is better at surviving the winter (lower $d$).
 Neither type has an advantage over a full environmental cycle, and they stably coexist.
-Stable coexistence was not possible in a constant environment (section ``''), and occurs in Fig.
-\ref{fig:fluctuatingselection} due to the storage effect.
+Stable coexistence between $b$ and $d$ specialists was not possible in a constant environment (section ``''), and is facilitated by some combination of the storage and bounded population size effects.
 

### 2017-04-13 manuscript.tex
latest

@@
 276]{ewens_2004}.
+The emphasis here is on predicting how the course of allele frequencies over time depends on their selective effect, genetic drift and linkage.
 However, the resulting picture of evolution does not include even basic elements of the ecological underpinnings of selection, including dynamic population size and density-dependence.
@@
 This approach is powerful enough to model eco-evolutionary feedbacks between co-evolving traits, but is generally problem-specific and restricted to only a few traits at a time.
+These models emphasize invasion from low frequencies and co-existence more than frequency trajectories over time, with the entire adaptive dynamics approach revolving around ``invasion fitness'' \citep{diekmann_2004}.
 
@@
 Nevertheless, there are strong indications there exist broader principles governing the operation of selection.
+
 In many groups of organisms, including corals \citep{darling_2012}, insects \citep{southwood_1977}, fishes \citep{winemiller_1992}, zooplankton \citep{allan_76} and plants \citep{grime_1988,westoby_1998}, different species can be divided into a small number of distinct trait clusters corresponding to fundamentally distinct ``primary strategies'' \citep{winemiller_2015}.
+
 The most famous example is Grime's plant trait classification scheme \citep{grime_1974,grime_1977,grime_1988}.
@@
 Here we start exploring its full time-dependent behaviour.
-We consider a situation where birth and death rates vary periodically with amplitude sufficent to cause large changes in population density.
+The broader context for this section is the recurring difficulties in population genetic inference caused by the confounding of population demography and selection --- selective sweeps generate similar genealogies to population bottlenecks \citep{barton_1998}.
+Current approaches attempt to treat selection and demography separately, typically fitting a demographic model to data on ostensibly neutral sites and then inferring selection against this demographic background, but with mixed success due to linkage \citep{schrider_2016}.
+Fitness models like Eq.
+\eqref{eq:master} may help to address to these difficulties since they do not artificially separate the dynamics of abundance and frequency.
+
+Here we will not directly address these inference challenges, but will instead illustrate the complicated time-dependent behavior that can arise from Eq.
+\eqref{eq:master} in an example where birth and death rates vary periodically with amplitude sufficent to cause large changes in population density.
 This is inspired by natural \textit{Drosophila} populations, which expand rapidly in the warmer months when fruit is abundant, but largely die off in the colder months.
 Within this seasonal population density cycle, hundreds of polymorphisms also cycle in frequency \citep{bergland_14}.
-Some of these polymorphisms may be adaptive and potentially millions of years old \citep{bergland_14}.
-In this situation, selection on allele frequencies occurs on the same time scale as population demography, a considerably more complicated scenario than classical sweeps in demographically stable populations \citep{messer_2016}.
+Some of these polymorphisms may be adaptive and potentially millions of years old, suggesting stable coexistence \citep{bergland_14,messer_2016}.
+Selection on allele frequencies thus occurs on the same time scale as population demography, a situation vastly more complicated than classical sweeps in demographically stable populations \citep{messer_2016}.
 
@@
 The first is the ``storage effect'', which introduces a form of frequency dependent selection that promotes coexistence in the presence of environmental fluctuations but not in a constant environment.
-The storage effect occurs when some individuals are protected from selection.
-This is the case in both the classic lottery model and our extension of it: a fraction $(1-d_i)n_i$ of each type's adults do not experience selection in a given iteration.
-Protection from selection favors rare types in fluctuating environments because abundant types cannot fully exploit environmental periods that favor them (since only some of the rare type can be displaced), whereas rare types gain the full benefits of their favorable periods (far more abundant type individuals can be displaced than the rare types could possibly displace) \citep{chesson_1981}.
+The storage effect occurs when some individuals are protected from selection; in the lottery model a fraction $(1-d_i)n_i$ of each type's adults do not experience selection in a given iteration.
+Protection from selection promotes coexistence in fluctuating environments because abundant types cannot fully exploit environmental periods that favor them (since only a fraction of the rare type can be displaced), whereas rare types gain the full benefits of their favorable periods (far more adults from the abundant type die than can possibly be replaced by the rare types) \citep{chesson_1981}.
 
@@
 \eqref{eq:master} for an example where $b$ and $d$ cycle between zero and positive values (``summers'' with rapid growth and no mortality, and ``winters'' with mortality and no growth).
-Both the storage effect (winter survivors are protected from selection when summer starts) and the bounded density effect (expansion to high density occurs every cycle) are present.
+Both the storage effect (adults are sheltered from selection during the summer growth phase) and the bounded density effect (expansion to high density occurs every cycle) are operating.
 Two types are present, a $b$ specialist, which is better at rapidly growing in the summer (higher $b$), and a $d$ specialist which is better at surviving the winter (lower $d$).
+
 Neither type has an advantage over a full environmental cycle, and they stably coexist.
-Stable coexistence between $b$ and $d$ specialists was not possible in a constant environment (section ``''), and is facilitated by some combination of the storage and bounded population size effects.
-
-Current approaches to population genetic inference treat demography and selection separately, typically fitting a demographic model to data on ostensibly neutral sites, and then inferring selection against this demographic background, despite the recurrent problem that the two are confounded \citep{schrider_2016}.
-in expanding (e.g.
-humans) or seasonally-cycling (e.g.
-.
+This is due to some combination of the storage and bounded population size effects (stable coexistence between $b$ and $d$ specialists was not possible in a constant environment).
+It is clear that the classic lottery model, which has a storage effect but no bounded population size effect, will give very different coexistence predictions from our extension of it, because population size will immediately return to capacity $N=T$ in the first summer iteration, after which type frequencies remain constant until the winter.
+The $d$ specialist thus effectively has infinitely many propagules to secure its winter frequency gains, an enormous advantage compared to the finite propagule density dynamics in Fig.
+\ref{fig:fluctuatingselection}.
+Similar difficulties arise in previous models of how the storage effect promotes genetic variation \citep{ellner_1994}, which assume that the total number of offspring per iteration is constant.
+Beyond this observation, disentangling the storage and bounded population size effects is not straightforward, and requires a more detailed discussion of each effect than we have space for here.
+Our model is well suited for such a disentangling since it extends the classic model of the storage effect to arbitrary population densities.
 
@@
 \includegraphics[scale=0.7]{fluctuatingselection.pdf}
-\caption{\label{fig:fluctuatingselection} Stable coexistence between $b$ and $d$ specialists when $b$ and $d$ fluctuate.
-(a) $b$ (solid lines) and $d$ (dashed lines) alternate being nonzero.
-The $b$ specialist (black) has higher $b$ and $d$ than the $d$ specialist (blue).
-(b) Both types grow during the positive $b$ phase, and decline during the positive $d$ phase, but the $d$ specialist (blue) does so at a slower rate than the $b$ specialist (black).
-Total height (blue+black) is population size $N$.
-(c) The positive $b$ and $d$ phases favor the $b$ and $d$ specialists, respectively, and they stably coexist due to the storage effect.}
+\caption{\label{fig:fluctuatingselection} Stable coexistence between $b$ and $d$ specialists in a fluctuating environment.
+(a) Birth and death rates seasonally alternate being nonzero (white for winter, green for summer).
+The $b$ specialist (black) has higher $b$ and $d$ ($b=0.5$, $d=0.2$) than the $d$ specialist ($b=0.2$, $d=0.1$) (blue).
+(b) Both types grow during the positive $b$ phase, and decline during the positive $d$ phase, but the $d$ specialist does so at a lower rate.
+Total height (blue+black) is population density $N/T$.
+(c) Summer favors the $b$ specialist, winter the $d$ specialist, and they stably coexist.}
 \end{figure}

### 2017-04-13 simulations.py

@@ -49,8 +49,8 @@ def A(m,c,U):
             +(L*(1-np.exp(-L))/(1-(1+L)*np.exp(-L))-l*(1-np.exp(-l))/(1-(1+l)*np.exp(-l)))/(L-l)*(sum(c*l)-c*l))
     
 def deltnplus(m,c,U):
-    if sum(m)>0:
-        L=sum(m)/U
+    if sum(m)>0 and U>0:
+        L=sum(m)/float(U)
         cbar=sum(m*c)/sum(m)
         return m*(np.exp(-L)+(R(m,c,U)+A(m,c,U))*c/cbar)
     else:
@@ -164,27 +164,26 @@ def d(t):
 c=np.array([1.,1.])
 T=100000
 g=40.     #iterations per seasonal cycle
-bparam=np.array([[0.0,0.0],[.5,.0423]])
+bparam=np.array([[0.0,0.0],[.5,.2]])
 dparam=np.array([[0.0,0.0],[0.2,0.1]])
-totaltime=100000
+totaltime=100
 
 n=np.array([60000.,25000.])
 nhist=[n];
 for t in range(totaltime):
     U=T-sum(n)
-    #n = n + deltnplus((b(t)*n*U/T).astype(np.int),c,U.astype(np.int))-d(t)*n
-    n = n + deltnplus((b(t)*n).astype(np.int),c,U.astype(np.int))-d(t)*n
+    n = n + deltnplus((b(t)*n*U/T).astype(np.int),c,U.astype(np.int))-d(t)*n
     nhist.append(list(n))
     
-n=np.array([60000.,25000.])
-nhistclassic=[n];
-for t in range(totaltime):
-    U=T-sum(n)
-    n = n + deltnplusclassic((b(t)*n*U/T).astype(np.int),c,U.astype(np.int))-d(t)*n
-    nhistclassic.append(list(n))
-
-plt.stackplot(range(totaltime+1),np.array(nhistclassic)[:,1],np.array(nhistclassic)[:,0],colors=['b','k'])
-plt.plot(np.array(nhistclassic)[:,0]/np.sum(nhistclassic,1),'k',lw=2)
+#n=np.array([60000.,25000.])
+#nhistclassic=[n];
+#for t in range(totaltime):
+#    U=T-sum(n)
+#    n = n + deltnplusclassic((b(t)*n*U/T).astype(np.int),c,U.astype(np.int))-d(t)*n
[…]

### 2017-04-13 manuscript.tex
latest

@@
 
-Nevertheless, there are strong indications there exist broader principles governing the operation of selection.
+Nevertheless,  broader principles governing the operation of selection.
 
-In many groups of organisms, including corals \citep{darling_2012}, insects \citep{southwood_1977}, fishes \citep{winemiller_1992}, zooplankton \citep{allan_76} and plants \citep{grime_1988,westoby_1998}, different species can be divided into a small number of distinct trait clusters corresponding to fundamentally distinct ``primary strategies'' \citep{winemiller_2015}.
+In many groups of organisms, including corals \citep{darling_2012}, insects \citep{southwood_1977}, fishes \citep{winemiller_1992}, zooplankton \citep{allan_76} and plants \citep{grime_1988,westoby_1998},
+
+different species can be divided into a small number of distinct trait clusters corresponding to fundamentally distinct ``primary strategies'' \citep{winemiller_2015}.
 

### 2017-04-13 manuscript.tex
latest

@@
 This approach is powerful enough to model eco-evolutionary feedbacks between co-evolving traits, but is generally problem-specific and restricted to only a few traits at a time.
-These models emphasize invasion from low frequencies and co-existence more than frequency trajectories over time, with the entire adaptive dynamics approach revolving around ``invasion fitness'' \citep{diekmann_2004}.
+These models emphasize invasion from low frequencies and co-existence over frequency trajectories over time.
+For instance, ``invasion fitness'' is the foundation of the adaptive dynamics approach \citep{diekmann_2004}.
 
@@
 
-Nevertheless,  broader principles governing the operation of selection.
-
-In many groups of organisms, including corals \citep{darling_2012}, insects \citep{southwood_1977}, fishes \citep{winemiller_1992}, zooplankton \citep{allan_76} and plants \citep{grime_1988,westoby_1998},
+Empirical studies of organismal traits corals \citep{darling_2012}, insects \citep{southwood_1977}, fishes \citep{winemiller_1992}, zooplankton \citep{allan_76} and plants \citep{grime_1988,westoby_1998},
 
@@
 Thus, plant species can be classified on a triangular C/S/R ternary plot \citep{grime_1974}.
+
 Trait classification schemes for other organisms are broadly analogous to Grime's scheme \citep{winemiller_2015}.

### 2017-04-14 manuscript.tex
latest

@@
 These models emphasize invasion from low frequencies and co-existence over frequency trajectories over time.
-For instance, ``invasion fitness'' is the foundation of the adaptive dynamics approach \citep{diekmann_2004}.
+For instance, the adaptive dynamics approach revolves around the analysis of ``invasion fitness'' \citep{diekmann_2004}.
 
@@
 
-Empirical studies of organismal traits corals \citep{darling_2012}, insects \citep{southwood_1977}, fishes \citep{winemiller_1992}, zooplankton \citep{allan_76} and plants \citep{grime_1988,westoby_1998},
-
-different species can be divided into a small number of distinct trait clusters corresponding to fundamentally distinct ``primary strategies'' \citep{winemiller_2015}.
-
-The most famous example is Grime's plant trait classification scheme \citep{grime_1974,grime_1977,grime_1988}.
+Empirical studies of organismal traits have suggested the existence of a few ``primary strategies'' which reflect how different species have responded to selection \citep{winemiller_2015}.
+The most famous example is Grime's triangle \citep{grime_1974,grime_1977,grime_1988,westoby_1998}.
 Grime considered two broad determinants of population density: stress (persistent hardship e.g.
@@
 Thus, plant species can be classified on a triangular C/S/R ternary plot \citep{grime_1974}.
-
-Trait classification schemes for other organisms are broadly analogous to Grime's scheme \citep{winemiller_2015}.
+Similar trait classification schemes were proposed at a similar time for  insects \citep{southwood_1977}, fishes \citep{winemiller_1992}, and zooplankton \citep{allan_76}.
+While these schemes were largely verbal, modern hierarchical clustering techniques have revealed distinct clusters analogous to Grime's primary strategies in corals \citep{darling_2012},
 

### 2017-04-14 manuscript.tex
latest

@@
 Survival is not possible in high-stress, high-disturbance environments.
-Grime showed that measures of C, S and R across a wide range of plant species are anti-correlated, so that strong C-strategists are weak S and R strategists, and so on.
-Thus, plant species can be classified on a triangular C/S/R ternary plot \citep{grime_1974}.
-Similar trait classification schemes were proposed at a similar time for  insects \citep{southwood_1977}, fishes \citep{winemiller_1992}, and zooplankton \citep{allan_76}.
-While these schemes were largely verbal, modern hierarchical clustering techniques have revealed distinct clusters analogous to Grime's primary strategies in corals \citep{darling_2012},
+Grime showed that measures of C, S and R across a wide range of plant species are anti-correlated, so that strong C-strategists are weak S and R strategists, and so on, creating a triangular C/S/R ternary plot \citep{grime_1974}.
+Similar schemes were proposed for insects \citep{southwood_1977}, fishes \citep{winemiller_1992}, and zooplankton \citep{allan_76}.
+More recently, modern hierarchical clustering techniques have revealed distinct trait clusters in corals analogous to Grime's primary strategies \citep{darling_2012}.
+Trait classification schemes suggest that, to a first approximation, functional traits contribute to fitness via a few key factors such as intrinsic reproductive rate or stress-tolerance.
 
@@
 
-Trait classification schemes show empirically that, beneath the complicated details of trait variation, even among closely-related species, fitness is predominantly determined by a few key factors such as intrinsic reproductive rate or stress-tolerance.
-However, while trait classification schemes are firmly grounded in trait data, they are verbal and descriptive rather than mathematical, a recognized hindrance to their broader applicability (e.g.
-\citealt{tilman_2007}).
+Here we explore the interplay between some of these ``key factors'' of fitness in a simplified, territorial model of growth, dispersal and competition, with the broader aim of adding some ecological realism to the time-depedendent, genetically-focused, population genetic models of evolution.
+To achieve this, we develop a generalized version of the classic lottery model of \cite{chesson_1981}, which has two features that make it well suited for this role.
 
-The aim of this paper is explore the interplay between some major dimensions of fitness in a simplified, territorial model of growth, dispersal and competition.
-Building on the earlier r/K and C/S/R schemes, a central question is how fitness depends on the interaction between population density, intrinsic birth/death rates and competitive ability.
+First, the lottery representation of competition is much simpler than it is in other commonly used ecological models, such as coefficients for the pairwise effects of types on each other (e.g.
+the $\alpha$ coefficients in the generalized Lotka-Volterra equations), or  explicit resource consumption \citep{tilman_1982}.
+Mature individuals (``adults'') each require their own territory, whereas newborn individuals (``propagules'') disperse to, and subsequently compete for, territories made available by the death of adults.
+Territorial contest among propagules leaves a single victorious adult per territory, the victor chosen at random from the propagules present, with probabilities weighted by a coefficient for each type representing competitive ability, akin to a lottery \citep{sale_77}.
+Second, as we will see, the lottery model is closely connected to one of the central models of population genetics, the Wright-Fisher model of genetic drift.
+
+However, the classic lottery model breaks down at low densities (few propagules dispersing to each territory; see ``Model'').
+This was not a limitation in its original application to reef fishes, where a huge number of larvae from each species compete to secure territories \citep{chesson_1981}, but is a critical limitation for studying density-dependent selection.
+We analytically extend the classic lottery model to correctly account for low density behavior.
 
 We broadly follow the spirit of MacArthur's $r$/$K$ selection scheme in that our model is intended to account for fundamentally different forms of selection without getting  entangled in the intricacies of particular ecological scenarios.
-However, rather than building directly on MacArthur's formalism and its later extensions using Lotka-Volterra equations to incorporate competition (``$\alpha$-selection'') \citep{gill_1974,case_1974,joshi_2001}, our model is devised more with Grime's C/S/R scheme in mind, and represents a quantitative formalization of how C/S/R manifests at the level of within-population genotypic evolution (as opposed to phenotypic divergence between species).
+
+However, rather than building directly on MacArthur's formalism and its later extensions using Lotka-Volterra equations to incorporate competition (``$\alpha$-selection'') \citep{gill_1974,case_1974,joshi_2001},
+
+ our model is devised more with Grime's C/S/R scheme in mind, and represents a quantitative formalization of how C/S/R manifests at the level of within-population genotypic evolution (as opposed to phenotypic divergence between species).
+
  This choice is motivated in part by the substantial empirical support for C/S/R-like schemes, and in part by the failings of the $r$/$K$ low/high density dichotomy --- many growth ability traits will confer advantages at both low and high densities (more details in the Discussion).
 
-As we will see, a generalized version of the classic lottery model of \cite{chesson_1981} is a convenient starting point for ecologically-grounded models of selection in evolutionary genetics.
-In the classic lottery model, mature individuals (``adults'') each require their own territory, whereas newborn individuals (``propagules'') disperse to, and subsequently compete for, territories made available by the death of adults.
-Territorial contest among propagules leaves a single victorious adult per territory, the victor chosen at random from the propagules present (akin to a lottery; \citealt{sale_77}), with probabilities weighted by a coefficient for each type representing competitive ability.
-This representation of competition is much simpler than having coefficients for the pairwise effects of types on each other (e.g.
-the $\alpha$ coefficients in the generalized Lotka-Volterra equations), or than modeling resource consumption explicitly \citep{tilman_1982}.
-The classic lottery model is also closely connected to one of the central models of population genetics, the Wright-Fisher model of genetic drift.
+Building on the r/K and C/S/R schemes, a central question is how fitness depends on population density, intrinsic birth/death rates and competitive ability.
 
-However, the classic lottery model breaks down at low densities (that is, if there are only a few propagules dispersing to each territory; see ``Model'').
-This was not a limitation in the lottery model's original application to reef fishes, where a huge number of larvae from each species compete to secure territories each generation \citep{chesson_1981}, but is a critical limitation for studying density-dependent selection.
-We analytically extend the classic lottery model to correctly account for low density behavior.
+ However, while trait classification schemes are  grounded in trait data, they are verbal and descriptive rather than mathematical, a recognized hindrance to their broader applicability (e.g.
+\citealt{tilman_2007}).
 

### 2017-04-14 manuscript.tex
latest

@@
 276]{ewens_2004}.
-The emphasis here is on predicting how the course of allele frequencies over time depends on their selective effect, genetic drift and linkage.
-However, the resulting picture of evolution does not include even basic elements of the ecological underpinnings of selection, including dynamic population size and density-dependence.
+The emphasis here is on predicting how the course of allele frequencies over time depends on their selective effect, genetic drift and linkage, and inferring past selection, drift and linkage given a sample of nucleotide sequences.
+The resulting picture of evolution does not include even basic elements of the ecological underpinnings of selection, such as density-dependence.
 
-By contrast, models of phenotypic trait evolution represent the change in phenotypic abundances over time using absolute fitness functions which describe how those traits affect survival and reproduction in particular ecological scenarios.
+By contrast, models of phenotypic trait evolution use absolute fitness functions which describe how some traits of interest affect survival and reproduction in particular ecological scenarios.
 This approach is powerful enough to model eco-evolutionary feedbacks between co-evolving traits, but is generally problem-specific and restricted to only a few traits at a time.
-These models emphasize invasion from low frequencies and co-existence over frequency trajectories over time.
-For instance, the adaptive dynamics approach revolves around the analysis of ``invasion fitness'' \citep{diekmann_2004}.
+The emphasis here is on invasion from low frequencies and co-existence, rather than frequency trajectories over time.
+For instance, adaptive dynamics revolves around the analysis of ``invasion fitness'' \citep{diekmann_2004}.
 
-Far less work has been done to generalize beyond particular traits or ecological scenarios to models of fitness that still capture key distinctions between different forms of selection.
+Less work has been done to generalize beyond particular traits or ecological scenarios to models of fitness that still capture key distinctions between different forms of selection.
 Perhaps this is not surprising given that fitness is such a complex quantity, dependent on all of a phenotype's functional traits \citep{violle_2007} as well as its biotic and abiotic environment.
@@
 For example, MacArthur's famous $r$/$K$ scheme \citep{macarthur_1962,macarthur_1967} is now almost exclusively known as a framework for understanding life-history traits, and judged on its failure in that role \citep{pianka_1970,stearns_1977,boyce_1984,reznick_2002}.
-The r/K scheme's original purpose was as an extension of the existing population-genetic treatment of selection to account for population density \citep{macarthur_1962}, but few attempts have been made to develop it further as a mathematical analysis of the major different forms of selection.
+The r/K scheme's original purpose was to extend the existing population-genetic treatment of selection to account for population density \citep{macarthur_1962}, but few attempts have been made to develop it further as a mathematical analysis of the major different forms of selection.
 
-Empirical studies of organismal traits have suggested the existence of a few ``primary strategies'' which reflect how different species have responded to selection \citep{winemiller_2015}.
+On the other hand, empirical studies of organismal traits have suggested the existence of a few ``primary strategies'', reflecting broadly distinct responses to selection \citep{winemiller_2015}.
 The most famous example is Grime's triangle \citep{grime_1974,grime_1977,grime_1988,westoby_1998}.
@@
 More recently, modern hierarchical clustering techniques have revealed distinct trait clusters in corals analogous to Grime's primary strategies \citep{darling_2012}.
-Trait classification schemes suggest that, to a first approximation, functional traits contribute to fitness via a few key factors such as intrinsic reproductive rate or stress-tolerance.
+These empirical findings suggest that, to a first approximation, functional traits contribute to fitness via a few key factors such as intrinsic reproductive rate or stress-tolerance.
 
@@
 
-Here we explore the interplay between some of these ``key factors'' of fitness in a simplified, territorial model of growth, dispersal and competition, with the broader aim of adding some ecological realism to the time-depedendent, genetically-focused, population genetic models of evolution.
-To achieve this, we develop a generalized version of the classic lottery model of \cite{chesson_1981}, which has two features that make it well suited for this role.
+Here we explore the interplay between some of these ``key factors'' of fitness in a simplified, territorial model of growth, dispersal and competition, with the broader aim of adding some ecological realism to population genetics' time-depedendent, genetically-focused view of evolution.
+To achieve this, we revisit the classic lottery model of \cite{chesson_1981}, which has two features that make it well suited for this role, and one critical flaw.
 
-First, the lottery representation of competition is much simpler than it is in other commonly used ecological models, such as coefficients for the pairwise effects of types on each other (e.g.
-the $\alpha$ coefficients in the generalized Lotka-Volterra equations), or  explicit resource consumption \citep{tilman_1982}.
+The first feature is that the lottery representation of competition is  particularly concise.
 Mature individuals (``adults'') each require their own territory, whereas newborn individuals (``propagules'') disperse to, and subsequently compete for, territories made available by the death of adults.
 Territorial contest among propagules leaves a single victorious adult per territory, the victor chosen at random from the propagules present, with probabilities weighted by a coefficient for each type representing competitive ability, akin to a lottery \citep{sale_77}.
-Second, as we will see, the lottery model is closely connected to one of the central models of population genetics, the Wright-Fisher model of genetic drift.
+By comparison, coefficients for the pairwise effects of types on each other (e.g.
+the $\alpha$ coefficients in the generalized Lotka-Volterra equations and the associated concept of ``$\alpha$-selection''; \citealt{gill_1974,case_1974,joshi_2001}), or  explicit resource consumption \citep{tilman_1982}, are much more complicated.
+The second feature, as we will see, is the close connection between the lottery model and one of the foundational models of population genetics, the Wright-Fisher model of genetic drift.
 
-However, the classic lottery model breaks down at low densities (few propagules dispersing to each territory; see ``Model'').
-This was not a limitation in its original application to reef fishes, where a huge number of larvae from each species compete to secure territories \citep{chesson_1981}, but is a critical limitation for studying density-dependent selection.
-We analytically extend the classic lottery model to correctly account for low density behavior.
+The critical flaw is that the classic lottery model breaks down at low densities (few propagules dispersing to each territory).
+This was not a limitation in its original application to reef fishes \citep{chesson_1981}, which have an enormous number of larvae, but is a critical limitation for studying density-dependent selection.
+Thus, our first task is to analytically extend the classic lottery model to correctly account for low density behavior (sections ``Model'' and ``Mean field approximation'').
 
-We broadly follow the spirit of MacArthur's $r$/$K$ selection scheme in that our model is intended to account for fundamentally different forms of selection without getting  entangled in the intricacies of particular ecological scenarios.
-
-However, rather than building directly on MacArthur's formalism and its later extensions using Lotka-Volterra equations to incorporate competition (``$\alpha$-selection'') \citep{gill_1974,case_1974,joshi_2001},
-
- our model is devised more with Grime's C/S/R scheme in mind, and represents a quantitative formalization of how C/S/R manifests at the level of within-population genotypic evolution (as opposed to phenotypic divergence between species).
-
- This choice is motivated in part by the substantial empirical support for C/S/R-like schemes, and in part by the failings of the $r$/$K$ low/high density dichotomy --- many growth ability traits will confer advantages at both low and high densities (more details in the Discussion).
-
-Building on the r/K and C/S/R schemes, a central question is how fitness depends on population density, intrinsic birth/death rates and competitive ability.
-
- However, while trait classification schemes are  grounded in trait data, they are verbal and descriptive rather than mathematical, a recognized hindrance to their broader applicability (e.g.
+Using our extended lottery model, we then revisit Grime's C/S/R scheme,  and evaluate how C/S/R manifests at the level of within-population genotypic evolution (as opposed to phenotypic divergence between species; sections ``Invasion of rare genotypes and coexistence'' and ``Primary strategies and Grime's triangle'').
+This represents a ``sanity check'' on our density-dependent lottery model.
+The resulting formulation of the C/S/R scheme is mathematical, in contrast to Grime's original verbal and descriptive approach, which is a recognized hindrance to the evaluation or broader application of the C/S/R scheme (e.g.
 \citealt{tilman_2007}).
 
-In the section ``Model'', we introduce the basic assumptions of our generalized lottery model.
-Analytical expressions for the change in genotype abundances over time are introduced in section ``Mean field approximation'', with mathematical details relegated to the Appendices.
-In the following two sections, we then discuss the behavior of rare mutants (invasion and coexistence) and our formalization of Grime's triangle.
+We then explore some time-dependent behavior of our extended lottery model.
+The broader context for this section is the recurring difficulties in population genetic inference caused by the similarity of genealogies generated by selective sweeps and population bottlenecks \citep{barton_1998}.
+
+Current approaches attempt to treat selection and demography separately, typically fitting a demographic model to data on ostensibly neutral sites and then inferring selection against this demographic background, but with mixed success due to linkage \citep{schrider_2016}.
+Fitness models like Eq.
+\eqref{eq:master} may help to address to these difficulties since they do not artificially separate the dynamics of abundance and frequency.
+Taking an example inspired by recent studies of rapid, seasonal evolution in \textit{Drosophila} \citep{bergland_14}, we discuss how environmental fluctuations might stabilize polymorphisms in the presence of cyclical population density.
 
@@
 The number of individuals of the $i$'th genotype landing in any particular territory is denoted $x_i$.
-Random dispersal implies that in the limit $T\rightarrow \infty$, with $n_i/T$ held fixed, $x_i$ is Poisson distributed with mean territorial propagule density $l_i=m_i/U$ (this dispersal Poisson distribution is denoted $p_i(x_i)=l_i^{x_i} e^{-l_i}/x_i!$).
+Random dispersal implies that in the limit $T\rightarrow \infty$, with $n_i$ adjusted to keep $n_i/T$ constant as the limit is taken, $x_i$ is Poisson distributed with mean territorial propagule density $l_i=m_i/U$ (this dispersal Poisson distribution is denoted $p_i(x_i)=l_i^{x_i} e^{-l_i}/x_i!$).
 Although $T$ is finite in our model, we assume that $T$ and the $n_i$ are large enough that $x_i$ is Poisson-distributed to a good approximation (details in Appendix A).
@@
 
-So we have only considered the low-frequency invasion behavior of Eq.
-\eqref{eq:master}.
-Here we start exploring its full time-dependent behaviour.
-The broader context for this section is the recurring difficulties in population genetic inference caused by the confounding of population demography and selection --- selective sweeps generate similar genealogies to population bottlenecks \citep{barton_1998}.
-Current approaches attempt to treat selection and demography separately, typically fitting a demographic model to data on ostensibly neutral sites and then inferring selection against this demographic background, but with mixed success due to linkage \citep{schrider_2016}.
-Fitness models like Eq.
-\eqref{eq:master} may help to address to these difficulties since they do not artificially separate the dynamics of abundance and frequency.
+So we have only considered the low-frequency invasion behavior of Eq. \eqref{eq:master}. Here we start exploring its full time-dependent behaviour.
 

### 2017-04-15 manuscript.tex
latest

@@
 Loss of propagules during dispersal is subsumed into $m_i$.
+
 In general, $m_i$ will increase with $n_i$, and will also depend on population density $N$.
 For example, if $b_i$ is the number of successfully dispersing propagules produced per genotype $i$ adult, then the loss of propagules due to dispersal to occupied territories implies $m_i=b_i(1-N/T)n_i$, akin to Levins' competition-colonization model \citep{levins_71,tilman_94}.
-Here we assume $m_i=b_i n_i$, where $b_i$ is a constant, meaning that all propagules land on unoccupied territories (a form of directed dispersal).
-This choice simplifies the mathematics without seriously restricting the generality of our analysis, since the results presented here are not sensitive to the specific functional form of $m_i$.
+In section ``Cyclical birth and death rates'' we evaluate Eq.
+\eqref{eq:master} numerically using this functional form for $m_i$.
+
+In the sections ``Invasion of rare genotypes and coexistence'' and ``Primary strategies and Grime's triangle'', we assume the simpler form $m_i=b_i n_i$, where $b_i$ is a constant, meaning that all propagules land on unoccupied territories (a form of directed dispersal).
+This simplifies the mathematics without affecting the results of those sections, which only depend on the low-frequency invasion behavior of Eq.
+\eqref{eq:master}.
 Note that due to our assumption of uniform dispersal, the parameter $b_i$ can be thought of as a measure of ``colonization ability'', which combines fecundity and dispersal ability \citep{levins_71,tilman_94,bolker_99}.
 
+We assume that the $n_i$ and $T$ are all so large that stochastic, finite population size effects (``drift'') can be ignored.
+We will thus obtain deterministic equations for the change in the $n_i$ over time, leaving the evaluation of drift in our model for future work.
+
 The number of individuals of the $i$'th genotype landing in any particular territory is denoted $x_i$.
-Random dispersal implies that in the limit $T\rightarrow \infty$, with $n_i$ adjusted to keep $n_i/T$ constant as the limit is taken, $x_i$ is Poisson distributed with mean territorial propagule density $l_i=m_i/U$ (this dispersal Poisson distribution is denoted $p_i(x_i)=l_i^{x_i} e^{-l_i}/x_i!$).
-Although $T$ is finite in our model, we assume that $T$ and the $n_i$ are large enough that $x_i$ is Poisson-distributed to a good approximation (details in Appendix A).
-Note that the large $n_i$, large $T$ approximation places no restrictions on our densities $n_i/T$, but it does preclude consideration of demographic stochasticity when $n_i$ itself is very small.
+We assume that $x_i$ is Poisson distributed with mean territorial propagule density $l_i=m_i/U$; this distribution is denoted $p_i(x_i)=l_i^{x_i} e^{-l_i}/x_i!$.
+
+While random dispersal does not exactly give a Poisson distribution,
+
+in many populations the value of $T$ would be large enough to make this an excellent approximation (Appendix A).
 
@@
 This is mathematically equivalent to the classic lottery model with non-overlapping generations ($d_i=1$ for all $i$) and $w_i=b_i c_i$.
-
-The classic lottery model allows us to replace the abstract Wright-Fisher relative fitnesses $w_i$ with more ecologically-grounded fecundity, competitive ability and mortality parameters $b_i$, $c_i$ and $d_i$, respectively.
+Thus, the classic lottery model allows us to replace the abstract Wright-Fisher relative fitnesses $w_i$ with more ecologically-grounded fecundity, competitive ability and mortality parameters $b_i$, $c_i$ and $d_i$, respectively.
 Since birth and death rates affect absolute abundances, this allows us to evaluate selection at different densities (after appropriate extensions are made), in an otherwise very similar model to the canonical Wright-Fisher model.
@@
 $\Delta_+ n_i/m_i$ from Eq.
-\eqref{eq:master} (and its separate components) are shown, along with direct simulations of random dispersal and lottery competition over one iteration over a range of propagule densities (varied by changing $U$ with the $m_i$ fixed).
+\eqref{eq:master} (and its separate components) are shown, along with direct simulations of random dispersal and lottery competition over one iteration over a range of propagule densities ($U$ is varied between $5\times 10^3$ and $10^6$ with $m_1=10^4$ and $m_2=9\times 10^4$).
 Two genotypes are present.
-(a) and (b) show the low-frequency genotype with $c$-advantage ($m_1/M=0.1$, $c_1=1.5$), (c) and (d) show the high-frequency predominant genotype ($m_2/M=0.9$, $c_2=1$).
+(a) and (b) show the low-frequency genotype with $c$-advantage ($c_1=1.5$), (c) and (d) show the high-frequency predominant genotype ($c_2=1$).
 Simulation points are almost invisible in (c) and (d) due to near exact agreement with Eq.
@@
 Total height (blue+black) is population density $N/T$.
-(c) Summer favors the $b$ specialist, winter the $d$ specialist, and they stably coexist.}
+(c) Summer favors the $b$ specialist, winter the $d$ specialist, and they stably coexist.
+For illustration, the propagule abundances are assumed to have the form $m_i=b_i(1-N/T)n_i$, reflecting non-directed disperal.}
 \end{figure}
@@
 
-For each genotype's dispersal, the counts of propagules across unnocupied territories follows a multinomial distribution with equal probabilities of landing in each territory.
+For simplicity of presentation, we have assumed a Poisson distribution for the $x_i$ as our model of dispersal.
+Strictly speaking, the total number of $i$ propagules $\sum x_i$ (summed over unoccupied territories) is then no longer a constant $m_i$, but fluctuates between generations for a given mean $m_i$, which is more biologically realistic.
+Nevertheless, since we do not consider the random fluctuations in type abundances here, and for ease of comparison with the classic lottery model, we ignore the fluctuations in $m_i$.
+Instead we focus, on Poisson fluctuations in propagule composition in each territory.
+
+In the exact model of random dispersal, the counts of a genotype's propagules across unnocupied territories follows a multinomial distribution with dimension $U$, total number of trials equal to $m_i$, and equal probabilities $1/U$ for a propagule to land in a given territory.
 Thus, the $x_i$ in different territories are not independent random variables.
-However, for sufficiently large $T$, holding $n_i/T$ fixed, the Poisson limit theorem implies that this multinomial distribution for the $x_i$ accross territories is closely approximated by a product of independent Poisson distributions for each territory, each with rate parameter $l_i$ (large $T$ implies large $U$ except in the biologically uninteresting case that there is vanishing population turnover $d_i \sim 1/T$).
-
-Alternatively, we could have assumed a Poisson distribution for the $x_i$ as our model of dispersal from the outset.
-The total number of genotype $i$ propagules $\sum x_i$ (summed over unoccupied territories) would then no longer be a constant $m_i$, but would fluctuate between generations for a given mean $m_i$, which is more biologically realistic.
-Nevertheless, for simplicity and ease of comparison with the classic lottery model, we ignore the possibility of fluctuations in $m_i$ and focus instead on Poisson fluctuations propagule composition in each territory.
+However, for sufficiently large $U$ and $m_i$, this multinomial distribution for the $x_i$ across territories is closely approximated by a product of independent Poisson distributions for each territory, each with rate parameter $l_i$ \citep[Theorem 1]{arenbaev_1977}.
+Since we are ignoring finite population size effects, we effectively have $T\rightarrow \infty$, in which case $U$ can be only be small enough to violate the Poisson approximation if there is vanishing population turnover, and then the dispersal distribution is irrelevant anyway.
+Likewise, in ignoring stochastic finite population size for the $n_i$, we have effectively already assumed that $m_i$ is large enough to justify the Poisson approximation (the error scales as $1/\sqrt{m_i}$; \citealt{arenbaev_1977}).
 

### 2017-04-17 manuscript.tex
latest

@@
 Evolutionary models differ greatly in their treatment of fitness.
-In models of genetic evolution, genotypes are typically assigned constant (or frequency-dependent) selection coefficients describing the change in their relative frequencies over time due to differences in viability.
-This considerably simplifies the mathematics of selection, facilitating greater genetic realism, and can be justified over sufficiently short time intervals \citep[p.
+In models of genetic evolution, genotypes are typically assigned constant (or frequency-dependent) selection coefficients describing the change in their relative frequencies over time.
+This simplified treatment of selection facilitates genetic realism, and can be justified over sufficiently short time intervals \citep[p.
 276]{ewens_2004}.
-The emphasis here is on predicting how the course of allele frequencies over time depends on their selective effect, genetic drift and linkage, and inferring past selection, drift and linkage given a sample of nucleotide sequences.
-The resulting picture of evolution does not include even basic elements of the ecological underpinnings of selection, such as density-dependence.
+The emphasis here is to predict how allele frequencies change over time based on their fitness effect, genetic drift and linkage, or to infer past selection, drift and linkage given a sample of nucleotide sequences.
+The resulting picture of evolution excludes basic elements of the ecological underpinnings of selection, including how selection affects population size, and density-dependence.
+This complicates the inference of past selection, because demographic changes can look geneaologically very similar to selective frequency changes \citep{barton_1998}.
+Separately fitting demographic models to data on ostensibly neutral sites, and then inferring selection against this demographic background, has had mixed success due to linkage \citep{schrider_2016}.
 
-By contrast, models of phenotypic trait evolution use absolute fitness functions which describe how some traits of interest affect survival and reproduction in particular ecological scenarios.
-This approach is powerful enough to model eco-evolutionary feedbacks between co-evolving traits, but is generally problem-specific and restricted to only a few traits at a time.
-The emphasis here is on invasion from low frequencies and co-existence, rather than frequency trajectories over time.
-For instance, adaptive dynamics revolves around the analysis of ``invasion fitness'' \citep{diekmann_2004}.
+By contrast, models of phenotypic trait evolution use absolute fitness functions to describe how some traits of interest affect survival and reproduction in particular ecological scenarios.
+These fitness functions can be quite problem-specific and often only account for a few traits at a time.
+The emphasis here is on invasion from low frequencies and co-existence, rather than frequency or abundance trajectories over time.
+For instance, adaptive dynamics uses ``invasion fitness'' to explore the consequences of eco-evolutionary feedbacks \citep{diekmann_2004}.
 
-Less work has been done to generalize beyond particular traits or ecological scenarios to models of fitness that still capture key distinctions between different forms of selection.
-Perhaps this is not surprising given that fitness is such a complex quantity, dependent on all of a phenotype's functional traits \citep{violle_2007} as well as its biotic and abiotic environment.
-In most cases, a detailed, trait-based, predictive model of fitness would be enormously complicated and have narrow applicability.
+Less work has been done to generalize beyond particular traits or ecological scenarios to model fundmentally different forms of selection.
+Perhaps this is not surprising given that fitness is such a complex quantity, dependent on all of a phenotype's functional traits \citep{violle_2007} and its environment.
+A detailed, trait-based, predictive model of fitness would be enormously complicated and situation-specific.
 It is therefore easy to doubt the feasibility of a simplified, general mathematical treatment of fitness \citep[p.
@@
 For example, MacArthur's famous $r$/$K$ scheme \citep{macarthur_1962,macarthur_1967} is now almost exclusively known as a framework for understanding life-history traits, and judged on its failure in that role \citep{pianka_1970,stearns_1977,boyce_1984,reznick_2002}.
-The r/K scheme's original purpose was to extend the existing population-genetic treatment of selection to account for population density \citep{macarthur_1962}, but few attempts have been made to develop it further as a mathematical analysis of the major different forms of selection.
+The r/K scheme's original purpose was to extend the existing population-genetic treatment of selection to account for population density \citep{macarthur_1962}, but few attempts have been made to develop it further along these lines.
 
-On the other hand, empirical studies of organismal traits have suggested the existence of a few ``primary strategies'', reflecting broadly distinct responses to selection \citep{winemiller_2015}.
-The most famous example is Grime's triangle \citep{grime_1974,grime_1977,grime_1988,westoby_1998}.
-Grime considered two broad determinants of population density: stress (persistent hardship e.g.
-due to resource scarcity, unfavorable temperatures or toxins) and disturbance (intermittent destruction of vegetation e.g.
+Empirical trait classification studies have suggested the existence of a few ``primary strategies'', reflecting broadly distinct responses to selection \citep{winemiller_2015}.
+Grime famously considered \citep{grime_1974,grime_1977,grime_1988,westoby_1998} two broad determinants of population density: stress (persistent hardship e.g.
+due to resource scarcity or unfavorable temperatures) and disturbance (intermittent destruction of vegetation e.g.
 due to trampling, herbivory, pathogens, extreme weather or fire).
@@
 More recently, modern hierarchical clustering techniques have revealed distinct trait clusters in corals analogous to Grime's primary strategies \citep{darling_2012}.
-These empirical findings suggest that, to a first approximation, functional traits contribute to fitness via a few key factors such as intrinsic reproductive rate or stress-tolerance.
+These empirical findings suggest that functional traits contribute to fitness predominantly via a few key factors such as intrinsic reproductive rate or stress-tolerance.
 
@@
 
-Here we explore the interplay between some of these ``key factors'' of fitness in a simplified, territorial model of growth, dispersal and competition, with the broader aim of adding some ecological realism to population genetics' time-depedendent, genetically-focused view of evolution.
-To achieve this, we revisit the classic lottery model of \cite{chesson_1981}, which has two features that make it well suited for this role, and one critical flaw.
+Here we explore the interplay between some ``key factors'' of fitness in a simplified, territorial model of growth, dispersal and competition.
+This broadly follows the original spirit of MacArthur's $r/K$ scheme, but more specifically, our aim is to begin adding some ecological realism to population genetics' time-depedendent, genetically-focused view of evolution.
+We revisit the classic lottery model of \cite{chesson_1981}, which has two features that make it well suited for this role, but one critical flaw that we address here.
 
@@
 the $\alpha$ coefficients in the generalized Lotka-Volterra equations and the associated concept of ``$\alpha$-selection''; \citealt{gill_1974,case_1974,joshi_2001}), or  explicit resource consumption \citep{tilman_1982}, are much more complicated.
-The second feature, as we will see, is the close connection between the lottery model and one of the foundational models of population genetics, the Wright-Fisher model of genetic drift.
+The second feature is the close connection between the lottery model and one of the foundational models of population genetics, the Wright-Fisher model of genetic drift, which we discuss further below.
 
-The critical flaw is that the classic lottery model breaks down at low densities (few propagules dispersing to each territory).
-This was not a limitation in its original application to reef fishes \citep{chesson_1981}, which have an enormous number of larvae, but is a critical limitation for studying density-dependent selection.
-Thus, our first task is to analytically extend the classic lottery model to correctly account for low density behavior (sections ``Model'' and ``Mean field approximation'').
+The critical flaw of the classic lottery model is that it breaks down at low densities (few propagules dispersing to each territory), precluding density-dependent behaviour.
+Our first task is to analytically extend the classic lottery model to correctly account for low density behavior (sections ``Model'' and ``Mean field approximation'').
 
@@
 We then explore some time-dependent behavior of our extended lottery model.
-The broader context for this section is the recurring difficulties in population genetic inference caused by the similarity of genealogies generated by selective sweeps and population bottlenecks \citep{barton_1998}.
+Taking an example inspired by recent studies of rapid, seasonal evolution in \textit{Drosophila} \citep{bergland_14}, we discuss how environmental fluctuations might stabilize polymorphisms in the presence of cyclical population density.
 
-Current approaches attempt to treat selection and demography separately, typically fitting a demographic model to data on ostensibly neutral sites and then inferring selection against this demographic background, but with mixed success due to linkage \citep{schrider_2016}.
-Fitness models like Eq.
-\eqref{eq:master} may help to address to these difficulties since they do not artificially separate the dynamics of abundance and frequency.
-Taking an example inspired by recent studies of rapid, seasonal evolution in \textit{Drosophila} \citep{bergland_14}, we discuss how environmental fluctuations might stabilize polymorphisms in the presence of cyclical population density.
+%Fitness models like Eq. \eqref{eq:master} may help to address to these difficulties since they do not artificially separate the dynamics of abundance and frequency.
 
@@
 
-We assume that the reproductively mature individuals in a population (``adults'') each require their own territory to survive and reproduce (Fig.~\ref{fig:lottery}).
+We assume that reproductively mature individuals (``adults'') each require their own territory to survive and reproduce (Fig.~\ref{fig:lottery}).
 All territories are identical, and the total number of territories is $T$.
@@
 In iteration $t$, the number of adults of the $i$'th genotype is $n_i(t)$, the total number of adults is $N(t)=\sum_i n_i(t)$, and the number of unoccupied territories is $U(t)=T-N(t)$.
+We assume that the $n_i$ and $T$ are large enough that stochastic fluctuations in the $n_i$ (``drift'') can be ignored.
+We derive deterministic equations for the expected change in the $n_i$ over time, leaving the evaluation of drift for future work.
 
@@
 Each iteration, adults produce new offspring (``propagules''), $m_i$ of which disperse to unoccupied territories.
-We assume that adults cannot be ousted from occupied territories, and so only propagules landing on unoccupied territories are included in $m_i$.
-Propagules disperse at random uniformly, and independently of each other; all propagules have the same probability of landing on any of the $U$ unoccupied territories.
-Thus, there is no interaction between propagules (e.g.
+We assume that adults cannot be ousted from their territories, so that $m_i$ only includes propagules landing on unoccupied territories.
+Propagules disperse at random over the unoccupied territories, regardless of distance from their parents, and independently of each other.
+There is no interaction between propagules (e.g.
 avoidance of territories crowded with propagules).
@@
 
-In general, $m_i$ will increase with $n_i$, and will also depend on population density $N$.
+In general, $m_i$ will increase with $n_i$, and will depend on population density $N$.
 For example, if $b_i$ is the number of successfully dispersing propagules produced per genotype $i$ adult, then the loss of propagules due to dispersal to occupied territories implies $m_i=b_i(1-N/T)n_i$, akin to Levins' competition-colonization model \citep{levins_71,tilman_94}.
 In section ``Cyclical birth and death rates'' we evaluate Eq.
-\eqref{eq:master} numerically using this functional form for $m_i$.
+\eqref{eq:master} numerically using this functional form for $m_i$, with $b_i$ assumed to be constant.
 
-In the sections ``Invasion of rare genotypes and coexistence'' and ``Primary strategies and Grime's triangle'', we assume the simpler form $m_i=b_i n_i$, where $b_i$ is a constant, meaning that all propagules land on unoccupied territories (a form of directed dispersal).
+In the sections ``Invasion of rare genotypes and coexistence'' and ``Primary strategies and Grime's triangle'', we assume the simpler form $m_i=b_i n_i$, with constant $b_i$, meaning that all propagules land on unoccupied territories (a form of directed dispersal).
 This simplifies the mathematics without affecting the results of those sections, which only depend on the low-frequency invasion behavior of Eq.
@@
 
-We assume that the $n_i$ and $T$ are all so large that stochastic, finite population size effects (``drift'') can be ignored.
-We will thus obtain deterministic equations for the change in the $n_i$ over time, leaving the evaluation of drift in our model for future work.
+The number of individuals of the $i$'th genotype landing in any particular territory is denoted $x_i$.
+We assume that $x_i$ follows a Poisson distribution $p_i(x_i)=l_i^{x_i} e^{-l_i}/x_i!$, where $l_i=m_i/U$ is the mean territorial propagule density.
+This only approximates uniform random dispersal, but is essentially exact provided that the $n_i$ are large enough that drift can be ignored (Appendix A).
 
-The number of individuals of the $i$'th genotype landing in any particular territory is denoted $x_i$.
-We assume that $x_i$ is Poisson distributed with mean territorial propagule density $l_i=m_i/U$; this distribution is denoted $p_i(x_i)=l_i^{x_i} e^{-l_i}/x_i!$.
-
-While random dispersal does not exactly give a Poisson distribution,
-
-in many populations the value of $T$ would be large enough to make this an excellent approximation (Appendix A).
-
-When multiple propagules land on the same territory, they compete to secure the territory as they develop.
-This territorial contest is modeled as a weighted lottery: the probability that genotype $i$ wins a territory by the next iteration, assuming that at least one of its propagules is present, is $c_i x_i/\sum_j c_j x_j$, where $c_i$ is a constant representing relative competitive ability (Fig.
+When multiple propagules land on the same territory, the victor is determined by lottery competition: genotype $i$ wins a territory with probability $c_i x_i/\sum_j c_j x_j$, where $c_i$ is a constant representing relative competitive ability (Fig.
 \ref{fig:lottery}).
 
-In the classic lottery model \citep{chesson_1981}, unoccupied territories are assumed to be saturated with propagules from every genotype i.e.
-$l_i\gg 1$ where $l_i=m_i/U$ is the mean propagule density.
+In the classic lottery model \citep{chesson_1981}, unoccupied territories are assumed to be saturated with propagules from every genotype $l_i\gg 1$.
 From the law of large numbers, the composition of propagules in each territory will then not deviate appreciably from the mean composition $l_1,l_2,\ldots,l_G$ ($G$ is the number of genotypes present), and so the probability that genotype $i$ wins any particular unoccupied territory is approximately $c_i l_i/\sum_j c_j l_j$.
@@
 Thus, the classic lottery model allows us to replace the abstract Wright-Fisher relative fitnesses $w_i$ with more ecologically-grounded fecundity, competitive ability and mortality parameters $b_i$, $c_i$ and $d_i$, respectively.
-Since birth and death rates affect absolute abundances, this allows us to evaluate selection at different densities (after appropriate extensions are made), in an otherwise very similar model to the canonical Wright-Fisher model.
-Note that the classic lottery model, and the present work, both ignore the stochastic drift in type frequencies (large $T$ approximation) which is the focus of the Wright-Fisher model.
+Since birth and death rates affect absolute abundances, this allows us to evaluate selection at different densities (after appropriate extensions are made), in an otherwise very similar model to the canonical Wright-Fisher.
+We therefore expect that drift in our extended lottery model should be similar to that in the Wright-Fisher model, but we leave this for future work.
 
@@
 
-Dispersal is Poisson, and so we expect that a fraction $p_1(x_1)\ldots p_G(x_G)$ of the $U$ unoccupied territories will have the propagule composition $x_1,\ldots,x_G$.
+We expect that a fraction $p_1(x_1)\ldots p_G(x_G)$ of the $U$ unoccupied territories will have the propagule composition $x_1,\ldots,x_G$.
 Genotype $i$ is expected to win $c_i x_i/\sum_j c_j x_j$ of these.
-Ignoring fluctuations about these two expectations (large $T$ approximation), genotype $i$'s territorial acquisition is given by
+Ignoring fluctuations about these two expectations (due to our no-drift, large $T$, large $n_i$ approximation), genotype $i$'s territorial acquisition is given by
 \begin{equation}
@@
 in our extended lottery model, where the sum only includes territories with at least one propagule present.
+Note that unlike the classic lottery model, not all unoccupied territories are claimed each iteration, since under Poisson dispersal a fraction $e^{-L}$ remain unoccupied.
 
@@
 
-Note that the competitive ability coefficients $c_i$ represent a strictly relative aspect of fitness in the sense that they can only influence population size $N$ indirectly by changing genotype frequencies.
-This can be seen by summing Eq.
-\eqref{eq:delttot} over genotypes to get the  change in population size $N$,
-\begin{equation}
-\Delta N=U(1-e^{-L})-\sum_i d_i n_i,\label{eq:deltN}
-\end{equation}
-which is independent of $c_i$ (here $L=\sum_j l_j$ is the overall propagule density).
-Unlike the classic lottery model, not all $U$ unoccupied territories are claimed each iteration; a fraction $e^{-L}$ remain unoccupied.
+%(the total number of terr change in the total  $\Delta_+ N=U(1-e^{-L})$ Summing Eq. \eqref{eq:delttot} over genotypes, we get the total number of territories occupied each iteration, 
+%\begin{equation}
+%.\label{eq:deltN}
+%\end{equation}
+
+%Note that the competitive ability coefficients $c_i$ represent a strictly relative aspect of fitness in the sense that they can only influence population size $N$ indirectly by changing genotype frequencies.
+%which is independent of $c_i$ (here $L=\sum_j l_j$ is the overall propagule density).
 
@@
 
-Our approximation is similar to the high-$l_i$ approximation behind the classic lottery model in that we replace the $x_i$ with appropriate mean values.
-However, we cannot simply replace $x_i$ with $l_i$ as in the classic lottery model.
-For a genotype with a low propagule density $l_i\ll 1$, we have $x_i=1$ in the few territories where its propagules land, and so its growth comes entirely from territories which deviate appreciably from $l_i$.
-In our more general approximation, territories with a single propagule from the focal genotype are handled separately.
-In place of the requirement of $l_i\gg 1$ for all $i$, our approximation only requires that there are no large discrepancies in competitive ability (specifically, that we do not have $c_i/c_j\gg 1$ for any two genotypes; further discussion in section ``Discussion'').
+Similarly to the high-$l_i$ approximation of classic lottery model, we replace the $x_i$ with appropriate mean values, although we cannot simply replace $x_i$ with $l_i$.
+For a genotype with low propagule density $l_i\ll 1$, we have $x_i=1$ in the territories where its propagules land, and so its growth comes entirely from territories which deviate appreciably from $l_i$.
+To account for this, we separate Eq.
+\eqref{eq:growthsumuncoupled} into $x_i=1$ and $x_i>1$ parts.
+Our more general mean field approximation only requires that there are no large discrepancies in competitive ability (i.e.
+we do not have $c_i/c_j\gg 1$ for any two genotypes).
 We obtain (details in Appendix B)
@@
 If $l_i\gg 1$ for all genotypes, we recover the classic lottery model (only the $A_ic_i/\overline{c}$ term remains, and $A_i\rightarrow 1/L$).
-Thus, Eq.
-\eqref{eq:master} generalizes the classic lottery model to account for arbitrary propagule densities for each genotype.
 
 Fig.~\ref{fig:simcomp} shows that Eq.
-\eqref{eq:master} (and its components) closely approximate direct simulations of random dispersal and lottery competition over a wide range of propagule densities (obtained by varying $U$).
-Two genotypes are present, one of which has a $c$-advantage and is at low frequency.
+\eqref{eq:master} (and its components) closely approximate direct simulations of random dispersal and lottery competition over a wide range of propagule densities.
+Two genotypes are present, one of which is at low frequency.
 The growth of the low-frequency genotype relies crucially on the low-density competition term $R_i c_i/\overline{c}$, and also to a lesser extent on the high density competition term $A_i c_i/\overline{c}$ if $l_1$ is large enough (Fig.~\ref{fig:simcomp}b).
@@
 Fig.
-3 also shows the breakdown of the classic lottery model at low densities (low $l_1$ and $l_2$).
+3 also shows the breakdown of the classic lottery model at low propagule densities.
 
@@
 \eqref{eq:master}.
-Dashed lines in (a) and (c) show the corresponding classic lottery model predictions.}
+Dashed lines in (a) and (c) show the breakdown of the classic lottery model.}
 \end{figure}
-
-\subsection*{Invasion of rare genotypes and coexistence}\label{sec:invas}
-
-In our model (section ``Model''), each genotype is defined by three traits: $b$, $c$ and $d$.
-To determine how these will evolve in a population where they are being modified by mutations, we need to know whether mutant lineages will grow (or decline) starting from low densities.
-In this section we discuss the behavior of rare genotypes predicted by Eq.
-\eqref{eq:master}.
-
-Suppose that a population with a single genotype $i$ is in equilibrium. Then $R_i=0$, $\overline{c}=c_i$ and $\Delta n_i = 0$, and so Eq. \eqref{eq:master} gives
-\begin{equation}
-b_i\left(e^{-L}+A_i\right)-d_i=0,\label{eq:equil}
-\end{equation}
-where $A_i=(1-(1+L)e^{-L})/L$.
-Now suppose that a new genotype $j$, which is initially rare, appears in the population.
-Then $A_j\ll R_j$, $l_j\approx 0$ and $\overline{c}\approx c_i$, and so, from Eq.
-\eqref{eq:master}, $n_j$ will increase if
-\begin{equation}
-b_j \left(e^{-L}+R_j\frac{c_j}{c_i}\right)-d_j>0,\label{eq:invad}
-\end{equation}
-where $R_j\approx (1-e^{-L})/\left(\frac{c_j}{c_i}+\frac{L-1-e^{-L}}{1-(1+L)e^{-L}}\right)$.
-
-Combining Eqs.
-\eqref{eq:equil} and \eqref{eq:invad}, we see that $j$ will invade if it is superior in any one of the three traits, but is otherwise identical to $i$.
-If the new genotype has the same competitive ability $c_j=c_i$, then $R_j\approx A_i$ and Eqs.
-\eqref{eq:equil} and \eqref{eq:invad} imply that invasion occurs when $b_jd_i-b_id_j>0$, and in particular when $b_j>b_i$ with $d_j=d_i$, or when $d_j<d_i$ with $b_j=b_i$.
-In the case that the new genotype has a different competitive ability but the same $b_i$ and $d_i$, Eqs.
-\eqref{eq:equil} and \eqref{eq:invad} imply that invasion occurs when $R_j c_j/c_i > A_i$; it is not hard to verify that this occurs if and only if $c_j>c_i$ using the simplified expressions for $A_i$ and $R_j$ given after Eqs.
-\eqref{eq:equil} and \eqref{eq:invad} respectively.
-Moreover, if $j$ invades in any of these cases, it will eventually exclude $i$, since it is strictly superior.
-
-Stable coexistence is possible between genotypes that are superior in different traits.
-Suppose that $j$ is better at securing territories ($c_j>c_i$), that $i$ is better at producing propagules ($b_i>b_j$), and that $d_i=d_j$.
-Coexistence occurs if $j$ will invade an $i$-dominated population, but $i$ will also invade a $j$-dominated population (``mutual invasion'').
-If $b_i$ is so large that $L\gg 1$ when $i$ is dominant, and $b_j$ is so small that $L\ll 1$ when $j$ is dominant, then, combining Eqs.
-\eqref{eq:equil} and \eqref{eq:invad}, we find that $i$ invades $j$ because $b_i>b_j$, while $j$ invades $i$ provided that
-\begin{equation}
-b_jc_jR_j-b_i c_i A_i>0. \label{eq:jinvadcoex}
-\end{equation}
-Thus, coexistence occurs if $c_j/c_i$ is large enough.
-This is a version of the classic competition-colonization trade-off \citep{tilman_94,levins_71}: the competitor ($c$-specialist) leaves many territories unoccupied (low $L$) due to its poor colonization ability (low $b$), which the colonizer ($b$-specialist) can then exploit.
-A similar argument applies for coexistence between high-$c$ and low-$d$ specialists; a ``competition-longevity'' trade-off \citep{tilman_94}.
-Mutual invasibility is not possible between $b$- and $d$-specialists.
-
-If the rare genotype $j$ arises due to mutation, then its initial low-density behavior is more complicated than the above invasion analysis suggests.
-The mutant lineage starts with one individual $n_j=1$, and remains at low abundance for many generations after its initial appearance.
-During this period, the mutant abundance $n_j$ will behave stochastically, and the deterministic equations \eqref{eq:growthsumuncoupled} and \eqref{eq:master} do not apply (section ``Model'').
-However, if $n_j$ becomes large enough, its behavior will become effectively deterministic, and governed by Eq.
-\eqref{eq:master}.
-For mutants with fitness greater than the population mean fitness, this occurs when $n_j$ is of order $1/s$ \citep{desai_2007}, where the selection coefficient $s$ is the mutant's fitness advantage (i.e.
-$s=\frac{\Delta n_i/n_i}{\sum_i\Delta n_i/n_i\times n_i/N}-1$).
-Here we do not consider the initial stochastic behavior of novel mutants, and have restricted our attention to the earliest deterministic behavior of rare genotypes.
-In particular, for beneficial mutations we have only considered the case where $s$ is large enough that deterministic behavior starts when $n_j \ll N$.
 
@@
 Also shown is the approximate dependence of $\Delta_+ n_i$ on $b_i$ and $c_i$ for each environment (fourth row).
-These can be used infer the expected direction of evolution for the traits $b$, $c$ and $d$ (fifth row) as follows.
+These can be used infer the expected direction of evolution for the traits $b$, $c$ and $d$ (fifth row) using a standard invasion analysis.
 
-As noted in the previous section, if beneficial mutations survive the low-abundance stochastic regime, they proceed to grow deterministically according to Eq.
+When a new mutant genotype $j$ appears, it starts with one individual $n_j=1$.
+While the mutant lineage remains at low-abundance, $n_j$ will behave stochastically, but if its expected growth rate is positive and $n_j$ becomes large enough, $n_j$ will effectively grow deterministically according to Eq.
 \eqref{eq:master}.
-The probability of surviving to deterministic abundances increases with the mutant fitness advantage, and is therefore typically on the order of one percent, whereas the fixation of neutral mutations is exceedingly unlikely (probability of order $1/N$).
-Consequently, the direction of evolutionary change is determined by which trait changes are both available, and confer an appreciable benefit, where availability is subject to constraints imposed by the environment.
+This transition occurs at an abundance of order $1/r$ \citep{uecker_2011}, where $r=\Delta n_i/n_i$ is the mutant lineage's growth rate.
+Since here we do not evaluate drift in the extended lottery model, we do not attempt to calculate the probability that mutants escape the initial stochastic phase, and restrict our attention to the earliest deterministic behavior of rare genotypes while they are still at negligibly low frequency.
+We simply invoke the well known result this probability is proportional to $r$, with a proportionality factor typically of order one (Haldane's formula; \citealt{uecker_2011}).
+The fixation of neutral mutations is exceedingly unlikely (probability of order $1/N$).
+Consequently, the direction of evolutionary change is determined by the mutational trait changes which are available and also confer an appreciable fitness benefit, where availability is subject to constraints imposed by the environment.
 
@@
 
-\section*{Cyclical birth and death rates}
+\subsection*{Coexistence in constant and cyclical environments}\label{sec:invas}
 
-So we have only considered the low-frequency invasion behavior of Eq. \eqref{eq:master}. Here we start exploring its full time-dependent behaviour.
+So far we have only considered the low-frequency invasion behavior of Eq. \eqref{eq:master}. Here we start exploring its full time-dependent behaviour.
+
+%each genotype is defined by three traits: $b$, $c$ and $d$.
+
+Conider a population with a single genotype $i$ in equilibrium. Then $R_i=0$, $\overline{c}=c_i$ and $\Delta n_i = 0$, and Eq. \eqref{eq:master} gives
+\begin{equation}
+b_i\left(e^{-L}+A_i\right)-d_i=0,\label{eq:equil}
+\end{equation}
+where $A_i=(1-(1+L)e^{-L})/L$.
+Now suppose that a new genotype $j$, which is initially rare, appears in the population.
+Then $A_j\ll R_j$, $l_j\approx 0$ and $\overline{c}\approx c_i$, and so, from Eq.
+\eqref{eq:master}, $n_j$ will increase if
+\begin{equation}
+b_j \left(e^{-L}+R_j\frac{c_j}{c_i}\right)-d_j>0,\label{eq:invad}
+\end{equation}
+where $R_j\approx (1-e^{-L})/\left(\frac{c_j}{c_i}+\frac{L-1-e^{-L}}{1-(1+L)e^{-L}}\right)$.
+
+Stable coexistence is possible between genotypes that are superior in different traits.
+Suppose that $j$ is better at securing territories ($c_j>c_i$), that $i$ is better at producing propagules ($b_i>b_j$), and that $d_i=d_j$.
+Coexistence occurs if $j$ will invade an $i$-dominated population, but $i$ will also invade a $j$-dominated population (``mutual invasion'').
+If $b_i$ is so large that $L\gg 1$ when $i$ is dominant, and $b_j$ is so small that $L\ll 1$ when $j$ is dominant, then, combining Eqs.
+\eqref{eq:equil} and \eqref{eq:invad}, we find that $i$ invades $j$ because $b_i>b_j$, while $j$ invades $i$ provided that
+\begin{equation}
+b_jc_jR_j-b_i c_i A_i>0. \label{eq:jinvadcoex}
+\end{equation}
+Thus, coexistence occurs if $c_j/c_i$ is large enough.
+This is a version of the classic competition-colonization trade-off \citep{tilman_94,levins_71}: the competitor ($c$-specialist) leaves many territories unoccupied (low $L$) due to its poor colonization ability (low $b$), which the colonizer ($b$-specialist) can then exploit.
+A similar argument applies for coexistence between high-$c$ and low-$d$ specialists; a ``competition-longevity'' trade-off \citep{tilman_94}.
+Mutual invasibility is not possible between $b$- and $d$-specialists.
 

### 2017-04-17 manuscript.tex
latest

@@
 \newpage{}
-
-``...the concept of fitness is probably too complex to allow of a useful mathematical development.
-Since it enters fundamentally into many population genetics considerations, it is remarkable how little attention has been paid to it.'' --- Warren J.
-Ewens, Mathematical Population Genetics I, 2004
 
@@
 
-%(the total number of terr change in the total  $\Delta_+ N=U(1-e^{-L})$ Summing Eq. \eqref{eq:delttot} over genotypes, we get the total number of territories occupied each iteration, 
-%\begin{equation}
-%.\label{eq:deltN}
-%\end{equation}
-
-%Note that the competitive ability coefficients $c_i$ represent a strictly relative aspect of fitness in the sense that they can only influence population size $N$ indirectly by changing genotype frequencies.
-%which is independent of $c_i$ (here $L=\sum_j l_j$ is the overall propagule density).
-
 \section*{Results}
@@
 
-We now discuss which changes in the traits $b, c$ and $d$ will be particularly favored under different environmental conditions.
+We now discuss the evolutionary change in $b, c$ and $d$ under different environmental conditions.
 Of specific interest are Grime's ``disturbed'', ``stressful'' and ``ideal'' environments.
@@
 
-The ideal environment is characterized by the near-absence of stress and disturbance.
-Consequently, $d_i\ll 1$, whereas $b_i$ is potentially much larger than $1$.
+The ideal environment is characterized by the near-absence of stress and disturbance: $d_i\ll 1$, whereas $b_i$ is potentially much larger than $1$.
 From Eq.
@@
 We assume that disturbances are equally damaging to adults and juveniles, so that only $(1-d_i)\Delta_+ n_i$ rather than $\Delta_+ n_i$ territories are secured by genotype $i$ each iteration.
-[More justification?] Disturbed environments then correspond to $d_i$ being close to $1$ for all genotypes (almost all adults and juveniles are killed each iteration).
+Disturbed environments then correspond to $d_i$ being close to $1$ for all genotypes (almost all adults and juveniles are killed each iteration).
 From Eq.
@@
 
-So far we have only considered the low-frequency invasion behavior of Eq. \eqref{eq:master}. Here we start exploring its full time-dependent behaviour.
+In the previous section we only considered the how $b$, $c$ and $d$ should respond in Grime's environmental extremes.
+Here we further explore the low frequency behavior of Eq.
+\eqref{eq:master} to determine which types can coexist in a constant environment, and then consider the full time-dependent behaviour of Eq.
+\eqref{eq:master} in a cyclical environment.
 
-%each genotype is defined by three traits: $b$, $c$ and $d$.
-
-Conider a population with a single genotype $i$ in equilibrium. Then $R_i=0$, $\overline{c}=c_i$ and $\Delta n_i = 0$, and Eq. \eqref{eq:master} gives
+In a population with a single genotype $i$ which is in equilibrium, $R_i=0$, $\overline{c}=c_i$ and $\Delta n_i = 0$, and Eq. \eqref{eq:master} gives
 \begin{equation}
@@
 
-Here we will not directly address these inference challenges, but will instead illustrate the complicated time-dependent behavior that can arise from Eq.
-\eqref{eq:master} in an example where birth and death rates vary periodically with amplitude sufficent to cause large changes in population density.
-This is inspired by natural \textit{Drosophila} populations, which expand rapidly in the warmer months when fruit is abundant, but largely die off in the colder months.
+Now suppose that birth and death rates vary periodically with amplitude sufficent to cause large changes in population density.
+This example is inspired by natural \textit{Drosophila} populations, which expand rapidly in the warmer months when fruit is abundant, but largely die off in the colder months.
 Within this seasonal population density cycle, hundreds of polymorphisms also cycle in frequency \citep{bergland_14}.
@@
 Beyond this observation, disentangling the storage and bounded population size effects is not straightforward, and requires a more detailed discussion of each effect than we have space for here.
-Our model is well suited for such a disentangling since it extends the classic model of the storage effect to arbitrary population densities.
+Our model is well suited for such a disentangling since it extends the canonical example of the storage effect to allow for density dependent effects.
 
@@
 
-As discussed in ``Primary strategies and Grime's triangle'', adaptive evolution in the direction predicted by our generalized lottery model produces traits consistent with Grime's scheme, with the possible exception of selection for $b$ at high density, because the corresponding trait data  is ambiguous.
-Selection for $b$ at high density is also is counter to the expectations of MacArthur's $r$/$K$ dichotomy \citep{macarthur_1967}, since $b$ is closely related to the maximal, low-density growth rate $r=b-d$ \citep{pianka_1972}, yet in the $r$/$K$ scheme, high density populations should be subject to $K$, not $r$, selection.
+In the introduction we mentioned the recurring difficulties with confounding selection and demography in population genetic inference.
+While we have not directly attempted to perform inference with our model, and Eq.
+\eqref{eq:master} may not be appropriate for particular inference problems, it seems that something similar (and hopefully more analytically tractable) to our extension of the lottery model is unavoidable because, fundamentally, selective births and deaths affect both  abundances and frequencies, not one or the other in isolation.
+Moreover, some aspects of allele frequency change are intrinsically density dependent.
+In the classic lottery model, which as we have seen is essentially the Wright-Fisher model with overlapping generations, $b_i$ and $c_i$ are  equivalent in the sense that the number of territorial victories only depends on the product $b_i c_i$ (see ``Model'').
+This is no longer the case in our extension, where $b$ and $c$ specialists can co-exist.
+This ``colonization-competition trade-off'' is well known in the co-existence literature \citep{tilman_94}.
+It and similar forms of ``spatial co-existence'' in stable environments have previously been modeled either with Levin's  qualitative representation of competition \citep{levins_71,tilman_94}, as opposed to the quantitative $c$ of lottery competition, or with a more sophisticated treatment of space (non-uniform dispersal; \citealt{shmida_84,bolker_99}).
+In cyclical environments, polymorphisms can be stablized by the bounded density effect which is  completely lost if there is an exclusive focus on allele frequencies \citep{yi_2013}.
+We leave thedetails of how our model might be applied in inference, including the crucial issue of its genetic drift predictions, for future work.
 
-It is not surprising that $b$ can matter at high densities.
+It is interesting to compare the predictions of the extended lottery model with earlier approaches, such as the $r$/$K$ scheme.
+While adaptive evolution in the direction predicted by our model does produce traits broadly consistent with Grime's scheme, trait data on selection for higher $b$ at high density was ambiguous.
+This prediction is also counter to the expectations of MacArthur's $r$/$K$ dichotomy \citep{macarthur_1967}, since $b$ is closely related to the maximal, low-density growth rate $r=b-d$ \citep{pianka_1972}, yet in the $r$/$K$ scheme, high density populations should be subject to $K$, not $r$, selection.
+Yet it is not surprising that $b$ can matter at high densities.
 In our model (or any lottery model of competition), $b$ matters at high densities because territorial contests among juveniles are intrinsically unpredictable.
@@
 $b$ in our model.
-Our $c$ is strictly competitive ability only --- as such, changes in $c$ do not directly affect population density (section ``Model'').
+Our $c$ is strictly competitive ability only --- as such, changes in $c$ do not directly affect population density (the total number of territories occupied in an iteration is $\Delta_+ N=U(1-e^{-L})$, which does not depend directly on the $c_i$).
 The clean separation of a strictly-relative $c$ parameter is particularly useful from an evolutionary genetics perspective, essentially embedding a zero-sum relative fitness trait within a non-zero-sum fitness model.
@@
 
-In the classic lottery model, $b_i$ and $c_i$ are essentially equivalent in the sense that the number of territorial victories only depends on the product $b_i c_i$ (see ``Model'').
-This is no longer the case in our density- and frequency-dependent generalization, where $b$ and $c$ specialists can co-exist.
-This ``colonization-competition trade-off'' is well known in the co-existence literature \citep{tilman_94}.
-It and similar forms of ``spatial co-existence'' in stable environments have previously been modeled either with Levin's  qualitative representation of competition \citep{levins_71,tilman_94}, as opposed to the quantitative $c$ of lottery competition, or with a more sophisticated treatment of space (non-uniform dispersal; \citealt{shmida_84,bolker_99}).
-In fluctuating environments our model would likely give similar co-existence predictions as \cite{chesson_1981}, which revolve around the ``storage effect'', since we retain overlapping generations and competition at the juvenile phase only.
-
 Our realization of Grime's triangle (Fig.~4) differs from approaches which identify primary strategies as trait combinations which can co-exist \citep{bolker_99}, referring instead to the direction of adaptive trait evolution under different regimes of stress and disturbance, which is closer in spirit to Grime's arguments \citep{grime_1974,grime_1977}.
@@
 Thus, while trade-offs may amplify specialization, and are sometimes invoked to explain primary strategy schemes \citep{macarthur_1967,winemiller_1992,aerts_1999}, they are not necessary for it.
-
-%In Fig.
-\ref{fig:table2} we compare our model with some of the other models and schemes touched upon here.
-In a sense this is an ``apples and oranges'' comparison: for instance, Grime's scheme was developed for an entirely different purpose (species classification by traits).
-As such, Fig.
-\ref{fig:table2} is not exhaustive and should be read more as a summary of our model's purpose.
-Like MacArthur's $r$/$K$ scheme, our model is motivated by the need to expand the treatment of selection in population genetics \citep{macarthur_1962}, i.e.
-to incorporate crucial ecological factors in our most genetically realistic models of evolution.
-Thus, viewing evolutionary ecology \citep{kokko_2007,pelletier_2009,schoener_2011} as a spectrum ranging from evolution-only to ecology-only, our model lies close to the understudied evolution-only end of the spectrum.
-By comparison, more familiar approaches to evolutionary ecology such as adaptive dynamics --- essentially ecology coupled with mutant invasion \citep{diekmann_2004} ---  lie close to the ecology-only end of the spectrum.
-
-%\begin{figure}
-%\centering
-%\begin{tabularx}{\linewidth}{lXXXXX}
-%&Formal model?  & Ecologically meaningful? & Empirically-grounded trait scheme? & Generality beyond specific scenarios? & Genetically flexible? \\ \hline
-%Density-dependent lottery & \ding{51} & \ding{51} & \ding{51} & \ding{51} & \ding{51} \\  
-%  MacArthur's $r$/$K$ + $\alpha$ & \ding{51} & \ding{51} & \ding{55}* & \ding{51} & \ding{51} \\
-%  Grime's C/S/R & \ding{55} & \ding{51} & \ding{51} & \ding{51} & NA \\
-%  Traditional pop. gen. & \ding{51} & \ding{55} & \ding{55} & \ding{51} & \ding{51} \\
-%  Eco-evo:  \\
-%  \quad Adaptive dynamics & \ding{51} & \ding{51} & \ding{51} & \ding{51}** & \ding{55} \\
-%  \quad Brute force simulation & \ding{51} & \ding{51} & NA & \ding{55} & \ding{51}
-%\end{tabularx}
-%\caption{\label{fig:table2} Comparison of our density-dependent lottery model with related models and schemes in ecology and evolutionary biology.
-*MacArthur's $r$- and $K$-, as well as $\alpha$ selection, were all derived theoretically.
-Applications to traits came later, and with mixed success (see ``Introduction'' and ``Discussion'').
-**In practice, most of the adaptive dynamics literature focuses on specific eco-evolutionary outcomes such as evolutionary ``branching'' \citep{geritz_1997} or mutualisms \citep{ferriere_2002}, but in principle it can be applied with any fitness model including our density-dependent lottery.}
-%\end{figure}
 

### 2017-04-25 manuscript.tex
latest

@@
 
+``...the concept of fitness is probably too complex to allow of a useful mathematical development.
+Since it enters fundamentally into many population genetics considerations, it is remarkable how little attention has been paid to it.'' --- Warren J.
+Ewens, Mathematical Population Genetics I, 2004
+
 Evolutionary models differ greatly in their treatment of fitness.
@@
 276]{ewens_2004}.
-The emphasis here is to predict how allele frequencies change over time based on their fitness effect, genetic drift and linkage, or to infer past selection, drift and linkage given a sample of nucleotide sequences.
-The resulting picture of evolution excludes basic elements of the ecological underpinnings of selection, including how selection affects population size, and density-dependence.
+The emphasis here is to infer past selection, migration and demographic change given a sample of nucleotide sequences, or to predict how allele frequencies change over time based on their fitness effects, population structure, genetic drift and linkage.
+The resulting picture of evolution excludes basic elements of the ecological underpinnings of selection, including density dependence, and how selection affects population size.
 This complicates the inference of past selection, because demographic changes can look geneaologically very similar to selective frequency changes \citep{barton_1998}.
-Separately fitting demographic models to data on ostensibly neutral sites, and then inferring selection against this demographic background, has had mixed success due to linkage \citep{schrider_2016}.
 
-By contrast, models of phenotypic trait evolution use absolute fitness functions to describe how some traits of interest affect survival and reproduction in particular ecological scenarios.
+%Separately fitting demographic models to data on ostensibly neutral sites, and then inferring selection against this demographic background, has had mixed success due to linkage \citep{schrider_2016}.
+
+By contrast, models of phenotypic trait evolution use absolute fitness functions to describe how some traits of interest affect survival and reproduction in particular ecological scenarios \citep{metz_1992,diekmann_2004}.
 These fitness functions can be quite problem-specific and often only account for a few traits at a time.
-The emphasis here is on invasion from low frequencies and co-existence, rather than frequency or abundance trajectories over time.
+The emphasis here is on the conditions for invasion from low frequencies and co-existence, rather than frequency or abundance trajectories over time.
 For instance, adaptive dynamics uses ``invasion fitness'' to explore the consequences of eco-evolutionary feedbacks \citep{diekmann_2004}.
 
-Less work has been done to generalize beyond particular traits or ecological scenarios to model fundmentally different forms of selection.
+It is challenging to generalize beyond particular traits or ecological scenarios to model fundamentally different forms of selection.
 Perhaps this is not surprising given that fitness is such a complex quantity, dependent on all of a phenotype's functional traits \citep{violle_2007} and its environment.
@@
 For example, MacArthur's famous $r$/$K$ scheme \citep{macarthur_1962,macarthur_1967} is now almost exclusively known as a framework for understanding life-history traits, and judged on its failure in that role \citep{pianka_1970,stearns_1977,boyce_1984,reznick_2002}.
-The r/K scheme's original purpose was to extend the existing population-genetic treatment of selection to account for population density \citep{macarthur_1962}, but few attempts have been made to develop it further along these lines.
+However, the r/K scheme's original purpose was to extend the existing population-genetic treatment of selection to account for population density \citep{macarthur_1962}.
+Few attempts have been made to develop it further along these lines.
 
@@
 \ref{fig:grimeschematic}): competitors ``C'' excel in low stress, low disturbance environments; stress tolerators ``S'' excel in high stress, low disturbance environments; and ruderals ``R''  excel in low stress, high disturbance environments.
-Survival is not possible in high-stress, high-disturbance environments.
+Population persistence is not possible in high-stress, high-disturbance environments.
 Grime showed that measures of C, S and R across a wide range of plant species are anti-correlated, so that strong C-strategists are weak S and R strategists, and so on, creating a triangular C/S/R ternary plot \citep{grime_1974}.
@@
 The two axes show increasing levels of environmental stress and disturbance, respectively.
-Survival is not possible if the combination of stress and disturbance is too large (dashed line).
+Population persistence is not possible if the combination of stress and disturbance is too large (dashed line).
 This creates a triangle, each corner of which corresponds to a ``primary strategy''.}
@@
 Here we explore the interplay between some ``key factors'' of fitness in a simplified, territorial model of growth, dispersal and competition.
-This broadly follows the original spirit of MacArthur's $r/K$ scheme, but more specifically, our aim is to begin adding some ecological realism to population genetics' time-depedendent, genetically-focused view of evolution.
-We revisit the classic lottery model of \cite{chesson_1981}, which has two features that make it well suited for this role, but one critical flaw that we address here.
+This broadly follows the original spirit of MacArthur's $r/K$ scheme.
+More specifically, our aim is to begin adding some ecological realism to population genetics' time-depedendent, genetically-focused view of evolution.
+We revisit the classic lottery model of \cite{chesson_1981}, which has two features that make it well suited for this role, but one critical flaw that we rectify here.
 
@@
 All territories are identical, and the total number of territories is $T$.
-Time $t$ advances in discrete iterations, each representing the average time from birth to reproductive maturity.
+Time $t$ advances in discrete iterations, each representing the time from birth to reproductive maturity.
 In iteration $t$, the number of adults of the $i$'th genotype is $n_i(t)$, the total number of adults is $N(t)=\sum_i n_i(t)$, and the number of unoccupied territories is $U(t)=T-N(t)$.
@@
 Intuitively, the cause of this discrepancy is that individuals are discrete.
-Genotypes with few propagules depend on the outcome of contests in territories where they have at least one propagule present, not some small fraction of a propagule as would be implied by low propagule density $l$ in the classic lottery model.
+Genotypes with few propagules depend on the outcome of contests in territories where they have at least one propagule present, not some small fraction of a propagule as would be implied by small $l_i$ in the classic lottery model.
 In other words, deviations from the mean propagule composition $l_1,l_2,\ldots,l_G$ are important at low density.
@@
 
-We now discuss the evolutionary change in $b, c$ and $d$ under different environmental conditions.
-Of specific interest are Grime's ``disturbed'', ``stressful'' and ``ideal'' environments.
+Here we describe how Grime's ``disturbed'', ``stressful'' and ``ideal'' environments can be captured mathematically in our scheme.
 To proceed, we need to map these verbally-defined environments to quantitative parameter regimes in our model.
@@
 The ideal environment is characterized by the near-absence of stress and disturbance: $d_i\ll 1$, whereas $b_i$ is potentially much larger than $1$.
-From Eq.
+
+%From Eq.
 \eqref{eq:deltN}, the equilibrium value of $L$ only depends on the ratio of birth and death rates.
@@
 
-Disturbed environments are characterized by unavoidably high extrinsic mortality caused by physical destruction.
-Environmental variability, and disturbances in particular, can be modeled as stochastic fluctuations in $b$, $c$ and $d$ \citep{chesson_1981}.
-For simplicity, we do not pursue this more complicated stochastic approach.
-Instead, we represent disturbance by high constant mortality rates $d_i$, and extend this mortality to juveniles in the process of territorial contest (since disturbances do not only affect adults as in Eq.
+Disturbed environments are characterized by short bursts of high elevated extrinsic mortality caused by physical destruction
+
+high extrinsic mortality , or, more accurately, .
+
+In the disturbance extreme of Grime's triangle we can
+
+, we instead represent disturbance by high constant mortality rates $d_i$, and extend this mortality to juveniles in the process of territorial contest (since disturbances do not only affect adults as in Eq.
 \eqref{eq:delttot}).
+
 We assume that disturbances are equally damaging to adults and juveniles, so that only $(1-d_i)\Delta_+ n_i$ rather than $\Delta_+ n_i$ territories are secured by genotype $i$ each iteration.
 Disturbed environments then correspond to $d_i$ being close to $1$ for all genotypes (almost all adults and juveniles are killed each iteration).
-From Eq.
+
+%From Eq.
 \eqref{eq:deltN}, the single genotype equilibrium is given by $L/(1-e^{-L})=d_i/[(1-d_i)b_i]$, and since $L\ll 1$ and $N/T\ll 1$ due to high mortality, we have $L\approx 2(1-d_i/[(1-d_i)b_i])$.

### 2017-04-25 manuscript.tex
latest

@@
 \end{equation}
-This is reasonable approximation in the absence of disturbances; when we come to consider the effects of disturbances (Section ``Primary strategies and Grime's triangle''), we will incorporate disturbance-induced mortality in competing juveniles (Fig.~\ref{fig:lottery}).
+This seems reasonable in the absence of disturbances; when we come to consider the effects of disturbances (Section ``Primary strategies and Grime's triangle''), we will incorporate disturbance-induced mortality in competing juveniles (Fig.~\ref{fig:lottery}).
 
@@
 
-The ideal environment is characterized by the near-absence of stress and disturbance: $d_i\ll 1$, whereas $b_i$ is potentially much larger than $1$.
+The ideal environment is characterized by the near-absence of stress and disturbance: $d\ll 1$, whereas $b$ is potentially much larger than $1$.
+Disturbed environments are characterized by short bursts of high extrinsic mortality $d$ caused by physical destruction.
+We assume that in the disturbance extreme of Grime's triangle, many such bursts (which could each be be spatially localized) will occur each iteration (i.e.
+over the time from birth to reproductive maturity), and so we can approximate extreme disturbance by high constant mortality rates $d_i$, which are now extended to affect juveniles as well.
+We assume that disturbances are equally damaging to adults and juveniles, so that only $(1-d_i)\Delta_+ n_i$ rather than $\Delta_+ n_i$ territories are secured by genotype $i$ each iteration.
+Disturbed environments then correspond to $d_i$ being close to $1$ for all genotypes (most adults and juveniles are killed over an iteration).
 
@@
 The population density is also high $N/T\approx 1$ (since $L=b_iN/(N-T)=b_i/(1-T/N)$).
-
-Disturbed environments are characterized by short bursts of high elevated extrinsic mortality caused by physical destruction
-
-high extrinsic mortality , or, more accurately, .
-
-In the disturbance extreme of Grime's triangle we can
-
-, we instead represent disturbance by high constant mortality rates $d_i$, and extend this mortality to juveniles in the process of territorial contest (since disturbances do not only affect adults as in Eq.
-\eqref{eq:delttot}).
-
-We assume that disturbances are equally damaging to adults and juveniles, so that only $(1-d_i)\Delta_+ n_i$ rather than $\Delta_+ n_i$ territories are secured by genotype $i$ each iteration.
-Disturbed environments then correspond to $d_i$ being close to $1$ for all genotypes (almost all adults and juveniles are killed each iteration).
 
@@
 Severe stress inhibits growth and reproduction, so that $b\ll 1$ \citep{grime_1974,grime_1977}.
-Mutations which appreciably improve $b$ will be either non-existent or extremely unlikely, so $b$ is constrained to remain low.
-In Grime's view, under these conditions the rate at which propagules successfully develop to adulthood cannot appreciably exceed the mortality rate.
-This implies $b/d\approx 1$ in our model, and so the propagule density $L$ is suppressed to such low levels that there are essentially no territorial contests occurring.
+In Grime's view, this means that the rate at which propagules successfully develop to adulthood cannot appreciably exceed the mortality rate $b/d\approx 1$.
+In our model, this implies that territorial contests are rare.
 
-The alternative view is that, while stressful environments imply lower $b$ and support a lower number of individuals per unit area compared what is attainable in ideal environments, stressed populations are actually at high densities relative to the environmental carrying capacity, and are highly competitive \citep{taylor_1990}.
-In the particular case that stress is caused by scarcity of consumable resources, we might expect intense resource competition (for empirical support, see \citealt{davis_1998}).
-Thus, $b$ may actually appreciably exceed $d$ under stressful conditions, even though the absolute value of $b$ is small.
+The alternative view is that stressed environments are highly competitive \citep{taylor_1990}.
+In particular, if consumable resources are scarce, we expect intense resource competition (for empirical support, see \citealt{davis_1998}).
+Stress does restrict growth and reproduction, but it also means that fewer individuals can be supported per unit area.
+Thus, stressed populations are actually at high density relative the environmental carrying capacity (but low density relative to ideal environments).
 
@@
 
-How does Fig.~4 compare to empirical analyses of Grime's C/S/R strategies?
+%How does Fig.~4 compare to empirical analyses of Grime's C/S/R strategies?
 In our comparison we will stick to fishes, corals and plants, for which three-way primary strategy schemes are well developed \citep{grime_1977,winemiller_1992,darling_2012}.
 The connection of our model to fish strategies is necessarily more tentative, given that fishes are motile and not all territorial.
-
-In disturbed environments, we predict evolution for higher $b$ and lower $d$, but not higher $c$.
+%
+%In disturbed environments, we predict evolution for higher $b$ and lower $d$, but not higher $c$.
 Higher $b$ means higher fecundity and/or dispersal ability (see ``Model'').
@@
 Thus, ruderals are characterized by both high $b$ and high $d$, but there is a constant pressure for a shorter life cycle, extending the limits of disturbance that can be tolerated.
-
-In stressful environments, we predict evolution for lower $d$, and also  for higher $c$ in the high-density interpretation of stressful environments.
+%
+%In stressful environments, we predict evolution for lower $d$, and also  for higher $c$ in the high-density interpretation of stressful environments.
 Low $d$ is essential when $b\ll 1$, and stress tolerant plants and corals have long life spans, allowing for long intervals between successful recruitments (and episodic broadcast spawning in corals).
@@
 This may reflect a high-$c$ strategy in the face of intense competition for severely limited resources (the high-density interpretation).
-
-In ideal environments, we predict evolution for higher $b$ and $c$, but not lower $d$.
+%
+%In ideal environments, we predict evolution for higher $b$ and $c$, but not lower $d$.
 In plants and corals, a key mechanism for winning territorial contests (higher $c$) is rapidly outgrowing and ``shading out'' competitors; empirically, rapid individual growth is a defining feature of the competitor trait cluster \citep{grime_1977,darling_2012}.

### 2017-04-26 manuscript.tex
latest

@@
 The ideal environment is characterized by the near-absence of stress and disturbance: $d\ll 1$, whereas $b$ is potentially much larger than $1$.
+
 Disturbed environments are characterized by short bursts of high extrinsic mortality $d$ caused by physical destruction.
@@
 We assume that disturbances are equally damaging to adults and juveniles, so that only $(1-d_i)\Delta_+ n_i$ rather than $\Delta_+ n_i$ territories are secured by genotype $i$ each iteration.
-Disturbed environments then correspond to $d_i$ being close to $1$ for all genotypes (most adults and juveniles are killed over an iteration).
+Heavily disturbed environments then correspond most adults and juveniles being killed every an iteration ($d_i\approx 1$).
 
@@
 
-The alternative view is that stressed environments are highly competitive \citep{taylor_1990}.
+The alternative view is that stressed environments are highly competitive.
 In particular, if consumable resources are scarce, we expect intense resource competition (for empirical support, see \citealt{davis_1998}).
 Stress does restrict growth and reproduction, but it also means that fewer individuals can be supported per unit area.
-Thus, stressed populations are actually at high density relative the environmental carrying capacity (but low density relative to ideal environments).
+Thus, in this view, stressed populations are actually at high density relative the environmental carrying capacity (but low density relative to ideal environments; \citealt{taylor_1990}).
+Territorial contests are important.
 

### 2017-04-26 manuscript.tex
latest

@@
 In iteration $t$, the number of adults of the $i$'th genotype is $n_i(t)$, the total number of adults is $N(t)=\sum_i n_i(t)$, and the number of unoccupied territories is $U(t)=T-N(t)$.
+
 We assume that the $n_i$ and $T$ are large enough that stochastic fluctuations in the $n_i$ (``drift'') can be ignored.
 We derive deterministic equations for the expected change in the $n_i$ over time, leaving the evaluation of drift for future work.
+In particular, we do not evaluate the initial stochastic behaviour of adaptive mutant lineages while they are at low abundances.
+We restrict our attention to the earliest deterministic behavior of mutant lineages while they are still at low frequency (the transition to deterministic growth occurs at an abundance of order equal to their inverse expected absolute growth rate; \citealt{uecker_2011}).
 
@@
 
-When a new mutant genotype $j$ appears, it starts with one individual $n_j=1$.
-While the mutant lineage remains at low-abundance, $n_j$ will behave stochastically, but if its expected growth rate is positive and $n_j$ becomes large enough, $n_j$ will effectively grow deterministically according to Eq.
-\eqref{eq:master}.
-This transition occurs at an abundance of order $1/r$ \citep{uecker_2011}, where $r=\Delta n_i/n_i$ is the mutant lineage's growth rate.
-Since here we do not evaluate drift in the extended lottery model, we do not attempt to calculate the probability that mutants escape the initial stochastic phase, and restrict our attention to the earliest deterministic behavior of rare genotypes while they are still at negligibly low frequency.
 We simply invoke the well known result this probability is proportional to $r$, with a proportionality factor typically of order one (Haldane's formula; \citealt{uecker_2011}).

### 2017-04-26 manuscript.tex
latest

@@
 We derive deterministic equations for the expected change in the $n_i$ over time, leaving the evaluation of drift for future work.
-In particular, we do not evaluate the initial stochastic behaviour of adaptive mutant lineages while they are at low abundances.
-We restrict our attention to the earliest deterministic behavior of mutant lineages while they are still at low frequency (the transition to deterministic growth occurs at an abundance of order equal to their inverse expected absolute growth rate; \citealt{uecker_2011}).
+In particular, we do not evaluate the initial stochastic behaviour of adaptive mutant lineages while they are at low abundance.
+When considering mutations, we restrict our attention to the earliest deterministic behavior of mutant lineages while they are still at low frequency (the transition to deterministic growth occurs at an abundance of order equal to their inverse expected absolute growth rate; \citealt{uecker_2011}).
 
@@
 In Grime's view, this means that the rate at which propagules successfully develop to adulthood cannot appreciably exceed the mortality rate $b/d\approx 1$.
-In our model, this implies that territorial contests are rare.
+In our model, this implies low propagule density and little territorial contest.
 
 The alternative view is that stressed environments are highly competitive.
-In particular, if consumable resources are scarce, we expect intense resource competition (for empirical support, see \citealt{davis_1998}).
-Stress does restrict growth and reproduction, but it also means that fewer individuals can be supported per unit area.
-Thus, in this view, stressed populations are actually at high density relative the environmental carrying capacity (but low density relative to ideal environments; \citealt{taylor_1990}).
-Territorial contests are important.
+In spite of inhibited growth and reproduction, population density $N/T$ is actually high since stressful environments also have lower environmental carrying capacity \citep{taylor_1990}.
+For example, if consumable resources are scarce, we expect intense resource competition (for empirical support, see \citealt{davis_1998}).
+In our model, territorial contests will then be important.
 
 The mapping of different environments to our model parameters is summarized in the first two rows of Fig.~\ref{fig:table}.
-Also shown is the approximate dependence of $\Delta_+ n_i$ on $b_i$ and $c_i$ for each environment (fourth row).
+Also shown is the approximate dependence of $\Delta_+ n_i$ on $b_i$ and $c_i$ for each environment (fourth row; details in Appendix).
 These can be used infer the expected direction of evolution for the traits $b$, $c$ and $d$ (fifth row) using a standard invasion analysis.
 
-We simply invoke the well known result this probability is proportional to $r$, with a proportionality factor typically of order one (Haldane's formula; \citealt{uecker_2011}).
+According to Haldane's formula, the probability that an adaptive mutant survives its initial low abundance phase and begins to grow deterministically is proportional to its expected absolute growth rate, with a proportionality factor typically of order one (\citealt{uecker_2011}) [find Haldane reference].
 The fixation of neutral mutations is exceedingly unlikely (probability of order $1/N$).
@@
 A similar argument applies for coexistence between high-$c$ and low-$d$ specialists; a ``competition-longevity'' trade-off \citep{tilman_94}.
+These forms of co-existence require density dependence, and are not present in the classic lottery model.
 Mutual invasibility is not possible between $b$- and $d$-specialists.
@@
 Allele frequencies are successively multiplied by relative fitness values for each environmental iteration, and so two alleles favored in different enviroments can only stably coexist if the product of fitnesses for one type exactly equals the product for the other \citep{dempster_1955}.
-Thus, stable coexistence still requires frequency dependent selection or heterozygote advantage (as is required in a constant environment).
+Thus, stable coexistence still requires frequency-dependent selection or heterozygote advantage (as is required in a constant environment).
 
-This classical argument overlooks two general mechanisms by which fluctuating selection promotes coexistence.
+This classical argument overlooks two effects that are present in our extended lottery model.
+The first is the storage effect, which requires part of the population to be protected from selection each iteration (as occurs when generations are non-overlapping) \citep{chesson_1981}, and which is also present in the classic lottery model.
+The second is the  bounded population size effect of \cite{yi_2013}, which requires fluctuations in population density and density-dependent regulation of population growth.
+
+%two general mechanisms by which fluctuating selection promotes coexistence.
 The first is the ``storage effect'', which introduces a form of frequency dependent selection that promotes coexistence in the presence of environmental fluctuations but not in a constant environment.
@@
 
-The second mechanism we will call the ``bounded density effect'', since it is a consequence of the inhibition of reproduction at high population densities (Dempster's (1955) argument ignores density-dependent effects).
+%The second mechanism we will call the ``bounded density effect'', since it is a consequence of the inhibition of reproduction at high population densities (Dempster's (1955) argument ignores density-dependent effects).
 If there is growth from low to high density in each environmental cycle, then types that are abundant determine the time available for growth every cycle, with less time spent growing in cycles where they are favored, and more in cycles where they are not.
-This promotes coexistence even in the absence of frequency dependent selection \citep{yi_2013}.
+This promotes coexistence even in the absence of frequency-dependent selection \citep{yi_2013}.
 
@@
 This is due to some combination of the storage and bounded population size effects (stable coexistence between $b$ and $d$ specialists was not possible in a constant environment).
-It is clear that the classic lottery model, which has a storage effect but no bounded population size effect, will give very different coexistence predictions from our extension of it, because population size will immediately return to capacity $N=T$ in the first summer iteration, after which type frequencies remain constant until the winter.
+The classic lottery model gives dramatically different co-existence predictions in this scenario (keeping all other parameters the same, the birth rate of the winter survivor must be about $5$ times smaller).
+This occurs because density returns immediately to capacity $N=T$ in the first summer iteration, after which type frequencies remain constant until the winter.
 The $d$ specialist thus effectively has infinitely many propagules to secure its winter frequency gains, an enormous advantage compared to the finite propagule density dynamics in Fig.

### 2017-04-27 manuscript.tex
latest

@@
 
+[Needs to be revised but won't do that now in case it has to become Nature introductory paragraph]
 Fitness is typically represented in heavily simplified terms in evolutionary genetics, often using constant selection coefficients.
@@
 Since it enters fundamentally into many population genetics considerations, it is remarkable how little attention has been paid to it.'' --- Warren J.
-Ewens, Mathematical Population Genetics I, 2004
+Ewens, Mathematical Population Genetics I, 2004 [Should be removed if we need space]
 
@@
 \end{equation}
+To supplement our analytical mean field derivation, we verified that Eq.
+\eqref{eq:master} closely approximates the exact behavior of our density-dependent lottery model using numerical simulations (Appendix B).
+Thus, Eq.
+\eqref{eq:master} describes how type abundances change over time in a lottery model where population density can itself vary with time.
 
@@
 If $l_i\gg 1$ for all genotypes, we recover the classic lottery model (only the $A_ic_i/\overline{c}$ term remains, and $A_i\rightarrow 1/L$).
-
-Fig.~\ref{fig:simcomp} shows that Eq.
-\eqref{eq:master} (and its components) closely approximate direct simulations of random dispersal and lottery competition over a wide range of propagule densities.
-Two genotypes are present, one of which is at low frequency.
-The growth of the low-frequency genotype relies crucially on the low-density competition term $R_i c_i/\overline{c}$, and also to a lesser extent on the high density competition term $A_i c_i/\overline{c}$ if $l_1$ is large enough (Fig.~\ref{fig:simcomp}b).
-On the other hand, $R_i c_i/\overline{c}$ is negligible for the high-frequency genotype, which depends instead on high density territorial victories (Fig.~\ref{fig:simcomp}d).
-Fig.
-3 also shows the breakdown of the classic lottery model at low propagule densities.
-
-\begin{figure}
-\centering
-\includegraphics[scale=0.7]{simulationcomparison.pdf}
-\caption{\label{fig:simcomp} The change in genotype abundances in a density dependent lottery model is closely approximated by Eq.
-\eqref{eq:master}.
-$\Delta_+ n_i/m_i$ from Eq.
-\eqref{eq:master} (and its separate components) are shown, along with direct simulations of random dispersal and lottery competition over one iteration over a range of propagule densities ($U$ is varied between $5\times 10^3$ and $10^6$ with $m_1=10^4$ and $m_2=9\times 10^4$).
-Two genotypes are present.
-(a) and (b) show the low-frequency genotype with $c$-advantage ($c_1=1.5$), (c) and (d) show the high-frequency predominant genotype ($c_2=1$).
-Simulation points are almost invisible in (c) and (d) due to near exact agreement with Eq.
-\eqref{eq:master}.
-Dashed lines in (a) and (c) show the breakdown of the classic lottery model.}
-\end{figure}
 
@@
 
-%From Eq.
-\eqref{eq:deltN}, the equilibrium value of $L$ only depends on the ratio of birth and death rates.
-For one genotype, $L/(1-e^{-L})=b_i/d_i$, and so the propagule density is high $L\approx b_i/d_i\gg 1$, and every unoccupied territory will be heavily contested.
-The population density is also high $N/T\approx 1$ (since $L=b_iN/(N-T)=b_i/(1-T/N)$).
-
-%From Eq.
-\eqref{eq:deltN}, the single genotype equilibrium is given by $L/(1-e^{-L})=d_i/[(1-d_i)b_i]$, and since $L\ll 1$ and $N/T\ll 1$ due to high mortality, we have $L\approx 2(1-d_i/[(1-d_i)b_i])$.
-Clearly $b_i$ must be exceptionally large to ensure population persistence.
-The terms proportional to $c_i/\overline{c}$ in Eq.
-\eqref{eq:master} are then negligible, and $\Delta_+ n_i$ depends primarily on $b_i$.
-
 Stressful environments are more ambiguous, and have been the subject of an extensive debate in the plant ecology literature (the ``Grime-Tilman'' debate; \citealt{aerts_1999} and references therein).
@@
 Also shown is the approximate dependence of $\Delta_+ n_i$ on $b_i$ and $c_i$ for each environment (fourth row; details in Appendix).
-These can be used infer the expected direction of evolution for the traits $b$, $c$ and $d$ (fifth row) using a standard invasion analysis.
+These can be used to infer the expected direction of evolution for the traits $b$, $c$ and $d$ (fifth row) using a standard invasion analysis.
 
-According to Haldane's formula, the probability that an adaptive mutant survives its initial low abundance phase and begins to grow deterministically is proportional to its expected absolute growth rate, with a proportionality factor typically of order one (\citealt{uecker_2011}) [find Haldane reference].
+According to Haldane's formula, the probability that an adaptive mutant survives its initial low abundance phase and begins to grow deterministically is proportional to its expected absolute growth rate \citep{haldane_1927,uecker_2011}, with a proportionality factor typically of order one.
 The fixation of neutral mutations is exceedingly unlikely (probability of order $1/N$).
-Consequently, the direction of evolutionary change is determined by the mutational trait changes which are available and also confer an appreciable fitness benefit, where availability is subject to constraints imposed by the environment.
+The direction of evolutionary change is determined by the mutational trait changes which are available and also confer an appreciable fitness benefit \citep{yampolsky_2001}, where availability is subject to constraints imposed by the environment.
 
@@
   Density $N/T$  & High & Low & Low & High \\
-  $\Delta_+ n_i\propto$ & $b_i c_i$ & $b_i$ & $b_i$ & $b_i c_i$ \\
   Evolution for & $\uparrow b$, $ \uparrow c$ & $\uparrow b$, $\downarrow d$ & $\downarrow d$ & $\uparrow c, \downarrow d$
@@
 
-In the previous section we only considered the how $b$, $c$ and $d$ should respond in Grime's environmental extremes.
+In the previous section we only considered how $b$, $c$ and $d$ should respond in Grime's environmental extremes.
 Here we further explore the low frequency behavior of Eq.
@@
 
-This classical argument overlooks two effects that are present in our extended lottery model.
-The first is the storage effect, which requires part of the population to be protected from selection each iteration (as occurs when generations are non-overlapping) \citep{chesson_1981}, and which is also present in the classic lottery model.
-The second is the  bounded population size effect of \cite{yi_2013}, which requires fluctuations in population density and density-dependent regulation of population growth.
+This classical argument overlooks two general effects that promote coexistence in fluctuating environments.
+The first is the storage effect, which occurs when part of the population is protected from selection (due to overlapping generations in the lottery model; \citealt{chesson_1981}).
+The second is the  bounded population size effect of \cite{yi_2013}, which occurs when each environmental cycle involves growth from low to high density, with the time spent growing each cycle dependent on the fitness of the types present.
 
@@
 
-Figure \ref{fig:fluctuatingselection} shows the behavior of Eq.
+Fig.
+\ref{fig:fluctuatingselection}a-c shows the behavior of Eq.
 \eqref{eq:master} for an example where $b$ and $d$ cycle between zero and positive values (``summers'' with rapid growth and no mortality, and ``winters'' with mortality and no growth).
@@
 Two types are present, a $b$ specialist, which is better at rapidly growing in the summer (higher $b$), and a $d$ specialist which is better at surviving the winter (lower $d$).
+Neither type has an advantage over a full environmental cycle, and they stably coexist.
+This is due to a combination of the storage and bounded density effects (recall that stable coexistence between $b$ and $d$ specialists was not possible in a constant environment).
 
-Neither type has an advantage over a full environmental cycle, and they stably coexist.
-This is due to some combination of the storage and bounded population size effects (stable coexistence between $b$ and $d$ specialists was not possible in a constant environment).
-The classic lottery model gives dramatically different co-existence predictions in this scenario (keeping all other parameters the same, the birth rate of the winter survivor must be about $5$ times smaller).
-This occurs because density returns immediately to capacity $N=T$ in the first summer iteration, after which type frequencies remain constant until the winter.
-The $d$ specialist thus effectively has infinitely many propagules to secure its winter frequency gains, an enormous advantage compared to the finite propagule density dynamics in Fig.
-\ref{fig:fluctuatingselection}.
-Similar difficulties arise in previous models of how the storage effect promotes genetic variation \citep{ellner_1994}, which assume that the total number of offspring per iteration is constant.
-Beyond this observation, disentangling the storage and bounded population size effects is not straightforward, and requires a more detailed discussion of each effect than we have space for here.
-Our model is well suited for such a disentangling since it extends the canonical example of the storage effect to allow for density dependent effects.
+The classic lottery model (Eq.
+1) gives dramatically different co-existence predictions because expansion to carrying capacity occurs immediately at the start of the summer (Fig.
+\ref{fig:fluctuatingselection}d-f).
+As a result, coexistence requires that the winter survivor's $b$ must be about $5$ times smaller than predicted when we properly account for the growth in the abundance of each type using Eq.
+\eqref{eq:master} (keeping the other parameters the same).
+Previous models of the promotion of genetic variation via the storage effect  \citep{ellner_1994} similarly assume that the total number of offspring per iteration is constant, and would produce a similar error.
+
+%Disentangling the storage and bounded population size effects is not straightforward, and requires a more detailed discussion of each effect than we have space for here.
+Our model is well suited for such a disentangling since it extends the canonical example of the storage effect to allow for density-dependent effects.
 
@@
 (a) Birth and death rates seasonally alternate being nonzero (white for winter, green for summer).
-The $b$ specialist (black) has higher $b$ and $d$ ($b=0.5$, $d=0.2$) than the $d$ specialist ($b=0.2$, $d=0.1$) (blue).
+The $b$ specialist (black) has higher $b$ and $d$ ($b=0.5$, $d=0.2$) than the $d$ specialist ($b=0.217$, $d=0.1$) (blue).
 (b) Both types grow during the positive $b$ phase, and decline during the positive $d$ phase, but the $d$ specialist does so at a lower rate.
@@
 (c) Summer favors the $b$ specialist, winter the $d$ specialist, and they stably coexist.
-For illustration, the propagule abundances are assumed to have the form $m_i=b_i(1-N/T)n_i$, reflecting non-directed disperal.}
+(d-f) Same as (a-c) for the classic lottery model, where coexistence now requires $b\approx 0.0421$ for the $d$ specialist.
+For illustration, the propagule abundances are assumed to have the form $m_i=b_i(1-N/T)n_i$, reflecting non-directed dispersal.}
 \end{figure}
@@
 \eqref{eq:master} may not be appropriate for particular inference problems, it seems that something similar (and hopefully more analytically tractable) to our extension of the lottery model is unavoidable because, fundamentally, selective births and deaths affect both  abundances and frequencies, not one or the other in isolation.
-Moreover, some aspects of allele frequency change are intrinsically density dependent.
+Moreover, some aspects of allele frequency change are intrinsically density-dependent.
 In the classic lottery model, which as we have seen is essentially the Wright-Fisher model with overlapping generations, $b_i$ and $c_i$ are  equivalent in the sense that the number of territorial victories only depends on the product $b_i c_i$ (see ``Model'').
@@
 It and similar forms of ``spatial co-existence'' in stable environments have previously been modeled either with Levin's  qualitative representation of competition \citep{levins_71,tilman_94}, as opposed to the quantitative $c$ of lottery competition, or with a more sophisticated treatment of space (non-uniform dispersal; \citealt{shmida_84,bolker_99}).
-In cyclical environments, polymorphisms can be stablized by the bounded density effect which is  completely lost if there is an exclusive focus on allele frequencies \citep{yi_2013}.
-We leave thedetails of how our model might be applied in inference, including the crucial issue of its genetic drift predictions, for future work.
+In cyclical environments, polymorphisms can be stabilized by the bounded density effect which is  completely lost if there is an exclusive focus on allele frequencies \citep{yi_2013}.
+We leave the details of how our model might be applied in inference, including the crucial issue of its genetic drift predictions, for future work.
 
-It is interesting to compare the predictions of the extended lottery model with earlier approaches, such as the $r$/$K$ scheme.
-While adaptive evolution in the direction predicted by our model does produce traits broadly consistent with Grime's scheme, trait data on selection for higher $b$ at high density was ambiguous.
+It is interesting to compare the predictions of the extended lottery model with earlier approaches, such as the $r$/$K$ scheme, where $r=b-d$ is the maximal, low-density growth rate \citep{pianka_1972}.
+Confusingly, the term ``$K$-selection'' sometimes refers generally to selection at high density \citep{pianka_1972}, encompassing both selection for higher saturation density \citep{macarthur_1967} and competitive ability \citep{gill_1974}.
+Contrary to an $r$/$K$ dichotomy, empirical studies have shown that maximal growth rate and saturation density (measured by abundance) are positively correlated, both between species/strains \citep{luckinbill_1979,kuno_1991,hendriks_2005,fitzsimmons_2010}, and as a result of experimental evolution \citep{luckinbill_1978,luckinbill_1979}.
+From the perspective of our model, this correlation is not surprising since the saturation density, which is determined by a balance between births and deaths, increases with $b$.
+
+%While adaptive evolution in the direction predicted by our model does produce traits broadly consistent with Grime's scheme, trait data on selection for higher $b$ at high density was ambiguous.
 This prediction is also counter to the expectations of MacArthur's $r$/$K$ dichotomy \citep{macarthur_1967}, since $b$ is closely related to the maximal, low-density growth rate $r=b-d$ \citep{pianka_1972}, yet in the $r$/$K$ scheme, high density populations should be subject to $K$, not $r$, selection.
@@
 
-Confusingly, the term ``$K$-selection'' sometimes refers generally to selection at high density \citep{pianka_1972}, encompassing both selection for higher saturation density \citep{macarthur_1967} and competitive ability \citep{gill_1974}.
-Contrary to an $r$/$K$ dichotomy, empirical studies have shown that maximal growth rate and saturation density (measured by abundance) are positively correlated, both between species/strains \citep{luckinbill_1979,kuno_1991,hendriks_2005,fitzsimmons_2010}, and as a result of experimental evolution \citep{luckinbill_1978,luckinbill_1979}.
-From the perspective of our model, this correlation is not surprising since the saturation density, which is determined by a balance between births and deaths, increases with $b$.
-
 There is support for a negative relationship between competitive success at high density and maximal growth rate \citep{luckinbill_1979}, consistent with an $r$/$K$ dichotomy.
@@
 $b$ in our model.
-Our $c$ is strictly competitive ability only --- as such, changes in $c$ do not directly affect population density (the total number of territories occupied in an iteration is $\Delta_+ N=U(1-e^{-L})$, which does not depend directly on the $c_i$).
+Our $c$ is strictly competitive ability only --- as such, changes in $c$ do not directly affect population density (the total number of territories occupied per iteration is $\Delta_+ N=U(1-e^{-L})$, which does not depend directly on the $c_i$).
 The clean separation of a strictly-relative $c$ parameter is particularly useful from an evolutionary genetics perspective, essentially embedding a zero-sum relative fitness trait within a non-zero-sum fitness model.
@@
 While our model can be applied to species rather than genotypes (e.g.
-ecological invasions), our focus is genotype evolution.
+ecological invasions), our focus is genotype evolution i.e.
+the change in allele frequencies over time.
 Our assumption that there are no large $c$ discrepancies (section ``Mean field approximation'') amounts to a restriction on the amount of genetic variation in $c$ in the population.
@@
 
-%\subsection{Logistic growth and classic nesting sites model}
+\subsection*{Comparison with simulations}
 
-%In the nesting sites model, it is assumed that propagules landing on occupied territory do not survive, while those landing on unoccupied territory immediately occupy the territory.
-Thus, the probability that a propagule survives is simply $(1-N/K)$, where $N=\sum_i n_i$ is the total population size.
-Consequently, $n_i$ increases by $1$ over the interval $[t,t+\Delta t]$ with probability $b_i n_i (1-N/K)\Delta t$.
-Provided that the abundances $n_i$ are large enough that demographic stochasticity is negligible (i.e.
-they have established), we can treat $n_i(t)$ as a continuous variable with growth given by the expectation of $b_i n_i (1-N/K)\Delta t$, and so, taking $\Delta t \rightarrow 0$ we obtain
-%\begin{equation}
-%\frac{dn_i}{dt}=b_i\left(1-\frac{N}{K}\right)n_i. \label{eq:logistic}
-%\end{equation}
+Fig.~\ref{fig:simcomp} shows that Eq.
+\eqref{eq:master} and its components closely approximate our density-dependent lottery model over a wide range of propagule densities (the latter is evaluated by direct simulations of uniform random dispersal and lottery competition).
+Two genotypes are present, one of which is at low frequency.
+The growth of the low-frequency genotype relies crucially on the low-density competition term $R_i c_i/\overline{c}$, and also to a lesser extent on the high density competition term $A_i c_i/\overline{c}$ if $l_1$ is large enough (Fig.~\ref{fig:simcomp}b).
+On the other hand, $R_i c_i/\overline{c}$ is negligible for the high-frequency genotype, which depends instead on high density territorial victories (Fig.~\ref{fig:simcomp}d).
+Fig.
+3 also shows the breakdown of the classic lottery model at low propagule densities.
 
-%\section{Territorial availability}
-%
-%In Sec.
-\ref{sec:c} we assumed that adults have the same territorial requirement, regardless of genotype.
-This can be generalized to allow genotypes to differ in their territorial requirements.
-Each adult from genotype $i$ now requires $t_i$ units of territory, $T$ is the total territory available to the population, and $T-\sum_i t_i$ units of territory are unoccupied.
-%
-%Competition between propagules with different territorial requirements is potentially much more complicated than the model in Sec.
-\ref{sec:c}, because competition is no longer neatly partitioned into a set of identical territories.
-The larger a propagule's territorial requirement, the more neighboring propagules it will interact with during its development to adulthood.
-%
-%
-%A simple way to avoid these complications is to partition the unoccupied territory into uniform units of territory with size given by the largest territorial requirement $t^*={\rm max}_i t_i$.
-The uniform territory approach used in Sec.
-\ref{sec:c}
-%
-%The biological intuition behind this 
-%
-%Thus $U=(T-\sum_i t_i n_i)/t^*$.
+\begin{figure}
+\centering
+\includegraphics[scale=0.7]{simulationcomparison.pdf}
+\caption{\label{fig:simcomp} The change in genotype abundances in a density dependent lottery model is closely approximated by Eq.
+\eqref{eq:master}.
+$\Delta_+ n_i/m_i$ from Eq.
+\eqref{eq:master} (and its separate components) are shown, along with direct simulations of random dispersal and lottery competition over one iteration over a range of propagule densities ($U$ is varied between $5\times 10^3$ and $10^6$ with $m_1=10^4$ and $m_2=9\times 10^4$).
+Two genotypes are present.
+(a) and (b) show the low-frequency genotype with $c$-advantage ($c_1=1.5$), (c) and (d) show the high-frequency predominant genotype ($c_2=1$).
+Simulation points are almost invisible in (c) and (d) due to near exact agreement with Eq.
+\eqref{eq:master}.
+Dashed lines in (a) and (c) show the breakdown of the classic lottery model.}
+\end{figure}
+
+\section*{Appendix C: Direction of evolution under different environmental extremes}
+
+Summing Eq. \eqref{eq:master} over genotypes, $N$ obeys 
+\begin{equation}
+\Delta N=U(1-e^{-L})-\sum_i d_i n_i,\label{eq:deltN}
+\end{equation}
+where $U=\sum_i b_i n_i/L$.
+In equilibrium ($\Delta N=0$), Eq.
+\eqref{eq:deltN} implies $L/(1-e^{-L})=\sum_i b_i n_i/\sum_i d_i n_i$, and so $L$ only depends on the ratio of population mean birth rate to population mean death rate.
+
+and so the propagule density is high $L\approx b_i/d_i\gg 1$, and every unoccupied territory will be heavily contested.
+The population density is also high $N/T\approx 1$ (since $L=b_iN/(N-T)=b_i/(1-T/N)$).
+
+%From Eq.
+\eqref{eq:deltN}, the single genotype equilibrium is given by $L/(1-e^{-L})=d_i/[(1-d_i)b_i]$, and since $L\ll 1$ and $N/T\ll 1$ due to high mortality, we have $L\approx 2(1-d_i/[(1-d_i)b_i])$.
+Clearly $b_i$ must be exceptionally large to ensure population persistence.
+The terms proportional to $c_i/\overline{c}$ in Eq.
+\eqref{eq:master} are then negligible, and $\Delta_+ n_i$ depends primarily on $b_i$.
 

### 2017-04-27 simulations.py

@@ -164,7 +164,7 @@ def d(t):
 c=np.array([1.,1.])
 T=100000
 g=40.     #iterations per seasonal cycle
-bparam=np.array([[0.0,0.0],[.5,.2]])
+bparam=np.array([[0.0,0.0],[.5,.217]])
 dparam=np.array([[0.0,0.0],[0.2,0.1]])
 totaltime=100
 
@@ -175,16 +175,6 @@ for t in range(totaltime):
     n = n + deltnplus((b(t)*n*U/T).astype(np.int),c,U.astype(np.int))-d(t)*n
     nhist.append(list(n))
     
-#n=np.array([60000.,25000.])
-#nhistclassic=[n];
-#for t in range(totaltime):
-#    U=T-sum(n)
-#    n = n + deltnplusclassic((b(t)*n*U/T).astype(np.int),c,U.astype(np.int))-d(t)*n
-#    nhistclassic.append(list(n))
-#
-#plt.stackplot(range(totaltime+1),np.array(nhistclassic)[:,1],np.array(nhistclassic)[:,0],colors=['b','k'])
-#plt.plot(np.array(nhistclassic)[:,0]/np.sum(nhistclassic,1),'k',lw=2)
-    
     
 #n=np.array([6000.,2500.])
 #nsimhist=[n]
@@ -193,7 +183,59 @@ for t in range(totaltime):
 #    n = n + sum(deltnplussim((b(t)*n*U/T).astype(np.int),c,U.astype(np.int)))-d(t)*n
 #    nsimhist.append(list(n))
 
-fig=plt.subplot(311)
+fig=plt.subplot(321)
+for t in range(0,100,40):
+    fig.plot(range(t,t+20),[dparam[1][0] for _ in range(t,t+20)],'k.',lw=2)
+    fig.plot(range(t,t+20),[dparam[1][1] for _ in range(t,t+20)],'b.',lw=2)
+    
+fig.fill([20,40,40,20],[0,0,1,1],'g',alpha=0.3,edgecolor='none')
+fig.fill([60,40+40,40+40,60],[0,0,1,1],'g',alpha=0.3,edgecolor='none')
+fig.plot(range(20,40),[bparam[1][0] for _ in range(t,t+20)],'k.',lw=2)
+fig.plot(range(20,40),[bparam[1][1] for _ in range(t,t+20)],'b.',lw=2)
[…]

### 2017-04-27 manuscript.tex
latest

@@
 
-The ideal environment is characterized by the near-absence of stress and disturbance: $d\ll 1$, whereas $b$ is potentially much larger than $1$.
+Ideal environments are characterized by the near-absence of stress and disturbance: $d\ll 1$, whereas $b$ could potentially be much larger than $1$.
 
@@
 The alternative view is that stressed environments are highly competitive.
-In spite of inhibited growth and reproduction, population density $N/T$ is actually high since stressful environments also have lower environmental carrying capacity \citep{taylor_1990}.
+In spite of inhibited growth and reproduction, density is actually high since stressful environments also have lower environmental carrying capacity \citep{taylor_1990}.
 For example, if consumable resources are scarce, we expect intense resource competition (for empirical support, see \citealt{davis_1998}).
@@
 The mapping of different environments to our model parameters is summarized in the first two rows of Fig.~\ref{fig:table}.
-Also shown is the approximate dependence of $\Delta_+ n_i$ on $b_i$ and $c_i$ for each environment (fourth row; details in Appendix).
-These can be used to infer the expected direction of evolution for the traits $b$, $c$ and $d$ (fifth row) using a standard invasion analysis.
-
+These can be used to infer the expected direction of evolution for the traits $b$, $c$ and $d$ (fourth row; Appendix C).
 According to Haldane's formula, the probability that an adaptive mutant survives its initial low abundance phase and begins to grow deterministically is proportional to its expected absolute growth rate \citep{haldane_1927,uecker_2011}, with a proportionality factor typically of order one.
@@
 The direction of evolutionary change is determined by the mutational trait changes which are available and also confer an appreciable fitness benefit \citep{yampolsky_2001}, where availability is subject to constraints imposed by the environment.
-
-For example, in Grime's interpretation of stressful environments, $L$ is low, so competition is not important, and only mutants with greater $b$ or lower $d$ will have an appreciably greater $\Delta n_i$.
-Mutations in $c$ are effectively neutral, and will rarely fix.
-However, $b$ is constrained to be small.
-Thus, while some rare mutations may produce small improvements in $b$, it is much more likely that mutations lower $d$, making this the expected direction of evolutionary change.
+We evaluate the  invasion fitness of mutants under Eq.
+\eqref{eq:master} for each environmental regime in Appendix C.
 
@@
   Other parameters & $b\gg d$ & $b\gg d$ & $b\approx d$ & $b>d$ \\
-  Density $N/T$  & High & Low & Low & High \\
+  Density $L$  & $\gg 1$ & $\ll 1$ & $\ll 1$ & $\gg 1$ \\
   Evolution for & $\uparrow b$, $ \uparrow c$ & $\uparrow b$, $\downarrow d$ & $\downarrow d$ & $\uparrow c, \downarrow d$
@@
 
-and so the propagule density is high $L\approx b_i/d_i\gg 1$, and every unoccupied territory will be heavily contested.
-The population density is also high $N/T\approx 1$ (since $L=b_iN/(N-T)=b_i/(1-T/N)$).
+We consider the invasion of mutants into a population with a single type $i$ in equilibrium. The fitness of low frequency invader types $j$ is
+\begin{equation}
+\Delta n_j/n_j = b_j \left(e^{-L}+R_j\frac{c_j}{c_i}\right)-d_j \label{eq:invad}
+\end{equation}
+where $R_j\approx (1-e^{-L})/\left(\frac{c_j}{c_i}+\frac{L-1-e^{-L}}{1-(1+L)e^{-L}}\right)$.
+
+In ideal environments, $d$ is much smaller than $1$ whereas $b$ is not. Thus, $1 \ll b_i/d_i = L/(1-e^{-L}) \approx L$, and Eq. \eqref{eq:invad} becomes 
+\begin{equation}
+\Delta n_j/n_j \approx \frac{b_j c_j}{c_j + L-1}-d_j, \label{eq:invad}
+\end{equation}
+where without loss of generality we set the reference competitiveness at $c_i=1$.
 

### 2017-04-28 manuscript.tex
latest

@@
 
-Stable coexistence is possible between genotypes that are superior in different traits.
-Suppose that $j$ is better at securing territories ($c_j>c_i$), that $i$ is better at producing propagules ($b_i>b_j$), and that $d_i=d_j$.
+Now suppose that $j$ is better at securing territories ($c_j>c_i$), that $i$ is better at producing propagules ($b_i>b_j$), with $d_i=d_j$.
 Coexistence occurs if $j$ will invade an $i$-dominated population, but $i$ will also invade a $j$-dominated population (``mutual invasion'').
@@
 These forms of co-existence require density dependence, and are not present in the classic lottery model.
-Mutual invasibility is not possible between $b$- and $d$-specialists.
+Mutual invasion is not possible between $b$- and $d$-specialists.
 
@@
 
-Summing Eq. \eqref{eq:master} over genotypes, $N$ obeys 
+In this Appendix we consider the invasion of mutants into a population with a single type $i$ in equilibrium under the different environments from section ``Primary strategies and Grime's triangle''.
+We first evaluate the ideal and stressful environments (the treatment of mortality is different for disturbed environments).
+
+Summing Eq. \eqref{eq:master} over genotypes, the change in $N$ is given by  
 \begin{equation}
@@
 \end{equation}
-where $U=\sum_i b_i n_i/L$.
-In equilibrium ($\Delta N=0$), Eq.
-\eqref{eq:deltN} implies $L/(1-e^{-L})=\sum_i b_i n_i/\sum_i d_i n_i$, and so $L$ only depends on the ratio of population mean birth rate to population mean death rate.
+where $U=\sum_i b_i n_i/L$. In equilibrium ($\Delta N=0$), this implies $L/(1-e^{-L})=b_i/d_i $. The fitness of a low frequency invader $j$ is
+\begin{equation}
+\Delta n_j/n_j \approx b_j \left(e^{-L}+R_jc_j\right)-d_j \label{eq:invad}
+\end{equation}
+where $R_j\approx (1-e^{-L})/\left(c_j+\frac{L-1-e^{-L}}{1-(1+L)e^{-L}}\right)$, and we set the reference competitiveness at $c_i=1$ without loss of generality.
 
-We consider the invasion of mutants into a population with a single type $i$ in equilibrium. The fitness of low frequency invader types $j$ is
+According our representation of Grime's stressful environment, $b\ll 1$, $d\approx b$ and thus $L\ll 1$. Then Eq. \eqref{eq:invad} becomes $\Delta n_j/n_j \approx b_j-d_j$.
+
+According to our definition of an ideal environment, $d$ is much smaller than $1$ whereas $b$ is not. Thus, $1 \ll b_i/d_i = L/(1-e^{-L}) \approx L$, and Eq. \eqref{eq:invad} becomes 
 \begin{equation}
-\Delta n_j/n_j = b_j \left(e^{-L}+R_j\frac{c_j}{c_i}\right)-d_j \label{eq:invad}
+\Delta n_j/n_j \approx \frac{b_j c_j}{c_j + L-1}-d_j, 
 \end{equation}
-where $R_j\approx (1-e^{-L})/\left(\frac{c_j}{c_i}+\frac{L-1-e^{-L}}{1-(1+L)e^{-L}}\right)$.
 
-In ideal environments, $d$ is much smaller than $1$ whereas $b$ is not. Thus, $1 \ll b_i/d_i = L/(1-e^{-L}) \approx L$, and Eq. \eqref{eq:invad} becomes 
-\begin{equation}
-\Delta n_j/n_j \approx \frac{b_j c_j}{c_j + L-1}-d_j, \label{eq:invad}
-\end{equation}
-where without loss of generality we set the reference competitiveness at $c_i=1$.
-
-%From Eq.
+From Eq.
 \eqref{eq:deltN}, the single genotype equilibrium is given by $L/(1-e^{-L})=d_i/[(1-d_i)b_i]$, and since $L\ll 1$ and $N/T\ll 1$ due to high mortality, we have $L\approx 2(1-d_i/[(1-d_i)b_i])$.

### 2017-04-29 manuscript.tex
latest

@@
 \end{equation}
-To supplement our analytical mean field derivation, we verified that Eq.
-\eqref{eq:master} closely approximates the exact behavior of our density-dependent lottery model using numerical simulations (Appendix B).
+To supplement our analytical mean field derivation, we did numerical simulations of our exact our density-dependent lottery model, and verified  that Eq.
+\eqref{eq:master} is a good approximation (Appendix B).
 Thus, Eq.
@@
 Here we describe how Grime's ``disturbed'', ``stressful'' and ``ideal'' environments can be captured mathematically in our scheme.
-To proceed, we need to map these verbally-defined environments to quantitative parameter regimes in our model.
+To proceed, we  map these verbally-defined environments to quantitative parameter regimes in our model.
 
-Ideal environments are characterized by the near-absence of stress and disturbance: $d\ll 1$, whereas $b$ could potentially be much larger than $1$.
+Ideal environments are characterized by the near-absence of stress and disturbance, so that survival is easy $d\ll 1$ and growth and reproduction can be rapid $b\gg d$.
+The propagule density is determined by the magnitude of $b/d$, and in this case $L\gg 1$ (Appendix C).
 
 Disturbed environments are characterized by short bursts of high extrinsic mortality $d$ caused by physical destruction.
-We assume that in the disturbance extreme of Grime's triangle, many such bursts (which could each be be spatially localized) will occur each iteration (i.e.
-over the time from birth to reproductive maturity), and so we can approximate extreme disturbance by high constant mortality rates $d_i$, which are now extended to affect juveniles as well.
-We assume that disturbances are equally damaging to adults and juveniles, so that only $(1-d_i)\Delta_+ n_i$ rather than $\Delta_+ n_i$ territories are secured by genotype $i$ each iteration.
-Heavily disturbed environments then correspond most adults and juveniles being killed every an iteration ($d_i\approx 1$).
+We assume that disturbances are equally damaging to adults and juveniles, so that only $(1-d_i)\Delta_+ n_i$ rather than $\Delta_+ n_i$ territories are secured by type $i$ each iteration.
+In the heavily disturbed extreme of Grime's triangle, many such bursts (which could each be spatially localized) will occur each iteration (i.e.
+over the time required for a propagule to reach reproductive maturity), which we approximate as almost all adults and juveniles dying every iteration $d\approx 1$.
+Population persistence is only possible if $(1-d)b>d$, and so $b\gg 1$ (Appendix C).
+We assume that fulfilling this requirement presents such a challenge that $(1-d)b$ does not greatly exceed $d$, which implies $L\ll 1$.
 
 Stressful environments are more ambiguous, and have been the subject of an extensive debate in the plant ecology literature (the ``Grime-Tilman'' debate; \citealt{aerts_1999} and references therein).
-Severe stress inhibits growth and reproduction, so that $b\ll 1$ \citep{grime_1974,grime_1977}.
-In Grime's view, this means that the rate at which propagules successfully develop to adulthood cannot appreciably exceed the mortality rate $b/d\approx 1$.
-In our model, this implies low propagule density and little territorial contest.
+Severe stress inhibits growth and reproduction, so that $b\ll 1$.
+In Grime's view, this means that the rate at which propagules successfully develop to adulthood cannot appreciably exceed the mortality rate $b/d\approx 1$; coping with the environmental stress is barely possible \citep{grime_1974,grime_1977}.
+In our model, this implies $L\ll 1$.
 
 The alternative view is that stressed environments are highly competitive.
-In spite of inhibited growth and reproduction, density is actually high since stressful environments also have lower environmental carrying capacity \citep{taylor_1990}.
-For example, if consumable resources are scarce, we expect intense resource competition (for empirical support, see \citealt{davis_1998}).
-In our model, territorial contests will then be important.
-
-The mapping of different environments to our model parameters is summarized in the first two rows of Fig.~\ref{fig:table}.
-These can be used to infer the expected direction of evolution for the traits $b$, $c$ and $d$ (fourth row; Appendix C).
-According to Haldane's formula, the probability that an adaptive mutant survives its initial low abundance phase and begins to grow deterministically is proportional to its expected absolute growth rate \citep{haldane_1927,uecker_2011}, with a proportionality factor typically of order one.
-The fixation of neutral mutations is exceedingly unlikely (probability of order $1/N$).
-The direction of evolutionary change is determined by the mutational trait changes which are available and also confer an appreciable fitness benefit \citep{yampolsky_2001}, where availability is subject to constraints imposed by the environment.
-We evaluate the  invasion fitness of mutants under Eq.
-\eqref{eq:master} for each environmental regime in Appendix C.
+In spite of inhibited growth and reproduction, density is actually high relative to carrying capacity of the stressful environment \citep{taylor_1990}.
+For example, if consumable resources are scarce, we expect intense resource competition \citep{davis_1998}.
+Thus, even though $b\ll 1$, we can still have $b/d\gg 1$ and $L\ll 1$.
 
@@
   Constraints & $d \ll 1$ & $d \approx 1$ & $b \ll 1$ & $b \ll 1$ \\
-  Other parameters & $b\gg d$ & $b\gg d$ & $b\approx d$ & $b>d$ \\
-  Density $L$  & $\gg 1$ & $\ll 1$ & $\ll 1$ & $\gg 1$ \\
-  Evolution for & $\uparrow b$, $ \uparrow c$ & $\uparrow b$, $\downarrow d$ & $\downarrow d$ & $\uparrow c, \downarrow d$
+  Other parameters & $b/d\gg 1$ & $(1-d)b\approx d$ & $b\approx d$ & $b/d\gg 1$ \\
+  Density $L$  & $\gg 1$ & $\ll 1$ & $\ll 1$ & $\gg 1$
 \end{tabular}
-\caption{\label{fig:table} The realization of Grime's three environmental extremes in our model, as well as the high-density variant of the stressful environment.
-Shown are the mapping of each environment to our parameters, the approximate dependence of $\Delta_+ n_i$ on $b_i$ and $c_i$, as well as the corresponding expected evolutionary changes in $b_i$, $c_i$ and $d_i$.
-*Mortality affects both adults and juveniles in the disturbed environment, with $\Delta_+ n_i$ replaced by $(1-d_i)\Delta_+ n_i$ in Eq.
-\eqref{eq:delttot}.}
+\caption{\label{fig:table} The mathematical realization of Grime's three environmental extremes in our model, as well as the high-density variant of the stressful environment.
+*Mortality affects both adults and juveniles in the disturbed environment, with $\Delta_+ n_i$ replaced by $(1-d_i)\Delta_+ n_i$.}
 \end{figure}
+
+Given our mathematical representation of Grime's environmental extremes, summarized in Fig.~\ref{fig:table}, the expected direction of evolution is determined by three factors.
+First, the probability that an adaptive mutant survives its initial low abundance phase and begins to grow deterministically is proportional to its expected absolute growth rate \citep{haldane_1927,uecker_2011}.
+By comparison, neutral mutations rarely fix (probability of order $1/N$).
+Thus $b$, $c$ and $d$ will tend to change in the direction of greatest fitness increase.
+Under Eq.
+\eqref{eq:master}, proportional changes of the form $b_j=(1+\epsilon) b_i$ or $d_j=(1+\epsilon) d_i$ have identical fitness effects
+
+When propagule density is low $L\ll 1$, $c$ does not contribute to fitness.
+
+Second, the fact that some mutations are more common than others can change the direction of evolution \citep{yampolsky_2001}, and in particular, very few mutations can perform difficult trait changes such as increasing $b$ under stress.
+These mutational constraints are different under different environments (Fig.~\ref{fig:table}).
+
+Finally, there may be pleiotropy between $b$, $c$, and $d$, which could be due to trade-offs between them.
+Here we do not attempt to account for the complications posed by pleiotropy, and treat $b$, $c$ and $d$ independently.
 
@@
 
-In this Appendix we consider the invasion of mutants into a population with a single type $i$ in equilibrium under the different environments from section ``Primary strategies and Grime's triangle''.
+In this Appendix we consider the invasion of mutants into a population with a single resident type $i$ in equilibrium under the different environments from section ``Primary strategies and Grime's triangle''.
 We first evaluate the ideal and stressful environments (the treatment of mortality is different for disturbed environments).
 
-Summing Eq. \eqref{eq:master} over genotypes, the change in $N$ is given by  
+Summing Eq. \eqref{eq:growthsumuncoupled} over genotypes, the change in $N$ is given by  
 \begin{equation}
@@
 
-According our representation of Grime's stressful environment, $b\ll 1$, $d\approx b$ and thus $L\ll 1$. Then Eq. \eqref{eq:invad} becomes $\Delta n_j/n_j \approx b_j-d_j$.
+According our representation of Grime's stressful environment, $b\ll 1$, $d\approx b$ and thus $L\ll 1$.
+Then Eq.
+\eqref{eq:invad} becomes $\Delta n_j/n_j \approx b_j-d_j$.
+Mutations which improve $b$ or $d$ by the same fraction $\epsilon$ (such that $b_j = (1+ \epsilon)b_i$ or $d_j = (1- \epsilon)d_i$) will therefore yield identical fitness benefits.
+Stressful environments will therefore favor $d$-specialist ``tolerators'', which bring their mortality rates down enough to enable survival in the face of low $b$, only if stress makes it harder to increase $b$ than lower $d$, such that adaptive mutations of a given effect size $\epsilon$ occur more often for $d$.
 
-According to our definition of an ideal environment, $d$ is much smaller than $1$ whereas $b$ is not. Thus, $1 \ll b_i/d_i = L/(1-e^{-L}) \approx L$, and Eq. \eqref{eq:invad} becomes 
+In Grime's ideal environment, $d$ is much smaller than $1$ whereas $b$ is not. Thus, $1 \ll b_i/d_i = L/(1-e^{-L}) \approx L$, and Eq. \eqref{eq:invad} becomes 
 \begin{equation}
-\Delta n_j/n_j \approx \frac{b_j c_j}{c_j + L-1}-d_j, 
+\Delta n_j/n_j \approx \frac{b_j c_j}{c_j + L-1}-d_j\approx \frac{b_j c_j}{L}-d_j, \label{eq:highLinvad}
 \end{equation}
+where the last approximation follows from the fact that $c_j$ will be of similar magnitude to $c_i=1$ (ignoring mutations with very large effect sizes; compare with Eq.
+\eqref{eq:lottery}, where the same expression comes from assuming that $L$ is effectively infinite).
+It is easily verified from Eq.
+\eqref{eq:highLinvad} that improving $b$, $c$, or $d$ by a fraction $\epsilon$ yields identical fitness benefits.
+This may seem counter-intuitive in that reducing an already tiny mortality rate by a fraction $\epsilon$ seems to yield marginal benefits compared to increasing a  potentially large propagule production $b$ by the same fraction.
+However, the relative magnitudes of $b$ and $d$ are irrelevant, because the ancestral resident type's density-regulated birth rate $b_i c_i/L$ exactly balances its mortality $d_i$ in equilibrium.
 
-From Eq.
-\eqref{eq:deltN}, the single genotype equilibrium is given by $L/(1-e^{-L})=d_i/[(1-d_i)b_i]$, and since $L\ll 1$ and $N/T\ll 1$ due to high mortality, we have $L\approx 2(1-d_i/[(1-d_i)b_i])$.
-Clearly $b_i$ must be exceptionally large to ensure population persistence.
-The terms proportional to $c_i/\overline{c}$ in Eq.
-\eqref{eq:master} are then negligible, and $\Delta_+ n_i$ depends primarily on $b_i$.
+Biases in the direction of evolution of $b$, $c$ or $d$ must again follow from biases in mutation availability or trade-offs/pleiotropy.
+The same argument applies for high-$L$ interpretation of the stress regime.
+
+To represent Grime's disturbed environment, we account for juvenile mortality due to disturbance by replacing $\Delta_+ n_i$ with $(1-d_i)\Delta_+ n_i$.
+For a single type $i$ in equilibrium, Eq.
+\eqref{eq:deltN} becomes
+\begin{equation}
+\Delta N=(1-d_i)U(1-e^{-L})-d_i N,\label{eq:deltNdist}
+\end{equation}
+which reaches equilibrium when $L/(1-e^{-L})=(1-d_i)b_i/d_i$ (note that population persistence now requires $(1-d_i)b_i>d_i$).
+We represent Grime's disturbed environment by $d_i\approx 1$ and $(1-d_i)b_i\approx d_i$, so that $L\ll 1$.
+A mutant's invasion fitness is then given by
+\begin{equation}
+\Delta n_j/n_j = (1-d_j)b_j - d_j.
+\end{equation}
+The effects of improving $b$ and $d$ are now no longer identical.
+Increasing $b$ by a fraction $\epsilon$ gives $\Delta n_j/n_j = \epsilon (1-d_i)b_i$, whereas reducing $d$ by the same fraction gives $\Delta n_j/n_j = \epsilon b_i$, a much greater advantage given that $d_i\approx 1$.
+Thus, selection intrinsically favors reducing $d$ in disturbed environments.
 

### 2017-05-02 manuscript.tex
latest

@@
 Since it enters fundamentally into many population genetics considerations, it is remarkable how little attention has been paid to it.'' --- Warren J.
-Ewens, Mathematical Population Genetics I, 2004 [Should be removed if we need space]
+Ewens, Mathematical Population Genetics I, 2004, pp.
+276
 
 Evolutionary models differ greatly in their treatment of fitness.
-In models of genetic evolution, genotypes are typically assigned constant (or frequency-dependent) selection coefficients describing the change in their relative frequencies over time.
-This simplified treatment of selection facilitates genetic realism, and can be justified over sufficiently short time intervals \citep[p.
+In models of genetic evolution, genotypes are typically assigned constant (or occasionally frequency-dependent) selection coefficients describing the change in their relative frequencies over time.
+This simplified treatment of selection facilitates explicit time-dependent treatment of genotype frequencies, and can be justified over sufficiently short time intervals \citep[p.
 276]{ewens_2004}.
-The emphasis here is to infer past selection, migration and demographic change given a sample of nucleotide sequences, or to predict how allele frequencies change over time based on their fitness effects, population structure, genetic drift and linkage.
+The emphasis of population genetics is to infer past selection, migration and demographic change given a sample of nucleotide sequences, or to predict how allele frequencies change over time based on their relative fitness effects together with population structure, genetic drift and linkage.
 The resulting picture of evolution excludes basic elements of the ecological underpinnings of selection, including density dependence, and how selection affects population size.
-This complicates the inference of past selection, because demographic changes can look geneaologically very similar to selective frequency changes \citep{barton_1998}.
+This complicates the inference of past selection, because demographic changes can look genealogically very similar to selective frequency changes \citep{barton_1998}.
 
@@
 This broadly follows the original spirit of MacArthur's $r/K$ scheme.
-More specifically, our aim is to begin adding some ecological realism to population genetics' time-depedendent, genetically-focused view of evolution.
+More specifically, our aim is to begin merging some ecological realism to population genetics' time-dependent, genetically-focused view of evolution.
 We revisit the classic lottery model of \cite{chesson_1981}, which has two features that make it well suited for this role, but one critical flaw that we rectify here.
@@
 
-Using our extended lottery model, we then revisit Grime's C/S/R scheme,  and evaluate how C/S/R manifests at the level of within-population genotypic evolution (as opposed to phenotypic divergence between species; sections ``Invasion of rare genotypes and coexistence'' and ``Primary strategies and Grime's triangle'').
+Using our extended lottery model, we then revisit Grime's C/S/R scheme,  and evaluate how C/S/R manifests at the level of within-population genotypic evolution (as opposed to phenotypic divergence between species; section ``Primary strategies and Grime's triangle'').
 This represents a ``sanity check'' on our density-dependent lottery model.
@@
 Taking an example inspired by recent studies of rapid, seasonal evolution in \textit{Drosophila} \citep{bergland_14}, we discuss how environmental fluctuations might stabilize polymorphisms in the presence of cyclical population density.
-
-%Fitness models like Eq. \eqref{eq:master} may help to address to these difficulties since they do not artificially separate the dynamics of abundance and frequency.
 
@@
 We derive deterministic equations for the expected change in the $n_i$ over time, leaving the evaluation of drift for future work.
-In particular, we do not evaluate the initial stochastic behaviour of adaptive mutant lineages while they are at low abundance.
-When considering mutations, we restrict our attention to the earliest deterministic behavior of mutant lineages while they are still at low frequency (the transition to deterministic growth occurs at an abundance of order equal to their inverse expected absolute growth rate; \citealt{uecker_2011}).
+This is an excellent approximation when the $n_i$ are all large.
+How ever, we also do not evaluate the initial stochastic behaviour of adaptive mutant lineages while they are at low abundance.
+When considering new mutations, we therefore restrict our attention to begin with the earliest (lowest $n_i$) deterministic behavior of mutant lineages (the transition to deterministic growth occurs at an abundance $n_i$ of order equal to their inverse expected absolute growth rate; \citealt{uecker_2011}).
 
@@
 We assume that $x_i$ follows a Poisson distribution $p_i(x_i)=l_i^{x_i} e^{-l_i}/x_i!$, where $l_i=m_i/U$ is the mean territorial propagule density.
-This only approximates uniform random dispersal, but is essentially exact provided that the $n_i$ are large enough that drift can be ignored (Appendix A).
+This is approximation becomes exact when the $n_i$ are large enough that drift in $n_i$ can be ignored (Appendix A).
 
@@
 Since birth and death rates affect absolute abundances, this allows us to evaluate selection at different densities (after appropriate extensions are made), in an otherwise very similar model to the canonical Wright-Fisher.
-We therefore expect that drift in our extended lottery model should be similar to that in the Wright-Fisher model, but we leave this for future work.
+We therefore expect that drift in realized values of $n_i$ in our extended lottery model should be similar to that in the Wright-Fisher model, but we leave this for future work.
 
@@
 Eq.
-\eqref{eq:lottery} is nonsensical at low densities ($l_i\ll 1$): genotype $i$ can win at most $m_i$ territories, yet Eq.
+\eqref{eq:lottery} is nonsensical if even a single type has low propagule density ($l_i\ll 1$): genotype $i$ can win at most $m_i$ territories, yet Eq.
 \eqref{eq:lottery} demands $c_i l_i/\sum_j c_j l_j$ of the $U$ unoccupied territories, for any value of $U$.
@@
 
-For the majority of this manuscript we assume that mortality only occurs in adults (setting aside the juvenile deaths implicit in territorial contest), and at a constant, genotype-specific per-capita rate $d_i$, so that the overall change in genotype abundances is
+For the majority of this manuscript we assume that mortality only occurs in adults (setting aside the juvenile deaths implicit in territorial contest), and at a constant, genotype-specific per-capita rate $0\leq d_i\leq 1$, so that the overall change in genotype abundances is
 \begin{equation}
@@
 
-In a population with a single genotype $i$ which is in equilibrium, $R_i=0$, $\overline{c}=c_i$ and $\Delta n_i = 0$, and Eq. \eqref{eq:master} gives
-\begin{equation}
-b_i\left(e^{-L}+A_i\right)-d_i=0,\label{eq:equil}
-\end{equation}
-where $A_i=(1-(1+L)e^{-L})/L$.
-Now suppose that a new genotype $j$, which is initially rare, appears in the population.
-Then $A_j\ll R_j$, $l_j\approx 0$ and $\overline{c}\approx c_i$, and so, from Eq.
-\eqref{eq:master}, $n_j$ will increase if
-\begin{equation}
-b_j \left(e^{-L}+R_j\frac{c_j}{c_i}\right)-d_j>0,\label{eq:invad}
-\end{equation}
-where $R_j\approx (1-e^{-L})/\left(\frac{c_j}{c_i}+\frac{L-1-e^{-L}}{1-(1+L)e^{-L}}\right)$.
+Stable coexistence is possible in our extended lottery model. Suppose that $j$ is better at securing territories ($c_j>c_i$) but that $i$ is better at producing propagules ($b_i>b_j$), with $d_i=d_j$.
 
-Now suppose that $j$ is better at securing territories ($c_j>c_i$), that $i$ is better at producing propagules ($b_i>b_j$), with $d_i=d_j$.
-Coexistence occurs if $j$ will invade an $i$-dominated population, but $i$ will also invade a $j$-dominated population (``mutual invasion'').
-If $b_i$ is so large that $L\gg 1$ when $i$ is dominant, and $b_j$ is so small that $L\ll 1$ when $j$ is dominant, then, combining Eqs.
-\eqref{eq:equil} and \eqref{eq:invad}, we find that $i$ invades $j$ because $b_i>b_j$, while $j$ invades $i$ provided that
-\begin{equation}
-b_jc_jR_j-b_i c_i A_i>0. \label{eq:jinvadcoex}
-\end{equation}
-Thus, coexistence occurs if $c_j/c_i$ is large enough.
  This is a version of the classic competition-colonization trade-off \citep{tilman_94,levins_71}: the competitor ($c$-specialist) leaves many territories unoccupied (low $L$) due to its poor colonization ability (low $b$), which the colonizer ($b$-specialist) can then exploit.
@@
 
-In this Appendix we consider the invasion of mutants into a population with a single resident type $i$ in equilibrium under the different environments from section ``Primary strategies and Grime's triangle''.
-We first evaluate the ideal and stressful environments (the treatment of mortality is different for disturbed environments).
+In this Appendix we consider mutant invasion of a resident type $i$ (such that $N=n_i$) which is in equilibrium.
 
-Summing Eq. \eqref{eq:growthsumuncoupled} over genotypes, the change in $N$ is given by  
+In a population with a single genotype $i$ which is in equilibrium ($\Delta n_i = 0$), $R_i=0$, $\overline{c}=c_i$, $A_i=(1-(1+L)e^{-L})/L$, and Eq. \eqref{eq:master} gives
 \begin{equation}
-\Delta N=U(1-e^{-L})-\sum_i d_i n_i,\label{eq:deltN}
+b_i(1-e^{-L})/L-d_i=0,\label{eq:equil}
 \end{equation}
-where $U=\sum_i b_i n_i/L$. In equilibrium ($\Delta N=0$), this implies $L/(1-e^{-L})=b_i/d_i $. The fitness of a low frequency invader $j$ is
+which implies $L\approx b_i/d_i$ when $b_i/d_i\gg 1$ and $L\ll 1$ when $b_i/d_i\approx 1$.
+(Eq.
+\eqref{eq:Da}{eq:equil} can also be seen directly from Eq.
+\eqref{eq:growthsumuncoupled}).
+Now suppose that a new genotype $j$, which is initially rare, appears in the population.
+Then $A_j/R_j\ll 0$, $l_j\approx 0$ and $\overline{c}\approx c_i$, and so, from Eq.
+\eqref{eq:master}, the mutant lineage's fitness is
 \begin{equation}
-\Delta n_j/n_j \approx b_j \left(e^{-L}+R_jc_j\right)-d_j \label{eq:invad}
+\Delta n_j/n_j \approx b_j \left(e^{-L}+R_j\frac{c_j}{c_i}\right)-d_j \label{eq:invad}
 \end{equation}
-where $R_j\approx (1-e^{-L})/\left(c_j+\frac{L-1-e^{-L}}{1-(1+L)e^{-L}}\right)$, and we set the reference competitiveness at $c_i=1$ without loss of generality.
+where $R_j\approx (1-e^{-L})/\left(\frac{c_j}{c_i}+\frac{L-1-e^{-L}}{1-(1+L)e^{-L}}\right)$.
 
-According our representation of Grime's stressful environment, $b\ll 1$, $d\approx b$ and thus $L\ll 1$.
-Then Eq.
+To see that coexistence is possible between $b$ and $c$ specialists, we check if $j$ will invade an $i$-dominated population, but $i$ will also invade a $j$-dominated population (``mutual invasion'').
+Suppose that $b_i$ is so large that $L\gg 1$ when $i$ is dominant, and $b_j$ is so small that $L\ll 1$ when $j$ is dominant.
+Then, combining Eqs.
+\eqref{eq:equil} and \eqref{eq:invad}, we find that $i$ invades $j$ because $b_i>b_j$, while $j$ invades $i$ provided that
+\begin{equation}
+b_jc_jR_j-b_i c_i A_i>0. \label{eq:jinvadcoex}
+\end{equation}
+Thus, coexistence occurs if $c_j/c_i$ is large enough.
+
+We now consider the different environments from section ``Primary strategies and Grime's triangle''.
+In our representation of Grime's stressful environment, $b\ll 1$, $d\approx b$ and $L\ll 1$, and so Eq.
 \eqref{eq:invad} becomes $\Delta n_j/n_j \approx b_j-d_j$.
-Mutations which improve $b$ or $d$ by the same fraction $\epsilon$ (such that $b_j = (1+ \epsilon)b_i$ or $d_j = (1- \epsilon)d_i$) will therefore yield identical fitness benefits.
-Stressful environments will therefore favor $d$-specialist ``tolerators'', which bring their mortality rates down enough to enable survival in the face of low $b$, only if stress makes it harder to increase $b$ than lower $d$, such that adaptive mutations of a given effect size $\epsilon$ occur more often for $d$.
+Mutations which improve $b$ or $d$ by the same fraction $\epsilon$, such that $b_j = (1+ \epsilon)b_i$ or $d_j = (1- \epsilon)d_i$,  yield identical fitness benefits.
 
-In Grime's ideal environment, $d$ is much smaller than $1$ whereas $b$ is not. Thus, $1 \ll b_i/d_i = L/(1-e^{-L}) \approx L$, and Eq. \eqref{eq:invad} becomes 
+In our representation of Grime's ideal environment, $d\ll 1$, $b_i/d_i \gg 1$ and $L\gg 1$, and so Eq. \eqref{eq:invad} becomes 
 \begin{equation}
@@
 \end{equation}
-where the last approximation follows from the fact that $c_j$ will be of similar magnitude to $c_i=1$ (ignoring mutations with very large effect sizes; compare with Eq.
-\eqref{eq:lottery}, where the same expression comes from assuming that $L$ is effectively infinite).
-It is easily verified from Eq.
-\eqref{eq:highLinvad} that improving $b$, $c$, or $d$ by a fraction $\epsilon$ yields identical fitness benefits.
-This may seem counter-intuitive in that reducing an already tiny mortality rate by a fraction $\epsilon$ seems to yield marginal benefits compared to increasing a  potentially large propagule production $b$ by the same fraction.
-However, the relative magnitudes of $b$ and $d$ are irrelevant, because the ancestral resident type's density-regulated birth rate $b_i c_i/L$ exactly balances its mortality $d_i$ in equilibrium.
-
+where the last approximation follows from the fact that $c_j$ will be of similar magnitude to $c_i=1$ (ignoring mutations with very large effect sizes; incidentally, Eq.
+\eqref{eq:lottery} yields the same expression by assuming $L\rightarrow \infty$ even though this is obviously unrealistic for a rare mutant).
+Similarly to Grime's stressful environment, improving $b$, $c$, or $d$ by a fraction $\epsilon$ yields identical fitness benefits, even though they may have very different magnitudes, because the ancestral resident type's density-regulated birth rate $b_i c_i/L$ exactly balances its mortality $d_i$ in equilibrium.
 Biases in the direction of evolution of $b$, $c$ or $d$ must again follow from biases in mutation availability or trade-offs/pleiotropy.
@@
 
-To represent Grime's disturbed environment, we account for juvenile mortality due to disturbance by replacing $\Delta_+ n_i$ with $(1-d_i)\Delta_+ n_i$.
-For a single type $i$ in equilibrium, Eq.
-\eqref{eq:deltN} becomes
-\begin{equation}
-\Delta N=(1-d_i)U(1-e^{-L})-d_i N,\label{eq:deltNdist}
-\end{equation}
-which reaches equilibrium when $L/(1-e^{-L})=(1-d_i)b_i/d_i$ (note that population persistence now requires $(1-d_i)b_i>d_i$).
-We represent Grime's disturbed environment by $d_i\approx 1$ and $(1-d_i)b_i\approx d_i$, so that $L\ll 1$.
+To represent Grime's disturbed environment, $\Delta_+ n_i$ is replaced with $(1-d_i)\Delta_+ n_i$, and a single type reaches equilibrium when $L/(1-e^{-L})=(1-d_i)b_i/d_i$ (population persistence now requires $(1-d_i)b_i>d_i$ rather than $b_i>d_i$).
+We assumed $d_i\approx 1$ and $(1-d_i)b_i\approx d_i$, so that $L\ll 1$.
 A mutant's invasion fitness is then given by
@@
 \end{equation}
+Another distinct feature of disturbed environments is that the bulk of mortality is environmental and unavoidable.
+That is, $d_i=d_{ie}+d_{ig}$, where the environmental part $d_{ie}\approx 1$ does not evolve, and the genetic part $d_{ig}\ll 1$ is affected by mutations in the same way as before $d_{jg}=(1-\epsilon) d_{ig}$.
+
 The effects of improving $b$ and $d$ are now no longer identical.
-Increasing $b$ by a fraction $\epsilon$ gives $\Delta n_j/n_j = \epsilon (1-d_i)b_i$, whereas reducing $d$ by the same fraction gives $\Delta n_j/n_j = \epsilon b_i$, a much greater advantage given that $d_i\approx 1$.
-Thus, selection intrinsically favors reducing $d$ in disturbed environments.
+Increasing $b$ by a fraction $\epsilon$ gives $\Delta n_j/n_j = \epsilon (1-d_i)b_i\approx \epsilon$, whereas reducing $d$ by the same fraction gives $\Delta n_j/n_j = \epsilon d_{ig}/(1-d_i)$.
+Thus, even though the genetic component of mortality $d_{g}$ is small compared to the environmental component $d_{e}$, if disturbance is intense enough ($1-d_i$ is small enough), selection for reduced mortality is potentially much stronger than that for propagule production.
 

### 2017-05-02 manuscript.tex
latest

@@
 
-In the previous section we only considered how $b$, $c$ and $d$ should respond in Grime's environmental extremes.
+In the previous section we only considered how $b$, $c$ and $d$ should respond in Grime's environmental extremes, based on invasion fitness.
 Here we further explore the low frequency behavior of Eq.
@@
 
-Stable coexistence is possible in our extended lottery model. Suppose that $j$ is better at securing territories ($c_j>c_i$) but that $i$ is better at producing propagules ($b_i>b_j$), with $d_i=d_j$.
-
+In a constant environment, stable coexistence is possible in our extended lottery model.
+A $b$ specialist $i$ and $c$-specialist $j$ ($b_i>b_j$, $c_j>c_i$) can co-exist because then propagule density $L$ is frequency dependent, and so is the importance of competitive ability (Appendix D).
 This is a version of the classic competition-colonization trade-off \citep{tilman_94,levins_71}: the competitor ($c$-specialist) leaves many territories unoccupied (low $L$) due to its poor colonization ability (low $b$), which the colonizer ($b$-specialist) can then exploit.
-A similar argument applies for coexistence between high-$c$ and low-$d$ specialists; a ``competition-longevity'' trade-off \citep{tilman_94}.
-These forms of co-existence require density dependence, and are not present in the classic lottery model.
-Mutual invasion is not possible between $b$- and $d$-specialists.
+A similar situation holds for coexistence between high-$c$ and low-$d$ specialists; a ``competition-longevity'' trade-off \citep{tilman_94}.
+These forms of co-existence require density dependence (being mediated by $L$), and are not present in the classic lottery model.
+Coexistence is not possible between $b$- and $d$-specialists (Appendix D).
 
@@
 \ref{fig:fluctuatingselection}d-f).
-As a result, coexistence requires that the winter survivor's $b$ must be about $5$ times smaller than predicted when we properly account for the growth in the abundance of each type using Eq.
+As a result, coexistence requires that the winter survivor's $b$ must be about $5$ times smaller than required when we properly account for the growth in the abundance of each type using Eq.
 \eqref{eq:master} (keeping the other parameters the same).
@@
 \includegraphics[scale=0.7]{fluctuatingselection.pdf}
-\caption{\label{fig:fluctuatingselection} Stable coexistence between $b$ and $d$ specialists in a fluctuating environment.
+\caption{\label{fig:fluctuatingselection} Stable coexistence between $b$ and $d$ specialists in a fluctuating environment requires a much greater $b$ advantage in the classic lottery model compared to our density-dependent extension of it when population density is seasonally cyclical.
 (a) Birth and death rates seasonally alternate being nonzero (white for winter, green for summer).
@@
 (c) Summer favors the $b$ specialist, winter the $d$ specialist, and they stably coexist.
-(d-f) Same as (a-c) for the classic lottery model, where coexistence now requires $b\approx 0.0421$ for the $d$ specialist.
+(d-f) Same as (a-c) for the classic lottery model; the types no longer coexist.
+(g-i) Same as (d-f) where now $b = 0.0421$ for the $d$ specialist and the types coexist.
 For illustration, the propagule abundances are assumed to have the form $m_i=b_i(1-N/T)n_i$, reflecting non-directed dispersal.}
@@
 
-In the introduction we mentioned the recurring difficulties with confounding selection and demography in population genetic inference.
-While we have not directly attempted to perform inference with our model, and Eq.
-\eqref{eq:master} may not be appropriate for particular inference problems, it seems that something similar (and hopefully more analytically tractable) to our extension of the lottery model is unavoidable because, fundamentally, selective births and deaths affect both  abundances and frequencies, not one or the other in isolation.
-Moreover, some aspects of allele frequency change are intrinsically density-dependent.
-In the classic lottery model, which as we have seen is essentially the Wright-Fisher model with overlapping generations, $b_i$ and $c_i$ are  equivalent in the sense that the number of territorial victories only depends on the product $b_i c_i$ (see ``Model'').
-This is no longer the case in our extension, where $b$ and $c$ specialists can co-exist.
-This ``colonization-competition trade-off'' is well known in the co-existence literature \citep{tilman_94}.
-It and similar forms of ``spatial co-existence'' in stable environments have previously been modeled either with Levin's  qualitative representation of competition \citep{levins_71,tilman_94}, as opposed to the quantitative $c$ of lottery competition, or with a more sophisticated treatment of space (non-uniform dispersal; \citealt{shmida_84,bolker_99}).
-In cyclical environments, polymorphisms can be stabilized by the bounded density effect which is  completely lost if there is an exclusive focus on allele frequencies \citep{yi_2013}.
-We leave the details of how our model might be applied in inference, including the crucial issue of its genetic drift predictions, for future work.
-
 It is interesting to compare the predictions of the extended lottery model with earlier approaches, such as the $r$/$K$ scheme, where $r=b-d$ is the maximal, low-density growth rate \citep{pianka_1972}.
 Confusingly, the term ``$K$-selection'' sometimes refers generally to selection at high density \citep{pianka_1972}, encompassing both selection for higher saturation density \citep{macarthur_1967} and competitive ability \citep{gill_1974}.
-Contrary to an $r$/$K$ dichotomy, empirical studies have shown that maximal growth rate and saturation density (measured by abundance) are positively correlated, both between species/strains \citep{luckinbill_1979,kuno_1991,hendriks_2005,fitzsimmons_2010}, and as a result of experimental evolution \citep{luckinbill_1978,luckinbill_1979}.
-From the perspective of our model, this correlation is not surprising since the saturation density, which is determined by a balance between births and deaths, increases with $b$.
+Contrary to predictions of an $r$/$K$ trade-off, empirical studies have shown that maximal growth rate and saturation density (measured by abundance) are positively correlated, both between species/strains \citep{luckinbill_1979,kuno_1991,hendriks_2005,fitzsimmons_2010}, and as a result of experimental evolution \citep{luckinbill_1978,luckinbill_1979}.
+From the perspective of our model, this positive correlation is not surprising since the saturation density, which is determined by a balance between births and deaths, increases with $b$.
 
@@
 
-There is support for a negative relationship between competitive success at high density and maximal growth rate \citep{luckinbill_1979}, consistent with an $r$/$K$ dichotomy.
+There is support for a negative relationship between competitive success at high density and maximal growth rate \citep{luckinbill_1979}, consistent with a tradeoff between $r$ and the competitive aspect of $K$.
 This could be driven by a tradeoff between individual size and reproductive rate.
@@
 
-Our realization of Grime's triangle (Fig.~4) differs from approaches which identify primary strategies as trait combinations which can co-exist \citep{bolker_99}, referring instead to the direction of adaptive trait evolution under different regimes of stress and disturbance, which is closer in spirit to Grime's arguments \citep{grime_1974,grime_1977}.
-In addition, we have not assumed any kind of trade-offs or pleitropy between $b$, $c$ and $d$, only constraints imposed by the environment on the order of magnitude of $b$ and $d$.
+Our realization of Grime's triangle (Fig.~4) differs from approaches which identify primary strategies as trait combinations which can stably co-exist \citep{bolker_99}, referring instead to the direction and rate of ongoing adaptive trait evolution under different regimes of stress and disturbance, which is closer in spirit to Grime's arguments \citep{grime_1974,grime_1977}.
+In addition, we have not assumed any kind of trade-offs or pleiotropy between $b$, $c$ and $d$, only constraints imposed by the environment on the order of magnitude of $b$ and $d$.
 As an example of a trade-off, corals which rapidly out-shade neighbors have a tall, branched morphology which is vulnerable to disturbances, and so, all else being equal, ideal environment $c$-strategists will suffer higher mortality from disturbances \citep{darling_2012}.
@@
 
-One limitation of our model as a general-purpose model of density-dependent selection is the restriction of competition to interference competition between juveniles for durable resources (lottery recruitment to adulthood), analogous to the ubiquitous assumption of viability selection in population genetics \citep[p.
+One potential limitation of our model as a general-purpose model of density-dependent selection is its restriction to interference competition between juveniles for durable resources (lottery recruitment to adulthood), analogous to the ubiquitous assumption of viability selection in population genetics \citep[p.
 45]{ewens_2004}.
-In some respects this is the complement of resource competition models, which restrict their attention to exploitation competition, typically without age structure \citep{tilman_1982}.
-In the particular case that resources are spatially localized (e.g.
+In some respects this is the complement of consumable resource competition models, which restrict their attention to indirect exploitation competition, typically without age structure \citep{tilman_1982}.
+In the particular case that consumable resources are spatially localized (e.g.
 due to restricted movement through soils), resource competition and territorial acquisition effectively coincide, and in principle resource competition could be represented by a competitive ability $c$ (or conversely, $c$ should be derivable from resource competition).
 The situation is more complicated if the resources are well-mixed, since, in general, resource levels then need to be explicitly tracked.
-It seems plausible that explicit resource tracking may not be necessary when the focus is on the evolution of similar genotypes rather than the stable co-existence of widely differing species \citep{ram_2016}.
+It seems plausible that explicit resource tracking may not be necessary when the focus is on the evolution of similar genotypes that use identical resources rather than the stable co-existence of widely differing species with different resource preferences \citep{ram_2016}.
 We are not aware of any attempts to delineate conditions under which explicit resource tracking is unnecessary even if it is assumed that community structure is ultimately determined by competition for consumable resources.
@@
 However, this restriction could become important when looking at species interactions rather than genotype evolution.
+
+In the introduction we mentioned the recurring difficulties with confounding selection and demography in population genetic inference.
+It seems that Eq.
+\eqref{eq:master} or something similar (and hopefully more analytically tractable) is unavoidable for the analysis of time-course genetic data because, fundamentally, selective births and deaths affect both abundances and frequencies, not one or the other in isolation.
+Moreover, some aspects of allele frequency change are intrinsically density-dependent.
+In the classic lottery model, which as we have seen is essentially the Wright-Fisher model with overlapping generations, $b_i$ and $c_i$ are  equivalent in the sense that the number of territorial victories only depends on the product $b_i c_i$ (see ``Model'').
+This is no longer the case in our extension, where $b$ and $c$ specialists can co-exist.
+This ``colonization-competition trade-off'' is well known in the co-existence literature \citep{tilman_94}.
+It and similar forms of ``spatial co-existence'' in stable environments have previously been modeled either with Levin's  qualitative representation of competition \citep{levins_71,tilman_94}, as opposed to the quantitative $c$ of lottery competition, or with a more sophisticated treatment of space (non-uniform dispersal; \citealt{shmida_84,bolker_99}).
+In cyclical environments, polymorphisms can be stabilized by the bounded density effect which is  completely lost if there is an exclusive focus on allele frequencies \citep{yi_2013}.
+We leave the details of how our model might be applied in inference, including the crucial issue of its genetic drift predictions (providing a null model for neutral sites), for future work.
 
@@
 
-\section*{Appendix C: Direction of evolution under different environmental extremes}
+\section*{Appendix C: Mutant invasion under different environments}
 
-In this Appendix we consider mutant invasion of a resident type $i$ (such that $N=n_i$) which is in equilibrium.
+In this Appendix we evaluate the invasion of novel mutants in a population with a single resident type $i$ (such that $N=n_i$), which is in equilibrium.
 
-In a population with a single genotype $i$ which is in equilibrium ($\Delta n_i = 0$), $R_i=0$, $\overline{c}=c_i$, $A_i=(1-(1+L)e^{-L})/L$, and Eq. \eqref{eq:master} gives
+For a single type $i$ in equilibrium ($\Delta n_i = 0$), we have $R_i=0$, $\overline{c}=c_i$, $A_i=(1-(1+L)e^{-L})/L$, and Eq. \eqref{eq:master} becomes
 \begin{equation}
-b_i(1-e^{-L})/L-d_i=0,\label{eq:equil}
+b_i(1-e^{-L})/L-d_i=0.\label{eq:equil}
 \end{equation}
-which implies $L\approx b_i/d_i$ when $b_i/d_i\gg 1$ and $L\ll 1$ when $b_i/d_i\approx 1$.
-(Eq.
-\eqref{eq:Da}{eq:equil} can also be seen directly from Eq.
+(Alternatively, Eq.
+\eqref{eq:equil} can be deduced directly from Eq.
 \eqref{eq:growthsumuncoupled}).
-Now suppose that a new genotype $j$, which is initially rare, appears in the population.
+This implies $L\approx b_i/d_i$ when $b_i/d_i\gg 1$ and $L\ll 1$ when $b_i/d_i\approx 1$.
+Now suppose that a novel mutant $j$, which is initially rare, appears in the population.
 Then $A_j/R_j\ll 0$, $l_j\approx 0$ and $\overline{c}\approx c_i$, and so, from Eq.
@@
 \end{equation}
-where $R_j\approx (1-e^{-L})/\left(\frac{c_j}{c_i}+\frac{L-1-e^{-L}}{1-(1+L)e^{-L}}\right)$.
+where $R_j\approx (1-e^{-L})/\left(\frac{c_j}{c_i}+\frac{L-1+e^{-L}}{1-(1+L)e^{-L}}\right)$ since $l_j\ll 1$.
 
-To see that coexistence is possible between $b$ and $c$ specialists, we check if $j$ will invade an $i$-dominated population, but $i$ will also invade a $j$-dominated population (``mutual invasion'').
-Suppose that $b_i$ is so large that $L\gg 1$ when $i$ is dominant, and $b_j$ is so small that $L\ll 1$ when $j$ is dominant.
-Then, combining Eqs.
-\eqref{eq:equil} and \eqref{eq:invad}, we find that $i$ invades $j$ because $b_i>b_j$, while $j$ invades $i$ provided that
-\begin{equation}
-b_jc_jR_j-b_i c_i A_i>0. \label{eq:jinvadcoex}
-\end{equation}
-Thus, coexistence occurs if $c_j/c_i$ is large enough.
-
-We now consider the different environments from section ``Primary strategies and Grime's triangle''.
+We now consider mutant invasion under the different environments from section ``Primary strategies and Grime's triangle''.
 In our representation of Grime's stressful environment, $b\ll 1$, $d\approx b$ and $L\ll 1$, and so Eq.
@@
 \begin{equation}
-\Delta n_j/n_j \approx \frac{b_j c_j}{c_j + L-1}-d_j\approx \frac{b_j c_j}{L}-d_j, \label{eq:highLinvad}
+\Delta n_j/n_j \approx \frac{b_j}{c_j/c_i + L-1}\frac{c_j}{c_i}-d_j\approx \frac{b_j}{L}\frac{c_j}{c_i}-d_j, \label{eq:highLinvad}
 \end{equation}
-where the last approximation follows from the fact that $c_j$ will be of similar magnitude to $c_i=1$ (ignoring mutations with very large effect sizes; incidentally, Eq.
-\eqref{eq:lottery} yields the same expression by assuming $L\rightarrow \infty$ even though this is obviously unrealistic for a rare mutant).
+where the last approximation follows from the fact that $c_j$ will be of similar magnitude to $c_i$ (ignoring mutations with very large effect sizes; incidentally, Eq.
+\eqref{eq:lottery} yields the same expression by assuming $L\rightarrow \infty$ even though this assumption is unrealistic for a rare mutant).
 Similarly to Grime's stressful environment, improving $b$, $c$, or $d$ by a fraction $\epsilon$ yields identical fitness benefits, even though they may have very different magnitudes, because the ancestral resident type's density-regulated birth rate $b_i c_i/L$ exactly balances its mortality $d_i$ in equilibrium.
@@
 
+\section*{Appendix D: Coexistence in a stable environment}
+
+To determine whether coexistence is possible in a constant environment, we check for ``mutual invasion'', that is, we check that $j$ will invade an $i$-dominated population, but $i$ will also invade a $j$-dominated population.
+
+We consider the case of coexistence between a $b$ specialist $i$ and a $c$ specialist $j$ ($b_i>b_j$, $c_j>c_i$ and $d_i=d_j$).
+Suppose that $b_i$ is so large that $L\gg 1$ when $i$ is dominant, and $b_j$ is so small that $L\ll 1$ when $j$ is dominant.
+Then, when $j$ is dominant, we have $\Delta n_i/n_i=b_i-d_i=b_i-d_j=b_i-b_j>0$.
+When $i$ is dominant, Eq.
+\eqref{eq:highLinvad} applies, where Eq.
+\eqref{eq:equil} implies $d_j=d_i=b_i(1-e^{-L})/L\approx b_i/L$, and so
+\begin{equation}
+\Delta n_j/n_j \approx \frac{b_j}{L}\frac{c_j}{c_i}-\frac{b_i}{L}.
+\end{equation}
+Therefore, coexistence occurs if $c_j/c_i$ is sufficiently large.
+The analogous argument for $d$ and $c$ specialists ($d_i<d_j$ with $L\gg 1$ when $i$ dominates, $L\ll 1$ when $j$ dominates, and $b_i=b_j$) gives $\Delta n_j/n_j \approx d_i\frac{c_j}{c_i}-d_j$, which again implies coexistence if $c_j/c_i$ is sufficiently large.
+
+For $b$ and $d$ specialists ($c_i=c_j$), we have $\Delta n_j/n_j\approx b_j d_i/b_i-d_j$ when $i$ dominates and $\Delta n_i/n_i\approx b_i d_j/b_j-d_i$ when $j$ dominates.
+Thus, either $i$ or $j$ grows when rare, but not both, and stable coexistence is not possible in a constant environment.
+
 \end{document}

### 2017-05-02 simulations.py

@@ -175,15 +175,7 @@ for t in range(totaltime):
     n = n + deltnplus((b(t)*n*U/T).astype(np.int),c,U.astype(np.int))-d(t)*n
     nhist.append(list(n))
     
-    
-#n=np.array([6000.,2500.])
-#nsimhist=[n]
-#for t in range(totaltime):
-#    U=T-sum(n)
-#    n = n + sum(deltnplussim((b(t)*n*U/T).astype(np.int),c,U.astype(np.int)))-d(t)*n
-#    nsimhist.append(list(n))
-
-fig=plt.subplot(321)
+fig=plt.subplot(331)
 for t in range(0,100,40):
     fig.plot(range(t,t+20),[dparam[1][0] for _ in range(t,t+20)],'k.',lw=2)
     fig.plot(range(t,t+20),[dparam[1][1] for _ in range(t,t+20)],'b.',lw=2)
@@ -206,21 +198,63 @@ fig.annotate(r'$d$',xy=(0.49,0.4),xycoords='axes fraction',fontsize=16)
 fig.annotate(r'$b$',xy=(0.69,0.4),xycoords='axes fraction',fontsize=16)
 fig.annotate(r'$d$',xy=(0.89,0.4),xycoords='axes fraction',fontsize=16)
 
-fig=plt.subplot(323)
+fig=plt.subplot(334)
 plt.stackplot(range(totaltime+1),np.array(nhist)[:,1],np.array(nhist)[:,0],colors=['b','k'])
 fig.set_xticklabels([])
 fig.set_yticklabels(['0.0','0.2','0.4','0.6','0.8','1.0'])
 plt.ylabel(r"Abundance$/T$",fontsize=14)
 fig.annotate(r'$(b)$',xy=(0.01,0.9),xycoords='axes fraction',fontsize=12)
 
-fig=plt.subplot(325)
+fig=plt.subplot(337)
 plt.stackplot(range(totaltime+1),np.array(nhist)[:,1]/np.sum(nhist,1),np.array(nhist)[:,0]/np.sum(nhist,1),colors=['b','k'])
 plt.ylim([0,1])
 plt.xlabel(r"$t$",fontsize=14)
 plt.ylabel("Frequency",fontsize=14)
 fig.annotate(r'$(c)$',xy=(0.01,0.9),xycoords='axes fraction',color='white',fontsize=12)
 
-#Classic lottery
+#Classic lottery same
[…]

### 2017-05-03 manuscript.tex
latest

@@
 
-Here we describe how Grime's ``disturbed'', ``stressful'' and ``ideal'' environments can be captured mathematically in our scheme.
-To proceed, we  map these verbally-defined environments to quantitative parameter regimes in our model.
+Here we describe how Grime's ``disturbed'', ``stressful'' and ``ideal'' environments can be captured mathematically in our model, and evaluate the direction of evolution for each.
+To proceed, we map these verbally-defined environments to quantitative parameter regimes in our model, summarized in Fig.~\ref{fig:table}.
 
@@
 Disturbed environments are characterized by short bursts of high extrinsic mortality $d$ caused by physical destruction.
-We assume that disturbances are equally damaging to adults and juveniles, so that only $(1-d_i)\Delta_+ n_i$ rather than $\Delta_+ n_i$ territories are secured by type $i$ each iteration.
+We assume that disturbances are equally damaging to adults and juveniles, so that only $(1-d_i)\Delta_+ n_i$ (rather than $\Delta_+ n_i$) territories are secured by type $i$ each iteration.
 In the heavily disturbed extreme of Grime's triangle, many such bursts (which could each be spatially localized) will occur each iteration (i.e.
-over the time required for a propagule to reach reproductive maturity), which we approximate as almost all adults and juveniles dying every iteration $d\approx 1$.
-Population persistence is only possible if $(1-d)b>d$, and so $b\gg 1$ (Appendix C).
-We assume that fulfilling this requirement presents such a challenge that $(1-d)b$ does not greatly exceed $d$, which implies $L\ll 1$.
+over the time required for a propagule to reach reproductive maturity), which we approximate as a high constant mortality rate $d\approx 1$.
+Population persistence requires $b>d/(1-d)\gg 1$ (Appendix C), i.e.
+severe disturbance comes with a high fecundity/dispersal requirement.
+We expect that it is not possible to greatly exceed this requirement ($b\approx d/(1-d)$), which implies $L\ll 1$.
 
@@
   Other parameters & $b/d\gg 1$ & $(1-d)b\approx d$ & $b\approx d$ & $b/d\gg 1$ \\
-  Density $L$  & $\gg 1$ & $\ll 1$ & $\ll 1$ & $\gg 1$
+  Propagule density $L$  & $\gg 1$ & $\ll 1$ & $\ll 1$ & $\gg 1$
 \end{tabular}
-\caption{\label{fig:table} The mathematical realization of Grime's three environmental extremes in our model, as well as the high-density variant of the stressful environment.
+\caption{\label{fig:table} The mathematical realization of Grime's three environmental extremes in our model, as well as the high-density (HD) variant of the stressful environment.
 *Mortality affects both adults and juveniles in the disturbed environment, with $\Delta_+ n_i$ replaced by $(1-d_i)\Delta_+ n_i$.}
@@
 
-Given our mathematical representation of Grime's environmental extremes, summarized in Fig.~\ref{fig:table}, the expected direction of evolution is determined by three factors.
-First, the probability that an adaptive mutant survives its initial low abundance phase and begins to grow deterministically is proportional to its expected absolute growth rate \citep{haldane_1927,uecker_2011}.
-By comparison, neutral mutations rarely fix (probability of order $1/N$).
+The expected direction of evolution is determined by three factors.
+First, there may be pleiotropy between $b$, $c$, and $d$, which could be due to trade-offs between them.
+Here we do not attempt to account for the complications posed by pleiotropy, and treat $b$, $c$ and $d$ independently.
+
+Second, some mutations are more common than others, which can change the direction of evolution \citep{yampolsky_2001}.
+In particular, very few mutations can perform difficult trait changes such as increasing resilience to intense physical disturbance.
+Quantifying which trait changes are more mutationally accessible than others is beyond the scope of this article, and so we will assume that $b$, $c$ and $d$ all have the same distribution of beneficial fitness effects.
+
+Finally, the probability that an adaptive mutant survives its initial low abundance phase and begins to grow deterministically is proportional to its expected absolute growth rate \citep{haldane_1927,uecker_2011}.
+Very few neutral mutations fix (probability of order $1/N$).
 Thus $b$, $c$ and $d$ will tend to change in the direction of greatest fitness increase.
+
 Under Eq.
-\eqref{eq:master}, proportional changes of the form $b_j=(1+\epsilon) b_i$ or $d_j=(1+\epsilon) d_i$ have identical fitness effects
+\eqref{eq:master}, proportional $b$- and $d$-mutations of the form $b_j=(1+\epsilon) b_i$ and $d_j=(1-\epsilon) d_i$ have identical fitness effects (here $j$ is the mutant type; Appendix C).
+When propagule density is high  $L\gg 1$, the same is true for $c$-mutations of the form $c_j=(1+\epsilon) c_i$.
+When propagule density is low $L\ll 1$, $c$ does not contribute to fitness and $c$-mutations are neutral.
+Thus, in ideal and stressed environments, selection under Eq.
+\eqref{eq:master} does not intrinsically favor one trait improvement over another, apart from the irrelevance of $c$ at low density.
 
-When propagule density is low $L\ll 1$, $c$ does not contribute to fitness.
-
-Second, the fact that some mutations are more common than others can change the direction of evolution \citep{yampolsky_2001}, and in particular, very few mutations can perform difficult trait changes such as increasing $b$ under stress.
-These mutational constraints are different under different environments (Fig.~\ref{fig:table}).
-
-Finally, there may be pleiotropy between $b$, $c$, and $d$, which could be due to trade-offs between them.
-Here we do not attempt to account for the complications posed by pleiotropy, and treat $b$, $c$ and $d$ independently.
+The situation in disturbed environments is more complicated.
+The bulk of the high mortality $d\approx 1$ is environmental and unavoidable.
+Thus, mutational changes in $d$ are intrinsically less effective than mutation change in $b$.
+However, this effect is easily swamped by the fact that disturbance mortality affects both juvenile recruitment and adult mortality.
 
@@
 Schematic representation of the triangular space bounded by the low/high extremes of stress/disturbance.
-The low-$T$ interpretation of stress is also shown.
+The high-density (HD) interpretation of stress is also shown.
 The vertices of the triangles correspond to different environmental extremes.

### 2017-05-03 manuscript.tex
latest

@@
 For example, if consumable resources are scarce, we expect intense resource competition \citep{davis_1998}.
-Thus, even though $b\ll 1$, we can still have $b/d\gg 1$ and $L\ll 1$.
+Thus, even though $b\ll 1$, we still have $b/d\gg 1$ and $L\ll 1$.
 
@@
 First, there may be pleiotropy between $b$, $c$, and $d$, which could be due to trade-offs between them.
-Here we do not attempt to account for the complications posed by pleiotropy, and treat $b$, $c$ and $d$ independently.
+Second, adaptive mutations in one trait may occur more frequently than in others, which can change the direction of evolution \citep{yampolsky_2001}.
+Quantifying how traits are affected by mutations is beyond the scope of this article, and so we will not evaluate either of these effects: mutations affect $b$, $c$ and $d$ independently, and they all have the same distribution of beneficial fitness effects.
 
-Second, some mutations are more common than others, which can change the direction of evolution \citep{yampolsky_2001}.
-In particular, very few mutations can perform difficult trait changes such as increasing resilience to intense physical disturbance.
-Quantifying which trait changes are more mutationally accessible than others is beyond the scope of this article, and so we will assume that $b$, $c$ and $d$ all have the same distribution of beneficial fitness effects.
-
-Finally, the probability that an adaptive mutant survives its initial low abundance phase and begins to grow deterministically is proportional to its expected absolute growth rate \citep{haldane_1927,uecker_2011}.
+The third effect is that selection can favor some trait changes over others.
+The probability that an adaptive mutant survives its initial low abundance phase and begins to grow deterministically is proportional to its expected absolute growth rate \citep{haldane_1927,uecker_2011}.
 Very few neutral mutations fix (probability of order $1/N$).

### 2017-05-03 manuscript.tex
latest

@@
 \end{tabular}
-\caption{\label{fig:table} The mathematical realization of Grime's three environmental extremes in our model, as well as the high-density (HD) variant of the stressful environment.
+\caption{\label{fig:table} The mathematical representation of Grime's three environmental extremes in our model, as well as the high-density (HD) variant of the stressful environment.
 *Mortality affects both adults and juveniles in the disturbed environment, with $\Delta_+ n_i$ replaced by $(1-d_i)\Delta_+ n_i$.}
@@
 First, there may be pleiotropy between $b$, $c$, and $d$, which could be due to trade-offs between them.
-Second, adaptive mutations in one trait may occur more frequently than in others, which can change the direction of evolution \citep{yampolsky_2001}.
-Quantifying how traits are affected by mutations is beyond the scope of this article, and so we will not evaluate either of these effects: mutations affect $b$, $c$ and $d$ independently, and they all have the same distribution of beneficial fitness effects.
+Second, adaptive mutations in one trait may occur more frequently than in others \citep{yampolsky_2001}.
+We do no doubt that these factors operate in nature and contribute towards ``primary strategy'' trait patterns.
+However, modeling how mutations affect traits subject to trade-offs is beyond the scope of this article, and so we will not evaluate these first two factors: we assume that mutations affect $b$, $c$ and $d$ independently, and that $b$, $c$ and $d$ have the same distribution of beneficial fitness effects.
 
-The third effect is that selection can favor some trait changes over others.
-The probability that an adaptive mutant survives its initial low abundance phase and begins to grow deterministically is proportional to its expected absolute growth rate \citep{haldane_1927,uecker_2011}.
+The third factor is that some traits may have a greater impact on fitness: $b$, $c$ and $d$ will tend to evolve in the direction of greatest fitness increase because the probability that an adaptive mutant survives its initial low abundance phase and begins to grow deterministically is proportional to its expected absolute growth rate \citep{haldane_1927,uecker_2011}.
 Very few neutral mutations fix (probability of order $1/N$).
-Thus $b$, $c$ and $d$ will tend to change in the direction of greatest fitness increase.
 
 Under Eq.
-\eqref{eq:master}, proportional $b$- and $d$-mutations of the form $b_j=(1+\epsilon) b_i$ and $d_j=(1-\epsilon) d_i$ have identical fitness effects (here $j$ is the mutant type; Appendix C).
-When propagule density is high  $L\gg 1$, the same is true for $c$-mutations of the form $c_j=(1+\epsilon) c_i$.
+\eqref{eq:master}, mutations which change $b$ or $d$ by the same  proportion (i.e.
+$b_j=(1+\epsilon) b_i$ or $d_j=(1-\epsilon) d_i$ where $j$ is the mutant type) have identical fitness effects (Appendix C).
+When propagule density is high  $L\gg 1$, the same is true for proportional $c$-mutations $c_j=(1+\epsilon) c_i$.
 When propagule density is low $L\ll 1$, $c$ does not contribute to fitness and $c$-mutations are neutral.
-Thus, in ideal and stressed environments, selection under Eq.
-\eqref{eq:master} does not intrinsically favor one trait improvement over another, apart from the irrelevance of $c$ at low density.
+Thus, selection under Eq.
+\eqref{eq:master}, which applies in ideal and stressed environments, does not intrinsically favor one trait improvement over another, apart from the irrelevance of $c$ at low density.
 
-The situation in disturbed environments is more complicated.
+The situation in disturbed environments, where mortality also affects juveniles, is more complicated.
 The bulk of the high mortality $d\approx 1$ is environmental and unavoidable.
-Thus, mutational changes in $d$ are intrinsically less effective than mutation change in $b$.
+Thus, proportional changes in the remaining genetic part of mortality can have a tiny impact on $d$, suggesting the primacy of $b$ under disturbance.
 However, this effect is easily swamped by the fact that disturbance mortality affects both juvenile recruitment and adult mortality.

### 2017-05-03 manuscript.tex
latest

@@
 
-\title{Density-dependent selection in evolutionary genetics: a lottery model of Grime's triangle}
+\title{A lottery model of density-dependent selection in evolutionary genetics}
 %\author{Jason Bertram $^{1,\ast}$ \\ 
@@
 Biases in the direction of evolution of $b$, $c$ or $d$ must again follow from biases in mutation availability or trade-offs/pleiotropy.
-The same argument applies for high-$L$ interpretation of the stress regime.
+The same argument applies for the high-$L$ interpretation of the stress regime.
 
@@
 \begin{equation}
-\Delta n_j/n_j = (1-d_j)b_j - d_j.
+\Delta n_j/n_j \approx (1-d_j)b_j - d_j.
 \end{equation}
@@
 
-The effects of improving $b$ and $d$ are now no longer identical.
-Increasing $b$ by a fraction $\epsilon$ gives $\Delta n_j/n_j = \epsilon (1-d_i)b_i\approx \epsilon$, whereas reducing $d$ by the same fraction gives $\Delta n_j/n_j = \epsilon d_{ig}/(1-d_i)$.
-Thus, even though the genetic component of mortality $d_{g}$ is small compared to the environmental component $d_{e}$, if disturbance is intense enough ($1-d_i$ is small enough), selection for reduced mortality is potentially much stronger than that for propagule production.
+The effects of improving $b$ and $d$ are now no longer identical for given values of $d_i$ and $d_{ig}$.
+Increasing $b$ by a fraction $\epsilon$ gives $\Delta n_j/n_j = \epsilon (1-d_i)b_i\approx \epsilon d_i$, whereas reducing $d_g$ by the same fraction gives $\Delta n_j/n_j = \epsilon d_{ig}/(1-d_i)$.
+
+Not surprisingly, selection for lower $d$ is weaker than for greater $b$ if $d_{ig}$ is sufficiently small. However, assuming that there is some non-negligible genetic contribution to mortality
+
+Thus, since $d_i\approx 1$, selection for lower $d$ is potentially much stronger than that for higher $b$.
+
+even though the genetic component of mortality $d_{g}$ is small compared to the environmental component $d_{e}$, if disturbance is intense enough ( is small enough), .
 

### 2017-05-04 manuscript.tex
latest

@@
 
-Here we describe how Grime's ``disturbed'', ``stressful'' and ``ideal'' environments can be captured mathematically in our model, and evaluate the direction of evolution for each.
+Here we describe how Grime's ``disturbed'', ``stressful'' and ``ideal'' environments can be captured mathematically in our model, and evaluate role of selection in shaping the corresponding ``primary strategies''.
 To proceed, we map these verbally-defined environments to quantitative parameter regimes in our model, summarized in Fig.~\ref{fig:table}.
@@
 over the time required for a propagule to reach reproductive maturity), which we approximate as a high constant mortality rate $d\approx 1$.
-Population persistence requires $b>d/(1-d)\gg 1$ (Appendix C), i.e.
-severe disturbance comes with a high fecundity/dispersal requirement.
-We expect that it is not possible to greatly exceed this requirement ($b\approx d/(1-d)$), which implies $L\ll 1$.
+Population persistence requires $b>d/(1-d)$, which implies $b\gg 1$ (severe disturbance comes with a high fecundity/dispersal requirement).
+We assume this requirement can just be met, so that $b\approx d/(1-d)$ and $L\ll 1$ (Appendix C).
 
@@
 Severe stress inhibits growth and reproduction, so that $b\ll 1$.
-In Grime's view, this means that the rate at which propagules successfully develop to adulthood cannot appreciably exceed the mortality rate $b/d\approx 1$; coping with the environmental stress is barely possible \citep{grime_1974,grime_1977}.
+In Grime's view, this means that the rate at which propagules successfully develop to adulthood cannot appreciably exceed the mortality rate $b/d\approx 1$; it is just possible to cope with the environmental stress \citep{grime_1974,grime_1977}.
 In our model, this implies $L\ll 1$.
@@
 Second, adaptive mutations in one trait may occur more frequently than in others \citep{yampolsky_2001}.
-We do no doubt that these factors operate in nature and contribute towards ``primary strategy'' trait patterns.
-However, modeling how mutations affect traits subject to trade-offs is beyond the scope of this article, and so we will not evaluate these first two factors: we assume that mutations affect $b$, $c$ and $d$ independently, and that $b$, $c$ and $d$ have the same distribution of beneficial fitness effects.
+Third, selection favors some trait changes over others: $b$, $c$ and $d$ will tend to evolve in the direction of greatest fitness increase because the probability that an adaptive mutant survives its initial low abundance phase and begins to grow deterministically is proportional to its expected absolute growth rate \citep{haldane_1927,uecker_2011}.
+Only a tiny proportion of neutral and deleterious mutations fix (probability of order $1/N$ for neutrals).
 
-The third factor is that some traits may have a greater impact on fitness: $b$, $c$ and $d$ will tend to evolve in the direction of greatest fitness increase because the probability that an adaptive mutant survives its initial low abundance phase and begins to grow deterministically is proportional to its expected absolute growth rate \citep{haldane_1927,uecker_2011}.
-Very few neutral mutations fix (probability of order $1/N$).
+Here we only consider the last of these factors.
+To account for the first two factors would require an evaluation of the effects of mutations on traits subject to trade-offs, which is beyond the scope of this article.
+Thus, we consider how selection alone shapes the direction of evolution for $b$, $c$ and $d$, assuming that mutations affect them independently and with the same distribution of beneficial fitness effects.
 
@@
 Another distinct feature of disturbed environments is that the bulk of mortality is environmental and unavoidable.
-That is, $d_i=d_{ie}+d_{ig}$, where the environmental part $d_{ie}\approx 1$ does not evolve, and the genetic part $d_{ig}\ll 1$ is affected by mutations in the same way as before $d_{jg}=(1-\epsilon) d_{ig}$.
+That is, $d_i=d_{ie}+d_{ig}$, where the environmental part $d_{ie}$ does not evolve, and the genetic part $d_{ig}$ is affected proportionally by mutations $d_{jg}=(1-\epsilon) d_{ig}$.
 
-The effects of improving $b$ and $d$ are now no longer identical for given values of $d_i$ and $d_{ig}$.
-Increasing $b$ by a fraction $\epsilon$ gives $\Delta n_j/n_j = \epsilon (1-d_i)b_i\approx \epsilon d_i$, whereas reducing $d_g$ by the same fraction gives $\Delta n_j/n_j = \epsilon d_{ig}/(1-d_i)$.
+The effects of improving $b$ and $d$ are now no longer identical for given values of $d_{ie}$ and $d_{ig}$.
+A $b-$mutation $b_j=(1+\epsilon)b_i$ gives $\Delta n_j/n_j = \epsilon (1-d_i)b_i\approx \epsilon d_i \approx \epsilon$, whereas a $d$-mutation $d_{jg}=(1-\epsilon) d_{ig}$ gives $\Delta n_j/n_j = \epsilon d_{ig}/(1-d_i)$.
+Since overall mortality is high $d_i\approx 1$, $d_{g}$ can be small compared to $d_{e}$ and yet still make a greater contribution to invasion fitness than $b$.
+The case where $d_{ig}$ is so small that $d_{ig}\ll 1-d_i$ (i.e.
+$d_{ig}$ is an order of magnitude smaller than $1-d_i$ which itself an order of magnitude smaller than $d_{ie}$) effectively represents a hard constraint that $d$ cannot evolve.
+We do not deny this possibility, but this bring us back to the issue of mutations affect traits (the first two factors controlling the direction of evolution in section ``Primary strategies and Grime's triangle''), which we do not address in this manuscript.
 
-Not surprisingly, selection for lower $d$ is weaker than for greater $b$ if $d_{ig}$ is sufficiently small. However, assuming that there is some non-negligible genetic contribution to mortality
-
-Thus, since $d_i\approx 1$, selection for lower $d$ is potentially much stronger than that for higher $b$.
-
-even though the genetic component of mortality $d_{g}$ is small compared to the environmental component $d_{e}$, if disturbance is intense enough ( is small enough), .
+Thus, assuming that $d$ can evolve in this sense, $d_{ig}$ is either comparable to or greater than $1-d_i$.
+Yet $d_{ig}$ cannot be appreciably greater than $1-d_i$ for long, because selection will then strongly favor reductions in $d_g$ over increases in $b$, thereby bringing $d_{ig}$ closer to $1-d_i$.
+Thus, we expect selection on $b$ and $d$ to be of comparable strength in disturbed environments as well.
 

### 2017-05-05 manuscript.tex
latest

@@
 
+\section*{Introduction}
+
 Evolutionary models differ greatly in their treatment of fitness.
@@
 This complicates the inference of past selection, because demographic changes can look genealogically very similar to selective frequency changes \citep{barton_1998}.
-
-%Separately fitting demographic models to data on ostensibly neutral sites, and then inferring selection against this demographic background, has had mixed success due to linkage \citep{schrider_2016}.
 
@@
 The extremes of these two factors define three primary strategies denoted by C/S/R respectively (Fig.
-\ref{fig:grimeschematic}): competitors ``C'' excel in low stress, low disturbance environments; stress tolerators ``S'' excel in high stress, low disturbance environments; and ruderals ``R''  excel in low stress, high disturbance environments.
+\ref{fig:axes}): competitors ``C'' excel in low stress, low disturbance environments; stress tolerators ``S'' excel in high stress, low disturbance environments; and ruderals ``R''  excel in low stress, high disturbance environments.
 Population persistence is not possible in high-stress, high-disturbance environments.
@@
 \centering
-\includegraphics[scale=1]{grimeschematic.pdf}
-\caption{\label{fig:grimeschematic} Schematic of Grime's triangle.
+\includegraphics[scale=1]{axes.pdf}
+\caption{\label{fig:axes} Schematic of Grime's triangle and the associated direction of selection in our model.
 The two axes show increasing levels of environmental stress and disturbance, respectively.
 Population persistence is not possible if the combination of stress and disturbance is too large (dashed line).
-This creates a triangle, each corner of which corresponds to a ``primary strategy''.}
+This creates a triangle, each corner of which corresponds to an environmental extreme and associated ``primary strategy''.
+The high-density (HD) interpretation of stress is shown as a rotation of the triangle with respect to density.
+Vertical arrows show how selection acts on $b$, $c$, and $d$: $c$ is irrelevant at low $L$, but otherwise $b$, $c$ and $d$ all experience the same selective force.}
 \end{figure}
@@
 
-Using our extended lottery model, we then revisit Grime's C/S/R scheme,  and evaluate how C/S/R manifests at the level of within-population genotypic evolution (as opposed to phenotypic divergence between species; section ``Primary strategies and Grime's triangle'').
-This represents a ``sanity check'' on our density-dependent lottery model.
-The resulting formulation of the C/S/R scheme is mathematical, in contrast to Grime's original verbal and descriptive approach, which is a recognized hindrance to the evaluation or broader application of the C/S/R scheme (e.g.
-\citealt{tilman_2007}).
+Using our extended lottery model, we then revisit Grime's C/S/R scheme, and evaluate the extent to which selection favors fecundity, competitive ability or lower mortality under Grime's environmental extremes (section ``Primary strategies and Grime's triangle'').
+This represents a simple ``sanity check'' on the invasion behavior of our extended lottery model in different extreme cases.
 
@@
 
-Here we only consider the last of these factors.
-To account for the first two factors would require an evaluation of the effects of mutations on traits subject to trade-offs, which is beyond the scope of this article.
-Thus, we consider how selection alone shapes the direction of evolution for $b$, $c$ and $d$, assuming that mutations affect them independently and with the same distribution of beneficial fitness effects.
+Accounting for the first two factors would require an evaluation of the effects of mutations on traits subject to trade-offs, which is beyond the scope of this article.
+We only consider how selection shapes the direction of evolution for $b$, $c$ and $d$, assuming that mutations affect them independently and with the same distribution of beneficial fitness effects.
+Specifically, we consider the effect of mutations that change $b$, $c$, or $d$ by a fixed proportion $b_j=(1+\epsilon) b_i$, $c_j=(1+\epsilon) c_i$ or $d_j=(1-\epsilon) d_i$, where $j$ is is a novel mutant (Appendix C).
+Mutations with magnitude $\epsilon$ are assumed to occur with the same frequency for all traits .
 
-Under Eq.
-\eqref{eq:master}, mutations which change $b$ or $d$ by the same  proportion (i.e.
-$b_j=(1+\epsilon) b_i$ or $d_j=(1-\epsilon) d_i$ where $j$ is the mutant type) have identical fitness effects (Appendix C).
-When propagule density is high  $L\gg 1$, the same is true for proportional $c$-mutations $c_j=(1+\epsilon) c_i$.
-When propagule density is low $L\ll 1$, $c$ does not contribute to fitness and $c$-mutations are neutral.
-Thus, selection under Eq.
-\eqref{eq:master}, which applies in ideal and stressed environments, does not intrinsically favor one trait improvement over another, apart from the irrelevance of $c$ at low density.
+Not surprisingly, mutations which increase $c$ are effectively neutral when $L\ll 1$ and adaptive when $L\gg 1$.
+However, beyond this, selection in our density-dependent lottery model does not favor one trait improvement over another (Fig.
+3).
+This applies regardless of the initial magnitude of $b$ or $d$.
+More specifically, in ideal and stressed environments, where Eq.
+\eqref{eq:master} applies, $b$ and $d$ mutations of size $\epsilon$ have identical fitness effects (Appendix C), and the same is true for proportional $c$-mutations $c_j=(1+\epsilon) c_i$ when $L\gg 1$.
 
-The situation in disturbed environments, where mortality also affects juveniles, is more complicated.
+In disturbed environments, $\Delta_+ n_i$ is replaced by $(1-d_i)\Delta_+ n_i$.
 The bulk of the high mortality $d\approx 1$ is environmental and unavoidable.
-Thus, proportional changes in the remaining genetic part of mortality can have a tiny impact on $d$, suggesting the primacy of $b$ under disturbance.
-However, this effect is easily swamped by the fact that disturbance mortality affects both juvenile recruitment and adult mortality.
-
-Following Grime's original argument for a triangular scheme \citep{grime_1977}, Fig.~\ref{fig:axes} represents each environmental extreme schematically as a vertex on a triangular space defined by perpendicular stress and disturbance axes.
-The ideal environment lies at the origin (no stress or disturbance), while the stressful and disturbed environments lie at the limits of survival on their respective axes.
-The hypotenuse connecting the stress and disturbance endpoints represents the limits of survival in the presence of a combination of stress and disturbance.
-The direction of evolutionary change is different at each vertex, leading to the emergence of different trait clusters or ``primary strategies''.
-
-\begin{figure}
-\centering
-\includegraphics[scale=1]{axes.pdf}
-\caption{\label{fig:axes} The realization of Grime's triangle in our model.
-Schematic representation of the triangular space bounded by the low/high extremes of stress/disturbance.
-The high-density (HD) interpretation of stress is also shown.
-The vertices of the triangles correspond to different environmental extremes.
-Selection favors different traits at each vertex, leading to different trait clusters.}
-\end{figure}
+This implies that $\epsilon$ mutations in the remaining genetic part of mortality have a reduced impact on $d$, suggesting the primacy of $b$ under disturbance.
+However, this effect is easily swamped by the fact that $d_i\approx 1$ and that disturbance mortality affects both juvenile recruitment and adult mortality.
+In Appendix C, we argue that $b$ and $d$ mutations should have comparable effects on fitness (assuming that $d$ can evolve at all) because the fitness benefit of improving $d$ declines as $d$ improves and so $d$ cannot run ahead of $b$ for long (diminishing returns).
 
@@
 The second is the  bounded population size effect of \cite{yi_2013}, which occurs when each environmental cycle involves growth from low to high density, with the time spent growing each cycle dependent on the fitness of the types present.
-
-%two general mechanisms by which fluctuating selection promotes coexistence.
-The first is the ``storage effect'', which introduces a form of frequency dependent selection that promotes coexistence in the presence of environmental fluctuations but not in a constant environment.
-The storage effect occurs when some individuals are protected from selection; in the lottery model a fraction $(1-d_i)n_i$ of each type's adults do not experience selection in a given iteration.
-Protection from selection promotes coexistence in fluctuating environments because abundant types cannot fully exploit environmental periods that favor them (since only a fraction of the rare type can be displaced), whereas rare types gain the full benefits of their favorable periods (far more adults from the abundant type die than can possibly be replaced by the rare types) \citep{chesson_1981}.
-
-%The second mechanism we will call the ``bounded density effect'', since it is a consequence of the inhibition of reproduction at high population densities (Dempster's (1955) argument ignores density-dependent effects).
-If there is growth from low to high density in each environmental cycle, then types that are abundant determine the time available for growth every cycle, with less time spent growing in cycles where they are favored, and more in cycles where they are not.
-This promotes coexistence even in the absence of frequency-dependent selection \citep{yi_2013}.
 
@@
 We leave these complications for future work.
+
+The resulting formulation of the C/S/R scheme is mathematical, in contrast to Grime's original verbal and descriptive approach, which is a recognized hindrance to the evaluation or broader application of the C/S/R scheme (e.g.
+\citealt{tilman_2007}).
 

### 2017-05-05 manuscript.tex
latest

@@
 
-Using our extended lottery model, we then revisit Grime's C/S/R scheme, and evaluate the extent to which selection favors fecundity, competitive ability or lower mortality under Grime's environmental extremes (section ``Primary strategies and Grime's triangle'').
-This represents a simple ``sanity check'' on the invasion behavior of our extended lottery model in different extreme cases.
+Using our extended lottery model, we then revisit Grime's C/S/R scheme, and evaluate the extent to which selection favors fecundity, competitive ability or lower mortality under Grime's environments (section ``Primary strategies and Grime's triangle'').
+This represents a simple ``sanity check'' on the invasion behavior of our extended lottery model under different extremes.
 
@@
 
-The resulting formulation of the C/S/R scheme is mathematical, in contrast to Grime's original verbal and descriptive approach, which is a recognized hindrance to the evaluation or broader application of the C/S/R scheme (e.g.
+Our realization of Grime's triangle (Fig.~\ref{fig:axes}) differs from approaches which identify primary strategies as trait combinations which can stably co-exist \citep{bolker_99}, referring instead to the direction and rate of ongoing adaptive trait evolution under different regimes of stress and disturbance, which is closer in spirit to Grime's arguments \citep{grime_1974,grime_1977}.
+Moreover, our formulation is mathematical, in contrast to Grime's original verbal and descriptive approach, which is a recognized hindrance to the evaluation or broader application of the C/S/R scheme (e.g.
 \citealt{tilman_2007}).
+However, section ``Primary strategies and Grime's triangle'' suggests that, apart from the obvious irrelevance of territorial contests at low density, selection alone has no preference for fecundity, competitive ability or longevity.
+This could mean that one or both of the factors we set aside --- pleiotropy/trade-offs and mutation bias --- are needed to get  true ``primary strategy'' trait differentiation.
+A separate model of trade-offs and mutation effects would be needed to explore this possibility.
 
-Our realization of Grime's triangle (Fig.~4) differs from approaches which identify primary strategies as trait combinations which can stably co-exist \citep{bolker_99}, referring instead to the direction and rate of ongoing adaptive trait evolution under different regimes of stress and disturbance, which is closer in spirit to Grime's arguments \citep{grime_1974,grime_1977}.
-In addition, we have not assumed any kind of trade-offs or pleiotropy between $b$, $c$ and $d$, only constraints imposed by the environment on the order of magnitude of $b$ and $d$.
-As an example of a trade-off, corals which rapidly out-shade neighbors have a tall, branched morphology which is vulnerable to disturbances, and so, all else being equal, ideal environment $c$-strategists will suffer higher mortality from disturbances \citep{darling_2012}.
-Fig.~\ref{fig:axes} gives the same conclusion without invoking trade-offs; mutations which reduce disturbance vulnerability are essentially neutral under ideal conditions, leading to no improvements in mortality from disturbances, whereas $c$ will tend to increase over time.
-Thus, while trade-offs may amplify specialization, and are sometimes invoked to explain primary strategy schemes \citep{macarthur_1967,winemiller_1992,aerts_1999}, they are not necessary for it.
+Nevertheless, it is interesting to note that ruderals, which are typically thought of as high fecundity dispersers ($b$ specialists), may also be strongly $d$-selected, consistent with our findings.
+An effective way to reduce $d$ in the face of unavoidable physical destruction is to shorten the time to reproductive maturity --- short life cycles are a characteristically ruderal trait.
+Moreover, a recent hierarchical cluster analysis of coral traits did find a distinct ``ruderal'' cluster, but high fecundity was not its distinguishing feature.
+Rather, ruderals used brood- (as opposed to broadcast-) spawning, which could plausibly be a mechanism for improving propagule survivorship in disturbed environments \citep{darling_2012}.
 

### 2017-05-06 manuscript.tex
latest

@@
 
-\textit{Keywords}: r/K selection, absolute fitness, eco-evo, interference competition, competition-colonization trade-off.
+\textit{Keywords}: r/K selection, absolute fitness, eco-evo, Grime's triangle, competition-colonization trade-off, fluctuating selection, storage effect.
 
@@
 
-[Needs to be revised but won't do that now in case it has to become Nature introductory paragraph]
-Fitness is typically represented in heavily simplified terms in evolutionary genetics, often using constant selection coefficients.
+Fitness is typically represented in heavily simplified terms in evolutionary genetics, often using constant selection coefficients, to make it easier to infer or predict how type frequencies change over time.
 This excludes fundamental ecological factors such as dynamic population size or density-dependence from the most genetically-realistic treatments of evolution, a problem that inspired MacArthur's influential but problematic $r$/$K$ theory.
-Following in the spirit of $r$/$K$-selection as a general-purpose theory of density-dependent selection, but grounding ourselves empirically in ``primary strategy'' trait classification schemes like Grime's triangle, we develop a new model of density-dependent selection which revolves around territorial contests.
-To do so, we generalize the classic lottery model of territorial acquisition, which has primarily been used for studying species co-existence questions, to accommodate arbitrary densities.
-We use this density-dependent lottery model to predict the direction of trait evolution under different environmental conditions and thereby provide a mathematical underpinning for Grime's verbal scheme.
-We revisit previous concepts of density-dependent selection, including $r$ and $K$ selection, and argue that our model distinguishes between different aspects of fitness in a more natural and intuitive manner.
+Following the spirit of $r$/$K$-selection as a general-purpose theory of density-dependent selection, we develop a new model of density-dependent selection by generalizing the fixed-density classic lottery model of territorial acquisition to accommodate arbitrary population densities.
+We show that, with density dependence, co-existence is possible in the lottery model in a stable environment.
+Inspired by natural \textit{Drosophila} populations, we consider co-existence under strong, seasonally-fluctuating selection coupled to large cycles in population density, and show that co-existence (stable polymorphism) is promoted via a combination of the classic storage effect and density-regulated population growth.
+We also show that the only significant bias introduced by selection at the different environmental extremes of Grime's triangle is the relative importance of competitive ability at different densities, confirming an important role for phenotypic constraints in shaping ``primary strategies''.
 
@@
 
-Empirical trait classification studies have suggested the existence of a few ``primary strategies'', reflecting broadly distinct responses to selection \citep{winemiller_2015}.
+However, empirical trait classification studies have suggested the existence of a few ``primary strategies'', reflecting broadly distinct responses to selection \citep{winemiller_2015}.
 Grime famously considered \citep{grime_1974,grime_1977,grime_1988,westoby_1998} two broad determinants of population density: stress (persistent hardship e.g.
@@
 Similar schemes were proposed for insects \citep{southwood_1977}, fishes \citep{winemiller_1992}, and zooplankton \citep{allan_76}.
-More recently, modern hierarchical clustering techniques have revealed distinct trait clusters in corals analogous to Grime's primary strategies \citep{darling_2012}.
+More recently, modern hierarchical clustering techniques have also revealed, with greater statistical rigor, distinct trait clusters in corals analogous to Grime's primary strategies \citep{darling_2012}.
 These empirical findings suggest that functional traits contribute to fitness predominantly via a few key factors such as intrinsic reproductive rate or stress-tolerance.
@@
 \caption{\label{fig:axes} Schematic of Grime's triangle and the associated direction of selection in our model.
-The two axes show increasing levels of environmental stress and disturbance, respectively.
-Population persistence is not possible if the combination of stress and disturbance is too large (dashed line).
+The two axes show increasing levels of environmental stress and disturbance.
+Population persistence is not possible if the combination of stress and disturbance is too large (below dashed line).
 This creates a triangle, each corner of which corresponds to an environmental extreme and associated ``primary strategy''.
 The high-density (HD) interpretation of stress is shown as a rotation of the triangle with respect to density.
-Vertical arrows show how selection acts on $b$, $c$, and $d$: $c$ is irrelevant at low $L$, but otherwise $b$, $c$ and $d$ all experience the same selective force.}
+The strength of selection on $c$ depends on density $L$, but $b$ and $d$ experience the same selective force in all three environmental extremes in both versions of the triangle.}
 \end{figure}

### 2017-05-06 manuscript.tex
latest

@@
 This broadly follows the original spirit of MacArthur's $r/K$ scheme.
-More specifically, our aim is to begin merging some ecological realism to population genetics' time-dependent, genetically-focused view of evolution.
+More specifically, our aim is to begin merging some ecological realism into population genetics' time-dependent, genetically-focused view of evolution.
 We revisit the classic lottery model of \cite{chesson_1981}, which has two features that make it well suited for this role, but one critical flaw that we rectify here.
@@
 We then explore some time-dependent behavior of our extended lottery model.
-Taking an example inspired by recent studies of rapid, seasonal evolution in \textit{Drosophila} \citep{bergland_14}, we discuss how environmental fluctuations might stabilize polymorphisms in the presence of cyclical population density.
+Taking an example inspired by recent studies of rapid, seasonal evolution in \textit{Drosophila} \citep{bergland_14}, we discuss how environmental fluctuations might stabilize polymorphisms when population density is cyclical.
 
@@
 This is an excellent approximation when the $n_i$ are all large.
-How ever, we also do not evaluate the initial stochastic behaviour of adaptive mutant lineages while they are at low abundance.
+However, we also do not evaluate the initial stochastic behaviour of adaptive mutant lineages while they are at low abundance.
 When considering new mutations, we therefore restrict our attention to begin with the earliest (lowest $n_i$) deterministic behavior of mutant lineages (the transition to deterministic growth occurs at an abundance $n_i$ of order equal to their inverse expected absolute growth rate; \citealt{uecker_2011}).
@@
 \caption{\label{fig:lottery} Each iteration of our lottery model has three main elements.
-First, propagules are produced by adults which are dispersed at random over the unoccupied territories (only propagules landing on unoccupied territories are shown).
+First, propagules are produced by adults and are dispersed at random over the unoccupied territories (only propagules landing on unoccupied territories are shown).
 Lottery competition then occurs in each unoccupied territory (competition in only one territory is illustrated): each genotype has a probability proportional to $b_i n_i c_i$ of securing the territory.
@@
 
-For the majority of this manuscript we assume that mortality only occurs in adults (setting aside the juvenile deaths implicit in territorial contest), and at a constant, genotype-specific per-capita rate $0\leq d_i\leq 1$, so that the overall change in genotype abundances is
+For the majority of this manuscript we assume that mortality only occurs in adults (Fig 2; setting aside the juvenile deaths implicit in territorial contest), and at a constant, genotype-specific per-capita rate $0\leq d_i\leq 1$, so that the overall change in genotype abundances is
 \begin{equation}
@@
 \end{equation}
-This seems reasonable in the absence of disturbances; when we come to consider the effects of disturbances (Section ``Primary strategies and Grime's triangle''), we will incorporate disturbance-induced mortality in competing juveniles (Fig.~\ref{fig:lottery}).
+This seems reasonable in the absence of disturbances; when we come to consider the effects of disturbances (Section ``Primary strategies and Grime's triangle''), we will incorporate disturbance-induced mortality in competing juveniles.
 
@@
 
-Here we describe how Grime's ``disturbed'', ``stressful'' and ``ideal'' environments can be captured mathematically in our model, and evaluate role of selection in shaping the corresponding ``primary strategies''.
+Here we describe how Grime's ``disturbed'', ``stressful'' and ``ideal'' environments can be captured mathematically in our model, and evaluate the role of selection in shaping the corresponding ``primary strategies''.
 To proceed, we map these verbally-defined environments to quantitative parameter regimes in our model, summarized in Fig.~\ref{fig:table}.
 
-Ideal environments are characterized by the near-absence of stress and disturbance, so that survival is easy $d\ll 1$ and growth and reproduction can be rapid $b\gg d$.
-The propagule density is determined by the magnitude of $b/d$, and in this case $L\gg 1$ (Appendix C).
+Ideal environments are characterized by the near-absence of stress and disturbance, so that survival is easy ($d\ll 1$) and growth and reproduction can be rapid ($b\gg d$).
+The propagule density is determined by the magnitude of $b/d$, and hence $L\gg 1$ (Appendix C).
 
@@
 Severe stress inhibits growth and reproduction, so that $b\ll 1$.
-In Grime's view, this means that the rate at which propagules successfully develop to adulthood cannot appreciably exceed the mortality rate $b/d\approx 1$; it is just possible to cope with the environmental stress \citep{grime_1974,grime_1977}.
+In Grime's view, this means that the rate at which propagules successfully develop to adulthood cannot appreciably exceed the mortality rate ($b/d\approx 1$); it is just possible to cope with the environmental stress \citep{grime_1974,grime_1977}.
 In our model, this implies $L\ll 1$.
@@
 The expected direction of evolution is determined by three factors.
-First, there may be pleiotropy between $b$, $c$, and $d$, which could be due to trade-offs between them.
+First, there may be pleiotropy between $b$, $c$, and $d$, which could be due to physiological trade-offs between them.
 Second, adaptive mutations in one trait may occur more frequently than in others \citep{yampolsky_2001}.
-Third, selection favors some trait changes over others: $b$, $c$ and $d$ will tend to evolve in the direction of greatest fitness increase because the probability that an adaptive mutant survives its initial low abundance phase and begins to grow deterministically is proportional to its expected absolute growth rate \citep{haldane_1927,uecker_2011}.
-Only a tiny proportion of neutral and deleterious mutations fix (probability of order $1/N$ for neutrals).
+Third, selection favors some trait changes over others: $b$, $c$ and $d$ will tend to evolve in the direction of greatest fitness increase because the probability that an adaptive mutant survives its initial low abundance phase and begins to grow deterministically rather than be lost to drift is proportional to its expected absolute growth rate \citep{haldane_1927,uecker_2011}.
 
-Accounting for the first two factors would require an evaluation of the effects of mutations on traits subject to trade-offs, which is beyond the scope of this article.
-We only consider how selection shapes the direction of evolution for $b$, $c$ and $d$, assuming that mutations affect them independently and with the same distribution of beneficial fitness effects.
+Accounting for the first two factors is beyond the scope of this article.
+We only consider how selection shapes the direction of evolution for $b$, $c$ and $d$, assuming that mutations affect them independently and with comparable distributions of beneficial fitness effects.
 Specifically, we consider the effect of mutations that change $b$, $c$, or $d$ by a fixed proportion $b_j=(1+\epsilon) b_i$, $c_j=(1+\epsilon) c_i$ or $d_j=(1-\epsilon) d_i$, where $j$ is is a novel mutant (Appendix C).
-Mutations with magnitude $\epsilon$ are assumed to occur with the same frequency for all traits .
+Mutations with magnitude $\epsilon$ are assumed to occur with the same frequency for all traits.
 
 Not surprisingly, mutations which increase $c$ are effectively neutral when $L\ll 1$ and adaptive when $L\gg 1$.
-However, beyond this, selection in our density-dependent lottery model does not favor one trait improvement over another (Fig.
-3).
-This applies regardless of the initial magnitude of $b$ or $d$.
-More specifically, in ideal and stressed environments, where Eq.
-\eqref{eq:master} applies, $b$ and $d$ mutations of size $\epsilon$ have identical fitness effects (Appendix C), and the same is true for proportional $c$-mutations $c_j=(1+\epsilon) c_i$ when $L\gg 1$.
+However, selection in our density-dependent lottery model does not favor improvement in $b$ versus $d$.
+The same is true for proportional $c$-mutations $c_j=(1+\epsilon) c_i$ when $L\gg 1$ in ideal and stressed (HD) environments.
 
-In disturbed environments, $\Delta_+ n_i$ is replaced by $(1-d_i)\Delta_+ n_i$.
+In disturbed environments, where $\Delta_+ n_i$ is replaced by $(1-d_i)\Delta_+ n_i$, the argument for equal selection on $b$ and $d$ is more subtle and contingent.
 The bulk of the high mortality $d\approx 1$ is environmental and unavoidable.
-This implies that $\epsilon$ mutations in the remaining genetic part of mortality have a reduced impact on $d$, suggesting the primacy of $b$ under disturbance.
-However, this effect is easily swamped by the fact that $d_i\approx 1$ and that disturbance mortality affects both juvenile recruitment and adult mortality.
-In Appendix C, we argue that $b$ and $d$ mutations should have comparable effects on fitness (assuming that $d$ can evolve at all) because the fitness benefit of improving $d$ declines as $d$ improves and so $d$ cannot run ahead of $b$ for long (diminishing returns).
+This implies that $\epsilon$ mutations in the remaining genetic part of mortality have a reduced impact on total $d$, suggesting the primacy of $b$ under disturbance.
+However, this effect is easily matched by the fact that $d_i\approx 1$ and that disturbance mortality affects both juvenile recruitment and adult mortality.
+In Appendix C, we argue that $b$ and $d$ mutations should contrive to have comparable effects on fitness in the disturbance case (assuming that $d$ can evolve at all) because the fitness benefit of improving $d$ declines as $d$ improves.
 
@@
 
-In the previous section we only considered how $b$, $c$ and $d$ should respond in Grime's environmental extremes, based on invasion fitness.
+In the previous section we only considered how $b$, $c$ and $d$ should respond to selection in Grime's environmental extremes, based on invasion fitness.
 Here we further explore the low frequency behavior of Eq.
@@
 In a constant environment, stable coexistence is possible in our extended lottery model.
-A $b$ specialist $i$ and $c$-specialist $j$ ($b_i>b_j$, $c_j>c_i$) can co-exist because then propagule density $L$ is frequency dependent, and so is the importance of competitive ability (Appendix D).
-This is a version of the classic competition-colonization trade-off \citep{tilman_94,levins_71}: the competitor ($c$-specialist) leaves many territories unoccupied (low $L$) due to its poor colonization ability (low $b$), which the colonizer ($b$-specialist) can then exploit.
+A $b$-specialist $i$ and $c$-specialist $j$ ($b_i>b_j$, $c_j>c_i$) can co-exist because then propagule density $L$ is frequency-dependent, and so is the importance of competitive ability (Appendix D).
+This is a version of the classic competition-colonization trade-off \citep{tilman_94,levins_71}; the competitor ($c$-specialist) leaves many territories unoccupied (low $L$) due to its poor colonization ability (low $b$), which the colonizer ($b$-specialist) can then exploit.
 A similar situation holds for coexistence between high-$c$ and low-$d$ specialists; a ``competition-longevity'' trade-off \citep{tilman_94}.
 These forms of co-existence require density dependence (being mediated by $L$), and are not present in the classic lottery model.
-Coexistence is not possible between $b$- and $d$-specialists (Appendix D).
+Coexistence is not possible between $b$- and $d$-specialists in a constant environment (Appendix D).
 
@@
 
-Fig.
-\ref{fig:fluctuatingselection}a-c shows the behavior of Eq.
+Fig.~\ref{fig:fluctuatingselection}a-c shows the behavior of Eq.
 \eqref{eq:master} for an example where $b$ and $d$ cycle between zero and positive values (``summers'' with rapid growth and no mortality, and ``winters'' with mortality and no growth).
 Both the storage effect (adults are sheltered from selection during the summer growth phase) and the bounded density effect (expansion to high density occurs every cycle) are operating.
-Two types are present, a $b$ specialist, which is better at rapidly growing in the summer (higher $b$), and a $d$ specialist which is better at surviving the winter (lower $d$).
+Two types are present, a $b$-specialist, which is better at rapidly growing in the summer (higher $b$), and a $d$-specialist which is better at surviving the winter (lower $d$).
 Neither type has an advantage over a full environmental cycle, and they stably coexist.
@@
 The classic lottery model (Eq.
-1) gives dramatically different co-existence predictions because expansion to carrying capacity occurs immediately at the start of the summer (Fig.
+1) fails to give co-existence for these parameters because expansion to carrying capacity occurs immediately at the start of the summer (Fig.
 \ref{fig:fluctuatingselection}d-f).
 As a result, coexistence requires that the winter survivor's $b$ must be about $5$ times smaller than required when we properly account for the growth in the abundance of each type using Eq.
-\eqref{eq:master} (keeping the other parameters the same).
+\eqref{eq:master} (keeping the other parameters the same; Fig.~\ref{fig:fluctuatingselection}g-i).
 Previous models of the promotion of genetic variation via the storage effect  \citep{ellner_1994} similarly assume that the total number of offspring per iteration is constant, and would produce a similar error.
@@
 (a) Birth and death rates seasonally alternate being nonzero (white for winter, green for summer).
-The $b$ specialist (black) has higher $b$ and $d$ ($b=0.5$, $d=0.2$) than the $d$ specialist ($b=0.217$, $d=0.1$) (blue).
-(b) Both types grow during the positive $b$ phase, and decline during the positive $d$ phase, but the $d$ specialist does so at a lower rate.
+The $b$-specialist (black) has higher $b$ and $d$ ($b=0.5$, $d=0.2$) than the $d$-specialist ($b=0.217$, $d=0.1$) (blue).
+(b) Both types grow during the positive $b$ phase, and decline during the positive $d$ phase, but the $d$-specialist does so at a lower rate.
 Total height (blue+black) is population density $N/T$.
-(c) Summer favors the $b$ specialist, winter the $d$ specialist, and they stably coexist.
+(c) Summer favors the $b$ specialist, winter the $d$-specialist, and they stably coexist.
 (d-f) Same as (a-c) for the classic lottery model; the types no longer coexist.
@@
 Confusingly, the term ``$K$-selection'' sometimes refers generally to selection at high density \citep{pianka_1972}, encompassing both selection for higher saturation density \citep{macarthur_1967} and competitive ability \citep{gill_1974}.
-Contrary to predictions of an $r$/$K$ trade-off, empirical studies have shown that maximal growth rate and saturation density (measured by abundance) are positively correlated, both between species/strains \citep{luckinbill_1979,kuno_1991,hendriks_2005,fitzsimmons_2010}, and as a result of experimental evolution \citep{luckinbill_1978,luckinbill_1979}.
+Contrary to predictions of an $r$/$K$ trade-off, empirical studies have shown that maximal growth rate at low density and the high density at which saturation occurs (measured by abundance) are positively correlated, both between species/strains \citep{luckinbill_1979,kuno_1991,hendriks_2005,fitzsimmons_2010}, and as a result of experimental evolution \citep{luckinbill_1978,luckinbill_1979}.
 From the perspective of our model, this positive correlation is not surprising since the saturation density, which is determined by a balance between births and deaths, increases with $b$.
@@
 
-$K$-selection in the narrow logistic sense of selection for a greater environmental carrying capacity for given birth and death rates, sometimes referred to as ``efficiency'' \citep{macarthur_1967}, could be represented in our model by smaller individual territorial requirements.
-To a first approximation, two co-occurring genotypes which differ by a small amount in their territorial requirements only should have the same fitness since the costs or benefits of a change in the amount of unocupied territory is shared equally among genotypes via the propagule density per territory $L$.
+$K$-selection in the narrow logistic sense of selection for a greater environmental carrying capacity for given $r$, sometimes referred to as ``efficiency'' \citep{macarthur_1967}, could be represented in our model by smaller individual territorial requirements.
+To a first approximation, two co-occurring genotypes which differ by a small amount in their territorial requirements only should have the same fitness, since the costs or benefits of a change in the amount of unocupied territory is shared equally among genotypes via the propagule density per territory $L$.
 The situation is more complicated when the differences in territorial requirements become large enough that territorial contests can occur on different scales for different genotypes.
@@
 
-Our realization of Grime's triangle (Fig.~\ref{fig:axes}) differs from approaches which identify primary strategies as trait combinations which can stably co-exist \citep{bolker_99}, referring instead to the direction and rate of ongoing adaptive trait evolution under different regimes of stress and disturbance, which is closer in spirit to Grime's arguments \citep{grime_1974,grime_1977}.
+Our realization of Grime's triangle (Fig.~\ref{fig:axes}) differs from approaches which identify primary strategies as trait combinations that can stably co-exist \citep{bolker_99}, referring instead to the direction and rate of ongoing adaptive trait evolution under different regimes of stress and disturbance, which is closer in spirit to Grime's arguments \citep{grime_1974,grime_1977}.
 Moreover, our formulation is mathematical, in contrast to Grime's original verbal and descriptive approach, which is a recognized hindrance to the evaluation or broader application of the C/S/R scheme (e.g.
@@
 However, section ``Primary strategies and Grime's triangle'' suggests that, apart from the obvious irrelevance of territorial contests at low density, selection alone has no preference for fecundity, competitive ability or longevity.
-This could mean that one or both of the factors we set aside --- pleiotropy/trade-offs and mutation bias --- are needed to get  true ``primary strategy'' trait differentiation.
-A separate model of trade-offs and mutation effects would be needed to explore this possibility.
+One or both of the factors we set aside --- pleiotropy/trade-offs and mutation bias --- are needed to get  true ``primary strategy'' trait differentiation.
 
-Nevertheless, it is interesting to note that ruderals, which are typically thought of as high fecundity dispersers ($b$ specialists), may also be strongly $d$-selected, consistent with our findings.
+Nevertheless, it is interesting to note that ruderals, which are typically thought of as high fecundity dispersers ($b$-specialists), may also be strongly $d$-selected, which while unintuitive, is consistent with our findings.
 An effective way to reduce $d$ in the face of unavoidable physical destruction is to shorten the time to reproductive maturity --- short life cycles are a characteristically ruderal trait.
@@
 We are not aware of any attempts to delineate conditions under which explicit resource tracking is unnecessary even if it is assumed that community structure is ultimately determined by competition for consumable resources.
-More work is needed connecting resource competition models to the density-dependent selection literature, since most of the former has to date been focused on narrower issues of the role of competition at low resource availability \citep{aerts_1999,davis_1998,tilman_2007}.
+More work is needed connecting resource competition models to the density-dependent selection literature, since most of the former has to date been focused on narrower issues of the role of competition at low resource availability and in the absence of direct interactions between organisms at the same trophic level \citep{aerts_1999,davis_1998,tilman_2007}.
 
@@
 It and similar forms of ``spatial co-existence'' in stable environments have previously been modeled either with Levin's  qualitative representation of competition \citep{levins_71,tilman_94}, as opposed to the quantitative $c$ of lottery competition, or with a more sophisticated treatment of space (non-uniform dispersal; \citealt{shmida_84,bolker_99}).
-In cyclical environments, polymorphisms can be stabilized by the bounded density effect which is  completely lost if there is an exclusive focus on allele frequencies \citep{yi_2013}.
-We leave the details of how our model might be applied in inference, including the crucial issue of its genetic drift predictions (providing a null model for neutral sites), for future work.
+In cyclical environments, polymorphisms can be stabilized by the bounded density effect, which is  completely lost if there is an exclusive focus on allele frequencies \citep{yi_2013}.
+We leave the details of how our model might be applied to inference problems, including the crucial issue of its genetic drift predictions (providing a null model for neutral sites), for future work.
 
@@
 \eqref{eq:lottery} yields the same expression by assuming $L\rightarrow \infty$ even though this assumption is unrealistic for a rare mutant).
-Similarly to Grime's stressful environment, improving $b$, $c$, or $d$ by a fraction $\epsilon$ yields identical fitness benefits, even though they may have very different magnitudes, because the ancestral resident type's density-regulated birth rate $b_i c_i/L$ exactly balances its mortality $d_i$ in equilibrium.
+Similarly to Grime's stressful environment, improving $b$, $c$, or $d$ by a fraction $\epsilon$ yields identical fitness benefits, even though they may have very different absolute magnitudes, because the ancestral resident type's density-regulated birth rate $b_i c_i/L$ exactly balances its mortality $d_i$ in equilibrium.
 Biases in the direction of evolution of $b$, $c$ or $d$ must again follow from biases in mutation availability or trade-offs/pleiotropy.
@@
 The effects of improving $b$ and $d$ are now no longer identical for given values of $d_{ie}$ and $d_{ig}$.
-A $b-$mutation $b_j=(1+\epsilon)b_i$ gives $\Delta n_j/n_j = \epsilon (1-d_i)b_i\approx \epsilon d_i \approx \epsilon$, whereas a $d$-mutation $d_{jg}=(1-\epsilon) d_{ig}$ gives $\Delta n_j/n_j = \epsilon d_{ig}/(1-d_i)$.
-Since overall mortality is high $d_i\approx 1$, $d_{g}$ can be small compared to $d_{e}$ and yet still make a greater contribution to invasion fitness than $b$.
+A $b$-mutation $b_j=(1+\epsilon)b_i$ gives $\Delta n_j/n_j = \epsilon (1-d_i)b_i\approx \epsilon d_i \approx \epsilon$, whereas a $d$-mutation $d_{jg}=(1-\epsilon) d_{ig}$ gives $\Delta n_j/n_j = \epsilon d_{ig}/(1-d_i)$.
+Since overall mortality is high ($d_i\approx 1$), $d_{g}$ can be small compared to $d_{e}$ and yet still make a greater contribution to invasion fitness than $b$.
 The case where $d_{ig}$ is so small that $d_{ig}\ll 1-d_i$ (i.e.
 $d_{ig}$ is an order of magnitude smaller than $1-d_i$ which itself an order of magnitude smaller than $d_{ie}$) effectively represents a hard constraint that $d$ cannot evolve.
-We do not deny this possibility, but this bring us back to the issue of mutations affect traits (the first two factors controlling the direction of evolution in section ``Primary strategies and Grime's triangle''), which we do not address in this manuscript.
+We do not deny this possibility, but this brings us back to issues of mutational bias and constraint (the first two factors controlling the direction of evolution in section ``Primary strategies and Grime's triangle''), which we do not address in this manuscript.
 
@@
 
-We consider the case of coexistence between a $b$ specialist $i$ and a $c$ specialist $j$ ($b_i>b_j$, $c_j>c_i$ and $d_i=d_j$).
+We consider the case of coexistence between a $b$-specialist $i$ and a $c$-specialist $j$ ($b_i>b_j$, $c_j>c_i$ and $d_i=d_j$).
 Suppose that $b_i$ is so large that $L\gg 1$ when $i$ is dominant, and $b_j$ is so small that $L\ll 1$ when $j$ is dominant.
@@
 Therefore, coexistence occurs if $c_j/c_i$ is sufficiently large.
-The analogous argument for $d$ and $c$ specialists ($d_i<d_j$ with $L\gg 1$ when $i$ dominates, $L\ll 1$ when $j$ dominates, and $b_i=b_j$) gives $\Delta n_j/n_j \approx d_i\frac{c_j}{c_i}-d_j$, which again implies coexistence if $c_j/c_i$ is sufficiently large.
+The analogous argument for $d$- and $c$-specialists ($d_i<d_j$ with $L\gg 1$ when $i$ dominates, $L\ll 1$ when $j$ dominates, and $b_i=b_j$) gives $\Delta n_j/n_j \approx d_i\frac{c_j}{c_i}-d_j$, which again implies coexistence if $c_j/c_i$ is sufficiently large.
 
-For $b$ and $d$ specialists ($c_i=c_j$), we have $\Delta n_j/n_j\approx b_j d_i/b_i-d_j$ when $i$ dominates and $\Delta n_i/n_i\approx b_i d_j/b_j-d_i$ when $j$ dominates.
+For $b$-and $d$-specialists ($c_i=c_j$), we have $\Delta n_j/n_j\approx b_j d_i/b_i-d_j$ when $i$ dominates and $\Delta n_i/n_i\approx b_i d_j/b_j-d_i$ when $j$ dominates.
 Thus, either $i$ or $j$ grows when rare, but not both, and stable coexistence is not possible in a constant environment.

### 2017-05-08 manuscript.tex
latest

@@
 Yet $d_{ig}$ cannot be appreciably greater than $1-d_i$ for long, because selection will then strongly favor reductions in $d_g$ over increases in $b$, thereby bringing $d_{ig}$ closer to $1-d_i$.
-Thus, we expect selection on $b$ and $d$ to be of comparable strength in disturbed environments as well.
+Thus, if $b$ and $d$ are subject to long-term evolution subject to external degradation at comparable rates \citep{bertram_2017}, then we expect selection on $b$ and $d$ to be of comparable strength in disturbed environments as well.
 

### 2017-05-16 manuscript.tex
latest

@@
 \title{A lottery model of density-dependent selection in evolutionary genetics}
-%\author{Jason Bertram $^{1,\ast}$ \\ 
-%Joanna Masel $^{1}$}
+\author{Jason Bertram $^{1,\ast}$ \\ 
+Joanna Masel $^{1}$}
 
@@
 
-%\noindent{}1. Department of Ecology and Evolutionary Biology, University of Arizona, Tucson, AZ 85721.
+\noindent{}1. Department of Ecology and Evolutionary Biology, University of Arizona, Tucson, AZ 85721.
 
-%\noindent{}$\ast$ Corresponding author; e-mail: jbertram@email.arizona.edu.
+\noindent{}$\ast$ Corresponding author; e-mail: jbertram@email.arizona.edu.
 
@@
 
-\textit{Manuscript elements}:
+%\textit{Manuscript elements}:
 
@@
 
-\textit{Manuscript type}: Article. 
+%\textit{Manuscript type}: Article. 
 % Or e-article, note, e-note, natural history miscellany,
@@
 
-\noindent{\footnotesize Prepared using the suggested \LaTeX{} 
-template for \textit{Am.\ Nat.}}
+%\noindent{\footnotesize Prepared using the suggested \LaTeX{} 
+%template for \textit{Am.\ Nat.}}
 
 \linenumbers{}
-\modulolinenumbers[3]
+\modulolinenumbers[1]
 
@@
 This example is inspired by natural \textit{Drosophila} populations, which expand rapidly in the warmer months when fruit is abundant, but largely die off in the colder months.
-Within this seasonal population density cycle, hundreds of polymorphisms also cycle in frequency \citep{bergland_14}.
+Along with this seasonal population density cycle, hundreds of polymorphisms exhibit frequency cycles that are in phase with the seasons \citep{bergland_14}.
 Some of these polymorphisms may be adaptive and potentially millions of years old, suggesting stable coexistence \citep{bergland_14,messer_2016}.
@@
 The classical population genetic treatment of fluctuating selection suggests that environmental fluctuations do not promote coexistence.
-Allele frequencies are successively multiplied by relative fitness values for each environmental iteration, and so two alleles favored in different enviroments can only stably coexist if the product of fitnesses for one type exactly equals the product for the other \citep{dempster_1955}.
+Allele frequencies are successively multiplied by relative fitness values for each environmental iteration, and so two alleles favored in different environments can only stably coexist if the product of fitnesses for one type exactly equals the product for the other \citep{dempster_1955}.
 Thus, stable coexistence still requires frequency-dependent selection or heterozygote advantage (as is required in a constant environment).
 
-This classical argument overlooks two general effects that promote coexistence in fluctuating environments.
-The first is the storage effect, which occurs when part of the population is protected from selection (due to overlapping generations in the lottery model; \citealt{chesson_1981}).
+This classical argument overlooks two general mechanisms that promote coexistence in fluctuating environments \citep{messer_2016}.
+The first is the classic version of the storage effect, which occurs when part of the population is protected from selection (due to overlapping generations in the lottery model; \citealt{chesson_1981}).
 The second is the bounded population size effect of \cite{yi_2013}, which occurs when each environmental cycle involves growth from low to high density, with the time spent growing each cycle dependent on the fitness of the types present.
@@
 
-%\section*{Acknowledgments}
+\section*{Acknowledgments}
 
-%We thank Peter Chesson and Joachim Hermisson for many constructive comments on this manuscript. This work was financially supported by the National Science Foundation (DEB-1348262).
+We thank Peter Chesson and Joachim Hermisson for many constructive comments on this manuscript. This work was financially supported by the National Science Foundation (DEB-1348262).
 
-\bibliographystyle{amnatnat}
+\bibliographystyle{plainnat}
 \bibliography{reference}

### 2017-05-16 manuscript.tex
latest

@@
 \title{A lottery model of density-dependent selection in evolutionary genetics}
-\author{Jason Bertram $^{1,\ast}$ \\ 
-Joanna Masel $^{1}$}
+%\author{Jason Bertram $^{1,\ast}$ \\ 
+%Joanna Masel $^{1}$}
 
@@
 
-\section*{Acknowledgments}
+%\section*{Acknowledgments}
 
-We thank Peter Chesson and Joachim Hermisson for many constructive comments on this manuscript. This work was financially supported by the National Science Foundation (DEB-1348262).
+%We thank Peter Chesson and Joachim Hermisson for many constructive comments on this manuscript. This work was financially supported by the National Science Foundation (DEB-1348262).
 

### 2017-09-22 manuscript.tex
latest

@@
 
-Fitness is typically represented in heavily simplified terms in evolutionary genetics, often using constant selection coefficients, to make it easier to infer or predict how type frequencies change over time.
+Fitness is typically represented in heavily simplified terms in evolutionary genetics, often using constant selection coefficients
+
 This excludes fundamental ecological factors such as dynamic population size or density-dependence from the most genetically-realistic treatments of evolution, a problem that inspired MacArthur's influential but problematic $r$/$K$ theory.
-Following the spirit of $r$/$K$-selection as a general-purpose theory of density-dependent selection, we develop a new model of density-dependent selection by generalizing the fixed-density classic lottery model of territorial acquisition to accommodate arbitrary population densities.
+
+Following the spirit of $r$/$K$-selection as a general-purpose theory of density-dependent selection
+
+new model of density-dependent selection by generalizing the fixed-density classic lottery model of territorial acquisition to accommodate arbitrary population densities.
+
 We show that, with density dependence, co-existence is possible in the lottery model in a stable environment.
+
 Inspired by natural \textit{Drosophila} populations, we consider co-existence under strong, seasonally-fluctuating selection coupled to large cycles in population density, and show that co-existence (stable polymorphism) is promoted via a combination of the classic storage effect and density-regulated population growth.
-We also show that the only significant bias introduced by selection at the different environmental extremes of Grime's triangle is the relative importance of competitive ability at different densities, confirming an important role for phenotypic constraints in shaping ``primary strategies''.
 
@@
 
-``...the concept of fitness is probably too complex to allow of a useful mathematical development.
-Since it enters fundamentally into many population genetics considerations, it is remarkable how little attention has been paid to it.'' --- Warren J.
-Ewens, Mathematical Population Genetics I, 2004, pp.
-276
-
 \section*{Introduction}
 
-Evolutionary models differ greatly in their treatment of fitness.
 In models of genetic evolution, genotypes are typically assigned constant (or occasionally frequency-dependent) selection coefficients describing the change in their relative frequencies over time.
@@
 
-It is challenging to generalize beyond particular traits or ecological scenarios to model fundamentally different forms of selection.
-Perhaps this is not surprising given that fitness is such a complex quantity, dependent on all of a phenotype's functional traits \citep{violle_2007} and its environment.
-A detailed, trait-based, predictive model of fitness would be enormously complicated and situation-specific.
-It is therefore easy to doubt the feasibility of a simplified, general mathematical treatment of fitness \citep[p.
-276]{ewens_2004}.
  For example, MacArthur's famous $r$/$K$ scheme \citep{macarthur_1962,macarthur_1967} is now almost exclusively known as a framework for understanding life-history traits, and judged on its failure in that role \citep{pianka_1970,stearns_1977,boyce_1984,reznick_2002}.
@@
 
-However, empirical trait classification studies have suggested the existence of a few ``primary strategies'', reflecting broadly distinct responses to selection \citep{winemiller_2015}.
-Grime famously considered \citep{grime_1974,grime_1977,grime_1988,westoby_1998} two broad determinants of population density: stress (persistent hardship e.g.
-due to resource scarcity or unfavorable temperatures) and disturbance (intermittent destruction of vegetation e.g.
-due to trampling, herbivory, pathogens, extreme weather or fire).
-The extremes of these two factors define three primary strategies denoted by C/S/R respectively (Fig.
-\ref{fig:axes}): competitors ``C'' excel in low stress, low disturbance environments; stress tolerators ``S'' excel in high stress, low disturbance environments; and ruderals ``R''  excel in low stress, high disturbance environments.
-Population persistence is not possible in high-stress, high-disturbance environments.
-Grime showed that measures of C, S and R across a wide range of plant species are anti-correlated, so that strong C-strategists are weak S and R strategists, and so on, creating a triangular C/S/R ternary plot \citep{grime_1974}.
-Similar schemes were proposed for insects \citep{southwood_1977}, fishes \citep{winemiller_1992}, and zooplankton \citep{allan_76}.
-More recently, modern hierarchical clustering techniques have also revealed, with greater statistical rigor, distinct trait clusters in corals analogous to Grime's primary strategies \citep{darling_2012}.
-These empirical findings suggest that functional traits contribute to fitness predominantly via a few key factors such as intrinsic reproductive rate or stress-tolerance.
+in a simplified, territorial model of growth, dispersal and competition.
 
-\begin{figure}
-\centering
-\includegraphics[scale=1]{axes.pdf}
-\caption{\label{fig:axes} Schematic of Grime's triangle and the associated direction of selection in our model.
-The two axes show increasing levels of environmental stress and disturbance.
-Population persistence is not possible if the combination of stress and disturbance is too large (below dashed line).
-This creates a triangle, each corner of which corresponds to an environmental extreme and associated ``primary strategy''.
-The high-density (HD) interpretation of stress is shown as a rotation of the triangle with respect to density.
-The strength of selection on $c$ depends on density $L$, but $b$ and $d$ experience the same selective force in all three environmental extremes in both versions of the triangle.}
-\end{figure}
-
-Here we explore the interplay between some ``key factors'' of fitness in a simplified, territorial model of growth, dispersal and competition.
-This broadly follows the original spirit of MacArthur's $r/K$ scheme.
-More specifically, our aim is to begin merging some ecological realism into population genetics' time-dependent, genetically-focused view of evolution.
 We revisit the classic lottery model of \cite{chesson_1981}, which has two features that make it well suited for this role, but one critical flaw that we rectify here.
@@
 
-Using our extended lottery model, we then revisit Grime's C/S/R scheme, and evaluate the extent to which selection favors fecundity, competitive ability or lower mortality under Grime's environments (section ``Primary strategies and Grime's triangle'').
-This represents a simple ``sanity check'' on the invasion behavior of our extended lottery model under different extremes.
-
-We then explore some time-dependent behavior of our extended lottery model.
 Taking an example inspired by recent studies of rapid, seasonal evolution in \textit{Drosophila} \citep{bergland_14}, we discuss how environmental fluctuations might stabilize polymorphisms when population density is cyclical.
@@
 
-\subsection*{Primary strategies and Grime's triangle}
-
-Here we describe how Grime's ``disturbed'', ``stressful'' and ``ideal'' environments can be captured mathematically in our model, and evaluate the role of selection in shaping the corresponding ``primary strategies''.
-To proceed, we map these verbally-defined environments to quantitative parameter regimes in our model, summarized in Fig.~\ref{fig:table}.
-
-Ideal environments are characterized by the near-absence of stress and disturbance, so that survival is easy ($d\ll 1$) and growth and reproduction can be rapid ($b\gg d$).
-The propagule density is determined by the magnitude of $b/d$, and hence $L\gg 1$ (Appendix C).
-
-Disturbed environments are characterized by short bursts of high extrinsic mortality $d$ caused by physical destruction.
-We assume that disturbances are equally damaging to adults and juveniles, so that only $(1-d_i)\Delta_+ n_i$ (rather than $\Delta_+ n_i$) territories are secured by type $i$ each iteration.
-In the heavily disturbed extreme of Grime's triangle, many such bursts (which could each be spatially localized) will occur each iteration (i.e.
-over the time required for a propagule to reach reproductive maturity), which we approximate as a high constant mortality rate $d\approx 1$.
-Population persistence requires $b>d/(1-d)$, which implies $b\gg 1$ (severe disturbance comes with a high fecundity/dispersal requirement).
-We assume this requirement can just be met, so that $b\approx d/(1-d)$ and $L\ll 1$ (Appendix C).
-
-Stressful environments are more ambiguous, and have been the subject of an extensive debate in the plant ecology literature (the ``Grime-Tilman'' debate; \citealt{aerts_1999} and references therein).
-Severe stress inhibits growth and reproduction, so that $b\ll 1$.
-In Grime's view, this means that the rate at which propagules successfully develop to adulthood cannot appreciably exceed the mortality rate ($b/d\approx 1$); it is just possible to cope with the environmental stress \citep{grime_1974,grime_1977}.
-In our model, this implies $L\ll 1$.
-
-The alternative view is that stressed environments are highly competitive.
-In spite of inhibited growth and reproduction, density is actually high relative to carrying capacity of the stressful environment \citep{taylor_1990}.
-For example, if consumable resources are scarce, we expect intense resource competition \citep{davis_1998}.
-Thus, even though $b\ll 1$, we still have $b/d\gg 1$ and $L\ll 1$.
-
-\begin{figure}
-\centering
-\begin{tabular}{l*{4}{c}}
-  & Ideal & Disturbance* & Stress (G) & Stress (HD) \\ \hline
-  Constraints & $d \ll 1$ & $d \approx 1$ & $b \ll 1$ & $b \ll 1$ \\
-  Other parameters & $b/d\gg 1$ & $(1-d)b\approx d$ & $b\approx d$ & $b/d\gg 1$ \\
-  Propagule density $L$  & $\gg 1$ & $\ll 1$ & $\ll 1$ & $\gg 1$
-\end{tabular}
-\caption{\label{fig:table} The mathematical representation of Grime's three environmental extremes in our model, as well as the high-density (HD) variant of the stressful environment.
-*Mortality affects both adults and juveniles in the disturbed environment, with $\Delta_+ n_i$ replaced by $(1-d_i)\Delta_+ n_i$.}
-\end{figure}
-
-The expected direction of evolution is determined by three factors.
-First, there may be pleiotropy between $b$, $c$, and $d$, which could be due to physiological trade-offs between them.
-Second, adaptive mutations in one trait may occur more frequently than in others \citep{yampolsky_2001}.
-Third, selection favors some trait changes over others: $b$, $c$ and $d$ will tend to evolve in the direction of greatest fitness increase because the probability that an adaptive mutant survives its initial low abundance phase and begins to grow deterministically rather than be lost to drift is proportional to its expected absolute growth rate \citep{haldane_1927,uecker_2011}.
-
-Accounting for the first two factors is beyond the scope of this article.
-We only consider how selection shapes the direction of evolution for $b$, $c$ and $d$, assuming that mutations affect them independently and with comparable distributions of beneficial fitness effects.
-Specifically, we consider the effect of mutations that change $b$, $c$, or $d$ by a fixed proportion $b_j=(1+\epsilon) b_i$, $c_j=(1+\epsilon) c_i$ or $d_j=(1-\epsilon) d_i$, where $j$ is is a novel mutant (Appendix C).
-Mutations with magnitude $\epsilon$ are assumed to occur with the same frequency for all traits.
-
-Not surprisingly, mutations which increase $c$ are effectively neutral when $L\ll 1$ and adaptive when $L\gg 1$.
-However, selection in our density-dependent lottery model does not favor improvement in $b$ versus $d$.
-The same is true for proportional $c$-mutations $c_j=(1+\epsilon) c_i$ when $L\gg 1$ in ideal and stressed (HD) environments.
-
-In disturbed environments, where $\Delta_+ n_i$ is replaced by $(1-d_i)\Delta_+ n_i$, the argument for equal selection on $b$ and $d$ is more subtle and contingent.
-The bulk of the high mortality $d\approx 1$ is environmental and unavoidable.
-This implies that $\epsilon$ mutations in the remaining genetic part of mortality have a reduced impact on total $d$, suggesting the primacy of $b$ under disturbance.
-However, this effect is easily matched by the fact that $d_i\approx 1$ and that disturbance mortality affects both juvenile recruitment and adult mortality.
-In Appendix C, we argue that $b$ and $d$ mutations should contrive to have comparable effects on fitness in the disturbance case (assuming that $d$ can evolve at all) because the fitness benefit of improving $d$ declines as $d$ improves.
-
-%How does Fig.~4 compare to empirical analyses of Grime's C/S/R strategies?
-In our comparison we will stick to fishes, corals and plants, for which three-way primary strategy schemes are well developed \citep{grime_1977,winemiller_1992,darling_2012}.
-The connection of our model to fish strategies is necessarily more tentative, given that fishes are motile and not all territorial.
-%
-%In disturbed environments, we predict evolution for higher $b$ and lower $d$, but not higher $c$.
-Higher $b$ means higher fecundity and/or dispersal ability (see ``Model'').
-This is consistent with a ruderal strategy.
-Plant ruderals devote a large proportion of their productivity to seed production \citep{grime_1977}, whereas the analogous ``opportunist'' fishes have large intrinsic growth rates \citep{winemiller_1992}.
-In corals, ruderals are distinguished by brood spawning (rather than broadcast spawning; \citealt{darling_2012}).
-This corresponds to higher parental investment and lower overall propagule production --- counter-intuitively, a stereotypical ``$K$-selected'', high-density trait \citep{pianka_1970}.
-However, since broadcast spawners are vulnerable to an Allee effect at the egg fertilization stage \citep{knowlton_2001}, brood spawning could actually be a way to ensure high $b$ at low densities \citep{darling_2012}.
-Lower $d$ could be achieved by improved individual resistance to physical destruction, but it is hard to reduce mortality in the face of severe disturbances.
-Alternatively, shortening the time to reproductive maturity (the iteration time in our model) is an effective way of reducing the chance of death per iteration, $d$, for a given frequency of disturbance.
-An exceptionally short life cycle is probably the most defining characteristic of ruderals \citep{grime_1977,winemiller_1992,darling_2012}.
-Note that if evolution manages to appreciably reduce $d$ for a given disturbance intensity, then the population no longer lies at the extreme disturbance vertex of Grime's triangle.
-Thus, ruderals are characterized by both high $b$ and high $d$, but there is a constant pressure for a shorter life cycle, extending the limits of disturbance that can be tolerated.
-%
-%In stressful environments, we predict evolution for lower $d$, and also  for higher $c$ in the high-density interpretation of stressful environments.
-Low $d$ is essential when $b\ll 1$, and stress tolerant plants and corals have long life spans, allowing for long intervals between successful recruitments (and episodic broadcast spawning in corals).
-For fishes, the ``equilibrium'' strategy is the analogue of Grime's stress tolerator.
-This strategy is associated with consumable resource limitation, and is also characterized by long life span, as well as high parental investment in tiny broods.
-This may reflect a high-$c$ strategy in the face of intense competition for severely limited resources (the high-density interpretation).
-%
-%In ideal environments, we predict evolution for higher $b$ and $c$, but not lower $d$.
-In plants and corals, a key mechanism for winning territorial contests (higher $c$) is rapidly outgrowing and ``shading out'' competitors; empirically, rapid individual growth is a defining feature of the competitor trait cluster \citep{grime_1977,darling_2012}.
-The situation for $b$ is more ambiguous.
-Competitor strategies in plants and corals span a range of $b$.
-For fishes, the analogous ``periodic'' strategy is characterized by enormous spawn sizes as well as rapid development \citep{winemiller_1992,winemiller_2015}.
-The role of $b$ in ideal environments will be discussed further below.
-
 \subsection*{Coexistence in constant and cyclical environments}\label{sec:invas}
@@
 
-%Disentangling the storage and bounded population size effects is not straightforward, and requires a more detailed discussion of each effect than we have space for here.
-Our model is well suited for such a disentangling since it extends the canonical example of the storage effect to allow for density-dependent effects.
-
 \begin{figure}
@@
 
-%While adaptive evolution in the direction predicted by our model does produce traits broadly consistent with Grime's scheme, trait data on selection for higher $b$ at high density was ambiguous.
-This prediction is also counter to the expectations of MacArthur's $r$/$K$ dichotomy \citep{macarthur_1967}, since $b$ is closely related to the maximal, low-density growth rate $r=b-d$ \citep{pianka_1972}, yet in the $r$/$K$ scheme, high density populations should be subject to $K$, not $r$, selection.
+%This prediction is also counter to the expectations of MacArthur's $r$/$K$ dichotomy \citep{macarthur_1967}, since $b$ is closely related to the maximal, low-density growth rate $r=b-d$ \citep{pianka_1972}, yet in the $r$/$K$ scheme, high density populations should be subject to $K$, not $r$, selection.
 Yet it is not surprising that $b$ can matter at high densities.
@@
 We leave these complications for future work.
-
-Our realization of Grime's triangle (Fig.~\ref{fig:axes}) differs from approaches which identify primary strategies as trait combinations that can stably co-exist \citep{bolker_99}, referring instead to the direction and rate of ongoing adaptive trait evolution under different regimes of stress and disturbance, which is closer in spirit to Grime's arguments \citep{grime_1974,grime_1977}.
-Moreover, our formulation is mathematical, in contrast to Grime's original verbal and descriptive approach, which is a recognized hindrance to the evaluation or broader application of the C/S/R scheme (e.g.
-\citealt{tilman_2007}).
-However, section ``Primary strategies and Grime's triangle'' suggests that, apart from the obvious irrelevance of territorial contests at low density, selection alone has no preference for fecundity, competitive ability or longevity.
-One or both of the factors we set aside --- pleiotropy/trade-offs and mutation bias --- are needed to get  true ``primary strategy'' trait differentiation.
 

### 2017-09-23 manuscript.tex
latest

@@
 
-In models of genetic evolution, genotypes are typically assigned constant (or occasionally frequency-dependent) selection coefficients describing the change in their relative frequencies over time.
+Density-dependent selection was one of the first topics to explicitly connect ecology and evolution \citep{travis_2013}.
+
+The stated aim of MacArthur's famous $r$/$K$ scheme was to extend the existing population-genetic treatment of selection to account for population density \citep{macarthur_1962}.
+Few attempts have been made to develop it further along these lines.
+
+ For example, MacArthur's famous $r$/$K$ scheme \citep{macarthur_1962,macarthur_1967} is now almost exclusively known as a framework for understanding life-history traits, and judged on its failure in that role \citep{pianka_1970,stearns_1977,boyce_1984,reznick_2002}.
+
+It is common practice in population genetics to assign constant selection coefficients to genotypes describing the change in their relative frequencies over time.
+
 This simplified treatment of selection facilitates explicit time-dependent treatment of genotype frequencies, and can be justified over sufficiently short time intervals \citep[p.
@@
 For instance, adaptive dynamics uses ``invasion fitness'' to explore the consequences of eco-evolutionary feedbacks \citep{diekmann_2004}.
-
- For example, MacArthur's famous $r$/$K$ scheme \citep{macarthur_1962,macarthur_1967} is now almost exclusively known as a framework for understanding life-history traits, and judged on its failure in that role \citep{pianka_1970,stearns_1977,boyce_1984,reznick_2002}.
-However, the r/K scheme's original purpose was to extend the existing population-genetic treatment of selection to account for population density \citep{macarthur_1962}.
-Few attempts have been made to develop it further along these lines.
 
@@
 \includegraphics[scale=0.8]{lottery.pdf}
-\caption{\label{fig:lottery} Each iteration of our lottery model has three main elements.
-First, propagules are produced by adults and are dispersed at random over the unoccupied territories (only propagules landing on unoccupied territories are shown).
-Lottery competition then occurs in each unoccupied territory (competition in only one territory is illustrated): each genotype has a probability proportional to $b_i n_i c_i$ of securing the territory.
-Then occupied territories are freed up by adult mortality.
-In Eq.
-\eqref{eq:delttot} and most of the paper, only adults can die (red crosses), but we will also consider the case where juveniles die (blue cross; section ``Primary strategies and Grime's triangle'').}
+\caption{\label{fig:lottery} Each iteration of our model has three elements.
+First, propagules are produced by adults and dispersed at random (only propagules landing on unoccupied territories are shown).
+Lottery competition then occurs in each unoccupied territory (only illustrated in one territory).
+Each genotype has a probability proportional to $c_i x_i$ of securing a given territory, where $c_i$ measures competitive ability and $x_i$ is the number of propagules that disperse there.
+In the illustrated territory, the black genotype disperses more propagules but is a poorer competitor.
+Territories are then made available by adult mortality (red crosses).}
 \end{figure}
@@
 
-In the sections ``Invasion of rare genotypes and coexistence'' and ``Primary strategies and Grime's triangle'', we assume the simpler form $m_i=b_i n_i$, with constant $b_i$, meaning that all propagules land on unoccupied territories (a form of directed dispersal).
+In ``Invasion of rare genotypes and coexistence'', we assume the simpler form $m_i=b_i n_i$, with constant $b_i$, meaning that all propagules land on unoccupied territories (a form of directed dispersal).
 This simplifies the mathematics without affecting the results of those sections, which only depend on the low-frequency invasion behavior of Eq.
@@
 
-For the majority of this manuscript we assume that mortality only occurs in adults (Fig 2; setting aside the juvenile deaths implicit in territorial contest), and at a constant, genotype-specific per-capita rate $0\leq d_i\leq 1$, so that the overall change in genotype abundances is
+We assume that mortality only occurs in adults (Fig.~\ref{fig:lottery}; setting aside the juvenile deaths implicit in territorial contest), and at a constant, genotype-specific per-capita rate $0\leq d_i\leq 1$, so that the overall change in genotype abundances is
 \begin{equation}
@@
 \end{equation}
-This seems reasonable in the absence of disturbances; when we come to consider the effects of disturbances (Section ``Primary strategies and Grime's triangle''), we will incorporate disturbance-induced mortality in competing juveniles.
 
@@
 
-\section*{Appendix C: Mutant invasion under different environments}
+\section*{Appendix C: Mutant invasion and coexistence in a constant environment}
 
-In this Appendix we evaluate the invasion of novel mutants in a population with a single resident type $i$ (such that $N=n_i$), which is in equilibrium.
+Here we evaluate the initial growth or decline of mutants in a population with a single resident type, which is in equilibrium.
+To determine whether coexistence is possible, we check for ``mutual invasion'', that is, we check that type $j$ will invade an $i$-dominated population, but type $i$ will also invade a $j$-dominated population.
 
-For a single type $i$ in equilibrium ($\Delta n_i = 0$), we have $R_i=0$, $\overline{c}=c_i$, $A_i=(1-(1+L)e^{-L})/L$, and Eq. \eqref{eq:master} becomes
+Solving for equilibrium when $i$ is the resident ($\Delta n_i = 0$), we have $R_i=0$, $\overline{c}=c_i$, $A_i=(1-(1+L)e^{-L})/L$, and Eq. \eqref{eq:master} becomes
 \begin{equation}
@@
 \end{equation}
-(Alternatively, Eq.
-\eqref{eq:equil} can be deduced directly from Eq.
-\eqref{eq:growthsumuncoupled}).
-This implies $L\approx b_i/d_i$ when $b_i/d_i\gg 1$ and $L\ll 1$ when $b_i/d_i\approx 1$.
+This implies $L\approx b_i/d_i$ if $b_i/d_i\gg 1$ and $L\ll 1$ if $b_i/d_i\approx 1$.
+
 Now suppose that a novel mutant $j$, which is initially rare, appears in the population.
@@
 \end{equation}
-where $R_j\approx (1-e^{-L})/\left(\frac{c_j}{c_i}+\frac{L-1+e^{-L}}{1-(1+L)e^{-L}}\right)$ since $l_j\ll 1$.
-
-We now consider mutant invasion under the different environments from section ``Primary strategies and Grime's triangle''.
-In our representation of Grime's stressful environment, $b\ll 1$, $d\approx b$ and $L\ll 1$, and so Eq.
-\eqref{eq:invad} becomes $\Delta n_j/n_j \approx b_j-d_j$.
-Mutations which improve $b$ or $d$ by the same fraction $\epsilon$, such that $b_j = (1+ \epsilon)b_i$ or $d_j = (1- \epsilon)d_i$,  yield identical fitness benefits.
-
-In our representation of Grime's ideal environment, $d\ll 1$, $b_i/d_i \gg 1$ and $L\gg 1$, and so Eq. \eqref{eq:invad} becomes 
-\begin{equation}
-\Delta n_j/n_j \approx \frac{b_j}{c_j/c_i + L-1}\frac{c_j}{c_i}-d_j\approx \frac{b_j}{L}\frac{c_j}{c_i}-d_j, \label{eq:highLinvad}
-\end{equation}
-where the last approximation follows from the fact that $c_j$ will be of similar magnitude to $c_i$ (ignoring mutations with very large effect sizes; incidentally, Eq.
-\eqref{eq:lottery} yields the same expression by assuming $L\rightarrow \infty$ even though this assumption is unrealistic for a rare mutant).
-Similarly to Grime's stressful environment, improving $b$, $c$, or $d$ by a fraction $\epsilon$ yields identical fitness benefits, even though they may have very different absolute magnitudes, because the ancestral resident type's density-regulated birth rate $b_i c_i/L$ exactly balances its mortality $d_i$ in equilibrium.
-Biases in the direction of evolution of $b$, $c$ or $d$ must again follow from biases in mutation availability or trade-offs/pleiotropy.
-The same argument applies for the high-$L$ interpretation of the stress regime.
-
-To represent Grime's disturbed environment, $\Delta_+ n_i$ is replaced with $(1-d_i)\Delta_+ n_i$, and a single type reaches equilibrium when $L/(1-e^{-L})=(1-d_i)b_i/d_i$ (population persistence now requires $(1-d_i)b_i>d_i$ rather than $b_i>d_i$).
-We assumed $d_i\approx 1$ and $(1-d_i)b_i\approx d_i$, so that $L\ll 1$.
-A mutant's invasion fitness is then given by
-\begin{equation}
-\Delta n_j/n_j \approx (1-d_j)b_j - d_j.
-\end{equation}
-Another distinct feature of disturbed environments is that the bulk of mortality is environmental and unavoidable.
-That is, $d_i=d_{ie}+d_{ig}$, where the environmental part $d_{ie}$ does not evolve, and the genetic part $d_{ig}$ is affected proportionally by mutations $d_{jg}=(1-\epsilon) d_{ig}$.
-
-The effects of improving $b$ and $d$ are now no longer identical for given values of $d_{ie}$ and $d_{ig}$.
-A $b$-mutation $b_j=(1+\epsilon)b_i$ gives $\Delta n_j/n_j = \epsilon (1-d_i)b_i\approx \epsilon d_i \approx \epsilon$, whereas a $d$-mutation $d_{jg}=(1-\epsilon) d_{ig}$ gives $\Delta n_j/n_j = \epsilon d_{ig}/(1-d_i)$.
-Since overall mortality is high ($d_i\approx 1$), $d_{g}$ can be small compared to $d_{e}$ and yet still make a greater contribution to invasion fitness than $b$.
-The case where $d_{ig}$ is so small that $d_{ig}\ll 1-d_i$ (i.e.
-$d_{ig}$ is an order of magnitude smaller than $1-d_i$ which itself an order of magnitude smaller than $d_{ie}$) effectively represents a hard constraint that $d$ cannot evolve.
-We do not deny this possibility, but this brings us back to issues of mutational bias and constraint (the first two factors controlling the direction of evolution in section ``Primary strategies and Grime's triangle''), which we do not address in this manuscript.
-
-Thus, assuming that $d$ can evolve in this sense, $d_{ig}$ is either comparable to or greater than $1-d_i$.
-Yet $d_{ig}$ cannot be appreciably greater than $1-d_i$ for long, because selection will then strongly favor reductions in $d_g$ over increases in $b$, thereby bringing $d_{ig}$ closer to $1-d_i$.
-Thus, if $b$ and $d$ are subject to long-term evolution subject to external degradation at comparable rates \citep{bertram_2017}, then we expect selection on $b$ and $d$ to be of comparable strength in disturbed environments as well.
-
-\section*{Appendix D: Coexistence in a stable environment}
-
-To determine whether coexistence is possible in a constant environment, we check for ``mutual invasion'', that is, we check that $j$ will invade an $i$-dominated population, but $i$ will also invade a $j$-dominated population.
+where $R_j\approx (1-e^{-L})/\left(\frac{c_j}{c_i}+\frac{L-1+e^{-L}}{1-(1+L)e^{-L}}\right)$.
 

### 2017-10-02 manuscript.tex
latest

@@
 
-\title{A lottery model of density-dependent selection in evolutionary genetics}
+\title{A lottery model of density-dependent selection}
 %\author{Jason Bertram $^{1,\ast}$ \\ 
@@
 
-Density-dependent selection was one of the first topics to explicitly connect ecology and evolution \citep{travis_2013}.
+One of the simplest and most widely used models of selection is to assign genotypes constant selection coefficients describing how their relative proportions change over time.
+This model is common in population genetics, where its simplicity is useful for analyzing the detailed time course of genotypic frequencies.
 
-The stated aim of MacArthur's famous $r$/$K$ scheme was to extend the existing population-genetic treatment of selection to account for population density \citep{macarthur_1962}.
-Few attempts have been made to develop it further along these lines.
+Yet selection does not only change relative genotypic proportions.
+Ultimately, it involves the births and deaths of individuals, and is therefore intimately connected to changes in absolute abundance \cite{evolution_demography_stuff}.
+This immediately poses a challenge: any description of evolution grounded in abundances needs to specify how population density is regulated, as well as how fitness depends on density, which may differ between genotypes (density-dependent selection).
+As a result, the early density-dependent selection literature was among the first to explicitly grapple with the interplay between ecology and evolution \citep{travis_2013}.
 
- For example, MacArthur's famous $r$/$K$ scheme \citep{macarthur_1962,macarthur_1967} is now almost exclusively known as a framework for understanding life-history traits, and judged on its failure in that role \citep{pianka_1970,stearns_1977,boyce_1984,reznick_2002}.
+Much of the work on density-dependent selection grew out of MacArthur's $r$/$K$-selection scheme \citep{macarthur_1962,macarthur_1967} in which $r$-selection favors faster growth at low density, whereas $K$-selection favors the ability to keep growing at high density.
+Although $r$ and $K$ are named after the common ``carrying capacity'' parameterization of the logistic model, the concepts of $r$- and $K$-selection are not restricted to that model \citep{macarthur_1967}.
+Nevertheless, much of the density-dependent selection literature has adopted variants of the logistic model, often in the setting of Mendelian inheritance \cite{roughgarden etc}.
 
-It is common practice in population genetics to assign constant selection coefficients to genotypes describing the change in their relative frequencies over time.
+with genotype-specific $K$ values parameterizing growth ability at high density.
+A closely related approach is to add Lotka-Volterra-type inter-genotypic competition terms with $\alpha$ competition coefficients, which adds ``$\alpha$-selection'' to the mix \citealt{gill_1974,case_1974,joshi_2001}.
 
-This simplified treatment of selection facilitates explicit time-dependent treatment of genotype frequencies, and can be justified over sufficiently short time intervals \citep[p.
-276]{ewens_2004}.
-The emphasis of population genetics is to infer past selection, migration and demographic change given a sample of nucleotide sequences, or to predict how allele frequencies change over time based on their relative fitness effects together with population structure, genetic drift and linkage.
-The resulting picture of evolution excludes basic elements of the ecological underpinnings of selection, including density dependence, and how selection affects population size.
-This complicates the inference of past selection, because demographic changes can look genealogically very similar to selective frequency changes \citep{barton_1998}.
+ variant that allows for inter-genotypic versus intra-genotypic
 
-By contrast, models of phenotypic trait evolution use absolute fitness functions to describe how some traits of interest affect survival and reproduction in particular ecological scenarios \citep{metz_1992,diekmann_2004}.
-These fitness functions can be quite problem-specific and often only account for a few traits at a time.
-The emphasis here is on the conditions for invasion from low frequencies and co-existence, rather than frequency or abundance trajectories over time.
-For instance, adaptive dynamics uses ``invasion fitness'' to explore the consequences of eco-evolutionary feedbacks \citep{diekmann_2004}.
+or  explicit resource consumption \citep{tilman_1982}, are much more complicated.
+
+$r$/$K$ scheme \citep{macarthur_1962,macarthur_1967} is now almost exclusively known as a framework for understanding life-history traits, and judged on its failure in that role \citep{pianka_1970,stearns_1977,boyce_1984,reznick_2002}.
 
@@
 Territorial contest among propagules leaves a single victorious adult per territory, the victor chosen at random from the propagules present, with probabilities weighted by a coefficient for each type representing competitive ability, akin to a lottery \citep{sale_77}.
-By comparison, coefficients for the pairwise effects of types on each other (e.g.
-the $\alpha$ coefficients in the generalized Lotka-Volterra equations and the associated concept of ``$\alpha$-selection''; \citealt{gill_1974,case_1974,joshi_2001}), or  explicit resource consumption \citep{tilman_1982}, are much more complicated.
+
 The second feature is the close connection between the lottery model and one of the foundational models of population genetics, the Wright-Fisher model of genetic drift, which we discuss further below.

### 2017-10-11 manuscript.tex
latest

@@
 
-%\textit{Manuscript type}: Article. 
-% Or e-article, note, e-note, natural history miscellany,
-% e-natural history miscellany, comment, reply, symposium, or
-% countdown to 150.
-
-\bigskip
-
-%\noindent{\footnotesize Prepared using the suggested \LaTeX{} 
-%template for \textit{Am.\ Nat.}}
-
 \linenumbers{}
@@
 
-One of the simplest and most widely used models of selection is to assign genotypes constant selection coefficients describing how their relative proportions change over time.
-This model is common in population genetics, where its simplicity is useful for analyzing the detailed time course of genotypic frequencies.
+Some of the earliest explicit discussions of the interplay between ecology and evolution appear in the density-dependent selection literature \citep{travis_2013}.
+
+Much of this work grew out of MacArthur's $r$/$K$-selection scheme \citep{macarthur_1962,macarthur_1967} in which $r$-selection favors faster growth at low density, whereas $K$-selection favors the ability to keep growing at high density.
+
+Although $r$ and $K$ are named after the common ``carrying capacity'' parameterization of the logistic model, the concepts of $r$- and $K$-selection are not restricted to that model \citep{macarthur_1967}.
 
@@
 This immediately poses a challenge: any description of evolution grounded in abundances needs to specify how population density is regulated, as well as how fitness depends on density, which may differ between genotypes (density-dependent selection).
-As a result, the early density-dependent selection literature was among the first to explicitly grapple with the interplay between ecology and evolution \citep{travis_2013}.
+As a result,
 
-Much of the work on density-dependent selection grew out of MacArthur's $r$/$K$-selection scheme \citep{macarthur_1962,macarthur_1967} in which $r$-selection favors faster growth at low density, whereas $K$-selection favors the ability to keep growing at high density.
-Although $r$ and $K$ are named after the common ``carrying capacity'' parameterization of the logistic model, the concepts of $r$- and $K$-selection are not restricted to that model \citep{macarthur_1967}.
+One of the simplest and most widely used models of selection is to assign genotypes constant selection coefficients describing how their relative proportions change over time.
+This model is common in population genetics, where its simplicity is useful for analyzing the detailed time course of genotypic frequencies.
+
  Nevertheless, much of the density-dependent selection literature has adopted variants of the logistic model, often in the setting of Mendelian inheritance \cite{roughgarden etc}.

### 2017-10-13 manuscript.tex
latest

@@
 
-Some of the earliest explicit discussions of the interplay between ecology and evolution appear in the density-dependent selection literature \citep{travis_2013}.
+The density-dependent selection literature contains some of the earliest explicit analyses of the interplay between ecology and evolution \citep{travis_2013}.
+This work is most famously associated with MacArthur's $r$/$K$ scheme \citep{macarthur_1962,macarthur_1967} in which $r$-selection favors faster growth at low density, whereas $K$-selection favors the ability to keep growing at high density.
 
-Much of this work grew out of MacArthur's $r$/$K$-selection scheme \citep{macarthur_1962,macarthur_1967} in which $r$-selection favors faster growth at low density, whereas $K$-selection favors the ability to keep growing at high density.
+Starting with Kostitzin's
 

### 2017-10-13 manuscript.tex
latest

@@
 \section*{Introduction}
+
+Selection has two broadly different connotations, which both appear in the simplest models of evolution.
+The first is that if one genotype grows faster than another, its frequency will increase.
+Mathematically, if one genotype's abundance increases at a constant per-capita rate $r_1$, and another at $r_2$, then the frequency of the first $p_1$ changes at rate $(r_1-r_2)p_1 p_2$.
+Here $s=r_1-r_2$ is the ``selection coefficient''.
+The second is directly phrased in terms of genotypic frequencies.
+Two genotypes are assigned relative fitnesss $w_1$ and $w_2$ respectively, and then $p_1$ changes at rate $(w_1-w_2)/\overline{w} p_1 p_2$, where $\overline{w}=w_1p_1+w_2p_2$ is mean relative fitness.
+While this formulation is agnostic about population size $N$, it is often assumed that $N$ is constant.
+Thus, in the first connotation, selection is simply differential growth, and does not require interactions between genotypes.
+The second, on the other hand, connotes the displacement of one genotypes by another, which is intimately connected to the ecological concept of competition \cite{gill_1974}.
+
+The existence of these two different forms of selection was perceived long ago by MacArthur in his $r$/$K$ scheme \citep{macarthur_1962,macarthur_1967}, and before that by Kostitzin \cite{kostitzin_book}.
 

### 2017-10-14 manuscript.tex
latest

--- 2017-10-13 manuscript.tex
@@
 
+In a huge class of evolutionary models, and in many introductory texts, selection is described in terms of shifts in genotypic frequencies (e.g.
+\cite[pp.
+468]{barton_2007}).
+Genotypes are assigned constant selection coefficients describing how their relative proportions change over time.
+These selective shifts in frequency result from two fundamentally different population processes: i) in the absence of interactions between individuals, genotypic growth rates differ; ii) as a result of competitive interactions between individuals, some genotypes displace others.
+Population density is the key factor that controls whether individuals interact, thus controlling the relative contributions of these forms of selection.
+
+The distinction between these forms of selection was perceived long ago by Kostitzin \citep{kostitzin_1939,christiansen_2004}, but is mostly associated with MacArthur's $r$/$K$ scheme \citep{macarthur_1962,macarthur_1967} and the following literature on density-dependent selection.
+
 The density-dependent selection literature contains some of the earliest explicit analyses of the interplay between ecology and evolution \citep{travis_2013}.
@@
 
-Starting with Kostitzin's
-
 Although $r$ and $K$ are named after the common ``carrying capacity'' parameterization of the logistic model, the concepts of $r$- and $K$-selection are not restricted to that model \citep{macarthur_1967}.
 
-Yet selection does not only change relative genotypic proportions.
-Ultimately, it involves the births and deaths of individuals, and is therefore intimately connected to changes in absolute abundance \cite{evolution_demography_stuff}.
  This immediately poses a challenge: any description of evolution grounded in abundances needs to specify how population density is regulated, as well as how fitness depends on density, which may differ between genotypes (density-dependent selection).
 As a result,
-
-One of the simplest and most widely used models of selection is to assign genotypes constant selection coefficients describing how their relative proportions change over time.
-This model is common in population genetics, where its simplicity is useful for analyzing the detailed time course of genotypic frequencies.
 

### 2017-10-15 manuscript.tex
latest

--- 2017-10-13 manuscript.tex
@@
 
-The density-dependent selection literature contains some of the earliest explicit analyses of the interplay between ecology and evolution \citep{travis_2013}.
-This work is most famously associated with MacArthur's $r$/$K$ scheme \citep{macarthur_1962,macarthur_1967} in which $r$-selection favors faster growth at low density, whereas $K$-selection favors the ability to keep growing at high density.
+MacArthur famously distinguished between ``$r$-selection'' for faster growth at low population density, and ``$K$-selection'' for the ability to keep growing at higher densities \citep{macarthur_1962,macarthur_1967}.
+The $r$/$K$ scheme is mostly known, and critiqued, for its role in the study of life-history evolution \citep{pianka_1970,boyce_1984,charlesworth_1994,reznick_2002}.
+The $r$/$K$ scheme also initiated a sizeable literature on density-dependent selection  (note that Kostitzin presented similar ideas much earlier \citep{kostitzin_1939,christiansen_2004}).
+This literature is remarkable in that it contains some of earliest explicit analyses of the interplay between ecology and evolution \citep{travis_2013}, yet it has left a complex legacy.
 
-Starting with Kostitzin's
+Although $r$ and $K$ are named after the parameters of the logistic model, $K$ generally refers to carrying capacity regardless of how the density regulation of population size occurs \citep{macarthur_1967}.
+Yet the ability to keep growing at ever higher densities is only one aspect of selection at high population density.
+Competition is critically important at high densities, and cannot be represented solely in terms of the density regulation of total population size.
+This led to the notion of ``$\alpha$-selection'' after the competition coefficients in the Lotka-Volterra equations \citep{gill_1974,case_1974,joshi_2001}.
 
-Although $r$ and $K$ are named after the common ``carrying capacity'' parameterization of the logistic model, the concepts of $r$- and $K$-selection are not restricted to that model \citep{macarthur_1967}.
+In the standard representation of selection, genotypes are assigned selection coefficients describing how their relative proportions change over time (e.g. \cite[pp. 468]{barton_2007}).
 
-Yet selection does not only change relative genotypic proportions.
-Ultimately, it involves the births and deaths of individuals, and is therefore intimately connected to changes in absolute abundance \cite{evolution_demography_stuff}.
-This immediately poses a challenge: any description of evolution grounded in abundances needs to specify how population density is regulated, as well as how fitness depends on density, which may differ between genotypes (density-dependent selection).
-As a result,
+It is thus broadly misleading to view $r$- and $K$-selection as two main contrasting forms of selection, and the widespread use of the logistic model in the density-dependent literature is more problematic than simply being a mathematical idealization [citations].
 
-One of the simplest and most widely used models of selection is to assign genotypes constant selection coefficients describing how their relative proportions change over time.
-This model is common in population genetics, where its simplicity is useful for analyzing the detailed time course of genotypic frequencies.
+The true contrast hinted at by the $r$/$K$ scheme is between interaction-dependent, and interaction-independent selection.
+That is, selective shifts in frequency are a result of differences in growth rates between genotypes, and genotypic growth rates may differ in the absence of interactions between individuals, or as a result of interactions between individuals.
 
- Nevertheless, much of the density-dependent selection literature has adopted variants of the logistic model, often in the setting of Mendelian inheritance \cite{roughgarden etc}.
+Population density is a key factor controling whether individuals interact, thereby setting the relative contributions of these forms of selection.
+Since a huge class of growth-affecting interactions are competitive, it seems clear that the study of density-dependent selection should be grounded a reasonable representation of competition and how its effects vary with density.
 
-with genotype-specific $K$ values parameterizing growth ability at high density.
-A closely related approach is to add Lotka-Volterra-type inter-genotypic competition terms with $\alpha$ competition coefficients, which adds ``$\alpha$-selection'' to the mix \citealt{gill_1974,case_1974,joshi_2001}.
-
- variant that allows for inter-genotypic versus intra-genotypic
-
-or  explicit resource consumption \citep{tilman_1982}, are much more complicated.
-
-$r$/$K$ scheme \citep{macarthur_1962,macarthur_1967} is now almost exclusively known as a framework for understanding life-history traits, and judged on its failure in that role \citep{pianka_1970,stearns_1977,boyce_1984,reznick_2002}.
+%including considerable effort devoted to showing that evolution under density-dependent selection optimizes population size in some sense, akin to a maximization of fitness.
 

### 2017-10-25 manuscript.tex
latest

@@
 MacArthur famously distinguished between ``$r$-selection'' for faster growth at low population density, and ``$K$-selection'' for the ability to keep growing at higher densities \citep{macarthur_1962,macarthur_1967}.
-The $r$/$K$ scheme is mostly known, and critiqued, for its role in the study of life-history evolution \citep{pianka_1970,boyce_1984,charlesworth_1994,reznick_2002}.
-The $r$/$K$ scheme also initiated a sizeable literature on density-dependent selection  (note that Kostitzin presented similar ideas much earlier \citep{kostitzin_1939,christiansen_2004}).
-This literature is remarkable in that it contains some of earliest explicit analyses of the interplay between ecology and evolution \citep{travis_2013}, yet it has left a complex legacy.
+While the $r$/$K$ scheme is primarily known for its seminal role in the study of life-history evolution \citep{pianka_1970,boyce_1984,charlesworth_1994,reznick_2002}, it
+also initiated a sizeable literature on density-dependent selection.
+This literature is remarkable in that it contains some of earliest models of the interplay between ecology and evolution \citep{roughgarden_1979,travis_2013},
 
-Although $r$ and $K$ are named after the parameters of the logistic model, $K$ generally refers to carrying capacity regardless of how the density regulation of population size occurs \citep{macarthur_1967}.
-Yet the ability to keep growing at ever higher densities is only one aspect of selection at high population density.
-Competition is critically important at high densities, and cannot be represented solely in terms of the density regulation of total population size.
+Identifying selection at high densities with ``$K$-selection'' is a compelling oversimplification. Starting from a very general description of density-regulated population growth
+\begin{equation}
+\frac{d n_1}{d t}=f_1(n_1,n_2)\qquad\frac{d n_2}{d t}=f_2(n_1,n_2)
+\end{equation}
+
+``population genetics of crowded populations where fitness is $K$''.
+
+ $K$ generally refers to carrying capacity regardless of how the density regulation of population size occurs \citep{macarthur_1967}.
+
+%Competition is critically important at high densities, and cannot be represented solely in terms of the density regulation of total population size.
 This led to the notion of ``$\alpha$-selection'' after the competition coefficients in the Lotka-Volterra equations \citep{gill_1974,case_1974,joshi_2001}.
 
+Yet selection at high population density clearly involves more than carrying capacity.
+
+and the widespread use of the logistic model in the density-dependent literature is more problematic than simply being a mathematical idealization [citations].
+
 In the standard representation of selection, genotypes are assigned selection coefficients describing how their relative proportions change over time (e.g. \cite[pp. 468]{barton_2007}).
-
-It is thus broadly misleading to view $r$- and $K$-selection as two main contrasting forms of selection, and the widespread use of the logistic model in the density-dependent literature is more problematic than simply being a mathematical idealization [citations].
 

### 2017-10-25 simulations.py

@@ -37,16 +37,25 @@ def R(m,c,U):
     l=m/float(U)
     L=sum(l)
     cbar=sum(m*c)/sum(m); 
-    return cbar*np.exp(-l)*(1-np.exp(-(L-l)))\
+    out=cbar*np.exp(-l)*(1-np.exp(-(L-l)))\
             /(c + (L-1+np.exp(-L))/(1-(1+L)*np.exp(-L))*(cbar*L-c*l)/(L-l))
+    
+    for i in range(len(out)):
+        if np.isnan(out)[i]: out[i]=0
+            
+    return out
 
 def A(m,c,U):
     l=m/float(U)
     L=sum(l)
     cbar=sum(m*c)/sum(m)
-    return (1-np.exp(-l))\
+    out=(1-np.exp(-l))\
             /((1-np.exp(-l))/(1-(1+l)*np.exp(-l))*c*l\
             +(L*(1-np.exp(-L))/(1-(1+L)*np.exp(-L))-l*(1-np.exp(-l))/(1-(1+l)*np.exp(-l)))/(L-l)*(sum(c*l)-c*l))
+    for i in range(len(out)):
+        if np.isnan(out)[i]: out[i]=0
+            
+    return out
     
 def deltnplus(m,c,U):
     if sum(m)>0 and U>0:
@@ -172,7 +181,7 @@ def d(t):
 c=np.array([1.,1.])
 T=100000
 g=40.     #iterations per seasonal cycle
-bparam=np.array([[0.0,0.0],[.5,.217]])
+bparam=np.array([[0.0,0.0],[2.,.5]])
 dparam=np.array([[0.0,0.0],[0.2,0.1]])
 totaltime=100
 
@@ -225,7 +234,7 @@ n=np.array([60000.,25000.])
 nhistclassic=[n];
 for t in range(totaltime):
     U=T-sum(n)
-    n = n + deltnplusclassic((b(t)*n*U/T).astype(np.int),c,U.astype(np.int))-d(t)*n
+    n = n + deltnplusclassicboundedm((b(t)*n*U/T).astype(np.int),c,U.astype(np.int))-d(t)*n
     nhistclassic.append(list(n))
 
 fig=plt.subplot(332)
@@ -274,7 +283,7 @@ n=np.array([60000.,25000.])
 nhistclassic=[n];
 for t in range(totaltime):
     U=T-sum(n)
[…]

### 2017-10-25 manuscript.tex
latest

@@
 also initiated a sizeable literature on density-dependent selection.
-This literature is remarkable in that it contains some of earliest models of the interplay between ecology and evolution \citep{roughgarden_1979,travis_2013},
+This literature is remarkable for containing some of earliest models of the interplay between ecology and evolution \citep{roughgarden_1979,travis_2013}.
 
-Identifying selection at high densities with ``$K$-selection'' is a compelling oversimplification. Starting from a very general description of density-regulated population growth
+In MacArthur's view, selection at high densities is ``$K$-selection'', such that in ``crowded populations ...
+fitness is $K$''.
+MacArthur's argument is based on general equations for density-regulated population growth
 \begin{equation}
-\frac{d n_1}{d t}=f_1(n_1,n_2)\qquad\frac{d n_2}{d t}=f_2(n_1,n_2)
+\frac{d n_1}{d t}=f_1(n_1,n_2)\qquad\frac{d n_2}{d t}=f_2(n_1,n_2),
 \end{equation}
-
-``population genetics of crowded populations where fitness is $K$''.
+where, by definition, each
 

### 2017-10-25 manuscript.tex
latest

@@
 
-In MacArthur's view, selection at high densities is ``$K$-selection'', such that in ``crowded populations ...
-fitness is $K$''.
-MacArthur's argument is based on general equations for density-regulated population growth
+Given the 
+For MacArthur, in ``crowded populations ... fitness is $K$'' \citep[pp. 149]{macarthur_1967}. The argument  is quite general,
+
+ for density-regulated population growth a community with two types
+
+MacArthur's argument is based on very general 
 \begin{equation}
@@
 \end{equation}
-where, by definition, each
+where $n_1$ and $n_2$ are the respective densities of the two types, and the growth rates $f_1$ and $f_2$ decrease with both $n_1$ and $n_2$.
+
+ the environmental carrying capacities for each type are defined by $f_1(K_1,0)=0$
 
  $K$ generally refers to carrying capacity regardless of how the density regulation of population size occurs \citep{macarthur_1967}.
+
+Yet the $r$/$K$ scheme is strikingly incompatible with the  model of selection
 

### 2017-10-26 manuscript.tex
latest

@@
 
-Given the 
+MacArthur justified the $r$/$K$ scheme as follows. Consider two types subject to density-dependent population regulation described by
+\begin{equation}
+\frac{d n_1}{d t}=f_1(n_1,n_2)\qquad\frac{d n_2}{d t}=f_2(n_1,n_2), \label{eq:macgeneral}
+\end{equation}
+where $n_1$ and $n_2$ are the respective densities of the two types, and the growth functions $f_1$ and $f_2$ decrease with both $n_1$ and $n_2$.
+Now suppose that the population starts growing from low density.
+Then $n_1$ and $n_2$ will increase until at least one type stops growing; it makes no difference which type, so assume $f_1(n_1,n_2)=0$.
+Then $n_1$ can never increase again,
+
+Then $n_2$ either keeps growing, or $f_2(n_1,n_2)=0$ as well. In either case, $n_2$ is guaranteed to persist,
+
+if the population is allowed to grow according Eq. \eqref{eq:macgeneral}
+
+persistence depends on being able to grow at higher values of $n_1$ and $n_2$ than the other type in some region. Formally, a type will be excluded unless there are some values of $n_1$ and $n_2$
+
+In the simplest case that each type suppresses its own growth and the growth of the other type equally, $f_1=f_1(N)$ and $f_2=f_2(N)$ only depend on total density $n_1+n_2=N$. If
+
+the environmental carrying capacities for each type are defined by $f_1(K_1,0)=0$
+
+ and any type that can keep growing
+
 For MacArthur, in ``crowded populations ... fitness is $K$'' \citep[pp. 149]{macarthur_1967}. The argument  is quite general,
@@
  for density-regulated population growth a community with two types
-
-MacArthur's argument is based on very general 
-\begin{equation}
-\frac{d n_1}{d t}=f_1(n_1,n_2)\qquad\frac{d n_2}{d t}=f_2(n_1,n_2),
-\end{equation}
-where $n_1$ and $n_2$ are the respective densities of the two types, and the growth rates $f_1$ and $f_2$ decrease with both $n_1$ and $n_2$.
-
- the environmental carrying capacities for each type are defined by $f_1(K_1,0)=0$
 

### 2017-10-26 manuscript.tex
latest

--- 2017-10-25 manuscript.tex
@@
 
-MacArthur famously distinguished between ``$r$-selection'' for faster growth at low population density, and ``$K$-selection'' for the ability to keep growing at higher densities \citep{macarthur_1962,macarthur_1967}.
+MacArthur famously distinguished between ``$r$-selection'' for faster growth at low population density, and ``$K$-selection'' for the ability to keep growing at high densities \citep{macarthur_1962,macarthur_1967}.
 While the $r$/$K$ scheme is primarily known for its seminal role in the study of life-history evolution \citep{pianka_1970,boyce_1984,charlesworth_1994,reznick_2002}, it
@@
 
-In MacArthur's view, selection at high densities is ``$K$-selection'', such that in ``crowded populations ...
-fitness is $K$''.
-MacArthur's argument is based on general equations for density-regulated population growth
+Verbal interpretations of the $r$/$K$ scheme, and the logistic model, tend to obscure the generality of MacArthur's argument, and so it is worth briefly reviewing that argument here.
+Consider two types with densities $n_1$ and $n_2$ subject to density-dependent population regulation described by
 \begin{equation}
-\frac{d n_1}{d t}=f_1(n_1,n_2)\qquad\frac{d n_2}{d t}=f_2(n_1,n_2),
+\frac{d n_1}{d t}=f_1(n_1,n_2)\qquad\frac{d n_2}{d t}=f_2(n_1,n_2), \label{eq:macgeneral}
 \end{equation}
-where, by definition, each
+where $f_1$ and $f_2$ are strictly decreasing functions of $n_1$ and $n_2$.
+By plotting the nullclines $f_1(n_1,n_2)=0$ and $f_2(n_1,n_2)=0$ in $(n_1,n_2)$ space, it becomes clear that for type $2$ to persist, its nullcline must enclose a region of $(n_1,n_2)$ space that type $1$'s nullcline does not.
+This region locally corresponds to greater $n_1$ and/or $n_2$ than can be tolerated by type $1$, because $f_1$ and $f_2$ are strictly decreasing.
+In this sense, selection at high density favors the ability to keep growing at ever greater densities.
 
- $K$ generally refers to carrying capacity regardless of how the density regulation of population size occurs \citep{macarthur_1967}.
+MacArthur then analyzed the possible outcomes of Eq.~\eqref{eq:macgeneral} in terms of "carrying capacities" (generally labeled ``$K$''), defined as the points where the nullclines intersect the axes (i.e.
+$f_1(K_{11},0)=0$, $f_1(0,K_{12})=0$, $f_2(0,K_{22})=0$, $f_2(K_{21},0)=0$).
+These intersection points neatly summarize the high density behavior of each type provided that the nullclines are approximately linear.
+He concluded that in ``crowded populations ...
+fitness is $K$'' \citep[pp.
+149]{macarthur_1967}.
+Note that only $K_{11}$ and $K_{22}$ are saturation densities; $K_{12}$ and $K_{21}$ measure the effects of competition between types.
+For instance, if $f_1(n_1,n_2)=r_1(1-\alpha_{11}n_1-\alpha_{12}n_2)n_1$, then  $K_{12}=\alpha_{12}$ is just the standard Lotka-Volterra competition coefficient.
+This led to the distinction between selection for greater saturation density ($K$-selection in the strict sense) and selection for greater competitive ability (``$\alpha$-selection'') at high density \citep{gill_1974,case_1974,joshi_2001}.
 
-%Competition is critically important at high densities, and cannot be represented solely in terms of the density regulation of total population size.
-This led to the notion of ``$\alpha$-selection'' after the competition coefficients in the Lotka-Volterra equations \citep{gill_1974,case_1974,joshi_2001}.
-
-Yet selection at high population density clearly involves more than carrying capacity.
-
-and the widespread use of the logistic model in the density-dependent literature is more problematic than simply being a mathematical idealization [citations].
+Yet the $r$/$K$ scheme is strikingly incompatible with the  model of selection
 

### 2017-10-27 manuscript.tex
latest

@@
 
-Verbal interpretations of the $r$/$K$ scheme, and the logistic model, tend to obscure the generality of MacArthur's argument, and so it is worth briefly reviewing that argument here.
+Verbal interpretations of the $r$/$K$ scheme, and the widespread use of the logistic model when discussing it, have tended to obscure the compelling generality of MacArthur's argument.
+We therefore  briefly review that argument now.
 Consider two types with densities $n_1$ and $n_2$ subject to density-dependent population regulation described by
@@
 By plotting the nullclines $f_1(n_1,n_2)=0$ and $f_2(n_1,n_2)=0$ in $(n_1,n_2)$ space, it becomes clear that for type $2$ to persist, its nullcline must enclose a region of $(n_1,n_2)$ space that type $1$'s nullcline does not.
-This region locally corresponds to greater $n_1$ and/or $n_2$ than can be tolerated by type $1$, because $f_1$ and $f_2$ are strictly decreasing.
-In this sense, selection at high density favors the ability to keep growing at ever greater densities.
+Since $f_1$ and $f_2$ are strictly decreasing, this corresponds to growth of type $2$ in a region of greater $n_1$ and/or $n_2$ than can be tolerated by type $1$.
+The term ``$K$-selection'' comes from four ``carrying capacities'' defined as the points where the nullclines intersect the axes (i.e.
+$f_1(K_{11},0)=0$, $f_1(0,K_{12})=0$, $f_2(0,K_{22})=0$, $f_2(K_{21},0)=0$).
+These intersection points summarize the nullcline structure provided that the nullclines are sufficiently linear, leading to the conclusion that in ``crowded populations ...
+fitness is $K$'' \citep[pp.
+149]{macarthur_1967}
 
-MacArthur then analyzed the possible outcomes of Eq.~\eqref{eq:macgeneral} in terms of "carrying capacities" (generally labeled ``$K$''), defined as the points where the nullclines intersect the axes (i.e.
-$f_1(K_{11},0)=0$, $f_1(0,K_{12})=0$, $f_2(0,K_{22})=0$, $f_2(K_{21},0)=0$).
-These intersection points neatly summarize the high density behavior of each type provided that the nullclines are approximately linear.
-He concluded that in ``crowded populations ...
-fitness is $K$'' \citep[pp.
-149]{macarthur_1967}.
-Note that only $K_{11}$ and $K_{22}$ are saturation densities; $K_{12}$ and $K_{21}$ measure the effects of competition between types.
-For instance, if $f_1(n_1,n_2)=r_1(1-\alpha_{11}n_1-\alpha_{12}n_2)n_1$, then  $K_{12}=\alpha_{12}$ is just the standard Lotka-Volterra competition coefficient.
-This led to the distinction between selection for greater saturation density ($K$-selection in the strict sense) and selection for greater competitive ability (``$\alpha$-selection'') at high density \citep{gill_1974,case_1974,joshi_2001}.
+Confusingly, only $K_{11}$ and $K_{22}$ are saturation densities; $K_{12}$ and $K_{21}$ measure the effects of competition between types.
+For instance, under Lotka-Volterra competition we have $f_1(n_1,n_2)=r_1(1-\alpha_{11}n_1-\alpha_{12}n_2)n_1$ and $K_{12}=1/\alpha_{12}$, where $\alpha_{12}$ is the per-capita competitive effect of the second type on the first.
+Thus, we can distinguish between selection for greater saturation density and selection for greater competitive ability (``$\alpha$-selection'') \citep{gill_1974,case_1974,joshi_2001}.
+Nevertheless, it seems hard not to agree that ``fitness is $K$'' when interpreted in the broad sense that selection at high density either favors the ability to keep growing at ever greater densities, or the ability to stop your competitors' growth at lower densities.
 
-Yet the $r$/$K$ scheme is strikingly incompatible with the  model of selection
+Yet  ``$K$-selection'' is incompatible with the most widely used model of selection.
+
+there is a deeper issue with the $r$/$K$ scheme as a description of selection in crowded versus uncrowded conditions ---
 

### 2017-10-28 manuscript.tex
latest

@@
 MacArthur famously distinguished between ``$r$-selection'' for faster growth at low population density, and ``$K$-selection'' for the ability to keep growing at high densities \citep{macarthur_1962,macarthur_1967}.
-While the $r$/$K$ scheme is primarily known for its seminal role in the study of life-history evolution \citep{pianka_1970,boyce_1984,charlesworth_1994,reznick_2002}, it
+While the $r$/$K$ scheme is primarily known in the life-history context \citep{pianka_1970,boyce_1984,charlesworth_1994,reznick_2002}, it
 also initiated a sizeable literature on density-dependent selection.
@@
 $f_1(K_{11},0)=0$, $f_1(0,K_{12})=0$, $f_2(0,K_{22})=0$, $f_2(K_{21},0)=0$).
-These intersection points summarize the nullcline structure provided that the nullclines are sufficiently linear, leading to the conclusion that in ``crowded populations ...
+These intersection points summarize the nullcline structure provided that the nullclines are sufficiently linear, and led MacArthur to conclude that in ``crowded populations ...
 fitness is $K$'' \citep[pp.
@@
 
-Confusingly, only $K_{11}$ and $K_{22}$ are saturation densities; $K_{12}$ and $K_{21}$ measure the effects of competition between types.
+Note that only $K_{11}$ and $K_{22}$ conform with the widespread logistic interpretation of ``$K$'' as representing saturation density; $K_{12}$ and $K_{21}$ measure the effects of competition between types.
 For instance, under Lotka-Volterra competition we have $f_1(n_1,n_2)=r_1(1-\alpha_{11}n_1-\alpha_{12}n_2)n_1$ and $K_{12}=1/\alpha_{12}$, where $\alpha_{12}$ is the per-capita competitive effect of the second type on the first.
-Thus, we can distinguish between selection for greater saturation density and selection for greater competitive ability (``$\alpha$-selection'') \citep{gill_1974,case_1974,joshi_2001}.
+To avoid this confusion we can distinguish selection for greater saturation density and selection for greater competitive ability (``$\alpha$-selection'') \citep{gill_1974,case_1974,joshi_2001}.
 Nevertheless, it seems hard not to agree that ``fitness is $K$'' when interpreted in the broad sense that selection at high density either favors the ability to keep growing at ever greater densities, or the ability to stop your competitors' growth at lower densities.

### 2017-10-30 manuscript.tex
latest

@@
 
-Verbal interpretations of the $r$/$K$ scheme, and the widespread use of the logistic model when discussing it, have tended to obscure the compelling generality of MacArthur's argument.
-We therefore  briefly review that argument now.
+The $r$/$K$ scheme is almost universally discussed in terms of the logistic model.
+This has obscured the remarkable generality of MacArthur's argument, which we thus now briefly review.
 Consider two types with densities $n_1$ and $n_2$ subject to density-dependent population regulation described by
@@
 where $f_1$ and $f_2$ are strictly decreasing functions of $n_1$ and $n_2$.
-By plotting the nullclines $f_1(n_1,n_2)=0$ and $f_2(n_1,n_2)=0$ in $(n_1,n_2)$ space, it becomes clear that for type $2$ to persist, its nullcline must enclose a region of $(n_1,n_2)$ space that type $1$'s nullcline does not.
-Since $f_1$ and $f_2$ are strictly decreasing, this corresponds to growth of type $2$ in a region of greater $n_1$ and/or $n_2$ than can be tolerated by type $1$.
+Plotting the nullclines $f_1(n_1,n_2)=0$ and $f_2(n_1,n_2)=0$ in $(n_1,n_2)$ space, it becomes clear that for a type to persist, its nullcline must enclose a region of $(n_1,n_2)$ space that the other type's nullcline does not.
+Since $f_1$ and $f_2$ are strictly decreasing, this corresponds to growth in a region of greater $n_1$ and/or $n_2$ than can be tolerated by the other type.
 The term ``$K$-selection'' comes from four ``carrying capacities'' defined as the points where the nullclines intersect the axes (i.e.
@@
 
-Note that only $K_{11}$ and $K_{22}$ conform with the widespread logistic interpretation of ``$K$'' as representing saturation density; $K_{12}$ and $K_{21}$ measure the effects of competition between types.
-For instance, under Lotka-Volterra competition we have $f_1(n_1,n_2)=r_1(1-\alpha_{11}n_1-\alpha_{12}n_2)n_1$ and $K_{12}=1/\alpha_{12}$, where $\alpha_{12}$ is the per-capita competitive effect of the second type on the first.
-To avoid this confusion we can distinguish selection for greater saturation density and selection for greater competitive ability (``$\alpha$-selection'') \citep{gill_1974,case_1974,joshi_2001}.
-Nevertheless, it seems hard not to agree that ``fitness is $K$'' when interpreted in the broad sense that selection at high density either favors the ability to keep growing at ever greater densities, or the ability to stop your competitors' growth at lower densities.
+Note that while $K$ is often interpreted as saturation density following the logistic model \citep{gill_1974}, MacArthur's ``$K$'' is broader: $K_{11}$ and $K_{22}$ are saturation densities while $K_{12}$ and $K_{21}$ represent the effects of competition between types.
+For instance, under Lotka-Volterra competition we have $f_1(n_1,n_2)=r_1(1-\alpha_{11}n_1-\alpha_{12}n_2)n_1$, where saturation density $\alpha_{11}=1/K_{11}$ measures intra-type competitive effects, and $\alpha_{12}=1/K_{12}$ measures inter-type competitive effects.
+Obviously intra- and inter-type competition need not be closely connected, but in the case of exploitation competition for consumable resources they are connected via resource use ``efficiency''  \citep{gill_1974,case_1974}.
+Nevertheless, given the generality of the preceding argument, it seems hard to disagree that ``fitness is $K$'' in the general sense that selection at high density either favors the ability to keep growing at ever higher densities, or the ability to stop the growth of competitors at lower densities.
 
-Yet  ``$K$-selection'' is incompatible with the most widely used model of selection.
+It is therefore remarkable that MacArthur's $K$-selection is incompatible with the standard representation of selection in evolutionary biology.
+This representation focuses on the change the proportion of different types (frequencies) over time,
 
-there is a deeper issue with the $r$/$K$ scheme as a description of selection in crowded versus uncrowded conditions ---
-
-In the standard representation of selection, genotypes are assigned selection coefficients describing how their relative proportions change over time (e.g. \cite[pp. 468]{barton_2007}).
+each type is assigned a selection coefficient describing the change in its relative proportion over time (e.g. \cite[pp. 468]{barton_2007}).
 

### 2017-11-01 manuscript.tex
latest

--- 2017-10-27 manuscript.tex
@@
 
-\textit{Keywords}: r/K selection, absolute fitness, eco-evo, Grime's triangle, competition-colonization trade-off, fluctuating selection, storage effect.
+\textit{Keywords}: r/K selection, absolute fitness, eco-evo, competition-colonization trade-off, fluctuating selection, storage effect.
 
@@
 
-MacArthur famously distinguished between ``$r$-selection'' for faster growth at low population density, and ``$K$-selection'' for the ability to keep growing at high densities \citep{macarthur_1962,macarthur_1967}.
-While the $r$/$K$ scheme is primarily known for its seminal role in the study of life-history evolution \citep{pianka_1970,boyce_1984,charlesworth_1994,reznick_2002}, it
-also initiated a sizeable literature on density-dependent selection.
-This literature is remarkable for containing some of earliest models of the interplay between ecology and evolution \citep{roughgarden_1979,travis_2013}.
+MacArthur's $r$/$K$ scheme \citep{macarthur_1962,macarthur_1967} is famous for its early role in life-history theory \citep{charlesworth_1994,reznick_2002}, and for initiating a sizeable literature on density-dependent selection \citep{christiansen_2004}, which contains some of earliest treatments of the interplay between ecology and evolution \citep{roughgarden_1979,travis_2013}.
+The central claim of the $r$/$K$ scheme is that in ``crowded populations ...
+fitness is $K$'' ($K$ will be defined below) \citep[pp.
+149]{macarthur_1967}.
+Remarkably, not only does MacArthur's argument for this claim seem to apply to essentially any population subject to density-regulated growth, the claim is also incompatible with the standard representation of selection in evolutionary biology.
+Since the general version of the ``fitness is $K$'' claim has been somewhat obscured by the widespread use of single-locus variants of the logistic model in the density-dependent selection literature, as well as well as the problems of the $r$/$K$ life-history literature \cite[pp.
+265]{charlesworth_1994}, we briefly review MacArthur's argument before proceeding.
 
-Verbal interpretations of the $r$/$K$ scheme, and the widespread use of the logistic model when discussing it, have tended to obscure the compelling generality of MacArthur's argument.
-We therefore  briefly review that argument now.
-Consider two types with densities $n_1$ and $n_2$ subject to density-dependent population regulation described by
+MacArthur considers two types subject to density-dependent population growth described by
 \begin{equation}
@@
 \end{equation}
-where $f_1$ and $f_2$ are strictly decreasing functions of $n_1$ and $n_2$.
-By plotting the nullclines $f_1(n_1,n_2)=0$ and $f_2(n_1,n_2)=0$ in $(n_1,n_2)$ space, it becomes clear that for type $2$ to persist, its nullcline must enclose a region of $(n_1,n_2)$ space that type $1$'s nullcline does not.
-Since $f_1$ and $f_2$ are strictly decreasing, this corresponds to growth of type $2$ in a region of greater $n_1$ and/or $n_2$ than can be tolerated by type $1$.
-The term ``$K$-selection'' comes from four ``carrying capacities'' defined as the points where the nullclines intersect the axes (i.e.
-$f_1(K_{11},0)=0$, $f_1(0,K_{12})=0$, $f_2(0,K_{22})=0$, $f_2(K_{21},0)=0$).
-These intersection points summarize the nullcline structure provided that the nullclines are sufficiently linear, leading to the conclusion that in ``crowded populations ...
-fitness is $K$'' \citep[pp.
-149]{macarthur_1967}
+where $f_1$ and $f_2$ are strictly decreasing functions of the population densities $n_1$ and $n_2$.
+Plotting the nullclines $f_1(n_1,n_2)=0$ and $f_2(n_1,n_2)=0$, it becomes clear that one type will be excluded unless it is able to grow at greater densities than the other type in some region of $(n_1,n_2)$ space (Fig.~\ref{fig:Ksel}).
+The term ``$K$-selection'' comes from four ``carrying capacities'' defined by $f_1(K_{11},0)=0$, $f_1(0,K_{12})=0$, $f_2(0,K_{22})=0$ and $f_2(K_{21},0)=0$.
+These $K$ values summarize the nullcline structure provided that the nullclines do not curve too much.
+Note that only $K_{11}$ and $K_{22}$ are saturation densities analogous to ``$K$'' in the logistic model; $K_{12}$ and $K_{21}$ represent the effects of competition between types.
+For instance, under Lotka-Volterra competition we have $f_1(n_1,n_2)=r_1(1-\alpha_{11}n_1-\alpha_{12}n_2)n_1$, where $\alpha_{11}=1/K_{11}$ measures intra-type competitive effects, and $\alpha_{12}=1/K_{12}$ measures inter-type competitive effects.
+In the case of exploitation competition for consumable resources, intra- and inter-type competition are connected via resource use ``efficiency'', but obviously this need not be true more generally \citep{gill_1974,case_1974}.
+Nevertheless, bearing in mind MacArthur's broad use of ``$K$'', it seems hard to disagree that ``fitness is $K$'' i.e.
+that selection in crowded populations either favors the ability to keep growing at ever higher densities, or the ability to stop the growth of competitors at lower densities.
 
-Confusingly, only $K_{11}$ and $K_{22}$ are saturation densities; $K_{12}$ and $K_{21}$ measure the effects of competition between types.
-For instance, under Lotka-Volterra competition we have $f_1(n_1,n_2)=r_1(1-\alpha_{11}n_1-\alpha_{12}n_2)n_1$ and $K_{12}=1/\alpha_{12}$, where $\alpha_{12}$ is the per-capita competitive effect of the second type on the first.
-Thus, we can distinguish between selection for greater saturation density and selection for greater competitive ability (``$\alpha$-selection'') \citep{gill_1974,case_1974,joshi_2001}.
-Nevertheless, it seems hard not to agree that ``fitness is $K$'' when interpreted in the broad sense that selection at high density either favors the ability to keep growing at ever greater densities, or the ability to stop your competitors' growth at lower densities.
+\begin{figure}
+\centering
+\includegraphics[scale=0.6]{Ksel.pdf}
+\caption{\label{fig:Ksel} MacArthur's argument for why ``fitness is $K$'' in crowded environments (adapted from \citep[pp.
+147]{macarthur_1967}).
+Arrows show the local direction of the dynamical trajectories of Eq.~\eqref{eq:macgeneral}.
+The nullclines where $f_1(n_1,n_2)=0$ and $f_2(n_1,n_2)=0$ need not be linear, but the ``carrying capacities'' $K$ are then less informative about the nullcline structure.
+Assuming these nullclines exist, exclusion is certain unless a type's nullcline extends out to greater densities in some region (here the outcome depends on initial condition).}
+\end{figure}
 
-Yet  ``$K$-selection'' is incompatible with the most widely used model of selection.
+By contrast, in the standard ``relative fitness'' representation of selection, each type is assigned a selection coefficient describing the change in its relative proportion over time (e.g.
+\cite[pp.
+468]{barton_2007}), with no consideration of population density.
+While this can describe selection in an uncrowded, expanding population, most commonly (e.g.
+in the Wright-Fisher model) it is assumed that total population density $N$ is fixed.
+In the latter case, the population is crowded in the sense that one type must displace another to grow.
+Selection happens along a line defined by $n_1+n_2=N$ and population density is otherwise irrelevant.
+Given this inconsistency with the $r$/$K$ scheme, the constant-$N$, shifting-frequency picture of selection does not seem compatible with the extremely broad variety of absolute population growth dynamics represented by Eq.~\eqref{eq:macgeneral}.
 
-there is a deeper issue with the $r$/$K$ scheme as a description of selection in crowded versus uncrowded conditions ---
+In fact, one step in MacArthur's argument considerably narrows its scope: assuming that the nullclines of $f_1$ and $f_2$ exist at all.
+The strength of this assumption is clear if we note that essence of the dichotomy underlying the $r$/$K$ scheme is between interaction-dependent selection and interaction-independent selection.
+That is, selective shifts in frequency are a result of differences in absolute growth rates, but these differences can arise in two logically distinct ways: 1) some types expand more rapidly the absence of interactions between individuals or 2) some types are superior in their interactions with other types.
+Population density is a key factor controlling whether individuals interact, thereby setting the relative contributions of these forms of selection.
+But there are obviously important forms of interaction that are not expressible in terms of slanted density isoclines.
+For instance, contests for territory and mates are predominantly relative.
+MacArthur's argument starts with assumptions about how selection depends on density, and then tries to make sense of what this says about the evolution of competitive traits such as ``efficiency'' of consumable resource consumption.
+The study of density-dependent selection should start from a description of the interactions between types, and then analyze how the effects of those interactions vary with density.
 
-In the standard representation of selection, genotypes are assigned selection coefficients describing how their relative proportions change over time (e.g. \cite[pp. 468]{barton_2007}).
+%This  also seems to be why the density-dependent selection literature expended such considerable effort to showing that evolution under density-dependent selection optimizes population size in some sense \citep{roughgarden_1979} --- a meaningless
 
-The true contrast hinted at by the $r$/$K$ scheme is between interaction-dependent, and interaction-independent selection.
-That is, selective shifts in frequency are a result of differences in growth rates between genotypes, and genotypic growth rates may differ in the absence of interactions between individuals, or as a result of interactions between individuals.
+Here we investigate the
 
-Population density is a key factor controling whether individuals interact, thereby setting the relative contributions of these forms of selection.
-Since a huge class of growth-affecting interactions are competitive, it seems clear that the study of density-dependent selection should be grounded a reasonable representation of competition and how its effects vary with density.
-
-%including considerable effort devoted to showing that evolution under density-dependent selection optimizes population size in some sense, akin to a maximization of fitness.
-
-in a simplified, territorial model of growth, dispersal and competition.
+territorial model of growth, dispersal and competition.
 

### 2017-11-03 Kplot.py
latest

@@ -1,0 +1,42 @@
+# -*- coding: utf-8 -*-
+"""
+Created on Thu Nov  2 13:39:18 2017
+
+@author: jbertram
+"""
+
+import numpy as np
+import matplotlib.pyplot as plt
+
+X, Y = np.mgrid[0:1:16j, 0:1:16j]
+U = 2*(1-X-1.5*Y)*X
+V = (1-Y-1.5*X)*Y
+speed = np.sqrt(U*U + V*V)
+
+fig1, ax1 = plt.subplots()
+ax1.set_aspect(1)
+ax1.quiver(X, Y, U/speed, V/speed,alpha=0.5)
+plt.xlim([0,1])
+plt.ylim([0,1])
+
+x=np.linspace(0,2,100)
+plt.plot(x,(1-x)/1.5,'k',linewidth=2)
+plt.plot(x,1-1.5*x,'k',linewidth=2)
+
+ax1.annotate(r'$f_2(n_1,n_2)=0$',xy=(0.15,0.8),xycoords='axes fraction',fontsize=16)
+ax1.annotate(r'$f_1(n_1,n_2)=0$',xy=(0.7,0.2),xycoords='axes fraction',fontsize=16)
+
+ax1.set_xticklabels([])
+ax1.set_yticklabels([])
+plt.xlabel(r"$n_1$",fontsize=16)
+plt.ylabel(r"$n_2$",fontsize=16)
+
+
+#fig1, ax1 = plt.subplots()
+#ax1.streamplot(X, Y, U, V,density=0.5)
+#plt.xlim([0,2])
+#plt.ylim([0,2])
+#
+#x=np.linspace(0,2,100)
+#plt.plot(x,1-0.5*x,'k',linewidth=2)
+#plt.plot(x,2-2*x,'k',linewidth=2)

### 2017-11-03 Kplot.py
latest

@@ -8,14 +8,16 @@ Created on Thu Nov  2 13:39:18 2017
 import numpy as np
 import matplotlib.pyplot as plt
 
-X, Y = np.mgrid[0:1:16j, 0:1:16j]
+Y, X = np.mgrid[0:1:1000j, 0:1:1000j]
 U = 2*(1-X-1.5*Y)*X
 V = (1-Y-1.5*X)*Y
 speed = np.sqrt(U*U + V*V)
 
 fig1, ax1 = plt.subplots()
 ax1.set_aspect(1)
-ax1.quiver(X, Y, U/speed, V/speed,alpha=0.5)
+#ax1.quiver(X, Y, U/speed, V/speed,alpha=0.5)
+seedpoints=np.array([[1,1]])
+ax1.streamplot(X, Y, U, V,color='k',start_points=)
 plt.xlim([0,1])
 plt.ylim([0,1])
 
@@ -23,13 +25,13 @@ x=np.linspace(0,2,100)
 plt.plot(x,(1-x)/1.5,'k',linewidth=2)
 plt.plot(x,1-1.5*x,'k',linewidth=2)
 
-ax1.annotate(r'$f_2(n_1,n_2)=0$',xy=(0.15,0.8),xycoords='axes fraction',fontsize=16)
-ax1.annotate(r'$f_1(n_1,n_2)=0$',xy=(0.7,0.2),xycoords='axes fraction',fontsize=16)
+ax1.annotate(r'$f_2(n_1,n_2)=0$',xy=(0.15,0.815),xycoords='axes fraction',fontsize=20)
+ax1.annotate(r'$f_1(n_1,n_2)=0$',xy=(0.6,0.27),xycoords='axes fraction',fontsize=20)
 
 ax1.set_xticklabels([])
 ax1.set_yticklabels([])
-plt.xlabel(r"$n_1$",fontsize=16)
-plt.ylabel(r"$n_2$",fontsize=16)
+plt.xlabel(r"$n_1$",fontsize=20)
+plt.ylabel(r"$n_2$",fontsize=20)
 
 
 #fig1, ax1 = plt.subplots()

### 2017-11-03 manuscript.tex
l

@@
 \centering
-\includegraphics[scale=0.6]{Ksel.pdf}
-\caption{\label{fig:Ksel} MacArthur's argument for why ``fitness is $K$'' in crowded environments (adapted from \citep[pp.
-147]{macarthur_1967}).
-Arrows show the local direction of the dynamical trajectories of Eq.~\eqref{eq:macgeneral}.
+\includegraphics[scale=0.8]{Kplot.pdf}
+\caption{\label{fig:Ksel} (a) MacArthur's dynamical argument for why ``fitness is $K$'' in crowded environments.
 The nullclines where $f_1(n_1,n_2)=0$ and $f_2(n_1,n_2)=0$ need not be linear, but the ``carrying capacities'' $K$ are then less informative about the nullcline structure.

### 2017-11-03 Kplot.py

@@ -8,30 +8,50 @@ Created on Thu Nov  2 13:39:18 2017
 import numpy as np
 import matplotlib.pyplot as plt
 
-Y, X = np.mgrid[0:1:1000j, 0:1:1000j]
+x,y=np.linspace(0,1,1000),np.linspace(0,1,1000)
+X, Y = np.meshgrid(x,y)
 U = 2*(1-X-1.5*Y)*X
 V = (1-Y-1.5*X)*Y
 speed = np.sqrt(U*U + V*V)
 
-fig1, ax1 = plt.subplots()
+fig1, (ax1, ax2) = plt.subplots(ncols=2,figsize=[8,4])
 ax1.set_aspect(1)
-#ax1.quiver(X, Y, U/speed, V/speed,alpha=0.5)
-seedpoints=np.array([[1,1]])
-ax1.streamplot(X, Y, U, V,color='k',start_points=)
-plt.xlim([0,1])
-plt.ylim([0,1])
+ax2.set_aspect(1)
+plt.tight_layout()
+seedpoints=np.concatenate([[[_,0.2-_] for _ in np.arange(0.02,0.2,0.02)],\
+            [[float(_)/10,1.-_*0.1] for _ in range(0,11) if _ not in [2,7]]])
+ax1.streamplot(x, y, U, V,density=10,start_points=seedpoints,linewidth=1.)
 
-x=np.linspace(0,2,100)
-plt.plot(x,(1-x)/1.5,'k',linewidth=2)
-plt.plot(x,1-1.5*x,'k',linewidth=2)
+z1=np.linspace(0,1,10)
+z2=1-z1
+ax1.plot(z1,z2/1.5,'k',linewidth=2)
+ax1.plot(z2/1.5,z1,'k',linewidth=2)
+ax1.set_xlim([0,1])
+ax1.set_ylim([0,1])
 
-ax1.annotate(r'$f_2(n_1,n_2)=0$',xy=(0.15,0.815),xycoords='axes fraction',fontsize=20)
-ax1.annotate(r'$f_1(n_1,n_2)=0$',xy=(0.6,0.27),xycoords='axes fraction',fontsize=20)
+ax1.annotate(r'$\frac{dn_2}{dt}=0$',xy=(0.15,0.81),xycoords='axes fraction',fontsize=16)
+ax1.annotate(r'$\frac{dn_1}{dt}=0$',xy=(0.62,0.28),xycoords='axes fraction',fontsize=16)
[…]

### 2017-11-08 manuscript.tex
l

@@
 
-Fitness is typically represented in heavily simplified terms in evolutionary genetics, often using constant selection coefficients
-
 This excludes fundamental ecological factors such as dynamic population size or density-dependence from the most genetically-realistic treatments of evolution, a problem that inspired MacArthur's influential but problematic $r$/$K$ theory.
@@
 
-MacArthur's $r$/$K$ scheme \citep{macarthur_1962,macarthur_1967} is famous for its early role in life-history theory \citep{charlesworth_1994,reznick_2002}, and for initiating a sizeable literature on density-dependent selection \citep{christiansen_2004}, which contains some of earliest treatments of the interplay between ecology and evolution \citep{roughgarden_1979,travis_2013}.
-The central claim of the $r$/$K$ scheme is that in ``crowded populations ...
-fitness is $K$'' ($K$ will be defined below) \citep[pp.
+A variety of different measures of fitness are used to quantify the effects of selection.
+Some widely used examples in evolutionary ecology are expected lifetime reproductive ratio $R_0$,  intrinsic growth rate $r$, saturation population density (often labeled ``$K$'') \citep{benton_2000}, and invasion fitness \citep{metz_1992}.
+In addition, ``relative fitness'' is the standard in much of evolutionary biology, particularly evolutionary genetics, where attention is generally restricted to relative genotypic proportions \cite[pp.
+468]{barton_2007}.
+
+In principle these fitness measures should all be grounded in population ecology, because ultimately selection operates through the environmentally-dependent differential reproduction and survival of individuals \citep{metcalf_2007}.
+Widely-used ``absolute'' measures of fitness like $r$ and $K$ are derived directly from population ecology.
+By contrast, relative fitness is almost entirely divorced from population ecology.
+This discrepancy runs deeper than a simple difference in emphasis on absolute densities rather than relative frequencies.
+
+In uncrowded populations there is no discrepancy.
+Relative fitness simply represents  differences in $r$ ($r$-selection) \citep[pp.
+26]{crow_1970}.
+The discrepancy emerges in crowded populations.
+Then the relative fitness description of selection assumes that $N$ is fixed, and fitness strictly involves the ability to increase in frequency at the expense of others (e.g.
+in the Wright-Fisher model).
+On the other hand, from quite general population ecological considerations, MacArthur famously argued that ``fitness is $K$'' in crowded populations ($K$-selection) \citep[pp.
 149]{macarthur_1967}.
-Remarkably, not only does MacArthur's argument for this claim seem to apply to essentially any population subject to density-regulated growth, the claim is also incompatible with the standard representation of selection in evolutionary biology.
-Since the general version of the ``fitness is $K$'' claim has been somewhat obscured by the widespread use of single-locus variants of the logistic model in the density-dependent selection literature, as well as well as the problems of the $r$/$K$ life-history literature \cite[pp.
-265]{charlesworth_1994}, we briefly review MacArthur's argument before proceeding.
+To appreciate the import of MacArthur's argument, and see the nature of the discrepancy between these views of selection, we now briefly review MacArthur's argument.
 
@@
 where $f_1$ and $f_2$ are strictly decreasing functions of the population densities $n_1$ and $n_2$.
-Plotting the nullclines $f_1(n_1,n_2)=0$ and $f_2(n_1,n_2)=0$, it becomes clear that one type will be excluded unless it is able to grow at greater densities than the other type in some region of $(n_1,n_2)$ space (Fig.~\ref{fig:Ksel}).
-The term ``$K$-selection'' comes from four ``carrying capacities'' defined by $f_1(K_{11},0)=0$, $f_1(0,K_{12})=0$, $f_2(0,K_{22})=0$ and $f_2(K_{21},0)=0$.
-These $K$ values summarize the nullcline structure provided that the nullclines do not curve too much.
+Plotting the nullclines $f_1(n_1,n_2)=0$ and $f_2(n_1,n_2)=0$, it can be seen that one type will be excluded unless it is able to grow at higher densities than the other type in some region of $(n_1,n_2)$ space (Fig.~\ref{fig:Ksel}a).
+MacArthur defines four ``carrying capacities'' generally labeled $K$, as the intersection points of the nullclines with the $n_1$ and $n_2$ axes (i.e.
+$f_1(K_{11},0)=0$, $f_1(0,K_{12})=0$, $f_2(0,K_{22})=0$ and $f_2(K_{21},0)=0$).
+Assuming that the nullclines are linear, these $K$ values determine whether a region of higher-density growth exists for each type.
+
 Note that only $K_{11}$ and $K_{22}$ are saturation densities analogous to ``$K$'' in the logistic model; $K_{12}$ and $K_{21}$ represent the effects of competition between types.
-For instance, under Lotka-Volterra competition we have $f_1(n_1,n_2)=r_1(1-\alpha_{11}n_1-\alpha_{12}n_2)n_1$, where $\alpha_{11}=1/K_{11}$ measures intra-type competitive effects, and $\alpha_{12}=1/K_{12}$ measures inter-type competitive effects.
-In the case of exploitation competition for consumable resources, intra- and inter-type competition are connected via resource use ``efficiency'', but obviously this need not be true more generally \citep{gill_1974,case_1974}.
-Nevertheless, bearing in mind MacArthur's broad use of ``$K$'', it seems hard to disagree that ``fitness is $K$'' i.e.
-that selection in crowded populations either favors the ability to keep growing at ever higher densities, or the ability to stop the growth of competitors at lower densities.
+For instance, under Lotka-Volterra competition we have $f_1(n_1,n_2)=r_1(1-\alpha_{11}n_1-\alpha_{12}n_2)n_1$.
+Then the nullclines are linear and $\alpha_{11}=1/K_{11}$ measures intra-type competitive effects, while $\alpha_{12}=1/K_{12}$ measures inter-type competitive effects.
+Thus, for MacArthur ``fitness is $K$'' in the sense that selection in crowded populations either favors the ability to keep growing at ever higher densities, or the ability to stop the growth of competitors at lower densities.
+This general idea applies even if the nullclines are so curved that the ``$K$'' values themselves do not give much information about the regions of high-density growth.
 
@@
 \includegraphics[scale=0.8]{Kplot.pdf}
-\caption{\label{fig:Ksel} (a) MacArthur's dynamical argument for why ``fitness is $K$'' in crowded environments.
-The nullclines where $f_1(n_1,n_2)=0$ and $f_2(n_1,n_2)=0$ need not be linear, but the ``carrying capacities'' $K$ are then less informative about the nullcline structure.
-Assuming these nullclines exist, exclusion is certain unless a type's nullcline extends out to greater densities in some region (here the outcome depends on initial condition).}
+\caption{\label{fig:Ksel} (a) MacArthur's dynamical argument for why ``fitness is $K$'' in crowded environments, using a Lotka-Volterra form for Eq.
+\eqref{eq:macgeneral}.
+Assuming the nullclines $f_1(n_1,n_2)=0$ and $f_2(n_1,n_2)=0$ exist, exclusion is certain unless a type's nullcline extends out to greater densities in some region.
+In this example, the outcome depends on initial condition.
+(b) The constant-$N$, relative fitness description of selection.}
 \end{figure}
 
-By contrast, in the standard ``relative fitness'' representation of selection, each type is assigned a selection coefficient describing the change in its relative proportion over time (e.g.
-\cite[pp.
-468]{barton_2007}), with no consideration of population density.
-While this can describe selection in an uncrowded, expanding population, most commonly (e.g.
-in the Wright-Fisher model) it is assumed that total population density $N$ is fixed.
-In the latter case, the population is crowded in the sense that one type must displace another to grow.
-Selection happens along a line defined by $n_1+n_2=N$ and population density is otherwise irrelevant.
-Given this inconsistency with the $r$/$K$ scheme, the constant-$N$, shifting-frequency picture of selection does not seem compatible with the extremely broad variety of absolute population growth dynamics represented by Eq.~\eqref{eq:macgeneral}.
+Clearly MacArthur's claim is not compatible with the constant-$N$, relative fitness description of selection, in which $n_1$ and $n_2$ change along a line defined by $n_1+n_2=N$, and population density is otherwise irrelevant (Fig.~\ref{fig:Ksel}b).
+In this case, all four $K$ values are equal to $N$ [not!], and the $f_1$ and $f_2$ nullclines do not even exist.
+Since $N$ is fixed, it is not possible to persist at higher densities than $N$, nor is it possible to suppress the density of a competitor at densities lower than $N$
+
+Given the extremely broad variety of absolute population growth dynamics represented by Eq.~\eqref{eq:macgeneral}, at this point we might start to doubt that the relative fitness, constant=$N$ description of selection has a plausible basis in population ecology.
+In particular, the uncoupling of selection and population demography could be problematic, because they are both driven by the same birth/death events.
 
@@
 %This  also seems to be why the density-dependent selection literature expended such considerable effort to showing that evolution under density-dependent selection optimizes population size in some sense \citep{roughgarden_1979} --- a meaningless
+
+%In the case of exploitation competition for consumable resources, intra- and inter-type competition are connected via resource use ``efficiency'', but obviously this need not be true more generally \citep{gill_1974,case_1974}.
 

### 2017-11-08 manuscript.tex
l

@@
 where $f_1$ and $f_2$ are strictly decreasing functions of the population densities $n_1$ and $n_2$.
-Plotting the nullclines $f_1(n_1,n_2)=0$ and $f_2(n_1,n_2)=0$, it can be seen that one type will be excluded unless it is able to grow at higher densities than the other type in some region of $(n_1,n_2)$ space (Fig.~\ref{fig:Ksel}a).
-MacArthur defines four ``carrying capacities'' generally labeled $K$, as the intersection points of the nullclines with the $n_1$ and $n_2$ axes (i.e.
+Plotting the nullclines $f_1(n_1,n_2)=0$ and $f_2(n_1,n_2)=0$, it can be seen that a type will be excluded if its nullcline is completely contained in the region bounded by the other type's nullcline.
+For a type to have the possibility of persisting, it must be able to grow in some region of $(n_1,n_2)$ space at higher total population density $N=n_1+n_2$ than the other type (Fig.~\ref{fig:Ksel}a).
+
+MacArthur defines four ``carrying capacities'' as the intersection points of the nullclines with the $n_1$ and $n_2$ axes (i.e.
 $f_1(K_{11},0)=0$, $f_1(0,K_{12})=0$, $f_2(0,K_{22})=0$ and $f_2(K_{21},0)=0$).
-Assuming that the nullclines are linear, these $K$ values determine whether a region of higher-density growth exists for each type.
-
+Assuming that the nullclines are close to linear, these $K$ values determine whether a region of higher-density growth exists for each type.
 Note that only $K_{11}$ and $K_{22}$ are saturation densities analogous to ``$K$'' in the logistic model; $K_{12}$ and $K_{21}$ represent the effects of competition between types.
@@
 Then the nullclines are linear and $\alpha_{11}=1/K_{11}$ measures intra-type competitive effects, while $\alpha_{12}=1/K_{12}$ measures inter-type competitive effects.
-Thus, for MacArthur ``fitness is $K$'' in the sense that selection in crowded populations either favors the ability to keep growing at ever higher densities, or the ability to stop the growth of competitors at lower densities.
-This general idea applies even if the nullclines are so curved that the ``$K$'' values themselves do not give much information about the regions of high-density growth.
+Thus, to summarize, ``fitness is $K$'' in the sense that selection in crowded populations either favors the ability to keep growing at ever higher densities (moving a type's own nullcline outwards), or the ability to suppress the growth of competitors at lower densities (moving the nullcline of competitors inwards).
+This general idea applies even if the nullclines are nonlinear to such an extent that the ``$K$'' values themselves do not give much information about the regions of high-density growth.
 
@@
 
-Clearly MacArthur's claim is not compatible with the constant-$N$, relative fitness description of selection, in which $n_1$ and $n_2$ change along a line defined by $n_1+n_2=N$, and population density is otherwise irrelevant (Fig.~\ref{fig:Ksel}b).
+Clearly MacArthur's argument is not compatible with the constant-$N$, relative fitness description of selection, in which $n_1$ and $n_2$ change along a line defined by $n_1+n_2=N$, and population density is otherwise irrelevant (Fig.~\ref{fig:Ksel}b).
+Since $N$ is fixed, it is not possible to persist at higher densities than $N$, nor is it possible to suppress the density of a competitor at densities lower than $N$
+
 In this case, all four $K$ values are equal to $N$ [not!], and the $f_1$ and $f_2$ nullclines do not even exist.
-Since $N$ is fixed, it is not possible to persist at higher densities than $N$, nor is it possible to suppress the density of a competitor at densities lower than $N$
 

### 2017-11-08 Kplot.py
l

@@ -12,7 +12,7 @@ x,y=np.linspace(0,1,1000),np.linspace(0,1,1000)
 X, Y = np.meshgrid(x,y)
 U = 2*(1-X-1.5*Y)*X
 V = (1-Y-1.5*X)*Y
-speed = np.sqrt(U*U + V*V)
+#speed = np.sqrt(U*U + V*V)
 
 fig1, (ax1, ax2) = plt.subplots(ncols=2,figsize=[8,4])
 ax1.set_aspect(1)

### 2017-11-08 simulations.py

@@ -64,7 +64,7 @@ def deltnplus(m,c,U):
         return m*(np.exp(-L)+(R(m,c,U)+A(m,c,U))*c/cbar)
     else:
         return np.zeros(len(m))
-    
+          
 def deltnplusclassic(m,c,U):
     if sum(m)>0:
         return U*m*c/sum(m*c)
@@ -159,6 +159,62 @@ plt.gca().annotate(r'$(d)$',xy=(0.9,0.9),xycoords='axes fraction',fontsize=12)
 plt.legend(loc='upper center',prop={'size':10})
 
 plt.savefig('/home/jbertram/repos/densitydependentlottery/simulationcomparison.pdf')
+
+#(n1,n2) vector field
+#=================================================================
+T=100000
+x,y=np.linspace(0.0,1.,201),np.linspace(0.,1.,201)
+xgrid, ygrid=np.meshgrid(x,y)
+X,Y=np.zeros([len(x),len(y)]),np.zeros([len(x),len(y)])
+b=np.array([3.,2.5])
+c=np.array([1.,5.])
+d=np.array([0.2,0.2])
+
+for i in xrange(len(x)):
+    for j in xrange(len(y)):
+        n=T*np.array([x[i],y[j]])
+        U=T-sum(n)
+        X[j,i],Y[j,i]=deltnplus(b*n*U/T,c,U)-d*n
+
+fig1, ax1 = plt.subplots()#(ncols=2,figsize=[8,4])
+ax1.set_aspect(1)
+#ax2.set_aspect(1)
+plt.tight_layout()
+#seedpoints=np.concatenate([[[_,0.2-_] for _ in np.arange(0.02,0.2,0.02)],\
+#            [[float(_)/10,1.-_*0.1] for _ in range(0,11) if _ not in [2,7]]])
+ax1.streamplot(x, y, X, Y,start_points=[[0.01,0.01]],linewidth=1.)
+#ax1.streamplot(x, y, X, Y,density=2,linewidth=1.)
+ax1.set_xlim([0,1])
+ax1.set_ylim([0,1])
+
+z1=np.linspace(0,1,10)
+z2=1-z1
+ax1.plot(z1,z2/1.5,'k',linewidth=2)
+ax1.plot(z2/1.5,z1,'k',linewidth=2)
+
+
[…]

### 2017-11-08 manuscript.tex
l

@@
 
-A variety of different measures of fitness are used to quantify the effects of selection.
+There are a variety of different measures of fitness.
 Some widely used examples in evolutionary ecology are expected lifetime reproductive ratio $R_0$, intrinsic growth rate $r$, saturation population density (often labeled ``$K$'') \citep{benton_2000}, and invasion fitness \citep{metz_1992}.
@@
 
-In principle these fitness measures should all be grounded in population ecology, because ultimately selection operates through the environmentally-dependent differential reproduction and survival of individuals \citep{metcalf_2007}.
-Widely-used ``absolute'' measures of fitness like $r$ and $K$ are derived directly from population ecology.
-By contrast, relative fitness is almost entirely divorced from population ecology.
+Despite this variety, fitness measures are different ways to environmentally-dependent quantify the differential reproduction and survival of individuals \citep{metcalf_2007}.
+
+relative fitness is almost entirely divorced from population ecology.
+
+In uncrowded populations, there is no discrepancy. Relative fitness simply represents  differences in $r$ ($r$-selection) \citep[pp. 26]{crow_1970}.
+
+Absolute measures of fitness like $r$ and $K$ are derived directly from population ecology. By contrast,
+
 This discrepancy runs deeper than a simple difference in emphasis on absolute densities rather than relative frequencies.
 
-In uncrowded populations there is no discrepancy.
-Relative fitness simply represents  differences in $r$ ($r$-selection) \citep[pp.
-26]{crow_1970}.
  The discrepancy emerges in crowded populations.

### 2017-11-09 simulations.py
l

@@ -163,12 +163,12 @@ plt.savefig('/home/jbertram/repos/densitydependentlottery/simulationcomparison.p
 #(n1,n2) vector field
 #=================================================================
 T=100000
-x,y=np.linspace(0.0,1.,201),np.linspace(0.,1.,201)
+x,y=np.linspace(0.0,1.,401),np.linspace(0.,1.,401)
 xgrid, ygrid=np.meshgrid(x,y)
 X,Y=np.zeros([len(x),len(y)]),np.zeros([len(x),len(y)])
-b=np.array([3.,2.5])
-c=np.array([1.,5.])
-d=np.array([0.2,0.2])
+b=np.array([1.,1.])
+c=np.array([1.,1.5])
+d=np.array([0.1,0.1])
 
 for i in xrange(len(x)):
     for j in xrange(len(y)):
@@ -182,15 +182,14 @@ ax1.set_aspect(1)
 plt.tight_layout()
 #seedpoints=np.concatenate([[[_,0.2-_] for _ in np.arange(0.02,0.2,0.02)],\
 #            [[float(_)/10,1.-_*0.1] for _ in range(0,11) if _ not in [2,7]]])
-ax1.streamplot(x, y, X, Y,start_points=[[0.01,0.01]],linewidth=1.)
-#ax1.streamplot(x, y, X, Y,density=2,linewidth=1.)
+seedpoints=np.array([[0.5*x[_],0.5*y[200-_]] for _ in range(0,201,10)])
+#ax1.streamplot(x, y, X, Y,start_points=seedpoints,linewidth=1.,density=10)
+ax1.streamplot(x, y, X, Y,start_points=[[ 0.025,  0.9]],linewidth=1.,density=100)
 ax1.set_xlim([0,1])
 ax1.set_ylim([0,1])
 
 z1=np.linspace(0,1,10)
-z2=1-z1
-ax1.plot(z1,z2/1.5,'k',linewidth=2)
-ax1.plot(z2/1.5,z1,'k',linewidth=2)
+
 
 
 ax1.annotate(r'$\frac{dn_2}{dt}=0$',xy=(0.15,0.81),xycoords='axes fraction',fontsize=16)

### 2017-11-09 simulations.py
l

@@ -166,26 +166,35 @@ T=100000
 x,y=np.linspace(0.0,1.,201),np.linspace(0.,1.,201)
 xgrid, ygrid=np.meshgrid(x,y)
 X,Y=np.zeros([len(x),len(y)]),np.zeros([len(x),len(y)])
-b=np.array([3.,2.5])
-c=np.array([1.,5.])
-d=np.array([0.2,0.2])
+b=np.array([10.,10.])
+c=np.array([1.,2.])
+d=np.array([0.1,0.1])
 
 for i in xrange(len(x)):
     for j in xrange(len(y)):
         n=T*np.array([x[i],y[j]])
         U=T-sum(n)
-        X[j,i],Y[j,i]=deltnplus(b*n*U/T,c,U)-d*n
+        X[j,i],Y[j,i]=deltnplus(b*n*U/float(T),c,U)-d*n
 
 fig1, ax1 = plt.subplots()#(ncols=2,figsize=[8,4])
 ax1.set_aspect(1)
 #ax2.set_aspect(1)
 plt.tight_layout()
-#seedpoints=np.concatenate([[[_,0.2-_] for _ in np.arange(0.02,0.2,0.02)],\
-#            [[float(_)/10,1.-_*0.1] for _ in range(0,11) if _ not in [2,7]]])
-ax1.streamplot(x, y, X, Y,start_points=[[0.01,0.01]],linewidth=1.)
-#ax1.streamplot(x, y, X, Y,density=2,linewidth=1.)
+seedpoints=[[x[_]/2,y[200-_]/2] for _ in range(0,201,10)]
+ax1.streamplot(x, y, X, Y,start_points=seedpoints,linewidth=1.,density=20)
 ax1.set_xlim([0,1])
 ax1.set_ylim([0,1])
+
+
+
+f1null=np.array([[x[[_ for _ in range(len(x)) if X[i,_]>0][-1]],y[i]] for i in range(1,len(y)) if max(X[i])>0])
+f2null=np.array([[x[[_ for _ in range(len(x)) if Y[i,_]>0][-1]],y[i]] for i in range(1,len(y)) if max(Y[i])>0])
+plt.plot(f1null[:,0],f1null[:,1],'k',linewidth=2.)
+plt.plot(f2null[:,0],f2null[:,1],'b',linewidth=2.)
+
+plt.plot([f2null[0,1],f2null[-1,1]],[f2null[-1,1],f2null[0,1]])
+
+
 
 z1=np.linspace(0,1,10)
 z2=1-z1

### 2017-11-09 manuscript.tex

@@
 
-Despite this variety, fitness measures are different ways to environmentally-dependent quantify the differential reproduction and survival of individuals \citep{metcalf_2007}.
-
-relative fitness is almost entirely divorced from population ecology.
+Despite this variety, in principle all fitness measures are ultimately ways to quantify an individual's potential for reproduction and survival \citep{metcalf_2007}.
+Yet the shifting-frequency relative view of fitness stands out as particularly divorced from population ecology.
 

### 2017-11-09 simulations.py
simulations conflict

@@ -1,0 +1,465 @@
+# -*- coding: utf-8 -*-
+"""
+Created on Thu Mar 24 12:01:54 2016
+
+@author: jbertram
+"""
+
+import numpy as np
+import matplotlib.pyplot as plt
+import bisect
+
+def deltnplussim(m,c,U):
+    scatter=[[0 for x in range(len(m))] for y in range(U)];
+    for i in range(len(m)):
+        for y in xrange(m[i]):
+            cell=int(U*np.random.rand());
+            scatter[cell][i]=scatter[cell][i]+1;
+    
+    winsnocomp=np.zeros(len(m)); wins1=np.zeros(len(m)); wins2=np.zeros(len(m));
+    comp=np.zeros(U);
+    for i in range(U):
+        comp[i]=sum(scatter[i])
+        if comp[i]>0:            
+            lotterycmf=np.cumsum(np.array(scatter[i])*c)
+            victor=bisect.bisect(lotterycmf,np.random.rand()*lotterycmf[-1])
+            
+            if scatter[i][victor]==1 and comp[i]==1:
+                winsnocomp[victor]=winsnocomp[victor]+1
+            elif scatter[i][victor]==1:
+                wins1[victor]=wins1[victor]+1
+            else: 
+                wins2[victor]=wins2[victor]+1
+        
+    return np.array([winsnocomp,wins1,wins2])
+
+def R(m,c,U):
+    l=m/float(U)
+    L=sum(l)
+    cbar=sum(m*c)/sum(m); 
+    out=cbar*np.exp(-l)*(1-np.exp(-(L-l)))\
+            /(c + (L-1+np.exp(-L))/(1-(1+L)*np.exp(-L))*(cbar*L-c*l)/(L-l))
+    
+    for i in range(len(out)):
+        if np.isnan(out)[i]: out[i]=0
+            
+    return out
+
+def A(m,c,U):
+    l=m/float(U)
+    L=sum(l)
+    cbar=sum(m*c)/sum(m)
+    out=(1-np.exp(-l))\
+            /((1-np.exp(-l))/(1-(1+l)*np.exp(-l))*c*l\
[…]

### 2017-11-10 Kplot.py
latest

@@ -10,8 +10,9 @@ import matplotlib.pyplot as plt
 
 x,y=np.linspace(0,1,1000),np.linspace(0,1,1000)
 X, Y = np.meshgrid(x,y)
-U = 2*(1-X-1.5*Y)*X
-V = (1-Y-1.5*X)*Y
+alphas=np.array([[1,1.5],[1.5,1]])
+U = 2*(1-alphas[0,0]*X-alphas[0,1]*Y)*X
+V = (1-alphas[1,1]*Y-alphas[1,0]*X)*Y
 #speed = np.sqrt(U*U + V*V)
 
 fig1, (ax1, ax2) = plt.subplots(ncols=2,figsize=[8,4])
@@ -20,12 +21,13 @@ ax2.set_aspect(1)
 plt.tight_layout()
 seedpoints=np.concatenate([[[_,0.2-_] for _ in np.arange(0.02,0.2,0.02)],\
             [[float(_)/10,1.-_*0.1] for _ in range(0,11) if _ not in [2,7]]])
-ax1.streamplot(x, y, U, V,density=10,start_points=seedpoints,linewidth=1.)
+ax1.streamplot(x, y, U, V,density=100,start_points=seedpoints,linewidth=1.)
 
 z1=np.linspace(0,1,10)
 z2=1-z1
-ax1.plot(z1,z2/1.5,'k',linewidth=2)
-ax1.plot(z2/1.5,z1,'k',linewidth=2)
+ax1.plot(z1,(1-alphas[0,0]*z1)/alphas[0,1],'k',linewidth=2)
+ax1.plot(z1,(1-alphas[1,0]*z1)/alphas[1,1],'k',linewidth=2)
+#ax1.plot([0,1/alphas[0,0]],[1/alphas[1,1],0],'y',linewidth=2)
 ax1.set_xlim([0,1])
 ax1.set_ylim([0,1])
 
@@ -52,6 +54,7 @@ ax2.set_xlabel(r"$n_1$",fontsize=20)
 ax2.set_ylabel(r"$n_2$",fontsize=20)
 
 plt.savefig('/home/jbertram/repos/densitydependentlottery/Kplot.pdf',bbox="tight")
+
 
 
 #fig1, ax1 = plt.subplots()

### 2017-11-10 simulations.py

@@ -49,9 +49,9 @@ def A(m,c,U):
     l=m/float(U)
     L=sum(l)
     cbar=sum(m*c)/sum(m)
-    out=(1-np.exp(-l))\
+    out=cbar*(1-np.exp(-l))\
             /((1-np.exp(-l))/(1-(1+l)*np.exp(-l))*c*l\
-            +(L*(1-np.exp(-L))/(1-(1+L)*np.exp(-L))-l*(1-np.exp(-l))/(1-(1+l)*np.exp(-l)))/(L-l)*(sum(c*l)-c*l))
+            +(L*(1-np.exp(-L))/(1-(1+L)*np.exp(-L))-l*(1-np.exp(-l))/(1-(1+l)*np.exp(-l)))/(L-l)*(cbar*L-c*l))
     for i in range(len(out)):
         if np.isnan(out)[i]: out[i]=0
             
@@ -93,22 +93,22 @@ nocomp=np.array([np.exp(-M/U) for U in Us]);
 comp1=np.array([R(m,c,U)[0]*c[0]/cbar for U in Us]);
 comp2=np.array([A(m,c,U)[0]*c[0]/cbar for U in Us]);
 
-subplot(221)
+plt.subplot(221)
 plt.plot(float(m[0])/Us,(exact[:,0,0]+exact[:,1,0]+exact[:,2,0])/m[0],'k.',markersize=1.,label="Simulation")
 plt.plot(float(m[0])/Us,(nocomp+comp1+comp2),'k',label=r"$\Delta_+ n_1/m_1$")
 plt.plot(float(m[0])/Us,totalclassic[:,0]/m[0],'k--',label=r"Classic lottery")
-plt.plot(float(m[0])/Us,totalclassicboundedm[:,0]/m[0],'kx',label=r"Classic lottery")
+#plt.plot(float(m[0])/Us,totalclassicboundedm[:,0]/m[0],'kx',label=r"Classic lottery")
 plt.xlim([0,maxl])
 plt.ylim([0,1.1])
-gca().xaxis.set_label_coords(0.5, -0.09)
-gca().set_xticklabels(['0','','1','','2','','3'])
-xlabel(r"$l_1$",fontsize=14)
-ylabel(r"$\Delta_+ n_1/m_1$",fontsize=14)
+plt.gca().xaxis.set_label_coords(0.5, -0.09)
+plt.gca().set_xticklabels(['0','','1','','2','','3'])
+plt.xlabel(r"$l_1$",fontsize=14)
[…]

### 2017-11-11 manuscript.tex
latest
l

--- 2017-11-08 manuscript.tex
@@
 
-\title{A lottery model of density-dependent selection}
+\title{Density-dependent selection with strictly relative contests}
 %\author{Jason Bertram $^{1,\ast}$ \\ 
@@
 468]{barton_2007}.
+This variety is not problematic by itself because different fitness measures serve different purposes.
+But ultimately any measure of fitness should be grounded in the processes of birth and death which govern population ecology and demography \citep{metcalf_2007}.
+While this connection is apparent for absolute fitness measures like $r$ and $K$, relative fitness seems largely divorced from population ecology.
 
-Despite this variety, fitness measures are different ways to environmentally-dependent quantify the differential reproduction and survival of individuals \citep{metcalf_2007}.
+In uncrowded populations, relative fitness simply represents differences in the intrinsic exponential growth rate $r$ \citep[pp.
+26]{crow_1970}, with selection favoring greater $r$ ($r$-selection).
+In crowded populations, relative fitness models almost universally assume that total population size $N$ is fixed, or has some externally imposed time course.
+This is exemplified by the Wright-Fisher model, in which time advances in discrete generations which do not overlap, and fitness can be interpreted as a product of fertility and juvenile viability \citep[pp.
+185]{crow_1970}.
+The limitations of these relative fitness models are openly acknowledged, usually with an emphasis on the difficulty of incorporating more of the life cycle ``components'' of fitness (e.g.
+\cite[pp.
+276]{ewens_2004}).
+But more fundamentally, the constant-$N$, relative fitness description uncouples selection and demography, at odds with the fact that the same birth/death events drive both.
 
-relative fitness is almost entirely divorced from population ecology.
+The issue can be expressed more formally by revisiting MacArthur's famous analysis of selection in crowded populations \citep{macarthur_1967}.
+MacArthur considers a population consisting of two types with densities $n_1$ and $n_2$ subject to density-dependent population growth described by general differential equations
+\begin{equation}
+\frac{d n_1}{d t}=f_1(n_1,n_2)\qquad\frac{d n_2}{d t}=f_2(n_1,n_2). \label{eq:macgeneral}
+\end{equation}
+Since no population has unlimited resources, the functions $f_1$ and $f_2$ must decline to zero if either $n_1$ or $n_2$ are made sufficiently large.
+The outcome of selection can then be determined by the shape of the nullclines $f_1(n_1,n_2)=0$ and $f_2(n_1,n_2)=0$.
+Specifically, a type will be excluded if its nullcline is completely contained in the region bounded by the other type's nullcline.
+For a type to have the possibility of persisting, it must be able to keep growing to higher densities than the other type can tolerate in some region of $(n_1,n_2)$ space (Fig.~\ref{fig:Ksel}a).
 
-In uncrowded populations, there is no discrepancy. Relative fitness simply represents  differences in $r$ ($r$-selection) \citep[pp. 26]{crow_1970}.
+MacArthur's argument is clearly incompatible with a narrow interpretation of relative fitness models as selection along lines defined by $n_1+n_2=N$ for each $N$ (Fig.~\ref{fig:Ksel}b).
+In this case the $f_1$ and $f_2$ nullclines are not even defined.
+Moreover, at face value MacArthur's, conclusion that ``fitness is $K$'' in crowded populations \citep[pp.
+149]{macarthur_1967} seems to be incompatible with constant $N$.
+In particular, the logistic model and its bi-allelic diploid variants are widely used to analyze MacArthur's ``$K$-selection'' in which case positive selection is necessarily tied to increases in $N$.
+More sophisticated models using the Lotka-Volterra competition equations, or explicit resource consumption, also intrinsically tie high-density selection to changes in population density.
+This raises the possibility that the ubiquitous constant-$N$, relative fitness description of selection could be incompatible with a huge class of population ecological processes occurring in nature and the lab.
 
-Absolute measures of fitness like $r$ and $K$ are derived directly from population ecology. By contrast,
+Here we introduce a novel density-dependent model of population growth based on territorial contests, and show that when this model reaches a demographic steady-state, the constant-$N$, relative fitness picture emerges.
+We show that our model is entirely consistent with the claim that ``fitness is $K$'' in crowded populations when MacArthur's very liberal use of ``$K$'' is taken into account.
+In particular, MacArthur's argument does not justify the widespread intuition that selection in crowded environments  attempts to maximize population size is a consequence of
 
-This discrepancy runs deeper than a simple difference in emphasis on absolute densities rather than relative frequencies.
+argument is widely interpreted to mean that positive selection in crowded populations involves the ability to keep growing at ever higher population densities [citations]
 
- The discrepancy emerges in crowded populations.
-Then the relative fitness description of selection assumes that $N$ is fixed, and fitness strictly involves the ability to increase in frequency at the expense of others (e.g.
-in the Wright-Fisher model).
-On the other hand, from quite general population ecological considerations, MacArthur famously argued that ``fitness is $K$'' in crowded populations ($K$-selection) \citep[pp.
-149]{macarthur_1967}.
-To appreciate the import of MacArthur's argument, and see the nature of the discrepancy between these views of selection, we now briefly review MacArthur's argument.
+Moreover, it is not compatible with most verbal interpretations of MacArthur's argument,
 
-MacArthur considers two types subject to density-dependent population growth described by
-\begin{equation}
-\frac{d n_1}{d t}=f_1(n_1,n_2)\qquad\frac{d n_2}{d t}=f_2(n_1,n_2), \label{eq:macgeneral}
-\end{equation}
-where $f_1$ and $f_2$ are strictly decreasing functions of the population densities $n_1$ and $n_2$.
-Plotting the nullclines $f_1(n_1,n_2)=0$ and $f_2(n_1,n_2)=0$, it can be seen that a type will be excluded if its nullcline is completely contained in the region bounded by the other type's nullcline.
-For a type to have the possibility of persisting, it must be able to grow in some region of $(n_1,n_2)$ space at higher total population density $N=n_1+n_2$ than the other type (Fig.~\ref{fig:Ksel}a).
+\citep{anderson_1971}
+
+ with the assumption of constant $N$, although as we will discuss below, it is a narrow interpretation of MacArthur's argument.
+
+We revisit the classic lottery model of \cite{chesson_1981}, which has two features that make it well suited for this role, but one critical flaw that we rectify here.
+
+The first feature is that the lottery representation of competition is  particularly concise.
+Mature individuals (``adults'') each require their own territory, whereas newborn individuals (``propagules'') disperse to, and subsequently compete for, territories made available by the death of adults.
+Territorial contest among propagules leaves a single victorious adult per territory, the victor chosen at random from the propagules present, with probabilities weighted by a coefficient for each type representing competitive ability, akin to a lottery \citep{sale_77}.
+
+The second feature is the close connection between the lottery model and one of the foundational models of population genetics, the Wright-Fisher model of genetic drift, which we discuss further below.
+
+The critical flaw of the classic lottery model is that it breaks down at low densities (few propagules dispersing to each territory), precluding density-dependent behaviour.
+Our first task is to analytically extend the classic lottery model to correctly account for low density behavior (sections ``Model'' and ``Mean field approximation'').
 
@@
 
-Clearly MacArthur's argument is not compatible with the constant-$N$, relative fitness description of selection, in which $n_1$ and $n_2$ change along a line defined by $n_1+n_2=N$, and population density is otherwise irrelevant (Fig.~\ref{fig:Ksel}b).
-Since $N$ is fixed, it is not possible to persist at higher densities than $N$, nor is it possible to suppress the density of a competitor at densities lower than $N$
-
-In this case, all four $K$ values are equal to $N$ [not!], and the $f_1$ and $f_2$ nullclines do not even exist.
-
-Given the extremely broad variety of absolute population growth dynamics represented by Eq.~\eqref{eq:macgeneral}, at this point we might start to doubt that the relative fitness, constant=$N$ description of selection has a plausible basis in population ecology.
-In particular, the uncoupling of selection and population demography could be problematic, because they are both driven by the same birth/death events.
-
-In fact, one step in MacArthur's argument considerably narrows its scope: assuming that the nullclines of $f_1$ and $f_2$ exist at all.
-The strength of this assumption is clear if we note that essence of the dichotomy underlying the $r$/$K$ scheme is between interaction-dependent selection and interaction-independent selection.
-That is, selective shifts in frequency are a result of differences in absolute growth rates, but these differences can arise in two logically distinct ways: 1) some types expand more rapidly the absence of interactions between individuals or 2) some types are superior in their interactions with other types.
-Population density is a key factor controlling whether individuals interact, thereby setting the relative contributions of these forms of selection.
-But there are obviously important forms of interaction that are not expressible in terms of slanted density isoclines.
-For instance, contests for territory and mates are predominantly relative.
-MacArthur's argument starts with assumptions about how selection depends on density, and then tries to make sense of what this says about the evolution of competitive traits such as ``efficiency'' of consumable resource consumption.
-The study of density-dependent selection should start from a description of the interactions between types, and then analyze how the effects of those interactions vary with density.
-
 %This  also seems to be why the density-dependent selection literature expended such considerable effort to showing that evolution under density-dependent selection optimizes population size in some sense \citep{roughgarden_1979} --- a meaningless
-
-%In the case of exploitation competition for consumable resources, intra- and inter-type competition are connected via resource use ``efficiency'', but obviously this need not be true more generally \citep{gill_1974,case_1974}.
-
-Here we investigate the
-
-territorial model of growth, dispersal and competition.
-
-We revisit the classic lottery model of \cite{chesson_1981}, which has two features that make it well suited for this role, but one critical flaw that we rectify here.
-
-The first feature is that the lottery representation of competition is  particularly concise.
-Mature individuals (``adults'') each require their own territory, whereas newborn individuals (``propagules'') disperse to, and subsequently compete for, territories made available by the death of adults.
-Territorial contest among propagules leaves a single victorious adult per territory, the victor chosen at random from the propagules present, with probabilities weighted by a coefficient for each type representing competitive ability, akin to a lottery \citep{sale_77}.
-
-The second feature is the close connection between the lottery model and one of the foundational models of population genetics, the Wright-Fisher model of genetic drift, which we discuss further below.
-
-The critical flaw of the classic lottery model is that it breaks down at low densities (few propagules dispersing to each territory), precluding density-dependent behaviour.
-Our first task is to analytically extend the classic lottery model to correctly account for low density behavior (sections ``Model'' and ``Mean field approximation'').
-
-Taking an example inspired by recent studies of rapid, seasonal evolution in \textit{Drosophila} \citep{bergland_14}, we discuss how environmental fluctuations might stabilize polymorphisms when population density is cyclical.
 
@@
 \begin{equation}
-R_i=\frac{\overline{c}e^{-l_i}(1-e^{-(L-l_i)})}{c_i +\frac{L-1+e^{-L}}{1-(1+L)e^{-L}}\frac{\overline{c}L- c_il_i}{L-l_i}},\label{eq:Dr}
+R_i=\frac{\overline{c}e^{-l_i}(1-e^{-(L-l_i)})}{c_i +\frac{\overline{c}L- c_il_i}{L-l_i}\frac{L-1+e^{-L}}{1-(1+L)e^{-L}}},\label{eq:Dr}
 \end{equation}
@@
 \begin{equation}
-A_i=\frac{\overline{c}(1-e^{-l_i})}{\frac{1-e^{-l_i}}{1-(1+l_i)e^{-l_i}}c_il_i+\frac{1}{L-l_i}\left(L\frac{1-e^{-L}}{1-(1+L)e^{-L}}-l_i\frac{1-e^{-l_i}}{1-(1+l_i)e^{-l_i}}\right)\sum_{j\neq i}c_jl_j}.\label{eq:Da}
+A_i=\frac{\overline{c}(1-e^{-l_i})}{\frac{1-e^{-l_i}}{1-(1+l_i)e^{-l_i}}c_il_i+\frac{\overline{c}L- c_il_i}{L-l_i}\left(L\frac{1-e^{-L}}{1-(1+L)e^{-L}}-l_i\frac{1-e^{-l_i}}{1-(1+l_i)e^{-l_i}}\right)}.\label{eq:Da}
 \end{equation}
@@
 The logistic model for which $r$ and $K$ are named, does not capture this intuition.
+
+%In fact, one step in MacArthur's argument considerably narrows its scope: assuming that the nullclines of $f_1$ and $f_2$ exist at all.
+The strength of this assumption is clear if we note that essence of the dichotomy underlying the $r$/$K$ scheme is between interaction-dependent selection and interaction-independent selection.
+That is, selective shifts in frequency are a result of differences in absolute growth rates, but these differences can arise in two logically distinct ways: 1) some types expand more rapidly the absence of interactions between individuals or 2) some types are superior in their interactions with other types.
+Population density is a key factor controlling whether individuals interact, thereby setting the relative contributions of these forms of selection.
+But there are obviously important forms of interaction that are not expressible in terms of slanted density isoclines.
+For instance, contests for territory and mates are predominantly relative.
+MacArthur's argument starts with assumptions about how selection depends on density, and then tries to make sense of what this says about the evolution of competitive traits such as ``efficiency'' of consumable resource consumption.
+The study of density-dependent selection should start from a description of the interactions between types, and then analyze how the effects of those interactions vary with density.
+
+%In the case of exploitation competition for consumable resources, intra- and inter-type competition are connected via resource use ``efficiency'', but obviously this need not be true more generally \citep{gill_1974,case_1974}.
 

### 2017-11-11 Kplot.py
l

@@ -44,7 +44,7 @@ ax1.set_yticklabels([])
 ax1.set_xlabel(r"$n_1$",fontsize=20)
 ax1.set_ylabel(r"$n_2$",fontsize=20)
 
-ax2.quiver(z1[:-1], z2[:-1], z1[1:]-z1[:-1], z2[1:]-z2[:-1], scale_units='xy', angles='xy', scale=1)
+ax2.quiver(z1[:-1], z2[:-1], z1[1:]-z1[:-1], z2[1:]-z2[:-1], color="b", scale_units='xy', angles='xy', scale=1)
 ax2.annotate(r'$\frac{dN}{dt}=0$',xy=(0.53,0.53),xycoords='axes fraction',fontsize=16)
 ax2.annotate(r'$(b)$',xy=(0.9,0.93),xycoords='axes fraction',fontsize=16)
 

### 2017-11-11 manuscript.tex
l

@@
 In crowded populations, relative fitness models almost universally assume that total population size $N$ is fixed, or has some externally imposed time course.
-This is exemplified by the Wright-Fisher model, in which time advances in discrete generations which do not overlap, and fitness can be interpreted as a product of fertility and juvenile viability \citep[pp.
+This is exemplified by the Wright-Fisher model, in which time advances in discrete non-overlapping generations, and fitness can be interpreted as a product of fertility and juvenile viability \citep[pp.
 185]{crow_1970}.
@@
 Since no population has unlimited resources, the functions $f_1$ and $f_2$ must decline to zero if either $n_1$ or $n_2$ are made sufficiently large.
-The outcome of selection can then be determined by the shape of the nullclines $f_1(n_1,n_2)=0$ and $f_2(n_1,n_2)=0$.
+The outcome of selection can then be determined by the shape of the nullclines $f_1(n_1,n_2)=0$ and $f_2(n_1,n_2)=0$ in $(n_1,n_2)$ space.
 Specifically, a type will be excluded if its nullcline is completely contained in the region bounded by the other type's nullcline.
-For a type to have the possibility of persisting, it must be able to keep growing to higher densities than the other type can tolerate in some region of $(n_1,n_2)$ space (Fig.~\ref{fig:Ksel}a).
+In other words, for a type to have the possibility of persisting, it must be able to keep growing to higher densities than the other type can tolerate in some region of $(n_1,n_2)$ space (Fig.~\ref{fig:Ksel}a).
 
@@
 More sophisticated models using the Lotka-Volterra competition equations, or explicit resource consumption, also intrinsically tie high-density selection to changes in population density.
-This raises the possibility that the ubiquitous constant-$N$, relative fitness description of selection could be incompatible with a huge class of population ecological processes occurring in nature and the lab.
-
-Here we introduce a novel density-dependent model of population growth based on territorial contests, and show that when this model reaches a demographic steady-state, the constant-$N$, relative fitness picture emerges.
-We show that our model is entirely consistent with the claim that ``fitness is $K$'' in crowded populations when MacArthur's very liberal use of ``$K$'' is taken into account.
-In particular, MacArthur's argument does not justify the widespread intuition that selection in crowded environments  attempts to maximize population size is a consequence of
-
-argument is widely interpreted to mean that positive selection in crowded populations involves the ability to keep growing at ever higher population densities [citations]
-
-Moreover, it is not compatible with most verbal interpretations of MacArthur's argument,
-
-\citep{anderson_1971}
-
- with the assumption of constant $N$, although as we will discuss below, it is a narrow interpretation of MacArthur's argument.
-
-We revisit the classic lottery model of \cite{chesson_1981}, which has two features that make it well suited for this role, but one critical flaw that we rectify here.
-
-The first feature is that the lottery representation of competition is  particularly concise.
-Mature individuals (``adults'') each require their own territory, whereas newborn individuals (``propagules'') disperse to, and subsequently compete for, territories made available by the death of adults.
-Territorial contest among propagules leaves a single victorious adult per territory, the victor chosen at random from the propagules present, with probabilities weighted by a coefficient for each type representing competitive ability, akin to a lottery \citep{sale_77}.
-
-The second feature is the close connection between the lottery model and one of the foundational models of population genetics, the Wright-Fisher model of genetic drift, which we discuss further below.
-
-The critical flaw of the classic lottery model is that it breaks down at low densities (few propagules dispersing to each territory), precluding density-dependent behaviour.
-Our first task is to analytically extend the classic lottery model to correctly account for low density behavior (sections ``Model'' and ``Mean field approximation'').
-
-MacArthur defines four ``carrying capacities'' as the intersection points of the nullclines with the $n_1$ and $n_2$ axes (i.e.
-$f_1(K_{11},0)=0$, $f_1(0,K_{12})=0$, $f_2(0,K_{22})=0$ and $f_2(K_{21},0)=0$).
-Assuming that the nullclines are close to linear, these $K$ values determine whether a region of higher-density growth exists for each type.
-Note that only $K_{11}$ and $K_{22}$ are saturation densities analogous to ``$K$'' in the logistic model; $K_{12}$ and $K_{21}$ represent the effects of competition between types.
-For instance, under Lotka-Volterra competition we have $f_1(n_1,n_2)=r_1(1-\alpha_{11}n_1-\alpha_{12}n_2)n_1$.
-Then the nullclines are linear and $\alpha_{11}=1/K_{11}$ measures intra-type competitive effects, while $\alpha_{12}=1/K_{12}$ measures inter-type competitive effects.
-Thus, to summarize, ``fitness is $K$'' in the sense that selection in crowded populations either favors the ability to keep growing at ever higher densities (moving a type's own nullcline outwards), or the ability to suppress the growth of competitors at lower densities (moving the nullcline of competitors inwards).
-This general idea applies even if the nullclines are nonlinear to such an extent that the ``$K$'' values themselves do not give much information about the regions of high-density growth.
+This raises the possibility that the ubiquitous constant-$N$, relative fitness description of selection could be incompatible with a huge class of population ecological processes occurring in nature and experiments.
 
@@
 
+Here we introduce a novel model of density-dependent population growth based on territorial contests, and show that when this model reaches a demographic steady-state, the constant-$N$, relative fitness picture emerges.
+Our model is firmly grounded in population ecology, with fundamental parameters given by birth and death rates, and competitive ability.
+We show that this model can also be interpreted as a density-dependent generalization of the Wright-Fisher model with overlapping generations.
+
+Futhermore, we show that our model is entirely consistent with the claim that ``fitness is $K$'' in crowded populations provided that MacArthur's liberal use of ``$K$'' is taken into account.
+In particular, we emphasize that MacArthur's argument does not justify the widespread intuition that selection in crowded environments is necessarily connected to achieving greater densities \citep{anderson_1971}.
+This is largely an artifact of the models historically used in the density-dependent selection literature, which ignore relative contests.
+
 %This  also seems to be why the density-dependent selection literature expended such considerable effort to showing that evolution under density-dependent selection optimizes population size in some sense \citep{roughgarden_1979} --- a meaningless
+
+Our model is essentially a density-dependent generalization of the classic ecological lottery model \cite{chesson_1981}.
+In the lottery model, mature individuals (``adults'') each require their own territory, whereas newborn individuals (``propagules'') disperse to, and subsequently compete for, territories made available by the death of adults.
+Territorial contest among propagules leaves a single victorious adult per territory, the victor chosen at random from the propagules present, with probabilities weighted by a coefficient for each type representing competitive ability, akin to a weighted lottery \citep{sale_77}.
+
+The classic lottery model assumes a saturated population with constant $N$, and a large number of propagules dispersing to each territory (the Wright-Fisher model makes a similar ``infinite propagule'' assumption to justify sampling with replacement).
+As such, the lottery model breaks down at low densities (few propagules dispersing to each territory).
+Our first task is to analytically extend the classic lottery model to correctly account for low density behavior (sections ``Model'' and ``Mean field approximation'').
+We then...
 
@@
 First, propagules are produced by adults and dispersed at random (only propagules landing on unoccupied territories are shown).
-Lottery competition then occurs in each unoccupied territory (only illustrated in one territory).
+Some territories may receive zero propagules.
+Lottery competition then occurs in each unoccupied territory that receives a propagule (only illustrated in one territory).
 Each genotype has a probability proportional to $c_i x_i$ of securing a given territory, where $c_i$ measures competitive ability and $x_i$ is the number of propagules that disperse there.
@@
 The classic lottery model (Eq.
-1) fails to give co-existence for these parameters because expansion to carrying capacity occurs immediately at the start of the summer (Fig.
+BLAH) fails to give co-existence for these parameters because expansion to carrying capacity occurs immediately at the start of the summer (Fig.
 \ref{fig:fluctuatingselection}d-f).
@@
 \end{figure}
+
+\subsection*{K-selection versus relative fitness}
+
+MacArthur defines four ``carrying capacities'' as the intersection points of the nullclines with the $n_1$ and $n_2$ axes (i.e.
+$f_1(K_{11},0)=0$, $f_1(0,K_{12})=0$, $f_2(0,K_{22})=0$ and $f_2(K_{21},0)=0$).
+Assuming that the nullclines are close to linear, these $K$ values determine whether a region of higher-density growth exists for each type.
+Note that only $K_{11}$ and $K_{22}$ are saturation densities analogous to ``$K$'' in the logistic model; $K_{12}$ and $K_{21}$ represent the effects of competition between types.
+For instance, under Lotka-Volterra competition we have $f_1(n_1,n_2)=r_1(1-\alpha_{11}n_1-\alpha_{12}n_2)n_1$.
+Then the nullclines are linear and $\alpha_{11}=1/K_{11}$ measures intra-type competitive effects, while $\alpha_{12}=1/K_{12}$ measures inter-type competitive effects.
+Thus, to summarize, ``fitness is $K$'' in the sense that selection in crowded populations either favors the ability to keep growing at ever higher densities (moving a type's own nullcline outwards), or the ability to suppress the growth of competitors at lower densities (moving the nullcline of competitors inwards).
+This general idea applies even if the nullclines are nonlinear to such an extent that the ``$K$'' values themselves do not give much information about the regions of high-density growth.
 

### 2017-11-15 manuscript.tex
l

@@
 Since no population has unlimited resources, the functions $f_1$ and $f_2$ must decline to zero if either $n_1$ or $n_2$ are made sufficiently large.
-The outcome of selection can then be determined by the shape of the nullclines $f_1(n_1,n_2)=0$ and $f_2(n_1,n_2)=0$ in $(n_1,n_2)$ space.
+The outcome of selection is then determined by the shape of the nullclines $f_1(n_1,n_2)=0$ and $f_2(n_1,n_2)=0$ in $(n_1,n_2)$ space.
 Specifically, a type will be excluded if its nullcline is completely contained in the region bounded by the other type's nullcline.
@@
 In this case the $f_1$ and $f_2$ nullclines are not even defined.
-Moreover, at face value MacArthur's, conclusion that ``fitness is $K$'' in crowded populations \citep[pp.
-149]{macarthur_1967} seems to be incompatible with constant $N$.
-In particular, the logistic model and its bi-allelic diploid variants are widely used to analyze MacArthur's ``$K$-selection'' in which case positive selection is necessarily tied to increases in $N$.
+Moreover, at face value MacArthur's conclusion that ``fitness is $K$'' in crowded populations \citep[pp.
+149]{macarthur_1967} seems to be fundamentally incompatible with constant $N$.
+In particular, the logistic model and its bi-allelic diploid variants are widely used to analyze MacArthur's ``$K$-selection'' in which case positive selection is necessarily tied to increases in equilibrium population size.
 More sophisticated models using the Lotka-Volterra competition equations, or explicit resource consumption, also intrinsically tie high-density selection to changes in population density.
@@
 \includegraphics[scale=0.8]{Kplot.pdf}
-\caption{\label{fig:Ksel} (a) MacArthur's dynamical argument for why ``fitness is $K$'' in crowded environments, using a Lotka-Volterra form for Eq.
+\caption{\label{fig:Ksel} (a) MacArthur's dynamical argument for why ``fitness is $K$'' in crowded environments, illustrated using the logistic model in Eq.
 \eqref{eq:macgeneral}.
-Assuming the nullclines $f_1(n_1,n_2)=0$ and $f_2(n_1,n_2)=0$ exist, exclusion is certain unless a type's nullcline extends out to greater densities in some region.
-In this example, the outcome depends on initial condition.
-(b) The constant-$N$, relative fitness description of selection.}
+The type that  (b) The constant-$N$, relative fitness description of selection.}
 \end{figure}
@@
 
+\subsection*{$K$-selection, $c$-selection and relative fitness}
+
+Here we compare $K$-selection and MacArthur's broader claims about selection in crowded populations with our density-dependent lottery model.
+To do so we first need to clarify MacArthur's particular use of ``$K$''.
+
+MacArthur defines four ``carrying capacities'' as the intersection points of the nullclines with the $n_1$ and $n_2$ axes (i.e.
+$f_1(K_{11},0)=0$, $f_1(0,K_{12})=0$, $f_2(0,K_{22})=0$ and $f_2(K_{21},0)=0$).
+Assuming that the nullclines are close to linear, these $K$ values determine whether a region of higher-density growth exists for each type.
+Note that only $K_{11}$ and $K_{22}$ are saturation densities analogous to ``$K$'' in the logistic model; $K_{12}$ and $K_{21}$ represent the effects of competition between types.
+For instance, under Lotka-Volterra competition we have $f_1(n_1,n_2)=r_1(1-\alpha_{11}n_1-\alpha_{12}n_2)n_1$.
+Then the nullclines are linear and $\alpha_{11}=1/K_{11}$ measures intra-type competitive effects, while $\alpha_{12}=1/K_{12}$ measures inter-type competitive effects.
+Thus, to summarize, ``fitness is $K$'' in the sense that selection in crowded populations either favors the ability to keep growing at ever higher densities (moving a type's own nullcline outwards), or the ability to suppress the growth of competitors at lower densities (moving the nullcline of competitors inwards).
+This general idea applies even if the nullclines are nonlinear to such an extent that the ``$K$'' values themselves do not give much information about the regions of high-density growth.
+
 \subsection*{Coexistence in constant and cyclical environments}\label{sec:invas}
@@
 \end{figure}
-
-\subsection*{K-selection versus relative fitness}
-
-MacArthur defines four ``carrying capacities'' as the intersection points of the nullclines with the $n_1$ and $n_2$ axes (i.e.
-$f_1(K_{11},0)=0$, $f_1(0,K_{12})=0$, $f_2(0,K_{22})=0$ and $f_2(K_{21},0)=0$).
-Assuming that the nullclines are close to linear, these $K$ values determine whether a region of higher-density growth exists for each type.
-Note that only $K_{11}$ and $K_{22}$ are saturation densities analogous to ``$K$'' in the logistic model; $K_{12}$ and $K_{21}$ represent the effects of competition between types.
-For instance, under Lotka-Volterra competition we have $f_1(n_1,n_2)=r_1(1-\alpha_{11}n_1-\alpha_{12}n_2)n_1$.
-Then the nullclines are linear and $\alpha_{11}=1/K_{11}$ measures intra-type competitive effects, while $\alpha_{12}=1/K_{12}$ measures inter-type competitive effects.
-Thus, to summarize, ``fitness is $K$'' in the sense that selection in crowded populations either favors the ability to keep growing at ever higher densities (moving a type's own nullcline outwards), or the ability to suppress the growth of competitors at lower densities (moving the nullcline of competitors inwards).
-This general idea applies even if the nullclines are nonlinear to such an extent that the ``$K$'' values themselves do not give much information about the regions of high-density growth.
 

### 2017-11-15 Kplot.py

@@ -10,6 +10,51 @@ import matplotlib.pyplot as plt
 
 x,y=np.linspace(0,1,1000),np.linspace(0,1,1000)
 X, Y = np.meshgrid(x,y)
+alphas=np.array([[1,1],[1.5,1.5]])
+U = (1-alphas[0,0]*X-alphas[0,1]*Y)*X
+V = (1-alphas[1,1]*Y-alphas[1,0]*X)*Y
+#speed = np.sqrt(U*U + V*V)
+
+fig1, (ax1, ax2) = plt.subplots(ncols=2,figsize=[8,4])
+ax1.set_aspect(1)
+ax2.set_aspect(1)
+plt.tight_layout()
+seedpoints=np.concatenate([[[_,0.2-_] for _ in np.arange(0.02,0.2,0.02)],\
+            [[float(_)/10,1.-_*0.1] for _ in range(0,11) if _ not in [1,7]]])
+ax1.streamplot(x, y, U, V,density=100,start_points=seedpoints,linewidth=1.)
+
+z1=np.linspace(0,1,10)
+z2=1-z1
+ax1.plot(z1,(1-alphas[0,0]*z1)/alphas[0,1],'k',linewidth=2)
+ax1.plot(z1,(1-alphas[1,0]*z1)/alphas[1,1],'k',linewidth=2)
+#ax1.plot([0,1/alphas[0,0]],[1/alphas[1,1],0],'y',linewidth=2)
+ax1.set_xlim([0,1])
+ax1.set_ylim([0,1])
+
+ax1.annotate(r'$\frac{dn_2}{dt}=0$',xy=(0.005,0.68),xycoords='axes fraction',fontsize=16)
+ax1.annotate(r'$\frac{dn_1}{dt}=0$',xy=(0.65,0.4),xycoords='axes fraction',fontsize=16)
+ax1.annotate(r'$K_{1}$',xy=(0.95,-0.07),xycoords='axes fraction',fontsize=16)
+ax1.annotate(r'$K_{2}$',xy=(-0.1,0.65),xycoords='axes fraction',fontsize=16)
+ax1.annotate(r'$(a)$',xy=(0.9,0.93),xycoords='axes fraction',fontsize=16)
+
+ax1.set_xticklabels([])
+ax1.set_yticklabels([])
+ax1.set_xlabel(r"$n_1$",fontsize=20)
+ax1.set_ylabel(r"$n_2$",fontsize=20)
+
[…]

### 2017-11-15 manuscript.tex
l

@@
 \includegraphics[scale=0.8]{Kplot.pdf}
-\caption{\label{fig:Ksel} (a) MacArthur's dynamical argument for why ``fitness is $K$'' in crowded environments, illustrated using the logistic model in Eq.
+\caption{\label{fig:Ksel} (a) MacArthur's dynamical argument for why ``fitness is $K$'' in crowded environments, illustrated using the logistic model $f_1=r_1(1-\frac{n_1+n_2}{K_1})n_1$ and $f_2=r_2(1-\frac{n_1+n_2}{K_2})n_1$ in Eq.
 \eqref{eq:macgeneral}.
-The type that  (b) The constant-$N$, relative fitness description of selection.}
+Here $r_1=r_2$ and $K_1>K_2$.
+(b) The constant-$N$, relative fitness description of selection.}
 \end{figure}
@@
 
-Here we compare $K$-selection and MacArthur's broader claims about selection in crowded populations with our density-dependent lottery model.
-To do so we first need to clarify MacArthur's particular use of ``$K$''.
+Here we compare MacArthur's claims about selection in crowded populations with our density-dependent lottery model.
 
-MacArthur defines four ``carrying capacities'' as the intersection points of the nullclines with the $n_1$ and $n_2$ axes (i.e.
-$f_1(K_{11},0)=0$, $f_1(0,K_{12})=0$, $f_2(0,K_{22})=0$ and $f_2(K_{21},0)=0$).
-Assuming that the nullclines are close to linear, these $K$ values determine whether a region of higher-density growth exists for each type.
-Note that only $K_{11}$ and $K_{22}$ are saturation densities analogous to ``$K$'' in the logistic model; $K_{12}$ and $K_{21}$ represent the effects of competition between types.
-For instance, under Lotka-Volterra competition we have $f_1(n_1,n_2)=r_1(1-\alpha_{11}n_1-\alpha_{12}n_2)n_1$.
-Then the nullclines are linear and $\alpha_{11}=1/K_{11}$ measures intra-type competitive effects, while $\alpha_{12}=1/K_{12}$ measures inter-type competitive effects.
-Thus, to summarize, ``fitness is $K$'' in the sense that selection in crowded populations either favors the ability to keep growing at ever higher densities (moving a type's own nullcline outwards), or the ability to suppress the growth of competitors at lower densities (moving the nullcline of competitors inwards).
+As discussed in the Introduction, MacArthur's general argument starts from Eq.~\eqref{eq:macgeneral} and then hinges on the shape of the nullclines $f_1(n_1,n_2)=0$ and $f_2(n_1,n_2)=0$.
+The symbol ``$K$'' enters the argument as the general label for the four intersection points of the nullclines with the $n_1$ and $n_2$ axes, specifically $f_1(K_{11},0)=0$, $f_1(0,K_{12})=0$, $f_2(0,K_{22})=0$ and $f_2(K_{21},0)=0$.
+These $K$ values determine whether a region of higher-density growth exists for each type, provided that the nullclines are close to being straight lines.
+
+We are now in a position to evaluate MacArthur's claim that ``fitness is $K$'' in crowded environments.
+Importantly, only $K_{11}$ and $K_{22}$ are saturation densities analogous to ``$K$'' in the logistic model; $K_{12}$ and $K_{21}$ represent the effects of competition between types.
+For instance, in the case of the Lotka-Volterra competition model we have $f_1(n_1,n_2)=r_1(1-\alpha_{11}n_1-\alpha_{12}n_2)n_1$ where $\alpha_{11}=1/K_{11}$ measures competitive effects between individuals of the first type, whereas $\alpha_{12}=1/K_{12}$ measures the effects of competition from individuals of the second type.
+Thus, ``fitness is $K$'' in the sense that selection in crowded populations either favors the ability to keep growing at ever higher densities (moving a type's own nullcline outwards), or the ability to suppress the growth of competitors at lower densities (moving the nullcline of competitors inwards).
 This general idea applies even if the nullclines are nonlinear to such an extent that the ``$K$'' values themselves do not give much information about the regions of high-density growth.

### 2017-11-16 manuscript.tex
l

@@
 468]{barton_2007}.
-This variety is not problematic by itself because different fitness measures serve different purposes.
-But ultimately any measure of fitness should be grounded in the processes of birth and death which govern population ecology and demography \citep{metcalf_2007}.
-While this connection is apparent for absolute fitness measures like $r$ and $K$, relative fitness seems largely divorced from population ecology.
+This variety is not necessarily problematic in itself, because different fitness may be more useful in different circumstances.
+But any measure of fitness should ultimately be grounded in the processes of birth and death which govern population ecology and demography \citep{metcalf_2007,doebeli_2017}.
+While this grounding is apparent for absolute fitness measures like $r$ and $K$, relative fitness seems largely divorced from population ecology.
 
@@
 
-As discussed in the Introduction, MacArthur's general argument starts from Eq.~\eqref{eq:macgeneral} and then hinges on the shape of the nullclines $f_1(n_1,n_2)=0$ and $f_2(n_1,n_2)=0$.
+In the Introduction we described MacArthur's general conclusions about the behaviour of Eq.~\eqref{eq:macgeneral} based on the shape of the nullclines $f_1(n_1,n_2)=0$ and $f_2(n_1,n_2)=0$, but did not specify the role of ``$K$''.
+
 The symbol ``$K$'' enters the argument as the general label for the four intersection points of the nullclines with the $n_1$ and $n_2$ axes, specifically $f_1(K_{11},0)=0$, $f_1(0,K_{12})=0$, $f_2(0,K_{22})=0$ and $f_2(K_{21},0)=0$.

### 2017-11-16 manuscript.tex
l

@@
 The issue can be expressed more formally by revisiting MacArthur's famous analysis of selection in crowded populations \citep{macarthur_1967}.
-MacArthur considers a population consisting of two types with densities $n_1$ and $n_2$ subject to density-dependent population growth described by general differential equations
+MacArthur considers a population consisting of two types with densities $n_1$ and $n_2$ subject to density-dependent population growth described by
 \begin{equation}
@@
 \end{equation}
-Since no population has unlimited resources, the functions $f_1$ and $f_2$ must decline to zero if either $n_1$ or $n_2$ are made sufficiently large.
-The outcome of selection is then determined by the shape of the nullclines $f_1(n_1,n_2)=0$ and $f_2(n_1,n_2)=0$ in $(n_1,n_2)$ space.
+Since no population has unlimited resources, the functions $f_1$ and $f_2$ must decline to zero if $n_1$ or $n_2$ are made sufficiently large.
+This defines the nullclines $f_1(n_1,n_2)=0$ and $f_2(n_1,n_2)=0$ in $(n_1,n_2)$ space.
+The outcome of selection is then determined by the shape of these nullclines.
 Specifically, a type will be excluded if its nullcline is completely contained in the region bounded by the other type's nullcline.
@@
 
-MacArthur's argument is clearly incompatible with a narrow interpretation of relative fitness models as selection along lines defined by $n_1+n_2=N$ for each $N$ (Fig.~\ref{fig:Ksel}b).
-In this case the $f_1$ and $f_2$ nullclines are not even defined.
-Moreover, at face value MacArthur's conclusion that ``fitness is $K$'' in crowded populations \citep[pp.
-149]{macarthur_1967} seems to be fundamentally incompatible with constant $N$.
-In particular, the logistic model and its bi-allelic diploid variants are widely used to analyze MacArthur's ``$K$-selection'' in which case positive selection is necessarily tied to increases in equilibrium population size.
-More sophisticated models using the Lotka-Volterra competition equations, or explicit resource consumption, also intrinsically tie high-density selection to changes in population density.
-This raises the possibility that the ubiquitous constant-$N$, relative fitness description of selection could be incompatible with a huge class of population ecological processes occurring in nature and experiments.
+MacArthur's argument is often presented using the logistic model and its bi-allelic diploid variants [citations].
+Positive selection in crowded populations is then necessarily tied to increases in equilibrium population size --- commonly referred to as ``$K$-selection''.
+Clearly this is  incompatible with the constant $N$ assumption of relative fitness models.
+While MacArthur's argument is not restricted to the logistic model, other standard ecological  models like the Lotka-Volterra competition equations, or consumable resource models, also tie selection in crowded populations to changes in $N$, albiet in more subtle ways (we return to this in section BLAH).
+More generally, it is not clear how relative fitness models fit in with MacArthur's argument at all, since they only allow for selection along lines defined by $n_1+n_2=N$ for each $N$ (Fig.~\ref{fig:Ksel}b); the $f_1$ and $f_2$ nullclines are thus not even defined.
+We are thus faced with the possibility that the ubiquitous constant-$N$, relative fitness description of selection could be incompatible with a huge class of population ecological processes occurring in nature and experiments.
 
@@
 
-Futhermore, we show that our model is entirely consistent with the claim that ``fitness is $K$'' in crowded populations provided that MacArthur's liberal use of ``$K$'' is taken into account.
+Futhermore, we show that our model is entirely consistent with MacArthur's analysis of selection in crowded poulations.
 In particular, we emphasize that MacArthur's argument does not justify the widespread intuition that selection in crowded environments is necessarily connected to achieving greater densities \citep{anderson_1971}.
@@
 
-In the Introduction we described MacArthur's general conclusions about the behaviour of Eq.~\eqref{eq:macgeneral} based on the shape of the nullclines $f_1(n_1,n_2)=0$ and $f_2(n_1,n_2)=0$, but did not specify the role of ``$K$''.
+In the Introduction we gave the essential details of MacArthur's general argument, but we did not specify the role of ``$K$''.
+
+revolving around the shape of the nullclines $f_1(n_1,n_2)=0$ and $f_2(n_1,n_2)=0$ of Eq.~\eqref{eq:macgeneral}
 

### 2017-11-17 manuscript.tex
l

@@
 
-Here we compare MacArthur's claims about selection in crowded populations with our density-dependent lottery model.
+In the Introduction we outlined MacArthur's general argument. Here we discuss out some further aspects of
 
-In the Introduction we gave the essential details of MacArthur's general argument, but we did not specify the role of ``$K$''.
+compare MacArthur's claims about selection in crowded populations with our density-dependent lottery model. but did not specify the role of ``$K$''.
 

### 2017-11-18 manuscript.tex
l

--- 2017-11-16 manuscript.tex
@@
 This defines the nullclines $f_1(n_1,n_2)=0$ and $f_2(n_1,n_2)=0$ in $(n_1,n_2)$ space.
-The outcome of selection is then determined by the shape of these nullclines.
+The outcome of selection is then determined by the relationship between these nullclines.
 Specifically, a type will be excluded if its nullcline is completely contained in the region bounded by the other type's nullcline.
@@
 
-Here we compare MacArthur's claims about selection in crowded populations with our density-dependent lottery model.
+We now compare MacArthur's claims about selection in crowded environments with our density-dependent lottery model.
 
-In the Introduction we gave the essential details of MacArthur's general argument, but we did not specify the role of ``$K$''.
+As shown in the Introduction, MacArthur's argument revolves around relationship between the nullclines $f_1(n_1,n_2)=0$ and $f_2(n_1,n_2)=0$ of Eq.~\eqref{eq:macgeneral}.
+To formalize this relationship, the familiar symbol ``$K$'' is introduced to label the four intersection points of the nullclines with the $n_1$ and $n_2$ axes, specifically $f_1(K_{11},0)=0$, $f_1(0,K_{12})=0$, $f_2(0,K_{22})=0$ and $f_2(K_{21},0)=0$.
+These $K$ values determine whether a region of higher-density growth exists for each type, provided that the nullclines are close to being straight lines.
+It is in this sense that MacArthur concludes that ``fitness is $K$'' in crowded populations \citep[pp.
+149]{macarthur_1967}.
 
-revolving around the shape of the nullclines $f_1(n_1,n_2)=0$ and $f_2(n_1,n_2)=0$ of Eq.~\eqref{eq:macgeneral}
-
-The symbol ``$K$'' enters the argument as the general label for the four intersection points of the nullclines with the $n_1$ and $n_2$ axes, specifically $f_1(K_{11},0)=0$, $f_1(0,K_{12})=0$, $f_2(0,K_{22})=0$ and $f_2(K_{21},0)=0$.
-These $K$ values determine whether a region of higher-density growth exists for each type, provided that the nullclines are close to being straight lines.
-
-We are now in a position to evaluate MacArthur's claim that ``fitness is $K$'' in crowded environments.
-Importantly, only $K_{11}$ and $K_{22}$ are saturation densities analogous to ``$K$'' in the logistic model; $K_{12}$ and $K_{21}$ represent the effects of competition between types.
-For instance, in the case of the Lotka-Volterra competition model we have $f_1(n_1,n_2)=r_1(1-\alpha_{11}n_1-\alpha_{12}n_2)n_1$ where $\alpha_{11}=1/K_{11}$ measures competitive effects between individuals of the first type, whereas $\alpha_{12}=1/K_{12}$ measures the effects of competition from individuals of the second type.
+Notice that this use of ``$K$'' is considerably more general than the $K$ parameter in the logistic model, including both the saturation densities $K_{11}$ and $K_{22}$ and high-density interaction terms $K_{12}$ and $K_{21}$.
+For instance, in the Lotka-Volterra competition model we have $f_1(n_1,n_2)=r_1(1-\alpha_{11}n_1-\alpha_{12}n_2)n_1$ where $\alpha_{11}=1/K_{11}$ measures competitive effects within the first type, whereas $\alpha_{12}=1/K_{12}$ measures competitive effects on the first type due to the second.
 Thus, ``fitness is $K$'' in the sense that selection in crowded populations either favors the ability to keep growing at ever higher densities (moving a type's own nullcline outwards), or the ability to suppress the growth of competitors at lower densities (moving the nullcline of competitors inwards).

### 2017-11-20 manuscript.tex
l

@@
 \section*{Abstract}
-
-This excludes fundamental ecological factors such as dynamic population size or density-dependence from the most genetically-realistic treatments of evolution, a problem that inspired MacArthur's influential but problematic $r$/$K$ theory.
-
-Following the spirit of $r$/$K$-selection as a general-purpose theory of density-dependent selection
 
@@
 \begin{equation}
-\Delta_+ n_i(t)=\frac{c_i l_i}{\sum_j c_j l_j}U(t)=b_i n_i\frac{1}{L}\frac{c_i}{\overline{c}}, \label{eq:lottery}
+\Delta_+ n_i(t)=\frac{c_i l_i}{\sum_j c_j l_j}U(t)=\frac{c_i l_i}{\overline{c}L}U(t), \label{eq:lottery}
 \end{equation}
@@
 
-As shown in the Introduction, MacArthur's argument revolves around relationship between the nullclines $f_1(n_1,n_2)=0$ and $f_2(n_1,n_2)=0$ of Eq.~\eqref{eq:macgeneral}.
-To formalize this relationship, the familiar symbol ``$K$'' is introduced to label the four intersection points of the nullclines with the $n_1$ and $n_2$ axes, specifically $f_1(K_{11},0)=0$, $f_1(0,K_{12})=0$, $f_2(0,K_{22})=0$ and $f_2(K_{21},0)=0$.
+As shown in the Introduction, MacArthur's argument revolves around the behaviour of  ecological models of the general form Eq.~\ref{eq:macgeneral}, which depends on the relationship between the nullclines $f_1(n_1,n_2)=0$ and $f_2(n_1,n_2)=0$.
+To formalize this relationship, MacArthur used the symbol ``$K$'' to label the four intersection points of the nullclines with the $n_1$ and $n_2$ axes, specifically $f_1(K_{11},0)=0$, $f_1(0,K_{12})=0$, $f_2(0,K_{22})=0$ and $f_2(K_{21},0)=0$.
 These $K$ values determine whether a region of higher-density growth exists for each type, provided that the nullclines are close to being straight lines.
-It is in this sense that MacArthur concludes that ``fitness is $K$'' in crowded populations \citep[pp.
-149]{macarthur_1967}.
+Note that, while $K_{11}$ and $K_{22}$ are saturation densities akin to the $K$ parameter in the logistic model, this is not true for $K_{12}$ and $K_{21}$, which are related to competition between types.
+For instance, in the Lotka-Volterra competition model we have $f_1(n_1,n_2)=r_1(1-\alpha_{11}n_1-\alpha_{12}n_2)n_1$ where $\alpha_{11}=1/K_{11}$ measures competitive effects within the first type, whereas $\alpha_{12}=1/K_{12}$ measures competitive effects on the first type due to the second.
 
-Notice that this use of ``$K$'' is considerably more general than the $K$ parameter in the logistic model, including both the saturation densities $K_{11}$ and $K_{22}$ and high-density interaction terms $K_{12}$ and $K_{21}$.
-For instance, in the Lotka-Volterra competition model we have $f_1(n_1,n_2)=r_1(1-\alpha_{11}n_1-\alpha_{12}n_2)n_1$ where $\alpha_{11}=1/K_{11}$ measures competitive effects within the first type, whereas $\alpha_{12}=1/K_{12}$ measures competitive effects on the first type due to the second.
-Thus, ``fitness is $K$'' in the sense that selection in crowded populations either favors the ability to keep growing at ever higher densities (moving a type's own nullcline outwards), or the ability to suppress the growth of competitors at lower densities (moving the nullcline of competitors inwards).
+Thus, when MacArthur concludes that  ``fitness is $K$'' in crowded populations \citep[pp.
+149]{macarthur_1967}, the meaning is that selection in crowded populations either favors the ability to keep growing at ever higher densities (moving a type's own nullcline outwards), or the ability to suppress the growth of competitors at lower densities (moving the nullcline of competitors inwards).
 This general idea applies even if the nullclines are nonlinear to such an extent that the ``$K$'' values themselves do not give much information about the regions of high-density growth.
+
+The nullclines of Eq.~\eqref{eq:master}
 

### 2017-11-20 simulations.py

@@ -166,22 +166,22 @@ T=100000
 x,y=np.linspace(0.0,1.,401),np.linspace(0.,1.,401)
 xgrid, ygrid=np.meshgrid(x,y)
 X,Y=np.zeros([len(x),len(y)]),np.zeros([len(x),len(y)])
-b=np.array([2.,1.])
-c=np.array([1.,10.])
-d=np.array([0.1,0.2])
+b=np.array([2.,.2])
+c=np.array([1.,1.])
+d=np.array([0.1,0.1])
 
 for i in xrange(len(x)):
     for j in xrange(len(y)):
         n=T*np.array([x[i],y[j]])
         U=T-sum(n)
-        X[j,i],Y[j,i]=deltnplus(b*n*U/float(T),c,U)-d*n
+        X[j,i],Y[j,i]=deltnplus(b*n,c,U)-d*n
 
 fig1, ax1 = plt.subplots()#(ncols=2,figsize=[8,4])
 ax1.set_aspect(1)
 #ax2.set_aspect(1)
 plt.tight_layout()
 seedpoints=[[x[_]/2,y[200-_]/2] for _ in range(0,201,10)]
-ax1.streamplot(x, y, X, Y,start_points=seedpoints,linewidth=1.,density=20)
+ax1.streamplot(x, y, X, Y,start_points=seedpoints,linewidth=1.,density=100)
 ax1.set_xlim([0,1])
 ax1.set_ylim([0,1])
 

### 2017-11-21 manuscript.tex
l

@@
 
-\title{Density-dependent selection with strictly relative contests}
+\title{Density-dependent selection with strictly relative competition}
 %\author{Jason Bertram $^{1,\ast}$ \\ 
@@
 
+This conclusion takes different forms in standard models of density-regulated population growth.
 MacArthur's argument is often presented using the logistic model and its bi-allelic diploid variants [citations].
-Positive selection in crowded populations is then necessarily tied to increases in equilibrium population size --- commonly referred to as ``$K$-selection''.
+Positive selection in crowded populations is then necessarily tied to increases in equilibrium population density --- commonly referred to as ``$K$-selection'' (Fig.~\ref{fig:Ksel}a).
+The more general Lotka-Volterra competition model, given by $f_1=r_1(1-\alpha_{11}n_1-\alpha_{12}n_2)n_1$ (and similarly for $f_2$) also ties selection in crowded populations to changes in $N$, albeit not solely via saturation density (we return to this in section BLAH).
+Under exploitation competition for a shared, homogeneous, consumable resource, the type able to deplete the resource to the lowest equilibrium density excludes the other ($R^*$-rule).
+
 Clearly this is incompatible with the constant $N$ assumption of relative fitness models.
-While MacArthur's argument is not restricted to the logistic model, other standard ecological  models like the Lotka-Volterra competition equations, or consumable resource models, also tie selection in crowded populations to changes in $N$, albiet in more subtle ways (we return to this in section BLAH).
+
+We are thus faced with the possibility that the ubiquitous constant-$N$, relative fitness description of selection could be incompatible with a huge class of population ecological processes occurring in nature and experiments.
+
 More generally, it is not clear how relative fitness models fit in with MacArthur's argument at all, since they only allow for selection along lines defined by $n_1+n_2=N$ for each $N$ (Fig.~\ref{fig:Ksel}b); the $f_1$ and $f_2$ nullclines are thus not even defined.
-We are thus faced with the possibility that the ubiquitous constant-$N$, relative fitness description of selection could be incompatible with a huge class of population ecological processes occurring in nature and experiments.
 

### 2017-11-22 manuscript.tex
l

--- 2017-11-20 manuscript.tex
@@
 468]{barton_2007}.
-This variety is not necessarily problematic in itself, because different fitness may be more useful in different circumstances.
-But any measure of fitness should ultimately be grounded in the processes of birth and death which govern population ecology and demography \citep{metcalf_2007,doebeli_2017}.
-While this grounding is apparent for absolute fitness measures like $r$ and $K$, relative fitness seems largely divorced from population ecology.
+This variety is not necessarily problematic in itself, because different measures of fitness may be more useful in different circumstances.
+But any measure of fitness should ultimately be grounded in the processes of birth and death which govern population biology \citep{metcalf_2007,doebeli_2017}.
+While this grounding is clear for absolute fitness measures like $r$ and $K$, relative fitness seems largely divorced from population ecology.
 
@@
 \end{equation}
-Since no population has unlimited resources, the functions $f_1$ and $f_2$ must decline to zero if $n_1$ or $n_2$ are made sufficiently large.
+Other than the type densities, the environment is assumed to remain constant.
+The functions $f_1$ and $f_2$ must decline to zero if $n_1$ or $n_2$ are sufficiently large, because no population has unlimited resources.
 This defines the nullclines $f_1(n_1,n_2)=0$ and $f_2(n_1,n_2)=0$ in $(n_1,n_2)$ space.
@@
 
+At face value, the preceding conclusion does not seem compatible with the constant $N$ assumption of relative fitness models.
+It is not even clear how to connect relative fitness models and MacArthur's argument, since the former only allows for selection along lines defined by $n_1+n_2=N$ for each $N$ (Fig.~\ref{fig:Ksel}b); the $f_1$ and $f_2$ nullclines are thus not even defined.
+The incompatibility is particularly apparent in some of the most widely-used models of population growth.
 MacArthur's argument is often presented using the logistic model and its bi-allelic diploid variants [citations].
-Positive selection in crowded populations is then necessarily tied to increases in equilibrium population size --- commonly referred to as ``$K$-selection''.
-Clearly this is  incompatible with the constant $N$ assumption of relative fitness models.
-While MacArthur's argument is not restricted to the logistic model, other standard ecological  models like the Lotka-Volterra competition equations, or consumable resource models, also tie selection in crowded populations to changes in $N$, albiet in more subtle ways (we return to this in section BLAH).
-More generally, it is not clear how relative fitness models fit in with MacArthur's argument at all, since they only allow for selection along lines defined by $n_1+n_2=N$ for each $N$ (Fig.~\ref{fig:Ksel}b); the $f_1$ and $f_2$ nullclines are thus not even defined.
-We are thus faced with the possibility that the ubiquitous constant-$N$, relative fitness description of selection could be incompatible with a huge class of population ecological processes occurring in nature and experiments.
+Positive selection in crowded populations then requires increases in equilibrium population density --- commonly referred to as ``$K$-selection'' (Fig.~\ref{fig:Ksel}a).
+The Lotka-Volterra competition model also couples selection in crowded populations to changes in total density $N$ (we return to this in BLAH).
+Moreover, exploitation competition for consumable resources is potentially a decisive factor in many crowded populations.
+On a single, homogeneous, limiting resource, the type able to deplete the resource to the lowest equilibrium density excludes the other ($R^*$-rule); except in very special cases, this will also entail differences in saturation density.
+Thus, it would seem that the ubiquitous constant-$N$, relative fitness description of selection is incompatible with a huge class of population ecological processes occurring in nature and experiments.
 
@@
 
+The constant-$N$, relative fitness description has historically been justified as a short-term linear approximation of a more complicated long-term eco-evolutionary process \cite[pp.
+277]{ewens_2004}.
+That is, within a sufficiently short time frame, population size can be treated as constant and selection quantified in terms of constant selection coefficients expressing relative fitness differences.
+Provided that selection is sufficiently weak and stable over time, this ``short-term''  assumption is not a major restriction.
+Yet it is increasingly recognized that selection is not always weak, that it can fluctuate dramatically over time, and coupled to this, that a population's density can vary by orders of magnitude over a few generations as a routine feature of its ecology.
+These are not rare exceptions, but occur widely in nature and the lab, including in wild \textit{Drosophila} \citep{messer_2016}.
+Moreover, the short-term approximation automatically precludes consideration of inherently long-term evolutionary processes like the management of genetic load and population extinction \citep{bertram_2017}.
+Nevertheless, relative fitness models like Wright-Fisher are the foundation for much of the population genetic literature, and are still widely used without considering the ``short-term'' restriction.
+Thus, to summarize, there are important theoretical and practical reasons to deepen our understanding of the population-ecological foundations of relative fitness models.
+
 Here we introduce a novel model of density-dependent population growth based on territorial contests, and show that when this model reaches a demographic steady-state, the constant-$N$, relative fitness picture emerges.
@@
 
-Futhermore, we show that our model is entirely consistent with MacArthur's analysis of selection in crowded poulations.
+Futhermore, we show that our model is entirely consistent with MacArthur's analysis of selection in crowded populations.
 In particular, we emphasize that MacArthur's argument does not justify the widespread intuition that selection in crowded environments is necessarily connected to achieving greater densities \citep{anderson_1971}.
@@
 
+\subsection*{Basic assumptions}
+
 We assume that reproductively mature individuals (``adults'') each require their own territory to survive and reproduce (Fig.~\ref{fig:lottery}).
@@
 Time $t$ advances in discrete iterations, each representing the time from birth to reproductive maturity.
-In iteration $t$, the number of adults of the $i$'th genotype is $n_i(t)$, the total number of adults is $N(t)=\sum_i n_i(t)$, and the number of unoccupied territories is $U(t)=T-N(t)$.
+In iteration $t$, the number of adults of the $i$'th type is $n_i(t)$, the total number of adults is $N(t)=\sum_i n_i(t)$, and the number of unoccupied territories is $U(t)=T-N(t)$.
 
-We assume that the $n_i$ and $T$ are large enough that stochastic fluctuations in the $n_i$ (``drift'') can be ignored.
+We assume that the $n_i$ are large enough that stochastic fluctuations in the $n_i$ (``drift'') can be ignored.
+In particular, we do not evaluate the initial stochastic behaviour of mutant lineages while they are at low abundance.
 We derive deterministic equations for the expected change in the $n_i$ over time, leaving the evaluation of drift for future work.
-This is an excellent approximation when the $n_i$ are all large.
-However, we also do not evaluate the initial stochastic behaviour of adaptive mutant lineages while they are at low abundance.
-When considering new mutations, we therefore restrict our attention to begin with the earliest (lowest $n_i$) deterministic behavior of mutant lineages (the transition to deterministic growth occurs at an abundance $n_i$ of order equal to their inverse expected absolute growth rate; \citealt{uecker_2011}).
+
+%When considering new mutations, we therefore restrict our attention to the earliest (lowest $n_i$) deterministic behavior of mutant lineages (the transition to deterministic growth occurs at an abundance $n_i$ of order equal to their inverse expected absolute growth rate; \citealt{uecker_2011}).
 
@@
 Lottery competition then occurs in each unoccupied territory that receives a propagule (only illustrated in one territory).
-Each genotype has a probability proportional to $c_i x_i$ of securing a given territory, where $c_i$ measures competitive ability and $x_i$ is the number of propagules that disperse there.
-In the illustrated territory, the black genotype disperses more propagules but is a poorer competitor.
-Territories are then made available by adult mortality (red crosses).}
+Each type has a probability proportional to $c_i x_i$ of securing a given territory, where $c_i$ measures competitive ability and $x_i$ is the number of propagules that disperse there.
+In the illustrated territory, the black type disperses more propagules but is a poorer competitor.
+Territories are then made available by deaths among those adults present at the start of the iteration (red crosses).}
 \end{figure}
@@
 Loss of propagules during dispersal is subsumed into $m_i$.
+We assume that each adult produces a constant number $b_i$ of successfully dispersing propagules; the loss of propagules due to dispersal to occupied territories then implies $m_i=b_i(1-N/T)n_i$.
+Note that due to our assumption of uniform dispersal, the parameter $b_i$ can be thought of as a measure of ``colonization ability'', which combines fecundity and dispersal ability \citep{levins_71,tilman_94,bolker_99}.
+In addition to random dispersal, we will compare our model to perfect directed dispersal, in which each propagule finds an unoccupied territory if one is available ($m_i=b_i$) \citep{chesson_1983}.
 
-In general, $m_i$ will increase with $n_i$, and will depend on population density $N$.
-For example, if $b_i$ is the number of successfully dispersing propagules produced per genotype $i$ adult, then the loss of propagules due to dispersal to occupied territories implies $m_i=b_i(1-N/T)n_i$, akin to Levins' competition-colonization model \citep{levins_71,tilman_94}.
-In section ``Cyclical birth and death rates'' we evaluate Eq.
-\eqref{eq:master} numerically using this functional form for $m_i$, with $b_i$ assumed to be constant.
-
-In ``Invasion of rare genotypes and coexistence'', we assume the simpler form $m_i=b_i n_i$, with constant $b_i$, meaning that all propagules land on unoccupied territories (a form of directed dispersal).
-This simplifies the mathematics without affecting the results of those sections, which only depend on the low-frequency invasion behavior of Eq.
-\eqref{eq:master}.
-Note that due to our assumption of uniform dispersal, the parameter $b_i$ can be thought of as a measure of ``colonization ability'', which combines fecundity and dispersal ability \citep{levins_71,tilman_94,bolker_99}.
-
-The number of individuals of the $i$'th genotype landing in any particular territory is denoted $x_i$.
+The number of individuals of the $i$'th type landing in any particular territory is denoted $x_i$.
 We assume that $x_i$ follows a Poisson distribution $p_i(x_i)=l_i^{x_i} e^{-l_i}/x_i!$, where $l_i=m_i/U$ is the mean territorial propagule density.
@@
 
-When multiple propagules land on the same territory, the victor is determined by lottery competition: genotype $i$ wins a territory with probability $c_i x_i/\sum_j c_j x_j$, where $c_i$ is a constant representing relative competitive ability (Fig.
+When multiple propagules land on the same territory, the victor is determined by lottery competition: type $i$ wins a territory with probability $c_i x_i/\sum_j c_j x_j$, where $c_i$ is a constant representing relative competitive ability (Fig.
 \ref{fig:lottery}).
+We expect that a fraction $p_1(x_1)\ldots p_G(x_G)$ of the $U$ unoccupied territories will have the propagule composition $x_1,\ldots,x_G$.
+type $i$ is expected to win $c_i x_i/\sum_j c_j x_j$ of these.
+Ignoring fluctuations about these two expectations (due to our no-drift, large $T$, large $n_i$ approximation), type $i$'s territorial acquisition is given by
+\begin{equation}
+\Delta_+ n_i(t)=U(t)\sum_{x_1,\ldots,x_G} \frac{c_i x_i}{\sum_j c_j x_j} p_1(x_1)\ldots p_G(x_G), \label{eq:growthsumuncoupled}
+\end{equation}
+in our extended lottery model, where the sum only includes territories with at least one propagule present.
 
-In the classic lottery model \citep{chesson_1981}, unoccupied territories are assumed to be saturated with propagules from every genotype $l_i\gg 1$.
-From the law of large numbers, the composition of propagules in each territory will then not deviate appreciably from the mean composition $l_1,l_2,\ldots,l_G$ ($G$ is the number of genotypes present), and so the probability that genotype $i$ wins any particular unoccupied territory is approximately $c_i l_i/\sum_j c_j l_j$.
-Let $\Delta_+ n_i$ denote the number of territories won by genotype $i$.
-Then $\Delta_+ n_1,\Delta_+ n_2,\ldots,\Delta_+ n_G$ follow a multinomial distribution with $U$ trials and success probabilities $\frac{c_1 l_1}{\sum_j c_j l_j},\frac{c_2 l_2}{\sum_j c_j l_j},\ldots,\frac{c_G l_G}{\sum_j c_j l_j}$, respectively.
-Genotype $i$ is expected to win $c_i l_i/\sum_j c_j l_j$ of the $U$ available territories, and deviations from this expected outcome are small (since $T$ is large by assumption), giving
+Finally, we assume that mortality only occurs in adults (Fig.~\ref{fig:lottery}; setting aside the juvenile deaths implicit in territorial contest), and at a constant, type-specific per-capita rate $0\leq d_i\leq 1$, so that the overall change in type abundances is
+\begin{equation}
+\Delta n_i(t)=\Delta_+ n_i(t)-d_i n_i(t). \label{eq:delttot}
+\end{equation}
+
+\subsection*{Connection to the Wright-Fisher and classic lottery models}
+
+In the classic lottery model \citep{chesson_1981}, unoccupied territories are assumed to be saturated with propagules from every type $l_i\gg 1$.
+From the law of large numbers, the composition of propagules in each territory will then not deviate appreciably from the mean composition $l_1,l_2,\ldots,l_G$ ($G$ is the number of types present), and so the probability that type $i$ wins any particular unoccupied territory is approximately $c_i l_i/\sum_j c_j l_j$.
+Then the numbers of territories won by each type $\Delta_+ n_1,\Delta_+ n_2,\ldots,\Delta_+ n_G$ follow a multinomial distribution with $U$ trials and success probabilities $\frac{c_1 l_1}{\sum_j c_j l_j},\frac{c_2 l_2}{\sum_j c_j l_j},\ldots,\frac{c_G l_G}{\sum_j c_j l_j}$, respectively.
+Type $i$ is expected to win $c_i l_i/\sum_j c_j l_j$ of the $U$ available territories, and deviations from this expected outcome are small (since $T$ is large by assumption), giving
 \begin{equation}
@@
 
+In our extension of the classic lottery model (Eq.~\ref{eq:delttot}), we do not restrict ourselves to high propagule densities.
+Eq.
+\eqref{eq:lottery} is nonsensical if even a single type has low propagule density ($l_i\ll 1$): type $i$ can win at most $m_i$ territories, yet Eq.~\eqref{eq:lottery} demands $c_i l_i/\sum_j c_j l_j$ of the $U$ unoccupied territories, for any value of $U$.
+Thus, deviations from the mean propagule composition $l_1,l_2,\ldots,l_G$ are important at low density.
+The reason for this lies in the fundamental discreteness of biological individuals: types with few propagules depend on the outcome of contests in territories where they have at least one propagule present, not some small fraction of a propagule as would be implied by small $l_i$ in Eq.~\eqref{eq:lottery}.
+
 There is a close connection between the classic lottery model and the Wright-Fisher model of genetic drift \citep{svardal_2015}.
-In the Wright-Fisher model, genotype abundances are sampled each generation from a multinomial distribution with success probabilities $w_i n_i/\sum_j w_j n_j$, where $w$ is relative fitness and the $n_i$ are  genotype abundances in the preceding generation.
+In the Wright-Fisher model, type abundances are sampled each generation from a multinomial distribution with success probabilities $w_i n_i/\sum_j w_j n_j$, where $w$ is relative fitness and the $n_i$ are  type abundances in the preceding generation.
 Population size $N$ remains constant.
-This is mathematically equivalent to the classic lottery model with non-overlapping generations ($d_i=1$ for all $i$) and $w_i=b_i c_i$.
-Thus, the classic lottery model allows us to replace the abstract Wright-Fisher relative fitnesses $w_i$ with more ecologically-grounded fecundity, competitive ability and mortality parameters $b_i$, $c_i$ and $d_i$, respectively.
-Since birth and death rates affect absolute abundances, this allows us to evaluate selection at different densities (after appropriate extensions are made), in an otherwise very similar model to the canonical Wright-Fisher.
-We therefore expect that drift in realized values of $n_i$ in our extended lottery model should be similar to that in the Wright-Fisher model, but we leave this for future work.
-
-In our extension of the classic lottery model, we do not restrict ourselves to high propagule densities.
-Eq.
-\eqref{eq:lottery} is nonsensical if even a single type has low propagule density ($l_i\ll 1$): genotype $i$ can win at most $m_i$ territories, yet Eq.
-\eqref{eq:lottery} demands $c_i l_i/\sum_j c_j l_j$ of the $U$ unoccupied territories, for any value of $U$.
-Intuitively, the cause of this discrepancy is that individuals are discrete.
-Genotypes with few propagules depend on the outcome of contests in territories where they have at least one propagule present, not some small fraction of a propagule as would be implied by small $l_i$ in the classic lottery model.
-In other words, deviations from the mean propagule composition $l_1,l_2,\ldots,l_G$ are important at low density.
-
-We expect that a fraction $p_1(x_1)\ldots p_G(x_G)$ of the $U$ unoccupied territories will have the propagule composition $x_1,\ldots,x_G$.
-Genotype $i$ is expected to win $c_i x_i/\sum_j c_j x_j$ of these.
-Ignoring fluctuations about these two expectations (due to our no-drift, large $T$, large $n_i$ approximation), genotype $i$'s territorial acquisition is given by
-\begin{equation}
-\Delta_+ n_i(t)=U(t)\sum_{x_1,\ldots,x_G} \frac{c_i x_i}{\sum_j c_j x_j} p_1(x_1)\ldots p_G(x_G), \label{eq:growthsumuncoupled}
-\end{equation}
-in our extended lottery model, where the sum only includes territories with at least one propagule present.
-Note that unlike the classic lottery model, not all unoccupied territories are claimed each iteration, since under Poisson dispersal a fraction $e^{-L}$ remain unoccupied.
-
-We assume that mortality only occurs in adults (Fig.~\ref{fig:lottery}; setting aside the juvenile deaths implicit in territorial contest), and at a constant, genotype-specific per-capita rate $0\leq d_i\leq 1$, so that the overall change in genotype abundances is
-\begin{equation}
-\Delta n_i(t)=\Delta_+ n_i(t)-d_i n_i(t). \label{eq:delttot}
-\end{equation}
+This is equivalent to the classic lottery model with non-overlapping generations ($d_i=1$ for all $i$) and relative fitness given by $w_i=b_i c_i$ i.e.
+a product of fecundity and viability \citep[pp.
+185]{crow_1970}.
+Thus, the classic lottery model is essentially the Wright-Fisher model extended to allow overlapping generations, but ignoring drift.
+This means that our extension of the classic lottery model to arbitrary densities represents a density-dependent generalization of the Wright-Fisher model.
 
@@
 Similarly to the high-$l_i$ approximation of classic lottery model, we replace the $x_i$ with appropriate mean values, although we cannot simply replace $x_i$ with $l_i$.
-For a genotype with low propagule density $l_i\ll 1$, we have $x_i=1$ in the territories where its propagules land, and so its growth comes entirely from territories which deviate appreciably from $l_i$.
+For a type with low propagule density $l_i\ll 1$, we have $x_i=1$ in the territories where its propagules land, and so its growth comes entirely from territories which deviate appreciably from $l_i$.
 To account for this, we separate Eq.
@@
 Our more general mean field approximation only requires that there are no large discrepancies in competitive ability (i.e.
-we do not have $c_i/c_j\gg 1$ for any two genotypes).
+we do not have $c_i/c_j\gg 1$ for any two types).
 We obtain (details in Appendix B)
 \begin{equation}
-\Delta_+ n_i(t)\approx b_i n_i\left[e^{-L}+(R_i+A_i)\frac{c_i}{\overline{c}}\right], \label{eq:master}
+\Delta_+ n_i(t)\approx \left[e^{-L}+(R_i+A_i)\frac{c_i}{\overline{c}}\right]l_i U(t), \label{eq:master}
 \end{equation}
@@
 \end{equation}
-To supplement our analytical mean field derivation, we did numerical simulations of our exact our density-dependent lottery model, and verified  that Eq.
-\eqref{eq:master} is a good approximation (Appendix B).
-Thus, Eq.
-\eqref{eq:master} describes how type abundances change over time in a lottery model where population density can itself vary with time.
 
@@
 The first, $e^{-L}$, accounts for propagules which land alone on unoccupied territories; these territories are won without contest.
-The second, $R_i c_i/\overline{c}$ represents competitive victories when the $i$ genotype is a rare invader in a high density population: from Eq.
-\eqref{eq:Dr}, $R_i\rightarrow 0$ when the $i$ genotype is abundant ($l_i\gg 1$), or other genotypes are collectively rare ($L-l_i\ll 1$).
-The third term, $A_ic_i/\overline{c}$, represents competitive victories when the $i$ genotype is abundant: $A_i\rightarrow 0$ if $l_i\ll 1$.
+The second, $R_i c_i/\overline{c}$ represents competitive victories when the $i$ type is a rare invader in a high density population, determining its invasion fitness \citep{metz_1992}.
+The third term, $A_ic_i/\overline{c}$, represents competitive victories when the $i$ type is abundant.
 The relative importance of these three terms varies with both the overall propagule density $L$ and the relative propagule frequencies $m_i/M$.
-If $l_i\gg 1$ for all genotypes, we recover the classic lottery model (only the $A_ic_i/\overline{c}$ term remains, and $A_i\rightarrow 1/L$).
+If $l_i\gg 1$ for all types, we recover the classic lottery model (only the $A_ic_i/\overline{c}$ term remains, and $A_i\rightarrow 1/L$).
+
+Fig.~\ref{fig:simcomp} shows that Eq.
+\eqref{eq:master} and its components closely approximate our density-dependent lottery model over a wide range of propagule densities (the latter is evaluated by direct simulations of uniform random dispersal and lottery competition).
+Two types are present, one of which is at low frequency.
+The growth of the low-frequency type relies crucially on the low-density competition term $R_i c_i/\overline{c}$, and also to a lesser extent on the high density competition term $A_i c_i/\overline{c}$ if $l_1$ is large enough (Fig.~\ref{fig:simcomp}b).
+On the other hand, $R_i c_i/\overline{c}$ is negligible for the high-frequency type, which depends instead on high density territorial victories (Fig.~\ref{fig:simcomp}d).
+Fig.
+3 also shows the breakdown of the classic lottery model at low propagule densities.
+
+\begin{figure}
+\centering
+\includegraphics[scale=0.7]{simulationcomparison.pdf}
+\caption{\label{fig:simcomp} The change in type abundances in a density dependent lottery model is closely approximated by Eq.
+\eqref{eq:master}.
+(a) and (c) per-propagule success probability $\Delta_+ n_i/l_i U$ from Eq.
+\eqref{eq:master}, the classic lottery model and  individual-based simulations of random dispersal and lottery competition.
+(b) and (d) the separate components of Eq.
+\eqref{eq:master}.
+Two types are present.
+(a) and (b) show a low-frequency type with a $c$-advantage ($c_1=1.5$), (c) and (d) show a high-frequency predominant type ($c_2=1$).
+Simulation points are almost invisible in (c) and (d) due to near exact agreement with Eq.
+\eqref{eq:master}.
+Dashed lines in (a) and (c) show the breakdown of the classic lottery model.
+Parameters: $m_1=10^4$ and $m_2=9\times 10^4$ and $U$ is varied between $5\times 10^3$ and $10^6$.}
+\end{figure}
+
+Note that unlike the classic lottery model, not all unoccupied territories are claimed each iteration, since under Poisson dispersal a fraction $e^{-L}$ remain unoccupied.
+
+$\Delta_+ N=U(1-e^{-L})$
 
@@
 For instance, in the Lotka-Volterra competition model we have $f_1(n_1,n_2)=r_1(1-\alpha_{11}n_1-\alpha_{12}n_2)n_1$ where $\alpha_{11}=1/K_{11}$ measures competitive effects within the first type, whereas $\alpha_{12}=1/K_{12}$ measures competitive effects on the first type due to the second.
-
 Thus, when MacArthur concludes that  ``fitness is $K$'' in crowded populations \citep[pp.
@@
 This general idea applies even if the nullclines are nonlinear to such an extent that the ``$K$'' values themselves do not give much information about the regions of high-density growth.
+
+The distinction between
 
@@
 This is a realistic feature of the model.
-Even if one genotype is guaranteed to win a territory in a ``fair'' contest (e.g.
+Even if one type is guaranteed to win a territory in a ``fair'' contest (e.g.
 it is the most efficient exploiter of a limiting consumable resource; \citealt{tilman_1982}), inferior competitors can win by chance.
 For example, an inferior competitor's propagules may happen to arrive first, gaining a decisive developmental advantage.
-First arrivals are more likely to occur for genotypes with a fecundity and/or dispersal advantage, as represented by higher $b$ in lottery models.
+First arrivals are more likely to occur for types with a fecundity and/or dispersal advantage, as represented by higher $b$ in lottery models.
 The analogous intuition in the Wright-Fisher model is that fecundity confers a relative fitness advantage, even though population size is not changing.
@@
 However, competitive success as measured by $\alpha$ (i.e.
-the per-capita effect of one genotype on another genotype's growth rate) is only partly determined by individual competitive ability --- in the presence of age-structured competition and territoriality, it also includes the ability of each genotype to produce contestants i.e.
+the per-capita effect of one type on another type's growth rate) is only partly determined by individual competitive ability --- in the presence of age-structured competition and territoriality, it also includes the ability of each type to produce contestants i.e.
 $b$ in our model.
@@
 $K$-selection in the narrow logistic sense of selection for a greater environmental carrying capacity for given $r$, sometimes referred to as ``efficiency'' \citep{macarthur_1967}, could be represented in our model by smaller individual territorial requirements.
-To a first approximation, two co-occurring genotypes which differ by a small amount in their territorial requirements only should have the same fitness, since the costs or benefits of a change in the amount of unocupied territory is shared equally among genotypes via the propagule density per territory $L$.
-The situation is more complicated when the differences in territorial requirements become large enough that territorial contests can occur on different scales for different genotypes.
+To a first approximation, two co-occurring types which differ by a small amount in their territorial requirements only should have the same fitness, since the costs or benefits of a change in the amount of unocupied territory is shared equally among types via the propagule density per territory $L$.
+The situation is more complicated when the differences in territorial requirements become large enough that territorial contests can occur on different scales for different types.
 We leave these complications for future work.
@@
 The situation is more complicated if the resources are well-mixed, since, in general, resource levels then need to be explicitly tracked.
-It seems plausible that explicit resource tracking may not be necessary when the focus is on the evolution of similar genotypes that use identical resources rather than the stable co-existence of widely differing species with different resource preferences \citep{ram_2016}.
+It seems plausible that explicit resource tracking may not be necessary when the focus is on the evolution of similar types that use identical resources rather than the stable co-existence of widely differing species with different resource preferences \citep{ram_2016}.
 We are not aware of any attempts to delineate conditions under which explicit resource tracking is unnecessary even if it is assumed that community structure is ultimately determined by competition for consumable resources.
@@
 
-While our model can be applied to species rather than genotypes (e.g.
-ecological invasions), our focus is genotype evolution i.e.
+While our model can be applied to species rather than types (e.g.
+ecological invasions), our focus is type evolution i.e.
 the change in allele frequencies over time.
@@
 Since beneficial mutation effect sizes will typically not be much larger than a few percent, large $c$ discrepancies can only arise if the mutation rate is extremely large, and so the assumption will not be violated in most cases.
-However, this restriction could become important when looking at species interactions rather than genotype evolution.
+However, this restriction could become important when looking at species interactions rather than type evolution.
 
@@
 
-In the exact model of random dispersal, the counts of a genotype's propagules across unnocupied territories follows a multinomial distribution with dimension $U$, total number of trials equal to $m_i$, and equal probabilities $1/U$ for a propagule to land in a given territory.
+In the exact model of random dispersal, the counts of a type's propagules across unnocupied territories follows a multinomial distribution with dimension $U$, total number of trials equal to $m_i$, and equal probabilities $1/U$ for a propagule to land in a given territory.
 Thus, the $x_i$ in different territories are not independent random variables.
@@
 However, the relative fluctuations in $\sum_{j\neq i} c_j x_j$ can be large if some of the $c_j$ are much larger than the others.
-Specifically, in the presence of a rare, extremely strong competitor ($c_j l_j\gg c_{j'} l_{j'}$ for all other nonfocal genotypes $j'$, and $l_j\ll 1$), then the RHS of Eq.
+Specifically, in the presence of a rare, extremely strong competitor ($c_j l_j\gg c_{j'} l_{j'}$ for all other nonfocal types $j'$, and $l_j\ll 1$), then the RHS of Eq.
 \eqref{eq:cvr} can be large and we cannot make the replacement Eq.~\eqref{eq:meanfieldr}.
@@
 \end{equation}
-Following a similar procedure as for $\Delta_r n_i$, where the vector of propagule abundances is denoted ${\mathbf x}$, the mean focal genotype abundance is, 
+Following a similar procedure as for $\Delta_r n_i$, where the vector of propagule abundances is denoted ${\mathbf x}$, the mean focal type abundance is, 
 \begin{align}
@@
 \end{align}
-For nonfocal genotypes $j\neq i$, we have
+For nonfocal types $j\neq i$, we have
 \begin{align}
@@
 To calculate the relative fluctuations in $\sum_{j\neq i} c_j x_j$, we use a similar approximation as for $\Delta_r n_i$: $\hat{p}$ is approximated by $\hat{q}$, defined as the ${\mathbf x}$ dispersal probabilities in a territory conditional on $x_i>2$ (that is, treating the $x_j$ as indepenent).
-All covariances between nonfocal genotypes are now zero, so that $\sigma_{\hat{q}}^2(\sum c_j x_j)=\sum c_j^2 \sigma_{\hat{q}}^2(x_j)$, where $\sigma_{\hat{q}}^2(x_j)=l_j$ for $j\neq i$, and
+All covariances between nonfocal types are now zero, so that $\sigma_{\hat{q}}^2(\sum c_j x_j)=\sum c_j^2 \sigma_{\hat{q}}^2(x_j)$, where $\sigma_{\hat{q}}^2(x_j)=l_j$ for $j\neq i$, and
 \begin{equation}
@@
 where $A_i$ is defined in Eq.~\eqref{eq:Da}.
-
-\subsection*{Comparison with simulations}
-
-Fig.~\ref{fig:simcomp} shows that Eq.
-\eqref{eq:master} and its components closely approximate our density-dependent lottery model over a wide range of propagule densities (the latter is evaluated by direct simulations of uniform random dispersal and lottery competition).
-Two genotypes are present, one of which is at low frequency.
-The growth of the low-frequency genotype relies crucially on the low-density competition term $R_i c_i/\overline{c}$, and also to a lesser extent on the high density competition term $A_i c_i/\overline{c}$ if $l_1$ is large enough (Fig.~\ref{fig:simcomp}b).
-On the other hand, $R_i c_i/\overline{c}$ is negligible for the high-frequency genotype, which depends instead on high density territorial victories (Fig.~\ref{fig:simcomp}d).
-Fig.
-3 also shows the breakdown of the classic lottery model at low propagule densities.
-
-\begin{figure}
-\centering
-\includegraphics[scale=0.7]{simulationcomparison.pdf}
-\caption{\label{fig:simcomp} The change in genotype abundances in a density dependent lottery model is closely approximated by Eq.
-\eqref{eq:master}.
-$\Delta_+ n_i/m_i$ from Eq.
-\eqref{eq:master} (and its separate components) are shown, along with direct simulations of random dispersal and lottery competition over one iteration over a range of propagule densities ($U$ is varied between $5\times 10^3$ and $10^6$ with $m_1=10^4$ and $m_2=9\times 10^4$).
-Two genotypes are present.
-(a) and (b) show the low-frequency genotype with $c$-advantage ($c_1=1.5$), (c) and (d) show the high-frequency predominant genotype ($c_2=1$).
-Simulation points are almost invisible in (c) and (d) due to near exact agreement with Eq.
-\eqref{eq:master}.
-Dashed lines in (a) and (c) show the breakdown of the classic lottery model.}
-\end{figure}
 

### 2017-11-22 Kplot.py

@@ -55,7 +55,8 @@ plt.savefig('/home/jbertram/repos/densitydependentlottery/Kplot.pdf',bbox="tight
 
 #===============================================================================
 X, Y = np.meshgrid(x,y)
-alphas=np.array([[1,1.5],[1.5,1]])
+#alphas=np.array([[1,1.5],[1.5,1]])
+alphas=np.array([[1,0.8],[1.2,1]])
 U = 2*(1-alphas[0,0]*X-alphas[0,1]*Y)*X
 V = (1-alphas[1,1]*Y-alphas[1,0]*X)*Y
 #speed = np.sqrt(U*U + V*V)

### 2017-11-22 simulations.py

@@ -83,47 +83,41 @@ m=np.array([10000,90000])
 M=float(sum(m)) 
 c=np.array([1.5,1.])
 cbar=sum(m*c)/sum(m)
-maxl=2; Us=np.array([int(m[0]/x) for x in np.linspace(0.001,maxl,100)]);
+maxl=2; Us=np.array([int(m[0]/x) for x in np.linspace(0.001,maxl,1000)]);
 exact=np.array([deltnplussim(m,c,U) for U in Us]);
 totalclassic=np.array([deltnplusclassic(m,c,U) for U in Us]);
 totalclassicboundedm=np.array([deltnplusclassicboundedm(m,c,U) for U in Us]);
+
+fig1, (ax1, ax2) = plt.subplots(nrows=2,figsize=[4,8])
 
 #First genotype
 nocomp=np.array([np.exp(-M/U) for U in Us]);
 comp1=np.array([R(m,c,U)[0]*c[0]/cbar for U in Us]);
 comp2=np.array([A(m,c,U)[0]*c[0]/cbar for U in Us]);
 
-plt.subplot(221)
-plt.plot(float(m[0])/Us,(exact[:,0,0]+exact[:,1,0]+exact[:,2,0])/m[0],'k.',markersize=1.,label="Simulation")
-plt.plot(float(m[0])/Us,(nocomp+comp1+comp2),'k',label=r"$\Delta_+ n_1/m_1$")
-plt.plot(float(m[0])/Us,totalclassic[:,0]/m[0],'k--',label=r"Classic lottery")
-#plt.plot(float(m[0])/Us,totalclassicboundedm[:,0]/m[0],'kx',label=r"Classic lottery")
-plt.xlim([0,maxl])
-plt.ylim([0,1.1])
-plt.gca().xaxis.set_label_coords(0.5, -0.09)
-plt.gca().set_xticklabels(['0','','1','','2','','3'])
-plt.xlabel(r"$l_1$",fontsize=14)
-plt.ylabel(r"$\Delta_+ n_1/m_1$",fontsize=14)
+ax1.plot(float(m[0])/Us,(exact[:,0,0]+exact[:,1,0]+exact[:,2,0])/m[0],'k.',markersize=1.,label="Simulation")
+ax1.plot(float(m[0])/Us,(nocomp+comp1+comp2),'k',label=r"Eq. (5)")
[…]

### 2017-11-22 manuscript.tex
l

@@
 If $l_i\gg 1$ for all types, we recover the classic lottery model (only the $A_ic_i/\overline{c}$ term remains, and $A_i\rightarrow 1/L$).
+Note that not all unoccupied territories are claimed each iteration, since under Poisson dispersal a fraction $e^{-L}$ remain unoccupied; the total number of territories gained is thus $\Delta_+ N=U(1-e^{-L})$.
 
 Fig.~\ref{fig:simcomp} shows that Eq.
-\eqref{eq:master} and its components closely approximate our density-dependent lottery model over a wide range of propagule densities (the latter is evaluated by direct simulations of uniform random dispersal and lottery competition).
+\eqref{eq:master} and its components closely approximate individual-based simulations of the density-dependent lottery model over a wide range of propagule densities $l_i$.
 Two types are present, one of which is at low frequency.
-The growth of the low-frequency type relies crucially on the low-density competition term $R_i c_i/\overline{c}$, and also to a lesser extent on the high density competition term $A_i c_i/\overline{c}$ if $l_1$ is large enough (Fig.~\ref{fig:simcomp}b).
-On the other hand, $R_i c_i/\overline{c}$ is negligible for the high-frequency type, which depends instead on high density territorial victories (Fig.~\ref{fig:simcomp}d).
-Fig.
-3 also shows the breakdown of the classic lottery model at low propagule densities.
+The growth of the low-frequency type relies crucially on the low-density competition term $R_i c_i/\overline{c}$.
+On the other hand, $R_i c_i/\overline{c}$ is negligible for the high-frequency type, which depends instead on high density territorial victories.
+Fig.~\ref{fig:simcomp} also shows the breakdown of the classic lottery model at low propagule densities.
 
@@
 \centering
-\includegraphics[scale=0.7]{simulationcomparison.pdf}
-\caption{\label{fig:simcomp} The change in type abundances in a density dependent lottery model is closely approximated by Eq.
+\includegraphics[scale=0.8]{simulationcomparison.pdf}
+\caption{\label{fig:simcomp} Comparison of mean field approximation Eq.
+\eqref{eq:master} with simulations.
+Per-propagule success probability $\Delta_+ n_i/l_i U$ from Eq.
+\eqref{eq:master} and its three components, the classic lottery model, and  individual-based simulations of random dispersal and lottery competition.
+Two types are present, a rare type with $c_1=1.5$, and a common type with $c_2=1$.
+Simulation points are almost invisible in for the common type due to near exact agreement with Eq.
 \eqref{eq:master}.
-(a) and (c) per-propagule success probability $\Delta_+ n_i/l_i U$ from Eq.
-\eqref{eq:master}, the classic lottery model and  individual-based simulations of random dispersal and lottery competition.
-(b) and (d) the separate components of Eq.
-\eqref{eq:master}.
-Two types are present.
-(a) and (b) show a low-frequency type with a $c$-advantage ($c_1=1.5$), (c) and (d) show a high-frequency predominant type ($c_2=1$).
-Simulation points are almost invisible in (c) and (d) due to near exact agreement with Eq.
-\eqref{eq:master}.
-Dashed lines in (a) and (c) show the breakdown of the classic lottery model.
+Dashed lines in show the breakdown of the classic lottery model.
 Parameters: $m_1=10^4$ and $m_2=9\times 10^4$ and $U$ is varied between $5\times 10^3$ and $10^6$.}
 \end{figure}
-
-Note that unlike the classic lottery model, not all unoccupied territories are claimed each iteration, since under Poisson dispersal a fraction $e^{-L}$ remain unoccupied.
-
-$\Delta_+ N=U(1-e^{-L})$
 

### 2017-11-22 Kplot.py

@@ -99,8 +99,6 @@ ax2.set_yticklabels([])
 ax2.set_xlabel(r"$n_1$",fontsize=20)
 ax2.set_ylabel(r"$n_2$",fontsize=20)
 
-plt.savefig('/home/jbertram/repos/densitydependentlottery/Kplot.pdf',bbox="tight")
-
 
 #fig1, ax1 = plt.subplots()
 #ax1.streamplot(X, Y, U, V,density=0.5)

### 2017-11-22 simulations.py

@@ -88,71 +88,62 @@ exact=np.array([deltnplussim(m,c,U) for U in Us]);
 totalclassic=np.array([deltnplusclassic(m,c,U) for U in Us]);
 totalclassicboundedm=np.array([deltnplusclassicboundedm(m,c,U) for U in Us]);
 
-fig1, (ax1, ax2) = plt.subplots(nrows=2,figsize=[4,8])
+fig1, (ax1, ax2) = plt.subplots(ncols=2,figsize=[8,4])
 
 #First genotype
 nocomp=np.array([np.exp(-M/U) for U in Us]);
 comp1=np.array([R(m,c,U)[0]*c[0]/cbar for U in Us]);
 comp2=np.array([A(m,c,U)[0]*c[0]/cbar for U in Us]);
 
+ax1.plot(float(m[0])/Us,nocomp,'k',label=r"$e^{-L}$")
+ax1.plot(float(m[0])/Us,comp1,'r',label=r"$R_i \frac{c_i}{\overline{c}}$")
+ax1.plot(float(m[0])/Us,comp2,'b',label=r"$A_i \frac{c_i}{\overline{c}}$")
+ax1.plot(float(m[0])/Us,(nocomp+comp1+comp2),'k',label=r"$e^{-L}+R_i \frac{c_i}{\overline{c}}+A_i \frac{c_i}{\overline{c}}$",linewidth=2)
+
 ax1.plot(float(m[0])/Us,(exact[:,0,0]+exact[:,1,0]+exact[:,2,0])/m[0],'k.',markersize=1.,label="Simulation")
-ax1.plot(float(m[0])/Us,(nocomp+comp1+comp2),'k',label=r"Eq. (5)")
-ax1.plot(float(m[0])/Us,totalclassic[:,0]/m[0],'k--',label=r"Classic lottery")
-ax1.set_xlim([0,maxl])
-ax1.set_ylim([0,1.1])
-ax1.xaxis.set_label_coords(0.5, -0.09)
-ax1.set_xticklabels(['0','','1','','2','','3'])
-ax1.set_xlabel(r"$l_1$",fontsize=14)
-ax1.set_ylabel("Per propagule success",fontsize=14)
-
-ax1.annotate(r'$(a)$',xy=(0.9,0.9),xycoords='axes fraction',fontsize=12)
-ax1.legend(loc='upper center',prop={'size':10})
-
[…]

### 2017-11-23 manuscript.tex
l

@@
 
-\title{Density-dependent selection with strictly relative competition}
+\title{Density-dependent selection and the population ecology of relative fitness}
 %\author{Jason Bertram $^{1,\ast}$ \\ 
@@
 
-\textit{Keywords}: r/K selection, absolute fitness, eco-evo, competition-colonization trade-off, fluctuating selection, storage effect.
+\textit{Keywords}: $r$/$K$ selection, absolute fitness, eco-evo, competition-colonization trade-off, fluctuating selection, storage effect.
 
@@
 This general idea applies even if the nullclines are nonlinear to such an extent that the ``$K$'' values themselves do not give much information about the regions of high-density growth.
+
+\begin{figure}
+\centering
+\includegraphics[scale=0.8]{LVvslottery.pdf}
+\caption{\label{fig:Ksel} Blah}
+\end{figure}
 

### 2017-11-23 Kplot.py

@@ -25,8 +25,8 @@ ax1.streamplot(x, y, U, V,density=100,start_points=seedpoints,linewidth=1.)
 
 z1=np.linspace(0,1,10)
 z2=1-z1
-ax1.plot(z1,(1-alphas[0,0]*z1)/alphas[0,1],'k',linewidth=2)
-ax1.plot(z1,(1-alphas[1,0]*z1)/alphas[1,1],'k',linewidth=2)
+ax1.plot(z1,(1-alphas[0,0]*z1)/alphas[0,1],'k--',linewidth=2)
+ax1.plot(z1,(1-alphas[1,0]*z1)/alphas[1,1],'k--',linewidth=2)
 #ax1.plot([0,1/alphas[0,0]],[1/alphas[1,1],0],'y',linewidth=2)
 ax1.set_xlim([0,1])
 ax1.set_ylim([0,1])

### 2017-11-23 simulations.py

@@ -151,7 +151,7 @@ T=100000
 x,y=np.linspace(0.0,1.,401),np.linspace(0.,1.,401)
 xgrid, ygrid=np.meshgrid(x,y)
 X,Y=np.zeros([len(x),len(y)]),np.zeros([len(x),len(y)])
-b=np.array([1.,1.])
+b=np.array([5.,5.])
 c=np.array([5.,1.])
 d=np.array([0.1,0.1])
 
@@ -159,60 +159,58 @@ for i in xrange(len(x)):
     for j in xrange(len(y)):
         n=T*np.array([x[i],y[j]])
         U=T-sum(n)
-        X[j,i],Y[j,i]=deltnplus(b*n,c,U)-d*n
+        X[j,i],Y[j,i]=deltnplus(b*n*U/float(T),c,U)-d*n
 
 
 fig1, (ax2, ax1) = plt.subplots(ncols=2,figsize=[8,4])
-ax1.set_aspect(1)
-ax2.set_aspect(1)
 
-ax1.plot([f2null[-1,0],f1null[0,0]],[f2null[-1,1],f1null[0,1]],'y',linewidth=2)
+f1null=np.array([[x[[_ for _ in range(len(x)) if X[i,_]>0][-1]],y[i]] for i in range(1,len(y)) if max(X[i])>0])
+f2null=np.array([[x[[_ for _ in range(len(x)) if Y[i,_]>0][-1]],y[i]] for i in range(1,len(y)) if max(Y[i])>0])
+ax1.plot(f1null[:,0],f1null[:,1],'k--',linewidth=2.)
+ax1.plot(f2null[:,0],f2null[:,1],'k--',linewidth=2.)
+#plt.plot([f2null[0,0],f2null[-1,0]],[f2null[0,1],f2null[-1,1]],'y',linewidth=2)
+ax1.plot([f2null[-1,0],f1null[0,0]],[f2null[-1,1],f1null[0,1]],'k',linewidth=2,alpha=0.8)
+ax1.set_xlim([0,f1null[0,0]])
+ax1.set_ylim([0,f1null[-1,1]])
+
+ax1.annotate(r'$(b)$',xy=(0.9,0.93),xycoords='axes fraction',fontsize=16)
 
 seedpoints=[[x[_]/2,y[200-_]/2] for _ in range(0,201,10)]
 seedpoints=np.concatenate([[[_,0.2-_] for _ in np.arange(0.02,0.2,0.02)],\
[…]

### 2017-11-28 manuscript.tex
l

@@
 
-The issue can be expressed more formally by revisiting MacArthur's famous analysis of selection in crowded populations \citep{macarthur_1967}.
+The issue can be expressed more formally by revisiting MacArthur's analysis of selection in crowded populations \citep{macarthur_1967}.
 MacArthur considers a population consisting of two types with densities $n_1$ and $n_2$ subject to density-dependent population growth described by
@@
 \end{equation}
-Other than the type densities, the environment is assumed to remain constant.
+Other than densities, the environment is assumed to remain constant.
 The functions $f_1$ and $f_2$ must decline to zero if $n_1$ or $n_2$ are sufficiently large, because no population has unlimited resources.
@@
 
-At face value, the preceding conclusion does not seem compatible with the constant $N$ assumption of relative fitness models.
-It is not even clear how to connect relative fitness models and MacArthur's argument, since the former only allows for selection along lines defined by $n_1+n_2=N$ for each $N$ (Fig.~\ref{fig:Ksel}b); the $f_1$ and $f_2$ nullclines are thus not even defined.
-The incompatibility is particularly apparent in some of the most widely-used models of population growth.
-MacArthur's argument is often presented using the logistic model and its bi-allelic diploid variants [citations].
-Positive selection in crowded populations then requires increases in equilibrium population density --- commonly referred to as ``$K$-selection'' (Fig.~\ref{fig:Ksel}a).
-The Lotka-Volterra competition model also couples selection in crowded populations to changes in total density $N$ (we return to this in BLAH).
-Moreover, exploitation competition for consumable resources is potentially a decisive factor in many crowded populations.
-On a single, homogeneous, limiting resource, the type able to deplete the resource to the lowest equilibrium density excludes the other ($R^*$-rule); except in very special cases, this will also entail differences in saturation density.
+This conclusion about selection in crowded populations does not seem compatible with the constant-$N$ assumption of relative fitness models.
+The incompatibility is particularly apparent in some of the most widely-used models of density-regulated population growth (to illustrate our point we stick to the simple haploid case): 1) The logistic model: the type with the greatest saturation population density excludes the others.
+2) The ``$R^*$ rule'', a central tenet of resource competition theory, states that when multiple types are exploiting a homogeneous consumable resource that limits their growth, the type able to deplete the resource to the lowest equilibrium density $R^*$ excludes the others \citep{grover_1997}.
+Except in special cases, differences in $R^*$ entails differences in saturation population density.
+3) The Lotka-Volterra competition model also couples selection in crowded populations to changes in total density $N$ (we return to this in BLAH).
 Thus, it would seem that the ubiquitous constant-$N$, relative fitness description of selection is incompatible with a huge class of population ecological processes occurring in nature and experiments.
+
+It is not even clear how to connect relative fitness models to density-regulated growth, since they only describe selection along lines defined by $n_1+n_2=N$ for each $N$ (Fig.~\ref{fig:Ksel}b); the $f_1$ and $f_2$ nullclines are thus not defined.
 
@@
 \includegraphics[scale=0.8]{Kplot.pdf}
-\caption{\label{fig:Ksel} (a) MacArthur's dynamical argument for why ``fitness is $K$'' in crowded environments, illustrated using the logistic model $f_1=r_1(1-\frac{n_1+n_2}{K_1})n_1$ and $f_2=r_2(1-\frac{n_1+n_2}{K_2})n_1$ in Eq.
+\caption{\label{fig:Ksel} (a) MacArthur's dynamical argument for how selection operates in crowded environments, illustrated using the logistic model $f_1=r_1(1-\frac{n_1+n_2}{K_1})n_1$ and $f_2=r_2(1-\frac{n_1+n_2}{K_2})n_1$ in Eq.
 \eqref{eq:macgeneral}.
@@
 
-In our extension of the classic lottery model (Eq.~\ref{eq:delttot}), we do not restrict ourselves to high propagule densities.
 Eq.
-\eqref{eq:lottery} is nonsensical if even a single type has low propagule density ($l_i\ll 1$): type $i$ can win at most $m_i$ territories, yet Eq.~\eqref{eq:lottery} demands $c_i l_i/\sum_j c_j l_j$ of the $U$ unoccupied territories, for any value of $U$.
-Thus, deviations from the mean propagule composition $l_1,l_2,\ldots,l_G$ are important at low density.
-The reason for this lies in the fundamental discreteness of biological individuals: types with few propagules depend on the outcome of contests in territories where they have at least one propagule present, not some small fraction of a propagule as would be implied by small $l_i$ in Eq.~\eqref{eq:lottery}.
+\eqref{eq:lottery} breaks down for types with low propagule density ($l_i\ll 1$) because territorial acquisition is then not correctly represented by a lottery in each territory with the mean propagule density.
+Instead, a rare type's propagules only make it to a few territories where at least one of its propagule present.
+In our extension of the classic lottery model, we correct (Eq.~\ref{eq:delttot}) to account for this.
+
+%type $i$ can win at most $m_i$ territories, yet Eq.~\eqref{eq:lottery} demands $c_i l_i/\sum_j c_j l_j$ of the $U$ unoccupied territories, for any value of $U$.
 
@@
 The second, $R_i c_i/\overline{c}$ represents competitive victories when the $i$ type is a rare invader in a high density population, determining its invasion fitness \citep{metz_1992}.
-The third term, $A_ic_i/\overline{c}$, represents competitive victories when the $i$ type is abundant.
+The third term, $A_i c_i/\overline{c}$, represents competitive victories when the $i$ type is abundant.
 The relative importance of these three terms varies with both the overall propagule density $L$ and the relative propagule frequencies $m_i/M$.
@@
 \eqref{eq:master} with simulations.
-Per-propagule success probability $\Delta_+ n_i/l_i U$ from Eq.
-\eqref{eq:master} and its three components, the classic lottery model, and  individual-based simulations of random dispersal and lottery competition.
+Per-propagule success probability $\Delta_+ n_i/l_i U$ from the classic lottery model, individual-based simulations of random dispersal and lottery competition, and Eq.
+\eqref{eq:master} and its three components.
 Two types are present, a rare type with $c_1=1.5$, and a common type with $c_2=1$.
@@
 
+--- commonly referred to as ``$K$-selection'' (Fig.~\ref{fig:Ksel}a)
+
 As shown in the Introduction, MacArthur's argument revolves around the behaviour of  ecological models of the general form Eq.~\ref{eq:macgeneral}, which depends on the relationship between the nullclines $f_1(n_1,n_2)=0$ and $f_2(n_1,n_2)=0$.
@@
 These $K$ values determine whether a region of higher-density growth exists for each type, provided that the nullclines are close to being straight lines.
-Note that, while $K_{11}$ and $K_{22}$ are saturation densities akin to the $K$ parameter in the logistic model, this is not true for $K_{12}$ and $K_{21}$, which are related to competition between types.
-For instance, in the Lotka-Volterra competition model we have $f_1(n_1,n_2)=r_1(1-\alpha_{11}n_1-\alpha_{12}n_2)n_1$ where $\alpha_{11}=1/K_{11}$ measures competitive effects within the first type, whereas $\alpha_{12}=1/K_{12}$ measures competitive effects on the first type due to the second.
+Note that only $K_{11}$ and $K_{22}$ are saturation densities akin to the $K$ parameter in the logistic model; $K_{12}$ and $K_{21}$ are related to competition between types.
+For instance, in the Lotka-Volterra competition model we have $f_1(n_1,n_2)=r_1(1-\alpha_{11}n_1-\alpha_{12}n_2)n_1$ where $\alpha_{11}=1/K_{11}$ measures competitive effects within the first type, whereas $\alpha_{12}=1/K_{12}$ measures competitive effects on the first type due to the second (Fig.~\ref{fig:LVvslottery}a).
 Thus, when MacArthur concludes that  ``fitness is $K$'' in crowded populations \citep[pp.
-149]{macarthur_1967}, the meaning is that selection in crowded populations either favors the ability to keep growing at ever higher densities (moving a type's own nullcline outwards), or the ability to suppress the growth of competitors at lower densities (moving the nullcline of competitors inwards).
+149]{macarthur_1967}, the meaning is that selection either favors the ability to keep growing at ever higher densities (moving a type's own nullcline outwards), or the ability to suppress the growth of competitors at lower densities (moving the nullcline of competitors inwards).
 This general idea applies even if the nullclines are nonlinear to such an extent that the ``$K$'' values themselves do not give much information about the regions of high-density growth.
@@
 \includegraphics[scale=0.8]{LVvslottery.pdf}
-\caption{\label{fig:Ksel} Blah}
+\caption{\label{fig:LVvslottery} Blah}
 \end{figure}
 
-The distinction between
-
-The nullclines of Eq.~\eqref{eq:master}
+In Fig.~\ref{fig:LVvslottery} we show a phase diagram for Eq.~\eqref{eq:master}
 

### 2017-11-28 manuscript.tex
l

@@
 \end{equation}
-Other than densities, the environment is assumed to remain constant.
+Apart from the type densities, the environment is assumed to remain constant.
 The functions $f_1$ and $f_2$ must decline to zero if $n_1$ or $n_2$ are sufficiently large, because no population has unlimited resources.
@@
 This conclusion about selection in crowded populations does not seem compatible with the constant-$N$ assumption of relative fitness models.
-The incompatibility is particularly apparent in some of the most widely-used models of density-regulated population growth (to illustrate our point we stick to the simple haploid case): 1) The logistic model: the type with the greatest saturation population density excludes the others.
+Some of the most widely-used models of density-regulated population growth are indeed not compatible: 1) In the logistic model, the type with the greatest saturation population density excludes the others.
 2) The ``$R^*$ rule'', a central tenet of resource competition theory, states that when multiple types are exploiting a homogeneous consumable resource that limits their growth, the type able to deplete the resource to the lowest equilibrium density $R^*$ excludes the others \citep{grover_1997}.
-Except in special cases, differences in $R^*$ entails differences in saturation population density.
+Differences in $R^*$ will often entail differences in saturation density.
 3) The Lotka-Volterra competition model also couples selection in crowded populations to changes in total density $N$ (we return to this in BLAH).
-Thus, it would seem that the ubiquitous constant-$N$, relative fitness description of selection is incompatible with a huge class of population ecological processes occurring in nature and experiments.
-
-It is not even clear how to connect relative fitness models to density-regulated growth, since they only describe selection along lines defined by $n_1+n_2=N$ for each $N$ (Fig.~\ref{fig:Ksel}b); the $f_1$ and $f_2$ nullclines are thus not defined.
+It would therefore seem that the ubiquitous constant-$N$, relative fitness description of selection is incompatible with a huge class of population ecological processes occurring in nature and experiments.
+It is not even clear how relative fitness models connect to general ecological models of the form Eq.~\eqref{eq:macgeneral}, since the former only describe selection along lines defined by $n_1+n_2=N$ for each $N$ (Fig.~\ref{fig:Ksel}b); the $f_1$ and $f_2$ nullclines are thus not defined.
 
@@
 
-The constant-$N$, relative fitness description has historically been justified as a short-term linear approximation of a more complicated long-term eco-evolutionary process \cite[pp.
+The constant-$N$, relative fitness description has historically been justified as a short-term linear approximation \cite[pp.
 277]{ewens_2004}.
@@
 Provided that selection is sufficiently weak and stable over time, this ``short-term''  assumption is not a major restriction.
-Yet it is increasingly recognized that selection is not always weak, that it can fluctuate dramatically over time, and coupled to this, that a population's density can vary by orders of magnitude over a few generations as a routine feature of its ecology.
+Yet it is increasingly recognized that selection is not always weak, that it can fluctuate considerably over time, and that $N$ can vary by orders of magnitude over a few generations as a routine feature of a population's ecology.
 These are not rare exceptions, but occur widely in nature and the lab, including in wild \textit{Drosophila} \citep{messer_2016}.
-Moreover, the short-term approximation automatically precludes consideration of inherently long-term evolutionary processes like the management of genetic load and population extinction \citep{bertram_2017}.
+The short-term approximation also precludes consideration of inherently long-term evolutionary processes like the management of genetic load and population extinction \citep{bertram_2017}.
 Nevertheless, relative fitness models like Wright-Fisher are the foundation for much of the population genetic literature, and are still widely used without considering the ``short-term'' restriction.
@@
 
---- commonly referred to as ``$K$-selection'' (Fig.~\ref{fig:Ksel}a)
-
-As shown in the Introduction, MacArthur's argument revolves around the behaviour of  ecological models of the general form Eq.~\ref{eq:macgeneral}, which depends on the relationship between the nullclines $f_1(n_1,n_2)=0$ and $f_2(n_1,n_2)=0$.
+As shown in the Introduction, MacArthur's argument revolves around the behaviour of  ecological models of the general form Eq.~\ref{eq:macgeneral}, behaviour which depends on the relationship between the nullclines $f_1(n_1,n_2)=0$ and $f_2(n_1,n_2)=0$.
 To formalize this relationship, MacArthur used the symbol ``$K$'' to label the four intersection points of the nullclines with the $n_1$ and $n_2$ axes, specifically $f_1(K_{11},0)=0$, $f_1(0,K_{12})=0$, $f_2(0,K_{22})=0$ and $f_2(K_{21},0)=0$.
 These $K$ values determine whether a region of higher-density growth exists for each type, provided that the nullclines are close to being straight lines.
-Note that only $K_{11}$ and $K_{22}$ are saturation densities akin to the $K$ parameter in the logistic model; $K_{12}$ and $K_{21}$ are related to competition between types.
+Note that only $K_{11}$ and $K_{22}$ are saturation densities akin to the $K$ parameter in the logistic model; selection on these saturation densities is commonly referred to as ``$K$-selection'' (Fig.~\ref{fig:Ksel}a).
+The other intersection points, $K_{12}$ and $K_{21}$, are related to competition between types.
 For instance, in the Lotka-Volterra competition model we have $f_1(n_1,n_2)=r_1(1-\alpha_{11}n_1-\alpha_{12}n_2)n_1$ where $\alpha_{11}=1/K_{11}$ measures competitive effects within the first type, whereas $\alpha_{12}=1/K_{12}$ measures competitive effects on the first type due to the second (Fig.~\ref{fig:LVvslottery}a).
+
 Thus, when MacArthur concludes that  ``fitness is $K$'' in crowded populations \citep[pp.
-149]{macarthur_1967}, the meaning is that selection either favors the ability to keep growing at ever higher densities (moving a type's own nullcline outwards), or the ability to suppress the growth of competitors at lower densities (moving the nullcline of competitors inwards).
+149]{macarthur_1967}, it does not imply a dichotomy between $r$-selection (uncrowded) and $K$-selection (crowded) in the sense of selection for greater saturation density.
+MacArthur's argument shows that selection either favors the ability to keep growing at ever higher densities (moving a type's own nullcline outwards), or the ability to suppress the growth of competitors at lower densities (moving the nullcline of competitors inwards).
 This general idea applies even if the nullclines are nonlinear to such an extent that the ``$K$'' values themselves do not give much information about the regions of high-density growth.
@@
 \includegraphics[scale=0.8]{LVvslottery.pdf}
-\caption{\label{fig:LVvslottery} Blah}
+\caption{\label{fig:LVvslottery} (a) Lotka-Volt}
 \end{figure}
 
-In Fig.~\ref{fig:LVvslottery} we show a phase diagram for Eq.~\eqref{eq:master}
+Fig.~\ref{fig:LVvslottery} shows two examples of selection when both of two types have the same saturation density $N=K_{11}=K_{22}$.
+The first shows exlusion by a superior Lotka-Volterra competitor.
+Note that the trajectories do not follow a line of constant  $N$.
+The reason for this is that inter-type competition must be stronger than intra-type competition for exclusion to occur, which causes a dip in $N$ over a selective sweep (except for the special case  .
+
+$c$-selection in the density-dependent lottery model, Eq.~\eqref{eq:master}.
 

### 2017-11-29 manuscript.tex
l

@@
 
-This conclusion about selection in crowded populations does not seem compatible with the constant-$N$ assumption of relative fitness models.
-Some of the most widely-used models of density-regulated population growth are indeed not compatible: 1) In the logistic model, the type with the greatest saturation population density excludes the others.
-2) The ``$R^*$ rule'', a central tenet of resource competition theory, states that when multiple types are exploiting a homogeneous consumable resource that limits their growth, the type able to deplete the resource to the lowest equilibrium density $R^*$ excludes the others \citep{grover_1997}.
+The connection between constant-$N$ relative fitness models and the broad class of density-regulated growth models considered by MacArthur (Eq.~\ref{eq:macgeneral}) is not clear, since the former only describe selection along lines defined by $n_1+n_2=N$ for each $N$ (Fig.~\ref{fig:Ksel}b).
+In particular, the $f_1$ and $f_2$ nullclines are not even defined.
+It is clear, however, that constant-$N$ relative fitness models are not compatible with some of the most widely-used models of density-regulated population growth: 1) In the logistic model, the type with the greatest saturation population density excludes the others.
+2) The ``$R^*$ rule'', a central tenet of resource competition theory, states that when growth is limited by a single homogeneous consumable resource, the type able to deplete the resource to the lowest equilibrium density $R^*$ excludes the others \citep{grover_1997}.
 Differences in $R^*$ will often entail differences in saturation density.
-3) The Lotka-Volterra competition model also couples selection in crowded populations to changes in total density $N$ (we return to this in BLAH).
+3) The Lotka-Volterra competition model also couples selection in crowded populations to changes in total density $N$ (we return to this in section ``$K$-selection, $c$-selection and relative fitness'').
 It would therefore seem that the ubiquitous constant-$N$, relative fitness description of selection is incompatible with a huge class of population ecological processes occurring in nature and experiments.
-It is not even clear how relative fitness models connect to general ecological models of the form Eq.~\eqref{eq:macgeneral}, since the former only describe selection along lines defined by $n_1+n_2=N$ for each $N$ (Fig.~\ref{fig:Ksel}b); the $f_1$ and $f_2$ nullclines are thus not defined.
 
@@
 \begin{equation}
-R_i=\frac{\overline{c}e^{-l_i}(1-e^{-(L-l_i)})}{c_i +\frac{\overline{c}L- c_il_i}{L-l_i}\frac{L-1+e^{-L}}{1-(1+L)e^{-L}}},\label{eq:Dr}
+R_i=\frac{\overline{c}e^{-l_i}(1-e^{-(L-l_i)})}{c_i +\frac{\overline{c}L- c_il_i}{L-l_i}\frac{L-1+e^{-L}}{1-(1+L)e^{-L}}},\nonumber \label{eq:Dr}
 \end{equation}
@@
 \begin{equation}
-A_i=\frac{\overline{c}(1-e^{-l_i})}{\frac{1-e^{-l_i}}{1-(1+l_i)e^{-l_i}}c_il_i+\frac{\overline{c}L- c_il_i}{L-l_i}\left(L\frac{1-e^{-L}}{1-(1+L)e^{-L}}-l_i\frac{1-e^{-l_i}}{1-(1+l_i)e^{-l_i}}\right)}.\label{eq:Da}
+A_i=\frac{\overline{c}(1-e^{-l_i})}{\frac{1-e^{-l_i}}{1-(1+l_i)e^{-l_i}}c_il_i+\frac{\overline{c}L- c_il_i}{L-l_i}\left(L\frac{1-e^{-L}}{1-(1+L)e^{-L}}-l_i\frac{1-e^{-l_i}}{1-(1+l_i)e^{-l_i}}\right)}.
+\nonumber \label{eq:Da}
 \end{equation}
@@
 If $l_i\gg 1$ for all types, we recover the classic lottery model (only the $A_ic_i/\overline{c}$ term remains, and $A_i\rightarrow 1/L$).
-Note that not all unoccupied territories are claimed each iteration, since under Poisson dispersal a fraction $e^{-L}$ remain unoccupied; the total number of territories gained is thus $\Delta_+ N=U(1-e^{-L})$.
+Note that not all unoccupied territories are claimed each iteration, since under Poisson dispersal a fraction $e^{-L}$ remain unoccupied; total population density thus obeys
+\begin{equation}
+\Delta N=(1-e^{-L})U-\sum_i d_i n_i \label{eq:Nmaster}
+\end{equation}
 
 Fig.~\ref{fig:simcomp} shows that Eq.
-\eqref{eq:master} and its components closely approximate individual-based simulations of the density-dependent lottery model over a wide range of propagule densities $l_i$.
+\eqref{eq:master} and its components closely approximate simulations of the density-dependent lottery model over a wide range of propagule densities.
 Two types are present, one of which is at low frequency.
@@
 Dashed lines in show the breakdown of the classic lottery model.
-Parameters: $m_1=10^4$ and $m_2=9\times 10^4$ and $U$ is varied between $5\times 10^3$ and $10^6$.}
+Parameters: $m_1=10^4$ and $m_2=9\times 10^4$ and $U$ varies between $5\times 10^3$ and $10^6$.}
 \end{figure}
@@
 
-We now compare MacArthur's claims about selection in crowded environments with our density-dependent lottery model.
+We now compare the density-dependent lottery model from the previous section with MacArthur's claims about selection in crowded environments.
 
@@
 These $K$ values determine whether a region of higher-density growth exists for each type, provided that the nullclines are close to being straight lines.
-Note that only $K_{11}$ and $K_{22}$ are saturation densities akin to the $K$ parameter in the logistic model; selection on these saturation densities is commonly referred to as ``$K$-selection'' (Fig.~\ref{fig:Ksel}a).
+Note that only $K_{11}$ and $K_{22}$ are saturation densities akin to the $K$ parameter in the logistic model; following widespread convention, we will refer to selection on these saturation densities as ``$K$-selection'' (Fig.~\ref{fig:Ksel}a).
 The other intersection points, $K_{12}$ and $K_{21}$, are related to competition between types.
-For instance, in the Lotka-Volterra competition model we have $f_1(n_1,n_2)=r_1(1-\alpha_{11}n_1-\alpha_{12}n_2)n_1$ where $\alpha_{11}=1/K_{11}$ measures competitive effects within the first type, whereas $\alpha_{12}=1/K_{12}$ measures competitive effects on the first type due to the second (Fig.~\ref{fig:LVvslottery}a).
+For instance, in the Lotka-Volterra competition model we have
+\begin{align}
+\frac{d n_1}{dt} = r_1(1-\alpha_{11}n_1-\alpha_{12}n_2)n_1 = f_1(n_1,n_2)\nonumber\\
+\frac{d n_2}{dt} = r_2(1-\alpha_{22}n_1-\alpha_{21}n_2)n_2 = f_2(n_1,n_2)\label{eq:LV}
+\end{align}
+where $\alpha_{11}=1/K_{11}$ and $\alpha_{22}=1/K_{22}$ measure competitive effects within each type, while $\alpha_{12}=1/K_{12}$ and $\alpha_{21}=1/K_{21}$ measure competitive effects on the first type due to the second (Fig.~\ref{fig:LVvslottery}a).
 
@@
 149]{macarthur_1967}, it does not imply a dichotomy between $r$-selection (uncrowded) and $K$-selection (crowded) in the sense of selection for greater saturation density.
-MacArthur's argument shows that selection either favors the ability to keep growing at ever higher densities (moving a type's own nullcline outwards), or the ability to suppress the growth of competitors at lower densities (moving the nullcline of competitors inwards).
+MacArthur's argument shows that selection either favors the ability to keep growing at ever higher densities (moving a type's own nullcline outwards), or the ability to suppress the growth of competitors at lower densities (moving the nullcline of competitors inwards) \citep{gill_1974}.
 This general idea applies even if the nullclines are nonlinear to such an extent that the ``$K$'' values themselves do not give much information about the regions of high-density growth.
@@
 \includegraphics[scale=0.8]{LVvslottery.pdf}
-\caption{\label{fig:LVvslottery} (a) Lotka-Volt}
+\caption{\label{fig:LVvslottery} Selection between types with identical saturation density but different inter-type competitive ability.
+(a) Lotka-Volterra competition (Eq.~\ref{eq:LV}) with $r_1=r_2=1$, $\alpha_{11}=\alpha_{22}=1$, $\alpha_{12}=0.9$ and $\alpha_{21}=1.2$.
+Trajectories do not follow the line $N=K_{11}=K_{22}$.
+(b) Lottery competition (Eq.~\ref{eq:master}) with $b_1=b_2=5$, $d_1=d_2=0.1$ and $c_1/c_2=5$.
+Trajectories converge on the line $N=K_{11}=K_{22}$.}
 \end{figure}
 
-Fig.~\ref{fig:LVvslottery} shows two examples of selection when both of two types have the same saturation density $N=K_{11}=K_{22}$.
-The first shows exlusion by a superior Lotka-Volterra competitor.
-Note that the trajectories do not follow a line of constant  $N$.
-The reason for this is that inter-type competition must be stronger than intra-type competition for exclusion to occur, which causes a dip in $N$ over a selective sweep (except for the special case  .
+Constant-$N$ relative fitness models are not compatible with $K$-selection, and must refer to the competitive effects summarized by $K_{12}$ and $K_{21}$.
+We thus now restrict our attention to the case $K_{11}=K_{22}$.
+Fig.~\ref{fig:LVvslottery} shows two examples of selection between two types with the same saturation density.
+Fig.~\ref{fig:LVvslottery}a shows exclusion by a superior Lotka-Volterra competitor ($\alpha_{21}>\alpha_{12}$).
+Even though the saturation densities are identical, the density trajectories do not follow a line of constant $N$.
+This is a general properly of the Lotka-Volterra competition model; only a highly restricted subset of $r$ and $\alpha$ values will keep $N$ constant over a selective sweep (further details in Appendix C).
+Intuitively, for one type to exclude another with the same saturation density, inter-type competitive effects must be stronger than intra-type competitive effects, causing a dip in $N$ over a selective sweep.
 
-$c$-selection in the density-dependent lottery model, Eq.~\eqref{eq:master}.
+By contrast, if one type in our density-dependent lottery model has a $c$ advantage but birth and death rates are identical, the density trajectories converge on the line of constant density equal to the saturation density.
+Selection then occurs purely along this line, uncoupled from the density regulation of $N$.
+In other words, once the population reaches demographic equilibrium, it behaves indistinguishably from a constant-$N$ relative fitness model.
+More generally, the competitive ability trait $c$ does not directly affect population density (this can be seen formally in Eq.~\eqref{eq:Nmaster}), since $c$ only affects which type wins a territory, not whether a territory is won at all.
+In other words, $c$ is a zero-sum relative fitness trait embedded within a non-zero-sum fitness model.
 
-\subsection*{Coexistence in constant and cyclical environments}\label{sec:invas}
+Thus, we can conclude that relative fitness models are perfectly consistent with MacArthur's claim that ``fitness is $K$'' in crowded populations, bearing in mind his broad use of the symbol ``$K$''.
+However, inconsistency remains with standard growth models such as the Lotka-Volterra model.
+We revisit this issue in the Discussion.
 
-In the previous section we only considered how $b$, $c$ and $d$ should respond to selection in Grime's environmental extremes, based on invasion fitness.
-Here we further explore the low frequency behavior of Eq.
-\eqref{eq:master} to determine which types can coexist in a constant environment, and then consider the full time-dependent behaviour of Eq.
-\eqref{eq:master} in a cyclical environment.
-
-In a constant environment, stable coexistence is possible in our extended lottery model.
-A $b$-specialist $i$ and $c$-specialist $j$ ($b_i>b_j$, $c_j>c_i$) can co-exist because then propagule density $L$ is frequency-dependent, and so is the importance of competitive ability (Appendix D).
-This is a version of the classic competition-colonization trade-off \citep{tilman_94,levins_71}; the competitor ($c$-specialist) leaves many territories unoccupied (low $L$) due to its poor colonization ability (low $b$), which the colonizer ($b$-specialist) can then exploit.
-A similar situation holds for coexistence between high-$c$ and low-$d$ specialists; a ``competition-longevity'' trade-off \citep{tilman_94}.
-These forms of co-existence require density dependence (being mediated by $L$), and are not present in the classic lottery model.
-Coexistence is not possible between $b$- and $d$-specialists in a constant environment (Appendix D).
-
-Now suppose that birth and death rates vary periodically with amplitude sufficent to cause large changes in population density.
-This example is inspired by natural \textit{Drosophila} populations, which expand rapidly in the warmer months when fruit is abundant, but largely die off in the colder months.
-Along with this seasonal population density cycle, hundreds of polymorphisms exhibit frequency cycles that are in phase with the seasons \citep{bergland_14}.
-Some of these polymorphisms may be adaptive and potentially millions of years old, suggesting stable coexistence \citep{bergland_14,messer_2016}.
-Selection on allele frequencies thus occurs on the same time scale as population demography, a situation vastly more complicated than classical sweeps in demographically stable populations \citep{messer_2016}.
-
-The classical population genetic treatment of fluctuating selection suggests that environmental fluctuations do not promote coexistence.
-Allele frequencies are successively multiplied by relative fitness values for each environmental iteration, and so two alleles favored in different environments can only stably coexist if the product of fitnesses for one type exactly equals the product for the other \citep{dempster_1955}.
-Thus, stable coexistence still requires frequency-dependent selection or heterozygote advantage (as is required in a constant environment).
-
-This classical argument overlooks two general mechanisms that promote coexistence in fluctuating environments \citep{messer_2016}.
-The first is the classic version of the storage effect, which occurs when part of the population is protected from selection (due to overlapping generations in the lottery model; \citealt{chesson_1981}).
-The second is the bounded population size effect of \cite{yi_2013}, which occurs when each environmental cycle involves growth from low to high density, with the time spent growing each cycle dependent on the fitness of the types present.
-
-Fig.~\ref{fig:fluctuatingselection}a-c shows the behavior of Eq.
-\eqref{eq:master} for an example where $b$ and $d$ cycle between zero and positive values (``summers'' with rapid growth and no mortality, and ``winters'' with mortality and no growth).
-Both the storage effect (adults are sheltered from selection during the summer growth phase) and the bounded density effect (expansion to high density occurs every cycle) are operating.
-Two types are present, a $b$-specialist, which is better at rapidly growing in the summer (higher $b$), and a $d$-specialist which is better at surviving the winter (lower $d$).
-Neither type has an advantage over a full environmental cycle, and they stably coexist.
-This is due to a combination of the storage and bounded density effects (recall that stable coexistence between $b$ and $d$ specialists was not possible in a constant environment).
-
-The classic lottery model (Eq.
-BLAH) fails to give co-existence for these parameters because expansion to carrying capacity occurs immediately at the start of the summer (Fig.
-\ref{fig:fluctuatingselection}d-f).
-As a result, coexistence requires that the winter survivor's $b$ must be about $5$ times smaller than required when we properly account for the growth in the abundance of each type using Eq.
-\eqref{eq:master} (keeping the other parameters the same; Fig.~\ref{fig:fluctuatingselection}g-i).
-Previous models of the promotion of genetic variation via the storage effect  \citep{ellner_1994} similarly assume that the total number of offspring per iteration is constant, and would produce a similar error.
-
-\begin{figure}
-\centering
-\includegraphics[scale=0.7]{fluctuatingselection.pdf}
-\caption{\label{fig:fluctuatingselection} Stable coexistence between $b$ and $d$ specialists in a fluctuating environment requires a much greater $b$ advantage in the classic lottery model compared to our density-dependent extension of it when population density is seasonally cyclical.
-(a) Birth and death rates seasonally alternate being nonzero (white for winter, green for summer).
-The $b$-specialist (black) has higher $b$ and $d$ ($b=0.5$, $d=0.2$) than the $d$-specialist ($b=0.217$, $d=0.1$) (blue).
-(b) Both types grow during the positive $b$ phase, and decline during the positive $d$ phase, but the $d$-specialist does so at a lower rate.
-Total height (blue+black) is population density $N/T$.
-(c) Summer favors the $b$ specialist, winter the $d$-specialist, and they stably coexist.
-(d-f) Same as (a-c) for the classic lottery model; the types no longer coexist.
-(g-i) Same as (d-f) where now $b = 0.0421$ for the $d$ specialist and the types coexist.
-For illustration, the propagule abundances are assumed to have the form $m_i=b_i(1-N/T)n_i$, reflecting non-directed dispersal.}
-\end{figure}
+%, we will evaluate the broader ecological plausibility of relative fitness models from the basis of our density-dependent lottery model.
 
@@
 
-It is interesting to compare the predictions of the extended lottery model with earlier approaches, such as the $r$/$K$ scheme, where $r=b-d$ is the maximal, low-density growth rate \citep{pianka_1972}.
-Confusingly, the term ``$K$-selection'' sometimes refers generally to selection at high density \citep{pianka_1972}, encompassing both selection for higher saturation density \citep{macarthur_1967} and competitive ability \citep{gill_1974}.
-Contrary to predictions of an $r$/$K$ trade-off, empirical studies have shown that maximal growth rate at low density and the high density at which saturation occurs (measured by abundance) are positively correlated, both between species/strains \citep{luckinbill_1979,kuno_1991,hendriks_2005,fitzsimmons_2010}, and as a result of experimental evolution \citep{luckinbill_1978,luckinbill_1979}.
-From the perspective of our model, this positive correlation is not surprising since the saturation density, which is determined by a balance between births and deaths, increases with $b$.
+If s is small it doesn't matter anyway
 
-%This prediction is also counter to the expectations of MacArthur's $r$/$K$ dichotomy \citep{macarthur_1967}, since $b$ is closely related to the maximal, low-density growth rate $r=b-d$ \citep{pianka_1972}, yet in the $r$/$K$ scheme, high density populations should be subject to $K$, not $r$, selection.
-Yet it is not surprising that $b$ can matter at high densities.
-In our model (or any lottery model of competition), $b$ matters at high densities because territorial contests among juveniles are intrinsically unpredictable.
-This is a realistic feature of the model.
-Even if one type is guaranteed to win a territory in a ``fair'' contest (e.g.
-it is the most efficient exploiter of a limiting consumable resource; \citealt{tilman_1982}), inferior competitors can win by chance.
-For example, an inferior competitor's propagules may happen to arrive first, gaining a decisive developmental advantage.
-First arrivals are more likely to occur for types with a fecundity and/or dispersal advantage, as represented by higher $b$ in lottery models.
-The analogous intuition in the Wright-Fisher model is that fecundity confers a relative fitness advantage, even though population size is not changing.
-The logistic model for which $r$ and $K$ are named, does not capture this intuition.
-
-%In fact, one step in MacArthur's argument considerably narrows its scope: assuming that the nullclines of $f_1$ and $f_2$ exist at all.
-The strength of this assumption is clear if we note that essence of the dichotomy underlying the $r$/$K$ scheme is between interaction-dependent selection and interaction-independent selection.
+%The strength of this assumption is clear if we note that essence of the dichotomy underlying the $r$/$K$ scheme is between interaction-dependent selection and interaction-independent selection.
 That is, selective shifts in frequency are a result of differences in absolute growth rates, but these differences can arise in two logically distinct ways: 1) some types expand more rapidly the absence of interactions between individuals or 2) some types are superior in their interactions with other types.
@@
 
-There is support for a negative relationship between competitive success at high density and maximal growth rate \citep{luckinbill_1979}, consistent with a tradeoff between $r$ and the competitive aspect of $K$.
-This could be driven by a tradeoff between individual size and reproductive rate.
-To avoid confusion with other forms of ''$K$-selection'', selection for competitive ability has been called ``$\alpha$-selection'' after the competition coefficients in the Lotka-Volterra equation \citep{gill_1974,case_1974,joshi_2001}.
+%To avoid confusion with other forms of ''$K$-selection'', selection for competitive ability has been called ``$\alpha$-selection'' after the competition coefficients in the Lotka-Volterra equation \citep{gill_1974,case_1974,joshi_2001}.
 However, competitive success as measured by $\alpha$ (i.e.
@@
 $b$ in our model.
-Our $c$ is strictly competitive ability only --- as such, changes in $c$ do not directly affect population density (the total number of territories occupied per iteration is $\Delta_+ N=U(1-e^{-L})$, which does not depend directly on the $c_i$).
-The clean separation of a strictly-relative $c$ parameter is particularly useful from an evolutionary genetics perspective, essentially embedding a zero-sum relative fitness trait within a non-zero-sum fitness model.
-This could have interesting applications for modeling the impacts of intra-specific competition on species extinction, for example due to clonal interference \citep{gerrish_1998,desai_2007} between $c$-strategists on the one hand, and $b$- and $d$- strategists on the other.
-
-$K$-selection in the narrow logistic sense of selection for a greater environmental carrying capacity for given $r$, sometimes referred to as ``efficiency'' \citep{macarthur_1967}, could be represented in our model by smaller individual territorial requirements.
-To a first approximation, two co-occurring types which differ by a small amount in their territorial requirements only should have the same fitness, since the costs or benefits of a change in the amount of unocupied territory is shared equally among types via the propagule density per territory $L$.
-The situation is more complicated when the differences in territorial requirements become large enough that territorial contests can occur on different scales for different types.
-We leave these complications for future work.
-
-Nevertheless, it is interesting to note that ruderals, which are typically thought of as high fecundity dispersers ($b$-specialists), may also be strongly $d$-selected, which while unintuitive, is consistent with our findings.
-An effective way to reduce $d$ in the face of unavoidable physical destruction is to shorten the time to reproductive maturity --- short life cycles are a characteristically ruderal trait.
-Moreover, a recent hierarchical cluster analysis of coral traits did find a distinct ``ruderal'' cluster, but high fecundity was not its distinguishing feature.
-Rather, ruderals used brood- (as opposed to broadcast-) spawning, which could plausibly be a mechanism for improving propagule survivorship in disturbed environments \citep{darling_2012}.
 
@@
 
-\section*{Appendix C: Mutant invasion and coexistence in a constant environment}
+\section*{Appendix C: Total density in the Lotka-Volterra competition model}
 
-Here we evaluate the initial growth or decline of mutants in a population with a single resident type, which is in equilibrium.
-To determine whether coexistence is possible, we check for ``mutual invasion'', that is, we check that type $j$ will invade an $i$-dominated population, but type $i$ will also invade a $j$-dominated population.
+Here we show that under the Lotka-Volterra model of competition, total density $N$ does not in general remain constant over a selective sweep in a crowded population even if the types have the same saturation density.
 
-Solving for equilibrium when $i$ is the resident ($\Delta n_i = 0$), we have $R_i=0$, $\overline{c}=c_i$, $A_i=(1-(1+L)e^{-L})/L$, and Eq. \eqref{eq:master} becomes
+We assume $\alpha_{11}=\alpha_{22}$ and $N=1/\alpha_{11}$ and check whether it is then possible for $\frac{dN}{dt}$ to be zero in the sweep ($n_1,n_2 \neq 0$).
+Substituting these conditions into Eq.~\eqref{eq:LV}, we obtain
+\begin{align}
+\frac{d n_1}{dt} = -r_1(\alpha_{12}-\alpha_{11})n_1n_2 \nonumber\\
+\frac{d n_2}{dt} = -r_2(\alpha_{21}-\alpha_{22})n_1n_2
+\end{align}
+Adding these together, $\frac{dN}{dt}$ can only be zero if 
 \begin{equation}
-b_i(1-e^{-L})/L-d_i=0.\label{eq:equil}
+r_1(\alpha_{12}-\alpha_{11})+r_2(\alpha_{21}-\alpha_{22})=0. \label{eq:constNcondition}
 \end{equation}
-This implies $L\approx b_i/d_i$ if $b_i/d_i\gg 1$ and $L\ll 1$ if $b_i/d_i\approx 1$.
-
-Now suppose that a novel mutant $j$, which is initially rare, appears in the population.
-Then $A_j/R_j\ll 0$, $l_j\approx 0$ and $\overline{c}\approx c_i$, and so, from Eq.
-\eqref{eq:master}, the mutant lineage's fitness is
-\begin{equation}
-\Delta n_j/n_j \approx b_j \left(e^{-L}+R_j\frac{c_j}{c_i}\right)-d_j \label{eq:invad}
-\end{equation}
-where $R_j\approx (1-e^{-L})/\left(\frac{c_j}{c_i}+\frac{L-1+e^{-L}}{1-(1+L)e^{-L}}\right)$.
-
-We consider the case of coexistence between a $b$-specialist $i$ and a $c$-specialist $j$ ($b_i>b_j$, $c_j>c_i$ and $d_i=d_j$).
-Suppose that $b_i$ is so large that $L\gg 1$ when $i$ is dominant, and $b_j$ is so small that $L\ll 1$ when $j$ is dominant.
-Then, when $j$ is dominant, we have $\Delta n_i/n_i=b_i-d_i=b_i-d_j=b_i-b_j>0$.
-When $i$ is dominant, Eq.
-\eqref{eq:highLinvad} applies, where Eq.
-\eqref{eq:equil} implies $d_j=d_i=b_i(1-e^{-L})/L\approx b_i/L$, and so
-\begin{equation}
-\Delta n_j/n_j \approx \frac{b_j}{L}\frac{c_j}{c_i}-\frac{b_i}{L}.
-\end{equation}
-Therefore, coexistence occurs if $c_j/c_i$ is sufficiently large.
-The analogous argument for $d$- and $c$-specialists ($d_i<d_j$ with $L\gg 1$ when $i$ dominates, $L\ll 1$ when $j$ dominates, and $b_i=b_j$) gives $\Delta n_j/n_j \approx d_i\frac{c_j}{c_i}-d_j$, which again implies coexistence if $c_j/c_i$ is sufficiently large.
-
-For $b$-and $d$-specialists ($c_i=c_j$), we have $\Delta n_j/n_j\approx b_j d_i/b_i-d_j$ when $i$ dominates and $\Delta n_i/n_i\approx b_i d_j/b_j-d_i$ when $j$ dominates.
-Thus, either $i$ or $j$ grows when rare, but not both, and stable coexistence is not possible in a constant environment.
+To get some intuition for Eq.~\eqref{eq:constNcondition}, suppose that a mutant arises with improved competitive ability but identical intrinsic growth rate and saturation density ($r_1=r_2$ and $\alpha_{11}=\alpha_{22}$).
+This could represent a mutation to an interference competition trait, for example \citep{gill_1974}.
+Then, according the above condition, for $N$ to remain constant over the sweep, the mutant must find the wildtype more tolerable than itself by exactly the same amount that the wildtype finds the mutant less tolerable than itself.
+This condition, and Eq.~\eqref{eq:constNcondition} more generally, are so restrictive that we can conclude that selective sweeps in the Lotka-Volterra competition model will generally involve non-constant $N$.
 

### 2017-11-29 fluctuating_selection.py
l

@@ -5,150 +5,6 @@ Created on Thu Mar 24 12:01:54 2016
 @author: jbertram
 """
 
-#Simulation comparison figure
-#================================================================
-m=np.array([10000,90000])
-M=float(sum(m)) 
-c=np.array([1.5,1.])
-cbar=sum(m*c)/sum(m)
-maxl=2; Us=np.array([int(m[0]/x) for x in np.linspace(0.001,maxl,1000)]);
-exact=np.array([deltnplussim(m,c,U) for U in Us]);
-totalclassic=np.array([deltnplusclassic(m,c,U) for U in Us]);
-totalclassicboundedm=np.array([deltnplusclassicboundedm(m,c,U) for U in Us]);
-
-fig1, (ax1, ax2) = plt.subplots(ncols=2,figsize=[8,4])
-
-#First genotype
-nocomp=np.array([np.exp(-M/U) for U in Us]);
-comp1=np.array([R(m,c,U)[0]*c[0]/cbar for U in Us]);
-comp2=np.array([A(m,c,U)[0]*c[0]/cbar for U in Us]);
-
-ax1.plot(float(m[0])/Us,nocomp,'k',label=r"$e^{-L}$")
-ax1.plot(float(m[0])/Us,comp1,'r',label=r"$R_i \frac{c_i}{\overline{c}}$")
-ax1.plot(float(m[0])/Us,comp2,'b',label=r"$A_i \frac{c_i}{\overline{c}}$")
-ax1.plot(float(m[0])/Us,(nocomp+comp1+comp2),'k',label=r"$e^{-L}+R_i \frac{c_i}{\overline{c}}+A_i \frac{c_i}{\overline{c}}$",linewidth=2)
-
-ax1.plot(float(m[0])/Us,(exact[:,0,0]+exact[:,1,0]+exact[:,2,0])/m[0],'k.',markersize=1.,label="Simulation")
-ax1.plot(float(m[0])/Us,exact[:,0,0]/m[0],'k.',markersize=1)
-ax1.plot(float(m[0])/Us,exact[:,1,0]/m[0],'k.',markersize=1)
-ax1.plot(float(m[0])/Us,exact[:,2,0]/m[0],'k.',markersize=1)
-
-ax1.plot(float(m[0])/Us,totalclassic[:,0]/m[0],'k--',label=r"Classic lottery",linewidth=2)
-ax1.set_xlim([0,maxl])
[…]

### 2017-11-29 function_definitions.py

@@ -1,0 +1,78 @@
+# -*- coding: utf-8 -*-
+"""
+Created on Wed Nov 29 12:25:02 2017
+
+@author: jbertram
+"""
+
+import numpy as np
+import matplotlib.pyplot as plt
+import bisect
+
+def deltnplussim(m,c,U):
+    scatter=[[0 for x in range(len(m))] for y in range(U)];
+    for i in range(len(m)):
+        for y in xrange(m[i]):
+            cell=int(U*np.random.rand());
+            scatter[cell][i]=scatter[cell][i]+1;
+    
+    winsnocomp=np.zeros(len(m)); wins1=np.zeros(len(m)); wins2=np.zeros(len(m));
+    comp=np.zeros(U);
+    for i in range(U):
+        comp[i]=sum(scatter[i])
+        if comp[i]>0:            
+            lotterycmf=np.cumsum(np.array(scatter[i])*c)
+            victor=bisect.bisect(lotterycmf,np.random.rand()*lotterycmf[-1])
+            
+            if scatter[i][victor]==1 and comp[i]==1:
+                winsnocomp[victor]=winsnocomp[victor]+1
+            elif scatter[i][victor]==1:
+                wins1[victor]=wins1[victor]+1
+            else: 
+                wins2[victor]=wins2[victor]+1
+        
+    return np.array([winsnocomp,wins1,wins2])
+
+def R(m,c,U):
+    l=m/float(U)
+    L=sum(l)
+    cbar=sum(m*c)/sum(m); 
+    out=cbar*np.exp(-l)*(1-np.exp(-(L-l)))\
+            /(c + (L-1+np.exp(-L))/(1-(1+L)*np.exp(-L))*(cbar*L-c*l)/(L-l))
+    
+    for i in range(len(out)):
+        if np.isnan(out)[i]: out[i]=0
+            
+    return out
+
+def A(m,c,U):
+    l=m/float(U)
+    L=sum(l)
+    cbar=sum(m*c)/sum(m)
+    out=cbar*(1-np.exp(-l))\
+            /((1-np.exp(-l))/(1-(1+l)*np.exp(-l))*c*l\
+            +(L*(1-np.exp(-L))/(1-(1+L)*np.exp(-L))-l*(1-np.exp(-l))/(1-(1+l)*np.exp(-l)))/(L-l)*(cbar*L-c*l))
+    for i in range(len(out)):
+        if np.isnan(out)[i]: out[i]=0
+            
[…]

### 2017-11-29 phase_plot_LV_lottery.py

@@ -1,0 +1,83 @@
+# -*- coding: utf-8 -*-
+"""
+Created on Wed Nov 29 12:26:40 2017
+
+@author: jbertram
+"""
+
+#(n1,n2) vector field
+#=================================================================
+T=100000
+x,y=np.linspace(0.0,1.,401),np.linspace(0.,1.,401)
+xgrid, ygrid=np.meshgrid(x,y)
+X,Y=np.zeros([len(x),len(y)]),np.zeros([len(x),len(y)])
+b=np.array([5.,5.])
+c=np.array([5.,1.])
+d=np.array([0.1,0.1])
+
+for i in xrange(len(x)):
+    for j in xrange(len(y)):
+        n=T*np.array([x[i],y[j]])
+        U=T-sum(n)
+        X[j,i],Y[j,i]=deltnplus(b*n*U/float(T),c,U)-d*n
+
+
+fig1, (ax2, ax1) = plt.subplots(ncols=2,figsize=[8,4])
+
+f1null=np.array([[x[[_ for _ in range(len(x)) if X[i,_]>0][-1]],y[i]] for i in range(1,len(y)) if max(X[i])>0])
+f2null=np.array([[x[[_ for _ in range(len(x)) if Y[i,_]>0][-1]],y[i]] for i in range(1,len(y)) if max(Y[i])>0])
+ax1.plot(f1null[:,0],f1null[:,1],'k--',linewidth=2.)
+ax1.plot(f2null[:,0],f2null[:,1],'k--',linewidth=2.)
+#plt.plot([f2null[0,0],f2null[-1,0]],[f2null[0,1],f2null[-1,1]],'y',linewidth=2)
+ax1.plot([f2null[-1,0],f1null[0,0]],[f2null[-1,1],f1null[0,1]],'k',linewidth=2,alpha=0.8)
+ax1.set_xlim([0,f1null[0,0]])
+ax1.set_ylim([0,f1null[-1,1]])
+
+ax1.annotate(r'$(b)$',xy=(0.9,0.93),xycoords='axes fraction',fontsize=16)
+
+seedpoints=[[x[_]/2,y[200-_]/2] for _ in range(0,201,10)]
+seedpoints=np.concatenate([[[_,0.2-_] for _ in np.arange(0.02,0.2,0.02)],\
[…]

### 2017-11-29 simulation_comparison.py

@@ -1,0 +1,74 @@
+# -*- coding: utf-8 -*-
+"""
+Created on Wed Nov 29 12:26:08 2017
+
+@author: jbertram
+"""
+
+#Simulation comparison figure
+#================================================================
+m=np.array([10000,90000])
+M=float(sum(m)) 
+c=np.array([1.5,1.])
+cbar=sum(m*c)/sum(m)
+maxl=2; Us=np.array([int(m[0]/x) for x in np.linspace(0.001,maxl,1000)]);
+exact=np.array([deltnplussim(m,c,U) for U in Us]);
+totalclassic=np.array([deltnplusclassic(m,c,U) for U in Us]);
+totalclassicboundedm=np.array([deltnplusclassicboundedm(m,c,U) for U in Us]);
+
+fig1, (ax1, ax2) = plt.subplots(ncols=2,figsize=[8,4])
+
+#First genotype
+nocomp=np.array([np.exp(-M/U) for U in Us]);
+comp1=np.array([R(m,c,U)[0]*c[0]/cbar for U in Us]);
+comp2=np.array([A(m,c,U)[0]*c[0]/cbar for U in Us]);
+
+ax1.plot(float(m[0])/Us,nocomp,'k',label=r"$e^{-L}$")
+ax1.plot(float(m[0])/Us,comp1,'r',label=r"$R_i \frac{c_i}{\overline{c}}$")
+ax1.plot(float(m[0])/Us,comp2,'b',label=r"$A_i \frac{c_i}{\overline{c}}$")
+ax1.plot(float(m[0])/Us,(nocomp+comp1+comp2),'k',label=r"$e^{-L}+R_i \frac{c_i}{\overline{c}}+A_i \frac{c_i}{\overline{c}}$",linewidth=2)
+
+ax1.plot(float(m[0])/Us,(exact[:,0,0]+exact[:,1,0]+exact[:,2,0])/m[0],'k.',markersize=1.,label="Simulation")
+ax1.plot(float(m[0])/Us,exact[:,0,0]/m[0],'k.',markersize=1)
+ax1.plot(float(m[0])/Us,exact[:,1,0]/m[0],'k.',markersize=1)
+ax1.plot(float(m[0])/Us,exact[:,2,0]/m[0],'k.',markersize=1)
+
[…]

### 2017-11-29 strength_of_selection.py

@@ -1,0 +1,55 @@
+# -*- coding: utf-8 -*-
+"""
+Created on Tue Nov 21 14:29:53 2017
+
+@author: jbertram
+"""
+
+import numpy as np
+import matplotlib.pyplot as plt
+from scipy.integrate import odeint
+
+#Logistic model
+#==========================================================
+
+r=[1,1];
+s=0.1
+K=[1e5,1e5*(1+s)];
+
+x=np.linspace(0,20/s,1000)
+
+def f(n,t):
+    return [r[0]*(1-(n[0]+n[1])/K[0])*n[0],r[0]*(1-(n[0]+n[1])/K[1])*n[1]]
+    
+def g(p,t):
+    return s*p*(1-p)
+
+solK=odeint(f,[K[0]-1.,1.],x)
+solrel=odeint(g,1/K[0],x)
+
+plt.plot(x,solK[:,1]/(solK[:,0]+solK[:,1]))
+plt.plot(x,solrel)
+plt.ylim([0,1])
+
+#lottery model
+#==========================================================
+
+T=100000
+totaltime=200
+b=np.array([5.,5.])
+c=np.array([1.,2.])
+d=np.array([0.1,0.1])
+
+n=np.array([90000.,100.])
+nhist=[n];
+for t in range(totaltime):
+    U=T-sum(n)
+    n=n+deltnplus(b*n*U/T,c,U)-d*n
+    nhist.append(list(n))
+    
+nhist=np.array(nhist)
+
+plt.plot(nhist[:,1])
+
+plt.figure()
+plt.plot(np.sum(nhist,1))

### 2017-11-29 function_definitions.py
l

@@ -10,15 +10,15 @@ import matplotlib.pyplot as plt
 import bisect
 
 def deltnplussim(m,c,U):
-    scatter=[[0 for x in range(len(m))] for y in range(U)];
+    scatter=np.zeros([int(U),len(m)])
     for i in range(len(m)):
-        for y in xrange(m[i]):
-            cell=int(U*np.random.rand());
-            scatter[cell][i]=scatter[cell][i]+1;
+        for y in xrange(int(m[i])):
+            cell=int(int(U)*np.random.rand());
+            scatter[cell,i]=scatter[cell,i]+1;
     
     winsnocomp=np.zeros(len(m)); wins1=np.zeros(len(m)); wins2=np.zeros(len(m));
     comp=np.zeros(U);
-    for i in range(U):
+    for i in range(int(U)):
         comp[i]=sum(scatter[i])
         if comp[i]>0:            
             lotterycmf=np.cumsum(np.array(scatter[i])*c)

### 2017-11-29 simulations.py

@@ -1,390 +1,0 @@
-# -*- coding: utf-8 -*-
-"""
-Created on Thu Mar 24 12:01:54 2016
-
-@author: jbertram
-"""
-
-#Simulation comparison figure
-#================================================================
-m=np.array([10000,90000])
-M=float(sum(m)) 
-c=np.array([1.5,1.])
-cbar=sum(m*c)/sum(m)
-maxl=2; Us=np.array([int(m[0]/x) for x in np.linspace(0.001,maxl,1000)]);
-exact=np.array([deltnplussim(m,c,U) for U in Us]);
-totalclassic=np.array([deltnplusclassic(m,c,U) for U in Us]);
-totalclassicboundedm=np.array([deltnplusclassicboundedm(m,c,U) for U in Us]);
-
-fig1, (ax1, ax2) = plt.subplots(ncols=2,figsize=[8,4])
-
-#First genotype
-nocomp=np.array([np.exp(-M/U) for U in Us]);
-comp1=np.array([R(m,c,U)[0]*c[0]/cbar for U in Us]);
-comp2=np.array([A(m,c,U)[0]*c[0]/cbar for U in Us]);
-
-ax1.plot(float(m[0])/Us,nocomp,'k',label=r"$e^{-L}$")
-ax1.plot(float(m[0])/Us,comp1,'r',label=r"$R_i \frac{c_i}{\overline{c}}$")
-ax1.plot(float(m[0])/Us,comp2,'b',label=r"$A_i \frac{c_i}{\overline{c}}$")
-ax1.plot(float(m[0])/Us,(nocomp+comp1+comp2),'k',label=r"$e^{-L}+R_i \frac{c_i}{\overline{c}}+A_i \frac{c_i}{\overline{c}}$",linewidth=2)
-
-ax1.plot(float(m[0])/Us,(exact[:,0,0]+exact[:,1,0]+exact[:,2,0])/m[0],'k.',markersize=1.,label="Simulation")
-ax1.plot(float(m[0])/Us,exact[:,0,0]/m[0],'k.',markersize=1)
-ax1.plot(float(m[0])/Us,exact[:,1,0]/m[0],'k.',markersize=1)
-ax1.plot(float(m[0])/Us,exact[:,2,0]/m[0],'k.',markersize=1)
-
-ax1.plot(float(m[0])/Us,totalclassic[:,0]/m[0],'k--',label=r"Classic lottery",linewidth=2)
[…]

### 2017-11-29 strength_of_selection.py

@@ -36,15 +36,40 @@ plt.ylim([0,1])
 
 T=100000
 totaltime=200
-b=np.array([5.,5.])
-c=np.array([1.,2.])
-d=np.array([0.1,0.1])
+s=0.5
+b=np.array([1.,1.*(1+s)])
+c=np.array([1.,1.])
+d=np.array([.1,.1])
 
 n=np.array([90000.,100.])
 nhist=[n];
 for t in range(totaltime):
     U=T-sum(n)
     n=n+deltnplus(b*n*U/T,c,U)-d*n
+    nhist.append(list(n))
+    
+nhist=np.array(nhist)
+
+plt.plot(nhist[:,1])
+
+plt.figure()
+plt.plot(np.sum(nhist,1))
+
+plt.ylim([0,T])
+
+plt.figure()
+plt.plot(map(np.log,nhist[:,1]))
+
+print (nhist[11,1]-nhist[10,1])/nhist[10,1]
+
+
+
+
+n=np.array([90000.,100.])
+nhist=[n];
+for t in range(totaltime):
+    U=T-sum(n)
+    n=n+np.sum(deltnplussim(b*n*U/T,c,U),0)-d*n
     nhist.append(list(n))
     
 nhist=np.array(nhist)

### 2017-11-30 strength_of_selection.py
l

@@ -35,13 +35,26 @@ plt.ylim([0,1])
 #==========================================================
 
 T=100000
-totaltime=200
-s=0.5
-b=np.array([1.,1.*(1+s)])
+totaltime=100
+b=np.array([5.,5.])
 c=np.array([1.,1.])
-d=np.array([.1,.1])
+d=np.array([1.,1.])
+n0=np.array([49000.,100.])
 
-n=np.array([90000.,100.])
+n=n0;
+nhist=[n];
+for t in range(totaltime/2):
+    U=T-sum(n)
+    n=n+deltnplus(b*n*U/T,c,U)-d*n
+    nhist.append(list(n))
+
+#plt.plot(np.sum(nhist,1))
+n0=np.array(n)
+print n0 #equilibrium
+
+s=0.4
+d=np.array([1.,1.*(1-s)])
+n=n0;
 nhist=[n];
 for t in range(totaltime):
     U=T-sum(n)
@@ -49,32 +62,37 @@ for t in range(totaltime):
     nhist.append(list(n))
     
 nhist=np.array(nhist)
+Nhist=np.sum(nhist,1)
+print "b absolute: ", nhist[1,1]/nhist[0,1]-1
+print "b relative: ", nhist[1,1]/nhist[0,1]*Nhist[0]/Nhist[1]-1
+print "N change: ", Nhist[-1]/Nhist[0]
 
-plt.plot(nhist[:,1])
-
-plt.figure()
-plt.plot(np.sum(nhist,1))
-
-plt.ylim([0,T])
-
-plt.figure()
-plt.plot(map(np.log,nhist[:,1]))
-
-print (nhist[11,1]-nhist[10,1])/nhist[10,1]
+plt.plot(nhist[:,1]/Nhist)
+#plt.plot(Nhist)
 
 
-
-
-n=np.array([90000.,100.])
+d=np.array([1.,1.])
+c=np.array([1.,2.01])
+n=n0;
 nhist=[n];
 for t in range(totaltime):
     U=T-sum(n)
-    n=n+np.sum(deltnplussim(b*n*U/T,c,U),0)-d*n
+    n=n+deltnplus(b*n*U/T,c,U)-d*n
     nhist.append(list(n))
     
 nhist=np.array(nhist)
+Nhist=np.sum(nhist,1)
+print nhist[1,1]/nhist[0,1]-1
[…]

### 2017-12-01 manuscript.tex
l

@@
 But any measure of fitness should ultimately be grounded in the processes of birth and death which govern population biology \citep{metcalf_2007,doebeli_2017}.
-While this grounding is clear for absolute fitness measures like $r$ and $K$, relative fitness seems largely divorced from population ecology.
+While this grounding is clear for absolute fitness measures like $r$ and saturation density, relative fitness seems largely divorced from population ecology.
 
@@
 26]{crow_1970}, with selection favoring greater $r$ ($r$-selection).
-In crowded populations, relative fitness models almost universally assume that total population size $N$ is fixed, or has some externally imposed time course.
+The situation is considerably more complicated in crowded populations.
+In this case, relative fitness models almost universally assume that total population size $N$ is fixed, or has some externally imposed time course.
 This is exemplified by the Wright-Fisher model, in which time advances in discrete non-overlapping generations, and fitness can be interpreted as a product of fertility and juvenile viability \citep[pp.
 185]{crow_1970}.
-The limitations of these relative fitness models are openly acknowledged, usually with an emphasis on the difficulty of incorporating more of the life cycle ``components'' of fitness (e.g.
+The limitations of relative fitness models are openly acknowledged, usually with an emphasis on the difficulty of incorporating multiple life cycle ``components'' of fitness (e.g.
 \cite[pp.
-276]{ewens_2004}).
+276]{ewens_2004} \citep{prout_1980}).
 But more fundamentally, the constant-$N$, relative fitness description uncouples selection and demography, at odds with the fact that the same birth/death events drive both.
 
-The issue can be expressed more formally by revisiting MacArthur's analysis of selection in crowded populations \citep{macarthur_1967}.
+The issue can be expressed formally by revisiting MacArthur's analysis of selection in crowded populations \citep{macarthur_1967}.
 MacArthur considers a population consisting of two types with densities $n_1$ and $n_2$ subject to density-dependent population growth described by
@@
 The short-term approximation also precludes consideration of inherently long-term evolutionary processes like the management of genetic load and population extinction \citep{bertram_2017}.
-Nevertheless, relative fitness models like Wright-Fisher are the foundation for much of the population genetic literature, and are still widely used without considering the ``short-term'' restriction.
-Thus, to summarize, there are important theoretical and practical reasons to deepen our understanding of the population-ecological foundations of relative fitness models.
+Nevertheless, relative fitness models like Wright-Fisher are the foundation for much of the population genetic literature, and are still widely used without considering the ``short-term'' restriction or the broader gulf with population ecology \citep{mallet_2012}.
+Thus, it is important that we deepen our population-ecological understanding of relative fitness models, both to gain insight into their domain of applicability, and as part of the broad challenge of synthesizing ecology and evolution.
+
+The recent rise of adaptive dynamics has provided a powerful framework for modeling the interplay between long-term evolution and population ecology, including complex couplings between evolutionary change and population density.
+However, the focus of this latter work is on trait evolution rather than the underlying genetics, and in particular, the details of selective sweeps are typically subsumed into effectively instantaneous ``trait substitutions'' determined by invasion fitness.
+We emphasize that our focus here is on the description of selection itself, which is critical for evolutionary genetics i.e.
+for making sense of the change in allele frequencies over time.
 
@@
 This is a general properly of the Lotka-Volterra competition model; only a highly restricted subset of $r$ and $\alpha$ values will keep $N$ constant over a selective sweep (further details in Appendix C).
-Intuitively, for one type to exclude another with the same saturation density, inter-type competitive effects must be stronger than intra-type competitive effects, causing a dip in $N$ over a selective sweep.
+Intuitively, for one type to exclude another with the same saturation density, inter-type competitive effects must be stronger than intra-type competitive effects, causing a dip in $N$ due to the selective sweep.
 

### 2017-12-01 strength_of_selection.py

@@ -13,7 +13,7 @@ from scipy.integrate import odeint
 #==========================================================
 
 r=[1,1];
-s=0.1
+s=0.3
 K=[1e5,1e5*(1+s)];
 
 x=np.linspace(0,20/s,1000)
@@ -30,6 +30,9 @@ solrel=odeint(g,1/K[0],x)
 plt.plot(x,solK[:,1]/(solK[:,0]+solK[:,1]))
 plt.plot(x,solrel)
 plt.ylim([0,1])
+
+s=np.linspace(0,0.9,100)
+plt.plot(s,s/(1-s)-s)
 
 #lottery model
 #==========================================================

### 2017-12-02 strength_of_selection.py
l

@@ -38,8 +38,8 @@ plt.plot(s,s/(1-s)-s)
 #==========================================================
 
 T=100000
-totaltime=100
-b=np.array([5.,5.])
+totaltime=80
+b=np.array([3.,3.])
 c=np.array([1.,1.])
 d=np.array([1.,1.])
 n0=np.array([49000.,100.])
@@ -70,12 +70,16 @@ print "b absolute: ", nhist[1,1]/nhist[0,1]-1
 print "b relative: ", nhist[1,1]/nhist[0,1]*Nhist[0]/Nhist[1]-1
 print "N change: ", Nhist[-1]/Nhist[0]
 
-plt.plot(nhist[:,1]/Nhist)
-#plt.plot(Nhist)
+
+fig1,ax1=plt.subplots()
+ax1.plot(nhist[:,1]/Nhist)
+
+fig2,ax2=plt.subplots()
+ax2.plot(Nhist)
 
 
 d=np.array([1.,1.])
-c=np.array([1.,2.01])
+c=np.array([1.,3.])
 n=n0;
 nhist=[n];
 for t in range(totaltime):
@@ -88,8 +92,8 @@ Nhist=np.sum(nhist,1)
 print nhist[1,1]/nhist[0,1]-1
 print nhist[1,1]/nhist[0,1]*Nhist[0]/Nhist[1]-1
 
-plt.plot(nhist[:,1]/Nhist)
-#plt.plot(Nhist)
+ax1.plot(nhist[:,1]/Nhist)
+ax2.plot(Nhist)
 
 
 

### 2017-12-04 manuscript.tex
l

--- 2017-11-29 manuscript.tex
@@
 
-\title{Density-dependent selection and the population ecology of relative fitness}
+\title{Density-dependent selection and the limits of relative fitness}
 %\author{Jason Bertram $^{1,\ast}$ \\ 
@@
 468]{barton_2007}.
-This variety is not necessarily problematic in itself, because different measures of fitness may be more useful in different circumstances.
-But any measure of fitness should ultimately be grounded in the processes of birth and death which govern population biology \citep{metcalf_2007,doebeli_2017}.
-While this grounding is clear for absolute fitness measures like $r$ and $K$, relative fitness seems largely divorced from population ecology.
+The variety of fitness measures is not problematic in itself, because different measures  may be more useful in different circumstances.
+But it should be clear how the measure being used is connected to the processes of birth and death which govern population biology \citep{metcalf_2007,doebeli_2017}.
+While such a connection is fairly clear for absolute fitness measures like $r$, relative fitness seems largely divorced from population biology.
+It has even been proposed that relative fitness be justified from more abstract measure-theoretical arguments, abandoning population biology altogether \citep{wagner_2010}.
 
-In uncrowded populations, relative fitness simply represents differences in the intrinsic exponential growth rate $r$ \citep[pp.
-26]{crow_1970}, with selection favoring greater $r$ ($r$-selection).
-In crowded populations, relative fitness models almost universally assume that total population size $N$ is fixed, or has some externally imposed time course.
-This is exemplified by the Wright-Fisher model, in which time advances in discrete non-overlapping generations, and fitness can be interpreted as a product of fertility and juvenile viability \citep[pp.
-185]{crow_1970}.
-The limitations of these relative fitness models are openly acknowledged, usually with an emphasis on the difficulty of incorporating more of the life cycle ``components'' of fitness (e.g.
-\cite[pp.
-276]{ewens_2004}).
-But more fundamentally, the constant-$N$, relative fitness description uncouples selection and demography, at odds with the fact that the same birth/death events drive both.
+In uncrowded populations, relative fitness simply represents differences in intrinsic growth rate. In continuous time, we have the canonical selection equation \citep[pp. 26]{crow_1970}
+\begin{equation}
+\frac{d p_i}{dt}=(r_i-\overline{r})p_i, \label{eq:canonical}
+\end{equation}
+where $r_i$ is the intrinsic exponential growth rate of type $i$, $p_i$ is its frequency, and $\overline{r}=\sum_i r_i p_i$ is the population mean $r$.
+If there are two types present, a wildtype and a mutant for instance, then Eq.~\eqref{eq:canonical} can be written as $\frac{d p_i}{dt}=s p_i(1-p_i)$, where the constant selection coefficient $s$ is the difference in $r$ between types; the corresponding adaptive sweeps follow a logistic curve.
 
-The issue can be expressed more formally by revisiting MacArthur's analysis of selection in crowded populations \citep{macarthur_1967}.
-MacArthur considers a population consisting of two types with densities $n_1$ and $n_2$ subject to density-dependent population growth described by
-\begin{equation}
-\frac{d n_1}{d t}=f_1(n_1,n_2)\qquad\frac{d n_2}{d t}=f_2(n_1,n_2). \label{eq:macgeneral}
-\end{equation}
-Apart from the type densities, the environment is assumed to remain constant.
-The functions $f_1$ and $f_2$ must decline to zero if $n_1$ or $n_2$ are sufficiently large, because no population has unlimited resources.
-This defines the nullclines $f_1(n_1,n_2)=0$ and $f_2(n_1,n_2)=0$ in $(n_1,n_2)$ space.
-The outcome of selection is then determined by the relationship between these nullclines.
-Specifically, a type will be excluded if its nullcline is completely contained in the region bounded by the other type's nullcline.
-In other words, for a type to have the possibility of persisting, it must be able to keep growing to higher densities than the other type can tolerate in some region of $(n_1,n_2)$ space (Fig.~\ref{fig:Ksel}a).
-
-The connection between constant-$N$ relative fitness models and the broad class of density-regulated growth models considered by MacArthur (Eq.~\ref{eq:macgeneral}) is not clear, since the former only describe selection along lines defined by $n_1+n_2=N$ for each $N$ (Fig.~\ref{fig:Ksel}b).
-In particular, the $f_1$ and $f_2$ nullclines are not even defined.
-It is clear, however, that constant-$N$ relative fitness models are not compatible with some of the most widely-used models of density-regulated population growth: 1) In the logistic model, the type with the greatest saturation population density excludes the others.
-2) The ``$R^*$ rule'', a central tenet of resource competition theory, states that when growth is limited by a single homogeneous consumable resource, the type able to deplete the resource to the lowest equilibrium density $R^*$ excludes the others \citep{grover_1997}.
+The difficulty with Eq.~\eqref{eq:canonical} arises in crowded populations.
+It is then commonly assumed that total population density $N$ has reached its equilibrium value, which is assumed to be a fixed constant.
+This is exemplified by the Wright-Fisher model.
+Yet MacArthur famously showed that when population growth is density-regulated, selection in crowded populations is intimately connected to the ability to keep growing at higher densities than other types can tolerate \citep{macarthur_1967}.
+The classic example is the logistic model, where the type with the greatest saturation population density ``$K$'' excludes the others.
+Similarly, the ``$R^*$ rule'', a central tenet of resource competition theory, states that when growth is limited by a single homogeneous consumable resource, the type able to deplete the resource to the lowest equilibrium density $R^*$ excludes the others \citep{grover_1997}.
 Differences in $R^*$ will often entail differences in saturation density.
-3) The Lotka-Volterra competition model also couples selection in crowded populations to changes in total density $N$ (we return to this in section ``$K$-selection, $c$-selection and relative fitness'').
+The Lotka-Volterra competition model also couples selection in crowded populations to changes in $N$ except in a few special cases (we return to this in section ``$K$-selection, $c$-selection and relative fitness'').
+Thus, $N$ changes during, and as a result of, the adaptive sweeps described by Eq.~\eqref{eq:canonical}.
 It would therefore seem that the ubiquitous constant-$N$, relative fitness description of selection is incompatible with a huge class of population ecological processes occurring in nature and experiments.
@@
 \includegraphics[scale=0.8]{Kplot.pdf}
-\caption{\label{fig:Ksel} (a) MacArthur's dynamical argument for how selection operates in crowded environments, illustrated using the logistic model $f_1=r_1(1-\frac{n_1+n_2}{K_1})n_1$ and $f_2=r_2(1-\frac{n_1+n_2}{K_2})n_1$ in Eq.
-\eqref{eq:macgeneral}.
-Here $r_1=r_2$ and $K_1>K_2$.
+\caption{\label{fig:Ksel} Selection in crowded environments shown as a phase diagram for the densities of two types $n_1$ and $n_2$.
+(a) The logistic model $\frac{dn_1}{dt}=r_1(1-\frac{n_1+n_2}{K_1})n_1$ and $\frac{dn_2}{dt}=r_2(1-\frac{n_1+n_2}{K_2})n_1$ with $r_1=r_2$ and $K_1>K_2$.
 (b) The constant-$N$, relative fitness description of selection.}
@@
 
-The constant-$N$, relative fitness description has historically been justified as a short-term linear approximation \cite[pp.
+The relative fitness description has been justified in broadly two different ways for crowded populations (we do not address Wagner's [\citeyear{wagner_2010}] measure theorical justification, which explicitly seeks to be independent of population biology and thus falls outside of our scope).
+First, if we only care about type frequencies, we can sometimes simply ignore $N$ \cite[pp.
+468]{barton_2007}, because Eq.~\eqref{eq:canonical} is unaffected by changes in $N$ if the mechanism regulating density applies equally to all types \citep{prout_1980}.
+Yet, while this does allow us to relax the restrictive assumption of constant $N$, it does not address the assumption of constant $s$.
+In the examples from the previous paragraph, different types \textit{are} affected by the regulation of density in different ways; indeed, the type-specific responses to density  are at the center of MacArthur's argument and the literature on density-dependent selection that grew out of it.
+
+Second, constant $N$ and $s$ can both be seen as a short-term linear approximation \cite[pp.
 277]{ewens_2004}.
-That is, within a sufficiently short time frame, population size can be treated as constant and selection quantified in terms of constant selection coefficients expressing relative fitness differences.
-Provided that selection is sufficiently weak and stable over time, this ``short-term''  assumption is not a major restriction.
-Yet it is increasingly recognized that selection is not always weak, that it can fluctuate considerably over time, and that $N$ can vary by orders of magnitude over a few generations as a routine feature of a population's ecology.
-These are not rare exceptions, but occur widely in nature and the lab, including in wild \textit{Drosophila} \citep{messer_2016}.
-The short-term approximation also precludes consideration of inherently long-term evolutionary processes like the management of genetic load and population extinction \citep{bertram_2017}.
-Nevertheless, relative fitness models like Wright-Fisher are the foundation for much of the population genetic literature, and are still widely used without considering the ``short-term'' restriction.
-Thus, to summarize, there are important theoretical and practical reasons to deepen our understanding of the population-ecological foundations of relative fitness models.
+That is, within a sufficiently short time frame, $N$ and $s$ can be treated as constant.
+Provided that selection is sufficiently weak and stable over time, this ``short-term''  assumption is not a major restriction \cite[PAGE NUMBER]{crow_1970}.
+Yet it is increasingly recognized that selection is not always weak, that it can fluctuate considerably over time \citep{messer_2016}, and that $N$ can vary by orders of magnitude over a few generations as a routine feature of a population's ecology These are not rare exceptions, but occur widely in nature and the lab, including in wild \textit{Drosophila} \citep{bergland_14}.
+Nevertheless, relative fitness models like Wright-Fisher are the foundation for much of the population genetic literature, and are still widely used without considering the ``short-term'' restriction or the gulf with population ecology \citep{mallet_2012}.
+Thus, it is important that we deepen our  understanding of the population ecological basis of relative fitness models, both to gain insight into their domain of applicability, and as part of the broader challenge of synthesizing ecology and evolution.
+
+%The constant-$N$ relative fitness description of selection also precludes consideration of inherently long-term evolutionary processes like the management of genetic load and population extinction \citep{bertram_2017}.
+Adaptive dynamics provides a powerful framework for modeling the interplay between long-term evolution and population ecology, including complex couplings between evolutionary change and population density.
+However, the focus of this latter work is on trait evolution rather than the underlying genetics, and in particular, selective sweeps are typically subsumed into effectively-instantaneous ``trait substitutions''.
+We emphasize that our focus here is the description of selection itself, which particularly critical for making sense of evolution at the genetic level.
+
+%The limitations of relative fitness models are openly acknowledged, usually with an emphasis on the difficulty of incorporating multiple life cycle ``components'' of fitness (e.g.
+\cite[pp.
+276]{ewens_2004} \citep{prout_1980}).
+
+%in which time advances in discrete non-overlapping generations and fitness is interpreted as a product of fertility and juvenile viability \citep[pp. 185]{crow_1970}.
+
+%But more fundamentally, the constant-$N$, relative fitness description uncouples selection and demography, at odds with the fact that the same birth/death events drive both.
+
+ %The connection between constant-$N$ relative fitness models and the broad class of density-regulated growth models considered by MacArthur (Eq.~\ref{eq:macgeneral}) is not clear, since the former only describe selection along lines defined by $n_1+n_2=N$ for each $N$ (Fig.~\ref{fig:Ksel}b).
+In particular, the $f_1$ and $f_2$ nullclines are not even defined.
 
@@
 Note that due to our assumption of uniform dispersal, the parameter $b_i$ can be thought of as a measure of ``colonization ability'', which combines fecundity and dispersal ability \citep{levins_71,tilman_94,bolker_99}.
-In addition to random dispersal, we will compare our model to perfect directed dispersal, in which each propagule finds an unoccupied territory if one is available ($m_i=b_i$) \citep{chesson_1983}.
 
@@
 
-%type $i$ can win at most $m_i$ territories, yet Eq.~\eqref{eq:lottery} demands $c_i l_i/\sum_j c_j l_j$ of the $U$ unoccupied territories, for any value of $U$.
-
 There is a close connection between the classic lottery model and the Wright-Fisher model of genetic drift \citep{svardal_2015}.
@@
 
-\subsection*{Mean Field Approximation}
+\subsection*{Mean-field approximation of the density-dependent lottery}
 
@@
 
-\subsection*{$K$-selection, $c$-selection and relative fitness}
+\subsection*{$K$-selection versus relative fitness}
 
-We now compare the density-dependent lottery model from the previous section with MacArthur's claims about selection in crowded environments.
+We now compare the density-dependent lottery model from the previous section with MacArthur's analysis of selection in crowded environments \citep{macarthur_1967}.
+MacArthur considers a population with two types that have densities $n_1$ and $n_2$ subject to density-dependent growth described by
+\begin{equation}
+\frac{d n_1}{d t}=f_1(n_1,n_2)\qquad\frac{d n_2}{d t}=f_2(n_1,n_2). \label{eq:macgeneral}
+\end{equation}
+The environment is assumed to remain constant apart from the type densities.
+The functions $f_1$ and $f_2$ must decline to zero if $n_1$ or $n_2$ are sufficiently large, because no population has unlimited resources.
+This defines the nullclines $f_1(n_1,n_2)=0$ and $f_2(n_1,n_2)=0$ in $(n_1,n_2)$ space.
+The outcome of selection is then determined by the relationship between these nullclines.
+Specifically, a type will be excluded if its nullcline is completely contained in the region bounded by the other type's nullcline.
+In other words, for a type to have the possibility of persisting, it must be able to keep growing to higher densities than the other type can tolerate in some region of $(n_1,n_2)$ space (Fig.~\ref{fig:Ksel}a).
 
-As shown in the Introduction, MacArthur's argument revolves around the behaviour of  ecological models of the general form Eq.~\ref{eq:macgeneral}, behaviour which depends on the relationship between the nullclines $f_1(n_1,n_2)=0$ and $f_2(n_1,n_2)=0$.
-To formalize this relationship, MacArthur used the symbol ``$K$'' to label the four intersection points of the nullclines with the $n_1$ and $n_2$ axes, specifically $f_1(K_{11},0)=0$, $f_1(0,K_{12})=0$, $f_2(0,K_{22})=0$ and $f_2(K_{21},0)=0$.
+To formalize the relationship between nullclines, MacArthur used the symbol ``$K$'' to label the four intersection points of the nullclines with the $n_1$ and $n_2$ axes, specifically $f_1(K_{11},0)=0$, $f_1(0,K_{12})=0$, $f_2(0,K_{22})=0$ and $f_2(K_{21},0)=0$.
 These $K$ values determine whether a region of higher-density growth exists for each type, provided that the nullclines are close to being straight lines.
@@
 \begin{align}
-\frac{d n_1}{dt} = r_1(1-\alpha_{11}n_1-\alpha_{12}n_2)n_1 = f_1(n_1,n_2)\nonumber\\
-\frac{d n_2}{dt} = r_2(1-\alpha_{22}n_1-\alpha_{21}n_2)n_2 = f_2(n_1,n_2)\label{eq:LV}
+f_1(n_1,n_2) = r_1(1-\alpha_{11}n_1-\alpha_{12}n_2)n_1\nonumber\\
+f_2(n_1,n_2) = r_2(1-\alpha_{22}n_1-\alpha_{21}n_2)n_2\label{eq:LV}
 \end{align}
@@
 Thus, when MacArthur concludes that  ``fitness is $K$'' in crowded populations \citep[pp.
-149]{macarthur_1967}, it does not imply a dichotomy between $r$-selection (uncrowded) and $K$-selection (crowded) in the sense of selection for greater saturation density.
-MacArthur's argument shows that selection either favors the ability to keep growing at ever higher densities (moving a type's own nullcline outwards), or the ability to suppress the growth of competitors at lower densities (moving the nullcline of competitors inwards) \citep{gill_1974}.
+149]{macarthur_1967}, the meaning is that selection either favors the ability to keep growing at ever higher densities (moving a type's own nullcline outwards), or the ability to suppress the growth of competitors at lower densities (moving the nullcline of competitors inwards) \citep{gill_1974}.
 This general idea applies even if the nullclines are nonlinear to such an extent that the ``$K$'' values themselves do not give much information about the regions of high-density growth.
@@
 
-Constant-$N$ relative fitness models are not compatible with $K$-selection, and must refer to the competitive effects summarized by $K_{12}$ and $K_{21}$.
-We thus now restrict our attention to the case $K_{11}=K_{22}$.
-Fig.~\ref{fig:LVvslottery} shows two examples of selection between two types with the same saturation density.
-Fig.~\ref{fig:LVvslottery}a shows exclusion by a superior Lotka-Volterra competitor ($\alpha_{21}>\alpha_{12}$).
-Even though the saturation densities are identical, the density trajectories do not follow a line of constant $N$.
-This is a general properly of the Lotka-Volterra competition model; only a highly restricted subset of $r$ and $\alpha$ values will keep $N$ constant over a selective sweep (further details in Appendix C).
-Intuitively, for one type to exclude another with the same saturation density, inter-type competitive effects must be stronger than intra-type competitive effects, causing a dip in $N$ over a selective sweep.
+It is obvious from Eq.~\eqref{eq:LV} that selection can favor a superior competitor in a crowded population even if its saturation density is the same as, or lower than that of the other types present.
+However, note that the Lotka-Volterra model still closely couples selection to population density.
+Fig.~\ref{fig:LVvslottery}a shows Lotka-Volterra selection between two types with the same saturation density ($\alpha_{11}>\alpha_{22}$, $\alpha_{21}>\alpha_{12}$).
+Even though the initial and final densities of a sweep are the same, density is not constant during a sweep.
+Only a highly restricted subset of $r$ and $\alpha$ values will keep $N$ constant over a selective sweep (further details in Appendix C).
+Intuitively, for one type to exclude another with the same saturation density, inter-type competitive effects must be stronger than intra-type competitive effects, causing a dip in $N$ over the sweep.
 
-By contrast, if one type in our density-dependent lottery model has a $c$ advantage but birth and death rates are identical, the density trajectories converge on the line of constant density equal to the saturation density.
+By contrast, if one type in our density-dependent lottery model has a $c$ advantage but birth and death rates are identical, the density trajectories converge on the line of constant density equal to the saturation density Fig.~\ref{fig:LVvslottery}b.
 Selection then occurs purely along this line, uncoupled from the density regulation of $N$.
@@
 More generally, the competitive ability trait $c$ does not directly affect population density (this can be seen formally in Eq.~\eqref{eq:Nmaster}), since $c$ only affects which type wins a territory, not whether a territory is won at all.
-In other words, $c$ is a zero-sum relative fitness trait embedded within a non-zero-sum fitness model.
+This is all perfectly consistent with MacArthur's general argument.
 
-Thus, we can conclude that relative fitness models are perfectly consistent with MacArthur's claim that ``fitness is $K$'' in crowded populations, bearing in mind his broad use of the symbol ``$K$''.
-However, inconsistency remains with standard growth models such as the Lotka-Volterra model.
-We revisit this issue in the Discussion.
+\section*{The limits of relative fitness}
 
-%, we will evaluate the broader ecological plausibility of relative fitness models from the basis of our density-dependent lottery model.
+We are now in a position to analyze the validity of relative fitness models more rigorously.
+To do so, we evaluate some objections to Eq.~\eqref{eq:canonical} in a crowded poulation using our density-dependent lottery model as a benchmark to better understand the population ecological implications.
+
+We first set aside the most obvious objection: the selection coefficient will often be density-dependent because crowded and uncrowded conditions are so different.
+This objection cannot be ignored if $N$ changes substantially during a sweep.
+We return to this in the Discussion.
+For now we restrict our attention to the case that $N$ is at its equilibrium value.
+
+With $N$ at equilibrium, the presence of density-dependent selection is beside the point.
+As we have seen in our lottery model, the importance of $c$ depends on density, yet since $c$ has no effect on density, $s$ is exactly constant for $c$ sweeps.
+For $b$ and $d$ sweeps, $s$ is also exactly consant, because the overall regulation of population density applies equally to all types \citep{prout_1980}.
+To see this formally, we set $c_i$ to be the same for all types in Eq.~\eqref{eq:master} which yields
+\begin{equation}
+\Delta n_i = \frac{m_i}{M}(1-e^{-L})U-d_i n_i = \left(b_i\frac{1-e^{-L}}{L}-d_i \right) n_i.
+\end{equation}
+Thus, the density-regulation of population size occurs via the juvenile recruitment fraction $(1-e^{L})/L$.
+This applies uniformly to all types and therefore does not induce density-dependence in $b$ or $d$ selective advantages.
+Thus, $b$, $c$ and $d$ sweeps all obey Eq.~\eqref{eq:canonical} exactly.
+That is, in our lottery model density will adjust itself as a result of selection, selection is density-dependent, and yet Eq.~\eqref{eq:canonical} applies exactly for pure $b$, $c$ and $d$ sweeps.
+
+As noted in the Introduction, widely used models of population growth predict that selection in crowded populations will intrinsically affect population density.
+In the previous section we revisited MacArthur's argument and saw that there is no reason to expect this to be true in general, and indeed the competitive ability trait $c$ in our lottery model does not have this property.
 
@@
 
-If s is small it doesn't matter anyway
-
-%The strength of this assumption is clear if we note that essence of the dichotomy underlying the $r$/$K$ scheme is between interaction-dependent selection and interaction-independent selection.
-That is, selective shifts in frequency are a result of differences in absolute growth rates, but these differences can arise in two logically distinct ways: 1) some types expand more rapidly the absence of interactions between individuals or 2) some types are superior in their interactions with other types.
+It is important to recognize that MacArthur's argument does not imply a dichotomy between $r$-selection (uncrowded) and $K$-selection (crowded), with the latter taken to mean selection for greater saturation density.
+A better dichotomy is between interaction-dependent selection and interaction-independent selection.
+That is, selective shifts in frequency are a result of differences in absolute growth rates, but these differences can arise in two logically distinct ways: 1) some types expand more rapidly in the absence of interactions between individuals or 2) some types are superior in their interactions with other types.
 Population density is a key factor controlling whether individuals interact, thereby setting the relative contributions of these forms of selection.
-But there are obviously important forms of interaction that are not expressible in terms of slanted density isoclines.
-For instance, contests for territory and mates are predominantly relative.
-MacArthur's argument starts with assumptions about how selection depends on density, and then tries to make sense of what this says about the evolution of competitive traits such as ``efficiency'' of consumable resource consumption.
-The study of density-dependent selection should start from a description of the interactions between types, and then analyze how the effects of those interactions vary with density.
 

### 2017-12-05 manuscript.tex
l

@@
 
-In uncrowded populations, relative fitness simply represents differences in intrinsic growth rate. In continuous time, we have the canonical selection equation \citep[pp. 26]{crow_1970}
+In uncrowded populations, relative fitness simply represents differences in intrinsic growth rate.
+In continuous time, we have the canonical selection equation $\frac{d p_i}{dt}=(r_i-\overline{r})p_i$, where $r_i$ is the intrinsic exponential growth rate of type $i$, $p_i$ is its frequency, and $\overline{r}=\sum_i r_i p_i$ is the population mean $r$ \citep[pp.
+26]{crow_1970}.
+If there are two types present, a wildtype and a mutant for instance, then Eq.~\eqref{eq:canonical} can be written as
 \begin{equation}
-\frac{d p_i}{dt}=(r_i-\overline{r})p_i, \label{eq:canonical}
+\frac{d p_i}{dt}=s p_i(1-p_i), \label{eq:canonical}
 \end{equation}
-where $r_i$ is the intrinsic exponential growth rate of type $i$, $p_i$ is its frequency, and $\overline{r}=\sum_i r_i p_i$ is the population mean $r$.
-If there are two types present, a wildtype and a mutant for instance, then Eq.~\eqref{eq:canonical} can be written as $\frac{d p_i}{dt}=s p_i(1-p_i)$, where the constant selection coefficient $s$ is the difference in $r$ between types; the corresponding adaptive sweeps follow a logistic curve.
+where the constant selection coefficient $s$ is the difference in $r$ between types. The corresponding adaptive sweeps follow a logistic curve.
 
@@
 The Lotka-Volterra competition model also couples selection in crowded populations to changes in $N$ except in a few special cases (we return to this in section ``$K$-selection, $c$-selection and relative fitness'').
-Thus, $N$ changes during, and as a result of, the adaptive sweeps described by Eq.~\eqref{eq:canonical}.
+In these examples, both $N$ and $s$ change during, and as a result of, adaptive sweeps.
 It would therefore seem that the ubiquitous constant-$N$, relative fitness description of selection is incompatible with a huge class of population ecological processes occurring in nature and experiments.
@@
 %in which time advances in discrete non-overlapping generations and fitness is interpreted as a product of fertility and juvenile viability \citep[pp. 185]{crow_1970}.
-
-%But more fundamentally, the constant-$N$, relative fitness description uncouples selection and demography, at odds with the fact that the same birth/death events drive both.
 
@@
 
-\section*{The limits of relative fitness}
+\section*{Type-specific responses to density and the strength of selection}
 
-We are now in a position to analyze the validity of relative fitness models more rigorously.
-To do so, we evaluate some objections to Eq.~\eqref{eq:canonical} in a crowded poulation using our density-dependent lottery model as a benchmark to better understand the population ecological implications.
+We are now in a position to analyze the validity of relative fitness models more explicitly.
+In the previous section we showed that selection and the regulation of population density can be completely independent of each other even if population growth is density-regulated, and moreover that MacArthur's argument for the $r$/$K$ scheme never precluded this possibility.
+Nevertheless, selection and density regulation \textit{are} intimately connected in widely used models of population growth.
+To understand why this poses a problem for Eq.~\eqref{eq:canonical}, consider the simple birth-death model \cite[pp.
+20]{kostitzin_1939}
+\begin{equation}
+\frac{d n_i}{dt}=(b_i -\delta_iN) n_i \label{eq:simplebirthdeath}
+\end{equation}
+where $\delta_i$ is the per-capita increase in type $i$'s mortality rate due to crowding (for simplicity, there are no deaths when uncrowded).
+Then, starting from a monomorphic population, the frequency of an adaptive $\delta$-variant $\delta_i\rightarrow \delta_i(1-\epsilon)$ obeys
+\begin{equation}
+\frac{d p_i}{dt}=\epsilon \delta_i N p_i(1-p_i). \label{eq:Ndependentsweep}
+\end{equation}
+The selection coefficient $s=\epsilon \delta_i N$ thus depends on density (compare \cite[pp.
+29]{crow_1970}).
+On the other hand, the frequency of an adaptive $b$-variant $b_i\rightarrow b_i(1+\epsilon)$ will exactly obey Eq.~\eqref{eq:canonical} with $s=\epsilon b_i$, independent of density.
 
-We first set aside the most obvious objection: the selection coefficient will often be density-dependent because crowded and uncrowded conditions are so different.
-This objection cannot be ignored if $N$ changes substantially during a sweep.
-We return to this in the Discussion.
-For now we restrict our attention to the case that $N$ is at its equilibrium value.
+In practice the density dependence in Eq.~\eqref{eq:Ndependentsweep} only matters if $N$ changes substantially during a sweep.
+This can easily occur if a population is far from demographic equilibrium (we return to this scenario in the Discussion).
+A more serious objection to Eq.~\eqref{eq:canonical} is that it is internally inconsistent even in crowded populations where $N$ has reached equilibrium, because adaptive sweeps in density-dependent traits then induce a change in $N$ and also change themselves in response to $N$.
+To quantify how serious this objection is, we need to account for how much $N$ changes over a sweep.
+In Eq.~\eqref{eq:simplebirthdeath}, the saturation density is $N=b_i/\delta_i$, and so the selection coefficient increases from $s_{\rm initial}$ to $s_{\rm final}=s_{\rm initial}/(1-\epsilon)$ over a sweep (Fig.~\ref{fig:strengthofselection}).
+Thus, Eq.~\eqref{eq:canonical} breaks down if selection is sufficiently strong on the density-dependent mortality rate, with proportional effects of $\epsilon>0.2$ inducing substantial deviations from constant selection.
 
-With $N$ at equilibrium, the presence of density-dependent selection is beside the point.
+\begin{figure}
+\centering
+\includegraphics[scale=0.8]{strengthofselection.pdf}
+\caption{\label{fig:strengthofselection} Proportional change in the selection coefficient over a ``$K$-selection''-type sweep for a type that experiences proportionally $1-\epsilon$ fewer deaths induced by crowding.
+The population is in demographic equilibrium at the start of the sweep.}
+\end{figure}
+
+Let us now contrast the simple linearly density-dependent model Eq.~\eqref{eq:Ndependentsweep} with our density-dependent lottery.
 As we have seen in our lottery model, the importance of $c$ depends on density, yet since $c$ has no effect on density, $s$ is exactly constant for $c$ sweeps.
@@
 Thus, the density-regulation of population size occurs via the juvenile recruitment fraction $(1-e^{L})/L$.
-This applies uniformly to all types and therefore does not induce density-dependence in $b$ or $d$ selective advantages.
-Thus, $b$, $c$ and $d$ sweeps all obey Eq.~\eqref{eq:canonical} exactly.
-That is, in our lottery model density will adjust itself as a result of selection, selection is density-dependent, and yet Eq.~\eqref{eq:canonical} applies exactly for pure $b$, $c$ and $d$ sweeps.
+This applies uniformly to all types and therefore does not induce density-dependence in selective advantages related to $b$ or $d$.
+Note that our lottery model accounts for two key aspects of the interaction between selection and demography: selection is density-dependent, and density will change as a result of selection on absolute fitness traits.
+And yet pure $b$, $c$ and $d$ sweeps all obey Eq.~\eqref{eq:canonical} exactly.
 
-As noted in the Introduction, widely used models of population growth predict that selection in crowded populations will intrinsically affect population density.
-In the previous section we revisited MacArthur's argument and saw that there is no reason to expect this to be true in general, and indeed the competitive ability trait $c$ in our lottery model does not have this property.
+The reason that our lottery model behaves so differently from Eq.~\eqref{eq:simplebirthdeath} is that each type can respond differently to density via $c$ without affecting the regulation of population density.
+That is, our lottery model separates density-dependent selection from ``selection-dependent density'' \citep{prout_1980}, two distinct effects that are conflated in simple growth models with linearly-density-dependent per-capita growth rates.
+If types in the lottery model differ in $c$ as well as $b$ or $d$, then the lottery model will exhibit similar complications as Fig.~\ref{fig:strengthofselection}.
 

### 2017-12-05 strength_of_selection.py

@@ -9,6 +9,23 @@ import numpy as np
 import matplotlib.pyplot as plt
 from scipy.integrate import odeint
 
+#SS figure
+#========================================================
+
+fig1,ax1=plt.subplots(figsize=[4,4])
+
+s=np.linspace(0,0.5,100)
+ax1.plot(s,1/(1-s),'k',linewidth=2)
+
+ax1.set_ylim([1,2])
+ax1.set_xlabel(r"$\epsilon$",fontsize=20)
+ax1.set_ylabel(r"$s_{\rm final}/s_{\rm initial}$",fontsize=20)
+
+plt.tight_layout()
+
+
+plt.savefig('/home/jbertram/repos/densitydependentlottery/strengthofselection.pdf',bbox="tight")
+
 #Logistic model
 #==========================================================
 
@@ -30,9 +47,6 @@ solrel=odeint(g,1/K[0],x)
 plt.plot(x,solK[:,1]/(solK[:,0]+solK[:,1]))
 plt.plot(x,solrel)
 plt.ylim([0,1])
-
-s=np.linspace(0,0.9,100)
-plt.plot(s,s/(1-s)-s)
 
 #lottery model
 #==========================================================

### 2017-12-05 manuscript.tex
l

@@
 Yet MacArthur famously showed that when population growth is density-regulated, selection in crowded populations is intimately connected to the ability to keep growing at higher densities than other types can tolerate \citep{macarthur_1967}.
-The classic example is the logistic model, where the type with the greatest saturation population density ``$K$'' excludes the others.
+The classic example is the logistic model, where the type with the greatest saturation population density ``$K$'' excludes the others (Fig.~\ref{fig:Ksel}a).
 Similarly, the ``$R^*$ rule'', a central tenet of resource competition theory, states that when growth is limited by a single homogeneous consumable resource, the type able to deplete the resource to the lowest equilibrium density $R^*$ excludes the others \citep{grover_1997}.
@@
 In these examples, both $N$ and $s$ change during, and as a result of, adaptive sweeps.
-It would therefore seem that the ubiquitous constant-$N$, relative fitness description of selection is incompatible with a huge class of population ecological processes occurring in nature and experiments.
+It would therefore seem that the ubiquitous constant-$N$, relative fitness description of selection is incompatible with a huge class of population ecological processes occurring in nature and experiments (Fig.~\ref{fig:Ksel}b).
 
@@
 468]{barton_2007}, because Eq.~\eqref{eq:canonical} is unaffected by changes in $N$ if the mechanism regulating density applies equally to all types \citep{prout_1980}.
-Yet, while this does allow us to relax the restrictive assumption of constant $N$, it does not address the assumption of constant $s$.
-In the examples from the previous paragraph, different types \textit{are} affected by the regulation of density in different ways; indeed, the type-specific responses to density  are at the center of MacArthur's argument and the literature on density-dependent selection that grew out of it.
+While this does allow us to relax the restrictive assumption of constant $N$, it does not address the assumption of constant $s$.
+In the examples from the previous paragraph, different types are affected by the regulation of density in different ways; indeed, the type-specific responses to density  are at the center of MacArthur's argument and the literature on density-dependent selection that grew out of it.
 
@@
 That is, within a sufficiently short time frame, $N$ and $s$ can be treated as constant.
-Provided that selection is sufficiently weak and stable over time, this ``short-term''  assumption is not a major restriction \cite[PAGE NUMBER]{crow_1970}.
-Yet it is increasingly recognized that selection is not always weak, that it can fluctuate considerably over time \citep{messer_2016}, and that $N$ can vary by orders of magnitude over a few generations as a routine feature of a population's ecology These are not rare exceptions, but occur widely in nature and the lab, including in wild \textit{Drosophila} \citep{bergland_14}.
+Provided that selection is sufficiently weak and stable over time, this ``short-term''  assumption is not a major restriction.
+Yet it is increasingly recognized that selection is not always weak, that it can fluctuate considerably over time, and that $N$ can vary by orders of magnitude over a few generations as a routine feature of a population's ecology \citep{messer_2016}.
+These are not rare exceptions, but occur widely in nature and the lab, including in wild \textit{Drosophila} \citep{bergland_14}.
 Nevertheless, relative fitness models like Wright-Fisher are the foundation for much of the population genetic literature, and are still widely used without considering the ``short-term'' restriction or the gulf with population ecology \citep{mallet_2012}.
-Thus, it is important that we deepen our  understanding of the population ecological basis of relative fitness models, both to gain insight into their domain of applicability, and as part of the broader challenge of synthesizing ecology and evolution.
+Thus, it is important to understand the population ecological basis of relative fitness models, both to gain insight into their domain of applicability, and as part of the broader challenge of synthesizing ecology and evolution.
 
-%The constant-$N$ relative fitness description of selection also precludes consideration of inherently long-term evolutionary processes like the management of genetic load and population extinction \citep{bertram_2017}.
-Adaptive dynamics provides a powerful framework for modeling the interplay between long-term evolution and population ecology, including complex couplings between evolutionary change and population density.
-However, the focus of this latter work is on trait evolution rather than the underlying genetics, and in particular, selective sweeps are typically subsumed into effectively-instantaneous ``trait substitutions''.
-We emphasize that our focus here is the description of selection itself, which particularly critical for making sense of evolution at the genetic level.
+Here we analyze the population ecology of relative fitness using a novel model of density-dependent population growth based on territorial contests.
+Rather than starting from the standard models of population growth mentioned above, our point of departure  is the classic lottery model \citep{sale_77,chesson_1981}.
+Like the Wright-Fisher model, the classic lottery assumes a saturated population with constant $N$, and fitness involves a product of fertility and juvenile viability \citep[pp.
+185]{crow_1970}.
+Unlike the Wright-Fisher model, generations can overlap in the lottery model.
+We generalize the lottery model to allow population density to vary, and show that this model can be interpreted as a density-dependent generalization of the Wright-Fisher model with overlapping generations.
 
-%The limitations of relative fitness models are openly acknowledged, usually with an emphasis on the difficulty of incorporating multiple life cycle ``components'' of fitness (e.g.
-\cite[pp.
-276]{ewens_2004} \citep{prout_1980}).
-
-%in which time advances in discrete non-overlapping generations and fitness is interpreted as a product of fertility and juvenile viability \citep[pp. 185]{crow_1970}.
-
- %The connection between constant-$N$ relative fitness models and the broad class of density-regulated growth models considered by MacArthur (Eq.~\ref{eq:macgeneral}) is not clear, since the former only describe selection along lines defined by $n_1+n_2=N$ for each $N$ (Fig.~\ref{fig:Ksel}b).
-In particular, the $f_1$ and $f_2$ nullclines are not even defined.
-
-Here we introduce a novel model of density-dependent population growth based on territorial contests, and show that when this model reaches a demographic steady-state, the constant-$N$, relative fitness picture emerges.
-Our model is firmly grounded in population ecology, with fundamental parameters given by birth and death rates, and competitive ability.
-We show that this model can also be interpreted as a density-dependent generalization of the Wright-Fisher model with overlapping generations.
-
+We show that when this model reaches a demographic steady-state, the constant-$N$, relative fitness picture emerges.
 Futhermore, we show that our model is entirely consistent with MacArthur's analysis of selection in crowded populations.
-In particular, we emphasize that MacArthur's argument does not justify the widespread intuition that selection in crowded environments is necessarily connected to achieving greater densities \citep{anderson_1971}.
+In particular, we emphasize that MacArthur's argument does not justify the widespread intuition that selection in crowded environments is necessarily connected to achieving greater densities.
 This is largely an artifact of the models historically used in the density-dependent selection literature, which ignore relative contests.
 
-%This  also seems to be why the density-dependent selection literature expended such considerable effort to showing that evolution under density-dependent selection optimizes population size in some sense \citep{roughgarden_1979} --- a meaningless
-
-Our model is essentially a density-dependent generalization of the classic ecological lottery model \cite{chesson_1981}.
-In the lottery model, mature individuals (``adults'') each require their own territory, whereas newborn individuals (``propagules'') disperse to, and subsequently compete for, territories made available by the death of adults.
-Territorial contest among propagules leaves a single victorious adult per territory, the victor chosen at random from the propagules present, with probabilities weighted by a coefficient for each type representing competitive ability, akin to a weighted lottery \citep{sale_77}.
-
-The classic lottery model assumes a saturated population with constant $N$, and a large number of propagules dispersing to each territory (the Wright-Fisher model makes a similar ``infinite propagule'' assumption to justify sampling with replacement).
-As such, the lottery model breaks down at low densities (few propagules dispersing to each territory).
-Our first task is to analytically extend the classic lottery model to correctly account for low density behavior (sections ``Model'' and ``Mean field approximation'').
-We then...
+ Our first task is to analytically extend the classic lottery model to correctly account for low density behavior (sections ``Model'' and ``Mean field approximation''). We then...
 
@@
 
-\subsection*{Basic assumptions}
+\subsection*{Model assumptions and definitions}
 
@@
 
+%This  also seems to be why the density-dependent selection literature expended such considerable effort to showing that evolution under density-dependent selection optimizes population size in some sense \citep{roughgarden_1979}
+
 It is important to recognize that MacArthur's argument does not imply a dichotomy between $r$-selection (uncrowded) and $K$-selection (crowded), with the latter taken to mean selection for greater saturation density.
@@
 
-In the introduction we mentioned the recurring difficulties with confounding selection and demography in population genetic inference.
-It seems that Eq.
-\eqref{eq:master} or something similar (and hopefully more analytically tractable) is unavoidable for the analysis of time-course genetic data because, fundamentally, selective births and deaths affect both abundances and frequencies, not one or the other in isolation.
-Moreover, some aspects of allele frequency change are intrinsically density-dependent.
-In the classic lottery model, which as we have seen is essentially the Wright-Fisher model with overlapping generations, $b_i$ and $c_i$ are  equivalent in the sense that the number of territorial victories only depends on the product $b_i c_i$ (see ``Model'').
-This is no longer the case in our extension, where $b$ and $c$ specialists can co-exist.
-This ``colonization-competition trade-off'' is well known in the co-existence literature \citep{tilman_94}.
-It and similar forms of ``spatial co-existence'' in stable environments have previously been modeled either with Levin's  qualitative representation of competition \citep{levins_71,tilman_94}, as opposed to the quantitative $c$ of lottery competition, or with a more sophisticated treatment of space (non-uniform dispersal; \citealt{shmida_84,bolker_99}).
-In cyclical environments, polymorphisms can be stabilized by the bounded density effect, which is  completely lost if there is an exclusive focus on allele frequencies \citep{yi_2013}.
-We leave the details of how our model might be applied to inference problems, including the crucial issue of its genetic drift predictions (providing a null model for neutral sites), for future work.
+Another issue with the constant-$N$ relative fitness description of selection is that it precludes consideration of longer-term aspects of the interplay between evolution and ecology such as population extinction.
+Adaptive dynamics currently provides a powerful framework for adresing the complex feedbacks between evolutionary change and population density.
+However, the focus of adatpive dynamics is trait evolution rather than the underlying genetics, and in particular, selective sweeps are typically subsumed into effectively-instantaneous ``trait substitutions''.
+We emphasize that our focus here has been the description of selection itself, which is particularly critical for making sense of evolution at the genetic level.
 

### 2017-12-05 strength_of_selection.py

@@ -12,7 +12,7 @@ from scipy.integrate import odeint
 #SS figure
 #========================================================
 
-fig1,ax1=plt.subplots(figsize=[4,4])
+fig1,ax1=plt.subplots(figsize=[5,4])
 
 s=np.linspace(0,0.5,100)
 ax1.plot(s,1/(1-s),'k',linewidth=2)
@@ -21,10 +21,12 @@ ax1.set_ylim([1,2])
 ax1.set_xlabel(r"$\epsilon$",fontsize=20)
 ax1.set_ylabel(r"$s_{\rm final}/s_{\rm initial}$",fontsize=20)
 
+ax1.xaxis.set_label_coords(0.5, -0.09)
+ax1.yaxis.set_label_coords(-0.1, 0.5)
+
 plt.tight_layout()
 
-
-plt.savefig('/home/jbertram/repos/densitydependentlottery/strengthofselection.pdf',bbox="tight")
+plt.savefig('/home/jason/repos/densitydependentlottery/strengthofselection.pdf',bbox="tight")
 
 #Logistic model
 #==========================================================

### 2017-12-06 manuscript.tex
l

@@
 \title{Density-dependent selection and the limits of relative fitness}
-%\author{Jason Bertram $^{1,\ast}$ \\ 
-%Joanna Masel $^{1}$}
+\author{Jason Bertram $^{1,\ast}$ \\ 
+Joanna Masel $^{1}$}
 
@@
 
-\textit{Keywords}: $r$/$K$ selection, absolute fitness, eco-evo, competition-colonization trade-off, fluctuating selection, storage effect.
+\textit{Keywords}: $r$/$K$ selection, eco-evolutionary synthesis.
 
@@
 
-new model of density-dependent selection by generalizing the fixed-density classic lottery model of territorial acquisition to accommodate arbitrary population densities.
-
-We show that, with density dependence, co-existence is possible in the lottery model in a stable environment.
-
-Inspired by natural \textit{Drosophila} populations, we consider co-existence under strong, seasonally-fluctuating selection coupled to large cycles in population density, and show that co-existence (stable polymorphism) is promoted via a combination of the classic storage effect and density-regulated population growth.
+TBA
 
@@
 In uncrowded populations, relative fitness simply represents differences in intrinsic growth rate.
-In continuous time, we have the canonical selection equation $\frac{d p_i}{dt}=(r_i-\overline{r})p_i$, where $r_i$ is the intrinsic exponential growth rate of type $i$, $p_i$ is its frequency, and $\overline{r}=\sum_i r_i p_i$ is the population mean $r$ \citep[pp.
+In discrete time, the change in frequency of type $i$ is $\Delta p_i=\left(\frac{W_i}{\overline{W}}-1\right) p_i$, where $W_i$ is the intrinsic absolute growth factor of type $i$,  $\overline{W}=\sum_i W_i p_i$ is the population mean $W$.
+Here we can rescale $W$ however we please and replace it with ``relative fitness'' $w$ without affecting the ratio $\frac{W_i}{\overline{W}}=\frac{w_i}{\overline{w}}$.
+In continuous time, the canonical selection equation is $\frac{d p_i}{dt}=(r_i-\overline{r}) p_i$, where $W$ is replaced by the intrinsic exponential growth rate $r$ \citep[pp.
 26]{crow_1970}.
-If there are two types present, a wildtype and a mutant for instance, then Eq.~\eqref{eq:canonical} can be written as
+If there are two types present, a wildtype $j$ and a mutant $i$ for instance, then the continuous time canonical selection equation can be written as
 \begin{equation}
@@
 The difficulty with Eq.~\eqref{eq:canonical} arises in crowded populations.
-It is then commonly assumed that total population density $N$ has reached its equilibrium value, which is assumed to be a fixed constant.
-This is exemplified by the Wright-Fisher model.
+Since crowded and uncrowded conditions are so different, $s$ will often depend on density and cannot be treated as a simple constant (the possible frequency-dependence of $s$ is a separate issue, because this still leaves a closed model in terms of the frequencies).
+This issue is evaded by simply assuming that total population density $N$ has reached its equilibrium value, which is assumed to be a fixed constant.
+The selection coefficient $s$ now parameterizes the rate at which selection changes relative frequencies, but no longer corresponds to differences in intrinsic growth rates $r$.
+
 Yet MacArthur famously showed that when population growth is density-regulated, selection in crowded populations is intimately connected to the ability to keep growing at higher densities than other types can tolerate \citep{macarthur_1967}.
@@
 In these examples, both $N$ and $s$ change during, and as a result of, adaptive sweeps.
-It would therefore seem that the ubiquitous constant-$N$, relative fitness description of selection is incompatible with a huge class of population ecological processes occurring in nature and experiments (Fig.~\ref{fig:Ksel}b).
+It would therefore seem that the ubiquitous constant-$N$, relative fitness description of selection is incompatible with a huge class of the population ecological processes driving selection (Fig.~\ref{fig:Ksel}b).
 
@@
 
-The relative fitness description has been justified in broadly two different ways for crowded populations (we do not address Wagner's [\citeyear{wagner_2010}] measure theorical justification, which explicitly seeks to be independent of population biology and thus falls outside of our scope).
-First, if we only care about type frequencies, we can sometimes simply ignore $N$ \cite[pp.
-468]{barton_2007}, because Eq.~\eqref{eq:canonical} is unaffected by changes in $N$ if the mechanism regulating density applies equally to all types \citep{prout_1980}.
-While this does allow us to relax the restrictive assumption of constant $N$, it does not address the assumption of constant $s$.
-In the examples from the previous paragraph, different types are affected by the regulation of density in different ways; indeed, the type-specific responses to density  are at the center of MacArthur's argument and the literature on density-dependent selection that grew out of it.
+The relative fitness description has been justified in broadly two different ways for crowded populations (we do not discuss Wagner's [\citeyear{wagner_2010}] measure-theoretical justification, which is explicitly independent of population biology and thus falls outside of our scope).
+The first is to assume that selection is independent of density but still allow density to be affected by selection \citep[pp.
+468]{barton_2007} \citep{prout_1980}.
+While this  allows us to relax the assumption of constant $N$, it does not address the problem that $s$  can depend on density.
+In the examples from the previous paragraph, selection is density-dependent; indeed, the type-specific responses to density are at the center of MacArthur's argument and the density-dependent selection literature that grew out of it \citep{roughgarden_1979}.
 
@@
 These are not rare exceptions, but occur widely in nature and the lab, including in wild \textit{Drosophila} \citep{bergland_14}.
-Nevertheless, relative fitness models like Wright-Fisher are the foundation for much of the population genetic literature, and are still widely used without considering the ``short-term'' restriction or the gulf with population ecology \citep{mallet_2012}.
+Nevertheless, relative fitness models are the foundation for much of the population genetic literature, and are still widely used without considering the ``short-term'' restriction or the lack of integration with population ecology \citep{mallet_2012}.
 Thus, it is important to understand the population ecological basis of relative fitness models, both to gain insight into their domain of applicability, and as part of the broader challenge of synthesizing ecology and evolution.
 
+Another issue with the constant-$N$ relative fitness description of selection is that it precludes consideration of longer-term aspects of the interplay between evolution and ecology such as population extinction.
+Adaptive dynamics currently provides a powerful framework for adresing the complex feedbacks between evolutionary change and population density \citep{diekmann_2004}.
+However, the focus of adaptive dynamics is trait evolution rather than the underlying genetics, and in particular, selective sweeps are typically subsumed into effectively-instantaneous ``trait substitutions''.
+We emphasize that our focus here is the description of selection itself, which is particularly critical for making sense of evolution at the genetic level.
+
 Here we analyze the population ecology of relative fitness using a novel model of density-dependent population growth based on territorial contests.
-Rather than starting from the standard models of population growth mentioned above, our point of departure  is the classic lottery model \citep{sale_77,chesson_1981}.
+Rather attempting to make sense of relative fitness in the standard models of population growth mentioned above (e.g.
+\cite{mallet_2012}), we instead do the reverse, and attempt to make population ecological sense of the widely-used Wright-Fisher, constant-$N$, relative fitness model.
+Our starting point is the classic lottery model of territorial contest \citep{sale_77,chesson_1981}.
 Like the Wright-Fisher model, the classic lottery assumes a saturated population with constant $N$, and fitness involves a product of fertility and juvenile viability \citep[pp.
@@
 It is obvious from Eq.~\eqref{eq:LV} that selection can favor a superior competitor in a crowded population even if its saturation density is the same as, or lower than that of the other types present.
-However, note that the Lotka-Volterra model still closely couples selection to population density.
+However, note that the Lotka-Volterra model still closely couples selection to population density \citep{smouse_1976}.
 Fig.~\ref{fig:LVvslottery}a shows Lotka-Volterra selection between two types with the same saturation density ($\alpha_{11}>\alpha_{22}$, $\alpha_{21}>\alpha_{12}$).
@@
 
-\section*{Type-specific responses to density and the strength of selection}
+\subsection*{Density-dependent selection and the strength of selection}
 
-We are now in a position to analyze the validity of relative fitness models more explicitly.
+We are now in a position to analyze the validity of Eq.~\eqref{eq:canonical} more explicitly.
 In the previous section we showed that selection and the regulation of population density can be completely independent of each other even if population growth is density-regulated, and moreover that MacArthur's argument for the $r$/$K$ scheme never precluded this possibility.
@@
 
-The reason that our lottery model behaves so differently from Eq.~\eqref{eq:simplebirthdeath} is that each type can respond differently to density via $c$ without affecting the regulation of population density.
-That is, our lottery model separates density-dependent selection from ``selection-dependent density'' \citep{prout_1980}, two distinct effects that are conflated in simple growth models with linearly-density-dependent per-capita growth rates.
-If types in the lottery model differ in $c$ as well as $b$ or $d$, then the lottery model will exhibit similar complications as Fig.~\ref{fig:strengthofselection}.
-
 \section*{Discussion}
 
-%This  also seems to be why the density-dependent selection literature expended such considerable effort to showing that evolution under density-dependent selection optimizes population size in some sense \citep{roughgarden_1979}
+It is widely recognized that the evolutionary description of selection is poorly integrated with the ecological literature on population growth \citep{mallet_2012}.
+Typically this is cast as a narrow focus on frequencies on the part of evolutionary geneticists, who, perhaps out of necessity, omit the complexities of population ecology.
+While there may be some truth to this, we argue that part of the reason for the disconnect between the evolutionary geneticist's selection and the ecologist's selection is that relative fitness models actually contain important ecological processes omitted in many ecological models, including the classical Lotka-Volterra and logistic models.
+To illustrate, we highlight two important properties of our extended lottery model.
 
-It is important to recognize that MacArthur's argument does not imply a dichotomy between $r$-selection (uncrowded) and $K$-selection (crowded), with the latter taken to mean selection for greater saturation density.
+First, our lottery model contains a ``reproductive excess'' [any good suggestions for citations here, Joanna?] that allows selection (and competition) to occur among juveniles without affecting population density at the adult stage.
+This excess appears when the number of propagules produced is greater than the number of available territories.
+By contrast, in classical ecological growth models only one life-history stage is represented and selection is intrinsically connected to the effects of each type's adult density on the other.
+These are restrictive assumptions --- it is often the case that competition and selection are concentrated at the more fragile juvenile stage \citep{chesson_1983}.
+Reproductive excesses tend to produce strictly-relative lottery-type contests in which fitter types can grow at the expense of others by preferentially filling the available adult ``slots''.
+The number of slots can remain fixed or change independently of the large amount of selection occurring at the juvenile stage.
+Reproductive excesses expand the domain of validity of Eq.~\eqref{eq:canonical} by facilitating selection-independent density \citep{prout_1980}.
+
+Second, density is regulated by fecundity $b$ and mortality $d$ traits that are affected equally by density.
+There are only so many territories, and all territories are equally available to all types.
+This removes the density-dependence that would make selection on $b$ and $d$ exhibit the difficulties shown in Fig.~\ref{fig:strengthofselection} without removing density-dependent selection from the model as whole via $c$ (by contrast, \cite[pp.
+468]{barton_2007} allows density to depend on selection but removes density-dependent selection entirely).
+Although standard ecological models do have type-specific regulation of density (setting aside special parameter choices \citep{mallet_2012}), it is not obvious that we should expect this as a general rule.
+Much of our intuition about type-specific regulation of density (and density-dependent selection in general) is grounded in the logistic model and its diploid bi-allelic variants \citep{roughgarden_1979}, which should hardly be regarded as universal \citep{christiansen_2004,mallet_2012}.
+Consumable resource models are a better guide, because they explicitly model the processes underlying an important form of competition.
+Density regulation is type-specific in well-mixed resource competition models, but for many populations consumable resources are not well-mixed.
+Spatial localization of consumable resources (e.g.
+due to restricted movement of  nutrients through soils) will tend to create a territorial situation similar to the lottery model, where resource competition only occurs locally and is subsumed into the competitive ability $c$.
+Density regulation will then apply at the population level similarly to the lottery model, allowing $N$ to change via $b$ or $d$ without affecting Eq.~\eqref{eq:canonical}.
+
+Relative fitness breaks down if strong selection occurs in a way that changes population density and is also density-dependent (Fig.~\ref{fig:strengthofselection}).
+In our lottery model, the latter can only occur if types differ in $c$ as well as $b$ or $d$.
+The reason for this is that density-dependent ($c$) and density-determining traits ($b$ and $d$) are completely separate.
+
+It is interesting to compare this separation of traits with the $r$/$K$ scheme, often presented as a dichotomy between $r$-selection (uncrowded) and $K$-selection (crowded), with the latter taken to mean selection for greater saturation density (e.g.
+\citep{gill_1974}).
+As we have seen, MacArthur's argument does not justify this dichotomy at all.
 A better dichotomy is between interaction-dependent selection and interaction-independent selection.
-That is, selective shifts in frequency are a result of differences in absolute growth rates, but these differences can arise in two logically distinct ways: 1) some types expand more rapidly in the absence of interactions between individuals or 2) some types are superior in their interactions with other types.
-Population density is a key factor controlling whether individuals interact, thereby setting the relative contributions of these forms of selection.
+That is, selective shifts in frequency are a result of differences in absolute growth rates, but these differences can arise in two logically distinct ways: i) some types expand more rapidly in the absence of interactions between individuals or ii) some types are superior in their interactions with other types.
+In our the lottery model, $b$ and $d$ are interaction-independent traits (both within and between types), whereas $c$ is entirely about interactions between types.
+This logical distinction captures the essence of crowding, namely that crowding forces you to displace others in order to grow.
+Composite parameters like $K$ and $R^{*}$ conflate the regulation of density (factors which cause density trajectories to move ``outwards'' in Fig.~\ref{fig:Ksel}a) with interaction-dependent displacement (factors which cause density trajectories to move diagonally in Fig.~\ref{fig:Ksel}b).
 
-%In the case of exploitation competition for consumable resources, intra- and inter-type competition are connected via resource use ``efficiency'', but obviously this need not be true more generally \citep{gill_1974,case_1974}.
+Relative fitness models truly breaks down if $N$ is far from equilibrium and selection is density-dependent.
+For example, wild \textit{Drosophila} experience huge seasonal boom-bust cycles in population density coupled to strong selection that drives large swings in allele frequency \citep{bergland_14}.
+In this case there is no choice but to abandon relative fitness.
+Our density-dependent lottery model may not be a particularly good description of \textit{Drosophila} ecology, but the close connection between our model and Wright-Fisher is particularly useful, because drift in our model should behave broadly similarly.
+Thus, our model it should provide a useful starting point for analyzing evolution in this and other from-from-equilibrium situations.
 
-%To avoid confusion with other forms of ''$K$-selection'', selection for competitive ability has been called ``$\alpha$-selection'' after the competition coefficients in the Lotka-Volterra equation \citep{gill_1974,case_1974,joshi_2001}.
-However, competitive success as measured by $\alpha$ (i.e.
-the per-capita effect of one type on another type's growth rate) is only partly determined by individual competitive ability --- in the presence of age-structured competition and territoriality, it also includes the ability of each type to produce contestants i.e.
-$b$ in our model.
+\section*{Acknowledgments}
 
-One potential limitation of our model as a general-purpose model of density-dependent selection is its restriction to interference competition between juveniles for durable resources (lottery recruitment to adulthood), analogous to the ubiquitous assumption of viability selection in population genetics \citep[p.
-45]{ewens_2004}.
-In some respects this is the complement of consumable resource competition models, which restrict their attention to indirect exploitation competition, typically without age structure \citep{tilman_1982}.
-In the particular case that consumable resources are spatially localized (e.g.
-due to restricted movement through soils), resource competition and territorial acquisition effectively coincide, and in principle resource competition could be represented by a competitive ability $c$ (or conversely, $c$ should be derivable from resource competition).
-The situation is more complicated if the resources are well-mixed, since, in general, resource levels then need to be explicitly tracked.
-It seems plausible that explicit resource tracking may not be necessary when the focus is on the evolution of similar types that use identical resources rather than the stable co-existence of widely differing species with different resource preferences \citep{ram_2016}.
-We are not aware of any attempts to delineate conditions under which explicit resource tracking is unnecessary even if it is assumed that community structure is ultimately determined by competition for consumable resources.
-More work is needed connecting resource competition models to the density-dependent selection literature, since most of the former has to date been focused on narrower issues of the role of competition at low resource availability and in the absence of direct interactions between organisms at the same trophic level \citep{aerts_1999,davis_1998,tilman_2007}.
-
-While our model can be applied to species rather than types (e.g.
-ecological invasions), our focus is type evolution i.e.
-the change in allele frequencies over time.
-Our assumption that there are no large $c$ discrepancies (section ``Mean field approximation'') amounts to a restriction on the amount of genetic variation in $c$ in the population.
-Since beneficial mutation effect sizes will typically not be much larger than a few percent, large $c$ discrepancies can only arise if the mutation rate is extremely large, and so the assumption will not be violated in most cases.
-However, this restriction could become important when looking at species interactions rather than type evolution.
-
-Another issue with the constant-$N$ relative fitness description of selection is that it precludes consideration of longer-term aspects of the interplay between evolution and ecology such as population extinction.
-Adaptive dynamics currently provides a powerful framework for adresing the complex feedbacks between evolutionary change and population density.
-However, the focus of adatpive dynamics is trait evolution rather than the underlying genetics, and in particular, selective sweeps are typically subsumed into effectively-instantaneous ``trait substitutions''.
-We emphasize that our focus here has been the description of selection itself, which is particularly critical for making sense of evolution at the genetic level.
-
-%\section*{Acknowledgments}
-
-%We thank Peter Chesson and Joachim Hermisson for many constructive comments on this manuscript. This work was financially supported by the National Science Foundation (DEB-1348262).
+We thank Peter Chesson and Joachim Hermisson for many constructive comments on this manuscript. This work was financially supported by the National Science Foundation (DEB-1348262).
 

### 2017-12-08 manuscript.tex
l

@@
 In uncrowded populations, relative fitness simply represents differences in intrinsic growth rate.
-In discrete time, the change in frequency of type $i$ is $\Delta p_i=\left(\frac{W_i}{\overline{W}}-1\right) p_i$, where $W_i$ is the intrinsic absolute growth factor of type $i$,  $\overline{W}=\sum_i W_i p_i$ is the population mean $W$.
+In discrete time, the change in frequency of type $i$ is $\Delta p_i=\left(\frac{W_i}{\overline{W}}-1\right) p_i$, where $W_i$ is the intrinsic absolute growth factor of type $i$, and $\overline{W}=\sum_i W_i p_i$ is the population mean $W$.
 Here we can rescale $W$ however we please and replace it with ``relative fitness'' $w$ without affecting the ratio $\frac{W_i}{\overline{W}}=\frac{w_i}{\overline{w}}$.
@@
 26]{crow_1970}.
-If there are two types present, a wildtype $j$ and a mutant $i$ for instance, then the continuous time canonical selection equation can be written as
+If there are two types present, a wildtype and a mutant for instance, then the continuous time canonical selection equation can be written as
 \begin{equation}
@@
 The difficulty with Eq.~\eqref{eq:canonical} arises in crowded populations.
-Since crowded and uncrowded conditions are so different, $s$ will often depend on density and cannot be treated as a simple constant (the possible frequency-dependence of $s$ is a separate issue, because this still leaves a closed model in terms of the frequencies).
-This issue is evaded by simply assuming that total population density $N$ has reached its equilibrium value, which is assumed to be a fixed constant.
+Since crowded and uncrowded conditions are so different, $s$ will often depend on density.
+Eq.~\eqref{eq:canonical} is then no longer a complete description of selection (we would also need to specify a model for density).
+Note that frequency-dependent selection does not raise similar problems; Eq.~\eqref{eq:canonical} is still a complete description of selection even if its behavior is more complicated compared to the constant-$s$ case.
+Standard population genetics evades the issue of density-dependent selection by simply assuming that total population density $N$ has reached its equilibrium value, which is assumed to be a fixed constant.
 The selection coefficient $s$ now parameterizes the rate at which selection changes relative frequencies, but no longer corresponds to differences in intrinsic growth rates $r$.
 
-Yet MacArthur famously showed that when population growth is density-regulated, selection in crowded populations is intimately connected to the ability to keep growing at higher densities than other types can tolerate \citep{macarthur_1967}.
+Counter to the assumption of constant $N$, MacArthur famously showed that when population growth is density-regulated, selection in crowded populations is intimately connected to the ability to keep growing at higher densities than other types can tolerate \citep{macarthur_1967}.
 The classic example is the logistic model, where the type with the greatest saturation population density ``$K$'' excludes the others (Fig.~\ref{fig:Ksel}a).
@@
 Differences in $R^*$ will often entail differences in saturation density.
-The Lotka-Volterra competition model also couples selection in crowded populations to changes in $N$ except in a few special cases (we return to this in section ``$K$-selection, $c$-selection and relative fitness'').
+The Lotka-Volterra competition model also couples selection in crowded populations to changes in $N$ except in a few special cases (we return to this in section ``$K$-selection versus relative fitness'').
 In these examples, both $N$ and $s$ change during, and as a result of, adaptive sweeps.
-It would therefore seem that the ubiquitous constant-$N$, relative fitness description of selection is incompatible with a huge class of the population ecological processes driving selection (Fig.~\ref{fig:Ksel}b).
+It would therefore seem that the ubiquitous constant-$N$, relative fitness description of selection is incompatible with a huge class of population ecological processes driving selection (Fig.~\ref{fig:Ksel}b).
 
@@
 468]{barton_2007} \citep{prout_1980}.
-While this  allows us to relax the assumption of constant $N$, it does not address the problem that $s$  can depend on density.
+While this  allows us to relax the assumption of constant $N$, it does not address the problem that $s$  can, in reality, depend on density.
 In the examples from the previous paragraph, selection is density-dependent; indeed, the type-specific responses to density are at the center of MacArthur's argument and the density-dependent selection literature that grew out of it \citep{roughgarden_1979}.
@@
 Another issue with the constant-$N$ relative fitness description of selection is that it precludes consideration of longer-term aspects of the interplay between evolution and ecology such as population extinction.
-Adaptive dynamics currently provides a powerful framework for adresing the complex feedbacks between evolutionary change and population density \citep{diekmann_2004}.
+Adaptive dynamics currently provides a powerful framework for addressing the complex feedbacks between evolutionary change and population density \citep{diekmann_2004}.
 However, the focus of adaptive dynamics is trait evolution rather than the underlying genetics, and in particular, selective sweeps are typically subsumed into effectively-instantaneous ``trait substitutions''.
-We emphasize that our focus here is the description of selection itself, which is particularly critical for making sense of evolution at the genetic level.
+We emphasize that our focus here is the description of the time-dependent process of selection itself, which is particularly critical for making sense of evolution at the genetic level for which we now have abundant data.
 
@@
 
-Our first task is to analytically extend the classic lottery model to correctly account for low density behavior (sections ``Model'' and ``Mean field approximation''). We then...
+Our first task is to analytically extend the classic lottery model to correctly account for low density behavior (sections ). We then...
 
@@
 In particular, we do not evaluate the initial stochastic behaviour of mutant lineages while they are at low abundance.
-We derive deterministic equations for the expected change in the $n_i$ over time, leaving the evaluation of drift for future work.
+We derive equations for the expected change in the $n_i$ over time, leaving the evaluation of drift for future work.
 
@@
 We assume that $x_i$ follows a Poisson distribution $p_i(x_i)=l_i^{x_i} e^{-l_i}/x_i!$, where $l_i=m_i/U$ is the mean territorial propagule density.
-This is approximation becomes exact when the $n_i$ are large enough that drift in $n_i$ can be ignored (Appendix A).
+This approximation becomes exact when the $n_i$ are large enough that drift in $n_i$ can be ignored (Appendix A).
 
@@
 
-\subsection*{Mean-field approximation of the density-dependent lottery}
+\subsection*{Analytical approximation of the density-dependent lottery}
 
@@
 \eqref{eq:growthsumuncoupled} involves an expectation over the time-dependent dispersal distributions $p_i$, and is thus too complicated to give intuition about the dynamics of density-dependent lottery competition.
-We now evaluate this expectation using a ``mean field'' approximation.
+We now evaluate this expectation.
 
@@
 \eqref{eq:growthsumuncoupled} into $x_i=1$ and $x_i>1$ parts.
-Our more general mean field approximation only requires that there are no large discrepancies in competitive ability (i.e.
+Our more general approximation only requires that there are no large discrepancies in competitive ability (i.e.
 we do not have $c_i/c_j\gg 1$ for any two types).
@@
 \includegraphics[scale=0.8]{simulationcomparison.pdf}
-\caption{\label{fig:simcomp} Comparison of mean field approximation Eq.
+\caption{\label{fig:simcomp} Comparison of the analytical approximation Eq.
 \eqref{eq:master} with simulations.
@@
 \end{align}
-where $\alpha_{11}=1/K_{11}$ and $\alpha_{22}=1/K_{22}$ measure competitive effects within each type, while $\alpha_{12}=1/K_{12}$ and $\alpha_{21}=1/K_{21}$ measure competitive effects on the first type due to the second (Fig.~\ref{fig:LVvslottery}a).
+where $\alpha_{11}=1/K_{11}$ and $\alpha_{22}=1/K_{22}$ measure competitive effects within types, while $\alpha_{12}=1/K_{12}$ and $\alpha_{21}=1/K_{21}$ measure competitive effects between types (Fig.~\ref{fig:LVvslottery}a).
 
@@
 However, note that the Lotka-Volterra model still closely couples selection to population density \citep{smouse_1976}.
-Fig.~\ref{fig:LVvslottery}a shows Lotka-Volterra selection between two types with the same saturation density ($\alpha_{11}>\alpha_{22}$, $\alpha_{21}>\alpha_{12}$).
+Fig.~\ref{fig:LVvslottery}a shows Lotka-Volterra selection between two types with the same saturation density ($\alpha_{11}=\alpha_{22}$, $\alpha_{21}>\alpha_{12}$).
 Even though the initial and final densities of a sweep are the same, density is not constant during a sweep.
@@
 
-By contrast, if one type in our density-dependent lottery model has a $c$ advantage but birth and death rates are identical, the density trajectories converge on the line of constant density equal to the saturation density Fig.~\ref{fig:LVvslottery}b.
-Selection then occurs purely along this line, uncoupled from the density regulation of $N$.
+By contrast, if one type in our density-dependent lottery model has a $c$ advantage but birth and death rates are identical (and hence so is saturation density), the density trajectories converge on the line of constant density equal to the saturation density (Fig.~\ref{fig:LVvslottery}b).
+Selection then occurs purely along this line, uncoupled from the regulation of density.
 In other words, once the population reaches demographic equilibrium, it behaves indistinguishably from a constant-$N$ relative fitness model.
 More generally, the competitive ability trait $c$ does not directly affect population density (this can be seen formally in Eq.~\eqref{eq:Nmaster}), since $c$ only affects which type wins a territory, not whether a territory is won at all.
-This is all perfectly consistent with MacArthur's general argument.
+While quite different from classical growth models like the Lotka-Volterra, this is all perfectly consistent with MacArthur's general argument.
 
-\subsection*{Density-dependent selection and the strength of selection}
+\subsection*{Density-dependence and the strength of selection}
 
 We are now in a position to analyze the validity of Eq.~\eqref{eq:canonical} more explicitly.
-In the previous section we showed that selection and the regulation of population density can be completely independent of each other even if population growth is density-regulated, and moreover that MacArthur's argument for the $r$/$K$ scheme never precluded this possibility.
+In the previous section we showed that selection and the regulation of population density can be completely independent of each other even if population growth is density-regulated, and moreover that MacArthur's general argument \citep{macarthur_1967} for the $r$/$K$ scheme never precluded this possibility.
 Nevertheless, selection and density regulation \textit{are} intimately connected in widely used models of population growth.
@@
 This can easily occur if a population is far from demographic equilibrium (we return to this scenario in the Discussion).
-A more serious objection to Eq.~\eqref{eq:canonical} is that it is internally inconsistent even in crowded populations where $N$ has reached equilibrium, because adaptive sweeps in density-dependent traits then induce a change in $N$ and also change themselves in response to $N$.
-To quantify how serious this objection is, we need to account for how much $N$ changes over a sweep.
-In Eq.~\eqref{eq:simplebirthdeath}, the saturation density is $N=b_i/\delta_i$, and so the selection coefficient increases from $s_{\rm initial}$ to $s_{\rm final}=s_{\rm initial}/(1-\epsilon)$ over a sweep (Fig.~\ref{fig:strengthofselection}).
-Thus, Eq.~\eqref{eq:canonical} breaks down if selection is sufficiently strong on the density-dependent mortality rate, with proportional effects of $\epsilon>0.2$ inducing substantial deviations from constant selection.
+A more serious objection to Eq.~\eqref{eq:canonical} is that it is internally inconsistent even when $N$ has reached equilibrium.
+To quantify how serious this objection is, we need to account for how much $N$ changes as a result of a sweep involving a density-regulating trait.
+In Eq.~\eqref{eq:simplebirthdeath}, the saturation density is $N_{\rm initial}=b_i/\delta_i$, which increases to $N_{\rm final}=b_i/\delta_i(1-\epsilon)=N_{\rm initial}/(1-\epsilon)$ over a $\delta$ sweep.
+Thus, the selection coefficient increases from $s_{\rm initial}= \epsilon b_i$ to $s_{\rm final}=s_{\rm initial}/(1-\epsilon)$ over a $\delta$ sweep (Fig.~\ref{fig:strengthofselection}).
+Consequently, Eq.~\eqref{eq:canonical} breaks down if selection is sufficiently strong on $\delta_i$, with proportional effects of $\epsilon>0.2$ inducing substantial deviations from constant selection.
+The cause of this effect is that the mortality rate $\delta_i N$ is both density-regulating and subject to density-dependent selection; thus selection on $\delta$ induces a change in $N$ but also changes in strength in response to $N$.
 
@@
 
-Let us now contrast the simple linearly density-dependent model Eq.~\eqref{eq:Ndependentsweep} with our density-dependent lottery.
+Let us now contrast the simple linearly density-dependent model Eq.~\eqref{eq:simplebirthdeath} with our density-dependent lottery.
 As we have seen in our lottery model, the importance of $c$ depends on density, yet since $c$ has no effect on density, $s$ is exactly constant for $c$ sweeps.
-For $b$ and $d$ sweeps, $s$ is also exactly consant, because the overall regulation of population density applies equally to all types \citep{prout_1980}.
+For $b$ and $d$ sweeps, $s$ is also exactly constant, because the overall regulation of population density applies equally to all types \citep{prout_1980}.
 To see this formally, we set $c_i$ to be the same for all types in Eq.~\eqref{eq:master} which yields
@@
 And yet pure $b$, $c$ and $d$ sweeps all obey Eq.~\eqref{eq:canonical} exactly.
+[Need to flesh this out a little more and address some subtleties with discrete vs.
+continuous time]
 
@@
 
-It is widely recognized that the evolutionary description of selection is poorly integrated with the ecological literature on population growth \citep{mallet_2012}.
-Typically this is cast as a narrow focus on frequencies on the part of evolutionary geneticists, who, perhaps out of necessity, omit the complexities of population ecology.
-While there may be some truth to this, we argue that part of the reason for the disconnect between the evolutionary geneticist's selection and the ecologist's selection is that relative fitness models actually contain important ecological processes omitted in many ecological models, including the classical Lotka-Volterra and logistic models.
-To illustrate, we highlight two important properties of our extended lottery model.
+It is widely recognized that the relative fitness description of selection widely used by evolutionists is poorly integrated with the ecological literature on population growth \citep{mallet_2012}.
+This is not simply a lack of ecological realism on the part of relative fitness models.
+In many population growth models, only one life-history stage is represented, and the competitive effects resulting from crowding appear as a reduction in absolute fitness that only depends on the type densities at this life-history stage (e.g.
+the $n_i^2$ and $n_in_j$ terms in the Lotka Volterra equation).
+However, many species have a ``reproductive excess'' of juveniles that are considerably more fragile than their adult counterparts \citep{chesson_1983}.
+As a result, competition and the selection it induces can concentrated at the juvenile phase, where a large reproductive excess allows for for most juveniles to be competitively eliminated.
+In our lottery model, this excess appears when the number of propagules is greater than the number of available territories.
+Then only $\approx 1/L$ of the juveniles contesting available territories survive to adulthood.
+Note that unlike the role of adult density $n_i$ in single-life-stage models, it is the propagule densities $l_i$ that represent the crowding drives competition.
+In general, reproductive excesses will tend to produce strictly-relative lottery-type contests in which fitter types can grow at the expense of others by preferentially filling the available adult ``slots''.
+The number of slots can remain fixed or change independently of selection at the juvenile stage.
+By ignoring reproductive excesses, single life-stage models are biased to have total population density be sensitive to ongoing selection, unfairly invalidating Eq.~\eqref{eq:canonical}.
+In this respect, the Wright-Fisher model and similar ``viability selection'' models actually capture an important ecological process.
 
-First, our lottery model contains a ``reproductive excess'' [any good suggestions for citations here, Joanna?] that allows selection (and competition) to occur among juveniles without affecting population density at the adult stage.
-This excess appears when the number of propagules produced is greater than the number of available territories.
-By contrast, in classical ecological growth models only one life-history stage is represented and selection is intrinsically connected to the effects of each type's adult density on the other.
-These are restrictive assumptions --- it is often the case that competition and selection are concentrated at the more fragile juvenile stage \citep{chesson_1983}.
-Reproductive excesses tend to produce strictly-relative lottery-type contests in which fitter types can grow at the expense of others by preferentially filling the available adult ``slots''.
-The number of slots can remain fixed or change independently of the large amount of selection occurring at the juvenile stage.
-Reproductive excesses expand the domain of validity of Eq.~\eqref{eq:canonical} by facilitating selection-independent density \citep{prout_1980}.
+Having defended Eq.~\eqref{eq:canonical}
 
-Second, density is regulated by fecundity $b$ and mortality $d$ traits that are affected equally by density.
-There are only so many territories, and all territories are equally available to all types.
-This removes the density-dependence that would make selection on $b$ and $d$ exhibit the difficulties shown in Fig.~\ref{fig:strengthofselection} without removing density-dependent selection from the model as whole via $c$ (by contrast, \cite[pp.
-468]{barton_2007} allows density to depend on selection but removes density-dependent selection entirely).
-Although standard ecological models do have type-specific regulation of density (setting aside special parameter choices \citep{mallet_2012}), it is not obvious that we should expect this as a general rule.
-Much of our intuition about type-specific regulation of density (and density-dependent selection in general) is grounded in the logistic model and its diploid bi-allelic variants \citep{roughgarden_1979}, which should hardly be regarded as universal \citep{christiansen_2004,mallet_2012}.
-Consumable resource models are a better guide, because they explicitly model the processes underlying an important form of competition.
+, relative fitness does indeed break down if strong selection occurs in a way that changes (adult) population density and is also density-dependent (Fig.~\ref{fig:strengthofselection}).
+In our lottery model, the this can only occur if types differ in $c$ as well as $b$ or $d$.
+The reason for this is that density-dependent ($c$) and density-determining traits ($b$ and $d$) are completely separate.
+This may seem odd given the behavior of most models in the density-dependent literature, so it worth investigating this point further.
+
+The essential distinction between selection in uncrowded and crowded conditions is the absence or presence of interactions between types (these interactions may be indirect, as in the case of consumable resource competition).
+That is, selective shifts in frequency are a result of differences in absolute growth rates, but these differences can arise in two logically distinct ways: (i) some types expand more rapidly in the absence of interactions between types or (ii) some types are superior in their interactions with other types.
+This distinction captures a key property of crowding, namely that you must displace others to grow.
+
+By contrast, the separation of traits in the $r$/$K$ scheme is presented as a dichotomy between $r$-selection (uncrowded) and $K$-selection (crowded), with the latter taken to mean selection for greater saturation density (e.g.
+\citep{gill_1974}).
+As we have seen, MacArthur's general argument \citep{macarthur_1967} does not justify this dichotomy, because selection in crowded populations also involves inter-type competition represented by the Lotka-Volterra $\alpha_{ij}$ coefficients; selection on the latter has been termed ``$\alpha$-selection'' \citep{gill_1974,joshi_2001}.
+In the Lotka-Volterra model, $r_i$ belongs to (i) and the inter-type coefficients $\alpha_{ij}$ ($i\neq j$) belong to (ii).
+But $\alpha_{ii}$, the inverse of type $i$'s carrying capacity $K_{ii}$, straddles both (i) and (ii).
+For this reason, $K$-selection --- selection for greater carrying capacity --- is a particularly confusing way to characterize selection in crowded populations.
+
+In our lottery model, $b$ and $d$ are interaction-independent traits, whereas $c$ is entirely about interactions between types.
+
+(factors which cause density trajectories to move ``outwards'' in Fig.~\ref{fig:Ksel}a)
+
+(factors which cause density trajectories to move diagonally in Fig.~\ref{fig:Ksel}b).
+
+\citep{chesson_2000}
+
+Another common feature of density-regulated growth models is the presence of traits which both regulate density and are subject to density-dependent selection, such as the $\delta$ trait in the simple birth-death model Eq.~\eqref{eq:simplebirthdeath}.
+The $b$/$\delta$ parameterization in Eq.~\eqref{eq:simplebirthdeath} emphasizes the obvious point that traits which control density need not be subject to density-dependent selection.
+The $K$ parameter in the standard $r/K$ parameterization of the logistic model has been particularly misleading in this respect, because it is solely responsible for the regulation of density and is subject to density-dependent selection.
+Nevertheless, $\delta$-like traits arise automatically in the simplest linear models of density-regulated growth, including the logistic and Lotka-Volterra models, regardless of how they are parameterized, and are thus deeply ingrained in the literature on density-dependent selection \citep{roughgarden_1979,christiansen_2004,mallet_2012}.
+In our lottery model, $b$ and $d$ control density, but do not exhibit density-dependent selection, and thus do not pose the same threat to Eq.~\eqref{eq:canonical} as a $\delta$-like trait.
+
+It is not obvious that we should expect this as a general rule.
+Much of our intuition about type-specific regulation of density (and density-dependent selection in general)  Consumable resource models are a better guide, because they explicitly model the processes underlying an important form of competition.
 Density regulation is type-specific in well-mixed resource competition models, but for many populations consumable resources are not well-mixed.
@@
 Density regulation will then apply at the population level similarly to the lottery model, allowing $N$ to change via $b$ or $d$ without affecting Eq.~\eqref{eq:canonical}.
-
-Relative fitness breaks down if strong selection occurs in a way that changes population density and is also density-dependent (Fig.~\ref{fig:strengthofselection}).
-In our lottery model, the latter can only occur if types differ in $c$ as well as $b$ or $d$.
-The reason for this is that density-dependent ($c$) and density-determining traits ($b$ and $d$) are completely separate.
-
-It is interesting to compare this separation of traits with the $r$/$K$ scheme, often presented as a dichotomy between $r$-selection (uncrowded) and $K$-selection (crowded), with the latter taken to mean selection for greater saturation density (e.g.
-\citep{gill_1974}).
-As we have seen, MacArthur's argument does not justify this dichotomy at all.
-A better dichotomy is between interaction-dependent selection and interaction-independent selection.
-That is, selective shifts in frequency are a result of differences in absolute growth rates, but these differences can arise in two logically distinct ways: i) some types expand more rapidly in the absence of interactions between individuals or ii) some types are superior in their interactions with other types.
-In our the lottery model, $b$ and $d$ are interaction-independent traits (both within and between types), whereas $c$ is entirely about interactions between types.
-This logical distinction captures the essence of crowding, namely that crowding forces you to displace others in order to grow.
-Composite parameters like $K$ and $R^{*}$ conflate the regulation of density (factors which cause density trajectories to move ``outwards'' in Fig.~\ref{fig:Ksel}a) with interaction-dependent displacement (factors which cause density trajectories to move diagonally in Fig.~\ref{fig:Ksel}b).
 
@@
 
-We thank Peter Chesson and Joachim Hermisson for many constructive comments on this manuscript. This work was financially supported by the National Science Foundation (DEB-1348262).
+We thank Peter Chesson and Joachim Hermisson for many constructive comments on an earlier and quite different version of this manuscript.
+This work was financially supported by the National Science Foundation (DEB-1348262) and the John Templeton Foundation (60814).
 
@@
 
-Here we show that under the Lotka-Volterra model of competition, total density $N$ does not in general remain constant over a selective sweep in a crowded population even if the types have the same saturation density.
+Here we show that under the Lotka-Volterra model of competition, total density $N$ does not in general remain constant over a selective sweep in a crowded population even if the types have the same saturation density (for a related discussion on the density- and frequency-dependence of selection in the Lotka-Volterra mode, see \citep{smouse_1976,mallet_2012}).
 
-We assume $\alpha_{11}=\alpha_{22}$ and $N=1/\alpha_{11}$ and check whether it is then possible for $\frac{dN}{dt}$ to be zero in the sweep ($n_1,n_2 \neq 0$).
+We assume equal effects of crowding within types $\alpha_{11}=\alpha_{22}=\alpha_{\rm intra}$ and $N=1/\alpha_{\rm intra}$ and check whether it is then possible for $\frac{dN}{dt}$ to be zero in the sweep ($n_1,n_2 \neq 0$).
 Substituting these conditions into Eq.~\eqref{eq:LV}, we obtain
 \begin{align}
-\frac{d n_1}{dt} = -r_1(\alpha_{12}-\alpha_{11})n_1n_2 \nonumber\\
-\frac{d n_2}{dt} = -r_2(\alpha_{21}-\alpha_{22})n_1n_2
+\frac{d n_1}{dt} = r_1(\alpha_{11}-\alpha_{12})n_1n_2 \nonumber\\
+\frac{d n_2}{dt} = r_2(\alpha_{22}-\alpha_{21})n_1n_2
 \end{align}
@@
 \begin{equation}
-r_1(\alpha_{12}-\alpha_{11})+r_2(\alpha_{21}-\alpha_{22})=0. \label{eq:constNcondition}
+r_1(\alpha_{\rm intra}-\alpha_{12})+r_2(\alpha_{\rm intra}-\alpha_{21})=0. \label{eq:constNcondition}
 \end{equation}
@@
 Then, according the above condition, for $N$ to remain constant over the sweep, the mutant must find the wildtype more tolerable than itself by exactly the same amount that the wildtype finds the mutant less tolerable than itself.
-This condition, and Eq.~\eqref{eq:constNcondition} more generally, are so restrictive that we can conclude that selective sweeps in the Lotka-Volterra competition model will generally involve non-constant $N$.
+
+Even if we persuaded ourselves that this balance of inter-type interactions is plausible in some circumstances, when multiple types are present the requirement for constant $N$ becomes
+\begin{equation}
+\sum_{ij}r_i(\alpha_{\rm intra}-\alpha_{ij})p_ip_j=0,
+\end{equation}
+which depends on frequency and thus cannot be satisfied in general for constant inter-type coefficients $\alpha_{ij}$.
+We conclude that selection in the Lotka-Volterra competition model will generally involve non-constant $N$.
 

### 2017-12-08 manuscript.tex
l

@@
 Since crowded and uncrowded conditions are so different, $s$ will often depend on density.
-Eq.~\eqref{eq:canonical} is then no longer a complete description of selection (we would also need to specify a model for density).
+Eq.~\eqref{eq:canonical} is then no longer a complete description of selection --- we would also need to specify a model for density.
 Note that frequency-dependent selection does not raise similar problems; Eq.~\eqref{eq:canonical} is still a complete description of selection even if its behavior is more complicated compared to the constant-$s$ case.
@@
 The number of slots can remain fixed or change independently of selection at the juvenile stage.
-By ignoring reproductive excesses, single life-stage models are biased to have total population density be sensitive to ongoing selection, unfairly invalidating Eq.~\eqref{eq:canonical}.
+By ignoring reproductive excesses, single life-stage models are biased to have total population density be sensitive to ongoing selection, artificially invalidating Eq.~\eqref{eq:canonical}.
 In this respect, the Wright-Fisher model and similar ``viability selection'' models actually capture an important ecological process.
 
-Having defended Eq.~\eqref{eq:canonical}
+Having highlighted this redeeming quality of Eq.~\eqref{eq:canonical}, we now turn to its breakdown.
+We first discuss the problem shown in Fig.~\ref{fig:strengthofselection}, which occurs when strong selection changes (adult) population density and is also density-dependent.
+In our lottery model, this can only occur if types differ in $c$ as well as $b$ or $d$.
+Why does our model need joint trait variation when the classical density-dependent selection literature is built on traits that behave like the $\delta$ in Eq.~\eqref{eq:canonical} \citep{christiansen_2004}?
+To briefly review: Based on a diploid, bi-allelic variant of the logistic model, the $r$/$K$ scheme posited a dichotomy between $r$-selection (uncrowded) and $K$-selection (crowded), with the latter taken to mean selection for greater saturation density \citep{gill_1974}.
+A more general Lotka-Volterra model introduces the inter-type $\alpha_{ij}$ competition coefficients, with selection on these termed ``$\alpha$-selection'' \citep{gill_1974,joshi_2001}.
+Thus we are left with $r$-selection (uncrowded) versus $K$- or $\alpha$-selection (crowded).
+The latter two forms of selection in crowded populations behave like selection on $\delta$ in that they are both density-dependent and cause density to change over a sweep (although $N$ only dips temporarily during an $\alpha$-sweep).
 
-, relative fitness does indeed break down if strong selection occurs in a way that changes (adult) population density and is also density-dependent (Fig.~\ref{fig:strengthofselection}).
-In our lottery model, the this can only occur if types differ in $c$ as well as $b$ or $d$.
-The reason for this is that density-dependent ($c$) and density-determining traits ($b$ and $d$) are completely separate.
-This may seem odd given the behavior of most models in the density-dependent literature, so it worth investigating this point further.
+By contrast, the density-dependent ($c$) and density-determining traits ($b$ and $d$) in our lottery model are independent.
+This trait separation highlights the two distinct directions in which density and selection interact: selection may depend on density, and density may depend on selection \citep{prout_1976}.
+Both are needed to pose a threat to Eq.~\eqref{eq:canonical} (recall that we are still assuming $N$ is not far from equilibrium).
+In this respect the simple linear models that have become standard in discussions of density-dependent selection \citep{roughgarden_1979,christiansen_2004,mallet_2012} actually represent a complicated form of the density-selection interaction that confounds a key underlying issue.
 
-The essential distinction between selection in uncrowded and crowded conditions is the absence or presence of interactions between types (these interactions may be indirect, as in the case of consumable resource competition).
-That is, selective shifts in frequency are a result of differences in absolute growth rates, but these differences can arise in two logically distinct ways: (i) some types expand more rapidly in the absence of interactions between types or (ii) some types are superior in their interactions with other types.
-This distinction captures a key property of crowding, namely that you must displace others to grow.
-
-By contrast, the separation of traits in the $r$/$K$ scheme is presented as a dichotomy between $r$-selection (uncrowded) and $K$-selection (crowded), with the latter taken to mean selection for greater saturation density (e.g.
-\citep{gill_1974}).
-As we have seen, MacArthur's general argument \citep{macarthur_1967} does not justify this dichotomy, because selection in crowded populations also involves inter-type competition represented by the Lotka-Volterra $\alpha_{ij}$ coefficients; selection on the latter has been termed ``$\alpha$-selection'' \citep{gill_1974,joshi_2001}.
-In the Lotka-Volterra model, $r_i$ belongs to (i) and the inter-type coefficients $\alpha_{ij}$ ($i\neq j$) belong to (ii).
-But $\alpha_{ii}$, the inverse of type $i$'s carrying capacity $K_{ii}$, straddles both (i) and (ii).
-For this reason, $K$-selection --- selection for greater carrying capacity --- is a particularly confusing way to characterize selection in crowded populations.
-
-In our lottery model, $b$ and $d$ are interaction-independent traits, whereas $c$ is entirely about interactions between types.
-
-(factors which cause density trajectories to move ``outwards'' in Fig.~\ref{fig:Ksel}a)
-
-(factors which cause density trajectories to move diagonally in Fig.~\ref{fig:Ksel}b).
-
-\citep{chesson_2000}
-
-Another common feature of density-regulated growth models is the presence of traits which both regulate density and are subject to density-dependent selection, such as the $\delta$ trait in the simple birth-death model Eq.~\eqref{eq:simplebirthdeath}.
-The $b$/$\delta$ parameterization in Eq.~\eqref{eq:simplebirthdeath} emphasizes the obvious point that traits which control density need not be subject to density-dependent selection.
-The $K$ parameter in the standard $r/K$ parameterization of the logistic model has been particularly misleading in this respect, because it is solely responsible for the regulation of density and is subject to density-dependent selection.
-Nevertheless, $\delta$-like traits arise automatically in the simplest linear models of density-regulated growth, including the logistic and Lotka-Volterra models, regardless of how they are parameterized, and are thus deeply ingrained in the literature on density-dependent selection \citep{roughgarden_1979,christiansen_2004,mallet_2012}.
-In our lottery model, $b$ and $d$ control density, but do not exhibit density-dependent selection, and thus do not pose the same threat to Eq.~\eqref{eq:canonical} as a $\delta$-like trait.
-
-It is not obvious that we should expect this as a general rule.
-Much of our intuition about type-specific regulation of density (and density-dependent selection in general)  Consumable resource models are a better guide, because they explicitly model the processes underlying an important form of competition.
-Density regulation is type-specific in well-mixed resource competition models, but for many populations consumable resources are not well-mixed.
-Spatial localization of consumable resources (e.g.
-due to restricted movement of  nutrients through soils) will tend to create a territorial situation similar to the lottery model, where resource competition only occurs locally and is subsumed into the competitive ability $c$.
-Density regulation will then apply at the population level similarly to the lottery model, allowing $N$ to change via $b$ or $d$ without affecting Eq.~\eqref{eq:canonical}.
-
-Relative fitness models truly breaks down if $N$ is far from equilibrium and selection is density-dependent.
+Relative fitness models truly break down if $N$ is far from equilibrium and selection is density-dependent.
 For example, wild \textit{Drosophila} experience huge seasonal boom-bust cycles in population density coupled to strong selection that drives large swings in allele frequency \citep{bergland_14}.
 In this case there is no choice but to abandon relative fitness.
-Our density-dependent lottery model may not be a particularly good description of \textit{Drosophila} ecology, but the close connection between our model and Wright-Fisher is particularly useful, because drift in our model should behave broadly similarly.
+Whether or not our density-dependent lottery model is a good description of \textit{Drosophila} ecology, the close connection between our model and Wright-Fisher is useful, because drift in our model should behave broadly similarly.
 Thus, our model it should provide a useful starting point for analyzing evolution in this and other from-from-equilibrium situations.

### 2017-12-12 manuscript.tex
l

@@
 
-It is widely recognized that the relative fitness description of selection widely used by evolutionists is poorly integrated with the ecological literature on population growth \citep{mallet_2012}.
+It is widely recognized that the relative fitness description of selection is poorly integrated with the ecological literature on population growth \citep{mallet_2012}.
 This is not simply a lack of ecological realism on the part of relative fitness models.
@@
 the $n_i^2$ and $n_in_j$ terms in the Lotka Volterra equation).
-However, many species have a ``reproductive excess'' of juveniles that are considerably more fragile than their adult counterparts \citep{chesson_1983}.
+However, many species have a ``reproductive excess'' of juveniles that are considerably more fragile than their adult counterparts \citep{chesson_1983,turner1968population,kimura1969natural,nei1971fertility}.
 As a result, competition and the selection it induces can concentrated at the juvenile phase, where a large reproductive excess allows for for most juveniles to be competitively eliminated.
@@
 
-Here we show that under the Lotka-Volterra model of competition, total density $N$ does not in general remain constant over a selective sweep in a crowded population even if the types have the same saturation density (for a related discussion on the density- and frequency-dependence of selection in the Lotka-Volterra mode, see \citep{smouse_1976,mallet_2012}).
+Here we show that under the Lotka-Volterra model of competition, total density $N$ does not in general remain constant over a selective sweep in a crowded population even if the types have the same saturation density (for a related discussion on the density- and frequency-dependence of selection in the Lotka-Volterra model, see \citep{smouse_1976,mallet_2012}).
 

### 2017-12-12 manuscript.tex
l

--- 2017-12-08 manuscript.tex
@@
 
-\textit{Keywords}: $r$/$K$ selection, eco-evolutionary synthesis.
+\textit{Keywords}: Lottery model, Wright-Fisher model, $r$/$K$-selection, $\alpha$-selection, eco-evolutionary synthesis.
 
@@
 
-TBA
+Evolutionary biologists frequently describe selection using relative fitness, and relative fitness models are foundational in population genetics.
+Yet the population ecological basis of relative fitness is poorly understood, and the classical ecology literature on selection in crowded populations seems to be incompatible with widely-used relative fitness models such as the Wright-Fisher.
+Here we develop a generalization of the Wright-Fisher model in which density depends dynamically on the demographic rates of the types present.
+We then explore the population ecology of relative-fitness using this model as a reference point.
+Although density-dependent selection causes relative fitness to break down, we emphasize that relative fitness models are fairly robust in populations close to demographic equilbrium.
+In particular, relative fitness only breaks down if selection is strong and density-dependent, and density is also selection-dependent.
+Our generalized Wright-Fisher model clearly distinguishes these effects, whereas the classical literature on density-dependent selection confounds them.
+We argue that selection-independent density is ecologically plausible for many species given the prevalence of reproductive excesses, an important aspect of selection omitted in many ecological models.
+Our model also offers a natural alternative to relative fitness when the latter is untenable, as is likely the case far from demographic equilbrium.
 
@@
 Another issue with the constant-$N$ relative fitness description of selection is that it precludes consideration of longer-term aspects of the interplay between evolution and ecology such as population extinction.
-Adaptive dynamics currently provides a powerful framework for addressing the complex feedbacks between evolutionary change and population density \citep{diekmann_2004}.
+[will make this sound less dismissive] Adaptive dynamics currently provides a powerful framework for addressing the complex feedbacks between evolutionary change and population density \citep{diekmann_2004}.
 However, the focus of adaptive dynamics is trait evolution rather than the underlying genetics, and in particular, selective sweeps are typically subsumed into effectively-instantaneous ``trait substitutions''.
@@
 the $n_i^2$ and $n_in_j$ terms in the Lotka Volterra equation).
-However, many species have a ``reproductive excess'' of juveniles that are considerably more fragile than their adult counterparts \citep{chesson_1983}.
+However, many species have a ``reproductive excess'' [thanks for the suggestion, still need to do references] of juveniles that are considerably more fragile than their adult counterparts \citep{chesson_1983}.
 As a result, competition and the selection it induces can concentrated at the juvenile phase, where a large reproductive excess allows for for most juveniles to be competitively eliminated.
@@
 
-Having highlighted this redeeming quality of Eq.~\eqref{eq:canonical}, we now turn to its breakdown.
+We now turn to the breakdown of Eq.~\eqref{eq:canonical}.
 We first discuss the problem shown in Fig.~\ref{fig:strengthofselection}, which occurs when strong selection changes (adult) population density and is also density-dependent.
@@
 By contrast, the density-dependent ($c$) and density-determining traits ($b$ and $d$) in our lottery model are independent.
-This trait separation highlights the two distinct directions in which density and selection interact: selection may depend on density, and density may depend on selection \citep{prout_1976}.
+This trait separation highlights the two distinct directions in which density and selection interact: selection may depend on density, and density may depend on selection \citep{prout_1980}.
 Both are needed to pose a threat to Eq.~\eqref{eq:canonical} (recall that we are still assuming $N$ is not far from equilibrium).
 In this respect the simple linear models that have become standard in discussions of density-dependent selection \citep{roughgarden_1979,christiansen_2004,mallet_2012} actually represent a complicated form of the density-selection interaction that confounds a key underlying issue.
+
+Note that while this is a conceptual reason to be wary of the classical density-dependent selection models, it is not clear how we should expect the trait variation in nature to align.
+For instance, should we expect mutations to generally affect $b$, $c$ and $d$ independently of each other, or pleiotropically such that $\delta$-like selection is prevalent?
+In the case of well-mixed consumable resources, the $R^*$ rule  suggests that $\delta$-like selection will be prevalent.
+However, for many populations consumable resources are not well-mixed.
+Spatial localization of consumable resources (e.g.
+due to restricted movement of  nutrients through soils) will tend to create a territorial situation similar to the lottery model, where resource competition only occurs locally and is subsumed into the competitive ability $c$, which does not affect $N$.
 

### 2017-12-13 manuscript.tex
l

--- 2017-12-12 manuscript.tex
@@
 
-\textit{Keywords}: $r$/$K$ selection, eco-evolutionary synthesis.
+\textit{Keywords}: Lottery model, competitive Lotka-Volterra, $r$/$K$-selection, interference competition, $R^*$ theory, eco-evo.
 
@@
 
-TBA
+Evolutionary biologists frequently describe selection using relative fitness, and relative fitness models are foundational in population genetics.
+Yet the population ecological basis of relative fitness is poorly understood, and the classical ecology literature on selection in crowded populations seems to be incompatible with widely-used relative fitness models such as the Wright-Fisher.
+Here we develop a generalization of the Wright-Fisher model in which density depends dynamically on the demographic rates of the types present.
+We then explore the population ecology of relative fitness using this model as a reference point.
+Although density-dependent selection can cause relative fitness to break down, relative fitness models are fairly robust in populations close to demographic equilbrium.
+In particular, relative fitness only breaks down if selection is strong and density-dependent, and density is also selection-dependent.
+Our generalized Wright-Fisher model clearly distinguishes between three demographic parameters, each of whose behavior is, on its own, correctly described by relative fitness models.
+In contrast, the classical literature on density-dependent selection confounds them.
+We argue that selection-independent density is ecologically plausible for many species given the prevalence of reproductive excesses, an important aspect of selection omitted in many ecological models.
+Our model also offers a natural alternative to relative fitness when the latter is untenable, as is likely the case far from demographic equilibrium.
 
@@
 There are a variety of different measures of fitness.
-Some widely used examples in evolutionary ecology are expected lifetime reproductive ratio $R_0$, intrinsic growth rate $r$, saturation population density (often labeled ``$K$'') \citep{benton_2000}, and invasion fitness \citep{metz_1992}.
+Some widely used examples in evolutionary ecology are expected lifetime reproductive ratio $R_0$, intrinsic population growth rate $r$, saturation population density (often labeled ``$K$'') \citep{benton_2000}, and invasion fitness \citep{metz_1992}.
 In addition, ``relative fitness'' is the standard in much of evolutionary biology, particularly evolutionary genetics, where attention is generally restricted to relative genotypic proportions \cite[pp.
@@
 
-In uncrowded populations, relative fitness simply represents differences in intrinsic growth rate.
+In uncrowded populations, relative fitness simply represents differences in intrinsic population growth rate.
 In discrete time, the change in frequency of type $i$ is $\Delta p_i=\left(\frac{W_i}{\overline{W}}-1\right) p_i$, where $W_i$ is the intrinsic absolute growth factor of type $i$, and $\overline{W}=\sum_i W_i p_i$ is the population mean $W$.
@@
 Another issue with the constant-$N$ relative fitness description of selection is that it precludes consideration of longer-term aspects of the interplay between evolution and ecology such as population extinction.
-Adaptive dynamics currently provides a powerful framework for addressing the complex feedbacks between evolutionary change and population density \citep{diekmann_2004}.
+[will make this sound less dismissive] Adaptive dynamics currently provides a powerful framework for addressing the complex feedbacks between evolutionary change and population density, including but not limited to extinction \citep{diekmann_2004}.
 However, the focus of adaptive dynamics is trait evolution rather than the underlying genetics, and in particular, selective sweeps are typically subsumed into effectively-instantaneous ``trait substitutions''.
-We emphasize that our focus here is the description of the time-dependent process of selection itself, which is particularly critical for making sense of evolution at the genetic level for which we now have abundant data.
+We emphasize that our focus here is the description of the time-dependent process by which selection changes allele frequencies, which is particularly critical for making sense of evolution at the genetic level for which we now have abundant data.
 
 Here we analyze the population ecology of relative fitness using a novel model of density-dependent population growth based on territorial contests.
-Rather attempting to make sense of relative fitness in the standard models of population growth mentioned above (e.g.
-\cite{mallet_2012}), we instead do the reverse, and attempt to make population ecological sense of the widely-used Wright-Fisher, constant-$N$, relative fitness model.
+Rather attempting to make sense of relative fitness in existing standard models of population growth mentioned above (e.g.
+\citep{kimura1969natural,mallet_2012}), we instead do the reverse, and attempt to make population ecological sense of the widely-used Wright-Fisher, constant-$N$, relative fitness model.
 Our starting point is the classic lottery model of territorial contest \citep{sale_77,chesson_1981}.
@@
 Unlike the Wright-Fisher model, generations can overlap in the lottery model.
-We generalize the lottery model to allow population density to vary, and show that this model can be interpreted as a density-dependent generalization of the Wright-Fisher model with overlapping generations.
+We generalize the lottery model to allow population density to vary, creating a density-dependent generalization of the Wright-Fisher model with overlapping generations.
 
@@
 In particular, we emphasize that MacArthur's argument does not justify the widespread intuition that selection in crowded environments is necessarily connected to achieving greater densities.
-This is largely an artifact of the models historically used in the density-dependent selection literature, which ignore relative contests.
+This intuition is largely an artifact of the models historically used in the density-dependent selection literature, which ignore the existence of relative contests.
 
-Our first task is to analytically extend the classic lottery model to correctly account for low density behavior (sections ). We then...
+Our first task is to analytically extend the classic lottery model to correctly account for low density behavior (sections ). We then...[BLAH]
 
@@
 We expect that a fraction $p_1(x_1)\ldots p_G(x_G)$ of the $U$ unoccupied territories will have the propagule composition $x_1,\ldots,x_G$.
-type $i$ is expected to win $c_i x_i/\sum_j c_j x_j$ of these.
+Type $i$ is expected to win $c_i x_i/\sum_j c_j x_j$ of these.
 Ignoring fluctuations about these two expectations (due to our no-drift, large $T$, large $n_i$ approximation), type $i$'s territorial acquisition is given by
 \begin{equation}
-\Delta_+ n_i(t)=U(t)\sum_{x_1,\ldots,x_G} \frac{c_i x_i}{\sum_j c_j x_j} p_1(x_1)\ldots p_G(x_G), \label{eq:growthsumuncoupled}
+\Delta_+ n_i(t)=U(t)\sum_{x_1,\ldots,x_G} \frac{c_i x_i}{\sum_j c_j x_j} p_1(x_1)\ldots p_G(x_G) \label{eq:growthsumuncoupled}
 \end{equation}
@@
 \eqref{eq:lottery} breaks down for types with low propagule density ($l_i\ll 1$) because territorial acquisition is then not correctly represented by a lottery in each territory with the mean propagule density.
-Instead, a rare type's propagules only make it to a few territories where at least one of its propagule present.
-In our extension of the classic lottery model, we correct (Eq.~\ref{eq:delttot}) to account for this.
+Instead, the propagules from a low density type only contest a small fraction of territories, and do so with at least one propagule present (not a small fraction of a propagule $l_i\ll 1$).
+In section ``Analytical approximation of the density-dependent lottery'', we present a generalization of \eqref{eq:lottery} that accounts for this low density behavior.
 
@@
 The first, $e^{-L}$, accounts for propagules which land alone on unoccupied territories; these territories are won without contest.
-The second, $R_i c_i/\overline{c}$ represents competitive victories when the $i$ type is a rare invader in a high density population, determining its invasion fitness \citep{metz_1992}.
+The second, $R_i c_i/\overline{c}$, represents competitive victories when the $i$ type is a rare invader in a high density population, determining its invasion fitness \citep{metz_1992}.
 The third term, $A_i c_i/\overline{c}$, represents competitive victories when the $i$ type is abundant.
@@
 We are now in a position to analyze the validity of Eq.~\eqref{eq:canonical} more explicitly.
-In the previous section we showed that selection and the regulation of population density can be completely independent of each other even if population growth is density-regulated, and moreover that MacArthur's general argument \citep{macarthur_1967} for the $r$/$K$ scheme never precluded this possibility.
+In the previous section we showed that selection and the regulation of population density can be completely independent of each other even if population growth is density-regulated, and moreover that MacArthur's general argument \citep{macarthur_1967} never precluded this possibility.
 Nevertheless, selection and density regulation \textit{are} intimately connected in widely used models of population growth.
@@
 \end{equation}
-The selection coefficient $s=\epsilon \delta_i N$ thus depends on density (compare \cite[pp.
-29]{crow_1970}).
+The selection coefficient $s=\epsilon \delta_i N$ thus depends on density (compare with Model III in \cite{kimura1969natural}).
 On the other hand, the frequency of an adaptive $b$-variant $b_i\rightarrow b_i(1+\epsilon)$ will exactly obey Eq.~\eqref{eq:canonical} with $s=\epsilon b_i$, independent of density.
@@
 This can easily occur if a population is far from demographic equilibrium (we return to this scenario in the Discussion).
-A more serious objection to Eq.~\eqref{eq:canonical} is that it is internally inconsistent even when $N$ has reached equilibrium.
-To quantify how serious this objection is, we need to account for how much $N$ changes as a result of a sweep involving a density-regulating trait.
-In Eq.~\eqref{eq:simplebirthdeath}, the saturation density is $N_{\rm initial}=b_i/\delta_i$, which increases to $N_{\rm final}=b_i/\delta_i(1-\epsilon)=N_{\rm initial}/(1-\epsilon)$ over a $\delta$ sweep.
-Thus, the selection coefficient increases from $s_{\rm initial}= \epsilon b_i$ to $s_{\rm final}=s_{\rm initial}/(1-\epsilon)$ over a $\delta$ sweep (Fig.~\ref{fig:strengthofselection}).
-Consequently, Eq.~\eqref{eq:canonical} breaks down if selection is sufficiently strong on $\delta_i$, with proportional effects of $\epsilon>0.2$ inducing substantial deviations from constant selection.
+However, even if $N$ has reached equilibrium, it will change substantially over a $\delta$-sweep if selection on $\delta$ is sufficiently strong.
+To quantify this effect, we need to account for how much $N$ changes as a result of a $\delta$-sweep beginning and ending in equilibrium \citep{kimura1969natural}.
+The saturation density in Eq.~\eqref{eq:simplebirthdeath} is $N_{\rm initial}=b_i/\delta_i$, which increases to $N_{\rm final}=b_i/\delta_i(1-\epsilon)=N_{\rm initial}/(1-\epsilon)$ over the $\delta$-sweep.
+Thus, the selection coefficient increases from $s_{\rm initial}= \epsilon b_i$ to $s_{\rm final}=s_{\rm initial}/(1-\epsilon)$.
+Consequently, substantial deviations from Eq.~\eqref{eq:canonical} occur with proportional changes to $\delta$ of order $\epsilon=0.1$ and upwards (Fig.~\ref{fig:strengthofselection}).
+
 The cause of this effect is that the mortality rate $\delta_i N$ is both density-regulating and subject to density-dependent selection; thus selection on $\delta$ induces a change in $N$ but also changes in strength in response to $N$.
@@
 This is not simply a lack of ecological realism on the part of relative fitness models.
-In many population growth models, only one life-history stage is represented, and the competitive effects resulting from crowding appear as a reduction in absolute fitness that only depends on the type densities at this life-history stage (e.g.
+Rather, in many population growth models, only one life-history stage is represented, and the competitive effects resulting from crowding appear as a reduction in absolute fitness that only depends on the type densities at this life-history stage (e.g.
 the $n_i^2$ and $n_in_j$ terms in the Lotka Volterra equation).
 However, many species have a ``reproductive excess'' of juveniles that are considerably more fragile than their adult counterparts \citep{chesson_1983,turner1968population,kimura1969natural,nei1971fertility}.
-As a result, competition and the selection it induces can concentrated at the juvenile phase, where a large reproductive excess allows for for most juveniles to be competitively eliminated.
+As a result, competition and the selection it induces can be concentrated at the juvenile phase, where a large reproductive excess allows for most juveniles to be competitively eliminated.
 In our lottery model, this excess appears when the number of propagules is greater than the number of available territories.
 Then only $\approx 1/L$ of the juveniles contesting available territories survive to adulthood.
-Note that unlike the role of adult density $n_i$ in single-life-stage models, it is the propagule densities $l_i$ that represent the crowding drives competition.
+Note that unlike the role of adult density $n_i$ in single-life-stage models, it is the propagule densities $l_i$ that represent the crowding that drives competition.
 In general, reproductive excesses will tend to produce strictly-relative lottery-type contests in which fitter types can grow at the expense of others by preferentially filling the available adult ``slots''.
@@
 
-Having highlighted this redeeming quality of Eq.~\eqref{eq:canonical}, we now turn to its breakdown.
+We now turn to the breakdown of Eq.~\eqref{eq:canonical}.
 We first discuss the problem shown in Fig.~\ref{fig:strengthofselection}, which occurs when strong selection changes (adult) population density and is also density-dependent.
-In our lottery model, this can only occur if types differ in $c$ as well as $b$ or $d$.
+In our lottery model, this can only occur if and only if types differ in $c$ as well as $b$ or $d$.
 Why does our model need joint trait variation when the classical density-dependent selection literature is built on traits that behave like the $\delta$ in Eq.~\eqref{eq:canonical} \citep{christiansen_2004}?
@@
 By contrast, the density-dependent ($c$) and density-determining traits ($b$ and $d$) in our lottery model are independent.
-This trait separation highlights the two distinct directions in which density and selection interact: selection may depend on density, and density may depend on selection \citep{prout_1976}.
-Both are needed to pose a threat to Eq.~\eqref{eq:canonical} (recall that we are still assuming $N$ is not far from equilibrium).
+This trait separation highlights the two distinct directions in which density and selection interact: selection may depend on density, and density may depend on selection \citep{prout_1980}.
+The combination is needed to pose a threat to Eq.~\eqref{eq:canonical} (recall that we are still assuming $N$ is not far from equilibrium).
 In this respect the simple linear models that have become standard in discussions of density-dependent selection \citep{roughgarden_1979,christiansen_2004,mallet_2012} actually represent a complicated form of the density-selection interaction that confounds a key underlying issue.
 
-Relative fitness models truly break down if $N$ is far from equilibrium and selection is density-dependent.
+Note that while this is a conceptual reason to be wary of the classical density-dependent selection models, it is not clear how we should expect the trait variation in nature to align.
+For instance, should we expect mutations to generally affect $b$, $c$ and $d$ independently of each other, or pleiotropically such that $\delta$-like selection is prevalent?
+In the case of well-mixed indirect exploitation competition for consumable resources, the $R^*$ rule  suggests that $\delta$-like selection will be prevalent.
+However, for many populations consumable resources are not well-mixed.
+Spatial localization of consumable resources (e.g.
+due to restricted movement of  nutrients through soils) will tend to create a territorial situation similar to the lottery model, where resource competition only occurs locally and both it any interference competition are subsumed into the competitive ability $c$, which does not affect $N$.
+
+Relative fitness models truly break down when $N$ is far from equilibrium and selection is density-dependent.
 For example, wild \textit{Drosophila} experience huge seasonal boom-bust cycles in population density coupled to strong selection that drives large swings in allele frequency \citep{bergland_14}.
-In this case there is no choice but to abandon relative fitness.
+In this case there is no choice but to abandon relative fitness, and our model provides one potentially suitable option.
 Whether or not our density-dependent lottery model is a good description of \textit{Drosophila} ecology, the close connection between our model and Wright-Fisher is useful, because drift in our model should behave broadly similarly.

### 2017-12-14 manuscript.tex
l

@@
 All territories are identical, and the total number of territories is $T$.
-Time $t$ advances in discrete iterations, each representing the time from birth to reproductive maturity.
-In iteration $t$, the number of adults of the $i$'th type is $n_i(t)$, the total number of adults is $N(t)=\sum_i n_i(t)$, and the number of unoccupied territories is $U(t)=T-N(t)$.
+Time advances in discrete iterations, each representing the time from birth to reproductive maturity.
+In a given iteration the number of adults of the $i$'th type will be denoted $n_i$, the total number of adults by $N=\sum_i n_i$, and the number of unoccupied territories by $U=T-N$.
 
@@
 \begin{equation}
-\Delta_+ n_i(t)=U(t)\sum_{x_1,\ldots,x_G} \frac{c_i x_i}{\sum_j c_j x_j} p_1(x_1)\ldots p_G(x_G) \label{eq:growthsumuncoupled}
+\Delta_+ n_i=U\sum_{x_1,\ldots,x_G} \frac{c_i x_i}{\sum_j c_j x_j} p_1(x_1)\ldots p_G(x_G) \label{eq:growthsumuncoupled}
 \end{equation}
@@
 
-Finally, we assume that mortality only occurs in adults (Fig.~\ref{fig:lottery}; setting aside the juvenile deaths implicit in territorial contest), and at a constant, type-specific per-capita rate $0\leq d_i\leq 1$, so that the overall change in type abundances is
+Finally, we assume that mortality only occurs in adults (setting aside the juvenile deaths implicit in territorial contest), and at a constant, type-specific per-capita rate $0\leq d_i\leq 1$, so that the overall change in type abundances is
 \begin{equation}
-\Delta n_i(t)=\Delta_+ n_i(t)-d_i n_i(t). \label{eq:delttot}
+\Delta n_i=\Delta_+ n_i-d_i n_i. \label{eq:delttot}
 \end{equation}
@@
 \begin{equation}
-\Delta_+ n_i(t)=\frac{c_i l_i}{\sum_j c_j l_j}U(t)=\frac{c_i l_i}{\overline{c}L}U(t), \label{eq:lottery}
+\Delta_+ n_i=\frac{c_i l_i}{\sum_j c_j l_j}U=\frac{c_i l_i}{\overline{c}L}U, \label{eq:lottery}
 \end{equation}
 where $\overline{c}=\sum_j c_j m_j/M$ is the mean propagule competitive ability for a randomly selected propagule, $L=M/U$ is the total propagule density and $M=\sum_j m_j$ is the total number of propagules.
-
-Eq.
-\eqref{eq:lottery} breaks down for types with low propagule density ($l_i\ll 1$) because territorial acquisition is then not correctly represented by a lottery in each territory with the mean propagule density.
-Instead, the propagules from a low density type only contest a small fraction of territories, and do so with at least one propagule present (not a small fraction of a propagule $l_i\ll 1$).
-In section ``Analytical approximation of the density-dependent lottery'', we present a generalization of \eqref{eq:lottery} that accounts for this low density behavior.
+In section ``Analytical approximation of the density-dependent lottery'', we derive a generalization of Eq.~\eqref{eq:lottery} that accommodates arbitrary propagule densities $l_i$.
 
@@
 Thus, the classic lottery model is essentially the Wright-Fisher model extended to allow overlapping generations, but ignoring drift.
-This means that our extension of the classic lottery model to arbitrary densities represents a density-dependent generalization of the Wright-Fisher model.
+This means that our extension of the classic lottery model to arbitrary densities represents a density-dependent generalization of the Wright-Fisher model (we also do not consider drift here).
 
@@
 
-Similarly to the high-$l_i$ approximation of classic lottery model, we replace the $x_i$ with appropriate mean values, although we cannot simply replace $x_i$ with $l_i$.
+Similarly to the high-$l_i$ approximation of the classic lottery model, we replace the $x_i$ with appropriate mean values, although we cannot simply replace $x_i$ with $l_i$ as in Eq.~\eqref{eq:lottery}.
+The classic lottery model breaks down for types with low propagule density ($l_i\ll 1$) because territorial acquisition is then not correctly represented by a lottery in each territory with the mean propagule density.
 For a type with low propagule density $l_i\ll 1$, we have $x_i=1$ in the territories where its propagules land, and so its growth comes entirely from territories which deviate appreciably from $l_i$.
@@
 \begin{equation}
-\Delta_+ n_i(t)\approx \left[e^{-L}+(R_i+A_i)\frac{c_i}{\overline{c}}\right]l_i U(t), \label{eq:master}
+\Delta_+ n_i\approx \left[e^{-L}+(R_i+A_i)\frac{c_i}{\overline{c}}\right]l_i U, \label{eq:master}
 \end{equation}
@@
 The first, $e^{-L}$, accounts for propagules which land alone on unoccupied territories; these territories are won without contest.
-The second, $R_i c_i/\overline{c}$, represents competitive victories when the $i$ type is a rare invader in a high density population, determining its invasion fitness \citep{metz_1992}.
+The second, $R_i c_i/\overline{c}$, represents competitive victories when the $i$ type is a rare invader in a high density population (i.e.
+it determines invasion fitness \citep{metz_1992}).
 The third term, $A_i c_i/\overline{c}$, represents competitive victories when the $i$ type is abundant.
-The relative importance of these three terms varies with both the overall propagule density $L$ and the relative propagule frequencies $m_i/M$.
+The relative importance of these three terms varies with both the overall propagule density $L$ and the relative propagule frequencies $l_i/L$.
 If $l_i\gg 1$ for all types, we recover the classic lottery model (only the $A_ic_i/\overline{c}$ term remains, and $A_i\rightarrow 1/L$).
-Note that not all unoccupied territories are claimed each iteration, since under Poisson dispersal a fraction $e^{-L}$ remain unoccupied; total population density thus obeys
-\begin{equation}
-\Delta N=(1-e^{-L})U-\sum_i d_i n_i \label{eq:Nmaster}
-\end{equation}
 
@@
 \end{figure}
+
+Eq.~\eqref{eq:master} takes a much simpler form if all types are competitively equivalent ($c_i=c$),
+\begin{equation}
+\Delta_+ n_i = \frac{l_i}{L}(1-e^{-L})U. \label{eq:masterequalc}
+\end{equation}
+Here $1-e^{-L}$ is the fraction of territories that receive at least one propagule under Poisson dispersal, $(1-e^{-L})U$ is the total number of territories gained, and type $i$ receives $l_i/L$ of these.
+Total population density thus grows according to
+\begin{equation}
+\Delta N=(1-e^{-L})U-\sum_i d_i n_i \label{eq:Nmaster}
+\end{equation}
 
@@
 Fig.~\ref{fig:LVvslottery}a shows Lotka-Volterra selection between two types with the same saturation density ($\alpha_{11}=\alpha_{22}$, $\alpha_{21}>\alpha_{12}$).
-Even though the initial and final densities of a sweep are the same, density is not constant during a sweep.
+Even though the initial and final densities of a sweep are the same, density is not constant over a sweep.
 Only a highly restricted subset of $r$ and $\alpha$ values will keep $N$ constant over a selective sweep (further details in Appendix C).
-Intuitively, for one type to exclude another with the same saturation density, inter-type competitive effects must be stronger than intra-type competitive effects, causing a dip in $N$ over the sweep.
+Intuitively, for one type to exclude another with the same saturation density, competitive suppression of growth between types must be stronger than competitive suppression of growth within types, causing a dip in $N$ over the sweep.
 
-By contrast, if one type in our density-dependent lottery model has a $c$ advantage but birth and death rates are identical (and hence so is saturation density), the density trajectories converge on the line of constant density equal to the saturation density (Fig.~\ref{fig:LVvslottery}b).
-Selection then occurs purely along this line, uncoupled from the regulation of density.
+By contrast, if one type in our density-dependent lottery model has a $c$ advantage but the types are otherwise identical (so that each type has the same saturation density), the density trajectories converge on the line of constant density equal to the saturation density (Fig.~\ref{fig:LVvslottery}b).
+Selection then occurs purely along this line, similarly to Fig.~\ref{fig:Ksel}b.
+This occurs because $c$ does not directly affect $N$: it only affects the relative likelihood for each type to win a contested territory, not whether a territory is contested in the first place (this can be seen formally in Eq.~\eqref{eq:Nmaster}).
 In other words, once the population reaches demographic equilibrium, it behaves indistinguishably from a constant-$N$ relative fitness model.
-More generally, the competitive ability trait $c$ does not directly affect population density (this can be seen formally in Eq.~\eqref{eq:Nmaster}), since $c$ only affects which type wins a territory, not whether a territory is won at all.
 While quite different from classical growth models like the Lotka-Volterra, this is all perfectly consistent with MacArthur's general argument.
@@
 However, even if $N$ has reached equilibrium, it will change substantially over a $\delta$-sweep if selection on $\delta$ is sufficiently strong.
-To quantify this effect, we need to account for how much $N$ changes as a result of a $\delta$-sweep beginning and ending in equilibrium \citep{kimura1969natural}.
-The saturation density in Eq.~\eqref{eq:simplebirthdeath} is $N_{\rm initial}=b_i/\delta_i$, which increases to $N_{\rm final}=b_i/\delta_i(1-\epsilon)=N_{\rm initial}/(1-\epsilon)$ over the $\delta$-sweep.
-Thus, the selection coefficient increases from $s_{\rm initial}= \epsilon b_i$ to $s_{\rm final}=s_{\rm initial}/(1-\epsilon)$.
+To quantify this effect, we need to account for how much $N$ changes as a result of a $\delta$-sweep beginning and ending in equilibrium \citep{kimura1969natural}; from Eq.~\eqref{eq:simplebirthdeath} we have an increase from $N_{\rm initial}=b_i/\delta_i$ to $N_{\rm final}=b_i/\delta_i(1-\epsilon)=N_{\rm initial}/(1-\epsilon)$.
+The corresponding selection coefficient increases from $s_{\rm initial}= \epsilon b_i$ to $s_{\rm final}=s_{\rm initial}/(1-\epsilon)$.
 Consequently, substantial deviations from Eq.~\eqref{eq:canonical} occur with proportional changes to $\delta$ of order $\epsilon=0.1$ and upwards (Fig.~\ref{fig:strengthofselection}).
-
-The cause of this effect is that the mortality rate $\delta_i N$ is both density-regulating and subject to density-dependent selection; thus selection on $\delta$ induces a change in $N$ but also changes in strength in response to $N$.
 
@@
 Let us now contrast the simple linearly density-dependent model Eq.~\eqref{eq:simplebirthdeath} with our density-dependent lottery.
-As we have seen in our lottery model, the importance of $c$ depends on density, yet since $c$ has no effect on density, $s$ is exactly constant for $c$ sweeps.
-For $b$ and $d$ sweeps, $s$ is also exactly constant, because the overall regulation of population density applies equally to all types \citep{prout_1980}.
-To see this formally, we set $c_i$ to be the same for all types in Eq.~\eqref{eq:master} which yields
+Selection on $c$ is clearly density-dependent, but since selection on $c$ does not affect density (section ``$K$-selection versus relative fitness''), selection on $c$ does not introduce density-dependence in Eq.~\eqref{eq:canonical}.
+
+Turning to selection on $b$ and $d$, recall that $m_i=b_i n_i U/T$, and so $L=M/U=\overline{b}N/T$ where $\overline{b}$ is the population mean $b$. Thus, from Eq.~\eqref{eq:masterequalc} we have
 \begin{equation}
-\Delta n_i = \frac{m_i}{M}(1-e^{-L})U-d_i n_i = \left(b_i\frac{1-e^{-L}}{L}-d_i \right) n_i.
+\Delta n_i = \left(\frac{b_i}{\overline{b}}\frac{1-e^{-\overline{b}N/T}}{N}(T-N)-d_i\right)n_i. \label{eq:bdensitydependence}
 \end{equation}
-Thus, the density-regulation of population size occurs via the juvenile recruitment fraction $(1-e^{L})/L$.
-This applies uniformly to all types and therefore does not induce density-dependence in selective advantages related to $b$ or $d$.
-Note that our lottery model accounts for two key aspects of the interaction between selection and demography: selection is density-dependent, and density will change as a result of selection on absolute fitness traits.
-And yet pure $b$, $c$ and $d$ sweeps all obey Eq.~\eqref{eq:canonical} exactly.
-[Need to flesh this out a little more and address some subtleties with discrete vs.
-continuous time]
+The mortality $d$ is akin to the birth rate in Eq.~\eqref{eq:simplebirthdeath}, and so, while $d$ does affect density, selection on $d$ is density independent, and so $d$ sweeps also follow the canonical relative fitness model exactly.
+
+At first glance $b$ in Eq.~\eqref{eq:bdensitydependence} appears to be analogous to the $\delta$ in Eq.~\eqref{eq:simplebirthdeath} because it regulates density and is multiplied by the density-dependent term $f(\overline{b},N)=\frac{1-e^{-\overline{b}N/T}}{N}(T-N)$.
+This term declines from $\overline{b}$ at low density to zero at high density and as a result, selection on $b$ is density-dependent.
+In words: the selective advantage from having greater $b$ depends on the number of territories being contested; if almost all are occupied, then there is little advantage to having a greater $b$.
+
+Nevertheless, the behavior of equilibrium-to-equilibrium $b$-sweeps are qualitatively different from the $\delta$ sweeps above.
+The reason is that $b$ does not regulate density via the $b_i$ term in front of $f(\overline{b},N)$, which is divided by $\overline{b}$ and thus actually reflects relative lottery contests.
+The density regulating effect of $b$ actually appears in the fraction of contested territories $1-e^{-\overline{b}N/T}$.
+This reflects the fact that $b$ sweeps can only increase density by causing more unoccupied territories to receive propagules.
+The net effect on $f(\overline{b},N)$ is precisely zero.
+To see this, note that in a single-type equilibrium we have $b_i/\overline{b}=1$ and so, setting Eq.~\eqref{eq:bdensitydependence} to zero, we $f(\overline{b},N)=d_i$ exactly at the beginning and end of a pure $b$ sweep, even though the density $N$ increases.
+Strictly speaking there is some deviation in $f(N)$ from $d_i$ during the sweep, but this deviation is an order of magnitude smaller than for a $\delta$ sweep (the deviation due to a sweep with proportional effect $b_i\rightarrow b_i(1+\epsilon)$ is only of order $\epsilon^2$, whereas the analogous effect in Fig.~\ref{fig:strengthofselection} is of order $\epsilon$; see Appendix D for details).
+Since selection must already be quite strong for a $\delta$ sweep to threaten Eq.~\eqref{eq:canonical}, we can safely conclude that $b$ sweeps obey Eq.~\eqref{eq:canonical} to a very close approximation.
+
+To summarize: (i) $c$-selection is density-dependent, but $c$ does not affect density (ii) $d$-selection is density-independent but $d$ affects density (iii) $b$ affects density and is subject to density-dependent selection.
+And yet pure $b$, $c$ and $d$ sweeps starting and ending at equilibrium all obey the canonical selection equation exactly (almost exactly for $b$).
 
@@
 
+\section*{Appendix D: Density-dependence of $b$-selection}
+
+In section ``Density-dependence and the strength of selection'' we argued that the density-dependent factor $f(\overline{b},N)$ is unchanged at the beginning and end points of an equilibrium-to-equilibrium $b$.
+Here we estimate the magnitude of the deviation in $f(\overline{b},N)$ during the sweep.
+
+For simplicity, we introduce the notation $D=N/T$ and assume that $D$ is small.
+We can thus make the approximation $1-e^{-\overline{b}D}\approx \overline{b}D$ and $f(\overline{b},N)\approx \overline{b}(1-D)$.
+We expect this to be a conservative approximate based on the worst case scenario, because $N$ is most sensitive to an increase in $b$ in this low-density linear regime.
+We first calculate the value of $f(\overline{b},N)$ at the halfway point in a sweep, where the halfway point is estimated with simple linear averages for $b$ and $N$.
+The sweep is driven by a $b$ variant with $b_j=b_i(1+\epsilon)$, and we denote the corresponding initial and final densities by $D_i$ and $D_j$ respectively, where we have $d_i=b_i(1-D_i)=b_j(1-D_j)$.
+We obtain
+\begin{align}
+f_{\rm{half}}=f(\frac{b_i+b_j}{2},\frac{N_i+N_j}{2})&=\frac{b_i+b_j}{2}\left(1-\frac{D_i+D_j}{2}\right) \nonumber \\
+&=\frac{1}{4} (b_i+b_j)(2-D_i-D_j) \nonumber \\
+&=\frac{1}{4} (2d_i+b_i(1-D_j)+b_j(1-N_i)).
+\end{align}
+Dividing by $d_i$, the proportional deviation in $f(N)$ at the midpoint of the sweep is
+\begin{align}
+\frac{f_{\rm{half}}}{d_i}&=\frac{1}{4}\left(2+\frac{b_i}{b_j}+\frac{b_j}{b_i}\right)\nonumber \\
+&=\frac{1}{4}\left(2+\frac{1}{1+\epsilon}+1+\epsilon\right)\nonumber \\
+&=1+\frac{1}{4}(\epsilon^2-\epsilon^3+\ldots),
+\end{align}
+where we have used the Taylor expansion $\frac{1}{1+\epsilon}=1-\epsilon+\epsilon^2-\epsilon^3+\ldots$.
+
+By contrast, for a $\delta$ sweep in Eq.~\eqref{eq:simplebirthdeath}, the density-dependent term $N$ increases by a factor of $\frac{1}{1-\epsilon}=1+\epsilon+\epsilon^2+\ldots$.
+Thus,  the deviations in $f(N)$ are an order of magnitude smaller than those shown in Fig.~\eqref{fig:strengthofselection}, and even proportional changes of order $\epsilon=0.1$ will cause a negligible deviation from the canonical selection equation.
+
 \end{document}

### 2017-12-15 manuscript.tex
l

@@
 We then explore the population ecology of relative fitness using this model as a reference point.
-Although density-dependent selection can cause relative fitness to break down, relative fitness models are fairly robust in populations close to demographic equilbrium.
+Although density-dependent selection can cause relative fitness to break down, relative fitness models are fairly robust in populations close to demographic equilibrium.
 In particular, relative fitness only breaks down if selection is strong and density-dependent, and density is also selection-dependent.
@@
 The difficulty with Eq.~\eqref{eq:canonical} arises in crowded populations.
-Since crowded and uncrowded conditions are so different, $s$ will often depend on density.
+Since crowded and uncrowded conditions are so different, it seems likely that $s$ will often depend on density (although empirical verification has proved challenging \citep{travis_2013}).
 Eq.~\eqref{eq:canonical} is then no longer a complete description of selection --- we would also need to specify a model for density.
-Note that frequency-dependent selection does not raise similar problems; Eq.~\eqref{eq:canonical} is still a complete description of selection even if its behavior is more complicated compared to the constant-$s$ case.
+Note that frequency-dependent selection does not raise similar problems; Eq.~\eqref{eq:canonical} is still a complete description of selection even if its behavior is more complicated due to $s$ depending on frequency.
 Standard population genetics evades the issue of density-dependent selection by simply assuming that total population density $N$ has reached its equilibrium value, which is assumed to be a fixed constant.
@@
 
-Counter to the assumption of constant $N$, MacArthur famously showed that when population growth is density-regulated, selection in crowded populations is intimately connected to the ability to keep growing at higher densities than other types can tolerate \citep{macarthur_1967}.
+Counter to the assumption of constant $N$, MacArthur famously argued that when population growth is density-regulated, selection in crowded populations is intimately connected to the ability to keep growing at higher densities than other types can tolerate \citep{macarthur_1967}.
 The classic example is the logistic model, where the type with the greatest saturation population density ``$K$'' excludes the others (Fig.~\ref{fig:Ksel}a).
@@
 Differences in $R^*$ will often entail differences in saturation density.
-The Lotka-Volterra competition model also couples selection in crowded populations to changes in $N$ except in a few special cases (we return to this in section ``$K$-selection versus relative fitness'').
+The Lotka-Volterra competition model also couples selection in crowded populations to density except in special cases \citep{smouse_1976,mallet_2012}.
 In these examples, both $N$ and $s$ change during, and as a result of, adaptive sweeps.
@@
 
-The relative fitness description has been justified in broadly two different ways for crowded populations (we do not discuss Wagner's [\citeyear{wagner_2010}] measure-theoretical justification, which is explicitly independent of population biology and thus falls outside of our scope).
-The first is to assume that selection is independent of density but still allow density to be affected by selection \citep[pp.
+In light of these difficulties, the relative fitness description has been justified in broadly three different ways for crowded populations (we do not discuss Wagner's [\citeyear{wagner_2010}] measure-theoretical justification, which is independent of population biology).
+The first is to simply assume that selection is density-independent but relax the assumption of constant $N$ by allowing density to change as a result of selective sweeps \citep[pp.
 468]{barton_2007} \citep{prout_1980}.
-While this  allows us to relax the assumption of constant $N$, it does not address the problem that $s$  can, in reality, depend on density.
-In the examples from the previous paragraph, selection is density-dependent; indeed, the type-specific responses to density are at the center of MacArthur's argument and the density-dependent selection literature that grew out of it \citep{roughgarden_1979}.
+Obviously this  does not address the problem that $s$ can, in reality, depend on density.
+Type-specific responses to density are at the center of MacArthur's argument and the density-dependent selection literature that grew out of it \citep{roughgarden_1979}.
 
-Second, constant $N$ and $s$ can both be seen as a short-term linear approximation \cite[pp.
+The second justification, which grew out of a controversy over Haldane's ``cost of selection'', is to appeal to the existence of a ``reproductive excess'' of juveniles that are more fragile than their adult counterparts \citep{turner1968population,kimura1969natural,nei1971fertility}.
+Selection can then be concentrated at the juvenile phase, uncoupling selection from population density at the adult phase unless it is so strong that the reproductive excess is depleted.
+This justifies the use of Eq.~\eqref{eq:canonical} because, for a population in demographic equilibrium, selective sweeps do not affect density, and so it does not matter if selection is density-dependent.
+However, this reproductive excesses literature was focused on the evaluation of genetic load, and was poorly integrated with population ecology.
+\cite{kimura1969natural} took constant $N$ as a requirement and then introduced some mathematical variants of the logistic model that satisfy this requirement.
+\cite{nei1971fertility} proposed a model with an explicit representation of reproductive excess, but used an unusual model of competition based on pair-wise interactions which was only defined for at most two different types.
+As a result, the role of reproductive excesses in justifying Eq.~\eqref{eq:canonical} is still largely verbal.
+
+Thirdly, Eq.~\eqref{eq:canonical} can justified as a short-term linear approximation \cite[pp.
 277]{ewens_2004}.
@@
 Provided that selection is sufficiently weak and stable over time, this ``short-term''  assumption is not a major restriction.
-Yet it is increasingly recognized that selection is not always weak, that it can fluctuate considerably over time, and that $N$ can vary by orders of magnitude over a few generations as a routine feature of a population's ecology \citep{messer_2016}.
+Yet it is increasingly recognized that selection is not always weak, that it can fluctuate strongly over time, and that $N$ can vary by orders of magnitude over a few generations as a routine feature of a population's ecology \citep{messer_2016}.
 These are not rare exceptions, but occur widely in nature and the lab, including in wild \textit{Drosophila} \citep{bergland_14}.
+
 Nevertheless, relative fitness models are the foundation for much of the population genetic literature, and are still widely used without considering the ``short-term'' restriction or the lack of integration with population ecology \citep{mallet_2012}.
 Thus, it is important to understand the population ecological basis of relative fitness models, both to gain insight into their domain of applicability, and as part of the broader challenge of synthesizing ecology and evolution.
-
-Another issue with the constant-$N$ relative fitness description of selection is that it precludes consideration of longer-term aspects of the interplay between evolution and ecology such as population extinction.
-[will make this sound less dismissive] Adaptive dynamics currently provides a powerful framework for addressing the complex feedbacks between evolutionary change and population density, including but not limited to extinction \citep{diekmann_2004}.
-However, the focus of adaptive dynamics is trait evolution rather than the underlying genetics, and in particular, selective sweeps are typically subsumed into effectively-instantaneous ``trait substitutions''.
-We emphasize that our focus here is the description of the time-dependent process by which selection changes allele frequencies, which is particularly critical for making sense of evolution at the genetic level for which we now have abundant data.
 
@@
 Unlike the Wright-Fisher model, generations can overlap in the lottery model.
-We generalize the lottery model to allow population density to vary, creating a density-dependent generalization of the Wright-Fisher model with overlapping generations.
+In the ``Model'' section and ``Analytical approximation of the density-dependent lottery'' part of the results section, we generalize the lottery model to allow population density to vary, creating a variable-density version of the Wright-Fisher model with overlapping generations.
 
-We show that when this model reaches a demographic steady-state, the constant-$N$, relative fitness picture emerges.
-Futhermore, we show that our model is entirely consistent with MacArthur's analysis of selection in crowded populations.
-In particular, we emphasize that MacArthur's argument does not justify the widespread intuition that selection in crowded environments is necessarily connected to achieving greater densities.
-This intuition is largely an artifact of the models historically used in the density-dependent selection literature, which ignore the existence of relative contests.
+In the ``$K$-selection and selection-dependent density'' section, we show that when our model reaches a demographic steady-state, the constant-$N$, relative fitness picture emerges.
 
-Our first task is to analytically extend the classic lottery model to correctly account for low density behavior (sections ). We then...[BLAH]
+Furthermore, we show that our model is entirely consistent with MacArthur's analysis of selection in crowded populations.
 
@@
 
-\subsection*{Model assumptions and definitions}
+\subsection*{Assumptions and definitions}
 
@@
 Time advances in discrete iterations, each representing the time from birth to reproductive maturity.
-In a given iteration the number of adults of the $i$'th type will be denoted $n_i$, the total number of adults by $N=\sum_i n_i$, and the number of unoccupied territories by $U=T-N$.
-
-We assume that the $n_i$ are large enough that stochastic fluctuations in the $n_i$ (``drift'') can be ignored.
-In particular, we do not evaluate the initial stochastic behaviour of mutant lineages while they are at low abundance.
-We derive equations for the expected change in the $n_i$ over time, leaving the evaluation of drift for future work.
-
-%When considering new mutations, we therefore restrict our attention to the earliest (lowest $n_i$) deterministic behavior of mutant lineages (the transition to deterministic growth occurs at an abundance $n_i$ of order equal to their inverse expected absolute growth rate; \citealt{uecker_2011}).
+In a given iteration, the number of adults of the $i$'th type will be denoted by $n_i$, the total number of adults by $N=\sum_i n_i$, and the number of unoccupied territories by $U=T-N$.
+We assume that the $n_i$ are large enough that stochastic fluctuations in the $n_i$ (``drift'') can be ignored (with $T$ also assumed large to allow for small type densities $n_i/T$).
 
@@
 
-Each iteration, adults produce new offspring (``propagules''), $m_i$ of which disperse to unoccupied territories.
-We assume that adults cannot be ousted from their territories, so that $m_i$ only includes propagules landing on unoccupied territories.
-Propagules disperse at random over the unoccupied territories, regardless of distance from their parents, and independently of each other.
-There is no interaction between propagules (e.g.
-avoidance of territories crowded with propagules).
-Loss of propagules during dispersal is subsumed into $m_i$.
-We assume that each adult produces a constant number $b_i$ of successfully dispersing propagules; the loss of propagules due to dispersal to occupied territories then implies $m_i=b_i(1-N/T)n_i$.
+Each iteration, adults produce new offspring (``propagules'').
+These disperse at random, regardless of distance from their parents, and independently of each other (e.g.
+there is no avoidance of territories crowded with propagules).
+We assume that adults cannot be ousted by juveniles, so that propagules landing on occupied territories are doomed.
+We assume that each adult from type $i$ produces a constant number $b_i$ of successfully dispersing propagules; the number of propagules dispersing to unoccupied territories is then given by $m_i=b_in_iU/T$.
 Note that due to our assumption of uniform dispersal, the parameter $b_i$ can be thought of as a measure of ``colonization ability'', which combines fecundity and dispersal ability \citep{levins_71,tilman_94,bolker_99}.
 
-The number of individuals of the $i$'th type landing in any particular territory is denoted $x_i$.
+The number of propagules of the $i$'th type landing in any particular territory is denoted $x_i$.
 We assume that $x_i$ follows a Poisson distribution $p_i(x_i)=l_i^{x_i} e^{-l_i}/x_i!$, where $l_i=m_i/U$ is the mean territorial propagule density.
-This approximation becomes exact when the $n_i$ are large enough that drift in $n_i$ can be ignored (Appendix A).
+This is strictly only an approximation of random dispersal, but it is an excellent approximation given our assumption that the $n_i$ are large enough that drift can be ignored (Appendix A).
 
@@
 Type $i$ is expected to win $c_i x_i/\sum_j c_j x_j$ of these.
-Ignoring fluctuations about these two expectations (due to our no-drift, large $T$, large $n_i$ approximation), type $i$'s territorial acquisition is given by
+Ignoring fluctuations about these two expectations (due to our no-drift, large $n_i$, large $T$ approximation), type $i$'s territorial acquisition is given by
 \begin{equation}
-\Delta_+ n_i=U\sum_{x_1,\ldots,x_G} \frac{c_i x_i}{\sum_j c_j x_j} p_1(x_1)\ldots p_G(x_G) \label{eq:growthsumuncoupled}
+\Delta_+ n_i=U\sum_{x_1,\ldots,x_G} \frac{c_i x_i}{\sum_j c_j x_j} p_1(x_1)\ldots p_G(x_G), \label{eq:growthsumuncoupled}
 \end{equation}
-in our extended lottery model, where the sum only includes territories with at least one propagule present.
+where the sum only includes territories with at least one propagule present.
 
-Finally, we assume that mortality only occurs in adults (setting aside the juvenile deaths implicit in territorial contest), and at a constant, type-specific per-capita rate $0\leq d_i\leq 1$, so that the overall change in type abundances is
+Finally, we assume that adult mortality only occurs in adults present at the start of the  iteration, and at a constant, type-specific per-capita rate $0\leq d_i\leq 1$.
+Thus, the overall change in type abundances is
 \begin{equation}
@@
 \end{equation}
-where $\overline{c}=\sum_j c_j m_j/M$ is the mean propagule competitive ability for a randomly selected propagule, $L=M/U$ is the total propagule density and $M=\sum_j m_j$ is the total number of propagules.
+where $\overline{c}=\sum_j c_j m_j/M$ is the mean competitive ability for a randomly selected propagule, $L=M/U$ is the total propagule density and $M=\sum_j m_j$ is the total number of propagules.
 In section ``Analytical approximation of the density-dependent lottery'', we derive a generalization of Eq.~\eqref{eq:lottery} that accommodates arbitrary propagule densities $l_i$.
@@
 This is equivalent to the classic lottery model with non-overlapping generations ($d_i=1$ for all $i$) and relative fitness given by $w_i=b_i c_i$ i.e.
-a product of fecundity and viability \citep[pp.
+a product of fertility and viability \citep[pp.
 185]{crow_1970}.
 Thus, the classic lottery model is essentially the Wright-Fisher model extended to allow overlapping generations, but ignoring drift.
-This means that our extension of the classic lottery model to arbitrary densities represents a density-dependent generalization of the Wright-Fisher model (we also do not consider drift here).
+This means that our extension of the classic lottery model to arbitrary densities represents a variable-density generalization of the Wright-Fisher model (we also do not consider drift here).
 
@@
 \end{equation}
-Here $1-e^{-L}$ is the fraction of territories that receive at least one propagule under Poisson dispersal, $(1-e^{-L})U$ is the total number of territories gained, and type $i$ receives $l_i/L$ of these.
+Here $1-e^{-L}$ is the fraction of territories that receive at least one propagule under Poisson dispersal, $(1-e^{-L})U$ is the total number of territories gained, and type $i$ receives a fraction $l_i/L$ of these.
 Total population density thus grows according to
@@
 
-\subsection*{$K$-selection versus relative fitness}
+\subsection*{Selection-dependent density and $K$-selection}
 
-We now compare the density-dependent lottery model from the previous section with MacArthur's analysis of selection in crowded environments \citep{macarthur_1967}.
+Equipped with our variable-density lottery model, we now start evaluating the validity of Eq.~\eqref{eq:canonical}.
+In this section we explore whether we should expect population density to vary as a result of selection \citep{prout_1980}.
+Since the idea that density does vary with selection is closely connected to the notion of ``$K$-selection'', we start by revisiting MacArthur's analysis of selection in crowded environments \citep{macarthur_1967}.
+
 MacArthur considers a population with two types that have densities $n_1$ and $n_2$ subject to density-dependent growth described by
@@
 149]{macarthur_1967}, the meaning is that selection either favors the ability to keep growing at ever higher densities (moving a type's own nullcline outwards), or the ability to suppress the growth of competitors at lower densities (moving the nullcline of competitors inwards) \citep{gill_1974}.
-This general idea applies even if the nullclines are nonlinear to such an extent that the ``$K$'' values themselves do not give much information about the regions of high-density growth.
+This general idea is much broader than ``$K$-selection'' in the sense of selection for greater saturation density, and applies even if the nullclines are nonlinear to such an extent that the ``$K$'' values themselves do not give much information about the regions of high-density growth.
 
@@
 It is obvious from Eq.~\eqref{eq:LV} that selection can favor a superior competitor in a crowded population even if its saturation density is the same as, or lower than that of the other types present.
-However, note that the Lotka-Volterra model still closely couples selection to population density \citep{smouse_1976}.
+However, note that the Lotka-Volterra model still couples selection to population density \citep{smouse_1976}.
 Fig.~\ref{fig:LVvslottery}a shows Lotka-Volterra selection between two types with the same saturation density ($\alpha_{11}=\alpha_{22}$, $\alpha_{21}>\alpha_{12}$).
@@
 
+Note that this behavior of $c$-selection arises directly from the role of $c$ as a trait determining relative competitive success in territorial contests, and as such this behavior is closely related to the classical appeal to reproductive excesses to justify Eq.~\eqref{eq:canonical}.
+By contrast, previous models of selection-independent density either used unusual models of competition \citep{kimura1969natural,nei1971fertility}, or made restrictive parameter choices in the Lotka-Volterra model (Appendix C; \citealt{smouse_1976,mallet_2012}).
+
 \subsection*{Density-dependence and the strength of selection}
 
-We are now in a position to analyze the validity of Eq.~\eqref{eq:canonical} more explicitly.
 In the previous section we showed that selection and the regulation of population density can be completely independent of each other even if population growth is density-regulated, and moreover that MacArthur's general argument \citep{macarthur_1967} never precluded this possibility.
@@
 
+In particular, we emphasize that MacArthur's argument does not justify the widespread intuition that selection in crowded environments is necessarily connected to achieving greater densities.
+This intuition is largely an artifact of the models historically used in the density-dependent selection literature, which ignore the existence of relative contests.
+
 We now turn to the breakdown of Eq.~\eqref{eq:canonical}.
@@
 Thus, our model it should provide a useful starting point for analyzing evolution in this and other from-from-equilibrium situations.
+
+Another issue with the constant-$N$ relative fitness description of selection is that it precludes consideration of longer-term aspects of the interplay between evolution and ecology such as population extinction.
+A variety of approaches have been developed for dealing with these issues in quantitative genetics \citep{burger1995evolution,engen_2013}, population genetics \citep{bertram2017predicting} and adaptive dynamics \citep{ferriere2013eco,dieckmann2004adaptive}.
+Although this manuscript, and density-dependent selection more broadly, are pertinent to the interplay between ecology and evolution \citep{travis_2013}, our focus here is the description of the time-dependent process by which selection changes allele frequencies.
+This is particularly critical for making sense of evolution at the genetic level, for which we now have abundant data.
 

### 2017-12-15 manuscript.tex
l

@@
 
-Thirdly, Eq.~\eqref{eq:canonical} can justified as a short-term linear approximation \cite[pp.
+Third, Eq.~\eqref{eq:canonical} can justified as a short-term linear approximation \cite[pp.
 277]{ewens_2004}.
@@
 Here we analyze the population ecology of relative fitness using a novel model of density-dependent population growth based on territorial contests.
-Rather attempting to make sense of relative fitness in existing standard models of population growth mentioned above (e.g.
-\citep{kimura1969natural,mallet_2012}), we instead do the reverse, and attempt to make population ecological sense of the widely-used Wright-Fisher, constant-$N$, relative fitness model.
+Rather than attempting to make sense of relative fitness in existing standard models of population growth (e.g.
+\citep{kimura1969natural,mallet_2012}), we instead do the reverse, and attempt to make population ecological sense of the widely-used Wright-Fisher relative-fitness model.
 Our starting point is the classic lottery model of territorial contest \citep{sale_77,chesson_1981}.
 Like the Wright-Fisher model, the classic lottery assumes a saturated population with constant $N$, and fitness involves a product of fertility and juvenile viability \citep[pp.
-185]{crow_1970}.
-Unlike the Wright-Fisher model, generations can overlap in the lottery model.
-In the ``Model'' section and ``Analytical approximation of the density-dependent lottery'' part of the results section, we generalize the lottery model to allow population density to vary, creating a variable-density version of the Wright-Fisher model with overlapping generations.
+185]{crow_1970}, but unlike the Wright-Fisher model, generations can overlap.
+Our first task is to generalize the lottery model to create a variable-density version of the Wright-Fisher model with overlapping generations (sections ``Model'' and ``Analytical approximation of the variable-density lottery'').
 
-In the ``$K$-selection and selection-dependent density'' section, we show that when our model reaches a demographic steady-state, the constant-$N$, relative fitness picture emerges.
-
-Furthermore, we show that our model is entirely consistent with MacArthur's analysis of selection in crowded populations.
+Equipped with our new model, we turn to the evaluation of Eq.~\eqref{eq:canonical}.
+We first discuss selection on the ability to contest territories, which behaves like a pure constant-$N$, relative fitness trait, and discuss how this fits with MacArthur's analysis of selection in crowded populations (section ``$K$-selection and selection-dependent density'').
+We then discuss selection on density-regulating traits, contrasting the results from the classical density-dependent selection literature with our model (section ``Density-regulating traits and the threat of strong selection'').
+We conclude [...]
 
@@
 
-\subsection*{Analytical approximation of the density-dependent lottery}
+\subsection*{Analytical approximation of the variable-density lottery}
 
@@
 It is obvious from Eq.~\eqref{eq:LV} that selection can favor a superior competitor in a crowded population even if its saturation density is the same as, or lower than that of the other types present.
-However, note that the Lotka-Volterra model still couples selection to population density \citep{smouse_1976}.
+However, the Lotka-Volterra model still couples selection to population density \citep{smouse_1976}.
 Fig.~\ref{fig:LVvslottery}a shows Lotka-Volterra selection between two types with the same saturation density ($\alpha_{11}=\alpha_{22}$, $\alpha_{21}>\alpha_{12}$).
@@
 
-Note that this behavior of $c$-selection arises directly from the role of $c$ as a trait determining relative competitive success in territorial contests, and as such this behavior is closely related to the classical appeal to reproductive excesses to justify Eq.~\eqref{eq:canonical}.
+The constant-$N$ behavior of $c$-selection arises from the role of $c$ as a trait determining relative competitive success in territorial contests.
+As such, this behavior is a result of a reproductive excesses Eq.~\eqref{eq:canonical}.
 By contrast, previous models of selection-independent density either used unusual models of competition \citep{kimura1969natural,nei1971fertility}, or made restrictive parameter choices in the Lotka-Volterra model (Appendix C; \citealt{smouse_1976,mallet_2012}).
 
-\subsection*{Density-dependence and the strength of selection}
+\subsection*{Density-regulating traits and the threat of strong selection}
 
-In the previous section we showed that selection and the regulation of population density can be completely independent of each other even if population growth is density-regulated, and moreover that MacArthur's general argument \citep{macarthur_1967} never precluded this possibility.
-Nevertheless, selection and density regulation \textit{are} intimately connected in widely used models of population growth.
-To understand why this poses a problem for Eq.~\eqref{eq:canonical}, consider the simple birth-death model \cite[pp.
-20]{kostitzin_1939}
+In the previous section we showed that selection and the regulation of population density can be independent even if population growth is density-regulated.
+Nevertheless, selection and density regulation \textit{are} intimately connected in widely used models of population growth, as well as for the $b$ and $d$ traits in our variable-density lottery model.
+
+To see why this potentially poses a threat to relative fitness, consider the simple birth-death model \cite[pp. 20]{kostitzin_1939} \citep{travis_2013} 
 \begin{equation}
@@
 \end{equation}
-where $\delta_i$ is the per-capita increase in type $i$'s mortality rate due to crowding (for simplicity, there are no deaths when uncrowded).
-Then, starting from a monomorphic population, the frequency of an adaptive $\delta$-variant $\delta_i\rightarrow \delta_i(1-\epsilon)$ obeys
+where $\delta_i$ is the per-capita increase in mortality rate due to crowding (for simplicity, there are no deaths when uncrowded), playing a similar role as $K$ in the logistic model.
+
+Starting from a monomorphic population, the frequency of a $\delta_i\rightarrow \delta_i(1-\epsilon)$ variant obeys 
 \begin{equation}
@@
 The selection coefficient $s=\epsilon \delta_i N$ thus depends on density (compare with Model III in \cite{kimura1969natural}).
-On the other hand, the frequency of an adaptive $b$-variant $b_i\rightarrow b_i(1+\epsilon)$ will exactly obey Eq.~\eqref{eq:canonical} with $s=\epsilon b_i$, independent of density.
+On the other hand, the frequency of a $b_i\rightarrow b_i(1+\epsilon)$ variant will exactly obey Eq.~\eqref{eq:canonical} with $s=\epsilon b_i$, independent of density.
 
@@
 \includegraphics[scale=0.8]{strengthofselection.pdf}
-\caption{\label{fig:strengthofselection} Proportional change in the selection coefficient over a ``$K$-selection''-type sweep for a type that experiences proportionally $1-\epsilon$ fewer deaths induced by crowding.
-The population is in demographic equilibrium at the start of the sweep.}
+\caption{\label{fig:strengthofselection} Proportional change in the selection coefficient over a ``$K$-like'' sweep for a type that experiences proportionally $1-\epsilon$ fewer deaths induced by crowding.
+The population is in demographic equilibrium at the start and end of the sweep.}
 \end{figure}
 
-Let us now contrast the simple linearly density-dependent model Eq.~\eqref{eq:simplebirthdeath} with our density-dependent lottery.
-Selection on $c$ is clearly density-dependent, but since selection on $c$ does not affect density (section ``$K$-selection versus relative fitness''), selection on $c$ does not introduce density-dependence in Eq.~\eqref{eq:canonical}.
-
-Turning to selection on $b$ and $d$, recall that $m_i=b_i n_i U/T$, and so $L=M/U=\overline{b}N/T$ where $\overline{b}$ is the population mean $b$. Thus, from Eq.~\eqref{eq:masterequalc} we have
+Let us now turn to selection on $b$ and $d$ in our lottery model.
+Recall that $m_i=b_i n_i U/T$, and so $L=M/U=\overline{b}N/T$ where $\overline{b}$ is the population mean $b$.
+Thus, from Eq.~\eqref{eq:masterequalc} we have
 \begin{equation}
@@
 \end{equation}
-The mortality $d$ is akin to the birth rate in Eq.~\eqref{eq:simplebirthdeath}, and so, while $d$ does affect density, selection on $d$ is density independent, and so $d$ sweeps also follow the canonical relative fitness model exactly.
+It can be seen that the mortality $d$ is akin to the birth rate in Eq.~\eqref{eq:simplebirthdeath}, and so, while $d$ does affect density, selection on $d$ is density independent.
+Thus, $d$ sweeps follow the canonical relative fitness model exactly.
 
-At first glance $b$ in Eq.~\eqref{eq:bdensitydependence} appears to be analogous to the $\delta$ in Eq.~\eqref{eq:simplebirthdeath} because it regulates density and is multiplied by the density-dependent term $f(\overline{b},N)=\frac{1-e^{-\overline{b}N/T}}{N}(T-N)$.
+At first glance, $b$ in Eq.~\eqref{eq:bdensitydependence} appears to be analogous to the $\delta$ in Eq.~\eqref{eq:simplebirthdeath} because it regulates density and is multiplied by the density-dependent term $f(\overline{b},N)=\frac{1-e^{-\overline{b}N/T}}{N}(T-N)$.
 This term declines from $\overline{b}$ at low density to zero at high density and as a result, selection on $b$ is density-dependent.
-In words: the selective advantage from having greater $b$ depends on the number of territories being contested; if almost all are occupied, then there is little advantage to having a greater $b$.
+The source of this density-depedence is that the selective advantage from having greater $b$ depends on the number of territories being contested; if almost all are occupied, then there is little advantage to having a greater $b$.
 
 Nevertheless, the behavior of equilibrium-to-equilibrium $b$-sweeps are qualitatively different from the $\delta$ sweeps above.
-The reason is that $b$ does not regulate density via the $b_i$ term in front of $f(\overline{b},N)$, which is divided by $\overline{b}$ and thus actually reflects relative lottery contests.
-The density regulating effect of $b$ actually appears in the fraction of contested territories $1-e^{-\overline{b}N/T}$.
-This reflects the fact that $b$ sweeps can only increase density by causing more unoccupied territories to receive propagules.
-The net effect on $f(\overline{b},N)$ is precisely zero.
-To see this, note that in a single-type equilibrium we have $b_i/\overline{b}=1$ and so, setting Eq.~\eqref{eq:bdensitydependence} to zero, we $f(\overline{b},N)=d_i$ exactly at the beginning and end of a pure $b$ sweep, even though the density $N$ increases.
+The reason is that $b$ regulates density by controlling how many unoccupied territories receive propagules.
+Thus, greater $b$ means more propagules contesting territories, but also more territories being contested.
+The net effect on $f(\overline{b},N)$ is precisely zero in equilibrium: in a single-type equilibrium we have $b_i/\overline{b}=1$ and so $f(\overline{b},N)=d_i$ exactly at the beginning and end of a pure $b$ sweep, even though the density $N$ increases.
 Strictly speaking there is some deviation in $f(N)$ from $d_i$ during the sweep, but this deviation is an order of magnitude smaller than for a $\delta$ sweep (the deviation due to a sweep with proportional effect $b_i\rightarrow b_i(1+\epsilon)$ is only of order $\epsilon^2$, whereas the analogous effect in Fig.~\ref{fig:strengthofselection} is of order $\epsilon$; see Appendix D for details).
-Since selection must already be quite strong for a $\delta$ sweep to threaten Eq.~\eqref{eq:canonical}, we can safely conclude that $b$ sweeps obey Eq.~\eqref{eq:canonical} to a very close approximation.
-
-To summarize: (i) $c$-selection is density-dependent, but $c$ does not affect density (ii) $d$-selection is density-independent but $d$ affects density (iii) $b$ affects density and is subject to density-dependent selection.
-And yet pure $b$, $c$ and $d$ sweeps starting and ending at equilibrium all obey the canonical selection equation exactly (almost exactly for $b$).
+Since selection must already be quite strong for a $\delta$ sweep to threaten Eq.~\eqref{eq:canonical}, we conclude that $b$ sweeps also obey the canonical selection equation (to a close approximation).
 
@@
 
-It is widely recognized that the relative fitness description of selection is poorly integrated with the ecological literature on population growth \citep{mallet_2012}.
-This is not simply a lack of ecological realism on the part of relative fitness models.
+Summarizing the behaviour of the three traits in the variable-density lottery: (i) $c$-selection is density-dependent, but $c$ does not regulate density; (ii) $d$ regulates density, but $d$-selection is density-independent; (iii) $b$ regulates density and $b$-selection is density-dependent.
+Despite their differences, pure $b$, $c$ and $d$ sweeps starting and ending at equilibrium all obey the canonical selection equation.
+This rich variety of behaviors in relation to density is quite different from that found in the classical density-dependent selection literature \citep{roughgarden_1979,christiansen_2004}.
+
+To briefly review: based on a diploid, bi-allelic variant of the logistic model, the $r$/$K$ scheme posited a dichotomy between $r$-selection (uncrowded) and $K$-selection (crowded) \citep{macarthur_1962}, with the latter taken to mean selection for greater saturation density \citep{gill_1974}.
+A more general Lotka-Volterra model introduces the inter-type $\alpha_{ij}$ competition coefficients, with selection on these termed ``$\alpha$-selection'' \citep{gill_1974,joshi_2001}.
+Thus, setting aside $r$ which confers no selective advantage at equilibrium, we are left with $K$ and $\alpha$, which both behave like $\delta$ in that they are both density-dependent and cause density to change over a sweep (although $N$ only dips temporarily during an $\alpha$-sweep).
+Thus, sufficiently strong selection is all that is needed for relative fitness to fall apart in the classical density-dependent selection literature.
+
+Our variable-density lottery model shows that this contrast is not simply a lack of ecological realism on the part of relative fitness models.
 Rather, in many population growth models, only one life-history stage is represented, and the competitive effects resulting from crowding appear as a reduction in absolute fitness that only depends on the type densities at this life-history stage (e.g.
 the $n_i^2$ and $n_in_j$ terms in the Lotka Volterra equation).
-However, many species have a ``reproductive excess'' of juveniles that are considerably more fragile than their adult counterparts \citep{chesson_1983,turner1968population,kimura1969natural,nei1971fertility}.
-As a result, competition and the selection it induces can be concentrated at the juvenile phase, where a large reproductive excess allows for most juveniles to be competitively eliminated.
-In our lottery model, this excess appears when the number of propagules is greater than the number of available territories.
+As noted in the introduction, this precludes the important possibility that selection is concentrated at a fragile juvenile stage as a result of a  reproductive excess \citep{chesson_1983,turner1968population,kimura1969natural,nei1971fertility}.
+
+In our variable-density lottery, reproductive excess appears when the number of propagules is greater than the number of available territories.
 Then only $\approx 1/L$ of the juveniles contesting available territories survive to adulthood.
-Note that unlike the role of adult density $n_i$ in single-life-stage models, it is the propagule densities $l_i$ that represent the crowding that drives competition.
+Unlike the role of adult density $n_i$ in single-life-stage models, it is the propagule densities $l_i$ that represent the crowding that drives competition.
 In general, reproductive excesses will tend to produce strictly-relative lottery-type contests in which fitter types can grow at the expense of others by preferentially filling the available adult ``slots''.
@@
 By ignoring reproductive excesses, single life-stage models are biased to have total population density be sensitive to ongoing selection, artificially invalidating Eq.~\eqref{eq:canonical}.
-In this respect, the Wright-Fisher model and similar ``viability selection'' models actually capture an important ecological process.
-
-In particular, we emphasize that MacArthur's argument does not justify the widespread intuition that selection in crowded environments is necessarily connected to achieving greater densities.
-This intuition is largely an artifact of the models historically used in the density-dependent selection literature, which ignore the existence of relative contests.
+In this respect, the Wright-Fisher model and similar viability selection heuristics actually capture an important ecological process.
 
@@
 We first discuss the problem shown in Fig.~\ref{fig:strengthofselection}, which occurs when strong selection changes (adult) population density and is also density-dependent.
-In our lottery model, this can only occur if and only if types differ in $c$ as well as $b$ or $d$.
-Why does our model need joint trait variation when the classical density-dependent selection literature is built on traits that behave like the $\delta$ in Eq.~\eqref{eq:canonical} \citep{christiansen_2004}?
-To briefly review: Based on a diploid, bi-allelic variant of the logistic model, the $r$/$K$ scheme posited a dichotomy between $r$-selection (uncrowded) and $K$-selection (crowded), with the latter taken to mean selection for greater saturation density \citep{gill_1974}.
-A more general Lotka-Volterra model introduces the inter-type $\alpha_{ij}$ competition coefficients, with selection on these termed ``$\alpha$-selection'' \citep{gill_1974,joshi_2001}.
-Thus we are left with $r$-selection (uncrowded) versus $K$- or $\alpha$-selection (crowded).
-The latter two forms of selection in crowded populations behave like selection on $\delta$ in that they are both density-dependent and cause density to change over a sweep (although $N$ only dips temporarily during an $\alpha$-sweep).
-
-By contrast, the density-dependent ($c$) and density-determining traits ($b$ and $d$) in our lottery model are independent.
-This trait separation highlights the two distinct directions in which density and selection interact: selection may depend on density, and density may depend on selection \citep{prout_1980}.
-The combination is needed to pose a threat to Eq.~\eqref{eq:canonical} (recall that we are still assuming $N$ is not far from equilibrium).
+In the variable-density lottery, this occurs if and only if types differ in more than one trait.
+The $c$ and $d$ traits highlight the two distinct directions in which density and selection interact: selection may depend on density, and density may depend on selection \citep{prout_1980}.
+The combination is needed to pose a threat to Eq.~\eqref{eq:canonical}, and even then, the $b$ trait remarkably demonstrates that both may not be sufficient (recall that we are still assuming $N$ is not far from equilibrium).
 In this respect the simple linear models that have become standard in discussions of density-dependent selection \citep{roughgarden_1979,christiansen_2004,mallet_2012} actually represent a complicated form of the density-selection interaction that confounds a key underlying issue.
 
-Note that while this is a conceptual reason to be wary of the classical density-dependent selection models, it is not clear how we should expect the trait variation in nature to align.
+While this is a conceptual reason to be wary of the classical density-dependent selection models, it is not clear how we should expect the trait variation in nature to align.
 For instance, should we expect mutations to generally affect $b$, $c$ and $d$ independently of each other, or pleiotropically such that $\delta$-like selection is prevalent?

### 2017-12-16 manuscript.tex
l

@@
 
-Evolutionary biologists frequently describe selection using relative fitness, and relative fitness models are foundational in population genetics.
+Relative fitness  models are the basis of  foundational in population genetics.
+
 Yet the population ecological basis of relative fitness is poorly understood, and the classical ecology literature on selection in crowded populations seems to be incompatible with widely-used relative fitness models such as the Wright-Fisher.
@@
 
+Another issue with the constant-$N$ relative fitness description of selection is that it precludes consideration of longer-term aspects of the interplay between evolution and ecology such as population extinction.
+A variety of approaches have been developed for dealing with these issues in quantitative genetics \citep{burger1995evolution,engen_2013}, population genetics \citep{bertram2017predicting} and adaptive dynamics \citep{ferriere2013eco,dieckmann2004adaptive}.
+Although density-dependent selection is  pertinent to these longer-term issues \citep{travis_2013}, our focus here is the description of the time-dependent process by which selection changes allele frequencies.
+This is particularly critical for making sense of evolution at the genetic level, for which we now have abundant data.
+
 Here we analyze the population ecology of relative fitness using a novel model of density-dependent population growth based on territorial contests.
@@
 
-Equipped with our new model, we turn to the evaluation of Eq.~\eqref{eq:canonical}.
+Equipped with this new model, we turn to the evaluation of Eq.~\eqref{eq:canonical}.
 We first discuss selection on the ability to contest territories, which behaves like a pure constant-$N$, relative fitness trait, and discuss how this fits with MacArthur's analysis of selection in crowded populations (section ``$K$-selection and selection-dependent density'').
-We then discuss selection on density-regulating traits, contrasting the results from the classical density-dependent selection literature with our model (section ``Density-regulating traits and the threat of strong selection'').
-We conclude [...]
+We then consider selection on density-regulating traits (section ``Density-regulating traits and the threat of strong selection''), and conclude by contrasting the classical density-dependent selection literature with our results (``Discussion'').
 
@@
 
-Summarizing the behaviour of the three traits in the variable-density lottery: (i) $c$-selection is density-dependent, but $c$ does not regulate density; (ii) $d$ regulates density, but $d$-selection is density-independent; (iii) $b$ regulates density and $b$-selection is density-dependent.
-Despite their differences, pure $b$, $c$ and $d$ sweeps starting and ending at equilibrium all obey the canonical selection equation.
+Summarizing the three traits in the variable-density lottery model: (i) $c$-selection is density-dependent, but $c$ does not regulate density; (ii) $d$ regulates density, but $d$-selection is density-independent; (iii) $b$ regulates density and $b$-selection is density-dependent.
+Despite these differences, pure $b$, $c$ and $d$ sweeps starting and ending at equilibrium all obey the canonical selection equation.
 This rich variety of behaviors in relation to density is quite different from that found in the classical density-dependent selection literature \citep{roughgarden_1979,christiansen_2004}.
 
-To briefly review: based on a diploid, bi-allelic variant of the logistic model, the $r$/$K$ scheme posited a dichotomy between $r$-selection (uncrowded) and $K$-selection (crowded) \citep{macarthur_1962}, with the latter taken to mean selection for greater saturation density \citep{gill_1974}.
+To briefly review: based on a diploid, bi-allelic variant of the logistic model, the $r$/$K$ scheme proposed a dichotomy between $r$-selection (uncrowded) and $K$-selection (crowded) \citep{macarthur_1962}, with the latter taken to mean selection for greater saturation density \citep{gill_1974}.
 A more general Lotka-Volterra model introduces the inter-type $\alpha_{ij}$ competition coefficients, with selection on these termed ``$\alpha$-selection'' \citep{gill_1974,joshi_2001}.
-Thus, setting aside $r$ which confers no selective advantage at equilibrium, we are left with $K$ and $\alpha$, which both behave like $\delta$ in that they are both density-dependent and cause density to change over a sweep (although $N$ only dips temporarily during an $\alpha$-sweep).
-Thus, sufficiently strong selection is all that is needed for relative fitness to fall apart in the classical density-dependent selection literature.
+Setting aside $r$ which confers no selective advantage at equilibrium, we are left with $K$ and $\alpha$, which both behave like $\delta$ in that they are density-dependent and cause density to change over a sweep (although $N$ only dips temporarily during an $\alpha$-sweep).
+Thus, strong selection is sufficient for relative fitness to break down in the classical view of density-dependent selection.
+Indeed, in the defense of Eq.~\eqref{eq:canonical} given by \cite{kimura1969natural}, it was necessary to assume that $s$ will be a few percent at most.
 
-Our variable-density lottery model shows that this contrast is not simply a lack of ecological realism on the part of relative fitness models.
+Our variable-density lottery model shows that it is not simply a lack of ecological realism that underlies the incompatibility of relative fitness with the classical view of density-dependent selection.
 Rather, in many population growth models, only one life-history stage is represented, and the competitive effects resulting from crowding appear as a reduction in absolute fitness that only depends on the type densities at this life-history stage (e.g.
 the $n_i^2$ and $n_in_j$ terms in the Lotka Volterra equation).
-As noted in the introduction, this precludes the important possibility that selection is concentrated at a fragile juvenile stage as a result of a  reproductive excess \citep{chesson_1983,turner1968population,kimura1969natural,nei1971fertility}.
+As noted in the introduction, this precludes selection concentrated at a fragile juvenile stage as a result of a reproductive excess \citep{chesson_1983,turner1968population,kimura1969natural,nei1971fertility}.
 
-In our variable-density lottery, reproductive excess appears when the number of propagules is greater than the number of available territories.
+Reproductive excesses appear in the variable-density lottery model when the number of propagules is greater than the number of available territories.
 Then only $\approx 1/L$ of the juveniles contesting available territories survive to adulthood.
 Unlike the role of adult density $n_i$ in single-life-stage models, it is the propagule densities $l_i$ that represent the crowding that drives competition.
-In general, reproductive excesses will tend to produce strictly-relative lottery-type contests in which fitter types can grow at the expense of others by preferentially filling the available adult ``slots''.
+In general, reproductive excesses will tend to produce strictly-relative lottery-type contests in which fitter types grow at the expense of others by preferentially filling the available adult ``slots''.
 The number of slots can remain fixed or change independently of selection at the juvenile stage.
-By ignoring reproductive excesses, single life-stage models are biased to have total population density be sensitive to ongoing selection, artificially invalidating Eq.~\eqref{eq:canonical}.
+By ignoring reproductive excesses, single life-stage models are biased to have total population density be sensitive to ongoing selection.
 In this respect, the Wright-Fisher model and similar viability selection heuristics actually capture an important ecological process.
@@
 We now turn to the breakdown of Eq.~\eqref{eq:canonical}.
-We first discuss the problem shown in Fig.~\ref{fig:strengthofselection}, which occurs when strong selection changes (adult) population density and is also density-dependent.
+We first discuss the problem shown in Fig.~\ref{fig:strengthofselection}, which occurs when strong selection changes population density and is also density-dependent.
 In the variable-density lottery, this occurs if and only if types differ in more than one trait.
-The $c$ and $d$ traits highlight the two distinct directions in which density and selection interact: selection may depend on density, and density may depend on selection \citep{prout_1980}.
-The combination is needed to pose a threat to Eq.~\eqref{eq:canonical}, and even then, the $b$ trait remarkably demonstrates that both may not be sufficient (recall that we are still assuming $N$ is not far from equilibrium).
-In this respect the simple linear models that have become standard in discussions of density-dependent selection \citep{roughgarden_1979,christiansen_2004,mallet_2012} actually represent a complicated form of the density-selection interaction that confounds a key underlying issue.
+The $c$ and $d$ traits represent the two distinct directions in which density and selection interact: selection may depend on density, and density may depend on selection \citep{prout_1980}.
+The combination is necessary to pose a threat to Eq.~\eqref{eq:canonical}.
+However, the $b$ trait remarkably demonstrates that the combination is not sufficient, since the density-dependence of $b$-selection disappears over equilibrium-to-equilibrium $b$-sweeps.
+Thus, the simple linear models that have become standard in discussions of density-dependent selection \citep{roughgarden_1979,christiansen_2004,mallet_2012,travis_2013} actually represent a complicated form of the interaction between density and selection, and their parameters confound the underlying issues.
 
@@
 
-Relative fitness models truly break down when $N$ is far from equilibrium and selection is density-dependent.
-For example, wild \textit{Drosophila} experience huge seasonal boom-bust cycles in population density coupled to strong selection that drives large swings in allele frequency \citep{bergland_14}.
+Relative fitness models truly break down when $N$ is far from equilibrium and selection is density-dependent (as seems likely \citep{travis_2013}).
+For example, wild \textit{Drosophila} experience large seasonal boom-bust cycles in population density coupled to strong selection that drives large swings in allele frequency \citep{bergland_14}.
 In this case there is no choice but to abandon relative fitness, and our model provides one potentially suitable option.
 Whether or not our density-dependent lottery model is a good description of \textit{Drosophila} ecology, the close connection between our model and Wright-Fisher is useful, because drift in our model should behave broadly similarly.
-Thus, our model it should provide a useful starting point for analyzing evolution in this and other from-from-equilibrium situations.
-
-Another issue with the constant-$N$ relative fitness description of selection is that it precludes consideration of longer-term aspects of the interplay between evolution and ecology such as population extinction.
-A variety of approaches have been developed for dealing with these issues in quantitative genetics \citep{burger1995evolution,engen_2013}, population genetics \citep{bertram2017predicting} and adaptive dynamics \citep{ferriere2013eco,dieckmann2004adaptive}.
-Although this manuscript, and density-dependent selection more broadly, are pertinent to the interplay between ecology and evolution \citep{travis_2013}, our focus here is the description of the time-dependent process by which selection changes allele frequencies.
-This is particularly critical for making sense of evolution at the genetic level, for which we now have abundant data.
+Thus, our model it should provide a useful starting point for analyzing evolution in this and other far-from-equilibrium situations.
 
@@
 
-In section ``Density-dependence and the strength of selection'' we argued that the density-dependent factor $f(\overline{b},N)$ is unchanged at the beginning and end points of an equilibrium-to-equilibrium $b$.
+In section ``Density-regulating traits and the threat of strong selection'' we argued that the density-dependent factor $f(\overline{b},N)$ is unchanged at the beginning and end points of an equilibrium-to-equilibrium $b$.
 Here we estimate the magnitude of the deviation in $f(\overline{b},N)$ during the sweep.

### 2017-12-19 manuscript.tex
l

@@
 
-In uncrowded populations, relative fitness simply represents differences in intrinsic population growth rate.
+In age-structured, sexual populations, selection cannot in general be represented in terms of a single type-specific fitness parameter \citep[Chap.
+4]{charlesworth_1994}.
+Relative fitness models can then be justified as a linear approximation in which selection is approximately constant \cite[pp.
+277]{ewens_2004} \citep[Chap.
+4]{charlesworth_1994}.
+Provided that selection is sufficiently weak and stable over time, this linear approximation is not a major restriction on the use of relative fitness.
+Yet it is increasingly recognized that selection is not always weak, that it can fluctuate strongly over time, and that $N$ can vary by orders of magnitude over a few generations as a routine feature of a population's ecology \citep{messer_2016}.
+These are not rare exceptions, but occur widely in nature and the lab, including in wild \textit{Drosophila} \citep{bergland_14}.
+
+Nevertheless, relative fitness models are the foundation for much of the population genetic literature, and are still widely used without considering the ``linear approximation'' restriction or the poor integration with population ecology \citep{mallet_2012}.
+Thus, it is important that we understand the population ecological basis of relative fitness models, both to gain insight into their domain of applicability, and as part of the broader challenge of synthesizing ecology and evolution.
+
+Our aim here is to explore in more depth where the population ecological limits of relative fitness lie.
+To do so, we set aside the complexities of age structure and mating, since we know that these can cause relative fitness to break down.
+In this simplest case of asexual haploids with no age structure, the primary threat to the relative fitness description of selection is the dependence of selection on population density.
+
+To see this, first consider an uncrowded population.
+Relative fitness then simply represents differences in intrinsic population growth rate.
 In discrete time, the change in frequency of type $i$ is $\Delta p_i=\left(\frac{W_i}{\overline{W}}-1\right) p_i$, where $W_i$ is the intrinsic absolute growth factor of type $i$, and $\overline{W}=\sum_i W_i p_i$ is the population mean $W$.
@@
 The difficulty with Eq.~\eqref{eq:canonical} arises in crowded populations.
-Since crowded and uncrowded conditions are so different, it seems likely that $s$ will often depend on density (although empirical verification has proved challenging \citep{travis_2013}).
-Eq.~\eqref{eq:canonical} is then no longer a complete description of selection --- we would also need to specify a model for density.
+Since crowded and uncrowded conditions are so different, we expect that $s$ will often depend on density (although empirical verification has proved challenging \citep{travis_2013}).
+Eq.~\eqref{eq:canonical} is then no longer a complete description of selection --- we would also need to specify a model for how density is changing.
 Note that frequency-dependent selection does not raise similar problems; Eq.~\eqref{eq:canonical} is still a complete description of selection even if its behavior is more complicated due to $s$ depending on frequency.
@@
 
-Counter to the assumption of constant $N$, MacArthur famously argued that when population growth is density-regulated, selection in crowded populations is intimately connected to the ability to keep growing at higher densities than other types can tolerate \citep{macarthur_1967}.
+However, MacArthur famously argued that when population growth is density-regulated, selection in crowded populations is intimately connected to the ability to keep growing at higher densities than other types can tolerate \citep{macarthur_1967}.
 The classic example is the logistic model, where the type with the greatest saturation population density ``$K$'' excludes the others (Fig.~\ref{fig:Ksel}a).
@@
 In these examples, both $N$ and $s$ change during, and as a result of, adaptive sweeps.
-It would therefore seem that the ubiquitous constant-$N$, relative fitness description of selection is incompatible with a huge class of population ecological processes driving selection (Fig.~\ref{fig:Ksel}b).
+It would therefore seem that the ubiquitous constant-$N$, relative fitness description of selection is incompatible with a huge class of population ecological processes driving selection (Fig.~\ref{fig:Ksel}b), even in the absence of age-structure and mating.
 
@@
 
-In light of these difficulties, the relative fitness description has been justified in broadly three different ways for crowded populations (we do not discuss Wagner's [\citeyear{wagner_2010}] measure-theoretical justification, which is independent of population biology).
+In light of these difficulties, the relative fitness description has been justified in broadly two different ways for crowded populations (we do not discuss Wagner's [\citeyear{wagner_2010}] measure-theoretical justification, which is independent of population biology).
 The first is to simply assume that selection is density-independent but relax the assumption of constant $N$ by allowing density to change as a result of selective sweeps \citep[pp.
@@
 As a result, the role of reproductive excesses in justifying Eq.~\eqref{eq:canonical} is still largely verbal.
-
-Third, Eq.~\eqref{eq:canonical} can justified as a short-term linear approximation \cite[pp.
-277]{ewens_2004}.
-That is, within a sufficiently short time frame, $N$ and $s$ can be treated as constant.
-Provided that selection is sufficiently weak and stable over time, this ``short-term''  assumption is not a major restriction.
-Yet it is increasingly recognized that selection is not always weak, that it can fluctuate strongly over time, and that $N$ can vary by orders of magnitude over a few generations as a routine feature of a population's ecology \citep{messer_2016}.
-These are not rare exceptions, but occur widely in nature and the lab, including in wild \textit{Drosophila} \citep{bergland_14}.
-
-Nevertheless, relative fitness models are the foundation for much of the population genetic literature, and are still widely used without considering the ``short-term'' restriction or the lack of integration with population ecology \citep{mallet_2012}.
-Thus, it is important to understand the population ecological basis of relative fitness models, both to gain insight into their domain of applicability, and as part of the broader challenge of synthesizing ecology and evolution.
 
@@
 Then only $\approx 1/L$ of the juveniles contesting available territories survive to adulthood.
-Unlike the role of adult density $n_i$ in single-life-stage models, it is the propagule densities $l_i$ that represent the crowding that drives competition.
+Unlike the role of adult density $n_i$ in single-life-stage models, it is the propagule densities $l_i$ that represent the crowding that drives competition (a ``critical age-group'' \cite[pp.
+54]{charlesworth_1994}).
 In general, reproductive excesses will tend to produce strictly-relative lottery-type contests in which fitter types grow at the expense of others by preferentially filling the available adult ``slots''.

### 2017-12-19 manuscript.tex
l

@@
 
-Relative fitness  models are the basis of  foundational in population genetics.
-
+It it common in evolutionary biology to describe selection in terms of constant relative fitness values.
 Yet the population ecological basis of relative fitness is poorly understood, and the classical ecology literature on selection in crowded populations seems to be incompatible with widely-used relative fitness models such as the Wright-Fisher.
@@
 We then explore the population ecology of relative fitness using this model as a reference point.
-Although density-dependent selection can cause relative fitness to break down, relative fitness models are fairly robust in populations close to demographic equilibrium.
-In particular, relative fitness only breaks down if selection is strong and density-dependent, and density is also selection-dependent.
-Our generalized Wright-Fisher model clearly distinguishes between three demographic parameters, each of whose behavior is, on its own, correctly described by relative fitness models.
-In contrast, the classical literature on density-dependent selection confounds them.
+Our model makes a clear distinction between density-dependent selection and selection-dependent density, both of which are necessary for relative fitness to break down.
+These two effects are confounded in the classical literature on density-dependent selection, where density-regulating traits are almost always subject to density-dependent selection.
+We show that the presence of both effects is necessary but not sufficient: even strong selection on a density-regulating trait that is subject to density-dependent selection can conform to the classical relative fitness description almost exactly.
 We argue that selection-independent density is ecologically plausible for many species given the prevalence of reproductive excesses, an important aspect of selection omitted in many ecological models.
-Our model also offers a natural alternative to relative fitness when the latter is untenable, as is likely the case far from demographic equilibrium.
+Our model also offers a possible alternative to relative fitness when the latter is untenable, as is likely the case far from demographic equilibrium.
 
@@
 While such a connection is fairly clear for absolute fitness measures like $r$, relative fitness seems largely divorced from population biology.
-It has even been proposed that relative fitness be justified from more abstract measure-theoretical arguments, abandoning population biology altogether \citep{wagner_2010}.
+It has even been proposed that relative fitness be justified from measure theory, abandoning population biology altogether \citep{wagner_2010}.
 
-In age-structured, sexual populations, selection cannot in general be represented in terms of a single type-specific fitness parameter \citep[Chap.
+In age-structured, sexual populations, selection cannot in general be represented in terms of type-specific relative-fitness constants \citep[Chap.
 4]{charlesworth_1994}.
-Relative fitness models can then be justified as a linear approximation in which selection is approximately constant \cite[pp.
+Constant relative fitness values are then be justified as a linear approximation \cite[pp.
 277]{ewens_2004} \citep[Chap.
 4]{charlesworth_1994}.
-Provided that selection is sufficiently weak and stable over time, this linear approximation is not a major restriction on the use of relative fitness.
-Yet it is increasingly recognized that selection is not always weak, that it can fluctuate strongly over time, and that $N$ can vary by orders of magnitude over a few generations as a routine feature of a population's ecology \citep{messer_2016}.
-These are not rare exceptions, but occur widely in nature and the lab, including in wild \textit{Drosophila} \citep{bergland_14}.
+Provided that selection is sufficiently weak and stable over time, this linear approximation is close to exact.
+Yet strong, temporally-variable selection occurs widely in nature and the lab, including in wild \textit{Drosophila}, where population density varies by orders of magnitude each seasonal cycle \citep{messer_2016,bergland_14}.
+Nevertheless, relative fitness models are the foundation for much of the evolutionary genetics literature, and are still widely used without considering the ``linear approximation'' restriction or the broader lack of integration with population ecology \citep{mallet_2012}.
 
-Nevertheless, relative fitness models are the foundation for much of the population genetic literature, and are still widely used without considering the ``linear approximation'' restriction or the poor integration with population ecology \citep{mallet_2012}.
-Thus, it is important that we understand the population ecological basis of relative fitness models, both to gain insight into their domain of applicability, and as part of the broader challenge of synthesizing ecology and evolution.
+The aim of this maniscript is to explore the population ecological limits of relative fitness models in the context of density-dependent selection.
+We set aside age structure and mating because, as noted above, relative fitness models break down in the presence of these complications.
+In asexual haploids with no age structure, even strong selection need not invalidate constant relative fitness.
+By studying this simpler case, we get a clearer picture of what is needed for relative fitness to break down.
 
-Our aim here is to explore in more depth where the population ecological limits of relative fitness lie.
-To do so, we set aside the complexities of age structure and mating, since we know that these can cause relative fitness to break down.
-In this simplest case of asexual haploids with no age structure, the primary threat to the relative fitness description of selection is the dependence of selection on population density.
-
-To see this, first consider an uncrowded population.
-Relative fitness then simply represents differences in intrinsic population growth rate.
+With these simplifications, the basis of relative fitness is straightforward in uncrowded populations: it simply represents differences in intrinsic population growth rate.
 In discrete time, the change in frequency of type $i$ is $\Delta p_i=\left(\frac{W_i}{\overline{W}}-1\right) p_i$, where $W_i$ is the intrinsic absolute growth factor of type $i$, and $\overline{W}=\sum_i W_i p_i$ is the population mean $W$.
@@
 The difficulty with Eq.~\eqref{eq:canonical} arises in crowded populations.
-Since crowded and uncrowded conditions are so different, we expect that $s$ will often depend on density (although empirical verification has proved challenging \citep{travis_2013}).
+Since crowded and uncrowded conditions are so different, we expect that $s$ will often depend on density \citep{travis_2013}.
 Eq.~\eqref{eq:canonical} is then no longer a complete description of selection --- we would also need to specify a model for how density is changing.
 Note that frequency-dependent selection does not raise similar problems; Eq.~\eqref{eq:canonical} is still a complete description of selection even if its behavior is more complicated due to $s$ depending on frequency.
-Standard population genetics evades the issue of density-dependent selection by simply assuming that total population density $N$ has reached its equilibrium value, which is assumed to be a fixed constant.
-The selection coefficient $s$ now parameterizes the rate at which selection changes relative frequencies, but no longer corresponds to differences in intrinsic growth rates $r$.
+Population genetics traditionally evades the issue of density-dependent selection by simply assuming that total population density $N$ has reached its equilibrium value, which is assumed to be a fixed constant.
+The selection coefficient $s$ now abstractly parameterizes the rate at which selection changes relative frequencies, and no longer corresponds to differences in intrinsic growth rates $r$.
 
@@
 
-In light of these difficulties, the relative fitness description has been justified in broadly two different ways for crowded populations (we do not discuss Wagner's [\citeyear{wagner_2010}] measure-theoretical justification, which is independent of population biology).
+In light of this difficulty, the relative fitness description has been justified in broadly two different ways for crowded populations (we do not discuss Wagner's [\citeyear{wagner_2010}] measure-theoretical justification, which is independent of population biology).
 The first is to simply assume that selection is density-independent but relax the assumption of constant $N$ by allowing density to change as a result of selective sweeps \citep[pp.
@@
 Obviously this does not address the problem that $s$ can, in reality, depend on density.
-Type-specific responses to density are at the center of MacArthur's argument and the density-dependent selection literature that grew out of it \citep{roughgarden_1979}.
+Type-specific responses to density are at the center of MacArthur's argument and the density-dependent selection literature that grew out of it (e.g.
+\citep{roughgarden_1979}).
 
-The second justification, which grew out of a controversy over Haldane's ``cost of selection'', is to appeal to the existence of a ``reproductive excess'' of juveniles that are more fragile than their adult counterparts \citep{turner1968population,kimura1969natural,nei1971fertility}.
+The second justification, which primarily grew out of a controversy over Haldane's ``cost of selection'', is to appeal to the existence of a ``reproductive excess'' of juveniles that are more fragile than their adult counterparts \citep{turner1968population,kimura1969natural,nei1971fertility}.
 Selection can then be concentrated at the juvenile phase, uncoupling selection from population density at the adult phase unless it is so strong that the reproductive excess is depleted.
-This justifies the use of Eq.~\eqref{eq:canonical} because, for a population in demographic equilibrium, selective sweeps do not affect density, and so it does not matter if selection is density-dependent.
-However, this reproductive excesses literature was focused on the evaluation of genetic load, and was poorly integrated with population ecology.
-\cite{kimura1969natural} took constant $N$ as a requirement and then introduced some mathematical variants of the logistic model that satisfy this requirement.
+This justifies Eq.~\eqref{eq:canonical} because, for a population in demographic equilibrium, selective sweeps do not affect density, and so the density-dependence of selection does not matter.
+Unfortuantely this reproductive excess literature is also poorly integrated with population ecology.
+\cite{kimura1969natural} took constant $N$ as a requirement and then derived some variants of the logistic model that satisfy this requirement.
 \cite{nei1971fertility} proposed a model with an explicit representation of reproductive excess, but used an unusual model of competition based on pair-wise interactions which was only defined for at most two different types.
@@
 
-Another issue with the constant-$N$ relative fitness description of selection is that it precludes consideration of longer-term aspects of the interplay between evolution and ecology such as population extinction.
-A variety of approaches have been developed for dealing with these issues in quantitative genetics \citep{burger1995evolution,engen_2013}, population genetics \citep{bertram2017predicting} and adaptive dynamics \citep{ferriere2013eco,dieckmann2004adaptive}.
-Although density-dependent selection is  pertinent to these longer-term issues \citep{travis_2013}, our focus here is the description of the time-dependent process by which selection changes allele frequencies.
-This is particularly critical for making sense of evolution at the genetic level, for which we now have abundant data.
-
-Here we analyze the population ecology of relative fitness using a novel model of density-dependent population growth based on territorial contests.
+Here we study the population ecology of relative fitness using a novel model of density-dependent population growth based on territorial contests.
 Rather than attempting to make sense of relative fitness in existing standard models of population growth (e.g.
@@
 
+Another issue with the constant-$N$ relative fitness description of selection is that it precludes consideration of longer-term aspects of the interplay between evolution and ecology such as population extinction.
+A variety of approaches have been developed for dealing with these issues in quantitative genetics \citep{burger1995evolution,engen_2013}, population genetics \citep{bertram2017predicting} and adaptive dynamics \citep{ferriere2013eco,dieckmann2004adaptive}.
+Although density-dependent selection is  pertinent to these longer-term issues \citep{travis_2013}, our focus here has been the description of the time-dependent process by which selection changes allele frequencies.
+This is particularly critical for making sense of evolution at the genetic level, for which we now have abundant data.
+
 \section*{Acknowledgments}

### 2017-12-19 manuscript.tex
l

@@
-\documentclass[11pt]{article}
-\usepackage[sc]{mathpazo}
-\usepackage{amsmath,pifont}
+\documentclass[12pt]{article}
+\usepackage{amsmath}
 \usepackage{fullpage}
-\usepackage[authoryear,sectionbib,sort]{natbib}
-\linespread{1.7}
-\usepackage[utf8]{inputenc}
+\usepackage[authoryear,round]{natbib}
 \usepackage{lineno}
-
 \usepackage{graphicx} 
-\usepackage{tabularx,setspace}
+\usepackage{setspace}
+\doublespacing
 
@@
 
-%\textit{Manuscript elements}:
+\textit{Keywords}: Lottery model, competitive Lotka-Volterra, $r$/$K$-selection, interference competition, $R^*$ theory, eco-evo.
 
@@
 
-\textit{Keywords}: Lottery model, competitive Lotka-Volterra, $r$/$K$-selection, interference competition, $R^*$ theory, eco-evo.
+\textit{Author contributions}: JB and JM conceptualized the manuscript. JB did the formal analysis. JB wrote the manuscript with review and editing from JM.
 
 \bigskip
+
+\textit{Running title}: Density-dependence and relative fitness
+
+\bigskip
+
+\textit{Acknowledgments}: We thank Peter Chesson and Joachim Hermisson for many constructive comments on an earlier and quite different version of this manuscript.
+This work was financially supported by the National Science Foundation (DEB-1348262) and the John Templeton Foundation (60814).
 
@@
 
-\section*{Abstract}
+\section*{\centering \huge  Density-dependent selection and the limits of relative fitness}
 
-It it common in evolutionary biology to describe selection in terms of constant relative fitness values.
-Yet the population ecological basis of relative fitness is poorly understood, and the classical ecology literature on selection in crowded populations seems to be incompatible with widely-used relative fitness models such as the Wright-Fisher.
-Here we develop a generalization of the Wright-Fisher model in which density depends dynamically on the demographic rates of the types present.
-We then explore the population ecology of relative fitness using this model as a reference point.
-Our model makes a clear distinction between density-dependent selection and selection-dependent density, both of which are necessary for relative fitness to break down.
-These two effects are confounded in the classical literature on density-dependent selection, where density-regulating traits are almost always subject to density-dependent selection.
-We show that the presence of both effects is necessary but not sufficient: even strong selection on a density-regulating trait that is subject to density-dependent selection can conform to the classical relative fitness description almost exactly.
-We argue that selection-independent density is ecologically plausible for many species given the prevalence of reproductive excesses, an important aspect of selection omitted in many ecological models.
+\bigskip
+
+\subsection*{Abstract}
+
+Selection is commonly described in terms of relative fitness.
+Yet much of the ecological theory of selection in density-regulated populations seems to be incompatible with widely-used, constant-density relative fitness models such as the Wright-Fisher.
+Here we analyze the population ecological limits of relative fitness using a novel generalization of the Wright-Fisher model in which population density depends dynamically on the demographic rates of the types present.
+Our model clearly distinguishes between density-dependent selection and selection-dependent density.
+These two effects are confounded in many standard models of density-regulated population growth.
+Both effects are necessary, in combination with strong selection, for relative fitness to break down in populations close to demographic equilibrium.
+Remarkably, both effects are not sufficient: we show that strong selection on a density-regulating trait that is subject to density-dependent selection can conform to the relative fitness description almost exactly.
+We reiterate the prevalence of reproductive excesses, an important aspect of selection omitted in many population growth models, which implies that even strong selection need not affect density in many species.
 Our model also offers a possible alternative to relative fitness when the latter is untenable, as is likely the case far from demographic equilibrium.
+
+\noindent (195 words)
 
@@
 It has even been proposed that relative fitness be justified from measure theory, abandoning population biology altogether \citep{wagner_2010}.
+Given the ubiquitous use of relative fitness, it is important that we understand its population ecological basis, both to clarify its domain of applicability, and as part of the broader challenge of synthesizing ecology and evolution.
 
-In age-structured, sexual populations, selection cannot in general be represented in terms of type-specific relative-fitness constants \citep[Chap.
+Constant relative fitness values can be justified as a linear approximation \cite[pp.
+277]{ewens_2004} \citep[Chap.
+4]{charlesworth_1994} which is close to exact provided that selection is sufficiently weak and stable over time.
+Yet strong, temporally-variable selection occurs widely in nature and the lab, including in wild \textit{Drosophila}, where population density varies by orders of magnitude each seasonal cycle \citep{messer_2016,bergland_14}.
+The question is whether relative fitness can be used when selection is not vanishingly weak.
+In age-structured populations that reproduce by outcrossing, selection cannot in general be represented in terms of type-specific relative-fitness constants unless selection is weak \citep[Chap.
 4]{charlesworth_1994}.
-Constant relative fitness values are then be justified as a linear approximation \cite[pp.
-277]{ewens_2004} \citep[Chap.
-4]{charlesworth_1994}.
-Provided that selection is sufficiently weak and stable over time, this linear approximation is close to exact.
-Yet strong, temporally-variable selection occurs widely in nature and the lab, including in wild \textit{Drosophila}, where population density varies by orders of magnitude each seasonal cycle \citep{messer_2016,bergland_14}.
-Nevertheless, relative fitness models are the foundation for much of the evolutionary genetics literature, and are still widely used without considering the ``linear approximation'' restriction or the broader lack of integration with population ecology \citep{mallet_2012}.
+We will therefore restrict our attention to asexual haploids with little or no age structure, where the relative fitness description may be valid even if selection is strong.
+This simplification also makes it easier to grasp the population ecological underpinnings of selection, as opposed to the role of mating systems or complex changes in age structure.
 
-The aim of this maniscript is to explore the population ecological limits of relative fitness models in the context of density-dependent selection.
-We set aside age structure and mating because, as noted above, relative fitness models break down in the presence of these complications.
-In asexual haploids with no age structure, even strong selection need not invalidate constant relative fitness.
-By studying this simpler case, we get a clearer picture of what is needed for relative fitness to break down.
-
-With these simplifications, the basis of relative fitness is straightforward in uncrowded populations: it simply represents differences in intrinsic population growth rate.
+The basis of relative fitness is straightforward in the absence of crowding: it simply represents differences in intrinsic population growth rate.
 In discrete time, the change in frequency of type $i$ is $\Delta p_i=\left(\frac{W_i}{\overline{W}}-1\right) p_i$, where $W_i$ is the intrinsic absolute growth factor of type $i$, and $\overline{W}=\sum_i W_i p_i$ is the population mean $W$.
@@
 
-Our variable-density lottery model shows that it is not simply a lack of ecological realism that underlies the incompatibility of relative fitness with the classical view of density-dependent selection.
+Our variable-density lottery model shows that it is not simply a lack of ecological realism that underlies the contrast between relative fitness with the classical view of density-dependent selection.
 Rather, in many population growth models, only one life-history stage is represented, and the competitive effects resulting from crowding appear as a reduction in absolute fitness that only depends on the type densities at this life-history stage (e.g.
@@
 Then only $\approx 1/L$ of the juveniles contesting available territories survive to adulthood.
-Unlike the role of adult density $n_i$ in single-life-stage models, it is the propagule densities $l_i$ that represent the crowding that drives competition (a ``critical age-group'' \cite[pp.
+Unlike the role of adult density $n_i$ in single-life-stage models, it is the propagule densities $l_i$ that represent the crowding that drives competition (a ``critical age-group''; \citealt[pp.
 54]{charlesworth_1994}).
@@
 
-Relative fitness models truly break down when $N$ is far from equilibrium and selection is density-dependent (as seems likely \citep{travis_2013}).
+Relative fitness models truly break down when $N$ is far from equilibrium and selection is density-dependent (as seems likely; \citealt{travis_2013}).
 For example, wild \textit{Drosophila} experience large seasonal boom-bust cycles in population density coupled to strong selection that drives large swings in allele frequency \citep{bergland_14}.
@@
 
-\section*{Acknowledgments}
-
-We thank Peter Chesson and Joachim Hermisson for many constructive comments on an earlier and quite different version of this manuscript.
-This work was financially supported by the National Science Foundation (DEB-1348262) and the John Templeton Foundation (60814).
-
-\bibliographystyle{plainnat}
+\bibliographystyle{abbrvnat}
 \bibliography{reference}

### 2017-12-20 manuscript.tex
l

@@
 Remarkably, both effects are not sufficient: we show that strong selection on a density-regulating trait that is subject to density-dependent selection can conform to the relative fitness description almost exactly.
-We reiterate the prevalence of reproductive excesses, an important aspect of selection omitted in many population growth models, which implies that even strong selection need not affect density in many species.
+We reiterate the importance of reproductive excesses, an aspect of selection omitted in many population growth models, which implies that even strong selection need not affect density in many species.
 Our model also offers a possible alternative to relative fitness when the latter is untenable, as is likely the case far from demographic equilibrium.
@@
 We assume that each adult from type $i$ produces a constant number $b_i$ of successfully dispersing propagules; the number of propagules dispersing to unoccupied territories is then given by $m_i=b_in_iU/T$.
+The total number of these propagules will be denoted $M=\sum_i m_i$.
 Note that due to our assumption of uniform dispersal, the parameter $b_i$ can be thought of as a measure of ``colonization ability'', which combines fecundity and dispersal ability \citep{levins_71,tilman_94,bolker_99}.
@@
 The number of propagules of the $i$'th type landing in any particular territory is denoted $x_i$.
-We assume that $x_i$ follows a Poisson distribution $p_i(x_i)=l_i^{x_i} e^{-l_i}/x_i!$, where $l_i=m_i/U$ is the mean territorial propagule density.
+We assume that $x_i$ follows a Poisson distribution $p_i(x_i)=l_i^{x_i} e^{-l_i}/x_i!$, where $l_i=m_i/U$ is the mean territorial propagule density for type $i$ (the total propagule density will be denoted $L=\sum_i l_i$).
 This is strictly only an approximation of random dispersal, but it is an excellent approximation given our assumption that the $n_i$ are large enough that drift can be ignored (Appendix A).
@@
 We expect that a fraction $p_1(x_1)\ldots p_G(x_G)$ of the $U$ unoccupied territories will have the propagule composition $x_1,\ldots,x_G$.
-Type $i$ is expected to win $c_i x_i/\sum_j c_j x_j$ of these.
+Type $i$ is expected to win a proportion $c_i x_i/\sum_j c_j x_j$ of these.
 Ignoring fluctuations about these two expectations (due to our no-drift, large $n_i$, large $T$ approximation), type $i$'s territorial acquisition is given by
 \begin{equation}
-\Delta_+ n_i=U\sum_{x_1,\ldots,x_G} \frac{c_i x_i}{\sum_j c_j x_j} p_1(x_1)\ldots p_G(x_G), \label{eq:growthsumuncoupled}
+\Delta_+ n_i=U\sum_{x_1,\ldots,x_G} \frac{c_i x_i}{\sum_j c_j x_j} p_1(x_1)\cdots p_G(x_G), \label{eq:growthsumuncoupled}
 \end{equation}
@@
 Then the numbers of territories won by each type $\Delta_+ n_1,\Delta_+ n_2,\ldots,\Delta_+ n_G$ follow a multinomial distribution with $U$ trials and success probabilities $\frac{c_1 l_1}{\sum_j c_j l_j},\frac{c_2 l_2}{\sum_j c_j l_j},\ldots,\frac{c_G l_G}{\sum_j c_j l_j}$, respectively.
-Type $i$ is expected to win $c_i l_i/\sum_j c_j l_j$ of the $U$ available territories, and deviations from this expected outcome are small (since $T$ is large by assumption), giving
+Type $i$ is expected to win a proportion $c_i l_i/\sum_j c_j l_j$ of the $U$ available territories, and deviations from this expected outcome are small (since $T$ is large by assumption), giving
 \begin{equation}
@@
 \end{equation}
-where $\overline{c}=\sum_j c_j m_j/M$ is the mean competitive ability for a randomly selected propagule, $L=M/U$ is the total propagule density and $M=\sum_j m_j$ is the total number of propagules.
+where $\overline{c}=\sum_j c_j m_j/M$ is the mean competitive ability for a randomly selected propagule.
 In section ``Analytical approximation of the density-dependent lottery'', we derive a generalization of Eq.~\eqref{eq:lottery} that accommodates arbitrary propagule densities $l_i$.
@@
 Fig.~\ref{fig:simcomp} shows that Eq.
-\eqref{eq:master} and its components closely approximate simulations of the density-dependent lottery model over a wide range of propagule densities.
+\eqref{eq:master} and its components closely approximate simulations of our variable-density lottery model over a wide range of propagule densities.
 Two types are present, one of which is at low frequency.
@@
 However, for sufficiently large $U$ and $m_i$, this multinomial distribution for the $x_i$ across territories is closely approximated by a product of independent Poisson distributions for each territory, each with rate parameter $l_i$ \citep[Theorem 1]{arenbaev_1977}.
-Since we are ignoring finite population size effects, we effectively have $T\rightarrow \infty$, in which case $U$ can be only be small enough to violate the Poisson approximation if there is vanishing population turnover, and then the dispersal distribution is irrelevant anyway.
+Since we are ignoring finite population size effects, we effectively have $T\rightarrow \infty$, in which case $U$ can only be small enough to violate the Poisson approximation if there is vanishing population turnover, and then the dispersal distribution is irrelevant anyway.
 Likewise, in ignoring stochastic finite population size for the $n_i$, we have effectively already assumed that $m_i$ is large enough to justify the Poisson approximation (the error scales as $1/\sqrt{m_i}$; \citealt{arenbaev_1977}).
@@
 
-Our ``mean field'' approximation is to replace $x_j$ with its mean in the last term in Eq.~\eqref{eq:deltr},
+Our approximation is to replace $x_j$ with its mean in the last term in Eq.~\eqref{eq:deltr},
 \begin{equation}
@@
 Below we justify this replacement by arguing that the standard deviation $\sigma_{\tilde{p}}(\sum_{j\neq i} c_j x_j)$ (with respect to $\tilde{p}$), is much smaller than $\langle\sum_{j\neq i} c_j x_j\rangle_{\tilde{p}}$.
+If this standard deviation is small as claimed, then the mean value gives an accurate representation of the values taken by $\sum_{j\neq i} c_j x_j$ across territories.
 
@@
 \begin{align}
-\langle x_j \rangle_{\tilde{p}}&=\frac{l_j}{1-(1+L)e^{-L}}\frac{1}{L-l_i}\sum_{k=2}^{\infty} P(X) (X-1)\nonumber\\
+\langle x_j \rangle_{\tilde{p}}&=\frac{l_j}{1-(1+L)e^{-L}}\frac{1}{L-l_i}\sum_{X=2}^{\infty} P(X) (X-1)\nonumber\\
 &=\frac{l_j}{1-(1+L)e^{-L}}\frac{L-1+e^{-L}}{L-l_i},
@@
 
-The exact analysis of the fluctuations in $\sum_{j\neq i} c_j x_j$ is complicated because the $x_j$ are not independent with respect to $\tilde{p}$.
-These fluctuations are part of the ``drift'' in type abundances which we leave for future work.
+Having evaluated the mean propagule numbers, we now evaluate the variance in propagule numbers to check that the mean value replacement in Eq.~\eqref{eq:meanfieldr} is justified.
+This is complicated because the $x_j$ are not independent with respect to $\tilde{p}$.
 Here we use the following approximation to give some insight into the magnitude of these fluctuations and also the nature of the correlations between the $x_j$.
@@
 The distinction between $\tilde{p}$ with $\tilde{q}$ will be discussed further below.
-The $\tilde{q}$ approximation gives $\langle x_j \rangle_{\tilde{q}}=\langle x_j \rangle_p/C=l_j/C$,
+The $\tilde{q}$ approximation gives $\langle x_j \rangle_{\tilde{q}}=\langle x_j \rangle_p/C=l_j/C$, where $C=1-e^{-(L-l_i)}$, with variances and covariances given by
 \begin{align}
@@
 &=\frac{1}{C}\langle x_j x_k \rangle_p-\frac{l_jl_k}{C^2}\nonumber\\
-&=\frac{l_j l_k}{C}\left(1-\frac{1}{C}\right),\label{eq:covr}
+&=\frac{l_j l_k}{C}\left(1-\frac{1}{C}\right)\qquad\qquad $j\neq k$. \label{eq:covr}
 \end{align}
-where $C=1-e^{-(L-l_i)}$ and $j\neq k$.
 
@@
 \eqref{eq:meanxjrare} correctly predicts that there are on average $\sum_{j\neq i}\langle x_j \rangle_{\tilde{p}}\approx L-1$ nonfocal propagules because $\tilde{p}$ accounts for potentially large negative covariances between the $x_j$ ``after dispersal''.
-By neglecting the latter covariences, $\tilde{q}$ overestimates the fluctuations in $\sum_{j\neq i} c_j x_j$; thus $\tilde{q}$ gives an upper bound on the fluctuations.
-The discrepancy between $\tilde{q}$ and $\tilde{p}$ will be largest when $L$ is of order $1$ or smaller, because then the propagule assumed to already be present under $\tilde{q}$ is comparable to, or greater than, the entire propgaule density.
+By neglecting the latter covariances, $\tilde{q}$ overestimates the fluctuations in $\sum_{j\neq i} c_j x_j$; thus $\tilde{q}$ gives an upper bound on the fluctuations.
+The discrepancy between $\tilde{q}$ and $\tilde{p}$ will be largest when $L$ is of order $1$ or smaller, because then the propagule assumed to already be present under $\tilde{q}$ is comparable to, or greater than, the entire propagule density.
 
@@
 \end{equation}
-where $\hat{p}$ is the probability distribution of both focal and nonfocal propagaule abundances \textit{after} dispersal in those territories where at least two focal propagules landed.
+where $\hat{p}$ is the probability distribution of both focal and nonfocal propagule abundances \textit{after} dispersal in those territories where at least two focal propagules landed.
 

### 2017-12-20 manuscript.tex
l

@@
 
-We assume that reproductively mature individuals (``adults'') each require their own territory to survive and reproduce (Fig.~\ref{fig:lottery}).
+We assume that reproductively mature individuals (``adults'') each require their own territory to survive and reproduce.
 All territories are identical, and the total number of territories is $T$.
@@
 \end{equation}
+Fig.~2 illustrates one iteration of the model.
 
@@
 \end{equation}
-where $\langle \rangle_{\tilde{p}}$ denotes the expectation with respect to $\tilde{p}$, and $\tilde{p}$ is the probability distribution of nonfocal propagule abundances $x_j$ \textit{after} dispersal, in those territories where exactly one focal propagule, and at least one non-focal propagule, landed.
+where $\langle \rangle_{\tilde{p}}$ denotes the expectation with respect to $\tilde{p}$, and $\tilde{p}$ is the probability distribution of nonfocal propagule abundances $x_j$ \textit{after} dispersal (the significance of evaluating after dispersal should become clearer below), in those territories where exactly one focal propagule, and at least one non-focal propagule, landed.
 
@@
 We first calculate $\langle x_j \rangle_{\tilde{p}}$.
-Let $X=\sum_j x_j$ denote the total number of propagules in a territory and ${\mathbf x_i}=(x_1,\ldots,x_{i-1},x_{i+1}\ldots,x_G)$ denote the vector of non-focal abundances, so that $p({\mathbf x_i})=p_1(x_1)\ldots p_{i-1}(x_{i-1})p_{i+1}(x_{i+1})\ldots p_G(x_G)$.
-Then, $\tilde{p}$ can be written as
-\begin{align}
-\tilde{p}({\mathbf x_i})&=p({\mathbf x_i}|X\geq 2,x_i=1)\nonumber\\
-&=\frac{P({\mathbf x_i},X\geq 2|x_i=1)}{P(X\geq 2)}\nonumber\\
-&=\frac{1}{1-(1+L)e^{-L}}\sum_{X=2}^{\infty} P(X) p({\mathbf x_i}|X_i=X-1),
-\end{align}
-and so
+Let $X=\sum_j x_j$ denote the total number of propagules in a territory and ${\mathbf x_i}=(x_1,\ldots,x_{i-1},x_{i+1}\ldots,x_G)$ denote the vector of non-focal abundances, so that $p({\mathbf x_i})=p_1(x_1)\cdots p_{i-1}(x_{i-1})p_{i+1}(x_{i+1})\cdots p_G(x_G)$.
+Then, $\tilde{p}$ is the probability distribution for ${\mathbf x_i}$ conditional on the territory having received $X\geq 2$ propagules, where $X_i=X-1$ of them are nonfocal, giving
+\begin{equation}
+\tilde{p}({\mathbf x_i})=\sum_{X=2}^{\infty}P(X|X\geq 2) p({\mathbf x_i}|X_i=X-1)
+\end{equation}
+Here the total number of propagules $X$ follows a Poisson distribution with mean $L$, and we have $P(X|X\geq 2)=P(X)/P(X\geq 2)=P(X)/(1-(1+L)e^{-L})$. Thus,
 \begin{align}
@@
 We replace $\tilde{p}$ with $\tilde{q}$, defined as the ${\mathbf x_i}$ Poisson dispersal probabilities conditional on $X_i\geq1$ (which are independent).
-The distinction between $\tilde{p}$ with $\tilde{q}$ will be discussed further below.
+The distinction between $\tilde{p}$ and $\tilde{q}$ will be discussed further below.
 The $\tilde{q}$ approximation gives $\langle x_j \rangle_{\tilde{q}}=\langle x_j \rangle_p/C=l_j/C$, where $C=1-e^{-(L-l_i)}$, with variances and covariances given by
@@
 &=\frac{1}{C}\langle x_j x_k \rangle_p-\frac{l_jl_k}{C^2}\nonumber\\
-&=\frac{l_j l_k}{C}\left(1-\frac{1}{C}\right)\qquad\qquad $j\neq k$. \label{eq:covr}
+&=\frac{l_j l_k}{C}\left(1-\frac{1}{C}\right)\qquad\qquad j\neq k. \label{eq:covr}
 \end{align}
 
-The exact distribution $\tilde{p}$ assumes that exactly one of the  propagules present in a given site after dispersal belongs to the focal type, whereas $\tilde{q}$ assumes that there is a focal propagule present before non-focal dispersal commences.
-As a result, $\tilde{q}$ predicts that the mean propagule density is greater than $L$ (in sites with only one focal propagule is present) when the focal type is rare and the propagule density is high.
-This is erroneous, because the mean number of propagules in every site is $L$ by definition.
-Specifically, if $L-l_i \approx L\gg 1$, then the mean propagule density predicted by $\tilde{q}$ is approximately $L+1$.
-The discrepancy causes rare invaders to have an intrinsic rarity disadvantage (territorial contests under $\tilde{q}$ are more intense than they should be).
-In contrast, Eq.
+The distinction between $\tilde{p}$ and $\tilde{q}$ is closely connected to why the exact dispersal probability distribution $\tilde{p}$ is evaluated \textit{after} dispersal.
+$\tilde{p}$ assumes that one of the propagules present in a given territory after dispersal belongs to the focal type (a statement about propagule identity), whereas $\tilde{q}$ assumes that there is a focal propagule present before non-focal dispersal commences (a statement about propagule presence).
+As a result, if the focal type is rare, $\tilde{q}$ predicts that the mean propagule density is $L+1$ in territories with one focal propagule.
+This is erroneous, because the mean number of propagules per territory when lottery competition commences is $L$, regardless of whether a focal propagule happens to be present.
+The discrepancy causes rare invaders evaluated under $\tilde{q}$ to have an intrinsic rarity disadvantage, which is absurd (i.e.
+even if $b$, $c$ and $d$ are identical, $\tilde{q}$ predicts that a rare type will decline in frequency).
+By contrast, Eq.
 \eqref{eq:meanxjrare} correctly predicts that there are on average $\sum_{j\neq i}\langle x_j \rangle_{\tilde{p}}\approx L-1$ nonfocal propagules because $\tilde{p}$ accounts for potentially large negative covariances between the $x_j$ ``after dispersal''.
@@
 \end{align}
+In going from line 3 to 4, we used the same logic used to evaluate the inner sum in Eq.~\eqref{eq:raremonster1}, and in going from 4 to 5 we have separately evaluated the contributions from the $X$ and $x_i$ terms in the numerator.
 
-To calculate the relative fluctuations in $\sum_{j\neq i} c_j x_j$, we use a similar approximation as for $\Delta_r n_i$: $\hat{p}$ is approximated by $\hat{q}$, defined as the ${\mathbf x}$ dispersal probabilities in a territory conditional on $x_i>2$ (that is, treating the $x_j$ as indepenent).
+To calculate the standard deviation in $\sum_{j\neq i} c_j x_j$, we use a similar approximation as for $\Delta_r n_i$: $\hat{p}$ is approximated by $\hat{q}$, defined as the ${\mathbf x}$ dispersal probabilities in a territory conditional on $x_i>2$ (that is, treating the $x_j$ as independent).
 All covariances between nonfocal types are now zero, so that $\sigma_{\hat{q}}^2(\sum c_j x_j)=\sum c_j^2 \sigma_{\hat{q}}^2(x_j)$, where $\sigma_{\hat{q}}^2(x_j)=l_j$ for $j\neq i$, and

### 2017-12-22 manuscript.tex
l

@@
 
-In the previous section we showed that selection and the regulation of population density can be independent even if population growth is density-regulated.
-Nevertheless, selection and density regulation \textit{are} intimately connected in widely used models of population growth, as well as for the $b$ and $d$ traits in our variable-density lottery model.
+In the previous section we showed that $c$-selection and the regulation of population density are independent even though population growth is density-regulated in our variable-density lottery.
+Nevertheless, selection and density regulation \textit{are} intimately connected in widely used models of population growth, as well as for the lottery $b$ and $d$ traits.
 
-To see why this potentially poses a threat to relative fitness, consider the simple birth-death model \cite[pp. 20]{kostitzin_1939} \citep{travis_2013} 
+To see why this connection potentially poses a threat to relative fitness, consider the simple birth-death model \cite[pp. 20]{kostitzin_1939} \citep{travis_2013} 
 \begin{equation}
@@
 The corresponding selection coefficient increases from $s_{\rm initial}= \epsilon b_i$ to $s_{\rm final}=s_{\rm initial}/(1-\epsilon)$.
-Consequently, substantial deviations from Eq.~\eqref{eq:canonical} occur with proportional changes to $\delta$ of order $\epsilon=0.1$ and upwards (Fig.~\ref{fig:strengthofselection}).
+Consequently, noticeable deviations from Eq.~\eqref{eq:canonical} occur with proportional changes to $\delta$ of order $\epsilon=0.2$ and upwards i.e.
+selection must be quite strong (Fig.~\ref{fig:strengthofselection}).
 
@@
 \includegraphics[scale=0.8]{strengthofselection.pdf}
-\caption{\label{fig:strengthofselection} Proportional change in the selection coefficient over a ``$K$-like'' sweep for a type that experiences proportionally $1-\epsilon$ fewer deaths induced by crowding.
-The population is in demographic equilibrium at the start and end of the sweep.}
+\caption{\label{fig:strengthofselection} (a) Proportional change in the selection coefficient over a ``$K$-like'' sweep for a type that experiences proportionally $1-\epsilon$ fewer deaths induced by crowding.
+The population is in demographic equilibrium at the start and end of the sweep.
+(b) Example equilibrium-to-equilibrium $\delta$-sweep (Eq.~\ref{eq:Ndependentsweep}) for $\epsilon=0.2$ showing a noticeable deviation from the canonical selection equation.}
 \end{figure}
@@
 It can be seen that the mortality $d$ is akin to the birth rate in Eq.~\eqref{eq:simplebirthdeath}, and so, while $d$ does affect density, selection on $d$ is density independent.
-Thus, $d$ sweeps follow the canonical relative fitness model exactly.
+Thus, $d$ sweeps follow the canonical relative fitness model exactly (Fig.~\ref{fig:DDS_lottery}).
 
@@
 Since selection must already be quite strong for a $\delta$ sweep to threaten Eq.~\eqref{eq:canonical}, we conclude that $b$ sweeps also obey the canonical selection equation (to a close approximation).
+
+\begin{figure}
+\centering
+\includegraphics[scale=0.8]{DDS_lottery.pdf}
+\caption{\label{fig:DDS_lottery} The density-dependence of selection in our variable-density lottery plotted as the selection coefficient $(\Delta n_j-\Delta n_i)/n_j$ experienced by an adaptive variant $j$ present at the same frequency as a wildtype $i$.
+Here $b_i=1$, $d_i=0.5$ and $c_i=1$.
+For $b$-selection we set $b_j=b_i(1+\epsilon)$, and similarly for $c$ and $d$, with $\epsilon=0.1$.
+$d$-selection is density-independent, $b$-selection gets weaker with lower territorial availability, while $c$-selection initially increases with density as territorial contests become more important, but eventually also declines as  available territories become scarce.
+The equilibrium density for the $i$ type is $\approx 0.4$.}
+\end{figure}
 

### 2017-12-22 DDS_lottery.py

@@ -1,0 +1,78 @@
+# -*- coding: utf-8 -*-
+"""
+Created on Wed Dec 13 11:40:51 2017
+
+@author: jbertram
+"""
+
+import numpy as np
+import matplotlib.pyplot as plt
+import scipy.optimize as optimize
+
+T=100000
+totaltime=80
+b=np.array([1.,1.])
+c=np.array([1.,1.])
+d=np.array([0.1,0.1])
+n0=np.array([49000.,100.])
+
+n=n0;
+nhist=[n];
+for t in range(totaltime/2):
+    U=T-sum(n)
+    n=n+deltnplus(b*n*U/T,c,U)-d*n
+    nhist.append(list(n))
+
+#plt.plot(np.sum(nhist,1))
+n0=np.array(n)
+print n0 #equilibrium
+
+s=1.
+b=np.array([1.,1.*(1+s)])
+n=n0;
+nhist=[n];
+for t in range(totaltime):
+    U=T-sum(n)
+    n=n+deltnplus(b*n*U/T,c,U)-d*n
+    nhist.append(list(n))
+    
+nhist=np.array(nhist)
+Nhist=np.sum(nhist,1)
+Dhist=Nhist/T
+bhist=np.array(np.sum(b*nhist,1)/Nhist)
+bhist[0]=1.
+
+plt.plot((1/Dhist-1)*(1-np.exp(-bhist*Dhist))/d[0])
+
+plt.plot(Dhist)
+
+#=======================================================
+fig1, ax1 = plt.subplots(figsize=[5,4])
+
+T=100000
+b=np.array([1.,1.])
+c=np.array([1.,1.])
+d=np.array([.5,.5])
+eps=0.1
+
+ns=np.array([_*np.array([1,1]) for _ in np.linspace(1,T/2,100)]) 
+
+bDDS=np.sum(np.array([-1,1])*np.array([deltnplus(np.array([1.,1+eps])*b*x*(1-sum(x)/T),c,T-sum(x))/x-d for x in ns]),1)
+cDDS=np.sum(np.array([-1,1])*np.array([deltnplus(b*x*(1-sum(x)/T),np.array([1.,1+eps])*c,T-sum(x))/x-d for x in ns]),1)
+dDDS=np.sum(np.array([-1,1])*np.array([-np.array([1.,1-eps])*d for x in ns]),1)
[…]

### 2017-12-22 strength_of_selection.py

@@ -12,7 +12,7 @@ from scipy.integrate import odeint
 #SS figure
 #========================================================
 
-fig1,ax1=plt.subplots(figsize=[5,4])
+fig1, (ax1, ax2) = plt.subplots(ncols=2,figsize=[8,4])
 
 s=np.linspace(0,0.5,100)
 ax1.plot(s,1/(1-s),'k',linewidth=2)
@@ -23,32 +23,36 @@ ax1.set_ylabel(r"$s_{\rm final}/s_{\rm initial}$",fontsize=20)
 
 ax1.xaxis.set_label_coords(0.5, -0.09)
 ax1.yaxis.set_label_coords(-0.1, 0.5)
+ax1.annotate(r'$(a)$',xy=(0.01,0.93),xycoords='axes fraction',fontsize=16)
 
-plt.tight_layout()
+b=np.array([1,1]);
+s=0.2
+d=np.array([1e-4,1e-4*(1-s)]);
 
-plt.savefig('/home/jason/repos/densitydependentlottery/strengthofselection.pdf',bbox="tight")
-
-#Logistic model
-#==========================================================
-
-r=[1,1];
-s=0.3
-K=[1e5,1e5*(1+s)];
-
-x=np.linspace(0,20/s,1000)
+x=np.linspace(0,10/s,1000)
 
 def f(n,t):
-    return [r[0]*(1-(n[0]+n[1])/K[0])*n[0],r[0]*(1-(n[0]+n[1])/K[1])*n[1]]
+    return (b-d*sum(n))*n
     
 def g(p,t):
     return s*p*(1-p)
 
-solK=odeint(f,[K[0]-1.,1.],x)
-solrel=odeint(g,1/K[0],x)
+solK=odeint(f,[0.99*b[0]/d[0],0.01*b[0]/d[0]],x)
+solrel=odeint(g,0.01,x)
 
-plt.plot(x,solK[:,1]/(solK[:,0]+solK[:,1]))
-plt.plot(x,solrel)
-plt.ylim([0,1])
+ax2.plot(x,solK[:,1]/(solK[:,0]+solK[:,1]),'k--', linewidth=2,label=r"$\epsilon=0.2$ (Eq. 11)")
+ax2.plot(x,solrel,'k', linewidth=2,label=r"Canonical (Eq. 1)")
+ax2.legend(loc='lower right',prop={'size':11})
[…]

### 2017-12-22 manuscript.tex
l

@@
 The question is whether relative fitness can be used when selection is not vanishingly weak.
-In age-structured populations that reproduce by outcrossing, selection cannot in general be represented in terms of type-specific relative-fitness constants unless selection is weak \citep[Chap.
+In general, age-structured populations that reproduce by outcrossing do not permit strong selection to be represented in terms of type-specific relative-fitness constants \citep[Chap.
 4]{charlesworth_1994}.
-We will therefore restrict our attention to asexual haploids with little or no age structure, where the relative fitness description may be valid even if selection is strong.
-This simplification also makes it easier to grasp the population ecological underpinnings of selection, as opposed to the role of mating systems or complex changes in age structure.
+We will therefore restrict our attention to asexual haploids with little or no age structure, where it is easier to evaluate how the success or failure of the relative fitness description is tied to the underlying population ecological assumptions.
 
@@
 Thus, strong selection is sufficient for relative fitness to break down in the classical view of density-dependent selection.
-Indeed, in the defense of Eq.~\eqref{eq:canonical} given by \cite{kimura1969natural}, it was necessary to assume that $s$ will be a few percent at most.
+Indeed, in the defense of Eq.~\eqref{eq:canonical} given by \cite{kimura1969natural}, it was assumed that $s$ will be a few percent at most.
+While this may be reasonable for adaptive mutations, there is no reason to expect selection on standing variation to be so weak, wild \textit{Drosophila} being an obvious counter-example \citep{bergland_2014}.
 

### 2017-12-22 manuscript.tex
l

@@
 
-\textit{Keywords}: Lottery model, competitive Lotka-Volterra, $r$/$K$-selection, interference competition, $R^*$ theory, eco-evo.
+\textit{Keywords}: Lottery model, competitive Lotka-Volterra, $r$/$K$-selection, interference competition, eco-evo.
 
@@
 Selection is commonly described in terms of relative fitness.
-Yet much of the ecological theory of selection in density-regulated populations seems to be incompatible with widely-used, constant-density relative fitness models such as the Wright-Fisher.
+Yet when selection is strong, the ecological view of selection in density-regulated populations seems to be incompatible with widely-used, constant-density relative fitness models such as the Wright-Fisher.
 Here we analyze the population ecological limits of relative fitness using a novel generalization of the Wright-Fisher model in which population density depends dynamically on the demographic rates of the types present.
-Our model clearly distinguishes between density-dependent selection and selection-dependent density.
-These two effects are confounded in many standard models of density-regulated population growth.
+Our model contains a ``reproductive excess’’, and clearly distinguishes between density-dependent selection and selection-dependent density.
+These two effects are confounded in standard models of density-regulated population growth.
 Both effects are necessary, in combination with strong selection, for relative fitness to break down in populations close to demographic equilibrium.
-Remarkably, both effects are not sufficient: we show that strong selection on a density-regulating trait that is subject to density-dependent selection can conform to the relative fitness description almost exactly.
-We reiterate the importance of reproductive excesses, an aspect of selection omitted in many population growth models, which implies that even strong selection need not affect density in many species.
+Remarkably, both effects are not sufficient: we give an example of strong selection on a density-regulating trait subject to density-dependent selection that conforms to the relative fitness description almost exactly.
+We reiterate the importance of reproductive excesses in many species, which allows even strong selection to have no effect on density.
 Our model also offers a possible alternative to relative fitness when the latter is untenable, as is likely the case far from demographic equilibrium.
 
-\noindent (195 words)
+\noindent (191 words)
 

### 2018-01-02 manuscript.tex
l

@@
 This justifies Eq.~\eqref{eq:canonical} because, for a population in demographic equilibrium, selective sweeps do not affect density, and so the density-dependence of selection does not matter.
-Unfortuantely this reproductive excess literature is also poorly integrated with population ecology.
+Unfortunately this reproductive excess literature is also poorly integrated with population ecology.
 \cite{kimura1969natural} took constant $N$ as a requirement and then derived some variants of the logistic model that satisfy this requirement.
@@
 Indeed, in the defense of Eq.~\eqref{eq:canonical} given by \cite{kimura1969natural}, it was assumed that $s$ will be a few percent at most.
-While this may be reasonable for adaptive mutations, there is no reason to expect selection on standing variation to be so weak, wild \textit{Drosophila} being an obvious counter-example \citep{bergland_2014}.
+While this may be reasonable for adaptive mutations, there is no reason to expect selection on standing variation to be so weak, wild \textit{Drosophila} being an obvious counter-example \citep{bergland_14}.
 
-Our variable-density lottery model shows that it is not simply a lack of ecological realism that underlies the contrast between relative fitness with the classical view of density-dependent selection.
+Our variable-density lottery model shows that it is not simply a lack of ecological realism that underlies the contrast between relative fitness and the classical view of density-dependent selection.
 Rather, in many population growth models, only one life-history stage is represented, and the competitive effects resulting from crowding appear as a reduction in absolute fitness that only depends on the type densities at this life-history stage (e.g.

### 2018-02-22 manuscript.tex
l

@@
 The constant-$N$ behavior of $c$-selection arises from the role of $c$ as a trait determining relative competitive success in territorial contests.
-As such, this behavior is a result of a reproductive excesses Eq.~\eqref{eq:canonical}.
+As such, this behavior is a result of a reproductive excesses.
 By contrast, previous models of selection-independent density either used unusual models of competition \citep{kimura1969natural,nei1971fertility}, or made restrictive parameter choices in the Lotka-Volterra model (Appendix C; \citealt{smouse_1976,mallet_2012}).
@@
 
-\section*{Appendix A: Poisson approximation}
+\section*{Appendix A: Poisson dispersal}
 
-For simplicity of presentation, we have assumed a Poisson distribution for the $x_i$ as our model of dispersal.
-Strictly speaking, the total number of $i$ propagules $\sum x_i$ (summed over unoccupied territories) is then no longer a constant $m_i$, but fluctuates between generations for a given mean $m_i$, which is more biologically realistic.
+For simplicity of presentation, we assume a Poisson distribution for the $x_i$ as our model of dispersal.
+Strictly speaking, the total number of $i$ propagules $\sum x_i$ (summed over unoccupied territories) is then no longer a constant $m_i$, but fluctuates between generations for a given mean $m_i$.
 Nevertheless, since we do not consider the random fluctuations in type abundances here, and for ease of comparison with the classic lottery model, we ignore the fluctuations in $m_i$.
-Instead we focus, on Poisson fluctuations in propagule composition in each territory.
+Instead we focus on Poisson fluctuations in propagule composition in each territory.
 
@@
 $x_i=1$ and $X_i\geq 1$ where $X_i=\sum_{j\neq i} x_j$ is the number of nonfocal propagules.
-The number of territories where this occurs is $Up_i(1)P(X_i\geq 1)=b_i n_i e^{-l_i}(1-e^{-(L-l_i)})$.
+The number of territories where this occurs is $Up_i(1)P(X_i\geq 1)=m_i e^{-l_i}(1-e^{-(L-l_i)})$.
 Thus

### 2018-02-23 manuscript.tex
l

@@
 \end{equation}
-where $\langle \rangle_{\tilde{p}}$ denotes the expectation with respect to $\tilde{p}$, and $\tilde{p}$ is the probability distribution of nonfocal propagule abundances $x_j$ \textit{after} dispersal (the significance of evaluating after dispersal should become clearer below), in those territories where exactly one focal propagule, and at least one non-focal propagule, landed.
+where $\langle \rangle_{\tilde{p}}$ denotes the expectation with respect to $\tilde{p}$, and $\tilde{p}$ is the probability distribution of nonfocal propagule abundances $x_j$, in those territories where exactly one focal propagule, and at least one non-focal propagule, landed.
 
-Our approximation is to replace $x_j$ with its mean in the last term in Eq.~\eqref{eq:deltr},
+To evaluate the expectation in Eq. \eqref{eq:deltr}, we replace the $x_j$ with  ``effective'' mean values $\langle x_j\rangle_{\tilde{q}}$, to obtain the approximation
 \begin{equation}
-\left\langle\frac{c_i}{c_i +\sum_{j\neq i} c_j x_j}\right\rangle_{\tilde{p}}\approx \frac{c_i}{c_i +\sum_{j\neq i} c_j \langle x_j\rangle_{\tilde{p}}}.\label{eq:meanfieldr}
+\left\langle\frac{c_i}{c_i +\sum_{j\neq i} c_j x_j}\right\rangle_{\tilde{p}}\approx \frac{c_i}{c_i +\sum_{j\neq i} c_j \langle x_j\rangle_{\tilde{q}}}.\label{eq:meanfieldr}
 \end{equation}
+Note that the effective means $\langle x_j\rangle_{\tilde{q}}$ are taken with respect to a new distribution $\tilde{q}$.
+We will define $\tilde{q}$ below; first we explain the motivation for introducing it.
+Suppose that we use $\tilde{p}$ instead of $\tilde{q}$ in Eq.
+\eqref{eq:meanfieldr} ($x_j\rightarrow \langle x_j\rangle_{\tilde{p}}$).
+Moreover, suppose that the focal type is rare ($l_i\ll 1$), that the total propagule density is high $L\gg 1$, and that all types have identical $b$, $c$ and $d$.
+Then the mean total propagule density in territories with one focal propagule is $1+\sum_j \langle x_j\rangle_{\tilde{p}}\approx L+1$ and the right hand side of Eq.
+\eqref{eq:meanfieldr} is approximately $1/(L+1)$.
+Since $L$ is large and the focal type is rare, a negligible fraction of territories will be have a single uncontested focal propagule or more than one focal propagule.
+Thus, $1/(L+1)$ is the probability that a random focal propagule successfully wins a territory in the Eq.
+\eqref{eq:meanfieldr} approximation.
+On the other hand, virtually all nonfocal propagules
+
+ territories
+
+\label{eq:meanfieldr}
+
+Since $x_i$ and the $x_j$ are independent under $\tilde{p}$
+
+If the focal type is rare, the mean propagule density is $L+1$ in territories with one focal propagule.
+
+The discrepancy causes rare invaders evaluated under $\tilde{q}$ to have an intrinsic rarity disadvantage, which is absurd (i.e.
+even if $b$, $c$ and $d$ are identical, $\tilde{q}$ predicts that a rare type will decline in frequency).
+
+By contrast, Eq.
+\eqref{eq:meanxjrare} correctly predicts that there are on average $\sum_{j\neq i}\langle x_j \rangle_{\tilde{p}}\approx L-1$ nonfocal propagules because $\tilde{p}$ accounts for potentially large negative covariances between the $x_j$ ``after dispersal''.
+By neglecting the latter covariances, $\tilde{q}$ overestimates the fluctuations in $\sum_{j\neq i} c_j x_j$; thus $\tilde{q}$ gives an upper bound on the fluctuations.
+The discrepancy between $\tilde{q}$ and $\tilde{p}$ will be largest when $L$ is of order $1$ or smaller, because then the propagule assumed to already be present under $\tilde{q}$ is comparable to, or greater than, the entire propagule density.
+
+$\tilde{p}$
+
+The distinction between $\tilde{p}$ and $\tilde{q}$ is closely connected to why the exact dispersal probability distribution $\tilde{p}$ is evaluated \textit{after} dispersal.
+$\tilde{p}$ assumes that one of the propagules present in a given territory after dispersal belongs to the focal type (a statement about propagule identity), whereas $\tilde{q}$ assumes that there is a focal propagule present before non-focal dispersal commences (a statement about propagule presence).
+
 Below we justify this replacement by arguing that the standard deviation $\sigma_{\tilde{p}}(\sum_{j\neq i} c_j x_j)$ (with respect to $\tilde{p}$), is much smaller than $\langle\sum_{j\neq i} c_j x_j\rangle_{\tilde{p}}$.
@@
 \end{align}
-
-The distinction between $\tilde{p}$ and $\tilde{q}$ is closely connected to why the exact dispersal probability distribution $\tilde{p}$ is evaluated \textit{after} dispersal.
-$\tilde{p}$ assumes that one of the propagules present in a given territory after dispersal belongs to the focal type (a statement about propagule identity), whereas $\tilde{q}$ assumes that there is a focal propagule present before non-focal dispersal commences (a statement about propagule presence).
-As a result, if the focal type is rare, $\tilde{q}$ predicts that the mean propagule density is $L+1$ in territories with one focal propagule.
-This is erroneous, because the mean number of propagules per territory when lottery competition commences is $L$, regardless of whether a focal propagule happens to be present.
-The discrepancy causes rare invaders evaluated under $\tilde{q}$ to have an intrinsic rarity disadvantage, which is absurd (i.e.
-even if $b$, $c$ and $d$ are identical, $\tilde{q}$ predicts that a rare type will decline in frequency).
-By contrast, Eq.
-\eqref{eq:meanxjrare} correctly predicts that there are on average $\sum_{j\neq i}\langle x_j \rangle_{\tilde{p}}\approx L-1$ nonfocal propagules because $\tilde{p}$ accounts for potentially large negative covariances between the $x_j$ ``after dispersal''.
-By neglecting the latter covariances, $\tilde{q}$ overestimates the fluctuations in $\sum_{j\neq i} c_j x_j$; thus $\tilde{q}$ gives an upper bound on the fluctuations.
-The discrepancy between $\tilde{q}$ and $\tilde{p}$ will be largest when $L$ is of order $1$ or smaller, because then the propagule assumed to already be present under $\tilde{q}$ is comparable to, or greater than, the entire propagule density.
 

### 2018-02-24 manuscript.tex
l

@@
 
-We separate the right hand side of Eq.~\eqref{eq:growthsumuncoupled} into three components $\Delta_+ n_i = \Delta_u n_i+\Delta_r n_i+\Delta_a n_i$ which vary in relative magnitude depending on the propagule densities $l_i$.
-Following the notation in the main text, the Poisson distributions for the $x_i$ (or some subset of the $x_i$) will be denoted $p$, and we use $P$ as a general shorthand for the probability of particular outcomes.
-
-\subsection*{Growth without competition}
-
-The first component, $\Delta_u n_i$, accounts for territories where only one focal propagule is present $x_i=1$ and $x_j=0$ for $j\neq i$ ($u$ stands for ``uncontested'').
+In this appendix we derive Eq.~\eqref{eq:master}. We start by separating the right hand side of Eq.~\eqref{eq:growthsumuncoupled} into three components
+\begin{equation}
+\Delta_+ n_i = \Delta_u n_i+\Delta_r n_i+\Delta_a n_i,
+\end{equation}
+which vary in relative magnitude depending on the propagule densities $l_i$.
+The first component, $\Delta_u n_i$, accounts for territories where only one focal propagule is present ($x_i=1$ and $x_j=0$ for $j\neq i$; $u$ stands for ``uncontested'').
 The proportion of territories where this occurs is $l_i e^{-L}$, and so
@@
 
-\subsection*{Competition when rare}
-
-The second component, $\Delta_r n_i$, accounts for territories where a single focal propagule is present along with at least one non-focal propagule ($r$ stands for ``rare'') i.e.
-$x_i=1$ and $X_i\geq 1$ where $X_i=\sum_{j\neq i} x_j$ is the number of nonfocal propagules.
+The second component, $\Delta_r n_i$, accounts for territories where a single focal propagule is present along with at least one non-focal propagule ($x_i=1$ and $X_i\geq 1$ where $X_i=\sum_{j\neq i} x_j$ is the number of nonfocal propagules; $r$ stands for ``rare'').
 The number of territories where this occurs is $Up_i(1)P(X_i\geq 1)=m_i e^{-l_i}(1-e^{-(L-l_i)})$.
@@
 \end{equation}
-where $\langle \rangle_{\tilde{p}}$ denotes the expectation with respect to $\tilde{p}$, and $\tilde{p}$ is the probability distribution of nonfocal propagule abundances $x_j$, in those territories where exactly one focal propagule, and at least one non-focal propagule, landed.
+where $\langle \rangle_{\tilde{p}}$ denotes the expectation with respect to the probability distribution $\tilde{p}$ of nonfocal propagule abundances $x_j$, in those territories where exactly one focal propagule, and at least one non-focal propagule, landed.
 
-To evaluate the expectation in Eq. \eqref{eq:deltr}, we replace the $x_j$ with  ``effective'' mean values $\langle x_j\rangle_{\tilde{q}}$, to obtain the approximation
+The final contribution, $\Delta_a n_i$, accounts for territories where two or more focal propagules are present ($x_i\geq 2$; $a$ stands for ``abundant"). Similar to Eq.~\eqref{eq:deltr}, we have 
+\begin{equation}
+\Delta_a n_i=U(1-(1+l_i)e^{-l_i})\left\langle \frac{c_i x_i}{\sum_j c_j x_j} \right\rangle_{\hat{p}}\label{eq:delta}
+\end{equation}
+where $\hat{p}$ is the probability distribution of both focal and nonfocal propagule abundances in those territories where at least two focal propagules landed.
+
+To derive Eq.~\eqref{eq:master} we approximate the expectations in Eq.~\eqref{eq:deltr} and Eq.~\eqref{eq:delta} by replacing $x_i$ and the $x_j$ with ``effective'' mean values as follows 
 \begin{equation}
@@
 \end{equation}
-Note that the effective means $\langle x_j\rangle_{\tilde{q}}$ are taken with respect to a new distribution $\tilde{q}$.
-We will define $\tilde{q}$ below; first we explain the motivation for introducing it.
-Suppose that we use $\tilde{p}$ instead of $\tilde{q}$ in Eq.
-\eqref{eq:meanfieldr} ($x_j\rightarrow \langle x_j\rangle_{\tilde{p}}$).
-Moreover, suppose that the focal type is rare ($l_i\ll 1$), that the total propagule density is high $L\gg 1$, and that all types have identical $b$, $c$ and $d$.
-Then the mean total propagule density in territories with one focal propagule is $1+\sum_j \langle x_j\rangle_{\tilde{p}}\approx L+1$ and the right hand side of Eq.
+\begin{equation}
+\left\langle \frac{c_i x_i}{\sum_j c_j x_j} \right\rangle_{\hat{p}}\approx  \frac{c_i \langle x_i \rangle_{\hat{q}}}{\sum_j c_j \langle x_j\rangle_{\hat{q}}}.\label{eq:meanfielda}
+\end{equation}
+Here the effective means $\langle \rangle_{\tilde{q}}$ and $\langle \rangle_{\hat{q}}$ are taken with respect to new distributions $\tilde{q}$ and $\hat{q}$, respectively.
+In the following subsection we define $\tilde{q}$ and $\hat{q}$ and explain our reasoning for using these distributions to take the effective means.
+
+\subsection*{The effective distributions $\tilde{q}$ and $\hat{q}$}
+
+When making approximations \eqref{eq:meanfieldr} and \eqref{eq:meanfielda}, we must ensure that the resulting growth equations are logically consistent.
+For instance, suppose that two identical types are present (same $b$, $c$ and $d$), one with low density $l_1\ll 1$, and the other with high density $l_2\approx L\gg 1$.
+Since $L$ is large, uncontested territories make a negligible contribution to growth.
+Consequently, the rare type grows almost entirely due to $\Delta_r n_1$, while the common type grows almost entirely due to $\Delta_a n_2$.
+The approximate per-capita growth rates $\Delta_r n_1/m_1 = \Delta_a n_2/m_2$ implied by \eqref{eq:meanfieldr} and \eqref{eq:meanfielda} must then be exactly equal for logical consistency.
+Even if \eqref{eq:meanfieldr} and \eqref{eq:meanfielda} are individually excellent approximations, without this consistency condition the resulting approximate model implies that the rare type grows or declines purely by virtue of being rare, but we could select any rare subset of a single-type population and predict the same behavior --- a contradiction.
+
+To be more concrete, suppose that we naively use $\tilde{p}$ and $\hat{p}$ instead of $\tilde{q}$ and $\hat{q}$ in Eqs.~\eqref{eq:meanfieldr} and \eqref{eq:meanfielda}, respectively.
+Then, for the rare type, the mean total propagule density in territories were it is present is $\langle x_2\rangle_{\tilde{p}}+1\approx L+1$ (since the rare and common types disperse independently), and the right hand side of Eq.
 \eqref{eq:meanfieldr} is approximately $1/(L+1)$.
-Since $L$ is large and the focal type is rare, a negligible fraction of territories will be have a single uncontested focal propagule or more than one focal propagule.
-Thus, $1/(L+1)$ is the probability that a random focal propagule successfully wins a territory in the Eq.
-\eqref{eq:meanfieldr} approximation.
+
+in the Eq. \eqref{eq:meanfieldr} approximation.
+
 On the other hand, virtually all nonfocal propagules
 
- territories
+$\tilde{q}$ and $\hat{q}$ are chosen to
 
-\label{eq:meanfieldr}
+$\langle x_j\rangle_{\tilde{q}}$, to obtain the approximation
+
+%Following the notation in the main text, the Poisson distributions for the $x_i$ (or some subset of the $x_i$) will be denoted $p$, and we use $P$ as a general shorthand for the probability of particular outcomes.
 
@@
 If the focal type is rare, the mean propagule density is $L+1$ in territories with one focal propagule.
-
-The discrepancy causes rare invaders evaluated under $\tilde{q}$ to have an intrinsic rarity disadvantage, which is absurd (i.e.
-even if $b$, $c$ and $d$ are identical, $\tilde{q}$ predicts that a rare type will decline in frequency).
 
@@
 
-The final contribution, $\Delta_a n_i$, accounts for territories where two or more focal propagules are present ($a$ stands for ``abundant"). Similarly to Eq.~\eqref{eq:deltr}, we have 
-\begin{equation}
-\Delta_a n_i=U(1-(1+l_i)e^{l_i})\left\langle \frac{c_i x_i}{\sum_j c_j x_j} \right\rangle_{\hat{p}}\label{eq:delta}
-\end{equation}
-where $\hat{p}$ is the probability distribution of both focal and nonfocal propagule abundances \textit{after} dispersal in those territories where at least two focal propagules landed.
+Again, we argue that the relative fluctuations in $\sum c_j x_j$ are much smaller than $1$ (with respect to $\hat{p}$), so that,
 
-Again, we argue that the relative fluctuations in $\sum c_j x_j$ are much smaller than $1$ (with respect to $\hat{p}$), so that,
-\begin{equation}
-\left\langle \frac{c_i x_i}{\sum_j c_j x_j} \right\rangle_{\hat{p}}\approx  \frac{c_i \langle x_i \rangle_{\hat{p}}}{\sum_j c_j \langle x_j\rangle_{\hat{p}}}.\label{eq:meanfielda}
-\end{equation}
 Following a similar procedure as for $\Delta_r n_i$, where the vector of propagule abundances is denoted ${\mathbf x}$, the mean focal type abundance is, 

### 2018-02-24 manuscript.tex
l

@@
 
-\section*{Appendix B: Derivation of growth equation}
+\section*{Appendix B: Growth equation derivation}
 
@@
 
-When making approximations \eqref{eq:meanfieldr} and \eqref{eq:meanfielda}, we must ensure that the resulting growth equations are logically consistent.
-For instance, suppose that two identical types are present (same $b$, $c$ and $d$), one with low density $l_1\ll 1$, and the other with high density $l_2\approx L\gg 1$.
-Since $L$ is large, uncontested territories make a negligible contribution to growth.
-Consequently, the rare type grows almost entirely due to $\Delta_r n_1$, while the common type grows almost entirely due to $\Delta_a n_2$.
-The approximate per-capita growth rates $\Delta_r n_1/m_1 = \Delta_a n_2/m_2$ implied by \eqref{eq:meanfieldr} and \eqref{eq:meanfielda} must then be exactly equal for logical consistency.
-Even if \eqref{eq:meanfieldr} and \eqref{eq:meanfielda} are individually excellent approximations, without this consistency condition the resulting approximate model implies that the rare type grows or declines purely by virtue of being rare, but we could select any rare subset of a single-type population and predict the same behavior --- a contradiction.
+The approximations \eqref{eq:meanfieldr} and \eqref{eq:meanfielda} must be done in a way that ensures consistency between rare and common types.
+Suppose that two identical types are present (same $b$, $c$ and $d$), one with low density $l_1\ll 1$, and the other with high density $l_2\approx L\gg 1$.
+Since $L$ is large, uncontested territories make a negligible contribution each type's growth.
+The rare type grows almost entirely due to $\Delta_r n_1$, while the common type grows almost entirely due to $\Delta_a n_2$.
+To ensure consistency, these identical types should retain constant relative frequencies, and so the approximate per-capita growth rates $\Delta_r n_1/m_1 = \Delta_a n_2/m_2$ implied by \eqref{eq:meanfieldr} and \eqref{eq:meanfielda} must be equal.
 
-To be more concrete, suppose that we naively use $\tilde{p}$ and $\hat{p}$ instead of $\tilde{q}$ and $\hat{q}$ in Eqs.~\eqref{eq:meanfieldr} and \eqref{eq:meanfielda}, respectively.
-Then, for the rare type, the mean total propagule density in territories were it is present is $\langle x_2\rangle_{\tilde{p}}+1\approx L+1$ (since the rare and common types disperse independently), and the right hand side of Eq.
-\eqref{eq:meanfieldr} is approximately $1/(L+1)$.
+This consistency requirement is not automatically satisfied even if \eqref{eq:meanfieldr} and \eqref{eq:meanfielda} are individually good approximations.
+Even small violations will cause one of the identical types to grow exponentially relative to the other.
+This pathological growth occurs purely because one group is treated as rare and the other as common.
+Since a single-type population can be divided into arbitrary rare and common subsets of identical types, and clearly not all such subsets can grow/decline, we have a logical contradiction.
 
-in the Eq. \eqref{eq:meanfieldr} approximation.
+To be more concrete, suppose that we use $\tilde{p}$ and $\hat{p}$ in place of $\tilde{q}$ and $\hat{q}$ in Eqs.~\eqref{eq:meanfieldr} and \eqref{eq:meanfielda}, respectively.
+Then, since $\langle x_2\rangle_{\tilde{p}}\approx L$, the right hand side of Eq.
+\eqref{eq:meanfieldr} is approximately $1/(L+1)$ for the rare type.
+And, since $l_1\ll 1$ and $L\gg 1$ in Eq.~\eqref{eq:deltr}, we have $\Delta_r n_1 \approx 1/(L+1)$.
+Similarly, for the common type, $\sum_j \langle x_j\rangle_{\hat{p}}\approx L$ in Eq.
+\eqref{eq:meanfielda}, and so $\Delta_a n_2 \approx 1/L$.
+Therefore, the consistency requirement is violated if we naively use the conditional distributions $\tilde{p}$ and $\hat{p}$ derived from the dispersal distribution $p$ to calculate the effective means.
 
-On the other hand, virtually all nonfocal propagules
+The effective distributions $\tilde{q}$ and $\hat{q}$ are chosen to ensure that this consistency requirement is satisfied.
 
-$\tilde{q}$ and $\hat{q}$ are chosen to
-
-$\langle x_j\rangle_{\tilde{q}}$, to obtain the approximation
+$\tilde{p}$ assumes that one of the propagules present in a given territory after dispersal belongs to the focal type (a statement about propagule identity), whereas $\tilde{q}$ assumes that there is a focal propagule present before non-focal dispersal commences (a statement about propagule presence).
 
 %Following the notation in the main text, the Poisson distributions for the $x_i$ (or some subset of the $x_i$) will be denoted $p$, and we use $P$ as a general shorthand for the probability of particular outcomes.
-
-Since $x_i$ and the $x_j$ are independent under $\tilde{p}$
-
-If the focal type is rare, the mean propagule density is $L+1$ in territories with one focal propagule.
 
@@
 The discrepancy between $\tilde{q}$ and $\tilde{p}$ will be largest when $L$ is of order $1$ or smaller, because then the propagule assumed to already be present under $\tilde{q}$ is comparable to, or greater than, the entire propagule density.
-
-$\tilde{p}$
-
-The distinction between $\tilde{p}$ and $\tilde{q}$ is closely connected to why the exact dispersal probability distribution $\tilde{p}$ is evaluated \textit{after} dispersal.
-$\tilde{p}$ assumes that one of the propagules present in a given territory after dispersal belongs to the focal type (a statement about propagule identity), whereas $\tilde{q}$ assumes that there is a focal propagule present before non-focal dispersal commences (a statement about propagule presence).
 

### 2018-02-26 manuscript.tex
l

@@
 Suppose that two identical types are present (same $b$, $c$ and $d$), one with low density $l_1\ll 1$, and the other with high density $l_2\approx L\gg 1$.
-Since $L$ is large, uncontested territories make a negligible contribution each type's growth.
+Since $L$ is large, uncontested territories contribute negligibly to each type's growth.
 The rare type grows almost entirely due to $\Delta_r n_1$, while the common type grows almost entirely due to $\Delta_a n_2$.
@@
 
-To be more concrete, suppose that we use $\tilde{p}$ and $\hat{p}$ in place of $\tilde{q}$ and $\hat{q}$ in Eqs.~\eqref{eq:meanfieldr} and \eqref{eq:meanfielda}, respectively.
+For example, suppose that we use $\tilde{p}$ and $\hat{p}$ in place of $\tilde{q}$ and $\hat{q}$ in Eqs.~\eqref{eq:meanfieldr} and \eqref{eq:meanfielda}, respectively.
 Then, since $\langle x_2\rangle_{\tilde{p}}\approx L$, the right hand side of Eq.

### 2018-02-26 manuscript.tex
l

@@
 The approximations \eqref{eq:meanfieldr} and \eqref{eq:meanfielda} must be done in a way that ensures consistency between rare and common types.
-Suppose that two identical types are present (same $b$, $c$ and $d$), one with low density $l_1\ll 1$, and the other with high density $l_2\approx L\gg 1$.
+To illustrate, suppose that two identical types are present (same $b$, $c$ and $d$), one with low density $l_1\ll 1$, and the other with high density $l_2\approx L\gg 1$.
 Since $L$ is large, uncontested territories contribute negligibly to each type's growth.
 The rare type grows almost entirely due to $\Delta_r n_1$, while the common type grows almost entirely due to $\Delta_a n_2$.
-To ensure consistency, these identical types should retain constant relative frequencies, and so the approximate per-capita growth rates $\Delta_r n_1/m_1 = \Delta_a n_2/m_2$ implied by \eqref{eq:meanfieldr} and \eqref{eq:meanfielda} must be equal.
-
-This consistency requirement is not automatically satisfied even if \eqref{eq:meanfieldr} and \eqref{eq:meanfielda} are individually good approximations.
-Even small violations will cause one of the identical types to grow exponentially relative to the other.
-This pathological growth occurs purely because one group is treated as rare and the other as common.
-Since a single-type population can be divided into arbitrary rare and common subsets of identical types, and clearly not all such subsets can grow/decline, we have a logical contradiction.
+To ensure consistency, the approximate per-capita growth rates $\Delta_r n_1/m_1 = \Delta_a n_2/m_2$ implied by \eqref{eq:meanfieldr} and \eqref{eq:meanfielda} must be equal.
+Even small violations of this consistency condition will cause one type to grow exponentially relative to the other.
+This is a logical contradiction: any single-type population can be divided at will into rare and common subsets of identical types, and clearly not all such subsets can grow/decline.
 

### 2018-02-27 manuscript.tex
l

@@
 The approximations \eqref{eq:meanfieldr} and \eqref{eq:meanfielda} must be done in a way that ensures consistency between rare and common types.
-To illustrate, suppose that two identical types are present (same $b$, $c$ and $d$), one with low density $l_1\ll 1$, and the other with high density $l_2\approx L\gg 1$.
+To illustrate, suppose that two identical (same $b$, $c$ and $d$) types are present, one with low density $l_1\ll 1$, and the other with high density $l_2\approx L\gg 1$.
 Since $L$ is large, uncontested territories contribute negligibly to each type's growth.
 The rare type grows almost entirely due to $\Delta_r n_1$, while the common type grows almost entirely due to $\Delta_a n_2$.
-To ensure consistency, the approximate per-capita growth rates $\Delta_r n_1/m_1 = \Delta_a n_2/m_2$ implied by \eqref{eq:meanfieldr} and \eqref{eq:meanfielda} must be equal.
-Even small violations of this consistency condition will cause one type to grow exponentially relative to the other.
-This is a logical contradiction: any single-type population can be divided at will into rare and common subsets of identical types, and clearly not all such subsets can grow/decline.
+To ensure consistency, the approximate per-capita growth rates $\Delta_r n_1/m_1 = \Delta_a n_2/m_2$ implied by the approximations \eqref{eq:meanfieldr} and \eqref{eq:meanfielda} must be equal.
+Even small violations of this consistency condition would mean exponential growth of one type relative to the other.
+This is a serious pathology: any single-type population can be arbitrarily partitioned into rare and common identical subtypes, and the predicted growth or decline of any small group of individuals would depend on whether it is regarded as part of a rare or a common type.
 
-For example, suppose that we use $\tilde{p}$ and $\hat{p}$ in place of $\tilde{q}$ and $\hat{q}$ in Eqs.~\eqref{eq:meanfieldr} and \eqref{eq:meanfielda}, respectively.
-Then, since $\langle x_2\rangle_{\tilde{p}}\approx L$, the right hand side of Eq.
-\eqref{eq:meanfieldr} is approximately $1/(L+1)$ for the rare type.
-And, since $l_1\ll 1$ and $L\gg 1$ in Eq.~\eqref{eq:deltr}, we have $\Delta_r n_1 \approx 1/(L+1)$.
+This pathology arises if we naively use $\tilde{p}$ and $\hat{p}$ in place of $\tilde{q}$ and $\hat{q}$ to calculate the effective means.
+In this case we have $\langle x_2\rangle_{\tilde{p}}\approx L$, the right hand side of Eq.
+\eqref{eq:meanfieldr} is approximately $1/(L+1)$ for the rare type, and, since $l_1\ll 1$ and $L\gg 1$ in Eq.~\eqref{eq:deltr}, we have $\Delta_r n_1 \approx 1/(L+1)$.
 Similarly, for the common type, $\sum_j \langle x_j\rangle_{\hat{p}}\approx L$ in Eq.
 \eqref{eq:meanfielda}, and so $\Delta_a n_2 \approx 1/L$.
-Therefore, the consistency requirement is violated if we naively use the conditional distributions $\tilde{p}$ and $\hat{p}$ derived from the dispersal distribution $p$ to calculate the effective means.
+Thus, the identical rare type is erroneously predicted to decline in frequency.
 
-The effective distributions $\tilde{q}$ and $\hat{q}$ are chosen to ensure that this consistency requirement is satisfied.
+The effective distributions $\tilde{q}$ and $\hat{q}$ ensure that the above consistency requirement is satisfied.
 

### 2018-02-28 manuscript.tex
l

@@
 \doublespacing
+\sloppy
 
@@
 
-In this appendix we derive Eq.~\eqref{eq:master}. We start by separating the right hand side of Eq.~\eqref{eq:growthsumuncoupled} into three components
+In this appendix we derive Eq.~\eqref{eq:master}.
+Following the notation in the main text, the Poisson distributions for the $x_i$ (or some subset of the $x_i$) will be denoted $p$, and we use $P$ as a general shorthand for the probability of particular outcomes.
+
+We start by separating the right hand side of Eq.~\eqref{eq:growthsumuncoupled} into three components
 \begin{equation}
-\Delta_+ n_i = \Delta_u n_i+\Delta_r n_i+\Delta_a n_i,
+\Delta_+ n_i = \Delta_u n_i+\Delta_r n_i+\Delta_a n_i,\label{eq:delt_decomp}
 \end{equation}
@@
 To illustrate, suppose that two identical (same $b$, $c$ and $d$) types are present, one with low density $l_1\ll 1$, and the other with high density $l_2\approx L\gg 1$.
-Since $L$ is large, uncontested territories contribute negligibly to each type's growth.
-The rare type grows almost entirely due to $\Delta_r n_1$, while the common type grows almost entirely due to $\Delta_a n_2$.
-To ensure consistency, the approximate per-capita growth rates $\Delta_r n_1/m_1 = \Delta_a n_2/m_2$ implied by the approximations \eqref{eq:meanfieldr} and \eqref{eq:meanfielda} must be equal.
+Since $L$ is large, uncontested territories make up a negligible fraction of the total, and the rare type grows almost entirely due to $\Delta_r n_1$, while the common type grows almost entirely due to $\Delta_a n_2$.
+To ensure consistency, the approximate per-capita growth rates implied by the approximations \eqref{eq:meanfieldr} and \eqref{eq:meanfielda} must be equal $\Delta_r n_1/m_1 = \Delta_a n_2/m_2$.
 Even small violations of this consistency condition would mean exponential growth of one type relative to the other.
-This is a serious pathology: any single-type population can be arbitrarily partitioned into rare and common identical subtypes, and the predicted growth or decline of any small group of individuals would depend on whether it is regarded as part of a rare or a common type.
+This behavior is pathological: any single-type population can be arbitrarily partitioned into identical rare and common subtypes, but a given small group of individuals would be predicted to either grow or shrink depending on whether it viewed as part of the rare or common type.
 
-This pathology arises if we naively use $\tilde{p}$ and $\hat{p}$ in place of $\tilde{q}$ and $\hat{q}$ to calculate the effective means.
-In this case we have $\langle x_2\rangle_{\tilde{p}}\approx L$, the right hand side of Eq.
-\eqref{eq:meanfieldr} is approximately $1/(L+1)$ for the rare type, and, since $l_1\ll 1$ and $L\gg 1$ in Eq.~\eqref{eq:deltr}, we have $\Delta_r n_1 \approx 1/(L+1)$.
-Similarly, for the common type, $\sum_j \langle x_j\rangle_{\hat{p}}\approx L$ in Eq.
+The above pathology arises if we naively use $\tilde{p}$ and $\hat{p}$ to calculate the effective means.
+The right hand side of Eq.
+\eqref{eq:meanfieldr} is then approximately $1/(1+L)$, and since $l_1\ll 1$ and $L\gg 1$ we have $\Delta_r n_1 \approx 1/(1+L)$ in Eq.~\eqref{eq:deltr}.
+Similarly, for the common type, $\sum_j \langle x_j\rangle_{\hat{p}} = L$ in Eq.
 \eqref{eq:meanfielda}, and so $\Delta_a n_2 \approx 1/L$.
-Thus, the identical rare type is erroneously predicted to decline in frequency.
+Thus, the identical rare type is  pathologically predicted to decline in frequency.
 
-The effective distributions $\tilde{q}$ and $\hat{q}$ ensure that the above consistency requirement is satisfied.
+The effective distributions $\tilde{q}$ and $\hat{q}$ are devised to avoid this pathology.
+Let ${\mathbf x}$ denote the vector of propagule abundances $(x_1,\ldots,x_G)$ in a given territory, and ${\mathbf x_i}=(x_1,\ldots,x_{i-1},x_{i+1}\ldots,x_G)$ similarly denote the vector of non-focal abundances, so that $p({\mathbf x_i})=p_1(x_1)\cdots p_{i-1}(x_{i-1})p_{i+1}(x_{i+1})\cdots p_G(x_G)$.
+The corresponding total propagule numbers are denoted $X=\sum_j x_j$ and $X_i=X-x_i$.
+Then, in territories where one focal propagule and at least one non-focal propagule were received, the effective distribution is defined by
+\begin{equation}
+\tilde{q}({\mathbf x_i})=\sum_{X=2}^{\infty}P(X|X\geq 2) p({\mathbf x_i}|X_i=X-1),
+\end{equation}
+where the total number of propagules $X$ follows a Poisson distribution with mean $L$.
+Similarly, in territories where more than one focal propagule was received, the effective distribution is defined by
+\begin{equation}
+\hat{q}({x})=\sum_{X=2}^{\infty}P(X|x_i\geq 2) p({\mathbf x}|x_i\geq 2, X).
+\end{equation}
+In other words, $\tilde{q}$ and $\hat{q}$ make the approximation that the distribution for the total number of propagules per territory is the same in all territories.
+In reality, conditioning on one or more focal propagules being present does affect the distribution of $X$ in the corresponding subset of territories.
+Thus, for $\tilde{p}$ and $\hat{p}$, which are simply the corresponding conditional distributions of the actual dispersal distribution $p$, the mean propagule density is $L$ but $\langle X \rangle_{\tilde{p}}=L+1$.
 
-$\tilde{p}$ assumes that one of the propagules present in a given territory after dispersal belongs to the focal type (a statement about propagule identity), whereas $\tilde{q}$ assumes that there is a focal propagule present before non-focal dispersal commences (a statement about propagule presence).
-
-%Following the notation in the main text, the Poisson distributions for the $x_i$ (or some subset of the $x_i$) will be denoted $p$, and we use $P$ as a general shorthand for the probability of particular outcomes.
-
-By contrast, Eq.
+%By contrast, Eq.
 \eqref{eq:meanxjrare} correctly predicts that there are on average $\sum_{j\neq i}\langle x_j \rangle_{\tilde{p}}\approx L-1$ nonfocal propagules because $\tilde{p}$ accounts for potentially large negative covariances between the $x_j$ ``after dispersal''.
@@
 
-Below we justify this replacement by arguing that the standard deviation $\sigma_{\tilde{p}}(\sum_{j\neq i} c_j x_j)$ (with respect to $\tilde{p}$), is much smaller than $\langle\sum_{j\neq i} c_j x_j\rangle_{\tilde{p}}$.
+%Below we justify this replacement by arguing that the standard deviation $\sigma_{\tilde{p}}(\sum_{j\neq i} c_j x_j)$ (with respect to $\tilde{p}$), is much smaller than $\langle\sum_{j\neq i} c_j x_j\rangle_{\tilde{p}}$.
 If this standard deviation is small as claimed, then the mean value gives an accurate representation of the values taken by $\sum_{j\neq i} c_j x_j$ across territories.
 
+\subsection*{Calculating the effective means}
+
 We first calculate $\langle x_j \rangle_{\tilde{p}}$.
-Let $X=\sum_j x_j$ denote the total number of propagules in a territory and ${\mathbf x_i}=(x_1,\ldots,x_{i-1},x_{i+1}\ldots,x_G)$ denote the vector of non-focal abundances, so that $p({\mathbf x_i})=p_1(x_1)\cdots p_{i-1}(x_{i-1})p_{i+1}(x_{i+1})\cdots p_G(x_G)$.
-Then, $\tilde{p}$ is the probability distribution for ${\mathbf x_i}$ conditional on the territory having received $X\geq 2$ propagules, where $X_i=X-1$ of them are nonfocal, giving
-\begin{equation}
-\tilde{p}({\mathbf x_i})=\sum_{X=2}^{\infty}P(X|X\geq 2) p({\mathbf x_i}|X_i=X-1)
-\end{equation}
-Here the total number of propagules $X$ follows a Poisson distribution with mean $L$, and we have $P(X|X\geq 2)=P(X)/P(X\geq 2)=P(X)/(1-(1+L)e^{-L})$. Thus,
+
+and we have $P(X|X\geq 2)=P(X)/P(X\geq 2)=P(X)/(1-(1+L)e^{-L})$
+ Thus,
 \begin{align}
@@
 where the last line follows from $\sum_{X=2}^{\infty} P(X)(X-1)=\sum_{X=1}^{\infty} P(X)(X-1)=\sum_{X=1}^{\infty} P(X)X-\sum_{X=1}^{\infty}P(X)$.
+
+Substituting Eqs. \eqref{eq:meanfieldr} and \eqref{eq:meanxjrare} into Eq.~\eqref{eq:deltr}, we obtain
+\begin{equation}
+\Delta_r n_i\approx m_i R_i\frac{c_i}{\overline{c}}, \label{eq:deltrfinal}
+\end{equation}
+where $R_i$ is defined in Eq.~\eqref{eq:Dr}.
+
+Following a similar procedure as for $\Delta_r n_i$, where the vector of propagule abundances is denoted ${\mathbf x}$, the mean focal type abundance is, 
+\begin{align}
+\langle x_i \rangle_{\hat{p}}&=\sum_{\mathbf x} x_i p(\mathbf x|x_i\geq 2)\nonumber \\
+&=\sum_{x_i} x_i p(x_i|x_i\geq 2) \nonumber\\
+&=\frac{1}{1-(1+l_i)e^{-l_i}}\sum_{x_i\geq 2} p(x_i)x_i\nonumber\\
+&=l_i\frac{1-e^{-l_i}}{1-(1+l_i)e^{-l_i}}.
+\end{align}
+For nonfocal types $j\neq i$, we have
+\begin{align}
+\langle x_j \rangle_{\hat{p}}&=\sum_{\mathbf x} x_j p(\mathbf x|x_i\geq 2)\nonumber \\
+&=\sum_{X}P(X|x_i\geq 2)\sum_{\mathbf x} x_j p({\mathbf x}|x_i\geq 2,X)\nonumber\\
+&=\sum_{X}P(X|x_i\geq 2)\sum_{x_i} p(x_i|x_i\geq 2,X) \sum_{\mathbf x_i} x_j p(\mathbf x_i|X_i=X-x_i)\nonumber\\
+&=\sum_{X}P(X|x_i\geq 2)\sum_{x_i}p(x_i|x_i\geq 2,X) \frac{l_j(X-x_i)}{L-l_i} \nonumber\\
+&=\frac{l_j}{L-l_i}\left[\sum_{X}P(X|x_i\geq 2)X - \sum_{x_i}p(x_i|x_i\geq 2) x_i \right]\nonumber\\
+&=\frac{l_j}{L-l_i}\left( L\frac{1-e^{-L}}{1-(1+L)e^{-L}}- l_i\frac{1-e^{-l_i}}{1-(1+l_i)e^{-l_i}}\right).
+\end{align}
+In going from line 3 to 4, we used the same logic used to evaluate the inner sum in Eq.~\eqref{eq:raremonster1}, and in going from 4 to 5 we have separately evaluated the contributions from the $X$ and $x_i$ terms in the numerator.
+
+Combining Eqs. \eqref{eq:delta} and \eqref{eq:meanfielda}, we obtain
+\begin{equation}
+\Delta_a n_i=m_i A_i \frac{c_i}{\overline{c}},
+\end{equation}
+where $A_i$ is defined in Eq.~\eqref{eq:Da}.
+
+\subsection*{Limits of the approximation}
 
@@
 
-Substituting Eqs. \eqref{eq:meanfieldr} and \eqref{eq:meanxjrare} into Eq.~\eqref{eq:deltr}, we obtain
-\begin{equation}
-\Delta_r n_i\approx m_i R_i\frac{c_i}{\overline{c}}, \label{eq:deltrfinal}
-\end{equation}
-where $R_i$ is defined in Eq.~\eqref{eq:Dr}.
-
-\subsection*{Competition when abundant}
-
 Again, we argue that the relative fluctuations in $\sum c_j x_j$ are much smaller than $1$ (with respect to $\hat{p}$), so that,
-
-Following a similar procedure as for $\Delta_r n_i$, where the vector of propagule abundances is denoted ${\mathbf x}$, the mean focal type abundance is, 
-\begin{align}
-\langle x_i \rangle_{\hat{p}}&=\sum_{\mathbf x} x_i p(\mathbf x|x_i\geq 2)\nonumber \\
-&=\sum_{x_i} x_i p(x_i|x_i\geq 2) \nonumber\\
-&=\frac{1}{1-(1+l_i)e^{-l_i}}\sum_{x_i\geq 2} p(x_i)x_i\nonumber\\
-&=l_i\frac{1-e^{-l_i}}{1-(1+l_i)e^{-l_i}}.
-\end{align}
-For nonfocal types $j\neq i$, we have
-\begin{align}
-\langle x_j \rangle_{\hat{p}}&=\sum_{\mathbf x} x_j p(\mathbf x|x_i\geq 2)\nonumber \\
-&=\sum_{X}P(X|x_i\geq 2)\sum_{\mathbf x} x_j p({\mathbf x}|x_i\geq 2,X)\nonumber\\
-&=\sum_{X}P(X|x_i\geq 2)\sum_{x_i} p(x_i|x_i\geq 2,X) \sum_{\mathbf x_i} x_j p(\mathbf x_i|X_i=X-x_i)\nonumber\\
-&=\sum_{X}P(X|x_i\geq 2)\sum_{x_i}p(x_i|x_i\geq 2,X) \frac{l_j(X-x_i)}{L-l_i} \nonumber\\
-&=\frac{l_j}{L-l_i}\left[\sum_{X}P(X|x_i\geq 2)X - \sum_{x_i}p(x_i|x_i\geq 2) x_i \right]\nonumber\\
-&=\frac{l_j}{L-l_i}\left( L\frac{1-e^{-L}}{1-(1+L)e^{-L}}- l_i\frac{1-e^{-l_i}}{1-(1+l_i)e^{-l_i}}\right).
-\end{align}
-In going from line 3 to 4, we used the same logic used to evaluate the inner sum in Eq.~\eqref{eq:raremonster1}, and in going from 4 to 5 we have separately evaluated the contributions from the $X$ and $x_i$ terms in the numerator.
 
@@
 Similarly to Eq.~\eqref{eq:cvr}, the RHS of \eqref{eq:cva} will not be $\ll 1$ in the presence of a rare, extremely strong competitor.
-
-Combining Eqs. \eqref{eq:delta} and \eqref{eq:meanfielda}, we obtain
-\begin{equation}
-\Delta_a n_i=m_i A_i \frac{c_i}{\overline{c}},
-\end{equation}
-where $A_i$ is defined in Eq.~\eqref{eq:Da}.
 

### 2018-02-28 manuscript.tex
l

@@
 
-The approximations \eqref{eq:meanfieldr} and \eqref{eq:meanfielda} must be done in a way that ensures consistency between rare and common types.
-To illustrate, suppose that two identical (same $b$, $c$ and $d$) types are present, one with low density $l_1\ll 1$, and the other with high density $l_2\approx L\gg 1$.
-Since $L$ is large, uncontested territories make up a negligible fraction of the total, and the rare type grows almost entirely due to $\Delta_r n_1$, while the common type grows almost entirely due to $\Delta_a n_2$.
+The approximations \eqref{eq:meanfieldr} and \eqref{eq:meanfielda} must be consistent between rare and common types.
+To illustrate, suppose that two identical types (same $b$, $c$ and $d$) are present, with low $l_1\ll 1$ and high density $l_2\approx L\gg 1$ respectively.
+Since $L$ is large, uncontested territories make up a negligible fraction of the total.
+The rare type grows almost entirely due to $\Delta_r n_1$, while the common type grows almost entirely due to $\Delta_a n_2$.
 To ensure consistency, the approximate per-capita growth rates implied by the approximations \eqref{eq:meanfieldr} and \eqref{eq:meanfielda} must be equal $\Delta_r n_1/m_1 = \Delta_a n_2/m_2$.
 Even small violations of this consistency condition would mean exponential growth of one type relative to the other.
-This behavior is pathological: any single-type population can be arbitrarily partitioned into identical rare and common subtypes, but a given small group of individuals would be predicted to either grow or shrink depending on whether it viewed as part of the rare or common type.
+This behavior is clearly pathological, because any single-type population can be arbitrarily partitioned into identical rare and common subtypes.
+Thus,   predicted growth or decline would depend on an arbitrary assignment of rarity.
 
-The above pathology arises if we naively use $\tilde{p}$ and $\hat{p}$ to calculate the effective means.
+For example, suppose that we use $\tilde{p}$ and $\hat{p}$ to calculate the effective means.
 The right hand side of Eq.
-\eqref{eq:meanfieldr} is then approximately $1/(1+L)$, and since $l_1\ll 1$ and $L\gg 1$ we have $\Delta_r n_1 \approx 1/(1+L)$ in Eq.~\eqref{eq:deltr}.
+\eqref{eq:meanfieldr} is then approximately $1/(L+1)$, and since $l_1\ll 1$ and $L\gg 1$ we have $\Delta_r n_1 \approx 1/(L+1)$ in Eq.~\eqref{eq:deltr}.
 Similarly, for the common type, $\sum_j \langle x_j\rangle_{\hat{p}} = L$ in Eq.
@@
 The effective distributions $\tilde{q}$ and $\hat{q}$ are devised to avoid this pathology.
-Let ${\mathbf x}$ denote the vector of propagule abundances $(x_1,\ldots,x_G)$ in a given territory, and ${\mathbf x_i}=(x_1,\ldots,x_{i-1},x_{i+1}\ldots,x_G)$ similarly denote the vector of non-focal abundances, so that $p({\mathbf x_i})=p_1(x_1)\cdots p_{i-1}(x_{i-1})p_{i+1}(x_{i+1})\cdots p_G(x_G)$.
+The idea is to make the approximation that the distribution for the total number of propagules per territory is the same in all territories.
+This is only an approximation because conditioning on focal propagules being present does change the distribution of $X$ in the corresponding subset of territories (in the above example, the mean propagule density across all territories is $L$, but in the territories responsible for the growth of the rare type we have $\langle X \rangle_{\tilde{p}}=L+1$).
+
+More formally, let ${\mathbf x}$ denote the vector of propagule abundances $(x_1,\ldots,x_G)$ in a given territory, and ${\mathbf x_i}=(x_1,\ldots,x_{i-1},x_{i+1}\ldots,x_G)$ similarly denote the vector of non-focal abundances, so that $p({\mathbf x_i})=p_1(x_1)\cdots p_{i-1}(x_{i-1})p_{i+1}(x_{i+1})\cdots p_G(x_G)$.
 The corresponding total propagule numbers are denoted $X=\sum_j x_j$ and $X_i=X-x_i$.
-Then, in territories where one focal propagule and at least one non-focal propagule were received, the effective distribution is defined by
+Then, in territories where one focal propagule and at least one non-focal propagule are present, the effective distribution is defined by
 \begin{equation}
@@
 \end{equation}
-where the total number of propagules $X$ follows a Poisson distribution with mean $L$.
-Similarly, in territories where more than one focal propagule was received, the effective distribution is defined by
+where the total number of propagules $X$ follows a Poisson distribution with mean $L$, and $P(X|X\geq 2)=P(X)/P(X\geq 2)=P(X)/(1-(1+L)e^{-L})$.
+Similarly, in territories where more than one focal propagule is present, the effective distribution is defined by
 \begin{equation}
-\hat{q}({x})=\sum_{X=2}^{\infty}P(X|x_i\geq 2) p({\mathbf x}|x_i\geq 2, X).
+\hat{q}({\mathbf x})=\sum_{X=2}^{\infty}P(X|X\geq 2) p({\mathbf x}|x_i\geq 2, X).
 \end{equation}
-In other words, $\tilde{q}$ and $\hat{q}$ make the approximation that the distribution for the total number of propagules per territory is the same in all territories.
-In reality, conditioning on one or more focal propagules being present does affect the distribution of $X$ in the corresponding subset of territories.
-Thus, for $\tilde{p}$ and $\hat{p}$, which are simply the corresponding conditional distributions of the actual dispersal distribution $p$, the mean propagule density is $L$ but $\langle X \rangle_{\tilde{p}}=L+1$.
-
-%By contrast, Eq.
-\eqref{eq:meanxjrare} correctly predicts that there are on average $\sum_{j\neq i}\langle x_j \rangle_{\tilde{p}}\approx L-1$ nonfocal propagules because $\tilde{p}$ accounts for potentially large negative covariances between the $x_j$ ``after dispersal''.
-By neglecting the latter covariances, $\tilde{q}$ overestimates the fluctuations in $\sum_{j\neq i} c_j x_j$; thus $\tilde{q}$ gives an upper bound on the fluctuations.
-The discrepancy between $\tilde{q}$ and $\tilde{p}$ will be largest when $L$ is of order $1$ or smaller, because then the propagule assumed to already be present under $\tilde{q}$ is comparable to, or greater than, the entire propagule density.
-
-%Below we justify this replacement by arguing that the standard deviation $\sigma_{\tilde{p}}(\sum_{j\neq i} c_j x_j)$ (with respect to $\tilde{p}$), is much smaller than $\langle\sum_{j\neq i} c_j x_j\rangle_{\tilde{p}}$.
-If this standard deviation is small as claimed, then the mean value gives an accurate representation of the values taken by $\sum_{j\neq i} c_j x_j$ across territories.
 
@@
 
-We first calculate $\langle x_j \rangle_{\tilde{p}}$.
-
-and we have $P(X|X\geq 2)=P(X)/P(X\geq 2)=P(X)/(1-(1+L)e^{-L})$
- Thus,
+Here we calculate the effective means, starting with the $\Delta_r n_i$ component. We have
 \begin{align}
-\langle x_j \rangle_{\tilde{p}}&=\sum_{\mathbf x_i} \tilde{p}({\mathbf x_i})x_j\nonumber\\
+\langle x_j \rangle_{\tilde{q}}&=\sum_{\mathbf x_i} \tilde{q}({\mathbf x_i})x_j\nonumber\\
 &=\frac{1}{1-(1+L)e^{-L}}\sum_{X=2}^{\infty} P(X) \sum_{\mathbf x_i} p({\mathbf x_i}|X_i=X-1)x_j.
@@
 \begin{align}
-\langle x_j \rangle_{\tilde{p}}&=\frac{l_j}{1-(1+L)e^{-L}}\frac{1}{L-l_i}\sum_{X=2}^{\infty} P(X) (X-1)\nonumber\\
+\langle x_j \rangle_{\tilde{q}}&=\frac{l_j}{1-(1+L)e^{-L}}\frac{1}{L-l_i}\sum_{X=2}^{\infty} P(X) (X-1)\nonumber\\
 &=\frac{l_j}{1-(1+L)e^{-L}}\frac{L-1+e^{-L}}{L-l_i},
@@
 where the last line follows from $\sum_{X=2}^{\infty} P(X)(X-1)=\sum_{X=1}^{\infty} P(X)(X-1)=\sum_{X=1}^{\infty} P(X)X-\sum_{X=1}^{\infty}P(X)$.
-
-Substituting Eqs. \eqref{eq:meanfieldr} and \eqref{eq:meanxjrare} into Eq.~\eqref{eq:deltr}, we obtain
+Substituting Eqs.~\eqref{eq:meanfieldr} and \eqref{eq:meanxjrare} into Eq.~\eqref{eq:deltr}, we obtain
 \begin{equation}
@@
 
-Following a similar procedure as for $\Delta_r n_i$, where the vector of propagule abundances is denoted ${\mathbf x}$, the mean focal type abundance is, 
+Turning now to the $\Delta_a n_i$ component, the mean focal abundance is 
 \begin{align}
-\langle x_i \rangle_{\hat{p}}&=\sum_{\mathbf x} x_i p(\mathbf x|x_i\geq 2)\nonumber \\
-&=\sum_{x_i} x_i p(x_i|x_i\geq 2) \nonumber\\
+\langle x_i \rangle_{\hat{q}}&=\sum_{\mathbf x} \hat{q}({\mathbf x}) x_i\nonumber \\
+&=\sum_{x_i} p(x_i|x_i\geq 2)x_i \nonumber\\
 &=\frac{1}{1-(1+l_i)e^{-l_i}}\sum_{x_i\geq 2} p(x_i)x_i\nonumber\\
@@
 \begin{align}
-\langle x_j \rangle_{\hat{p}}&=\sum_{\mathbf x} x_j p(\mathbf x|x_i\geq 2)\nonumber \\
-&=\sum_{X}P(X|x_i\geq 2)\sum_{\mathbf x} x_j p({\mathbf x}|x_i\geq 2,X)\nonumber\\
-&=\sum_{X}P(X|x_i\geq 2)\sum_{x_i} p(x_i|x_i\geq 2,X) \sum_{\mathbf x_i} x_j p(\mathbf x_i|X_i=X-x_i)\nonumber\\
-&=\sum_{X}P(X|x_i\geq 2)\sum_{x_i}p(x_i|x_i\geq 2,X) \frac{l_j(X-x_i)}{L-l_i} \nonumber\\
-&=\frac{l_j}{L-l_i}\left[\sum_{X}P(X|x_i\geq 2)X - \sum_{x_i}p(x_i|x_i\geq 2) x_i \right]\nonumber\\
+\langle x_j \rangle_{\hat{q}}&=\sum_{X=2}^{\infty}P(X|X\geq 2)\sum_{\mathbf x}  p({\mathbf x}|x_i\geq 2,X) x_j\nonumber\\
+&=\sum_{X=2}^{\infty}P(X|X\geq 2)\sum_{x_i} p(x_i|x_i\geq 2,X) \sum_{\mathbf x_i}  p(\mathbf x_i|X_i=X-x_i) x_j\nonumber\\
+&=\sum_{X=2}^{\infty}P(X|X\geq 2)\sum_{x_i}p(x_i|x_i\geq 2,X) \frac{l_j(X-x_i)}{L-l_i} \nonumber\\
+&=\frac{l_j}{L-l_i}\left[\sum_{X=2}^{\infty}P(X|X\geq 2)X - \sum_{x_i}p(x_i|x_i\geq 2) x_i \right]\nonumber\\
 &=\frac{l_j}{L-l_i}\left( L\frac{1-e^{-L}}{1-(1+L)e^{-L}}- l_i\frac{1-e^{-l_i}}{1-(1+l_i)e^{-l_i}}\right).
 \end{align}
-In going from line 3 to 4, we used the same logic used to evaluate the inner sum in Eq.~\eqref{eq:raremonster1}, and in going from 4 to 5 we have separately evaluated the contributions from the $X$ and $x_i$ terms in the numerator.
-
-Combining Eqs. \eqref{eq:delta} and \eqref{eq:meanfielda}, we obtain
+In going from line 2 to 3, we used the same logic used to evaluate the inner sum in Eq.~\eqref{eq:raremonster1}, and in going from 3 to 4 we have separately evaluated the contributions from the $X$ and $x_i$ terms in the numerator.
+Combining these results with Eqs.~\eqref{eq:delta} and \eqref{eq:meanfielda}, we obtain
 \begin{equation}
@@
 \subsection*{Limits of the approximation}
+
+The validity of the approximations in Eq.~\eqref{eq:meanfieldr} and \eqref{eq:meanfielda}
+
+If this standard deviation is small as claimed, then the mean value gives an accurate representation of the values taken by $\sum_{j\neq i} c_j x_j$ across territories.
+
+are relies on the standard deviation $\sigma_{\tilde{q}}(\sum_{j\neq i} c_j x_j)$ (with respect to $\tilde{q}$) being much smaller than the corresponding mean $\langle\sum_{j\neq i} c_j x_j\rangle_{\tilde{p}}$.
 

### 2018-03-03 manuscript.tex
l

@@
 
-\subsection*{Limits of the approximation}
+\subsection*{Approximation limits}
 
-The validity of the approximations in Eq.~\eqref{eq:meanfieldr} and \eqref{eq:meanfielda}
+Eq.~\eqref{eq:meanfieldr} and \eqref{eq:meanfielda} must not only be consistent with each other, they must also be individually good approximations. Here we evaluate these approximations.
 
-If this standard deviation is small as claimed, then the mean value gives an accurate representation of the values taken by $\sum_{j\neq i} c_j x_j$ across territories.
+The fundamental requirement for making the replacement in Eqs.~\eqref{eq:meanfieldr} and \eqref{eq:meanfielda} is that we can ignore the fluctuations in the $x_i$ and hence replace them with a constant effective mean value.
+Mathematically, we require that the standard deviations $\sigma_{\tilde{q}}(\sum_{j\neq i} c_j x_j)$ and $\sigma_{\hat{q}}(\sum_j c_j x_j)$ must be sufficiently small compared to the corresponding means $\langle\sum_{j\neq i} c_j x_j\rangle_{\tilde{q}}$ and $\langle\sum_j c_j x_j\rangle_{\hat{q}}$ in Eqs.~\eqref{eq:meanfieldr} and \eqref{eq:meanfielda} respectively.
 
-are relies on the standard deviation $\sigma_{\tilde{q}}(\sum_{j\neq i} c_j x_j)$ (with respect to $\tilde{q}$) being much smaller than the corresponding mean $\langle\sum_{j\neq i} c_j x_j\rangle_{\tilde{p}}$.
+To evaluate these standard deviations, we will work with $\tilde{p}$ and $\hat{p}$ distributions instead of $\tilde{q}$ and $\hat{q}$.
+This is mathematically much simpler because the $x_i$ are independent under $\tilde{p}$ and $\hat{p}$, and is justified by the fact that $\tilde{p}$ and $\hat{p}$ are closely related to $\tilde{q}$ and $\hat{q}$ respectively, and so we expect the relevant means and standard deviations will be similar.
 
-Having evaluated the mean propagule numbers, we now evaluate the variance in propagule numbers to check that the mean value replacement in Eq.~\eqref{eq:meanfieldr} is justified.
-This is complicated because the $x_j$ are not independent with respect to $\tilde{p}$.
-Here we use the following approximation to give some insight into the magnitude of these fluctuations and also the nature of the correlations between the $x_j$.
-We replace $\tilde{p}$ with $\tilde{q}$, defined as the ${\mathbf x_i}$ Poisson dispersal probabilities conditional on $X_i\geq1$ (which are independent).
-The distinction between $\tilde{p}$ and $\tilde{q}$ will be discussed further below.
-The $\tilde{q}$ approximation gives $\langle x_j \rangle_{\tilde{q}}=\langle x_j \rangle_p/C=l_j/C$, where $C=1-e^{-(L-l_i)}$, with variances and covariances given by
+\begin{figure}
+\centering
+\includegraphics[scale=0.8]{approx_details.pdf}
+\caption{\label{fig:approx_details} Horse.}
+\end{figure}
+
+Starting with Eq.~\eqref{eq:meanfieldr}, we have $\langle x_j \rangle_{\tilde{p}}=l_j/C$, where $C=1-e^{-(L-l_i)}$, and the corresponding variance is given by
 \begin{align}
-\sigma_{\tilde{q}}^2(x_j)&=\langle x_j^2 \rangle_{\tilde{q}}-\langle x_j \rangle_{\tilde{q}}^2\nonumber\\
-&=\frac{1}{C}\langle x_j^2 \rangle_p-\frac{l_j^2}{C^2}\nonumber \\
-&=\frac{1}{C}(l_j^2 + l_j)-\frac{l_j^2}{C^2}\nonumber \\
-&=\frac{l_j^2}{C}\left(1-\frac{1}{C}\right)+\frac{l_j}{C},\label{eq:varr}
+\sigma_{\tilde{p}}^2(x_j)&=\langle x_j^2 \rangle_{\tilde{p}}-\langle x_j \rangle_{\tilde{p}}^2\nonumber\\
+&=\frac{l_j^2 + l_j}{C}-\frac{l_j^2}{C^2}\nonumber \\
+&=\left(1-\frac{1}{C}\right)\frac{l_j^2}{C}+\frac{l_j}{C},\label{eq:varr}
 \end{align}
-and 
-\begin{align}
-\sigma_{\tilde{q}}(x_j,x_k)&=\langle x_j x_k \rangle_{\tilde{q}}-\langle x_j \rangle_{\tilde{q}}\langle x_k \rangle_{\tilde{q}}\nonumber\\
-&=\frac{1}{C}\langle x_j x_k \rangle_p-\frac{l_jl_k}{C^2}\nonumber\\
-&=\frac{l_j l_k}{C}\left(1-\frac{1}{C}\right)\qquad\qquad j\neq k. \label{eq:covr}
-\end{align}
-
-Decomposing the variance in $\sum_{j\neq i} c_j x_j$,
+Since $\sigma^2_{\tilde{p}}(\sum_{j\neq i} c_j x_j)=\sum_{j\neq i} c_j^2\sigma^2_{\tilde{p}}(x_j)$, and the first term in Eq.
+\eqref{eq:varr} is negative (because $C<1$), we obtain an upper bound on the relative fluctuations in $\sum_{j\neq i} c_j x_j$,
 \begin{equation}
-\sigma_{\tilde{q}}^2(\sum_{j\neq i} c_j x_j)=\sum_{j\neq i}\left[c_j^2\sigma_{\tilde{q}}^2(x_j)+2\sum_{k>j, k\neq i}c_j c_k\sigma_{\tilde{q}}(x_j,x_k)\right],\label{eq:vartotr}
-\end{equation}
-and using the fact that $\sigma_{\tilde{q}}(x_j,x_k)$ and the first term in Eq.
-\eqref{eq:varr} are negative because $C<1$, we obtain an upper bound on the relative fluctuations in $\sum_{j\neq i} c_j x_j$,
-\begin{equation}
-\frac{\sigma(\sum_{j\neq i} c_j x_j)}{\langle\sum_{j\neq i} c_j x_j\rangle}=C^{1/2}\frac{\left(\sum_{j\neq i}c_j^2 l_j+(1-1/C)\left(\sum_{j\neq i}c_j l_j\right)^2 \right)^{1/2}}{\sum_{j\neq i}c_j l_j}<C^{1/2}\frac{\left(\sum_{j\neq i}c_j^2 l_j\right)^{1/2}}{\sum_{j\neq i}c_j l_j}.
+\frac{\sigma(\sum_{j\neq i} c_j x_j)}{\langle\sum_{j\neq i} c_j x_j\rangle}=C^{1/2}\frac{\left(\sum_{j\neq i}c_j^2 ((1-\frac{1}{C})l_j^2+l_j)\right)^{1/2}}{\sum_{j\neq i}c_j l_j}<C^{1/2}\frac{\left(\sum_{j\neq i}c_j^2 l_j\right)^{1/2}}{\sum_{j\neq i}c_j l_j}.
 \label{eq:cvr}
@@
 
-Suppose that the $c_j$ are all of similar magnitude (their ratios are of order one).
-Then Eq.~\eqref{eq:cvr} is $\ll 1$ for the case when $L-l_i \ll 1$ (due to the factor of $C^{1/2}$), and also for the case when at least some of the nonfocal propagule densities are large $l_j\gg 1$ (since it is then of order $1/\sqrt{L-l_i}$).
+Suppose that the $c_j$ have similar magnitudes (their ratios are of order one).
+Then the right hand side of Eq.~\eqref{eq:cvr} is approximately equal to $\sqrt{\frac{1-e^{-(L-l_i)}}{L-l_i}}$.
+This quantity is small for both low and high nonfocal densities.
 The worst case scenario occurs when $L-l_i$ is of order one.
-Then Eq.~\eqref{eq:cvr} gives a relative error of approximately $50\%$, which from our earlier discussion we know to be a substantial overestimate when $L$ is of order $1$.
-Our numerical results (Fig.
-\ref{fig:simcomp}) confirm that the relative errors are indeed small.
+Then the  Eq.~\eqref{eq:cvr} gives a relative error of approximately $50\%$
+
+Our numerical results (Fig. \ref{fig:simcomp}) confirm that the relative errors are indeed small.
 

### 2018-03-03 approximation_details.py

@@ -1,0 +1,126 @@
+# -*- coding: utf-8 -*-
+"""
+Created on Mon Feb 19 14:37:21 2018
+
+@author: jbertram
+"""
+
+import numpy as np
+import bisect
+import matplotlib.pyplot as plt
+
+def deltnplusnaive(m,c,U):
+    if sum(m)>0 and U>0:
+        L=sum(m)/float(U)
+        l=m/float(U)
+        cbar=sum(m*c)/sum(m)
+        dr=m*np.exp(-l)*(1-np.exp(-(L-l)))*c/((cbar*L-c*l)/(1-np.exp(-(L-l)))+c)
+        da=m*c*(1-np.exp(-l))/(cbar*L+c*l**2*np.exp(-l)/(1-(1+l)*np.exp(-l)))
+        return m*np.exp(-L) + dr + da
+    else:
+        return np.zeros(len(m))
+
+def wins(c):
+    winsnocomp=np.zeros(len(c)); wins1=np.zeros(len(c)); wins2=np.zeros(len(c));
+    comp=np.zeros(T);
+    for i in range(int(T)):
+        comp[i]=sum(scatter[i])
+        if comp[i]>0:            
+            lotterycmf=np.cumsum(np.array(scatter[i])*c)
+            victor=bisect.bisect(lotterycmf,np.random.rand()*lotterycmf[-1])
+            
+            if scatter[i][victor]==1 and comp[i]==1:
+                winsnocomp[victor]=winsnocomp[victor]+1
+            elif scatter[i][victor]==1:
+                wins1[victor]=wins1[victor]+1
+            else: 
+                wins2[victor]=wins2[victor]+1
+                
+    return winsnocomp+wins1+wins2
+            
+
+
+T=1000000
+cs=np.array([[10.**i,1.] for i in np.linspace(-2,2,101)])
+
+fig1, (ax1, ax2) = plt.subplots(nrows=2,figsize=[5,8])
+
+l1=3.; l2=3.; L=l1+l2
+scatter=np.random.poisson(lam=[l1,l2],size=[T,2])
+m=np.array([l1*T,l2*T])
+
+#deltaovermsim1=np.array(map(wins,cs))/m
[…]

### 2018-03-03 manuscript.tex
l

@@
 
-\begin{figure}
-\centering
-\includegraphics[scale=0.8]{approx_details.pdf}
-\caption{\label{fig:approx_details} Horse.}
-\end{figure}
-
-Starting with Eq.~\eqref{eq:meanfieldr}, we have $\langle x_j \rangle_{\tilde{p}}=l_j/C$, where $C=1-e^{-(L-l_i)}$, and the corresponding variance is given by
+Starting with Eq.~\eqref{eq:meanfieldr}, we have $\langle x_j \rangle_{\tilde{p}}=l_j/C$, where $C=1-e^{-(L-l_i)}$, and the corresponding variances and covariances are given by
 \begin{align}
@@
 \end{align}
-Since $\sigma^2_{\tilde{p}}(\sum_{j\neq i} c_j x_j)=\sum_{j\neq i} c_j^2\sigma^2_{\tilde{p}}(x_j)$, and the first term in Eq.
-\eqref{eq:varr} is negative (because $C<1$), we obtain an upper bound on the relative fluctuations in $\sum_{j\neq i} c_j x_j$,
+and
+\begin{align}
+\sigma_{\tilde{p}}(x_j,x_k)&=\langle x_j x_k \rangle_{\tilde{p}}-\langle x_j \rangle_{\tilde{p}}\langle x_k \rangle_{\tilde{p}}\nonumber\\
+&=\frac{1}{C}\langle x_j x_k \rangle_p-\frac{l_jl_k}{C^2}\nonumber\\
+&=\left(1-\frac{1}{C}\right)\frac{l_j l_k}{C}\qquad\qquad j\neq k. \label{eq:covr}
+\end{align} 
+Note that $1-1/C$ is negative because $C<1$. Decomposing the variance in $\sum_{j\neq i} c_j x_j$,
 \begin{equation}
-\frac{\sigma(\sum_{j\neq i} c_j x_j)}{\langle\sum_{j\neq i} c_j x_j\rangle}=C^{1/2}\frac{\left(\sum_{j\neq i}c_j^2 ((1-\frac{1}{C})l_j^2+l_j)\right)^{1/2}}{\sum_{j\neq i}c_j l_j}<C^{1/2}\frac{\left(\sum_{j\neq i}c_j^2 l_j\right)^{1/2}}{\sum_{j\neq i}c_j l_j}.
+\sigma_{\tilde{p}}^2(\sum_{j\neq i} c_j x_j)=\sum_{j\neq i}\left[c_j^2\sigma_{\tilde{p}}^2(x_j)+2\sum_{k>j, k\neq i}c_j c_k\sigma_{\tilde{p}}(x_j,x_k)\right],\label{eq:vartotr}
+\end{equation}
+we obtain 
+\begin{equation}
+\frac{\sigma(\sum_{j\neq i} c_j x_j)}{\langle\sum_{j\neq i} c_j x_j\rangle}=C^{1/2}\frac{\left(\sum_{j\neq i}c_j^2 l_j+(1-\frac{1}{C})\left(\sum_{j\neq i}c_j l_j\right)^2 \right)^{1/2}}{\sum_{j\neq i}c_j l_j}.
 \label{eq:cvr}
@@
 
-Suppose that the $c_j$ have similar magnitudes (their ratios are of order one).
-Then the right hand side of Eq.~\eqref{eq:cvr} is approximately equal to $\sqrt{\frac{1-e^{-(L-l_i)}}{L-l_i}}$.
-This quantity is small for both low and high nonfocal densities.
-The worst case scenario occurs when $L-l_i$ is of order one.
-Then the  Eq.~\eqref{eq:cvr} gives a relative error of approximately $50\%$
+Eq.~\eqref{eq:cvr} reveals two key points.
+First, when the $c_j$ have similar magnitudes (their ratios are of order one), Eq.~\eqref{eq:meanfieldr} is an excellent approximation.
+In this case, the right hand side of Eq.~\eqref{eq:cvr} is approximately equal to $C^{1/2}\left(\frac{1}{L-l_i}+1-\frac{1}{C}\right)^{1/2}$, which is small for both low and high nonfocal densities.
+The worst case scenario occurs when $L-l_i$ is of order one, and it can be directly verified that Eq.~\eqref{eq:meanfieldr} is then still a good approximation (see Fig.~\ref{fig:approx_details}).
+Second, if some of the $c_j$ are much larger than the others, the relative fluctuations in $\sum_{j\neq i} c_j x_j$ can be large.
+Specifically, in the presence of a rare,  strong competitor ($c_j l_j\gg c_{j'} l_{j'}$ for all other nonfocal types $j'$, and $l_j\ll 1$), then the right hand side of Eq.
+\eqref{eq:cvr} can be large and we cannot make the replacement Eq.~\eqref{eq:meanfieldr}.
+Fig.~\ref{fig:approx_details} shows the breakdown of the effective mean approximation when the are large differences in $c$.
 
-Our numerical results (Fig. \ref{fig:simcomp}) confirm that the relative errors are indeed small.
+\begin{figure}
+\centering
+\includegraphics[scale=0.8]{approx_details.pdf}
+\caption{\label{fig:approx_details} Comparison of our approximation with simulations and the naive $\tilde{p}$,$\hat{p}$ approxmation as a function of the relative $c$-advantage between two types.
+Our approximation breaks down in the presence of large relative $c$-advantages.
+The inset shows the pathology of the $\tilde{p}$,$\hat{p}$ approxmation --- in the neutral case of $c=1$ one type grows faster than the other.
+}
+\end{figure}
 
-However, the relative fluctuations in $\sum_{j\neq i} c_j x_j$ can be large if some of the $c_j$ are much larger than the others.
-Specifically, in the presence of a rare, extremely strong competitor ($c_j l_j\gg c_{j'} l_{j'}$ for all other nonfocal types $j'$, and $l_j\ll 1$), then the RHS of Eq.
-\eqref{eq:cvr} can be large and we cannot make the replacement Eq.~\eqref{eq:meanfieldr}.
-
-Again, we argue that the relative fluctuations in $\sum c_j x_j$ are much smaller than $1$ (with respect to $\hat{p}$), so that,
-
-To calculate the standard deviation in $\sum_{j\neq i} c_j x_j$, we use a similar approximation as for $\Delta_r n_i$: $\hat{p}$ is approximated by $\hat{q}$, defined as the ${\mathbf x}$ dispersal probabilities in a territory conditional on $x_i>2$ (that is, treating the $x_j$ as independent).
-All covariances between nonfocal types are now zero, so that $\sigma_{\hat{q}}^2(\sum c_j x_j)=\sum c_j^2 \sigma_{\hat{q}}^2(x_j)$, where $\sigma_{\hat{q}}^2(x_j)=l_j$ for $j\neq i$, and
+Turning now to Eq.~\eqref{eq:meanfielda}, all covariances between nonfocal types are now zero, so that $\sigma_{\hat{p}}^2(\sum c_j x_j)=\sum c_j^2 \sigma_{\hat{p}}^2(x_j)$, where $\sigma_{\hat{p}}^2(x_j)=l_j$ for $j\neq i$.
+Here
 \begin{equation}
-\sigma_{\hat{q}}^2(x_i)=\frac{l_i}{D}\left(l_i+1-e^{-l_i}-\frac{l_i}{D}\left(1-e^{-l_i}\right)^2\right),
+\sigma_{\hat{p}}^2(x_i)=\frac{l_i}{D}\left(l_i+1-e^{-l_i}-\frac{l_i}{D}\left(1-e^{-l_i}\right)^2\right),
 \end{equation}
@@
 \begin{equation}
-\frac{\sigma_{\hat{q}}(\sum c_j x_j)}{\langle\sum c_j x_j\rangle} = \frac{\left(\sum_{j\neq i} c_j^2 l_j + c_i^2 \sigma_{\hat{q}}^2(x_i)\right)^{1/2}}{\sum_{j\neq i} c_j l_j + c_i l_i (1-e^{-l_i})/D} \label{eq:cva}.
+\frac{\sigma_{\hat{p}}(\sum c_j x_j)}{\langle\sum c_j x_j\rangle} = \frac{\left(\sum_{j\neq i} c_j^2 l_j + c_i^2 \sigma_{\hat{p}}^2(x_i)\right)^{1/2}}{\sum_{j\neq i} c_j l_j + c_i l_i (1-e^{-l_i})/D} \label{eq:cva}.
 \end{equation}
 
-Similarly to Eq.~\eqref{eq:cvr}, the RHS of Eq.
-\eqref{eq:cva} is $\ll 1$ for the case that $L \ll 1$ (due to a factor of $D^{1/2}$), and also for the case when at least some of the propagule densities (focal or nonfocal) are large --- provided that $c_i$ and the $c_j$ are all of similar magnitude.
-Again, the worst case scenario occurs when $l_i$ and $L-l_i$ are of order $1$, in which case Eq.
-\eqref{eq:cva} is around $35\%$, which is again where the $\hat{q}$ approximation produces the biggest overestimate of the fluctuations in ${\mathbf x}$.
-Similarly to Eq.~\eqref{eq:cvr}, the RHS of \eqref{eq:cva} will not be $\ll 1$ in the presence of a rare, extremely strong competitor.
+Similarly to Eq.~\eqref{eq:cvr}, the right hand side of Eq.
+\eqref{eq:cva} is small for both low and high nonfocal densities.
+Again, the worst case scenario occurs when $l_i$ and $L-l_i$ are of order $1$, but Eq.~\eqref{eq:meanfielda} is still a good approximationin this case.
+Again, the approxmation breaks down in the presence of a rare, strong competitor (Fig.~\ref{fig:approx_details}).
 

### 2018-03-03 approximation_details.py

@@ -40,12 +40,12 @@ def wins(c):
             
 
 
-T=1000000
+T=10000
 cs=np.array([[10.**i,1.] for i in np.linspace(-2,2,101)])
 
 fig1, (ax1, ax2) = plt.subplots(nrows=2,figsize=[5,8])
 
-l1=3.; l2=3.; L=l1+l2
+l1=.1; l2=1.; L=l1+l2
 scatter=np.random.poisson(lam=[l1,l2],size=[T,2])
 m=np.array([l1*T,l2*T])
 

### 2018-03-06 manuscript.tex
l

@@
 There are a variety of different measures of fitness.
-Some widely used examples in evolutionary ecology are expected lifetime reproductive ratio $R_0$, intrinsic population growth rate $r$, saturation population density (often labeled ``$K$'') \citep{benton_2000}, and invasion fitness \citep{metz_1992}.
+Some widely used examples in evolutionary ecology are expected lifetime reproductive ratio $R_0$, intrinsic population growth rate $r$, saturation population density/carrying capacity (often labeled ``$K$'') \citep{benton_2000}, and invasion fitness \citep{metz_1992}.
 In addition, ``relative fitness'' is the standard in much of evolutionary biology, particularly evolutionary genetics, where attention is generally restricted to relative genotypic proportions \cite[pp.
@@
 But it should be clear how the measure being used is connected to the processes of birth and death which govern population biology \citep{metcalf_2007,doebeli_2017}.
-While such a connection is fairly clear for absolute fitness measures like $r$, relative fitness seems largely divorced from population biology.
+While such a connection is fairly clear for absolute fitness measures like $r$, relative fitness seems largely divorced from population ecology.
 It has even been proposed that relative fitness be justified from measure theory, abandoning population biology altogether \citep{wagner_2010}.
@@
 
-Constant relative fitness values can be justified as a linear approximation \cite[pp.
+Constant relative fitness values can be justified as an approximation which holds when selection is sufficiently weak and stable over time.
+Formally, these values are the ``zeroth order'' components of the actual relative fitnesses, which are not constant but depend on density, frequency, age structure, and so on \cite[pp.
 277]{ewens_2004} \citep[Chap.
-4]{charlesworth_1994} which is close to exact provided that selection is sufficiently weak and stable over time.
-Yet strong, temporally-variable selection occurs widely in nature and the lab, including in wild \textit{Drosophila}, where population density varies by orders of magnitude each seasonal cycle \citep{messer_2016,bergland_14}.
+4]{charlesworth_1994}.
+Yet strong, temporally-variable selection occurs widely in nature and the lab, including in wild \textit{Drosophila}, where population density also varies by orders of magnitude each seasonal cycle \citep{messer_2016,bergland_14}.
 The question is whether relative fitness can be used when selection is not vanishingly weak.
@@
 
-The basis of relative fitness is straightforward in the absence of crowding: it simply represents differences in intrinsic population growth rate.
+In the absence of crowding, relative fitness simply represents differences in intrinsic population growth rate.
 In discrete time, the change in frequency of type $i$ is $\Delta p_i=\left(\frac{W_i}{\overline{W}}-1\right) p_i$, where $W_i$ is the intrinsic absolute growth factor of type $i$, and $\overline{W}=\sum_i W_i p_i$ is the population mean $W$.
-Here we can rescale $W$ however we please and replace it with ``relative fitness'' $w$ without affecting the ratio $\frac{W_i}{\overline{W}}=\frac{w_i}{\overline{w}}$.
-In continuous time, the canonical selection equation is $\frac{d p_i}{dt}=(r_i-\overline{r}) p_i$, where $W$ is replaced by the intrinsic exponential growth rate $r$ \citep[pp.
-26]{crow_1970}.
-If there are two types present, a wildtype and a mutant for instance, then the continuous time canonical selection equation can be written as
+The change in frequency depends only the relative ratio $\frac{W_i}{\overline{W}}$, not the absolute values of the growth factors $W_i$.
+Thus, we can replace the $W_i$ with any set of values $w_i$ that are proportional to the $W_i$; these values are called ``relative fitnesses''.
+The analogous formula in continuous time is $\frac{d p_i}{dt}=(r_i-\overline{r}) p_i$, where $W_i$ is replaced by the intrinsic exponential growth rate $r_i$ \citep[pp.
+26]{crow_1970}, and the ratio $\frac{W_i}{\overline{W}}$ is replaced by the difference $r_i-\overline{r}$.
+In the particular case that there are two types present, a wildtype $i$ and a mutant $j$ for instance, then the continuous time equation takes the familiar form
 \begin{equation}
@@
 \end{equation}
-where the constant selection coefficient $s$ is the difference in $r$ between types. The corresponding adaptive sweeps follow a logistic curve.
+where $s=r_i-r_j$ is the selection coefficient.
 
-The difficulty with Eq.~\eqref{eq:canonical} arises in crowded populations.
+The simple interpretation of relative fitness as a difference in intrinsic growth rates breaks down when we allow for population crowding.
 Since crowded and uncrowded conditions are so different, we expect that $s$ will often depend on density \citep{travis_2013}.
@@
 there is no avoidance of territories crowded with propagules).
+To model the dispersal process, we assume that the number of propagules of the $i$'th type landing in any particular territory, denoted $x_i$, follows a Poisson distribution $p_i(x_i)=l_i^{x_i} e^{-l_i}/x_i!$.
+Here $l_i=m_i/U$ is the mean territorial propagule density, where $m_i$ is the mean number of propagules produced by type $i$ (the total propagule density will be denoted by $L=\sum_i l_i$ and the mean total number of propagules by $M=\sum_i m_i$).
 We assume that adults cannot be ousted by juveniles, so that propagules landing on occupied territories are doomed.
-We assume that each adult from type $i$ produces a constant number $b_i$ of successfully dispersing propagules; the number of propagules dispersing to unoccupied territories is then given by $m_i=b_in_iU/T$.
-The total number of these propagules will be denoted $M=\sum_i m_i$.
+Thus, the number of propagules dispersing to unoccupied territories can be written as $m_i=b_in_iU/T$, where $b_i$ is the number of propagules produced by each adult from type $i$.
 Note that due to our assumption of uniform dispersal, the parameter $b_i$ can be thought of as a measure of ``colonization ability'', which combines fecundity and dispersal ability \citep{levins_71,tilman_94,bolker_99}.
-
-The number of propagules of the $i$'th type landing in any particular territory is denoted $x_i$.
-We assume that $x_i$ follows a Poisson distribution $p_i(x_i)=l_i^{x_i} e^{-l_i}/x_i!$, where $l_i=m_i/U$ is the mean territorial propagule density for type $i$ (the total propagule density will be denoted $L=\sum_i l_i$).
-This is strictly only an approximation of random dispersal, but it is an excellent approximation given our assumption that the $n_i$ are large enough that drift can be ignored (Appendix A).
 
@@
 \ref{fig:lottery}).
-We expect that a fraction $p_1(x_1)\ldots p_G(x_G)$ of the $U$ unoccupied territories will have the propagule composition $x_1,\ldots,x_G$.
+We expect that a fraction $p_1(x_1)\cdots p_G(x_G)$ of the $U$ unoccupied territories will have the propagule composition $x_1,\ldots,x_G$.
 Type $i$ is expected to win a proportion $c_i x_i/\sum_j c_j x_j$ of these.
@@
 \eqref{eq:master}.
-Dashed lines in show the breakdown of the classic lottery model.
+Dashed lines show the breakdown of the classic lottery model.
 Parameters: $m_1=10^4$ and $m_2=9\times 10^4$ and $U$ varies between $5\times 10^3$ and $10^6$.}
@@
 \end{equation}
-where $\delta_i$ is the per-capita increase in mortality rate due to crowding (for simplicity, there are no deaths when uncrowded), playing a similar role as $K$ in the logistic model.
+where $\delta_i$ is the per-capita increase in mortality rate due to crowding (for simplicity, there are no deaths when uncrowded), playing a similar role as $1/K$ in the logistic model.
 
-Starting from a monomorphic population, the frequency of a $\delta_i\rightarrow \delta_i(1-\epsilon)$ variant obeys 
+Starting from a monomorphic population, the frequency of a variant with $\delta_j=\delta_i(1-\epsilon)$ obeys 
 \begin{equation}
-\frac{d p_i}{dt}=\epsilon \delta_i N p_i(1-p_i). \label{eq:Ndependentsweep}
+\frac{d p_j}{dt}=\epsilon \delta_i N p_j(1-p_j). \label{eq:Ndependentsweep}
 \end{equation}
 The selection coefficient $s=\epsilon \delta_i N$ thus depends on density (compare with Model III in \cite{kimura1969natural}).
-On the other hand, the frequency of a $b_i\rightarrow b_i(1+\epsilon)$ variant will exactly obey Eq.~\eqref{eq:canonical} with $s=\epsilon b_i$, independent of density.
+On the other hand, the frequency of a variant with $b_j= b_i(1+\epsilon)$ will exactly obey Eq.~\eqref{eq:canonical} with $s=\epsilon b_i$, independent of density.
 
@@
 However, even if $N$ has reached equilibrium, it will change substantially over a $\delta$-sweep if selection on $\delta$ is sufficiently strong.
-To quantify this effect, we need to account for how much $N$ changes as a result of a $\delta$-sweep beginning and ending in equilibrium \citep{kimura1969natural}; from Eq.~\eqref{eq:simplebirthdeath} we have an increase from $N_{\rm initial}=b_i/\delta_i$ to $N_{\rm final}=b_i/\delta_i(1-\epsilon)=N_{\rm initial}/(1-\epsilon)$.
+To quantify this effect, we need to account for how much $N$ changes as a result of a $\delta$-sweep beginning and ending in equilibrium \citep{kimura1969natural}; from Eq.~\eqref{eq:simplebirthdeath} we have an increase from $N_{\rm initial}=b_i/\delta_i$ to $N_{\rm final}=b_i/(\delta_i(1-\epsilon))=N_{\rm initial}/(1-\epsilon)$.
 The corresponding selection coefficient increases from $s_{\rm initial}= \epsilon b_i$ to $s_{\rm final}=s_{\rm initial}/(1-\epsilon)$.
@@
 The net effect on $f(\overline{b},N)$ is precisely zero in equilibrium: in a single-type equilibrium we have $b_i/\overline{b}=1$ and so $f(\overline{b},N)=d_i$ exactly at the beginning and end of a pure $b$ sweep, even though the density $N$ increases.
-Strictly speaking there is some deviation in $f(N)$ from $d_i$ during the sweep, but this deviation is an order of magnitude smaller than for a $\delta$ sweep (the deviation due to a sweep with proportional effect $b_i\rightarrow b_i(1+\epsilon)$ is only of order $\epsilon^2$, whereas the analogous effect in Fig.~\ref{fig:strengthofselection} is of order $\epsilon$; see Appendix D for details).
+Strictly speaking there is some deviation in $f(\overline{b},N)$ from $d_i$ during the sweep, but this deviation is an order of magnitude smaller than for a $\delta$ sweep (the deviation due to a sweep with proportional effect $b_j= b_i(1+\epsilon)$ is only of order $\epsilon^2$, whereas the analogous effect in Fig.~\ref{fig:strengthofselection} is of order $\epsilon$; see Appendix D for details).
 Since selection must already be quite strong for a $\delta$ sweep to threaten Eq.~\eqref{eq:canonical}, we conclude that $b$ sweeps also obey the canonical selection equation (to a close approximation).
@@
 A more general Lotka-Volterra model introduces the inter-type $\alpha_{ij}$ competition coefficients, with selection on these termed ``$\alpha$-selection'' \citep{gill_1974,joshi_2001}.
-Setting aside $r$ which confers no selective advantage at equilibrium, we are left with $K$ and $\alpha$, which both behave like $\delta$ in that they are density-dependent and cause density to change over a sweep (although $N$ only dips temporarily during an $\alpha$-sweep).
+Setting aside $r$ which confers no selective advantage at equilibrium, we are left with $K$ and $\alpha$, which both behave like $\delta$ in Eq.~\eqref{eq:simplebirthdeath} in that they are density-dependent and cause density to change over a sweep (although $N$ only dips transiently during an $\alpha$-sweep).
 Thus, strong selection is sufficient for relative fitness to break down in the classical view of density-dependent selection.
@@
 Spatial localization of consumable resources (e.g.
-due to restricted movement of  nutrients through soils) will tend to create a territorial situation similar to the lottery model, where resource competition only occurs locally and both it any interference competition are subsumed into the competitive ability $c$, which does not affect $N$.
+due to restricted movement of  nutrients through soils) will tend to create a territorial situation similar to the lottery model, where resource competition only occurs locally and both it and interference competition are subsumed into the competitive ability $c$, which does not affect $N$.
 
@@
 Whether or not our density-dependent lottery model is a good description of \textit{Drosophila} ecology, the close connection between our model and Wright-Fisher is useful, because drift in our model should behave broadly similarly.
-Thus, our model it should provide a useful starting point for analyzing evolution in this and other far-from-equilibrium situations.
+Thus, our model should provide a useful starting point for analyzing evolution in this and other far-from-equilibrium situations.
 
@@
 
-\section*{Appendix A: Poisson dispersal}
-
-For simplicity of presentation, we assume a Poisson distribution for the $x_i$ as our model of dispersal.
-Strictly speaking, the total number of $i$ propagules $\sum x_i$ (summed over unoccupied territories) is then no longer a constant $m_i$, but fluctuates between generations for a given mean $m_i$.
-Nevertheless, since we do not consider the random fluctuations in type abundances here, and for ease of comparison with the classic lottery model, we ignore the fluctuations in $m_i$.
-Instead we focus on Poisson fluctuations in propagule composition in each territory.
-
-In the exact model of random dispersal, the counts of a type's propagules across unnocupied territories follows a multinomial distribution with dimension $U$, total number of trials equal to $m_i$, and equal probabilities $1/U$ for a propagule to land in a given territory.
-Thus, the $x_i$ in different territories are not independent random variables.
-However, for sufficiently large $U$ and $m_i$, this multinomial distribution for the $x_i$ across territories is closely approximated by a product of independent Poisson distributions for each territory, each with rate parameter $l_i$ \citep[Theorem 1]{arenbaev_1977}.
-Since we are ignoring finite population size effects, we effectively have $T\rightarrow \infty$, in which case $U$ can only be small enough to violate the Poisson approximation if there is vanishing population turnover, and then the dispersal distribution is irrelevant anyway.
-Likewise, in ignoring stochastic finite population size for the $n_i$, we have effectively already assumed that $m_i$ is large enough to justify the Poisson approximation (the error scales as $1/\sqrt{m_i}$; \citealt{arenbaev_1977}).
-
-\section*{Appendix B: Growth equation derivation}
+\section*{Appendix A: Growth equation derivation}
 
@@
 
-\section*{Appendix C: Total density in the Lotka-Volterra competition model}
+\section*{Appendix B: Total density in the Lotka-Volterra competition model}
 
@@
 
-\section*{Appendix D: Density-dependence of $b$-selection}
+\section*{Appendix C: Density-dependence of $b$-selection}
 

### 2018-03-07 manuscript.tex
l

@@
 
-We assume that reproductively mature individuals (``adults'') each require their own territory to survive and reproduce.
+We assume that reproductively mature individuals (``adults'') require their own territory to survive and reproduce.
 All territories are identical, and the total number of territories is $T$.
@@
 In a given iteration, the number of adults of the $i$'th type will be denoted by $n_i$, the total number of adults by $N=\sum_i n_i$, and the number of unoccupied territories by $U=T-N$.
-We assume that the $n_i$ are large enough that stochastic fluctuations in the $n_i$ (``drift'') can be ignored (with $T$ also assumed large to allow for small type densities $n_i/T$).
+We assume that the $n_i$ are large enough that stochastic fluctuations in the $n_i$ (``drift'') can be ignored (with $T$ also assumed large to allow for low type densities $n_i/T$).
 
@@
 \includegraphics[scale=0.8]{lottery.pdf}
-\caption{\label{fig:lottery} Each iteration of our model has three elements.
-First, propagules are produced by adults and dispersed at random (only propagules landing on unoccupied territories are shown).
-Some territories may receive zero propagules.
-Lottery competition then occurs in each unoccupied territory that receives a propagule (only illustrated in one territory).
-Each type has a probability proportional to $c_i x_i$ of securing a given territory, where $c_i$ measures competitive ability and $x_i$ is the number of propagules that disperse there.
-In the illustrated territory, the black type disperses more propagules but is a poorer competitor.
-Territories are then made available by deaths among those adults present at the start of the iteration (red crosses).}
+\caption{\label{fig:lottery} One iteration of our model.
+Propagules are dispersed by adults at random (only propagules landing on unoccupied territories are shown).
+Territories can receive zero propagules.
+Lottery competition then occurs in each territory that receives more than one propagule (only illustrated in one territory).
+In a given territory, each type has probability proportional to $c_i x_i$ of winning the territory, where $c_i$ measures competitive ability and $x_i$ is the number of $i$ propagules present.
+In the illustrated territory, more black propagules are present, but white is a stronger competitor and has a higher probability of winning.
+Territories are made available for the next iteration by the death of adults present at the start of the iteration (red crosses).}
 \end{figure}
 
-Each iteration, adults produce new offspring (``propagules'').
-These disperse at random, regardless of distance from their parents, and independently of each other (e.g.
-there is no avoidance of territories crowded with propagules).
-To model the dispersal process, we assume that the number of propagules of the $i$'th type landing in any particular territory, denoted $x_i$, follows a Poisson distribution $p_i(x_i)=l_i^{x_i} e^{-l_i}/x_i!$.
-Here $l_i=m_i/U$ is the mean territorial propagule density, where $m_i$ is the mean number of propagules produced by type $i$ (the total propagule density will be denoted by $L=\sum_i l_i$ and the mean total number of propagules by $M=\sum_i m_i$).
-We assume that adults cannot be ousted by juveniles, so that propagules landing on occupied territories are doomed.
-Thus, the number of propagules dispersing to unoccupied territories can be written as $m_i=b_in_iU/T$, where $b_i$ is the number of propagules produced by each adult from type $i$.
-Note that due to our assumption of uniform dispersal, the parameter $b_i$ can be thought of as a measure of ``colonization ability'', which combines fecundity and dispersal ability \citep{levins_71,tilman_94,bolker_99}.
+Each iteration, adults produce propagules which disperse at random, independently of distance from their parents, and independently of each other.
+We assume that each adult from type $i$ produces $b_i$ propagules on average, so that the mean number of $i$ propagules dispersing to unoccupied territories is $m_i=b_in_iU/T$.
+The parameter $b_i$ can be thought of as a measure of ``colonization ability'', which combines fecundity and dispersal ability \citep{levins_71,tilman_94,bolker_99}.
+Random dispersal is then modeled using a Poisson distribution $p_i(x_i)=l_i^{x_i} e^{-l_i}/x_i!$ for the number $x_i$ of $i$ propagules dispersing to any particular unoccupied territory, where $l_i=m_i/U$ is the mean propagule density in unoccupied territories.
+The total propagule density will be denoted $L=\sum_i l_i$ and the mean total number of propagules by $M=\sum_i m_i$.
 
-When multiple propagules land on the same territory, the victor is determined by lottery competition: type $i$ wins a territory with probability $c_i x_i/\sum_j c_j x_j$, where $c_i$ is a constant representing relative competitive ability (Fig.
+We assume that adults cannot be ousted by juveniles, so that recruitment to adulthood occurs exclusively in unoccupied territories.
+When multiple propagules land on the same unoccupied territory, the winner is determined by lottery competition: type $i$ wins a territory with probability $c_i x_i/\sum_j c_j x_j$, where $c_i$ is a constant representing relative competitive ability (Fig.
 \ref{fig:lottery}).
-We expect that a fraction $p_1(x_1)\cdots p_G(x_G)$ of the $U$ unoccupied territories will have the propagule composition $x_1,\ldots,x_G$.
-Type $i$ is expected to win a proportion $c_i x_i/\sum_j c_j x_j$ of these.
-Ignoring fluctuations about these two expectations (due to our no-drift, large $n_i$, large $T$ approximation), type $i$'s territorial acquisition is given by
+Since the expected fraction of unoccupied territories with propagule composition $x_1,\ldots,x_G$ is $p_1(x_1)\cdots p_G(x_G)$, and type $i$ is expected to win a proportion $c_i x_i/\sum_j c_j x_j$ of these, type $i$'s expected territorial acquisition is given by
 \begin{equation}
@@
 \end{equation}
-where the sum only includes territories with at least one propagule present.
+where the sum only includes territories with at least one propagule present. Since we do not consider drift here, we will not analyze the fluctuations about these two expectations.
 
-Finally, we assume that adult mortality only occurs in adults present at the start of the  iteration, and at a constant, type-specific per-capita rate $0\leq d_i\leq 1$.
-Thus, the overall change in type abundances is
+Adult mortality only occurs in adults present at the start of the  iteration, and at a constant, type-specific per-capita rate $0\leq d_i\leq 1$ (Fig.~\ref{fig:lottery}).
+This gives an overall change in type abundances of
 \begin{equation}
@@
 \end{equation}
-Fig.~2 illustrates one iteration of the model.
 
-\subsection*{Connection to the Wright-Fisher and classic lottery models}
+\subsection*{Connection to the classic lottery model}
 
@@
 From the law of large numbers, the composition of propagules in each territory will then not deviate appreciably from the mean composition $l_1,l_2,\ldots,l_G$ ($G$ is the number of types present), and so the probability that type $i$ wins any particular unoccupied territory is approximately $c_i l_i/\sum_j c_j l_j$.
-Then the numbers of territories won by each type $\Delta_+ n_1,\Delta_+ n_2,\ldots,\Delta_+ n_G$ follow a multinomial distribution with $U$ trials and success probabilities $\frac{c_1 l_1}{\sum_j c_j l_j},\frac{c_2 l_2}{\sum_j c_j l_j},\ldots,\frac{c_G l_G}{\sum_j c_j l_j}$, respectively.
+The numbers of territories won by each type $\Delta_+ n_1,\Delta_+ n_2,\ldots,\Delta_+ n_G$ then follow a multinomial distribution with $U$ trials and success probabilities $\frac{c_1 l_1}{\sum_j c_j l_j},\frac{c_2 l_2}{\sum_j c_j l_j},\ldots,\frac{c_G l_G}{\sum_j c_j l_j}$, respectively.
 Type $i$ is expected to win a proportion $c_i l_i/\sum_j c_j l_j$ of the $U$ available territories, and deviations from this expected outcome are small (since $T$ is large by assumption), giving
@@
 where $\overline{c}=\sum_j c_j m_j/M$ is the mean competitive ability for a randomly selected propagule.
-In section ``Analytical approximation of the density-dependent lottery'', we derive a generalization of Eq.~\eqref{eq:lottery} that accommodates arbitrary propagule densities $l_i$.
+Note that all unoccupied territories are filled in a single iteration of the classic lottery model, whereas our more general model Eq.~\eqref{eq:growthsumuncoupled} allows for territories to be left unoccupied and hence for arbitrary propagule densities $l_i$.
 
-There is a close connection between the classic lottery model and the Wright-Fisher model of genetic drift \citep{svardal_2015}.
+%There is a close connection between the classic lottery model and the Wright-Fisher model of genetic drift \citep{svardal_2015}.
 In the Wright-Fisher model, type abundances are sampled each generation from a multinomial distribution with success probabilities $w_i n_i/\sum_j w_j n_j$, where $w$ is relative fitness and the $n_i$ are  type abundances in the preceding generation.
@@
 
-Eq.
-\eqref{eq:growthsumuncoupled} involves an expectation over the time-dependent dispersal distributions $p_i$, and is thus too complicated to give intuition about the dynamics of density-dependent lottery competition.
-We now evaluate this expectation.
+Eq.~\eqref{eq:growthsumuncoupled} involves an expectation over the time-dependent dispersal distributions $p_i$.
+Here we evaluate this expectation to obtain a better intuition about the dynamics of density-dependent lottery competition.
 
-Similarly to the high-$l_i$ approximation of the classic lottery model, we replace the $x_i$ with appropriate mean values, although we cannot simply replace $x_i$ with $l_i$ as in Eq.~\eqref{eq:lottery}.
-The classic lottery model breaks down for types with low propagule density ($l_i\ll 1$) because territorial acquisition is then not correctly represented by a lottery in each territory with the mean propagule density.
-For a type with low propagule density $l_i\ll 1$, we have $x_i=1$ in the territories where its propagules land, and so its growth comes entirely from territories which deviate appreciably from $l_i$.
-To account for this, we separate Eq.
-\eqref{eq:growthsumuncoupled} into $x_i=1$ and $x_i>1$ parts.
+Similarly to the high-$l_i$ approximation of the classic lottery model, we replace the $x_i$ in Eq.~\eqref{eq:growthsumuncoupled} with ``effective'' mean values, although we cannot simply use the means $l_i$ as in the classic lottery.
+For a rare type, growth comes almost entirely from territories with $x_i=1$, for which its mean density $l_i\ll 1$ is not representative.
+We therefore separate Eq.~\eqref{eq:growthsumuncoupled} into $x_i=1$ and $x_i>1$ components.
 Our more general approximation only requires that there are no large discrepancies in competitive ability (i.e.
@@
 The second, $R_i c_i/\overline{c}$, represents competitive victories when the $i$ type is a rare invader in a high density population (i.e.
-it determines invasion fitness \citep{metz_1992}).
+it determines invasion fitness; \citealt{metz_1992}).
 The third term, $A_i c_i/\overline{c}$, represents competitive victories when the $i$ type is abundant.
@@
 \includegraphics[scale=0.8]{simulationcomparison.pdf}
-\caption{\label{fig:simcomp} Comparison of the analytical approximation Eq.
-\eqref{eq:master} with simulations.
-Per-propagule success probability $\Delta_+ n_i/l_i U$ from the classic lottery model, individual-based simulations of random dispersal and lottery competition, and Eq.
-\eqref{eq:master} and its three components.
+\caption{\label{fig:simcomp} Comparison of the per-propagule success probability $\Delta_+ n_i/m_i$ in simulations, in Eq.
+\eqref{eq:master} and its three components, and in the classic lottery model.
+In the simulations, $x_i$ values are sampled from a Poisson distribution with mean $l_i$ in $U$ territories, and then the victor in each territory is decided by random sampling weighted by the lottery win probabilities $c_ix_i/\sum_j c_j x_j$.
 Two types are present, a rare type with $c_1=1.5$, and a common type with $c_2=1$.
-Simulation points are almost invisible in for the common type due to near exact agreement with Eq.
+Simulation points are almost invisible for the common type due to near exact agreement with Eq.
 \eqref{eq:master}.

### 2018-03-08 manuscript.tex
l

@@
 There are a variety of different measures of fitness.
-Some widely used examples in evolutionary ecology are expected lifetime reproductive ratio $R_0$, intrinsic population growth rate $r$, saturation population density/carrying capacity (often labeled ``$K$'') \citep{benton_2000}, and invasion fitness \citep{metz_1992}.
-In addition, ``relative fitness'' is the standard in much of evolutionary biology, particularly evolutionary genetics, where attention is generally restricted to relative genotypic proportions \cite[pp.
+Some widely used examples are expected lifetime reproductive ratio $R_0$, intrinsic population growth rate $r$, saturation population density/carrying capacity (often labeled ``$K$'') \citep{benton_2000}, and invasion fitness \citep{metz_1992}.
+In addition, ``relative fitness'' is the standard in much of evolutionary biology, particularly evolutionary genetics, where the focus is relative genotypic proportions \cite[pp.
 468]{barton_2007}.
-The variety of fitness measures is not problematic in itself, because different measures may be more useful in different circumstances.
-But it should be clear how the measure being used is connected to the processes of birth and death which govern population biology \citep{metcalf_2007,doebeli_2017}.
-While such a connection is fairly clear for absolute fitness measures like $r$, relative fitness seems largely divorced from population ecology.
+The variety of fitness measures is not problematic in itself, because different measures have different uses.
+But it should be clear how these measures are connected to the processes of birth and death which ultimately drive selection \citep{metcalf_2007,doebeli_2017}.
+While such a connection is fairly clear for absolute fitness measures like $r$, relative fitness is largely divorced from population ecology.
 It has even been proposed that relative fitness be justified from measure theory, abandoning population biology altogether \citep{wagner_2010}.
@@
 \citep{kimura1969natural,mallet_2012}), we instead do the reverse, and attempt to make population ecological sense of the widely-used Wright-Fisher relative-fitness model.
+
 Our starting point is the classic lottery model of territorial contest \citep{sale_77,chesson_1981}.
-Like the Wright-Fisher model, the classic lottery assumes a saturated population with constant $N$, and fitness involves a product of fertility and juvenile viability \citep[pp.
+The classic lottery assumes a saturated population with constant $N$, and fitness involves a product of fertility and juvenile viability \citep[pp.
 185]{crow_1970}, but unlike the Wright-Fisher model, generations can overlap.
+
 Our first task is to generalize the lottery model to create a variable-density version of the Wright-Fisher model with overlapping generations (sections ``Model'' and ``Analytical approximation of the variable-density lottery'').
@@
 Note that all unoccupied territories are filled in a single iteration of the classic lottery model, whereas our more general model Eq.~\eqref{eq:growthsumuncoupled} allows for territories to be left unoccupied and hence for arbitrary propagule densities $l_i$.
-
-%There is a close connection between the classic lottery model and the Wright-Fisher model of genetic drift \citep{svardal_2015}.
-In the Wright-Fisher model, type abundances are sampled each generation from a multinomial distribution with success probabilities $w_i n_i/\sum_j w_j n_j$, where $w$ is relative fitness and the $n_i$ are  type abundances in the preceding generation.
-Population size $N$ remains constant.
-This is equivalent to the classic lottery model with non-overlapping generations ($d_i=1$ for all $i$) and relative fitness given by $w_i=b_i c_i$ i.e.
-a product of fertility and viability \citep[pp.
-185]{crow_1970}.
-Thus, the classic lottery model is essentially the Wright-Fisher model extended to allow overlapping generations, but ignoring drift.
-This means that our extension of the classic lottery model to arbitrary densities represents a variable-density generalization of the Wright-Fisher model (we also do not consider drift here).
 

### 2018-03-09 manuscript.tex
l

@@
 Random dispersal is then modeled using a Poisson distribution $p_i(x_i)=l_i^{x_i} e^{-l_i}/x_i!$ for the number $x_i$ of $i$ propagules dispersing to any particular unoccupied territory, where $l_i=m_i/U$ is the mean propagule density in unoccupied territories.
-The total propagule density will be denoted $L=\sum_i l_i$ and the mean total number of propagules by $M=\sum_i m_i$.
+The total propagule density will be denoted $L=\sum_i l_i$.
 
 We assume that adults cannot be ousted by juveniles, so that recruitment to adulthood occurs exclusively in unoccupied territories.
-When multiple propagules land on the same unoccupied territory, the winner is determined by lottery competition: type $i$ wins a territory with probability $c_i x_i/\sum_j c_j x_j$, where $c_i$ is a constant representing relative competitive ability (Fig.
+When multiple propagules land on the same unoccupied territory, the winner is determined by lottery competition: type $i$ wins a territory with probability $c_i x_i/\sum_i c_i x_i$, where $c_i$ is a constant representing relative competitive ability (Fig.
 \ref{fig:lottery}).
-Since the expected fraction of unoccupied territories with propagule composition $x_1,\ldots,x_G$ is $p_1(x_1)\cdots p_G(x_G)$, and type $i$ is expected to win a proportion $c_i x_i/\sum_j c_j x_j$ of these, type $i$'s expected territorial acquisition is given by
+Since the expected fraction of unoccupied territories with propagule composition $x_1,\ldots,x_G$ is $p_1(x_1)\cdots p_G(x_G)$ where $G$ is the number of types present, and type $i$ is expected to win a proportion $c_i x_i/\sum_i c_i x_i$ of these, type $i$'s expected territorial acquisition is given by
 \begin{equation}
-\Delta_+ n_i=U\sum_{x_1,\ldots,x_G} \frac{c_i x_i}{\sum_j c_j x_j} p_1(x_1)\cdots p_G(x_G), \label{eq:growthsumuncoupled}
+\Delta_+ n_i=U\sum_{x_1,\ldots,x_G} \frac{c_i x_i}{\sum_i c_i x_i} p_1(x_1)\cdots p_G(x_G). \label{eq:growthsumuncoupled}
 \end{equation}
-where the sum only includes territories with at least one propagule present. Since we do not consider drift here, we will not analyze the fluctuations about these two expectations.
+Here the sum only includes territories with at least one propagule present. Since we do not consider drift here, we will not analyze the fluctuations around these two expectations.
 
@@
 
-In the classic lottery model \citep{chesson_1981}, unoccupied territories are assumed to be saturated with propagules from every type $l_i\gg 1$.
-From the law of large numbers, the composition of propagules in each territory will then not deviate appreciably from the mean composition $l_1,l_2,\ldots,l_G$ ($G$ is the number of types present), and so the probability that type $i$ wins any particular unoccupied territory is approximately $c_i l_i/\sum_j c_j l_j$.
-The numbers of territories won by each type $\Delta_+ n_1,\Delta_+ n_2,\ldots,\Delta_+ n_G$ then follow a multinomial distribution with $U$ trials and success probabilities $\frac{c_1 l_1}{\sum_j c_j l_j},\frac{c_2 l_2}{\sum_j c_j l_j},\ldots,\frac{c_G l_G}{\sum_j c_j l_j}$, respectively.
-Type $i$ is expected to win a proportion $c_i l_i/\sum_j c_j l_j$ of the $U$ available territories, and deviations from this expected outcome are small (since $T$ is large by assumption), giving
+In the classic lottery model \citep{chesson_1981}, unoccupied territories are assumed to be saturated with propagules from every type ($l_i\rightarrow \infty$ for all $i$).
+From the law of large numbers, the composition of propagules in each territory will not deviate appreciably from the mean composition $l_1,l_2,\ldots,l_G$.
+Type $i$ is thus expected to win a proportion $c_i l_i/\sum_i c_i l_i$ of the $U$ available territories,
 \begin{equation}
-\Delta_+ n_i=\frac{c_i l_i}{\sum_j c_j l_j}U=\frac{c_i l_i}{\overline{c}L}U, \label{eq:lottery}
+\Delta_+ n_i=\frac{c_i l_i}{\sum_i c_i l_i}U=\frac{c_i l_i}{\overline{c}L}U, \label{eq:lottery}
 \end{equation}
-where $\overline{c}=\sum_j c_j m_j/M$ is the mean competitive ability for a randomly selected propagule.
-Note that all unoccupied territories are filled in a single iteration of the classic lottery model, whereas our more general model Eq.~\eqref{eq:growthsumuncoupled} allows for territories to be left unoccupied and hence for arbitrary propagule densities $l_i$.
+where $\overline{c}=\sum_i c_i m_i/\sum_i m_i$ is the mean competitive ability for a randomly selected propagule.
+Note that all unoccupied territories are filled in a single iteration of the classic lottery model, whereas our more general model Eq.~\eqref{eq:growthsumuncoupled} allows for territories to be left unoccupied and hence also accommodates low propagule densities.
 
@@
 \eqref{eq:lottery}, the classic lottery per-propagule success rate $c_i/\overline{c}L$ has been replaced by three separate terms.
-The first, $e^{-L}$, accounts for propagules which land alone on unoccupied territories; these territories are won without contest.
-The second, $R_i c_i/\overline{c}$, represents competitive victories when the $i$ type is a rare invader in a high density population (i.e.
-it determines invasion fitness; \citealt{metz_1992}).
-The third term, $A_i c_i/\overline{c}$, represents competitive victories when the $i$ type is abundant.
+The first, $e^{-L}$, accounts for propagules which land alone on unoccupied territories; these propagules secure the territories without contest.
+The second, $R_i c_i/\overline{c}$, represents competitive victories on territories where only a single $i$ propagule lands, and at least one other propagule from a different type (this term dominates the growth of a rare invader in a high density population and determines invasion fitness).
+The third term, $A_i c_i/\overline{c}$, represents competitive victories in territories where two or more $i$ type propagules are present.
 The relative importance of these three terms varies with both the overall propagule density $L$ and the relative propagule frequencies $l_i/L$.
@@
 \includegraphics[scale=0.8]{simulationcomparison.pdf}
-\caption{\label{fig:simcomp} Comparison of the per-propagule success probability $\Delta_+ n_i/m_i$ in simulations, in Eq.
-\eqref{eq:master} and its three components, and in the classic lottery model.
-In the simulations, $x_i$ values are sampled from a Poisson distribution with mean $l_i$ in $U$ territories, and then the victor in each territory is decided by random sampling weighted by the lottery win probabilities $c_ix_i/\sum_j c_j x_j$.
+\caption{\label{fig:simcomp} Comparison of Eq.~\eqref{eq:master}, the classic lottery model, and simulations.
+The vertical axis is per-propagule success rate for all propagules $\Delta_+ n_i/m_i$, and for the three separate components in Eq.~\eqref{eq:master}.
+Simulations are 	conducted as follows: $x_1,\ldots,x_G$ values are sampled $U$ times from Poisson distributions with respective means $l_1,\ldots,l_G$, and the victor in each territory is then decided by random sampling weighted by the lottery win probabilities $c_ix_i/\sum_j c_j x_j$.
 Two types are present, a rare type with $c_1=1.5$, and a common type with $c_2=1$.
-Simulation points are almost invisible for the common type due to near exact agreement with Eq.
-\eqref{eq:master}.
+Simulation points are almost invisible for the common type due to near exact agreement with Eq.~\eqref{eq:master}.
 Dashed lines show the breakdown of the classic lottery model.
-Parameters: $m_1=10^4$ and $m_2=9\times 10^4$ and $U$ varies between $5\times 10^3$ and $10^6$.}
+Parameters: $U=10^5$ and $l_1/l_2=0.1$ is kept fixed while varying the total density $L$.}
 \end{figure}
 
-Eq.~\eqref{eq:master} takes a much simpler form if all types are competitively equivalent ($c_i=c$),
+In the special case that all types are competitively equivalent (identical $c_i$), Eq.~\eqref{eq:master} takes a simpler form,
 \begin{equation}
@@
 \end{equation}
-Here $1-e^{-L}$ is the fraction of territories that receive at least one propagule under Poisson dispersal, $(1-e^{-L})U$ is the total number of territories gained, and type $i$ receives a fraction $l_i/L$ of these.
+This formula can also be deduced directly from Eq.~\eqref{eq:growthsumuncoupled}: $1-e^{-L}$ is the fraction of territories that receive at least one propagule under Poisson dispersal, $(1-e^{-L})U$ is the total number of such territories, and type $i$ is expected to receive a fraction $l_i/L$ of these.
 Total population density thus grows according to
@@
 
+\subsection*{Selection in the variable-density lottery}
+
+We now outline the properties of selection in the variable-density lottery, starting with selection on competitive ability $c$.
+Eq.~\eqref{eq:Nmaster} shows that $c$ is not a density-regulating trait: $c$ only affects the relative likelihood for each type to win a contested territory, not whether a territory is contested in the first place.
+Selection between types which only differ in $c$ occurs without changing density $N$.
+However, $c$-selection is density-dependent, with the strength of selection (measured by the difference between types in the per-capita growth rate $\Delta n_i/n_i$) peaking at an intermediate density (Fig.~\ref{fig:DDS_lottery}).
+This intermediate peak occurs because most territories are claimed without contest at low density, whereas few unoccupied territories are available to be contested at high density.
+
+Turning to selection on the density-regulating traits $b$ and $d$, we write Eq.~\eqref{eq:masterequalc} in the form
+\begin{equation}
+\frac{\Delta n_i}{n_i} = \frac{b_i}{\overline{b}}\frac{1-e^{-\overline{b}N/T}}{N}(T-N)-d_i, \label{eq:bdensitydependence}
+\end{equation}
+where we have used that fact that $L=\overline{b}N/T$, and $\overline{b}$ is the population mean $b$.
+It is clear from Eq.~\eqref{eq:bdensitydependence} that selection on adult mortality $d$ is independent of density.
+On the other hand, $b$-selection is density-dependent.
+The advantage of having greater $b$ declines with density because fewer territories are available to to be claimed (Fig.~\ref{fig:DDS_lottery}).
+
+\begin{figure}
+\centering
+\includegraphics[scale=0.8]{DDS_lottery.pdf}
+\caption{\label{fig:DDS_lottery} The density-dependence of selection in our variable-density lottery plotted as difference in propagule success rate $\Delta n_j/m_j-\Delta n_i/n_i$ between an adaptive variant $j$ and a wildtype $i$ present at the same frequency.
+Here $b_i=1$, $d_i=0.5$ and $c_i=1$.
+For $b$-selection we set $b_j=b_i(1+\epsilon)$, and similarly for $c$ and $d$, with $\epsilon=0.1$.
+$d$-selection is density-independent, $b$-selection gets weaker with lower territorial availability, while $c$-selection initially increases with density as territorial contests become more important, but eventually also declines as  available territories become scarce.
+The equilibrium density for the $i$ type is $\approx 0.4$.}
+\end{figure}
+
 \subsection*{Selection-dependent density and $K$-selection}
 
-Equipped with our variable-density lottery model, we now start evaluating the validity of Eq.~\eqref{eq:canonical}.
-In this section we explore whether we should expect population density to vary as a result of selection \citep{prout_1980}.
-Since the idea that density does vary with selection is closely connected to the notion of ``$K$-selection'', we start by revisiting MacArthur's analysis of selection in crowded environments \citep{macarthur_1967}.
+We now start evaluating the limits of Eq.~\eqref{eq:canonical} in the context of the variable-density lottery model, focusing in this section on the issue of whether density varies as a result of selection \citep{prout_1980}.
+We showed in the previous section that $c$-selection in the variable-density lottery does not affect density.
+However, this is not the case in most previous models of density-dependent selection, which have been strongly influenced by the notion of ``$K$-selection'' and the associated questions about whether selection in crowded populations will tend to increase population density \citep{anderson_1971,roughgarden_1979,lande_2009}.
+To clarify the connection between ``$K$-selection'' and the variable density lottery model, we thus revisit MacArthur's analysis of selection in crowded populations \citep{macarthur_1967}.
 
@@
 The functions $f_1$ and $f_2$ must decline to zero if $n_1$ or $n_2$ are sufficiently large, because no population has unlimited resources.
-This defines the nullclines $f_1(n_1,n_2)=0$ and $f_2(n_1,n_2)=0$ in $(n_1,n_2)$ space.
+This defines nullclines $f_1(n_1,n_2)=0$ and $f_2(n_1,n_2)=0$ in $(n_1,n_2)$ space.
 The outcome of selection is then determined by the relationship between these nullclines.
@@
 
-Thus, when MacArthur concludes that  ``fitness is $K$'' in crowded populations \citep[pp.
-149]{macarthur_1967}, the meaning is that selection either favors the ability to keep growing at ever higher densities (moving a type's own nullcline outwards), or the ability to suppress the growth of competitors at lower densities (moving the nullcline of competitors inwards) \citep{gill_1974}.
+In summary, MacArthur's conclusion that  ``fitness is $K$'' in crowded populations \citep[pp.
+149]{macarthur_1967} means that selection either favors the ability to keep growing at ever higher densities (moving a type's own nullcline outwards), or the ability to suppress the growth of competitors at lower densities (moving the nullcline of competitors inwards) \citep{gill_1974}.
 This general idea is much broader than ``$K$-selection'' in the sense of selection for greater saturation density, and applies even if the nullclines are nonlinear to such an extent that the ``$K$'' values themselves do not give much information about the regions of high-density growth.
@@
 
-It is obvious from Eq.~\eqref{eq:LV} that selection can favor a superior competitor in a crowded population even if its saturation density is the same as, or lower than that of the other types present.
-However, the Lotka-Volterra model still couples selection to population density \citep{smouse_1976}.
+It is obvious in the Lotka-Volterra model that selection can favor a superior competitor in a crowded population even if its saturation density is lower than that of the other types present.
+Nevertheless, the Lotka-Volterra model still couples population density to selection \citep{smouse_1976}.
 Fig.~\ref{fig:LVvslottery}a shows Lotka-Volterra selection between two types with the same saturation density ($\alpha_{11}=\alpha_{22}$, $\alpha_{21}>\alpha_{12}$).
-Even though the initial and final densities of a sweep are the same, density is not constant over a sweep.
-Only a highly restricted subset of $r$ and $\alpha$ values will keep $N$ constant over a selective sweep (further details in Appendix C).
+Even though the initial and final densities of a sweep are the same, density is not constant over a sweep (constant density is possible but only occurs for a highly restricted subset of $r$ and $\alpha$ values; further details in Appendix C; also see \citealt{mallet_2012}).
 Intuitively, for one type to exclude another with the same saturation density, competitive suppression of growth between types must be stronger than competitive suppression of growth within types, causing a dip in $N$ over the sweep.
 
-By contrast, if one type in our density-dependent lottery model has a $c$ advantage but the types are otherwise identical (so that each type has the same saturation density), the density trajectories converge on the line of constant density equal to the saturation density (Fig.~\ref{fig:LVvslottery}b).
-Selection then occurs purely along this line, similarly to Fig.~\ref{fig:Ksel}b.
-This occurs because $c$ does not directly affect $N$: it only affects the relative likelihood for each type to win a contested territory, not whether a territory is contested in the first place (this can be seen formally in Eq.~\eqref{eq:Nmaster}).
+By contrast, density trajectories in the variable-density lottery model converge on the saturation density line (Fig.~\ref{fig:LVvslottery}b).
+Selection then occurs along this line, similarly to Fig.~\ref{fig:Ksel}b.
 In other words, once the population reaches demographic equilibrium, it behaves indistinguishably from a constant-$N$ relative fitness model.
-While quite different from classical growth models like the Lotka-Volterra, this is all perfectly consistent with MacArthur's general argument.
-
-The constant-$N$ behavior of $c$-selection arises from the role of $c$ as a trait determining relative competitive success in territorial contests.
-As such, this behavior is a result of a reproductive excesses.
-By contrast, previous models of selection-independent density either used unusual models of competition \citep{kimura1969natural,nei1971fertility}, or made restrictive parameter choices in the Lotka-Volterra model (Appendix C; \citealt{smouse_1976,mallet_2012}).
+This behavior occurs because $c$ determines relative competitive success in territorial contests.
+As such, it is a manifestation of a reproductive excess of propagules.
+Previous models of selection-independent density have either relied on unusual models of competition \citep{kimura1969natural,nei1971fertility}, or made restrictive Lotka-Volterra parameter choices (Appendix C; \citealt{smouse_1976,mallet_2012}).
 
@@
 
-Let us now turn to selection on $b$ and $d$ in our lottery model.
-Recall that $m_i=b_i n_i U/T$, and so $L=M/U=\overline{b}N/T$ where $\overline{b}$ is the population mean $b$.
-Thus, from Eq.~\eqref{eq:masterequalc} we have
-\begin{equation}
-\Delta n_i = \left(\frac{b_i}{\overline{b}}\frac{1-e^{-\overline{b}N/T}}{N}(T-N)-d_i\right)n_i. \label{eq:bdensitydependence}
-\end{equation}
-It can be seen that the mortality $d$ is akin to the birth rate in Eq.~\eqref{eq:simplebirthdeath}, and so, while $d$ does affect density, selection on $d$ is density independent.
  Thus, $d$ sweeps follow the canonical relative fitness model exactly (Fig.~\ref{fig:DDS_lottery}).
@@
  At first glance, $b$ in Eq.~\eqref{eq:bdensitydependence} appears to be analogous to the $\delta$ in Eq.~\eqref{eq:simplebirthdeath} because it regulates density and is multiplied by the density-dependent term $f(\overline{b},N)=\frac{1-e^{-\overline{b}N/T}}{N}(T-N)$.
-This term declines from $\overline{b}$ at low density to zero at high density and as a result, selection on $b$ is density-dependent.
-The source of this density-depedence is that the selective advantage from having greater $b$ depends on the number of territories being contested; if almost all are occupied, then there is little advantage to having a greater $b$.
 
@@
 Since selection must already be quite strong for a $\delta$ sweep to threaten Eq.~\eqref{eq:canonical}, we conclude that $b$ sweeps also obey the canonical selection equation (to a close approximation).
-
-\begin{figure}
-\centering
-\includegraphics[scale=0.8]{DDS_lottery.pdf}
-\caption{\label{fig:DDS_lottery} The density-dependence of selection in our variable-density lottery plotted as the selection coefficient $(\Delta n_j-\Delta n_i)/n_j$ experienced by an adaptive variant $j$ present at the same frequency as a wildtype $i$.
-Here $b_i=1$, $d_i=0.5$ and $c_i=1$.
-For $b$-selection we set $b_j=b_i(1+\epsilon)$, and similarly for $c$ and $d$, with $\epsilon=0.1$.
-$d$-selection is density-independent, $b$-selection gets weaker with lower territorial availability, while $c$-selection initially increases with density as territorial contests become more important, but eventually also declines as  available territories become scarce.
-The equilibrium density for the $i$ type is $\approx 0.4$.}
-\end{figure}
 

### 2018-03-09 DDS_lottery.py

@@ -69,7 +69,7 @@ ax1.legend(loc='upper center',prop={'size':11})
 #ax1.plot([eq_dens,eq_dens],[0,0.1],'k',linewidth=2)
 
 ax1.set_xlabel(r"Density $N/T$")
-ax1.set_ylabel(r"$(\Delta n_j-\Delta n_i)/n_j$")
+ax1.set_ylabel(r"$\Delta n_j/n_j-\Delta n_i/n_i$")
 
 plt.tight_layout()
 

### 2018-03-09 function_definitions.py

@@ -10,11 +10,14 @@ import matplotlib.pyplot as plt
 import bisect
 
 def deltnplussim(m,c,U):
-    scatter=np.zeros([int(U),len(m)])
-    for i in range(len(m)):
-        for y in xrange(int(m[i])):
-            cell=int(int(U)*np.random.rand());
-            scatter[cell,i]=scatter[cell,i]+1;
+#    scatter=np.zeros([int(U),len(m)])
+#    for i in range(len(m)):
+#        for y in xrange(int(m[i])):
+#            cell=int(int(U)*np.random.rand());
+#            scatter[cell,i]=scatter[cell,i]+1;
+
+    l=m/float(U)        
+    scatter=np.random.poisson(lam=l,size=[U,len(m)])
     
     winsnocomp=np.zeros(len(m)); wins1=np.zeros(len(m)); wins2=np.zeros(len(m));
     comp=np.zeros(U);

### 2018-03-09 simulation_comparison.py

@@ -7,33 +7,33 @@ Created on Wed Nov 29 12:26:08 2017
 
 #Simulation comparison figure
 #================================================================
-m=np.array([10000,90000])
-M=float(sum(m)) 
 c=np.array([1.5,1.])
-cbar=sum(m*c)/sum(m)
-maxl=2; Us=np.array([int(m[0]/x) for x in np.linspace(0.001,maxl,1000)]);
-exact=np.array([deltnplussim(m,c,U) for U in Us]);
-totalclassic=np.array([deltnplusclassic(m,c,U) for U in Us]);
-totalclassicboundedm=np.array([deltnplusclassicboundedm(m,c,U) for U in Us]);
+cbar=sum(np.array([0.1,0.9])*c)
+U=100000
+ls=np.array([[x*0.1,x] for x in np.linspace(0.01,20,1000)])
+
+exact=np.array([deltnplussim(m,c,U) for m in U*ls]);
+totalclassic=np.array([deltnplusclassic(m,c,U) for m in U*ls]);
+#totalclassicboundedm=np.array([deltnplusclassicboundedm(m,c,U) for m in U*ls]);
 
 fig1, (ax1, ax2) = plt.subplots(ncols=2,figsize=[8,4])
 
 #First genotype
-nocomp=np.array([np.exp(-M/U) for U in Us]);
-comp1=np.array([R(m,c,U)[0]*c[0]/cbar for U in Us]);
-comp2=np.array([A(m,c,U)[0]*c[0]/cbar for U in Us]);
+nocomp=np.array([np.exp(-np.sum(m)/U) for m in U*ls]);
+comp1=np.array([R(m,c,U)[0]*c[0]/cbar for m in U*ls]);
+comp2=np.array([A(m,c,U)[0]*c[0]/cbar for m in U*ls]);
 
-ax1.plot(float(m[0])/Us,nocomp,'k',label=r"$e^{-L}$")
-ax1.plot(float(m[0])/Us,comp1,'r',label=r"$R_i \frac{c_i}{\overline{c}}$")
-ax1.plot(float(m[0])/Us,comp2,'b',label=r"$A_i \frac{c_i}{\overline{c}}$")
[…]

### 2018-03-09 manuscript.tex
l

@@
 
-We now outline the properties of selection in the variable-density lottery, starting with selection on competitive ability $c$.
-Eq.~\eqref{eq:Nmaster} shows that $c$ is not a density-regulating trait: $c$ only affects the relative likelihood for each type to win a contested territory, not whether a territory is contested in the first place.
-Selection between types which only differ in $c$ occurs without changing density $N$.
+We now outline the basic properties of selection on $b$, $c$ and $d$ in the variable-density lottery.
+
+Eq.~\eqref{eq:Nmaster} shows that $c$ is not a density-regulating trait: $c$ only affects the relative likelihood for each type to win a contested territory, not the number of territories that are contested in the first place.
+Thus, selection between types which only differ in $c$ occurs without causing $N$ to change.
 However, $c$-selection is density-dependent, with the strength of selection (measured by the difference between types in the per-capita growth rate $\Delta n_i/n_i$) peaking at an intermediate density (Fig.~\ref{fig:DDS_lottery}).
-This intermediate peak occurs because most territories are claimed without contest at low density, whereas few unoccupied territories are available to be contested at high density.
+This intermediate peak occurs because at low density most territories are claimed without contest, whereas at high density few unoccupied territories are available to be contested.
 
-Turning to selection on the density-regulating traits $b$ and $d$, we write Eq.~\eqref{eq:masterequalc} in the form
+The birth and mortality rates $b$ and $d$ do regulate density. To see how selection on $b$ and $d$ depends on density, we write Eq.~\eqref{eq:masterequalc} in the form
 \begin{equation}
@@
 It is clear from Eq.~\eqref{eq:bdensitydependence} that selection on adult mortality $d$ is independent of density.
-On the other hand, $b$-selection is density-dependent.
-The advantage of having greater $b$ declines with density because fewer territories are available to to be claimed (Fig.~\ref{fig:DDS_lottery}).
+On the other hand, $b$-selection is density-dependent; the advantage of having greater $b$ declines with density because fewer territories are available to to be claimed (Fig.~\ref{fig:DDS_lottery}).
 
@@
 \includegraphics[scale=0.8]{DDS_lottery.pdf}
-\caption{\label{fig:DDS_lottery} The density-dependence of selection in our variable-density lottery plotted as difference in propagule success rate $\Delta n_j/m_j-\Delta n_i/n_i$ between an adaptive variant $j$ and a wildtype $i$ present at the same frequency.
+\caption{\label{fig:DDS_lottery} The density-dependence of selection in our variable-density lottery plotted as the difference in propagule success rate $\Delta n_j/m_j-\Delta n_i/n_i$ between an adaptive variant $j$ and a wildtype $i$ present at the same frequency.
 Here $b_i=1$, $d_i=0.5$ and $c_i=1$.
@@
 
-We now start evaluating the limits of Eq.~\eqref{eq:canonical} in the context of the variable-density lottery model, focusing in this section on the issue of whether density varies as a result of selection \citep{prout_1980}.
-We showed in the previous section that $c$-selection in the variable-density lottery does not affect density.
-However, this is not the case in most previous models of density-dependent selection, which have been strongly influenced by the notion of ``$K$-selection'' and the associated questions about whether selection in crowded populations will tend to increase population density \citep{anderson_1971,roughgarden_1979,lande_2009}.
-To clarify the connection between ``$K$-selection'' and the variable density lottery model, we thus revisit MacArthur's analysis of selection in crowded populations \citep{macarthur_1967}.
+Equipped with an understanding of selection in the variable-density lottery model, we now turn to the broader issue of how density responds to selection \citep{prout_1980}.
+Going back to some of the earliest work on the topic \citep{kostitzin_1939,christiansen_2004}, most models of selection in density-regulated populations have some form of ``$K$-selection'' in the sense that selection in crowded populations increases saturation density.
+This makes sense intuitively because births balance deaths at equilibrium, and adaptive mutants must overcome the existing balance to expand.
+Yet, as we saw in the previous section, $c$-selection has no effect on population density in the variable-density lottery model.
+To make sense of how this property of $c$-selection fits with more familar population growth models, we revisit MacArthur's analysis of $K$-selection \citep{macarthur_1967}.
 
-MacArthur considers a population with two types that have densities $n_1$ and $n_2$ subject to density-dependent growth described by
+MacArthur considered a population with two types that have densities $n_1$ and $n_2$ subject to density-dependent growth,
 \begin{equation}
@@
 \end{equation}
-The environment is assumed to remain constant apart from the type densities.
-The functions $f_1$ and $f_2$ must decline to zero if $n_1$ or $n_2$ are sufficiently large, because no population has unlimited resources.
+The environment is assumed to remain constant apart from changing type densities.
+The functions $f_1$ and $f_2$ must decline to zero if $n_1$ or $n_2$ are sufficiently large, because the resources required for growth are limited.
 This defines nullclines $f_1(n_1,n_2)=0$ and $f_2(n_1,n_2)=0$ in $(n_1,n_2)$ space.
@@
 Specifically, a type will be excluded if its nullcline is completely contained in the region bounded by the other type's nullcline.
-In other words, for a type to have the possibility of persisting, it must be able to keep growing to higher densities than the other type can tolerate in some region of $(n_1,n_2)$ space (Fig.~\ref{fig:Ksel}a).
+Thus, for a type to have the possibility of persisting, it must be able to keep growing to higher densities than the other type can tolerate in some region of $(n_1,n_2)$ space (Fig.~\ref{fig:Ksel}a).
 
-To formalize the relationship between nullclines, MacArthur used the symbol ``$K$'' to label the four intersection points of the nullclines with the $n_1$ and $n_2$ axes, specifically $f_1(K_{11},0)=0$, $f_1(0,K_{12})=0$, $f_2(0,K_{22})=0$ and $f_2(K_{21},0)=0$.
+MacArthur used ``$K$'' to label the four intersection points of the nullclines with the $n_1$ and $n_2$ axes, specifically $f_1(K_{11},0)=0$, $f_1(0,K_{12})=0$, $f_2(K_{21},0)=0$ and $f_2(0,K_{22})=0$.
 These $K$ values determine whether a region of higher-density growth exists for each type, provided that the nullclines are close to being straight lines.
-Note that only $K_{11}$ and $K_{22}$ are saturation densities akin to the $K$ parameter in the logistic model; following widespread convention, we will refer to selection on these saturation densities as ``$K$-selection'' (Fig.~\ref{fig:Ksel}a).
+Note that only $K_{11}$ and $K_{22}$ are saturation densities akin to the $K$ parameter in the logistic model (Fig.~\ref{fig:Ksel}a).
 The other intersection points, $K_{12}$ and $K_{21}$, are related to competition between types.
-For instance, in the Lotka-Volterra competition model we have
+To be more concrete, in the Lotka-Volterra competition model we have
 \begin{align}
@@
 \end{align}
-where $\alpha_{11}=1/K_{11}$ and $\alpha_{22}=1/K_{22}$ measure competitive effects within types, while $\alpha_{12}=1/K_{12}$ and $\alpha_{21}=1/K_{21}$ measure competitive effects between types (Fig.~\ref{fig:LVvslottery}a).
-
-In summary, MacArthur's conclusion that  ``fitness is $K$'' in crowded populations \citep[pp.
-149]{macarthur_1967} means that selection either favors the ability to keep growing at ever higher densities (moving a type's own nullcline outwards), or the ability to suppress the growth of competitors at lower densities (moving the nullcline of competitors inwards) \citep{gill_1974}.
-This general idea is much broader than ``$K$-selection'' in the sense of selection for greater saturation density, and applies even if the nullclines are nonlinear to such an extent that the ``$K$'' values themselves do not give much information about the regions of high-density growth.
+where $\alpha_{11}=1/K_{11}$ and $\alpha_{22}=1/K_{22}$ measure competitive effects within types, while $\alpha_{12}=1/K_{12}$ and $\alpha_{21}=1/K_{21}$ measure competitive effects between types.
+Hence, ``fitness is $K$'' in crowded populations \citep[pp.
+149]{macarthur_1967} in the sense that selection either favors the ability to keep growing at ever higher densities (moving a type's own nullcline outwards), or the ability to suppress the growth of competitors at lower densities (moving the nullcline of competitors inwards) \citep{gill_1974}.
+This general idea is much broader than ``$K$-selection'' in the sense of selection for greater saturation density.
 
@@
 
-It is obvious in the Lotka-Volterra model that selection can favor a superior competitor in a crowded population even if its saturation density is lower than that of the other types present.
+The Lotka-Volterra competition model makes it is clear that selection in crowded populations need not favor ever greater saturation density \citep{gill_1974}.
+For instance, Fig.~\ref{fig:LVvslottery}a shows Lotka-Volterra selection between two types with the same saturation density ($\alpha_{11}=\alpha_{22}$, $\alpha_{21}>\alpha_{12}$).
+
 Nevertheless, the Lotka-Volterra model still couples population density to selection \citep{smouse_1976}.
-Fig.~\ref{fig:LVvslottery}a shows Lotka-Volterra selection between two types with the same saturation density ($\alpha_{11}=\alpha_{22}$, $\alpha_{21}>\alpha_{12}$).
 Even though the initial and final densities of a sweep are the same, density is not constant over a sweep (constant density is possible but only occurs for a highly restricted subset of $r$ and $\alpha$ values; further details in Appendix C; also see \citealt{mallet_2012}).
 Intuitively, for one type to exclude another with the same saturation density, competitive suppression of growth between types must be stronger than competitive suppression of growth within types, causing a dip in $N$ over the sweep.
+
+(Fig.~\ref{fig:LVvslottery}a)
 

### 2018-03-09 manuscript.tex
l

@@
 
-Selection is commonly described in terms of relative fitness.
-Yet when selection is strong, the ecological view of selection in density-regulated populations seems to be incompatible with widely-used, constant-density relative fitness models such as the Wright-Fisher.
-Here we analyze the population ecological limits of relative fitness using a novel generalization of the Wright-Fisher model in which population density depends dynamically on the demographic rates of the types present.
+Selection is commonly described in terms of constant relative fitnesses.
+Yet when selection is strong, the ecological view of selection in density-regulated populations seems to be incompatible with constant-density relative fitnesses.
+Here we analyze the population ecological limits of relative fitness using a novel [BLAH] in which population density depends dynamically on the demographic rates of the types present.
 Our model contains a ``reproductive excess’’, and clearly distinguishes between density-dependent selection and selection-dependent density.
@@
 There are a variety of different measures of fitness.
-Some widely used examples are expected lifetime reproductive ratio $R_0$, intrinsic population growth rate $r$, saturation population density/carrying capacity (often labeled ``$K$'') \citep{benton_2000}, and invasion fitness \citep{metz_1992}.
+Some widely used examples are expected lifetime reproductive ratio $R_0$, intrinsic population growth rate $r$, equilibrium population density/carrying capacity (often labeled ``$K$'') \citep{benton_2000}, and invasion fitness \citep{metz_1992}.
 In addition, ``relative fitness'' is the standard in much of evolutionary biology, particularly evolutionary genetics, where the focus is relative genotypic proportions \cite[pp.
@@
 However, MacArthur famously argued that when population growth is density-regulated, selection in crowded populations is intimately connected to the ability to keep growing at higher densities than other types can tolerate \citep{macarthur_1967}.
-The classic example is the logistic model, where the type with the greatest saturation population density ``$K$'' excludes the others (Fig.~\ref{fig:Ksel}a).
+The classic example is the logistic model, where the type with the greatest equilibrium population density ``$K$'' excludes the others (Fig.~\ref{fig:Ksel}a).
 Similarly, the ``$R^*$ rule'', a central tenet of resource competition theory, states that when growth is limited by a single homogeneous consumable resource, the type able to deplete the resource to the lowest equilibrium density $R^*$ excludes the others \citep{grover_1997}.
-Differences in $R^*$ will often entail differences in saturation density.
+Differences in $R^*$ will often entail differences in equilibrium density.
 The Lotka-Volterra competition model also couples selection in crowded populations to density except in special cases \citep{smouse_1976,mallet_2012}.
@@
 Rather than attempting to make sense of relative fitness in existing standard models of population growth (e.g.
-\citep{kimura1969natural,mallet_2012}), we instead do the reverse, and attempt to make population ecological sense of the widely-used Wright-Fisher relative-fitness model.
-
+\citep{kimura1969natural,mallet_2012}), we instead do the reverse, and attempt to make population ecological sense [BLAH].
 Our starting point is the classic lottery model of territorial contest \citep{sale_77,chesson_1981}.
 The classic lottery assumes a saturated population with constant $N$, and fitness involves a product of fertility and juvenile viability \citep[pp.
-185]{crow_1970}, but unlike the Wright-Fisher model, generations can overlap.
-
-Our first task is to generalize the lottery model to create a variable-density version of the Wright-Fisher model with overlapping generations (sections ``Model'' and ``Analytical approximation of the variable-density lottery'').
+185]{crow_1970}, [BLAH] generations can overlap.
+Our first task is to generalize the lottery model to create a variable-density version [BLAH] (sections ``Model'' and ``Analytical approximation of the variable-density lottery'').
 
@@
 We assume that each adult from type $i$ produces $b_i$ propagules on average, so that the mean number of $i$ propagules dispersing to unoccupied territories is $m_i=b_in_iU/T$.
-The parameter $b_i$ can be thought of as a measure of ``colonization ability'', which combines fecundity and dispersal ability \citep{levins_71,tilman_94,bolker_99}.
+The parameter $b_i$ can be thought of as a measure of ``colonization ability'', which combines fecundity and dispersal ability \citep{levins_71,tilman_94}.
 Random dispersal is then modeled using a Poisson distribution $p_i(x_i)=l_i^{x_i} e^{-l_i}/x_i!$ for the number $x_i$ of $i$ propagules dispersing to any particular unoccupied territory, where $l_i=m_i/U$ is the mean propagule density in unoccupied territories.
@@
 
-\subsection*{Analytical approximation of the variable-density lottery}
+\subsection*{Solving the variable-density lottery}
 
-Eq.~\eqref{eq:growthsumuncoupled} involves an expectation over the time-dependent dispersal distributions $p_i$.
-Here we evaluate this expectation to obtain a better intuition about the dynamics of density-dependent lottery competition.
-
-Similarly to the high-$l_i$ approximation of the classic lottery model, we replace the $x_i$ in Eq.~\eqref{eq:growthsumuncoupled} with ``effective'' mean values, although we cannot simply use the means $l_i$ as in the classic lottery.
-For a rare type, growth comes almost entirely from territories with $x_i=1$, for which its mean density $l_i\ll 1$ is not representative.
-We therefore separate Eq.~\eqref{eq:growthsumuncoupled} into $x_i=1$ and $x_i>1$ components.
-Our more general approximation only requires that there are no large discrepancies in competitive ability (i.e.
+Here we evaluate the expectation in Eq.~\eqref{eq:growthsumuncoupled} to better understand the dynamics of density-dependent lottery competition.
+Similarly to the classic lottery model, we replace the $x_i$, which take different values in different territories, with ``effective'' mean values.
+However, since we want to allow for low propagule densities, we cannot simply replace the $x_i$ with the means $l_i$ as in the classic lottery.
+For a low density type, growth comes almost entirely from territories with $x_i=1$, for which its mean density $l_i\ll 1$ is not representative.
+We therefore separate Eq.~\eqref{eq:growthsumuncoupled} into $x_i=1$ and $x_i>1$ components, taking care to ensure that the effective mean approximations for these components are consistent with each other (details in Appendix B).
+The resulting variable-density approximation only requires that there are no large discrepancies in competitive ability (i.e.
 we do not have $c_i/c_j\gg 1$ for any two types).
-We obtain (details in Appendix B)
+We obtain
 \begin{equation}
@@
 
-\subsection*{Selection in the variable-density lottery}
+\subsection*{Density-dependent selection in the variable-density lottery}
 
-We now outline the basic properties of selection on $b$, $c$ and $d$ in the variable-density lottery.
+We now outline the basic properties of selection on $b$, $c$ and $d$.
+The birth and mortality rates $b$ and $d$ are the traits which regulate density; $b$ controls the fraction of unoccupied territories that are contested, while $d$ controls adult mortality.
+Competitive ability $c$ does not regulate density since it only affects the relative likelihood for each type to win a contested territory.
+Thus, selection between types which only differ in $c$ occurs without causing $N$ to change (Eq.~\eqref{eq:Nmaster} shows this formally).
 
-Eq.~\eqref{eq:Nmaster} shows that $c$ is not a density-regulating trait: $c$ only affects the relative likelihood for each type to win a contested territory, not the number of territories that are contested in the first place.
-Thus, selection between types which only differ in $c$ occurs without causing $N$ to change.
-However, $c$-selection is density-dependent, with the strength of selection (measured by the difference between types in the per-capita growth rate $\Delta n_i/n_i$) peaking at an intermediate density (Fig.~\ref{fig:DDS_lottery}).
+Selection in the variable density lottery model is density-dependent, by which we mean that  the selection coefficient, measured by the difference in per-capita growth rate $\Delta n_i/n_i$ between types, depends on $N$.
+Density-dependent selection is sometimes taken to mean a qualitative change in which types are fitter than others at different densities \citep{travis_2013}.
+While reversal in the order of fitnesses and co-existence driven by density-regulation are possible in the variable-density lottery (a special case of the competition-colonization trade-off; \citealt{levins_71,tilman_94,bolker_99}), questions related to co-existence are tangential to our aims and will not be pursued further here.
+
+Selection on $c$ is density-dependent, with the strength of selection  peaking at an intermediate density (Fig.~\ref{fig:DDS_lottery}).
 This intermediate peak occurs because at low density most territories are claimed without contest, whereas at high density few unoccupied territories are available to be contested.
-
-The birth and mortality rates $b$ and $d$ do regulate density. To see how selection on $b$ and $d$ depends on density, we write Eq.~\eqref{eq:masterequalc} in the form
+To see how selection on $b$ and $d$ depend on density, we write Eq.~\eqref{eq:masterequalc} in the alternative form
 \begin{equation}
@@
 where we have used that fact that $L=\overline{b}N/T$, and $\overline{b}$ is the population mean $b$.
-It is clear from Eq.~\eqref{eq:bdensitydependence} that selection on adult mortality $d$ is independent of density.
-On the other hand, $b$-selection is density-dependent; the advantage of having greater $b$ declines with density because fewer territories are available to to be claimed (Fig.~\ref{fig:DDS_lottery}).
+It is clear $d$-selection is independent of density.
+On the other hand, the strength of $b$-selection declines with density due to the presence of the $T-N$ factor; in words, the advantage of having greater $b$ gets smaller the fewer territories there are to be claimed (Fig.~\ref{fig:DDS_lottery}).
 
@@
 
-\subsection*{Selection-dependent density and $K$-selection}
+\subsection*{The response of density to selection; $c$-selection versus $K$-selection}
 
-Equipped with an understanding of selection in the variable-density lottery model, we now turn to the broader issue of how density responds to selection \citep{prout_1980}.
-Going back to some of the earliest work on the topic \citep{kostitzin_1939,christiansen_2004}, most models of selection in density-regulated populations have some form of ``$K$-selection'' in the sense that selection in crowded populations increases saturation density.
-This makes sense intuitively because births balance deaths at equilibrium, and adaptive mutants must overcome the existing balance to expand.
-Yet, as we saw in the previous section, $c$-selection has no effect on population density in the variable-density lottery model.
-To make sense of how this property of $c$-selection fits with more familar population growth models, we revisit MacArthur's analysis of $K$-selection \citep{macarthur_1967}.
+We now turn to the issue of how density responds to selection, comparing the variable-density lottery and previous models of selection in density-regulated populations  \citep{prout_1980}.
+Starting with some of the earliest work on the topic \citep{kostitzin_1939,christiansen_2004}, most previous models exhibit some form of ``$K$-selection'' i.e.
+selection in crowded populations increases equilibrium density.
+This idea is intuitively appealing because births balance deaths at equilibrium, and adaptive mutants must overcome the existing balance to expand \citep[pp.
+227]{kostitzin_1939}.
+Yet, as we saw in the previous section, $c$-selection has no effect on population density in the variable-density lottery.
+To make sense of how $c$-selection fits with previous population growth models, we now revisit MacArthur's general treatment of $K$-selection \citep{macarthur_1967}.
 
@@
 
-MacArthur used ``$K$'' to label the four intersection points of the nullclines with the $n_1$ and $n_2$ axes, specifically $f_1(K_{11},0)=0$, $f_1(0,K_{12})=0$, $f_2(K_{21},0)=0$ and $f_2(0,K_{22})=0$.
+MacArthur used ``$K$'' to label the four intersection points of the nullclines with the axes, specifically $f_1(K_{11},0)=0$, $f_1(0,K_{12})=0$, $f_2(K_{21},0)=0$ and $f_2(0,K_{22})=0$.
 These $K$ values determine whether a region of higher-density growth exists for each type, provided that the nullclines are close to being straight lines.
-Note that only $K_{11}$ and $K_{22}$ are saturation densities akin to the $K$ parameter in the logistic model (Fig.~\ref{fig:Ksel}a).
+Note that only $K_{11}$ and $K_{22}$ are equilibrium densities akin to the $K$ parameter in the logistic model (Fig.~\ref{fig:Ksel}a).
 The other intersection points, $K_{12}$ and $K_{21}$, are related to competition between types.
@@
 Hence, ``fitness is $K$'' in crowded populations \citep[pp.
-149]{macarthur_1967} in the sense that selection either favors the ability to keep growing at ever higher densities (moving a type's own nullcline outwards), or the ability to suppress the growth of competitors at lower densities (moving the nullcline of competitors inwards) \citep{gill_1974}.
-This general idea is much broader than ``$K$-selection'' in the sense of selection for greater saturation density.
+149]{macarthur_1967} in the sense that selection either favors the ability to keep growing at ever higher densities (moving a type's own nullcline outwards), or the ability to suppress the growth of competitors at lower densities (moving the nullcline of competitors inwards).
+This general idea is much broader than selection for greater equilibrium density \citep{gill_1974}.
 
@@
 \includegraphics[scale=0.8]{LVvslottery.pdf}
-\caption{\label{fig:LVvslottery} Selection between types with identical saturation density but different inter-type competitive ability.
+\caption{\label{fig:LVvslottery} Selection between types with identical equilibrium density but different inter-type competitive ability.
 (a) Lotka-Volterra competition (Eq.~\ref{eq:LV}) with $r_1=r_2=1$, $\alpha_{11}=\alpha_{22}=1$, $\alpha_{12}=0.9$ and $\alpha_{21}=1.2$.
@@
 
-The Lotka-Volterra competition model makes it is clear that selection in crowded populations need not favor ever greater saturation density \citep{gill_1974}.
-For instance, Fig.~\ref{fig:LVvslottery}a shows Lotka-Volterra selection between two types with the same saturation density ($\alpha_{11}=\alpha_{22}$, $\alpha_{21}>\alpha_{12}$).
+Compared to simple birth-death models \citep{kostitzin_1939} or variants of the logistic \citep{roughgarden_1979}, the Lotka-Volterra model clearly distinguishes between intra- and inter-type competitive effects.
+Thus, one type can displace another without having a greater equilibrium density (Fig.~\ref{fig:LVvslottery}a).
+Nevertheless, selection drives transient changes in density in the Lotka-Volterra model even if the initial and final densities of a sweep are the same (constant density only occurs for a highly restricted subset of $r$ and $\alpha$ values; further details in Appendix C; also see \citealt{mallet_2012,smouse_1976}).
+Intuitively, for one type to exclude the other, competitive suppression of growth between types must be stronger than competitive suppression of growth within types, causing $N$ to dip over a sweep (Fig.~\ref{fig:LVvslottery}a).
 
-Nevertheless, the Lotka-Volterra model still couples population density to selection \citep{smouse_1976}.
-Even though the initial and final densities of a sweep are the same, density is not constant over a sweep (constant density is possible but only occurs for a highly restricted subset of $r$ and $\alpha$ values; further details in Appendix C; also see \citealt{mallet_2012}).
-Intuitively, for one type to exclude another with the same saturation density, competitive suppression of growth between types must be stronger than competitive suppression of growth within types, causing a dip in $N$ over the sweep.
+By contrast, density trajectories for $c$-selection in the variable-density lottery converge on a line of constant equilibrium density (Fig.~\ref{fig:LVvslottery}b).
+This means that once the population reaches demographic equilibrium, it behaves indistinguishably from a constant-$N$ relative fitness model (Fig.~\ref{fig:Ksel}b).
+This complete uncoupling of density from $c$-selection arises due to the presence of an excess of propagules which pay the cost of selection without affecting adult density.
+As a result, Eq.~\eqref{eq:canonical} holds in equilibrium even though $c$-selection is density-dependent.
+We discuss the non-equilibrium case in [BLAH].
 
-(Fig.~\ref{fig:LVvslottery}a)
+\subsection*{Density-dependent selection on density-regulating traits}
 
-By contrast, density trajectories in the variable-density lottery model converge on the saturation density line (Fig.~\ref{fig:LVvslottery}b).
-Selection then occurs along this line, similarly to Fig.~\ref{fig:Ksel}b.
-In other words, once the population reaches demographic equilibrium, it behaves indistinguishably from a constant-$N$ relative fitness model.
-This behavior occurs because $c$ determines relative competitive success in territorial contests.
-As such, it is a manifestation of a reproductive excess of propagules.
-Previous models of selection-independent density have either relied on unusual models of competition \citep{kimura1969natural,nei1971fertility}, or made restrictive Lotka-Volterra parameter choices (Appendix C; \citealt{smouse_1976,mallet_2012}).
+The previous section underscores the fact that density must be affected by ongoing selection to threaten Eq.~\eqref{eq:canonical}.
+We now turn to selection on density-regulating traits such as $b$ and $d$ in the variable density lottery.
 
-\subsection*{Density-regulating traits and the threat of strong selection}
+As shown in Fig.~\ref{fig:DDS_lottery}, $d$-selection is independent of density; the selection coefficient for a $d$-variant with $d_j= d_i(1-\epsilon)$ is a constant $s=\epsilon d_i$.
+Thus, to threaten the constant-$s$ assumption in an equilibrium population we require selection to act on a trait that is both density-dependent and density-regulating.
 
-In the previous section we showed that $c$-selection and the regulation of population density are independent even though population growth is density-regulated in our variable-density lottery.
-Nevertheless, selection and density regulation \textit{are} intimately connected in widely used models of population growth, as well as for the lottery $b$ and $d$ traits.
-
-To see why this connection potentially poses a threat to relative fitness, consider the simple birth-death model \cite[pp. 20]{kostitzin_1939} \citep{travis_2013} 
+This is the case for $b$ in the variable-density lottery.
+However, before we discuss this relatively complicated trait, it is helpful to summarize the threat to Eq.~\eqref{eq:canonical} in simpler models of density-regulated growth, as exemplified by the birth-death model \citep{kostitzin_1939}
 \begin{equation}
-\frac{d n_i}{dt}=(b_i -\delta_iN) n_i \label{eq:simplebirthdeath}
+\frac{d n_i}{dt}=(b_i -\delta_iN) n_i. \label{eq:simplebirthdeath}
 \end{equation}
-where $\delta_i$ is the per-capita increase in mortality rate due to crowding (for simplicity, there are no deaths when uncrowded), playing a similar role as $1/K$ in the logistic model.
-
-Starting from a monomorphic population, the frequency of a variant with $\delta_j=\delta_i(1-\epsilon)$ obeys 
-\begin{equation}
-\frac{d p_j}{dt}=\epsilon \delta_i N p_j(1-p_j). \label{eq:Ndependentsweep}
-\end{equation}
-The selection coefficient $s=\epsilon \delta_i N$ thus depends on density (compare with Model III in \cite{kimura1969natural}).
-On the other hand, the frequency of a variant with $b_j= b_i(1+\epsilon)$ will exactly obey Eq.~\eqref{eq:canonical} with $s=\epsilon b_i$, independent of density.
-
-In practice the density dependence in Eq.~\eqref{eq:Ndependentsweep} only matters if $N$ changes substantially during a sweep.
-This can easily occur if a population is far from demographic equilibrium (we return to this scenario in the Discussion).
-However, even if $N$ has reached equilibrium, it will change substantially over a $\delta$-sweep if selection on $\delta$ is sufficiently strong.
-To quantify this effect, we need to account for how much $N$ changes as a result of a $\delta$-sweep beginning and ending in equilibrium \citep{kimura1969natural}; from Eq.~\eqref{eq:simplebirthdeath} we have an increase from $N_{\rm initial}=b_i/\delta_i$ to $N_{\rm final}=b_i/(\delta_i(1-\epsilon))=N_{\rm initial}/(1-\epsilon)$.
-The corresponding selection coefficient increases from $s_{\rm initial}= \epsilon b_i$ to $s_{\rm final}=s_{\rm initial}/(1-\epsilon)$.
-Consequently, noticeable deviations from Eq.~\eqref{eq:canonical} occur with proportional changes to $\delta$ of order $\epsilon=0.2$ and upwards i.e.
-selection must be quite strong (Fig.~\ref{fig:strengthofselection}).
+Here $\delta_i$ is per-capita mortality due to crowding (for simplicity, there are no deaths when uncrowded).
+Starting from a type $i$ population in equilibrium (we consider the non-equilibrium case in BLAH), a variant with $\delta_j=\delta_i(1-\epsilon)$ has density-dependent selection coefficient $s=\epsilon \delta_i N$ in Eq.~\eqref{eq:canonical}.
+This only affects the sweep if $N$ changes substantially as the population shifts to the new $j$ type equilibrium.
+From Eq.~\eqref{eq:simplebirthdeath} we have $N_{\rm initial}=b_i/\delta_i$ and $N_{\rm final}=b_i/(\delta_i(1-\epsilon))=N_{\rm initial}/(1-\epsilon)$, and so $s_{\rm initial}= \epsilon b_i$ and $s_{\rm final}=s_{\rm initial}/(1-\epsilon)$.
+Consequently, substantial deviations from Eq.~\eqref{eq:canonical} requires strong selection on $\delta$ (Fig.~\ref{fig:strengthofselection}; \citealt{kimura1969natural,crow_1970}).
 
@@
 \includegraphics[scale=0.8]{strengthofselection.pdf}
-\caption{\label{fig:strengthofselection} (a) Proportional change in the selection coefficient over a ``$K$-like'' sweep for a type that experiences proportionally $1-\epsilon$ fewer deaths induced by crowding.
+\caption{\label{fig:strengthofselection} (a) Proportional change in the selection coefficient for a type that experiences proportionally $1-\epsilon$ fewer deaths induced by crowding.
 The population is in demographic equilibrium at the start and end of the sweep.
-(b) Example equilibrium-to-equilibrium $\delta$-sweep (Eq.~\ref{eq:Ndependentsweep}) for $\epsilon=0.2$ showing a noticeable deviation from the canonical selection equation.}
+(b) Example equilibrium-to-equilibrium sweep.}
 \end{figure}
 
- Thus, $d$ sweeps follow the canonical relative fitness model exactly (Fig.~\ref{fig:DDS_lottery}).
-
 At first glance, $b$ in Eq.~\eqref{eq:bdensitydependence} appears to be analogous to the $\delta$ in Eq.~\eqref{eq:simplebirthdeath} because it regulates density and is multiplied by the density-dependent term $f(\overline{b},N)=\frac{1-e^{-\overline{b}N/T}}{N}(T-N)$.
-
-Nevertheless, the behavior of equilibrium-to-equilibrium $b$-sweeps are qualitatively different from the $\delta$ sweeps above.
-The reason is that $b$ regulates density by controlling how many unoccupied territories receive propagules.
-Thus, greater $b$ means more propagules contesting territories, but also more territories being contested.
-The net effect on $f(\overline{b},N)$ is precisely zero in equilibrium: in a single-type equilibrium we have $b_i/\overline{b}=1$ and so $f(\overline{b},N)=d_i$ exactly at the beginning and end of a pure $b$ sweep, even though the density $N$ increases.
-Strictly speaking there is some deviation in $f(\overline{b},N)$ from $d_i$ during the sweep, but this deviation is an order of magnitude smaller than for a $\delta$ sweep (the deviation due to a sweep with proportional effect $b_j= b_i(1+\epsilon)$ is only of order $\epsilon^2$, whereas the analogous effect in Fig.~\ref{fig:strengthofselection} is of order $\epsilon$; see Appendix D for details).
-Since selection must already be quite strong for a $\delta$ sweep to threaten Eq.~\eqref{eq:canonical}, we conclude that $b$ sweeps also obey the canonical selection equation (to a close approximation).
+However, equilibrium-to-equilibrium $b$-sweeps are qualitatively different from $\delta$ sweeps.
+The reason is that $b$ is not a simple per-capita birth rate.
+Greater $b$ means more propagules contesting territories, but also more territories being contested.
+The net effect on $f(\overline{b},N)$ is precisely zero in equilibrium: in a single-type equilibrium we have $b_i/\overline{b}=1$ and so $f(\overline{b},N)=d_i$ exactly at the beginning and end of a $b$-sweep, even though $N$ increases.
+Strictly speaking there is some deviation in $f(\overline{b},N)$ from $d_i$ during the sweep, but this deviation is an order of magnitude smaller than for a $\delta$ sweep (deviation from Eq.~\eqref{eq:canonical} in Fig.~\ref{fig:strengthofselection} is of order $\epsilon$, whereas the analogous effect for $b$ sweep in the variable-density lottery is only of order $\epsilon^2$; see Appendix D for details).
+Since selection must already be quite strong for a $\delta$-sweep to threaten Eq.~\eqref{eq:canonical}, the constant-$s$ model effectively applies exactly for equilibrium $b$-sweeps.
 
@@
 
+%Previous models of selection-independent density have either relied on unusual models of competition \citep{kimura1969natural,nei1971fertility}, or made restrictive Lotka-Volterra parameter choices (Appendix C; \citealt{smouse_1976,mallet_2012}).
+
 Summarizing the three traits in the variable-density lottery model: (i) $c$-selection is density-dependent, but $c$ does not regulate density; (ii) $d$ regulates density, but $d$-selection is density-independent; (iii) $b$ regulates density and $b$-selection is density-dependent.
-Despite these differences, pure $b$, $c$ and $d$ sweeps starting and ending at equilibrium all obey the canonical selection equation.
 This rich variety of behaviors in relation to density is quite different from that found in the classical density-dependent selection literature \citep{roughgarden_1979,christiansen_2004}.
+Yet despite these differences, pure $b$, $c$ and $d$ sweeps starting and ending at equilibrium all obey the canonical constant-$s$ relative fitness description of selection.
 
-To briefly review: based on a diploid, bi-allelic variant of the logistic model, the $r$/$K$ scheme proposed a dichotomy between $r$-selection (uncrowded) and $K$-selection (crowded) \citep{macarthur_1962}, with the latter taken to mean selection for greater saturation density \citep{gill_1974}.
+To briefly review: based on a diploid, bi-allelic variant of the logistic model, the $r$/$K$ scheme proposed a dichotomy between $r$-selection (uncrowded) and $K$-selection (crowded) \citep{macarthur_1962}, with the latter taken to mean selection for greater equilibrium density \citep{gill_1974}.
 A more general Lotka-Volterra model introduces the inter-type $\alpha_{ij}$ competition coefficients, with selection on these termed ``$\alpha$-selection'' \citep{gill_1974,joshi_2001}.
@@
 By ignoring reproductive excesses, single life-stage models are biased to have total population density be sensitive to ongoing selection.
-In this respect, the Wright-Fisher model and similar viability selection heuristics actually capture an important ecological process.
+In this respect, the viability selection heuristics that are common in population genetics \cite[pp.
+61]{gillespie_2004} actually capture an important ecological process.
 
@@
 In this case there is no choice but to abandon relative fitness, and our model provides one potentially suitable option.
-Whether or not our density-dependent lottery model is a good description of \textit{Drosophila} ecology, the close connection between our model and Wright-Fisher is useful, because drift in our model should behave broadly similarly.
+Whether or not our density-dependent lottery model is a good description of \textit{Drosophila} ecology, the close connection between our model and Wright-Fisher is useful [BLAH], because drift in our model should behave broadly similarly.
 Thus, our model should provide a useful starting point for analyzing evolution in this and other far-from-equilibrium situations.
@@
 \includegraphics[scale=0.8]{approx_details.pdf}
-\caption{\label{fig:approx_details} Comparison of our approximation with simulations and the naive $\tilde{p}$,$\hat{p}$ approxmation as a function of the relative $c$-advantage between two types.
-Our approximation breaks down in the presence of large relative $c$-advantages.
-The inset shows the pathology of the $\tilde{p}$,$\hat{p}$ approxmation --- in the neutral case of $c=1$ one type grows faster than the other.
-}
+\caption{\label{fig:approx_details} Comparison of our $\tilde{q}$,$\hat{q}$ approximation with simulations, and also with the naive $\tilde{p}$,$\hat{p}$ approximation, as a function of the relative $c$ difference between two types.
+Our approximation breaks down in the presence of large $c$ differences.
+The inset shows the pathology of the $\tilde{p}$,$\hat{p}$ approximation --- growth rates are not equal in the neutral case $c=1$.
+Simulation procedure is the same as in Fig.~\ref{fig:simcomp}, with $U=10^5$.}
 \end{figure}
@@
 \eqref{eq:cva} is small for both low and high nonfocal densities.
-Again, the worst case scenario occurs when $l_i$ and $L-l_i$ are of order $1$, but Eq.~\eqref{eq:meanfielda} is still a good approximationin this case.
-Again, the approxmation breaks down in the presence of a rare, strong competitor (Fig.~\ref{fig:approx_details}).
+Again, the worst case scenario occurs when $l_i$ and $L-l_i$ are of order $1$, but Eq.~\eqref{eq:meanfielda} is still a good approximation in this case.
+Again, the approximation breaks down in the presence of a rare, strong competitor (Fig.~\ref{fig:approx_details}).
 
@@
 
-Here we show that under the Lotka-Volterra model of competition, total density $N$ does not in general remain constant over a selective sweep in a crowded population even if the types have the same saturation density (for a related discussion on the density- and frequency-dependence of selection in the Lotka-Volterra model, see \citep{smouse_1976,mallet_2012}).
+Here we show that under the Lotka-Volterra model of competition, total density $N$ does not in general remain constant over a selective sweep in a crowded population even if the types have the same equilibrium density (for a related discussion on the density- and frequency-dependence of selection in the Lotka-Volterra model, see \citep{smouse_1976,mallet_2012}).
 
@@
 \end{equation}
-To get some intuition for Eq.~\eqref{eq:constNcondition}, suppose that a mutant arises with improved competitive ability but identical intrinsic growth rate and saturation density ($r_1=r_2$ and $\alpha_{11}=\alpha_{22}$).
+To get some intuition for Eq.~\eqref{eq:constNcondition}, suppose that a mutant arises with improved competitive ability but identical intrinsic growth rate and equilibrium density ($r_1=r_2$ and $\alpha_{11}=\alpha_{22}$).
 This could represent a mutation to an interference competition trait, for example \citep{gill_1974}.
@@
 \end{equation}
-which depends on frequency and thus cannot be satisfied in general for constant inter-type coefficients $\alpha_{ij}$.
-We conclude that selection in the Lotka-Volterra competition model will generally involve non-constant $N$.
+which depends on frequency and thus cannot be satisfied in general for constant inter-type coefficients $\alpha_{ij}$. Therefore, Lotka-Volterra selection will generally involve non-constant $N$.
 
@@
 By contrast, for a $\delta$ sweep in Eq.~\eqref{eq:simplebirthdeath}, the density-dependent term $N$ increases by a factor of $\frac{1}{1-\epsilon}=1+\epsilon+\epsilon^2+\ldots$.
-Thus,  the deviations in $f(N)$ are an order of magnitude smaller than those shown in Fig.~\eqref{fig:strengthofselection}, and even proportional changes of order $\epsilon=0.1$ will cause a negligible deviation from the canonical selection equation.
+Thus,  the deviations in $f(N)$ are an order of magnitude smaller than those shown in Fig.~\eqref{fig:strengthofselection}.
 

### 2018-03-09 approximation_details.py

@@ -8,6 +8,7 @@ Created on Mon Feb 19 14:37:21 2018
 import numpy as np
 import bisect
 import matplotlib.pyplot as plt
+from mpl_toolkits.axes_grid1.inset_locator import mark_inset
 
 def deltnplusnaive(m,c,U):
     if sum(m)>0 and U>0:
@@ -39,19 +40,18 @@ def wins(c):
     return winsnocomp+wins1+wins2
             
 
-
-T=10000
+T=100000
 cs=np.array([[10.**i,1.] for i in np.linspace(-2,2,101)])
 
 fig1, (ax1, ax2) = plt.subplots(nrows=2,figsize=[5,8])
 
-l1=.1; l2=1.; L=l1+l2
+l1=3.; l2=3.; L=l1+l2
 scatter=np.random.poisson(lam=[l1,l2],size=[T,2])
 m=np.array([l1*T,l2*T])
 
-#deltaovermsim1=np.array(map(wins,cs))/m
-#deltaovermnaive1=np.array(map(lambda x: deltnplusnaive(m,x,T),cs))/m
-#deltaovermapprox1=np.array(map(lambda x: deltnplus(m,x,T),cs))/m
+deltaovermsim1=np.array(map(wins,cs))/m
+deltaovermnaive1=np.array(map(lambda x: deltnplusnaive(m,x,T),cs))/m
+deltaovermapprox1=np.array(map(lambda x: deltnplus(m,x,T),cs))/m
 
 ax1.plot(deltaovermsim1[:,0],'k.',markersize=3,label=r'Simulation')
 ax1.plot(deltaovermsim1[:,1],'k.',markersize=3)
@@ -72,9 +72,9 @@ l1=.1; l2=3.; L=l1+l2
 scatter=np.random.poisson(lam=[l1,l2],size=[T,2])
 m=np.array([l1*T,l2*T])
 
-#deltaovermsim2=np.array(map(wins,cs))/m
-#deltaovermnaive2=np.array(map(lambda x: deltnplusnaive(m,x,T),cs))/m
-#deltaovermapprox2=np.array(map(lambda x: deltnplus(m,x,T),cs))/m
+deltaovermsim2=np.array(map(wins,cs))/m
+deltaovermnaive2=np.array(map(lambda x: deltnplusnaive(m,x,T),cs))/m
+deltaovermapprox2=np.array(map(lambda x: deltnplus(m,x,T),cs))/m
 
[…]

### 2018-03-10 manuscript.tex
latest

@@
 
-\subsection*{Solving the variable-density lottery}
+\subsection*{Analytical approximation of the variable-density lottery}
 
@@
 The vertical axis is per-propagule success rate for all propagules $\Delta_+ n_i/m_i$, and for the three separate components in Eq.~\eqref{eq:master}.
-Simulations are 	conducted as follows: $x_1,\ldots,x_G$ values are sampled $U$ times from Poisson distributions with respective means $l_1,\ldots,l_G$, and the victor in each territory is then decided by random sampling weighted by the lottery win probabilities $c_ix_i/\sum_j c_j x_j$.
-Two types are present, a rare type with $c_1=1.5$, and a common type with $c_2=1$.
+Simulations are 	conducted as follows: $x_i,x_j$ values are sampled $U$ times from Poisson distributions with respective means $l_i,l_j$, and the victor in each territory is then decided by random sampling weighted by the lottery win probabilities $c_ix_i/\sum_i c_i x_i$.
+Two types are present, a rare type with $c_j=1.5$, and a common type with $c_i=1$.
 Simulation points are almost invisible for the common type due to near exact agreement with Eq.~\eqref{eq:master}.
-Dashed lines show the breakdown of the classic lottery model.
-Parameters: $U=10^5$ and $l_1/l_2=0.1$ is kept fixed while varying the total density $L$.}
+Dashed lines show the failure of the classic lottery model at low density.
+Parameters: $U=10^5$ and $l_j/l_i=0.1$.}
 \end{figure}
@@
 It is clear $d$-selection is independent of density.
-On the other hand, the strength of $b$-selection declines with density due to the presence of the $T-N$ factor; in words, the advantage of having greater $b$ gets smaller the fewer territories there are to be claimed (Fig.~\ref{fig:DDS_lottery}).
+On the other hand, the strength of $b$-selection declines with density because the advantage of having greater $b$ gets smaller the fewer territories there are to be claimed (Fig.~\ref{fig:DDS_lottery}).
 
@@
 As a result, Eq.~\eqref{eq:canonical} holds in equilibrium even though $c$-selection is density-dependent.
-We discuss the non-equilibrium case in [BLAH].
 
-\subsection*{Density-dependent selection on density-regulating traits}
+\subsection*{Density-regulating traits at equilibrium}
 
@@
 
-At first glance, $b$ in Eq.~\eqref{eq:bdensitydependence} appears to be analogous to the $\delta$ in Eq.~\eqref{eq:simplebirthdeath} because it regulates density and is multiplied by the density-dependent term $f(\overline{b},N)=\frac{1-e^{-\overline{b}N/T}}{N}(T-N)$.
-However, equilibrium-to-equilibrium $b$-sweeps are qualitatively different from $\delta$ sweeps.
-The reason is that $b$ is not a simple per-capita birth rate.
+In the variable-density lottery, equilibrium-to-equilibrium $b$-sweeps are qualitatively different from $\delta$ sweeps, because $b$ is not a simple birth rate.
 Greater $b$ means more propagules contesting territories, but also more territories being contested.
+The density-dependence of $b$-selection is represented by the term $f(\overline{b},N)=\frac{1-e^{-\overline{b}N/T}}{N}(T-N)$ in Eq.~\eqref{eq:bdensitydependence}.
 The net effect on $f(\overline{b},N)$ is precisely zero in equilibrium: in a single-type equilibrium we have $b_i/\overline{b}=1$ and so $f(\overline{b},N)=d_i$ exactly at the beginning and end of a $b$-sweep, even though $N$ increases.
 Strictly speaking there is some deviation in $f(\overline{b},N)$ from $d_i$ during the sweep, but this deviation is an order of magnitude smaller than for a $\delta$ sweep (deviation from Eq.~\eqref{eq:canonical} in Fig.~\ref{fig:strengthofselection} is of order $\epsilon$, whereas the analogous effect for $b$ sweep in the variable-density lottery is only of order $\epsilon^2$; see Appendix D for details).
-Since selection must already be quite strong for a $\delta$-sweep to threaten Eq.~\eqref{eq:canonical}, the constant-$s$ model effectively applies exactly for equilibrium $b$-sweeps.
+Since selection must already be quite strong for a $\delta$-sweep to threaten Eq.~\eqref{eq:canonical}, the density-independent model effectively applies exactly for equilibrium $b$-sweeps (however, selection is frequency-dependent because of the $1/\overline{b}$ factor).
+
+If selection simultaneously acts on more than one trait in the variable-density lottery, then evolution in a density-regulating trait can drive changes in the strength of selection on a trait subject to density-dependent selection (Fig.~\ref{fig:multiple}).
+This can produce behavior reminiscent of Fig.~\ref{fig:strengthofselection}.
+
+\begin{figure}
+\centering
+\includegraphics[scale=0.8]{multiple.pdf}
+\caption{\label{fig:multiple} Simultaneous selection on $d$ and $c$ as predicted by Eq.~\eqref{eq:master}. Selection is not constant over the sweep.}
+\end{figure}
 

### 2018-03-10 simulation_comparison.py

@@ -34,12 +34,12 @@ ax1.plot(ls[:,0],exact[:,1,0]/(ls[:,0]*U),'k.',markersize=1)
 ax1.plot(ls[:,0],exact[:,2,0]/(ls[:,0]*U),'k.',markersize=1)
 
 ax1.plot(ls[:,0],totalclassic[:,0]/(ls[:,0]*U),'k--',label=r"Classic lottery",linewidth=2)
-ax1.set_xlim([0,maxl])
+ax1.set_xlim([0,2])
 ax1.set_ylim([0,1])
 ax1.xaxis.set_label_coords(0.5, -0.09)
 ax1.yaxis.set_label_coords(-0.1, 0.5)
-ax1.set_xticklabels(['0','','1','','2','','3'])
-ax1.set_xlabel(r"$l_1$",fontsize=14)
+ax1.set_xticklabels([0,'',10,'',20])
+ax1.set_xlabel(r"$L$",fontsize=14)
 ax1.set_ylabel("Success per propagule",fontsize=12)
 
 ax1.annotate('Rare type',xy=(0.65,0.32),xycoords='axes fraction',fontsize=12)
@@ -61,7 +61,9 @@ ax2.set_xlim([0,20])
 ax2.set_ylim([0,1])
 ax2.xaxis.set_label_coords(0.5, -0.09)
 ax2.yaxis.set_label_coords(-0.1, 0.5)
-ax2.set_xlabel(r"$l_2$",fontsize=14)
+ax2.set_xlabel(r"$L$",fontsize=14)
+ax2.set_xticklabels([0,'',10,'',20])
+
 
 ax2.plot(ls[:,1],exact[:,0,1]/(ls[:,1]*U),'k.',markersize=1,label="Simulation")
 ax2.plot(ls[:,1],exact[:,1,1]/(ls[:,1]*U),'k.',markersize=1)
@@ -71,4 +73,4 @@ ax2.annotate('Common type',xy=(0.6,0.32),xycoords='axes fraction',fontsize=12)
 
 plt.tight_layout()
 
-plt.savefig('/home/jbertram/repos/densitydependentlottery/simulationcomparison.pdf',bbox="tight")
+plt.savefig('/home/jason/repos/densitydependentlottery/simulationcomparison.pdf',bbox="tight")

### 2018-03-10 strength_of_selection.py

@@ -51,18 +51,18 @@ ax2.set_ylabel("Frequency",fontsize=14)
 
 plt.tight_layout()
 
-plt.savefig('/home/jbertram/repos/densitydependentlottery/strengthofselection.pdf',bbox="tight")
+plt.savefig('/home/jason/repos/densitydependentlottery/strengthofselection.pdf',bbox="tight")
 
 
-#lottery model
+#lottery multiple traits
 #==========================================================
 
 T=100000
-totaltime=80
-b=np.array([3.,3.])
+totaltime=100
+b=np.array([1.,1.])
 c=np.array([1.,1.])
-d=np.array([1.,1.])
-n0=np.array([49000.,100.])
+d=np.array([.9,.9])
+n0=np.array([1000.,1000.])
 
 n=n0;
 nhist=[n];
@@ -71,55 +71,47 @@ for t in range(totaltime/2):
     n=n+deltnplus(b*n*U/T,c,U)-d*n
     nhist.append(list(n))
 
-#plt.plot(np.sum(nhist,1))
-n0=np.array(n)
-print n0 #equilibrium
+print nhist[-1]
 
-s=0.4
-d=np.array([1.,1.*(1-s)])
+totaltime=80
+b=np.array([1.,1.])
+c=np.array([1.,2.])
+d=np.array([.9,.7])
+n0=np.array([sum(nhist[-1]),10.])
+
 n=n0;
 nhist=[n];
 for t in range(totaltime):
     U=T-sum(n)
     n=n+deltnplus(b*n*U/T,c,U)-d*n
     nhist.append(list(n))
-    
+
 nhist=np.array(nhist)
 Nhist=np.sum(nhist,1)
-print "b absolute: ", nhist[1,1]/nhist[0,1]-1
-print "b relative: ", nhist[1,1]/nhist[0,1]*Nhist[0]/Nhist[1]-1
-print "N change: ", Nhist[-1]/Nhist[0]
 
+fig1, (ax1, ax2) = plt.subplots(ncols=2,figsize=[8,4])
 
-fig1,ax1=plt.subplots()
-ax1.plot(nhist[:,1]/Nhist)
+ax1.plot(nhist[:,1],"k", linewidth=2,label=r"$d_j=0.7$, $c_j=2$")
[…]

### 2018-03-10 manuscript.tex
l

@@
 
+[Needs revising]
 Selection is commonly described in terms of constant relative fitnesses.
 Yet when selection is strong, the ecological view of selection in density-regulated populations seems to be incompatible with constant-density relative fitnesses.
-Here we analyze the population ecological limits of relative fitness using a novel [BLAH] in which population density depends dynamically on the demographic rates of the types present.
-Our model contains a ``reproductive excess’’, and clearly distinguishes between density-dependent selection and selection-dependent density.
-These two effects are confounded in standard models of density-regulated population growth.
-Both effects are necessary, in combination with strong selection, for relative fitness to break down in populations close to demographic equilibrium.
-Remarkably, both effects are not sufficient: we give an example of strong selection on a density-regulating trait subject to density-dependent selection that conforms to the relative fitness description almost exactly.
+Here we analyze the population ecological limits of relative fitness using a novel of density-dependent selection which contains a ``reproductive excess’’.
+Our model clearly distinguishes between density-dependent selection and changes in density driven by selection.
+These two effects are confounded in standard models of density-regulated population growth, but both are necessary, in combination with strong selection, for relative fitness to break down in populations close to demographic equilibrium.
+Remarkably, both effects are not sufficient: we give an example of strong selection on a density-regulating trait subject to density-dependent selection that conforms to the density-independent relative fitness description almost exactly.
 We reiterate the importance of reproductive excesses in many species, which allows even strong selection to have no effect on density.
@@
 It has even been proposed that relative fitness be justified from measure theory, abandoning population biology altogether \citep{wagner_2010}.
-Given the ubiquitous use of relative fitness, it is important that we understand its population ecological basis, both to clarify its domain of applicability, and as part of the broader challenge of synthesizing ecology and evolution.
 
-Constant relative fitness values can be justified as an approximation which holds when selection is sufficiently weak and stable over time.
-Formally, these values are the ``zeroth order'' components of the actual relative fitnesses, which are not constant but depend on density, frequency, age structure, and so on \cite[pp.
-277]{ewens_2004} \citep[Chap.
-4]{charlesworth_1994}.
-Yet strong, temporally-variable selection occurs widely in nature and the lab, including in wild \textit{Drosophila}, where population density also varies by orders of magnitude each seasonal cycle \citep{messer_2016,bergland_14}.
-The question is whether relative fitness can be used when selection is not vanishingly weak.
-In general, age-structured populations that reproduce by outcrossing do not permit strong selection to be represented in terms of type-specific relative-fitness constants \citep[Chap.
-4]{charlesworth_1994}.
-We will therefore restrict our attention to asexual haploids with little or no age structure, where it is easier to evaluate how the success or failure of the relative fitness description is tied to the underlying population ecological assumptions.
+Constant relative fitness values are typically justified as an approximation that holds when selection is sufficiently weak and stable over time (\citealt{kimura1969natural}; \citealt[pp.
+277]{ewens_2004}; \citealt[Chap.
+4]{charlesworth_1994}) and population density is either changing either much faster or much slower than type frequencies \citep[Fig.
+9.5]{otto_2011}.
+Formally, these constant values are the leading terms in an expansion of a potentially complicated relative fitness function which depends on density, frequency, age structure, and so on.
+However, knowing that strong selection can break the relative fitness approximation in toy models like the logistic \citep{kimura1969natural}, or in highly complex age-structured models \citep[Chap.
+4]{charlesworth_1994} leaves important issues unresolved.
+For instance, must we simply abandon relative fitness in important model systems like wild \textit{Drosophila}, where strong seasonally-alternating selection occurs in unison with ``boom-bust'' cycles in population density \citep{messer_2016,bergland_14}?
+And if we do abandon relative fitness, are we compelled to develop customized absolute fitness models that capture the specific ecology of the system in question?
+Being able to assign relative fitness values to genotypes is a powerful idealization.
+It is therefore important that we understand the population ecological limits of this idealization when selection is not weak and demography is not stable, both to clarify its domain of applicability, and as part of the broader challenge of synthesizing ecology and evolution.
 
@@
 
-The simple interpretation of relative fitness as a difference in intrinsic growth rates breaks down when we allow for population crowding.
-Since crowded and uncrowded conditions are so different, we expect that $s$ will often depend on density \citep{travis_2013}.
+This simple interpretation of relative fitness breaks down when we allow for population crowding.
+Since crowded and uncrowded conditions are so different, $s$ will often depend on density \citep{travis_2013}.
 Eq.~\eqref{eq:canonical} is then no longer a complete description of selection --- we would also need to specify a model for how density is changing.
 Note that frequency-dependent selection does not raise similar problems; Eq.~\eqref{eq:canonical} is still a complete description of selection even if its behavior is more complicated due to $s$ depending on frequency.
-Population genetics traditionally evades the issue of density-dependent selection by simply assuming that total population density $N$ has reached its equilibrium value, which is assumed to be a fixed constant.
-The selection coefficient $s$ now abstractly parameterizes the rate at which selection changes relative frequencies, and no longer corresponds to differences in intrinsic growth rates $r$.
+This issue does not arise in many population genetics models because total population density $N$ is assumed to have reached some fixed equilibrium value.
+The selection coefficient $s$ then abstractly parameterizes the rate at which selection changes relative frequencies, and no longer corresponds to differences in intrinsic growth rates $r$.
 
-However, MacArthur famously argued that when population growth is density-regulated, selection in crowded populations is intimately connected to the ability to keep growing at higher densities than other types can tolerate \citep{macarthur_1967}.
+Yet MacArthur famously argued that when population growth is density-regulated, selection in crowded populations is intimately connected to the ability to keep growing at higher densities than other types can tolerate \citep{macarthur_1967}.
 The classic example is the logistic model, where the type with the greatest equilibrium population density ``$K$'' excludes the others (Fig.~\ref{fig:Ksel}a).
@@
 In these examples, both $N$ and $s$ change during, and as a result of, adaptive sweeps.
-It would therefore seem that the ubiquitous constant-$N$, relative fitness description of selection is incompatible with a huge class of population ecological processes driving selection (Fig.~\ref{fig:Ksel}b), even in the absence of age-structure and mating.
+It would therefore seem that the ubiquitous constant-$N$, relative fitness description of selection is incompatible with a huge class of population ecological processes driving selection (Fig.~\ref{fig:Ksel}b).
 
@@
 
-In light of this difficulty, the relative fitness description has been justified in broadly two different ways for crowded populations (we do not discuss Wagner's [\citeyear{wagner_2010}] measure-theoretical justification, which is independent of population biology).
+In light of this, the relative fitness description has been justified in broadly two different ways for crowded populations in the strong-selection, non-equilibrium demography regime.
 The first is to simply assume that selection is density-independent but relax the assumption of constant $N$ by allowing density to change as a result of selective sweeps \citep[pp.
@@
 \citep{roughgarden_1979}).
-
 The second justification, which primarily grew out of a controversy over Haldane's ``cost of selection'', is to appeal to the existence of a ``reproductive excess'' of juveniles that are more fragile than their adult counterparts \citep{turner1968population,kimura1969natural,nei1971fertility}.
 Selection can then be concentrated at the juvenile phase, uncoupling selection from population density at the adult phase unless it is so strong that the reproductive excess is depleted.
-This justifies Eq.~\eqref{eq:canonical} because, for a population in demographic equilibrium, selective sweeps do not affect density, and so the density-dependence of selection does not matter.
+Selective sweeps then need not affect density, and so the density-dependence of selection does not matter.
 Unfortunately this reproductive excess literature is also poorly integrated with population ecology.
-\cite{kimura1969natural} took constant $N$ as a requirement and then derived some variants of the logistic model that satisfy this requirement.
+For instance, \cite{kimura1969natural} simply took constant $N$ as a requirement and then derived some variants of the logistic model that satisfy this requirement.
 \cite{nei1971fertility} proposed a model with an explicit representation of reproductive excess, but used an unusual model of competition based on pair-wise interactions which was only defined for at most two different types.
-As a result, the role of reproductive excesses in justifying Eq.~\eqref{eq:canonical} is still largely verbal.
+As a result, the role of reproductive excesses in justifying the use of relative fitness is still largely verbal.
 
 Here we study the population ecology of relative fitness using a novel model of density-dependent population growth based on territorial contests.
-Rather than attempting to make sense of relative fitness in existing standard models of population growth (e.g.
-\citep{kimura1969natural,mallet_2012}), we instead do the reverse, and attempt to make population ecological sense [BLAH].
-Our starting point is the classic lottery model of territorial contest \citep{sale_77,chesson_1981}.
-The classic lottery assumes a saturated population with constant $N$, and fitness involves a product of fertility and juvenile viability \citep[pp.
-185]{crow_1970}, [BLAH] generations can overlap.
-Our first task is to generalize the lottery model to create a variable-density version [BLAH] (sections ``Model'' and ``Analytical approximation of the variable-density lottery'').
+We restrict our attention to asexual haploids with little or no age structure, since it is then easier to evaluate how the success or failure of the relative fitness description is tied to the underlying population ecological assumptions.
+Our starting point is the classic lottery model which was developed by ecologists to study competition driven by territorial contests in reef fishes and plants \citep{sale_77,chesson_1981}.
+The classic lottery incorporates a reproductive excess, and fitness involves a product of fertility and juvenile viability akin to the population simplest population genetic models of selection (e.g.
+\citealt[pp.
+185]{crow_1970}).
+The classic lottery also assumes constant $N$, and generations can overlap.
+Our first task is to generalize the lottery model to create a variable-density lottery (sections ``Model'' and ``Analytical approximation of the variable-density lottery'').
 
-Equipped with this new model, we turn to the evaluation of Eq.~\eqref{eq:canonical}.
-We first discuss selection on the ability to contest territories, which behaves like a pure constant-$N$, relative fitness trait, and discuss how this fits with MacArthur's analysis of selection in crowded populations (section ``$K$-selection and selection-dependent density'').
-We then consider selection on density-regulating traits (section ``Density-regulating traits and the threat of strong selection''), and conclude by contrasting the classical density-dependent selection literature with our results (``Discussion'').
+Equipped with this model, we turn to the evaluation of relative fitness. [Need to write outline, will do when structure is finalized]
 
@@
 The vertical axis is per-propagule success rate for all propagules $\Delta_+ n_i/m_i$, and for the three separate components in Eq.~\eqref{eq:master}.
-Simulations are 	conducted as follows: $x_i,x_j$ values are sampled $U$ times from Poisson distributions with respective means $l_i,l_j$, and the victor in each territory is then decided by random sampling weighted by the lottery win probabilities $c_ix_i/\sum_i c_i x_i$.
-Two types are present, a rare type with $c_j=1.5$, and a common type with $c_i=1$.
+Two types are present with $c_1=1$, $c_2=1.5$ and $l_2/l_1=0.1$.
+Simulations are conducted as follows: $x_1,x_2$ values are sampled $U=10^5$ times from Poisson distributions with respective means $l_1,l_2$, and the victorious type in each territory is then decided by random sampling weighted by the lottery win probabilities $c_ix_i/(c_1 x_1 + c_2 x_2)$.
 Simulation points are almost invisible for the common type due to near exact agreement with Eq.~\eqref{eq:master}.
-Dashed lines show the failure of the classic lottery model at low density.
-Parameters: $U=10^5$ and $l_j/l_i=0.1$.}
+Dashed lines show the failure of the classic lottery model at low density.}
 \end{figure}
@@
 \includegraphics[scale=0.8]{DDS_lottery.pdf}
-\caption{\label{fig:DDS_lottery} The density-dependence of selection in our variable-density lottery plotted as the difference in propagule success rate $\Delta n_j/m_j-\Delta n_i/n_i$ between an adaptive variant $j$ and a wildtype $i$ present at the same frequency.
-Here $b_i=1$, $d_i=0.5$ and $c_i=1$.
-For $b$-selection we set $b_j=b_i(1+\epsilon)$, and similarly for $c$ and $d$, with $\epsilon=0.1$.
+\caption{\label{fig:DDS_lottery} The density-dependence of selection in our variable-density lottery plotted as the difference in propagule success rate $\Delta n_2/m_2-\Delta n_1/n_1$ between an adaptive variant $2$ and a wildtype $1$ with equal frequencies.
+Here $b_1=1$, $d_1=0.5$ and $c_1=1$.
+For $b$-selection we set $b_2=b_1(1+\epsilon)$, and similarly for $c$ and $d$, with $\epsilon=0.1$.
 $d$-selection is density-independent, $b$-selection gets weaker with lower territorial availability, while $c$-selection initially increases with density as territorial contests become more important, but eventually also declines as  available territories become scarce.
-The equilibrium density for the $i$ type is $\approx 0.4$.}
+The wildtype equilibrium density is $n_1/T\approx 0.4$.}
 \end{figure}
@@
 
-As shown in Fig.~\ref{fig:DDS_lottery}, $d$-selection is independent of density; the selection coefficient for a $d$-variant with $d_j= d_i(1-\epsilon)$ is a constant $s=\epsilon d_i$.
-Thus, to threaten the constant-$s$ assumption in an equilibrium population we require selection to act on a trait that is both density-dependent and density-regulating.
+As shown in Fig.~\ref{fig:DDS_lottery}, $d$-selection is independent of density; the selection coefficient for a $d$-variant with $d_2= d_1(1-\epsilon)$ is a constant $s=\epsilon d_1$.
+Thus, to threaten the assumption of density-independent $s$ in an equilibrium population we require selection to act on a trait that is both density-dependent and density-regulating.
 
@@
 Here $\delta_i$ is per-capita mortality due to crowding (for simplicity, there are no deaths when uncrowded).
-Starting from a type $i$ population in equilibrium (we consider the non-equilibrium case in BLAH), a variant with $\delta_j=\delta_i(1-\epsilon)$ has density-dependent selection coefficient $s=\epsilon \delta_i N$ in Eq.~\eqref{eq:canonical}.
-This only affects the sweep if $N$ changes substantially as the population shifts to the new $j$ type equilibrium.
-From Eq.~\eqref{eq:simplebirthdeath} we have $N_{\rm initial}=b_i/\delta_i$ and $N_{\rm final}=b_i/(\delta_i(1-\epsilon))=N_{\rm initial}/(1-\epsilon)$, and so $s_{\rm initial}= \epsilon b_i$ and $s_{\rm final}=s_{\rm initial}/(1-\epsilon)$.
-Consequently, substantial deviations from Eq.~\eqref{eq:canonical} requires strong selection on $\delta$ (Fig.~\ref{fig:strengthofselection}; \citealt{kimura1969natural,crow_1970}).
+Starting from a type $1$ population in equilibrium (we consider the non-equilibrium case in BLAH), a variant with $\delta_2=\delta_1(1-\epsilon)$ has density-dependent selection coefficient $s=\epsilon \delta_1 N$ in Eq.~\eqref{eq:canonical}.
+This only affects the sweep if $N$ changes substantially as the population shifts to the new equilibrium.
+From Eq.~\eqref{eq:simplebirthdeath} we have $N_{\rm initial}=b_1/\delta_1$ and $N_{\rm final}=b_1/(\delta_1(1-\epsilon))=N_{\rm initial}/(1-\epsilon)$, and so $s_{\rm initial}= \epsilon b_1$ and $s_{\rm final}=s_{\rm initial}/(1-\epsilon)$.
+Consequently, substantial deviations from Eq.~\eqref{eq:canonical} occurs if there is sufficiently strong selection on $\delta$ (Fig.~\ref{fig:strengthofselection}; \citealt{kimura1969natural,crow_1970}).
 
@@
 
-In the variable-density lottery, equilibrium-to-equilibrium $b$-sweeps are qualitatively different from $\delta$ sweeps, because $b$ is not a simple birth rate.
-Greater $b$ means more propagules contesting territories, but also more territories being contested.
-The density-dependence of $b$-selection is represented by the term $f(\overline{b},N)=\frac{1-e^{-\overline{b}N/T}}{N}(T-N)$ in Eq.~\eqref{eq:bdensitydependence}.
-The net effect on $f(\overline{b},N)$ is precisely zero in equilibrium: in a single-type equilibrium we have $b_i/\overline{b}=1$ and so $f(\overline{b},N)=d_i$ exactly at the beginning and end of a $b$-sweep, even though $N$ increases.
-Strictly speaking there is some deviation in $f(\overline{b},N)$ from $d_i$ during the sweep, but this deviation is an order of magnitude smaller than for a $\delta$ sweep (deviation from Eq.~\eqref{eq:canonical} in Fig.~\ref{fig:strengthofselection} is of order $\epsilon$, whereas the analogous effect for $b$ sweep in the variable-density lottery is only of order $\epsilon^2$; see Appendix D for details).
-Since selection must already be quite strong for a $\delta$-sweep to threaten Eq.~\eqref{eq:canonical}, the density-independent model effectively applies exactly for equilibrium $b$-sweeps (however, selection is frequency-dependent because of the $1/\overline{b}$ factor).
+Equilibrium-to-equilibrium $b$-sweeps in the variable-density lottery are qualitatively different from $\delta$ sweeps, because greater $b$ not only means more propagules contesting territories, but also more territories being contested.
+Together, the net density-dependent effect on $b$-selection is zero; in Eq.~\eqref{eq:bdensitydependence}, since $b_i/\overline{b}=1$ in a single-type equilibrium, the density-dependence factor $f(\overline{b},N)=\frac{1-e^{-\overline{b}N/T}}{N}(T-N)$ is exactly equal to the constant mortality rate at the beginning and end of a $b$-sweep, even though $b$ and density change.
+During the sweep there is some deviation in $f(\overline{b},N)$, but this deviation is an order of magnitude smaller than for a $\delta$ sweep (the density-dependent deviation constant $s$ in Fig.~\ref{fig:strengthofselection} is of order $\epsilon$, whereas the analogous effect for $b$ sweep in the variable-density lottery is only of order $\epsilon^2$; see Appendix D for details).
+Since selection must already be strong for a $\delta$-sweep to threaten Eq.~\eqref{eq:canonical}, the density-independent model applies effectively exactly for equilibrium $b$-sweeps.
+Note, however, that the selection coefficient for $b$-sweeps (as  defined by differences in $\Delta n_i/n_i$) does depend on frequency because of the $1/\overline{b}$ factor.
 
-If selection simultaneously acts on more than one trait in the variable-density lottery, then evolution in a density-regulating trait can drive changes in the strength of selection on a trait subject to density-dependent selection (Fig.~\ref{fig:multiple}).
-This can produce behavior reminiscent of Fig.~\ref{fig:strengthofselection}.
+If selection acts simultaneously on more than one trait in the variable-density lottery, then evolution in a density-regulating trait can drive changes in the strength of selection on a trait subject to density-dependent selection (Fig.~\ref{fig:multiple}).
+This can produce behavior analogous to selection on $\delta$ in Fig.~\ref{fig:strengthofselection}.
 
@@
 \includegraphics[scale=0.8]{multiple.pdf}
-\caption{\label{fig:multiple} Simultaneous selection on $d$ and $c$ as predicted by Eq.~\eqref{eq:master}. Selection is not constant over the sweep.}
+\caption{\label{fig:multiple} Simultaneous selection on $d$ and $c$ in the variable-density lottery model as predicted by Eq.~\eqref{eq:master}.
+Selection is not constant over the sweep because $d$ is density-regulating and $c$ is density-dependent.}
 \end{figure}
@@
 Summarizing the three traits in the variable-density lottery model: (i) $c$-selection is density-dependent, but $c$ does not regulate density; (ii) $d$ regulates density, but $d$-selection is density-independent; (iii) $b$ regulates density and $b$-selection is density-dependent.
-This rich variety of behaviors in relation to density is quite different from that found in the classical density-dependent selection literature \citep{roughgarden_1979,christiansen_2004}.
-Yet despite these differences, pure $b$, $c$ and $d$ sweeps starting and ending at equilibrium all obey the canonical constant-$s$ relative fitness description of selection.
+Yet despite these differences, pure $b$, $c$ and $d$ sweeps starting and ending at equilibrium all obey the density-independent relative fitness description of selection almost exactly.
+This behavior is quite different from that found in the classical density-dependent selection literature \citep{roughgarden_1979,christiansen_2004}.
 
@@
 Thus, strong selection is sufficient for relative fitness to break down in the classical view of density-dependent selection.
-Indeed, in the defense of Eq.~\eqref{eq:canonical} given by \cite{kimura1969natural}, it was assumed that $s$ will be a few percent at most.
-While this may be reasonable for adaptive mutations, there is no reason to expect selection on standing variation to be so weak, wild \textit{Drosophila} being an obvious counter-example \citep{bergland_14}.
 
@@
 In this respect, the viability selection heuristics that are common in population genetics \cite[pp.
-61]{gillespie_2004} actually capture an important ecological process.
+61]{gillespie_2004} actually capture an important ecological process without jumping to the full complexity of age-structured models.
 
-We now turn to the breakdown of Eq.~\eqref{eq:canonical}.
-We first discuss the problem shown in Fig.~\ref{fig:strengthofselection}, which occurs when strong selection changes population density and is also density-dependent.
-In the variable-density lottery, this occurs if and only if types differ in more than one trait.
+We now turn to the breakdown of the density-independent relative fitness assumption which occurs when strong selection changes population density and is also density-dependent (Fig.~\ref{fig:strengthofselection}).
+In the variable-density lottery, this occurs if and only if types differ in more than one trait (Fig.~\ref{fig:multiple}).
 The $c$ and $d$ traits represent the two distinct directions in which density and selection interact: selection may depend on density, and density may depend on selection \citep{prout_1980}.
@@
 However, the $b$ trait remarkably demonstrates that the combination is not sufficient, since the density-dependence of $b$-selection disappears over equilibrium-to-equilibrium $b$-sweeps.
-Thus, the simple linear models that have become standard in discussions of density-dependent selection \citep{roughgarden_1979,christiansen_2004,mallet_2012,travis_2013} actually represent a complicated form of the interaction between density and selection, and their parameters confound the underlying issues.
+Thus, the simple differential/difference equations that have become standard in discussions of density-dependent selection \citep{roughgarden_1979,christiansen_2004,mallet_2012,travis_2013} actually represent a confounded form of the interaction between density and selection.
 
-While this is a conceptual reason to be wary of the classical density-dependent selection models, it is not clear how we should expect the trait variation in nature to align.
-For instance, should we expect mutations to generally affect $b$, $c$ and $d$ independently of each other, or pleiotropically such that $\delta$-like selection is prevalent?
+While this is a conceptual reason to be wary of the classical density-dependent selection models, it is not clear what trait variation to expect in nature.
+Should we expect mutations to generally affect $b$, $c$ and $d$ independently of each other, or pleiotropically such that $\delta$-like selection is prevalent?
 In the case of well-mixed indirect exploitation competition for consumable resources, the $R^*$ rule  suggests that $\delta$-like selection will be prevalent.
@@
 
-Relative fitness models truly break down when $N$ is far from equilibrium and selection is density-dependent (as seems likely; \citealt{travis_2013}).
-For example, wild \textit{Drosophila} experience large seasonal boom-bust cycles in population density coupled to strong selection that drives large swings in allele frequency \citep{bergland_14}.
-In this case there is no choice but to abandon relative fitness, and our model provides one potentially suitable option.
-Whether or not our density-dependent lottery model is a good description of \textit{Drosophila} ecology, the close connection between our model and Wright-Fisher is useful [BLAH], because drift in our model should behave broadly similarly.
-Thus, our model should provide a useful starting point for analyzing evolution in this and other far-from-equilibrium situations.
+The above findings underscore that the most serious threat to relative fitness arises from deviations from demographic equilibrium.
+This can dramatically alter frequency trajectories for individual sweeps (e.g.
+Fig.
+9.5 in \cite{otto_2011}; Fig.
+5 in \cite{mallet_2012}), as well as the long-term outcomes of selection in fluctuating environments \citep{lande_2009}.
+This would suggest that in the wild \textit{Drosophila} example mentioned in the Introduction, there is indeed no choice but to abandon relative fitness.
+The variable-density lottery could provide a useful starting point for analyzing evolution in this and other far-from-equilibrium situations.
+One convenient feature of our variable-density lottery (and lottery models in general) is a formal similarity to the Wright-Fisher model, which should facilitate the analysis of ``density-dependent drift''.
 

### 2018-03-10 DDS_lottery.py

@@ -69,7 +69,7 @@ ax1.legend(loc='upper center',prop={'size':11})
 #ax1.plot([eq_dens,eq_dens],[0,0.1],'k',linewidth=2)
 
 ax1.set_xlabel(r"Density $N/T$")
-ax1.set_ylabel(r"$\Delta n_j/n_j-\Delta n_i/n_i$")
+ax1.set_ylabel(r"$\Delta n_2/n_2-\Delta n_1/n_1$")
 
 plt.tight_layout()
 

### 2018-03-10 strength_of_selection.py

@@ -91,8 +91,8 @@ Nhist=np.sum(nhist,1)
 
 fig1, (ax1, ax2) = plt.subplots(ncols=2,figsize=[8,4])
 
-ax1.plot(nhist[:,1],"k", linewidth=2,label=r"$d_j=0.7$, $c_j=2$")
-ax1.plot(nhist[:,0],"k--", linewidth=2,label=r"$d_i=0.9$, $c_i=1$")
+ax1.plot(nhist[:,1],"k", linewidth=2,label=r"$d_2=0.7$, $c_2=2$")
+ax1.plot(nhist[:,0],"k--", linewidth=2,label=r"$d_1=0.9$, $c_1=1$")
 ax1.plot(Nhist,"k:", linewidth=2,label=r"Total")
 ax1.annotate(r'$(a)$',xy=(0.01,0.93),xycoords='axes fraction',fontsize=16)
 ax1.set_xticklabels([])
@@ -109,9 +109,9 @@ ax2.set_xticklabels([])
 ax2.set_xlabel("Time",fontsize=14)
 ax2.set_ylim([0.2,0.25])
 #ax2.set_yticklabels(['0','','','','','1'])
-ax2.set_ylabel(r"$\Delta n_j/n_j-\Delta n_i/n_i$")
+ax2.set_ylabel(r"$\Delta n_2/n_2-\Delta n_1/n_1$")
 
 plt.tight_layout()
 
-plt.savefig('/home/jason/repos/densitydependentlottery/multiple.pdf',bbox="tight")
+plt.savefig('/home/jbertram/repos/densitydependentlottery/multiple.pdf',bbox="tight")
 

### 2018-03-14 manuscript.tex
l

@@
 
-[Needs revising]
-Selection is commonly described in terms of constant relative fitnesses.
+[I'm going to revise this after your next round of comments]
+Selection is commonly described by assigning relative fitness values to genotypes.
 Yet when selection is strong, the ecological view of selection in density-regulated populations seems to be incompatible with constant-density relative fitnesses.
@@
 Some widely used examples are expected lifetime reproductive ratio $R_0$, intrinsic population growth rate $r$, equilibrium population density/carrying capacity (often labeled ``$K$'') \citep{benton_2000}, and invasion fitness \citep{metz_1992}.
-In addition, ``relative fitness'' is the standard in much of evolutionary biology, particularly evolutionary genetics, where the focus is relative genotypic proportions \cite[pp.
+In addition, ``relative fitness'' is the standard in much of evolutionary biology, particularly evolutionary genetics, where the focus is on relative genotypic proportions \cite[pp.
 468]{barton_2007}.
@@
 But it should be clear how these measures are connected to the processes of birth and death which ultimately drive selection \citep{metcalf_2007,doebeli_2017}.
-While such a connection is fairly clear for absolute fitness measures like $r$, relative fitness is largely divorced from population ecology.
+While such a connection is fairly clear for absolute fitness measures like $r$ or $R_0$, relative fitness has only weak justification from population ecology.
 It has even been proposed that relative fitness be justified from measure theory, abandoning population biology altogether \citep{wagner_2010}.
+Given the ubiquitous use of relative fitness, it is important that we understand its population ecological basis, both to clarify its domain of applicability, and as part of the broader challenge of synthesizing ecology and evolution.
 
-Constant relative fitness values are typically justified as an approximation that holds when selection is sufficiently weak and stable over time (\citealt{kimura1969natural}; \citealt[pp.
-277]{ewens_2004}; \citealt[Chap.
-4]{charlesworth_1994}) and population density is either changing either much faster or much slower than type frequencies \citep[Fig.
-9.5]{otto_2011}.
-Formally, these constant values are the leading terms in an expansion of a potentially complicated relative fitness function which depends on density, frequency, age structure, and so on.
-However, knowing that strong selection can break the relative fitness approximation in toy models like the logistic \citep{kimura1969natural}, or in highly complex age-structured models \citep[Chap.
-4]{charlesworth_1994} leaves important issues unresolved.
-For instance, must we simply abandon relative fitness in important model systems like wild \textit{Drosophila}, where strong seasonally-alternating selection occurs in unison with ``boom-bust'' cycles in population density \citep{messer_2016,bergland_14}?
-And if we do abandon relative fitness, are we compelled to develop customized absolute fitness models that capture the specific ecology of the system in question?
-Being able to assign relative fitness values to genotypes is a powerful idealization.
-It is therefore important that we understand the population ecological limits of this idealization when selection is not weak and demography is not stable, both to clarify its domain of applicability, and as part of the broader challenge of synthesizing ecology and evolution.
+For haploids growing in discrete time steps, the change in the abundance $n_i$ of type $i$ over a time step can be expressed as $\Delta n_i=(W_i - 1)n_i$ where $W_i$ is absolute fitness.
+The corresponding change in frequency is $\Delta p_i=\left(\frac{W_i}{\overline{W}}-1\right) p_i$, where $\overline{W}=\sum_i W_i p_i$.
+In continuous time, the Malthusian parameter $r_i$ replaces $W_i$ and we have $\frac{d n_i}{dt}=r_in_i$ and $\frac{d p_i}{dt}=(r_i-\overline{r}) p_i$ \citep[pp.
+26]{crow_1970}.
+In the particular case that two types are present, a mutant labeled $1$ and a wildtype labeled $2$ for instance, then the continuous time equation for the mutant takes the familiar form
+\begin{equation}
+\frac{d p_1}{dt}=s p_1(1-p_1), \label{eq:canonical}
+\end{equation}
+where $s=r_1-r_2$ is the selection coefficient.
+Note that in the discrete case we can replace the $W_i$ with any set of values proportional to the $W_i$ and the changes in frequency $\Delta p_i$ will be unaffected.
+These values, which give no information about the dynamics of total population density $N=\sum_i n_i$, are called ``relative fitnesses'' \citep[pp.
+468]{barton_2007}.
+Similarly, the selection coefficient and frequency dynamics in Eq.~\eqref{eq:canonical} would be unaffected if we were to add an arbitrary constant to both $r_1$ and $r_2$ (these would be relative log fitnesses).
 
-In the absence of crowding, relative fitness simply represents differences in intrinsic population growth rate.
-In discrete time, the change in frequency of type $i$ is $\Delta p_i=\left(\frac{W_i}{\overline{W}}-1\right) p_i$, where $W_i$ is the intrinsic absolute growth factor of type $i$, and $\overline{W}=\sum_i W_i p_i$ is the population mean $W$.
-The change in frequency depends only the relative ratio $\frac{W_i}{\overline{W}}$, not the absolute values of the growth factors $W_i$.
-Thus, we can replace the $W_i$ with any set of values $w_i$ that are proportional to the $W_i$; these values are called ``relative fitnesses''.
-The analogous formula in continuous time is $\frac{d p_i}{dt}=(r_i-\overline{r}) p_i$, where $W_i$ is replaced by the intrinsic exponential growth rate $r_i$ \citep[pp.
-26]{crow_1970}, and the ratio $\frac{W_i}{\overline{W}}$ is replaced by the difference $r_i-\overline{r}$.
-In the particular case that there are two types present, a wildtype $i$ and a mutant $j$ for instance, then the continuous time equation takes the familiar form
-\begin{equation}
-\frac{d p_i}{dt}=s p_i(1-p_i), \label{eq:canonical}
-\end{equation}
-where $s=r_i-r_j$ is the selection coefficient.
+In the absence of crowding, each type grows exponentially at a constant ``instrinsic'' rate $r_i$ (for concreteness, we will primarily use Eq.~\eqref{eq:canonical} as our point of comparison but similar logic applies in the discrete time case).
+The selection coefficient $s$ is then a constant equal to the difference in intrinsic population growth rates.
+The situation is more complicated when we allow for population crowding.
+Since crowded and uncrowded conditions can favor quite different traits \citep{travis_2013}, the Malthusian parameters $r_i$ may respond differently to changes in population density $N$, in which case the selection coefficient will be density-dependent.
+Eq.~\eqref{eq:canonical} would then need to supplemented with a model for the dynamics of $N$ to describe how the allele frequencies will change over time.
+The density and type frequency dynamics will in general be coupled \citep{travis_2013}.
+Thus, in a general crowded population the simple procedure of assigning constant relative fitness values to different types has to be replaced with an ecological description of absolute growth rates --- a substantially more complicated problem.
+Note that frequency-dependent selection does not raise such a severe problem, because we still only need to model the type frequencies, and not also the ecological variable $N$.
 
-This simple interpretation of relative fitness breaks down when we allow for population crowding.
-Since crowded and uncrowded conditions are so different, $s$ will often depend on density \citep{travis_2013}.
-Eq.~\eqref{eq:canonical} is then no longer a complete description of selection --- we would also need to specify a model for how density is changing.
-Note that frequency-dependent selection does not raise similar problems; Eq.~\eqref{eq:canonical} is still a complete description of selection even if its behavior is more complicated due to $s$ depending on frequency.
-This issue does not arise in many population genetics models because total population density $N$ is assumed to have reached some fixed equilibrium value.
-The selection coefficient $s$ then abstractly parameterizes the rate at which selection changes relative frequencies, and no longer corresponds to differences in intrinsic growth rates $r$.
-
-Yet MacArthur famously argued that when population growth is density-regulated, selection in crowded populations is intimately connected to the ability to keep growing at higher densities than other types can tolerate \citep{macarthur_1967}.
+In practice, many population genetics models assume that $N$ has reached some fixed equilibrium value, and each type is assigned a constant relative fitness.
+Selection is typically assumed to operate through viability (e.g.
+\citealt[pp.
+61]{gillespie_2004}), but other important population ecological assumptions --- such as how density is regulated --- are left unspecified.
+Beyond being conceptually unsatisfactory \citep{doebeli_2017}, we might also doubt whether this density-independent-$s$, constant-$N$ setup is consistent with the ecology of many populations.
+Much of the seminal literature on density-dependent selection revolves around the connection between selection in crowded populations and the ability to keep growing at higher densities \citep{macarthur_1967,roughgarden_1979}.
 The classic example is the logistic model, where the type with the greatest equilibrium population density ``$K$'' excludes the others (Fig.~\ref{fig:Ksel}a).
-Similarly, the ``$R^*$ rule'', a central tenet of resource competition theory, states that when growth is limited by a single homogeneous consumable resource, the type able to deplete the resource to the lowest equilibrium density $R^*$ excludes the others \citep{grover_1997}.
-Differences in $R^*$ will often entail differences in equilibrium density.
+Similarly, the ``$R^*$ rule'', a central tenet of resource competition theory, states that when growth is limited by a single homogeneous consumable resource, the type able to deplete the resource to the lowest equilibrium density $R^*$ (which will often entail a different equilibrium population density) excludes the others \citep{grover_1997}.
+In these examples, both $N$ and $s$ change during, and as a result of, adaptive sweeps.
 The Lotka-Volterra competition model also couples selection in crowded populations to density except in special cases \citep{smouse_1976,mallet_2012}.
-In these examples, both $N$ and $s$ change during, and as a result of, adaptive sweeps.
-It would therefore seem that the ubiquitous constant-$N$, relative fitness description of selection is incompatible with a huge class of population ecological processes driving selection (Fig.~\ref{fig:Ksel}b).
+It would therefore seem that the ubiquitous density-independent-$s$, constant-$N$ description of selection is incompatible with a huge class of population ecological processes driving selection (Fig.~\ref{fig:Ksel}b).
 
@@
 
-In light of this, the relative fitness description has been justified in broadly two different ways for crowded populations in the strong-selection, non-equilibrium demography regime.
-The first is to simply assume that selection is density-independent but relax the assumption of constant $N$ by allowing density to change as a result of selective sweeps \citep[pp.
-468]{barton_2007} \citep{prout_1980}.
-Obviously this does not address the problem that $s$ can, in reality, depend on density.
-Type-specific responses to density are at the center of MacArthur's argument and the density-dependent selection literature that grew out of it (e.g.
-\citep{roughgarden_1979}).
-The second justification, which primarily grew out of a controversy over Haldane's ``cost of selection'', is to appeal to the existence of a ``reproductive excess'' of juveniles that are more fragile than their adult counterparts \citep{turner1968population,kimura1969natural,nei1971fertility}.
-Selection can then be concentrated at the juvenile phase, uncoupling selection from population density at the adult phase unless it is so strong that the reproductive excess is depleted.
-Selective sweeps then need not affect density, and so the density-dependence of selection does not matter.
-Unfortunately this reproductive excess literature is also poorly integrated with population ecology.
-For instance, \cite{kimura1969natural} simply took constant $N$ as a requirement and then derived some variants of the logistic model that satisfy this requirement.
+In light of this, and the related difficulties posed by frequency-dependent selection, relative fitness values that are independent of type frequencies and population density are typically justified as an approximation that holds when selection is weak (\citealt{kimura1969natural}; \citealt[pp.
+277]{ewens_2004}; \citealt[Chap.
+4]{charlesworth_1994}) and density is changing much slower than type frequencies \citep[Fig.
+9.5]{otto_2011}.
+Under these conditions, the selection coefficient may depend on density and type frequencies, but since these change slowly the selection coefficient may remain approximately constant over several generations.
+In the case of density-dependence, this time interval of constancy could include entire selective sweeps.
+On the other hand, it is known in both simple models like the logistic \citep{kimura1969natural}, and in complex age-structured models \citep[Chap.
+4]{charlesworth_1994}, that frequency trajectories under strong density- or frequency-dependent selection can differ substantially from the constant-$s$ approximation.
+
+Yet knowing that the constant-$s$ approximation can fail \textit{in general} when selection is strong gives only a rough picture of the domain and manner in which the constant-$s$ approximation breaks down.
+For instance, wild \textit{Drosophila} is an important model system in genetics, and exhibits strong seasonally-alternating selection in unison with large ``boom-bust'' density cycles \citep{messer_2016,bergland_14}.
+Are we compelled to switch to a more ecologically-detailed evolutionary model based on Malthusian parameters or birth/death rates in this system?
+And if we make this switch, how much ecological detail do we need?
+Perhaps the strongest justification of constant relative fitness models is that they are a powerful idealization.
+Thus, before abandoning them it is important to have a better understanding when and how we should do so when selection is not weak, and demography is not stable.
+
+The rough picture given by the weak-selection approximation is missing at least one key element: virtually all models of density-dependent selection do not have a ``reproductive excess'' of juveniles that are more fragile than their adult counterparts \citep{turner1968population,kimura1969natural,nei1971fertility}.
+Reproductive excesses allow selection to be concentrated at a juvenile ``bottleneck'', which means that the density of adults can remain constant even if strong selection is occurring on juveniles --- rather different from the classic $K$-selection scenario.
+Although similar ideas featured prominently in   early debates on the regulation of population density (coincidentally grounded in experiments on flies; \citealt{nicholson_1954}), much of the reproductive excess literature revolves around Haldane's ``cost of selection'' \citep{haldane_1957}, and is also poorly integrated with population ecology.
+For instance, \cite{kimura1969natural} simply took constant $N$ as a requirement and then derived variants of the logistic model consistent with this.
 \cite{nei1971fertility} proposed a model with an explicit representation of reproductive excess, but used an unusual model of competition based on pair-wise interactions which was only defined for at most two different types.
-As a result, the role of reproductive excesses in justifying the use of relative fitness is still largely verbal.
+As a result, the role of reproductive excesses in justifying the use of constant relative finesses is largely verbal.
 
 Here we study the population ecology of relative fitness using a novel model of density-dependent population growth based on territorial contests.
-We restrict our attention to asexual haploids with little or no age structure, since it is then easier to evaluate how the success or failure of the relative fitness description is tied to the underlying population ecological assumptions.
+We restrict our attention to asexual haploids with little or no age structure, since it is then easier to evaluate how the success or failure of the constant-$s$ description is tied to the underlying population ecological assumptions.
 Our starting point is the classic lottery model which was developed by ecologists to study competition driven by territorial contests in reef fishes and plants \citep{sale_77,chesson_1981}.
-The classic lottery incorporates a reproductive excess, and fitness involves a product of fertility and juvenile viability akin to the population simplest population genetic models of selection (e.g.
+The classic lottery incorporates a reproductive excess, and fitness involves a product of fertility and juvenile viability akin to standard population genetic models of selection (e.g.
 \citealt[pp.
@@
 
-Equipped with this model, we turn to the evaluation of relative fitness. [Need to write outline, will do when structure is finalized]
+Equipped with our new variable-density lottery model, we outline its selection properties (section ``Density-dependent selection in the variable-density lottery''). Next, we discuss
+selection on the ability to contest territories, which behaves like a pure constant-$N$, relative fitness trait, and discuss how this fits with the classical notion of $K$-selection  (section ''The response of density to selection; $c$-selection versus $K$-selection'').
+We then consider selection on density-regulating traits (section ``Density-regulating traits and the threat of strong selection''), and conclude by contrasting the classical density-dependent selection literature with our results (``Discussion'').
 
@@
 
-\subsection*{Density-regulating traits at equilibrium}
+\subsection*{Density-regulating traits and the threat of strong selection}
 

### 2018-03-15 manuscript.tex
l

@@
 
-For haploids growing in discrete time steps, the change in the abundance $n_i$ of type $i$ over a time step can be expressed as $\Delta n_i=(W_i - 1)n_i$ where $W_i$ is absolute fitness.
+For haploids growing in discrete time steps, the change in the abundance $n_i$ of type $i$ over a time step can be expressed as $\Delta n_i=(W_i - 1)n_i$ where $W_i$ is ``absolute fitness''.
 The corresponding change in frequency is $\Delta p_i=\left(\frac{W_i}{\overline{W}}-1\right) p_i$, where $\overline{W}=\sum_i W_i p_i$.
@@
 26]{crow_1970}.
-In the particular case that two types are present, a mutant labeled $1$ and a wildtype labeled $2$ for instance, then the continuous time equation for the mutant takes the familiar form
+In the particular case that two types are present, the selection equation for the first type takes the familiar form
 \begin{equation}
@@
 where $s=r_1-r_2$ is the selection coefficient.
-Note that in the discrete case we can replace the $W_i$ with any set of values proportional to the $W_i$ and the changes in frequency $\Delta p_i$ will be unaffected.
-These values, which give no information about the dynamics of total population density $N=\sum_i n_i$, are called ``relative fitnesses'' \citep[pp.
+Note that, in the discrete case, we can replace the $W_i$ with any set of values proportional to the $W_i$ without affecting the  ratio $W_i/\overline{W}$ or $\Delta p_i$.
+These ``relative fitness'' values tell us how type frequencies change, but give no information about the dynamics of total population density $N=\sum_i n_i$ \citep[pp.
 468]{barton_2007}.
-Similarly, the selection coefficient and frequency dynamics in Eq.~\eqref{eq:canonical} would be unaffected if we were to add an arbitrary constant to both $r_1$ and $r_2$ (these would be relative log fitnesses).
+Similarly, the selection coefficient and frequency dynamics in Eq.~\eqref{eq:canonical} would be unaffected if we were to add an arbitrary constant to $r_1$ and $r_2$ (these would be relative log fitnesses).
 
-In the absence of crowding, each type grows exponentially at a constant ``instrinsic'' rate $r_i$ (for concreteness, we will primarily use Eq.~\eqref{eq:canonical} as our point of comparison but similar logic applies in the discrete time case).
-The selection coefficient $s$ is then a constant equal to the difference in intrinsic population growth rates.
+In a constant environment, and in the absence of crowding, $r_i$ is a constant ``intrinsic growth rate (for concreteness we will use Eq.~\eqref{eq:canonical} as our point of comparison, but the discrete time case is similar).
+The selection coefficient $s$ is then simply the difference in intrinsic population growth rates.
 The situation is more complicated when we allow for population crowding.
-Since crowded and uncrowded conditions can favor quite different traits \citep{travis_2013}, the Malthusian parameters $r_i$ may respond differently to changes in population density $N$, in which case the selection coefficient will be density-dependent.
-Eq.~\eqref{eq:canonical} would then need to supplemented with a model for the dynamics of $N$ to describe how the allele frequencies will change over time.
-The density and type frequency dynamics will in general be coupled \citep{travis_2013}.
-Thus, in a general crowded population the simple procedure of assigning constant relative fitness values to different types has to be replaced with an ecological description of absolute growth rates --- a substantially more complicated problem.
-Note that frequency-dependent selection does not raise such a severe problem, because we still only need to model the type frequencies, and not also the ecological variable $N$.
+Since crowded and uncrowded conditions can favor quite different traits \citep{travis_2013}, the Malthusian parameters $r_i$ may respond differently to changes in $N$.
+The selection coefficient would then be density-dependent, and we would need to also know the dynamics of $N$ to use Eq.~\eqref{eq:canonical}.
+In general we cannot just independently specify the dynamics of $N$, because those ecological dynamics are coupled with the evolutionary dynamics of type frequency \citep{travis_2013}.
+Thus, in a general crowded population the simple procedure of assigning constant relative fitness values to different types has to be replaced with an ecological description of absolute growth rates.
+Note that frequency-dependent selection does not raise such a severe problem, because we only need to model the type frequencies, not also the ecological variable $N$.
 
@@
 In these examples, both $N$ and $s$ change during, and as a result of, adaptive sweeps.
-The Lotka-Volterra competition model also couples selection in crowded populations to density except in special cases \citep{smouse_1976,mallet_2012}.
-It would therefore seem that the ubiquitous density-independent-$s$, constant-$N$ description of selection is incompatible with a huge class of population ecological processes driving selection (Fig.~\ref{fig:Ksel}b).
+The Lotka-Volterra competition model also couples density to selection in crowded populations except in special cases \citep{smouse_1976,mallet_2012}.
+It would therefore seem that the ubiquitous density-independent-$s$, constant-$N$ setup is incompatible with a broad class of population ecological processes driving selection (Fig.~\ref{fig:Ksel}b).
 
@@
 \includegraphics[scale=0.8]{Kplot.pdf}
-\caption{\label{fig:Ksel} Selection in crowded environments shown as a phase diagram for the densities of two types $n_1$ and $n_2$.
+\caption{\label{fig:Ksel} Phase diagram for the densities of two types $n_1$ and $n_2$ undergoing selection.
 (a) The logistic model $\frac{dn_1}{dt}=r_1(1-\frac{n_1+n_2}{K_1})n_1$ and $\frac{dn_2}{dt}=r_2(1-\frac{n_1+n_2}{K_2})n_1$ with $r_1=r_2$ and $K_1>K_2$.
@@
 
-In light of this, and the related difficulties posed by frequency-dependent selection, relative fitness values that are independent of type frequencies and population density are typically justified as an approximation that holds when selection is weak (\citealt{kimura1969natural}; \citealt[pp.
+In light of these difficulties, the assignment of a density- and frequency-independent relative fitness value to each type is typically justified as an approximation that holds when selection is weak (\citealt{kimura1969natural}; \citealt[pp.
 277]{ewens_2004}; \citealt[Chap.
-4]{charlesworth_1994}) and density is changing much slower than type frequencies \citep[Fig.
+4]{charlesworth_1994}) and $N$ changes slowly \citep[Fig.
 9.5]{otto_2011}.
-Under these conditions, the selection coefficient may depend on density and type frequencies, but since these change slowly the selection coefficient may remain approximately constant over several generations.
-In the case of density-dependence, this time interval of constancy could include entire selective sweeps.
-On the other hand, it is known in both simple models like the logistic \citep{kimura1969natural}, and in complex age-structured models \citep[Chap.
-4]{charlesworth_1994}, that frequency trajectories under strong density- or frequency-dependent selection can differ substantially from the constant-$s$ approximation.
+Under these conditions, $s$ is approximately constant in Eq.~\eqref{eq:canonical}, at least for some number of generations.
+If $s$ depends only on density, not frequency, this approximate constancy can hold over entire selective sweeps.
 
-Yet knowing that the constant-$s$ approximation can fail \textit{in general} when selection is strong gives only a rough picture of the domain and manner in which the constant-$s$ approximation breaks down.
-For instance, wild \textit{Drosophila} is an important model system in genetics, and exhibits strong seasonally-alternating selection in unison with large ``boom-bust'' density cycles \citep{messer_2016,bergland_14}.
-Are we compelled to switch to a more ecologically-detailed evolutionary model based on Malthusian parameters or birth/death rates in this system?
+However, our understanding of the strong-selection, unstable-$N$ regime is generally limited to a few prominent examples such as the classical density-dependent selection literature discussed above, and complex age-structured models \citep[Chap.
+4]{charlesworth_1994}.
+Since constant relative fitness models are such a powerful idealization, it is important to understand the specifics of how and when they fail when selection is not weak, and $N$ is not stable.
+For instance, in wild \textit{Drosophila}, strong seasonally-alternating selection happens concurrently with large ``boom-bust'' density cycles \citep{messer_2016,bergland_14}.
+Are we compelled to switch to a more ecologically-detailed model of selection based on Malthusian parameters or birth/death rates in this important model system?
 And if we make this switch, how much ecological detail do we need?
-Perhaps the strongest justification of constant relative fitness models is that they are a powerful idealization.
-Thus, before abandoning them it is important to have a better understanding when and how we should do so when selection is not weak, and demography is not stable.
 
-The rough picture given by the weak-selection approximation is missing at least one key element: virtually all models of density-dependent selection do not have a ``reproductive excess'' of juveniles that are more fragile than their adult counterparts \citep{turner1968population,kimura1969natural,nei1971fertility}.
+Most models of density-dependent selection are missing one key element that has important ramifications for the validity of the constant-$s$ approximation: they do not have a ``reproductive excess'' of juveniles that are more fragile than their adult counterparts \citep{turner1968population}.
 Reproductive excesses allow selection to be concentrated at a juvenile ``bottleneck'', which means that the density of adults can remain constant even if strong selection is occurring on juveniles --- rather different from the classic $K$-selection scenario.
-Although similar ideas featured prominently in   early debates on the regulation of population density (coincidentally grounded in experiments on flies; \citealt{nicholson_1954}), much of the reproductive excess literature revolves around Haldane's ``cost of selection'' \citep{haldane_1957}, and is also poorly integrated with population ecology.
+Although similar ideas featured prominently in   early debates on the regulation of population density (grounded in experiments on flies; \citealt{nicholson_1954}), much of the reproductive excess literature revolves around Haldane's ``cost of selection'' \citep{haldane_1957}, and is also poorly integrated with population ecology.
 For instance, \cite{kimura1969natural} simply took constant $N$ as a requirement and then derived variants of the logistic model consistent with this.
@@
 Here we study the population ecology of relative fitness using a novel model of density-dependent population growth based on territorial contests.
-We restrict our attention to asexual haploids with little or no age structure, since it is then easier to evaluate how the success or failure of the constant-$s$ description is tied to the underlying population ecological assumptions.
+We restrict our attention to asexual haploids, since it is then clearer how the success or failure of the constant-$s$ description is tied to the underlying population ecological assumptions.
 Our starting point is the classic lottery model which was developed by ecologists to study competition driven by territorial contests in reef fishes and plants \citep{sale_77,chesson_1981}.
@@
 The classic lottery also assumes constant $N$, and generations can overlap.
-Our first task is to generalize the lottery model to create a variable-density lottery (sections ``Model'' and ``Analytical approximation of the variable-density lottery'').
+Our first task is to relax the constant-$N$ requirement of the lottery model to create a variable-density lottery (sections ``Model'' and ``Analytical approximation of the variable-density lottery'').
 
-Equipped with our new variable-density lottery model, we outline its selection properties (section ``Density-dependent selection in the variable-density lottery''). Next, we discuss
-selection on the ability to contest territories, which behaves like a pure constant-$N$, relative fitness trait, and discuss how this fits with the classical notion of $K$-selection  (section ''The response of density to selection; $c$-selection versus $K$-selection'').
-We then consider selection on density-regulating traits (section ``Density-regulating traits and the threat of strong selection''), and conclude by contrasting the classical density-dependent selection literature with our results (``Discussion'').
+We then outline the selection properties of our new variable-density lottery model (section ``Density-dependent selection in the variable-density lottery''). Next, we discuss
+selection on the ability to contest territories, which behaves like a pure constant-$N$, constant-$s$ trait, and discuss how this fits with the classical notion of $K$-selection  (section ''The response of density to selection; $c$-selection versus $K$-selection'').
+We then consider selection on traits that regulate population density (section ``Density-regulating traits and the threat of strong selection''), and conclude by contrasting the classical density-dependent selection literature with our results (``Discussion'').
 
@@
 
-%Previous models of selection-independent density have either relied on unusual models of competition \citep{kimura1969natural,nei1971fertility}, or made restrictive Lotka-Volterra parameter choices (Appendix C; \citealt{smouse_1976,mallet_2012}).
+Summarizing the three traits in the variable-density lottery model: (i) $c$-selection is density-dependent, but $c$ does not regulate density; (ii) $d$ regulates density, but $d$-selection is density-independent; (iii) $b$ regulates density and $b$-selection is density-dependent.
+Yet despite their differences, pure $b$, $c$ and $d$ sweeps starting and ending at equilibrium all obey the density-independent relative fitness description of selection almost exactly.
 
-Summarizing the three traits in the variable-density lottery model: (i) $c$-selection is density-dependent, but $c$ does not regulate density; (ii) $d$ regulates density, but $d$-selection is density-independent; (iii) $b$ regulates density and $b$-selection is density-dependent.
-Yet despite these differences, pure $b$, $c$ and $d$ sweeps starting and ending at equilibrium all obey the density-independent relative fitness description of selection almost exactly.
-This behavior is quite different from that found in the classical density-dependent selection literature \citep{roughgarden_1979,christiansen_2004}.
+This behavior is quite different from that found in the classical density-dependent selection literature \citep{roughgarden_1979,christiansen_2004}, where strong selection is sufficient for relative fitness to break down.
+To briefly review: based on a diploid, bi-allelic variant of the logistic model, the $r$/$K$ scheme proposed a dichotomy between $r$-selection (uncrowded) and $K$-selection (crowded) \citep{macarthur_1962}, with the latter taken to mean selection for greater equilibrium density \citep{gill_1974}.
+The more general Lotka-Volterra competition model introduces the inter-type $\alpha_{ij}$ competition coefficients, with selection on these termed ``$\alpha$-selection'' \citep{gill_1974,joshi_2001}.
+Setting aside $r$ which confers no selective advantage at equilibrium, we are left with $K$ and $\alpha$.
+These traits both behave like $\delta$ in Eq.~\eqref{eq:simplebirthdeath} in that they are density-dependent and cause density to change over a sweep (Fig.~\ref{fig:strengthofselection}).
 
-To briefly review: based on a diploid, bi-allelic variant of the logistic model, the $r$/$K$ scheme proposed a dichotomy between $r$-selection (uncrowded) and $K$-selection (crowded) \citep{macarthur_1962}, with the latter taken to mean selection for greater equilibrium density \citep{gill_1974}.
-A more general Lotka-Volterra model introduces the inter-type $\alpha_{ij}$ competition coefficients, with selection on these termed ``$\alpha$-selection'' \citep{gill_1974,joshi_2001}.
-Setting aside $r$ which confers no selective advantage at equilibrium, we are left with $K$ and $\alpha$, which both behave like $\delta$ in Eq.~\eqref{eq:simplebirthdeath} in that they are density-dependent and cause density to change over a sweep (although $N$ only dips transiently during an $\alpha$-sweep).
-Thus, strong selection is sufficient for relative fitness to break down in the classical view of density-dependent selection.
+In the variable-density lottery, this occurs if and only if types differ in more than one trait (Fig.~\ref{fig:multiple}).
+The $c$ and $d$ traits represent the two distinct directions in which density and selection interact: selection may depend on density, and density may depend on selection \citep{prout_1980}.
+The combination of both is necessary to threaten the constant-$s$ approximation.
+However, remarkably, the $b$ trait demonstrates that the combination is not sufficient, since the density-dependence of $b$-selection effectively disappears over equilibrium-to-equilibrium $b$-sweeps.
+Thus, the simple differential/difference equations that have become standard in discussions of density-dependent selection \citep{roughgarden_1979,christiansen_2004,mallet_2012,travis_2013} actually confound important aspects of the interaction between density and selection.
 
-Our variable-density lottery model shows that it is not simply a lack of ecological realism that underlies the contrast between relative fitness and the classical view of density-dependent selection.
+While this is a conceptual reason to be wary of the classical density-dependent selection models, it is not clear what trait variation should be expected in nature.
+Should we expect mutations to generally affect $b$, $c$ and $d$ independently of each other, or pleiotropically such that $\delta$-like selection is prevalent?
+In the case of well-mixed indirect exploitation competition for consumable resources, the $R^*$ rule  suggests that $\delta$-like selection will be prevalent.
+However, for many populations consumable resources are not well-mixed.
+Spatial localization of consumable resources (e.g.
+due to restricted movement of nutrients through soils) will tend to create a territorial situation similar to the lottery model, where resource competition only occurs locally and both it and interference competition are subsumed into the competitive ability $c$, which does not affect $N$.
+
+Our variable-density lottery model shows that it is not simply a lack of ecological realism on the part of constant relative fitness models that underlies its contrast with the classical view of density-dependent selection.
 Rather, in many population growth models, only one life-history stage is represented, and the competitive effects resulting from crowding appear as a reduction in absolute fitness that only depends on the type densities at this life-history stage (e.g.
 the $n_i^2$ and $n_in_j$ terms in the Lotka Volterra equation).
-As noted in the introduction, this precludes selection concentrated at a fragile juvenile stage as a result of a reproductive excess \citep{chesson_1983,turner1968population,kimura1969natural,nei1971fertility}.
+As This precludes selection concentrated at a fragile juvenile stage facilitated by a reproductive excess \citep{chesson_1983,turner1968population,kimura1969natural,nei1971fertility}.
 
 Reproductive excesses appear in the variable-density lottery model when the number of propagules is greater than the number of available territories.
-Then only $\approx 1/L$ of the juveniles contesting available territories survive to adulthood.
+Then only $\approx 1/L$ of the juveniles contesting unoccupied territories survive to adulthood.
 Unlike the role of adult density $n_i$ in single-life-stage models, it is the propagule densities $l_i$ that represent the crowding that drives competition (a ``critical age-group''; \citealt[pp.
@@
 In this respect, the viability selection heuristics that are common in population genetics \cite[pp.
-61]{gillespie_2004} actually capture an important ecological process without jumping to the full complexity of age-structured models.
+61]{gillespie_2004} actually capture an important ecological process without making the full leap to complex age-structured models.
 
-We now turn to the breakdown of the density-independent relative fitness assumption which occurs when strong selection changes population density and is also density-dependent (Fig.~\ref{fig:strengthofselection}).
-In the variable-density lottery, this occurs if and only if types differ in more than one trait (Fig.~\ref{fig:multiple}).
-The $c$ and $d$ traits represent the two distinct directions in which density and selection interact: selection may depend on density, and density may depend on selection \citep{prout_1980}.
-The combination is necessary to pose a threat to Eq.~\eqref{eq:canonical}.
-However, the $b$ trait remarkably demonstrates that the combination is not sufficient, since the density-dependence of $b$-selection disappears over equilibrium-to-equilibrium $b$-sweeps.
-Thus, the simple differential/difference equations that have become standard in discussions of density-dependent selection \citep{roughgarden_1979,christiansen_2004,mallet_2012,travis_2013} actually represent a confounded form of the interaction between density and selection.
-
-While this is a conceptual reason to be wary of the classical density-dependent selection models, it is not clear what trait variation to expect in nature.
-Should we expect mutations to generally affect $b$, $c$ and $d$ independently of each other, or pleiotropically such that $\delta$-like selection is prevalent?
-In the case of well-mixed indirect exploitation competition for consumable resources, the $R^*$ rule  suggests that $\delta$-like selection will be prevalent.
-However, for many populations consumable resources are not well-mixed.
-Spatial localization of consumable resources (e.g.
-due to restricted movement of nutrients through soils) will tend to create a territorial situation similar to the lottery model, where resource competition only occurs locally and both it and interference competition are subsumed into the competitive ability $c$, which does not affect $N$.
-
-The above findings underscore that the most serious threat to relative fitness arises from deviations from demographic equilibrium.
-This can dramatically alter frequency trajectories for individual sweeps (e.g.
+The above findings underscore that the most serious threat to the constant-$s$ approximation arises due to deviations from demographic equilibrium as a result of changes in the demographic rates of the types already present i.e.
+as a result of a temporally-variable environment.
+While transient deviations from demographic equilibrium driven by the appearance of new types can also threaten the constant-$s$ approximation, it requires strong selection that is density-dependent and effects a density-regulating trait (and even then the constant-$s$ approximation may hold).
+Temporally-variable environments can dramatically alter frequency trajectories for individual sweeps (e.g.
 Fig.
 9.5 in \cite{otto_2011}; Fig.
-5 in \cite{mallet_2012}), as well as the long-term outcomes of selection in fluctuating environments \citep{lande_2009}.
-This would suggest that in the wild \textit{Drosophila} example mentioned in the Introduction, there is indeed no choice but to abandon relative fitness.
-The variable-density lottery could provide a useful starting point for analyzing evolution in this and other far-from-equilibrium situations.
-One convenient feature of our variable-density lottery (and lottery models in general) is a formal similarity to the Wright-Fisher model, which should facilitate the analysis of ``density-dependent drift''.
+5 in \cite{mallet_2012}), as well as the long-term outcomes of selection \citep{lande_2009}.
 
-Another issue with the constant-$N$ relative fitness description of selection is that it precludes consideration of longer-term aspects of the interplay between evolution and ecology such as population extinction.
+This suggests that in systems like the wild \textit{Drosophila} example mentioned in the Introduction, there is indeed no choice but to abandon relative fitness.
+The variable-density lottery could provide a useful starting point for analyzing evolution in this and other far-from-equilibrium situations for two reasons: 1) the $b$, $c$, $d$ trait scheme neatly distinguishes between different aspects of the interplay between density and selection; 2)  lottery models in general are mathematically similarity to the Wright-Fisher model, which should facilitate the analysis of genetic when $N$ is unstable.
+
+[Kind of an odd way to end, but it breaks the flow if I put it elsewhere] Another issue with the constant-$N$, constant-$s$ description of selection is that it precludes consideration of longer-term aspects of the interplay between evolution and ecology such as population extinction.
 A variety of approaches have been developed for dealing with these issues in quantitative genetics \citep{burger1995evolution,engen_2013}, population genetics \citep{bertram2017predicting} and adaptive dynamics \citep{ferriere2013eco,dieckmann2004adaptive}.

### 2018-03-16 manuscript.tex
l

@@
 The variety of fitness measures is not problematic in itself, because different measures have different uses.
-But it should be clear how these measures are connected to the processes of birth and death which ultimately drive selection \citep{metcalf_2007,doebeli_2017}.
+But it should be clear how these measures are connected to the processes of birth and death which ultimately drive selection (\citealt{metcalf_2007,doebeli_2017}; \citealt[pp.
+178]{charlesworth_1994}).
 While such a connection is fairly clear for absolute fitness measures like $r$ or $R_0$, relative fitness has only weak justification from population ecology.
 It has even been proposed that relative fitness be justified from measure theory, abandoning population biology altogether \citep{wagner_2010}.
-Given the ubiquitous use of relative fitness, it is important that we understand its population ecological basis, both to clarify its domain of applicability, and as part of the broader challenge of synthesizing ecology and evolution.
+Given the ubiquitous use of relative fitness in evolutionary genetics, it is important that we understand its population ecological basis, both to clarify its domain of applicability, and as part of the broader challenge of synthesizing ecology and evolution.
 
@@
 The selection coefficient $s$ is then simply the difference in intrinsic population growth rates.
-The situation is more complicated when we allow for population crowding.
-Since crowded and uncrowded conditions can favor quite different traits \citep{travis_2013}, the Malthusian parameters $r_i$ may respond differently to changes in $N$.
-The selection coefficient would then be density-dependent, and we would need to also know the dynamics of $N$ to use Eq.~\eqref{eq:canonical}.
-In general we cannot just independently specify the dynamics of $N$, because those ecological dynamics are coupled with the evolutionary dynamics of type frequency \citep{travis_2013}.
+However, growth cannot continue at a constant rate indefinitely.
+As population density increases, crowding will cause the Malthusian parameters $r_i$ to decline (e.g.
+\citealt[pp.
+203]{begon_1990}).
+Since different traits are advantageous in crowded and uncrowded conditions, selection in the presence of crowding can be density-dependent, and we would then need to know the dynamics of $N$ to use Eq.~\eqref{eq:canonical}.
+In general we cannot simply independently specify the dynamics of $N$, because those ecological dynamics are coupled with the evolutionary dynamics of type frequency \citep{travis_2013}.
 Thus, in a general crowded population the simple procedure of assigning constant relative fitness values to different types has to be replaced with an ecological description of absolute growth rates.
-Note that frequency-dependent selection does not raise such a severe problem, because we only need to model the type frequencies, not also the ecological variable $N$.
+Note that frequency-dependent selection does not raise such a severe problem, because a complete description of selection only requires us to model the type frequencies, not also the ecological variable $N$.
 
@@
 In these examples, both $N$ and $s$ change during, and as a result of, adaptive sweeps.
-The Lotka-Volterra competition model also couples density to selection in crowded populations except in special cases \citep{smouse_1976,mallet_2012}.
-It would therefore seem that the ubiquitous density-independent-$s$, constant-$N$ setup is incompatible with a broad class of population ecological processes driving selection (Fig.~\ref{fig:Ksel}b).
+Density also changes in response to selection in the Lotka-Volterra competition model except in special cases \citep{smouse_1976,mallet_2012}.
+It would therefore seem that the density-independent-$s$, constant-$N$ setup (Fig.~\ref{fig:Ksel}b) is incompatible with a broad class of population ecological processes driving selection.
 
@@
 
-In light of these difficulties, the assignment of a density- and frequency-independent relative fitness value to each type is typically justified as an approximation that holds when selection is weak (\citealt{kimura1969natural}; \citealt[pp.
+In light of these difficulties, the assignment of a density- and frequency-independent relative fitness value to each type is typically justified as an approximation that holds when selection is weak and $N$ changes slowly (\citealt{kimura1969natural}; \citealt[pp.
 277]{ewens_2004}; \citealt[Chap.
-4]{charlesworth_1994}) and $N$ changes slowly \citep[Fig.
+4]{charlesworth_1994}).
+Under these conditions, $s$ is approximately constant in Eq.~\eqref{eq:canonical}, at least for some number of generations.
+If $s$ depends only on density, not frequency, this approximate constancy can hold over entire selective sweeps \citep[Fig.
 9.5]{otto_2011}.
-Under these conditions, $s$ is approximately constant in Eq.~\eqref{eq:canonical}, at least for some number of generations.
-If $s$ depends only on density, not frequency, this approximate constancy can hold over entire selective sweeps.
 
-However, our understanding of the strong-selection, unstable-$N$ regime is generally limited to a few prominent examples such as the classical density-dependent selection literature discussed above, and complex age-structured models \citep[Chap.
-4]{charlesworth_1994}.
-Since constant relative fitness models are such a powerful idealization, it is important to understand the specifics of how and when they fail when selection is not weak, and $N$ is not stable.
+However, our understanding of the strong-selection, unstable-$N$ regime is
+
+generally limited to a few prominent examples such as the classical density-dependent selection literature discussed above,
+
+and complex age-structured models \citep[Chap. 4]{charlesworth_1994}.
+
+Since constant relative fitness models are such a powerful idealization, it is important to understand the specifics of how and when they fail when selection is not weak, or $N$ is not stable.
 For instance, in wild \textit{Drosophila}, strong seasonally-alternating selection happens concurrently with large ``boom-bust'' density cycles \citep{messer_2016,bergland_14}.
@@
 We then outline the selection properties of our new variable-density lottery model (section ``Density-dependent selection in the variable-density lottery''). Next, we discuss
-selection on the ability to contest territories, which behaves like a pure constant-$N$, constant-$s$ trait, and discuss how this fits with the classical notion of $K$-selection  (section ''The response of density to selection; $c$-selection versus $K$-selection'').
+selection on the ability to contest territories, which behaves like a pure constant-$N$, constant-$s$ trait, and discuss how this fits with the classical notion of $K$-selection  (section ``The response of density to selection; $c$-selection versus $K$-selection'').
 We then consider selection on traits that regulate population density (section ``Density-regulating traits and the threat of strong selection''), and conclude by contrasting the classical density-dependent selection literature with our results (``Discussion'').

### 2018-03-17 manuscript.tex
l

@@
 203]{begon_1990}).
-Since different traits are advantageous in crowded and uncrowded conditions, selection in the presence of crowding can be density-dependent, and we would then need to know the dynamics of $N$ to use Eq.~\eqref{eq:canonical}.
-In general we cannot simply independently specify the dynamics of $N$, because those ecological dynamics are coupled with the evolutionary dynamics of type frequency \citep{travis_2013}.
-Thus, in a general crowded population the simple procedure of assigning constant relative fitness values to different types has to be replaced with an ecological description of absolute growth rates.
-Note that frequency-dependent selection does not raise such a severe problem, because a complete description of selection only requires us to model the type frequencies, not also the ecological variable $N$.
+Selection can then be density-dependent, and we would need to know the dynamics of $N$ to use Eq.~\eqref{eq:canonical}.
+In general we cannot simply independently specify the dynamics of $N$, because those ecological dynamics are coupled with the evolutionary dynamics of type frequency \citep{nagylaki_1979,travis_2013}.
+Thus, the simple procedure of assigning constant relative fitness values to different types has to be replaced with an ecological description of absolute growth rates.
+Note that frequency-dependent selection does not raise such a severe problem, because a complete description of selection only requires us to model the type frequencies, not $N$ as well.
 
-In practice, many population genetics models assume that $N$ has reached some fixed equilibrium value, and each type is assigned a constant relative fitness.
+In practice, many population genetic models assume that $N$ has reached some fixed equilibrium value, and each type is assigned a constant relative fitness.
 Selection is typically assumed to operate through viability (e.g.
@@
 61]{gillespie_2004}), but other important population ecological assumptions --- such as how density is regulated --- are left unspecified.
-Beyond being conceptually unsatisfactory \citep{doebeli_2017}, we might also doubt whether this density-independent-$s$, constant-$N$ setup is consistent with the ecology of many populations.
-Much of the seminal literature on density-dependent selection revolves around the connection between selection in crowded populations and the ability to keep growing at higher densities \citep{macarthur_1967,roughgarden_1979}.
-The classic example is the logistic model, where the type with the greatest equilibrium population density ``$K$'' excludes the others (Fig.~\ref{fig:Ksel}a).
-Similarly, the ``$R^*$ rule'', a central tenet of resource competition theory, states that when growth is limited by a single homogeneous consumable resource, the type able to deplete the resource to the lowest equilibrium density $R^*$ (which will often entail a different equilibrium population density) excludes the others \citep{grover_1997}.
-In these examples, both $N$ and $s$ change during, and as a result of, adaptive sweeps.
-Density also changes in response to selection in the Lotka-Volterra competition model except in special cases \citep{smouse_1976,mallet_2012}.
+However, there are reasons to doubt that this setup is ecologically valid.
+Many absolute fitness models of density-regulated growth exhibit a strong connection between selection and the ability to keep growing at higher densities \citep{fisher_1930,kostitzin_1939,macarthur_1967,roughgarden_1979,christiansen_2004}.
+This includes simple birth-death \citep{kostitzin_1939} and logistic models (Fig.~\ref{fig:Ksel}a; \citealt{macarthur_1962,kimura1969natural,roughgarden_1979,boyce_1984}), more general functional forms for the absolute fitness penalties of crowding \citep{kimura_1978,charlesworth_1971,lande_2009,nagylaki_1979}, and the ``$R^*$ rule'' of resource competition theory (which states that the type able to deplete a shared, homogeneous, limiting consumable resource to the lowest equilibrium density $R^*$ excludes the others; \citealt{grover_1997}).
+Density also changes in response to selection (even if only transiently) in the Lotka-Volterra competition model except in special cases \citep{smouse_1976,mallet_2012}.
+This behavior implies that both $N$ and $s$ change during, and as a result of, adaptive sweeps, and has been studied extensively following Fisher's suggestion that evolution tends to increase density in the long term \citep{fisher_1930, macarthur_1962, roughgarden_1979, lande_2009}.
 It would therefore seem that the density-independent-$s$, constant-$N$ setup (Fig.~\ref{fig:Ksel}b) is incompatible with a broad class of population ecological processes driving selection.
@@
 
-However, our understanding of the strong-selection, unstable-$N$ regime is
+However, the preceding arguments do not imply that the constant-$s$ approximation \textit{only} applies under weak selection and stable $N$.
+All of the models of density-dependent selection mentioned above are missing one key element that has important ramifications for the validity of the constant-$s$ approximation: a ``reproductive excess'' of juveniles that are more fragile than their adult counterparts \citep{turner1968population}.
+In complex age-structured models, dependence on density is assumed to come from a ``critical age-group'', and this age group's density responds to selection in much the same way as found in the above models without age structure \citep[Chap.
+4]{charlesworth_1994} (e.g.
+adaptive evolution proceeds in the direction of increasing equilibrium density in the critical age-group).
+However, in the constant-$s$ approximation, reproductive excesses allow selection to be concentrated at a juvenile ``bottleneck'', which means that the density of adults can remain constant even if strong selection is occurring on juveniles.
 
-generally limited to a few prominent examples such as the classical density-dependent selection literature discussed above,
+Although similar ideas featured prominently in early debates on the regulation of population density (grounded in experiments on flies; \citealt{nicholson_1954}), much of the reproductive excess literature revolves around Haldane's ``cost of selection'' \citep{haldane_1957}, and is also poorly integrated with population ecology.
+For instance, \cite{kimura1969natural} simply took constant $N$ as a requirement and then derived variants of the logistic model consistent with this.
+\cite{nei1971fertility} proposed a model with an explicit representation of reproductive excess, but used an unusual model of competition based on pair-wise interactions which was only defined for at most two different types.
+As a result, the role of reproductive excesses in justifying the use of constant relative finesses is largely verbal.
 
-and complex age-structured models \citep[Chap. 4]{charlesworth_1994}.
+This behavior is intuitively appealing because births balance deaths at equilibrium, and adaptive mutants must overcome the existing balance to expand \citep[pp. 227]{kostitzin_1939}.
 
@@
 
-Most models of density-dependent selection are missing one key element that has important ramifications for the validity of the constant-$s$ approximation: they do not have a ``reproductive excess'' of juveniles that are more fragile than their adult counterparts \citep{turner1968population}.
-Reproductive excesses allow selection to be concentrated at a juvenile ``bottleneck'', which means that the density of adults can remain constant even if strong selection is occurring on juveniles --- rather different from the classic $K$-selection scenario.
-Although similar ideas featured prominently in   early debates on the regulation of population density (grounded in experiments on flies; \citealt{nicholson_1954}), much of the reproductive excess literature revolves around Haldane's ``cost of selection'' \citep{haldane_1957}, and is also poorly integrated with population ecology.
-For instance, \cite{kimura1969natural} simply took constant $N$ as a requirement and then derived variants of the logistic model consistent with this.
-\cite{nei1971fertility} proposed a model with an explicit representation of reproductive excess, but used an unusual model of competition based on pair-wise interactions which was only defined for at most two different types.
-As a result, the role of reproductive excesses in justifying the use of constant relative finesses is largely verbal.
+\citep{charlesworth_1971,kimura_1978,leon_1978,nagylaki_1979}
 
@@
 
-We now turn to the issue of how density responds to selection, comparing the variable-density lottery and previous models of selection in density-regulated populations  \citep{prout_1980}.
-Starting with some of the earliest work on the topic \citep{kostitzin_1939,christiansen_2004}, most previous models exhibit some form of ``$K$-selection'' i.e.
-selection in crowded populations increases equilibrium density.
-This idea is intuitively appealing because births balance deaths at equilibrium, and adaptive mutants must overcome the existing balance to expand \citep[pp.
-227]{kostitzin_1939}.
-Yet, as we saw in the previous section, $c$-selection has no effect on population density in the variable-density lottery.
+We now turn to the issue of how density responds to selection, comparing the variable-density lottery to previous studies of density-regulated populations \citep{prout_1980}.
+As we saw in the previous section, $c$-selection has no effect on population density in the variable-density lottery.
 To make sense of how $c$-selection fits with previous population growth models, we now revisit MacArthur's general treatment of $K$-selection \citep{macarthur_1967}.

### 2018-03-17 manuscript.tex
latest

@@
 This includes simple birth-death \citep{kostitzin_1939} and logistic models (Fig.~\ref{fig:Ksel}a; \citealt{macarthur_1962,kimura1969natural,roughgarden_1979,boyce_1984}), more general functional forms for the absolute fitness penalties of crowding \citep{kimura_1978,charlesworth_1971,lande_2009,nagylaki_1979}, and the ``$R^*$ rule'' of resource competition theory (which states that the type able to deplete a shared, homogeneous, limiting consumable resource to the lowest equilibrium density $R^*$ excludes the others; \citealt{grover_1997}).
-Density also changes in response to selection (even if only transiently) in the Lotka-Volterra competition model except in special cases \citep{smouse_1976,mallet_2012}.
+Density also changes in response to selection in the Lotka-Volterra competition model except in special cases (even if only transiently; \citealt{smouse_1976,mallet_2012}).
 This behavior implies that both $N$ and $s$ change during, and as a result of, adaptive sweeps, and has been studied extensively following Fisher's suggestion that evolution tends to increase density in the long term \citep{fisher_1930, macarthur_1962, roughgarden_1979, lande_2009}.
@@
 
-In light of these difficulties, the assignment of a density- and frequency-independent relative fitness value to each type is typically justified as an approximation that holds when selection is weak and $N$ changes slowly (\citealt{kimura1969natural}; \citealt[pp.
+In light of this difficulty, the assignment of a density- and frequency-independent relative fitness value to each type is typically justified as an approximation that holds when selection is weak and $N$ changes slowly (\citealt{kimura1969natural}; \citealt[pp.
 277]{ewens_2004}; \citealt[Chap.
@@
 However, the preceding arguments do not imply that the constant-$s$ approximation \textit{only} applies under weak selection and stable $N$.
-All of the models of density-dependent selection mentioned above are missing one key element that has important ramifications for the validity of the constant-$s$ approximation: a ``reproductive excess'' of juveniles that are more fragile than their adult counterparts \citep{turner1968population}.
+Every model of density-regulated growth mentioned above is missing an important aspect of selection which has ramifications for the constant-$s$ approximation: a ``reproductive excess'' of juveniles that are more fragile than their adult counterparts \citep{turner1968population}.
+Reproductive excesses allow selection to be concentrated at a juvenile ``bottleneck'', which means that the density of adults can remain constant even if strong selection is occurring on juveniles.
+Although similar ideas featured prominently in early debates on the regulation of population density (e.g.
+\citealt{nicholson_1954}),
+
+much of the later reproductive excess literature revolves around Haldane's ``cost of selection'' \citep{haldane_1957} rather than the issues surrounding the constant-$s$ approximation.
+
+For instance, \cite{kimura1969natural} simply assumed that $N$ should remain constant
+
+as a requirement and then derived variants of the logistic model consistent with this.
+\cite{nei1971fertility} proposed a model with an explicit representation of reproductive excess, but used an unusual model of competition based on pair-wise interactions which was only defined for at most two different types.
+
 In complex age-structured models, dependence on density is assumed to come from a ``critical age-group'', and this age group's density responds to selection in much the same way as found in the above models without age structure \citep[Chap.
@@
 adaptive evolution proceeds in the direction of increasing equilibrium density in the critical age-group).
-However, in the constant-$s$ approximation, reproductive excesses allow selection to be concentrated at a juvenile ``bottleneck'', which means that the density of adults can remain constant even if strong selection is occurring on juveniles.
 
-Although similar ideas featured prominently in early debates on the regulation of population density (grounded in experiments on flies; \citealt{nicholson_1954}), much of the reproductive excess literature revolves around Haldane's ``cost of selection'' \citep{haldane_1957}, and is also poorly integrated with population ecology.
-For instance, \cite{kimura1969natural} simply took constant $N$ as a requirement and then derived variants of the logistic model consistent with this.
-\cite{nei1971fertility} proposed a model with an explicit representation of reproductive excess, but used an unusual model of competition based on pair-wise interactions which was only defined for at most two different types.
   As a result, the role of reproductive excesses in justifying the use of constant relative finesses is largely verbal.

### 2018-03-19 manuscript.tex
latest

@@
 Some widely used examples are expected lifetime reproductive ratio $R_0$, intrinsic population growth rate $r$, equilibrium population density/carrying capacity (often labeled ``$K$'') \citep{benton_2000}, and invasion fitness \citep{metz_1992}.
-In addition, ``relative fitness'' is the standard in much of evolutionary biology, particularly evolutionary genetics, where the focus is on relative genotypic proportions \cite[pp.
+In addition, ``relative fitness'' is widely used in evolutionary genetics, where the focus is on relative genotypic frequencies \cite[pp.
 468]{barton_2007}.
-The variety of fitness measures is not problematic in itself, because different measures have different uses.
-But it should be clear how these measures are connected to the processes of birth and death which ultimately drive selection (\citealt{metcalf_2007,doebeli_2017}; \citealt[pp.
+The variety of fitness measures is not problematic in itself, but it should be clear how these measures are connected to the processes of birth and death which ultimately drive selection (\citealt{metcalf_2007,doebeli_2017}; \citealt[pp.
 178]{charlesworth_1994}).
@@
 It has even been proposed that relative fitness be justified from measure theory, abandoning population biology altogether \citep{wagner_2010}.
-Given the ubiquitous use of relative fitness in evolutionary genetics, it is important that we understand its population ecological basis, both to clarify its domain of applicability, and as part of the broader challenge of synthesizing ecology and evolution.
+Given the ubiquitous use of relative fitness in evolutionary genetics, it is important to understand its population ecological basis, both to clarify its domain of applicability, and as part of the broader challenge of synthesizing ecology and evolution.
 
@@
 26]{crow_1970}.
-In the particular case that two types are present, the selection equation for the first type takes the familiar form
+In the particular case that two types are present, the selection equation takes the familiar form
 \begin{equation}
-\frac{d p_1}{dt}=s p_1(1-p_1), \label{eq:canonical}
+\frac{d p_i}{dt}=s p_i(1-p_i), \label{eq:canonical}
 \end{equation}
-where $s=r_1-r_2$ is the selection coefficient.
+where $s=r_i-r_j$ is the selection coefficiento of type $i$ relative to type $j$.
 Note that, in the discrete case, we can replace the $W_i$ with any set of values proportional to the $W_i$ without affecting the  ratio $W_i/\overline{W}$ or $\Delta p_i$.
@@
 468]{barton_2007}.
-Similarly, the selection coefficient and frequency dynamics in Eq.~\eqref{eq:canonical} would be unaffected if we were to add an arbitrary constant to $r_1$ and $r_2$ (these would be relative log fitnesses).
+Similarly, the selection coefficient and frequency dynamics in Eq.~\eqref{eq:canonical} would be unaffected if we added an arbitrary constant to the Malthusian parameters $r_i$ (these would be relative log fitnesses).
 
-In a constant environment, and in the absence of crowding, $r_i$ is a constant ``intrinsic growth rate (for concreteness we will use Eq.~\eqref{eq:canonical} as our point of comparison, but the discrete time case is similar).
+In a constant environment, and in the absence of crowding, $r_i$ is a constant ``intrinsic growth rate (for concreteness we use Eq.~\eqref{eq:canonical} as our point of comparison, but the discrete time case is similar).
 The selection coefficient $s$ is then simply the difference in intrinsic population growth rates.
@@
 Selection can then be density-dependent, and we would need to know the dynamics of $N$ to use Eq.~\eqref{eq:canonical}.
-In general we cannot simply independently specify the dynamics of $N$, because those ecological dynamics are coupled with the evolutionary dynamics of type frequency \citep{nagylaki_1979,travis_2013}.
+In general we cannot simply independently specify the dynamics of $N$, because those ecological dynamics are coupled with the evolutionary dynamics of type frequency \citep{travis_2013}.
 Thus, the simple procedure of assigning constant relative fitness values to different types has to be replaced with an ecological description of absolute growth rates.
-Note that frequency-dependent selection does not raise such a severe problem, because a complete description of selection only requires us to model the type frequencies, not $N$ as well.
+Note that frequency-dependent selection does not raise such a severe problem, because a complete description of selection still only requires us to model the type frequencies, not $N$ as well.
+
+and has been studied extensively following Fisher's suggestion that evolution tends to increase density in the long term \citep{fisher_1930, macarthur_1962, roughgarden_1979, lande_2009}.
+
+A major focus of the literature on density-regulated population growth focuses on the connection between selection and the ability to keep growing at higher densities  \citep{kostitzin_1939,macarthur_1967,roughgarden_1979,christiansen_2004}.
+
+This includes simple birth-death \citep{kostitzin_1939} and logistic models (Fig.~\ref{fig:Ksel}a; \citealt{macarthur_1962,kimura1969natural,roughgarden_1979,boyce_1984}), more general functional forms for the absolute fitness penalties of crowding \citep{kimura_1978,charlesworth_1971,lande_2009,nagylaki_1979}, and the ``$R^*$ rule'' of resource competition theory (which states that the type able to deplete a shared, homogeneous, limiting consumable resource to the lowest equilibrium density $R^*$ excludes the others; \citealt{grover_1997}).
+Density also changes in response to selection in the Lotka-Volterra competition model except in special cases (even if only transiently; \citealt{smouse_1976,mallet_2012}).
+This behavior implies that both $N$ and $s$ change during, and as a result of, adaptive sweeps,
 
 In practice, many population genetic models assume that $N$ has reached some fixed equilibrium value, and each type is assigned a constant relative fitness.
-Selection is typically assumed to operate through viability (e.g.
-\citealt[pp.
-61]{gillespie_2004}), but other important population ecological assumptions --- such as how density is regulated --- are left unspecified.
-However, there are reasons to doubt that this setup is ecologically valid.
-Many absolute fitness models of density-regulated growth exhibit a strong connection between selection and the ability to keep growing at higher densities \citep{fisher_1930,kostitzin_1939,macarthur_1967,roughgarden_1979,christiansen_2004}.
-This includes simple birth-death \citep{kostitzin_1939} and logistic models (Fig.~\ref{fig:Ksel}a; \citealt{macarthur_1962,kimura1969natural,roughgarden_1979,boyce_1984}), more general functional forms for the absolute fitness penalties of crowding \citep{kimura_1978,charlesworth_1971,lande_2009,nagylaki_1979}, and the ``$R^*$ rule'' of resource competition theory (which states that the type able to deplete a shared, homogeneous, limiting consumable resource to the lowest equilibrium density $R^*$ excludes the others; \citealt{grover_1997}).
-Density also changes in response to selection in the Lotka-Volterra competition model except in special cases (even if only transiently; \citealt{smouse_1976,mallet_2012}).
-This behavior implies that both $N$ and $s$ change during, and as a result of, adaptive sweeps, and has been studied extensively following Fisher's suggestion that evolution tends to increase density in the long term \citep{fisher_1930, macarthur_1962, roughgarden_1979, lande_2009}.
+Selection is typically assumed to operate through viability, but other important population ecological assumptions --- such as how density is regulated --- are left unspecified (e.g.
+\citealt{gillespie_2004}; \citealt{nagylaki_1992}; \citealt{ewens_2004}).
+
  It would therefore seem that the density-independent-$s$, constant-$N$ setup (Fig.~\ref{fig:Ksel}b) is incompatible with a broad class of population ecological processes driving selection.

### 2018-03-19 manuscript.tex
l

@@
 178]{charlesworth_1994}).
-While such a connection is fairly clear for absolute fitness measures like $r$ or $R_0$, relative fitness has only weak justification from population ecology.
+While such a connection is clear for absolute fitness measures like $r$ or $R_0$, relative fitness has only weak justification from population ecology.
 It has even been proposed that relative fitness be justified from measure theory, abandoning population biology altogether \citep{wagner_2010}.
-Given the ubiquitous use of relative fitness in evolutionary genetics, it is important to understand its population ecological basis, both to clarify its domain of applicability, and as part of the broader challenge of synthesizing ecology and evolution.
+Given the widespread use of relative fitness in evolutionary genetics, it is important to understand its population ecological basis, both to clarify its domain of applicability, and as part of the broader challenge of synthesizing ecology and evolution.
 
@@
 The corresponding change in frequency is $\Delta p_i=\left(\frac{W_i}{\overline{W}}-1\right) p_i$, where $\overline{W}=\sum_i W_i p_i$.
-In continuous time, the Malthusian parameter $r_i$ replaces $W_i$ and we have $\frac{d n_i}{dt}=r_in_i$ and $\frac{d p_i}{dt}=(r_i-\overline{r}) p_i$ \citep[pp.
-26]{crow_1970}.
+In continuous time, the Malthusian parameter $r_i$ replaces $W_i$ and we have $\frac{d n_i}{dt}=r_in_i$ and $\frac{d p_i}{dt}=(r_i-\overline{r}) p_i$ \citep{crow_1970}.
 In the particular case that two types are present, the selection equation takes the familiar form
@@
 \end{equation}
-where $s=r_i-r_j$ is the selection coefficiento of type $i$ relative to type $j$.
+where $s=r_i-r_j$ is the selection coefficient of type $i$ relative to type $j$.
 Note that, in the discrete case, we can replace the $W_i$ with any set of values proportional to the $W_i$ without affecting the  ratio $W_i/\overline{W}$ or $\Delta p_i$.
@@
 
-In a constant environment, and in the absence of crowding, $r_i$ is a constant ``intrinsic growth rate (for concreteness we use Eq.~\eqref{eq:canonical} as our point of comparison, but the discrete time case is similar).
+In a constant environment, and in the absence of crowding, $r_i$ is a constant ``intrinsic'' growth rate (for concreteness we use Eq.~\eqref{eq:canonical} as our point of comparison, but the discrete time case is similar).
 The selection coefficient $s$ is then simply the difference in intrinsic population growth rates.
@@
 203]{begon_1990}).
-Selection can then be density-dependent, and we would need to know the dynamics of $N$ to use Eq.~\eqref{eq:canonical}.
-In general we cannot simply independently specify the dynamics of $N$, because those ecological dynamics are coupled with the evolutionary dynamics of type frequency \citep{travis_2013}.
+It is then possible that selection is density-dependent, and indeed this is probably not uncommon, because crowded and uncrowded conditions can favor very different traits \citep{travis_2013}.
+We would then need to know the dynamics of $N$ to use Eq.~\eqref{eq:canonical}.
+In general we cannot simply specify the dynamics of $N$ independently, because those ecological dynamics are coupled with the evolutionary dynamics of type frequency \citep{travis_2013}.
 Thus, the simple procedure of assigning constant relative fitness values to different types has to be replaced with an ecological description of absolute growth rates.
-Note that frequency-dependent selection does not raise such a severe problem, because a complete description of selection still only requires us to model the type frequencies, not $N$ as well.
+Note that frequency-dependent selection does not raise a similar problem, because a complete description of selection still only requires us to model the type frequencies, not the ecological variable $N$ as well.
 
-and has been studied extensively following Fisher's suggestion that evolution tends to increase density in the long term \citep{fisher_1930, macarthur_1962, roughgarden_1979, lande_2009}.
+Many previous studies of selection in density-regulated populations have focused on the evolution of traits that affect the ability to keep growing at higher densities  \citep{kostitzin_1939,macarthur_1967,roughgarden_1979,christiansen_2004}.
+This focus stems in part from Fisher's suggestion that evolution tends to increase density in the long term \citep{fisher_1930, lande_2009}, and MacArthur's concept of $K$-selection \citep{macarthur_1962}.
+But the principle that density changes in response to selection is also mathematically ingrained in our models of density-regulated population growth, including simple birth-death \citep{kostitzin_1939} and logistic models (Fig.~\ref{fig:Ksel}a; \citealt{macarthur_1962,roughgarden_1979,boyce_1984}), similar ``toy'' models with more general functional forms for the absolute fitness penalties of crowding \citep{kimura_1978,charlesworth_1971,lande_2009,nagylaki_1979,lande_2009}, the ``$R^*$ rule'' of resource competition theory (which states that the type able to deplete a shared limiting consumable resource to the lowest equilibrium density $R^*$ excludes the others; \citealt{grover_1997}), and the Lotka-Volterra competition model (at least during a sweep, except in special cases; \citealt{gill_1974,smouse_1976,mallet_2012}).
+This behavior implies that both $N$ and $s$ change during, and as a result of, adaptive sweeps.
 
-A major focus of the literature on density-regulated population growth focuses on the connection between selection and the ability to keep growing at higher densities  \citep{kostitzin_1939,macarthur_1967,roughgarden_1979,christiansen_2004}.
+In practice, many population genetic models simply ignore density-dependent selection, and and each type is assigned a constant relative fitness.
+Density either does not enter the model at all, or if finite-population size effects (``drift'') are important, then $N$ is assumed to have reached some fixed equilibrium value.
+Selection is typically assumed to operate through viability, but the ecological processes underlying the regulation of population density are left unspecified (e.g.
+\citealt{gillespie_2004}; \citealt{nagylaki_1992}; \citealt{ewens_2004}).
+It would therefore seem that the density-independent-$s$, constant-$N$ setup (Fig.~\ref{fig:Ksel}b) is incompatible with a broad class of population ecological processes driving selection.
 
-This includes simple birth-death \citep{kostitzin_1939} and logistic models (Fig.~\ref{fig:Ksel}a; \citealt{macarthur_1962,kimura1969natural,roughgarden_1979,boyce_1984}), more general functional forms for the absolute fitness penalties of crowding \citep{kimura_1978,charlesworth_1971,lande_2009,nagylaki_1979}, and the ``$R^*$ rule'' of resource competition theory (which states that the type able to deplete a shared, homogeneous, limiting consumable resource to the lowest equilibrium density $R^*$ excludes the others; \citealt{grover_1997}).
-Density also changes in response to selection in the Lotka-Volterra competition model except in special cases (even if only transiently; \citealt{smouse_1976,mallet_2012}).
-This behavior implies that both $N$ and $s$ change during, and as a result of, adaptive sweeps,
-
-In practice, many population genetic models assume that $N$ has reached some fixed equilibrium value, and each type is assigned a constant relative fitness.
-Selection is typically assumed to operate through viability, but other important population ecological assumptions --- such as how density is regulated --- are left unspecified (e.g.
-\citealt{gillespie_2004}; \citealt{nagylaki_1992}; \citealt{ewens_2004}).
-
- It would therefore seem that the density-independent-$s$, constant-$N$ setup (Fig.~\ref{fig:Ksel}b) is incompatible with a broad class of population ecological processes driving selection.
+%This behavior is intuitively appealing because births balance deaths at equilibrium, and adaptive mutants must overcome the existing balance to expand \citep[pp. 227]{kostitzin_1939}.
 
@@
   As a result, the role of reproductive excesses in justifying the use of constant relative finesses is largely verbal.
-
-This behavior is intuitively appealing because births balance deaths at equilibrium, and adaptive mutants must overcome the existing balance to expand \citep[pp. 227]{kostitzin_1939}.
 

### 2018-03-19 manuscript.tex
latest

@@
 In general we cannot simply specify the dynamics of $N$ independently, because those ecological dynamics are coupled with the evolutionary dynamics of type frequency \citep{travis_2013}.
-Thus, the simple procedure of assigning constant relative fitness values to different types has to be replaced with an ecological description of absolute growth rates.
+Thus, in the presence of density-dependent selection, the simple procedure of assigning constant relative fitness values to different types has to be replaced with an ecological description of absolute growth rates.
 Note that frequency-dependent selection does not raise a similar problem, because a complete description of selection still only requires us to model the type frequencies, not the ecological variable $N$ as well.
 
-Many previous studies of selection in density-regulated populations have focused on the evolution of traits that affect the ability to keep growing at higher densities  \citep{kostitzin_1939,macarthur_1967,roughgarden_1979,christiansen_2004}.
+In practice, many population genetic models simply assign a constant relative fitness to each type, and ignore density dependence.
+Selection is typically interpreted as operating through viability, but the ecological processes underlying the regulation of population density are left unspecified (e.g.
+\citealt{gillespie_2004}; \citealt{nagylaki_1992}; \citealt{ewens_2004}).
+Density either does not enter the model at all, or if finite-population size effects (``drift'') are important, then $N$ is assumed to have reached some fixed equilibrium value (Fig.~\ref{fig:Ksel}b).
+
+A rather different picture emerges in more ecologically-explicit studies of selection in density-regulated populations.
+A major focus of previous studies in this area is the evolution of traits that affect the ability to keep growing at higher densities \citep{kostitzin_1939,macarthur_1967,roughgarden_1979,christiansen_2004}.
 This focus stems in part from Fisher's suggestion that evolution tends to increase density in the long term \citep{fisher_1930, lande_2009}, and MacArthur's concept of $K$-selection \citep{macarthur_1962}.
-But the principle that density changes in response to selection is also mathematically ingrained in our models of density-regulated population growth, including simple birth-death \citep{kostitzin_1939} and logistic models (Fig.~\ref{fig:Ksel}a; \citealt{macarthur_1962,roughgarden_1979,boyce_1984}), similar ``toy'' models with more general functional forms for the absolute fitness penalties of crowding \citep{kimura_1978,charlesworth_1971,lande_2009,nagylaki_1979,lande_2009}, the ``$R^*$ rule'' of resource competition theory (which states that the type able to deplete a shared limiting consumable resource to the lowest equilibrium density $R^*$ excludes the others; \citealt{grover_1997}), and the Lotka-Volterra competition model (at least during a sweep, except in special cases; \citealt{gill_1974,smouse_1976,mallet_2012}).
+But more broadly, density changes in response to selection in many models of density-regulated growth, including simple birth-death \citep{kostitzin_1939} and logistic models (Fig.~\ref{fig:Ksel}a; \citealt{macarthur_1962,roughgarden_1979,boyce_1984}), similar ``toy'' models with more general functional forms for the absolute fitness penalties of crowding \citep{kimura_1978,charlesworth_1971,lande_2009,nagylaki_1979,lande_2009}, the ``$R^*$ rule'' of resource competition theory (which states that the type able to deplete a shared limiting consumable resource to the lowest equilibrium density $R^*$ excludes the others; \citealt{grover_1997}), and the Lotka-Volterra competition model (at least during a sweep, except in special cases; \citealt{gill_1974,smouse_1976,mallet_2012}).
 This behavior implies that both $N$ and $s$ change during, and as a result of, adaptive sweeps.
-
-In practice, many population genetic models simply ignore density-dependent selection, and and each type is assigned a constant relative fitness.
-Density either does not enter the model at all, or if finite-population size effects (``drift'') are important, then $N$ is assumed to have reached some fixed equilibrium value.
-Selection is typically assumed to operate through viability, but the ecological processes underlying the regulation of population density are left unspecified (e.g.
-\citealt{gillespie_2004}; \citealt{nagylaki_1992}; \citealt{ewens_2004}).
-It would therefore seem that the density-independent-$s$, constant-$N$ setup (Fig.~\ref{fig:Ksel}b) is incompatible with a broad class of population ecological processes driving selection.
-
-%This behavior is intuitively appealing because births balance deaths at equilibrium, and adaptive mutants must overcome the existing balance to expand \citep[pp. 227]{kostitzin_1939}.
 
@@
 
-However, the preceding arguments do not imply that the constant-$s$ approximation \textit{only} applies under weak selection and stable $N$.
+The preceding arguments do not imply that the constant-$s$ approximation \textit{only} applies under weak selection and stable $N$.
 Every model of density-regulated growth mentioned above is missing an important aspect of selection which has ramifications for the constant-$s$ approximation: a ``reproductive excess'' of juveniles that are more fragile than their adult counterparts \citep{turner1968population}.
@@
 \citealt{nicholson_1954}),
+
+%This behavior is intuitively appealing because births balance deaths at equilibrium, and adaptive mutants must overcome the existing balance to expand \citep[pp. 227]{kostitzin_1939}.
 

### 2018-03-20 manuscript.tex
l

@@
 
-In practice, many population genetic models simply assign a constant relative fitness to each type, and ignore density dependence.
+In practice, many population genetic models simply ignore density dependence and assign a constant relative fitness to each type.
 Selection is typically interpreted as operating through viability, but the ecological processes underlying the regulation of population density are left unspecified (e.g.
@@
 A rather different picture emerges in more ecologically-explicit studies of selection in density-regulated populations.
-A major focus of previous studies in this area is the evolution of traits that affect the ability to keep growing at higher densities \citep{kostitzin_1939,macarthur_1967,roughgarden_1979,christiansen_2004}.
-This focus stems in part from Fisher's suggestion that evolution tends to increase density in the long term \citep{fisher_1930, lande_2009}, and MacArthur's concept of $K$-selection \citep{macarthur_1962}.
-But more broadly, density changes in response to selection in many models of density-regulated growth, including simple birth-death \citep{kostitzin_1939} and logistic models (Fig.~\ref{fig:Ksel}a; \citealt{macarthur_1962,roughgarden_1979,boyce_1984}), similar ``toy'' models with more general functional forms for the absolute fitness penalties of crowding \citep{kimura_1978,charlesworth_1971,lande_2009,nagylaki_1979,lande_2009}, the ``$R^*$ rule'' of resource competition theory (which states that the type able to deplete a shared limiting consumable resource to the lowest equilibrium density $R^*$ excludes the others; \citealt{grover_1997}), and the Lotka-Volterra competition model (at least during a sweep, except in special cases; \citealt{gill_1974,smouse_1976,mallet_2012}).
-This behavior implies that both $N$ and $s$ change during, and as a result of, adaptive sweeps.
+Following Fisher's suggestion that evolution tends to increase density in the long term \citep{fisher_1930, leon_1978, lande_2009}, as well as the influential concept of $K$-selection (specifically, the idea that selection in crowded conditions favors greater equilibrium density; \citealt{macarthur_1962}), many studies of density-regulated growth have focused on the response of density to selection \citep{kostitzin_1939,macarthur_1967,roughgarden_1979,christiansen_2004}.
+Indeed, selection in crowded conditions is intimately connected to equilibrium density in many of the most widely used models of density-regulated population growth, including simple birth-death \citep{kostitzin_1939} and logistic models (Fig.~\ref{fig:Ksel}a; \citealt{macarthur_1962,roughgarden_1979,boyce_1984}), variants of these models using other functional forms for the absolute fitness penalties of crowding \citep{kimura_1978,charlesworth_1971,lande_2009,nagylaki_1979,lande_2009}, and the ``$R^*$ rule'' of resource competition theory (which states that the type able to deplete a shared limiting consumable resource to the lowest equilibrium density $R^*$ excludes the others; \citealt{grover_1997}).
+In these examples, both $N$ and $s$ change during, and as a result of, adaptive sweeps.
+Density also changes in response to selection in the Lotka-Volterra competition model, at least during a sweep (except in special cases; \citealt{gill_1974,smouse_1976,mallet_2012}).
 
@@
 
-In light of this difficulty, the assignment of a density- and frequency-independent relative fitness value to each type is typically justified as an approximation that holds when selection is weak and $N$ changes slowly (\citealt{kimura1969natural}; \citealt[pp.
+In light of the complications of density-dependence, the assignment of a density-independent relative fitness to each type can be justified as an approximation that holds when selection is weak and $N$ changes slowly (\citealt{kimura1969natural}; \citealt[pp.
 277]{ewens_2004}; \citealt[Chap.
@@
 
-The preceding arguments do not imply that the constant-$s$ approximation \textit{only} applies under weak selection and stable $N$.
-Every model of density-regulated growth mentioned above is missing an important aspect of selection which has ramifications for the constant-$s$ approximation: a ``reproductive excess'' of juveniles that are more fragile than their adult counterparts \citep{turner1968population}.
-Reproductive excesses allow selection to be concentrated at a juvenile ``bottleneck'', which means that the density of adults can remain constant even if strong selection is occurring on juveniles.
-Although similar ideas featured prominently in early debates on the regulation of population density (e.g.
-\citealt{nicholson_1954}),
-
-%This behavior is intuitively appealing because births balance deaths at equilibrium, and adaptive mutants must overcome the existing balance to expand \citep[pp. 227]{kostitzin_1939}.
-
-much of the later reproductive excess literature revolves around Haldane's ``cost of selection'' \citep{haldane_1957} rather than the issues surrounding the constant-$s$ approximation.
-
-For instance, \cite{kimura1969natural} simply assumed that $N$ should remain constant
-
-as a requirement and then derived variants of the logistic model consistent with this.
-\cite{nei1971fertility} proposed a model with an explicit representation of reproductive excess, but used an unusual model of competition based on pair-wise interactions which was only defined for at most two different types.
-
-In complex age-structured models, dependence on density is assumed to come from a ``critical age-group'', and this age group's density responds to selection in much the same way as found in the above models without age structure \citep[Chap.
-4]{charlesworth_1994} (e.g.
-adaptive evolution proceeds in the direction of increasing equilibrium density in the critical age-group).
-
-  As a result, the role of reproductive excesses in justifying the use of constant relative finesses is largely verbal.
-
-Since constant relative fitness models are such a powerful idealization, it is important to understand the specifics of how and when they fail when selection is not weak, or $N$ is not stable.
+The preceding arguments are well-known, and appear in various forms in standard textbooks (e.g.
+\citealt{crow_1970,begon_1990,nagylaki_1992,otto_2011,ewens_2004}).
+However, the constant relative fitness idealization of population genetics need not \textit{only} apply when selection is weak (or when selection in genuinely density-independent).
+Since this idealization is so powerful, it is important to understand the specifics of how and when it fails when selection is not weak, or $N$ is not stable.
 For instance, in wild \textit{Drosophila}, strong seasonally-alternating selection happens concurrently with large ``boom-bust'' density cycles \citep{messer_2016,bergland_14}.
@@
 
-\citep{charlesworth_1971,kimura_1978,leon_1978,nagylaki_1979}
+The models of density-regulated growth mentioned so far are all missing an aspect of selection with major ramifications for the constant-$s$ description of selection: a ``reproductive excess'' of juveniles that are more fragile than their adult counterparts \citep{turner1968population}.
+Reproductive excesses allow selection to be concentrated at a juvenile ``bottleneck'', which means that the density of adults can remain constant even if strong selection is occurring on juveniles.
+Although similar ideas featured prominently in early debates on the regulation of population density (e.g.
+\citealt{nicholson_1954}), few attempts have been made to explicitly model density-dependent selection in the presence of reproductive excesses.
+\cite{nei1971fertility} proposed a model with an explicit reproductive excess as part of a discussion of Haldane's ``cost of selection'' \citep{haldane_1957}, but used an unusual model of competition which was defined for at most two different types.
+In more complex models with detailed age structure (as opposed to juvenile vs adult), it is usually assumed that the density of a ``critical age-group'' controls the population's response to crowding, but otherwise the view of density dependence is surprisingly similar to that found in the above models without age structure \citep[Chap.
+4]{charlesworth_1994}.
+For example, selection proceeds in the direction of increasing equilibrium density in the critical age-group \citep[pp.
+148]{charlesworth_1994}.
+As a result, the role of reproductive excesses in justifying the use of constant relative finesses is still largely verbal.
 
-Here we study the population ecology of relative fitness using a novel model of density-dependent population growth based on territorial contests.
+Here we investigate the population ecology of relative fitness using a novel model of density-regulated population growth based on territorial contests, which explicitly incorporates a reproductive excess.
 We restrict our attention to asexual haploids, since it is then clearer how the success or failure of the constant-$s$ description is tied to the underlying population ecological assumptions.
 Our starting point is the classic lottery model which was developed by ecologists to study competition driven by territorial contests in reef fishes and plants \citep{sale_77,chesson_1981}.
-The classic lottery incorporates a reproductive excess, and fitness involves a product of fertility and juvenile viability akin to standard population genetic models of selection (e.g.
+The classic lottery already incorporates a reproductive excess, and fitness involves a product of fertility and juvenile viability akin to standard population genetic models of selection (e.g.
 \citealt[pp.
 185]{crow_1970}).
-The classic lottery also assumes constant $N$, and generations can overlap.
-Our first task is to relax the constant-$N$ requirement of the lottery model to create a variable-density lottery (sections ``Model'' and ``Analytical approximation of the variable-density lottery'').
+However, the classic lottery also assumes constant $N$.
+Our first task is to relax the constant-$N$ assumption of the lottery model to create a variable-density lottery (sections ``Model'' and ``Analytical approximation of the variable-density lottery'').
 

### 2018-03-20 manuscript.tex
l

@@
 
-The preceding arguments are well-known, and appear in various forms in standard textbooks (e.g.
-\citealt{crow_1970,begon_1990,nagylaki_1992,otto_2011,ewens_2004}).
-However, the constant relative fitness idealization of population genetics need not \textit{only} apply when selection is weak (or when selection in genuinely density-independent).
-Since this idealization is so powerful, it is important to understand the specifics of how and when it fails when selection is not weak, or $N$ is not stable.
+However, the preceding arguments do not imply that the constant relative fitness idealization of population genetics \textit{only} applies when selection is weak and $N$ is stable (or when selection is actually density-independent).
+This idealization is powerful, and so it is important to understand the specifics of how it succeeds or fails when selection is not weak, or $N$ is not stable.
 For instance, in wild \textit{Drosophila}, strong seasonally-alternating selection happens concurrently with large ``boom-bust'' density cycles \citep{messer_2016,bergland_14}.
@@
 
-The models of density-regulated growth mentioned so far are all missing an aspect of selection with major ramifications for the constant-$s$ description of selection: a ``reproductive excess'' of juveniles that are more fragile than their adult counterparts \citep{turner1968population}.
+Here we argue that the simplified models of density-regulated growth mentioned above give a narrow view of density-dependent selection that is misleading in some important ways.
+These ultimately derive from their failure to account for the presence of ``reproductive excesses'', that is, an excess of juveniles that are more fragile than their adult counterparts \citep{turner1968population}.
 Reproductive excesses allow selection to be concentrated at a juvenile ``bottleneck'', which means that the density of adults can remain constant even if strong selection is occurring on juveniles.
@@
 \cite{nei1971fertility} proposed a model with an explicit reproductive excess as part of a discussion of Haldane's ``cost of selection'' \citep{haldane_1957}, but used an unusual model of competition which was defined for at most two different types.
-In more complex models with detailed age structure (as opposed to juvenile vs adult), it is usually assumed that the density of a ``critical age-group'' controls the population's response to crowding, but otherwise the view of density dependence is surprisingly similar to that found in the above models without age structure \citep[Chap.
+In more complex models with detailed age structure (as opposed to juvenile vs adult), it is usually assumed that the density of a ``critical age-group'' controls the population's response to crowding, but otherwise the view of density dependence is surprisingly similar to that found in the models mentioned above, which have no age structure \citep[Chap.
 4]{charlesworth_1994}.
@@
 148]{charlesworth_1994}.
-As a result, the role of reproductive excesses in justifying the use of constant relative finesses is still largely verbal.
+As a result, the extent to which reproductive excesses justify the use of constant relative finesses is unclear beyond heuristic verbal arguments.
 
-Here we investigate the population ecology of relative fitness using a novel model of density-regulated population growth based on territorial contests, which explicitly incorporates a reproductive excess.
+We investigate the population ecology of relative fitness using a novel model of density-regulated population growth based on territorial contests, which explicitly incorporates a reproductive excess.
 We restrict our attention to asexual haploids, since it is then clearer how the success or failure of the constant-$s$ description is tied to the underlying population ecological assumptions.

### 2018-03-23 manuscript.tex
l

@@
 
-There are a variety of different measures of fitness.
-Some widely used examples are expected lifetime reproductive ratio $R_0$, intrinsic population growth rate $r$, equilibrium population density/carrying capacity (often labeled ``$K$'') \citep{benton_2000}, and invasion fitness \citep{metz_1992}.
+There are a variety of different measures of fitness, such as expected lifetime reproductive ratio $R_0$, intrinsic population growth rate $r$, equilibrium population density/carrying capacity (often labeled ``$K$'') \citep{benton_2000}, and invasion fitness \citep{metz_1992}.
 In addition, ``relative fitness'' is widely used in evolutionary genetics, where the focus is on relative genotypic frequencies \cite[pp.
@@
 where $s=r_i-r_j$ is the selection coefficient of type $i$ relative to type $j$.
-Note that, in the discrete case, we can replace the $W_i$ with any set of values proportional to the $W_i$ without affecting the  ratio $W_i/\overline{W}$ or $\Delta p_i$.
+Note that we can replace the $W_i$ with any set of values proportional to the $W_i$ without affecting the  ratio $W_i/\overline{W}$ or $\Delta p_i$.
 These ``relative fitness'' values tell us how type frequencies change, but give no information about the dynamics of total population density $N=\sum_i n_i$ \citep[pp.
 468]{barton_2007}.
-Similarly, the selection coefficient and frequency dynamics in Eq.~\eqref{eq:canonical} would be unaffected if we added an arbitrary constant to the Malthusian parameters $r_i$ (these would be relative log fitnesses).
+Similarly in the continuous case, the selection coefficient and frequency dynamics in Eq.~\eqref{eq:canonical} would be unaffected if we added an arbitrary constant to the Malthusian parameters $r_i$ (these would be relative log fitnesses).
 
-In a constant environment, and in the absence of crowding, $r_i$ is a constant ``intrinsic'' growth rate (for concreteness we use Eq.~\eqref{eq:canonical} as our point of comparison, but the discrete time case is similar).
-The selection coefficient $s$ is then simply the difference in intrinsic population growth rates.
+In a constant environment, and in the absence of crowding, $r_i$ is an ``intrinsic'' population growth rate (for concreteness we use Eq.~\eqref{eq:canonical} as our point of comparison, but the discrete time case is similar).
+The selection coefficient $s$ is then simply the difference in intrinsic growth rates.
 However, growth cannot continue at a constant rate indefinitely.
-As population density increases, crowding will cause the Malthusian parameters $r_i$ to decline (e.g.
+As population density increases, the Malthusian parameters $r_i$ will eventually start to decline due to crowding (e.g.
 \citealt[pp.
 203]{begon_1990}).
-It is then possible that selection is density-dependent, and indeed this is probably not uncommon, because crowded and uncrowded conditions can favor very different traits \citep{travis_2013}.
-We would then need to know the dynamics of $N$ to use Eq.~\eqref{eq:canonical}.
+When population growth is density-regulated, selection can be density-dependent, and indeed this is probably not uncommon, because crowded and uncrowded conditions can favor very different traits \citep{travis_2013}.
+Eq.~\eqref{eq:canonical} is then not a complete description of selection --- it lacks an additional coupled equation describing the dynamics of $N$, on which $s$ in Eq.~\eqref{eq:canonical} depends.
 In general we cannot simply specify the dynamics of $N$ independently, because those ecological dynamics are coupled with the evolutionary dynamics of type frequency \citep{travis_2013}.
@@
 
-In light of the complications of density-dependence, the assignment of a density-independent relative fitness to each type can be justified as an approximation that holds when selection is weak and $N$ changes slowly (\citealt{kimura1969natural}; \citealt[pp.
+In light of the complications of density-dependence, the assignment of a density-independent relative fitness to each type has been justified as an approximation that holds when selection is weak and $N$ changes slowly (\citealt{kimura1969natural}; \citealt[pp.
 277]{ewens_2004}; \citealt[Chap.
@@
 However, the preceding arguments do not imply that the constant relative fitness idealization of population genetics \textit{only} applies when selection is weak and $N$ is stable (or when selection is actually density-independent).
-This idealization is powerful, and so it is important to understand the specifics of how it succeeds or fails when selection is not weak, or $N$ is not stable.
+The idealization of assigning relative fitness values to genotypes is powerful, and so it is important to understand the specifics of when and how it succeeds or fails when selection is not weak, or $N$ is not stable.
 For instance, in wild \textit{Drosophila}, strong seasonally-alternating selection happens concurrently with large ``boom-bust'' density cycles \citep{messer_2016,bergland_14}.

### 2018-03-26 manuscript.tex
l

@@
 
-Here we argue that the simplified models of density-regulated growth mentioned above give a narrow view of density-dependent selection that is misleading in some important ways.
-These ultimately derive from their failure to account for the presence of ``reproductive excesses'', that is, an excess of juveniles that are more fragile than their adult counterparts \citep{turner1968population}.
-Reproductive excesses allow selection to be concentrated at a juvenile ``bottleneck'', which means that the density of adults can remain constant even if strong selection is occurring on juveniles.
-Although similar ideas featured prominently in early debates on the regulation of population density (e.g.
-\citealt{nicholson_1954}), few attempts have been made to explicitly model density-dependent selection in the presence of reproductive excesses.
+Here we argue that the simplified models of density-regulated growth mentioned above are misleading in some important ways.
+These ultimately derive from their failure to account for ``reproductive excesses'', that is, an excess of juveniles that are more fragile than their adult counterparts \citep{turner1968population}.
+Reproductive excesses have major repercussions for the interplay between density and selection, because they allow selection to be concentrated at a juvenile ``bottleneck'', which means that the density of adults can remain constant even if strong selection is occurring on juveniles.
+This concept is familiar to geneticists, as exemplified by the Wright-Fisher model, where each generation involves the production of an infinite number of zygotes, the death of all $N$ adults, followed by the sampling of $N$ of these zygotes to form the next generation of adults.
+Similarly, in the Moran model, there is always a juvenile ready to replace an adult death, and so $N$ remains constant.
+
+The Wright-Fisher and Moran models have infinite reproductive excess --- population density $N$ remains constant no matter how strong the selection on juveniles.
+In reality,  reproductive excesses are finite, and the dynamics of density-dependent selection will depend on the magnitude of excess.
+Modeling finite excesses are not straightforward
+
+Moreover, similar ideas featured prominently in early debates on the regulation of population density (e.g. \citealt{nicholson_1954}).
+
+few attempts have been made to explicitly  It is not straightforward do so, because
+
 \cite{nei1971fertility} proposed a model with an explicit reproductive excess as part of a discussion of Haldane's ``cost of selection'' \citep{haldane_1957}, but used an unusual model of competition which was defined for at most two different types.

### 2018-03-26 manuscript.tex
l

@@
 These ultimately derive from their failure to account for ``reproductive excesses'', that is, an excess of juveniles that are more fragile than their adult counterparts \citep{turner1968population}.
-Reproductive excesses have major repercussions for the interplay between density and selection, because they allow selection to be concentrated at a juvenile ``bottleneck'', which means that the density of adults can remain constant even if strong selection is occurring on juveniles.
-This concept is familiar to geneticists, as exemplified by the Wright-Fisher model, where each generation involves the production of an infinite number of zygotes, the death of all $N$ adults, followed by the sampling of $N$ of these zygotes to form the next generation of adults.
-Similarly, in the Moran model, there is always a juvenile ready to replace an adult death, and so $N$ remains constant.
+This has major repercussions for the interplay between density and selection, because selection can be concentrated at a juvenile ``bottleneck'', which means that the density of adults can remain constant even if strong selection is occurring on juveniles.
+Reproductive excesses featured prominently in early debates on the regulation of population density (e.g.
+\citealt{nicholson_1954}).
+They are also conceptually familiar in evolutionary biology, particularly in relation to Haldane's ``cost of selection'' \citep{haldane_1957,turner1968population}.
+Reproductive excesses even appear implicitly in foundational evolutionary-genetic models like the Wright-Fisher, where each non-overlapping generation involves the production of an infinite number of zygotes, a constant number $N$ of which are sampled to form the next generation of adults.
+Likewise, in the Moran model, there are always juveniles available to replace dead adults, and so $N$ remains constant.
+Nevertheless, few attempts have been made to study density-dependent selection in the presence of a reproductive excess.
 
-The Wright-Fisher and Moran models have infinite reproductive excess --- population density $N$ remains constant no matter how strong the selection on juveniles.
-In reality,  reproductive excesses are finite, and the dynamics of density-dependent selection will depend on the magnitude of excess.
+More specifically, what is missing is the presence of a finite reproductive excess, rather than an infinite excess as in the Wright-Fisher.
+
+the dynamics of both selection and density depend on the magnitude of a
+
+Yet the properties of density-dependent selection in the presence of reproductive excesses
+
+Wright-Fisher and Moran models both have inifinite excesses,
+
+few attempts have been made to model the consequences of finite reproductive excesses,
+
+ explicit models of finite reproductive excesses
+
+Reproductive excesses are challenging to model in this context,
+
+In reality, reproductive excesses are finite,
+
+ Modeling density-dependent selection in the presence of finite reproductive excesses
+
 Modeling finite excesses are not straightforward
 
-Moreover, similar ideas featured prominently in early debates on the regulation of population density (e.g. \citealt{nicholson_1954}).
-
-few attempts have been made to explicitly  It is not straightforward do so, because
-
-\cite{nei1971fertility} proposed a model with an explicit reproductive excess as part of a discussion of Haldane's ``cost of selection'' \citep{haldane_1957}, but used an unusual model of competition which was defined for at most two different types.
+\cite{nei1971fertility} proposed a model with an explicit reproductive excess as part of a , but used an unusual model of competition which was defined for at most two different types.
 In more complex models with detailed age structure (as opposed to juvenile vs adult), it is usually assumed that the density of a ``critical age-group'' controls the population's response to crowding, but otherwise the view of density dependence is surprisingly similar to that found in the models mentioned above, which have no age structure \citep[Chap.

### 2018-03-27 manuscript.tex
l

--- 2018-03-23 manuscript.tex
@@
 
-Here we argue that the simplified models of density-regulated growth mentioned above give a narrow view of density-dependent selection that is misleading in some important ways.
-These ultimately derive from their failure to account for the presence of ``reproductive excesses'', that is, an excess of juveniles that are more fragile than their adult counterparts \citep{turner1968population}.
-Reproductive excesses allow selection to be concentrated at a juvenile ``bottleneck'', which means that the density of adults can remain constant even if strong selection is occurring on juveniles.
-Although similar ideas featured prominently in early debates on the regulation of population density (e.g.
-\citealt{nicholson_1954}), few attempts have been made to explicitly model density-dependent selection in the presence of reproductive excesses.
-\cite{nei1971fertility} proposed a model with an explicit reproductive excess as part of a discussion of Haldane's ``cost of selection'' \citep{haldane_1957}, but used an unusual model of competition which was defined for at most two different types.
-In more complex models with detailed age structure (as opposed to juvenile vs adult), it is usually assumed that the density of a ``critical age-group'' controls the population's response to crowding, but otherwise the view of density dependence is surprisingly similar to that found in the models mentioned above, which have no age structure \citep[Chap.
+Here we argue that the simplified models of density-regulated growth mentioned above are misleading in important ways.
+These ultimately derive from their failure to account for ``reproductive excesses'', that is, an excess of juveniles that are more fragile than their adult counterparts \citep{turner1968population}.
+Reproductive excesses allow selection to be concentrated at a juvenile ``bottleneck'', potentially keeping the density of adults constant even if selection is strong.
+The reproductive excess concept featured prominently in early debates on the regulation of population density (e.g.
+\citealt{nicholson_1954}), and is also familiar to evolutionists, particularly in relation to Haldane's ``cost of selection'' \citep{haldane_1957,turner1968population}.
+Reproductive excesses are also implicit in foundational evolutionary-genetic models like the Wright-Fisher, where each non-overlapping generation involves the production of an infinite number of zygotes, a constant number $N$ of which are sampled to form the next generation of adults.
+Likewise, in the Moran model, a juvenile is available to replace a dead adult every iteration no matter how rapidly adults are dying, and as a result $N$ remains constant.
+
+Nevertheless, few studies of density-dependent selection have incorporated reproductive excesses.
+This requires density-regulated population growth in which the excess is finite and, in general, density-dependent.
+Such models are potentially much more complicated than those with no excess (e.g.
+logistic) or an infinite excess (e.g.
+Wright-Fisher).
+Nei's competitive selection model incorporated a finite reproductive excess to help clarify the ``cost of selection'' \citep{nei1971fertility,nagylaki_1992}, but used an unusual representation of competition based on pairwise interactions defined for at most two different types, and was also restricted to equal fertilities for each type.
+In more complex models with detailed age structure (as opposed to simply juveniles vs adults), it is usually assumed that the density of a ``critical age-group'' controls the population's response to crowding.
+The focus is then on the density in the critical age group,  and the role of density dependence is surprisingly similar to that found in models without any age structure \citep[Chap.
 4]{charlesworth_1994}.
-For example, selection proceeds in the direction of increasing equilibrium density in the critical age-group \citep[pp.
+For example, selection proceeds in the direction of increasing equilibrium density in the critical age-group, akin to classical $K$-selection \citep[pp.
 148]{charlesworth_1994}.
-As a result, the extent to which reproductive excesses justify the use of constant relative finesses is unclear beyond heuristic verbal arguments.
+This gives little insight into the role of reproductive excesses in stabilizing adult densities.
 
@@
 Our first task is to relax the constant-$N$ assumption of the lottery model to create a variable-density lottery (sections ``Model'' and ``Analytical approximation of the variable-density lottery'').
-
-We then outline the selection properties of our new variable-density lottery model (section ``Density-dependent selection in the variable-density lottery''). Next, we discuss
-selection on the ability to contest territories, which behaves like a pure constant-$N$, constant-$s$ trait, and discuss how this fits with the classical notion of $K$-selection  (section ``The response of density to selection; $c$-selection versus $K$-selection'').
-We then consider selection on traits that regulate population density (section ``Density-regulating traits and the threat of strong selection''), and conclude by contrasting the classical density-dependent selection literature with our results (``Discussion'').
 
@@
 We assume that each adult from type $i$ produces $b_i$ propagules on average, so that the mean number of $i$ propagules dispersing to unoccupied territories is $m_i=b_in_iU/T$.
-The parameter $b_i$ can be thought of as a measure of ``colonization ability'', which combines fecundity and dispersal ability \citep{levins_71,tilman_94}.
+The parameter $b_i$ can be thought of as a measure of ``colonization ability'', which combines fertility and dispersal ability \citep{levins_71,tilman_94}.
 Random dispersal is then modeled using a Poisson distribution $p_i(x_i)=l_i^{x_i} e^{-l_i}/x_i!$ for the number $x_i$ of $i$ propagules dispersing to any particular unoccupied territory, where $l_i=m_i/U$ is the mean propagule density in unoccupied territories.

### 2018-03-27 manuscript.tex
l

@@
 
-In light of the complications of density-dependence, the assignment of a density-independent relative fitness to each type has been justified as an approximation that holds when selection is weak and $N$ changes slowly (\citealt{kimura1969natural}; \citealt[pp.
+In light of the complications of density-dependence, the assignment of density-independent relative fitnesses has been justified as an approximation that holds when selection is weak and $N$ changes slowly (\citealt{kimura1969natural}; \citealt[pp.
 277]{ewens_2004}; \citealt[Chap.
@@
 
-Here we argue that the simplified models of density-regulated growth mentioned above are misleading in important ways.
-These ultimately derive from their failure to account for ``reproductive excesses'', that is, an excess of juveniles that are more fragile than their adult counterparts \citep{turner1968population}.
-Reproductive excesses allow selection to be concentrated at a juvenile ``bottleneck'', potentially keeping the density of adults constant even if selection is strong.
-The reproductive excess concept featured prominently in early debates on the regulation of population density (e.g.
-\citealt{nicholson_1954}), and is also familiar to evolutionists, particularly in relation to Haldane's ``cost of selection'' \citep{haldane_1957,turner1968population}.
-Reproductive excesses are also implicit in foundational evolutionary-genetic models like the Wright-Fisher, where each non-overlapping generation involves the production of an infinite number of zygotes, a constant number $N$ of which are sampled to form the next generation of adults.
-Likewise, in the Moran model, a juvenile is available to replace a dead adult every iteration no matter how rapidly adults are dying, and as a result $N$ remains constant.
+Here we argue that the simplified models of density-regulated growth mentioned above are misleading in their representation of the interplay between selection and density.
+This ultimately derives from their failure to account for ``reproductive excess'', that is, an excess of juveniles that experience stronger selection than their adult counterparts \citep{turner1968population}.
+This allows selection to be concentrated at a juvenile ``bottleneck'', potentially keeping the density of adults constant even under strong selection.
+Reproductive excess featured prominently in early debates about the regulation of population density (e.g.
+\citealt{nicholson_1954}), and also has a long history in evolutionary theory, particularly related to Haldane's ``cost of selection'' \citep{haldane_1957,turner1968population}.
+Additionally, reproductive excess is implicit in foundational evolutionary-genetic models like the Wright-Fisher, where each non-overlapping generation involves the production of an infinite number of zygotes, of which a constant number $N$ are sampled to form the next generation of adults.
+Likewise, in the Moran model, a juvenile is always available to replace a dead adult every iteration no matter how rapidly adults are dying, and as a result $N$ remains constant.
 
-Nevertheless, few studies of density-dependent selection have incorporated reproductive excesses.
-This requires density-regulated population growth in which the excess is finite and, in general, density-dependent.
-Such models are potentially much more complicated than those with no excess (e.g.
-logistic) or an infinite excess (e.g.
+Nevertheless, studies of density-dependent selection rarely incorporate reproductive excess.
+This requires that we model a finite, density-dependent excess, which is significantly more complicated zero (e.g.
+logistic) or infinite excess (e.g.
 Wright-Fisher).
-Nei's competitive selection model incorporated a finite reproductive excess to help clarify the ``cost of selection'' \citep{nei1971fertility,nagylaki_1992}, but used an unusual representation of competition based on pairwise interactions defined for at most two different types, and was also restricted to equal fertilities for each type.
-In more complex models with detailed age structure (as opposed to simply juveniles vs adults), it is usually assumed that the density of a ``critical age-group'' controls the population's response to crowding.
-The focus is then on the density in the critical age group,  and the role of density dependence is surprisingly similar to that found in models without any age structure \citep[Chap.
-4]{charlesworth_1994}.
-For example, selection proceeds in the direction of increasing equilibrium density in the critical age-group, akin to classical $K$-selection \citep[pp.
-148]{charlesworth_1994}.
-This gives little insight into the role of reproductive excesses in stabilizing adult densities.
+Nei's competitive selection model incorporated a finite reproductive excess to help clarify the ``cost of selection'' \citep{nei1971fertility,nagylaki_1992}, but used an unusual representation of competition based on pairwise interactions defined for at most two different genotypes, and was also restricted to equal fertilities for each type.
+In models with detailed age structure, the role of density is usually absorbed into one ``critical age-group'' that mediates the population's response to crowding.
+The existence of a reproductive excess implies that the critical age-group would have to occur at a pre-reproductive age.
+As a result, the focus shifts to juvenile density, obscuring the crucial connection between pre-reproductive-age selection, and post-reproductive-age density.
+For example, the result that selection proceeds in the direction of increasing equilibrium density in the critical age-group \citep[pp.
+148]{charlesworth_1994} is a form of the classical $K$-selection ideas discussed above applied to the critical age-group, and says nothing about other age-groups.
 
-We investigate the population ecology of relative fitness using a novel model of density-regulated population growth based on territorial contests, which explicitly incorporates a reproductive excess.
-We restrict our attention to asexual haploids, since it is then clearer how the success or failure of the constant-$s$ description is tied to the underlying population ecological assumptions.
+We develop a novel model of density-regulated population growth based on territorial contests, which explicitly incorporates a reproductive excess.
 Our starting point is the classic lottery model which was developed by ecologists to study competition driven by territorial contests in reef fishes and plants \citep{sale_77,chesson_1981}.
@@
 185]{crow_1970}).
-However, the classic lottery also assumes constant $N$.
-Our first task is to relax the constant-$N$ assumption of the lottery model to create a variable-density lottery (sections ``Model'' and ``Analytical approximation of the variable-density lottery'').
+However, the classic lottery also assumes constant $N$ (and infinite reproductive excess).
+Our first task is to relax the constant-$N$ assumption of the lottery model to create a variable-density lottery.
+
+We restrict our attention to asexual haploids, since it is then clearer how the success or failure of the constant-$s$ description is tied to the underlying population ecological assumptions.
 

### 2018-04-04 manuscript.tex
l

@@
 
-For haploids growing in discrete time steps, the change in the abundance $n_i$ of type $i$ over a time step can be expressed as $\Delta n_i=(W_i - 1)n_i$ where $W_i$ is ``absolute fitness''.
+For haploids tracked in discrete time, the change in the abundance $n_i$ of type $i$ over a time step can be expressed as $\Delta n_i=(W_i - 1)n_i$ where $W_i$ is ``absolute fitness''.
 The corresponding change in frequency is $\Delta p_i=\left(\frac{W_i}{\overline{W}}-1\right) p_i$, where $\overline{W}=\sum_i W_i p_i$.
@@
 468]{barton_2007}.
-Similarly in the continuous case, the selection coefficient and frequency dynamics in Eq.~\eqref{eq:canonical} would be unaffected if we added an arbitrary constant to the Malthusian parameters $r_i$ (these would be relative log fitnesses).
+Similarly in the continuous case, the selection coefficient and frequency dynamics in Eq.~\eqref{eq:canonical} would be unaffected if we added an arbitrary constant to the Malthusian parameters $r_i$ (these would then be relative log fitnesses).
 
-In a constant environment, and in the absence of crowding, $r_i$ is an ``intrinsic'' population growth rate (for concreteness we use Eq.~\eqref{eq:canonical} as our point of comparison, but the discrete time case is similar).
+In a constant environment, and in the absence of crowding, $r_i$ is an ``intrinsic'' population growth rate (for concreteness we use the continuous time Eq.~\eqref{eq:canonical} as our point of comparison, but the discrete time case is similar).
 The selection coefficient $s$ is then simply the difference in intrinsic growth rates.
-However, growth cannot continue at a constant rate indefinitely.
-As population density increases, the Malthusian parameters $r_i$ will eventually start to decline due to crowding (e.g.
+However, growth cannot continue at a non-zero constant rate indefinitely: the population is not viable if $r_i<0$, whereas  $r_i>0$ implies endlessly increasing population density.
+Thus, assuming viability, the increase in population density must be checked by crowding, implying that the Malthusian parameters $r_i$ eventually decline to zero (e.g.
 \citealt[pp.
 203]{begon_1990}).
-When population growth is density-regulated, selection can be density-dependent, and indeed this is probably not uncommon, because crowded and uncrowded conditions can favor very different traits \citep{travis_2013}.
+Selection can then be density-dependent, and indeed this is probably not uncommon, because crowded and uncrowded conditions can favor very different traits \citep{travis_2013}.
 Eq.~\eqref{eq:canonical} is then not a complete description of selection --- it lacks an additional coupled equation describing the dynamics of $N$, on which $s$ in Eq.~\eqref{eq:canonical} depends.
@@
 \citealt{gillespie_2004}; \citealt{nagylaki_1992}; \citealt{ewens_2004}).
-Density either does not enter the model at all, or if finite-population size effects (``drift'') are important, then $N$ is assumed to have reached some fixed equilibrium value (Fig.~\ref{fig:Ksel}b).
+Density either does not enter the model at all, or if finite-population size effects (``random genetic drift'') are important, then $N$ is assumed to have reached some fixed equilibrium value (Fig.~\ref{fig:Ksel}b).
 
-A rather different picture emerges in more ecologically-explicit studies of selection in density-regulated populations.
+A rather different picture emerges in more ecologically explicit studies of selection in density-regulated populations.
 Following Fisher's suggestion that evolution tends to increase density in the long term \citep{fisher_1930, leon_1978, lande_2009}, as well as the influential concept of $K$-selection (specifically, the idea that selection in crowded conditions favors greater equilibrium density; \citealt{macarthur_1962}), many studies of density-regulated growth have focused on the response of density to selection \citep{kostitzin_1939,macarthur_1967,roughgarden_1979,christiansen_2004}.
-Indeed, selection in crowded conditions is intimately connected to equilibrium density in many of the most widely used models of density-regulated population growth, including simple birth-death \citep{kostitzin_1939} and logistic models (Fig.~\ref{fig:Ksel}a; \citealt{macarthur_1962,roughgarden_1979,boyce_1984}), variants of these models using other functional forms for the absolute fitness penalties of crowding \citep{kimura_1978,charlesworth_1971,lande_2009,nagylaki_1979,lande_2009}, and the ``$R^*$ rule'' of resource competition theory (which states that the type able to deplete a shared limiting consumable resource to the lowest equilibrium density $R^*$ excludes the others; \citealt{grover_1997}).
-In these examples, both $N$ and $s$ change during, and as a result of, adaptive sweeps.
+Indeed, both $N$ and $s$ change during, and as a result of, adaptive sweeps in many of the most widely used models of density-regulated population growth, including simple birth-death \citep{kostitzin_1939} and logistic models (Fig.~\ref{fig:Ksel}a; \citealt{macarthur_1962,roughgarden_1979,boyce_1984}), variants of these models using other functional forms for the absolute fitness penalties of crowding \citep{kimura_1978,charlesworth_1971,lande_2009,nagylaki_1979,lande_2009}, and the ``$R^*$ rule'' of resource competition theory (which states that the type able to deplete a shared limiting consumable resource to the lowest equilibrium density $R^*$ excludes the others; \citealt{grover_1997}).
 Density also changes in response to selection in the Lotka-Volterra competition model, at least during a sweep (except in special cases; \citealt{gill_1974,smouse_1976,mallet_2012}).
+
+Another issue with the constant-$N$, constant-$s$ description of selection is that it precludes consideration of longer-term aspects of the interplay between evolution and ecology such as population extinction.
+A variety of approaches have been developed to address this in quantitative genetics \citep{burger1995evolution,engen_2013}, population genetics \citep{bertram2017predicting} and adaptive dynamics \citep{ferriere2013eco,dieckmann2004adaptive}.
+Although density-dependent selection is  pertinent to this longer-term issue, our focus here is the description of the time-dependent process by which selection changes allele frequencies.
+This is particularly critical for making sense of evolution at the genetic level, for which we now have abundant data.
 
@@
 \citealt{nicholson_1954}), and also has a long history in evolutionary theory, particularly related to Haldane's ``cost of selection'' \citep{haldane_1957,turner1968population}.
-Additionally, reproductive excess is implicit in foundational evolutionary-genetic models like the Wright-Fisher, where each non-overlapping generation involves the production of an infinite number of zygotes, of which a constant number $N$ are sampled to form the next generation of adults.
-Likewise, in the Moran model, a juvenile is always available to replace a dead adult every iteration no matter how rapidly adults are dying, and as a result $N$ remains constant.
+Additionally, reproductive excess is implicit in foundational evolutionary-genetic models like the Wright-Fisher, where each generation involves the production of an infinite number of zygotes, of which a constant number $N$ are sampled to form the next generation of adults.
+Likewise in the Moran model, a juvenile is always available to replace a dead adult every iteration no matter how rapidly adults are dying, and as a result $N$ remains constant.
 
 Nevertheless, studies of density-dependent selection rarely incorporate reproductive excess.
-This requires that we model a finite, density-dependent excess, which is significantly more complicated zero (e.g.
-logistic) or infinite excess (e.g.
-Wright-Fisher).
-Nei's competitive selection model incorporated a finite reproductive excess to help clarify the ``cost of selection'' \citep{nei1971fertility,nagylaki_1992}, but used an unusual representation of competition based on pairwise interactions defined for at most two different genotypes, and was also restricted to equal fertilities for each type.
-In models with detailed age structure, the role of density is usually absorbed into one ``critical age-group'' that mediates the population's response to crowding.
-The existence of a reproductive excess implies that the critical age-group would have to occur at a pre-reproductive age.
-As a result, the focus shifts to juvenile density, obscuring the crucial connection between pre-reproductive-age selection, and post-reproductive-age density.
-For example, the result that selection proceeds in the direction of increasing equilibrium density in the critical age-group \citep[pp.
-148]{charlesworth_1994} is a form of the classical $K$-selection ideas discussed above applied to the critical age-group, and says nothing about other age-groups.
+This requires that we model a finite, density-dependent excess, which is substantially more complicated than modeling either zero (e.g.
+logistic) or infinite (e.g.
+Wright-Fisher) reproductive excess.
+Nei's ``competitive selection'' model incorporated a finite reproductive excess to help clarify the ``cost of selection'' \citep{nei1971fertility,nagylaki_1992}, but used an unusual representation of competition based on pairwise interactions defined for at most two different genotypes, and was also restricted to equal fertilities for each genotype.
 
-We develop a novel model of density-regulated population growth based on territorial contests, which explicitly incorporates a reproductive excess.
-Our starting point is the classic lottery model which was developed by ecologists to study competition driven by territorial contests in reef fishes and plants \citep{sale_77,chesson_1981}.
-The classic lottery already incorporates a reproductive excess, and fitness involves a product of fertility and juvenile viability akin to standard population genetic models of selection (e.g.
+In models with detailed age structure, it is usually assumed that the density of a ``critical age group'' mediates the population's response to crowding \citep[pp.
+54]{charlesworth_1994}.
+Reproductive excess is a special case corresponding to a critical pre-reproductive age group.
+A central result of the theory of density-regulated age-structured populations is that selection proceeds in the direction of increasing equilibrium density in the critical age group \citep[pp.
+148]{charlesworth_1994}.
+This is a form of the classical $K$-selection ideas discussed above, but restricted to the critical age group (juveniles, in this case).
+By focusing on density in the critical age group, this literature loses the interdependence of pre-reproductive selection and post-reproductive density.
+
+We develop a novel model of density-regulated population growth based on territorial contests, which explicitly incorporates a finite reproductive excess.
+Our starting point is the classic lottery model, which was developed by ecologists to study competition driven by territorial contests in reef fishes and plants \citep{sale_77,chesson_1981}.
+Each type is assumed to have three traits: fecundity $b$, mortality $d$, and competitive ability $c$.
+In each iteration of the classic lottery model, each type produces a large number of juveniles, such that $N$ remains constant (infinite reproductive excess).
+Competitive ability $c$ affects the probability of winning a territory, and behaves like a pure relative fitness trait.
+Thus, fitness involves a product of fertility and juvenile viability akin to standard population genetic models of selection (e.g.
 \citealt[pp.
 185]{crow_1970}).
-However, the classic lottery also assumes constant $N$ (and infinite reproductive excess).
-Our first task is to relax the constant-$N$ assumption of the lottery model to create a variable-density lottery.
+Our first task is to relax the large-juvenile-number assumption of the lottery model to create a variable-density lottery with a finite, density-dependent reproductive excess.
 
-We restrict our attention to asexual haploids, since it is then clearer how the success or failure of the constant-$s$ description is tied to the underlying population ecological assumptions.
+The properties of density-dependent selection in our model are strikingly different from the classical literature discussed above.
+The strong connection between crowding and selection for greater equilibrium density is broken: selection need not affect density, and even when it does, Eq.~\eqref{eq:canonical} is almost exact even for strong selection, provided that only one of the traits $b$, $c$, or $d$ is involved.
 
@@
 
+We restrict our attention to asexual haploids, since it is then clearer how the properties of selection are tied to the underlying population ecological assumptions.
 We assume that reproductively mature individuals (``adults'') require their own territory to survive and reproduce.
@@
 In a given iteration, the number of adults of the $i$'th type will be denoted by $n_i$, the total number of adults by $N=\sum_i n_i$, and the number of unoccupied territories by $U=T-N$.
-We assume that the $n_i$ are large enough that stochastic fluctuations in the $n_i$ (``drift'') can be ignored (with $T$ also assumed large to allow for low type densities $n_i/T$).
+We assume that the $n_i$ are large enough that stochastic fluctuations in the $n_i$ (drift) can be ignored (with $T$ also assumed large to allow for low type densities $n_i/T$).
 
@@
 \end{equation}
-Here the sum only includes territories with at least one propagule present. Since we do not consider drift here, we will not analyze the fluctuations around these two expectations.
+Here the sum only includes territories with at least one propagule present. Since we do not consider random genetic drift here, we will not analyze the fluctuations around these two expectations.
 
@@
 The first, $e^{-L}$, accounts for propagules which land alone on unoccupied territories; these propagules secure the territories without contest.
-The second, $R_i c_i/\overline{c}$, represents competitive victories on territories where only a single $i$ propagule lands, and at least one other propagule from a different type (this term dominates the growth of a rare invader in a high density population and determines invasion fitness).
+The second, $R_i c_i/\overline{c}$, represents competitive victories on territories where only a single $i$ propagule lands, together with at least one other propagule from a different type (this term dominates the growth of a rare invader in a high density population and determines invasion fitness).
 The third term, $A_i c_i/\overline{c}$, represents competitive victories in territories where two or more $i$ type propagules are present.
@@
 Simulations are conducted as follows: $x_1,x_2$ values are sampled $U=10^5$ times from Poisson distributions with respective means $l_1,l_2$, and the victorious type in each territory is then decided by random sampling weighted by the lottery win probabilities $c_ix_i/(c_1 x_1 + c_2 x_2)$.
-Simulation points are almost invisible for the common type due to near exact agreement with Eq.~\eqref{eq:master}.
 Dashed lines show the failure of the classic lottery model at low density.}
@@
 
-Selection in the variable density lottery model is density-dependent, by which we mean that  the selection coefficient, measured by the difference in per-capita growth rate $\Delta n_i/n_i$ between types, depends on $N$.
+Selection in our variable density lottery model is density-dependent, by which we mean that  the selection coefficient, measured by the difference in per-capita growth rate $\Delta n_i/n_i$ between types, depends on $N$.
 Density-dependent selection is sometimes taken to mean a qualitative change in which types are fitter than others at different densities \citep{travis_2013}.
-While reversal in the order of fitnesses and co-existence driven by density-regulation are possible in the variable-density lottery (a special case of the competition-colonization trade-off; \citealt{levins_71,tilman_94,bolker_99}), questions related to co-existence are tangential to our aims and will not be pursued further here.
+While reversal in the order of fitnesses and co-existence driven by density-regulation are possible in our variable-density lottery (a special case of the competition-colonization trade-off; \citealt{levins_71,tilman_94,bolker_99}), questions related to co-existence are tangential to our aims and will not be pursued further here.
 
@@
 
-We now turn to the issue of how density responds to selection, comparing the variable-density lottery to previous studies of density-regulated populations \citep{prout_1980}.
-As we saw in the previous section, $c$-selection has no effect on population density in the variable-density lottery.
+We now turn to the issue of how density responds to selection, comparing our variable-density lottery to previous studies of density-regulated populations \citep{prout_1980}.
+As we saw in the previous section, $c$-selection has no effect on population density in our variable-density lottery.
 To make sense of how $c$-selection fits with previous population growth models, we now revisit MacArthur's general treatment of $K$-selection \citep{macarthur_1967}.
@@
 
-By contrast, density trajectories for $c$-selection in the variable-density lottery converge on a line of constant equilibrium density (Fig.~\ref{fig:LVvslottery}b).
+By contrast, density trajectories for $c$-selection in our variable-density lottery converge on a line of constant equilibrium density (Fig.~\ref{fig:LVvslottery}b).
 This means that once the population reaches demographic equilibrium, it behaves indistinguishably from a constant-$N$ relative fitness model (Fig.~\ref{fig:Ksel}b).
-This complete uncoupling of density from $c$-selection arises due to the presence of an excess of propagules which pay the cost of selection without affecting adult density.
-As a result, Eq.~\eqref{eq:canonical} holds in equilibrium even though $c$-selection is density-dependent.
+This complete uncoupling of density from $c$-selection arises due to the presence of an excess of propagules which pay the cost of selection without affecting adult density \citep{nei1971fertility}.
+As a result, Eq.~\eqref{eq:canonical} holds in demographic equilibrium even though $c$-selection is density-dependent.
 
@@
 
-The previous section underscores the fact that density must be affected by ongoing selection to threaten Eq.~\eqref{eq:canonical}.
-We now turn to selection on density-regulating traits such as $b$ and $d$ in the variable density lottery.
+For Eq.~\eqref{eq:canonical} to break down, the selection coefficient $s$ must depend on density.
+As shown in Fig.~\ref{fig:DDS_lottery}, this is not the case for $d$; the selection coefficient for a $d$-variant with $d_2= d_1(1-\epsilon)$ is a constant $s=\epsilon d_1$.
+Eq.~\eqref{eq:canonical} also holds when the population is at demographic equilibrium and density is unaffected by the outcome of selection; as discussed in the previous section, the latter is the case for $c$-selection in our model.
+Thus, to threaten Eq.~\eqref{eq:canonical}, we require selection to be density-dependent, and also density to be changing.
+This can obviously occur if density-dependent selection is occurring in a population far from demographic equilibrium.
+In this case the validity of Eq.~\eqref{eq:canonical} depends on the specifics of the rate and magnitude of demographic change (we return to this in the Discussion).
+However, Eq.~\eqref{eq:canonical} can be threatened even in demographically-stable populations if a density-regulating trait is subject to density-dependent selection, as is the case for $b$ in our variable-density lottery.
 
-As shown in Fig.~\ref{fig:DDS_lottery}, $d$-selection is independent of density; the selection coefficient for a $d$-variant with $d_2= d_1(1-\epsilon)$ is a constant $s=\epsilon d_1$.
-Thus, to threaten the assumption of density-independent $s$ in an equilibrium population we require selection to act on a trait that is both density-dependent and density-regulating.
-
-This is the case for $b$ in the variable-density lottery.
-However, before we discuss this relatively complicated trait, it is helpful to summarize the threat to Eq.~\eqref{eq:canonical} in simpler models of density-regulated growth, as exemplified by the birth-death model \citep{kostitzin_1939}
+Before we discuss the $b$ trait, it is helpful to summarize the threat to Eq.~\eqref{eq:canonical} in simpler models of density-regulated growth, as exemplified by the birth-death model \citep{kostitzin_1939}
 \begin{equation}
@@
 Here $\delta_i$ is per-capita mortality due to crowding (for simplicity, there are no deaths when uncrowded).
-Starting from a type $1$ population in equilibrium (we consider the non-equilibrium case in BLAH), a variant with $\delta_2=\delta_1(1-\epsilon)$ has density-dependent selection coefficient $s=\epsilon \delta_1 N$ in Eq.~\eqref{eq:canonical}.
-This only affects the sweep if $N$ changes substantially as the population shifts to the new equilibrium.
-From Eq.~\eqref{eq:simplebirthdeath} we have $N_{\rm initial}=b_1/\delta_1$ and $N_{\rm final}=b_1/(\delta_1(1-\epsilon))=N_{\rm initial}/(1-\epsilon)$, and so $s_{\rm initial}= \epsilon b_1$ and $s_{\rm final}=s_{\rm initial}/(1-\epsilon)$.
-Consequently, substantial deviations from Eq.~\eqref{eq:canonical} occurs if there is sufficiently strong selection on $\delta$ (Fig.~\ref{fig:strengthofselection}; \citealt{kimura1969natural,crow_1970}).
+Starting from a type $1$ population in equilibrium, a variant with $\delta_2=\delta_1(1-\epsilon)$ has density-dependent selection coefficient $s=\epsilon \delta_1 N$ in Eq.~\eqref{eq:canonical}, which will change over the course of the sweep as $N$ shifts from its initial type $1$ equilibrium to a type $2$ equilibrium.
+From Eq.~\eqref{eq:simplebirthdeath}, the equilibrium densities at the beginning and end of the sweep are $N_{\rm initial}=b_1/\delta_1$ and $N_{\rm final}=b_1/(\delta_1(1-\epsilon))=N_{\rm initial}/(1-\epsilon)$ respectively, and so $s_{\rm initial}= \epsilon b_1$ and $s_{\rm final}=s_{\rm initial}/(1-\epsilon)$.
+Consequently, substantial deviations from Eq.~\eqref{eq:canonical} occur if there is sufficiently strong selection on $\delta$ (Fig.~\ref{fig:strengthofselection}; \citealt{kimura1969natural,crow_1970}).
 
@@
 \includegraphics[scale=0.8]{strengthofselection.pdf}
-\caption{\label{fig:strengthofselection} (a) Proportional change in the selection coefficient for a type that experiences proportionally $1-\epsilon$ fewer deaths induced by crowding.
+\caption{\label{fig:strengthofselection} (a) Change in the selection coefficient between the beginning and end of a sweep of a type that experiences proportionally $1-\epsilon$ fold fewer crowding-induced deaths.
 The population is in demographic equilibrium at the start and end of the sweep.
@@
 
-Equilibrium-to-equilibrium $b$-sweeps in the variable-density lottery are qualitatively different from $\delta$ sweeps, because greater $b$ not only means more propagules contesting territories, but also more territories being contested.
-Together, the net density-dependent effect on $b$-selection is zero; in Eq.~\eqref{eq:bdensitydependence}, since $b_i/\overline{b}=1$ in a single-type equilibrium, the density-dependence factor $f(\overline{b},N)=\frac{1-e^{-\overline{b}N/T}}{N}(T-N)$ is exactly equal to the constant mortality rate at the beginning and end of a $b$-sweep, even though $b$ and density change.
-During the sweep there is some deviation in $f(\overline{b},N)$, but this deviation is an order of magnitude smaller than for a $\delta$ sweep (the density-dependent deviation constant $s$ in Fig.~\ref{fig:strengthofselection} is of order $\epsilon$, whereas the analogous effect for $b$ sweep in the variable-density lottery is only of order $\epsilon^2$; see Appendix D for details).
+Equilibrium-to-equilibrium $b$-sweeps in our variable-density lottery are qualitatively different from $\delta$ sweeps in this simpler birth-death model, because greater $b$ not only means more propagules contesting territories, but also more territories being contested.
+Together, the net density-dependent effect on $b$-selection coefficients is zero; in Eq.~\eqref{eq:bdensitydependence}, since $b_i/\overline{b}=1$ in a single-type equilibrium, the density-dependence factor $f(\overline{b},N)=\frac{1-e^{-\overline{b}N/T}}{N}(T-N)$ is exactly equal to the constant mortality rate at the beginning and end of a $b$-sweep, even though $b$ and density change.
+During the sweep there is some deviation in $f(\overline{b},N)$, but this deviation is an order of magnitude smaller than for a $\delta$ sweep (the density-dependent deviation constant $s$ in Fig.~\ref{fig:strengthofselection} is of order $\epsilon$, whereas the analogous effect for $b$ sweep in our variable-density lottery is only of order $\epsilon^2$; see Appendix D for details).
 Since selection must already be strong for a $\delta$-sweep to threaten Eq.~\eqref{eq:canonical}, the density-independent model applies effectively exactly for equilibrium $b$-sweeps.
@@
 
-If selection acts simultaneously on more than one trait in the variable-density lottery, then evolution in a density-regulating trait can drive changes in the strength of selection on a trait subject to density-dependent selection (Fig.~\ref{fig:multiple}).
+If selection acts simultaneously on more than one trait in our variable-density lottery, then evolution in a density-regulating trait can drive changes in the strength of selection on a trait subject to density-dependent selection (Fig.~\ref{fig:multiple}).
 This can produce behavior analogous to selection on $\delta$ in Fig.~\ref{fig:strengthofselection}.
@@
 \includegraphics[scale=0.8]{multiple.pdf}
-\caption{\label{fig:multiple} Simultaneous selection on $d$ and $c$ in the variable-density lottery model as predicted by Eq.~\eqref{eq:master}.
+\caption{\label{fig:multiple} Simultaneous selection on $d$ and $c$ in our variable-density lottery model as predicted by Eq.~\eqref{eq:master}.
 Selection is not constant over the sweep because $d$ is density-regulating and $c$ is density-dependent.}
@@
 
-Summarizing the three traits in the variable-density lottery model: (i) $c$-selection is density-dependent, but $c$ does not regulate density; (ii) $d$ regulates density, but $d$-selection is density-independent; (iii) $b$ regulates density and $b$-selection is density-dependent.
+Summarizing the three traits in our variable-density lottery model: (i) $c$-selection is density-dependent, but $c$ does not regulate density; (ii) $d$ regulates density, but $d$-selection is density-independent; (iii) $b$ regulates density and $b$-selection is density-dependent.
 Yet despite their differences, pure $b$, $c$ and $d$ sweeps starting and ending at equilibrium all obey the density-independent relative fitness description of selection almost exactly.
+This behavior is quite different from that found in the classical density-dependent selection literature (see ``Introduction'' and ``The response of density to selection; $c$-selection versus $K$-selection'').
+In the latter, selection in crowded populations takes broadly one of two forms: selection for greater carrying capacity ($K$-selection) or selection on competition coefficients (``$\alpha$-selection''; \citealt{gill_1974,joshi_2001}).
+$K$ and $\alpha$ both behave like $\delta$ in Eq.~\eqref{eq:simplebirthdeath} in that strong selection is sufficient for Eq.~\eqref{eq:canonical} to break down (Fig.~\ref{fig:strengthofselection}).
 
-This behavior is quite different from that found in the classical density-dependent selection literature \citep{roughgarden_1979,christiansen_2004}, where strong selection is sufficient for relative fitness to break down.
-To briefly review: based on a diploid, bi-allelic variant of the logistic model, the $r$/$K$ scheme proposed a dichotomy between $r$-selection (uncrowded) and $K$-selection (crowded) \citep{macarthur_1962}, with the latter taken to mean selection for greater equilibrium density \citep{gill_1974}.
-The more general Lotka-Volterra competition model introduces the inter-type $\alpha_{ij}$ competition coefficients, with selection on these termed ``$\alpha$-selection'' \citep{gill_1974,joshi_2001}.
-Setting aside $r$ which confers no selective advantage at equilibrium, we are left with $K$ and $\alpha$.
-These traits both behave like $\delta$ in Eq.~\eqref{eq:simplebirthdeath} in that they are density-dependent and cause density to change over a sweep (Fig.~\ref{fig:strengthofselection}).
+In our variable-density lottery, $\delta$-like behavior occurs if and only if types differ in more than one of the three traits $b$, $c$ and $d$ (Fig.~\ref{fig:multiple}).
+The $c$ and $d$ traits embody the two distinct directions in which density and selection can interact: selection may depend on density, and density may depend on selection \citep{prout_1980}.
+The combination of both is necessary to threaten the constant-$s$ approximation.
+The simple differential/difference equations that have become standard in discussions of density-dependent selection \citep{roughgarden_1979,christiansen_2004,mallet_2012,travis_2013} confound this distinction.
+Remarkably, the $b$ trait demonstrates that the combination is not sufficient, since the density-dependence of $b$-selection effectively disappears over equilibrium-to-equilibrium $b$-sweeps.
 
-In the variable-density lottery, this occurs if and only if types differ in more than one trait (Fig.~\ref{fig:multiple}).
-The $c$ and $d$ traits represent the two distinct directions in which density and selection interact: selection may depend on density, and density may depend on selection \citep{prout_1980}.
-The combination of both is necessary to threaten the constant-$s$ approximation.
-However, remarkably, the $b$ trait demonstrates that the combination is not sufficient, since the density-dependence of $b$-selection effectively disappears over equilibrium-to-equilibrium $b$-sweeps.
-Thus, the simple differential/difference equations that have become standard in discussions of density-dependent selection \citep{roughgarden_1979,christiansen_2004,mallet_2012,travis_2013} actually confound important aspects of the interaction between density and selection.
-
-While this is a conceptual reason to be wary of the classical density-dependent selection models, it is not clear what trait variation should be expected in nature.
+While this is a conceptual reason to be wary of the traits analyzed in classical density-dependent selection models, it is not clear what trait variation should be expected in nature.
 Should we expect mutations to generally affect $b$, $c$ and $d$ independently of each other, or pleiotropically such that $\delta$-like selection is prevalent?
@@
 Spatial localization of consumable resources (e.g.
-due to restricted movement of nutrients through soils) will tend to create a territorial situation similar to the lottery model, where resource competition only occurs locally and both it and interference competition are subsumed into the competitive ability $c$, which does not affect $N$.
+due to restricted movement of nutrients through soils) will tend to create a territorial situation similar to the lottery model, where resource competition only occurs locally.
+In this case, both resource competition and interference competition are subsumed into the competitive ability $c$, which, in the absence of pleiotropic effects on $b$ and/or $d$, does not affect $N$.
 
@@
 Rather, in many population growth models, only one life-history stage is represented, and the competitive effects resulting from crowding appear as a reduction in absolute fitness that only depends on the type densities at this life-history stage (e.g.
-the $n_i^2$ and $n_in_j$ terms in the Lotka Volterra equation).
-As This precludes selection concentrated at a fragile juvenile stage facilitated by a reproductive excess \citep{chesson_1983,turner1968population,kimura1969natural,nei1971fertility}.
+the $n_i^2$ and $n_in_j$ terms in the Lotka-Volterra equation).
+This precludes selection concentrated at a fragile juvenile stage facilitated by a reproductive excess \citep{chesson_1983,turner1968population,kimura1969natural,nei1971fertility}.
 
-Reproductive excesses appear in the variable-density lottery model when the number of propagules is greater than the number of available territories.
+Reproductive excesses appear in our variable-density lottery model when the number of propagules is greater than the number of available territories.
 Then only $\approx 1/L$ of the juveniles contesting unoccupied territories survive to adulthood.
-Unlike the role of adult density $n_i$ in single-life-stage models, it is the propagule densities $l_i$ that represent the crowding that drives competition (a ``critical age-group''; \citealt[pp.
-54]{charlesworth_1994}).
+Unlike the role of adult density $n_i$ in single-life-stage models, it is the propagule densities $l_i$ that represent the crowding that drives competition.
 In general, reproductive excesses will tend to produce strictly-relative lottery-type contests in which fitter types grow at the expense of others by preferentially filling the available adult ``slots''.
-The number of slots can remain fixed or change independently of selection at the juvenile stage.
-By ignoring reproductive excesses, single life-stage models are biased to have total population density be sensitive to ongoing selection.
+The number of available slots can remain fixed or change independently of selection at the juvenile stage.
+By ignoring reproductive excesses, single life-stage models are biased to have total population density be more sensitive to ongoing selection.
 In this respect, the viability selection heuristics that are common in population genetics \cite[pp.
@@
 as a result of a temporally-variable environment.
-While transient deviations from demographic equilibrium driven by the appearance of new types can also threaten the constant-$s$ approximation, it requires strong selection that is density-dependent and effects a density-regulating trait (and even then the constant-$s$ approximation may hold).
-Temporally-variable environments can dramatically alter frequency trajectories for individual sweeps (e.g.
+While transient deviations from demographic equilibrium driven by the appearance of new types can also threaten the constant-$s$ approximation, they require strong selection that is both density-dependent and affects a density-regulating trait (and even then the constant-$s$ approximation may hold).
+In contrast, temporally-variable environments can dramatically alter frequency trajectories for individual sweeps (e.g.
 Fig.
@@
 This suggests that in systems like the wild \textit{Drosophila} example mentioned in the Introduction, there is indeed no choice but to abandon relative fitness.
-The variable-density lottery could provide a useful starting point for analyzing evolution in this and other far-from-equilibrium situations for two reasons: 1) the $b$, $c$, $d$ trait scheme neatly distinguishes between different aspects of the interplay between density and selection; 2)  lottery models in general are mathematically similarity to the Wright-Fisher model, which should facilitate the analysis of genetic when $N$ is unstable.
-
-[Kind of an odd way to end, but it breaks the flow if I put it elsewhere] Another issue with the constant-$N$, constant-$s$ description of selection is that it precludes consideration of longer-term aspects of the interplay between evolution and ecology such as population extinction.
-A variety of approaches have been developed for dealing with these issues in quantitative genetics \citep{burger1995evolution,engen_2013}, population genetics \citep{bertram2017predicting} and adaptive dynamics \citep{ferriere2013eco,dieckmann2004adaptive}.
-Although density-dependent selection is  pertinent to these longer-term issues \citep{travis_2013}, our focus here has been the description of the time-dependent process by which selection changes allele frequencies.
-This is particularly critical for making sense of evolution at the genetic level, for which we now have abundant data.
+Our variable-density lottery could provide a useful starting point for analyzing evolution in this and other far-from-equilibrium situations for two reasons: 1) the $b$, $c$, $d$ trait scheme neatly distinguishes between different aspects of the interplay between density and selection; 2)  lottery models in general are mathematically similarity to the Wright-Fisher model, which should facilitate the analysis of genetic when $N$ is unstable.
 

### 2018-04-04 simulation_comparison.py

@@ -10,7 +10,7 @@ Created on Wed Nov 29 12:26:08 2017
 c=np.array([1.5,1.])
 cbar=sum(np.array([0.1,0.9])*c)
 U=100000
-ls=np.array([[x*0.1,x] for x in np.linspace(0.01,20,1000)])
+ls=np.array([[x*0.1,x] for x in np.linspace(0.01,20,50)])
 
 exact=np.array([deltnplussim(m,c,U) for m in U*ls]);
 totalclassic=np.array([deltnplusclassic(m,c,U) for m in U*ls]);
@@ -28,10 +28,10 @@ ax1.plot(ls[:,0],comp1,'r',label=r"$R_i \frac{c_i}{\overline{c}}$")
 ax1.plot(ls[:,0],comp2,'b',label=r"$A_i \frac{c_i}{\overline{c}}$")
 ax1.plot(ls[:,0],(nocomp+comp1+comp2),'k',label=r"$e^{-L}+R_i \frac{c_i}{\overline{c}}+A_i \frac{c_i}{\overline{c}}$",linewidth=2)
 
-ax1.plot(ls[:,0],(exact[:,0,0]+exact[:,1,0]+exact[:,2,0])/(ls[:,0]*U),'k.',markersize=1.,label="Simulation")
-ax1.plot(ls[:,0],exact[:,0,0]/(ls[:,0]*U),'k.',markersize=1)
-ax1.plot(ls[:,0],exact[:,1,0]/(ls[:,0]*U),'k.',markersize=1)
-ax1.plot(ls[:,0],exact[:,2,0]/(ls[:,0]*U),'k.',markersize=1)
+ax1.plot(ls[:,0],(exact[:,0,0]+exact[:,1,0]+exact[:,2,0])/(ls[:,0]*U),'ko',markersize=6,markerfacecolor='None',label="Simulation")
+ax1.plot(ls[:,0],exact[:,0,0]/(ls[:,0]*U),'ko',markersize=6,markerfacecolor='None')
+ax1.plot(ls[:,0],exact[:,1,0]/(ls[:,0]*U),'ko',markersize=6,markerfacecolor='None')
+ax1.plot(ls[:,0],exact[:,2,0]/(ls[:,0]*U),'ko',markersize=6,markerfacecolor='None')
 
 ax1.plot(ls[:,0],totalclassic[:,0]/(ls[:,0]*U),'k--',label=r"Classic lottery",linewidth=2)
 ax1.set_xlim([0,2])
[…]

### 2018-04-13 manuscript.tex
l

@@
 
-Another issue with the constant-$N$, constant-$s$ description of selection is that it precludes consideration of longer-term aspects of the interplay between evolution and ecology such as population extinction.
+The constant-$N$, constant-$s$ description of selection also precludes consideration of longer-term aspects of the interplay between evolution and ecology such as population extinction.
 A variety of approaches have been developed to address this in quantitative genetics \citep{burger1995evolution,engen_2013}, population genetics \citep{bertram2017predicting} and adaptive dynamics \citep{ferriere2013eco,dieckmann2004adaptive}.
@@
 
-In light of the complications of density-dependence, the assignment of density-independent relative fitnesses has been justified as an approximation that holds when selection is weak and $N$ changes slowly (\citealt{kimura1969natural}; \citealt[pp.
+In light of the complications arising from density-dependence, the assignment of density-independent relative fitnesses has been justified as an approximation that holds when selection is weak and $N$ changes slowly (\citealt{kimura1969natural}; \citealt[pp.
 277]{ewens_2004}; \citealt[Chap.
@@
 This ultimately derives from their failure to account for ``reproductive excess'', that is, an excess of juveniles that experience stronger selection than their adult counterparts \citep{turner1968population}.
-This allows selection to be concentrated at a juvenile ``bottleneck'', potentially keeping the density of adults constant even under strong selection.
+By allowing selection to be concentrated at a juvenile ``bottleneck'', reproductive excess makes it possible for the density of adults to remain constant even under strong selection.
 Reproductive excess featured prominently in early debates about the regulation of population density (e.g.
@@
 This is a form of the classical $K$-selection ideas discussed above, but restricted to the critical age group (juveniles, in this case).
-By focusing on density in the critical age group, this literature loses the interdependence of pre-reproductive selection and post-reproductive density.
+The interdependence of pre-reproductive selection and reproductive density is thus overlooked in the process of focusing on density in the critical age group.
 
-We develop a novel model of density-regulated population growth based on territorial contests, which explicitly incorporates a finite reproductive excess.
-Our starting point is the classic lottery model, which was developed by ecologists to study competition driven by territorial contests in reef fishes and plants \citep{sale_77,chesson_1981}.
+We re-evaluate the validity of the constant relative fitness description of selection in a novel model of density-regulated population growth that has a finite reproductive excess.
+Our model is inspired by the classic lottery model, which was developed by ecologists to study competition driven by territorial contests in reef fishes and plants \citep{sale_77,chesson_1981}, and which has some similarities to the Wright-Fisher model \cite{svardal_2015}.
 Each type is assumed to have three traits: fecundity $b$, mortality $d$, and competitive ability $c$.
@@
 185]{crow_1970}).
-Our first task is to relax the large-juvenile-number assumption of the lottery model to create a variable-density lottery with a finite, density-dependent reproductive excess.
+We relax the large-juvenile-number assumption of the lottery model to derive a variable-density lottery with a finite, density-dependent reproductive excess.
 
 The properties of density-dependent selection in our model are strikingly different from the classical literature discussed above.
-The strong connection between crowding and selection for greater equilibrium density is broken: selection need not affect density, and even when it does, Eq.~\eqref{eq:canonical} is almost exact even for strong selection, provided that only one of the traits $b$, $c$, or $d$ is involved.
+The strong connection between crowding and selection for greater equilibrium density is broken: selection need not affect density at all.
+And when it does, Eq.~\eqref{eq:canonical} is almost exact even for strong selection, provided that any changes in density are driven only by selection (as opposed to large deviations from demogaphic equilibrium), and that selection occurs on only one of the traits $b$, $c$, or $d$.
+On the flip side, the constant relative fitness approximation fails when strong selection acts concurrently on two or more of these traits, or when the population is far from demographic equilibrium.
 
@@
 $d$-selection is density-independent, $b$-selection gets weaker with lower territorial availability, while $c$-selection initially increases with density as territorial contests become more important, but eventually also declines as available territories become scarce.
-The wildtype equilibrium density is $n_1/T\approx 0.4$.}
+The wildtype equilibrium density is $n_1/T\approx 0.4$, suggesting that this population would rarely experience declining strength of $c$-selection with increasing $N$.}
 \end{figure}
@@
 Compared to simple birth-death models \citep{kostitzin_1939} or variants of the logistic \citep{roughgarden_1979}, the Lotka-Volterra model clearly distinguishes between intra- and inter-type competitive effects.
-Thus, one type can displace another without having a greater equilibrium density (Fig.~\ref{fig:LVvslottery}a).
-Nevertheless, selection drives transient changes in density in the Lotka-Volterra model even if the initial and final densities of a sweep are the same (constant density only occurs for a highly restricted subset of $r$ and $\alpha$ values; further details in Appendix C; also see \citealt{mallet_2012,smouse_1976}).
+Thus, when selection acts on inter-type competitive effects, one type can displace another without having a greater equilibrium density (Fig.~\ref{fig:LVvslottery}a).
+This has been termed ``$\alpha$-selection'' to distinguish it from $K$-selection, which involves intra-type competitive effects and changes in equilibrium density \cite{gill_1974,joshi_2001}.
+Although the initial and final densities of an $\alpha$-selection sweep are the same, density nevertheless does change transiently in the Lotka-Volterra model (constant density only occurs for a highly restricted subset of $r$ and $\alpha$ values; further details in Appendix C; also see \citealt{mallet_2012,smouse_1976}).
 Intuitively, for one type to exclude the other, competitive suppression of growth between types must be stronger than competitive suppression of growth within types, causing $N$ to dip over a sweep (Fig.~\ref{fig:LVvslottery}a).
@@
 
-Summarizing the three traits in our variable-density lottery model: (i) $c$-selection is density-dependent, but $c$ does not regulate density; (ii) $d$ regulates density, but $d$-selection is density-independent; (iii) $b$ regulates density and $b$-selection is density-dependent.
-Yet despite their differences, pure $b$, $c$ and $d$ sweeps starting and ending at equilibrium all obey the density-independent relative fitness description of selection almost exactly.
-This behavior is quite different from that found in the classical density-dependent selection literature (see ``Introduction'' and ``The response of density to selection; $c$-selection versus $K$-selection'').
-In the latter, selection in crowded populations takes broadly one of two forms: selection for greater carrying capacity ($K$-selection) or selection on competition coefficients (``$\alpha$-selection''; \citealt{gill_1974,joshi_2001}).
-$K$ and $\alpha$ both behave like $\delta$ in Eq.~\eqref{eq:simplebirthdeath} in that strong selection is sufficient for Eq.~\eqref{eq:canonical} to break down (Fig.~\ref{fig:strengthofselection}).
+Summarizing the properties of selection in our variable-density lottery model: (i) $c$-selection is density-dependent, but $c$ does not regulate density; (ii) $d$ regulates density, but $d$-selection is density-independent; (iii) $b$ regulates density and $b$-selection is density-dependent.
+Yet, despite the differences between $b$, $c$ and $d$, in a constant environment selection that only involves one of these traits obeys the density-independent relative fitness description of selection almost exactly.
+When strong selection acts on more than one of $b$, $c$ and $d$ (Fig.~\ref{fig:multiple}), the constant-$s$ approximation breaks down.
+The $c$ and $d$ traits exemplify the two distinct directions in which density and selection can interact: selection may depend on density, and density may change in response to ongoing selection \citep{prout_1980}.
+The combination of both is necessary to threaten the constant-$s$ approximation.
+Remarkably, the $b$ trait demonstrates that the combination is not sufficient; the density-dependence of $b$-selection effectively disappears over equilibrium-to-equilibrium $b$-sweeps.
 
-In our variable-density lottery, $\delta$-like behavior occurs if and only if types differ in more than one of the three traits $b$, $c$ and $d$ (Fig.~\ref{fig:multiple}).
-The $c$ and $d$ traits embody the two distinct directions in which density and selection can interact: selection may depend on density, and density may depend on selection \citep{prout_1980}.
-The combination of both is necessary to threaten the constant-$s$ approximation.
-The simple differential/difference equations that have become standard in discussions of density-dependent selection \citep{roughgarden_1979,christiansen_2004,mallet_2012,travis_2013} confound this distinction.
-Remarkably, the $b$ trait demonstrates that the combination is not sufficient, since the density-dependence of $b$-selection effectively disappears over equilibrium-to-equilibrium $b$-sweeps.
+Selection in the variable-density lottery is quite different from classical density-dependent selection (see ``Introduction'' and ``The response of density to selection; $c$-selection versus $K$-selection'').
+In the latter, only one life-history stage is represented, and the effects of crowding appear as a reduction in absolute fitness that only depends on the type densities at this life-history stage (e.g.
+the $n_i^2$ and $n_in_j$ terms in the Lotka-Volterra equation).
+Selection in crowded populations takes broadly one of two forms: selection for greater carrying capacity ($K$-selection) or selection on competition coefficients ($\alpha$-selection).
+These are both ``$\delta$-like''  in the sense that selection depends on density and also causes density to change ($\delta$ is defined in Eq.~\eqref{eq:simplebirthdeath}).
+Strong selection is therefore sufficient for Eq.~\eqref{eq:canonical} to break down (Fig.~\ref{fig:strengthofselection}), and no distinction is made between density-regulating and density-dependent traits.
 
-While this is a conceptual reason to be wary of the traits analyzed in classical density-dependent selection models, it is not clear what trait variation should be expected in nature.
-Should we expect mutations to generally affect $b$, $c$ and $d$ independently of each other, or pleiotropically such that $\delta$-like selection is prevalent?
-In the case of well-mixed indirect exploitation competition for consumable resources, the $R^*$ rule  suggests that $\delta$-like selection will be prevalent.
-However, for many populations consumable resources are not well-mixed.
-Spatial localization of consumable resources (e.g.
-due to restricted movement of nutrients through soils) will tend to create a territorial situation similar to the lottery model, where resource competition only occurs locally.
-In this case, both resource competition and interference competition are subsumed into the competitive ability $c$, which, in the absence of pleiotropic effects on $b$ and/or $d$, does not affect $N$.
-
-Our variable-density lottery model shows that it is not simply a lack of ecological realism on the part of constant relative fitness models that underlies its contrast with the classical view of density-dependent selection.
-Rather, in many population growth models, only one life-history stage is represented, and the competitive effects resulting from crowding appear as a reduction in absolute fitness that only depends on the type densities at this life-history stage (e.g.
-the $n_i^2$ and $n_in_j$ terms in the Lotka-Volterra equation).
-This precludes selection concentrated at a fragile juvenile stage facilitated by a reproductive excess \citep{chesson_1983,turner1968population,kimura1969natural,nei1971fertility}.
-
-Reproductive excesses appear in our variable-density lottery model when the number of propagules is greater than the number of available territories.
+The distinctive properties of selection in the variable-density lottery arise from a reproductive excess which appears when the number of propagules is greater than the number of available territories.
 Then only $\approx 1/L$ of the juveniles contesting unoccupied territories survive to adulthood.
 Unlike the role of adult density $n_i$ in single-life-stage models, it is the propagule densities $l_i$ that represent the crowding that drives competition.
-In general, reproductive excesses will tend to produce strictly-relative lottery-type contests in which fitter types grow at the expense of others by preferentially filling the available adult ``slots''.
+Reproductive excess produces relative contests in which fitter types grow at the expense of others by preferentially filling the available adult ``slots''.
 The number of available slots can remain fixed or change independently of selection at the juvenile stage.
-By ignoring reproductive excesses, single life-stage models are biased to have total population density be more sensitive to ongoing selection.
+By ignoring reproductive excess, single life-stage models are biased to have total population density be more sensitive to ongoing selection.
 In this respect, the viability selection heuristics that are common in population genetics \cite[pp.
 61]{gillespie_2004} actually capture an important ecological process without making the full leap to complex age-structured models.
+
+Looking beyond the variable-density lottery, it is not clear which forms of crowding-induced selection are more likely to occur in nature.
+Even if reproductive excesses are ubquitous, strictly relative $c$-like traits could pleiotropically interact with density-regulating traits so often that $\delta$-like behavior is prevalent.
+For instance, in the case in the case of well-mixed indirect exploitation competition for consumable resources, the $R^*$ rule suggests that competitive ability is intimately linked to equilibrium resource density, and hence that $\delta$-like behavior would be prevalent.
+However, this conclusion is sensititive to the assumptions of well-mixed resource competition models.
+Spatial localization of consumable resources (e.g.
+for plants due to restricted movement of nutrients through soils) will tend to create territorial contests similar to the lottery model, where resource competition only occurs locally and can be sensitive to contingencies such as the timing of propagule arrival \cite{bolker_99}.
+In this case, resource competition is effectively subsumed into a territorial competitive ability trait akin to $c$, which would likely affect $N$ much more weakly than suggested by the $R^*$ rule (assuming no pleiotropic interactions with $b$ or $d$).
+
+Moreover, even in well mixed populations, competition does not only involve indirect exploitation of shared resources, but also direct interference.
+Interference competition can dramatically alter the dynamics of resource exploitation \cite{case_1974,amarasekare_2002}, and is more likely than the exploitation of shared resource pools to involve relative contests akin to $c$-selection.
+For instance, sexual selection can be viewed as a form of relative interference competition between genotypes.
+Thus, \textit{a priori} we should not expect crowding in nature to only involve selection that is $\delta$-like.
+Other forms of selection like $c$-selection (that is, strictly relative traits in density-regulated populations) are also likely to be important.
+Note that in the classical density-dependent selection literature, interference competition is closely associated with $\alpha$-selection and the idea that selection need not affect equilibrium density \cite{gill_1974}.
+However, as explained above, $\alpha$-selection does transiently affect population density and can therefore be regarded as $\delta$-like.
 
@@
 This suggests that in systems like the wild \textit{Drosophila} example mentioned in the Introduction, there is indeed no choice but to abandon relative fitness.
-Our variable-density lottery could provide a useful starting point for analyzing evolution in this and other far-from-equilibrium situations for two reasons: 1) the $b$, $c$, $d$ trait scheme neatly distinguishes between different aspects of the interplay between density and selection; 2)  lottery models in general are mathematically similarity to the Wright-Fisher model, which should facilitate the analysis of genetic when $N$ is unstable.
+Our variable-density lottery could provide a useful starting point for analyzing evolution in this and other far-from-equilibrium situations for two reasons: 1) the $b$, $c$, $d$ trait scheme neatly distinguishes between different aspects of the interplay between density and selection; 2) lottery models in general are mathematically similar to the Wright-Fisher model, which should facilitate the analysis of genetic drift when $N$ is unstable.
 

### 2018-04-23 manuscript.tex
l

@@
 In continuous time, the Malthusian parameter $r_i$ replaces $W_i$ and we have $\frac{d n_i}{dt}=r_in_i$ and $\frac{d p_i}{dt}=(r_i-\overline{r}) p_i$ \citep{crow_1970}.
-In the particular case that two types are present, the selection equation takes the familiar form
+In the particular case that two types are present, the discrete and continuous changes in frequency can be written respectively as $\Delta p_i=\frac{W_i-W_j}{\overline{W}} p_i p_j$ and $\frac{d p_i}{dt}=s p_i p_j$
 \begin{equation}
-\frac{d p_i}{dt}=s p_i(1-p_i), \label{eq:canonical}
+\frac{d p_i}{dt}=s p_i p_j \label{eq:canonical}
 \end{equation}
 where $s=r_i-r_j$ is the selection coefficient of type $i$ relative to type $j$.
+
 Note that we can replace the $W_i$ with any set of values proportional to the $W_i$ without affecting the  ratio $W_i/\overline{W}$ or $\Delta p_i$.
@@
 
-In a constant environment, and in the absence of crowding, $r_i$ is an ``intrinsic'' population growth rate (for concreteness we use the continuous time Eq.~\eqref{eq:canonical} as our point of comparison, but the discrete time case is similar).
+Henceforth our point of comparison we will be the continuous time Eq.~\eqref{eq:canonical}.
+
+In a constant environment, and in the absence of crowding, $r_i$ is an ``intrinsic'' population growth rate.
 The selection coefficient $s$ is then simply the difference in intrinsic growth rates.

### 2018-04-24 manuscript.tex
l

--- 2018-04-13 manuscript.tex
@@
 
-For haploids tracked in discrete time, the change in the abundance $n_i$ of type $i$ over a time step can be expressed as $\Delta n_i=(W_i - 1)n_i$ where $W_i$ is ``absolute fitness''.
+For haploids tracked in discrete time, the change in the abundance $n_i$ of type $i$ over a time step can be expressed as $\Delta n_i=(W_i - 1)n_i$ where $W_i$ is ``absolute fitness'' (i.e.~the abundance after one time step is $n_i'=W_i n_i$).
 The corresponding change in frequency is $\Delta p_i=\left(\frac{W_i}{\overline{W}}-1\right) p_i$, where $\overline{W}=\sum_i W_i p_i$.
 In continuous time, the Malthusian parameter $r_i$ replaces $W_i$ and we have $\frac{d n_i}{dt}=r_in_i$ and $\frac{d p_i}{dt}=(r_i-\overline{r}) p_i$ \citep{crow_1970}.
-In the particular case that two types are present, the selection equation takes the familiar form
-\begin{equation}
-\frac{d p_i}{dt}=s p_i(1-p_i), \label{eq:canonical}
-\end{equation}
-where $s=r_i-r_j$ is the selection coefficient of type $i$ relative to type $j$.
 Note that we can replace the $W_i$ with any set of values proportional to the $W_i$ without affecting the  ratio $W_i/\overline{W}$ or $\Delta p_i$.
@@
 468]{barton_2007}.
-Similarly in the continuous case, the selection coefficient and frequency dynamics in Eq.~\eqref{eq:canonical} would be unaffected if we added an arbitrary constant to the Malthusian parameters $r_i$ (these would then be relative log fitnesses).
+Similarly in the continuous case, adding an arbitrary constant to the Malthusian parameters $r_i$ has no effect on $\frac{d p_i}{dt}$ (these would then be relative log fitnesses).
 
-In a constant environment, and in the absence of crowding, $r_i$ is an ``intrinsic'' population growth rate (for concreteness we use the continuous time Eq.~\eqref{eq:canonical} as our point of comparison, but the discrete time case is similar).
-The selection coefficient $s$ is then simply the difference in intrinsic growth rates.
+Relative fitness is often parameterized in terms of selection coefficients which represent the advantages of different types relative to each other.
+For instance, in continuous time $s=r_i-r_j$ is the selection coefficient of type $i$ relative to type $j$.
+Assuming that only $i$ and $j$ are present, the change in frequency can be written as
+\begin{equation}
+\frac{d p_i}{dt}=s p_i (1-p_i). \label{eq:canonicalcont}
+\end{equation}
+Thus, if $r_i$ and $r_j$ are constant, the frequency of the $i$ type will grow logistically with a constant rate parameter $s$.
+We then say that selection is independent of frequency and density.
+The discrete time case is more complicated.
+Defining the selection coefficient by $W_i=(1+s)W_j$, and again assuming $i$ and $j$ are the only types present, we have
+\begin{equation}
+\Delta p_i=\frac{W_i-W_j}{\overline{W}} p_i (1-p_i)=\frac{s}{1+sp_i} p_i (1-p_i). \label{eq:canonicaldisc}
+\end{equation}
+Hence, even in the simplest case that $W_i$ and $W_j$ are constant, selection is frequency-dependent in discrete time (note that this frequency dependence is negligible when $s$ is small compared to $1$; see \citealt{frank_2011}).
+We will refer to both the continuous and discrete time selection equations \eqref{eq:canonicalcont} and \eqref{eq:canonicaldisc} throughout this paper, but the simpler continuous time case will be our point of comparison for the rest of this section.
+
+In a constant environment, and in the absence of crowding, $r_i$ is a constant ``intrinsic'' population growth rate.
+The interpretation of Eq.~\eqref{eq:canonicalcont} is then simple: the selection coefficient $s$ is simply the difference in intrinsic growth rates.
 However, growth cannot continue at a non-zero constant rate indefinitely: the population is not viable if $r_i<0$, whereas  $r_i>0$ implies endlessly increasing population density.
-Thus, assuming viability, the increase in population density must be checked by crowding, implying that the Malthusian parameters $r_i$ eventually decline to zero (e.g.
+Thus, setting aside unviable populations, the increase in population density must be checked by crowding.
+This implies that the Malthusian parameters $r_i$ eventually decline to zero (e.g.
 \citealt[pp.
@@
 Selection can then be density-dependent, and indeed this is probably not uncommon, because crowded and uncrowded conditions can favor very different traits \citep{travis_2013}.
-Eq.~\eqref{eq:canonical} is then not a complete description of selection --- it lacks an additional coupled equation describing the dynamics of $N$, on which $s$ in Eq.~\eqref{eq:canonical} depends.
+Eq.~\eqref{eq:canonicalcont} is then not a complete description of selection --- it lacks an additional coupled equation describing the dynamics of $N$, on which $s$ in Eq.~\eqref{eq:canonicalcont} now depends.
 In general we cannot simply specify the dynamics of $N$ independently, because those ecological dynamics are coupled with the evolutionary dynamics of type frequency \citep{travis_2013}.
@@
 In practice, many population genetic models simply ignore density dependence and assign a constant relative fitness to each type.
-Selection is typically interpreted as operating through viability, but the ecological processes underlying the regulation of population density are left unspecified (e.g.
+Selection is typically interpreted as operating through viability, but the ecological processes underlying the regulation of population density are frequently left unspecified (e.g.
 \citealt{gillespie_2004}; \citealt{nagylaki_1992}; \citealt{ewens_2004}).
@@
 Following Fisher's suggestion that evolution tends to increase density in the long term \citep{fisher_1930, leon_1978, lande_2009}, as well as the influential concept of $K$-selection (specifically, the idea that selection in crowded conditions favors greater equilibrium density; \citealt{macarthur_1962}), many studies of density-regulated growth have focused on the response of density to selection \citep{kostitzin_1939,macarthur_1967,roughgarden_1979,christiansen_2004}.
-Indeed, both $N$ and $s$ change during, and as a result of, adaptive sweeps in many of the most widely used models of density-regulated population growth, including simple birth-death \citep{kostitzin_1939} and logistic models (Fig.~\ref{fig:Ksel}a; \citealt{macarthur_1962,roughgarden_1979,boyce_1984}), variants of these models using other functional forms for the absolute fitness penalties of crowding \citep{kimura_1978,charlesworth_1971,lande_2009,nagylaki_1979,lande_2009}, and the ``$R^*$ rule'' of resource competition theory (which states that the type able to deplete a shared limiting consumable resource to the lowest equilibrium density $R^*$ excludes the others; \citealt{grover_1997}).
+Indeed, both $N$ and $s$ change during, and as a result of, adaptive sweeps in many of the most widely used models of density-regulated population growth.
+The latter includes simple birth-death \citep{kostitzin_1939} and logistic models (Fig.~\ref{fig:Ksel}a; \citealt{macarthur_1962,roughgarden_1979,boyce_1984}), variants of these models using other functional forms for the absolute fitness penalties of crowding \citep{kimura_1978,charlesworth_1971,lande_2009,nagylaki_1979,lande_2009}, and the ``$R^*$ rule'' of resource competition theory (which states that the type able to deplete a shared limiting consumable resource to the lowest equilibrium density $R^*$ excludes the others; \citealt{grover_1997}).
 Density also changes in response to selection in the Lotka-Volterra competition model, at least during a sweep (except in special cases; \citealt{gill_1974,smouse_1976,mallet_2012}).
@@
 4]{charlesworth_1994}).
-Under these conditions, $s$ is approximately constant in Eq.~\eqref{eq:canonical}, at least for some number of generations.
-If $s$ depends only on density, not frequency, this approximate constancy can hold over entire selective sweeps \citep[Fig.
-9.5]{otto_2011}.
+Under these conditions, $s$ is approximately constant in Eq.~\eqref{eq:canonicalcont}, at least for some number of generations.
+If $s$ depends only on density, not frequency, this approximate constancy can hold over entire selective sweeps \citep{otto_2011}.
 
@@
 This is a form of the classical $K$-selection ideas discussed above, but restricted to the critical age group (juveniles, in this case).
-The interdependence of pre-reproductive selection and reproductive density is thus overlooked in the process of focusing on density in the critical age group.
+The interdependence of pre-reproductive selection and reproductive density is thus overlooked as a result of focusing on density in the critical age group.
 
 We re-evaluate the validity of the constant relative fitness description of selection in a novel model of density-regulated population growth that has a finite reproductive excess.
-Our model is inspired by the classic lottery model, which was developed by ecologists to study competition driven by territorial contests in reef fishes and plants \citep{sale_77,chesson_1981}, and which has some similarities to the Wright-Fisher model \cite{svardal_2015}.
+Our model is inspired by the classic discrete-time lottery model, which was developed by ecologists to study competition driven by territorial contests in reef fishes and plants \citep{sale_77,chesson_1981}, and which has some similarities to the Wright-Fisher model \citep{svardal_2015}.
 Each type is assumed to have three traits: fecundity $b$, mortality $d$, and competitive ability $c$.
@@
 The strong connection between crowding and selection for greater equilibrium density is broken: selection need not affect density at all.
-And when it does, Eq.~\eqref{eq:canonical} is almost exact even for strong selection, provided that any changes in density are driven only by selection (as opposed to large deviations from demogaphic equilibrium), and that selection occurs on only one of the traits $b$, $c$, or $d$.
+And when it does, the density-independent discrete-time selection equation \eqref{eq:canonicaldisc} is almost exact even for strong selection, provided that any changes in density are driven only by selection (as opposed to large deviations from demogaphic equilibrium), and that selection occurs on only one of the traits $b$, $c$, or $d$.
 On the flip side, the constant relative fitness approximation fails when strong selection acts concurrently on two or more of these traits, or when the population is far from demographic equilibrium.
@@
 In a given iteration, the number of adults of the $i$'th type will be denoted by $n_i$, the total number of adults by $N=\sum_i n_i$, and the number of unoccupied territories by $U=T-N$.
-We assume that the $n_i$ are large enough that stochastic fluctuations in the $n_i$ (drift) can be ignored (with $T$ also assumed large to allow for low type densities $n_i/T$).
+We assume that the $n_i$ are large enough that stochastic fluctuations in the $n_i$ (drift) can be ignored (with $T$ also assumed large to allow for low type densities $n_i/T\ll 1$).
 
@@
 
-Adult mortality only occurs in adults present at the start of the  iteration, and at a constant, type-specific per-capita rate $0\leq d_i\leq 1$ (Fig.~\ref{fig:lottery}).
-This gives an overall change in type abundances of
+Adult mortality occurs after lottery recruitment at a constant, type-specific per-capita rate $d_i\geq 1$, and can affect adults recruited in the current iteration, such that the new abundance at the end of the iteration is $(n_i+\Delta_+ n_i)/d_i$ (Fig.~\ref{fig:lottery}).
+In terms of absolute fitness, this can be written as
 \begin{equation}
-\Delta n_i=\Delta_+ n_i-d_i n_i. \label{eq:delttot}
+W_i=\frac{1}{d_i}\left(1+\frac{\Delta_+ n_i}{n_i}\right). 
 \end{equation}
+Here $\frac{\Delta_+ n_i}{n_i}$ is the per-capita rate of territorial acquisition, and $1/d_i$ is the fraction of type $i$ adults surviving to the next iteration.
 
@@
 \includegraphics[scale=0.8]{DDS_lottery.pdf}
-\caption{\label{fig:DDS_lottery} The density-dependence of selection in our variable-density lottery plotted as the difference in propagule success rate $\Delta n_2/m_2-\Delta n_1/n_1$ between an adaptive variant $2$ and a wildtype $1$ with equal frequencies.
+\caption{\label{fig:DDS_lottery} The density-dependence of selection in our variable-density lottery plotted as the difference in propagule success rate $\Delta n_2/n_2-\Delta n_1/n_1$ between an adaptive variant $2$ and a wildtype $1$ with equal frequencies.
 Here $b_1=1$, $d_1=0.5$ and $c_1=1$.

### 2018-04-24 DDS_lottery.py

@@ -57,13 +57,17 @@ eps=0.1
 
 ns=np.array([_*np.array([1,1]) for _ in np.linspace(1,T/2,100)]) 
 
-bDDS=np.sum(np.array([-1,1])*np.array([deltnplus(np.array([1.,1+eps])*b*x*(1-sum(x)/T),c,T-sum(x))/x-d for x in ns]),1)
-cDDS=np.sum(np.array([-1,1])*np.array([deltnplus(b*x*(1-sum(x)/T),np.array([1.,1+eps])*c,T-sum(x))/x-d for x in ns]),1)
-dDDS=np.sum(np.array([-1,1])*np.array([-np.array([1.,1-eps])*d for x in ns]),1)
+bWs=np.array([deltnplus(np.array([1.,1+eps])*b*x*(1-sum(x)/T),c,T-sum(x))/x-d for x in ns])
+bDDS=np.sum(np.array([-1,1])*bWs,1)/(1+np.mean(bWs,1))
+cWs=np.array([deltnplus(b*x*(1-sum(x)/T),np.array([1.,1+eps])*c,T-sum(x))/x-d for x in ns])
+cDDS=np.sum(np.array([-1,1])*cWs,1)/(1+np.mean(cWs,1))
+dWs=np.array([deltnplus(b*x*(1-sum(x)/T),c,T-sum(x))/x-np.array([1.+eps,1.])*d for x in ns])
+dDDS=np.sum(np.array([-1,1])*dWs,1)/(1+np.mean(dWs,1))
 ax1.plot(np.sum(ns,1)/T,bDDS,'k',linewidth=2,label=r"$b$ selection")
 ax1.plot(np.sum(ns,1)/T,cDDS,'k--',linewidth=2,label=r"$c$ selection")
 ax1.plot(np.sum(ns,1)/T,dDDS,'k:',linewidth=2,label=r"$d$ selection")
 ax1.legend(loc='upper center',prop={'size':11})
+#ax1.set_ylim([0,0.1])
 
 #eq_dens=optimize.fsolve(lambda D: (1-np.exp(-b[0]*D))*(1/D-1)-d[0],0.5)[0]
 #ax1.plot([eq_dens,eq_dens],[0,0.1],'k',linewidth=2)

### 2018-04-24 manuscript.tex
l

@@
 \caption{\label{fig:lottery} One iteration of our model.
-Propagules are dispersed by adults at random (only propagules landing on unoccupied territories are shown).
-Territories can receive zero propagules.
+Propagules are dispersed by adults at random (only those propagules landing on unoccupied territories are shown).
+Some territories may receive zero propagules.
 Lottery competition then occurs in each territory that receives more than one propagule (only illustrated in one territory).
-In a given territory, each type has probability proportional to $c_i x_i$ of winning the territory, where $c_i$ measures competitive ability and $x_i$ is the number of $i$ propagules present.
+In a given territory, type $i$ has probability proportional to $c_i x_i$ of winning the territory, where $c_i$ measures competitive ability and $x_i$ is the number of $i$ propagules present.
 In the illustrated territory, more black propagules are present, but white is a stronger competitor and has a higher probability of winning.
-Territories are made available for the next iteration by the death of adults present at the start of the iteration (red crosses).}
+Adult deaths make new territories available for the next iteration (red crosses).}
 \end{figure}
@@
 
-Selection in our variable density lottery model is density-dependent, by which we mean that  the selection coefficient, measured by the difference in per-capita growth rate $\Delta n_i/n_i$ between types, depends on $N$.
-Density-dependent selection is sometimes taken to mean a qualitative change in which types are fitter than others at different densities \citep{travis_2013}.
+Selection in our variable density lottery model is density-dependent.
+Since this a discrete time model, this means that when two types are present, the selection factor $(W_i-W_j)/\overline{W}$ in Eq.~\eqref{eq:canonicaldisc} depends on $N$.
+Note that density-dependent selection is sometimes taken to mean a qualitative change in which types are fitter than others at different densities \citep{travis_2013}.
 While reversal in the order of fitnesses and co-existence driven by density-regulation are possible in our variable-density lottery (a special case of the competition-colonization trade-off; \citealt{levins_71,tilman_94,bolker_99}), questions related to co-existence are tangential to our aims and will not be pursued further here.
@@
 \begin{equation}
-\frac{\Delta n_i}{n_i} = \frac{b_i}{\overline{b}}\frac{1-e^{-\overline{b}N/T}}{N}(T-N)-d_i, \label{eq:bdensitydependence}
+\frac{\Delta_+ n_i}{n_i} = \frac{b_i}{\overline{b}}\frac{1-e^{-\overline{b}N/T}}{N}(T-N), 
 \end{equation}
 where we have used that fact that $L=\overline{b}N/T$, and $\overline{b}$ is the population mean $b$. 
+\label{eq:bdensitydependence}
+
 It is clear $d$-selection is independent of density.
@@
 \includegraphics[scale=0.8]{DDS_lottery.pdf}
-\caption{\label{fig:DDS_lottery} The density-dependence of selection in our variable-density lottery plotted as the difference in propagule success rate $\Delta n_2/n_2-\Delta n_1/n_1$ between an adaptive variant $2$ and a wildtype $1$ with equal frequencies.
+\caption{\label{fig:DDS_lottery} The density-dependence of selection in our variable-density lottery between an adaptive variant $2$ and a wildtype $1$ with equal frequencies.
 Here $b_1=1$, $d_1=0.5$ and $c_1=1$.

### 2018-04-24 DDS_lottery.py

@@ -57,8 +57,8 @@ eps=0.1
 
 ns=np.array([_*np.array([1,1]) for _ in np.linspace(1,T/2,100)]) 
 
-bWs=np.array([deltnplus(np.array([1.,1+eps])*b*x*(1-sum(x)/T),c,T-sum(x))/x-d for x in ns])
-bDDS=np.sum(np.array([-1,1])*bWs,1)/(1+np.mean(bWs,1))
+bWs=np.array([(1+deltnplus(np.array([1.,1+eps])*b*x*(1-sum(x)/T),c,T-sum(x))/x)/d for x in ns])
+bDDS=np.sum(np.array([-1,1])*bWs,1)/np.mean(bWs,1)
 cWs=np.array([deltnplus(b*x*(1-sum(x)/T),np.array([1.,1+eps])*c,T-sum(x))/x-d for x in ns])
 cDDS=np.sum(np.array([-1,1])*cWs,1)/(1+np.mean(cWs,1))
 dWs=np.array([deltnplus(b*x*(1-sum(x)/T),c,T-sum(x))/x-np.array([1.+eps,1.])*d for x in ns])

### 2018-04-25 manuscript.tex
l

@@
 \begin{equation}
-W_i=\frac{1}{d_i}\left(1+\frac{\Delta_+ n_i}{n_i}\right). 
+W_i=\frac{1}{d_i}\left(1+\frac{\Delta_+ n_i}{n_i}\right). \label{eq:absfit}
 \end{equation}
@@
 Selection in our variable density lottery model is density-dependent.
-Since this a discrete time model, this means that when two types are present, the selection factor $(W_i-W_j)/\overline{W}$ in Eq.~\eqref{eq:canonicaldisc} depends on $N$.
+This means that when two types are present, the discrete-time selection factor $(W_i-W_j)/\overline{W}$ in Eq.~\eqref{eq:canonicaldisc} may depend on $N$.
 Note that density-dependent selection is sometimes taken to mean a qualitative change in which types are fitter than others at different densities \citep{travis_2013}.
@@
 Selection on $c$ is density-dependent, with the strength of selection  peaking at an intermediate density (Fig.~\ref{fig:DDS_lottery}).
-This intermediate peak occurs because at low density most territories are claimed without contest, whereas at high density few unoccupied territories are available to be contested.
+This intermediate peak occurs because most territories are claimed without contest at low density, whereas at high density few unoccupied territories are available to be contested.
+
 To see how selection on $b$ and $d$ depend on density, we write Eq.~\eqref{eq:masterequalc} in the alternative form
 \begin{equation}
-\frac{\Delta_+ n_i}{n_i} = \frac{b_i}{\overline{b}}\frac{1-e^{-\overline{b}N/T}}{N}(T-N), 
+\frac{\Delta_+ n_i}{n_i} = \frac{b_i}{\overline{b}}\frac{1-e^{-\overline{b}N/T}}{N}(T-N), \label{eq:bdensitydependence}
 \end{equation}
 where we have used that fact that $L=\overline{b}N/T$, and $\overline{b}$ is the population mean $b$.
-\label{eq:bdensitydependence}
-
-It is clear $d$-selection is independent of density.
+It is then clear that $d$-selection is independent of density, because the density-dependent factor $1 + \frac{\Delta_+ n_i}{n_i}$ in Eq.~\eqref{eq:absfit} is the same for types that differ in $d$ only.
 On the other hand, the strength of $b$-selection declines with density because the advantage of having greater $b$ gets smaller the fewer territories there are to be claimed (Fig.~\ref{fig:DDS_lottery}).
@@
 \caption{\label{fig:DDS_lottery} The density-dependence of selection in our variable-density lottery between an adaptive variant $2$ and a wildtype $1$ with equal frequencies.
-Here $b_1=1$, $d_1=0.5$ and $c_1=1$.
+Here $b_1=1$, $d_1=2$ and $c_1=1$.
 For $b$-selection we set $b_2=b_1(1+\epsilon)$, and similarly for $c$ and $d$, with $\epsilon=0.1$.
-$d$-selection is density-independent, $b$-selection gets weaker with lower territorial availability, while $c$-selection initially increases with density as territorial contests become more important, but eventually also declines as available territories become scarce.
-The wildtype equilibrium density is $n_1/T\approx 0.4$, suggesting that this population would rarely experience declining strength of $c$-selection with increasing $N$.}
+$d$-selection is density-independent, $b$-selection gets weaker with lower territorial availability, while $c$-selection initially increases with density as territorial contests become more important, but eventually also declines as available territories become scarce.}
 \end{figure}
@@
 
-We now turn to the issue of how density responds to selection, comparing our variable-density lottery to previous studies of density-regulated populations \citep{prout_1980}.
-As we saw in the previous section, $c$-selection has no effect on population density in our variable-density lottery.
-To make sense of how $c$-selection fits with previous population growth models, we now revisit MacArthur's general treatment of $K$-selection \citep{macarthur_1967}.
+In this section we discuss the response of density to selection, comparing our variable-density lottery to previous studies of density-regulated populations.
+Our focus will be on the competitive ability trait $c$, because, as mentioned in the previous section, $c$-selection has no effect on population density even though it affects fitness under crowded conditions.
+Since this behavior is quite different from selection on more commonly used crowding-sensitivity parameters like $K$ in the logistic model, we start by revisiting MacArthur's $K$-selection argument \citep{macarthur_1967}.
 
@@
 Trajectories do not follow the line $N=K_{11}=K_{22}$.
-(b) Lottery competition (Eq.~\ref{eq:master}) with $b_1=b_2=5$, $d_1=d_2=0.1$ and $c_1/c_2=5$.
+(b) Lottery competition (Eq.~\ref{eq:master}) with $b_1=b_2=5$, $d_1=d_2=1.1$ and $c_1/c_2=5$.
 Trajectories converge on the line $N=K_{11}=K_{22}$.}
@@
 This complete uncoupling of density from $c$-selection arises due to the presence of an excess of propagules which pay the cost of selection without affecting adult density \citep{nei1971fertility}.
-As a result, Eq.~\eqref{eq:canonical} holds in demographic equilibrium even though $c$-selection is density-dependent.
+As a result, the density-independent Eq.~\eqref{eq:canonicaldisc} holds in demographic equilibrium even though $c$-selection is density-dependent.
 
@@
 
-For Eq.~\eqref{eq:canonical} to break down, the selection coefficient $s$ must depend on density.
-As shown in Fig.~\ref{fig:DDS_lottery}, this is not the case for $d$; the selection coefficient for a $d$-variant with $d_2= d_1(1-\epsilon)$ is a constant $s=\epsilon d_1$.
-Eq.~\eqref{eq:canonical} also holds when the population is at demographic equilibrium and density is unaffected by the outcome of selection; as discussed in the previous section, the latter is the case for $c$-selection in our model.
-Thus, to threaten Eq.~\eqref{eq:canonical}, we require selection to be density-dependent, and also density to be changing.
+For Eq.~\eqref{eq:canonicaldisc} to break down, the selection factor $(W_i-W_j)/\overline{W}$ must depend on density.
+As shown in Fig.~\ref{fig:DDS_lottery}, this is not the case for $d$.
+Eq.~\eqref{eq:canonicaldisc} also holds when the population is at demographic equilibrium and density is unaffected by the outcome of selection; as discussed in the previous section, the latter is the case for $c$-selection in our model.
+Thus, to threaten Eq.~\eqref{eq:canonicaldisc}, we require selection to be density-dependent, and also density to be changing.
 This can obviously occur if density-dependent selection is occurring in a population far from demographic equilibrium.
-In this case the validity of Eq.~\eqref{eq:canonical} depends on the specifics of the rate and magnitude of demographic change (we return to this in the Discussion).
-However, Eq.~\eqref{eq:canonical} can be threatened even in demographically-stable populations if a density-regulating trait is subject to density-dependent selection, as is the case for $b$ in our variable-density lottery.
+In this case the validity of Eq.~\eqref{eq:canonicaldisc} depends on the specifics of the rate and magnitude of demographic change (we return to this in the Discussion).
+However, Eq.~\eqref{eq:canonicaldisc} can be threatened even in demographically-stable populations if a density-regulating trait is subject to density-dependent selection, as is the case for $b$ in our variable-density lottery.
 
-Before we discuss the $b$ trait, it is helpful to summarize the threat to Eq.~\eqref{eq:canonical} in simpler models of density-regulated growth, as exemplified by the birth-death model \citep{kostitzin_1939}
+Before we discuss the $b$ trait, it is helpful to summarize how Eq.~\eqref{eq:canonicalcont} breaks down in a simpler continuous-time birth-death model \citep{kostitzin_1939} 
 \begin{equation}
@@
 Here $\delta_i$ is per-capita mortality due to crowding (for simplicity, there are no deaths when uncrowded).
-Starting from a type $1$ population in equilibrium, a variant with $\delta_2=\delta_1(1-\epsilon)$ has density-dependent selection coefficient $s=\epsilon \delta_1 N$ in Eq.~\eqref{eq:canonical}, which will change over the course of the sweep as $N$ shifts from its initial type $1$ equilibrium to a type $2$ equilibrium.
+Starting from a type $1$ population in equilibrium, a variant with $\delta_2=\delta_1(1-\epsilon)$ has density-dependent selection coefficient $s=\epsilon \delta_1 N$ in Eq.~\eqref{eq:canonicalcont}, which will change over the course of the sweep as $N$ shifts from its initial type $1$ equilibrium to a type $2$ equilibrium.
 From Eq.~\eqref{eq:simplebirthdeath}, the equilibrium densities at the beginning and end of the sweep are $N_{\rm initial}=b_1/\delta_1$ and $N_{\rm final}=b_1/(\delta_1(1-\epsilon))=N_{\rm initial}/(1-\epsilon)$ respectively, and so $s_{\rm initial}= \epsilon b_1$ and $s_{\rm final}=s_{\rm initial}/(1-\epsilon)$.
-Consequently, substantial deviations from Eq.~\eqref{eq:canonical} occur if there is sufficiently strong selection on $\delta$ (Fig.~\ref{fig:strengthofselection}; \citealt{kimura1969natural,crow_1970}).
+Consequently, substantial deviations from Eq.~\eqref{eq:canonicalcont} occur if there is sufficiently strong selection on $\delta$ (Fig.~\ref{fig:strengthofselection}; \citealt{kimura1969natural,crow_1970}).
 
@@
 Equilibrium-to-equilibrium $b$-sweeps in our variable-density lottery are qualitatively different from $\delta$ sweeps in this simpler birth-death model, because greater $b$ not only means more propagules contesting territories, but also more territories being contested.
-Together, the net density-dependent effect on $b$-selection coefficients is zero; in Eq.~\eqref{eq:bdensitydependence}, since $b_i/\overline{b}=1$ in a single-type equilibrium, the density-dependence factor $f(\overline{b},N)=\frac{1-e^{-\overline{b}N/T}}{N}(T-N)$ is exactly equal to the constant mortality rate at the beginning and end of a $b$-sweep, even though $b$ and density change.
-During the sweep there is some deviation in $f(\overline{b},N)$, but this deviation is an order of magnitude smaller than for a $\delta$ sweep (the density-dependent deviation constant $s$ in Fig.~\ref{fig:strengthofselection} is of order $\epsilon$, whereas the analogous effect for $b$ sweep in our variable-density lottery is only of order $\epsilon^2$; see Appendix D for details).
-Since selection must already be strong for a $\delta$-sweep to threaten Eq.~\eqref{eq:canonical}, the density-independent model applies effectively exactly for equilibrium $b$-sweeps.
-Note, however, that the selection coefficient for $b$-sweeps (as  defined by differences in $\Delta n_i/n_i$) does depend on frequency because of the $1/\overline{b}$ factor.
+Together, the net density-dependent effect on $b$-selection is negligible: in a single-type equilibrium we have $W_i=1$ and $b_i/\overline{b}=1$, and hence the density-dependence factor $f(\overline{b},N)=\frac{1-e^{-\overline{b}N/T}}{N}(T-N)$ in Eq.~\eqref{eq:bdensitydependence} has the same value $d_i-1$ at the beginning and end of a $b$-sweep.
+During the sweep there is some deviation in $f(\overline{b},N)$, but this deviation is an order of magnitude smaller than for a $\delta$ sweep (the density-dependent deviation in Fig.~\ref{fig:strengthofselection} is of order $\epsilon$, whereas the analogous effect for $b$ sweep in our variable-density lottery is only of order $\epsilon^2$; see Appendix D for details).
+Since selection must already be strong for a $\delta$-sweep to threaten Eq.~\eqref{eq:canonicalcont}, the density-independent model applies effectively exactly for equilibrium $b$-sweeps.
 
@@
 
-In section ``Density-regulating traits and the threat of strong selection'' we argued that the density-dependent factor $f(\overline{b},N)$ is unchanged at the beginning and end points of an equilibrium-to-equilibrium $b$.
+In section ``Density-regulating traits and the threat of strong selection'' we argued that the density-dependent factor $f(\overline{b},N)=\frac{1-e^{-\overline{b}N/T}}{N}(T-N)$ is unchanged at the beginning and end points of an equilibrium-to-equilibrium $b$-sweep.
 Here we estimate the magnitude of the deviation in $f(\overline{b},N)$ during the sweep.
@@
 We first calculate the value of $f(\overline{b},N)$ at the halfway point in a sweep, where the halfway point is estimated with simple linear averages for $b$ and $N$.
-The sweep is driven by a $b$ variant with $b_j=b_i(1+\epsilon)$, and we denote the corresponding initial and final densities by $D_i$ and $D_j$ respectively, where we have $d_i=b_i(1-D_i)=b_j(1-D_j)$.
+The sweep is driven by a $b$ variant with $b_j=b_i(1+\epsilon)$, and we denote the corresponding initial and final densities by $D_i$ and $D_j$ respectively, where we have $f_{\rm initial}=b_i(1-D_i)=d_i-1=f_{\rm final}=b_j(1-D_j)$.
 We obtain
@@
 &=\frac{1}{4} (b_i+b_j)(2-D_i-D_j) \nonumber \\
-&=\frac{1}{4} (2d_i+b_i(1-D_j)+b_j(1-N_i)).
+&=\frac{1}{4} (2(d_i-1)+b_i(1-D_j)+b_j(1-D_i)).
 \end{align}
-Dividing by $d_i$, the proportional deviation in $f(N)$ at the midpoint of the sweep is
+Dividing by $d_i-1$, the proportional deviation in $f(N)$ at the midpoint of the sweep is
 \begin{align}
-\frac{f_{\rm{half}}}{d_i}&=\frac{1}{4}\left(2+\frac{b_i}{b_j}+\frac{b_j}{b_i}\right)\nonumber \\
+\frac{f_{\rm{half}}}{d_i-1}&=\frac{1}{4}\left(2+\frac{b_i}{b_j}+\frac{b_j}{b_i}\right)\nonumber \\
 &=\frac{1}{4}\left(2+\frac{1}{1+\epsilon}+1+\epsilon\right)\nonumber \\

### 2018-04-25 DDS_lottery.py

@@ -52,28 +52,28 @@ fig1, ax1 = plt.subplots(figsize=[5,4])
 T=100000
 b=np.array([1.,1.])
 c=np.array([1.,1.])
-d=np.array([.5,.5])
+d=np.array([2.,2.])
 eps=0.1
 
 ns=np.array([_*np.array([1,1]) for _ in np.linspace(1,T/2,100)]) 
 
 bWs=np.array([(1+deltnplus(np.array([1.,1+eps])*b*x*(1-sum(x)/T),c,T-sum(x))/x)/d for x in ns])
 bDDS=np.sum(np.array([-1,1])*bWs,1)/np.mean(bWs,1)
-cWs=np.array([deltnplus(b*x*(1-sum(x)/T),np.array([1.,1+eps])*c,T-sum(x))/x-d for x in ns])
-cDDS=np.sum(np.array([-1,1])*cWs,1)/(1+np.mean(cWs,1))
-dWs=np.array([deltnplus(b*x*(1-sum(x)/T),c,T-sum(x))/x-np.array([1.+eps,1.])*d for x in ns])
-dDDS=np.sum(np.array([-1,1])*dWs,1)/(1+np.mean(dWs,1))
+cWs=np.array([(1+deltnplus(b*x*(1-sum(x)/T),np.array([1.,1+eps])*c,T-sum(x))/x)/d for x in ns])
+cDDS=np.sum(np.array([-1,1])*cWs,1)/np.mean(cWs,1)
+dWs=np.array([(1+deltnplus(b*x*(1-sum(x)/T),c,T-sum(x))/x)/(np.array([1.+eps,1.])*d) for x in ns])
+dDDS=np.sum(np.array([-1,1])*dWs,1)/np.mean(dWs,1)
 ax1.plot(np.sum(ns,1)/T,bDDS,'k',linewidth=2,label=r"$b$ selection")
 ax1.plot(np.sum(ns,1)/T,cDDS,'k--',linewidth=2,label=r"$c$ selection")
 ax1.plot(np.sum(ns,1)/T,dDDS,'k:',linewidth=2,label=r"$d$ selection")
 ax1.legend(loc='upper center',prop={'size':11})
-#ax1.set_ylim([0,0.1])
+ax1.set_ylim([0,0.14])
 
 #eq_dens=optimize.fsolve(lambda D: (1-np.exp(-b[0]*D))*(1/D-1)-d[0],0.5)[0]
[…]

### 2018-04-25 phase_plot_LV_lottery.py

@@ -8,18 +8,18 @@ Created on Wed Nov 29 12:26:40 2017
 #(n1,n2) vector field
 #=================================================================
 T=100000
-x,y=np.linspace(0.0,1.,401),np.linspace(0.,1.,401)
+x,y=np.linspace(1e-5,1.,401),np.linspace(1e-5,1.,401)
 xgrid, ygrid=np.meshgrid(x,y)
 X,Y=np.zeros([len(x),len(y)]),np.zeros([len(x),len(y)])
 b=np.array([5.,5.])
 c=np.array([5.,1.])
-d=np.array([0.1,0.1])
+d=np.array([1.1,1.1])
 
 for i in xrange(len(x)):
     for j in xrange(len(y)):
         n=T*np.array([x[i],y[j]])
         U=T-sum(n)
-        X[j,i],Y[j,i]=deltnplus(b*n*U/float(T),c,U)-d*n
+        X[j,i],Y[j,i]=((1+deltnplus(b*n*U/float(T),c,U)/n)/d-1)*n
 
 
 fig1, (ax2, ax1) = plt.subplots(ncols=2,figsize=[8,4])
@@ -28,7 +28,6 @@ f1null=np.array([[x[[_ for _ in range(len(x)) if X[i,_]>0][-1]],y[i]] for i in r
 f2null=np.array([[x[[_ for _ in range(len(x)) if Y[i,_]>0][-1]],y[i]] for i in range(1,len(y)) if max(Y[i])>0])
 ax1.plot(f1null[:,0],f1null[:,1],'k--',linewidth=2.)
 ax1.plot(f2null[:,0],f2null[:,1],'k--',linewidth=2.)
-#plt.plot([f2null[0,0],f2null[-1,0]],[f2null[0,1],f2null[-1,1]],'y',linewidth=2)
 ax1.plot([f2null[-1,0],f1null[0,0]],[f2null[-1,1],f1null[0,1]],'k',linewidth=2,alpha=0.8)
 ax1.set_xlim([0,f1null[0,0]])
 ax1.set_ylim([0,f1null[-1,1]])

### 2018-04-25 strength_of_selection.py

@@ -61,33 +61,35 @@ T=100000
 totaltime=100
 b=np.array([1.,1.])
 c=np.array([1.,1.])
-d=np.array([.9,.9])
+d=np.array([1.5,1.5])
 n0=np.array([1000.,1000.])
 
 n=n0;
 nhist=[n];
 for t in range(totaltime/2):
     U=T-sum(n)
-    n=n+deltnplus(b*n*U/T,c,U)-d*n
+    n=n*(1+deltnplus(b*n*U/T,c,U)/n)/d
     nhist.append(list(n))
 
 print nhist[-1]
 
-totaltime=80
-b=np.array([1.,1.])
-c=np.array([1.,2.])
-d=np.array([.9,.7])
+totaltime=200
+b=np.array([1.,1.5])
+c=np.array([1.,1.])
+d=np.array([1.5,1.5])
 n0=np.array([sum(nhist[-1]),10.])
 
 n=n0;
-nhist=[n];
+nhist=[n]; Whist=[]
 for t in range(totaltime):
     U=T-sum(n)
-    n=n+deltnplus(b*n*U/T,c,U)-d*n
+    Whist.append((1+deltnplus(b*n*U/T,c,U)/n)/d)
+    n=n*Whist[-1]
     nhist.append(list(n))
 
 nhist=np.array(nhist)
 Nhist=np.sum(nhist,1)
+Whist=np.array(Whist)
 
 fig1, (ax1, ax2) = plt.subplots(ncols=2,figsize=[8,4])
 
@@ -97,19 +99,22 @@ ax1.plot(Nhist,"k:", linewidth=2,label=r"Total")
 ax1.annotate(r'$(a)$',xy=(0.01,0.93),xycoords='axes fraction',fontsize=16)
 ax1.set_xticklabels([])
 ax1.set_xlabel("Time",fontsize=14)
-ax1.set_yticklabels([0,0.05,0.1,0.15,0.2,0.25,0.3])
+#ax1.set_yticklabels([0,0.05,0.1,0.15,0.2,0.25,0.3])
 ax1.set_ylabel(r"Density/$T$",fontsize=14)
-ax1.set_ylim([0,30000])
+#ax1.set_ylim([0,30000])
 ax1.legend(loc='upper center',prop={'size':11})
 
[…]

### 2018-04-25 strength_of_selection.py
l

@@ -106,7 +106,7 @@ ax1.legend(loc='upper center',prop={'size':11})
 
 Wbarhist=np.sum((nhist[:-1]*Whist),1)/Nhist[:-1]
 ax2.plot((Whist[:,1]-Whist[:,0])/Wbarhist,'k',linewidth=2)
-s=Whist[0,1]-Whist[0,0]
+s=Whist[0,1]/Whist[0,0]-1
 ax2.plot(s/(1+s*nhist[:-1,1]/Nhist[:-1]),'b',linewidth=2)
 #plt.ylim([0,0.5])
 ax2.annotate(r'$(b)$',xy=(0.01,0.93),xycoords='axes fraction',fontsize=16)

### 2018-04-26 manuscript.tex
l

@@
 Relative fitness is often parameterized in terms of selection coefficients which represent the advantages of different types relative to each other.
-For instance, in continuous time $s=r_i-r_j$ is the selection coefficient of type $i$ relative to type $j$.
-Assuming that only $i$ and $j$ are present, the change in frequency can be written as
+For instance, in continuous time $s=r_2-r_1$ is the selection coefficient of type $2$ relative to type $1$.
+Assuming that only $2$ and $1$ are present, the change in frequency can be written as
 \begin{equation}
-\frac{d p_i}{dt}=s p_i (1-p_i). \label{eq:canonicalcont}
+\frac{d p_2}{dt}=s p_2 (1-p_2). \label{eq:canonicalcont}
 \end{equation}
-Thus, if $r_i$ and $r_j$ are constant, the frequency of the $i$ type will grow logistically with a constant rate parameter $s$.
+Thus, if $r_1$ and $r_2$ are constant, the frequency of the second type will grow logistically with a constant rate parameter $s$.
 We then say that selection is independent of frequency and density.
 The discrete time case is more complicated.
-Defining the selection coefficient by $W_i=(1+s)W_j$, and again assuming $i$ and $j$ are the only types present, we have
+Defining the selection coefficient by $W_2=(1+s)W_1$, and again assuming $1$ and $2$ are the only types present, we have
 \begin{equation}
-\Delta p_i=\frac{W_i-W_j}{\overline{W}} p_i (1-p_i)=\frac{s}{1+sp_i} p_i (1-p_i). \label{eq:canonicaldisc}
+\Delta p_2=\frac{W_2-W_1}{\overline{W}} p_2 (1-p_2)=\frac{s}{1+sp_2} p_2 (1-p_2). \label{eq:canonicaldisc}
 \end{equation}
-Hence, even in the simplest case that $W_i$ and $W_j$ are constant, selection is frequency-dependent in discrete time (note that this frequency dependence is negligible when $s$ is small compared to $1$; see \citealt{frank_2011}).
+Hence, even in the simplest case that $W_1$ and $W_2$ are constant, selection is frequency-dependent in discrete time (note that this frequency dependence is negligible when $s$ is small compared to $1$; see \citealt{frank_2011}).
 We will refer to both the continuous and discrete time selection equations \eqref{eq:canonicalcont} and \eqref{eq:canonicaldisc} throughout this paper, but the simpler continuous time case will be our point of comparison for the rest of this section.
@@
 \caption{\label{fig:Ksel} Phase diagram for the densities of two types $n_1$ and $n_2$ undergoing selection.
-(a) The logistic model $\frac{dn_1}{dt}=r_1(1-\frac{n_1+n_2}{K_1})n_1$ and $\frac{dn_2}{dt}=r_2(1-\frac{n_1+n_2}{K_2})n_1$ with $r_1=r_2$ and $K_1>K_2$.
+(a) The logistic model $\frac{dn_1}{dt}=r_1(1-\frac{n_1+n_2}{K_1})n_1$ and $\frac{dn_2}{dt}=r_2(1-\frac{n_1+n_2}{K_2})n_1$ with $r_1=r_2$ and $K_2>K_1$.
 (b) The constant-$N$, relative fitness description of selection.}
@@
 \begin{equation}
-\Delta_+ n_i = \frac{l_i}{L}(1-e^{-L})U. \label{eq:masterequalc}
+\Delta_+ n_i = \frac{l_i}{L}(1-e^{-L})U=\frac{b_i}{\overline{b}}\frac{1-e^{-\overline{b}N/T}}{N}(T-N), \label{eq:masterequalc}
 \end{equation}
+where we have used the fact that $L=\overline{b}N/T$ to make the dependence on $b$ and $N$ explicit ($\overline{b}$ is the population mean $b$).
 This formula can also be deduced directly from Eq.~\eqref{eq:growthsumuncoupled}: $1-e^{-L}$ is the fraction of territories that receive at least one propagule under Poisson dispersal, $(1-e^{-L})U$ is the total number of such territories, and type $i$ is expected to receive a fraction $l_i/L$ of these.
-Total population density thus grows according to
+
+Similarly, the total number of territories acquired is
 \begin{equation}
-\Delta N=(1-e^{-L})U-\sum_i d_i n_i \label{eq:Nmaster}
+\Delta_+ N=(1-e^{-L})U=(1-e^{-\overline{b}N/T})(T-N) \label{eq:Nmaster}
 \end{equation}
 
-\subsection*{Density-dependent selection in the variable-density lottery}
+\subsection*{Density regulation and selection in the variable-density lottery}
 
-We now outline the basic properties of selection on $b$, $c$ and $d$.
-The birth and mortality rates $b$ and $d$ are the traits which regulate density; $b$ controls the fraction of unoccupied territories that are contested, while $d$ controls adult mortality.
-Competitive ability $c$ does not regulate density since it only affects the relative likelihood for each type to win a contested territory.
-Thus, selection between types which only differ in $c$ occurs without causing $N$ to change (Eq.~\eqref{eq:Nmaster} shows this formally).
+Equipped with Eq.~\eqref{eq:master} we now outline the basic properties of the $b$, $c$ and $d$ traits.
+Adult density $N$ is regulated by the birth and mortality rates $b$ and $d$; $b$ controls the fraction of unoccupied territories that are contested (see Eq.~\eqref{eq:Nmaster}), while $d$ controls adult mortality.
+Competitive ability $c$ does not enter Eq.~\eqref{eq:Nmaster}, and therefore does not regulate total adult density: $c$ only affects the relative likelihood of winning a contested territory.
 
-Selection in our variable density lottery model is density-dependent.
-This means that when two types are present, the discrete-time selection factor $(W_i-W_j)/\overline{W}$ in Eq.~\eqref{eq:canonicaldisc} may depend on $N$.
+Selection in our variable-density lottery model is in general density-dependent, by which we mean that the discrete-time selection factor $(W_2-W_1)/\overline{W}$ from Eq.~\eqref{eq:canonicaldisc} may depend on $N$.
+More specifically, as we show below, $b$ and $c$ selection are density dependent, but $d$ selection is not.
 Note that density-dependent selection is sometimes taken to mean a qualitative change in which types are fitter than others at different densities \citep{travis_2013}.
@@
 
-Selection on $c$ is density-dependent, with the strength of selection  peaking at an intermediate density (Fig.~\ref{fig:DDS_lottery}).
-This intermediate peak occurs because most territories are claimed without contest at low density, whereas at high density few unoccupied territories are available to be contested.
+The strength of $b$ selection declines with increasing density.
+When types differ in $b$ only ($b$-selection), Eq.~\eqref{eq:master} simplifies to Eq.~\eqref{eq:masterequalc}, and absolute fitness can be written as $W_i=(1+\frac{b_i}{\overline{b}}f(\overline{b},N))/d_i$ where $f(\overline{b},N)=\frac{1-e^{-\overline{b}N/T}}{N}(T-N)$ is a decreasing function of $N$.
+Thus, the selection factor $\frac{W_2-W_1}{\overline{W}}=\frac{f(\overline{b},N)}{1+f(\overline{b},N)}\frac{b_2-b_1}{\overline{b}}$ declines with increasing density: the advantage of having greater $b$ gets smaller the fewer territories there are to be claimed (Fig.~\ref{fig:DDS_lottery}).
 
-To see how selection on $b$ and $d$ depend on density, we write Eq.~\eqref{eq:masterequalc} in the alternative form
-\begin{equation}
-\frac{\Delta_+ n_i}{n_i} = \frac{b_i}{\overline{b}}\frac{1-e^{-\overline{b}N/T}}{N}(T-N), \label{eq:bdensitydependence}
-\end{equation}
-where we have used that fact that $L=\overline{b}N/T$, and $\overline{b}$ is the population mean $b$.
-It is then clear that $d$-selection is independent of density, because the density-dependent factor $1 + \frac{\Delta_+ n_i}{n_i}$ in Eq.~\eqref{eq:absfit} is the same for types that differ in $d$ only.
-On the other hand, the strength of $b$-selection declines with density because the advantage of having greater $b$ gets smaller the fewer territories there are to be claimed (Fig.~\ref{fig:DDS_lottery}).
+In the case of $c$-selection, Eq.~\eqref{eq:master} implies that $W_2-W_1$ is proportional to $\frac{T-N}{T} \left[(R_2+A_2)c_2-(R_1+A_1)c_1\right]/\overline{c}$.
+The strength of $c$-selection thus peaks at an intermediate density (Fig.~\ref{fig:DDS_lottery}), because most territories are claimed without contest at low density ($R_1,R_2,A_1,A_2\rightarrow 0$), whereas at high density few unoccupied territories are available to be contested ($T-N\rightarrow 0$).
+
+Selection on $d$ is independent of density, because the density-dependent factor $1 + \frac{\Delta_+ n_i}{n_i}$ in Eq.~\eqref{eq:absfit} is the same for types that differ in $d$ only.
 
@@
 
-In this section we discuss the response of density to selection, comparing our variable-density lottery to previous studies of density-regulated populations.
-Our focus will be on the competitive ability trait $c$, because, as mentioned in the previous section, $c$-selection has no effect on population density even though it affects fitness under crowded conditions.
-Since this behavior is quite different from selection on more commonly used crowding-sensitivity parameters like $K$ in the logistic model, we start by revisiting MacArthur's $K$-selection argument \citep{macarthur_1967}.
+We now turn to the issue of how density responds to selection in our variable-density lottery and previous models of selection in density-regulated populations.
+In the latter, selection under crowded conditions typically induces changes in equilibrium density (see Introduction).
+In our variable-density lottery model, however, the competitive ability trait $c$ is not density-regulating, even though $c$ contributes to fitness under crowded conditions.
+Consequently, $c$-selection does not cause density to change.
+In this section we compare this $c$-selection behavior with the previous literature, which we take to be exemplified by MacArthur's $K$-selection argument \citep{macarthur_1967}.
 
@@
 
-By contrast, density trajectories for $c$-selection in our variable-density lottery converge on a line of constant equilibrium density (Fig.~\ref{fig:LVvslottery}b).
-This means that once the population reaches demographic equilibrium, it behaves indistinguishably from a constant-$N$ relative fitness model (Fig.~\ref{fig:Ksel}b).
-This complete uncoupling of density from $c$-selection arises due to the presence of an excess of propagules which pay the cost of selection without affecting adult density \citep{nei1971fertility}.
-As a result, the density-independent Eq.~\eqref{eq:canonicaldisc} holds in demographic equilibrium even though $c$-selection is density-dependent.
+In contrast to both $K$ and $\alpha$ selection, density trajectories for $c$-selection in our variable-density lottery converge on a line of constant equilibrium density (Fig.~\ref{fig:LVvslottery}b).
+This means that once $N$ reaches demographic equilibrium, selective sweeps behave indistinguishably from a constant-$N$ relative fitness model (Fig.~\ref{fig:Ksel}b).
+Thus, for $c$ sweeps, the selection factor $(W_2-W_1)/\overline{W}$ in Eq.~\eqref{eq:canonicaldisc} depends on frequency only, not density, provided that nothing else pushes $N$ out of demographic equilibrium over the sweep.
+This uncoupling of density from ongoing $c$-selection arises due to the presence of an excess of propagules which pay the cost of selection without affecting adult density \citep{nei1971fertility}.
 
@@
 
-For Eq.~\eqref{eq:canonicaldisc} to break down, the selection factor $(W_i-W_j)/\overline{W}$ must depend on density.
-As shown in Fig.~\ref{fig:DDS_lottery}, this is not the case for $d$.
-Eq.~\eqref{eq:canonicaldisc} also holds when the population is at demographic equilibrium and density is unaffected by the outcome of selection; as discussed in the previous section, the latter is the case for $c$-selection in our model.
-Thus, to threaten Eq.~\eqref{eq:canonicaldisc}, we require selection to be density-dependent, and also density to be changing.
-This can obviously occur if density-dependent selection is occurring in a population far from demographic equilibrium.
-In this case the validity of Eq.~\eqref{eq:canonicaldisc} depends on the specifics of the rate and magnitude of demographic change (we return to this in the Discussion).
+For the relative fitness model Eq.~\eqref{eq:canonicaldisc} to break down, the selection factor $(W_2-W_1)/\overline{W}$ must depend on density.
+As shown in Fig.~\ref{fig:DDS_lottery}, $(W_2-W_1)/\overline{W}$ is independent of $N$ in the case of $d$-selection.
+Selection is also independent of $N$ when the population is at demographic equilibrium and $N$ is unaffected by ongoing selection; as is the case for $c$-selection.
+Thus, to threaten Eq.~\eqref{eq:canonicaldisc}, we require that selection is density-dependent, and also that density is changing.
+This can obviously occur if density-dependent selection happens in a population far from demographic equilibrium, in which case the validity of Eq.~\eqref{eq:canonicaldisc} depends on the specifics of the rate and magnitude of demographic change (we return to this in the Discussion).
 However, Eq.~\eqref{eq:canonicaldisc} can be threatened even in demographically-stable populations if a density-regulating trait is subject to density-dependent selection, as is the case for $b$ in our variable-density lottery.
 
-Before we discuss the $b$ trait, it is helpful to summarize how Eq.~\eqref{eq:canonicalcont} breaks down in a simpler continuous-time birth-death model \citep{kostitzin_1939} 
+Before we discuss the $b$ trait, it is helpful to summarize how density-dependent selection on a density-regulating trait threatens Eq.~\eqref{eq:canonicalcont} in simpler continuous-time models.
+This applies, for example, to $K$-selection in the logistic \citep{kimura1969natural,crow_1970}.
+We consider the simple birth-death model \citep{kostitzin_1939}
 \begin{equation}
-\frac{d n_i}{dt}=(b_i -\delta_iN) n_i. \label{eq:simplebirthdeath}
+\frac{d n_i}{dt}=(b_i -\delta_iN) n_i, \label{eq:simplebirthdeath}
 \end{equation}
-Here $\delta_i$ is per-capita mortality due to crowding (for simplicity, there are no deaths when uncrowded).
+where $\delta_i$ is per-capita mortality due to crowding (for simplicity, there are no deaths when uncrowded).
 Starting from a type $1$ population in equilibrium, a variant with $\delta_2=\delta_1(1-\epsilon)$ has density-dependent selection coefficient $s=\epsilon \delta_1 N$ in Eq.~\eqref{eq:canonicalcont}, which will change over the course of the sweep as $N$ shifts from its initial type $1$ equilibrium to a type $2$ equilibrium.
 From Eq.~\eqref{eq:simplebirthdeath}, the equilibrium densities at the beginning and end of the sweep are $N_{\rm initial}=b_1/\delta_1$ and $N_{\rm final}=b_1/(\delta_1(1-\epsilon))=N_{\rm initial}/(1-\epsilon)$ respectively, and so $s_{\rm initial}= \epsilon b_1$ and $s_{\rm final}=s_{\rm initial}/(1-\epsilon)$.
-Consequently, substantial deviations from Eq.~\eqref{eq:canonicalcont} occur if there is sufficiently strong selection on $\delta$ (Fig.~\ref{fig:strengthofselection}; \citealt{kimura1969natural,crow_1970}).
+Consequently, substantial deviations from Eq.~\eqref{eq:canonicalcont} occur if there is sufficiently strong selection on $\delta$ (Fig.~\ref{fig:strengthofselection}).
 
@@
 Equilibrium-to-equilibrium $b$-sweeps in our variable-density lottery are qualitatively different from $\delta$ sweeps in this simpler birth-death model, because greater $b$ not only means more propagules contesting territories, but also more territories being contested.
-Together, the net density-dependent effect on $b$-selection is negligible: in a single-type equilibrium we have $W_i=1$ and $b_i/\overline{b}=1$, and hence the density-dependence factor $f(\overline{b},N)=\frac{1-e^{-\overline{b}N/T}}{N}(T-N)$ in Eq.~\eqref{eq:bdensitydependence} has the same value $d_i-1$ at the beginning and end of a $b$-sweep.
+Together, the net density-dependent effect on $b$-selection is negligible: in a single-type equilibrium we have $W_i=1$ and $b_i/\overline{b}=1$, and hence the density-dependence factor $f(\overline{b},N)=\frac{1-e^{-\overline{b}N/T}}{N}(T-N)$ in Eq.~\eqref{eq:masterequalc} has the same value $d_i-1$ at the beginning and end of a $b$-sweep (recall that $\frac{W_2-W_1}{\overline{W}}=\frac{f(\overline{b},N)}{1+f(\overline{b},N)}\frac{b_2-b_1}{\overline{b}}$ for $b$-selection).
 During the sweep there is some deviation in $f(\overline{b},N)$, but this deviation is an order of magnitude smaller than for a $\delta$ sweep (the density-dependent deviation in Fig.~\ref{fig:strengthofselection} is of order $\epsilon$, whereas the analogous effect for $b$ sweep in our variable-density lottery is only of order $\epsilon^2$; see Appendix D for details).
-Since selection must already be strong for a $\delta$-sweep to threaten Eq.~\eqref{eq:canonicalcont}, the density-independent model applies effectively exactly for equilibrium $b$-sweeps.
-
-If selection acts simultaneously on more than one trait in our variable-density lottery, then evolution in a density-regulating trait can drive changes in the strength of selection on a trait subject to density-dependent selection (Fig.~\ref{fig:multiple}).
-This can produce behavior analogous to selection on $\delta$ in Fig.~\ref{fig:strengthofselection}.
+Since selection must already be strong for a $\delta$-sweep to threaten Eq.~\eqref{eq:canonicalcont}, the density-independent model applies effectively exactly for equilibrium $b$-sweeps (Fig.~\ref{fig:bsweep}).
 
@@
 \centering
-\includegraphics[scale=0.8]{multiple.pdf}
-\caption{\label{fig:multiple} Simultaneous selection on $d$ and $c$ in our variable-density lottery model as predicted by Eq.~\eqref{eq:master}.
-Selection is not constant over the sweep because $d$ is density-regulating and $c$ is density-dependent.}
+\includegraphics[scale=0.8]{bsweep.pdf}
+\caption{\label{fig:bsweep} Equilibrium $b$ sweeps behave as though selection is independent of density even though $b$-selection is density-dependent in general.
+Panel (b) shows the density-dependent selection factor $(W_2-W_1)/\overline{W}$ predicted by Eq.~\eqref{eq:master} (solid line) compared to the same selection factor with the density-dependence term $f(\overline{b},N)$ held constant at its initial value (dashed line).}
+\end{figure}
+
+However, if selection acts simultaneously on more than one trait in our variable-density lottery, then evolution in a density-regulating trait can drive changes in the strength of selection on another trait subject to density-dependent selection.
+For instance, if selection acts simultaneously on $b$ and $d$, then $f(\overline{b},N)$ changes value from $d_1-1$ to $d_2-1$ over a sweep.
+The dynamics of density will then affect the selection factor $(W_2-W_1)/\overline{W}$ and cause deviations analogous to selection on $\delta$ in the continuous time case (Fig.~\ref{fig:bdsweep}).
+
+\begin{figure}
+\centering
+\includegraphics[scale=0.8]{bdsweep.pdf}
+\caption{\label{fig:bdsweep} Simultaneous selection on $b$ and $d$ induces density-dependence in the selection factor $(W_2-W_1)/\overline{W}$.
+Panel (b) shows the predictions of Eq.~\eqref{eq:masterequalc} (solid line) versus the same with the density-dependence factor$f(\overline{b},N)$ held constant at its initial value.}
 \end{figure}
@@
 Summarizing the properties of selection in our variable-density lottery model: (i) $c$-selection is density-dependent, but $c$ does not regulate density; (ii) $d$ regulates density, but $d$-selection is density-independent; (iii) $b$ regulates density and $b$-selection is density-dependent.
-Yet, despite the differences between $b$, $c$ and $d$, in a constant environment selection that only involves one of these traits obeys the density-independent relative fitness description of selection almost exactly.
-When strong selection acts on more than one of $b$, $c$ and $d$ (Fig.~\ref{fig:multiple}), the constant-$s$ approximation breaks down.
+Yet, despite the differences between $b$, $c$ and $d$,  selection in a constant environment that only involves one of these traits obeys the density-independent relative fitness description of selection almost exactly (that is, $(W_2-W_1)/\overline{W}$ in Eq.~\eqref{eq:canonicaldisc} is effectively independent of density).
+This density-independence breaks down when strong selection acts on more than one of $b$, $c$ and $d$ (Fig.~\ref{fig:bdsweep}).
 The $c$ and $d$ traits exemplify the two distinct directions in which density and selection can interact: selection may depend on density, and density may change in response to ongoing selection \citep{prout_1980}.
@@
 These are both ``$\delta$-like''  in the sense that selection depends on density and also causes density to change ($\delta$ is defined in Eq.~\eqref{eq:simplebirthdeath}).
-Strong selection is therefore sufficient for Eq.~\eqref{eq:canonical} to break down (Fig.~\ref{fig:strengthofselection}), and no distinction is made between density-regulating and density-dependent traits.
+Strong selection is therefore sufficient for Eq.~\eqref{eq:canonicalcont} to break down (Fig.~\ref{fig:strengthofselection}), and no distinction is made between density-regulating and density-dependent traits.
 
@@
 Spatial localization of consumable resources (e.g.
-for plants due to restricted movement of nutrients through soils) will tend to create territorial contests similar to the lottery model, where resource competition only occurs locally and can be sensitive to contingencies such as the timing of propagule arrival \cite{bolker_99}.
+for plants due to restricted movement of nutrients through soils) will tend to create territorial contests similar to the lottery model, where resource competition only occurs locally and can be sensitive to contingencies such as the timing of propagule arrival \citep{bolker_99}.
 In this case, resource competition is effectively subsumed into a territorial competitive ability trait akin to $c$, which would likely affect $N$ much more weakly than suggested by the $R^*$ rule (assuming no pleiotropic interactions with $b$ or $d$).
@@
 Moreover, even in well mixed populations, competition does not only involve indirect exploitation of shared resources, but also direct interference.
-Interference competition can dramatically alter the dynamics of resource exploitation \cite{case_1974,amarasekare_2002}, and is more likely than the exploitation of shared resource pools to involve relative contests akin to $c$-selection.
+Interference competition can dramatically alter the dynamics of resource exploitation \citep{case_1974,amarasekare_2002}, and is more likely than the exploitation of shared resource pools to involve relative contests akin to $c$-selection.
 For instance, sexual selection can be viewed as a form of relative interference competition between genotypes.
@@
 Other forms of selection like $c$-selection (that is, strictly relative traits in density-regulated populations) are also likely to be important.
-Note that in the classical density-dependent selection literature, interference competition is closely associated with $\alpha$-selection and the idea that selection need not affect equilibrium density \cite{gill_1974}.
-However, as explained above, $\alpha$-selection does transiently affect population density and can therefore be regarded as $\delta$-like.
+Note that in the classical density-dependent selection literature, interference competition is closely associated with $\alpha$-selection and the idea that selection need not affect equilibrium density \citep{gill_1974}.
+However, $\alpha$-selection does transiently affect population density and therefore retains $\delta$-like features.
 
-The above findings underscore that the most serious threat to the constant-$s$ approximation arises due to deviations from demographic equilibrium as a result of changes in the demographic rates of the types already present i.e.
+The above findings underscore that the most serious threat to the density-independent models of selection (Eqs.~\eqref{eq:canonicalcont} and \eqref{eq:canonicalcont}) arises due to deviations from demographic equilibrium as a result of changes in the demographic rates of the types already present i.e.
 as a result of a temporally-variable environment.
-While transient deviations from demographic equilibrium driven by the appearance of new types can also threaten the constant-$s$ approximation, they require strong selection that is both density-dependent and affects a density-regulating trait (and even then the constant-$s$ approximation may hold).
-In contrast, temporally-variable environments can dramatically alter frequency trajectories for individual sweeps (e.g.
+While transient deviations from demographic equilibrium driven by the appearance of new types can also threaten the density-independent approximation, this requires strong selection that is both density-dependent and affects a density-regulating trait (and, as exemplified by $b$-selection, even then the approximation may hold).
+By contrast, temporally-variable environments can dramatically alter frequency trajectories for individual sweeps (e.g.
 Fig.
@@
 
-This suggests that in systems like the wild \textit{Drosophila} example mentioned in the Introduction, there is indeed no choice but to abandon relative fitness.
+This suggests that in systems like the wild \textit{Drosophila} example mentioned in the Introduction, there may indeed be no choice but to abandon relative fitness.
 Our variable-density lottery could provide a useful starting point for analyzing evolution in this and other far-from-equilibrium situations for two reasons: 1) the $b$, $c$, $d$ trait scheme neatly distinguishes between different aspects of the interplay between density and selection; 2) lottery models in general are mathematically similar to the Wright-Fisher model, which should facilitate the analysis of genetic drift when $N$ is unstable.
@@
 Since $L$ is large, uncontested territories make up a negligible fraction of the total.
-The rare type grows almost entirely due to $\Delta_r n_1$, while the common type grows almost entirely due to $\Delta_a n_2$.
+The rare type's territorial acquisition is almost entirely due to $\Delta_r n_1$, while the common type's territorial acquisition entirely due to $\Delta_a n_2$.
 To ensure consistency, the approximate per-capita growth rates implied by the approximations \eqref{eq:meanfieldr} and \eqref{eq:meanfielda} must be equal $\Delta_r n_1/m_1 = \Delta_a n_2/m_2$.
@@
 \eqref{eq:cvr} can be large and we cannot make the replacement Eq.~\eqref{eq:meanfieldr}.
-Fig.~\ref{fig:approx_details} shows the breakdown of the effective mean approximation when the are large differences in $c$.
+Fig.~\ref{fig:approx_details} shows the breakdown of the effective mean approximation when there are large differences in $c$.
 
@@
 Our approximation breaks down in the presence of large $c$ differences.
-The inset shows the pathology of the $\tilde{p}$,$\hat{p}$ approximation --- growth rates are not equal in the neutral case $c=1$.
+The inset shows the pathology of the $\tilde{p}$,$\hat{p}$ approximation --- growth rates are not equal in the neutral case $c_1=c_2$.
 Simulation procedure is the same as in Fig.~\ref{fig:simcomp}, with $U=10^5$.}
@@
 
-In section ``Density-regulating traits and the threat of strong selection'' we argued that the density-dependent factor $f(\overline{b},N)=\frac{1-e^{-\overline{b}N/T}}{N}(T-N)$ is unchanged at the beginning and end points of an equilibrium-to-equilibrium $b$-sweep.
+In section ``Density-regulating traits and the threat of strong selection'' we argued that the density-dependent factor $f(\overline{b},N)=\frac{1-e^{-\overline{b}N/T}}{N}(T-N)$ is unchanged at the beginning and end points of an equilibrium to equilibrium sweep of a type with higher $b$.
 Here we estimate the magnitude of the deviation in $f(\overline{b},N)$ during the sweep.
@@
 We can thus make the approximation $1-e^{-\overline{b}D}\approx \overline{b}D$ and $f(\overline{b},N)\approx \overline{b}(1-D)$.
-We expect this to be a conservative approximate based on the worst case scenario, because $N$ is most sensitive to an increase in $b$ in this low-density linear regime.
+We expect this to be a conservative approximation based on the worst case scenario, because $N$ is most sensitive to an increase in $b$ in this low-density linear regime.
 We first calculate the value of $f(\overline{b},N)$ at the halfway point in a sweep, where the halfway point is estimated with simple linear averages for $b$ and $N$.
-The sweep is driven by a $b$ variant with $b_j=b_i(1+\epsilon)$, and we denote the corresponding initial and final densities by $D_i$ and $D_j$ respectively, where we have $f_{\rm initial}=b_i(1-D_i)=d_i-1=f_{\rm final}=b_j(1-D_j)$.
+The sweep is driven by a $b$ variant with $b_2=b_1(1+\epsilon)$, and we denote the initial and final densities by $D_1$ and $D_2$ respectively, where we have $f_{\rm initial}=b_1(1-D_1)=d_1-1=f_{\rm final}=b_2(1-D_2)$.
 We obtain
 \begin{align}
-f_{\rm{half}}=f(\frac{b_i+b_j}{2},\frac{N_i+N_j}{2})&=\frac{b_i+b_j}{2}\left(1-\frac{D_i+D_j}{2}\right) \nonumber \\
-&=\frac{1}{4} (b_i+b_j)(2-D_i-D_j) \nonumber \\
-&=\frac{1}{4} (2(d_i-1)+b_i(1-D_j)+b_j(1-D_i)).
+f_{\rm{half}}=f(\frac{b_1+b_2}{2},\frac{N_1+N_2}{2})&=\frac{b_1+b_2}{2}\left(1-\frac{D_1+D_2}{2}\right) \nonumber \\
+&=\frac{1}{4} (b_1+b_2)(2-D_1-D_2) \nonumber \\
+&=\frac{1}{4} (2(d_1-1)+b_1(1-D_2)+b_2(1-D_1)).
 \end{align}
-Dividing by $d_i-1$, the proportional deviation in $f(N)$ at the midpoint of the sweep is
+Dividing by $d_1-1$, the proportional deviation in $f(N)$ at the midpoint of the sweep is
 \begin{align}
-\frac{f_{\rm{half}}}{d_i-1}&=\frac{1}{4}\left(2+\frac{b_i}{b_j}+\frac{b_j}{b_i}\right)\nonumber \\
+\frac{f_{\rm{half}}}{d_1-1}&=\frac{1}{4}\left(2+\frac{b_1}{b_2}+\frac{b_2}{b_1}\right)\nonumber \\
 &=\frac{1}{4}\left(2+\frac{1}{1+\epsilon}+1+\epsilon\right)\nonumber \\

### 2018-04-26 DDS_lottery.py

@@ -63,11 +63,11 @@ cWs=np.array([(1+deltnplus(b*x*(1-sum(x)/T),np.array([1.,1+eps])*c,T-sum(x))/x)/
 cDDS=np.sum(np.array([-1,1])*cWs,1)/np.mean(cWs,1)
 dWs=np.array([(1+deltnplus(b*x*(1-sum(x)/T),c,T-sum(x))/x)/(np.array([1.+eps,1.])*d) for x in ns])
 dDDS=np.sum(np.array([-1,1])*dWs,1)/np.mean(dWs,1)
-ax1.plot(np.sum(ns,1)/T,bDDS,'k',linewidth=2,label=r"$b$ selection")
-ax1.plot(np.sum(ns,1)/T,cDDS,'k--',linewidth=2,label=r"$c$ selection")
-ax1.plot(np.sum(ns,1)/T,dDDS,'k:',linewidth=2,label=r"$d$ selection")
-ax1.legend(loc='upper center',prop={'size':11})
-ax1.set_ylim([0,0.14])
+ax1.plot(np.sum(ns,1)/T,bDDS,'k',linewidth=2,label=r"$b$")
+ax1.plot(np.sum(ns,1)/T,cDDS,'k--',linewidth=2,label=r"$c$")
+ax1.plot(np.sum(ns,1)/T,dDDS,'k:',linewidth=2,label=r"$d$")
+ax1.legend(loc='upper center',prop={'size':11},ncol=3)
+ax1.set_ylim([0,0.11])
 
 #eq_dens=optimize.fsolve(lambda D: (1-np.exp(-b[0]*D))*(1/D-1)-d[0],0.5)[0]
 #ax1.plot([eq_dens,eq_dens],[0,0.1],'k',linewidth=2)

### 2018-04-26 phase_plot_LV_lottery.py

@@ -42,8 +42,8 @@ ax1.streamplot(x, y, X, Y,start_points=seedpoints,linewidth=1.,density=100)
 
 ax1.set_xticklabels([])
 ax1.set_yticklabels([])
-ax1.set_xlabel(r"$n_1$",fontsize=20)
-ax1.set_ylabel(r"$n_2$",fontsize=20)
+ax1.set_xlabel(r"$n_2$",fontsize=20)
+ax1.set_ylabel(r"$n_1$",fontsize=20)
 
 x,y=np.linspace(0,1.5,1000),np.linspace(0,1.5,1000)
 X, Y = np.meshgrid(x,y)
@@ -66,16 +66,16 @@ ax2.streamplot(x, y, U, V,density=100,start_points=seedpoints,linewidth=1.)
 
 #ax2.annotate(r'$\frac{dn_2}{dt}=0$',xy=(0.15,0.81),xycoords='axes fraction',fontsize=16)
 #ax2.annotate(r'$\frac{dn_1}{dt}=0$',xy=(0.62,0.28),xycoords='axes fraction',fontsize=16)
-ax2.annotate(r'$K_{21}$',xy=(0.82,-0.07),xycoords='axes fraction',fontsize=16)
-ax2.annotate(r'$K_{11}$',xy=(0.95,-0.07),xycoords='axes fraction',fontsize=16)
-ax2.annotate(r'$K_{22}$',xy=(-0.13,0.89),xycoords='axes fraction',fontsize=16)
-ax2.annotate(r'$K_{12}$',xy=(-0.13,.97),xycoords='axes fraction',fontsize=16)
+ax2.annotate(r'$K_{12}$',xy=(0.82,-0.07),xycoords='axes fraction',fontsize=16)
+ax2.annotate(r'$K_{22}$',xy=(0.95,-0.07),xycoords='axes fraction',fontsize=16)
+ax2.annotate(r'$K_{11}$',xy=(-0.13,0.89),xycoords='axes fraction',fontsize=16)
+ax2.annotate(r'$K_{21}$',xy=(-0.13,.97),xycoords='axes fraction',fontsize=16)
 ax2.annotate(r'$(a)$',xy=(0.9,0.93),xycoords='axes fraction',fontsize=16)
 
 ax2.set_xticklabels([])
 ax2.set_yticklabels([])
-ax2.set_xlabel(r"$n_1$",fontsize=20)
[…]

### 2018-04-26 phase_plot_logistic.py

@@ -31,16 +31,16 @@ ax1.plot(z1,(1-alphas[1,0]*z1)/alphas[1,1],'k--',linewidth=2)
 ax1.set_xlim([0,1])
 ax1.set_ylim([0,1])
 
-ax1.annotate(r'$\frac{dn_2}{dt}=0$',xy=(0.005,0.68),xycoords='axes fraction',fontsize=16)
-ax1.annotate(r'$\frac{dn_1}{dt}=0$',xy=(0.65,0.4),xycoords='axes fraction',fontsize=16)
-ax1.annotate(r'$K_{1}$',xy=(0.95,-0.07),xycoords='axes fraction',fontsize=16)
-ax1.annotate(r'$K_{2}$',xy=(-0.1,0.65),xycoords='axes fraction',fontsize=16)
+ax1.annotate(r'$\frac{dn_1}{dt}=0$',xy=(0.005,0.68),xycoords='axes fraction',fontsize=16)
+ax1.annotate(r'$\frac{dn_2}{dt}=0$',xy=(0.65,0.4),xycoords='axes fraction',fontsize=16)
+ax1.annotate(r'$K_{2}$',xy=(0.95,-0.07),xycoords='axes fraction',fontsize=16)
+ax1.annotate(r'$K_{1}$',xy=(-0.1,0.65),xycoords='axes fraction',fontsize=16)
 ax1.annotate(r'$(a)$',xy=(0.9,0.93),xycoords='axes fraction',fontsize=16)
 
 ax1.set_xticklabels([])
 ax1.set_yticklabels([])
-ax1.set_xlabel(r"$n_1$",fontsize=20)
-ax1.set_ylabel(r"$n_2$",fontsize=20)
+ax1.set_xlabel(r"$n_2$",fontsize=20)
+ax1.set_ylabel(r"$n_1$",fontsize=20)
 
 ax2.quiver(z1[:-1], z2[:-1], z1[1:]-z1[:-1], z2[1:]-z2[:-1], color="b", scale_units='xy', angles='xy', scale=1)
 ax2.annotate(r'$\frac{dN}{dt}=0$',xy=(0.53,0.53),xycoords='axes fraction',fontsize=16)
@@ -48,8 +48,8 @@ ax2.annotate(r'$(b)$',xy=(0.9,0.93),xycoords='axes fraction',fontsize=16)
 
 ax2.set_xticklabels([])
 ax2.set_yticklabels([])
[…]

### 2018-04-26 strength_of_selection.py

@@ -54,7 +54,7 @@ plt.tight_layout()
 plt.savefig('/home/jason/repos/densitydependentlottery/strengthofselection.pdf',bbox="tight")
 
 
-#lottery multiple traits
+#lottery bsweep
 #==========================================================
 
 T=100000
@@ -73,7 +73,7 @@ for t in range(totaltime/2):
 
 print nhist[-1]
 
-totaltime=200
+totaltime=100
 b=np.array([1.,1.5])
 c=np.array([1.,1.])
 d=np.array([1.5,1.5])
@@ -93,30 +93,105 @@ Whist=np.array(Whist)
 
 fig1, (ax1, ax2) = plt.subplots(ncols=2,figsize=[8,4])
 
-ax1.plot(nhist[:,1],"k", linewidth=2,label=r"$d_2=0.7$, $c_2=2$")
-ax1.plot(nhist[:,0],"k--", linewidth=2,label=r"$d_1=0.9$, $c_1=1$")
+ax1.plot(nhist[:,1],"k", linewidth=2,label=r"$b_2=1.5$")
+ax1.plot(nhist[:,0],"k--", linewidth=2,label=r"$b_1=1$")
 ax1.plot(Nhist,"k:", linewidth=2,label=r"Total")
 ax1.annotate(r'$(a)$',xy=(0.01,0.93),xycoords='axes fraction',fontsize=16)
 ax1.set_xticklabels([])
 ax1.set_xlabel("Time",fontsize=14)
-#ax1.set_yticklabels([0,0.05,0.1,0.15,0.2,0.25,0.3])
+ax1.set_yticklabels([0,0.1,0.2,0.3,0.4,0.5,0.6,0.7])
 ax1.set_ylabel(r"Density/$T$",fontsize=14)
-#ax1.set_ylim([0,30000])
+ax1.set_ylim([0,70000])
 ax1.legend(loc='upper center',prop={'size':11})
 
 Wbarhist=np.sum((nhist[:-1]*Whist),1)/Nhist[:-1]
-ax2.plot((Whist[:,1]-Whist[:,0])/Wbarhist,'k',linewidth=2)
-s=Whist[0,1]/Whist[0,0]-1
-ax2.plot(s/(1+s*nhist[:-1,1]/Nhist[:-1]),'b',linewidth=2)
-#plt.ylim([0,0.5])
[…]

### 2018-04-26 manuscript.tex
l

@@
 
-[I'm going to revise this after your next round of comments]
 Selection is commonly described by assigning relative fitness values to genotypes.
-Yet when selection is strong, the ecological view of selection in density-regulated populations seems to be incompatible with constant-density relative fitnesses.
-Here we analyze the population ecological limits of relative fitness using a novel of density-dependent selection which contains a ``reproductive excess’’.
-Our model clearly distinguishes between density-dependent selection and changes in density driven by selection.
-These two effects are confounded in standard models of density-regulated population growth, but both are necessary, in combination with strong selection, for relative fitness to break down in populations close to demographic equilibrium.
-Remarkably, both effects are not sufficient: we give an example of strong selection on a density-regulating trait subject to density-dependent selection that conforms to the density-independent relative fitness description almost exactly.
-We reiterate the importance of reproductive excesses in many species, which allows even strong selection to have no effect on density.
-Our model also offers a possible alternative to relative fitness when the latter is untenable, as is likely the case far from demographic equilibrium.
+Yet many populations are density-regulated, and the strength of selection can then depend on density.
+Additionally, selection can induce changes in density.
+When these two effects occur together, the resulting selection dynamics are not describable by density-independent relative fitness values, particularly if selection is strong.
+In most previous models of density-dependent selection, these effects are assumed to occur together (e.g.
+``$K$-selection’’), and as a result relative fitness models seem to require weak selection.
+Here we argue that constant relative fitness models are valid under a substantially broader range of conditions.
+We develop a novel model of density-regulated population growth with three traits: fecundity, mortality, and competitive ability.
+Our model contains a finite ``reproductive excess’’, which allows selection on competitive ability to occur without affecting density.
+Remarkably, we find that even strong selection is compatible with density-independent relative fitnesses provided that only one trait evolves at a time.
+Our findings suggest that deviations from demographic equilibrium are the most serious threat to relative fitness models.
+In such cases our model offers a possible alternative to relative fitness.
 
-\noindent (191 words)
+\noindent (187 words)
 

### 2018-04-26 manuscript.tex
l

@@
 
-Selection is commonly described by assigning relative fitness values to genotypes.
-Yet many populations are density-regulated, and the strength of selection can then depend on density.
-Additionally, selection can induce changes in density.
-When these two effects occur together, the resulting selection dynamics are not describable by density-independent relative fitness values, particularly if selection is strong.
-In most previous models of density-dependent selection, these effects are assumed to occur together (e.g.
-``$K$-selection’’), and as a result relative fitness models seem to require weak selection.
-Here we argue that constant relative fitness models are valid under a substantially broader range of conditions.
+Selection is commonly described by assigning constant relative fitness values to genotypes.
+Yet population density is often regulated by crowding.
+Relative fitness may then depend on density, and density will change when selection acts on a density-regulating trait.
+When strong density-dependent selection acts on a density-regulating trait, selection is no longer describable by density-independent relative fitnesses, even in demographically-stable populations.
+In most previous models of density-dependent selection, density-regulating traits are subject to density-dependent selection (e.g.
+``$K$-selection’’), suggesting that density-independent relative fitnesses must be replaced with more ecologically-explicit absolute fitnesses unless selection is weak.
+Here we show that density-independent relative fitnesses can also accurately describe strong density-dependent selection.
 We develop a novel model of density-regulated population growth with three traits: fecundity, mortality, and competitive ability.
-Our model contains a finite ``reproductive excess’’, which allows selection on competitive ability to occur without affecting density.
-Remarkably, we find that even strong selection is compatible with density-independent relative fitnesses provided that only one trait evolves at a time.
+Our model incorporates a density-dependent juvenile ``reproductive excess’’, which largely decouples density-dependent selection from the regulation of density.
+We find that density-independent relative fitnesses accurately describe strong selection on each trait separately, even for fecundity, which is both density-regulating and subject to density-dependent selection.
 Our findings suggest that deviations from demographic equilibrium are the most serious threat to relative fitness models.
@@
 
-\noindent (187 words)
+\noindent (197 words)
 

### 2018-04-26 manuscript.tex
l

@@
 
-The above findings underscore that the most serious threat to the density-independent models of selection (Eqs.~\eqref{eq:canonicalcont} and \eqref{eq:canonicalcont}) arises due to deviations from demographic equilibrium as a result of changes in the demographic rates of the types already present i.e.
+The above findings underscore that the most serious threat to the density-independent models of selection (Eqs.~\eqref{eq:canonicalcont} and \eqref{eq:canonicaldisc}) arises due to deviations from demographic equilibrium as a result of changes in the demographic rates of the types already present i.e.
 as a result of a temporally-variable environment.

### 2018-04-28 manuscript.tex
l

@@
 Yet population density is often regulated by crowding.
-Relative fitness may then depend on density, and density will change when selection acts on a density-regulating trait.
-When strong density-dependent selection acts on a density-regulating trait, selection is no longer describable by density-independent relative fitnesses, even in demographically-stable populations.
-In most previous models of density-dependent selection, density-regulating traits are subject to density-dependent selection (e.g.
-``$K$-selection’’), suggesting that density-independent relative fitnesses must be replaced with more ecologically-explicit absolute fitnesses unless selection is weak.
-Here we show that density-independent relative fitnesses can also accurately describe strong density-dependent selection.
-We develop a novel model of density-regulated population growth with three traits: fecundity, mortality, and competitive ability.
-Our model incorporates a density-dependent juvenile ``reproductive excess’’, which largely decouples density-dependent selection from the regulation of density.
-We find that density-independent relative fitnesses accurately describe strong selection on each trait separately, even for fecundity, which is both density-regulating and subject to density-dependent selection.
+Relative fitness may then depend on density, and selection can change density when it acts on a density-regulating trait.
+When strong density-dependent selection acts on a density-regulating trait, selection is no longer describable by density-independent relative fitnesses, even in demographically stable populations.
+These conditions are met in most previous models of density-dependent selection when selection is strong (e.g.~``$K$-selection'' in the logistic and Lotka-Volterra models), suggesting that density-independent relative fitnesses must be replaced with more ecologically-explicit absolute fitnesses unless selection is weak.
+Here we show that density-independent relative fitnesses can also accurately describe strong density-dependent selection under some conditions.
+We develop a novel model of density-regulated population growth with three ecologically-intuitive traits: fecundity, mortality, and competitive ability.
+Our model, unlike the logistic or Lotka-Volterra, incorporates a density-dependent juvenile ``reproductive excess'', which largely decouples density-dependent selection from the regulation of density.
+We find that density-independent relative fitnesses accurately describe strong selection acting on any one trait, even fecundity, which is both density-regulating and subject to density-dependent selection.
 Our findings suggest that deviations from demographic equilibrium are the most serious threat to relative fitness models.
@@
 
-\noindent (197 words)
+\noindent (214 words)
 
@@
 Selection in our variable-density lottery model is in general density-dependent, by which we mean that the discrete-time selection factor $(W_2-W_1)/\overline{W}$ from Eq.~\eqref{eq:canonicaldisc} may depend on $N$.
-More specifically, as we show below, $b$ and $c$ selection are density dependent, but $d$ selection is not.
+More specifically, as we show below, $b$- and $c$- selection are density-dependent, but $d$-selection is not.
 Note that density-dependent selection is sometimes taken to mean a qualitative change in which types are fitter than others at different densities \citep{travis_2013}.
@@
 
-The strength of $b$ selection declines with increasing density.
+The strength of $b$-selection declines with increasing density.
 When types differ in $b$ only ($b$-selection), Eq.~\eqref{eq:master} simplifies to Eq.~\eqref{eq:masterequalc}, and absolute fitness can be written as $W_i=(1+\frac{b_i}{\overline{b}}f(\overline{b},N))/d_i$ where $f(\overline{b},N)=\frac{1-e^{-\overline{b}N/T}}{N}(T-N)$ is a decreasing function of $N$.
@@
 \includegraphics[scale=0.8]{DDS_lottery.pdf}
-\caption{\label{fig:DDS_lottery} The density-dependence of selection in our variable-density lottery between an adaptive variant $2$ and a wildtype $1$ with equal frequencies.
+\caption{\label{fig:DDS_lottery} The density-dependence of selection in our variable-density lottery between an adaptive variant $2$ and a wildtype variant $1$ with at equal frequencies.
 Here $b_1=1$, $d_1=2$ and $c_1=1$.
@@
 
-We now turn to the issue of how density responds to selection in our variable-density lottery and previous models of selection in density-regulated populations.
+We now turn to the issue of how density changes as a consequence of selection in our variable-density lottery, and in previous models of selection in density-regulated populations.
 In the latter, selection under crowded conditions typically induces changes in equilibrium density (see Introduction).
@@
 \includegraphics[scale=0.8]{bsweep.pdf}
-\caption{\label{fig:bsweep} Equilibrium $b$ sweeps behave as though selection is independent of density even though $b$-selection is density-dependent in general.
+\caption{\label{fig:bsweep} Equilibrium $b$-sweeps behave as though selection is independent of density even though $b$-selection is density-dependent in general.
 Panel (b) shows the density-dependent selection factor $(W_2-W_1)/\overline{W}$ predicted by Eq.~\eqref{eq:master} (solid line) compared to the same selection factor with the density-dependence term $f(\overline{b},N)$ held constant at its initial value (dashed line).}
@@
 Looking beyond the variable-density lottery, it is not clear which forms of crowding-induced selection are more likely to occur in nature.
-Even if reproductive excesses are ubquitous, strictly relative $c$-like traits could pleiotropically interact with density-regulating traits so often that $\delta$-like behavior is prevalent.
+Even if reproductive excesses are ubiquitous, strictly relative $c$-like traits could pleiotropically interact with density-regulating traits so often that $\delta$-like behavior is prevalent.
 For instance, in the case in the case of well-mixed indirect exploitation competition for consumable resources, the $R^*$ rule suggests that competitive ability is intimately linked to equilibrium resource density, and hence that $\delta$-like behavior would be prevalent.
@@
 
-Moreover, even in well mixed populations, competition does not only involve indirect exploitation of shared resources, but also direct interference.
+Moreover, even in well-mixed populations, competition does not only involve indirect exploitation of shared resources, but also direct interference.
 Interference competition can dramatically alter the dynamics of resource exploitation \citep{case_1974,amarasekare_2002}, and is more likely than the exploitation of shared resource pools to involve relative contests akin to $c$-selection.

### 2018-05-01 manuscript.tex
l

@@
 When strong density-dependent selection acts on a density-regulating trait, selection is no longer describable by density-independent relative fitnesses, even in demographically stable populations.
-These conditions are met in most previous models of density-dependent selection when selection is strong (e.g.~``$K$-selection'' in the logistic and Lotka-Volterra models), suggesting that density-independent relative fitnesses must be replaced with more ecologically-explicit absolute fitnesses unless selection is weak.
+These conditions are met in most previous models of density-dependent selection (e.g.~``$K$-selection'' in the logistic and Lotka-Volterra models), suggesting that density-independent relative fitnesses must be replaced with more ecologically explicit absolute fitnesses unless selection is weak.
 Here we show that density-independent relative fitnesses can also accurately describe strong density-dependent selection under some conditions.
-We develop a novel model of density-regulated population growth with three ecologically-intuitive traits: fecundity, mortality, and competitive ability.
+We develop a novel model of density-regulated population growth with three ecologically intuitive traits: fecundity, mortality, and competitive ability.
 Our model, unlike the logistic or Lotka-Volterra, incorporates a density-dependent juvenile ``reproductive excess'', which largely decouples density-dependent selection from the regulation of density.
@@
 
-\noindent (214 words)
+\noindent (210 words)
 

### 2018-07-10 manuscript.tex
l

@@
 
-Each iteration, adults produce propagules which disperse at random, independently of distance from their parents, and independently of each other.
+Each iteration, adults produce propagules which disperse at random, independently of distance from their parents, and independently of each other (undirected dispersal).
 We assume that each adult from type $i$ produces $b_i$ propagules on average, so that the mean number of $i$ propagules dispersing to unoccupied territories is $m_i=b_in_iU/T$.
@@
 where we have used the fact that $L=\overline{b}N/T$ to make the dependence on $b$ and $N$ explicit ($\overline{b}$ is the population mean $b$).
-This formula can also be deduced directly from Eq.~\eqref{eq:growthsumuncoupled}: $1-e^{-L}$ is the fraction of territories that receive at least one propagule under Poisson dispersal, $(1-e^{-L})U$ is the total number of such territories, and type $i$ is expected to receive a fraction $l_i/L$ of these.
+This formula can also be deduced directly from Eq.~\eqref{eq:growthsumuncoupled}, and is exact: $1-e^{-L}$ is the fraction of territories that receive at least one propagule under Poisson dispersal, $(1-e^{-L})U$ is the total number of such territories, and type $i$ is expected to receive a fraction $l_i/L$ of these.
 
-Similarly, the total number of territories acquired is
+By similar reasoning, the total number of territories is given by
 \begin{equation}
-\Delta_+ N=(1-e^{-L})U=(1-e^{-\overline{b}N/T})(T-N) \label{eq:Nmaster}
+\Delta_+ N=(1-e^{-L})U=(1-e^{-\overline{b}N/T})(T-N), \label{eq:Nmaster}
 \end{equation}
+which is exact and applies for arbitrary $b_i$, $c_i$ and $d_i$.
 

### 2018-07-11 manuscript.tex
latest

@@
 468]{barton_2007}.
-The variety of fitness measures is not problematic in itself, but it should be clear how these measures are connected to the processes of birth and death which ultimately drive selection (\citealt{metcalf_2007,doebeli_2017}; \citealt[pp.
+The justification of any measure of fitness ultimately derives from how it is connected to the processes of birth and death which drive selection (\citealt{metcalf_2007,doebeli_2017}; \citealt[pp.
 178]{charlesworth_1994}).
@@
 Each iteration, adults produce propagules which disperse at random, independently of distance from their parents, and independently of each other (undirected dispersal).
-We assume that each adult from type $i$ produces $b_i$ propagules on average, so that the mean number of $i$ propagules dispersing to unoccupied territories is $m_i=b_in_iU/T$.
-The parameter $b_i$ can be thought of as a measure of ``colonization ability'', which combines fertility and dispersal ability \citep{levins_71,tilman_94}.
-Random dispersal is then modeled using a Poisson distribution $p_i(x_i)=l_i^{x_i} e^{-l_i}/x_i!$ for the number $x_i$ of $i$ propagules dispersing to any particular unoccupied territory, where $l_i=m_i/U$ is the mean propagule density in unoccupied territories.
-The total propagule density will be denoted $L=\sum_i l_i$.
+We assume that each adult from type $i$ produces $b_i$ propagules on average, so that the mean number of $i$ propagules dispersing to unoccupied territories is $m_i=b_in_iU/T$ (the factor $U/T$ represents the loss of those propagules landing on occupied territories).
+Random dispersal is modeled using a Poisson distribution $p_i(x_i)=l_i^{x_i} e^{-l_i}/x_i!$ for the number $x_i$ of $i$ propagules dispersing to any particular unoccupied territory, where $l_i=m_i/U$ is the mean propagule density of type $i$ per unoccupied territory.
+The total propagule density per unoccupied territory will be denoted $L=\sum_i l_i$.
 
@@
 
-Equilibrium-to-equilibrium $b$-sweeps in our variable-density lottery are qualitatively different from $\delta$ sweeps in this simpler birth-death model, because greater $b$ not only means more propagules contesting territories, but also more territories being contested.
+Equilibrium-to-equilibrium $b$-sweeps in our variable-density lottery are qualitatively different from $\delta$ sweeps in this simpler birth-death model.
+Greater $b$ means not only that more propagules contest the available territories, but also that a greater fraction of unoccupied territories receive propagules.
 Together, the net density-dependent effect on $b$-selection is negligible: in a single-type equilibrium we have $W_i=1$ and $b_i/\overline{b}=1$, and hence the density-dependence factor $f(\overline{b},N)=\frac{1-e^{-\overline{b}N/T}}{N}(T-N)$ in Eq.~\eqref{eq:masterequalc} has the same value $d_i-1$ at the beginning and end of a $b$-sweep (recall that $\frac{W_2-W_1}{\overline{W}}=\frac{f(\overline{b},N)}{1+f(\overline{b},N)}\frac{b_2-b_1}{\overline{b}}$ for $b$-selection).

### 2018-07-12 manuscript.tex
l

@@
 where we have used the fact that $L=\overline{b}N/T$ to make the dependence on $b$ and $N$ explicit ($\overline{b}$ is the population mean $b$).
-This formula can also be deduced directly from Eq.~\eqref{eq:growthsumuncoupled}, and is exact: $1-e^{-L}$ is the fraction of territories that receive at least one propagule under Poisson dispersal, $(1-e^{-L})U$ is the total number of such territories, and type $i$ is expected to receive a fraction $l_i/L$ of these.
-
-By similar reasoning, the total number of territories is given by
+Eq.~\eqref{eq:masterequalc} happens to be exact even though it is a special case of the approximation Eq.~\eqref{eq:master}.
+This can be deduced directly from Eq.~\eqref{eq:growthsumuncoupled}: $1-e^{-L}$ is the fraction of territories that receive at least one propagule under Poisson dispersal, $(1-e^{-L})U$ is the total number of such territories, and type $i$ is expected to receive a fraction $l_i/L$ of these.
+By similar reasoning, the total number of territories acquired is given by
 \begin{equation}
-\Delta_+ N=(1-e^{-L})U=(1-e^{-\overline{b}N/T})(T-N), \label{eq:Nmaster}
+\Delta_+ N=(1-e^{-L})U=(1-e^{-\overline{b}N/T})(T-N). \label{eq:Nmaster}
 \end{equation}
-which is exact and applies for arbitrary $b_i$, $c_i$ and $d_i$.
+This formula is also exact, but unlike Eq.~\eqref{eq:masterequalc}, it also applies when the $c_i$ differ between types.
 
@@
 The approximations \eqref{eq:meanfieldr} and \eqref{eq:meanfielda} must be consistent between rare and common types.
-To illustrate, suppose that two identical types (same $b$, $c$ and $d$) are present, with low $l_1\ll 1$ and high density $l_2\approx L\gg 1$ respectively.
-Since $L$ is large, uncontested territories make up a negligible fraction of the total.
-The rare type's territorial acquisition is almost entirely due to $\Delta_r n_1$, while the common type's territorial acquisition entirely due to $\Delta_a n_2$.
-To ensure consistency, the approximate per-capita growth rates implied by the approximations \eqref{eq:meanfieldr} and \eqref{eq:meanfielda} must be equal $\Delta_r n_1/m_1 = \Delta_a n_2/m_2$.
+To illustrate, suppose that two identical types (same $b$, $c$ and $d$) are present, that the first type is rare and has vanishingly small density $l_1\rightarrow 0$ while the second is common and has high density $l_2\rightarrow \infty$.
+Since $l_2\rightarrow \infty$, uncontested territories make up a negligible fraction of the total.
+The rare type's territorial acquisition is entirely due to $\Delta_r n_1$, while the common type's territorial acquisition is entirely due to $\Delta_a n_2$.
+To ensure consistency, the approximate per-capita growth rates implied by the approximations \eqref{eq:meanfieldr} and \eqref{eq:meanfielda} must therefore be equal $\Delta_r n_1/m_1 = \Delta_a n_2/m_2$.
 Even small violations of this consistency condition would mean exponential growth of one type relative to the other.
@@
 
-For example, suppose that we use $\tilde{p}$ and $\hat{p}$ to calculate the effective means.
-The right hand side of Eq.
-\eqref{eq:meanfieldr} is then approximately $1/(L+1)$, and since $l_1\ll 1$ and $L\gg 1$ we have $\Delta_r n_1 \approx 1/(L+1)$ in Eq.~\eqref{eq:deltr}.
-Similarly, for the common type, $\sum_j \langle x_j\rangle_{\hat{p}} = L$ in Eq.
-\eqref{eq:meanfielda}, and so $\Delta_a n_2 \approx 1/L$.
+In general, the above consistency condition is not satisfied when the conditional Poisson distributions $\tilde{p}$ and $\hat{p}$ are used to calculate the effective means.
+Under the above limits ($l_1\rightarrow 0$, $l_2\rightarrow \infty$), the right hand side of Eq.~\eqref{eq:meanfieldr} tends to $1/(l_2+1)$, and so $\Delta_r n_1/m_1 \rightarrow 1/(l_2+1)$ in Eq.~\eqref{eq:deltr}.
+Similarly, for the common type, $\sum_j \langle x_j\rangle_{\hat{p}} = l_2$ in Eq.
+\eqref{eq:meanfielda}, and so $\Delta_a n_2/m_2 \rightarrow 1/L$.
 Thus, the identical rare type is pathologically predicted to decline in frequency.

### 2018-07-31 manuscript.tex
l

@@
 Following the notation in the main text, the Poisson distributions for the $x_i$ (or some subset of the $x_i$) will be denoted $p$, and we use $P$ as a general shorthand for the probability of particular outcomes.
+We denote the vector of propagule abundances by ${\mathbf x} = (x_1,\ldots,x_G)$ in a given territory, and the analogous vector of nonfocal abundances by ${\mathbf x_i}=(x_1,\ldots,x_{i-1},x_{i+1}\ldots,x_G)$.
+The corresponding total propagule numbers are denoted $X=\sum_j x_j$ and $X_i=X-x_i$.
 
-We start by separating the right hand side of Eq.~\eqref{eq:growthsumuncoupled} into three components
+Similar to the classic lottery model, our approximation involves replacing the $x_i$ with effective mean values.
+However, as discussed in the text preceding Eq.~\eqref{eq:master}, it is important to treat the $x_i=1$ case separately when allowing for low propagule densities.
+We thus start by separating the right hand side of Eq.~\eqref{eq:growthsumuncoupled} into three components
 \begin{equation}
-\Delta_+ n_i = \Delta_u n_i+\Delta_r n_i+\Delta_a n_i,\label{eq:delt_decomp}
+\Delta_+ n_i = \Delta_u n_i+\Delta_r n_i+\Delta_a n_i.\label{eq:delt_decomp}
 \end{equation}
-which vary in relative magnitude depending on the propagule densities $l_i$.
+The relative magnitude of these components depends on the propagule densities $l_i$.
 The first component, $\Delta_u n_i$, accounts for territories where only one focal propagule is present ($x_i=1$ and $x_j=0$ for $j\neq i$; $u$ stands for ``uncontested'').
@@
 
-The second component, $\Delta_r n_i$, accounts for territories where a single focal propagule is present along with at least one non-focal propagule ($x_i=1$ and $X_i\geq 1$ where $X_i=\sum_{j\neq i} x_j$ is the number of nonfocal propagules; $r$ stands for ``rare'').
+The second component, $\Delta_r n_i$, accounts for territories where a single focal propagule is present along with at least one non-focal propagule ($r$ stands for ``rare'').
 The number of territories where this occurs is $Up_i(1)P(X_i\geq 1)=m_i e^{-l_i}(1-e^{-(L-l_i)})$.
@@
 \begin{equation}
-\Delta_r n_i = m_i e^{-l_i}(1-e^{-(L-l_i)})\left\langle  \frac{c_i}{c_i +\sum_{j\neq i} c_j x_j } \right\rangle_{\tilde{p}},  \label{eq:deltr}
+\Delta_r n_i = m_i e^{-l_i}(1-e^{-(L-l_i)})\left\langle  \frac{c_i}{c_i +\sum_{j\neq i} c_j x_j } \right\rangle_{p({\mathbf x}|x_i=1,X_i\geq 1)},  \label{eq:deltr}
 \end{equation}
-where $\langle \rangle_{\tilde{p}}$ denotes the expectation with respect to the probability distribution $\tilde{p}$ of nonfocal propagule abundances $x_j$, in those territories where exactly one focal propagule, and at least one non-focal propagule, landed.
+where $\langle \rangle_{p({\mathbf x}|x_i=1,X_i\geq 1)}$ denotes the expectation with respect to the conditional probability distribution $p({\mathbf x}|x_i=1,X_i\geq 1)$ of propagule abundances in those territories where exactly one focal propagule, and at least one non-focal propagule, landed.
 
-The final contribution, $\Delta_a n_i$, accounts for territories where two or more focal propagules are present ($x_i\geq 2$; $a$ stands for ``abundant"). Similar to Eq.~\eqref{eq:deltr}, we have 
+The final contribution, $\Delta_a n_i$, accounts for territories where two or more focal propagules are present ($a$ stands for ``abundant"). Similar to Eq.~\eqref{eq:deltr}, we have 
 \begin{equation}
-\Delta_a n_i=U(1-(1+l_i)e^{-l_i})\left\langle \frac{c_i x_i}{\sum_j c_j x_j} \right\rangle_{\hat{p}}\label{eq:delta}
+\Delta_a n_i=U(1-(1+l_i)e^{-l_i})\left\langle \frac{c_i x_i}{\sum_j c_j x_j} \right\rangle_{p({\mathbf x}|x_i\geq 2)}.\label{eq:delta}
 \end{equation}
-where $\hat{p}$ is the probability distribution of both focal and nonfocal propagule abundances in those territories where at least two focal propagules landed.
 
@@
 \begin{equation}
-\left\langle\frac{c_i}{c_i +\sum_{j\neq i} c_j x_j}\right\rangle_{\tilde{p}}\approx \frac{c_i}{c_i +\sum_{j\neq i} c_j \langle x_j\rangle_{\tilde{q}}}.\label{eq:meanfieldr}
+\left\langle\frac{c_i}{c_i +\sum_{j\neq i} c_j x_j}\right\rangle_{p({\mathbf x}|x_i=1,X_i\geq 1)}\approx \frac{c_i}{c_i +\sum_{j\neq i} c_j \langle x_j\rangle_r}.\label{eq:meanfieldr}
 \end{equation}
 \begin{equation}
-\left\langle \frac{c_i x_i}{\sum_j c_j x_j} \right\rangle_{\hat{p}}\approx  \frac{c_i \langle x_i \rangle_{\hat{q}}}{\sum_j c_j \langle x_j\rangle_{\hat{q}}}.\label{eq:meanfielda}
+\left\langle \frac{c_i x_i}{\sum_j c_j x_j} \right\rangle_{p({\mathbf x}|x_i\geq 2)}\approx  \frac{c_i \langle x_i \rangle_a}{\sum_j c_j \langle x_j\rangle_a}.\label{eq:meanfielda}
 \end{equation}
-Here the effective means $\langle \rangle_{\tilde{q}}$ and $\langle \rangle_{\hat{q}}$ are taken with respect to new distributions $\tilde{q}$ and $\hat{q}$, respectively.
-In the following subsection we define $\tilde{q}$ and $\hat{q}$ and explain our reasoning for using these distributions to take the effective means.
+Here $\langle \rangle_r$ and $\langle \rangle_a$ are the effective means, which are defined in the following subsection.
 
-\subsection*{The effective distributions $\tilde{q}$ and $\hat{q}$}
+\subsection*{The effective means $\langle \rangle_r$ and $\langle \rangle_a$}
 
-The approximations \eqref{eq:meanfieldr} and \eqref{eq:meanfielda} must be consistent between rare and common types.
-To illustrate, suppose that two identical types (same $b$, $c$ and $d$) are present, that the first type is rare and has vanishingly small density $l_1\rightarrow 0$ while the second is common and has high density $l_2\rightarrow \infty$.
-Since $l_2\rightarrow \infty$, uncontested territories make up a negligible fraction of the total.
-The rare type's territorial acquisition is entirely due to $\Delta_r n_1$, while the common type's territorial acquisition is entirely due to $\Delta_a n_2$.
-To ensure consistency, the approximate per-capita growth rates implied by the approximations \eqref{eq:meanfieldr} and \eqref{eq:meanfielda} must therefore be equal $\Delta_r n_1/m_1 = \Delta_a n_2/m_2$.
+The decomposition Eq.~\eqref{eq:delt_decomp} is exact and involves no additional assumptions.
+However this decomposition complicates our approximation procedure because the separate components in Eq.~\eqref{eq:delt_decomp} must be approximated in a consistent manner.
+
+To illustrate this consistency requirement, suppose that two identical types (same $b$, $c$ and $d$) are present, the first with small density $l_1\ll 1$ and the second with large density $l_2 \gg 1$.
+In this case, uncontested territories make up a negligible fraction of $U$, the first type's territorial acquisition is almost entirely due to $\Delta_r n_1$, and the second type's territorial acquisition is almost entirely due to $\Delta_a n_2$.
+For consistency, the approximate per-capita growth rates in \eqref{eq:meanfieldr} and \eqref{eq:meanfielda} must be equal $\Delta_r n_1/m_1 = \Delta_a n_2/m_2$.
 Even small violations of this consistency condition would mean exponential growth of one type relative to the other.
-This behavior is clearly pathological, because any single-type population can be arbitrarily partitioned into identical rare and common subtypes.
+This is pathological, because any single-type population can be arbitrarily partitioned into identical rare and common subtypes.
 Thus, predicted growth or decline would depend on an arbitrary assignment of rarity.
 
-In general, the above consistency condition is not satisfied when the conditional Poisson distributions $\tilde{p}$ and $\hat{p}$ are used to calculate the effective means.
-Under the above limits ($l_1\rightarrow 0$, $l_2\rightarrow \infty$), the right hand side of Eq.~\eqref{eq:meanfieldr} tends to $1/(l_2+1)$, and so $\Delta_r n_1/m_1 \rightarrow 1/(l_2+1)$ in Eq.~\eqref{eq:deltr}.
-Similarly, for the common type, $\sum_j \langle x_j\rangle_{\hat{p}} = l_2$ in Eq.
-\eqref{eq:meanfielda}, and so $\Delta_a n_2/m_2 \rightarrow 1/L$.
-Thus, the identical rare type is pathologically predicted to decline in frequency.
+Suppose that we simply used the conditional distributions $p({\mathbf x}|x_i=1,X_i\geq 1)$ and $p({\mathbf x}|x_i\geq 2)$ to calculate the effective means, such that $\langle \rangle_r=\langle \rangle_{p({\mathbf x}|x_i=1,X_i\geq 1)}$ and $\langle \rangle_a=\langle \rangle_{p({\mathbf x}|x_i\geq 2)}$.
+Then, again using the example from the previous paragraph ($l_1\ll 1$, $l_2\gg 1$), the right hand side of Eq.~\eqref{eq:meanfieldr} would be $\approx 1/(l_2+1)$, and so $\Delta_r n_1/m_1 \approx 1/(l_2+1)$ in Eq.~\eqref{eq:deltr}.
+Similarly, $\sum_j \langle x_j\rangle_a = l_2$ in Eq.~\eqref{eq:meanfielda}, and so $\Delta_a n_2/m_2 \approx 1/l_2$.
+Thus, the rare type is pathologically predicted to decline in frequency under this approximation.
+This inconsistency occurs because the expected total numbers of propagules in the respective groups of territories are different; $\langle X \rangle_{p({\mathbf x}|x_1=1,X_1\geq 1)}\approx l_2+1 > \langle X \rangle_{p({\mathbf x}|x_2\geq 2)}\approx l_2 $.
+By approximating the expected outcome of lottery competition with effective means, the rare type is effectively assumed to experience more intense lottery competition.
 
-The effective distributions $\tilde{q}$ and $\hat{q}$ are devised to avoid this pathology.
-The idea is to make the approximation that the distribution for the total number of propagules per territory is the same in all territories.
-This is only an approximation because conditioning on focal propagules being present does change the distribution of $X$ in the corresponding subset of territories (in the above example, the mean propagule density across all territories is $L$, but in the territories responsible for the growth of the rare type we have $\langle X \rangle_{\tilde{p}}=L+1$).
+The effective distributions $\hat{p}$ are devised to avoid this pathology by forcing $\langle X \rangle$ to be same in all territories.
+They are defined as follows.
+For each propagule number $x_i$, $\hat{p}(x_i)=p$ which is independent of the other number
 
-More formally, let ${\mathbf x}$ denote the vector of propagule abundances $(x_1,\ldots,x_G)$ in a given territory, and ${\mathbf x_i}=(x_1,\ldots,x_{i-1},x_{i+1}\ldots,x_G)$ similarly denote the vector of non-focal abundances, so that $p({\mathbf x_i})=p_1(x_1)\cdots p_{i-1}(x_{i-1})p_{i+1}(x_{i+1})\cdots p_G(x_G)$.
-The corresponding total propagule numbers are denoted $X=\sum_j x_j$ and $X_i=X-x_i$.
+ by making the approximation that the distribution for $X$ is the same in all territories.
+
  Then, in territories where one focal propagule and at least one non-focal propagule are present, the effective distribution is defined by 

### 2018-08-04 manuscript.tex
l

@@
 
-Suppose that we simply used the conditional distributions $p({\mathbf x}|x_i=1,X_i\geq 1)$ and $p({\mathbf x}|x_i\geq 2)$ to calculate the effective means, such that $\langle \rangle_r=\langle \rangle_{p({\mathbf x}|x_i=1,X_i\geq 1)}$ and $\langle \rangle_a=\langle \rangle_{p({\mathbf x}|x_i\geq 2)}$.
-Then, again using the example from the previous paragraph ($l_1\ll 1$, $l_2\gg 1$), the right hand side of Eq.~\eqref{eq:meanfieldr} would be $\approx 1/(l_2+1)$, and so $\Delta_r n_1/m_1 \approx 1/(l_2+1)$ in Eq.~\eqref{eq:deltr}.
+Suppose that we used the conditional distributions $p({\mathbf x}|x_i=1,X_i\geq 1)$ and $p({\mathbf x}|x_i\geq 2)$ to calculate the effective means, such that $\langle \rangle_r=\langle \rangle_{p({\mathbf x}|x_i=1,X_i\geq 1)}$ and $\langle \rangle_a=\langle \rangle_{p({\mathbf x}|x_i\geq 2)}$.
+Then, in the example from the previous paragraph ($l_1\ll 1$, $l_2\gg 1$), the right hand side of Eq.~\eqref{eq:meanfieldr} would be $\approx 1/(l_2+1)$, and so $\Delta_r n_1/m_1 \approx 1/(l_2+1)$ in Eq.~\eqref{eq:deltr}.
 Similarly, $\sum_j \langle x_j\rangle_a = l_2$ in Eq.~\eqref{eq:meanfielda}, and so $\Delta_a n_2/m_2 \approx 1/l_2$.
-Thus, the rare type is pathologically predicted to decline in frequency under this approximation.
-This inconsistency occurs because the expected total numbers of propagules in the respective groups of territories are different; $\langle X \rangle_{p({\mathbf x}|x_1=1,X_1\geq 1)}\approx l_2+1 > \langle X \rangle_{p({\mathbf x}|x_2\geq 2)}\approx l_2 $.
-By approximating the expected outcome of lottery competition with effective means, the rare type is effectively assumed to experience more intense lottery competition.
+Thus, the rare type would be pathologically predicted to decline in frequency.
+This pathological behavior occurs because the expected total density of propagules in the respective groups of territories are different; $\langle X \rangle_{p({\mathbf x}|x_1=1,X_1\geq 1)}\approx l_2+1 > \langle X \rangle_{p({\mathbf x}|x_2\geq 2)}\approx l_2 $.
+As a result, the rare type's behavior is approximated as though it experiences more intense lottery competition than the common type.
+To avoid this, the effective means must be taken in a way that ensures that the total propagule density $\langle X \rangle$ is the same in Eq.~\eqref{eq:meanfieldr} and Eq.~\eqref{eq:meanfielda}.
 
-The effective distributions $\hat{p}$ are devised to avoid this pathology by forcing $\langle X \rangle$ to be same in all territories.
-They are defined as follows.
-For each propagule number $x_i$, $\hat{p}(x_i)=p$ which is independent of the other number
-
- by making the approximation that the distribution for $X$ is the same in all territories.
-
- Then, in territories where one focal propagule and at least one non-focal propagule are present, the effective distribution is defined by 
+For the focal type $i$, we use the conditional dispersal distributions $p$ to calculate the effective mean, 
 \begin{equation}
-\tilde{q}({\mathbf x_i})=\sum_{X=2}^{\infty}P(X|X\geq 2) p({\mathbf x_i}|X_i=X-1),
+\langle x_i \rangle_r=1,\qquad \langle x_i \rangle_a = \langle x_i \rangle_{p(x_i|x_i\geq 2)}.
 \end{equation}
-where the total number of propagules $X$ follows a Poisson distribution with mean $L$, and $P(X|X\geq 2)=P(X)/P(X\geq 2)=P(X)/(1-(1+L)e^{-L})$.
-Similarly, in territories where more than one focal propagule is present, the effective distribution is defined by
-\begin{equation}
-\hat{q}({\mathbf x})=\sum_{X=2}^{\infty}P(X|X\geq 2) p({\mathbf x}|x_i\geq 2, X).
-\end{equation}
-
-\subsection*{Calculating the effective means}
-
-Here we calculate the effective means, starting with the $\Delta_r n_i$ component. We have
+However, for nonfocal types $j\neq i$, we explicitly evaluate the $X$-dependence of the conditional dispersal probabilities, assuming that $X$ has the same conditional Poisson distribution $P(X|X\geq 2)$ for both $\langle \rangle_r$ and $\langle \rangle_a$.
+Starting with $\langle \rangle_r$, we first imagine that $X$ is a fixed constant to obtain
 \begin{align}
-\langle x_j \rangle_{\tilde{q}}&=\sum_{\mathbf x_i} \tilde{q}({\mathbf x_i})x_j\nonumber\\
-&=\frac{1}{1-(1+L)e^{-L}}\sum_{X=2}^{\infty} P(X) \sum_{\mathbf x_i} p({\mathbf x_i}|X_i=X-1)x_j.
+\sum_{x_j} p(x_j|x_i=1,X)x_j=\frac{l_j}{L-l_i}(X-1)
 \label{eq:raremonster1}
 \end{align}
-The inner sum over ${\mathbf x_i}$ is the mean number of propagules of a given nonfocal type $j$ that will be found in a territory which received $X-1$ nonfocal propagules in total, which is equal to $\frac{l_j}{L-l_i}(X-1)$.
-Thus,
+The right hand side is obtained by observing that the sum is the expected number of propagules with type $j$ that will be found in a territory which received $X-1$ nonfocal propagules in total.
+Applying our assumption about the distribution of $X$ gives
 \begin{align}
-\langle x_j \rangle_{\tilde{q}}&=\frac{l_j}{1-(1+L)e^{-L}}\frac{1}{L-l_i}\sum_{X=2}^{\infty} P(X) (X-1)\nonumber\\
-&=\frac{l_j}{1-(1+L)e^{-L}}\frac{L-1+e^{-L}}{L-l_i},
+\langle x_j \rangle_r&=\frac{l_j}{L-l_i}\sum_{X=2}^{\infty} P(X|X\geq 2) (X-1)\nonumber\\
+&=\frac{l_j}{L-l_i}\frac{L-1+e^{-L}}{1-(1+L)e^{-L}},
 \label{eq:meanxjrare}
 \end{align}
-where the last line follows from $\sum_{X=2}^{\infty} P(X)(X-1)=\sum_{X=1}^{\infty} P(X)(X-1)=\sum_{X=1}^{\infty} P(X)X-\sum_{X=1}^{\infty}P(X)$.
+where the last line follows from $P(X|X\geq 2) =\frac{1}{{1-(1+L)e^{-L}}}P(X)$ and $\sum_{X=2}^{\infty} P(X)(X-1)=\sum_{X=1}^{\infty} P(X)(X-1)=L-1+e^{-L}$.
 Substituting Eqs.~\eqref{eq:meanfieldr} and \eqref{eq:meanxjrare} into Eq.~\eqref{eq:deltr}, we obtain
@@
 
-Turning now to the $\Delta_a n_i$ component, the mean focal abundance is 
+Turning now to $\langle \rangle_a$, the mean focal abundance is 
 \begin{align}
-\langle x_i \rangle_{\hat{q}}&=\sum_{\mathbf x} \hat{q}({\mathbf x}) x_i\nonumber \\
-&=\sum_{x_i} p(x_i|x_i\geq 2)x_i \nonumber\\
+\langle x_i \rangle_{a}&=\sum_{x_i} p(x_i|x_i\geq 2)x_i \nonumber\\
 &=\frac{1}{1-(1+l_i)e^{-l_i}}\sum_{x_i\geq 2} p(x_i)x_i\nonumber\\
@@
 \end{align}
-For nonfocal types $j\neq i$, we have
+For nonfocal types $j\neq i$, we have analogously to Eq.~\eqref{eq:raremonster1},
 \begin{align}
-\langle x_j \rangle_{\hat{q}}&=\sum_{X=2}^{\infty}P(X|X\geq 2)\sum_{\mathbf x}  p({\mathbf x}|x_i\geq 2,X) x_j\nonumber\\
-&=\sum_{X=2}^{\infty}P(X|X\geq 2)\sum_{x_i} p(x_i|x_i\geq 2,X) \sum_{\mathbf x_i}  p(\mathbf x_i|X_i=X-x_i) x_j\nonumber\\
-&=\sum_{X=2}^{\infty}P(X|X\geq 2)\sum_{x_i}p(x_i|x_i\geq 2,X) \frac{l_j(X-x_i)}{L-l_i} \nonumber\\
-&=\frac{l_j}{L-l_i}\left[\sum_{X=2}^{\infty}P(X|X\geq 2)X - \sum_{x_i}p(x_i|x_i\geq 2) x_i \right]\nonumber\\
+\sum_{\mathbf x_i}  p(\mathbf x_i|x_i\geq 2,X) x_j&= \sum_{\mathbf x_i}  p(\mathbf x_i|X_i=X-x_i) x_j\nonumber\\
+&= \frac{l_j(X-x_i)}{L-l_i}.
+\end{align}
+Assuming again that $X$ has the conditional distribution $P(X|X\geq 2)$ yields 
+\begin{align}
+\langle x_j \rangle_a&=\frac{l_j}{L-l_i}\left[\sum_{X=2}^{\infty}P(X|X\geq 2)X - \langle x_i \rangle_a \right]\nonumber\\
 &=\frac{l_j}{L-l_i}\left( L\frac{1-e^{-L}}{1-(1+L)e^{-L}}- l_i\frac{1-e^{-l_i}}{1-(1+l_i)e^{-l_i}}\right).
 \end{align}
-In going from line 2 to 3, we used the same logic used to evaluate the inner sum in Eq.~\eqref{eq:raremonster1}, and in going from 3 to 4 we have separately evaluated the contributions from the $X$ and $x_i$ terms in the numerator.
 Combining these results with Eqs.~\eqref{eq:delta} and \eqref{eq:meanfielda}, we obtain
@@
 
-Eq.~\eqref{eq:meanfieldr} and \eqref{eq:meanfielda} must not only be consistent with each other, they must also be individually good approximations. Here we evaluate these approximations.
+Having derived the approximation Eq.~\eqref{eq:master}, we now evaluate its domain of validity.
+Eq.~\eqref{eq:master} relies on ignoring the fluctuations in $x_i$ and $x_j$, such that we can replace them with constant effective mean values.
+Mathematically, this requires that the standard deviations $\sigma_{p({\mathbf x}|x_i=1,X_i\geq 1)}(\sum_{j\neq i} c_j x_j)$ and $\sigma_{p({\mathbf x}|x_i\geq 2)}(\sum_j c_j x_j)$ are sufficiently small compared to the corresponding means $\langle\sum_{j\neq i} c_j x_j\rangle_{p({\mathbf x}|x_i=1,X_i\geq 1)}$ and $\langle\sum_j c_j x_j\rangle_{p({\mathbf x}|x_i\geq 2)}$ in Eqs.~\eqref{eq:meanfieldr} and \eqref{eq:meanfielda} respectively.
+We now evaluate these conditions.
 
-The fundamental requirement for making the replacement in Eqs.~\eqref{eq:meanfieldr} and \eqref{eq:meanfielda} is that we can ignore the fluctuations in the $x_i$ and hence replace them with a constant effective mean value.
-Mathematically, we require that the standard deviations $\sigma_{\tilde{q}}(\sum_{j\neq i} c_j x_j)$ and $\sigma_{\hat{q}}(\sum_j c_j x_j)$ must be sufficiently small compared to the corresponding means $\langle\sum_{j\neq i} c_j x_j\rangle_{\tilde{q}}$ and $\langle\sum_j c_j x_j\rangle_{\hat{q}}$ in Eqs.~\eqref{eq:meanfieldr} and \eqref{eq:meanfielda} respectively.
-
-To evaluate these standard deviations, we will work with $\tilde{p}$ and $\hat{p}$ distributions instead of $\tilde{q}$ and $\hat{q}$.
-This is mathematically much simpler because the $x_i$ are independent under $\tilde{p}$ and $\hat{p}$, and is justified by the fact that $\tilde{p}$ and $\hat{p}$ are closely related to $\tilde{q}$ and $\hat{q}$ respectively, and so we expect the relevant means and standard deviations will be similar.
-
-Starting with Eq.~\eqref{eq:meanfieldr}, we have $\langle x_j \rangle_{\tilde{p}}=l_j/C$, where $C=1-e^{-(L-l_i)}$, and the corresponding variances and covariances are given by
+We first calculate $\sigma_{p({\mathbf x}|x_i=1,X_i\geq 1)}(\sum_{j\neq i} c_j x_j)$.
+We have $\langle x_j \rangle_{p({\mathbf x}|x_i=1,X_i\geq 1)}=l_j/C$, where $C=1-e^{-(L-l_i)}$, and the corresponding variances and covariances are given by
 \begin{align}
-\sigma_{\tilde{p}}^2(x_j)&=\langle x_j^2 \rangle_{\tilde{p}}-\langle x_j \rangle_{\tilde{p}}^2\nonumber\\
+\sigma_{p({\mathbf x}|x_i=1,X_i\geq 1)}^2(x_j)&=\langle x_j^2 \rangle_{p({\mathbf x}|x_i=1,X_i\geq 1)}-\langle x_j \rangle_{p({\mathbf x}|x_i=1,X_i\geq 1)}^2\nonumber\\
 &=\frac{l_j^2 + l_j}{C}-\frac{l_j^2}{C^2}\nonumber \\
@@
 \begin{align}
-\sigma_{\tilde{p}}(x_j,x_k)&=\langle x_j x_k \rangle_{\tilde{p}}-\langle x_j \rangle_{\tilde{p}}\langle x_k \rangle_{\tilde{p}}\nonumber\\
-&=\frac{1}{C}\langle x_j x_k \rangle_p-\frac{l_jl_k}{C^2}\nonumber\\
+\sigma_{p({\mathbf x}|x_i=1,X_i\geq 1)}(x_j,x_k)&=\langle x_j x_k \rangle_{p({\mathbf x}|x_i=1,X_i\geq 1)}-\langle x_j \rangle_{p({\mathbf x}|x_i=1,X_i\geq 1)}\langle x_k \rangle_{p({\mathbf x}|x_i=1,X_i\geq 1)}\nonumber\\
+&=\frac{1}{C}\langle x_j x_k \rangle_{p({\mathbf x}|x_i=1,X_i\geq 1)}-\frac{l_jl_k}{C^2}\nonumber\\
 &=\left(1-\frac{1}{C}\right)\frac{l_j l_k}{C}\qquad\qquad j\neq k. \label{eq:covr}
@@
 \begin{equation}
-\sigma_{\tilde{p}}^2(\sum_{j\neq i} c_j x_j)=\sum_{j\neq i}\left[c_j^2\sigma_{\tilde{p}}^2(x_j)+2\sum_{k>j, k\neq i}c_j c_k\sigma_{\tilde{p}}(x_j,x_k)\right],\label{eq:vartotr}
+\sigma^2(\sum_{j\neq i} c_j x_j)=\sum_{j\neq i}\left[c_j^2\sigma^2(x_j)+2\sum_{k>j, k\neq i}c_j c_k\sigma(x_j,x_k)\right],\label{eq:vartotr}
 \end{equation}
@@
 \begin{equation}
-\frac{\sigma(\sum_{j\neq i} c_j x_j)}{\langle\sum_{j\neq i} c_j x_j\rangle}=C^{1/2}\frac{\left(\sum_{j\neq i}c_j^2 l_j+(1-\frac{1}{C})\left(\sum_{j\neq i}c_j l_j\right)^2 \right)^{1/2}}{\sum_{j\neq i}c_j l_j}.
+\frac{\sigma_{p({\mathbf x}|x_i=1,X_i\geq 1)}(\sum_{j\neq i} c_j x_j)}{\langle\sum_{j\neq i} c_j x_j\rangle_{p({\mathbf x}|x_i=1,X_i\geq 1)}}=C^{1/2}\frac{\left(\sum_{j\neq i}c_j^2 l_j+(1-\frac{1}{C})\left(\sum_{j\neq i}c_j l_j\right)^2 \right)^{1/2}}{\sum_{j\neq i}c_j l_j}.
 \label{eq:cvr}
@@
 
-Eq.~\eqref{eq:cvr} reveals two key points.
+Eq.~\eqref{eq:cvr} reveals two important points.
 First, when the $c_j$ have similar magnitudes (their ratios are of order one), Eq.~\eqref{eq:meanfieldr} is an excellent approximation.
@@
 
-Turning now to Eq.~\eqref{eq:meanfielda}, all covariances between nonfocal types are now zero, so that $\sigma_{\hat{p}}^2(\sum c_j x_j)=\sum c_j^2 \sigma_{\hat{p}}^2(x_j)$, where $\sigma_{\hat{p}}^2(x_j)=l_j$ for $j\neq i$.
+Turning now to Eq.~\eqref{eq:meanfielda}, all covariances between nonfocal types are now zero, so that $\sigma_{_{p({\mathbf x}|x_i\geq 2)}}^2(\sum c_j x_j)=\sum c_j^2 \sigma_{\hat{p}}^2(x_j)$, where $\sigma_{\hat{p}}^2(x_j)=l_j$ for $j\neq i$.
 Here

### 2018-08-27 manuscript.tex
l

@@
 To illustrate this consistency requirement, suppose that two identical types (same $b$, $c$ and $d$) are present, the first with small density $l_1\ll 1$ and the second with large density $l_2 \gg 1$.
-In this case, uncontested territories make up a negligible fraction of $U$, the first type's territorial acquisition is almost entirely due to $\Delta_r n_1$, and the second type's territorial acquisition is almost entirely due to $\Delta_a n_2$.
+In this case, uncontested territories make up a negligible fraction of $U$; the first type's territorial acquisition is almost entirely due to $\Delta_r n_1$; and the second type's territorial acquisition is almost entirely due to $\Delta_a n_2$.
 For consistency, the approximate per-capita growth rates in \eqref{eq:meanfieldr} and \eqref{eq:meanfielda} must be equal $\Delta_r n_1/m_1 = \Delta_a n_2/m_2$.
 Even small violations of this consistency condition would mean exponential growth of one type relative to the other.
-This is pathological, because any single-type population can be arbitrarily partitioned into identical rare and common subtypes.
+This behavior is pathological, because any single-type population can be arbitrarily partitioned into identical rare and common subtypes.
 Thus, predicted growth or decline would depend on an arbitrary assignment of rarity.
 
-Suppose that we used the conditional distributions $p({\mathbf x}|x_i=1,X_i\geq 1)$ and $p({\mathbf x}|x_i\geq 2)$ to calculate the effective means, such that $\langle \rangle_r=\langle \rangle_{p({\mathbf x}|x_i=1,X_i\geq 1)}$ and $\langle \rangle_a=\langle \rangle_{p({\mathbf x}|x_i\geq 2)}$.
+Suppose that we naively used the conditional distributions $p({\mathbf x}|x_i=1,X_i\geq 1)$ and $p({\mathbf x}|x_i\geq 2)$ to calculate the effective means, such that $\langle \rangle_r=\langle \rangle_{p({\mathbf x}|x_i=1,X_i\geq 1)}$ and $\langle \rangle_a=\langle \rangle_{p({\mathbf x}|x_i\geq 2)}$.
 Then, in the example from the previous paragraph ($l_1\ll 1$, $l_2\gg 1$), the right hand side of Eq.~\eqref{eq:meanfieldr} would be $\approx 1/(l_2+1)$, and so $\Delta_r n_1/m_1 \approx 1/(l_2+1)$ in Eq.~\eqref{eq:deltr}.
-Similarly, $\sum_j \langle x_j\rangle_a = l_2$ in Eq.~\eqref{eq:meanfielda}, and so $\Delta_a n_2/m_2 \approx 1/l_2$.
-Thus, the rare type would be pathologically predicted to decline in frequency.
+Similarly, $\sum_j \langle x_j\rangle_a \approx l_2$ in Eq.~\eqref{eq:meanfielda}, and so $\Delta_a n_2/m_2 \approx 1/l_2$.
+Thus, the rare type would be predicted to decline in frequency even though it has identical traits.
 This pathological behavior occurs because the expected total density of propagules in the respective groups of territories are different; $\langle X \rangle_{p({\mathbf x}|x_1=1,X_1\geq 1)}\approx l_2+1 > \langle X \rangle_{p({\mathbf x}|x_2\geq 2)}\approx l_2 $.
@@
 \begin{equation}
-\langle x_i \rangle_r=1,\qquad \langle x_i \rangle_a = \langle x_i \rangle_{p(x_i|x_i\geq 2)}.
+\langle x_i \rangle_r=1,\qquad \langle x_i \rangle_a = \langle x_i \rangle_{p(x_i|x_i\geq 2)}. \label{eq:effectivefocal}
 \end{equation}
-However, for nonfocal types $j\neq i$, we explicitly evaluate the $X$-dependence of the conditional dispersal probabilities, assuming that $X$ has the same conditional Poisson distribution $P(X|X\geq 2)$ for both $\langle \rangle_r$ and $\langle \rangle_a$.
-Starting with $\langle \rangle_r$, we first imagine that $X$ is a fixed constant to obtain
+However, for nonfocal types $j\neq i$, we separately evaluate the $X$-dependence of the conditional dispersal probabilities to ensure that $X$ has the same distribution for both $\langle \rangle_r$ and $\langle \rangle_a$.
+Specifically, we assume that $X$ follows a  Poisson distribution with rate parameter $L$, conditional on $X\geq 2$; this distribution will be denoted $P(X|X\geq 2)$.
+
+Starting with Eq.~\eqref{eq:meanfieldr}, we only need to evaluate $\langle x_j \rangle_r$ since $\langle x_i \rangle_r=1$. To evaluate the $X$-dependence separately, we first hold $X$ fixed to obtain
 \begin{align}
-\sum_{x_j} p(x_j|x_i=1,X)x_j=\frac{l_j}{L-l_i}(X-1)
+\sum_{x_j} p(x_j|x_i=1,X)x_j=\frac{l_j}{L-l_i}(X-1) \qquad j\neq i.
 \label{eq:raremonster1}
 \end{align}
-The right hand side is obtained by observing that the sum is the expected number of propagules with type $j$ that will be found in a territory which received $X-1$ nonfocal propagules in total.
-Applying our assumption about the distribution of $X$ gives
+The right hand side is obtained by observing that the sum on the left is the expected number of propagules with type $j$ that will be found in a territory which received $X-1$ nonfocal propagules in total.
+We then take the with respect to $P(X|X\geq 2)$ to give
 \begin{align}
@@
 
-Turning now to $\langle \rangle_a$, the mean focal abundance is 
+Turning now to Eq.~\eqref{eq:meanfielda}, from Eq.~\eqref{eq:effectivefocal} the mean focal abundance is 
 \begin{align}
@@
 \end{align}
-Assuming again that $X$ has the conditional distribution $P(X|X\geq 2)$ yields 
+Again taking the expection with respect to $P(X|X\geq 2)$ yields 
 \begin{align}
@@
 Eq.~\eqref{eq:master} relies on ignoring the fluctuations in $x_i$ and $x_j$, such that we can replace them with constant effective mean values.
-Mathematically, this requires that the standard deviations $\sigma_{p({\mathbf x}|x_i=1,X_i\geq 1)}(\sum_{j\neq i} c_j x_j)$ and $\sigma_{p({\mathbf x}|x_i\geq 2)}(\sum_j c_j x_j)$ are sufficiently small compared to the corresponding means $\langle\sum_{j\neq i} c_j x_j\rangle_{p({\mathbf x}|x_i=1,X_i\geq 1)}$ and $\langle\sum_j c_j x_j\rangle_{p({\mathbf x}|x_i\geq 2)}$ in Eqs.~\eqref{eq:meanfieldr} and \eqref{eq:meanfielda} respectively.
-We now evaluate these conditions.
+Mathematically, this requires that the standard deviations $\sigma_{p({\mathbf x}|x_i=1,X_i\geq 1)}(\sum_{j\neq i} c_j x_j)$ and $\sigma_{p({\mathbf x}|x_i\geq 2)}(\sum_j c_j x_j)$ are sufficiently small compared to the corresponding means $\langle\sum_{j\neq i} c_j x_j\rangle_{p({\mathbf x}|x_i=1,X_i\geq 1)}$ and $\langle\sum_j c_j x_j\rangle_{p({\mathbf x}|x_i\geq 2)}$ in Eqs.~\eqref{eq:meanfieldr} and \eqref{eq:meanfielda}.
 
-We first calculate $\sigma_{p({\mathbf x}|x_i=1,X_i\geq 1)}(\sum_{j\neq i} c_j x_j)$.
+We first consider the means and standard deviations in Eq.~\eqref{eq:meanfieldr}.
 We have $\langle x_j \rangle_{p({\mathbf x}|x_i=1,X_i\geq 1)}=l_j/C$, where $C=1-e^{-(L-l_i)}$, and the corresponding variances and covariances are given by
@@
 \end{equation}
+Eq.~\eqref{eq:cvr} shows that, when the $c_j$ have similar magnitudes (their ratios are of order one), Eq.~\eqref{eq:meanfieldr} is an excellent approximation.
+The right hand side of Eq.~\eqref{eq:cvr} is then approximately equal to $C^{1/2}\left(\frac{1}{L-l_i}+1-\frac{1}{C}\right)^{1/2}$, which is small for both low and high nonfocal densities.
+The worst case scenario occurs when $L-l_i$ is of order one, and it can be directly verified that Eq.~\eqref{eq:meanfieldr} is then still a good approximation (see Fig.~\ref{fig:approx_details}).
 
-Eq.~\eqref{eq:cvr} reveals two important points.
-First, when the $c_j$ have similar magnitudes (their ratios are of order one), Eq.~\eqref{eq:meanfieldr} is an excellent approximation.
-In this case, the right hand side of Eq.~\eqref{eq:cvr} is approximately equal to $C^{1/2}\left(\frac{1}{L-l_i}+1-\frac{1}{C}\right)^{1/2}$, which is small for both low and high nonfocal densities.
-The worst case scenario occurs when $L-l_i$ is of order one, and it can be directly verified that Eq.~\eqref{eq:meanfieldr} is then still a good approximation (see Fig.~\ref{fig:approx_details}).
-Second, if some of the $c_j$ are much larger than the others, the relative fluctuations in $\sum_{j\neq i} c_j x_j$ can be large.
-Specifically, in the presence of a rare,  strong competitor ($c_j l_j\gg c_{j'} l_{j'}$ for all other nonfocal types $j'$, and $l_j\ll 1$), then the right hand side of Eq.
-\eqref{eq:cvr} can be large and we cannot make the replacement Eq.~\eqref{eq:meanfieldr}.
+Turning to Eq.~\eqref{eq:meanfielda}, all covariances between nonfocal types are now zero, so that $\sigma_{p({\mathbf x}|x_i\geq 2)}^2(\sum c_j x_j)=\sum c_j^2 \sigma_{p({\mathbf x}|x_i\geq 2)}^2(x_j)$.
+For nonfocal types ($j\neq i$) $\sigma_{p({\mathbf x}|x_i\geq 2)}^2(x_j)=l_j$, whereas for the focal type we have
+\begin{equation}
+\sigma_{p({\mathbf x}|x_i\geq 2)}^2(x_i)=\frac{l_i}{D}\left(l_i+1-e^{-l_i}-\frac{l_i}{D}\left(1-e^{-l_i}\right)^2\right),
+\end{equation}
+where $D= 1-(1+l_i)e^{-l_i}$, and 
+\begin{equation}
+\frac{\sigma_{p({\mathbf x}|x_i\geq 2)}(\sum c_j x_j)}{\langle\sum c_j x_j\rangle_{p({\mathbf x}|x_i\geq 2)}} = \frac{\left(\sum_{j\neq i} c_j^2 l_j + c_i^2 \sigma_{\hat{p}}^2(x_i)\right)^{1/2}}{\sum_{j\neq i} c_j l_j + c_i l_i (1-e^{-l_i})/D} \label{eq:cva}.
+\end{equation}
+Similarly to Eq.~\eqref{eq:cvr}, the right hand side of Eq.
+\eqref{eq:cva} is small for both low and high nonfocal densities provided that the $c_j$ have similar magnitudes.
+Again, the worst case scenario occurs when $l_i$ and $L-l_i$ are of order $1$, but Eq.~\eqref{eq:meanfielda} is still a good approximation in this case (Fig.~\ref{fig:approx_details}).
+
+In both Eqs.~\eqref{eq:cvr} and \eqref{eq:cva}, the standard deviation in $\sum_{j\neq i} c_j x_j$ can be large relative to its mean if some of the $c_j$ are much larger than the others.
+Specifically, in the presence of a rare, strong competitor ($c_j l_j\gg c_{j'} l_{j'}$ for all other nonfocal types $j'$, and $l_j\ll 1$), then the right hand side of Eqs.~\eqref{eq:cvr} and ~\eqref{eq:cva} can be large and we cannot make the replacement Eq.~\eqref{eq:meanfieldr}.
 Fig.~\ref{fig:approx_details} shows the breakdown of the effective mean approximation when there are large differences in $c$.
@@
 \includegraphics[scale=0.8]{approx_details.pdf}
-\caption{\label{fig:approx_details} Comparison of our $\tilde{q}$,$\hat{q}$ approximation with simulations, and also with the naive $\tilde{p}$,$\hat{p}$ approximation, as a function of the relative $c$ difference between two types.
-Our approximation breaks down in the presence of large $c$ differences.
-The inset shows the pathology of the $\tilde{p}$,$\hat{p}$ approximation --- growth rates are not equal in the neutral case $c_1=c_2$.
+\caption{\label{fig:approx_details} Comparison of our effective mean approximation Eq.~\eqref{eq:master} with simulations, and also with the naive $\langle \rangle_r=\langle\rangle_{p({\mathbf x}|x_i=1,X_i\geq 1)}$ and $\langle\rangle_a=\langle\rangle_{p({\mathbf x}|x_i\geq 2)}$ approximation, as a function of the relative $c$ difference between two types.
+Eq.~\eqref{eq:master} breaks down in the presence of large $c$ differences.
+The inset shows the pathology of the naive approximation --- growth rates for rare and common types are not equal in the neutral case $c_1=c_2$.
 Simulation procedure is the same as in Fig.~\ref{fig:simcomp}, with $U=10^5$.}
@@
 
-Turning now to Eq.~\eqref{eq:meanfielda}, all covariances between nonfocal types are now zero, so that $\sigma_{_{p({\mathbf x}|x_i\geq 2)}}^2(\sum c_j x_j)=\sum c_j^2 \sigma_{\hat{p}}^2(x_j)$, where $\sigma_{\hat{p}}^2(x_j)=l_j$ for $j\neq i$.
-Here
-\begin{equation}
-\sigma_{\hat{p}}^2(x_i)=\frac{l_i}{D}\left(l_i+1-e^{-l_i}-\frac{l_i}{D}\left(1-e^{-l_i}\right)^2\right),
-\end{equation}
-where $D= 1-(1+l_i)e^{-l_i}$, and 
-\begin{equation}
-\frac{\sigma_{\hat{p}}(\sum c_j x_j)}{\langle\sum c_j x_j\rangle} = \frac{\left(\sum_{j\neq i} c_j^2 l_j + c_i^2 \sigma_{\hat{p}}^2(x_i)\right)^{1/2}}{\sum_{j\neq i} c_j l_j + c_i l_i (1-e^{-l_i})/D} \label{eq:cva}.
-\end{equation}
-
-Similarly to Eq.~\eqref{eq:cvr}, the right hand side of Eq.
-\eqref{eq:cva} is small for both low and high nonfocal densities.
-Again, the worst case scenario occurs when $l_i$ and $L-l_i$ are of order $1$, but Eq.~\eqref{eq:meanfielda} is still a good approximation in this case.
-Again, the approximation breaks down in the presence of a rare, strong competitor (Fig.~\ref{fig:approx_details}).
-
-\section*{Appendix B: Total density in the Lotka-Volterra competition model}
+\section*{Appendix B: Total density under Lotka-Volterra competition}
 

### 2018-08-27 approximation_details.py

@@ -55,9 +55,9 @@ deltaovermapprox1=np.array(map(lambda x: deltnplus(m,x,T),cs))/m
 
 ax1.plot(deltaovermsim1[:,0],'k.',markersize=3,label=r'Simulation')
 ax1.plot(deltaovermsim1[:,1],'k.',markersize=3)
-ax1.plot(deltaovermnaive1[:,0],'r',linewidth=2,label=r'$\tilde{p}$,$\hat{p}$ approximation')
+ax1.plot(deltaovermnaive1[:,0],'r',linewidth=2,label=r'Naive approx.')
 ax1.plot(deltaovermnaive1[:,1],'r',linewidth=2)
-ax1.plot(deltaovermapprox1[:,0],'b',linewidth=2,label=r'$\tilde{q}$,$\hat{q}$ approximation')
+ax1.plot(deltaovermapprox1[:,0],'b',linewidth=2,label=r'Eq.~(6)')
 ax1.plot(deltaovermapprox1[:,1],'b',linewidth=2)
 ax1.plot([50,50],[0,1],'k',linewidth=2)
 ax1.set_xticklabels([])

### 2018-08-30 manuscript.tex
l

@@
 We find that density-independent relative fitnesses accurately describe strong selection acting on any one trait, even fecundity, which is both density-regulating and subject to density-dependent selection.
-Our findings suggest that deviations from demographic equilibrium are the most serious threat to relative fitness models.
+Our findings suggest that deviations from demographic equilibrium pose the most serious difficulties for relative fitness models.
 In such cases our model offers a possible alternative to relative fitness.
@@
 
-\subsection*{Density-regulating traits and the threat of strong selection}
+\subsection*{Density-regulating traits under strong selection}
 
@@
 Selection is also independent of $N$ when the population is at demographic equilibrium and $N$ is unaffected by ongoing selection; as is the case for $c$-selection.
-Thus, to threaten Eq.~\eqref{eq:canonicaldisc}, we require that selection is density-dependent, and also that density is changing.
+Thus, to invalidate Eq.~\eqref{eq:canonicaldisc}, we require that selection is density-dependent, and also that density is changing.
 This can obviously occur if density-dependent selection happens in a population far from demographic equilibrium, in which case the validity of Eq.~\eqref{eq:canonicaldisc} depends on the specifics of the rate and magnitude of demographic change (we return to this in the Discussion).
-However, Eq.~\eqref{eq:canonicaldisc} can be threatened even in demographically-stable populations if a density-regulating trait is subject to density-dependent selection, as is the case for $b$ in our variable-density lottery.
+However, Eq.~\eqref{eq:canonicaldisc} can break down even in demographically-stable populations if a density-regulating trait is subject to density-dependent selection, as is the case for the $b$ trait in our variable-density lottery.
 
-Before we discuss the $b$ trait, it is helpful to summarize how density-dependent selection on a density-regulating trait threatens Eq.~\eqref{eq:canonicalcont} in simpler continuous-time models.
+Before we discuss the $b$ trait, it is helpful to summarize how density-dependent selection on a density-regulating trait can invalidate Eq.~\eqref{eq:canonicalcont} in simpler continuous-time models.
 This applies, for example, to $K$-selection in the logistic \citep{kimura1969natural,crow_1970}.
@@
 During the sweep there is some deviation in $f(\overline{b},N)$, but this deviation is an order of magnitude smaller than for a $\delta$ sweep (the density-dependent deviation in Fig.~\ref{fig:strengthofselection} is of order $\epsilon$, whereas the analogous effect for $b$ sweep in our variable-density lottery is only of order $\epsilon^2$; see Appendix D for details).
-Since selection must already be strong for a $\delta$-sweep to threaten Eq.~\eqref{eq:canonicalcont}, the density-independent model applies effectively exactly for equilibrium $b$-sweeps (Fig.~\ref{fig:bsweep}).
+Since selection must already be strong for a $\delta$-sweep to invalidate Eq.~\eqref{eq:canonicalcont}, the density-independent model applies effectively exactly for equilibrium $b$-sweeps (Fig.~\ref{fig:bsweep}).
 
@@
 The $c$ and $d$ traits exemplify the two distinct directions in which density and selection can interact: selection may depend on density, and density may change in response to ongoing selection \citep{prout_1980}.
-The combination of both is necessary to threaten the constant-$s$ approximation.
+The combination of both is necessary to invalidate the constant-$s$ approximation.
 Remarkably, the $b$ trait demonstrates that the combination is not sufficient; the density-dependence of $b$-selection effectively disappears over equilibrium-to-equilibrium $b$-sweeps.
@@
 
-The above findings underscore that the most serious threat to the density-independent models of selection (Eqs.~\eqref{eq:canonicalcont} and \eqref{eq:canonicaldisc}) arises due to deviations from demographic equilibrium as a result of changes in the demographic rates of the types already present i.e.
-as a result of a temporally-variable environment.
-While transient deviations from demographic equilibrium driven by the appearance of new types can also threaten the density-independent approximation, this requires strong selection that is both density-dependent and affects a density-regulating trait (and, as exemplified by $b$-selection, even then the approximation may hold).
+The above findings suggest that the largest deviations from the approximation of density-independent selection (as represented by Eqs.~\eqref{eq:canonicalcont} and \eqref{eq:canonicaldisc}) will occur in populations far from demographic equilibrium e.g.~as a result of a temporally-variable environment.
+While transient deviations from demographic equilibrium driven by the appearance of new types can also cause the density-independent approximation to break down, this requires strong selection that is both density-dependent and affects a density-regulating trait (and, as exemplified by $b$-selection, even then the approximation may hold).
 By contrast, temporally-variable environments can dramatically alter frequency trajectories for individual sweeps (e.g.
-Fig.
-9.5 in \cite{otto_2011}; Fig.
-5 in \cite{mallet_2012}), as well as the long-term outcomes of selection \citep{lande_2009}.
+Fig.~9.5 in \cite{otto_2011}; Fig.~5 in \cite{mallet_2012}), as well as the long-term outcomes of selection \citep{lande_2009}.
 
@@
 Thus, the rare type would be predicted to decline in frequency even though it has identical traits.
+
 This pathological behavior occurs because the expected total density of propagules in the respective groups of territories are different; $\langle X \rangle_{p({\mathbf x}|x_1=1,X_1\geq 1)}\approx l_2+1 > \langle X \rangle_{p({\mathbf x}|x_2\geq 2)}\approx l_2 $.
-As a result, the rare type's behavior is approximated as though it experiences more intense lottery competition than the common type.
-To avoid this, the effective means must be taken in a way that ensures that the total propagule density $\langle X \rangle$ is the same in Eq.~\eqref{eq:meanfieldr} and Eq.~\eqref{eq:meanfielda}.
+As a result, the rare type's behavior is approximated as though it experiences more intense lottery competition than the common type, which cannot be the case since the two types are identical.
+The effective means must thus be taken in a way that ensures that the expected total propagule density is the same in Eq.~\eqref{eq:meanfieldr} and Eq.~\eqref{eq:meanfielda}.
 
-For the focal type $i$, we use the conditional dispersal distributions $p$ to calculate the effective mean, 
+We achieve this as follows.
+For nonfocal types $j\neq i$, we separately evaluate the $X$-dependence of the conditional dispersal probabilities to ensure that $X$ has the same distribution for both $\langle \rangle_r$ and $\langle \rangle_a$.
+Specifically, we assume that $X$ follows a Poisson distribution with rate parameter $L$, conditional on $X\geq 2$; this distribution will be denoted $P(X|X\geq 2)$.
+However, for the focal type $i$, we use the exact conditional dispersal distributions $p$ to calculate the effective mean,
 \begin{equation}
@@
 \end{equation}
-However, for nonfocal types $j\neq i$, we separately evaluate the $X$-dependence of the conditional dispersal probabilities to ensure that $X$ has the same distribution for both $\langle \rangle_r$ and $\langle \rangle_a$.
-Specifically, we assume that $X$ follows a  Poisson distribution with rate parameter $L$, conditional on $X\geq 2$; this distribution will be denoted $P(X|X\geq 2)$.
+This assumption is mathematically convenient
 
@@
 
-In section ``Density-regulating traits and the threat of strong selection'' we argued that the density-dependent factor $f(\overline{b},N)=\frac{1-e^{-\overline{b}N/T}}{N}(T-N)$ is unchanged at the beginning and end points of an equilibrium to equilibrium sweep of a type with higher $b$.
+In section ``Density-regulating traits under strong selection'' we argued that the density-dependent factor $f(\overline{b},N)=\frac{1-e^{-\overline{b}N/T}}{N}(T-N)$ is unchanged at the beginning and end points of an equilibrium to equilibrium sweep of a type with higher $b$.
 Here we estimate the magnitude of the deviation in $f(\overline{b},N)$ during the sweep.

### 2018-08-30 strength_of_selection.py
l

@@ -12,7 +12,7 @@ from scipy.integrate import odeint
 #SS figure
 #========================================================
 
-fig1, (ax1, ax2) = plt.subplots(ncols=2,figsize=[8,4])
+fig1, ax1 = plt.subplots(ncols=1,figsize=[4,4])
 
 s=np.linspace(0,0.5,100)
 ax1.plot(s,1/(1-s),'k',linewidth=2)
@@ -31,27 +31,27 @@ d=np.array([1e-4,1e-4*(1-s)]);
 
 x=np.linspace(0,10/s,1000)
 
-def f(n,t):
-    return (b-d*sum(n))*n
-    
-def g(p,t):
-    return s*p*(1-p)
-
-solK=odeint(f,[0.99*b[0]/d[0],0.01*b[0]/d[0]],x)
-solrel=odeint(g,0.01,x)
-
-ax2.plot(x,solK[:,1]/(solK[:,0]+solK[:,1]),'k--', linewidth=2,label=r"$\epsilon=0.2$ (Eq. 11)")
-ax2.plot(x,solrel,'k', linewidth=2,label=r"Canonical (Eq. 1)")
-ax2.legend(loc='lower right',prop={'size':11})
-ax2.annotate(r'$(b)$',xy=(0.01,0.93),xycoords='axes fraction',fontsize=16)
-ax2.set_xticklabels([])
-ax2.set_xlabel("Time",fontsize=14)
-ax2.set_yticklabels(['0','','','','','1'])
-ax2.set_ylabel("Frequency",fontsize=14)
+#def f(n,t):
+#    return (b-d*sum(n))*n
+#    
+#def g(p,t):
+#    return s*p*(1-p)
+#
+#solK=odeint(f,[0.99*b[0]/d[0],0.01*b[0]/d[0]],x)
+#solrel=odeint(g,0.01,x)
+#
+#ax2.plot(x,solK[:,1]/(solK[:,0]+solK[:,1]),'k--', linewidth=2,label=r"$\epsilon=0.2$ (Eq. 11)")
+#ax2.plot(x,solrel,'k', linewidth=2,label=r"Canonical (Eq. 1)")
+#ax2.legend(loc='lower right',prop={'size':11})
+#ax2.annotate(r'$(b)$',xy=(0.01,0.93),xycoords='axes fraction',fontsize=16)
+#ax2.set_xticklabels([])
+#ax2.set_xlabel("Time",fontsize=14)
+#ax2.set_yticklabels(['0','','','','','1'])
[…]

### 2018-08-31 manuscript.tex
l

@@
 \end{equation}
-Hence, even in the simplest case that $W_1$ and $W_2$ are constant, selection is frequency-dependent in discrete time (note that this frequency dependence is negligible when $s$ is small compared to $1$; see \citealt{frank_2011}).
 We will refer to both the continuous and discrete time selection equations \eqref{eq:canonicalcont} and \eqref{eq:canonicaldisc} throughout this paper, but the simpler continuous time case will be our point of comparison for the rest of this section.
@@
 Thus, setting aside unviable populations, the increase in population density must be checked by crowding.
-This implies that the Malthusian parameters $r_i$ eventually decline to zero (e.g.
-\citealt[pp.
+This implies that the Malthusian parameters $r_i$ eventually decline to zero (e.g.~\citealt[pp.
 203]{begon_1990}).
@@
 \citealt{gillespie_2004}; \citealt{nagylaki_1992}; \citealt{ewens_2004}).
-Density either does not enter the model at all, or if finite-population size effects (``random genetic drift'') are important, then $N$ is assumed to have reached some fixed equilibrium value (Fig.~\ref{fig:Ksel}b).
+Density either does not enter the model at all, or if finite-population size effects (``random genetic drift'') are important, then $N$ is typically assumed to have reached some fixed equilibrium value (Fig.~\ref{fig:Ksel}b; for some approaches to relaxing the constant $N$ assumption in finite populations, see \citealt{lambert_2005,parsons_2007,chotibut_2017,constable_2017}).
 
@@
 
-The constant-$N$, constant-$s$ description of selection also precludes consideration of longer-term aspects of the interplay between evolution and ecology such as population extinction.
+The constant-$N$, constant-$s$ description of selection also limits consideration of longer-term aspects of the interplay between evolution and ecology such as population extinction and trait evolution.
 A variety of approaches have been developed to address this in quantitative genetics \citep{burger1995evolution,engen_2013}, population genetics \citep{bertram2017predicting} and adaptive dynamics \citep{ferriere2013eco,dieckmann2004adaptive}.
-Although density-dependent selection is pertinent to this longer-term issue, our focus here is the description of the time-dependent process by which selection changes allele frequencies.
+Although density-dependent selection is pertinent to these longer-term issues, our focus here is the description of the time-dependent process by which selection changes allele frequencies.
 This is particularly critical for making sense of evolution at the genetic level, for which we now have abundant data.
@@
 
-Here we argue that the simplified models of density-regulated growth mentioned above are misleading in their representation of the interplay between selection and density.
+Here we argue that the simplified models of density-regulated growth mentioned above are potentially misleading in their representation of the interplay between selection and density.
 This ultimately derives from their failure to account for ``reproductive excess'', that is, an excess of juveniles that experience stronger selection than their adult counterparts \citep{turner1968population}.
@@
 
-In models with detailed age structure, it is usually assumed that the density of a ``critical age group'' mediates the population's response to crowding \citep[pp.
+In models with detailed age structure, it is often assumed that the density of a ``critical age group'' mediates the population's response to crowding \citep[pp.
 54]{charlesworth_1994}.
@@
 In a given iteration, the number of adults of the $i$'th type will be denoted by $n_i$, the total number of adults by $N=\sum_i n_i$, and the number of unoccupied territories by $U=T-N$.
-We assume that the $n_i$ are large enough that stochastic fluctuations in the $n_i$ (drift) can be ignored (with $T$ also assumed large to allow for low type densities $n_i/T\ll 1$).
+We assume that the $n_i$ are large enough that stochastic fluctuations in the $n_i$ (drift) can be ignored, with $T$ also assumed large to allow for low type densities $n_i/T\ll 1$.
 
@@
 \end{equation}
-Here the sum only includes territories with at least one propagule present. Since we do not consider random genetic drift here, we will not analyze the fluctuations around these two expectations.
+Here the sum only includes territories with at least one propagule present.
+Note that $\Delta_+ n_i$ denotes the \textit{expected} territorial acquisition.
+Fluctuations about $\Delta_+ n_i$ (i.e.~drift) will not be analyzed in this manuscript.
+Note that drift can become important if $U$ is not sufficiently large even though $n_i$ and $T$ are large (by assumption); we do not consider this scenario on biological grounds, since it implies negligible population turnover.
 
@@
 For a low density type, growth comes almost entirely from territories with $x_i=1$, for which its mean density $l_i\ll 1$ is not representative.
-We therefore separate Eq.~\eqref{eq:growthsumuncoupled} into $x_i=1$ and $x_i>1$ components, taking care to ensure that the effective mean approximations for these components are consistent with each other (details in Appendix B).
+We therefore separate Eq.~\eqref{eq:growthsumuncoupled} into $x_i=1$ and $x_i>1$ components, taking care to ensure that the effective mean approximations for these components are consistent with each other (details in Appendix A).
 The resulting variable-density approximation only requires that there are no large discrepancies in competitive ability (i.e.
@@
 
-We now turn to the issue of how density changes as a consequence of selection in our variable-density lottery, and in previous models of selection in density-regulated populations.
+We now turn to the issue of how density changes as a consequence of selection in our variable-density lottery, and in more familiar models of selection in density-regulated populations.
 In the latter, selection under crowded conditions typically induces changes in equilibrium density (see Introduction).
@@
 
-MacArthur considered a population with two types that have densities $n_1$ and $n_2$ subject to density-dependent growth,
+MacArthur considered two types (with densities $n_1$ and $n_2$) in a constant environment subject to density-dependent growth,
 \begin{equation}
@@
 \end{equation}
-The environment is assumed to remain constant apart from changing type densities.
-The functions $f_1$ and $f_2$ must decline to zero if $n_1$ or $n_2$ are sufficiently large, because the resources required for growth are limited.
-This defines nullclines $f_1(n_1,n_2)=0$ and $f_2(n_1,n_2)=0$ in $(n_1,n_2)$ space.
-The outcome of selection is then determined by the relationship between these nullclines.
+The outcome of selection is determined by the relationship between the nullclines $f_1(n_1,n_2)=0$ and $f_2(n_1,n_2)=0$.
 Specifically, a type will be excluded if its nullcline is completely contained in the region bounded by the other type's nullcline.
-Thus, for a type to have the possibility of persisting, it must be able to keep growing to higher densities than the other type can tolerate in some region of $(n_1,n_2)$ space (Fig.~\ref{fig:Ksel}a).
 
-MacArthur used ``$K$'' to label the four intersection points of the nullclines with the axes, specifically $f_1(K_{11},0)=0$, $f_1(0,K_{12})=0$, $f_2(K_{21},0)=0$ and $f_2(0,K_{22})=0$.
-These $K$ values determine whether a region of higher-density growth exists for each type, provided that the nullclines are close to being straight lines.
-Note that only $K_{11}$ and $K_{22}$ are equilibrium densities akin to the $K$ parameter in the logistic model (Fig.~\ref{fig:Ksel}a).
-The other intersection points, $K_{12}$ and $K_{21}$, are related to competition between types.
-To be more concrete, in the Lotka-Volterra competition model we have
+MacArthur used the four intersection points of the nullclines with the axes, defined by $f_1(K_{11},0)=0$, $f_1(0,K_{12})=0$, $f_2(K_{21},0)=0$ and $f_2(0,K_{22})=0$, to analyze each type's exclusion or persistence.
+Note that only $K_{11}$ and $K_{22}$ are equilibrium densities akin to the $K$ parameter in the logistic model; the other intersection points, $K_{12}$ and $K_{21}$, are related to competition between types.
+For instance, in the Lotka-Volterra competition model we have
 \begin{align}
@@
 Hence, ``fitness is $K$'' in crowded populations \citep[pp.
-149]{macarthur_1967} in the sense that selection either favors the ability to keep growing at ever higher densities (moving a type's own nullcline outwards), or the ability to suppress the growth of competitors at lower densities (moving the nullcline of competitors inwards).
-This general idea is much broader than selection for greater equilibrium density \citep{gill_1974}.
+149]{macarthur_1967} in the sense that selection either favors the ability to keep growing at ever higher densities (moving a type's own nullcline outwards), or the ability to suppress the growth of competitors at lower densities (moving the nullcline of competitors inwards; \citealt{gill_1974}).
+However, even if the initial and final densities of an adaptive sweep in the Lotka-Volterra model are the same, $N$ nevertheless does change transiently (Fig.~\ref{fig:LVvslottery}a).
+Constant-$N$ over a sweep only occurs for a highly restricted subset of $r$ and $\alpha$ values (Appendix B; \citealt{mallet_2012,gill_1974,smouse_1976}).
 
@@
 (b) Lottery competition (Eq.~\ref{eq:master}) with $b_1=b_2=5$, $d_1=d_2=1.1$ and $c_1/c_2=5$.
-Trajectories converge on the line $N=K_{11}=K_{22}$.}
+The lottery model nullclines are defined by $W_1=1$ (lower nullcline) and $W_2=1$ respectively in Eq.~\eqref{eq:absfit}.
+For the discrete-time Eq.~\eqref{eq:master}, trajectories are the flow lines of the vector field obtained by evaluating the direction of the local changes in $n_1$ and $n_2$; these converge on the line $N=K_{11}=K_{22}$.}
 \end{figure}
 
-Compared to simple birth-death models \citep{kostitzin_1939} or variants of the logistic \citep{roughgarden_1979}, the Lotka-Volterra model clearly distinguishes between intra- and inter-type competitive effects.
-Thus, when selection acts on inter-type competitive effects, one type can displace another without having a greater equilibrium density (Fig.~\ref{fig:LVvslottery}a).
-This has been termed ``$\alpha$-selection'' to distinguish it from $K$-selection, which involves intra-type competitive effects and changes in equilibrium density \cite{gill_1974,joshi_2001}.
-Although the initial and final densities of an $\alpha$-selection sweep are the same, density nevertheless does change transiently in the Lotka-Volterra model (constant density only occurs for a highly restricted subset of $r$ and $\alpha$ values; further details in Appendix C; also see \citealt{mallet_2012,smouse_1976}).
-Intuitively, for one type to exclude the other, competitive suppression of growth between types must be stronger than competitive suppression of growth within types, causing $N$ to dip over a sweep (Fig.~\ref{fig:LVvslottery}a).
-
-In contrast to both $K$ and $\alpha$ selection, density trajectories for $c$-selection in our variable-density lottery converge on a line of constant equilibrium density (Fig.~\ref{fig:LVvslottery}b).
-This means that once $N$ reaches demographic equilibrium, selective sweeps behave indistinguishably from a constant-$N$ relative fitness model (Fig.~\ref{fig:Ksel}b).
-Thus, for $c$ sweeps, the selection factor $(W_2-W_1)/\overline{W}$ in Eq.~\eqref{eq:canonicaldisc} depends on frequency only, not density, provided that nothing else pushes $N$ out of demographic equilibrium over the sweep.
-This uncoupling of density from ongoing $c$-selection arises due to the presence of an excess of propagules which pay the cost of selection without affecting adult density \citep{nei1971fertility}.
+In contrast, density trajectories for $c$-selection in our variable-density lottery converge on a line of constant equilibrium density (Fig.~\ref{fig:LVvslottery}b).
+This means that once $N$ reaches demographic equilibrium, selective sweeps behave indistinguishably from a constant-$N$ relative fitness model(Fig.~\ref{fig:Ksel}b).
+Thus, for $c$-sweeps in a constant environment, the selection factor $(W_2-W_1)/\overline{W}$ in Eq.~\eqref{eq:canonicaldisc} is density-independent.
+This uncoupling of density from $c$-selection arises due to the presence of an excess of propagules which pay the cost of selection without affecting adult density \citep{nei1971fertility}.
 
@@
 
-For the relative fitness model Eq.~\eqref{eq:canonicaldisc} to break down, the selection factor $(W_2-W_1)/\overline{W}$ must depend on density.
-As shown in Fig.~\ref{fig:DDS_lottery}, $(W_2-W_1)/\overline{W}$ is independent of $N$ in the case of $d$-selection.
-Selection is also independent of $N$ when the population is at demographic equilibrium and $N$ is unaffected by ongoing selection; as is the case for $c$-selection.
-Thus, to invalidate Eq.~\eqref{eq:canonicaldisc}, we require that selection is density-dependent, and also that density is changing.
-This can obviously occur if density-dependent selection happens in a population far from demographic equilibrium, in which case the validity of Eq.~\eqref{eq:canonicaldisc} depends on the specifics of the rate and magnitude of demographic change (we return to this in the Discussion).
-However, Eq.~\eqref{eq:canonicaldisc} can break down even in demographically-stable populations if a density-regulating trait is subject to density-dependent selection, as is the case for the $b$ trait in our variable-density lottery.
-
-Before we discuss the $b$ trait, it is helpful to summarize how density-dependent selection on a density-regulating trait can invalidate Eq.~\eqref{eq:canonicalcont} in simpler continuous-time models.
-This applies, for example, to $K$-selection in the logistic \citep{kimura1969natural,crow_1970}.
-We consider the simple birth-death model \citep{kostitzin_1939}
+For density to matter in Eq.~\eqref{eq:canonicaldisc}, selection must be density-dependent and density must be changing.
+This can occur in a constant environment if selection acts on a density-regulating trait.
+Consider the simple birth-death model \citep{kostitzin_1939}
 \begin{equation}
@@
 \end{equation}
-where $\delta_i$ is per-capita mortality due to crowding (for simplicity, there are no deaths when uncrowded).
-Starting from a type $1$ population in equilibrium, a variant with $\delta_2=\delta_1(1-\epsilon)$ has density-dependent selection coefficient $s=\epsilon \delta_1 N$ in Eq.~\eqref{eq:canonicalcont}, which will change over the course of the sweep as $N$ shifts from its initial type $1$ equilibrium to a type $2$ equilibrium.
-From Eq.~\eqref{eq:simplebirthdeath}, the equilibrium densities at the beginning and end of the sweep are $N_{\rm initial}=b_1/\delta_1$ and $N_{\rm final}=b_1/(\delta_1(1-\epsilon))=N_{\rm initial}/(1-\epsilon)$ respectively, and so $s_{\rm initial}= \epsilon b_1$ and $s_{\rm final}=s_{\rm initial}/(1-\epsilon)$.
-Consequently, substantial deviations from Eq.~\eqref{eq:canonicalcont} occur if there is sufficiently strong selection on $\delta$ (Fig.~\ref{fig:strengthofselection}).
+where $\delta_i$ is per-capita mortality due to crowding.
+Starting from a type $1$ population in equilibrium, a variant with $\delta_2=\delta_1(1-\epsilon)$ has density-dependent selection coefficient $s=\epsilon \delta_1 N$, which will change over the course of the sweep as $N$ shifts from its initial type $1$ equilibrium to a type $2$ equilibrium.
+The equilibrium densities at the beginning and end of the sweep are $N_{\rm initial}=b_1/\delta_1$ and $N_{\rm final}=b_1/(\delta_1(1-\epsilon))=N_{\rm initial}/(1-\epsilon)$ respectively, and so $s_{\rm initial}= \epsilon b_1$ and $s_{\rm final}=s_{\rm initial}/(1-\epsilon)$.
+Consequently, substantial deviations from Eq.~\eqref{eq:canonicalcont} occur if there is sufficiently strong selection on $\delta$ (Fig.~\ref{fig:strengthofselection}; \citealt{kimura_1978,kimura1969natural}).
 
@@
 \caption{\label{fig:strengthofselection} (a) Change in the selection coefficient between the beginning and end of a sweep of a type that experiences proportionally $1-\epsilon$ fold fewer crowding-induced deaths.
-The population is in demographic equilibrium at the start and end of the sweep.
-(b) Example equilibrium-to-equilibrium sweep.}
+The population is in demographic equilibrium at the start and end of the sweep.}
 \end{figure}
 
-Equilibrium-to-equilibrium $b$-sweeps in our variable-density lottery are qualitatively different from $\delta$ sweeps in this simpler birth-death model.
+In our variable density lottery, $b$ regulates density and is subject to density-dependent selection, yet $b$-sweeps are qualitatively different from $\delta$ sweeps in the above example.
 Greater $b$ means not only that more propagules contest the available territories, but also that a greater fraction of unoccupied territories receive propagules.
 Together, the net density-dependent effect on $b$-selection is negligible: in a single-type equilibrium we have $W_i=1$ and $b_i/\overline{b}=1$, and hence the density-dependence factor $f(\overline{b},N)=\frac{1-e^{-\overline{b}N/T}}{N}(T-N)$ in Eq.~\eqref{eq:masterequalc} has the same value $d_i-1$ at the beginning and end of a $b$-sweep (recall that $\frac{W_2-W_1}{\overline{W}}=\frac{f(\overline{b},N)}{1+f(\overline{b},N)}\frac{b_2-b_1}{\overline{b}}$ for $b$-selection).
-During the sweep there is some deviation in $f(\overline{b},N)$, but this deviation is an order of magnitude smaller than for a $\delta$ sweep (the density-dependent deviation in Fig.~\ref{fig:strengthofselection} is of order $\epsilon$, whereas the analogous effect for $b$ sweep in our variable-density lottery is only of order $\epsilon^2$; see Appendix D for details).
-Since selection must already be strong for a $\delta$-sweep to invalidate Eq.~\eqref{eq:canonicalcont}, the density-independent model applies effectively exactly for equilibrium $b$-sweeps (Fig.~\ref{fig:bsweep}).
+During the sweep there is some deviation in $f(\overline{b},N)$, but this deviation is an order of magnitude smaller than for a $\delta$ sweep (the density-dependent deviation in Fig.~\ref{fig:strengthofselection} is of order $\epsilon$, whereas the analogous effect for $b$ sweep in our variable-density lottery is only of order $\epsilon^2$; see Appendix C for details).
+Since selection must already be strong for a $\delta$-sweep to invalidate Eq.~\eqref{eq:canonicalcont}, the density-independent model applies almost exactly for equilibrium $b$-sweeps (Fig.~\ref{fig:bsweep}).
 
@@
 
-In this appendix we derive Eq.~\eqref{eq:master}.
+Here we derive Eq.~\eqref{eq:master}.
 Following the notation in the main text, the Poisson distributions for the $x_i$ (or some subset of the $x_i$) will be denoted $p$, and we use $P$ as a general shorthand for the probability of particular outcomes.
@@
 
-This pathological behavior occurs because the expected total density of propagules in the respective groups of territories are different; $\langle X \rangle_{p({\mathbf x}|x_1=1,X_1\geq 1)}\approx l_2+1 > \langle X \rangle_{p({\mathbf x}|x_2\geq 2)}\approx l_2 $.
+This pathological behavior occurs because the expected total density of propagules in the respective groups of territories are different; $\langle X \rangle_{p({\mathbf x}|x_1=1,X_1\geq 1)}\approx l_2+1 > \langle X \rangle_{p({\mathbf x}|x_2\geq 2)}=\approx l_2 $.
 As a result, the rare type's behavior is approximated as though it experiences more intense lottery competition than the common type, which cannot be the case since the two types are identical.
@@
 \end{equation}
-This assumption is mathematically convenient
+As we will see, these effective means are straightforward to calculate analytically, and ensure that the expected total propagule density $\langle x_i \rangle + \sum_{j\neq i} \langle x_j \rangle$ is the same in Eq.~\eqref{eq:meanfieldr} and Eq.~\eqref{eq:meanfielda}.
 
@@
 The right hand side is obtained by observing that the sum on the left is the expected number of propagules with type $j$ that will be found in a territory which received $X-1$ nonfocal propagules in total.
-We then take the with respect to $P(X|X\geq 2)$ to give
+We then take the expectation with respect to $P(X|X\geq 2)$ to give
 \begin{align}
@@
 &=\frac{1}{1-(1+l_i)e^{-l_i}}\sum_{x_i\geq 2} p(x_i)x_i\nonumber\\
-&=l_i\frac{1-e^{-l_i}}{1-(1+l_i)e^{-l_i}}.
+&=l_i\frac{1-e^{-l_i}}{1-(1+l_i)e^{-l_i}}. \label{eq:meanxicommon}
 \end{align}
@@
 \end{align}
-Again taking the expection with respect to $P(X|X\geq 2)$ yields 
+Again taking the expectation with respect to $P(X|X\geq 2)$ yields 
 \begin{align}
 \langle x_j \rangle_a&=\frac{l_j}{L-l_i}\left[\sum_{X=2}^{\infty}P(X|X\geq 2)X - \langle x_i \rangle_a \right]\nonumber\\
-&=\frac{l_j}{L-l_i}\left( L\frac{1-e^{-L}}{1-(1+L)e^{-L}}- l_i\frac{1-e^{-l_i}}{1-(1+l_i)e^{-l_i}}\right).
+&=\frac{l_j}{L-l_i}\left( L\frac{1-e^{-L}}{1-(1+L)e^{-L}}- l_i\frac{1-e^{-l_i}}{1-(1+l_i)e^{-l_i}}\right). \label{eq:meanxjcommon}
 \end{align}
@@
 
+It is easily verified from Eqs.~\eqref{eq:meanxjrare}, \eqref{eq:meanxicommon} and \eqref{eq:meanxjcommon} that the total expected propagule density is the same in in Eq.~\eqref{eq:meanfieldr} and Eq.~\eqref{eq:meanfielda} i.e.~$\langle x_i \rangle_r + \sum_{j\neq i} \langle x_j \rangle_r=\langle x_i \rangle_a + \sum_{j\neq i} \langle x_j \rangle_a=\langle X \rangle_{P(X|X\geq 2)}$.
+As a result, Eq.~\eqref{eq:master} satisfies the consistency requirement (see Fig.~\ref{fig:approx_details}).
+
 \subsection*{Approximation limits}
@@
 Eq.~\eqref{eq:master} relies on ignoring the fluctuations in $x_i$ and $x_j$, such that we can replace them with constant effective mean values.
-Mathematically, this requires that the standard deviations $\sigma_{p({\mathbf x}|x_i=1,X_i\geq 1)}(\sum_{j\neq i} c_j x_j)$ and $\sigma_{p({\mathbf x}|x_i\geq 2)}(\sum_j c_j x_j)$ are sufficiently small compared to the corresponding means $\langle\sum_{j\neq i} c_j x_j\rangle_{p({\mathbf x}|x_i=1,X_i\geq 1)}$ and $\langle\sum_j c_j x_j\rangle_{p({\mathbf x}|x_i\geq 2)}$ in Eqs.~\eqref{eq:meanfieldr} and \eqref{eq:meanfielda}.
+To justify this, we show that the standard deviations $\sigma_{p({\mathbf x}|x_i=1,X_i\geq 1)}(\sum_{j\neq i} c_j x_j)$ and $\sigma_{p({\mathbf x}|x_i\geq 2)}(\sum_j c_j x_j)$ are small compared to the corresponding means $\langle\sum_{j\neq i} c_j x_j\rangle_{p({\mathbf x}|x_i=1,X_i\geq 1)}$ and $\langle\sum_j c_j x_j\rangle_{p({\mathbf x}|x_i\geq 2)}$ in Eqs.~\eqref{eq:meanfieldr} and \eqref{eq:meanfielda}.
+This result means that using the exact distributions $p({\mathbf x}|x_i=1,X_i\geq 1)$ and $p({\mathbf x}|x_i\geq 2)$ for the effective means would produce an accurate approximation of the components in \eqref{eq:delt_decomp} (though, as we have seen, not a consistent one).
+It is then clear that the effective means derived in the previous section will also give an accurate approximation since their magnitudes are similar to the exact means; this is obvious from the fact that the expected total number of propagules is of order $\min\{L,2\}$ in both cases.
 

### 2018-08-31 manuscript.tex
l

@@
 We find that density-independent relative fitnesses accurately describe strong selection acting on any one trait, even fecundity, which is both density-regulating and subject to density-dependent selection.
-Our findings suggest that deviations from demographic equilibrium pose the most serious difficulties for relative fitness models.
-In such cases our model offers a possible alternative to relative fitness.
+Pleiotropic interactions between these traits recovers the familiar $K$-selection behavior.
+In such cases, or when the population is maintained far from demographic equilibrium, our model offers a possible alternative to relative fitness.
 
-\noindent (210 words)
+\noindent (215 words)
 
@@
 \includegraphics[scale=0.8]{strengthofselection.pdf}
-\caption{\label{fig:strengthofselection} (a) Change in the selection coefficient between the beginning and end of a sweep of a type that experiences proportionally $1-\epsilon$ fold fewer crowding-induced deaths.
+\caption{\label{fig:strengthofselection} Change in the selection coefficient between the beginning and end of a sweep of a type that experiences proportionally $1-\epsilon$ fold fewer crowding-induced deaths.
 The population is in demographic equilibrium at the start and end of the sweep.}
@@
 Summarizing the properties of selection in our variable-density lottery model: (i) $c$-selection is density-dependent, but $c$ does not regulate density; (ii) $d$ regulates density, but $d$-selection is density-independent; (iii) $b$ regulates density and $b$-selection is density-dependent.
-Yet, despite the differences between $b$, $c$ and $d$,  selection in a constant environment that only involves one of these traits obeys the density-independent relative fitness description of selection almost exactly (that is, $(W_2-W_1)/\overline{W}$ in Eq.~\eqref{eq:canonicaldisc} is effectively independent of density).
+Yet, despite the differences between $b$, $c$ and $d$,  selection in a constant environment that only involves one of these traits obeys the density-independent relative fitness description of selection almost exactly (that is, $(W_2-W_1)/\overline{W}$ in Eq.~\eqref{eq:canonicaldisc} is approximately independent of density).
 This density-independence breaks down when strong selection acts on more than one of $b$, $c$ and $d$ (Fig.~\ref{fig:bdsweep}).
@@
 Remarkably, the $b$ trait demonstrates that the combination is not sufficient; the density-dependence of $b$-selection effectively disappears over equilibrium-to-equilibrium $b$-sweeps.
-
-Selection in the variable-density lottery is quite different from classical density-dependent selection (see ``Introduction'' and ``The response of density to selection; $c$-selection versus $K$-selection'').
-In the latter, only one life-history stage is represented, and the effects of crowding appear as a reduction in absolute fitness that only depends on the type densities at this life-history stage (e.g.
-the $n_i^2$ and $n_in_j$ terms in the Lotka-Volterra equation).
-Selection in crowded populations takes broadly one of two forms: selection for greater carrying capacity ($K$-selection) or selection on competition coefficients ($\alpha$-selection).
-These are both ``$\delta$-like''  in the sense that selection depends on density and also causes density to change ($\delta$ is defined in Eq.~\eqref{eq:simplebirthdeath}).
-Strong selection is therefore sufficient for Eq.~\eqref{eq:canonicalcont} to break down (Fig.~\ref{fig:strengthofselection}), and no distinction is made between density-regulating and density-dependent traits.
 
@@
 
-Looking beyond the variable-density lottery, it is not clear which forms of crowding-induced selection are more likely to occur in nature.
-Even if reproductive excesses are ubiquitous, strictly relative $c$-like traits could pleiotropically interact with density-regulating traits so often that $\delta$-like behavior is prevalent.
-For instance, in the case in the case of well-mixed indirect exploitation competition for consumable resources, the $R^*$ rule suggests that competitive ability is intimately linked to equilibrium resource density, and hence that $\delta$-like behavior would be prevalent.
-However, this conclusion is sensititive to the assumptions of well-mixed resource competition models.
-Spatial localization of consumable resources (e.g.
+In practice, traits like $b$, $c$ and $d$ will often have pleiotropic interactions which mean that adaptive sweeps will behave similarly to the familiar ``$\delta$ sweeps'' (``Density regulating traits under strong selection'').
+Thus, our analysis of the variable density lottery does not suggest that Eq.~\eqref{eq:canonicalcont} and \eqref{eq:canonicaldisc} should apply more broadly than previously thought.
+Rather, $b$, $c$ and $d$ represent a possible set of idealized fitness components mediating the interplay between selection and density in density-regulated populations.
+The conceptual distinctions they highlight will hopefully be useful regardless of the biological realism of the variable density lottery model.
+
+Apart from familiar examples such as reef fishes \citep{chesson_1981}, many biological systems do not obviously satisfy the assumptions of our variable density lottery.
+However, even if competition occurs primarily via consumable resource exploitation, spatial localization of consumable resources (e.g.
 for plants due to restricted movement of nutrients through soils) will tend to create territorial contests similar to the lottery model, where resource competition only occurs locally and can be sensitive to contingencies such as the timing of propagule arrival \citep{bolker_99}.
 In this case, resource competition is effectively subsumed into a territorial competitive ability trait akin to $c$, which would likely affect $N$ much more weakly than suggested by the $R^*$ rule (assuming no pleiotropic interactions with $b$ or $d$).
-
 Moreover, even in well-mixed populations, competition does not only involve indirect exploitation of shared resources, but also direct interference.
@@
 For instance, sexual selection can be viewed as a form of relative interference competition between genotypes.
-Thus, \textit{a priori} we should not expect crowding in nature to only involve selection that is $\delta$-like.
-Other forms of selection like $c$-selection (that is, strictly relative traits in density-regulated populations) are also likely to be important.
-Note that in the classical density-dependent selection literature, interference competition is closely associated with $\alpha$-selection and the idea that selection need not affect equilibrium density \citep{gill_1974}.
-However, $\alpha$-selection does transiently affect population density and therefore retains $\delta$-like features.
 
-The above findings suggest that the largest deviations from the approximation of density-independent selection (as represented by Eqs.~\eqref{eq:canonicalcont} and \eqref{eq:canonicaldisc}) will occur in populations far from demographic equilibrium e.g.~as a result of a temporally-variable environment.
-While transient deviations from demographic equilibrium driven by the appearance of new types can also cause the density-independent approximation to break down, this requires strong selection that is both density-dependent and affects a density-regulating trait (and, as exemplified by $b$-selection, even then the approximation may hold).
+In the analysis presented here we have restricted our attention to selection in demographically stable populations.
+The largest deviations from the approximation of density-independent selection (as represented by Eqs.~\eqref{eq:canonicalcont} and \eqref{eq:canonicaldisc}) will likely occur in populations far from demographic equilibrium e.g.~as a result of a temporally-variable environment.
+This is because extremely strong selection to change population density by an amount comparable to environmental varability (see Fig.~\ref{fig:strengthofselection}).
 By contrast, temporally-variable environments can dramatically alter frequency trajectories for individual sweeps (e.g.

### 2018-09-01 manuscript.tex
l

@@
 The strong connection between crowding and selection for greater equilibrium density is broken: selection need not affect density at all.
-And when it does, the density-independent discrete-time selection equation \eqref{eq:canonicaldisc} is almost exact even for strong selection, provided that any changes in density are driven only by selection (as opposed to large deviations from demogaphic equilibrium), and that selection occurs on only one of the traits $b$, $c$, or $d$.
+And when it does, the density-independent discrete-time selection equation \eqref{eq:canonicaldisc} is almost exact even for strong selection, provided that any changes in density are driven only by selection (as opposed to large deviations from demographic equilibrium), and that selection occurs on only one of the traits $b$, $c$, or $d$.
 On the flip side, the constant relative fitness approximation fails when strong selection acts concurrently on two or more of these traits, or when the population is far from demographic equilibrium.
@@
 Here the sum only includes territories with at least one propagule present.
-Note that $\Delta_+ n_i$ denotes the \textit{expected} territorial acquisition.
+Note that $\Delta_+ n_i$ denotes the expected territorial acquisition.
 Fluctuations about $\Delta_+ n_i$ (i.e.~drift) will not be analyzed in this manuscript.
@@
 Here $\frac{\Delta_+ n_i}{n_i}$ is the per-capita rate of territorial acquisition, and $1/d_i$ is the fraction of type $i$ adults surviving to the next iteration.
+Note that our model of mortality differs from the classic lottery model \citep{chesson_1981}, where mortality affects adults only and occurs after propagule production but before juvenile recruitment.
+In the latter case, selection on mortality exhibits some density-dependence, although this reflects the fact that newly recruited adults are guaranteed to reproduce before dying, which is not interesting for our purposes here.
+Our mortality model ensures that selection on $d_i$ is density-independent, allowing us to more clearly separate different sources of density-dependence and density regulation.
 
@@
 Note that density-dependent selection is sometimes taken to mean a qualitative change in which types are fitter than others at different densities \citep{travis_2013}.
-While reversal in the order of fitnesses and co-existence driven by density-regulation are possible in our variable-density lottery (a special case of the competition-colonization trade-off; \citealt{levins_71,tilman_94,bolker_99}), questions related to co-existence are tangential to our aims and will not be pursued further here.
+While reversal in the order of fitnesses and co-existence driven by density-regulation are possible in our variable-density lottery (a special case of the competition-colonization trade-off; \citealt{levins_71,tilman_94,bolker_99}), questions related to co-existence are tangential to the aims of the current manuscript and will not be pursued further here.
 
@@
 \includegraphics[scale=0.8]{strengthofselection.pdf}
-\caption{\label{fig:strengthofselection} Change in the selection coefficient between the beginning and end of a sweep of a type that experiences proportionally $1-\epsilon$ fold fewer crowding-induced deaths.
+\caption{\label{fig:strengthofselection} Change in the selection coefficient between the beginning and end of a sweep of a type that experiences proportionally $1-\epsilon$ fewer crowding-induced deaths.
 The population is in demographic equilibrium at the start and end of the sweep.}
@@
 Together, the net density-dependent effect on $b$-selection is negligible: in a single-type equilibrium we have $W_i=1$ and $b_i/\overline{b}=1$, and hence the density-dependence factor $f(\overline{b},N)=\frac{1-e^{-\overline{b}N/T}}{N}(T-N)$ in Eq.~\eqref{eq:masterequalc} has the same value $d_i-1$ at the beginning and end of a $b$-sweep (recall that $\frac{W_2-W_1}{\overline{W}}=\frac{f(\overline{b},N)}{1+f(\overline{b},N)}\frac{b_2-b_1}{\overline{b}}$ for $b$-selection).
-During the sweep there is some deviation in $f(\overline{b},N)$, but this deviation is an order of magnitude smaller than for a $\delta$ sweep (the density-dependent deviation in Fig.~\ref{fig:strengthofselection} is of order $\epsilon$, whereas the analogous effect for $b$ sweep in our variable-density lottery is only of order $\epsilon^2$; see Appendix C for details).
+During the sweep there is some deviation in $f(\overline{b},N)$, but this deviation is an order of magnitude smaller than for a $\delta$-sweep (the density-dependent deviation in Fig.~\ref{fig:strengthofselection} is of order $\epsilon$, whereas the analogous effect for a $b$-sweep in our variable-density lottery is only of order $\epsilon^2$; see Appendix C for details).
 Since selection must already be strong for a $\delta$-sweep to invalidate Eq.~\eqref{eq:canonicalcont}, the density-independent model applies almost exactly for equilibrium $b$-sweeps (Fig.~\ref{fig:bsweep}).
@@
 
-In practice, traits like $b$, $c$ and $d$ will often have pleiotropic interactions which mean that adaptive sweeps will behave similarly to the familiar ``$\delta$ sweeps'' (``Density regulating traits under strong selection'').
-Thus, our analysis of the variable density lottery does not suggest that Eq.~\eqref{eq:canonicalcont} and \eqref{eq:canonicaldisc} should apply more broadly than previously thought.
+Traits like $b$, $c$ and $d$ will often have pleiotropic interactions which mean that adaptive sweeps will behave similarly to the familiar ``$\delta$ sweeps'' (``Density regulating traits under strong selection'').
+Thus, our analysis of the variable density lottery does not necessarily imply that Eq.~\eqref{eq:canonicalcont} and \eqref{eq:canonicaldisc} should apply more broadly than previously thought.
 Rather, $b$, $c$ and $d$ represent a possible set of idealized fitness components mediating the interplay between selection and density in density-regulated populations.
-The conceptual distinctions they highlight will hopefully be useful regardless of the biological realism of the variable density lottery model.
+The conceptual distinctions they highlight will hopefully be useful regardless of the biological realism of our variable density lottery model.
 
@@
 The largest deviations from the approximation of density-independent selection (as represented by Eqs.~\eqref{eq:canonicalcont} and \eqref{eq:canonicaldisc}) will likely occur in populations far from demographic equilibrium e.g.~as a result of a temporally-variable environment.
-This is because extremely strong selection to change population density by an amount comparable to environmental varability (see Fig.~\ref{fig:strengthofselection}).
+This is because extremely strong selection is needed to change population density by an amount comparable to environmental varability (see Fig.~\ref{fig:strengthofselection}).
 By contrast, temporally-variable environments can dramatically alter frequency trajectories for individual sweeps (e.g.
@@
 
-This pathological behavior occurs because the expected total density of propagules in the respective groups of territories are different; $\langle X \rangle_{p({\mathbf x}|x_1=1,X_1\geq 1)}\approx l_2+1 > \langle X \rangle_{p({\mathbf x}|x_2\geq 2)}=\approx l_2 $.
+This pathological behavior occurs because the expected total density of propagules in the respective groups of territories are different: $\langle X \rangle_{p({\mathbf x}|x_1=1,X_1\geq 1)}\approx l_2+1 > \langle X \rangle_{p({\mathbf x}|x_2\geq 2)}\approx l_2 $.
 As a result, the rare type's behavior is approximated as though it experiences more intense lottery competition than the common type, which cannot be the case since the two types are identical.
@@
 This result means that using the exact distributions $p({\mathbf x}|x_i=1,X_i\geq 1)$ and $p({\mathbf x}|x_i\geq 2)$ for the effective means would produce an accurate approximation of the components in \eqref{eq:delt_decomp} (though, as we have seen, not a consistent one).
-It is then clear that the effective means derived in the previous section will also give an accurate approximation since their magnitudes are similar to the exact means; this is obvious from the fact that the expected total number of propagules is of order $\min\{L,2\}$ in both cases.
+It is then clear that the effective means derived in the previous section will also give an accurate approximation since their magnitudes are similar to the exact means; this is obvious from the fact that the expected total number of propagules is of order $\max\{L,2\}$ in both cases.
 
